### 2023-01-08 LICS2023.tex
Add files via upload

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{tikz-cd}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{amsthm}
\usetikzlibrary{cd}
\usepackage{stmaryrd}

\input{macros}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Tropical Geometry and the Lambda Calculus}

\author{\IEEEauthorblockN{Davide Barbarossa}
\IEEEauthorblockA{\textit{Universit\`a di Bologna}\\
Bologna, Italy \\
davide.barbarossa2@unibo.it}
\and
\IEEEauthorblockN{Paolo Pistone}
\IEEEauthorblockA{\textit{Universit\`a Roma Tre}\\
Rome, Italy\\
paolo.pistone@uniroma3.it}
}

\maketitle

\begin{abstract}

\end{abstract}

\begin{IEEEkeywords}
\end{IEEEkeywords}

\section{Introduction}

In the recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability, 
approximation bounds, 
as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence.

Notably, a significative effort has been made in order to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties. We can mention, for example, 
intersection type systems aimed at capturing time or space resources \cite{Beniamino}, or convergence probabilities \cite{UgoBreuvart, LICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthes} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for 
probabilistic \cite{PCOH, QBS} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
Usually, the main reason to resort to such logical, type-theoretical or denotational approaches is to obtain methods which are \emph{modular} and \emph{compositional}, that is, where properties of complex algorithms can be deduced from properties of their component parts.

\subsection{Two kinds of quantitative approaches}

Among such quantitative approaches, two different directions have received a considerable attention in recent years.

On the one hand there is the approach based on \emph{non-idempotent} intersection types \cite{} and their semantical counterpart, the so-called \emph{relational semantics} \cite{}, which have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{}, and have also been shown suitable to capture probabilistic properties \cite{Manzo}.

On the other hand there is the approach of \emph{program metrics} \cite{} and \emph{quantitative equational theories} \cite{Mardare}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}. This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{} and coinductive methods \cite{}, and was recently extended to account for the the full $\lambda$-calculus \cite{LICS2021, FSCD2022}.

In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{linearlogic} (i.e.~of using its input exactly once), plays a crucial role.
In the relational semantics linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider \emph{analytic} maps, i.e.~formal infinite series of monomials of the form $a_{n}\cdot x^{n}$.
By contrast, in metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances, while the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gabo2017}, and general programs as maps satisfying coarser 
continuity conditions (see e.g.~\cite{LICS2021}).

A natural question is thus whether these two, apparently unrelated, ways of interpreting linearity and duplication can be somehow connected. At a first glance, there seems to be a fundamental ``logarithmic'' gap between the two approaches.
In the relational semantics a typical ``duplicating'' map is obtained by composing the diagonal with multiplication:
$$
\begin{tikzcd}
\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
 & &  &
 \mathbb R\times \mathbb R 
 \ar{rrr}{\langle x,y\rangle \mapsto x\cdot y}
 & & & \mathbb R
\end{tikzcd}
$$
yielding the square product function $\lambda x.x^{2}$.
However, in metric semantics this function needs not even exist (as these models are often restricted to Lipschitz-continuous maps \cite{Gabo2017})! Instead, a typical ``duplicating'' map can be obtained by composing the diagonal with the sum 
$$
\begin{tikzcd}
\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
 & &  &
 \mathbb R\times \mathbb R 
 \ar{rrr}{\langle x,y\rangle \mapsto x+y}
 & & & \mathbb R
\end{tikzcd}
$$
yielding the linear (and Lipschitz) function $\lambda x.2x$.

As this example seems to suggest, there seems to be a sort of ``logarithmic'' gap between the two approaches. Can this be made explicit?

\subsection{Tropical Geometry: a ``logarithmic'' version of usual geometry}

In very simple words, tropical geometry describes what happens to usual algebra and geometry when we replace ``$\cdot$'' by  ``$+$'', and ``$+$'' by ``$\min$''. It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
Tropical geometry is a kind of \emph{idempotent} mathematics, since the role of addition is 
played by the idempotent operation $\min$.
For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$ corresponds in the tropical word correspond to the \emph{linear} function
$$
\trop p(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
$$

The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials). For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning \cite{}.

Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for: a way to see polynomials as linear (i.e.~Lipschitz) functions.
At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories, have also been observed \cite{}.

The goal of this paper is to make such intuitive connections precise, yielding a way to relate, conceptually, two different quantitative approaches to higher-order algorithms, as well as suggesting new potential applications of tropical geometry to the study of quantitative properties of programs.

\subsection{Contributions}

Our contributions in this paper are threefold:
\begin{itemize}

\item We study the relational model over the tropical semi-ring, showing that
\emph{bounded} simply typed terms correspond to tropical polynomials, while general
 simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{}. Moreover, we show that such maps are concave and locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
Moreover, we exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps.

\item Using the relational model as our main reference, we suggest a list of potential applications of tropical methods to the study of quantitative properties of non-deterministic/probabilistic functional programs like differential privacy, convergence \emph{log-probabilities} and the number of computation steps in the best case.

\item We conclude by putting the correspondence between differential and metric properties made possible by the tropical viewpoint at the right level of generality. 
By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{} and modules over the tropical semi-ring \cite{}, we show that the category of \emph{complete}  GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
\end{itemize}
%
%\section{Bounded and Differential $\lambda$-Calculi}
%
%
%Bounded Simply Typed $\lambda$-calculus $\BSTLC$:
%$$
%A::= o \mid !_{n}A \multimap A
%$$
%
%
%Resource Simply Typed $\lambda$-calculus $\RSTLC$:
%$$
%A::= o \mid [A, \dots , A] \multimap A
%$$
%
%
%Define a translation of types $(-)^{\C R}$ from $\BSTLC$ to $\RSTLC$ by $o^{\C R}=o$ and $(!_{n}A\multimap B)^{\C R}=
%[\underbrace{A^{\C R},\dots, A^{\C R}}_{n\text{ times}}]\multimap B^{\C R}$.
%
%\begin{proposition}
%$\Gamma \vdash_{\BSTLC} M:A$ implies 
%$\Gamma^{\C R}\vdash_{\RSTLC}M:A^{\C R}$.
%\end{proposition}
%

\section{Tropical Geometry}

- The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.

- tropical linear algebra: the category $\LREL$

- tropical polynomials and tropical roots

- tropical Laurent series and tropical roots

- tropicalization of analytic functions and their roots: 
$\trop P$

- the differential operator

\section{The Tropical Relational Model}

Describe the model in a parametric way over a continuous semi-ring, as in Manzonetto et al.

Define the tropicalization $\trop$ from the $\BB R$-based model
to the $\Lawv$-based model.

Focus on \emph{bounded} simply typed $\lambda$-calculus $\BSTLC$:
$$
A::= o \mid !_{n}A \multimap A
$$
The interpretation of a bounded type is defined as 
$\model A= \Lawv^{\nodel A}$, where 
$\nodel A$ is defined by $\nodel o=\B 1$ and 
$\nodel{!_{n}A\multimap B}:= \C M_{\leq n}(\nodel A) \times \nodel B$.

Bounded types are interesting
because of the following proposition:
\begin{proposition}
For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) are polynomials.

\end{proposition}
\begin{proof}
It suffices to check that $\nodel A$ is finite for all bounded types $A$. Indeed this implies that a morphism $f:\model A\to \model B$ corresponds to a finite matrix $\widehat f: \nodel A \times \nodel B \to \Lawv$.
\end{proof}

For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
\begin{itemize}
\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;

\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.

\end{itemize}

Discuss the tropical Taylor expansion $\trop\C T(M)$.

\section{Analytic Results}

- the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric

- concavity

- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.

- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
\begin{itemize}

\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally

\item singular points: the famous example

\item tropical roots

\end{itemize}

- discussion of the tropical Taylor expansion:

- characterization of the functional metric using derivatives

\section{Applications}

General discussion: optimization properties behave in a Lipschitz way.

- differential privacy and Lipschitzness

- log-probabilities and tropical roots

- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)

- measuring duplications of discrete functions (needs finiteness!)

\section{Quasi-metric Spaces and $\Lawv$-Modules}

- correspondence between $\Lawv$-modules and $\Lawv$-categories.

- exponential structure of $\LCAT$.

- $\LCAT_{!}$ is a cartesian closed differential category

\section{Related Work}

\section{Conclusion}

\end{document}

### 2023-01-08 main.tex
section split

--- 2023-01-08 LICS2023.tex
@@
 
-In the recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability,
-approximation bounds, 
-as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence.
-
-Notably, a significative effort has been made in order to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties.
-We can mention, for example,
-intersection type systems aimed at capturing time or space resources \cite{Beniamino}, or convergence probabilities \cite{UgoBreuvart, LICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthes} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for
-probabilistic \cite{PCOH, QBS} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
-Usually, the main reason to resort to such logical, type-theoretical or denotational approaches is to obtain methods which are \emph{modular} and \emph{compositional}, that is, where properties of complex algorithms can be deduced from properties of their component parts.
-
-\subsection{Two kinds of quantitative approaches}
-
-Among such quantitative approaches, two different directions have received a considerable attention in recent years.
-
-On the one hand there is the approach based on \emph{non-idempotent} intersection types \cite{} and their semantical counterpart, the so-called \emph{relational semantics} \cite{}, which have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{}, and have also been shown suitable to capture probabilistic properties \cite{Manzo}.
-
-On the other hand there is the approach of \emph{program metrics} \cite{} and \emph{quantitative equational theories} \cite{Mardare}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}.
-This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{} and coinductive methods \cite{}, and was recently extended to account for the the full $\lambda$-calculus \cite{LICS2021, FSCD2022}.
-
-In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{linearlogic} (i.e.~of using its input exactly once), plays a crucial role.
-In the relational semantics linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider \emph{analytic} maps, i.e.~formal infinite series of monomials of the form $a_{n}\cdot x^{n}$.
-By contrast, in metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances, while the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gabo2017}, and general programs as maps satisfying coarser
-continuity conditions (see e.g.~\cite{LICS2021}).
-
-A natural question is thus whether these two, apparently unrelated, ways of interpreting linearity and duplication can be somehow connected.
-At a first glance, there seems to be a fundamental ``logarithmic'' gap between the two approaches.
-In the relational semantics a typical ``duplicating'' map is obtained by composing the diagonal with multiplication:
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x\cdot y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the square product function $\lambda x.x^{2}$.
-However, in metric semantics this function needs not even exist (as these models are often restricted to Lipschitz-continuous maps \cite{Gabo2017})!
-Instead, a typical ``duplicating'' map can be obtained by composing the diagonal with the sum
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x+y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the linear (and Lipschitz) function $\lambda x.2x$.
-
-As this example seems to suggest, there seems to be a sort of ``logarithmic'' gap between the two approaches. Can this be made explicit?
-
-\subsection{Tropical Geometry: a ``logarithmic'' version of usual geometry}
-
-In very simple words, tropical geometry describes what happens to usual algebra and geometry when we replace ``$\cdot$'' by  ``$+$'', and ``$+$'' by ``$\min$''.
-It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
-Tropical geometry is a kind of \emph{idempotent} mathematics, since the role of addition is 
-played by the idempotent operation $\min$.
-For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$ corresponds in the tropical word correspond to the \emph{linear} function
-$$
-\trop p(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
-$$
-
-The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials).
-For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning \cite{}.
-
-Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for: a way to see polynomials as linear (i.e.~Lipschitz) functions.
-At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail.
-Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories, have also been observed \cite{}.
-
-The goal of this paper is to make such intuitive connections precise, yielding a way to relate, conceptually, two different quantitative approaches to higher-order algorithms, as well as suggesting new potential applications of tropical geometry to the study of quantitative properties of programs.
-
-\subsection{Contributions}
-
-Our contributions in this paper are threefold:
-\begin{itemize}
-
-\item We study the relational model over the tropical semi-ring, showing that
-\emph{bounded} simply typed terms correspond to tropical polynomials, while general
- simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{}.
-Moreover, we show that such maps are concave and locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
-Moreover, we exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps.
-
-\item Using the relational model as our main reference, we suggest a list of potential applications of tropical methods to the study of quantitative properties of non-deterministic/probabilistic functional programs like differential privacy, convergence \emph{log-probabilities} and the number of computation steps in the best case.
-
-\item We conclude by putting the correspondence between differential and metric properties made possible by the tropical viewpoint at the right level of generality. 
-By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{} and modules over the tropical semi-ring \cite{}, we show that the category of \emph{complete}  GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
-\end{itemize}
-%
-%\section{Bounded and Differential $\lambda$-Calculi}
-%
-%
-%Bounded Simply Typed $\lambda$-calculus $\BSTLC$:
-%$$
-%A::= o \mid !_{n}A \multimap A
-%$$
-%
-%
-%Resource Simply Typed $\lambda$-calculus $\RSTLC$:
-%$$
-%A::= o \mid [A, \dots , A] \multimap A
-%$$
-%
-%
-%Define a translation of types $(-)^{\C R}$ from $\BSTLC$ to $\RSTLC$ by $o^{\C R}=o$ and $(!_{n}A\multimap B)^{\C R}=
-%[\underbrace{A^{\C R},\dots, A^{\C R}}_{n\text{ times}}]\multimap B^{\C R}$.
-%
-%\begin{proposition}
-%$\Gamma \vdash_{\BSTLC} M:A$ implies 
-%$\Gamma^{\C R}\vdash_{\RSTLC}M:A^{\C R}$.
-%\end{proposition}
-%
+\input{section1}
 
 \section{Tropical Geometry}
-
-- The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.
-
-- tropical linear algebra: the category $\LREL$
-
-- tropical polynomials and tropical roots
-
-- tropical Laurent series and tropical roots
-
-- tropicalization of analytic functions and their roots: 
-$\trop P$
-
-- the differential operator
+\input{section2}
 
 \section{The Tropical Relational Model}
-
-Describe the model in a parametric way over a continuous semi-ring, as in Manzonetto et al.
-
-Define the tropicalization $\trop$ from the $\BB R$-based model
-to the $\Lawv$-based model.
-
-Focus on \emph{bounded} simply typed $\lambda$-calculus $\BSTLC$:
-$$
-A::= o \mid !_{n}A \multimap A
-$$
-The interpretation of a bounded type is defined as 
-$\model A= \Lawv^{\nodel A}$, where 
-$\nodel A$ is defined by $\nodel o=\B 1$ and 
-$\nodel{!_{n}A\multimap B}:= \C M_{\leq n}(\nodel A) \times \nodel B$.
-
-Bounded types are interesting
-because of the following proposition:
-\begin{proposition}
-For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) are polynomials.
-
-\end{proposition}
-\begin{proof}
-It suffices to check that $\nodel A$ is finite for all bounded types $A$.
-Indeed this implies that a morphism $f:\model A\to \model B$ corresponds to a finite matrix $\widehat f: \nodel A \times \nodel B \to \Lawv$.
-\end{proof}
-
-For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
-\begin{itemize}
-\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
-
-\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
-
-\end{itemize}
-
-Discuss the tropical Taylor expansion $\trop\C T(M)$.
+\input{section3}
 
 \section{Analytic Results}
-
-- the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric
-
-- concavity
-
-- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.
-
-- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
-\begin{itemize}
-
-\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally
-
-\item singular points: the famous example
-
-\item tropical roots
-
-\end{itemize}
-
-- discussion of the tropical Taylor expansion:
-
-- characterization of the functional metric using derivatives
+\input{section4}
 
 \section{Applications}
-
-General discussion: optimization properties behave in a Lipschitz way.
-
-- differential privacy and Lipschitzness
-
-- log-probabilities and tropical roots
-
-- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)
-
-- measuring duplications of discrete functions (needs finiteness!)
+\input{section5}
 
 \section{Quasi-metric Spaces and $\Lawv$-Modules}
-
-- correspondence between $\Lawv$-modules and $\Lawv$-categories.
-
-- exponential structure of $\LCAT$.
-
-- $\LCAT_{!}$ is a cartesian closed differential category
+\input{section6}
 
 \section{Related Work}
+\input{section7}
 
 \section{Conclusion}
+\input{section8}
 

### 2023-01-08 section1.tex
section split

--- 2023-01-08 LICS2023.tex
@@
-\documentclass[conference]{IEEEtran}
-\IEEEoverridecommandlockouts
-% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
-\usepackage{cite}
-\usepackage{amsmath,amssymb,amsfonts}
-\usepackage{algorithmic}
-\usepackage{graphicx}
-\usepackage{tikz-cd}
-\usepackage{textcomp}
-\usepackage{xcolor}
-\usepackage{amsthm}
-\usetikzlibrary{cd}
-\usepackage{stmaryrd}
-
-\input{macros}
-
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
-    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-\begin{document}
-
-\title{Tropical Geometry and the Lambda Calculus}
-
-\author{\IEEEauthorblockN{Davide Barbarossa}
-\IEEEauthorblockA{\textit{Universit\`a di Bologna}\\
-Bologna, Italy \\
-davide.barbarossa2@unibo.it}
-\and
-\IEEEauthorblockN{Paolo Pistone}
-\IEEEauthorblockA{\textit{Universit\`a Roma Tre}\\
-Rome, Italy\\
-paolo.pistone@uniroma3.it}
-}
-
-\maketitle
-
-\begin{abstract}
-
-\end{abstract}
-
-\begin{IEEEkeywords}
-\end{IEEEkeywords}
-
-\section{Introduction}
-
 In the recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability,
@@
 %
-
-\section{Tropical Geometry}
-
-- The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.
-
-- tropical linear algebra: the category $\LREL$
-
-- tropical polynomials and tropical roots
-
-- tropical Laurent series and tropical roots
-
-- tropicalization of analytic functions and their roots: 
-$\trop P$
-
-- the differential operator
-
-\section{The Tropical Relational Model}
-
-Describe the model in a parametric way over a continuous semi-ring, as in Manzonetto et al.
-
-Define the tropicalization $\trop$ from the $\BB R$-based model
-to the $\Lawv$-based model.
-
-Focus on \emph{bounded} simply typed $\lambda$-calculus $\BSTLC$:
-$$
-A::= o \mid !_{n}A \multimap A
-$$
-The interpretation of a bounded type is defined as 
-$\model A= \Lawv^{\nodel A}$, where 
-$\nodel A$ is defined by $\nodel o=\B 1$ and 
-$\nodel{!_{n}A\multimap B}:= \C M_{\leq n}(\nodel A) \times \nodel B$.
-
-Bounded types are interesting
-because of the following proposition:
-\begin{proposition}
-For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) are polynomials.
-
-\end{proposition}
-\begin{proof}
-It suffices to check that $\nodel A$ is finite for all bounded types $A$.
-Indeed this implies that a morphism $f:\model A\to \model B$ corresponds to a finite matrix $\widehat f: \nodel A \times \nodel B \to \Lawv$.
-\end{proof}
-
-For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
-\begin{itemize}
-\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
-
-\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
-
-\end{itemize}
-
-Discuss the tropical Taylor expansion $\trop\C T(M)$.
-
-\section{Analytic Results}
-
-- the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric
-
-- concavity
-
-- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.
-
-- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
-\begin{itemize}
-
-\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally
-
-\item singular points: the famous example
-
-\item tropical roots
-
-\end{itemize}
-
-- discussion of the tropical Taylor expansion:
-
-- characterization of the functional metric using derivatives
-
-\section{Applications}
-
-General discussion: optimization properties behave in a Lipschitz way.
-
-- differential privacy and Lipschitzness
-
-- log-probabilities and tropical roots
-
-- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)
-
-- measuring duplications of discrete functions (needs finiteness!)
-
-\section{Quasi-metric Spaces and $\Lawv$-Modules}
-
-- correspondence between $\Lawv$-modules and $\Lawv$-categories.
-
-- exponential structure of $\LCAT$.
-
-- $\LCAT_{!}$ is a cartesian closed differential category
-
-\section{Related Work}
-
-\section{Conclusion}
-
-\end{document}

### 2023-01-08 section2.tex
section split

--- 2023-01-08 LICS2023.tex
@@
-\documentclass[conference]{IEEEtran}
-\IEEEoverridecommandlockouts
-% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
-\usepackage{cite}
-\usepackage{amsmath,amssymb,amsfonts}
-\usepackage{algorithmic}
-\usepackage{graphicx}
-\usepackage{tikz-cd}
-\usepackage{textcomp}
-\usepackage{xcolor}
-\usepackage{amsthm}
-\usetikzlibrary{cd}
-\usepackage{stmaryrd}
-
-\input{macros}
-
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
-    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-\begin{document}
-
-\title{Tropical Geometry and the Lambda Calculus}
-
-\author{\IEEEauthorblockN{Davide Barbarossa}
-\IEEEauthorblockA{\textit{Universit\`a di Bologna}\\
-Bologna, Italy \\
-davide.barbarossa2@unibo.it}
-\and
-\IEEEauthorblockN{Paolo Pistone}
-\IEEEauthorblockA{\textit{Universit\`a Roma Tre}\\
-Rome, Italy\\
-paolo.pistone@uniroma3.it}
-}
-
-\maketitle
-
-\begin{abstract}
-
-\end{abstract}
-
-\begin{IEEEkeywords}
-\end{IEEEkeywords}
-
-\section{Introduction}
-
-In the recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability,
-approximation bounds, 
-as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence.
-
-Notably, a significative effort has been made in order to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties.
-We can mention, for example,
-intersection type systems aimed at capturing time or space resources \cite{Beniamino}, or convergence probabilities \cite{UgoBreuvart, LICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthes} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for
-probabilistic \cite{PCOH, QBS} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
-Usually, the main reason to resort to such logical, type-theoretical or denotational approaches is to obtain methods which are \emph{modular} and \emph{compositional}, that is, where properties of complex algorithms can be deduced from properties of their component parts.
-
-\subsection{Two kinds of quantitative approaches}
-
-Among such quantitative approaches, two different directions have received a considerable attention in recent years.
-
-On the one hand there is the approach based on \emph{non-idempotent} intersection types \cite{} and their semantical counterpart, the so-called \emph{relational semantics} \cite{}, which have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{}, and have also been shown suitable to capture probabilistic properties \cite{Manzo}.
-
-On the other hand there is the approach of \emph{program metrics} \cite{} and \emph{quantitative equational theories} \cite{Mardare}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}.
-This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{} and coinductive methods \cite{}, and was recently extended to account for the the full $\lambda$-calculus \cite{LICS2021, FSCD2022}.
-
-In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{linearlogic} (i.e.~of using its input exactly once), plays a crucial role.
-In the relational semantics linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider \emph{analytic} maps, i.e.~formal infinite series of monomials of the form $a_{n}\cdot x^{n}$.
-By contrast, in metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances, while the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gabo2017}, and general programs as maps satisfying coarser
-continuity conditions (see e.g.~\cite{LICS2021}).
-
-A natural question is thus whether these two, apparently unrelated, ways of interpreting linearity and duplication can be somehow connected.
-At a first glance, there seems to be a fundamental ``logarithmic'' gap between the two approaches.
-In the relational semantics a typical ``duplicating'' map is obtained by composing the diagonal with multiplication:
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x\cdot y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the square product function $\lambda x.x^{2}$.
-However, in metric semantics this function needs not even exist (as these models are often restricted to Lipschitz-continuous maps \cite{Gabo2017})!
-Instead, a typical ``duplicating'' map can be obtained by composing the diagonal with the sum
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x+y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the linear (and Lipschitz) function $\lambda x.2x$.
-
-As this example seems to suggest, there seems to be a sort of ``logarithmic'' gap between the two approaches. Can this be made explicit?
-
-\subsection{Tropical Geometry: a ``logarithmic'' version of usual geometry}
-
-In very simple words, tropical geometry describes what happens to usual algebra and geometry when we replace ``$\cdot$'' by  ``$+$'', and ``$+$'' by ``$\min$''.
-It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
-Tropical geometry is a kind of \emph{idempotent} mathematics, since the role of addition is 
-played by the idempotent operation $\min$.
-For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$ corresponds in the tropical word correspond to the \emph{linear} function
-$$
-\trop p(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
-$$
-
-The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials).
-For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning \cite{}.
-
-Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for: a way to see polynomials as linear (i.e.~Lipschitz) functions.
-At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail.
-Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories, have also been observed \cite{}.
-
-The goal of this paper is to make such intuitive connections precise, yielding a way to relate, conceptually, two different quantitative approaches to higher-order algorithms, as well as suggesting new potential applications of tropical geometry to the study of quantitative properties of programs.
-
-\subsection{Contributions}
-
-Our contributions in this paper are threefold:
-\begin{itemize}
-
-\item We study the relational model over the tropical semi-ring, showing that
-\emph{bounded} simply typed terms correspond to tropical polynomials, while general
- simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{}.
-Moreover, we show that such maps are concave and locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
-Moreover, we exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps.
-
-\item Using the relational model as our main reference, we suggest a list of potential applications of tropical methods to the study of quantitative properties of non-deterministic/probabilistic functional programs like differential privacy, convergence \emph{log-probabilities} and the number of computation steps in the best case.
-
-\item We conclude by putting the correspondence between differential and metric properties made possible by the tropical viewpoint at the right level of generality. 
-By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{} and modules over the tropical semi-ring \cite{}, we show that the category of \emph{complete}  GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
-\end{itemize}
-%
-%\section{Bounded and Differential $\lambda$-Calculi}
-%
-%
-%Bounded Simply Typed $\lambda$-calculus $\BSTLC$:
-%$$
-%A::= o \mid !_{n}A \multimap A
-%$$
-%
-%
-%Resource Simply Typed $\lambda$-calculus $\RSTLC$:
-%$$
-%A::= o \mid [A, \dots , A] \multimap A
-%$$
-%
-%
-%Define a translation of types $(-)^{\C R}$ from $\BSTLC$ to $\RSTLC$ by $o^{\C R}=o$ and $(!_{n}A\multimap B)^{\C R}=
-%[\underbrace{A^{\C R},\dots, A^{\C R}}_{n\text{ times}}]\multimap B^{\C R}$.
-%
-%\begin{proposition}
-%$\Gamma \vdash_{\BSTLC} M:A$ implies 
-%$\Gamma^{\C R}\vdash_{\RSTLC}M:A^{\C R}$.
-%\end{proposition}
-%
-
-\section{Tropical Geometry}
-
 - The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.
@@
 - the differential operator
-
-\section{The Tropical Relational Model}
-
-Describe the model in a parametric way over a continuous semi-ring, as in Manzonetto et al.
-
-Define the tropicalization $\trop$ from the $\BB R$-based model
-to the $\Lawv$-based model.
-
-Focus on \emph{bounded} simply typed $\lambda$-calculus $\BSTLC$:
-$$
-A::= o \mid !_{n}A \multimap A
-$$
-The interpretation of a bounded type is defined as 
-$\model A= \Lawv^{\nodel A}$, where 
-$\nodel A$ is defined by $\nodel o=\B 1$ and 
-$\nodel{!_{n}A\multimap B}:= \C M_{\leq n}(\nodel A) \times \nodel B$.
-
-Bounded types are interesting
-because of the following proposition:
-\begin{proposition}
-For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) are polynomials.
-
-\end{proposition}
-\begin{proof}
-It suffices to check that $\nodel A$ is finite for all bounded types $A$.
-Indeed this implies that a morphism $f:\model A\to \model B$ corresponds to a finite matrix $\widehat f: \nodel A \times \nodel B \to \Lawv$.
-\end{proof}
-
-For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
-\begin{itemize}
-\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
-
-\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
-
-\end{itemize}
-
-Discuss the tropical Taylor expansion $\trop\C T(M)$.
-
-\section{Analytic Results}
-
-- the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric
-
-- concavity
-
-- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.
-
-- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
-\begin{itemize}
-
-\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally
-
-\item singular points: the famous example
-
-\item tropical roots
-
-\end{itemize}
-
-- discussion of the tropical Taylor expansion:
-
-- characterization of the functional metric using derivatives
-
-\section{Applications}
-
-General discussion: optimization properties behave in a Lipschitz way.
-
-- differential privacy and Lipschitzness
-
-- log-probabilities and tropical roots
-
-- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)
-
-- measuring duplications of discrete functions (needs finiteness!)
-
-\section{Quasi-metric Spaces and $\Lawv$-Modules}
-
-- correspondence between $\Lawv$-modules and $\Lawv$-categories.
-
-- exponential structure of $\LCAT$.
-
-- $\LCAT_{!}$ is a cartesian closed differential category
-
-\section{Related Work}
-
-\section{Conclusion}
-
-\end{document}

### 2023-01-08 section3.tex
section split

--- 2023-01-08 LICS2023.tex
@@
-\documentclass[conference]{IEEEtran}
-\IEEEoverridecommandlockouts
-% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
-\usepackage{cite}
-\usepackage{amsmath,amssymb,amsfonts}
-\usepackage{algorithmic}
-\usepackage{graphicx}
-\usepackage{tikz-cd}
-\usepackage{textcomp}
-\usepackage{xcolor}
-\usepackage{amsthm}
-\usetikzlibrary{cd}
-\usepackage{stmaryrd}
-
-\input{macros}
-
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
-    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-\begin{document}
-
-\title{Tropical Geometry and the Lambda Calculus}
-
-\author{\IEEEauthorblockN{Davide Barbarossa}
-\IEEEauthorblockA{\textit{Universit\`a di Bologna}\\
-Bologna, Italy \\
-davide.barbarossa2@unibo.it}
-\and
-\IEEEauthorblockN{Paolo Pistone}
-\IEEEauthorblockA{\textit{Universit\`a Roma Tre}\\
-Rome, Italy\\
-paolo.pistone@uniroma3.it}
-}
-
-\maketitle
-
-\begin{abstract}
-
-\end{abstract}
-
-\begin{IEEEkeywords}
-\end{IEEEkeywords}
-
-\section{Introduction}
-
-In the recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability,
-approximation bounds, 
-as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence.
-
-Notably, a significative effort has been made in order to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties.
-We can mention, for example,
-intersection type systems aimed at capturing time or space resources \cite{Beniamino}, or convergence probabilities \cite{UgoBreuvart, LICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthes} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for
-probabilistic \cite{PCOH, QBS} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
-Usually, the main reason to resort to such logical, type-theoretical or denotational approaches is to obtain methods which are \emph{modular} and \emph{compositional}, that is, where properties of complex algorithms can be deduced from properties of their component parts.
-
-\subsection{Two kinds of quantitative approaches}
-
-Among such quantitative approaches, two different directions have received a considerable attention in recent years.
-
-On the one hand there is the approach based on \emph{non-idempotent} intersection types \cite{} and their semantical counterpart, the so-called \emph{relational semantics} \cite{}, which have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{}, and have also been shown suitable to capture probabilistic properties \cite{Manzo}.
-
-On the other hand there is the approach of \emph{program metrics} \cite{} and \emph{quantitative equational theories} \cite{Mardare}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}.
-This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{} and coinductive methods \cite{}, and was recently extended to account for the the full $\lambda$-calculus \cite{LICS2021, FSCD2022}.
-
-In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{linearlogic} (i.e.~of using its input exactly once), plays a crucial role.
-In the relational semantics linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider \emph{analytic} maps, i.e.~formal infinite series of monomials of the form $a_{n}\cdot x^{n}$.
-By contrast, in metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances, while the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gabo2017}, and general programs as maps satisfying coarser
-continuity conditions (see e.g.~\cite{LICS2021}).
-
-A natural question is thus whether these two, apparently unrelated, ways of interpreting linearity and duplication can be somehow connected.
-At a first glance, there seems to be a fundamental ``logarithmic'' gap between the two approaches.
-In the relational semantics a typical ``duplicating'' map is obtained by composing the diagonal with multiplication:
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x\cdot y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the square product function $\lambda x.x^{2}$.
-However, in metric semantics this function needs not even exist (as these models are often restricted to Lipschitz-continuous maps \cite{Gabo2017})!
-Instead, a typical ``duplicating'' map can be obtained by composing the diagonal with the sum
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x+y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the linear (and Lipschitz) function $\lambda x.2x$.
-
-As this example seems to suggest, there seems to be a sort of ``logarithmic'' gap between the two approaches. Can this be made explicit?
-
-\subsection{Tropical Geometry: a ``logarithmic'' version of usual geometry}
-
-In very simple words, tropical geometry describes what happens to usual algebra and geometry when we replace ``$\cdot$'' by  ``$+$'', and ``$+$'' by ``$\min$''.
-It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
-Tropical geometry is a kind of \emph{idempotent} mathematics, since the role of addition is 
-played by the idempotent operation $\min$.
-For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$ corresponds in the tropical word correspond to the \emph{linear} function
-$$
-\trop p(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
-$$
-
-The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials).
-For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning \cite{}.
-
-Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for: a way to see polynomials as linear (i.e.~Lipschitz) functions.
-At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail.
-Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories, have also been observed \cite{}.
-
-The goal of this paper is to make such intuitive connections precise, yielding a way to relate, conceptually, two different quantitative approaches to higher-order algorithms, as well as suggesting new potential applications of tropical geometry to the study of quantitative properties of programs.
-
-\subsection{Contributions}
-
-Our contributions in this paper are threefold:
-\begin{itemize}
-
-\item We study the relational model over the tropical semi-ring, showing that
-\emph{bounded} simply typed terms correspond to tropical polynomials, while general
- simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{}.
-Moreover, we show that such maps are concave and locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
-Moreover, we exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps.
-
-\item Using the relational model as our main reference, we suggest a list of potential applications of tropical methods to the study of quantitative properties of non-deterministic/probabilistic functional programs like differential privacy, convergence \emph{log-probabilities} and the number of computation steps in the best case.
-
-\item We conclude by putting the correspondence between differential and metric properties made possible by the tropical viewpoint at the right level of generality. 
-By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{} and modules over the tropical semi-ring \cite{}, we show that the category of \emph{complete}  GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
-\end{itemize}
-%
-%\section{Bounded and Differential $\lambda$-Calculi}
-%
-%
-%Bounded Simply Typed $\lambda$-calculus $\BSTLC$:
-%$$
-%A::= o \mid !_{n}A \multimap A
-%$$
-%
-%
-%Resource Simply Typed $\lambda$-calculus $\RSTLC$:
-%$$
-%A::= o \mid [A, \dots , A] \multimap A
-%$$
-%
-%
-%Define a translation of types $(-)^{\C R}$ from $\BSTLC$ to $\RSTLC$ by $o^{\C R}=o$ and $(!_{n}A\multimap B)^{\C R}=
-%[\underbrace{A^{\C R},\dots, A^{\C R}}_{n\text{ times}}]\multimap B^{\C R}$.
-%
-%\begin{proposition}
-%$\Gamma \vdash_{\BSTLC} M:A$ implies 
-%$\Gamma^{\C R}\vdash_{\RSTLC}M:A^{\C R}$.
-%\end{proposition}
-%
-
-\section{Tropical Geometry}
-
-- The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.
-
-- tropical linear algebra: the category $\LREL$
-
-- tropical polynomials and tropical roots
-
-- tropical Laurent series and tropical roots
-
-- tropicalization of analytic functions and their roots: 
-$\trop P$
-
-- the differential operator
-
-\section{The Tropical Relational Model}
-
 Describe the model in a parametric way over a continuous semi-ring, as in Manzonetto et al.
@@
 Discuss the tropical Taylor expansion $\trop\C T(M)$.
-
-\section{Analytic Results}
-
-- the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric
-
-- concavity
-
-- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.
-
-- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
-\begin{itemize}
-
-\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally
-
-\item singular points: the famous example
-
-\item tropical roots
-
-\end{itemize}
-
-- discussion of the tropical Taylor expansion:
-
-- characterization of the functional metric using derivatives
-
-\section{Applications}
-
-General discussion: optimization properties behave in a Lipschitz way.
-
-- differential privacy and Lipschitzness
-
-- log-probabilities and tropical roots
-
-- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)
-
-- measuring duplications of discrete functions (needs finiteness!)
-
-\section{Quasi-metric Spaces and $\Lawv$-Modules}
-
-- correspondence between $\Lawv$-modules and $\Lawv$-categories.
-
-- exponential structure of $\LCAT$.
-
-- $\LCAT_{!}$ is a cartesian closed differential category
-
-\section{Related Work}
-
-\section{Conclusion}
-
-\end{document}

### 2023-01-08 section4.tex
section split

--- 2023-01-08 LICS2023.tex
@@
-\documentclass[conference]{IEEEtran}
-\IEEEoverridecommandlockouts
-% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
-\usepackage{cite}
-\usepackage{amsmath,amssymb,amsfonts}
-\usepackage{algorithmic}
-\usepackage{graphicx}
-\usepackage{tikz-cd}
-\usepackage{textcomp}
-\usepackage{xcolor}
-\usepackage{amsthm}
-\usetikzlibrary{cd}
-\usepackage{stmaryrd}
-
-\input{macros}
-
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
-    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-\begin{document}
-
-\title{Tropical Geometry and the Lambda Calculus}
-
-\author{\IEEEauthorblockN{Davide Barbarossa}
-\IEEEauthorblockA{\textit{Universit\`a di Bologna}\\
-Bologna, Italy \\
-davide.barbarossa2@unibo.it}
-\and
-\IEEEauthorblockN{Paolo Pistone}
-\IEEEauthorblockA{\textit{Universit\`a Roma Tre}\\
-Rome, Italy\\
-paolo.pistone@uniroma3.it}
-}
-
-\maketitle
-
-\begin{abstract}
-
-\end{abstract}
-
-\begin{IEEEkeywords}
-\end{IEEEkeywords}
-
-\section{Introduction}
-
-In the recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability,
-approximation bounds, 
-as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence.
-
-Notably, a significative effort has been made in order to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties.
-We can mention, for example,
-intersection type systems aimed at capturing time or space resources \cite{Beniamino}, or convergence probabilities \cite{UgoBreuvart, LICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthes} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for
-probabilistic \cite{PCOH, QBS} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
-Usually, the main reason to resort to such logical, type-theoretical or denotational approaches is to obtain methods which are \emph{modular} and \emph{compositional}, that is, where properties of complex algorithms can be deduced from properties of their component parts.
-
-\subsection{Two kinds of quantitative approaches}
-
-Among such quantitative approaches, two different directions have received a considerable attention in recent years.
-
-On the one hand there is the approach based on \emph{non-idempotent} intersection types \cite{} and their semantical counterpart, the so-called \emph{relational semantics} \cite{}, which have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{}, and have also been shown suitable to capture probabilistic properties \cite{Manzo}.
-
-On the other hand there is the approach of \emph{program metrics} \cite{} and \emph{quantitative equational theories} \cite{Mardare}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}.
-This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{} and coinductive methods \cite{}, and was recently extended to account for the the full $\lambda$-calculus \cite{LICS2021, FSCD2022}.
-
-In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{linearlogic} (i.e.~of using its input exactly once), plays a crucial role.
-In the relational semantics linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider \emph{analytic} maps, i.e.~formal infinite series of monomials of the form $a_{n}\cdot x^{n}$.
-By contrast, in metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances, while the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gabo2017}, and general programs as maps satisfying coarser
-continuity conditions (see e.g.~\cite{LICS2021}).
-
-A natural question is thus whether these two, apparently unrelated, ways of interpreting linearity and duplication can be somehow connected.
-At a first glance, there seems to be a fundamental ``logarithmic'' gap between the two approaches.
-In the relational semantics a typical ``duplicating'' map is obtained by composing the diagonal with multiplication:
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x\cdot y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the square product function $\lambda x.x^{2}$.
-However, in metric semantics this function needs not even exist (as these models are often restricted to Lipschitz-continuous maps \cite{Gabo2017})!
-Instead, a typical ``duplicating'' map can be obtained by composing the diagonal with the sum
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x+y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the linear (and Lipschitz) function $\lambda x.2x$.
-
-As this example seems to suggest, there seems to be a sort of ``logarithmic'' gap between the two approaches. Can this be made explicit?
-
-\subsection{Tropical Geometry: a ``logarithmic'' version of usual geometry}
-
-In very simple words, tropical geometry describes what happens to usual algebra and geometry when we replace ``$\cdot$'' by  ``$+$'', and ``$+$'' by ``$\min$''.
-It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
-Tropical geometry is a kind of \emph{idempotent} mathematics, since the role of addition is 
-played by the idempotent operation $\min$.
-For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$ corresponds in the tropical word correspond to the \emph{linear} function
-$$
-\trop p(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
-$$
-
-The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials).
-For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning \cite{}.
-
-Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for: a way to see polynomials as linear (i.e.~Lipschitz) functions.
-At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail.
-Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories, have also been observed \cite{}.
-
-The goal of this paper is to make such intuitive connections precise, yielding a way to relate, conceptually, two different quantitative approaches to higher-order algorithms, as well as suggesting new potential applications of tropical geometry to the study of quantitative properties of programs.
-
-\subsection{Contributions}
-
-Our contributions in this paper are threefold:
-\begin{itemize}
-
-\item We study the relational model over the tropical semi-ring, showing that
-\emph{bounded} simply typed terms correspond to tropical polynomials, while general
- simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{}.
-Moreover, we show that such maps are concave and locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
-Moreover, we exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps.
-
-\item Using the relational model as our main reference, we suggest a list of potential applications of tropical methods to the study of quantitative properties of non-deterministic/probabilistic functional programs like differential privacy, convergence \emph{log-probabilities} and the number of computation steps in the best case.
-
-\item We conclude by putting the correspondence between differential and metric properties made possible by the tropical viewpoint at the right level of generality. 
-By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{} and modules over the tropical semi-ring \cite{}, we show that the category of \emph{complete}  GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
-\end{itemize}
-%
-%\section{Bounded and Differential $\lambda$-Calculi}
-%
-%
-%Bounded Simply Typed $\lambda$-calculus $\BSTLC$:
-%$$
-%A::= o \mid !_{n}A \multimap A
-%$$
-%
-%
-%Resource Simply Typed $\lambda$-calculus $\RSTLC$:
-%$$
-%A::= o \mid [A, \dots , A] \multimap A
-%$$
-%
-%
-%Define a translation of types $(-)^{\C R}$ from $\BSTLC$ to $\RSTLC$ by $o^{\C R}=o$ and $(!_{n}A\multimap B)^{\C R}=
-%[\underbrace{A^{\C R},\dots, A^{\C R}}_{n\text{ times}}]\multimap B^{\C R}$.
-%
-%\begin{proposition}
-%$\Gamma \vdash_{\BSTLC} M:A$ implies 
-%$\Gamma^{\C R}\vdash_{\RSTLC}M:A^{\C R}$.
-%\end{proposition}
-%
-
-\section{Tropical Geometry}
-
-- The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.
-
-- tropical linear algebra: the category $\LREL$
-
-- tropical polynomials and tropical roots
-
-- tropical Laurent series and tropical roots
-
-- tropicalization of analytic functions and their roots: 
-$\trop P$
-
-- the differential operator
-
-\section{The Tropical Relational Model}
-
-Describe the model in a parametric way over a continuous semi-ring, as in Manzonetto et al.
-
-Define the tropicalization $\trop$ from the $\BB R$-based model
-to the $\Lawv$-based model.
-
-Focus on \emph{bounded} simply typed $\lambda$-calculus $\BSTLC$:
-$$
-A::= o \mid !_{n}A \multimap A
-$$
-The interpretation of a bounded type is defined as 
-$\model A= \Lawv^{\nodel A}$, where 
-$\nodel A$ is defined by $\nodel o=\B 1$ and 
-$\nodel{!_{n}A\multimap B}:= \C M_{\leq n}(\nodel A) \times \nodel B$.
-
-Bounded types are interesting
-because of the following proposition:
-\begin{proposition}
-For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) are polynomials.
-
-\end{proposition}
-\begin{proof}
-It suffices to check that $\nodel A$ is finite for all bounded types $A$.
-Indeed this implies that a morphism $f:\model A\to \model B$ corresponds to a finite matrix $\widehat f: \nodel A \times \nodel B \to \Lawv$.
-\end{proof}
-
-For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
-\begin{itemize}
-\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
-
-\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
-
-\end{itemize}
-
-Discuss the tropical Taylor expansion $\trop\C T(M)$.
-
-\section{Analytic Results}
-
 - the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric
@@
 - characterization of the functional metric using derivatives
-
-\section{Applications}
-
-General discussion: optimization properties behave in a Lipschitz way.
-
-- differential privacy and Lipschitzness
-
-- log-probabilities and tropical roots
-
-- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)
-
-- measuring duplications of discrete functions (needs finiteness!)
-
-\section{Quasi-metric Spaces and $\Lawv$-Modules}
-
-- correspondence between $\Lawv$-modules and $\Lawv$-categories.
-
-- exponential structure of $\LCAT$.
-
-- $\LCAT_{!}$ is a cartesian closed differential category
-
-\section{Related Work}
-
-\section{Conclusion}
-
-\end{document}

### 2023-01-08 section5.tex
section split

--- 2023-01-08 LICS2023.tex
@@
-\documentclass[conference]{IEEEtran}
-\IEEEoverridecommandlockouts
-% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
-\usepackage{cite}
-\usepackage{amsmath,amssymb,amsfonts}
-\usepackage{algorithmic}
-\usepackage{graphicx}
-\usepackage{tikz-cd}
-\usepackage{textcomp}
-\usepackage{xcolor}
-\usepackage{amsthm}
-\usetikzlibrary{cd}
-\usepackage{stmaryrd}
-
-\input{macros}
-
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
-    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-\begin{document}
-
-\title{Tropical Geometry and the Lambda Calculus}
-
-\author{\IEEEauthorblockN{Davide Barbarossa}
-\IEEEauthorblockA{\textit{Universit\`a di Bologna}\\
-Bologna, Italy \\
-davide.barbarossa2@unibo.it}
-\and
-\IEEEauthorblockN{Paolo Pistone}
-\IEEEauthorblockA{\textit{Universit\`a Roma Tre}\\
-Rome, Italy\\
-paolo.pistone@uniroma3.it}
-}
-
-\maketitle
-
-\begin{abstract}
-
-\end{abstract}
-
-\begin{IEEEkeywords}
-\end{IEEEkeywords}
-
-\section{Introduction}
-
-In the recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability,
-approximation bounds, 
-as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence.
-
-Notably, a significative effort has been made in order to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties.
-We can mention, for example,
-intersection type systems aimed at capturing time or space resources \cite{Beniamino}, or convergence probabilities \cite{UgoBreuvart, LICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthes} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for
-probabilistic \cite{PCOH, QBS} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
-Usually, the main reason to resort to such logical, type-theoretical or denotational approaches is to obtain methods which are \emph{modular} and \emph{compositional}, that is, where properties of complex algorithms can be deduced from properties of their component parts.
-
-\subsection{Two kinds of quantitative approaches}
-
-Among such quantitative approaches, two different directions have received a considerable attention in recent years.
-
-On the one hand there is the approach based on \emph{non-idempotent} intersection types \cite{} and their semantical counterpart, the so-called \emph{relational semantics} \cite{}, which have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{}, and have also been shown suitable to capture probabilistic properties \cite{Manzo}.
-
-On the other hand there is the approach of \emph{program metrics} \cite{} and \emph{quantitative equational theories} \cite{Mardare}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}.
-This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{} and coinductive methods \cite{}, and was recently extended to account for the the full $\lambda$-calculus \cite{LICS2021, FSCD2022}.
-
-In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{linearlogic} (i.e.~of using its input exactly once), plays a crucial role.
-In the relational semantics linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider \emph{analytic} maps, i.e.~formal infinite series of monomials of the form $a_{n}\cdot x^{n}$.
-By contrast, in metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances, while the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gabo2017}, and general programs as maps satisfying coarser
-continuity conditions (see e.g.~\cite{LICS2021}).
-
-A natural question is thus whether these two, apparently unrelated, ways of interpreting linearity and duplication can be somehow connected.
-At a first glance, there seems to be a fundamental ``logarithmic'' gap between the two approaches.
-In the relational semantics a typical ``duplicating'' map is obtained by composing the diagonal with multiplication:
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x\cdot y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the square product function $\lambda x.x^{2}$.
-However, in metric semantics this function needs not even exist (as these models are often restricted to Lipschitz-continuous maps \cite{Gabo2017})!
-Instead, a typical ``duplicating'' map can be obtained by composing the diagonal with the sum
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x+y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the linear (and Lipschitz) function $\lambda x.2x$.
-
-As this example seems to suggest, there seems to be a sort of ``logarithmic'' gap between the two approaches. Can this be made explicit?
-
-\subsection{Tropical Geometry: a ``logarithmic'' version of usual geometry}
-
-In very simple words, tropical geometry describes what happens to usual algebra and geometry when we replace ``$\cdot$'' by  ``$+$'', and ``$+$'' by ``$\min$''.
-It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
-Tropical geometry is a kind of \emph{idempotent} mathematics, since the role of addition is 
-played by the idempotent operation $\min$.
-For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$ corresponds in the tropical word correspond to the \emph{linear} function
-$$
-\trop p(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
-$$
-
-The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials).
-For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning \cite{}.
-
-Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for: a way to see polynomials as linear (i.e.~Lipschitz) functions.
-At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail.
-Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories, have also been observed \cite{}.
-
-The goal of this paper is to make such intuitive connections precise, yielding a way to relate, conceptually, two different quantitative approaches to higher-order algorithms, as well as suggesting new potential applications of tropical geometry to the study of quantitative properties of programs.
-
-\subsection{Contributions}
-
-Our contributions in this paper are threefold:
-\begin{itemize}
-
-\item We study the relational model over the tropical semi-ring, showing that
-\emph{bounded} simply typed terms correspond to tropical polynomials, while general
- simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{}.
-Moreover, we show that such maps are concave and locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
-Moreover, we exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps.
-
-\item Using the relational model as our main reference, we suggest a list of potential applications of tropical methods to the study of quantitative properties of non-deterministic/probabilistic functional programs like differential privacy, convergence \emph{log-probabilities} and the number of computation steps in the best case.
-
-\item We conclude by putting the correspondence between differential and metric properties made possible by the tropical viewpoint at the right level of generality. 
-By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{} and modules over the tropical semi-ring \cite{}, we show that the category of \emph{complete}  GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
-\end{itemize}
-%
-%\section{Bounded and Differential $\lambda$-Calculi}
-%
-%
-%Bounded Simply Typed $\lambda$-calculus $\BSTLC$:
-%$$
-%A::= o \mid !_{n}A \multimap A
-%$$
-%
-%
-%Resource Simply Typed $\lambda$-calculus $\RSTLC$:
-%$$
-%A::= o \mid [A, \dots , A] \multimap A
-%$$
-%
-%
-%Define a translation of types $(-)^{\C R}$ from $\BSTLC$ to $\RSTLC$ by $o^{\C R}=o$ and $(!_{n}A\multimap B)^{\C R}=
-%[\underbrace{A^{\C R},\dots, A^{\C R}}_{n\text{ times}}]\multimap B^{\C R}$.
-%
-%\begin{proposition}
-%$\Gamma \vdash_{\BSTLC} M:A$ implies 
-%$\Gamma^{\C R}\vdash_{\RSTLC}M:A^{\C R}$.
-%\end{proposition}
-%
-
-\section{Tropical Geometry}
-
-- The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.
-
-- tropical linear algebra: the category $\LREL$
-
-- tropical polynomials and tropical roots
-
-- tropical Laurent series and tropical roots
-
-- tropicalization of analytic functions and their roots: 
-$\trop P$
-
-- the differential operator
-
-\section{The Tropical Relational Model}
-
-Describe the model in a parametric way over a continuous semi-ring, as in Manzonetto et al.
-
-Define the tropicalization $\trop$ from the $\BB R$-based model
-to the $\Lawv$-based model.
-
-Focus on \emph{bounded} simply typed $\lambda$-calculus $\BSTLC$:
-$$
-A::= o \mid !_{n}A \multimap A
-$$
-The interpretation of a bounded type is defined as 
-$\model A= \Lawv^{\nodel A}$, where 
-$\nodel A$ is defined by $\nodel o=\B 1$ and 
-$\nodel{!_{n}A\multimap B}:= \C M_{\leq n}(\nodel A) \times \nodel B$.
-
-Bounded types are interesting
-because of the following proposition:
-\begin{proposition}
-For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) are polynomials.
-
-\end{proposition}
-\begin{proof}
-It suffices to check that $\nodel A$ is finite for all bounded types $A$.
-Indeed this implies that a morphism $f:\model A\to \model B$ corresponds to a finite matrix $\widehat f: \nodel A \times \nodel B \to \Lawv$.
-\end{proof}
-
-For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
-\begin{itemize}
-\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
-
-\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
-
-\end{itemize}
-
-Discuss the tropical Taylor expansion $\trop\C T(M)$.
-
-\section{Analytic Results}
-
-- the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric
-
-- concavity
-
-- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.
-
-- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
-\begin{itemize}
-
-\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally
-
-\item singular points: the famous example
-
-\item tropical roots
-
-\end{itemize}
-
-- discussion of the tropical Taylor expansion:
-
-- characterization of the functional metric using derivatives
-
-\section{Applications}
-
 General discussion: optimization properties behave in a Lipschitz way.
@@
 - measuring duplications of discrete functions (needs finiteness!)
-
-\section{Quasi-metric Spaces and $\Lawv$-Modules}
-
-- correspondence between $\Lawv$-modules and $\Lawv$-categories.
-
-- exponential structure of $\LCAT$.
-
-- $\LCAT_{!}$ is a cartesian closed differential category
-
-\section{Related Work}
-
-\section{Conclusion}
-
-\end{document}

### 2023-01-08 main.tex
commit

@@
 Bologna, Italy \\
-davide.barbarossa2@unibo.it}
+davide.barbarossa@unibo.it}
 \and
@@
 
-\section{Tropical Geometry}
+\section{Two Quantitative Approaches to the $\lambda$-calculus}
 \input{section2}
 
-\section{The Tropical Relational Model}
+\section{Tropical Linear Algebra and Linear Programs}
 \input{section3}
 
-\section{Analytic Results}
+\section{Tropical Polynomials and Non-Linear Programs  }
 \input{section4}
 
-\section{Applications}
+\section{Analytic Results}
 \input{section5}
 
-\section{Quasi-metric Spaces and $\Lawv$-Modules}
+\section{Applications}
 \input{section6}
 
-\section{Related Work}
+\section{Quasi-metric Spaces and $\Lawv$-Modules}
 \input{section7}
 
-\section{Conclusion}
+\section{Related Work}
 \input{section8}
 
+\section{Conclusion}
+
+\input{section9}
+
 \end{document}

### 2023-01-08 section2.tex
commit

@@
-- The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.
+Recall the two approaches with more details on lambda-calculus and on existing challenges.
 
-- tropical linear algebra: the category $\LREL$
+- Metric models: functions as Lipschitz maps, differential privacy.
 
-- tropical polynomials and tropical roots
+- Problem with unrestricted duplications
 
-- tropical Laurent series and tropical roots
+- Bounded Linear Logic/lambda-calculus, why Met is only a model of BLL, and why it is difficult (but not impossible!) to construct models of full lambda
 
-- tropicalization of analytic functions and their roots: 
-$\trop P$
+- Differential lambda-calculus: differential operator, Taylor expansion (vs. 1)
 
-- the differential operator
+- resource lambda-terms, Taylor expansion (vs. 2)
+
+- resource lambdas are polynomials, so not linear. Why we need tropical geometry here.

### 2023-01-08 section3.tex
commit

@@
-Describe the model in a parametric way over a continuous semi-ring, as in Manzonetto et al.
+- matrices over a continuous semi-ring
 
-Define the tropicalization $\trop$ from the $\BB R$-based model
-to the $\Lawv$-based model.
+- The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.
 
-Focus on \emph{bounded} simply typed $\lambda$-calculus $\BSTLC$:
+- tropical linear algebra: the category $\LREL$
+
+- linear computation inside $\LREL$: linear lambda-terms through the symmetric monoidal adjunction 
 $$
-A::= o \mid !_{n}A \multimap A
+\LREL(Z\otimes X, Y) \simeq \LREL (Z, X\multimap Y)
 $$
-The interpretation of a bounded type is defined as 
-$\model A= \Lawv^{\nodel A}$, where 
-$\nodel A$ is defined by $\nodel o=\B 1$ and 
-$\nodel{!_{n}A\multimap B}:= \C M_{\leq n}(\nodel A) \times \nodel B$.
-
-Bounded types are interesting
-because of the following proposition:
-\begin{proposition}
-For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) are polynomials.
-
-\end{proposition}
-\begin{proof}
-It suffices to check that $\nodel A$ is finite for all bounded types $A$.
-Indeed this implies that a morphism $f:\model A\to \model B$ corresponds to a finite matrix $\widehat f: \nodel A \times \nodel B \to \Lawv$.
-\end{proof}
-
-For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
-\begin{itemize}
-\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
-
-\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
-
-\end{itemize}
-
-Discuss the tropical Taylor expansion $\trop\C T(M)$.
+(with $\LREL(X,Y)\simeq \Lawv^{X\times Y}$.

### 2023-01-08 section4.tex
commit

@@
-- the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric
+- non-linear programs correspond to non-linear maps.
 
-- concavity
+- tropical polynomials and tropical roots -> bounded lambda-terms
+(the operation $\C M_{\leq n}(-)$ defines a graded linear exponential comonad) via the adjunction
+$$
+\LREL(Z\otimes !_{\leq n}X,Y) \simeq \LREL(Z, !_{\leq n}X\multimap Y) 
+$$
 
-- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.
+- tropical Laurent series and tropical roots -> unbounded lambda-terms (the operation $\C M(-)$ defines a linear exponential comonad) via the adjunction
+$$
+\LREL(!(Z\& X),Y) \simeq \LREL(!Z, !X\multimap Y) 
+$$
 
-- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
+- tropicalization of analytic functions and their roots: 
+$\trop P$
+
+- the differential operator: formal differentiation of tropical Laurent series. Differentiation in $\LREL_{!}$.
+
+- $\LREL_{!}$ is a differential CCC satisfying Taylor. 
+Discuss the tropical Taylor expansion $\trop\C T(M)$.
+
+Focus on \emph{bounded} simply typed $\lambda$-calculus $\BSTLC$:
+$$
+A::= o \mid !_{n}A \multimap A
+$$
+The interpretation of a bounded type is defined as 
+$\model A= \Lawv^{\nodel A}$, where 
+$\nodel A$ is defined by $\nodel o=\B 1$ and 
+$\nodel{!_{n}A\multimap B}:= \C M_{\leq n}(\nodel A) \times \nodel B$.
+
+Bounded types are interesting
+because of the following proposition:
+\begin{proposition}
+For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) are polynomials.
+
+\end{proposition}
+\begin{proof}
+It suffices to check that $\nodel A$ is finite for all bounded types $A$.
+Indeed this implies that a morphism $f:\model A\to \model B$ corresponds to a finite matrix $\widehat f: \nodel A \times \nodel B \to \Lawv$.
+\end{proof}
+
+For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
 \begin{itemize}
+\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
 
-\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally
-
-\item singular points: the famous example
-
-\item tropical roots
+\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
 
 \end{itemize}
-
-- discussion of the tropical Taylor expansion:
-
-- characterization of the functional metric using derivatives

### 2023-01-08 section5.tex
commit

@@
-General discussion: optimization properties behave in a Lipschitz way.
+- the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric
 
-- differential privacy and Lipschitzness
+- concavity
 
-- log-probabilities and tropical roots
+- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.
 
-- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)
+- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
+\begin{itemize}
 
-- measuring duplications of discrete functions (needs finiteness!)
+\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally
+
+\item singular points: the famous example
+
+\item tropical roots
+
+\end{itemize}
+
+- discussion of the tropical Taylor expansion:
+
+- characterization of the functional metric using derivatives

### 2023-01-08 section7.tex
commit

--- 2023-01-08 LICS2023.tex
@@
-\documentclass[conference]{IEEEtran}
-\IEEEoverridecommandlockouts
-% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
-\usepackage{cite}
-\usepackage{amsmath,amssymb,amsfonts}
-\usepackage{algorithmic}
-\usepackage{graphicx}
-\usepackage{tikz-cd}
-\usepackage{textcomp}
-\usepackage{xcolor}
-\usepackage{amsthm}
-\usetikzlibrary{cd}
-\usepackage{stmaryrd}
-
-\input{macros}
-
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
-    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-\begin{document}
-
-\title{Tropical Geometry and the Lambda Calculus}
-
-\author{\IEEEauthorblockN{Davide Barbarossa}
-\IEEEauthorblockA{\textit{Universit\`a di Bologna}\\
-Bologna, Italy \\
-davide.barbarossa2@unibo.it}
-\and
-\IEEEauthorblockN{Paolo Pistone}
-\IEEEauthorblockA{\textit{Universit\`a Roma Tre}\\
-Rome, Italy\\
-paolo.pistone@uniroma3.it}
-}
-
-\maketitle
-
-\begin{abstract}
-
-\end{abstract}
-
-\begin{IEEEkeywords}
-\end{IEEEkeywords}
-
-\section{Introduction}
-
-In the recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability,
-approximation bounds, 
-as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence.
-
-Notably, a significative effort has been made in order to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties.
-We can mention, for example,
-intersection type systems aimed at capturing time or space resources \cite{Beniamino}, or convergence probabilities \cite{UgoBreuvart, LICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthes} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for
-probabilistic \cite{PCOH, QBS} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
-Usually, the main reason to resort to such logical, type-theoretical or denotational approaches is to obtain methods which are \emph{modular} and \emph{compositional}, that is, where properties of complex algorithms can be deduced from properties of their component parts.
-
-\subsection{Two kinds of quantitative approaches}
-
-Among such quantitative approaches, two different directions have received a considerable attention in recent years.
-
-On the one hand there is the approach based on \emph{non-idempotent} intersection types \cite{} and their semantical counterpart, the so-called \emph{relational semantics} \cite{}, which have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{}, and have also been shown suitable to capture probabilistic properties \cite{Manzo}.
-
-On the other hand there is the approach of \emph{program metrics} \cite{} and \emph{quantitative equational theories} \cite{Mardare}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}.
-This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{} and coinductive methods \cite{}, and was recently extended to account for the the full $\lambda$-calculus \cite{LICS2021, FSCD2022}.
-
-In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{linearlogic} (i.e.~of using its input exactly once), plays a crucial role.
-In the relational semantics linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider \emph{analytic} maps, i.e.~formal infinite series of monomials of the form $a_{n}\cdot x^{n}$.
-By contrast, in metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances, while the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gabo2017}, and general programs as maps satisfying coarser
-continuity conditions (see e.g.~\cite{LICS2021}).
-
-A natural question is thus whether these two, apparently unrelated, ways of interpreting linearity and duplication can be somehow connected.
-At a first glance, there seems to be a fundamental ``logarithmic'' gap between the two approaches.
-In the relational semantics a typical ``duplicating'' map is obtained by composing the diagonal with multiplication:
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x\cdot y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the square product function $\lambda x.x^{2}$.
-However, in metric semantics this function needs not even exist (as these models are often restricted to Lipschitz-continuous maps \cite{Gabo2017})!
-Instead, a typical ``duplicating'' map can be obtained by composing the diagonal with the sum
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x+y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the linear (and Lipschitz) function $\lambda x.2x$.
-
-As this example seems to suggest, there seems to be a sort of ``logarithmic'' gap between the two approaches. Can this be made explicit?
-
-\subsection{Tropical Geometry: a ``logarithmic'' version of usual geometry}
-
-In very simple words, tropical geometry describes what happens to usual algebra and geometry when we replace ``$\cdot$'' by  ``$+$'', and ``$+$'' by ``$\min$''.
-It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
-Tropical geometry is a kind of \emph{idempotent} mathematics, since the role of addition is 
-played by the idempotent operation $\min$.
-For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$ corresponds in the tropical word correspond to the \emph{linear} function
-$$
-\trop p(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
-$$
-
-The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials).
-For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning \cite{}.
-
-Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for: a way to see polynomials as linear (i.e.~Lipschitz) functions.
-At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail.
-Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories, have also been observed \cite{}.
-
-The goal of this paper is to make such intuitive connections precise, yielding a way to relate, conceptually, two different quantitative approaches to higher-order algorithms, as well as suggesting new potential applications of tropical geometry to the study of quantitative properties of programs.
-
-\subsection{Contributions}
-
-Our contributions in this paper are threefold:
-\begin{itemize}
-
-\item We study the relational model over the tropical semi-ring, showing that
-\emph{bounded} simply typed terms correspond to tropical polynomials, while general
- simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{}.
-Moreover, we show that such maps are concave and locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
-Moreover, we exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps.
-
-\item Using the relational model as our main reference, we suggest a list of potential applications of tropical methods to the study of quantitative properties of non-deterministic/probabilistic functional programs like differential privacy, convergence \emph{log-probabilities} and the number of computation steps in the best case.
-
-\item We conclude by putting the correspondence between differential and metric properties made possible by the tropical viewpoint at the right level of generality. 
-By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{} and modules over the tropical semi-ring \cite{}, we show that the category of \emph{complete}  GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
-\end{itemize}
-%
-%\section{Bounded and Differential $\lambda$-Calculi}
-%
-%
-%Bounded Simply Typed $\lambda$-calculus $\BSTLC$:
-%$$
-%A::= o \mid !_{n}A \multimap A
-%$$
-%
-%
-%Resource Simply Typed $\lambda$-calculus $\RSTLC$:
-%$$
-%A::= o \mid [A, \dots , A] \multimap A
-%$$
-%
-%
-%Define a translation of types $(-)^{\C R}$ from $\BSTLC$ to $\RSTLC$ by $o^{\C R}=o$ and $(!_{n}A\multimap B)^{\C R}=
-%[\underbrace{A^{\C R},\dots, A^{\C R}}_{n\text{ times}}]\multimap B^{\C R}$.
-%
-%\begin{proposition}
-%$\Gamma \vdash_{\BSTLC} M:A$ implies 
-%$\Gamma^{\C R}\vdash_{\RSTLC}M:A^{\C R}$.
-%\end{proposition}
-%
-
-\section{Tropical Geometry}
-
-- The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.
-
-- tropical linear algebra: the category $\LREL$
-
-- tropical polynomials and tropical roots
-
-- tropical Laurent series and tropical roots
-
-- tropicalization of analytic functions and their roots: 
-$\trop P$
-
-- the differential operator
-
-\section{The Tropical Relational Model}
-
-Describe the model in a parametric way over a continuous semi-ring, as in Manzonetto et al.
-
-Define the tropicalization $\trop$ from the $\BB R$-based model
-to the $\Lawv$-based model.
-
-Focus on \emph{bounded} simply typed $\lambda$-calculus $\BSTLC$:
-$$
-A::= o \mid !_{n}A \multimap A
-$$
-The interpretation of a bounded type is defined as 
-$\model A= \Lawv^{\nodel A}$, where 
-$\nodel A$ is defined by $\nodel o=\B 1$ and 
-$\nodel{!_{n}A\multimap B}:= \C M_{\leq n}(\nodel A) \times \nodel B$.
-
-Bounded types are interesting
-because of the following proposition:
-\begin{proposition}
-For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) are polynomials.
-
-\end{proposition}
-\begin{proof}
-It suffices to check that $\nodel A$ is finite for all bounded types $A$.
-Indeed this implies that a morphism $f:\model A\to \model B$ corresponds to a finite matrix $\widehat f: \nodel A \times \nodel B \to \Lawv$.
-\end{proof}
-
-For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
-\begin{itemize}
-\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
-
-\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
-
-\end{itemize}
-
-Discuss the tropical Taylor expansion $\trop\C T(M)$.
-
-\section{Analytic Results}
-
-- the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric
-
-- concavity
-
-- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.
-
-- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
-\begin{itemize}
-
-\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally
-
-\item singular points: the famous example
-
-\item tropical roots
-
-\end{itemize}
-
-- discussion of the tropical Taylor expansion:
-
-- characterization of the functional metric using derivatives
-
-\section{Applications}
-
-General discussion: optimization properties behave in a Lipschitz way.
-
-- differential privacy and Lipschitzness
-
-- log-probabilities and tropical roots
-
-- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)
-
-- measuring duplications of discrete functions (needs finiteness!)
-
-\section{Quasi-metric Spaces and $\Lawv$-Modules}
-
 - correspondence between $\Lawv$-modules and $\Lawv$-categories.
@@
 - $\LCAT_{!}$ is a cartesian closed differential category
-
-\section{Related Work}
-
-\section{Conclusion}
-
-\end{document}

### 2023-01-09 main.tex
Nuove modifiche

@@
 
-\title{Tropical Geometry and the Lambda Calculus}
+\title{Lambda Calculus through Tropical Mathematics}
 
@@
 \section{Two Quantitative Approaches to the $\lambda$-calculus}
+
 \input{section2}
 
-\section{Tropical Linear Algebra and Linear Programs}
+\section{Tropical (non-)Linear Algebra and computation}
+
+Di cosa stiamo parlando. Ovvero:
+
+- matrices over a continuous semi-ring
+
+- The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.
+
+- tropical linear algebra: the category $\LREL$
+
+\subsection{As a model of linear programs}
+
 \input{section3}
 
-\section{Tropical Polynomials and Non-Linear Programs  }
+\subsection{As a model of non-linear programs}
+
 \input{section4}
+
+\subsection{As a model of differential programs}
+
+- differential categories
+
+- the tropical differential operator: formal differentiation of tropical Laurent series. Differentiation in $\LREL_{!}$.
+
+- $\LREL_{!}$ is a differential CCC satisfying Taylor. 
+Discuss the tropical Taylor expansion $\trop\C T(M)$.
 
@@
 
-\section{Related Work}
+\section{Related Work (vale la pena? O meglio nelle conclusioni?)}
 \input{section8}

### 2023-01-09 section2.tex
Nuove modifiche

@@
-Recall the two approaches with more details on lambda-calculus and on existing challenges.
+%Recall the two approaches with more details on lambda-calculus and on existing challenges.
+
+\subsection{Metric approach: bounded $\lambda$-terms as Lipschitz functions}
 
@@
 
+\subsection{Resource approach: differential $\lambda$-terms as polynomials}
+
 - Differential lambda-calculus: differential operator, Taylor expansion (vs. 1)
@@
 
-- resource lambdas are polynomials, so not linear. Why we need tropical geometry here.
+- resource lambdas are polynomials, so not linear.
+
+\subsection{Tropical mathematics: a possible synthesis}
+
+- generalities about tropical maths (tropicalisation $\trop P$  of polynomials and of Laurent series, and their roots -- all that without $\LREL$)

### 2023-01-09 section4.tex
Nuove modifiche

@@
+\subsubsection{Bounded}
+
+Quanto vogliamo essere precisi in questa sottosottosezione ?
+
 - non-linear programs correspond to non-linear maps.
-
-- tropical polynomials and tropical roots -> bounded lambda-terms
-(the operation $\C M_{\leq n}(-)$ defines a graded linear exponential comonad) via the adjunction
-$$
-\LREL(Z\otimes !_{\leq n}X,Y) \simeq \LREL(Z, !_{\leq n}X\multimap Y) 
-$$
-
-- tropical Laurent series and tropical roots -> unbounded lambda-terms (the operation $\C M(-)$ defines a linear exponential comonad) via the adjunction
-$$
-\LREL(!(Z\& X),Y) \simeq \LREL(!Z, !X\multimap Y) 
-$$
-
-- tropicalization of analytic functions and their roots: 
-$\trop P$
-
-- the differential operator: formal differentiation of tropical Laurent series. Differentiation in $\LREL_{!}$.
-
-- $\LREL_{!}$ is a differential CCC satisfying Taylor. 
-Discuss the tropical Taylor expansion $\trop\C T(M)$.
 
@@
 
+- tropical polynomials and tropical roots $->$ bounded lambda-terms
+(the operation $\C M_{\leq n}(-)$ defines a graded linear exponential comonad) via the adjunction
+$$
+\LREL(Z\otimes !_{\leq n}X,Y) \simeq \LREL(Z, !_{\leq n}X\multimap Y) 
+$$
+
+\subsubsection{Unbounded}
+
+- tropical Laurent series and tropical roots $->$ unbounded lambda-terms (the operation $\C M(-)$ defines a linear exponential comonad) via the adjunction
+$$
+\LREL(!(Z\& X),Y) \simeq \LREL(!Z, !X\multimap Y) 
+$$
+
+That is:
+
+- $\LREL_!$ is CCC (Giulio)
+
 \end{itemize}

### 2023-01-09 section3C.tex
Sezioni riorganizzate
Ora tutte le sezioni del sorgente dovrebbero corrispondere alle sezioni del pdf

- differential categories

- the tropical differential operator: formal differentiation of tropical Laurent series. Differentiation in $\LREL_{!}$.

- $\LREL_{!}$ is a differential CCC satisfying Taylor. 
Discuss the tropical Taylor expansion $\trop\C T(M)$.

### 2023-01-09 section3intro.tex
Sezioni riorganizzate
Ora tutte le sezioni del sorgente dovrebbero corrispondere alle sezioni del pdf

--- 2023-01-08 LICS2023.tex
@@
-\documentclass[conference]{IEEEtran}
-\IEEEoverridecommandlockouts
-% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
-\usepackage{cite}
-\usepackage{amsmath,amssymb,amsfonts}
-\usepackage{algorithmic}
-\usepackage{graphicx}
-\usepackage{tikz-cd}
-\usepackage{textcomp}
-\usepackage{xcolor}
-\usepackage{amsthm}
-\usetikzlibrary{cd}
-\usepackage{stmaryrd}
+Di cosa stiamo parlando. Ovvero:
 
-\input{macros}
-
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
-    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-\begin{document}
-
-\title{Tropical Geometry and the Lambda Calculus}
-
-\author{\IEEEauthorblockN{Davide Barbarossa}
-\IEEEauthorblockA{\textit{Universit\`a di Bologna}\\
-Bologna, Italy \\
-davide.barbarossa2@unibo.it}
-\and
-\IEEEauthorblockN{Paolo Pistone}
-\IEEEauthorblockA{\textit{Universit\`a Roma Tre}\\
-Rome, Italy\\
-paolo.pistone@uniroma3.it}
-}
-
-\maketitle
-
-\begin{abstract}
-
-\end{abstract}
-
-\begin{IEEEkeywords}
-\end{IEEEkeywords}
-
-\section{Introduction}
-
-In the recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability,
-approximation bounds, 
-as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence.
-
-Notably, a significative effort has been made in order to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties.
-We can mention, for example,
-intersection type systems aimed at capturing time or space resources \cite{Beniamino}, or convergence probabilities \cite{UgoBreuvart, LICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthes} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for
-probabilistic \cite{PCOH, QBS} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
-Usually, the main reason to resort to such logical, type-theoretical or denotational approaches is to obtain methods which are \emph{modular} and \emph{compositional}, that is, where properties of complex algorithms can be deduced from properties of their component parts.
-
-\subsection{Two kinds of quantitative approaches}
-
-Among such quantitative approaches, two different directions have received a considerable attention in recent years.
-
-On the one hand there is the approach based on \emph{non-idempotent} intersection types \cite{} and their semantical counterpart, the so-called \emph{relational semantics} \cite{}, which have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{}, and have also been shown suitable to capture probabilistic properties \cite{Manzo}.
-
-On the other hand there is the approach of \emph{program metrics} \cite{} and \emph{quantitative equational theories} \cite{Mardare}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}.
-This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{} and coinductive methods \cite{}, and was recently extended to account for the the full $\lambda$-calculus \cite{LICS2021, FSCD2022}.
-
-In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{linearlogic} (i.e.~of using its input exactly once), plays a crucial role.
-In the relational semantics linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider \emph{analytic} maps, i.e.~formal infinite series of monomials of the form $a_{n}\cdot x^{n}$.
-By contrast, in metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances, while the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gabo2017}, and general programs as maps satisfying coarser
-continuity conditions (see e.g.~\cite{LICS2021}).
-
-A natural question is thus whether these two, apparently unrelated, ways of interpreting linearity and duplication can be somehow connected.
-At a first glance, there seems to be a fundamental ``logarithmic'' gap between the two approaches.
-In the relational semantics a typical ``duplicating'' map is obtained by composing the diagonal with multiplication:
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x\cdot y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the square product function $\lambda x.x^{2}$.
-However, in metric semantics this function needs not even exist (as these models are often restricted to Lipschitz-continuous maps \cite{Gabo2017})!
-Instead, a typical ``duplicating'' map can be obtained by composing the diagonal with the sum
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x+y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the linear (and Lipschitz) function $\lambda x.2x$.
-
-As this example seems to suggest, there seems to be a sort of ``logarithmic'' gap between the two approaches. Can this be made explicit?
-
-\subsection{Tropical Geometry: a ``logarithmic'' version of usual geometry}
-
-In very simple words, tropical geometry describes what happens to usual algebra and geometry when we replace ``$\cdot$'' by  ``$+$'', and ``$+$'' by ``$\min$''.
-It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
-Tropical geometry is a kind of \emph{idempotent} mathematics, since the role of addition is 
-played by the idempotent operation $\min$.
-For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$ corresponds in the tropical word correspond to the \emph{linear} function
-$$
-\trop p(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
-$$
-
-The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials).
-For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning \cite{}.
-
-Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for: a way to see polynomials as linear (i.e.~Lipschitz) functions.
-At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail.
-Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories, have also been observed \cite{}.
-
-The goal of this paper is to make such intuitive connections precise, yielding a way to relate, conceptually, two different quantitative approaches to higher-order algorithms, as well as suggesting new potential applications of tropical geometry to the study of quantitative properties of programs.
-
-\subsection{Contributions}
-
-Our contributions in this paper are threefold:
-\begin{itemize}
-
-\item We study the relational model over the tropical semi-ring, showing that
-\emph{bounded} simply typed terms correspond to tropical polynomials, while general
- simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{}.
-Moreover, we show that such maps are concave and locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
-Moreover, we exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps.
-
-\item Using the relational model as our main reference, we suggest a list of potential applications of tropical methods to the study of quantitative properties of non-deterministic/probabilistic functional programs like differential privacy, convergence \emph{log-probabilities} and the number of computation steps in the best case.
-
-\item We conclude by putting the correspondence between differential and metric properties made possible by the tropical viewpoint at the right level of generality. 
-By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{} and modules over the tropical semi-ring \cite{}, we show that the category of \emph{complete}  GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
-\end{itemize}
-%
-%\section{Bounded and Differential $\lambda$-Calculi}
-%
-%
-%Bounded Simply Typed $\lambda$-calculus $\BSTLC$:
-%$$
-%A::= o \mid !_{n}A \multimap A
-%$$
-%
-%
-%Resource Simply Typed $\lambda$-calculus $\RSTLC$:
-%$$
-%A::= o \mid [A, \dots , A] \multimap A
-%$$
-%
-%
-%Define a translation of types $(-)^{\C R}$ from $\BSTLC$ to $\RSTLC$ by $o^{\C R}=o$ and $(!_{n}A\multimap B)^{\C R}=
-%[\underbrace{A^{\C R},\dots, A^{\C R}}_{n\text{ times}}]\multimap B^{\C R}$.
-%
-%\begin{proposition}
-%$\Gamma \vdash_{\BSTLC} M:A$ implies 
-%$\Gamma^{\C R}\vdash_{\RSTLC}M:A^{\C R}$.
-%\end{proposition}
-%
-
-\section{Tropical Geometry}
+- matrices over a continuous semi-ring
 
@@
 - tropical linear algebra: the category $\LREL$
-
-- tropical polynomials and tropical roots
-
-- tropical Laurent series and tropical roots
-
-- tropicalization of analytic functions and their roots: 
-$\trop P$
-
-- the differential operator
-
-\section{The Tropical Relational Model}
-
-Describe the model in a parametric way over a continuous semi-ring, as in Manzonetto et al.
-
-Define the tropicalization $\trop$ from the $\BB R$-based model
-to the $\Lawv$-based model.
-
-Focus on \emph{bounded} simply typed $\lambda$-calculus $\BSTLC$:
-$$
-A::= o \mid !_{n}A \multimap A
-$$
-The interpretation of a bounded type is defined as 
-$\model A= \Lawv^{\nodel A}$, where 
-$\nodel A$ is defined by $\nodel o=\B 1$ and 
-$\nodel{!_{n}A\multimap B}:= \C M_{\leq n}(\nodel A) \times \nodel B$.
-
-Bounded types are interesting
-because of the following proposition:
-\begin{proposition}
-For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) are polynomials.
-
-\end{proposition}
-\begin{proof}
-It suffices to check that $\nodel A$ is finite for all bounded types $A$.
-Indeed this implies that a morphism $f:\model A\to \model B$ corresponds to a finite matrix $\widehat f: \nodel A \times \nodel B \to \Lawv$.
-\end{proof}
-
-For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
-\begin{itemize}
-\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
-
-\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
-
-\end{itemize}
-
-Discuss the tropical Taylor expansion $\trop\C T(M)$.
-
-\section{Analytic Results}
-
-- the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric
-
-- concavity
-
-- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.
-
-- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
-\begin{itemize}
-
-\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally
-
-\item singular points: the famous example
-
-\item tropical roots
-
-\end{itemize}
-
-- discussion of the tropical Taylor expansion:
-
-- characterization of the functional metric using derivatives
-
-\section{Applications}
-
-General discussion: optimization properties behave in a Lipschitz way.
-
-- differential privacy and Lipschitzness
-
-- log-probabilities and tropical roots
-
-- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)
-
-- measuring duplications of discrete functions (needs finiteness!)
-
-\section{Quasi-metric Spaces and $\Lawv$-Modules}
-
-- correspondence between $\Lawv$-modules and $\Lawv$-categories.
-
-- exponential structure of $\LCAT$.
-
-- $\LCAT_{!}$ is a cartesian closed differential category
-
-\section{Related Work}
-
-\section{Conclusion}
-
-\end{document}

### 2023-01-10 main.tex
Sezioni riorganizzate
Ora tutte le sezioni del sorgente dovrebbero corrispondere alle sezioni del pdf
Push serale
Prime parti di section3.
Inoltre ho aggiunto un \label in main.tex ed qualche macros utile.

--- 2023-01-08 main.tex
@@
 \section{Introduction}
-
 \input{section1}
@@
 
-\section{Tropical Linear Algebra and Linear Programs}
-\input{section3}
+\section{Tropical (non-)Linear Algebra and computation}
 
-\section{Tropical Polynomials and Non-Linear Programs  }
-\input{section4}
+\input{section3intro}%
+
+\subsection{As a model of linear programs}
+\input{section3A}
+
+\subsection{As a model of non-linear programs}
+\input{section3B}%
+
+\subsection{As a model of differential programs}
+\input{section3C}%
 
 \section{Analytic Results}
-\input{section5}
+\input{section4}%
 
 \section{Applications}
-\input{section6}
+\input{section5}%
 
-\section{Quasi-metric Spaces and $\Lawv$-Modules}
-\input{section7}
+\section{Quasi-metric Spaces and $\Lawv$-Modules}\label{sec:QMod}
+\input{section6}%
 
@@
 
-\section{Conclusion}
-
+\section{Conclusion and future work}
 \input{section9}

### 2023-01-10 section3A.tex
Push serale
Prime parti di section3.
Inoltre ho aggiunto un \label in main.tex ed qualche macros utile.

[Section III.A, Weighted], immediately gives:

\begin{fact}\label{fact:LREL_SMCC}
 $\LREL$ is a Symmetric Monoidal Closed Category.
\end{fact}

Its monoidal structures is given by a tensor product $\otimes$ acting on the objects as the Cartesian product of sets, and as the Kronecker product of matrices on morphisms.
Its closed structures is given by $X\multimap Y:=X\times Y$ with the usual evaluation and curry morphisms yielding a symmetric monoidal adjunction: $\LREL(Z\otimes X, Y) \simeq \LREL (Z, X\multimap Y)$.

Fact~\ref{fact:LREL_SMCC} means that $\LREL$ is a model of Intuitionistic Multiplicative Linear Logic (IMLL), and thus we can interpret \emph{linear} $\lam$-terms in it.

### 2023-01-10 section3intro.tex
Push serale
Prime parti di section3.
Inoltre ho aggiunto un \label in main.tex ed qualche macros utile.

@@
-Di cosa stiamo parlando. Ovvero:
+For a fixed \emph{continuous} semi-ring $Q$ [Def.
+II.5, Weighted], the well-known \emph{quantitative relational model}, is the category $\QREL$ of sets and set-indexed matrices with coefficients in $Q$ ([Weighted] would call it $Q^\Pi$).
+We must fix a convention for matrices: following [Weighted, MonTop, Ehrhard], we fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum\limits_{b\in Y} s_{b,c}t_{a,b}$.
+In linear algebra, a map $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns.
+In particular, the product of such a matrix for a vector defines a map $Q^Y\to Q^X$.
+Instead, we prefer to see a $t\in\HOM{\QREL}{X}{Y}$ as giving rise to a map $\hat t:Q^X\to Q^Y$ defined by $\hat t(x)_a:=\sum\limits_{b\in Y} t_{a,b}x_a$.
+In $\QREL^{op}$ (which corresponds to systematically taking transpose matrices), composition coincides with the product matrix/matrix and $\hat{(\cdot)}$ with the product matrix/vector.
+In order to avoid confusion, we will refer to a $t\in Q^{X\times Y}$ just as a \emph{matrix from $X$ to $Y$}.
+As it is expected, $Q^X$ is a $Q$-semimodule and the bijection $\hat{(\cdot)}$ identifies $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.
 
-- matrices over a continuous semi-ring
+\begin{remark}
+ The category $\QREL$ is (equivalent to) a (full) subcategory of the category $Q\SF{Mod}$ of $Q$-modules.
+ If $\QREL$ corresponds to considering semimodules (the $Q^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $Q\SF{Mod}$ corresponds to considering semimodules in abstract, without fixing a base.
+ We take this viewpoint in Section~\ref{sec:QMod}.
+\end{remark}
 
-- The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.
+The assumption of $Q$ being continuous (which can be weakened, see [??]) is used in order to make the potentially infinite sum, appearing in the formula for the composition, converge.
 
-- tropical linear algebra: the category $\LREL$
+In the following, we always consider the category $\LREL$, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,\cdot,0)$, which is one of the so-called \emph{tropical semirings}.
+The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring w.r.t.\ its quantale order $\preceq$, which is the reversed order $\geq$ of the usual order $\leq$ on $\BB R_{\geq0}\cup\set{\infty}$.
+% Qui dobbiamo aver gia detto che un quantale puo' essere dato come idempotent complete semiring o come complete lattice con un compatible monoid.
+This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
+
+\begin{remark}
+ In $\LREL$, the formula for the composition becomes the tropicalisation of the one defining it in $\QREL$, i.e.\ $(st)_{a,c}:=\inf\limits_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
+ similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$, which we call \emph{tropical linear}, are exactly those of shape $f(x)=\inf\limits_{b\in Y} \set{\check f_{a,b}+x_a}$, for some matrix $\check f$ from $X$ to $Y$.
+\end{remark}
+
+Since $\Lawv$ is a continuous (commutative) semiring, [Proposition III.3, Weighted] immediately applies and gives:
+
+\begin{fact}
+ $\LREL$ is a linear $\Lawv$-category.
+\end{fact}
+
+Unwrapping [Definition II.9, Weighted], this means that:
+$\HOM{\LREL}{X}{Y}$ is a continuous $\Lawv$-semimodule, with semimodule operations defined pointwise;
+$\LREL$ is a continuous $\Lawv$-category, i.e.\ composition of morphisms commutes with $\inf$'s;
+$\LREL$ is linear, i.e.\ pre- and post-composition with any morphism in any $\HOM{\LREL}{X}{Y}$ are automorphisms on it.
+
+In the remainder of this section we will see how tropical linear algebra -- expressed as $\LREL$ -- gives rise to denotational models of computation -- expressed as \lamcalc.

### 2023-01-10 section2.tex
commit

@@
 %Recall the two approaches with more details on lambda-calculus and on existing challenges.
+
+In this section we discuss in some more detail the two approaches in quantitative semantics recalled in the Introduction, at the same time providing an overview of how we aim at bridging them using tropical mathematics.
 
@@
 - Metric models: functions as Lipschitz maps, differential privacy.
+As recalled in the Introduction, in many situations (typically, when dealing with computationally expensive function) one does not look for algorithms to compute some function \emph{exactly}, but rather to approximate it (e.g.~in an efficient way) within some error bound.
+In other common situations (typically, in differential privacy \cite{}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, that a small error in the input will produce a comparably small error in the output.
+
+A natural way to account for these situations is to consider denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which measures differences of behavior.
+A fundamental insight coming from this works is that \emph{affine} programs, that is, programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
+$d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
+A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive 
+map \emph{symmetric monoidal closed}, and provides a model of the \emph{affine} simply typed $\lambda$-calculus.
+
+As observed in \cite{Gabo2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications, by considering \emph{Lipschitz-continuous} functions.
+For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the functional input $f$, can be interpreted as a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.
+By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $r\cdot X$ (i.e.~with distance $d_{r}(x,y)=r\cdot d(x,y)$) and $Y$, the program $M$ above
+can thus be interpreted as a non-expansive map from $2\cdot(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.
+
+These observations have led to the study of \emph{bounded} typed $\lambda$-calculi (like the system $\mathsf{Fuzz}$ \cite{Reed_2010}),
+which have been applied in the study of higher-order differential privacy \cite{}.
+In these calculi, which are inspired from Girard's \emph{bounded linear logic} \cite{}, function types are of the form $!_{r}X \multimap Y$, and programs are interpreted as $r$-Lipschitz maps.
+In categorical terms, the re-scaling operation $!_{r}(-)$ can be characterized as a  \emph{graded linear exponential comonad} \cite{}.
 
 - Problem with unrestricted duplications
+Then, what about the good old, ``unbounded'', simply typed $\lambda$-calculus?
+Actually, by using unbounded duplications, one might lose the Lipschitz property.
+For instance, while the functions $M_{k}=\lambda x.
+k\cdot x: \BB R\to \BB R$ are all Lipschitz-continuous, with Lipschitz constant $k$, the function $M=\lambda x.x^{2}$ obtained by ``duplicating'' $x$ is not Lipschitz anymore: $M$ is, so to say, \emph{too} sensitive to errors.
+More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of the simply typed $\lambda$-calculus (yet, $\Met$ contains several cartesian closed \emph{sub-}categories, see e.g.~\cite{}).
+Still, one might observe that the program $M$ above is actually Lipschitz-continuous, if not globally, at least \emph{locally} (i.e.~ over any compact set).
+However, constructing cartesian closed categories of locally Lipschitz maps is challenging.
+Yet, some examples exist in the literature (see \cite{EhrhardPCOHdiff, PistoneLICS2021}), and a new example will be exhibited in this paper.
 
 - Bounded Linear Logic/lambda-calculus, why Met is only a model of BLL, and why it is difficult (but not impossible!) to construct models of full lambda
+% - Metric models: functions as Lipschitz maps, differential privacy.
+%
+%- Problem with unrestricted duplications
+%
+%- Bounded Linear Logic/lambda-calculus, why Met is only a model of BLL, and why it is difficult (but not impossible!) to construct models of full lambda
 
@@
 - Differential lambda-calculus: differential operator, Taylor expansion (vs. 1)
+A different analysis of duplication and linearity in higher-order programs arises from works on \emph{differential} and \emph{resource} $\lambda$-calculi \cite{}, and the related field of non-idempotent intersection types (which, however, we will not consider in this paper, for space reasons).
+Notably, these approaches provide an elegant way to relate bounded and unbounded duplications through a notion of  \emph{Taylor expansion} of $\lambda$-terms.
 
@@
 - resource lambdas are polynomials, so not linear. 
+The key ingredient of these approaches is a notion of \emph{linear application}, which in the syntax of the differential $\lambda$-calculus is written $\Der[M]\cdot N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once.
+More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument.
+%Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
+%For example, if $M$ is the term $\lambda fx.f(fx)$ considered before, $\Der[M]$ takes a first input $N$ and passes it linearly to $M$.
+Notice that there are two ways of doing so, corresponding to the two bound occurrences of $f$ in $M$: either by applying $N$ to $fx$, or by
+%applying $f$ linearly to $Nx$ (indeed, if $f$ were applied in an unrestricted way, it might duplicate $Nx$, so that $N$ would not be used linearly).
+This justifies the equation below, in which $\Der[M]$ is identified with the non-deterministic sum of the two possible linear choices:
+%\begin{align}
+%\Der\left[\lambda f x.f(fx)\right]\cdot N = 
+%\lambda fx. N(fx) + \left(\Der[f]\cdot (Nx)\right)(fx)
+%\end{align}
+More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
+%
+
+The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy.
+\frac{\mathsf df(y)}{\mathsf dy}\cdot x$.
+Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
+Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:
+\begin{align}
+MN  =  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot (\Der^{(k)}[M] \cdot N^{k})0
+\end{align}
+Now, given our previous discussion of bounded duplications and Lipschitz-continuity, a natural question would be whether the differential approach to the $\lambda$-calculus may lead to decompose an unbounded $\lambda$-term
+into a sum of Lipschitz functions, thus bridging the metric and differential approach.  
+While suggestive, this idea seems to contradict the fact that, in the canonical semantics of the differential $\lambda$-calculus, the relational semantics \cite{}, the bounded terms appearing in the Taylor expansion correspond to \emph{polynomials}, and thus to non-Lipschitz functions.
+
+More generally, the relational semantics interprets unbounded programs as analytic functions, that is, as functions admitting a representation as power series.
+For instance, observing that an analytic map $f: \BB R\to \BB R$, where $f(x)=\sum_{n}\widehat f_{n}\cdot x^{n}$ is uniquely determined by the sequence $\widehat f_{n}$, the program $M_{\infty}:=\lambda fx.fx: (\BB R\To \BB R)\To (\BB R\To \BB R)$ is represented by the power series below:
+\begin{align}
+F_{\infty}(f,x)= \sum_{n=0}^{\infty} \widehat f_{n} x^{n}
+\end{align}
+By restricting ourselves to bounded applications, the terms in the power series become finite, that is, the interpretation becomes a \emph{polynomial}: for instance, the program $M_{2}:=\lambda fx.
+\sum_{i=0}^{2}\Der^{(i)}[f]\cdot x^{i}$, corresponding to passing $x$ \emph{at most twice} to $f$, is represented by the polynomial
+\begin{align}
+F_{\leq 2}(f,x)=\widehat f_{2} x^{2}  + \widehat f_{1}x +  \widehat f_{0} 
+\end{align}
+ In this framework the differential operator is naturally represented by formal differentiation of polynomials, where, as one would expect, 
+ $\Der[\sum_{n}a_{n}x^{n}]=\sum_{n}\Der[a_{n}x^{n}]$ and $\Der[a_{0}x^{0}]=0$ and $\Der[a_{n+1}x^{n+1}]= (n+1)a_{n+1}x^{n}$, so that power series can be Taylor expanded.
+
+Going back to our previous question, we can now reformulate it as follows: can we make the relational semantics \emph{Lipschitz}, hence amenable to metric and sensitivity analysis?
+The goal of this paper is to show that this is indeed possible and actually quite natural, by appealing to tropical mathematics.
 
@@
 - generalities about tropical maths (tropicalisation $\trop P$  of polynomials and of Laurent series, and their roots -- all that without $\LREL$)
+The fundamental observation that led to the study of mathematics over the \emph{tropical semi-ring} $\Lawv=([0,\infty],\min,+)$ was that, by replacing everywhere the ``$+$'' by the ``$\min$'' and the ``$\times$'' by the ``$+$'', the computation of many algebraic and geometric objects becomes combinatorial and simple.
+
+For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
+\begin{align}
+\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
+\end{align}
+Notably, the \emph{tropical roots} (whose definition is recalled in Section \ref{section3}) of $\trop p(\alpha)$ can be found through a rather simple (indeed polytime \cite{}) algorithm, and can be used to \emph{approximate} the actual roots of $p(x)$ \cite{}.
+More generally, the tropicalization of a power series $f(x)=\sum_{n}\widehat f_{n}x^{n}$ yields a \emph{tropical Laurent series} \cite{} 
+\begin{align}
+\trop f(\alpha)= \inf_{n}\left\{n\alpha+ \widehat f_{n}\right\}
+\end{align}
+a class of functions that we will study in detail in Section \ref{section4}.
+
+Since polynomials correspond to linear (hence Lipschitz) functions in tropical mathematics, the reconstruction of the relational semantics over the tropical semi-ring, presented in Section \ref{section3} and Section \ref{section4}, will provide a metric semantics of the differential $\lambda$-calculus, bridging sensitivity and resource analysis.
+In Section \ref{section5} we will suggest potential applications of this approach, relating well-studied applications of program metrics, resource analysis with current uses of tropical mathematics in computer science.
+Finally, in Section \ref{section6} we will show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between the theory of
+\emph{generalized metric spaces} (in the sense of Lawvere \cite{}) and the theory of modules over the tropical semi-ring.
+
+%- generalities about tropical maths (tropicalisation $\trop P$  of polynomials and of Laurent series, and their roots -- all that without $\LREL$)

### 2023-01-10 section2.tex
commit

@@
 
-- Metric models: functions as Lipschitz maps, differential privacy.
 As recalled in the Introduction, in many situations (typically, when dealing with computationally expensive function) one does not look for algorithms to compute some function \emph{exactly}, but rather to approximate it (e.g.~in an efficient way) within some error bound.
@@
 
-- Problem with unrestricted duplications
 Then, what about the good old, ``unbounded'', simply typed $\lambda$-calculus?
@@
 
-- Bounded Linear Logic/lambda-calculus, why Met is only a model of BLL, and why it is difficult (but not impossible!) to construct models of full lambda
 % - Metric models: functions as Lipschitz maps, differential privacy.
@@
 
-- Differential lambda-calculus: differential operator, Taylor expansion (vs. 1)
-A different analysis of duplication and linearity in higher-order programs arises from works on \emph{differential} and \emph{resource} $\lambda$-calculi \cite{}, and the related field of non-idempotent intersection types (which, however, we will not consider in this paper, for space reasons).
-Notably, these approaches provide an elegant way to relate bounded and unbounded duplications through a notion of  \emph{Taylor expansion} of $\lambda$-terms.
-
-- resource lambda-terms, Taylor expansion (vs. 2)
-
-- resource lambdas are polynomials, so not linear. 
 The key ingredient of these approaches is a notion of \emph{linear application}, which in the syntax of the differential $\lambda$-calculus is written $\Der[M]\cdot N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once.
@@
 \subsection{Tropical mathematics: a possible synthesis}
-
-- generalities about tropical maths (tropicalisation $\trop P$  of polynomials and of Laurent series, and their roots -- all that without $\LREL$)
 The fundamental observation that led to the study of mathematics over the \emph{tropical semi-ring} $\Lawv=([0,\infty],\min,+)$ was that, by replacing everywhere the ``$+$'' by the ``$\min$'' and the ``$\times$'' by the ``$+$'', the computation of many algebraic and geometric objects becomes combinatorial and simple.

### 2023-01-10 section2.tex
commit

@@
 In Section \ref{section5} we will suggest potential applications of this approach, relating well-studied applications of program metrics, resource analysis with current uses of tropical mathematics in computer science.
-Finally, in Section \ref{section6} we will show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between the theory of
-\emph{generalized metric spaces} (in the sense of Lawvere \cite{}) and the theory of modules over the tropical semi-ring.
+Finally, in Section \ref{section6} we will show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between
+\emph{generalized metric spaces} (in the sense of Lawvere \cite{}) and modules over the tropical semi-ring.
 

### 2023-01-11 section3B.tex
Push serale
Non so se dopo mangiato avrò le forze per scrivere qualcosina in più.
La sezione 3 va ricontrollata, aggiustata ecc, ma ora ci sono tutti gli elementi.

\subsubsection{Bounded duplications}

It is known ([reference??]) that the maps $X\mapsto \C M_{\leq n}(X)$ ($n\in\BB N$) lift to functors $!_n:\LREL\to\LREL$.
The sequence $(!_n)_{n\in\N}$ can be then turned into a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, satisfying the adjunction:
$\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
Therefore, $\LREL$, together with $(!_n)_{n\in\N}$, is a model for $\BSTLC$. 
Concretely, a type $A::= o \mid !_{n}A \multimap A$ (i tipi meglio dirli in Section II) is in interpreted as $\model o=\B 1$ and 
$\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$.

Bounded types are interesting because of the following proposition:
\begin{proposition}
For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) correspond to polynomials.
\end{proposition}
\begin{proof}
It suffices to check that $\model A$ is finite for all bounded types $A$. Indeed this implies that a morphism $t:\model A\to \model B$ is a finite matrix $t: \model A \times \model B \to \Lawv$.
Hence, its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is a polynomial.
\end{proof}

For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
\begin{itemize}
\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
\end{itemize}

- tropical polynomials and tropical roots ?? Meglio in Section II

\subsubsection{Unbounded duplications}

If in the previous subsection we considered \emph{bounded} multisets via the graded comonad $(!_n)_{n\in\N}$, it is immediate to see that \emph{unbounded} multisets yield a comonad able to interpret the $\STLC$:
following [Corollary III.6, Weighted] we have:

\begin{fact}
 $\LREL$ is Lafont.
\end{fact}

Therefore, it is well known that the map $X\mapsto \finMS{X}$ lifts to a functor $!:\LREL\to\LREL$ which %satisfies the adjunction $\LREL(!(Z\& X),Y) \simeq \LREL(!Z, !X\multimap Y)$, at 
is a Lafont-exponential.
In particular, [Theorem III.7, Weighted] says that:
\begin{fact}
 The coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.
\end{fact}

It is instructive at this point to see what its CCC-structure looks like in our tropical world.
In particular, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t:\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
\[
 \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1*\cdots *\mu_n}
 \set{s_{[b_1,\dots,b_n],c} + \sum\limits_{i=1}^n t_{\mu_i,b_i}}.
\]
The exponential object $X\multimap Y$ is $!X+Y$ (where $+$ is the disjoint union).
Remember that in $\Lawv$ the neutral element for addition is $\infty$ and the neutral for multiplication is $0$, so for instance the evaluation morphism is the matrix $\RM{eval}^{X,Y}\in\Lawv^{!((X\multimap Y) + X)\times Y}\simeq\Lawv^{(!!X\times !Y\times !X)\times Y}$ given by $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=0$ if $\rho_1=[\mu]$ and $\rho_2=[b]$, and $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=\infty$ otherwise.

\begin{remark}[Tropical Laurent series]
 As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ operates as a linear map $\widehat t:\Lawv^{!X}\to\Lawv^Y$.
 But we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_! x$ (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
 This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
 Concretely, we have:
 $t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
 Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
\end{remark}

### 2023-01-11 section3C.tex
Push serale
Non so se dopo mangiato avrò le forze per scrivere qualcosina in più.
La sezione 3 va ricontrollata, aggiustata ecc, ma ora ci sono tutti gli elementi.

@@
-- differential categories
+Until now we simply specialised well-known results in our tropical case, with the intent of showing how do things read in this particular case.
+Now we go further, by showing that $\LREL_!$ actually admits a \emph{differential structure}, turning it into a model of the $\STDLC$.
+This is where the \emph{metric} and the \emph{differential} viewpoints converge, as explained in the Introduction and Section II, and it will be generalised in Section 6.
 
-- the tropical differential operator: formal differentiation of tropical Laurent series. Differentiation in $\LREL_{!}$.
+A model of the $\STDLC$ is usually understood as so-called \emph{Cartesian closed differential categories} (CC$\partial$C), see [Giulio,?] for details.
+In order to treat the $+$ and the constructor $D[\_]\cdot (\_)$ of $\STDLC$, the main features of a CC$\partial$C $\C C$ are that:
 
-- $\LREL_{!}$ is a differential CCC satisfying Taylor. 
-Discuss the tropical Taylor expansion $\trop\C T(M)$.
+1) $\C C$ is a left-additive-CCC, i.e.\ its Homsets are commutative monoids and its Cartesian closed structure is well behaved w.r.t.\ this monoid structure;
+
+2) $\C C$ is equipped with a differential operator map $D:\HOM{\C C}{X}{Y}\to \HOM{\C C}{X\times X}{Y}$ (here $\times$ is the Cartesian product of $\C C$) satisfying $8$ axioms, called D1, ..., D7, D-curry.
+
+Let us show the differential structure of $\LREL_!$ (remember that the Cartesian product of $\LREL_!$ is the disjoint union $+$): 
+\begin{definition}
+ The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$,
+ $(Dt)_{\mu_1\oplus\mu_2,b}:=t_{\mu_2*\mu_1,b}$ if $\#\mu_1=1$ and $(Dt)_{\mu_1\oplus\mu_2,b}:=\infty$ otherwise.
+\end{definition}
+
+\begin{example}
+ \`E BENE METTERE UN ESEMPIO DI COME OPERA QUESTO DIFFERENZIALE (senn\`o non si capisce).
+ Secondo me conviene con un $D^2$ di qualcosa.
+\end{example}
+
+\begin{theorem}\label{thm:LREL!CCDC}
+ $\LREL_!$ equipped with $D$, is a CC$\partial$C.
+\end{theorem}
+\begin{proof}
+ The properties 1) and 2) are tedious technical checks.
+ If the first are quite straightforward, while the second deserve more attention.
+ Let us give the idea for ?
+\end{proof}
+
+There is more: not only $\LREL_!$ is a model of the $\STDLC$, but this model is also well-behaved w.r.t.\ to the \lamcalc Taylor expansion.
+This is expressed by the following property (see [Definition 4.22, Giulio], where it is called the fact of ``modeling the Taylor expansion'').
+
+\begin{theorem}\label{thm:modelsTaylor}
+ In $\LREL_!$ equipped with $D$, all morphisms can be Taylor expanded, i.e.\ in $\LREL_!$ we have the following equality:
+ \[
+  \RM{eval}^{X,Y}\circ_!\langle t,s\rangle =
+  \inf\limits_{n\in\N}
+  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ \langle \RM{id},\infty \rangle}.
+ \]
+ for all $t\in\HOM{\LREL_!}{Z}{X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$.
+\end{theorem}
+Here, $\Lambda^-:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$ is the uncarry operator and $\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is an operator corresponding to the syntactical \emph{differential substitution}.
+\begin{proof}
+ Che ci scriviamo ??
+\end{proof}
+
+This property is interesting because of [Theorem 4.23,Giulio], which says that, then, in the model the series corresponding to the Taylor expansion of an ordinary $\lam$-term $M$ converges to $M$.
+
+- Discuss the tropical Taylor expansion $\trop\C T(M)$ ??

### 2023-01-11 section6.tex

As we have seen, the morphisms of $\LREL$ can be seen as continuous functions between the $\Lawv$-modules $\Lawv^{X}$, when the latter are taken with the metric induced by the $\infty$-norm. This viewpoint gives a metric flavor to $\LREL$, and allowed us to relate differential and metric structure. Yet, how far can this correspondence between tropical linear/non-linear algebra and metric topology be pushed?

For instance, natural questions are: (1) Is this correspondence restricted to $\Lawv$-modules of the form $\Lawv^{X}$ (i.e.~with a fixed base), or does it hold in some sense for arbitrary $\Lawv$-modules? (2) Is this correspondence restricted to the $\infty$-norm metric, or does it hold for other metrics too?

An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{}. 
This correspondence, recalled in this section, and the fact that it leads to the definition of a linear $\Lawv$-category (i.e.~a model of the linear $\lambda$-calculus) of genealized metric spaces extending $\LREL$, is mostly a \emph{collage} of  \emph{folklore} results with more abstract results contained in recent literature in enriched category theory \cite{}.
Beyond this reconstruction, our main contribution is to show that this correspondence can be lifted to a model of the full differential $\lambda$-calculus, by a suitable generalization of the construction of the exponential comonad of $\LREL$.

%
%also extends exploit this correspondence to prove that arbitrary $\Lawv$-modules (equivalently, arbitrary \emph{complete} generalized metric spaces) provide a model of the differential $\lambda$-calculus, hence generalizing the tropical relational model.
%

\subsection{$\Lawv$-Modules}

A $\Lawv$-module is a triple $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$ (notice that this is the usual notion of module over the Lawvere quantale $\Lawv$, not that of module over the tropical semi-ring).
A $\Lawv$-module homomorphism is a map $f:M\to N$ commuting with both joins and the action. We let $\Mod$ indicate the category of $\Lawv$-modules and their homomorphisms.

$\Lawv$ is the most basic example of $\Lawv$-module.
Any $\Lawv$-module $M$ has a dual $M^{\op}$, with reversed order and (right-)action $x\multimapinv \epsilon= \bigvee\{y\mid \epsilon \star y\succeq x\}$.
 Other basic examples of $\Lawv$-modules are the sets $\Lawv^{X}$, with order and action defined pointwise.

 While the $\Lawv^{X}$ have a fixed base, for an arbitrary $\Lawv$-module one can retrieve a base via the \emph{Yoneda embedding}
$\Yon: M \to \Hom(M^{\op},\Lawv)$, where $\Yon(x)(y)=\inf\{\epsilon\mid \epsilon\star y\succeq x\}$.

\begin{proposition}\label{prop:yonemod}
For any $\Lawv$-module $M$, the Yoneda embedding has a left-adjoint $\sup(f)=\bigvee_{x\in M}f(x)\star x$.
\end{proposition}

Like $\LREL$, the category $\Mod$ has the relevant structure to interpret the linear $\lambda$-calculus:
\begin{proposition}
$\Mod$ is a linear $\Lawv$-category.
\end{proposition}
\begin{proof}[Proof sketch]
The hom-sets $\Hom(M,N)$ have a natural $\Lawv$-module structure, defined pointwise. Moreover, the tensor product of $\Lawv$-modules $M\otimes N$ \cite{} can be defined as the quotient of the usual tensor product of sup-lattices (see e.g.~\cite{}), under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$.
Notably, any element of $M\otimes N$ can be identified with a join of basic tensors $x\otimes y$, corresponding to the equivalence class of the pair $\langle x,y\rangle$.

Beyond the required adjointness of the internal hom and the tensor, one can check that $\Mod$ is actually \emph{$^{*}$-autonomous}, since it satisfies $(M^{\op})^{\op}\simeq M$ and 
$\Hom(M,N^{\op})\simeq (M\otimes N)^{\op}$.
Finally, both products and coproducts in $\Mod$ are given by the Cartesian products of the underlying posets, with action defined pointwise.
%
%, similarly to the case of sup-lattices \cite{}, as the quotient of the free sup-lattice $\C P(M\times N)$ under the smallest congruence containing all pairs $((\bigvee A,y),\bigcup_{a\in A}\{(a,y)\})$, for $A\subseteq M, y\in N$, 
%$((x,\bigvee B),\bigcup_{b\in B}\{(x,b)\})$, for all $x\in A$, $B\subseteq N$, and all pairws $\{
\end{proof}

\begin{remark}
The structure of linear $\Lawv$-category of $\Mod$ coincides with that of $\LREL$ for the modules $\Lawv^{X}$. Indeed, one can prove that $\Hom(\Lawv^{X},\Lawv^{Y})\simeq \Lawv^{X\times Y} $ (using the fact that module homomorphisms are in bijection with $\Lawv$-matrices), and 
moreover $\Lawv^{X}\otimes \Lawv^{Y}\simeq \Lawv^{X\times Y} $ and 
$\Pi_{i\in I}\Lawv^{X_{i}}\simeq \Lawv^{\coprod_{i\in I}X_{i}}$.
\end{remark}

\subsection{$\Lawv$-Categories}

Lawvere was the first to observe that a metric space can be described as a \emph{$\Lawv$-enriched} category. Indeed, spelling out the definition, a $\Lawv$-enriched category (in short, a $\Lawv$-category) is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$, satisfying 
\begin{align}
0  & \geq X(x,x) \tag{$\Lawv$-cat 1}\\
X(y,z)+X(x,y)&\geq  X(x,z) \tag{$\Lawv$-cat 2}
\end{align}
This structure is often referred as a \emph{generalized metric space} \cite{}, or as a \emph{quasi-metric space} \cite{}. 
Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, i.e.~it must satisfy $Y(f(x),f(y))\leq X(x,y)$.

A $\Lawv$-category is \emph{skeletal} \cite{} when 
$X(x,y)=0$ implies $x=y$, and 
 \emph{symmetric} when it coincides with its opposite category $X^{\op}(x,y):=X(y,x)$, i.e.~when $X(x,y)=X(y,x)$. 
 A metric space, in the usual sense, is thus the same as a skeletal and symmetric $\Lawv$-category.
  Notice that any $\Lawv$-category $X$ induces a skeletal category $X^{\sk}$ by quotienting points under $X(x,y)=0$, and a symmetric one by letting $X^{\sym}(x,y)=\max\{X(x,y),X^{\op}(x,y)\}$.

 $\Lawv$ has a canonical $\Lawv$-enriched structure given by 
 $\Lawv(r,s)=s \menus r$ (where ``$\dotdiv$'' indicates truncated subtraction), and the Euclidean distance coincides with its symmetrization $\Lawv^{\sym}(x,y)$.

For any $\Lawv$-category $X$, the presheafs $[X^{\op},\Lawv]$ on $X$ form another $\Lawv$-category, with metric $[X^{\op},\Lawv](f,g)= \sup_{x\in X}\Lawv(f(x),g(x))$.
Notice that, when $X$ has the discrete metric, $[X,\Lawv]$ coincides with $\Lawv^{X}$.
The \emph{Yoneda embedding} is the faithful functor $\Yon: X\to [X^{\op},\Lawv]$ given by $\Yon(x)(y)=X(y,x)$.

Interestingly, a fundamental example of $\Lawv$-categories is provided by $\Lawv$-modules.

\begin{proposition}
Any $\Lawv$-module $(M,\preceq, \star)$ is a $\Lawv$-category via
\begin{align}
M(x,y) = \inf\left\{ \epsilon \mid \epsilon \star x\geq y\right\}
\end{align}
Moreover, a homomorphism of $\Lawv$-modules is a functor of the associated $\Lawv$-categories.
\end{proposition}

Observe that, since the distance $M(x,y)$ coincides with the Yoneda embedding $\B Y$ in $\Mod$, the latter also coincides with the Yoneda embedding of the associated $\Lawv$-category (this justifies the use of a unique symbol for both embeddings).

\subsection{Complete $\Lawv$-Categories Correspond to $\Lawv$-Modules}

Lawvere also observed that Cauchy-completeness can be expressed in categorical language as the representability in $X$ of certain presheaves in $[X^{\op},\Lawv]$ \cite{Lawvere1973}. Category theory suggests a yet stronger notion of completeness, corresponding to the existence of all \emph{weighted colimits} (for a comparison of different notions of completeness on $\Lawv$-categories, see \cite{}).

First, let us recall that functors of shape $\Phi: X\times Y^{\op}\to \Lawv$ are called \emph{distributors} and usually noted $\Phi: Y \pfun X$.

\begin{definition}
Let $X,Y,Z$ be $\Lawv$-enriched category,
$\Phi: Z\pfun Y$ be a distributor and  $f:Y\to X$ be a functor.
A functor $g:Z\to X$ is the \emph{$\Phi$-weighted colimit of $f$ over $X$}, noted $\colim(\Phi,f)$, if for all $z\in z$ and $x\in X$
\begin{align}
X(g(z),x)= \sup_{y\in Y}\left\{X(f(y),x)\menus \Phi(y,z)\right\}
\end{align} 
A functor $f:X\to Y$ is said \emph{continuous} if it commutes with all existing weighted colimits in $X$, i.e.~$f(\colim(\Phi,g))=\colim(\Phi,f\circ g)$. A $\Lawv$-enriched category 
$X$ is said \emph{categorically complete} (or just \emph{complete}) if all weighted colimits over $X$ exist. 
\end{definition}

We let $\GMet$ indicate the category of complete and skeletal $\Lawv$-categories and continuous functors.

A useful alternative characterization of complete $\Lawv$-categories is the following:
\begin{proposition}
A $\Lawv$-category is complete iff $\Yon$ has a left-adjoint. 
\end{proposition}
Indeed, using this fact, together with Proposition \ref{prop:yonemod}, we arrive at the following:
\begin{proposition}
For any $\Lawv$-module, the associated $\Lawv$-category is complete. 
\end{proposition}

Beyond limits of Cauchy sequences, another important example of colimit is the following:
\begin{definition}
Let $X$ be a $\Lawv$-category, $x\in X$ and $\epsilon \in \Lawv$. The \emph{tensor of $x$ and $\epsilon$}, if it exists, is the colimit $\epsilon \otimes x:= \colim( [\epsilon],\Delta x)$, where
$[\epsilon]: \B 1\pfun \B 1$ is the constantly $\epsilon$ distributor
and $\Delta x:\B 1\to X$ is the constant functor. 
\end{definition}

In a complete $\Lawv$-category all tensors exist and give rise to a $\Lawv$-module structure:
\begin{proposition}
Any complete $\Lawv$-category $X$ is a $\Lawv$-module, with order given by $x\preceq_{X}y $ iff $X(y,x)=0$, and 
action given by tensors $\epsilon \otimes x$. Moreover, a continuous functor between complete $\Lawv$-categories is the same as a homomorphism of the associated $\Lawv$-modules. 
\end{proposition}

To conclude our correspondence between $\Lawv$-modules and complete $\Lawv$-categories, it remains to observe that the 
two constructions leading from one structure to the other are one the inverse of the other: for any $\Lawv$-module $(M,\preceq,\star)$,
$x\preceq_{M}y$ iff $M(y,x)=0$ iff $x\preceq y=0\star y$, and, from  
$M(\epsilon \star x, y)= M(x,y)\dotdiv \epsilon$, we deduce $\epsilon\otimes x=\epsilon \star x$. 
Conversely, 
for any complete $\Lawv$-category $X$ and $x,y\in X$, one can check that 
$X(y,x)=\inf\{ \epsilon \mid X(\epsilon\otimes y,x )=0\}$.

This leads to the following:

\begin{theorem}
$\Mod$ and $\GMet$ are isomorphic categories.
\end{theorem}

Since $\Mod$ (and thus $\GMet$ too) is a linear $\Lawv$-category, it is worth making its metric structure explicit. Given complete $\Lawv$-categories $X,X_{i},Y$, we have that:
\begin{itemize} 
\item the distance on the hom-set $\Hom(X,Y)$ is given by 
\begin{align}
\Hom(X,Y)(f,g)= \sup_{x\in X}Y(f(x),g(x));
\end{align}
\item the distance on the tensor $X\otimes Y$ is given by
\begin{align}
(X\otimes Y)(\alpha, \beta)=
\sup_{i}\inf_{j}X(x_{i},x'_{j})+Y(y_{j},y'_{j}),
\end{align}
where $\alpha=\bigvee_{i}x_{i}\otimes y_{i}$ and 
$\beta=\bigvee_{j}x'_{j}\otimes y'_{j}$, 
and thus coincides with the sum-metric over basic tensors;
\item the distance on the bi-product $\Pi_{i\in i}X_{i}$ is given by
\begin{align}
(\Pi_{i\in I}X_{i})(f,g)= \sup_{i\in I}X_{i}(f(x),g(x));
\end{align}

\end{itemize}

\subsection{Exponential and Differential Structure}

We now want to show how the correspondence $\Mod\simeq \GMet$ lifts to a model of the differential $\lambda$-calculus, extending the co-Kleisli category $\LREL_{!}$.

FOLLOW SECTION III.C for definitions of models of LL and DiLL.

For any $\Lawv$-module $M$, $n\in \BB N$ and permutation $\sigma\in \F S_{n}$, define the homomorphism $\langle \sigma\rangle: M^{\otimes_{n}}\to M^{\otimes_{n}}$ by letting 
$\langle\sigma\rangle (x_{1}\otimes \dots \otimes x_{n})=x_{\sigma(1)}\otimes \dots \otimes x_{\sigma(n)}$ on basic tensors, and extending by continuity on the whole tensor module.

\begin{definition}[symmetric tensor algebra]

For any $\Lawv$-module $M$ and $n\in \BB N$, let $\Sym_{n}(M)$ indicate the $\Lawv$-module obtained by quotienting 
$M^{\otimes_{n}}$ via the least congruence generated by the action $\langle \sigma \rangle$ of permutations $\sigma\in \F S_{n}$.
\end{definition}

To prove that the map $h:\Sym_{n}(M)\to M^{\otimes_{n}} $ is the equalizer of the diagram formed by all homomorphisms $\langle \sigma\rangle$, it is useful to provide an alternative characterization of it.

\begin{definition}
Let $M$ be a $\Lawv$-module and $n\in \BB N$. An element $x\in M^{\otimes_{n}}$ is said \emph{permutation-invariant} (in short, \emph{p-invariant}) if for all $\sigma\in \F S_{n}$, 
$\langle \sigma \rangle (x)=x$.

 A \emph{$\Lawv$-multiset} (with $n$ elements) is an element of $M^{\otimes_{n}}$ of the form 
 \begin{align}
 [x_{1},\dots, x_{n}]:= \bigvee_{\sigma\in \F S_{n}}
 x_{\sigma(1)}\otimes \dots \otimes x_{\sigma(n)}
 \end{align}
where $x_{1},\dots, x_{n}\in M$.
\end{definition}

\begin{proposition}
Any $\Lawv$-multiset is p-invariant. Moreover, the set $!_{n}M$ of p-invariant elements of $M^{\otimes_{n}}$ is a $\Lawv$-submodule of $M^{\otimes_{n}}$, in which element can be written as a join of $\Lawv$-multisets.
\end{proposition}

Since $!_{n}M$ is included in $M$, using the properties of p-invariant one can easily deduce that $!_{n}M$ provides the desired equalizer. It suffices then to show that $!_{n}M$ is isomorphic to the symmetric algebra.

\begin{proposition}
The inclusion morphism $!_{n}M \to M^{\otimes_{n}}$ is the equalizer of the diagram formed by all $M^{\otimes_{n}}\stackrel{\langle \sigma\rangle}{\to} M^{\otimes_{n}}$. Moreover, 
$!_{n}M \simeq \Sym_{n}(M)$.
\end{proposition}

The module $!_{n}M$ is a complete $\Lawv$-category with distance function defined on $\Lawv$-multisets as below:
\begin{align}
(!_{n}M)(\alpha,\beta)=
\sup_{\sigma\in \F S_{n}}\inf_{\tau\in \F S_{n}}\sum_{i=1}^{n}
X(x_{\sigma(i)},y_{\tau(i)})
\end{align}
where $\alpha=[x_{1},\dots, x_{n}]$ and $\beta= [y_{1},\dots, y_{n}]$.

Also in this case, the construction above generalizes the situation in $\LREL$: for the modules of the form $\Lawv^{X}$, the symmetric algebra construction coincides with the usual one, provided by multisets.
\begin{proposition}
For any set $X$, 
$!_{n}(\Lawv^{X})\simeq \Lawv^{\C M_{n}(X)}$, where
$\C M_{n}(X)$ is the set of multisets of $X$ of cardinality $n$. 
\end{proposition}

At this point, we can define the exponential comonad as 
$!M:= \prod_{n\in \BB N}!_{n}M$, obtaining a Lafont category.
ECC....

### 2023-01-12 section5.tex
Sezioni riorganizzate
Ora tutte le sezioni del sorgente dovrebbero corrispondere alle sezioni del pdf
commit

--- 2023-01-08 section5.tex
@@
-General discussion: optimization properties behave in a Lipschitz way.
+In this section we illustrate a few directions in which the tropical semantics just introduced could be used to analyze quantitative properties of higher-order programs. 
+The general pattern is to consider a program $M$, which can be expressed in the form of a polynomial or a power series, and 
+investigate what quantitative properties of $M$ can be deduced from properties of its tropicalization $\trop M$ (which are often easier to study).
+By the way, this is the approach taken in several well-studied applications of tropical geometry (e.g.~for optimization problems related to machine learning \cite{}).
 
-- differential privacy and Lipschitzness
+\subsection{Best Case Analysis and Metric Reasoning}
 
-- log-probabilities and tropical roots
+The possibility of using the relational model over the tropical semi-ring for ``best case'' resource analysis has already been explored in \cite{Manzo}.
+Notably, they considered an interpretation of a language for $\B{PCF}$ with non-deterministic choice in which each $\lambda$-abstraction and each occurrence of the fixpoint operator $Y$ is assigned a ``weight'' 1, and showed that for any program $M$ of type $\B{nat}$,
+the value of the interpretation $\model{M}\in \Lawv^{\BB N}$ on a particular natural number $k$, i.e.~$\model{M}(k)\in \Lawv$, corresponds to the \emph{minimum} number of $\beta$- or $\TT{fix}$-redexes reduced in a reductions sequence from $M$ to $\underline n$.
+In the next paragraph we will illustrate an analogous ``best case'' analysis for probabilistic programs.
 
-- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)
+What does the metric analysis of the tropical model in the previous sections add to that?
+Firstly, the possibility of \emph{comparing} different programs with respect to their quantitative properties.
+For example, in the $\B{PCF}$ semantics recalled above, the distance between two programs $M$ and $N$ of type $\B{nat}$, provides a bound on the difference between the  ``best case'' computation time of $M$ and that of $N$.
+Typically, by taking, instead of the $\infty$-norm metric on $\Lawv^{\BB N}$,
+a \emph{non symmetric} distance (a viewpoint we explicitly take in Section \ref{section6}) $q(\B x, \B y)=\sup_{n}\{\B y_{n}\dotdiv \B x_{n}\}$, a ``distance'' $q(\model{M},\model{N})\leq \epsilon$ would indicate that $\model{M}$ \emph{improves} on $\model{N}$ of at most $\epsilon$ steps at each computation.
 
-- measuring duplications of discrete functions (needs finiteness!)
+Secondly, the Lipschitz conditions studied in the previous section allow us to reason on program distances in a \emph{compositional} way: suppose, as before, that $M,N:A$ are two programs such that $M$ improves on $N$ by $\epsilon$, and let $\TT C[-]:A \to \B{nat}$ indicate a context; knowing that the interpretation of $\TT C$ is $k$-Lipschitz-continuous on some open set containing both $\model M$ and $\model N$, allows us to immediately deduce that $\TT C[M]$ improves on $\TT C[N]$ by $k \epsilon$.
+Observe that this will typically be the case when the Taylor expansions of $\TT C[M]$ and $\TT C[N]$ actually yields a \emph{finite} sum of at most $k$ terms, i.e.~when 
+\begin{align}
+\TT C[M] = \sum_{i=0}^{k} \TT D^{(k)}\Big[\lambda x.\TT C[x]\Big]\cdot M^{k}
+\end{align}
+and similarly for $\TT C[N]$.
+It is here worth recalling that, for $\STDLC$, a well-known result \cite{} is that the Taylor expansion of a closed $\lambda$-term is always finite, although its non-zero coefficients may be arbitrarily high.
+Notably, these observations suggest to study tropical versions of \emph{finiteness spaces} \cite{}, 
+a variant of the relational semantics modeling strongly normalizing programs via \emph{finite} power series.
+
+\subsection{Tropical Polynomials and Log-Probabilities}
+
+\subsection{Tropical Laurent Series and Differential Privacy}
+
+%
+%General discussion: optimization properties behave in a Lipschitz way.
+%
+%
+%- differential privacy and Lipschitzness
+%
+%
+%- log-probabilities and tropical roots 
+%
+%
+%- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)
+%
+%
+%- measuring duplications of discrete functions (needs finiteness!)

### 2023-01-12 section3B.tex
Ho aggiustato delle cose in Sezione III
In particolare, ho introdotto lì il famoso esempio

@@
  Concretely, we have:
- $t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
+ \[t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}\] where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
  Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
 \end{remark}
+
+\begin{remark}
+ Indentifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tropical Laurent series generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf\limits_{n\in\N}\set{nx+\widehat f(n)}$, for some $\widehat f:\N\to\Lawv$.
+Remark that we find usual \emph{tropical polynomials} of tropical geometry as a particular case: they corrispond to the ones for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}.
+\end{remark}
+
+\begin{proposition}\label{prop:descrete}
+ The interpretation $\model M$ of a $\lam$-term $M$ in $\LREL$ is a \emph{descrete} matrix, i.e.\ its coefficients are either $0$ or $\infty$.
+\end{proposition}
+\begin{proof}
+ First easily prove that composition preserves descreteness.
+ Then go by straightforward induction on $M$, using that the projections and evaluation of  $\LREL$ are descrete.
+\end{proof}
+
+\begin{example}
+ The function $f:\Lawv\to\Lawv$, $f(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ is a tropical Laurent series: it is of shape $f=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
+By Proposition~\ref{prop:descrete}, $f$ is not the interpretation of a $\lam$-term, because its matrix is not descrete. Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
+\end{example}

### 2023-01-12 section3C.tex
Ho aggiustato delle cose in Sezione III
In particolare, ho introdotto lì il famoso esempio

@@
 Let us show the differential structure of $\LREL_!$ (remember that the Cartesian product of $\LREL_!$ is the disjoint union $+$):
+
 \begin{definition}
- The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$,
- $(Dt)_{\mu_1\oplus\mu_2,b}:=t_{\mu_2*\mu_1,b}$ if $\#\mu_1=1$ and $(Dt)_{\mu_1\oplus\mu_2,b}:=\infty$ otherwise.
+ Let $\widetilde{(\cdot)}:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$, $\widetilde t_{\mu\oplus\rho,b}:= t_{\mu*\rho,b}$, and $d:\HOM{\LREL}{!(X+X)}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$, $(dt)_{\mu\oplus\rho,b}:=t_{\rho\oplus\mu,b}$ if $\#\mu_1=1$ and $(dt)_{\mu\oplus\rho,b}:=\infty$ otherwise.
+ The \emph{tropical differential operator} is the map $D:=d\circ\widetilde{(\cdot)}:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$.
 \end{definition}
@@
 \begin{example}
- \`E BENE METTERE UN ESEMPIO DI COME OPERA QUESTO DIFFERENZIALE (senn\`o non si capisce).
- Secondo me conviene con un $D^2$ di qualcosa.
+For $t\in\HOM{\LREL}{!X}{Y}$, we have:
+ $(Dt)_{\mu\oplus\rho,b}=t_{\rho*\mu,b}$ if $\#\mu_1=1$ and it is $\infty$ otherwise.
+ Analogously, we have:
+ $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu*\nu'*\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho*\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.
 \end{example}
@@
 \begin{proof}
- The properties 1) and 2) are tedious technical checks.
- If the first are quite straightforward, while the second deserve more attention.
- Let us give the idea for ?
+ The properties 1) are quite straightforward.
+ Properties 2) are tedious technical checks.
 \end{proof}
@@
   \inf\limits_{n\in\N}
-  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ \langle \RM{id},\infty \rangle}.
+  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
  \]
@@
 This property is interesting because of [Theorem 4.23,Giulio], which says that, then, in the model the series corresponding to the Taylor expansion of an ordinary $\lam$-term $M$ converges to $M$.
-
-- Discuss the tropical Taylor expansion $\trop\C T(M)$ ??

### 2023-01-12 section5.tex
commit

@@
 
-What does the metric analysis of the tropical model in the previous sections add to that?
+What does the metric analysis from the previous sections add to that?
 Firstly, the possibility of \emph{comparing} different programs with respect to their quantitative properties.
 For example, in the $\B{PCF}$ semantics recalled above, the distance between two programs $M$ and $N$ of type $\B{nat}$, provides a bound on the difference between the  ``best case'' computation time of $M$ and that of $N$.
-Typically, by taking, instead of the $\infty$-norm metric on $\Lawv^{\BB N}$,
-a \emph{non symmetric} distance (a viewpoint we explicitly take in Section \ref{section6}) $q(\B x, \B y)=\sup_{n}\{\B y_{n}\dotdiv \B x_{n}\}$, a ``distance'' $q(\model{M},\model{N})\leq \epsilon$ would indicate that $\model{M}$ \emph{improves} on $\model{N}$ of at most $\epsilon$ steps at each computation.
+For instance, by taking, instead of the $\infty$-norm metric on $\Lawv^{\BB N}$,
+the \emph{non symmetric} distance (or quasi-metric, a viewpoint we explicitly take in Section \ref{section6}) $q(\B x, \B y)=\sup_{n}\{\B y_{n}\dotdiv \B x_{n}\}$, a ``distance'' $q(\model{M},\model{N})\leq \epsilon$ would indicate that $\model{M}$ \emph{improves} on $\model{N}$ of at most $\epsilon$ steps at each computation.
 
@@
 and similarly for $\TT C[N]$.
-It is here worth recalling that, for $\STDLC$, a well-known result \cite{} is that the Taylor expansion of a closed $\lambda$-term is always finite, although its non-zero coefficients may be arbitrarily high.
+It is here worth recalling that, for $\STDLC$, a well-known result \cite{} is that the Taylor expansion of a closed application $MN$ is always finite, although its non-zero coefficients may be arbitrarily high.
 Notably, these observations suggest to study tropical versions of \emph{finiteness spaces} \cite{}, 
@@
 
-\subsection{Tropical Laurent Series and Differential Privacy}
+{\color{red}MI SA CHE QUESTO PEZZETTO SERVE ANCHE QUI, O NO?}
+First, let us recall that for any positive real $t$, the functions $\phi_{t}(x)=-t\log x$ and $\psi_{t}(\alpha)=e^{-\alpha/t}$ define continuous bijections between the space of probabilities $[0,1]$ and $\Lawv$.
+Moreover, the two tropical  operations $\min$ and $+$ can be obtained via these bijections by letting $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)+\psi_{t}(\beta)$ and $\alpha\prodt{t} \beta=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))$, and observing that $\lim_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$ and $\lim_{t\to 0}\alpha\prodt{t}y= \alpha+\beta$.
+
+\subsection{Resource Analysis for Differential Privacy}
+
+The typical situation in differential privacy is where one considers a probabilistic query $f: \mathsf{db}\to [0,1]^{X}$ over some database, and one requires that $f$ should not be \emph{too sensitive} to small changes in the output, in other to prevent potential leaks of private information about individual items in $\mathsf{db}$ (for instance, an element $x\in\mathsf{db}$ could be the list of students of some university and $f(x)$ indicates the percentage of female students$\mathsf{db}$).
+
+More formally, differential privacy is defined as follows \cite{}:
+\begin{definition}
+Let $f: \mathsf{db}\to [0,1]^{X}$ and $\epsilon \in \BB R_{\geq 0}$. $f$ is said \emph{$\epsilon$-differentially private} if for all $x,x'\in \mathsf{db}$
+differing by $L$ items, for all $s\in X$, 
+$$
+f(x)(s) \ = \ e^{\epsilon L} f(x')(s)
+$$
+\end{definition}
+
+A well-studied approach to ensure differential privacy for higher-order programs is to use bounded type systems like $\mathsf{Fuzz}$ \cite{}.
+Indeed, such systems come with an \emph{a priori} warrant that well-typed programs are Lipschitz.
+
+The use of tropical semantics suggests how resource analysis could also be used to provide bounds for differential privacy. 
+Suppose our probabilistic query $f$ can be expressed as a power series (this is what happens e.g.~in \emph{probabilistic coherent spaces} \cite{}).
+Then, if we discover, either by studying differential properties of $f$, or using methods from convex analysis as suggested in the previous section (e.g.~Theorem \ref{??}),
+ that the tropicalization $\trop f$ satisfies a Lipschitz condition, we may use this fact to deduce that $f$ is differentially private, as shown by the result below.
 
 %
+%We have seen how a higher-order probabilistic program, which is expressed as a polynomial, can be tropicalized.
+This operation can be extended to programs expressed as power series by taking the limit.
+%
+\begin{proposition}
+Let $f: [0,1]^{X}\to [0,1]^{Y}$ be an analytic function. If $\trop f$ is Lipschitz over some open set $U$, then $f$ is differentially private over $\psi_{t}(U)$, for small enough $t$.
+\end{proposition}
+\begin{proof}
+We consider the case $X=Y=\B 1$ for simplicity. 
+Express $f$ as a power series $f(x)=\sum_{n}\psi_{e}(\widehat f_{n})x^{n}$, so that $\trop f(\alpha)= \inf_{n}\left \{
+n\alpha + \widehat f_{n}
+\right\}$.
+Let us define a family of intermediate functions $\trop_{t}f(\alpha)=\phi_{t}(\stackrel{t}{\sum_{n}}\widehat{f}_{n}\times_{t}n\alpha)$. These, by construction, satisfy
+\begin{align}\label{eq:phit}
+f(\phi_{t}(\alpha)) = \phi_{t}(\trop_{t}f(\alpha))
+\end{align}
+and, using the fact that $\alpha\sumt{t}\beta\to\min\{\alpha,\beta\}$ and $\alpha\prodt{t}\beta\to \alpha+\beta$, we can deduce that $\trop f(\alpha)=\lim_{t\to 0}\trop_{t}f(\alpha)$ (for this, we use the fact that for all $m\in \BB N$,
+$\min_{i\leq m}\alpha_{i}\leq \stackrel{t}{\sum}_{i\leq m}\alpha_{i}\leq 
+(\min_{i\leq m}\alpha_{i})+t\log m$).
+
+Now, using the fact that $\trop f$ and $\trop_{t} f$ come arbitrarily close with $t$ small enough, together with \eqref{eq:phit}, the Lipschitz condition  
+$|\trop f(\phi_{t}(x))-\trop f(\phi_{t}(y))|\leq L\cdot |\phi_{t}(x)-\phi_{t}(y)|$ translates into the differential privacy condition 
+$f(x) \leq e^{L\cdot |\log x-\log y|} f(y)$ (supposing $y\geq x$ and $f(y)\neq 0$).
+\end{proof}
+
 %General discussion: optimization properties behave in a Lipschitz way.

### 2023-01-12 section6.tex
commit

@@
 
-FOLLOW SECTION III.C for definitions of models of LL and DiLL.
+First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies, Weight}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting the fact that tensors and products commute in $\Mod$, we define $!$ as the product of the symmetric algebras $!_{n}$.
 
@@
 
-Also in this case, the construction above generalizes the situation in $\LREL$: for the modules of the form $\Lawv^{X}$, the symmetric algebra construction coincides with the usual one, provided by multisets.
+Using the fact that $\prod_{i}M_{i}\otimes  N\simeq (\prod_{i}M_{i})\otimes N$ holds in $\Mod$ [\cite{}, p.~???], we obtain the following:
+\begin{theorem}
+The functor $!M:= \prod_{n\in \BB N}!_{n}M$ is a Lafont exponential in $\Mod$.
+Hence, $\Mod_{!}$ (equivalently, $\GMet_{!}$) is cartesian closed. 
+\end{theorem}
+
+Also in this case of exponentials, the constructions above generalize the situation in $\LREL$:
 \begin{proposition}
@@
 $!_{n}(\Lawv^{X})\simeq \Lawv^{\C M_{n}(X)}$, where
-$\C M_{n}(X)$ is the set of multisets of $X$ of cardinality $n$. 
+$\C M_{n}(X)$ is the set of multisets of $X$ of cardinality $n$, and 
+$!(\Lawv^{X})\simeq \Lawv^{\multiset(X)}$. 
 \end{proposition}
 
-At this point, we can define the exponential comonad as 
-$!M:= \prod_{n\in \BB N}!_{n}M$, obtaining a Lafont category.
-ECC....
+To conclude, we mush show that $\Mod$ is actually a model of $\STDLC$, that is, a cartesian closed differential category:
+\begin{enumerate}
+\item $\Mod$ is a left-additive-CCC, since each hom-set is a commutative monoid with respect to $\infty$ and $\min$, and the cartesian closed structure is well-behaved with respect to that;
+
+\item $\Mod$ can be equipped with a differential operator defined, for $f:!M\to N$, by 
+\begin{align}\label{eq:dermod}
+\Der[f](\alpha)=
+\bigvee\left\{
+f(\beta\cup [x]) \ \Big \vert  \ 
+\iota_{n}(\beta)\otimes \iota_{1}(x) \leq S(\alpha)
+\right\}
+\end{align}
+where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
+and $S: !(X\times Y)\to !X\otimes !Y$ is the Seely isomorphism \cite{}, 
+satisfying all axioms (D1)-(D7) and (DCurry), see the Appendix.
+\end{enumerate}
+
+This leads then to:
+\begin{theorem}
+$\Mod_{!}$ (equivalently, $\GMet_{!}$), equipped with $\Der$, is a $CC\partial C$.
+\end{theorem}
+
+Notice that, when $f:!(\Lawv^{X})\to \Lawv^{Y}$, so that $f$ is described by a matrix $\widehat f: \multiset(X)\times Y\to \Lawv$, one can check that \eqref{eq:dermod} coincides with the definition of the differential operator in $\LREL$.

### 2023-01-12 section6.tex
commit

@@
 where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
-and $S: !(X\times Y)\to !X\otimes !Y$ is the Seely isomorphism \cite{}, 
+and $S: !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{}, 
 satisfying all axioms (D1)-(D7) and (DCurry), see the Appendix.

### 2023-01-12 section6.tex
commit

@@
 
-Notice that, when $f:!(\Lawv^{X})\to \Lawv^{Y}$, so that $f$ is described by a matrix $\widehat f: \multiset(X)\times Y\to \Lawv$, one can check that \eqref{eq:dermod} coincides with the definition of the differential operator in $\LREL$.
+Notice that, when $f:!(\Lawv^{X})\to \Lawv^{Y}$, so that $f$ is described by a matrix $\widehat f: \multiset(X)\times Y\to \Lawv$, one can check that the matrix associated with \eqref{eq:dermod} coincides with the definition of the differential operator in $\LREL$.

### 2023-01-12 main.tex
commit
Push serale
Non so se dopo mangiato avrò le forze per scrivere qualcosina in più.
La sezione 3 va ricontrollata, aggiustata ecc, ma ora ci sono tutti gli elementi.

@@
 \usetikzlibrary{cd}
+\usepackage{mathabx}
+\usepackage{cmll}
 \usepackage{stmaryrd}
@@
 
-\title{Lambda Calculus through Tropical Mathematics}
+\title{Tropical Mathematics for the Lambda Calculus}
 
@@
 
-\section{Two Quantitative Approaches to the $\lambda$-calculus}
+\section{A Bridge Between Two Quantitative Approaches to the $\lambda$-calculus}
 \input{section2}
@@
 
-\section{Analytic Results}
+\section{Tropical Laurent series}
 \input{section4}%
@@
 
-\section{Quasi-metric Spaces and $\Lawv$-Modules}\label{sec:QMod}
+\section{Generalized Metric Spaces and Modules over the Lawvere Quantale}\label{sec:QMod}
 \input{section6}%

### 2023-01-12 section3B.tex

@@
 
-It is known ([reference??]) that the maps $X\mapsto \C M_{\leq n}(X)$ ($n\in\BB N$) lift to functors $!_n:\LREL\to\LREL$.
+It is known {\color{red}(([reference??])} that the maps $X\mapsto \C M_{\leq n}(X)$ ($n\in\BB N$) lift to functors $!_n:\LREL\to\LREL$.
 The sequence $(!_n)_{n\in\N}$ can be then turned into a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, satisfying the adjunction:
@@
 Therefore, $\LREL$, together with $(!_n)_{n\in\N}$, is a model for $\BSTLC$. 
-Concretely, a type $A::= o \mid !_{n}A \multimap A$ (i tipi meglio dirli in Section II) is in interpreted as $\model o=\B 1$ and 
+Concretely, a type %$A::= o \mid !_{n}A \multimap A$
+{\color{red}(i tipi meglio dirli in Section II)} is in interpreted as $\model o=\B 1$ and 
 $\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$.
@@
 It is instructive at this point to see what its CCC-structure looks like in our tropical world.
-In particular, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t:\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
+In particular, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
 \[
@@
 The exponential object $X\multimap Y$ is $!X+Y$ (where $+$ is the disjoint union).
-Remember that in $\Lawv$ the neutral element for addition is $\infty$ and the neutral for multiplication is $0$, so for instance the evaluation morphism is the matrix $\RM{eval}^{X,Y}\in\Lawv^{!((X\multimap Y) + X)\times Y}\simeq\Lawv^{(!!X\times !Y\times !X)\times Y}$ given by $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=0$ if $\rho_1=[\mu]$ and $\rho_2=[b]$, and $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=\infty$ otherwise.
+Remember that in $\Lawv$ the neutral element for addition is $\infty$ and the neutral for multiplication is $0$, so for instance the evaluation map is the matrix $\RM{eval}^{X,Y}\in\Lawv^{!((X\multimap Y) + X)\times Y}\simeq\Lawv^{(!!X\times !Y\times !X)\times Y}$ given by $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=0$ if $\rho_1=[\mu]$ and $\rho_2=[b]$, and $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=\infty$ otherwise.
 
@@
 
-\begin{example}
+\begin{example}\label{ex:famous_ex}
  The function $f:\Lawv\to\Lawv$, $f(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ is a tropical Laurent series: it is of shape $f=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.

### 2023-01-12 section4.tex
Sezioni riorganizzate
Ora tutte le sezioni del sorgente dovrebbero corrispondere alle sezioni del pdf
commit

--- 2023-01-08 section4.tex
@@
+In this section we study the tropical Laurent series $\Lawv^X\to \Lawv ^Y$ from the viewpoint of analysis.
+
+Let us start by recalling Example~\ref{ex:famous_ex}.
+By plotting its graph {\color{red}vogliamo plottarlo?}, we see first of all that the function is non-decreasing and concave.
+It is easy to see that this is actually always the case:
+
+\begin{proposition}
+ Any tropical Laurent series $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
+\end{proposition}
+
+The $f$ of Example~\ref{ex:famous_ex} is continuous on $\BB R_{\geq0}=\Lawv-\set{\infty}$ (w.r.t.\ the usual norm of real numbers).
+By considering the usual norm $\norm{x}_\infty:=\sup\limits_{a\in X} \absv{x_a}$ on $\Lawv^X$, we could generalise this property by dropping the case of $x$ having some $0$ coordinate:
+
+\begin{theorem}\label{thm:cont}
+ Any tropical Laurent series $f:\Lawv^X\to\Lawv$ is continuous on $\BB R_{>0}$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
+\end{theorem}
+\begin{proof}
+ The result follows after proving that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
+\end{proof}
+
+Not only the $f$ of Example~\ref{ex:famous_ex} is continuous, but it is also locally Lipschitz on $\BB R_{>0}$.
+Actually, we can prove the following:
+
+\begin{theorem}
+ Let $f:\Lawv\to\Lawv$ a tropical Laurent series with matrix $\widehat f:\N\to\Lawv$.
+ For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N$ s.t.:
+ \begin{enumerate}
+  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv$;
+  \item If $f(x_0)=\infty$ for some $x_0<\infty$, then $\C F_\epsilon=\emptyset$;
+  \item On all $[\epsilon,\infty]$, $f$ coincides with the tropical \emph{polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\widehat f(n)}$.
+ \end{enumerate}
+ In particular, $f$ is locally Lipschitz on all $\BB R_{>0}$.
+\end{theorem}
+\begin{proof}
+ We can let $\C F_\epsilon:=\set{n\in\N \mid
+ \widehat f(n)<\infty \textit{ and } \widehat f(m)> \widehat f(m)+\epsilon \textit{ for all } m<n}$.
+\end{proof}
+
+Remark that, in coherence with the previous result, in Example~\ref{ex:famous_ex} $f(x)$ is indeed a $\min$ for all $x>0$.
+At $x=0$ we have $f(x=0)=\inf\limits_{n\in\N} \frac{1}{2^n}=0$ which is \emph{not} a min.
+Also, while the derivative of $f$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
+This phenomenon is reminiscent of [Example 7, PCoh]
+%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
+, which actually motivated our first investigations.
+
+We could generalise the locally Lipschitz property as follows:
+\begin{theorem}
+ Let $f:\Lawv^X\to\Lawv$.
+ If $f$ is non-decreasing, concave and continuous, then it is locally Lipschitz.
+\end{theorem}
+\begin{proof}
+ Refinement of the arguments used for Theorem~\ref{thm:cont}.
+ {\color{red}Che ci scriviamo?}
+\end{proof}
+
+Highlight compositional reasoning based on Lipschitz.\\
+
 - the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric
 
-- concavity
-
-- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.
-
-- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
-\begin{itemize}
-
-\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally
-
-\item singular points: the famous example
-
-\item tropical roots
-
-\end{itemize}
+{\color{red}
+- extend this result to functions $\Lawv^{X}\to \Lawv^{Y}$ with $X,Y$ finite (as this will be useful in section V) Quale ?}
 
@@
 
-- characterization of the functional metric using derivatives
+- characterization of the functional metric using derivatives\\
+
+Let us end this section by mentioning another point of view on tropical Laurent series.
+$\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
+Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete normed cone.
+The normed cone structure induces an order on it, called its \emph{cone order}, by setting:
+$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
+This order makes it a Scott-continuous dcpo.
+Furthermore we have:
+
+\begin{proposition}
+  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}$, w.r.t.\ the cone orders on the domain and codomain.
+\end{proposition}

### 2023-01-13 section7.tex
commit

@@
-- correspondence between $\Lawv$-modules and $\Lawv$-categories.
+The connections between the differential $\lambda$-calculus (and differential linear logic), the relational semantics, and the theory of non-idempotent intersection types is very well-studied (see \cite{}, and more recently, \cite{} for a more abstract perspective, and \cite{} for a 2-categorical, or proof-relevant, extension).
+As we said, the relational semantics over the tropical semi-ring was quickly explored in \cite{}, to provide a ``best case'' resource analysis of a $\B{PCF}$-like language with non-deterministic choice.
+\emph{Probabilistic coherent spaces} \cite{}, a variant of  the relational semantics, provide an interpretation of higher-order probabilistic programs
+as analytic functions. In \cite{} it was observed that such functions satisfy a local Lipschitz condition somehow reminiscent of our examples in Section \ref{section4}.
 
-- exponential structure of $\LCAT$.
+The study of linear or bounded type systems for sensitivity analysis was initiated in \cite{} and later developed \cite{}.
+As recalled in the paper, the use bounded exponentials ensures that well-typed programs satisfy a Lipschitz condition.
+Related approaches, although not based on metrics, are provided by \emph{differential logical relations} \cite{} and \emph{change action} models \cite{}.
 
-- $\LCAT_{!}$ is a cartesian closed differential category
+More generally, the literature on program metrics in denotational semantics is vast.
+Since [6] metric spaces, also in Lawvere's generalized sense \cite{}, have been exploited as an alternative framework to standard, domain-theoretic, denotational semantics.
+While standard categories of metric spaces are not models of the full simply typed $\lambda$-calculus, several constructions of cartesian closed categories of metric spaces can be found in the literature.
+For instance,
+\emph{ultra}-metric spaces \cite{} form a CCC, and have been shown to model $\B{PCF}$ \cite{}.
+Also \emph{partial} metrics, introduced in \cite{}, have been shown to provide models of $\STLC$, under suitable generalizations \cite{}.
+More generally, \cite{} provides a general characterization of exponentiable objects in categories of (generalized) metric spaces, and \cite{} provides other ways to construct CCCs of
+(generalized) metrics, including one based on locally Lipschitz maps, 
+using ideas from {differential logical relations}.
+
+Motivated by connections with computer science and fuzzy set-theory, 
+the abstract study of generalized metric spaces in the framework of \emph{quantale}- or even \emph{quantaloid}-enriched categories has led to a vast literature in recent years \cite{}, 
+and its connections with tropical mathematics are discussed in \cite{}. Moreover, applications of quantale-modules to both logic and computer science have also been studied \cite{}.
+
+Connections between program metrics and the differential $\lambda$-calculus have been already suggested in \cite{}; moreover, \emph{cartesian difference categories} \cite{} have been proposed as a way to relate derivatives in differential categories with those found in change action models.
+
+Finally, applications of tropical mathematics in computer science abound, notably in optimization methods for machine learning \cite{} (typically, for neural networks with peicewise linear activation), linear regression \cite{}, convex analysis \cite{}, as well as finite automata \cite{}.
+While our discussion in Section \ref{section5} is inspired by a well-known application of tropical polynomials \cite{},  
+the vast literature in this domain lets us think that other ways to 
+apply tropical semantics to the analysis of higher-order programs might be studied.
+
+%
+%Other connections tropical/metric -> Fuji, ??
+%Applications of tropical to computer science.
+%Log-probabilities.
+%Quantale-modules -> Abramski?
+%
+%

### 2023-01-14 section3C.tex
Mini-commit
Ho aggiustato minime cose, ed ho elencato alcuni punti da dire nelle conclusions

@@
 \end{theorem}
-Here, $\Lambda^-:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$ is the uncarry operator and $\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is an operator corresponding to the syntactical \emph{differential substitution}.
+Here, $\Lambda^-:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$ is the uncarry operator and $\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is an operator corresponding to the syntactical \emph{differential substitution} of $\STDLC$.
 \begin{proof}
- Che ci scriviamo ??
+ {\color{red}(Che ci scriviamo ??}
 \end{proof}

### 2023-01-14 section3intro.tex
Mini-commit
Ho aggiustato minime cose, ed ho elencato alcuni punti da dire nelle conclusions

@@
 
-In the remainder of this section we will see how tropical linear algebra -- expressed as $\LREL$ -- gives rise to denotational models of computation -- expressed as \lamcalc.
+In the remainder of this section we will see how tropical linear algebra -- expressed as $\LREL$ -- gives rise to denotational models of computation -- expressed as variants of the \lamcalc.

### 2023-01-14 section4.tex
Mini-commit
Ho aggiustato minime cose, ed ho elencato alcuni punti da dire nelle conclusions

@@
 
-\begin{proposition}
+\begin{proposition}\label{prop:nondecr+conc}
  Any tropical Laurent series $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
@@
 \begin{theorem}\label{thm:cont}
- Any tropical Laurent series $f:\Lawv^X\to\Lawv$ is continuous on $\BB R_{>0}$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
+ Any tropical Laurent series $f:\Lawv^X\to\Lawv$ is continuous on $\BB R_{>0}^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
 \end{theorem}
@@
 
+We believe that a variant of such result holds in $\Lawv^X$ for $X$ finite.
+We do not know what happens if $X$ is infinite.
+
 Remark that, in coherence with the previous result, in Example~\ref{ex:famous_ex} $f(x)$ is indeed a $\min$ for all $x>0$.
@@
 \end{proof}
+
+Proposition~\ref{prop:nondecr+conc} and Theorem~\ref{thm:cont} immediately entail:
+\begin{corollary}\label{cor:TLSlocLip}
+ Tropical Laurent series $\Lawv^X\to\Lawv$ are locally Lipschitz on on $\BB R_{>0}$.
+\end{corollary}
 
@@
 \begin{proposition}
-  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}$, w.r.t.\ the cone orders on the domain and codomain.
+  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders on the domain and codomain.
 \end{proposition}

### 2023-01-14 section8.tex
Mini-commit
Ho aggiustato minime cose, ed ho elencato alcuni punti da dire nelle conclusions

We have shown that:

- intuitively, the interplay between tropical mathematics and $\lam$-calculus \emph{could} relate the metric and differential approaches on approximations of $\lam$-calc (introduction+section 2)

- this relation \emph{can} take place, because the natural category $\LREL$ and its generalised versions are metric models of the differential $\lam$-calculus (section 3 and 6)

- this relation \emph{seems} to provide applications in different fields (section 5).

Therefore, we mainly set the basis for future interplays between all these areas, hopefully motivating the interest in such an interplay.
For instance, the general questions are:

- Can we improve the results of section 4 ?

- Can we develop and make the applications of section 5 useful ?

- What does the general setting of section 6 give in terms of theoretical and applied results ?

- Do tools from tropical \emph{geometry} provide something for $\lam$-calculus ? (for instance, the role of tropical roots, tropical varieties,...)

- Finally, there is a last point which we think of interest and we did not mention through the paper: the inclusion of finiteness spaces in the picture.
Say what could they do and why.

### 2023-01-14 section5.tex
Commit temporaneo
Ho cominciato a scrivere la storia delle log-proba, ma non ho finito e per ora l'ho scritta così come viene.

@@
 a variant of the relational semantics modeling strongly normalizing programs via \emph{finite} power series.
+We mention this point in the Section~\ref{sec:concl}
+
+\subsection{Maslov dequantisation}
+
+{\color{red}MI SA CHE QUESTO PEZZETTO SERVE ANCHE QUI, O NO?}
+{\color{red}L'ho scritta qui cos\`i ma non merita una sezione tutta per se, non so se/come vogliamo dire Maslov}
+First, let us recall that for any non-negative real $t$, the functions $\phi_{t}(x)=-t\log x$ and $\psi_{t}(\alpha)=e^{-\alpha/t}$ are inverse of each other and define thus continuous (w.r.t.\ the usual topologies) bijections between the space of probabilities $[0,1]$ and $\BB R_{\geq 0}\cup\set{\infty}$ (we write $\log$ for the natural logarithm).
+Moreover, if we set $\alpha \widetilde+ \beta:= \frac{\alpha+\beta}{2}$, $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta:=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$, it is known that: $\lim\limits_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$.
+In this sense, setting $\Lawv_t:=([0,\infty],\sumt{t},\prodt{t})$, one says that $\Lawv_t\to_{t\to 0^+}\Lawv$.
+Moreover, setting $\widetilde\Lawv:=([0,\infty],\widetilde+,\cdot)$, it can be shown that $\Lawv_t\simeq\widetilde\Lawv$ for all $t>0$, so the $\Lawv_t$ are all isomorphic, whereas at the limit we have a discontinuity: it can be shown that $\Lawv_t\not\simeq\Lawv$.
+These considerations are called the \emph{Maslov dequantisation} of $\Lawv$ [??].
+%Moreover, the two tropical  operations $\min$ and $+$ on $\BB R_{\geq 0}\cup\set{\infty}$ can be obtained via these bijections as the limit of the operations $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$; we have: $\lim_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$ and $\lim_{t\to 0}\alpha\prodt{t}y= \alpha+\beta$.
 
@@
 
-{\color{red}MI SA CHE QUESTO PEZZETTO SERVE ANCHE QUI, O NO?}
-First, let us recall that for any positive real $t$, the functions $\phi_{t}(x)=-t\log x$ and $\psi_{t}(\alpha)=e^{-\alpha/t}$ define continuous bijections between the space of probabilities $[0,1]$ and $\Lawv$.
-Moreover, the two tropical  operations $\min$ and $+$ can be obtained via these bijections by letting $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)+\psi_{t}(\beta)$ and $\alpha\prodt{t} \beta=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))$, and observing that $\lim_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$ and $\lim_{t\to 0}\alpha\prodt{t}y= \alpha+\beta$.
+Let us consider in this subsection a probabilistic extension of $\lam$-calculus, adding as usual terms of shape $pM+qN$ and $M\oplus_p N$, for $p,q\in[0,1]$.
+These terms are typed via the rule:
+\[
+ \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
+\]
+and similar for $\Gamma\vdash pM+qN:A$.
+We add the rule:
+\[
+ M\oplus_p N \to pM+(1-p)N
+\]
+so that such terms play the role of probabilistic choices of parameter $p$, as well as the rule:
+\[
+ pM+qM\to (p+q)M.
+\]
+
+\begin{remark}
+{\color{red}By using [Weighted], it can be shown that $\LREL_!$ still is a model of such calculus. GIUSTIFICARLO BENE ?}
+\end{remark}
+
+Let us consider $M:=(I\oplus_p\Omega)\oplus_ps((I\oplus_p\Omega)\oplus_p(I\oplus_p\Omega))$.
+Reducing to normal form, we have:
+\[
+ M\twoheadrightarrow (p^2+(1-p)p^2+(1-p)^3)I+(p(1-p)+2(1-p)^2p)\Omega.
+\]
+Setting $q:=1-p$, we can consider the two weights of $I$ and $\Omega$ as polynomials in the variables $p$ and $q$, and call
+$Q_1(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ the weight of $I$, while
+$Q_2(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ is the weight of $\Omega$,
+where the $P$'s are the following monomials in $p,q$:
+$P_{00}(p,q):=p^2$,
+$P_{100}(p,q):=qp^2$,
+$P_{111}(p,q):=q^3$,
+$P_{01}(p,q):=pq$,
+$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
+The index $\omega\in\set{00,100,111,01,110,101}$ of each $P_\omega$ indicate the path of the reduction that led from $M$ to the respective occurrence of $I_\omega$ of $I$ or $\Omega_\omega$ of $\Omega$ from $M$ to its normal form ($0$ means ``left'' and $1$ means ``right'').
+For instance, in order to reach $I_{100}$, i.e.\ the second occurrence of $I$ from the left in $M$, we have to take the right path in the outer $\oplus_p$ of $M$, then two times the left path in the new outer $\oplus_p$'s that we encounter during reduction.
+The $P_{100}(p,q):=qp^2$ gives then the probability (as a function of $p,q$) of obtaining such occurrence of $I$ in the normal form if we were to sample at each time we reduce a $\oplus_p$.
+Moreover, the polynomials $Q_{1,2}(p,q)$ give the whole probability of obtaining respectively $I$ or $\Omega$, in the normal form after such samplings.
+
+The tropical point of view allows us to express two natural questions about this situation:
+\begin{enumerate}
+ \item What is the maximum likelihood estimator for the event ``$M$ produces $I_\omega$''?
+This means asking what is the choice of the parameters $p,q$ that maximizes the probability of getting $I_\omega$ by reducing $M$.
+ \item Knowing that $M$ produced $I$ (similarly for $\Omega$), which occurrence $I_\omega$ is the most likely to have been produced ? I.e., which is the most likely path $\omega$ to have been taken?
+\end{enumerate}
 

### 2023-01-15 section2.tex
Commit
Ti ho messo la lista delle cose da dire in Sezione II.C (se ho scordato qualcosa vedi tu ovviamente!)

@@
 \subsection{Tropical mathematics: a possible synthesis}
-The fundamental observation that led to the study of mathematics over the \emph{tropical semi-ring} $\Lawv=([0,\infty],\min,+)$ was that, by replacing everywhere the ``$+$'' by the ``$\min$'' and the ``$\times$'' by the ``$+$'', the computation of many algebraic and geometric objects becomes combinatorial and simple.
+
+{\color{red}Lista delle cose da dire:}
+
+1) Def di quantale, come lattice e come complete idempotent semiring.
+Among the the so-called \emph{tropical semirings}, we consider the \emph{Lawvere quantale/semiring}.
+
+2) Def di $\Lawv$, the \emph{Lawvere quantale}: seen as the idempotent complete semiring, it is $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,\cdot,0)$.
+Seen as lattice it is defined by the order $\preceq$, which is the reversed order $\geq$ of the usual order $\leq$ on $\BB R_{\geq0}\cup\set{\infty}$.
+
+3) Maslov dequantisation:
+
+First, let us recall that for any non-negative real $t$, the functions $\phi_{t}(x)=-t\log x$ and $\psi_{t}(\alpha)=e^{-\alpha/t}$ are inverse of each other and define thus continuous (w.r.t.\ the usual topologies) bijections between the space of probabilities $[0,1]$ and $\BB R_{\geq 0}\cup\set{\infty}$ (we write $\log$ for the natural logarithm).
+Moreover, if we set $\alpha \widetilde+ \beta:= \frac{\alpha+\beta}{2}$, $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta:=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$, it is known that: $\lim\limits_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$.
+In this sense, setting $\Lawv_t:=([0,\infty],\sumt{t},\prodt{t})$, one says that $\Lawv_t\to_{t\to 0^+}\Lawv$.
+Moreover, setting $\widetilde\Lawv:=([0,\infty],\widetilde+,\cdot)$, it can be shown that $\Lawv_t\simeq\widetilde\Lawv$ for all $t>0$, so the $\Lawv_t$ are all isomorphic, whereas at the limit we have a discontinuity: it can be shown that $\Lawv_t\not\simeq\Lawv$.
+
+4) Def di $\trop$ di un polinomio/serie, (\`e la stessa formula, dipende solo se gli indici sono finiti/infiniti).
+Come sta scritto sotto, giusto un po' pi\`u formale (per esempio, scriverlo come Definizione).
+
+The fundamental observation that led to the study of mathematics over the \emph{tropical semi-ring} $\Lawv=([0,\infty],\min,+)$ was that, by replacing everywhere the ``$+$'' by the ``$\min$'' and the ``$\times$'' by the ``$+$'', many algebraic and geometric objects becomes combinatorial and their computation simpler.
 
@@
 
-Since polynomials correspond to linear (hence Lipschitz) functions in tropical mathematics, the reconstruction of the relational semantics over the tropical semi-ring, presented in Section \ref{section3} and Section \ref{section4}, will provide a metric semantics of the differential $\lambda$-calculus, bridging sensitivity and resource analysis.
+\subsection{Outline of the paper}
+
+Since polynomials correspond to piecewise linear (hence Lipschitz) functions in tropical mathematics, the reconstruction of the relational semantics over the tropical semi-ring, presented in Section \ref{section3} and Section \ref{section4}, will provide a metric semantics of the differential $\lambda$-calculus, bridging sensitivity and resource analysis.
 In Section \ref{section5} we will suggest potential applications of this approach, relating well-studied applications of program metrics, resource analysis with current uses of tropical mathematics in computer science.

### 2023-01-15 section5.tex
Commit
Ti ho messo la lista delle cose da dire in Sezione II.C (se ho scordato qualcosa vedi tu ovviamente!)

@@
 We mention this point in the Section~\ref{sec:concl}
-
-\subsection{Maslov dequantisation}
-
-{\color{red}MI SA CHE QUESTO PEZZETTO SERVE ANCHE QUI, O NO?}
-{\color{red}L'ho scritta qui cos\`i ma non merita una sezione tutta per se, non so se/come vogliamo dire Maslov}
-First, let us recall that for any non-negative real $t$, the functions $\phi_{t}(x)=-t\log x$ and $\psi_{t}(\alpha)=e^{-\alpha/t}$ are inverse of each other and define thus continuous (w.r.t.\ the usual topologies) bijections between the space of probabilities $[0,1]$ and $\BB R_{\geq 0}\cup\set{\infty}$ (we write $\log$ for the natural logarithm).
-Moreover, if we set $\alpha \widetilde+ \beta:= \frac{\alpha+\beta}{2}$, $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta:=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$, it is known that: $\lim\limits_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$.
-In this sense, setting $\Lawv_t:=([0,\infty],\sumt{t},\prodt{t})$, one says that $\Lawv_t\to_{t\to 0^+}\Lawv$.
-Moreover, setting $\widetilde\Lawv:=([0,\infty],\widetilde+,\cdot)$, it can be shown that $\Lawv_t\simeq\widetilde\Lawv$ for all $t>0$, so the $\Lawv_t$ are all isomorphic, whereas at the limit we have a discontinuity: it can be shown that $\Lawv_t\not\simeq\Lawv$.
-These considerations are called the \emph{Maslov dequantisation} of $\Lawv$ [??].
-%Moreover, the two tropical  operations $\min$ and $+$ on $\BB R_{\geq 0}\cup\set{\infty}$ can be obtained via these bijections as the limit of the operations $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$; we have: $\lim_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$ and $\lim_{t\to 0}\alpha\prodt{t}y= \alpha+\beta$.
 

### 2023-01-15 section2.tex
commit

--- 2023-01-10 section2.tex
@@
 A natural way to account for these situations is to consider denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which measures differences of behavior.
-A fundamental insight coming from this works is that \emph{affine} programs, that is, programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
+A fundamental insight coming from this line of works is that \emph{affine} programs, that is, programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
 $d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
-A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive 
-map \emph{symmetric monoidal closed}, and provides a model of the \emph{affine} simply typed $\lambda$-calculus.
+A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive provides a model of the \emph{affine} simply typed $\lambda$-calculus, being both a \emph{symmetric monoidal closed} category (i.e.~a model of \emph{linear} $\STLC$) and a cartesian category in which cartesian and monoidal unit coincide.
 
-As observed in \cite{Gabo2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications, by considering \emph{Lipschitz-continuous} functions.
-For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the functional input $f$, can be interpreted as a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.
+As observed in \cite{Gabo2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications.
+The fundamental intuition is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz-continuous} map.
+For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the functional input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.
 By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $r\cdot X$ (i.e.~with distance $d_{r}(x,y)=r\cdot d(x,y)$) and $Y$, the program $M$ above
@@
 
-These observations have led to the study of \emph{bounded} typed $\lambda$-calculi (like the system $\mathsf{Fuzz}$ \cite{Reed_2010}),
-which have been applied in the study of higher-order differential privacy \cite{}.
-In these calculi, which are inspired from Girard's \emph{bounded linear logic} \cite{}, function types are of the form $!_{r}X \multimap Y$, and programs are interpreted as $r$-Lipschitz maps.
-In categorical terms, the re-scaling operation $!_{r}(-)$ can be characterized as a  \emph{graded linear exponential comonad} \cite{}.
+These observations have led to the study of \emph{bounded} typed $\lambda$-calculi (like the system $\mathsf{Fuzz}$ \cite{Reed_2010}), which have been applied in the study of higher-order differential privacy \cite{}.
+The types of such systems are defined \cite{} by combining linear constructors with a \emph{graded exponential comonad} $!_{r}(-)$ \cite{}.
+In the following sections we will sometimes make reference to a basic type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined as below
+\begin{align}
+A::= o \  \mid \  (!_{\leq n}A) \multimap A  
+\end{align}
+where $n$ indicates a natural number.
+Intuitively, $!_{\leq n}A\multimap B$ is the type of functions from $A$ to $B$ that may use their input at most $n$ times.
+More details about $\BSTLC$ are provided in the Appendix.
 
-Then, what about the good old, ``unbounded'', simply typed $\lambda$-calculus?
+Now, what about the good old, ``unbounded'', simply typed $\lambda$-calculus?
 Actually, by using unbounded duplications, one might lose the Lipschitz property.
@@
 k\cdot x: \BB R\to \BB R$ are all Lipschitz-continuous, with Lipschitz constant $k$, the function $M=\lambda x.x^{2}$ obtained by ``duplicating'' $x$ is not Lipschitz anymore: $M$ is, so to say, \emph{too} sensitive to errors.
-More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of the simply typed $\lambda$-calculus (yet, $\Met$ contains several cartesian closed \emph{sub-}categories, see e.g.~\cite{}).
+More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of $\STLC$ (yet, $\Met$ contains several cartesian closed \emph{sub-}categories, see e.g.~\cite{}).
 Still, one might observe that the program $M$ above is actually Lipschitz-continuous, if not globally, at least \emph{locally} (i.e.~ over any compact set).
@@
 
-% - Metric models: functions as Lipschitz maps, differential privacy.
-%
-%- Problem with unrestricted duplications
-%
-%- Bounded Linear Logic/lambda-calculus, why Met is only a model of BLL, and why it is difficult (but not impossible!) to construct models of full lambda
-
 \subsection{Resource approach: differential $\lambda$-terms as polynomials}
 
-The key ingredient of these approaches is a notion of \emph{linear application}, which in the syntax of the differential $\lambda$-calculus is written $\Der[M]\cdot N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once.
-More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument.
+A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} (and differential linear logic) and its categorical models. 
+The key ingredient is that a \emph{differential operator} $\Der$ is added to the usual syntax of the $\lambda$-calculus, with the intuition that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M]\cdot N$, which still has type $A\to B$, corresponds to a \emph{linear application} of $N$ to $M$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M]\cdot N$ has type $A\to B$, since $M$ might still ask for other copies of an input of type $A$).
+
+Interestingly, the categorical study of such differential operators has led to introduce a family of categories, called \emph{cartesian differential categories} $C\partial C$ \cite{}, which provide an abstract axiomatization of differentiation, in the usual mathematical sense.
+More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
+\begin{itemize}
+\item $\C C$ is left-additive, i.e.~its hom-sets are commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
+\item $\C C$ is equipped with a differential operator $D:
+\C C(X,Y)\to \C C(X\times X,Y)$ satisfying 7 axioms $D1,\dots, D7$, which translate usual properties of the differential (e.g.~the linearity of $D$ in one of its two variables, the chain rule,  Schwarz's theorem etc.).
+\end{itemize}
+
+Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with the differential operator $\Der[-]\cdot (-)$, together with a notion of \emph{linear substitution} (see \cite{} for details).
+The models of $\STDLC$ are the 
+cartesian \emph{closed} differential categories ($CC\partial C$), 
+which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential are both well-behaved with respect to the closed structure (see \cite{}).
+
+Another suggestive similarity between program derivatives and 
+actual derivatives is provided by the \emph{Taylor expansion}:
+in $\STDLC$ one can prove that any application $MN$ can be expanded as a sum of \emph{linear} applications
+$\Der^{(k)}[M]\cdot N^{k}$, i.e.~where $N$ is passed exactly $k$ times to $M$, as expressed below:
+\begin{align}\label{eq:taylor}
+MN  =  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot (\Der^{(k)}[M] \cdot N^{k})0
+\end{align}
+In other words, terms with unbounded duplications can be seen as some sort of ``limits'' of bounded, but arbitrarily large, duplications.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
+
+%More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument. 
 %Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
@@
 %\end{align}
-More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
+%More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
+%
+%
+%The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy.
+\frac{\mathsf df(y)}{\mathsf dy}\cdot x$.
+Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
+%Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:
+
+At this point, given our previous discussion of bounded duplications and Lipschitz-continuity, a natural question would be whether the Taylor formula
+could be used to decompose an unbounded $\lambda$-term 
+into a sum of Lipschitz maps, thus bridging the metric and differential approach.  
+While suggestive, an obstacle to realize this idea is that,
+in general, a $C\partial C$ needs not satisfy the Taylor formula
+\eqref{eq:taylor}, and in the \emph{relational semantics}, 
+a well-studied and sort of canonical ``Taylor'' semantics for $\STDLC$, the bounded terms appearing in the Taylor expansion correspond to \emph{polynomials}, and thus to non-Lipschitz functions.
+
+Our question can thus be reformulated as follows: can we make the relational semantics \emph{Lipschitz}, hence amenable to metric and sensitivity analysis?
+The goal of this paper is to show that, by appealing to tropical mathematics, this is indeed possible and leads to a surprising
+bridge between the metric and differential study to higher-order programs.
+%
+%More generally, the relational semantics interprets unbounded programs as \emph{analytic functions}, that is, as functions admitting a representation as power series.
+For instance, observing that an analytic map $f: \BB R\to \BB R$, where $f(x)=\sum_{n}\widehat f_{n}\cdot x^{n}$ is uniquely determined by the sequence $\widehat f_{n}$, the program $M_{\infty}:=\lambda fx.fx: (\BB R\To \BB R)\To (\BB R\To \BB R)$ is represented by the power series below:
+%\begin{align}
+%F_{\infty}(f,x)= \sum_{n=0}^{\infty} \widehat f_{n} x^{n}
+%\end{align}
+%By restricting ourselves to bounded applications, the terms in the power series become finite, that is, the interpretation becomes a \emph{polynomial}: for instance, the program $M_{2}:=\lambda fx.
+\sum_{i=0}^{2}\Der^{(i)}[f]\cdot x^{i}$, corresponding to passing $x$ \emph{at most twice} to $f$, is represented by the polynomial
+%\begin{align}
+%F_{\leq 2}(f,x)=\widehat f_{2} x^{2}  + \widehat f_{1}x +  \widehat f_{0} 
+%\end{align}
+% In this framework the differential operator is naturally represented by formal differentiation of polynomials, where, as one would expect, 
+% $\Der[\sum_{n}a_{n}x^{n}]=\sum_{n}\Der[a_{n}x^{n}]$ and $\Der[a_{0}x^{0}]=0$ and $\Der[a_{n+1}x^{n+1}]= (n+1)a_{n+1}x^{n}$, so that power series can be Taylor expanded.
+
+\subsection{Outline of the paper}
+
+In the next section we recall some basic ideas from tropical mathematics, and its connection with the study of the Lawvere quantale.
+%Since polynomials correspond to piecewise linear (hence Lipschitz) functions in tropical mathematics,
+The reconstruction of the relational semantics over the tropical semi-ring, presented in Section \ref{section3} and Section \ref{section4}, will provide a metric semantics of the differential $\lambda$-calculus, bridging sensitivity and resource analysis.
+In Section \ref{section5} we suggest potential applications of this approach, relating well-studied applications of program metrics, resource analysis with current uses of tropical mathematics in computer science.
+Finally, in Section \ref{section6} we show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between
+\emph{generalized metric spaces} (in the sense of Lawvere \cite{}) and modules over the tropical semi-ring.
+
+\section{Tropical mathematics: a possible synthesis}
+
+At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv$ \cite{}, which is at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{}.
+Let us recall that a quantale is a complete lattice endowed with a continuous monoid action.
+In our case we have $\Lawv=([0,\infty], \geq,+)$.
+Observe that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$.
+In particular, as discussed in Section \ref{section3}, tropical linear algebra coincides with the study of matrices with values taken over $\Lawv$.
+At the same time, a (pseudo-)metric is a function $d:X\times X\to \Lawv$ satisfying usual axioms, e.g.~the triangular law (indeed, such $d$ can be seen as a $\Lawv$-\emph{enriched category}, see Section \ref{section6}).
+
+A \emph{tropical polynomial} is a piecewise linear function $\varphi:\Lawv\to \Lawv$ of the form 
+\begin{align}\label{eq:polytrop}
+\varphi(\alpha)= \min_{i=1,\dots,k}\left\{ i\alpha + c_{i}\right\}
+\end{align}
+with coefficients $c_{i}$ taken from $\Lawv$. A value $\alpha\in \Lawv$ is a \emph{root} of the polynomial $P$ when
+the minimum at $\varphi(\alpha)$ is attained at least twice (equivalently, when 
+ $\varphi$ is not differentiable at $\alpha$). In other words, the tropical roots of $\varphi$ coincide with the points where the slope of $\varphi$ changes.
+
+Intuitively, tropical polynomials look much like standard polynomials, although with ``$+$'' replaced by ``$\min$'', and ``$\times$'' replaced by ``$+$''. 
+In fact, this intuition can be made precise as follows: for any positive real $t$, the tropical polynomials are in one to one correspondence with the functions $f:[0,1]\to [0,1]$ which can be written as a \emph{parameterized} polynomial
+$f_{t}(x)= \sum_{i=1}^{n}t^{c_{i}}x^{n}$, with the $c_{i}\in [0,\infty]$. 
+%Hence, for any polynomial $p(x)= 
+%For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
+%\begin{align}
+%\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
+%\end{align}
+
+More generally, for any function $f:[0,1]\to [0,1]$ which can be written as a (parameterized) \emph{power series} of the form $f(x)= \sum_{n}t^{c_{n}}x^{n}$ (such functions arise naturally from the interpretation of probabilistic programs, as we'll recall in Section \ref{section5}), we can define its \emph{tropicalization} $\trop f: \Lawv \to \Lawv$
+as the function 
+\begin{align}
+\trop f(\alpha) =\inf_{n}\left\{ i\alpha+c_{i}\right\}
+\end{align}
+Such functions, called \emph{tropical Laurent series} \cite{}, will be discussed in more detail in Section \ref{section5}.
+
+The functions $\varphi$ and $f$ can be related by a limit passage as follows: the functions $\phi_{t}(x)= -\log_{t}x$ and $\varphi_{t}(\alpha)= t^{-\alpha}$ define continuous bijections between $[0,1]$ and $[0,\infty]$ and, by letting
+$\varphi_{t}: [0,\infty]\to [0,\infty]$ be defined by 
+$\varphi_{t}(\alpha)= \phi_{t}\circ f \circ \psi_{t}$, one has that 
+$\varphi= \lim_{t\to 0}\varphi_{t}$. 
+Indeed, one can check that the ``parameterized'' sums and product $\alpha \sumt{t}\beta:= \phi_{t}(\psi_{t}(\alpha)+\psi_{t}(\beta))= \log_{t}(t^{-\alpha}+t^{-\beta})$ and 
+$\alpha \prodt{t}\beta:= \phi_{t}( \psi_{t}(\alpha)\psi_{t}(\beta))=
+\log_{t}(t^{-\alpha}t^{-\beta})$ converge respectively to $\min\{\alpha,\beta\}$ and $\alpha+\beta$, when $t\to 0$ (this remark is an example of the so-called \emph{Maslov dequantization}, relating the semi-ring $\BB R$ with the tropical semiring $((-\infty, \infty], \min, +)$, see e.g.~\cite{}).
+
+Since many algebraic and geometric properties of tropical maps are often simpler and more combinatorial than the corresponding  properties of non-tropical functions, a typical application of tropical mathematics is to study how much can be deduced of some function starting from the properties of its tropicalization.
+In Section \ref{section5} we will follow a similar direction, investigating what quantitative properties of a higher-order programs are revealed by the study of its tropical interpretation.
+
+%
+%
+%or more generally as a power series $f(x)=\sum_{n}\widehat f_{n}x^{n}$ with coefficient $\widehat f_{n}\in [0,1]$, we can define its \emph{tropicalization} $\trop f: \Lawv \to \Lawv$ as the function 
+%\begin{align}
+%\trop f(\alpha)= \inf_{n}\left\{ 
+%\end{align}
+%
+%This correspondence can be made precise through the the so-called
+% \emph{de Maslov dequantization} \cite{}.
+% For each positive real $t$, any polynomial in $\BB R[x]$ can be written under the \emph{$t$-parameterized} form:
+% \begin{align}
+% p_{t}(x)= \sum_{i=1}^{k}t^{c_{i}}x^{i}
+% \end{align}
+% with the coefficients $c_{i}$ taken from $\Lawv$. 
+% It is clear then that tropical polynomials and $t$-parameterized polynomials admit a one to one correspondence between their presentations.
+% 
+% Actually, the $\varphi$s and the $p_{t}$s can be related by passing through some intermediate functions $\varphi_{t}$ introduced by Maslov.
+%For any $t>1$, the functions $\phi_{t}(x)=-\log_{t}x$ and $\psi_{t}(\alpha)=t^{-\alpha}$ are inverse of each other and define thus continuous (w.r.t.\ the usual topologies) bijections between the space of probabilities $[0,1]$ and $\BB R_{\geq 0}\cup\set{\infty}$ (we write $\log$ for the natural logarithm).
+%Moreover, if we set $\alpha \widetilde+ \beta:= \frac{\alpha+\beta}{2}$, $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta:=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$, it is known that: $\lim\limits_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$.
+%In this sense, setting $\Lawv_t:=([0,\infty],\sumt{t},\prodt{t})$, one says that $\Lawv_t\to_{t\to 0^+}\Lawv$.
+%Moreover, setting $\widetilde\Lawv:=([0,\infty],\widetilde+,\cdot)$, it can be shown that $\Lawv_t\simeq\widetilde\Lawv$ for all $t>0$, so the $\Lawv_t$ are all isomorphic, whereas at the limit we have a discontinuity: it can be shown that $\Lawv_t\not\simeq\Lawv$.
 %
 
-The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy.
-\frac{\mathsf df(y)}{\mathsf dy}\cdot x$.
-Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
-Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:
-\begin{align}
-MN  =  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot (\Der^{(k)}[M] \cdot N^{k})0
-\end{align}
-Now, given our previous discussion of bounded duplications and Lipschitz-continuity, a natural question would be whether the differential approach to the $\lambda$-calculus may lead to decompose an unbounded $\lambda$-term
-into a sum of Lipschitz functions, thus bridging the metric and differential approach.  
-While suggestive, this idea seems to contradict the fact that, in the canonical semantics of the differential $\lambda$-calculus, the relational semantics \cite{}, the bounded terms appearing in the Taylor expansion correspond to \emph{polynomials}, and thus to non-Lipschitz functions.
-
-More generally, the relational semantics interprets unbounded programs as analytic functions, that is, as functions admitting a representation as power series.
-For instance, observing that an analytic map $f: \BB R\to \BB R$, where $f(x)=\sum_{n}\widehat f_{n}\cdot x^{n}$ is uniquely determined by the sequence $\widehat f_{n}$, the program $M_{\infty}:=\lambda fx.fx: (\BB R\To \BB R)\To (\BB R\To \BB R)$ is represented by the power series below:
-\begin{align}
-F_{\infty}(f,x)= \sum_{n=0}^{\infty} \widehat f_{n} x^{n}
-\end{align}
-By restricting ourselves to bounded applications, the terms in the power series become finite, that is, the interpretation becomes a \emph{polynomial}: for instance, the program $M_{2}:=\lambda fx.
-\sum_{i=0}^{2}\Der^{(i)}[f]\cdot x^{i}$, corresponding to passing $x$ \emph{at most twice} to $f$, is represented by the polynomial
-\begin{align}
-F_{\leq 2}(f,x)=\widehat f_{2} x^{2}  + \widehat f_{1}x +  \widehat f_{0} 
-\end{align}
- In this framework the differential operator is naturally represented by formal differentiation of polynomials, where, as one would expect, 
- $\Der[\sum_{n}a_{n}x^{n}]=\sum_{n}\Der[a_{n}x^{n}]$ and $\Der[a_{0}x^{0}]=0$ and $\Der[a_{n+1}x^{n+1}]= (n+1)a_{n+1}x^{n}$, so that power series can be Taylor expanded.
-
-Going back to our previous question, we can now reformulate it as follows: can we make the relational semantics \emph{Lipschitz}, hence amenable to metric and sensitivity analysis?
-The goal of this paper is to show that this is indeed possible and actually quite natural, by appealing to tropical mathematics.
-
-\subsection{Tropical mathematics: a possible synthesis}
-The fundamental observation that led to the study of mathematics over the \emph{tropical semi-ring} $\Lawv=([0,\infty],\min,+)$ was that, by replacing everywhere the ``$+$'' by the ``$\min$'' and the ``$\times$'' by the ``$+$'', the computation of many algebraic and geometric objects becomes combinatorial and simple.
-
-For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
-\begin{align}
-\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
-\end{align}
-Notably, the \emph{tropical roots} (whose definition is recalled in Section \ref{section3}) of $\trop p(\alpha)$ can be found through a rather simple (indeed polytime \cite{}) algorithm, and can be used to \emph{approximate} the actual roots of $p(x)$ \cite{}.
-More generally, the tropicalization of a power series $f(x)=\sum_{n}\widehat f_{n}x^{n}$ yields a \emph{tropical Laurent series} \cite{} 
-\begin{align}
-\trop f(\alpha)= \inf_{n}\left\{n\alpha+ \widehat f_{n}\right\}
-\end{align}
-a class of functions that we will study in detail in Section \ref{section4}.
-
-Since polynomials correspond to linear (hence Lipschitz) functions in tropical mathematics, the reconstruction of the relational semantics over the tropical semi-ring, presented in Section \ref{section3} and Section \ref{section4}, will provide a metric semantics of the differential $\lambda$-calculus, bridging sensitivity and resource analysis.
-In Section \ref{section5} we will suggest potential applications of this approach, relating well-studied applications of program metrics, resource analysis with current uses of tropical mathematics in computer science.
-Finally, in Section \ref{section6} we will show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between
-\emph{generalized metric spaces} (in the sense of Lawvere \cite{}) and modules over the tropical semi-ring.
-
-%- generalities about tropical maths (tropicalisation $\trop P$  of polynomials and of Laurent series, and their roots -- all that without $\LREL$)
+%
+%{\color{red}Lista delle cose da dire:}
+%
+%1) Def di quantale, come lattice e come complete idempotent semiring.
+%Among the the so-called \emph{tropical semirings}, we consider the \emph{Lawvere quantale/semiring}.
+%
+%2) Def di $\Lawv$, the \emph{Lawvere quantale}: seen as the idempotent complete semiring, it is $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,\cdot,0)$.
+%Seen as lattice it is defined by the order $\preceq$, which is the reversed order $\geq$ of the usual order $\leq$ on $\BB R_{\geq0}\cup\set{\infty}$.
+%
+%3) Maslov dequantisation:
+%
+%First, let us recall that for any non-negative real $t$, the functions $\phi_{t}(x)=-t\log x$ and $\psi_{t}(\alpha)=e^{-\alpha/t}$ are inverse of each other and define thus continuous (w.r.t.\ the usual topologies) bijections between the space of probabilities $[0,1]$ and $\BB R_{\geq 0}\cup\set{\infty}$ (we write $\log$ for the natural logarithm).
+%Moreover, if we set $\alpha \widetilde+ \beta:= \frac{\alpha+\beta}{2}$, $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta:=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$, it is known that: $\lim\limits_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$.
+%In this sense, setting $\Lawv_t:=([0,\infty],\sumt{t},\prodt{t})$, one says that $\Lawv_t\to_{t\to 0^+}\Lawv$.
+%Moreover, setting $\widetilde\Lawv:=([0,\infty],\widetilde+,\cdot)$, it can be shown that $\Lawv_t\simeq\widetilde\Lawv$ for all $t>0$, so the $\Lawv_t$ are all isomorphic, whereas at the limit we have a discontinuity: it can be shown that $\Lawv_t\not\simeq\Lawv$.
+%
+%4) Def di $\trop$ di un polinomio/serie, (\`e la stessa formula, dipende solo se gli indici sono finiti/infiniti).
+%Come sta scritto sotto, giusto un po' pi\`u formale (per esempio, scriverlo come Definizione).
+%
+%The fundamental observation that led to the study of mathematics over the \emph{tropical semi-ring} $\Lawv=([0,\infty],\min,+)$ was that, by replacing everywhere the ``$+$'' by the ``$\min$'' and the ``$\times$'' by the ``$+$'', many algebraic and geometric objects becomes combinatorial and their computation simpler.
+%
+%For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
+%\begin{align}
+%\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
+%\end{align}
+%Notably, the \emph{tropical roots} (whose definition is recalled in Section \ref{section3}) of $\trop p(\alpha)$ can be found through a rather simple (indeed polytime \cite{}) algorithm, and can be used to \emph{approximate} the actual roots of $p(x)$ \cite{}.
+%More generally, the tropicalization of a power series $f(x)=\sum_{n}\widehat f_{n}x^{n}$ yields a \emph{tropical Laurent series} \cite{} 
+%\begin{align}
+%\trop f(\alpha)= \inf_{n}\left\{n\alpha+ \widehat f_{n}\right\}
+%\end{align}
+%a class of functions that we will study in detail in Section \ref{section4}.
+%
+%%- generalities about tropical maths (tropicalisation $\trop P$  of polynomials and of Laurent series, and their roots -- all that without $\LREL$)
+%
+%
+%

### 2023-01-15 section2.tex
commit

@@
 
-At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv$ \cite{}, which is at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{}.
+At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{}.
 Let us recall that a quantale is a complete lattice endowed with a continuous monoid action.
-In our case we have $\Lawv=([0,\infty], \geq,+)$.
-Observe that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$.
-In particular, as discussed in Section \ref{section3}, tropical linear algebra coincides with the study of matrices with values taken over $\Lawv$.
-At the same time, a (pseudo-)metric is a function $d:X\times X\to \Lawv$ satisfying usual axioms, e.g.~the triangular law (indeed, such $d$ can be seen as a $\Lawv$-\emph{enriched category}, see Section \ref{section6}).
+In the case of $\Lawv$ the lattice is defined by the reverse order $\geq$, and the monoid action is provided by addition.
+Notice that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$.
+A consequence of these observation is that, as discussed in Section \ref{section3}, the tropical approach to linear algebra coincides with the study of ``$\Lawv$-valued matrices'', i.e.~of maps of the form $s: X\times Y\to \Lawv$ .
+In particular, a (pseudo-)metric on a set $X$ is nothing but a ``$\Lawv$-valued matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we discuss in Section \ref{section6}).
 
-A \emph{tropical polynomial} is a piecewise linear function $\varphi:\Lawv\to \Lawv$ of the form 
+Going beyond linear algebra, a \emph{tropical polynomial} is defined as a piecewise linear function $\varphi:\Lawv\to \Lawv$ of the form 
 \begin{align}\label{eq:polytrop}
@@
 \end{align}
-with coefficients $c_{i}$ taken from $\Lawv$. A value $\alpha\in \Lawv$ is a \emph{root} of the polynomial $P$ when
-the minimum at $\varphi(\alpha)$ is attained at least twice (equivalently, when 
- $\varphi$ is not differentiable at $\alpha$). In other words, the tropical roots of $\varphi$ coincide with the points where the slope of $\varphi$ changes.
-
+where the coefficients $c_{i}$ are taken from $\Lawv$. 
+%A value $\alpha\in \Lawv$ is a \emph{root} of the polynomial $P$ when
+%the minimum at $\varphi(\alpha)$ is attained at least twice (equivalently, when 
+% $\varphi$ is not differentiable at $\alpha$). In other words, the tropical roots of $\varphi$ coincide with the points where the slope of $\varphi$ changes. 
+%
 Intuitively, tropical polynomials look much like standard polynomials, although with ``$+$'' replaced by ``$\min$'', and ``$\times$'' replaced by ``$+$''. 
-In fact, this intuition can be made precise as follows: for any positive real $t$, the tropical polynomials are in one to one correspondence with the functions $f:[0,1]\to [0,1]$ which can be written as a \emph{parameterized} polynomial
+In fact, this intuition can be made precise as follows: for any positive real $t$, the tropical polynomials like $\varphi$ are in one-to-one correspondence with the functions $f:[0,1]\to [0,1]$ which can be written as a \emph{parameterized} polynomial
 $f_{t}(x)= \sum_{i=1}^{n}t^{c_{i}}x^{n}$, with the $c_{i}\in [0,\infty]$. 
@@
 
-More generally, for any function $f:[0,1]\to [0,1]$ which can be written as a (parameterized) \emph{power series} of the form $f(x)= \sum_{n}t^{c_{n}}x^{n}$ (such functions arise naturally from the interpretation of probabilistic programs, as we'll recall in Section \ref{section5}), we can define its \emph{tropicalization} $\trop f: \Lawv \to \Lawv$
-as the function 
+More generally, for any function $f:[0,1]\to [0,1]$ which can be written as a (parameterized) \emph{power series} of the form $f_{t}(x)= \sum_{n}t^{c_{n}}x^{n}$ (as we'll see, such functions arise naturally from the interpretation of probabilistic programs, as we'll recall in Section \ref{section5}), we can define its \emph{tropicalization} $\trop f: \Lawv \to \Lawv$
+as the function defined as follows:
 \begin{align}
-\trop f(\alpha) =\inf_{n}\left\{ i\alpha+c_{i}\right\}
+\trop f(\alpha) =\inf_{n}\left\{ n\alpha+c_{n}\right\}
 \end{align}
@@
 
-The functions $\varphi$ and $f$ can be related by a limit passage as follows: the functions $\phi_{t}(x)= -\log_{t}x$ and $\varphi_{t}(\alpha)= t^{-\alpha}$ define continuous bijections between $[0,1]$ and $[0,\infty]$ and, by letting
-$\varphi_{t}: [0,\infty]\to [0,\infty]$ be defined by 
-$\varphi_{t}(\alpha)= \phi_{t}\circ f \circ \psi_{t}$, one has that 
-$\varphi= \lim_{t\to 0}\varphi_{t}$. 
+The functions $f$ and $\trop f$ can be related by a limit passage as follows: the functions $\phi_{t}(x)= -\log_{t}x$ and $\varphi_{t}(\alpha)= t^{-\alpha}$ define continuous bijections between $[0,1]$ and $[0,\infty]$ and, by letting
+$\trop_{t}f: [0,\infty]\to [0,\infty]$ be defined by 
+$\trop_{t}f(\alpha)= \phi_{t}\circ f \circ \psi_{t}$, one has that 
+$\trop f= \lim_{t\to 0}\trop_{t}f$. 
 Indeed, one can check that the ``parameterized'' sums and product $\alpha \sumt{t}\beta:= \phi_{t}(\psi_{t}(\alpha)+\psi_{t}(\beta))= \log_{t}(t^{-\alpha}+t^{-\beta})$ and 

### 2023-01-15 section2.tex
commit

@@
 Finally, in Section \ref{section6} we show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between
-\emph{generalized metric spaces} (in the sense of Lawvere \cite{}) and modules over the tropical semi-ring.
+\emph{generalized metric spaces} and modules over the tropical semiring.
 
@@
 Notice that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$.
-A consequence of these observation is that, as discussed in Section \ref{section3}, the tropical approach to linear algebra coincides with the study of ``$\Lawv$-valued matrices'', i.e.~of maps of the form $s: X\times Y\to \Lawv$ .
-In particular, a (pseudo-)metric on a set $X$ is nothing but a ``$\Lawv$-valued matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we discuss in Section \ref{section6}).
+A consequence of these observations is that, as discussed in Section \ref{section3}, the tropical approach to linear algebra coincides with the study of ``$\Lawv$-valued matrices'', i.e.~of maps of the form $s: X\times Y\to \Lawv$ .
+In particular, a (pseudo-)metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we discuss in Section \ref{section6}).
 

### 2023-01-15 section1.tex
commit

@@
-In the recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability,
+In recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability,
 approximation bounds, 
 as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence. 
-
-Notably, a significative effort has been made in order to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties.
+Notably, a significative effort has been made to extend, or adapt, well-established qualitative methods, like e.g.~type systems, relational logics or denotational semantics, to account for quantitative properties.
 We can mention, for example,
@@
 probabilistic \cite{PCOH, QBS} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
-Usually, the main reason to resort to such logical, type-theoretical or denotational approaches is to obtain methods which are \emph{modular} and \emph{compositional}, that is, where properties of complex algorithms can be deduced from properties of their component parts.
+The main reason to look for methods relying on (quantitative extensions of) type-theory or denotational semantics is that these approaches yield \emph{modular} and \emph{compositional} techniques, that is, allow one to deduce properties of complex algorithms from the properties of their component parts.
 
@@
 
-Among such quantitative approaches, two different directions have received a considerable attention in recent years.
+Among such quantitative approaches, two different directions have received a considerable attention.
 
-On the one hand there is the approach based on \emph{non-idempotent} intersection types \cite{} and their semantical counterpart, the so-called \emph{relational semantics} \cite{}, which have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{}, and have also been shown suitable to capture probabilistic properties \cite{Manzo}.
-
-On the other hand there is the approach of \emph{program metrics} \cite{} and \emph{quantitative equational theories} \cite{Mardare}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}.
+On the one hand there is the approach of \emph{program metrics} \cite{} and \emph{quantitative equational theories} \cite{Mardare}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}.
 This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{} and coinductive methods \cite{}, and was recently extended to account for the the full $\lambda$-calculus \cite{LICS2021, FSCD2022}.
 
+On the other hand there is the approach based on \emph{differential} or \emph{resource-aware} extensions of the $\lambda$-calculus, which is well-connected with the \emph{relational semantics} \cite{} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{}.
+This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{}, and have also been shown suitable to capture probabilistic properties \cite{Manzo}.
+
 In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{linearlogic} (i.e.~of using its input exactly once), plays a crucial role.
-In the relational semantics linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider \emph{analytic} maps, i.e.~formal infinite series of monomials of the form $a_{n}\cdot x^{n}$.
-By contrast, in metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances, while the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gabo2017}, and general programs as maps satisfying coarser
-continuity conditions (see e.g.~\cite{LICS2021}).
+In the metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gabo2017}.
+By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
 
 A natural question is thus whether these two, apparently unrelated, ways of interpreting linearity and duplication can be somehow connected.
-At a first glance, there seems to be a fundamental ``logarithmic'' gap between the two approaches.
-In the relational semantics a typical ``duplicating'' map is obtained by composing the diagonal with multiplication:
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x\cdot y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the square product function $\lambda x.x^{2}$.
-However, in metric semantics this function needs not even exist (as these models are often restricted to Lipschitz-continuous maps \cite{Gabo2017})!
+At a first glance, there seems to be a fundamental ``logarithmic'' gap between the two approaches:
+in metric models $n$ times duplication results in a linear (hence Lipschitz) function $n\cdot x$, while in differential models this results to
+a monomial $x^{n}$, which is not Lipschitz. The fundamental insight of this article is that a natural way to overcome 
+this obstacle 
+is to work in the framework of 
+tropical mathematics.
+%
+%''
+%
+%s
+%emantics a typical ``duplicating'' map is obtained by composing the diagonal with multiplication:
+%$$
+%\begin{tikzcd}
+%\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
+% & &  &
+% \mathbb R\times \mathbb R 
+% \ar{rrr}{\langle x,y\rangle \mapsto x\cdot y}
+% & & & \mathbb R
+%\end{tikzcd}
+%$$
+%yielding the square product function $\lambda x.x^{2}$.
+%However, in metric semantics this function needs not even exist (as these models are often restricted to Lipschitz-continuous maps \cite{Gabo2017})!
 Instead, a typical ``duplicating'' map can be obtained by composing the diagonal with the sum
-$$
-\begin{tikzcd}
-\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
- & &  &
- \mathbb R\times \mathbb R 
- \ar{rrr}{\langle x,y\rangle \mapsto x+y}
- & & & \mathbb R
-\end{tikzcd}
-$$
-yielding the linear (and Lipschitz) function $\lambda x.2x$.
+%$$
+%\begin{tikzcd}
+%\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
+% & &  &
+% \mathbb R\times \mathbb R 
+% \ar{rrr}{\langle x,y\rangle \mapsto x+y}
+% & & & \mathbb R
+%\end{tikzcd}
+%$$
+%yielding the linear (and Lipschitz) function $\lambda x.2x$.
+%
+%As this example seems to suggest, there seems to be a sort of ``logarithmic'' gap between the two approaches. Can this be made explicit?
 
-As this example seems to suggest, there seems to be a sort of ``logarithmic'' gap between the two approaches. Can this be made explicit?
+\subsection{Tropical mathematics and program semantics }
 
-\subsection{Tropical Geometry: a ``logarithmic'' version of usual geometry}
+Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon (whose origin is, as far as we know, the only reason for the ``tropical'' adjective) as a way to do algebra and geometry by replacing the usual ring structure of numbers based on addition and multiplication by the semi-ring structure given by ``$\min$'' and ``$+$''.
+%
+%
+% interpreting the usual ``$\times$'' and  ``$+$'' operations by  ``$+$'' by ``$\min$''.
+It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
+Tropical mathematics is a form of \emph{idempotent} mathematics, since the role of addition is 
+played by the idempotent operation $\min$.
+For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$, when interpreted over the tropical semi-ring, becomes the piecewise linear function
+$
+\varphi(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
+$.
 
-In very simple words, tropical geometry describes what happens to usual algebra and geometry when we replace ``$\cdot$'' by  ``$+$'', and ``$+$'' by ``$\min$''.
-It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
-Tropical geometry is a kind of \emph{idempotent} mathematics, since the role of addition is 
-played by the idempotent operation $\min$.
-For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$ corresponds in the tropical word correspond to the \emph{linear} function
-$$
-\trop p(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
-$$
-
-The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials).
+The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials, or studying geometric invariants like e.g.~the genus).
 For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning \cite{}.
 
-Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for: a way to see polynomials as linear (i.e.~Lipschitz) functions.
+Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for, since polynomials become Lipschitz functions.
 At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail.
@@
 
-The goal of this paper is to make such intuitive connections precise, yielding a way to relate, conceptually, two different quantitative approaches to higher-order algorithms, as well as suggesting new potential applications of tropical geometry to the study of quantitative properties of programs.
+The goal of this paper is to study the tropical semantics of the differential $\lambda$-calculus and use it to connect it with metric semantics.
+This bridge does not only provide a conceptual unification of two different quantitative approaches to higher-order algorithms, but it also
+suggests ways in which techniques for resource-analysis could be used for sensitivity analysis and vice-versa, opening the way for new potential applications of tropical geometry to the study of quantitative properties of higher-order programs.
 

### 2023-01-15 section1.tex
commit

@@
 A natural question is thus whether these two, apparently unrelated, ways of interpreting linearity and duplication can be somehow connected.
-At a first glance, there seems to be a fundamental ``logarithmic'' gap between the two approaches:
-in metric models $n$ times duplication results in a linear (hence Lipschitz) function $n\cdot x$, while in differential models this results to
-a monomial $x^{n}$, which is not Lipschitz. The fundamental insight of this article is that a natural way to overcome 
-this obstacle 
-is to work in the framework of 
+At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
+in metric models $n$ times duplication results in a linear (hence Lipschitz) function $n\cdot x$, while in differential models this results in a monomial $x^{n}$, which is not Lipschitz.
+The fundamental insight of this article is that a natural way to overcome
+this obstacle and bridge the two viewpoints 
+is to develop differential semantics in the framework of 
 tropical mathematics.

### 2023-01-15 section5.tex
Commit parte proba

--- 2023-01-12 section5.tex
@@
 a variant of the relational semantics modeling strongly normalizing programs via \emph{finite} power series.
+We mention this point in the Section~\ref{sec:concl}
 
@@
 
-{\color{red}MI SA CHE QUESTO PEZZETTO SERVE ANCHE QUI, O NO?}
-First, let us recall that for any positive real $t$, the functions $\phi_{t}(x)=-t\log x$ and $\psi_{t}(\alpha)=e^{-\alpha/t}$ define continuous bijections between the space of probabilities $[0,1]$ and $\Lawv$.
-Moreover, the two tropical  operations $\min$ and $+$ can be obtained via these bijections by letting $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)+\psi_{t}(\beta)$ and $\alpha\prodt{t} \beta=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))$, and observing that $\lim_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$ and $\lim_{t\to 0}\alpha\prodt{t}y= \alpha+\beta$.
+Let us consider in this subsection a probabilistic extension of $\lam$-calculus, call it $\STLC_\oplus$, adding as usual terms of shape $pM+qN$ and $M\oplus_p N$, for $p,q\in[0,1]$.
+These terms are typed via the rule:
+\[
+ \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
+\]
+and similar for $\Gamma\vdash pM+qN:A$.
+We add the reduction rule:
+\[
+ M\oplus_p N \to pM+(1-p)N
+\]
+so that such terms play the role of probabilistic choices of parameter $p$, as well as the rule:
+\[
+ pM+qM\to (p+q)M.
+\]
+
+Let us consider $M:=(I\oplus_p\Omega)\oplus_p((I\oplus_p\Omega)\oplus_p(I\oplus_p\Omega))$.
+Reducing to normal form, we have:
+\[
+ M\twoheadrightarrow (p^2+(1-p)p^2+(1-p)^3)I+(p(1-p)+2(1-p)^2p)\Omega.
+\]
+Setting $q:=1-p$, we can consider the two weights of $I$ and $\Omega$ as polynomials in the variables $p$ and $q$.
+Let us call
+$Q_1(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ the weight of $I$ and
+$Q_2(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the weight of $\Omega$,
+where the $P$'s are the following monomials in $p,q$:
+$P_{00}(p,q):=p^2$,
+$P_{100}(p,q):=qp^2$,
+$P_{111}(p,q):=q^3$,
+$P_{01}(p,q):=pq$,
+$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
+The index $\omega\in\set{00,100,111,01,110,101}$ of each $P_\omega$ indicates the path of the reduction that led from $M$ to the respective occurrence $I_\omega$ of $I$ or $\Omega_\omega$ of $\Omega$ from $M$ to its normal form ($0$ means ``left'' and $1$ means ``right'').
+For instance, in order to reach $I_{100}$, i.e.\ the second occurrence of $I$ from the left in $M$, we have to take the right path in the outer $\oplus_p$ of $M$, then two times the left path in the new outer $\oplus_p$'s that we encounter during reduction.
+$P_{\omega}(p,q)$ gives then the probability (as a function of $p,q$) of obtaining the respective occurrence $I_{\omega}$ or $\Omega_\omega$ in the normal form, if we were to sample at each time we reduce a $\oplus_p$.
+It can thus also be read as the likelihood function of such event.
+The polynomials $Q_{1,2}(p,q)$ give instead the whole probability of obtaining respectively $I$ or $\Omega$, in the normal form after such samplings.
+
+The tropical point of view allows us to express two natural questions about this situation:
+\begin{enumerate}
+ \item What is the maximum likelihood estimator for the event ``$M$ produces $I_\omega$'' (similarly for $\Omega$)?
+This means asking what is the choice of the parameters $p,q$ that maximizes the probability of getting $I_\omega$ by reducing $M$.
+ \item Knowing that $M$ produced $I$ (similarly for $\Omega$), which is the choice of the parameters $p,q$ that maximizes the probability that the produced occurrence of $I$ was a fixed $I_{\omega_0}$ (i.e.
+that among the paths leading to $I$, the one taken was $\omega_0$)?
+\end{enumerate}
+
+For 1), we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)$ belong to:
+\begin{equation}
+  \begin{array}{r@{}c@{}l@{}}
+   \underset{(x,y)}{\RM{arg max}} P_\omega (x,y)
+   & = &
+   \underset{(x,y)}{\RM{arg min}}\set{-\log_c P_\omega (x,y)}
+   \\
+   & = &
+   \underset{(x,y)}{\RM{arg max}}\set{(\trop P_\omega) (-\log_c x,-\log_c y)} \label{eq:argmax}
+  \end{array}
+\end{equation}
+where the equalities hold for all $0<c\neq1$.
+
+For 2), %the $\omega_0$ is s.t.\ $\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$. So
+we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
+$\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$.
+This last condition is equivalent to ask that:
+\begin{equation}
+  \begin{array}{l@{}l@{}}
+    \min\limits_{\omega\in\set{00,100,111}} \, -\log_c P_\omega(x,y) = -\log_c P_{\omega_0}(x,y)
+   & \textit{ i.e.: }
+   \\
+   (\trop Q_1)( -\log_c x, -\log_c y) = (\trop P_{\omega_0})( -\log_c x, -\log_c y). &  \label{eq:max}
+  \end{array}
+\end{equation}
+In both cases, the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.
+
+\begin{fact}
+ By using [Weighted], it can be shown that $\LREL_!$ is a model of such calculus $\STLC_\oplus$. {\color{red}GIUSTIFICARLO BENE ?}
+\end{fact}
+
+Now, in $\LREL$, seen as a model of such probabilistic $\lam$-calculus, the interpretation of a term already computes the tropicalisation of the polynomials expressing the probabilities, because the underlying semiring of the model is tropical.
+For instance, for our running example $M$:
+\[\model M = \min\left\{\trop{Q_1}(p,1-p) \cdot \model I, \trop{Q_2}(p,1-p) \cdot \model \Omega\right\}.\]
+So, assuming the interpretation of a term known and assuming the tropical polynomials are extracted from it, one could study the optimisation problems \ref{eq:argmax}, \ref{eq:max} \emph{without} having to reduce the term.
+
+\begin{example}\label{ex:study}
+ For our running example $M$, we have $\trop Q_1(x,y)=\min\set{2x,y+2x,3y}$ and $\trop Q_2(x,y)=\min\set{x+y,2y+x}$.
+ Studying $\trop Q_1$ we see that {\color{red}Vogliamo plottare $\trop Q_1$ ?}
+ $\trop Q_1(x,y)=3y$ iff $y\leq \frac{2}{3}x$, and it coincides with $2x$ otherwise.
+ Remembering that $3y=P_{111}(x,y)$, we can now solve the optimisation problem~\ref{eq:max} for $\omega_0=111$:
+ via the substitution $x:=-\log_c p$, $y:=-\log_c (1-p)$, Equation~\ref{eq:max} is equivalent to $-\log_c (1-p)\leq -\frac{2}{3}\log_c p$, i.e.\ $1-p\geq p^{\frac{2}{3}}$.
+ This means that, for $p\in[0,1]$ s.t.\ $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), the most likely occurrence of $I$ to obtain, knowing that $M$ produced $I$, is $I_{111}$.
+ Remembering that $2x=P_{00}(x,y)$, for the other values of $p$ (for example, $p=\frac{1}{2}$), the most likely $I$ to be sampled is the occurrence $I_{00}$.
+ We have thus answered question 2) above.
+\end{example}
+
+\begin{remark}
+ Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_1$ or of $\trop Q_2$ give, by definition of tropical root, the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable paths of samples during the reduction of $M$ to its normal form.
+ Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log_c p,-\log_c(1-p))$ is a tropical root of, say, $\trop Q_1$ are the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable occurrences of $I$ that are sampled by $M$ during the reduction, knowing that some $I$ was sampled.
+\end{remark}
 

### 2023-01-15 section4.tex
commit

@@
 In this section we study the tropical Laurent series $\Lawv^X\to \Lawv ^Y$ from the viewpoint of analysis.
+
+\subsection{Results on Tropical Laurent Series}
+{\color{red}
+
+- recall Trop Laurent Series (TLS) tropical roots
+
+- Example 1
+
+- The morphisms of $\LREL_{!}$ can be seen as generalizations of TLS
+
+- Remark: for $f:\Lawv^{X}\to \Lawv^{Y}$, if $X$ is finite, $f$ is a finite variable TLS; if $f$ is written as a finite $\min$, it is a tropical polynomial.
+
+- Maybe Theorem 7 could even go here??
+Saying something like: after the remark above, we can prove a result that confirms what seen in Example 1, for the case when $X$ is finite: $f$ can LOCALLY be seen as a polynomial, the degree of such polynomials potentially diverging when $\epsilon\to 0$.
+Indeed...
+
+- bisognerebbe anche dire due parole sul differenziale e dell'espansione di Taylor come operazione sulle TLS invece che sulle matrici. 
+}
 
@@
 \end{proposition}
+
+\subsection{$\Lawv^{X}$ as a normed space.}
+
+{\color{red}Move here also the cone part}
 
@@
 
-\begin{theorem}
+\begin{theorem}\label{theorem:fepsilon}
  Let $f:\Lawv\to\Lawv$ a tropical Laurent series with matrix $\widehat f:\N\to\Lawv$.
@@
 
-We could generalise the locally Lipschitz property as follows:
-\begin{theorem}
- Let $f:\Lawv^X\to\Lawv$.
- If $f$ is non-decreasing, concave and continuous, then it is locally Lipschitz.
-\end{theorem}
+\subsection{$\Lawv^{X}$ as a metric space.}
+
+The norm $\|-\|_\infty$ naturally induces a metric $\|\B x-\B y\|_{\infty}$ over the spaces $\Lawv^{X}$. 
+We now study the Lipschitz properties of tropical Laurent series w.r.t.~these metrics.
+
+A first result is the following:
+\begin{proposition}\label{prop:troplinear}
+All tropical linear functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive.  
+\end{proposition}
+\begin{proof}[Proof sketch]
+Using the fact that $f(\B x)_{b}= \inf_{a\in X}\check f_{a,b}+\B x_{a}$,
+the problem reduces to checking that $|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
+This result shows that, in analogy with that happens in usual metric semantics, linear programs are interpreted by non-expansive functions. 
+%\begin{proof}
+%Using $f(\B x)_{b}= \inf_{a\in X}\check f_{a,b}+\B x_{a}$,
+%first observe that $|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$; we now have
+%$|f(\B x)_{b}-f(\B y)_{b}| \leq |(\inf_{a\in X}\check f_{a,b}-\B x_{a})-(\inf_{a\in X}\check f_{a,b}-\B y_{a})| \leq
+%\sup_{a\in X}|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})|\leq 
+% \| \B x-\B y\|_{\infty}$.
+%\end{proof}
+
+Before looking at what happens in the case of non-linear programs, let us make the metric structure of $\LREL$ explicit.
+
+Using the fact that the tropical linear functions from $\Lawv^{X}$ to $\Lawv^{Y}$ are in bijection with the elements of the hom-set $\LREL(X,Y)\simeq \Lawv^{X\times Y}$ through the map $f\mapsto \check f$, the following proposition provides a useful characterization of the functional metrics in $\LREL$:
+\begin{proposition}
+For all sets $X,Y$ and tropical linear functions $f,g:\Lawv^{X}\to \Lawv^{Y}$, $\| \check f-\check g\|_{\infty} =  \sup_{\B x\in \Lawv^{X}}
+\| f(\B x)-g(\B x)\|_{\infty}$.\end{proposition}
+
+To study what happens in the case of tropical Laurent series, let us first consider the case of bounded exponentials $!_{\leq n}X$:
+\begin{proposition}\label{prop:boundedlip}
+All tropical linear functions $f: \Lawv^{\C M_{\leq n}(X)}\to \Lawv^{Y}$ are $n$-Lipschitz-continuous.
+\end{proposition}
+\begin{proof}[Proof sketch]
+Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \check f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
+$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
+it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
+\end{proof}
+This result is perfectly analogous to what happens in standard models for sensitivity analysis based on the re-scaling trick recalled in Section \ref{section2}.
+
+We can now look at what happens in the case of the full exponential comonad, i.e.~with the tropical Laurent series.
+First, let us observe that, as expected, tropical polynomials 
+correspond to Lipschitz functions:
+\begin{proposition}
+Any tropical polynomial $\varphi:\Lawv^{\{1,\dots, K\}}\to\Lawv$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
+\end{proposition}
 \begin{proof}
- Refinement of the arguments used for Theorem~\ref{thm:cont}.
- {\color{red}Che ci scriviamo?}
+This immediately follows from Prop.~\ref{prop:boundedlip}, since a tropical polynomial can be represented as a tropical linear function from $\Lawv^{\C M_{\mathrm{deg}(\varphi)(\{1,\dots, K\})}}$ to $\Lawv$.
 \end{proof}
 
-Proposition~\ref{prop:nondecr+conc} and Theorem~\ref{thm:cont} immediately entail:
-\begin{corollary}\label{cor:TLSlocLip}
- Tropical Laurent series $\Lawv^X\to\Lawv$ are locally Lipschitz on on $\BB R_{>0}$.
+The proposition above, together with Theorem \ref{theorem:fepsilon}, can be used to show that tropical Laurent series with \emph{finitely many} variables are always locally Lipschitz over $(0,\infty)^{X}$.
+Yet, we can prove a more general statement, covering also the infinitary case.
+
+\begin{theorem}
+ All tropical Laurent series $\Lawv^X\to\Lawv$ are locally Lipschitz on $\BB R_{>0}$.
+\end{theorem}
+\begin{proof}[Proof sketch]
+The core of the proof is a convex analysis argument (see the Appendix) showing that an arbitrary function $f:\Lawv^X\to\Lawv$ which is non-decreasing, concave and continuous, must be locally Lipschitz.
+\end{proof}
+
+The results just presented translate into the following facts about the interpretation of higher-order programs:
+\begin{corollary}
+For any $\lambda$-term $M$:
+\begin{enumerate}
+\item if $\Gamma \vdash_{\BSTLC} M:A$, then $\model M:\model\Gamma \to \model A$ is a Lipschitz map.
+\item if $\Gamma \vdash_{\STLC}M:A$, then $\model M: \model\Gamma \to \model A$ is a locally Lipschitz map.
+\end{enumerate}
 \end{corollary}
 
-Highlight compositional reasoning based on Lipschitz.\\
+Finally, let us discuss the differential structure. We show that the distance between two tropical maps can be approximated using the terms appearing in their Taylor expansions:
+\begin{proposition}
+For all tropical Laurent series $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
+the functions
+$\delta^{(n)}f, \delta^{(n)}g: \Lawv^{\C M_{\leq n}(X)}\to \Lawv ^{Y}$, with 
+ $\delta^{(n)}f(\B x)= \Der^{(n)}f(\B x, \infty)$, appearing in their Taylor expansions, are $n$-Lipschitz-continuous, and moreover 
+\begin{align}
+\| \check f-\check g\|_{\infty}= \sup_{n} \| \check{\delta^{(n)}f}- \check{\delta^{(n)}g}\|_{\infty}
+\end{align}
+%where $\delta^{(n)}f:( \Lawv^{X})^{n}\to \Lawv^{X}$ is the tropical linear function $\delta^{(n)}f(\B x_{1},\dots, \B x_{n})=
+%(\Der^{(n)}f)(\B x_{1},\dots, \B x_{n}, \infty)$. 
+\end{proposition}
 
-- the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric
+%A consequence of this result is that the distance between two differential programs can always be approximated via bounded programs:
+%\begin{corollary}
+%For all $\lambda$-terms $M,N: A\to B$ of $\STDLC$, for all $\epsilon >0$, there exists $K\in \BB N$ such that 
+%$|\|\model{M}-\model{N}\|_{\infty}- | \| \model{
+%\end{corollary}
 
-{\color{red}
-- extend this result to functions $\Lawv^{X}\to \Lawv^{Y}$ with $X,Y$ finite (as this will be useful in section V) Quale ?}
-
-- discussion of the tropical Taylor expansion:
-
-- characterization of the functional metric using derivatives\\
-
+{\color{red}MOVE:
 Let us end this section by mentioning another point of view on tropical Laurent series.
@@
 \end{proposition}
+}

### 2023-01-15 section4.tex
commit

@@
 \begin{proposition}\label{prop:boundedlip}
-All tropical linear functions $f: \Lawv^{\C M_{\leq n}(X)}\to \Lawv^{Y}$ are $n$-Lipschitz-continuous.
+All tropical Laurent series $f: \Lawv^{X}\to \Lawv^{Y}$ for which $\check f:{\C M_{\leq n}(X)}\times Y\to \Lawv$, are $n$-Lipschitz-continuous.
 \end{proposition}

### 2023-01-16 section3B.tex
Proposal commit

@@
-\subsubsection{Bounded duplications}
-
-It is known {\color{red}(([reference??])} that the maps $X\mapsto \C M_{\leq n}(X)$ ($n\in\BB N$) lift to functors $!_n:\LREL\to\LREL$.
-The sequence $(!_n)_{n\in\N}$ can be then turned into a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, satisfying the adjunction:
-$\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
-Therefore, $\LREL$, together with $(!_n)_{n\in\N}$, is a model for $\BSTLC$. 
-Concretely, a type %$A::= o \mid !_{n}A \multimap A$
-{\color{red}(i tipi meglio dirli in Section II)} is in interpreted as $\model o=\B 1$ and 
-$\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$.
-
-Bounded types are interesting because of the following proposition:
-\begin{proposition}
-For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) correspond to polynomials.
-\end{proposition}
-\begin{proof}
-It suffices to check that $\model A$ is finite for all bounded types $A$. Indeed this implies that a morphism $t:\model A\to \model B$ is a finite matrix $t: \model A \times \model B \to \Lawv$.
-Hence, its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is a polynomial.
-\end{proof}
-
-For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
-\begin{itemize}
-\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
-\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
-\end{itemize}
-
-- tropical polynomials and tropical roots ?? Meglio in Section II
-
-\subsubsection{Unbounded duplications}
+%\subsubsection{Unbounded duplications}
 
@@
 
+{\color{red} Point to Sec VI}
+
 \begin{example}\label{ex:famous_ex}
@@
 \end{example}
+
+It is known {\color{red}(([reference??] e dire meglio)} that the maps $X\mapsto \C M_{\leq n}(X)$ ($n\in\BB N$) lift to functors $!_n:\LREL\to\LREL$.
+The sequence $(!_n)_{n\in\N}$ can be then turned into a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, satisfying the adjunction:
+$\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
+Therefore, $\LREL$, together with $(!_n)_{n\in\N}$, is a model for $\BSTLC$. 
+Concretely, a type %$A::= o \mid !_{n}A \multimap A$
+{\color{red}(i tipi meglio dirli in Section II)} is in interpreted as $\model o=\B 1$ and 
+$\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$.
+
+%Bounded types are interesting because of the following proposition:
+%\begin{proposition}
+%For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) correspond to polynomials.
+%\end{proposition}
+%\begin{proof}
+%It suffices to check that $\model A$ is finite for all bounded types $A$.
+Indeed this implies that a morphism $t:\model A\to \model B$ is a finite matrix $t: \model A \times \model B \to \Lawv$.Hence, its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is a polynomial.
+%\end{proof}
+
+%For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
+%\begin{itemize}
+%\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
+%\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
+%\end{itemize}

### 2023-01-16 section3A.tex
Proposal commit
Commit sul linear lam-calc

@@
+The linear simply typed $\lam$-calculus is a restriction of the ordinary $\lam$-calculus in which programs can only use their arguments \emph{exatly once} during the computation.
+This is the notion of linearity taken into account by linear logic, and indeed this calculus can be embedded in the Intuitionistic Multiplicative fragment of it (called \emph{IMLL}).
+
+More precisely, the linear $\lam$-calculus is obtained from the ordinary one by adding the constraint that each $\lam$-abstracted variable appears exactly once in the scope of the $\lam$-abstraction.
+
+In order to frame it in a category $\C C$, one needs a symmetric tensor product $\otimes$ together with exponential objects $X\multimap Y$ s.t.\ the evaluation and curry maps yield a \emph{symmetric monoidal adjunction}: $\C C(Z\otimes X, Y) \simeq \C C (Z, X\multimap Y)$.
+This is the notion of Symmetric Monoidal Closed Category (SMCC).
+
 [Section III.A, Weighted], immediately gives:
@@
 \begin{fact}\label{fact:LREL_SMCC}
- $\LREL$ is a Symmetric Monoidal Closed Category.
+ $\LREL$ is a SMCC, thus a model of the linear $\STLC$.
 \end{fact}
 
-Its monoidal structures is given by a tensor product $\otimes$ acting on the objects as the Cartesian product of sets, and as the Kronecker product of matrices on morphisms.
-Its closed structures is given by $X\multimap Y:=X\times Y$ with the usual evaluation and curry morphisms yielding a symmetric monoidal adjunction: $\LREL(Z\otimes X, Y) \simeq \LREL (Z, X\multimap Y)$.
-
-Fact~\ref{fact:LREL_SMCC} means that $\LREL$ is a model of Intuitionistic Multiplicative Linear Logic (IMLL), and thus we can interpret \emph{linear} $\lam$-terms in it.
+Its monoidal structures is given by a tensor product $\otimes$ acting on the objects as the Cartesian product of sets, and as the \emph{Kronecker product} {(\color{red} Ref?)} of matrices on morphisms.
+Its closed structures is given by $X\multimap Y:=X\times Y$ with the usual evaluation and curry maps.
+Actually, the SMCC $\LREL$ is even a model of IMLL.

### 2023-01-16 section3B.tex
Piccolo committino

@@
 
-{\color{red} Point to Sec VI}
+The fact that ordinary $\lam$-terms become only descrete matrices may seem a restrictive property.
+In Section~\ref{sec:app} we will however discuss how we can use this model in order to make it keep track of quantitative effects of programs.
 
@@
 
+\begin{remark}\label{rmk:ModelsOfBSTLC}
 It is known {\color{red}(([reference??] e dire meglio)} that the maps $X\mapsto \C M_{\leq n}(X)$ ($n\in\BB N$) lift to functors $!_n:\LREL\to\LREL$.
@@
 $\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$.
-
+\end{remark}
 %Bounded types are interesting because of the following proposition:

### 2023-01-16 section4.tex
Piccolo committino

@@
 
-\begin{theorem}
+\begin{theorem}\label{thmTLSlocLip}
  All tropical Laurent series $\Lawv^X\to\Lawv$ are locally Lipschitz on $\BB R_{>0}$.
@@
 \end{corollary} 
+\begin{proof}
+\begin{enumerate}
+\item It suffices to remark that if $A$ is a type of the $\BSTLC$, then $\model A$ is a finite set, by definition of the interpretation (Remark~\ref{rmk:ModelsOfBSTLC}).
+Therefore a morphism $t:\model A\to \model B$ is a finite matrix $t: \model A \times \model B \to \Lawv$.
+Hence, its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is a (tropical) polynomial, which is a Lipschitz functions.
+\item Trivial from Theorem~\ref{thmTLSlocLip}.
+\end{enumerate}
+\end{proof}
 

### 2023-01-16 section3B.tex
Piccolo commit pre-pappa

@@
 is a Lafont-exponential.
+
+{\color{red} Recall coKleisli}
 In particular, [Theorem III.7, Weighted] says that:
@@
 
-\begin{remark}[Tropical Laurent series]
- As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ operates as a linear map $\widehat t:\Lawv^{!X}\to\Lawv^Y$.
- But we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_!
-x$ (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
- This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
- Concretely, we have:
- \[t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}\] where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
- Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
-\end{remark}
-
-\begin{remark}
- Indentifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tropical Laurent series generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf\limits_{n\in\N}\set{nx+\widehat f(n)}$, for some $\widehat f:\N\to\Lawv$.
-Remark that we find usual \emph{tropical polynomials} of tropical geometry as a particular case: they corrispond to the ones for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}.
-\end{remark}
-
 \begin{proposition}\label{prop:descrete}
@@
 
-\begin{example}\label{ex:famous_ex}
- The function $f:\Lawv\to\Lawv$, $f(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ is a tropical Laurent series: it is of shape $f=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
-By Proposition~\ref{prop:descrete}, $f$ is not the interpretation of a $\lam$-term, because its matrix is not descrete. Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
-\end{example}
-
 \begin{remark}\label{rmk:ModelsOfBSTLC}
@@
 Therefore, $\LREL$, together with $(!_n)_{n\in\N}$, is a model for $\BSTLC$. 
-Concretely, a type %$A::= o \mid !_{n}A \multimap A$
-{\color{red}(i tipi meglio dirli in Section II)} is in interpreted as $\model o=\B 1$ and 
+Concretely, the types of this calculus (Equation~\ref{eq:Btyped}) are interpreted as: $\model o=\B 1$ and 
 $\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$.
+Remark that such types are thus interpreted as finite sets.
 \end{remark}

### 2023-01-16 section4.tex
Piccolo commit pre-pappa

@@
 It is easy to see that this is actually always the case:
+
+\begin{remark}[Tropical Laurent series]
+ As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ operates as a linear map $\widehat t:\Lawv^{!X}\to\Lawv^Y$.
+ But we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_!
+x$ (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
+ This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
+ Concretely, we have:
+ \[t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}\] where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
+ Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
+\end{remark}
+
+\begin{remark}
+ Indentifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tropical Laurent series generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf\limits_{n\in\N}\set{nx+\widehat f(n)}$, for some $\widehat f:\N\to\Lawv$.
+Remark that we find usual \emph{tropical polynomials} of tropical geometry as a particular case: they corrispond to the ones for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}.
+\end{remark}
+
+\begin{example}\label{ex:famous_ex}
+ The function $f:\Lawv\to\Lawv$, $f(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ is a tropical Laurent series: it is of shape $f=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
+By Proposition~\ref{prop:descrete}, $f$ is not the interpretation of a $\lam$-term, because its matrix is not descrete. Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
+\end{example}
 

### 2023-01-16 section3B.tex
Parte su Lafont e coKleisli

@@
 
-If in the previous subsection we considered \emph{bounded} multisets via the graded comonad $(!_n)_{n\in\N}$, it is immediate to see that \emph{unbounded} multisets yield a comonad able to interpret the $\STLC$:
-following [Corollary III.6, Weighted] we have:
+In order to interpret the $\STLC$, we need a Cartesia closed category (CCC).
+It is well known ({\color{red}Ref?}) that it is always possible to construct a CCC by taking the \emph{co-Kleisli} $\C C_!$ of a so-called \emph{Lafont category} $\C C$, where $!$ is its \emph{Lafont exponential} comonad.
+Let us quickly recall the ideas behind these notions.
+
+A SMCC is Lafont when it has finite products and it is equipped with a comonad $!$ (its \emph{Lafont exponential}) which, at level of objects, sends $X$ to an object $!X$ being the free commutative comonoid on $X$.
+Such object $!X$ represents the \emph{bang} connective of linear logic:
+the infinite duplications granted by $!X$ are provided by the infinite product $X^0\otimes X\otimes X^2\otimes X^3\otimes\cdots$, each factor representing a possible number of duplications.
+Is well known that, under mild conditions satisfied by $\QREL$, one can explicit this idea via the fact that the map $X\mapsto \finMS{X}$ lifts to a functor $!:\QREL\to\QREL$ which is a Lafont-exponential comonad.
+In our case, specialising [Corollary III.6, Weighted], we have:
 
@@
 
-Therefore, it is well known that the map $X\mapsto \finMS{X}$ lifts to a functor $!:\LREL\to\LREL$ which %satisfies the adjunction $\LREL(!(Z\& X),Y) \simeq \LREL(!Z, !X\multimap Y)$, at 
-is a Lafont-exponential.
+Let us recall that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
+We will explicit this constructions in our tropical setting in the next lines.
 
-{\color{red} Recall coKleisli}
-In particular, [Theorem III.7, Weighted] says that:
+As already mentioned, one obtains a CCC from a Lafont category by defining the exponential $X\to Y:=!X \multimap Y$.
+In our case, specialising [Theorem III.7, Weighted] we have indeed:
 \begin{fact}
@@
 
-It is instructive at this point to see what its CCC-structure looks like in our tropical world.
-In particular, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
+%It is instructive what its CCC-structure looks like in our tropical world.
+In our tropical setting, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
 \[
@@
 Remember that in $\Lawv$ the neutral element for addition is $\infty$ and the neutral for multiplication is $0$, so for instance the evaluation map is the matrix $\RM{eval}^{X,Y}\in\Lawv^{!((X\multimap Y) + X)\times Y}\simeq\Lawv^{(!!X\times !Y\times !X)\times Y}$ given by $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=0$ if $\rho_1=[\mu]$ and $\rho_2=[b]$, and $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=\infty$ otherwise.
+
+As it is well known, the Cartesian closed structure %of a category $\C C$ 
+allows to define a sound interpretation $\model{\Gamma\vdash M:A}\in\HOM{}{\model \Gamma}{\model A}$ of terms as morphisms.
+In our case, we have:
 

### 2023-01-16 section3intro.tex
Commit sul linear lam-calc
Parte su Lafont e coKleisli

--- 2023-01-10 section3intro.tex
@@
 \begin{remark}
- The category $\QREL$ is (equivalent to) a (full) subcategory of the category $Q\SF{Mod}$ of $Q$-modules.
+ The category $\QREL$ is (equivalent to) a subcategory of the category $Q\SF{Mod}$ of \emph{complete} $Q$-semimodules.
  If $\QREL$ corresponds to considering semimodules (the $Q^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $Q\SF{Mod}$ corresponds to considering semimodules in abstract, without fixing a base.
@@
 
-In the following, we always consider the category $\LREL$, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,\cdot,0)$, which is one of the so-called \emph{tropical semirings}.
-The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring w.r.t.\ its quantale order $\preceq$, which is the reversed order $\geq$ of the usual order $\leq$ on $\BB R_{\geq0}\cup\set{\infty}$.
-% Qui dobbiamo aver gia detto che un quantale puo' essere dato come idempotent complete semiring o come complete lattice con un compatible monoid.
+In the following, we always consider the category $\LREL$, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
+The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring (w.r.t.\ its quantale order $\preceq$.
 This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
@@
 
-In the remainder of this section we will see how tropical linear algebra -- expressed as $\LREL$ -- gives rise to denotational models of computation -- expressed as \lamcalc.
+In the next sections we will see how $\LREL$ gives rise to denotational models of variants of the $\lam$-calculus.

### 2023-01-16 section3C.tex
Ho rimesso il differenziale nella sua forma originale

@@
 \begin{definition}
- Let $\widetilde{(\cdot)}:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$, $\widetilde t_{\mu\oplus\rho,b}:= t_{\mu*\rho,b}$, and $d:\HOM{\LREL}{!(X+X)}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$, $(dt)_{\mu\oplus\rho,b}:=t_{\rho\oplus\mu,b}$ if $\#\mu_1=1$ and $(dt)_{\mu\oplus\rho,b}:=\infty$ otherwise.
- The \emph{tropical differential operator} is the map $D:=d\circ\widetilde{(\cdot)}:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$.
+ The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho*\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise.
 \end{definition}
 
-\begin{example}
+\begin{remark}
 For $t\in\HOM{\LREL}{!X}{Y}$, we have:
- $(Dt)_{\mu\oplus\rho,b}=t_{\rho*\mu,b}$ if $\#\mu_1=1$ and it is $\infty$ otherwise.
- Analogously, we have:
  $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu*\nu'*\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho*\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.
-\end{example}
+\end{remark}
 

### 2023-01-16 main.tex
commit
Commit temporaneo
Ho cominciato a scrivere la storia delle log-proba, ma non ho finito e per ora l'ho scritta così come viene.
Proposal commit
Piccolo committino
Piccolo commit pre-pappa
Parte su Lafont e coKleisli

@@
 \usepackage{graphicx}
+\usepackage{hyperref}
 \usepackage{tikz-cd}
@@
 
-\section{A Bridge Between Two Quantitative Approaches to the $\lambda$-calculus}
+\section{A Bridge Between Two Quantitative Approaches to the $\lambda$-calculus}\label{section2}
 \input{section2}
 
-\section{Tropical (non-)Linear Algebra and computation}
+%\section{Tropical (non-)Linear Algebra and computation}
+\subsection{Tropical Linear Algebra% and computation
+}
 
@@
 
-\subsection{As a model of linear programs}
+%\subsection{As a model of the \emph{linear} $\STLC$}
+\section{Tropical models of several $\lam$-calculi}\label{section3}
+\subsection{Prohibited duplication/erasure: the \emph{linear} $\STLC$}
 \input{section3A}
 
-\subsection{As a model of non-linear programs}
+%\subsection{As a model of the $\STLC$}
+\subsection{Allowed duplication/erasure: the $\STLC$}
 \input{section3B}%
 
-\subsection{As a model of differential programs}
+%\subsection{As a model of the $\STDLC$}
+\subsection{Controlled duplication/erasure: the $\STDLC$}
 \input{section3C}%
 
-\section{Tropical Laurent series}
+\section{Tropical Laurent series}\label{section4}
 \input{section4}%
 
-\section{Applications}
+\section{Applications}\label{sec:app}\label{section5}
 \input{section5}%
 
-\section{Generalized Metric Spaces and Modules over the Lawvere Quantale}\label{sec:QMod}
+\section{Generalized Metric Spaces and Modules over the Lawvere Quantale}\label{section6}
 \input{section6}%
 
-\section{Related Work}
+\section{Related Work}\label{section7}
+\input{section7}
+
+\section{Conclusion and future work}\label{section8}
 \input{section8}
 
-\section{Conclusion and future work}
-\input{section9}
+\bibliographystyle{plain}
+\bibliography{tropical.bib}
 

### 2023-01-16 section1.tex
commit

@@
 We can mention, for example,
-intersection type systems aimed at capturing time or space resources \cite{Beniamino}, or convergence probabilities \cite{UgoBreuvart, LICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthes} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for
-probabilistic \cite{PCOH, QBS} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
+intersection type systems aimed at capturing time or space resources \cite{decarvalho2018, Accattoli2022}, or convergence probabilities \cite{Breuvart2018, PistoneLICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthe_2012} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for
+probabilistic \cite{Ehrhard2011, Staton2017} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
 The main reason to look for methods relying on (quantitative extensions of) type-theory or denotational semantics is that these approaches yield \emph{modular} and \emph{compositional} techniques, that is, allow one to deduce properties of complex algorithms from the properties of their component parts.
@@
 
-On the one hand there is the approach of \emph{program metrics} \cite{} and \emph{quantitative equational theories} \cite{Mardare}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}.
-This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{} and coinductive methods \cite{}, and was recently extended to account for the the full $\lambda$-calculus \cite{LICS2021, FSCD2022}.
+On the one hand there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}.
+This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
 
-On the other hand there is the approach based on \emph{differential} or \emph{resource-aware} extensions of the $\lambda$-calculus, which is well-connected with the \emph{relational semantics} \cite{} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{}.
-This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{}, and have also been shown suitable to capture probabilistic properties \cite{Manzo}.
+On the other hand there is the approach based on \emph{differential} or \emph{resource-aware} extensions of the $\lambda$-calculus, which is well-connected with the \emph{relational semantics} \cite{Manzo2010, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018}.
+This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable to capture probabilistic properties \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
 
-In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{linearlogic} (i.e.~of using its input exactly once), plays a crucial role.
-In the metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gabo2017}.
+In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using its input exactly once), plays a crucial role.
+In the metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
 By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
@@
 
-Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon (whose origin is, as far as we know, the only reason for the ``tropical'' adjective) as a way to do algebra and geometry by replacing the usual ring structure of numbers based on addition and multiplication by the semi-ring structure given by ``$\min$'' and ``$+$''.
+Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} (whose origin stands behind the adjective ``tropical'') as a way to do algebra and geometry by replacing the usual ring structure of numbers based on addition and multiplication by the semi-ring structure given by ``$\min$'' and ``$+$''.
 %
@@
 The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials, or studying geometric invariants like e.g.~the genus).
-For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning \cite{}.
+For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).
 
 Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for, since polynomials become Lipschitz functions.
-At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail.
-Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories, have also been observed \cite{}.
+At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail.
+Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}, have also been observed \cite{Fuji}.
 
@@
 \emph{bounded} simply typed terms correspond to tropical polynomials, while general
- simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{}.
+ simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{Porzio2021}.
 Moreover, we show that such maps are concave and locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
@@
 \item We conclude by putting the correspondence between differential and metric properties made possible by the tropical viewpoint at the right level of generality. 
-By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{} and modules over the tropical semi-ring \cite{}, we show that the category of \emph{complete}  GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
+By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007}, we show that the category of \emph{complete}  GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
 \end{itemize}

### 2023-01-16 section2.tex
Proposal commit
Piccolo commit pre-pappa
commit

@@
 As recalled in the Introduction, in many situations (typically, when dealing with computationally expensive function) one does not look for algorithms to compute some function \emph{exactly}, but rather to approximate it (e.g.~in an efficient way) within some error bound.
-In other common situations (typically, in differential privacy \cite{}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, that a small error in the input will produce a comparably small error in the output.
+In other common situations (typically, in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, that a small error in the input will produce a comparably small error in the output.
 
@@
 
-As observed in \cite{Gabo2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications.
+As observed in \cite{Reed2010, Gaboardi2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications.
 The fundamental intuition is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz-continuous} map.
@@
 
-These observations have led to the study of \emph{bounded} typed $\lambda$-calculi (like the system $\mathsf{Fuzz}$ \cite{Reed_2010}), which have been applied in the study of higher-order differential privacy \cite{}.
-The types of such systems are defined \cite{} by combining linear constructors with a \emph{graded exponential comonad} $!_{r}(-)$ \cite{}.
+These observations have led to the study of \emph{bounded} typed $\lambda$-calculi (like the system $\mathsf{Fuzz}$ \cite{Reed2010}), which have been applied in the study of higher-order differential privacy \cite{Gaboardi2013, Gaboardi2017}.
+The types of such systems are defined by combining linear constructors with a \emph{graded exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
 In the following sections we will sometimes make reference to a basic type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined as below
-\begin{align}
+\begin{align}\label{eq:Btyped}
 A::= o \  \mid \  (!_{\leq n}A) \multimap A  
@@
 k\cdot x: \BB R\to \BB R$ are all Lipschitz-continuous, with Lipschitz constant $k$, the function $M=\lambda x.x^{2}$ obtained by ``duplicating'' $x$ is not Lipschitz anymore: $M$ is, so to say, \emph{too} sensitive to errors.
-More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of $\STLC$ (yet, $\Met$ contains several cartesian closed \emph{sub-}categories, see e.g.~\cite{}).
+More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of $\STLC$ (yet, $\Met$ contains several cartesian closed \emph{sub-}categories, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
 Still, one might observe that the program $M$ above is actually Lipschitz-continuous, if not globally, at least \emph{locally} (i.e.~ over any compact set).
 However, constructing cartesian closed categories of locally Lipschitz maps is challenging.
-Yet, some examples exist in the literature (see \cite{EhrhardPCOHdiff, PistoneLICS2021}), and a new example will be exhibited in this paper.
+Yet, some examples exist in the literature (see \cite{Ehrhard2011, PistoneLICS}), and a new example will be exhibited in this paper.
 
@@
 
-A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} (and differential linear logic) and its categorical models. 
+A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic, see e.g.~\cite{dill}) and its categorical models.
 The key ingredient is that a \emph{differential operator} $\Der$ is added to the usual syntax of the $\lambda$-calculus, with the intuition that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M]\cdot N$, which still has type $A\to B$, corresponds to a \emph{linear application} of $N$ to $M$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M]\cdot N$ has type $A\to B$, since $M$ might still ask for other copies of an input of type $A$).
 
-Interestingly, the categorical study of such differential operators has led to introduce a family of categories, called \emph{cartesian differential categories} $C\partial C$ \cite{}, which provide an abstract axiomatization of differentiation, in the usual mathematical sense.
+Interestingly, the categorical study of such differential operators has led to introduce a family of categories, called \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, which provide an abstract axiomatization of differentiation, in the usual mathematical sense.
 More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
@@
 
-Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with the differential operator $\Der[-]\cdot (-)$, together with a notion of \emph{linear substitution} (see \cite{} for details).
+Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with the differential operator $\Der[-]\cdot (-)$, together with a notion of \emph{linear substitution} (see \cite{difflambda} for details).
 The models of $\STDLC$ are the 
 cartesian \emph{closed} differential categories ($CC\partial C$), 
-which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential are both well-behaved with respect to the closed structure (see \cite{}).
+which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential are both well-behaved with respect to the closed structure (see \cite{Manzo2010}).
 
@@
 
-\section{Tropical mathematics: a possible synthesis}
+\section{Tropical mathematics%: a possible synthesis
+}
 
-At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{}.
+\subsection{Tropical numbers}
+
+At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}.
 Let us recall that a quantale is a complete lattice endowed with a continuous monoid action.
@@
 \end{align}
-Such functions, called \emph{tropical Laurent series} \cite{}, will be discussed in more detail in Section \ref{section5}.
+Such functions, called \emph{tropical Laurent series} \cite{Porzio2021}, will be discussed in more detail in Section \ref{section5}.
 
@@
 $\alpha \prodt{t}\beta:= \phi_{t}( \psi_{t}(\alpha)\psi_{t}(\beta))=
-\log_{t}(t^{-\alpha}t^{-\beta})$ converge respectively to $\min\{\alpha,\beta\}$ and $\alpha+\beta$, when $t\to 0$ (this remark is an example of the so-called \emph{Maslov dequantization}, relating the semi-ring $\BB R$ with the tropical semiring $((-\infty, \infty], \min, +)$, see e.g.~\cite{}).
+\log_{t}(t^{-\alpha}t^{-\beta})$ converge respectively to $\min\{\alpha,\beta\}$ and $\alpha+\beta$, when $t\to 0$ (this remark is an example of the so-called \emph{Maslov dequantization}, relating the semi-ring $\BB R$ with the tropical semiring $((-\infty, \infty], \min, +)$, see e.g.~\cite{Litvinov2007}).
 

### 2023-01-16 section3B.tex
commit

@@
 \begin{proposition}\label{prop:descrete}
- The interpretation $\model M$ of a $\lam$-term $M$ in $\LREL$ is a \emph{descrete} matrix, i.e.\ its coefficients are either $0$ or $\infty$.
+ The interpretation $\model M$ of a $\lam$-term $M$ in $\LREL$ is a \emph{discrete} matrix, i.e.\ its coefficients are either $0$ or $\infty$.
 \end{proposition}

### 2023-01-16 section3C.tex
commit

@@
 
-A model of the $\STDLC$ is usually understood as so-called \emph{Cartesian closed differential categories} (CC$\partial$C), see [Giulio,?] for details.
+A model of the $\STDLC$ is usually understood as so-called \emph{Cartesian closed differential categories} (CC$\partial$C), see \cite{Manzo2010} for details.
 In order to treat the $+$ and the constructor $D[\_]\cdot (\_)$ of $\STDLC$, the main features of a CC$\partial$C $\C C$ are that:
@@
 There is more: not only $\LREL_!$ is a model of the $\STDLC$, but this model is also well-behaved w.r.t.\ to the \lamcalc Taylor expansion.
-This is expressed by the following property (see [Definition 4.22, Giulio], where it is called the fact of ``modeling the Taylor expansion'').
+This is expressed by the following property (see [Definition 4.22, \cite{Manzo2010}], where it is called the fact of ``modeling the Taylor expansion'').
 
@@
 
-This property is interesting because of [Theorem 4.23,Giulio], which says that, then, in the model the series corresponding to the Taylor expansion of an ordinary $\lam$-term $M$ converges to $M$.
+This property is interesting because of [Theorem 4.23, \cite{Manzo2010}], which says that, then, in the model the series corresponding to the Taylor expansion of an ordinary $\lam$-term $M$ converges to $M$.

### 2023-01-16 section3intro.tex
commit

@@
 For a fixed \emph{continuous} semi-ring $Q$ [Def.
-II.5, Weighted], the well-known \emph{quantitative relational model}, is the category $\QREL$ of sets and set-indexed matrices with coefficients in $Q$ ([Weighted] would call it $Q^\Pi$).
-We must fix a convention for matrices: following [Weighted, MonTop, Ehrhard], we fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum\limits_{b\in Y} s_{b,c}t_{a,b}$.
+II.5, \cite{Manzo2013}], the well-known \emph{quantitative relational model}, is the category $\QREL$ of sets and set-indexed matrices with coefficients in $Q$ (\cite{Manzo2013} would call it $Q^\Pi$).
+We must fix a convention for matrices: following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum\limits_{b\in Y} s_{b,c}t_{a,b}$.
 In linear algebra, a map $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns.
@@
  If $\QREL$ corresponds to considering semimodules (the $Q^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $Q\SF{Mod}$ corresponds to considering semimodules in abstract, without fixing a base.
- We take this viewpoint in Section~\ref{sec:QMod}.
+ We take this viewpoint in Section~\ref{section6}.
 \end{remark}
 
-The assumption of $Q$ being continuous (which can be weakened, see [??]) is used in order to make the potentially infinite sum, appearing in the formula for the composition, converge.
+The assumption of $Q$ being continuous (which can be weakened, see \cite{???}) is used in order to make the potentially infinite sum, appearing in the formula for the composition, converge.
 
@@
 
-Since $\Lawv$ is a continuous (commutative) semiring, [Proposition III.3, Weighted] immediately applies and gives:
+Since $\Lawv$ is a continuous (commutative) semiring, [Proposition III.3, \cite{Manzo2013}] immediately applies and gives:
 
@@
 
-Unwrapping [Definition II.9, Weighted], this means that:
+Unwrapping [Definition II.9, \cite{Manzo2013}], this means that:
 $\HOM{\LREL}{X}{Y}$ is a continuous $\Lawv$-semimodule, with semimodule operations defined pointwise;

### 2023-01-16 section5.tex
commit

@@
 investigate what quantitative properties of $M$ can be deduced from properties of its tropicalization $\trop M$ (which are often easier to study).
-By the way, this is the approach taken in several well-studied applications of tropical geometry (e.g.~for optimization problems related to machine learning \cite{}).
+By the way, this is the approach taken in several well-studied applications of tropical geometry (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}).
 
@@
 
-The possibility of using the relational model over the tropical semi-ring for ``best case'' resource analysis has already been explored in \cite{Manzo}.
+The possibility of using the relational model over the tropical semi-ring for ``best case'' resource analysis has already been explored in \cite{Manzo2013}.
 Notably, they considered an interpretation of a language for $\B{PCF}$ with non-deterministic choice in which each $\lambda$-abstraction and each occurrence of the fixpoint operator $Y$ is assigned a ``weight'' 1, and showed that for any program $M$ of type $\B{nat}$,
@@
 and similarly for $\TT C[N]$.
-It is here worth recalling that, for $\STDLC$, a well-known result \cite{} is that the Taylor expansion of a closed application $MN$ is always finite, although its non-zero coefficients may be arbitrarily high.
-Notably, these observations suggest to study tropical versions of \emph{finiteness spaces} \cite{}, 
+It is here worth recalling that, for $\STDLC$, a well-known result \cite{difflambda} is that the Taylor expansion of a closed application $MN$ is always finite, although its non-zero coefficients may be arbitrarily high.
+Notably, these observations suggest to study tropical versions of \emph{finiteness spaces} \cite{Ehrhard2005}, 
 a variant of the relational semantics modeling strongly normalizing programs via \emph{finite} power series.
@@
 \begin{fact}
- By using [Weighted], it can be shown that $\LREL_!$ is a model of such calculus $\STLC_\oplus$. {\color{red}GIUSTIFICARLO BENE ?}
+ By using \cite{Manzo2013}, it can be shown that $\LREL_!$ is a model of such calculus $\STLC_\oplus$. {\color{red}GIUSTIFICARLO BENE ?}
 \end{fact}
@@
 
-More formally, differential privacy is defined as follows \cite{}:
+More formally, differential privacy is defined as follows \cite{Reed2010}:
 \begin{definition}
@@
 
-A well-studied approach to ensure differential privacy for higher-order programs is to use bounded type systems like $\mathsf{Fuzz}$ \cite{}.
+A well-studied approach to ensure differential privacy for higher-order programs is to use bounded type systems like $\mathsf{Fuzz}$ \cite{Reed2010}.
 Indeed, such systems come with an \emph{a priori} warrant that well-typed programs are Lipschitz.
@@
 The use of tropical semantics suggests how resource analysis could also be used to provide bounds for differential privacy. 
-Suppose our probabilistic query $f$ can be expressed as a power series (this is what happens e.g.~in \emph{probabilistic coherent spaces} \cite{}).
+Suppose our probabilistic query $f$ can be expressed as a power series (this is what happens e.g.~in \emph{probabilistic coherent spaces} \cite{Ehrhard2011}).
 Then, if we discover, either by studying differential properties of $f$, or using methods from convex analysis as suggested in the previous section (e.g.~Theorem \ref{??}),

### 2023-01-16 section6.tex
commit

@@
 
-An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{}.
-This correspondence, recalled in this section, and the fact that it leads to the definition of a linear $\Lawv$-category (i.e.~a model of the linear $\lambda$-calculus) of genealized metric spaces extending $\LREL$, is mostly a \emph{collage} of  \emph{folklore} results with more abstract results contained in recent literature in enriched category theory \cite{}.
+An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}.
+This correspondence, recalled in this section, and the fact that it leads to the definition of a linear $\Lawv$-category (i.e.~a model of the linear $\lambda$-calculus) of genealized metric spaces extending $\LREL$, is mostly a \emph{collage} of  \emph{folklore} results with more abstract results contained in recent literature in enriched category theory \cite{Fuji, Stubbe2006, Shen2014}.
 Beyond this reconstruction, our main contribution is to show that this correspondence can be lifted to a model of the full differential $\lambda$-calculus, by a suitable generalization of the construction of the exponential comonad of $\LREL$.
@@
 The hom-sets $\Hom(M,N)$ have a natural $\Lawv$-module structure, defined pointwise.
-Moreover, the tensor product of $\Lawv$-modules $M\otimes N$ \cite{} can be defined as the quotient of the usual tensor product of sup-lattices (see e.g.~\cite{}), under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$.
+Moreover, the tensor product of $\Lawv$-modules $M\otimes N$ can be defined as the quotient of the usual tensor product of sup-lattices (see e.g.~\cite{Russo2007}), under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$.
 Notably, any element of $M\otimes N$ can be identified with a join of basic tensors $x\otimes y$, corresponding to the equivalence class of the pair $\langle x,y\rangle$.
@@
 \end{align}
-This structure is often referred as a \emph{generalized metric space} \cite{}, or as a \emph{quasi-metric space} \cite{}. 
+This structure is often referred as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}, or as a \emph{quasi-metric space}. 
 Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, i.e.~it must satisfy $Y(f(x),f(y))\leq X(x,y)$.
 
-A $\Lawv$-category is \emph{skeletal} \cite{} when 
+A $\Lawv$-category is \emph{skeletal} \cite{Stubbe2014} when 
 $X(x,y)=0$ implies $x=y$, and 
@@
 Lawvere also observed that Cauchy-completeness can be expressed in categorical language as the representability in $X$ of certain presheaves in $[X^{\op},\Lawv]$ \cite{Lawvere1973}.
-Category theory suggests a yet stronger notion of completeness, corresponding to the existence of all \emph{weighted colimits} (for a comparison of different notions of completeness on $\Lawv$-categories, see \cite{}).
+Category theory suggests a yet stronger notion of completeness, corresponding to the existence of all \emph{weighted colimits} (for a comparison of different notions of completeness on $\Lawv$-categories, see \cite{Willerton2013, Rutten1998}).
 
@@
 
-First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies, Weight}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting the fact that tensors and products commute in $\Mod$, we define $!$ as the product of the symmetric algebras $!_{n}$.
+First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies2018, Manzo2013}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting the fact that tensors and products commute in $\Mod$, we define $!$ as the product of the symmetric algebras $!_{n}$.
 
@@
 
-Using the fact that $\prod_{i}M_{i}\otimes  N\simeq (\prod_{i}M_{i})\otimes N$ holds in $\Mod$ [\cite{}, p.~???], we obtain the following:
+Using the fact that $\prod_{i}M_{i}\otimes  N\simeq (\prod_{i}M_{i})\otimes N$ holds in $\Mod$ (see e.g.~\cite{Russo2007}), we obtain the following:
 \begin{theorem}
@@
 where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
-and $S: !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{}, 
+and $S: !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{Mellies2018}, 
 satisfying all axioms (D1)-(D7) and (DCurry), see the Appendix.

### 2023-01-16 section7.tex
commit

--- 2023-01-08 section7.tex
@@
-- correspondence between $\Lawv$-modules and $\Lawv$-categories.
+The connections between the differential $\lambda$-calculus (and differential linear logic), the relational semantics, and the theory of non-idempotent intersection types is very well-studied (see \cite{decarvalho2018}, and more recently, \cite{Mazza2016} for a more abstract perspective, and \cite{Olimpieri2021, Galal2021} for a 2-categorical, or proof-relevant, extension).
+As we said, the relational semantics over the tropical semi-ring was quickly explored in \cite{Manzo2013}, to provide a ``best case'' resource analysis of a $\B{PCF}$-like language with non-deterministic choice.
+\emph{Probabilistic coherent spaces} \cite{Ehrhard2011}, a variant of  the relational semantics, provide an interpretation of higher-order probabilistic programs
+as analytic functions. In \cite{Ehrhard2022} it was observed that such functions satisfy a local Lipschitz condition somehow reminiscent of our examples in Section \ref{section4}.
 
-- exponential structure of $\LCAT$.
+The study of linear or bounded type systems for sensitivity analysis was initiated in \cite{Girard92tcs} and later developed \cite{Schopp, SchoppDalLago, Reed2010}.
+As recalled in the paper, the use bounded exponentials ensures that well-typed programs satisfy a Lipschitz condition.
+Related approaches, although not based on metrics, are provided by \emph{differential logical relations} \cite{dallago} and \emph{change action} models \cite{Picallo2019}.
 
-- $\LCAT_{!}$ is a cartesian closed differential category
+More generally, the literature on program metrics in denotational semantics is vast.
+Since at least \cite{VANBREUGEL20011} metric spaces, also in Lawvere's generalized sense \cite{Lawvere1973}, have been exploited as an alternative framework to standard, domain-theoretic, denotational semantics.
+While standard categories of metric spaces are not models of the full simply typed $\lambda$-calculus, several constructions of cartesian closed categories of metric spaces can be found in the literature.
+For instance,
+\emph{ultra}-metric spaces form a CCC, and have been shown to model $\B{PCF}$ \cite{Escardo1999}.
+Also \emph{partial} metrics, introduced in \cite{matthews}, have been shown to provide models of $\STLC$, under suitable generalizations \cite{Geoffroy2020}.
+More generally, \cite{Clementino2006} provides a general characterization of exponentiable objects in categories of (generalized) metric spaces, and \cite{PistoneLICS, PistoneFSCD2022} provide other ways to construct CCCs of
+(generalized) metrics, including one based on locally Lipschitz maps, 
+using ideas from {differential logical relations} and quantitative equational theories \cite{Plotk}.
+
+Motivated by connections with computer science and fuzzy set-theory, 
+the abstract study of generalized metric spaces in the framework of \emph{quantale}- or even \emph{quantaloid}-enriched categories has led to a vast literature in recent years \cite{Hofmann2014, Stubbe2014},
+and its connections with tropical mathematics have been explored e.g.~in \cite{Fuji, Willerton2013}.
+Moreover, applications of quantale-modules to both logic and computer science have also been studied \cite{Abramsky1993b, Russo2007}.
+
+Connections between program metrics and the differential $\lambda$-calculus have been already suggested in \cite{PistoneLICS}; moreover, \emph{cartesian difference categories} \cite{Picallo2020} have been proposed as a way to relate derivatives in differential categories with those found in change action models.
+
+Finally, applications of tropical mathematics in computer science abound, originally for automata theory \cite{Chua1992, Simon}, and more recently
+ in optimization methods for machine learning and other statistical models (see e.g.~\cite{Maragos2021, Pachter2004, Zhang2018}), optimization \cite{Akian2011, Akian2012}, and convex analysis \cite{Lucet2009}.
+While our discussion in Section \ref{section5} is inspired by a well-known application of tropical polynomials to Hidden Markov Models \cite{Pachter2004},
+the vast literature in this domain lets us think that other ways to 
+apply tropical semantics to the analysis of higher-order programs might be studied.
+
+%
+%Other connections tropical/metric -> Fuji, ??
+%Applications of tropical to computer science.
+%Log-probabilities.
+%Quantale-modules -> Abramski?
+%
+%

### 2023-01-16 section4.tex
Commit serale

--- 2023-01-16 section4.tex
@@
 \subsection{Results on Tropical Laurent Series}
+
+Recall the definition of \emph{tropical polynomials} and \emph{tropical Laurent series} (of one variable $x\in\Lawv$) given respectively by Equation~\ref{eq:polytrop} and Equation~\ref{eq:defTLS}.
+A $x\in[0,\infty)$ is said to be a (finite) \emph{tropical root} of a tropical Laurent series $f:\Lawv\to\Lawv$ iff $f$ is not differentiable at $x$ (w.r.t.\ the usual topology on $\BB R_{\geq0}$.
+This is equivalent to ask that the $\inf$ defining $f$ at $x$ is obtained \emph{at least} twice.{\color{red} \`e vero anche per TLS o solo per poly questo?
+Inoltre, Robol parla anche di finite end-points of the domain: nel nostro caso sarebbe $x=0$}
+
+\begin{example}\label{ex:famous_ex}
+ The function $f:\Lawv\to\Lawv$ defined by $f(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ is a tropical Laurent series.
+\end{example}
+
+By plotting its graph {\color{red}vogliamo plottarlo?}, we observe several properties that we will lift to the more general case that we will consider in the next lines, and this example will serve as running one.
+
+\begin{remark}%[Tropical Laurent series]
+ As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ operates as a linear map $\widehat t:\Lawv^{!X}\to\Lawv^Y$.
+ But we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_!
+x$ (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
+ This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
+ Concretely, we have:
+ \[t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}\] where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
+ The functions of such shape are generalisations of usual tropical Laurent series: they are the case for infinite variables.
+ We will call them simply \emph{tropical Laurent series (tLs)}.
+ %Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
+\end{remark}
+
+We find the usual notion of tLs of one variable as follows:
+
+\begin{remark}
+ Indentifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tropical Laurent series generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf\limits_{n\in\N}\set{nx+\widehat f(n)}$, for some $\widehat f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
+Remark that we also find usual \emph{tropical polynomials} of tropical geometry as a particular case: they corrispond to the ones for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}.
+Actually, for us \emph{tropical polynomial} will mean a function $f:\Lawv^X\to\Lawv^Y$ of shape $f(x)_b:=\min\limits_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, for a \emph{finite} set $\C F\subseteq \, !X$.
+This is again the generalisation of usual tropical polynomials to the case of infinitely many variables.
+\end{remark}
+
+\begin{remark}
+ The $f$ of Example~\ref{ex:famous_ex} is indeed of shape $f=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
+ By Proposition~\ref{prop:descrete}, $f$ is not the interpretation of a $\lam$-term, because its matrix is not descrete.
+ Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
+\end{remark}
+
+Looking at Example~\ref{ex:famous_ex}, we see that the function is non-decreasing and concave.
+It is easy to see that this is actually always the case:
+
+\begin{proposition}\label{prop:nondecr+conc}
+ Any tLs $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
+\end{proposition}
+
+In Example~\ref{ex:famous_ex}, it can be proven by hand that $f(x)$ is a $\min$ for all $x>0$.
+At $x=0$ we have $f(x=0)=\inf\limits_{n\in\N} \frac{1}{2^n}=0$ which is thus the only point \emph{not} being a $\min$.
+Also, while the derivative of $f$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
+This phenomenon is reminiscent of [Example 7, PCoh],
+%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
+which actually motivated our first investigations.
+In fact, these properties are shared by all tLs with \emph{finitely} many variables, as shown by the following result.
+Remark that $!\set{1,\dots,k}$ can be identified with $\N^k$.
+So the matrix of a tLs $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one variable $f(x)$ in output) can be given as a $\widehat f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf\limits_{n\in \N^k}\set{nx+\widehat f(n)}$, where $nx$ is the scalar product.
+
+\begin{theorem}\label{theorem:fepsilon}
+ Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\widehat f:\N^k\to\Lawv$.
+ For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ s.t.:
+ \begin{enumerate}
+  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
+  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
+  \item On all $[\epsilon,\infty]^k$, $f$ coincides with the tropical \emph{polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\widehat f(n)}$.
+ \end{enumerate}
+\end{theorem}
+\begin{proof}
+ Call $\leq$ the well-founded pointwise order on $\N^k$.
+ Now we can let $\C F_\epsilon:=\set{n\in\N \mid
+ \widehat f(n)<\infty \textit{ and } \widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$.
+\end{proof}
+
 {\color{red}
-
-- recall Trop Laurent Series (TLS) tropical roots
-
-- Example 1
-
-- The morphisms of $\LREL_{!}$ can be seen as generalizations of TLS
-
-- Remark: for $f:\Lawv^{X}\to \Lawv^{Y}$, if $X$ is finite, $f$ is a finite variable TLS; if $f$ is written as a finite $\min$, it is a tropical polynomial.
-
-- Maybe Theorem 7 could even go here??
-Saying something like: after the remark above, we can prove a result that confirms what seen in Example 1, for the case when $X$ is finite: $f$ can LOCALLY be seen as a polynomial, the degree of such polynomials potentially diverging when $\epsilon\to 0$.
-Indeed...
-
 - bisognerebbe anche dire due parole sul differenziale e dell'espansione di Taylor come operazione sulle TLS invece che sulle matrici. 
@@
 
-Let us start by recalling Example~\ref{ex:famous_ex}.
-By plotting its graph {\color{red}vogliamo plottarlo?}, we see first of all that the function is non-decreasing and concave.
-It is easy to see that this is actually always the case:
+\subsection{$\Lawv^{X}$ as a normed cone}
 
-\begin{proposition}\label{prop:nondecr+conc}
- Any tropical Laurent series $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
+$\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
+Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete normed cone.
+The normed cone structure induces an order on it, called its \emph{cone order}, by setting:
+$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
+This order makes it a Scott-continuous dcpo.
+Furthermore we have:
+
+\begin{proposition}
+  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders on the domain and codomain.
 \end{proposition}
-
-\subsection{$\Lawv^{X}$ as a normed space.}
-
-{\color{red}Move here also the cone part}
 
@@
 
-Not only the $f$ of Example~\ref{ex:famous_ex} is continuous, but it is also locally Lipschitz on $\BB R_{>0}$.
-Actually, we can prove the following:
+\subsection{Metric results}%$\Lawv^{X}$ as a metric space.}
 
-\begin{theorem}\label{theorem:fepsilon}
- Let $f:\Lawv\to\Lawv$ a tropical Laurent series with matrix $\widehat f:\N\to\Lawv$.
- For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N$ s.t.:
- \begin{enumerate}
-  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv$;
-  \item If $f(x_0)=\infty$ for some $x_0<\infty$, then $\C F_\epsilon=\emptyset$;
-  \item On all $[\epsilon,\infty]$, $f$ coincides with the tropical \emph{polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\widehat f(n)}$.
- \end{enumerate}
- In particular, $f$ is locally Lipschitz on all $\BB R_{>0}$.
-\end{theorem}
-\begin{proof}
- We can let $\C F_\epsilon:=\set{n\in\N \mid
- \widehat f(n)<\infty \textit{ and } \widehat f(m)> \widehat f(m)+\epsilon \textit{ for all } m<n}$.
-\end{proof}
-
-We believe that a variant of such result holds in $\Lawv^X$ for $X$ finite.
-We do not know what happens if $X$ is infinite.
-
-Remark that, in coherence with the previous result, in Example~\ref{ex:famous_ex} $f(x)$ is indeed a $\min$ for all $x>0$.
-At $x=0$ we have $f(x=0)=\inf\limits_{n\in\N} \frac{1}{2^n}=0$ which is \emph{not} a min.
-Also, while the derivative of $f$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
-This phenomenon is reminiscent of [Example 7, PCoh]
-%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
-, which actually motivated our first investigations.
-
-\subsection{$\Lawv^{X}$ as a metric space.}
-
-The norm $\|-\|_\infty$ naturally induces a metric $\|\B x-\B y\|_{\infty}$ over the spaces $\Lawv^{X}$. 
+The norm $\norm{\cdot}_\infty$ naturally induces a metric $\|\B x-\B y\|_{\infty}$ over the spaces $\Lawv^{X}$. 
 We now study the Lipschitz properties of tropical Laurent series w.r.t.~these metrics.
 
+First of all, remark that Theorem~\ref{theorem:fepsilon} implies that tLs $f$ of finitely many variables are locally Lipschitz (and thus continuous) on all $\BB R_{>0}$.
+In this subsection we will refine and generalise this result.
+
 A first result is the following:
+
 \begin{proposition}\label{prop:troplinear}
@@
 %\end{corollary}
-
-{\color{red}MOVE:
-Let us end this section by mentioning another point of view on tropical Laurent series.
-$\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
-Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete normed cone.
-The normed cone structure induces an order on it, called its \emph{cone order}, by setting:
-$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
-This order makes it a Scott-continuous dcpo.
-Furthermore we have:
-
-\begin{proposition}
-  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders on the domain and codomain.
-\end{proposition}
-}

### 2023-01-17 main.tex
commit

@@
 
-\author{\IEEEauthorblockN{Davide Barbarossa}
-\IEEEauthorblockA{\textit{Universit\`a di Bologna}\\
-Bologna, Italy \\
-davide.barbarossa@unibo.it}
-\and
-\IEEEauthorblockN{Paolo Pistone}
-\IEEEauthorblockA{\textit{Universit\`a Roma Tre}\\
-Rome, Italy\\
-paolo.pistone@uniroma3.it}
-}
+%\author{\IEEEauthorblockN{Davide Barbarossa}
+%\IEEEauthorblockA{\textit{Universit\`a di Bologna}\\
+%Bologna, Italy \\
+%davide.barbarossa@unibo.it}
+%\and
+%\IEEEauthorblockN{Paolo Pistone}
+%\IEEEauthorblockA{\textit{Universit\`a Roma Tre}\\
+%Rome, Italy\\
+%paolo.pistone@uniroma3.it}
+%}
 
@@
 \begin{abstract}
+In this paper we study a quantitative semantics for the $\lambda$-calculus based on tropical mathematics. Notably, we show that the relational model with values over the tropical semiring 
+provides a unified framework for both 
+program metrics, as based on the analysis of program sensitivity via Lipschitz-conditions, and for resource analysis, as based on higher-order program differentiation.
+We discuss applications of this tropical semantics for ``best case'' resource analysis as well as for other quantitative properties like log-probabilities and differential privacy.
+Finally, we show that a general foundation for this semantic approach is provided by an abstract correspondence between Lawvere's theory  
+of generalized metric spaces and tropical algebra.
 
@@
 \begin{IEEEkeywords}
+Differential lambda calculus, Tropical semiring, Lawvere quantale, Program metrics, Relational semantics.
 \end{IEEEkeywords}
@@
 %\section{Tropical (non-)Linear Algebra and computation}
-\subsection{Tropical Linear Algebra% and computation
+\subsection{Tropical linear algebra via the relational model% and computation
 }

### 2023-01-17 section1.tex
commit

@@
 
-On the other hand there is the approach based on \emph{differential} or \emph{resource-aware} extensions of the $\lambda$-calculus, which is well-connected with the \emph{relational semantics} \cite{Manzo2010, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018}.
+On the other hand there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected with the \emph{relational semantics} \cite{Manzo2010, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
 This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable to capture probabilistic properties \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
@@
 At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
-in metric models $n$ times duplication results in a linear (hence Lipschitz) function $n\cdot x$, while in differential models this results in a monomial $x^{n}$, which is not Lipschitz.
+in metric models $n$ times duplication results in a linear (hence Lipschitz) function $n\cdot x$, while in differential models this results in a polynomial $x^{n}$, which is not Lipschitz.
 The fundamental insight of this article is that a natural way to overcome
@@
 It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
-Tropical mathematics is a form of \emph{idempotent} mathematics, since the role of addition is 
-played by the idempotent operation $\min$.
+%Tropical mathematics is a form of \emph{idempotent} mathematics, since the role of addition is 
+%played by the idempotent operation $\min$.
 For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$, when interpreted over the tropical semi-ring, becomes the piecewise linear function
@@
 
-The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials, or studying geometric invariants like e.g.~the genus).
+As working on the idempotent $\min,+$-semiring is often easier that working with addition and multiplication, the tropicalization of  algebraic concepts makes several intractable properties (e.g.~finding roots of polynomials, or studying geometric invariants like e.g.~the genus) become tractable.
 For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).
 
-Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for, since polynomials become Lipschitz functions.
-At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail.
+Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for, as the monomials $x^{n}$ turn precisely into the Lipschitz functions $nx$.
+At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.
 Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}, have also been observed \cite{Fuji}.
@@
  simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{Porzio2021}.
-Moreover, we show that such maps are concave and locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
-Moreover, we exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps.
+Moreover, we show that such maps are locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
+%Moreover, we exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps.
 
-\item Using the relational model as our main reference, we suggest a list of potential applications of tropical methods to the study of quantitative properties of non-deterministic/probabilistic functional programs like differential privacy, convergence \emph{log-probabilities} and the number of computation steps in the best case.
+\item Using the relational model as our main reference, we suggest a few potential applications of tropical methods to the study of quantitative properties of non-deterministic/probabilistic functional programs like counting best-case computation steps,
+measuring convergenge log-probabilities, and 
+differential privacy.
 

### 2023-01-17 section2.tex
Commit serale
commit

@@
 $d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
-A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive provides a model of the \emph{affine} simply typed $\lambda$-calculus, being both a \emph{symmetric monoidal closed} category (i.e.~a model of \emph{linear} $\STLC$) and a cartesian category in which cartesian and monoidal unit coincide.
+A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the {affine} simply typed $\lambda$-calculus, being both a \emph{symmetric monoidal closed} category (i.e.~a model of \emph{linear} $\STLC$) and a cartesian category in which cartesian and monoidal unit coincide.
 
 As observed in \cite{Reed2010, Gaboardi2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications.
-The fundamental intuition is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz-continuous} map.
+The fundamental intuition is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz} map.
 For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the functional input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.
@@
 
-These observations have led to the study of \emph{bounded} typed $\lambda$-calculi (like the system $\mathsf{Fuzz}$ \cite{Reed2010}), which have been applied in the study of higher-order differential privacy \cite{Gaboardi2013, Gaboardi2017}.
+These observations have led to the study of \emph{bounded} typed $\lambda$-calculi, like the system $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied in the study of higher-order differential privacy \cite{Gaboardi2013, Gaboardi2017}.
 The types of such systems are defined by combining linear constructors with a \emph{graded exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
-In the following sections we will sometimes make reference to a basic type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined as below
-\begin{align}\label{eq:Btyped}
-A::= o \  \mid \  (!_{\leq n}A) \multimap A  
-\end{align}
-where $n$ indicates a natural number.
-Intuitively, $!_{\leq n}A\multimap B$ is the type of functions from $A$ to $B$ that may use their input at most $n$ times.
-More details about $\BSTLC$ are provided in the Appendix.
+In the following sections we will sometimes make reference to a basic type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined via $A::= o \  \mid \  (!_{\leq n}A) \multimap A  $,
+where $n\in \BB N$. Intuitively, $!_{\leq n}A\multimap B$ is the type of functions from $A$ to $B$ that may use their input at most $n$ times. More details about $\BSTLC$ are provided in the Appendix.
 
@@
 More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of $\STLC$ (yet, $\Met$ contains several cartesian closed \emph{sub-}categories, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
-Still, one might observe that the program $M$ above is actually Lipschitz-continuous, if not globally, at least \emph{locally} (i.e.~ over any compact set).
-However, constructing cartesian closed categories of locally Lipschitz maps is challenging.
-Yet, some examples exist in the literature (see \cite{Ehrhard2011, PistoneLICS}), and a new example will be exhibited in this paper.
+Still, one might observe that the program $M$ above is actually Lipschitz-continuous, if not globally, at least \emph{locally} (i.e.~over any compact set).
+Indeed, some cartesian closed categories of locally Lipschitz maps have been produced in the literature (see \cite{Ehrhard2011, PistoneLICS}), and a new example will be exhibited in this paper.
 
@@
 
-A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic, see e.g.~\cite{dill}) and its categorical models.
-The key ingredient is that a \emph{differential operator} $\Der$ is added to the usual syntax of the $\lambda$-calculus, with the intuition that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M]\cdot N$, which still has type $A\to B$, corresponds to a \emph{linear application} of $N$ to $M$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M]\cdot N$ has type $A\to B$, since $M$ might still ask for other copies of an input of type $A$).
+A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models.
+The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus.
+The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M]\cdot N$, still of type $A\to B$, corresponds to the \emph{linear application} of $N$ to $M$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M]\cdot N$ has type $A\to B$, since $M$ might still need \emph{other} copies of an input of type $A$).
 
-Interestingly, the categorical study of such differential operators has led to introduce a family of categories, called \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, which provide an abstract axiomatization of differentiation, in the usual mathematical sense.
+Interestingly, the categorical study of these differential operators has led to the introduction of \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, a class of categories providing an abstract axiomatization of differentiation, in the usual mathematical sense.
 More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
@@
 \end{align}
-In other words, terms with unbounded duplications can be seen as some sort of ``limits'' of bounded, but arbitrarily large, duplications.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
+In other words, terms with unbounded duplications can be seen as some sort of limits of bounded, but arbitrarily large, duplications.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
 
@@
 
-At this point, given our previous discussion of bounded duplications and Lipschitz-continuity, a natural question would be whether the Taylor formula
-could be used to decompose an unbounded $\lambda$-term 
-into a sum of Lipschitz maps, thus bridging the metric and differential approach.  
-While suggestive, an obstacle to realize this idea is that,
-in general, a $C\partial C$ needs not satisfy the Taylor formula
-\eqref{eq:taylor}, and in the \emph{relational semantics}, 
-a well-studied and sort of canonical ``Taylor'' semantics for $\STDLC$, the bounded terms appearing in the Taylor expansion correspond to \emph{polynomials}, and thus to non-Lipschitz functions.
+At this point, given our previous discussion of bounded duplications and Lipschitz-continuity, it is natural to ask whether the Taylor formula
+may somehow be seen as a decomposition of a $\lambda$-term 
+into a limit of Lipschitz maps, thus bridging the metric and differential approach.  
+While suggestive, an obstacle to realize this idea is that, as anticipated in the Introduction, 
+%in general, a $C\partial C$ needs not satisfy the Taylor formula
+%\eqref{eq:taylor}, and 
+in the canonical ``Taylor'' semantics for $\STDLC$, the relational semantics, the bounded terms appearing in the Taylor expansion correspond to \emph{polynomials}, and thus to non-Lipschitz functions.
 
@@
 Finally, in Section \ref{section6} we show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between
-\emph{generalized metric spaces} and modules over the tropical semiring.
+{generalized metric spaces} and modules over the tropical semiring.
 
@@
 Let us recall that a quantale is a complete lattice endowed with a continuous monoid action.
-In the case of $\Lawv$ the lattice is defined by the reverse order $\geq$, and the monoid action is provided by addition.
+In the case of $\Lawv$ the lattice is defined by the reverse order $\geq$ on $\BB R$, and the monoid action is provided by addition.
 Notice that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$.
-A consequence of these observations is that, as discussed in Section \ref{section3}, the tropical approach to linear algebra coincides with the study of ``$\Lawv$-valued matrices'', i.e.~of maps of the form $s: X\times Y\to \Lawv$ .
-In particular, a (pseudo-)metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we discuss in Section \ref{section6}).
+A consequence of these observations is that, as we discussed below, the tropical approach to linear algebra coincides with the study of ``$\Lawv$-valued matrices'', i.e.~of maps of the form $s: X\times Y\to \Lawv$ .
+In particular, a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we explicitly take in Section \ref{section6}).
 
@@
 
-More generally, for any function $f:[0,1]\to [0,1]$ which can be written as a (parameterized) \emph{power series} of the form $f_{t}(x)= \sum_{n}t^{c_{n}}x^{n}$ (as we'll see, such functions arise naturally from the interpretation of probabilistic programs, as we'll recall in Section \ref{section5}), we can define its \emph{tropicalization} $\trop f: \Lawv \to \Lawv$
-as the function defined as follows:
-\begin{align}
+More generally, for any function $f:[0,1]\to [0,1]$ which can be written as a (parameterized) \emph{power series} of the form $f_{t}(x)= \sum_{n}t^{c_{n}}x^{n}$ (as we'll see in Section \ref{section5}, such functions arise naturally from the interpretation of probabilistic programs), we can define its \emph{tropicalization} as the function $\trop f: \Lawv \to \Lawv$ defined as follows:
+\begin{align}\label{eq:defTLS}
 \trop f(\alpha) =\inf_{n}\left\{ n\alpha+c_{n}\right\}
@@
 
-Since many algebraic and geometric properties of tropical maps are often simpler and more combinatorial than the corresponding  properties of non-tropical functions, a typical application of tropical mathematics is to study how much can be deduced of some function starting from the properties of its tropicalization.
-In Section \ref{section5} we will follow a similar direction, investigating what quantitative properties of a higher-order programs are revealed by the study of its tropical interpretation.
+%
+%Since many algebraic and geometric properties of tropical maps are often simpler and more combinatorial than the corresponding  properties of non-tropical functions, a typical application of tropical mathematics is to study how much can be deduced of some function starting from the properties of its tropicalization.
+%In Section \ref{section5} we will follow a similar direction, investigating what quantitative properties of a higher-order programs are revealed by the study of its tropical interpretation.
+%
+%
 

### 2023-01-17 section3A.tex
Piccolo commit pre-pappa
commit

--- 2023-01-10 section3A.tex
@@
-[Section III.A, Weighted], immediately gives:
+The linear simply typed $\lam$-calculus is a restriction of the ordinary $\lam$-calculus in which programs can only use their arguments \emph{exactly once}.
+This is the notion of linearity taken into account by linear logic, and indeed this calculus can be embedded in its \emph{intuitionistic multiplicative} fragment \emph{IMLL}.
+
+More precisely, the linear $\lam$-calculus is obtained from the ordinary one by adding the constraint that each $\lam$-abstracted variable appears exactly once in the scope of the $\lam$-abstraction.
+
+In order to frame it in a category $\C C$, one needs a symmetric tensor product $\otimes$ together with internal hom-set objects $X\multimap Y$ s.t.\ the evaluation and curry maps yield a \emph{symmetric monoidal adjunction}: $\C C(Z\otimes X, Y) \simeq \C C (Z, X\multimap Y)$.
+This gives the notion of Symmetric Monoidal Closed Category (SMCC).
+
+[Section III.A, \cite{Manzo2013}], immediately gives:
 
 \begin{fact}\label{fact:LREL_SMCC}
- $\LREL$ is a Symmetric Monoidal Closed Category.
+ $\LREL$ is a SMCC, thus a model of the linear $\STLC$.
 \end{fact}
 
-Its monoidal structures is given by a tensor product $\otimes$ acting on the objects as the Cartesian product of sets, and as the Kronecker product of matrices on morphisms.
-Its closed structures is given by $X\multimap Y:=X\times Y$ with the usual evaluation and curry morphisms yielding a symmetric monoidal adjunction: $\LREL(Z\otimes X, Y) \simeq \LREL (Z, X\multimap Y)$.
-
-Fact~\ref{fact:LREL_SMCC} means that $\LREL$ is a model of Intuitionistic Multiplicative Linear Logic (IMLL), and thus we can interpret \emph{linear} $\lam$-terms in it.
+Its monoidal structure is given by a tensor product $\otimes$ acting on the objects as the Cartesian product of sets, and as the \emph{Kronecker product} {(\color{red} Ref?)} of matrices on morphisms.
+Its closed structures is also given by the Cartesian product on sets, with the usual evaluation and curry maps.
+Actually, the SMCC $\LREL$ is even a model of IMLL.

### 2023-01-17 section3B.tex
commit

@@
 
-In order to interpret the $\STLC$, we need a Cartesia closed category (CCC).
-It is well known ({\color{red}Ref?}) that it is always possible to construct a CCC by taking the \emph{co-Kleisli} $\C C_!$ of a so-called \emph{Lafont category} $\C C$, where $!$ is its \emph{Lafont exponential} comonad.
-Let us quickly recall the ideas behind these notions.
-
+In order to interpret the $\STLC$, we need a Cartesian closed category (CCC).
+It is well known ({\color{red}Ref?}) that it is always possible to construct a CCC by taking the \emph{co-Kleisli} $\C C_!$ of a so-called \emph{Lafont category} $\C C$.
+%, a construction we now quickly recall.
+%defined via a \emph{Lafont exponential} comonad $!$.
+%Let us quickly recall the ideas behind these notions.
 A SMCC is Lafont when it has finite products and it is equipped with a comonad $!$ (its \emph{Lafont exponential}) which, at level of objects, sends $X$ to an object $!X$ being the free commutative comonoid on $X$.
-Such object $!X$ represents the \emph{bang} connective of linear logic:
-the infinite duplications granted by $!X$ are provided by the infinite product $X^0\otimes X\otimes X^2\otimes X^3\otimes\cdots$, each factor representing a possible number of duplications.
-Is well known that, under mild conditions satisfied by $\QREL$, one can explicit this idea via the fact that the map $X\mapsto \finMS{X}$ lifts to a functor $!:\QREL\to\QREL$ which is a Lafont-exponential comonad.
-In our case, specialising [Corollary III.6, Weighted], we have:
+Such object $!X$ represents the \emph{bang} connective of linear logic, granting infinite duplications via the infinite product $X^0\otimes X\otimes X^2\otimes X^3\otimes\cdots$, each factor representing a possible number of duplications.
+It is well known that, under mild conditions satisfied by $\QREL$, one can explicit this idea via the fact that the map $X\mapsto \finMS{X}$ (where $\finMS{X}$ is the set of finite multi-sets on $X$) lifts to a functor $!:\QREL\to\QREL$ which is a Lafont-exponential comonad.
+Specialising [Corollary III.6, \cite{Manzo2013}] to our case, we have:
 
@@
 Let us recall that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
-We will explicit this constructions in our tropical setting in the next lines.
-
-As already mentioned, one obtains a CCC from a Lafont category by defining the exponential $X\to Y:=!X \multimap Y$.
-In our case, specialising [Theorem III.7, Weighted] we have indeed:
+%We will explicit this constructions in our tropical setting in the next lines.
+As already mentioned, one obtains a CCC from a Lafont category by defining the exponential objects as $X\to Y:=!X \multimap Y$.
+In our case, specialising [Theorem III.7, \cite{Manzo2013}] we have indeed:
 \begin{fact}
@@
 %It is instructive what its CCC-structure looks like in our tropical world.
-In our tropical setting, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
-\[
+In our tropical setting, 
+the exponential object $X\multimap Y$ is $!X\times Y$, and 
+the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
+\begin{align}
  \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1*\cdots *\mu_n}
- \set{s_{[b_1,\dots,b_n],c} + \sum\limits_{i=1}^n t_{\mu_i,b_i}}.
-\]
-The exponential object $X\multimap Y$ is $!X+Y$ (where $+$ is the disjoint union).
-Remember that in $\Lawv$ the neutral element for addition is $\infty$ and the neutral for multiplication is $0$, so for instance the evaluation map is the matrix $\RM{eval}^{X,Y}\in\Lawv^{!((X\multimap Y) + X)\times Y}\simeq\Lawv^{(!!X\times !Y\times !X)\times Y}$ given by $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=0$ if $\rho_1=[\mu]$ and $\rho_2=[b]$, and $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=\infty$ otherwise.
+ \left\{s_{[b_1,\dots,b_n],c} + \sum\limits_{i=1}^n t_{\mu_i,b_i}\right\}.
+\end{align}
+% (where $+$ is the disjoint union).
+
+%Remember that in $\Lawv$ the neutral element for addition is $\infty$ and the neutral for multiplication is $0$, so for instance the evaluation map is the matrix $\RM{eval}^{X,Y}\in\Lawv^{!((X\multimap Y) \times X)\times Y}\simeq\Lawv^{(!!X\times !Y\times !X)\times Y}$ given by $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=0$ if $\rho_1=[\mu]$ and $\rho_2=[b]$, and $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=\infty$ otherwise.
 
@@
 \begin{proof}
- First easily prove that composition preserves descreteness.
- Then go by straightforward induction on $M$, using that the projections and evaluation of  $\LREL$ are descrete.
+ First easily prove that composition preserves discreteness.
+ Then go by straightforward induction on $M$, using that the projections and evaluation of  $\LREL$ are discrete.
 \end{proof}
 
-The fact that ordinary $\lam$-terms become only descrete matrices may seem a restrictive property.
-In Section~\ref{sec:app} we will however discuss how we can use this model in order to make it keep track of quantitative effects of programs.
+While ordinary $\lam$-terms yield discrete matrices, 
+as discussed in Section~\ref{sec:app}, the other coefficients from $[0,\infty]$ will appear when considering quantitative effects like computation steps or probabilities.
+%, give rise to  we will however discuss how 
+%we can use this model in order to make it keep track of quantitative effects of programs.
 
 \begin{remark}\label{rmk:ModelsOfBSTLC}
-It is known {\color{red}(([reference??] e dire meglio)} that the maps $X\mapsto \C M_{\leq n}(X)$ ($n\in\BB N$) lift to functors $!_n:\LREL\to\LREL$.
+%Similarly to It is known {\color{red}(([reference??] e dire meglio)} 
+The exponential $!$ can be ``decomposed'' into a family of \emph{graded} exponentials $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined by functors $X\mapsto \C M_{\leq n}(X)$. %  lift to functors 
 The sequence $(!_n)_{n\in\N}$ can be then turned into a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, satisfying the adjunction:
@@
 Therefore, $\LREL$, together with $(!_n)_{n\in\N}$, is a model for $\BSTLC$. 
-Concretely, the types of this calculus (Equation~\ref{eq:Btyped}) are interpreted as: $\model o=\B 1$ and 
+In particular, bounded arrow types are interpreted via
+%Concretely, the types of this calculus (Equation~\ref{eq:Btyped}) are interpreted as: $\model o=\B 1$ and 
 $\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$.
-Remark that such types are thus interpreted as finite sets.
+Notice that, for any type $A$ of $\BSTLC$, its interpretation $\model{A}$ is a finite set.
 \end{remark}

### 2023-01-17 section3C.tex
commit

@@
-Until now we simply specialised well-known results in our tropical case, with the intent of showing how do things read in this particular case.
-Now we go further, by showing that $\LREL_!$ actually admits a \emph{differential structure}, turning it into a model of the $\STDLC$.
-This is where the \emph{metric} and the \emph{differential} viewpoints converge, as explained in the Introduction and Section II, and it will be generalised in Section 6.
+Until now we simply specialised well-known results in our tropical case, with the intent of showing how things read in this particular case.
+Now we go further, by showing that $\LREL_!$ actually admits a \emph{differential structure}, turning it into a model of the $\STDLC$, i.e.~a $CC\partial C$.
+This viewpoint
+%, is where the \emph{metric} and the \emph{differential} viewpoints converge, as explained in the Introduction and Section II, and it 
+will be further generalised in Section \ref{section6}.
 
-A model of the $\STDLC$ is usually understood as so-called \emph{Cartesian closed differential categories} (CC$\partial$C), see \cite{Manzo2010} for details.
-In order to treat the $+$ and the constructor $D[\_]\cdot (\_)$ of $\STDLC$, the main features of a CC$\partial$C $\C C$ are that:
-
-1) $\C C$ is a left-additive-CCC, i.e.\ its Homsets are commutative monoids and its Cartesian closed structure is well behaved w.r.t.\ this monoid structure;
-
-2) $\C C$ is equipped with a differential operator map $D:\HOM{\C C}{X}{Y}\to \HOM{\C C}{X\times X}{Y}$ (here $\times$ is the Cartesian product of $\C C$) satisfying $8$ axioms, called D1, ..., D7, D-curry.
+%A model of the $\STDLC$ is usually understood as so-called \emph{Cartesian closed differential categories} (CC$\partial$C), see \cite{Manzo2010} for details.
+%In order to treat the $+$ and the constructor $D[\_]\cdot (\_)$ of $\STDLC$, the main features of a CC$\partial$C $\C C$ are that:
+%
+%1) $\C C$ is a left-additive-CCC, i.e.\ its Homsets are commutative monoids and its Cartesian closed structure is well behaved w.r.t.\ this monoid structure;
+%
+%2) $\C C$ is equipped with a differential operator map $D:\HOM{\C C}{X}{Y}\to \HOM{\C C}{X\times X}{Y}$ (here $\times$ is the Cartesian product of $\C C$) satisfying $8$ axioms, called D1, ..., D7, D-curry.
 
@@
 \begin{proof}
- The properties 1) are quite straightforward.
- Properties 2) are tedious technical checks.
+ Left-additivity (with respect to $\min$) is quite straightforward.
+ The other axioms of CC$\partial$C's are checked in the Appendix.
 \end{proof}
 
-There is more: not only $\LREL_!$ is a model of the $\STDLC$, but this model is also well-behaved w.r.t.\ to the \lamcalc Taylor expansion.
+There is more: not only $\LREL_!$ is a model of the $\STDLC$, but this model is also well-behaved w.r.t.\ to the Taylor expansion of $\lam$-terms.
 This is expressed by the following property (see [Definition 4.22, \cite{Manzo2010}], where it is called the fact of ``modeling the Taylor expansion'').
@@
 \begin{theorem}\label{thm:modelsTaylor}
- In $\LREL_!$ equipped with $D$, all morphisms can be Taylor expanded, i.e.\ in $\LREL_!$ we have the following equality:
- \[
-  \RM{eval}^{X,Y}\circ_!\langle t,s\rangle =
+ In $\LREL_!$ equipped with $D$, all morphisms can be Taylor expanded, i.e.\ in $\LREL_!$ we have that, for all
+$t\in\HOM{\LREL_!}{Z}{X\multimap Y}$ and $s\in\HOM{\LREL_!}{Z}{X}$, the evaluation of $t$ over $s$ yields 
+ \begin{align}\label{eq:taylorcat}
+  \RM{ev}\circ_!\langle t,s\rangle =
   \inf\limits_{n\in\N}
   \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
- \]
- for all $t\in\HOM{\LREL_!}{Z}{X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$.
+ \end{align}
  \end{theorem}
-Here, $\Lambda^-:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$ is the uncarry operator and $\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is an operator corresponding to the syntactical \emph{differential substitution} of $\STDLC$.
-\begin{proof}
- {\color{red}(Che ci scriviamo ??}
-\end{proof}
+It is worth discussing the formula above a bit more. 
+Here, $\Lambda^-$
+%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
+ is the uncurry operator and 
+%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
+$u\star s= (Du)\circ \langle s,\infty\rangle$ corresponds to linear application of $s$ to $u$.
+Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$,  i.e.~it coincides with the tropical interpretation of the Taylor formula \eqref{eq:taylor}.
+In other words, \eqref{eq:taylorcat} says that the series corresponding to the Taylor expansion of an ordinary $\lam$-term $M$ converges to $M$ [Theorem 4.23, \cite{Manzo2010}].
 
-This property is interesting because of [Theorem 4.23, \cite{Manzo2010}], which says that, then, in the model the series corresponding to the Taylor expansion of an ordinary $\lam$-term $M$ converges to $M$.
+%
+% corresponding to the syntactical \emph{differential substitution} of $\STDLC$.
+%%\begin{proof}
+% {\color{red}(Che ci scriviamo ??}
+%\end{proof}

### 2023-01-17 section3intro.tex
commit

@@
+The study of matrices with values over the tropical semi-ring can be seen as a special case of the
+\emph{quantitative relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus.
+
 For a fixed \emph{continuous} semi-ring $Q$ [Def.
-II.5, \cite{Manzo2013}], the well-known \emph{quantitative relational model}, is the category $\QREL$ of sets and set-indexed matrices with coefficients in $Q$ (\cite{Manzo2013} would call it $Q^\Pi$).
-We must fix a convention for matrices: following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum\limits_{b\in Y} s_{b,c}t_{a,b}$.
+II.5, \cite{Manzo2013}], the category $\QREL$ has sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms (\cite{Manzo2013} would call it $Q^\Pi$).
+The identity morphism of $\QREL$ is the matrix $i:X\times X\to Q$ given by $i_{a,a}=1$ and $i_{a,b\neq a}=0$, and composition of morphisms $t:X\times Y\to Q$ and $s:Y\times Z\to Q$ is given by 
+$(st)_{a,c}:=\sum\limits_{b\in Y} s_{b,c}t_{a,b}$.
+Notice that the assumption of $Q$ being continuous 
+%(which can be weakened, see \cite{???})
+ is used in order to make this potentially infinite sum converge.
+
+ As it is expected, $Q^X$ is a $Q$-semimodule and the bijection $\hat{(\cdot)}$ identifies $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$ via
+ \begin{equation}
+ \hat t(x)_a:=\sum\limits_{b\in Y} t_{a,b}x_a
+ \end{equation}
+
+ \begin{remark}
+ Following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we 
+ chose to see a matrix from $X$ to $Y$ as a map $s:X\times Y\to Q$.
+% 
+% fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum\limits_{b\in Y} s_{b,c}t_{a,b}$.
 In linear algebra, a map $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns.
-In particular, the product of such a matrix for a vector defines a map $Q^Y\to Q^X$.
-Instead, we prefer to see a $t\in\HOM{\QREL}{X}{Y}$ as giving rise to a map $\hat t:Q^X\to Q^Y$ defined by $\hat t(x)_a:=\sum\limits_{b\in Y} t_{a,b}x_a$.
-In $\QREL^{op}$ (which corresponds to systematically taking transpose matrices), composition coincides with the product matrix/matrix and $\hat{(\cdot)}$ with the product matrix/vector.
-In order to avoid confusion, we will refer to a $t\in Q^{X\times Y}$ just as a \emph{matrix from $X$ to $Y$}.
-As it is expected, $Q^X$ is a $Q$-semimodule and the bijection $\hat{(\cdot)}$ identifies $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.
+In particular, the product of such a matrix for a vector defines a map $Q^Y\to Q^X$, hence the opposite of what shown above.
+\end{remark}
+%
+%In $\QREL^{op}$ (which corresponds to systematically taking transpose matrices), composition coincides with the product matrix/matrix and $\hat{(\cdot)}$ with the product matrix/vector.
+%In order to avoid confusion, we will refer to a $t\in Q^{X\times Y}$ just as a \emph{matrix from $X$ to $Y$}.
+%
+%
+%
+%We must fix a convention for matrices: following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum\limits_{b\in Y} s_{b,c}t_{a,b}$.
+%In linear algebra, a map $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns.
+%In particular, the product of such a matrix for a vector defines a map $Q^Y\to Q^X$.
+%Instead, we prefer to see a $t\in\HOM{\QREL}{X}{Y}$ as giving rise to a map $\hat t:Q^X\to Q^Y$ defined by $\hat t(x)_a:=\sum\limits_{b\in Y} t_{a,b}x_a$.
+%In $\QREL^{op}$ (which corresponds to systematically taking transpose matrices), composition coincides with the product matrix/matrix and $\hat{(\cdot)}$ with the product matrix/vector.
+%In order to avoid confusion, we will refer to a $t\in Q^{X\times Y}$ just as a \emph{matrix from $X$ to $Y$}.
+%As it is expected, $Q^X$ is a $Q$-semimodule and the bijection $\hat{(\cdot)}$ identifies $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.
 
@@
 
-The assumption of $Q$ being continuous (which can be weakened, see \cite{???}) is used in order to make the potentially infinite sum, appearing in the formula for the composition, converge.
-
-In the following, we always consider the category $\LREL$, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
-The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring (w.r.t.\ its quantale order $\preceq$.
-This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
-
-\begin{remark}
- In $\LREL$, the formula for the composition becomes the tropicalisation of the one defining it in $\QREL$, i.e.\ $(st)_{a,c}:=\inf\limits_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
- similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$, which we call \emph{tropical linear}, are exactly those of shape $f(x)=\inf\limits_{b\in Y} \set{\check f_{a,b}+x_a}$, for some matrix $\check f$ from $X$ to $Y$.
-\end{remark}
+The tropical relational semantics is thus provided by the category $\LREL$ of matrices with values over $\Lawv$ (which, being a quantale, is indeed a continuous semi-ring).
+%, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
+%The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring (w.r.t.\ its quantale order $\preceq$.
+%This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
+It is worth observing that the formula for composition in $\LREL$ is the tropicalisation of the one defining it in $\QREL$, i.e.\ $(st)_{a,c}:=\inf\limits_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
+ similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$ induced by matrices, which we call \emph{tropical linear}, are exactly those of shape $f(x)=\inf\limits_{b\in Y} \set{\check f_{a,b}+x_a}$, for some matrix $\check f$ from $X$ to $Y$.
+%\end{remark}
 
@@
 
-In the next sections we will see how $\LREL$ gives rise to denotational models of variants of the $\lam$-calculus.
+In the next sections we will see how $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus.

### 2023-01-17 section4.tex
commit

@@
-In this section we study the tropical Laurent series $\Lawv^X\to \Lawv ^Y$ from the viewpoint of analysis.
+As morphisms in $\LREL$ correspond to tropical linear functions, morphisms in the coKleisli category $\LREL_{!}$ 
+correspond to a generalization of tropical Laurent series, i.e.~the tropicalization of power series. 
+The study of power series in tropical mathematics is very recent (see e.g.~\cite{Porzio2021}) and many properties of these functions still have to be well-understood.
 
-\subsection{Results on Tropical Laurent Series}
+In this section we recall some known facts about tropical Laurent series, and we establish some new properties, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
 
-Recall the definition of \emph{tropical polynomials} and \emph{tropical Laurent series} (of one variable $x\in\Lawv$) given respectively by Equation~\ref{eq:polytrop} and Equation~\ref{eq:defTLS}.
-A $x\in[0,\infty)$ is said to be a (finite) \emph{tropical root} of a tropical Laurent series $f:\Lawv\to\Lawv$ iff $f$ is not differentiable at $x$ (w.r.t.\ the usual topology on $\BB R_{\geq0}$.
-This is equivalent to ask that the $\inf$ defining $f$ at $x$ is obtained \emph{at least} twice.{\color{red} \`e vero anche per TLS o solo per poly questo?
+%
+%In this section we study the tropical Laurent series $\Lawv^X\to \Lawv ^Y$ from the viewpoint of analysis.
+
+\subsection{From Tropical Polynomials to tLs}
+
+A tropical polynomial, as we saw, is a piece-wise linear function of the form $\varphi(x)=\min_{i=1,\dots,k}\{ix+c_{i}\}$. For instance, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{in+2^{-i}\}$
+are plotted in Fig.~\ref{fig:plot1} for $n=2,3,4$.
+A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable.
+In other words, the roots of $\varphi$ correspond with the points where the slope of $\varphi$ changes (i.e.~the points where the minimum defining $\varphi$ is attained twice).
+A simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.
+
+As anticipated in Section \ref{sectionnew}, a tropical Laurent series (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function that can be expressed as $f(x)=\inf_{n}\{nx+\check f_{n}\}$, with $\check  f_{n}$ a sequence in $\Lawv$.
+In other words, a tLs is a ``limit'' of tropical polynomials of higher and higher degree.
+For instance, the function $\varphi(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ (illustrated in Fig.~\ref{fig:plot1}), that we will take as our running example,
+is the ``limit'' of the polynomials $\varphi_{n}$.
+Since $\inf$s are not in general $\min$s, the behavior of tLS may be less predictable than that of tropical polynomials.
+For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points.
+
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[scale=0.6]
+\begin{axis}[samples=250]
+\addplot[red,domain=0:0.8] {min(2*x+1/4, x+1/2, 1)};
+\addplot[blue,domain=0:0.8] {min(3*x+1/8,2*x+1/4, x+1/2, 1)-0.01};
+\addplot[orange,domain=0:0.8] {min(4*x+1/16,3*x+1/8,2*x+1/4, x+1/2, 1)-0.02};
+
+\addplot[violet,domain=0:0.8] {min(
+10*x+1/1424,
+9*x+1/712,
+8*x+1/356,
+7*x+1/128,
+6*x+1/64,
+5*x+1/32,
+4*x+1/16,3*x+1/8,2*x+1/4, x+1/2, 1)-.03};
+
+\end{axis}
+
+\end{tikzpicture}
+\end{center}
+\label{fig:plot1}
+\caption{Plot of the tropical polynomials $\varphi_{2}$ (in red), $\varphi_{3}$ (in blue), $\varphi_{4}$ (in orange) and of their limit tLs $\varphi$ (in violet).}
+\end{figure}
+
+%
+%
+%We will take as our running example
+%%given respectively by Equation~\ref{eq:polytrop} and Equation~\ref{eq:defTLS}.
+%A positive $x\in[0,\infty)$ is said to be a (finite) \emph{tropical root} of a tropical Laurent series $f:\Lawv\to\Lawv$ iff $f$ is not differentiable at $x$ (w.r.t.\ the usual topology on $\BB R_{\geq0}$.
+%This is equivalent to ask that the $\inf$ defining $f$ at $x$ is obtained \emph{at least} twice.{\color{red} \`e vero anche per TLS o solo per poly questo?
 Inoltre, Robol parla anche di finite end-points of the domain: nel nostro caso sarebbe $x=0$}
 
-\begin{example}\label{ex:famous_ex}
- The function $f:\Lawv\to\Lawv$ defined by $f(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ is a tropical Laurent series.
-\end{example}
+%\begin{example}\label{ex:famous_ex}
+% The function $f:\Lawv\to\Lawv$ defined by $f(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ is a tropical Laurent series.
+%\end{example}
 
-By plotting its graph {\color{red}vogliamo plottarlo?}, we observe several properties that we will lift to the more general case that we will consider in the next lines, and this example will serve as running one.
+%By plotting its graph {\color{red}vogliamo plottarlo?}, we observe several properties that we will lift to the more general case that we will consider in the next lines, and this example will serve as running one.
 
-\begin{remark}%[Tropical Laurent series]
- As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ operates as a linear map $\widehat t:\Lawv^{!X}\to\Lawv^Y$.
- But we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_!
-x$ (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
+%\begin{remark}%[Tropical Laurent series]
+ As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a linear map $\widehat t:\Lawv^{!X}\to\Lawv^Y$, but we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_!
+x$.
+%  (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
  This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
- Concretely, we have:
- \[t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}\] where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
- The functions of such shape are generalisations of usual tropical Laurent series: they are the case for infinite variables.
- We will call them simply \emph{tropical Laurent series (tLs)}.
- %Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
+ Concretely, we have
+ \begin{equation}
+ t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}
+ \end{equation}
+ where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
+ These functions correspond then to tLs with possibly infinitely many variables (in fact, as many as the elements of $X$). %In the following we will also refer to them as tLs.
+
+% 
+% We will call them simply \emph{tropical Laurent series (tLs)}.
+% %Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
+%\end{remark}
+%
+%We find the usual notion of tLs of one variable as follows:
+
+%\begin{remark}
+Notice that, by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf\limits_{n\in\N}\set{nx+\widehat f(n)}$, for some $\widehat f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
+% \end{remark}
+ \begin{remark}
+ Our running example is indeed of shape $\varphi=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
+% By Proposition~\ref{prop:descrete}, $f$ is not the interpretation of a $\lam$-term, because its matrix is not discrete.
+% Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
 \end{remark}
 
-We find the usual notion of tLs of one variable as follows:
+ In a similar way, the tropical polynomials can be identified with the tLs
+$f:\Lawv^X\to\Lawv^Y$ 
+ for which the support $\C F=\set{\mu\in!X\mid\check f_{\mu,b}\neq\infty}$ is \emph{finite}, which have thus shape $f(x)_b:=\min\limits_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$.
+ This is again the generalisation of usual tropical polynomials to the case of infinitely many variables.
 
-\begin{remark}
- Indentifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tropical Laurent series generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf\limits_{n\in\N}\set{nx+\widehat f(n)}$, for some $\widehat f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
-Remark that we also find usual \emph{tropical polynomials} of tropical geometry as a particular case: they corrispond to the ones for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}.
-Actually, for us \emph{tropical polynomial} will mean a function $f:\Lawv^X\to\Lawv^Y$ of shape $f(x)_b:=\min\limits_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, for a \emph{finite} set $\C F\subseteq \, !X$.
-This is again the generalisation of usual tropical polynomials to the case of infinitely many variables.
-\end{remark}
+% 
+% , i.e.~for a \emph{finite} set $\C F\subseteq \, !X$.
+%%Remark that we also find usual \emph{tropical polynomials} of tropical geometry as a particular case: they 
+%correspond to the tLs for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}. 
+%Actually, for us \emph{tropical polynomial} will mean a function 
+%\end{remark}
 
-\begin{remark}
- The $f$ of Example~\ref{ex:famous_ex} is indeed of shape $f=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
- By Proposition~\ref{prop:descrete}, $f$ is not the interpretation of a $\lam$-term, because its matrix is not descrete.
- Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
-\end{remark}
-
-Looking at Example~\ref{ex:famous_ex}, we see that the function is non-decreasing and concave.
-It is easy to see that this is actually always the case:
+Looking at Fig~\ref{fig:plot1}, we see that the function $\varphi$, just like the polynomials $\varphi_{1}$, is non-decreasing and concave.
+It can be shown that this is indeed always the case:
 
@@
 
-In Example~\ref{ex:famous_ex}, it can be proven by hand that $f(x)$ is a $\min$ for all $x>0$.
-At $x=0$ we have $f(x=0)=\inf\limits_{n\in\N} \frac{1}{2^n}=0$ which is thus the only point \emph{not} being a $\min$.
+%In Example~\ref{ex:famous_ex}, 
+Again, by looking at Fig~\ref{fig:plot1}, it appears that $\varphi$ behaves \emph{locally} like the polynomials $\varphi_{n}$.
+In particular, for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$.
+Indeed, we can compute  by hand that $\varphi$ coincides with $\varphi_{n}$
+for $\epsilon \geq \log2/2^{n+1}$.
+However, at
+%
+%$x\in [\epsilon,\infty]$, with $\epsilon>0$
+%It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
+ $x=0$ we have that $\varphi(x=0)=\inf\limits_{n\in\N} \frac{1}{2^n}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
 Also, while the derivative of $f$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
-This phenomenon is reminiscent of [Example 7, PCoh],
+This phenomenon is reminiscent of [Example 7, \cite{Ehrhard2005}],
 %Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
@@
 In fact, these properties are shared by all tLs with \emph{finitely} many variables, as shown by the following result.
+
 Remark that $!\set{1,\dots,k}$ can be identified with $\N^k$.
@@
 \end{theorem}
-\begin{proof}
- Call $\leq$ the well-founded pointwise order on $\N^k$.
- Now we can let $\C F_\epsilon:=\set{n\in\N \mid
- \widehat f(n)<\infty \textit{ and } \widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$.
-\end{proof}
-
-{\color{red}
-- bisognerebbe anche dire due parole sul differenziale e dell'espansione di Taylor come operazione sulle TLS invece che sulle matrici. 
-}
+%\begin{proof}
+%Call $\leq$ the well-founded pointwise order on $\N^k$.
+%Now we can let $\C F_\epsilon:=\set{n\in\N \mid
+%\widehat f(n)<\infty \textit{ and } \widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$.
+%\end{proof}
 

### 2023-01-17 section4.tex
commit

@@
 \begin{figure}
-\begin{center}
-\begin{tikzpicture}[scale=0.6]
+\begin{subfigure}{0.24\textwidth}
+\begin{tikzpicture}[scale=0.55]
 \begin{axis}[samples=250]
@@
 \end{tikzpicture}
-\end{center}
+\caption{}
 \label{fig:plot1}
-\caption{Plot of the tropical polynomials $\varphi_{2}$ (in red), $\varphi_{3}$ (in blue), $\varphi_{4}$ (in orange) and of their limit tLs $\varphi$ (in violet).}
+\end{subfigure}
+\begin{subfigure}{0.24\textwidth}
+\begin{tikzpicture}[scale=0.55]
+\begin{axis}[samples=50, view={15}{45}]
+\addplot3[orange,domain=0:2] {min(2*x, 2*x+y, 3*y)};
+
+\end{axis}
+\end{tikzpicture}
+\caption{}
+\label{fig:plot2}
+\end{subfigure}
+\label{fig:plot12}
+\caption{
+%\ref{fig:plot1}
+ (a) Plot of the tropical polynomials $\varphi_{2}$ (in red), $\varphi_{3}$ (in blue), $\varphi_{4}$ (in orange) and of their limit tLs $\varphi$ (in violet).\\
+%\ref{fig:plot2}
+(b) Plot of
+$\varphi(x,y)=\min\{2x, 2x+y,3y\}$.}
 \end{figure}
@@
 \begin{proposition}
-  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders on the domain and codomain.
+  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders.
+  % on the domain and codomain.
 \end{proposition}
-
-The $f$ of Example~\ref{ex:famous_ex} is continuous on $\BB R_{\geq0}=\Lawv-\set{\infty}$ (w.r.t.\ the usual norm of real numbers).
+The tLs $\varphi$ is continuous on $\BB R_{\geq0}=\Lawv-\set{\infty}$ (w.r.t.\ the usual norm of real numbers).
 By considering the usual norm $\norm{x}_\infty:=\sup\limits_{a\in X} \absv{x_a}$ on $\Lawv^X$, we could generalise this property by dropping the case of $x$ having some $0$ coordinate:
@@
 The norm $\norm{\cdot}_\infty$ naturally induces a metric $\|\B x-\B y\|_{\infty}$ over the spaces $\Lawv^{X}$. 
-We now study the Lipschitz properties of tropical Laurent series w.r.t.~these metrics.
+A consequence of Theorem~\ref{theorem:fepsilon} is that tLs with finitely many variables are locally Lipschitz on all $\BB R_{>0}$.
+In this subsection we will refine and generalise this result.
+%
+%We will show that tLs satisfy suitable Lipschitz properties  w.r.t.~these metrics.
 
-First of all, remark that Theorem~\ref{theorem:fepsilon} implies that tLs $f$ of finitely many variables are locally Lipschitz (and thus continuous) on all $\BB R_{>0}$.
-In this subsection we will refine and generalise this result.
-
-A first result is the following:
+Let us first look at tropical linear functions:
 
@@
 \end{proposition}
-\begin{proof}[Proof sketch]
-Using the fact that $f(\B x)_{b}= \inf_{a\in X}\check f_{a,b}+\B x_{a}$,
-the problem reduces to checking that $|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
+%\begin{proof}[Proof sketch]
+%Using the fact that $f(\B x)_{b}= \inf_{a\in X}\check f_{a,b}+\B x_{a}$,
+%the problem reduces to checking that $|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
 This result shows that, in analogy with that happens in usual metric semantics, linear programs are interpreted by non-expansive functions. 
@@
 Before looking at what happens in the case of non-linear programs, let us make the metric structure of $\LREL$ explicit. 
-
-Using the fact that the tropical linear functions from $\Lawv^{X}$ to $\Lawv^{Y}$ are in bijection with the elements of the hom-set $\LREL(X,Y)\simeq \Lawv^{X\times Y}$ through the map $f\mapsto \check f$, the following proposition provides a useful characterization of the functional metrics in $\LREL$:
+The following proposition provides a useful characterization of the functional metrics in $\LREL$, relying on 
+the bijection between $\HOM{\LREL}{X}{Y}$ and set of tropical linear functions from $\Lawv^{X}$ to $\Lawv^{Y}$.
 \begin{proposition}
-For all sets $X,Y$ and tropical linear functions $f,g:\Lawv^{X}\to \Lawv^{Y}$, $\| \check f-\check g\|_{\infty} =  \sup_{\B x\in \Lawv^{X}}
+For all tropical linear functions $f,g:\Lawv^{X}\to \Lawv^{Y}$, $\| \check f-\check g\|_{\infty} =  \sup_{\B x\in \Lawv^{X}}
 \| f(\B x)-g(\B x)\|_{\infty}$.\end{proposition}
 
-To study what happens in the case of tropical Laurent series, let us first consider the case of bounded exponentials $!_{\leq n}X$:
+Let us now consider the case of bounded exponentials $!_{\leq n}X$:
 \begin{proposition}\label{prop:boundedlip}
-All tropical Laurent series $f: \Lawv^{X}\to \Lawv^{Y}$ for which $\check f:{\C M_{\leq n}(X)}\times Y\to \Lawv$, are $n$-Lipschitz-continuous.
+If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\check f:{\C M_{\leq n}(X)}\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
 \end{proposition}
 \begin{proof}[Proof sketch]
-Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \check f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
-$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
-it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
+This follows from Proposition \ref{prop:troplinear} and the remark that, for all $x\in \Lawv^{X}$, $\| (!_{n} x)-(!_{n} y)\|_{\infty}\leq n\cdot \| x- y\|_{\infty}$, where $!_{n} x$ is the restriction of $!
+x$ to $\C M_{\leq n}(X)$.%
+%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \check f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
+%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
+%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
 \end{proof}
-This result is perfectly analogous to what happens in standard models for sensitivity analysis based on the re-scaling trick recalled in Section \ref{section2}.
+This result is perfectly analogous to what happens in the metric models discussed in Section \ref{section2}, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
 
-We can now look at what happens in the case of the full exponential comonad, i.e.~with the tropical Laurent series.
-First, let us observe that, as expected, tropical polynomials 
-correspond to Lipschitz functions:
-\begin{proposition}
-Any tropical polynomial $\varphi:\Lawv^{\{1,\dots, K\}}\to\Lawv$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
+We can now look at what happens with tLs, i.e.~when considering the full exponential $!$.
+First, since a tropical polynomial can be represented as a tropical linear function from $\Lawv^{\C M_{\mathrm{deg}(\varphi)(X)}}$ to $\Lawv$, from Proposition \ref{prop:boundedlip} we deduce:
+\begin{proposition}\label{prop:polylip}
+Any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
 \end{proposition}
-\begin{proof}
-This immediately follows from Prop.~\ref{prop:boundedlip}, since a tropical polynomial can be represented as a tropical linear function from $\Lawv^{\C M_{\mathrm{deg}(\varphi)(\{1,\dots, K\})}}$ to $\Lawv$.
-\end{proof}
 
-The proposition above, together with Theorem \ref{theorem:fepsilon}, can be used to show that tropical Laurent series with \emph{finitely many} variables are always locally Lipschitz over $(0,\infty)^{X}$.
+The proposition above, together with Theorem \ref{theorem:fepsilon}, can be used to show that tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz over $(0,\infty)^{X}$.
 Yet, we can prove a more general statement, covering also the infinitary case.
@@
 \begin{proof}
-\begin{enumerate}
-\item It suffices to remark that if $A$ is a type of the $\BSTLC$, then $\model A$ is a finite set, by definition of the interpretation (Remark~\ref{rmk:ModelsOfBSTLC}).
-Therefore a morphism $t:\model A\to \model B$ is a finite matrix $t: \model A \times \model B \to \Lawv$.
-Hence, its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is a (tropical) polynomial, which is a Lipschitz functions.
-\item Trivial from Theorem~\ref{thmTLSlocLip}.
-\end{enumerate}
+As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be a finite  and  its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Prop.~\ref{prop:polylip}.
+Claim 2) follows directly from Theorem~\ref{thmTLSlocLip}.
 \end{proof}
 
-Finally, let us discuss the differential structure. We show that the distance between two tropical maps can be approximated using the terms appearing in their Taylor expansions:
+Finally, let us discuss the differential structure.
+The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tLs $f:\Lawv^{X}\to \Lawv^{Y}$ into a tLs $D_{!}(f):(\Lawv^{X})^{2}\to \Lawv^{Y}$, linear in its first variable, and given by
+\begin{equation}
+D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\check f_{\mu*a}+x_{a}+\mu y\right\}
+\end{equation}
+Moreover, the Taylor formula \eqref{eq:taylorcat} yields 
+a Taylor formula for tLs of the form 
+$f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}$.
+The following result shows that the distance between two tropical maps can be approximated using the terms appearing in their Taylor expansions:
 \begin{proposition}
 For all tropical Laurent series $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
-the functions
-$\delta^{(n)}f, \delta^{(n)}g: \Lawv^{\C M_{\leq n}(X)}\to \Lawv ^{Y}$, with 
- $\delta^{(n)}f(\B x)= \Der^{(n)}f(\B x, \infty)$, appearing in their Taylor expansions, are $n$-Lipschitz-continuous, and moreover 
-\begin{align}
-\| \check f-\check g\|_{\infty}= \sup_{n} \| \check{\delta^{(n)}f}- \check{\delta^{(n)}g}\|_{\infty}
-\end{align}
+the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz-continuous, and moreover 
+$\| \check f-\check g\|_{\infty}= \sup_{n} \| {\delta^{(n)}(f)}- {\delta^{(n)}(g)}\|_{\infty}$, 
+where $\delta^{(n)}(h)$ indicates the matrix of $D_{!}^{(n)}(h)(!_{n}x,\infty)$.
 %where $\delta^{(n)}f:( \Lawv^{X})^{n}\to \Lawv^{X}$ is the tropical linear function $\delta^{(n)}f(\B x_{1},\dots, \B x_{n})=

### 2023-01-17 section4.tex
commit

@@
 
-Let us now consider the case of bounded exponentials $!_{\leq n}X$:
+Let us now consider the case of bounded exponentials:
 \begin{proposition}\label{prop:boundedlip}
-If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\check f:{\C M_{\leq n}(X)}\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
+If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\check f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
 \end{proposition}

### 2023-01-17 section4.tex
commit

@@
 \end{equation}
+One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
 Moreover, the Taylor formula \eqref{eq:taylorcat} yields 
@@
 \begin{proposition}
-For all tropical Laurent series $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
-the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz-continuous, and moreover 
-$\| \check f-\check g\|_{\infty}= \sup_{n} \| {\delta^{(n)}(f)}- {\delta^{(n)}(g)}\|_{\infty}$, 
-where $\delta^{(n)}(h)$ indicates the matrix of $D_{!}^{(n)}(h)(!_{n}x,\infty)$.
+For all tLs $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
+the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz. Moreover 
+$\| \check f-\check g\|_{\infty}= \sup_{n} \| {\delta^{(n)}f}- {\delta^{(n)}g}\|_{\infty}$, 
+where $\delta^{(n)}h$ indicates the matrix of $D_{!}^{(n)}h$.
 %where $\delta^{(n)}f:( \Lawv^{X})^{n}\to \Lawv^{X}$ is the tropical linear function $\delta^{(n)}f(\B x_{1},\dots, \B x_{n})=

### 2023-01-17 section3C.tex
commit

@@
 
-%A model of the $\STDLC$ is usually understood as so-called \emph{Cartesian closed differential categories} (CC$\partial$C), see \cite{Manzo2010} for details.
+%A model of the $\STDLC$ is usually understood as so-called \emph{Cartesian closed differential categories} (CC$\partial$C), see \cite{Manzo2012} for details.
 %In order to treat the $+$ and the constructor $D[\_]\cdot (\_)$ of $\STDLC$, the main features of a CC$\partial$C $\C C$ are that:
@@
 There is more: not only $\LREL_!$ is a model of the $\STDLC$, but this model is also well-behaved w.r.t.\ to the Taylor expansion of $\lam$-terms.
-This is expressed by the following property (see [Definition 4.22, \cite{Manzo2010}], where it is called the fact of ``modeling the Taylor expansion'').
+This is expressed by the following property (see [Definition 4.22, \cite{Manzo2012}], where it is called the fact of ``modeling the Taylor expansion'').
 
@@
 %$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
-$u\star s= (Du)\circ \langle s,\infty\rangle$ corresponds to linear application of $s$ to $u$.
+$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to linear application of $s$ to $u$.
 Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$,  i.e.~it coincides with the tropical interpretation of the Taylor formula \eqref{eq:taylor}.
-In other words, \eqref{eq:taylorcat} says that the series corresponding to the Taylor expansion of an ordinary $\lam$-term $M$ converges to $M$ [Theorem 4.23, \cite{Manzo2010}].
+In other words, \eqref{eq:taylorcat} says that the series corresponding to the Taylor expansion of an ordinary $\lam$-term $M$ converges to $M$ [Theorem 4.23, \cite{Manzo2012}].
 

### 2023-01-17 section3intro.tex
commit

@@
  \begin{equation}
- \hat t(x)_a:=\sum\limits_{b\in Y} t_{a,b}x_a
+ \hat t(x)_b:=\sum\limits_{a\in X} t_{a,b}x_a
  \end{equation}
@@
 % fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum\limits_{b\in Y} s_{b,c}t_{a,b}$.
-In linear algebra, a map $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns.
-In particular, the product of such a matrix for a vector defines a map $Q^Y\to Q^X$, hence the opposite of what shown above.
+Notice that usual linear algebra conventions correspond to working in $\LREL^{\op}$: a matrix $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns, and the usual matrix-vector product defines a map $Q^Y\to Q^X$.
 \end{remark}

### 2023-01-17 section4.tex
commit

@@
 
-A tropical polynomial, as we saw, is a piece-wise linear function of the form $\varphi(x)=\min_{i=1,\dots,k}\{ix+c_{i}\}$. For instance, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{in+2^{-i}\}$
+A tropical polynomial, as we saw, is a piece-wise linear function of the form $f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}$.
+For instance, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{in+2^{-i}\}$
 are plotted in Fig.~\ref{fig:plot1} for $n=2,3,4$.
 A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable.
-In other words, the roots of $\varphi$ correspond with the points where the slope of $\varphi$ changes (i.e.~the points where the minimum defining $\varphi$ is attained twice).
-A simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.
+In other words, the roots of $\varphi$ correspond with the points where the slope of $\varphi$ changes (i.e.~the points where the minimum defining $\varphi$ is attained at least twice).
+It can be shown that this corresponds to the usual factorization property of roots: for any root $x_{0}$, $f$ factorizes as
+$f(x)=\min\{x,x_{0}\}+ g(x)$.
+%
+%A
+% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.
 
@@
 %\ref{fig:plot1}
- (a) Plot of the tropical polynomials $\varphi_{2}$ (in red), $\varphi_{3}$ (in blue), $\varphi_{4}$ (in orange) and of their limit tLs $\varphi$ (in violet).\\
+ (a) Plot of the tropical polynomials $\varphi_{2}$ (in red), $\varphi_{3}$ (in blue), $\varphi_{4}$ (in orange) and of their limit tLs $\varphi$ (in violet).
+The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x$ where $ix+2^{-i}=(i+1)x+2^{-i+1}$, given by $x=2^{-i+1}$.
 %\ref{fig:plot2}
@@
 %%given respectively by Equation~\ref{eq:polytrop} and Equation~\ref{eq:defTLS}.
-%A positive $x\in[0,\infty)$ is said to be a (finite) \emph{tropical root} of a tropical Laurent series $f:\Lawv\to\Lawv$ iff $f$ is not differentiable at $x$ (w.r.t.\ the usual topology on $\BB R_{\geq0}$.
-%This is equivalent to ask that the $\inf$ defining $f$ at $x$ is obtained \emph{at least} twice.{\color{red} \`e vero anche per TLS o solo per poly questo?
+%A positive $x\in[0,\infty)$ is said to be a (finite) \emph{tropical root} of a tropical Laurent series $f:\Lawv\to\Lawv$ iff $f$ is not differentiable at $x$ (w.r.t.\ the usual topology on $\BB R_{\geq0}$.This is equivalent to ask that the $\inf$ defining $f$ at $x$ is obtained \emph{at least} twice.{\color{red} \`e vero anche per TLS o solo per poly questo?
 Inoltre, Robol parla anche di finite end-points of the domain: nel nostro caso sarebbe $x=0$}
@@
 %It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
- $x=0$ we have that $\varphi(x=0)=\inf\limits_{n\in\N} \frac{1}{2^n}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
+ $x=0$ we have that $\varphi(x=0)=\inf_{n\in\N} \frac{1}{2^n}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
 Also, while the derivative of $f$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
@@
 Remark that $!\set{1,\dots,k}$ can be identified with $\N^k$.
-So the matrix of a tLs $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one variable $f(x)$ in output) can be given as a $\widehat f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf\limits_{n\in \N^k}\set{nx+\widehat f(n)}$, where $nx$ is the scalar product.
+So the matrix of a tLs $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one variable $f(x)$ in output) can be given as a $\widehat f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf_{n\in \N^k}\set{nx+\widehat f(n)}$, where $nx$ is the scalar product.
 
@@
 
-\subsection{$\Lawv^{X}$ as a normed cone}
+\subsection{Continuity of tLs}%$\Lawv^{X}$ as a normed cone}
 
@@
 
-\subsection{Metric results}%$\Lawv^{X}$ as a metric space.}
+\subsection{Lipschitz-Continuity of tLs}%$\Lawv^{X}$ as a metric space.}
 

### 2023-01-17 section5.tex
commit

@@
 
-Secondly, the Lipschitz conditions studied in the previous section allow us to reason on program distances in a \emph{compositional} way: suppose, as before, that $M,N:A$ are two programs such that $M$ improves on $N$ by $\epsilon$, and let $\TT C[-]:A \to \B{nat}$ indicate a context; knowing that the interpretation of $\TT C$ is $k$-Lipschitz-continuous on some open set containing both $\model M$ and $\model N$, allows us to immediately deduce that $\TT C[M]$ improves on $\TT C[N]$ by $k \epsilon$.
+Secondly, the Lipschitz conditions from Section \ref{section4} allow us to reason on program distances in a \emph{compositional} way: suppose, as before, that $M,N:A$ are two programs such that $M$ improves on $N$ by $\epsilon$, and let $\TT C[-]:A \to \B{nat}$ indicate a context; knowing that the interpretation of $\TT C$ is $k$-Lipschitz-continuous on some open set containing both $\model M$ and $\model N$, allows us to immediately deduce that $\TT C[M]$ improves on $\TT C[N]$ by $k \epsilon$.
 Observe that this will typically be the case when the Taylor expansions of $\TT C[M]$ and $\TT C[N]$ actually yields a \emph{finite} sum of at most $k$ terms, i.e.~when 
@@
 a variant of the relational semantics modeling strongly normalizing programs via \emph{finite} power series.
-We mention this point in the Section~\ref{sec:concl}
+We mention this point in Section~\ref{section8}
 
@@
 \end{remark}
-
-\subsection{Resource Analysis for Differential Privacy}
-
-The typical situation in differential privacy is where one considers a probabilistic query $f: \mathsf{db}\to [0,1]^{X}$ over some database, and one requires that $f$ should not be \emph{too sensitive} to small changes in the output, in other to prevent potential leaks of private information about individual items in $\mathsf{db}$ (for instance, an element $x\in\mathsf{db}$ could be the list of students of some university and $f(x)$ indicates the percentage of female students$\mathsf{db}$).
-
-More formally, differential privacy is defined as follows \cite{Reed2010}:
-\begin{definition}
-Let $f: \mathsf{db}\to [0,1]^{X}$ and $\epsilon \in \BB R_{\geq 0}$. $f$ is said \emph{$\epsilon$-differentially private} if for all $x,x'\in \mathsf{db}$
-differing by $L$ items, for all $s\in X$, 
-$$
-f(x)(s) \ = \ e^{\epsilon L} f(x')(s)
-$$
-\end{definition}
-
-A well-studied approach to ensure differential privacy for higher-order programs is to use bounded type systems like $\mathsf{Fuzz}$ \cite{Reed2010}.
-Indeed, such systems come with an \emph{a priori} warrant that well-typed programs are Lipschitz.
-
-The use of tropical semantics suggests how resource analysis could also be used to provide bounds for differential privacy. 
-Suppose our probabilistic query $f$ can be expressed as a power series (this is what happens e.g.~in \emph{probabilistic coherent spaces} \cite{Ehrhard2011}).
-Then, if we discover, either by studying differential properties of $f$, or using methods from convex analysis as suggested in the previous section (e.g.~Theorem \ref{??}),
- that the tropicalization $\trop f$ satisfies a Lipschitz condition, we may use this fact to deduce that $f$ is differentially private, as shown by the result below.
-
-%
-%We have seen how a higher-order probabilistic program, which is expressed as a polynomial, can be tropicalized.
-This operation can be extended to programs expressed as power series by taking the limit.
-%
-\begin{proposition}
-Let $f: [0,1]^{X}\to [0,1]^{Y}$ be an analytic function. If $\trop f$ is Lipschitz over some open set $U$, then $f$ is differentially private over $\psi_{t}(U)$, for small enough $t$.
-\end{proposition}
-\begin{proof}
-We consider the case $X=Y=\B 1$ for simplicity. 
-Express $f$ as a power series $f(x)=\sum_{n}\psi_{e}(\widehat f_{n})x^{n}$, so that $\trop f(\alpha)= \inf_{n}\left \{
-n\alpha + \widehat f_{n}
-\right\}$.
-Let us define a family of intermediate functions $\trop_{t}f(\alpha)=\phi_{t}(\stackrel{t}{\sum_{n}}\widehat{f}_{n}\times_{t}n\alpha)$. These, by construction, satisfy
-\begin{align}\label{eq:phit}
-f(\phi_{t}(\alpha)) = \phi_{t}(\trop_{t}f(\alpha))
-\end{align}
-and, using the fact that $\alpha\sumt{t}\beta\to\min\{\alpha,\beta\}$ and $\alpha\prodt{t}\beta\to \alpha+\beta$, we can deduce that $\trop f(\alpha)=\lim_{t\to 0}\trop_{t}f(\alpha)$ (for this, we use the fact that for all $m\in \BB N$,
-$\min_{i\leq m}\alpha_{i}\leq \stackrel{t}{\sum}_{i\leq m}\alpha_{i}\leq 
-(\min_{i\leq m}\alpha_{i})+t\log m$).
-
-Now, using the fact that $\trop f$ and $\trop_{t} f$ come arbitrarily close with $t$ small enough, together with \eqref{eq:phit}, the Lipschitz condition  
-$|\trop f(\phi_{t}(x))-\trop f(\phi_{t}(y))|\leq L\cdot |\phi_{t}(x)-\phi_{t}(y)|$ translates into the differential privacy condition 
-$f(x) \leq e^{L\cdot |\log x-\log y|} f(y)$ (supposing $y\geq x$ and $f(y)\neq 0$).
-\end{proof}
-
-%General discussion: optimization properties behave in a Lipschitz way.
-%
-%
-%- differential privacy and Lipschitzness
-%
-%
-%- log-probabilities and tropical roots 
-%
-%
-%- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)
-%
-%
-%- measuring duplications of discrete functions (needs finiteness!)

### 2023-01-17 section5C.tex
commit

\subsection{Resource Analysis for Differential Privacy}

The typical situation in differential privacy is where one considers a probabilistic query $f: \mathsf{db}\to [0,1]^{X}$ over some database, and one requires that $f$ should not be \emph{too sensitive} to small changes in the output, in other to prevent potential leaks of private information about individual items in $\mathsf{db}$ (for instance, an element $x\in\mathsf{db}$ could be the list of students of some university and $f(x)$ indicates the percentage of female students$\mathsf{db}$).

More formally, differential privacy is defined as follows \cite{Reed2010}:
\begin{definition}
Let $f: \mathsf{db}\to [0,1]^{X}$ and $\epsilon \in \BB R_{\geq 0}$. $f$ is said \emph{$\epsilon$-differentially private} if for all $x,x'\in \mathsf{db}$
differing by $L$ items, for all $s\in X$, 
$$
f(x)(s) \ = \ e^{\epsilon L} f(x')(s)
$$
\end{definition}

A well-studied approach to ensure differential privacy for higher-order programs is to use bounded type systems like $\mathsf{Fuzz}$ \cite{Reed2010}. Indeed, such systems come with an \emph{a priori} warrant that well-typed programs are Lipschitz.

The use of tropical semantics suggests how resource analysis could also be used to provide bounds for differential privacy. 
Suppose our probabilistic query $f$ can be expressed as a power series (this is what happens e.g.~in \emph{probabilistic coherent spaces} \cite{Ehrhard2011}). Then, if we discover, either by studying differential properties of $f$, or using methods from convex analysis as suggested in the previous section (e.g.~Theorem \ref{??}), 
 that the tropicalization $\trop f$ satisfies a Lipschitz condition, we may use this fact to deduce that $f$ is differentially private, as shown by the result below.

%
%We have seen how a higher-order probabilistic program, which is expressed as a polynomial, can be tropicalized. This operation can be extended to programs expressed as power series by taking the limit. 
%
\begin{proposition}
Let $f: [0,1]^{X}\to [0,1]^{Y}$ be an analytic function. If $\trop f$ is Lipschitz over some open set $U$, then $f$ is differentially private over $\psi_{t}(U)$, for small enough $t$.
\end{proposition}
\begin{proof}
We consider the case $X=Y=\B 1$ for simplicity. 
Express $f$ as a power series $f(x)=\sum_{n}\psi_{e}(\widehat f_{n})x^{n}$, so that $\trop f(\alpha)= \inf_{n}\left \{
n\alpha + \widehat f_{n}
\right\}$.
Let us define a family of intermediate functions $\trop_{t}f(\alpha)=\phi_{t}(\stackrel{t}{\sum_{n}}\widehat{f}_{n}\times_{t}n\alpha)$. These, by construction, satisfy
\begin{align}\label{eq:phit}
f(\phi_{t}(\alpha)) = \phi_{t}(\trop_{t}f(\alpha))
\end{align}
and, using the fact that $\alpha\sumt{t}\beta\to\min\{\alpha,\beta\}$ and $\alpha\prodt{t}\beta\to \alpha+\beta$, we can deduce that $\trop f(\alpha)=\lim_{t\to 0}\trop_{t}f(\alpha)$ (for this, we use the fact that for all $m\in \BB N$,
$\min_{i\leq m}\alpha_{i}\leq \stackrel{t}{\sum}_{i\leq m}\alpha_{i}\leq 
(\min_{i\leq m}\alpha_{i})+t\log m$).

Now, using the fact that $\trop f$ and $\trop_{t} f$ come arbitrarily close with $t$ small enough, together with \eqref{eq:phit}, the Lipschitz condition  
$|\trop f(\phi_{t}(x))-\trop f(\phi_{t}(y))|\leq L\cdot |\phi_{t}(x)-\phi_{t}(y)|$ translates into the differential privacy condition 
$f(x) \leq e^{L\cdot |\log x-\log y|} f(y)$ (supposing $y\geq x$ and $f(y)\neq 0$).
\end{proof}

%General discussion: optimization properties behave in a Lipschitz way.
%
%
%- differential privacy and Lipschitzness
%
%
%- log-probabilities and tropical roots 
%
%
%- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)
%
%
%- measuring duplications of discrete functions (needs finiteness!)

### 2023-01-17 section6.tex
commit

@@
 This viewpoint gives a metric flavor to $\LREL$, and allowed us to relate differential and metric structure.
-Yet, how far can this correspondence between tropical linear/non-linear algebra and metric topology be pushed?
-
-For instance, natural questions are: (1) Is this correspondence restricted to $\Lawv$-modules of the form $\Lawv^{X}$ (i.e.~with a fixed base), or does it hold in some sense for arbitrary $\Lawv$-modules?
-(2) Is this correspondence restricted to the $\infty$-norm metric, or does it hold for other metrics too?
+Yet, how far can this correspondence be pushed?
+In particular, is this correspondence restricted to $\Lawv$-modules of the form $\Lawv^{X}$ (i.e.~with a fixed base), or does it hold in some sense for arbitrary $\Lawv$-modules?
+Is this correspondence restricted to the $\infty$-norm metric, or does it hold for other metrics too?
 
 An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}.
-This correspondence, recalled in this section, and the fact that it leads to the definition of a linear $\Lawv$-category (i.e.~a model of the linear $\lambda$-calculus) of genealized metric spaces extending $\LREL$, is mostly a \emph{collage} of  \emph{folklore} results with more abstract results contained in recent literature in enriched category theory \cite{Fuji, Stubbe2006, Shen2014}.
-Beyond this reconstruction, our main contribution is to show that this correspondence can be lifted to a model of the full differential $\lambda$-calculus, by a suitable generalization of the construction of the exponential comonad of $\LREL$.
+In this section we first reconstruct this correspondence, by combining some {folklore} results with more abstract ones from recent literature in enriched category theory \cite{Fuji, Stubbe2006, Shen2014}.
+Then we show that this correspondence can be lifted to a model of the full differential $\lambda$-calculus, by suitably generalizing the construction of the exponential comonad of $\LREL$.
 
@@
 
-A $\Lawv$-module is a triple $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$ (notice that this is the usual notion of module over the Lawvere quantale $\Lawv$, not that of module over the tropical semi-ring).
-A $\Lawv$-module homomorphism is a map $f:M\to N$ commuting with both joins and the action. We let $\Mod$ indicate the category of $\Lawv$-modules and their homomorphisms.
+A $\Lawv$-module is a triple $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$.% (notice that this is indeed the usual notion of module over the Lawvere quantale $\Lawv$).
+A $\Lawv$-module homomorphism is a map $f:M\to N$ commuting with both joins and the $\Lawv$-action. We let $\Mod$ indicate the category of $\Lawv$-modules and their homomorphisms.
 
@@
  While the $\Lawv^{X}$ have a fixed base, for an arbitrary $\Lawv$-module one can retrieve a base via the \emph{Yoneda embedding}
-$\Yon: M \to \Hom(M^{\op},\Lawv)$, where $\Yon(x)(y)=\inf\{\epsilon\mid \epsilon\star y\succeq x\}$.
+$\Yon: M \to \Mod(M^{\op},\Lawv)$, where $\Yon(x)(y)=\inf\{\epsilon\mid \epsilon\star y\succeq x\}$.
 
-\begin{proposition}\label{prop:yonemod}
+\begin{proposition}[cf.~\cite{Stubbe2014, Shen2014}]\label{prop:yonemod}
 For any $\Lawv$-module $M$, the Yoneda embedding has a left-adjoint $\sup(f)=\bigvee_{x\in M}f(x)\star x$.
@@
 \begin{proposition}
-$\Mod$ is a linear $\Lawv$-category.
+$\Mod$ is a SMCC.
 \end{proposition}
-\begin{proof}[Proof sketch]
-The hom-sets $\Hom(M,N)$ have a natural $\Lawv$-module structure, defined pointwise.
-Moreover, the tensor product of $\Lawv$-modules $M\otimes N$ can be defined as the quotient of the usual tensor product of sup-lattices (see e.g.~\cite{Russo2007}), under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$.
+%\begin{proof}
+The hom-sets $\Mod(M,N)$ have a natural $\Lawv$-module structure, defined pointwise.
+Moreover, the tensor product of $\Lawv$-modules $M\otimes N$ can be defined as the quotient of the usual tensor product of sup-lattices under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$ (see e.g.~\cite{Russo2007}).
 Notably, any element of $M\otimes N$ can be identified with a join of basic tensors $x\otimes y$, corresponding to the equivalence class of the pair $\langle x,y\rangle$.
-
 Beyond the required adjointness of the internal hom and the tensor, one can check that $\Mod$ is actually \emph{$^{*}$-autonomous}, since it satisfies $(M^{\op})^{\op}\simeq M$ and 
@@
 %$((x,\bigvee B),\bigcup_{b\in B}\{(x,b)\})$, for all $x\in A$, $B\subseteq N$, and all pairws $\{
-\end{proof}
+%\end{proof}
 
 \begin{remark}
-The structure of linear $\Lawv$-category of $\Mod$ coincides with that of $\LREL$ for the modules $\Lawv^{X}$.
-Indeed, one can prove that $\Hom(\Lawv^{X},\Lawv^{Y})\simeq \Lawv^{X\times Y} $ (using the fact that module homomorphisms are in bijection with $\Lawv$-matrices), and
-moreover $\Lawv^{X}\otimes \Lawv^{Y}\simeq \Lawv^{X\times Y} $ and 
-$\Pi_{i\in I}\Lawv^{X_{i}}\simeq \Lawv^{\coprod_{i\in I}X_{i}}$.
+The SMCC structure of $\Mod$ coincides with that of $\LREL$ for the modules $\Lawv^{X}$: we already know that $\Mod(\Lawv^{X},\Lawv^{Y})\simeq \Lawv^{X\times Y} $, and 
+one can prove $\Lawv^{X}\otimes \Lawv^{Y}\simeq \Lawv^{X\times Y} $ (cf.~\cite{Russo2007}).% and 
+%$\Pi_{i\in I}\Lawv^{X_{i}}\simeq \Lawv^{\coprod_{i\in I}X_{i}}$.
 \end{remark}
@@
 \end{align}
-This structure is often referred as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}, or as a \emph{quasi-metric space}. 
+This structure is often referred as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}.%, or as a \emph{quasi-metric space}. 
 Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, i.e.~it must satisfy $Y(f(x),f(y))\leq X(x,y)$.
@@
 For any $\Lawv$-category $X$, the presheafs $[X^{\op},\Lawv]$ on $X$ form another $\Lawv$-category, with metric $[X^{\op},\Lawv](f,g)= \sup_{x\in X}\Lawv(f(x),g(x))$.
-Notice that, when $X$ has the discrete metric, $[X,\Lawv]$ coincides with $\Lawv^{X}$.
+Notice that, when $X$ has the discrete metric, the metric space $[X,\Lawv]^{\sym}$ coincides with $\Lawv^{X}$ with the norm metric.
 The \emph{Yoneda embedding} is the faithful functor $\Yon: X\to [X^{\op},\Lawv]$ given by $\Yon(x)(y)=X(y,x)$.
 
-Interestingly, a fundamental example of $\Lawv$-categories is provided by $\Lawv$-modules.
+Actually, a well-studied example of $\Lawv$-categories is provided by $\Lawv$-modules.
 
@@
 
-Lawvere also observed that Cauchy-completeness can be expressed in categorical language as the representability in $X$ of certain presheaves in $[X^{\op},\Lawv]$ \cite{Lawvere1973}.
+Lawvere also observed that Cauchy-completeness can be expressed in categorical language as the representability in $X$ of certain presheaves in $[X^{\op},\Lawv]$ \cite{Lawvere1973, Rutten1998}.
 Category theory suggests a yet stronger notion of completeness, corresponding to the existence of all \emph{weighted colimits} (for a comparison of different notions of completeness on $\Lawv$-categories, see \cite{Willerton2013, Rutten1998}).
@@
 \begin{definition}
-Let $X,Y,Z$ be $\Lawv$-enriched category,
+Let $X,Y,Z$ be $\Lawv$-categories,
 $\Phi: Z\pfun Y$ be a distributor and  $f:Y\to X$ be a functor.
-A functor $g:Z\to X$ is the \emph{$\Phi$-weighted colimit of $f$ over $X$}, noted $\colim(\Phi,f)$, if for all $z\in z$ and $x\in X$
+A functor $g:Z\to X$ is the \emph{$\Phi$-weighted colimit of $f$ over $X$}, noted $\colim(\Phi,f)$, if for all $z\in Z$ and $x\in X$
 \begin{align}
@@
 \end{align} 
-A functor $f:X\to Y$ is said \emph{continuous} if it commutes with all existing weighted colimits in $X$, i.e.~$f(\colim(\Phi,g))=\colim(\Phi,f\circ g)$. A $\Lawv$-enriched category 
+A functor $f:X\to Y$ is \emph{continuous} if it commutes with all existing weighted colimits in $X$, i.e.~$f(\colim(\Phi,g))=\colim(\Phi,f\circ g)$. A $\Lawv$-enriched category 
 $X$ is said \emph{categorically complete} (or just \emph{complete}) if all weighted colimits over $X$ exist. 
@@
 A useful alternative characterization of complete $\Lawv$-categories is the following:
-\begin{proposition}
+\begin{proposition}[cf.~\cite{Stubbe2006, Shen2014}]
 A $\Lawv$-category is complete iff $\Yon$ has a left-adjoint. 
 \end{proposition}
+\begin{proof}
+For one side, if $X$ is complete, one can define $\sup : [X^{\op},\Lawv]\to X$ as a weighted colimit via $X(\sup x,b)=\sup_{a\in X}
+X(a,b)-x_{a}$. Conversely, if a left-adjoint $\sup $ exists, one can define $\mathrm{colim}(\Phi,f):= \sup \Psi$, where $\Psi_{a}=\inf_{b\in X}X(a,f(b))+\Phi_{b}$. 
+\end{proof}
+
 Indeed, using this fact, together with Proposition \ref{prop:yonemod}, we arrive at the following:
@@
 
-\begin{theorem}
+\begin{theorem}[cf.~\cite{Stubbe2006}]
 $\Mod$ and $\GMet$ are isomorphic categories.
@@
 
-Since $\Mod$ (and thus $\GMet$ too) is a linear $\Lawv$-category, it is worth making its metric structure explicit. Given complete $\Lawv$-categories $X,X_{i},Y$, we have that:
-\begin{itemize} 
-\item the distance on the hom-set $\Hom(X,Y)$ is given by 
+Since $\Mod$ (and thus $\GMet$ too) is a SMCC, it is worth making its metric structure explicit.
+Given complete $\Lawv$-categories $X,Y$, we have that the distance on the hom-set $\Mod(X,Y)$ is given by
 \begin{align}
-\Hom(X,Y)(f,g)= \sup_{x\in X}Y(f(x),g(x));
+\Mod(X,Y)(f,g)= \sup_{x\in X}Y(f(x),g(x));
 \end{align}
-\item the distance on the tensor $X\otimes Y$ is given by
+and the distance on basic tensors 
+of $X\otimes Y$ is given by
 \begin{align}
-(X\otimes Y)(\alpha, \beta)=
-\sup_{i}\inf_{j}X(x_{i},x'_{j})+Y(y_{j},y'_{j}),
+%(X\otimes Y)(\alpha, \beta)=
+%\sup_{i}\inf_{j}X(x_{i},x'_{j})+Y(y_{j},y'_{j}),
+(X\otimes Y)(x\otimes y, x'\otimes y')=
+X(x,x')+Y(y,y')
 \end{align}
-where $\alpha=\bigvee_{i}x_{i}\otimes y_{i}$ and 
-$\beta=\bigvee_{j}x'_{j}\otimes y'_{j}$, 
-and thus coincides with the sum-metric over basic tensors;
-\item the distance on the bi-product $\Pi_{i\in i}X_{i}$ is given by
-\begin{align}
-(\Pi_{i\in I}X_{i})(f,g)= \sup_{i\in I}X_{i}(f(x),g(x));
-\end{align}
+and is extended to arbitrary joins by continuity.
+%where $\alpha=\bigvee_{i}x_{i}\otimes y_{i}$ and 
+%$\beta=\bigvee_{j}x'_{j}\otimes y'_{j}$, 
+%and thus coincides with the sum-metric over basic tensors;
+%\item the distance on the bi-product $\Pi_{i\in i}X_{i}$ is given by
+%\begin{align}
+%(\Pi_{i\in I}X_{i})(f,g)= \sup_{i\in I}X_{i}(f(x),g(x));
+%\end{align}
 
-\end{itemize}
+%\end{itemize}
 
@@
 
-First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies2018, Manzo2013}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting the fact that tensors and products commute in $\Mod$, we define $!$ as the product of the symmetric algebras $!_{n}$.
+First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies2018, Manzo2013}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting suitable properties of $\Mod$, we may define $!$ as the product of the symmetric algebras.
 
@@
 
-To prove that the map $h:\Sym_{n}(M)\to M^{\otimes_{n}} $ is the equalizer of the diagram formed by all homomorphisms $\langle \sigma\rangle$, it is useful to provide an alternative characterization of it.
+To prove that the map $[x]\mapsto x:\Sym_{n}(M)\to M^{\otimes_{n}} $ is the equalizer of the diagram formed by all morphisms $\langle \sigma\rangle$, it is useful to provide an alternative characterization of it.
 
@@
 $\langle \sigma \rangle (x)=x$. 
-
  A \emph{$\Lawv$-multiset} (with $n$ elements) is an element of $M^{\otimes_{n}}$ of the form 
- \begin{align}
- [x_{1},\dots, x_{n}]:= \bigvee_{\sigma\in \F S_{n}}
- x_{\sigma(1)}\otimes \dots \otimes x_{\sigma(n)}
- \end{align}
-where $x_{1},\dots, x_{n}\in M$.
+$ [x_{1},\dots, x_{n}]:= \bigvee_{\sigma\in \F S_{n}}
+ x_{\sigma(1)}\otimes \dots \otimes x_{\sigma(n)}$, where $x_{1},\dots, x_{n}\in M$.
 \end{definition}
@@
 Any $\Lawv$-multiset is p-invariant.
-Moreover, the set $!_{n}M$ of p-invariant elements of $M^{\otimes_{n}}$ is a $\Lawv$-submodule of $M^{\otimes_{n}}$, in which element can be written as a join of $\Lawv$-multisets.
+Moreover, the set $!_{n}M$ of p-invariant elements of $M^{\otimes_{n}}$ is a $\Lawv$-submodule of $M^{\otimes_{n}}$, whose elements can be written as joins of $\Lawv$-multisets.
 \end{proposition}
@@
 
-Also in this case of exponentials, the constructions above generalize the situation in $\LREL$:
+Also in the case of exponentials, the constructions for $\Mod$ generalize those of $\LREL$:
 \begin{proposition}
-For any set $X$, 
-$!_{n}(\Lawv^{X})\simeq \Lawv^{\C M_{n}(X)}$, where
-$\C M_{n}(X)$ is the set of multisets of $X$ of cardinality $n$, and 
-$!(\Lawv^{X})\simeq \Lawv^{\multiset(X)}$. 
+$!_{n}(\Lawv^{X})\simeq \Lawv^{\C M_{\leq n}(X)}$, $!(\Lawv^{X})\simeq \Lawv^{\multiset(X)}$. In particular, $\Mod_{!}(\Lawv^{X},\Lawv^{Y})\simeq \LREL_{!}(X,Y)$.
 \end{proposition}
 
-To conclude, we mush show that $\Mod$ is actually a model of $\STDLC$, that is, a cartesian closed differential category:
-\begin{enumerate}
-\item $\Mod$ is a left-additive-CCC, since each hom-set is a commutative monoid with respect to $\infty$ and $\min$, and the cartesian closed structure is well-behaved with respect to that;
+To conclude, we mush show that $\Mod$ is a CC$\partial$C.
+Firstly, $\Mod$ is a left-additive-CCC, since each hom-set is a commutative monoid with respect to $\infty$ and $\min$, and the cartesian closed structure is well-behaved with respect to that.
 
-\item $\Mod$ can be equipped with a differential operator defined, for $f:!M\to N$, by 
+Secondly, $\Mod$ can be equipped with a differential operator $E$ defined, for $f:!M\to N$, by 
 \begin{align}\label{eq:dermod}
-\Der[f](\alpha)=
+Ef(\alpha)=
 \bigvee\left\{
@@
 where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
-and $S: !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{Mellies2018}, 
-satisfying all axioms (D1)-(D7) and (DCurry), see the Appendix.
-\end{enumerate}
+and $S: !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{Mellies2018}, and $E$ satisfies all required axioms.
+%\end{enumerate}
 
-This leads then to:
+Notice that, when $f\in \Mod_{!}(\Lawv^{X},\Lawv^{Y})\simeq \LREL_{!}(X,Y)$, its derivative $E f$ coincides with the derivative $D_{!}f$ described in Section \ref{section4}.
+
+All this leads then to:
 \begin{theorem}
-$\Mod_{!}$ (equivalently, $\GMet_{!}$), equipped with $\Der$, is a $CC\partial C$.
+$\Mod_{!}$ (equivalently, $\GMet_{!}$), equipped with $E$, is a $CC\partial C$.
 \end{theorem}
-
-Notice that, when $f:!(\Lawv^{X})\to \Lawv^{Y}$, so that $f$ is described by a matrix $\widehat f: \multiset(X)\times Y\to \Lawv$, one can check that the matrix associated with \eqref{eq:dermod} coincides with the definition of the differential operator in $\LREL$.

### 2023-01-17 section8.tex
commit

@@
-We have shown that:
+In our opinion, the main goals of this paper are two. Firstly,  to
+demonstrate the existence of a conceptual bridge between two different well-studied quantitative approaches to higher-order programs, and to highlight the possibility of transferring results and techniques from one approach to the other.
+Secondly, to suggest that tropical mathematics, a
+field which has been largely and successfully applied in computer science, could be used to study quantitative properties of higher-order programs.
 
-- intuitively, the interplay between tropical mathematics and $\lam$-calculus \emph{could} relate the metric and differential approaches on approximations of $\lam$-calc (introduction+section 2)
+While the first goal has been developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions (best-case analysis, log-probabilities, differential privacy).
+We believe that exploring in more depth these ideas could be a fruitful direction; moreover,
+since both generalized metrics and quantale-modules have been largely applied in computer science, 
+a natural question is whether the generalized approach of Section \ref{section6} could  lead to new applications of metric and tropical methods to the $\lambda$-calculus.
 
-- this relation \emph{can} take place, because the natural category $\LREL$ and its generalised versions are metric models of the differential $\lam$-calculus (section 3 and 6)
-
-- this relation \emph{seems} to provide applications in different fields (section 5).
-
-Therefore, we mainly set the basis for future interplays between all these areas, hopefully motivating the interest in such an interplay.
-For instance, the general questions are:
-
-- Can we improve the results of section 4 ?
-
-- Can we develop and make the applications of section 5 useful ?
-
-- What does the general setting of section 6 give in terms of theoretical and applied results ?
-
-- Do tools from tropical \emph{geometry} provide something for $\lam$-calculus ? (for instance, the role of tropical roots, tropical varieties,...)
-
-- Finally, there is a last point which we think of interest and we did not mention through the paper: the inclusion of finiteness spaces in the picture.
-Say what could they do and why.
+Finally, as suggested in Section \ref{section5}, an interesting direction to look at is the semantics of finiteness spaces \cite{Ehrhard2005}, which allows to model typable terms by means of \emph{finitary} Taylor expansions.
+Indeed, knowing that the application of a program $M$ to $N$ expands as a finite sum of linear applications (hence by a Lipschitz function in our setting), may allow one to predict how sensitive $M$ will be ``around $N$''.
+%
+%We have shown that:
+%
+%- intuitively, the interplay between tropical mathematics and $\lam$-calculus \emph{could} relate the metric and differential approaches on approximations of $\lam$-calc (introduction+section 2)
+%
+%- this relation \emph{can} take place, because the natural category $\LREL$ and its generalised versions are metric models of the differential $\lam$-calculus (section 3 and 6)
+%
+%- this relation \emph{seems} to provide applications in different fields (section 5).
+%
+%Therefore, we mainly set the basis for future interplays between all these areas, hopefully motivating the interest in such an interplay.
+%For instance, the general questions are:
+%
+%- Can we improve the results of section 4 ?
+%
+%- Can we develop and make the applications of section 5 useful ?
+%
+%- What does the general setting of section 6 give in terms of theoretical and applied results ?
+%
+%- Do tools from tropical \emph{geometry} provide something for $\lam$-calculus ? (for instance, the role of tropical roots, tropical varieties,...)
+%
+%- Finally, there is a last point which we think of interest and we did not mention through the paper: the inclusion of finiteness spaces in the picture.
+%Say what could they do and why.

### 2023-01-17 section5C.tex
commit

@@
 
-The typical situation in differential privacy is where one considers a probabilistic query $f: \mathsf{db}\to [0,1]^{X}$ over some database, and one requires that $f$ should not be \emph{too sensitive} to small changes in the output, in other to prevent potential leaks of private information about individual items in $\mathsf{db}$ (for instance, an element $x\in\mathsf{db}$ could be the list of students of some university and $f(x)$ indicates the percentage of female students$\mathsf{db}$).
+The typical situation in differential privacy is where one considers a probabilistic query $f: \mathsf{db}\to [0,1]^{X}$ over some database, and one requires that $f$ should not be \emph{too sensitive} to small changes in the output, in other to prevent potential leaks of private information about individual items in $\mathsf{db}$ (for instance, an element $x\in\mathsf{db}$ could be the list of students of some university and $f(x)$ could indicate the percentage of female students$\mathsf{db}$).
 
-More formally, differential privacy is defined as follows \cite{Reed2010}:
+More formally, differential privacy is defined as follows:
 \begin{definition}
@@
 
-The use of tropical semantics suggests how resource analysis could also be used to provide bounds for differential privacy. 
+Tropical semantics suggests that resource analysis could also be used to provide bounds for differential privacy. 
 Suppose our probabilistic query $f$ can be expressed as a power series (this is what happens e.g.~in \emph{probabilistic coherent spaces} \cite{Ehrhard2011}).
-Then, if we discover, either by studying differential properties of $f$, or using methods from convex analysis as suggested in the previous section (e.g.~Theorem \ref{??}),
+Then, if we discover, either by studying differential properties of $f$, or by using methods from convex analysis as suggested in the previous section (e.g.~Theorem \ref{theorem:fepsilon}),
  that the tropicalization $\trop f$ satisfies a Lipschitz condition, we may use this fact to deduce that $f$ is differentially private, as shown by the result below.
@@
 %
+
+Let us equip the space $[0,1]^{X}$ with the metric $d(x,y)=\sup_{a\in X}|\log x_{a}-\log y_{a}|$. Then we have the following:
 \begin{proposition}
-Let $f: [0,1]^{X}\to [0,1]^{Y}$ be an analytic function. If $\trop f$ is Lipschitz over some open set $U$, then $f$ is differentially private over $\psi_{t}(U)$, for small enough $t$.
+Let $f: [0,1]^{X}\to [0,1]^{Y}$ be a function expressed by a power series.
+If $\trop f$ is $\epsilon$-Lipschitz over some open set $U$, then $f$ is $\epsilon$-differentially private over $\psi_{t}(U)$, for small enough $t$.
 \end{proposition}
-\begin{proof}
-We consider the case $X=Y=\B 1$ for simplicity. 
-Express $f$ as a power series $f(x)=\sum_{n}\psi_{e}(\widehat f_{n})x^{n}$, so that $\trop f(\alpha)= \inf_{n}\left \{
-n\alpha + \widehat f_{n}
-\right\}$.
-Let us define a family of intermediate functions $\trop_{t}f(\alpha)=\phi_{t}(\stackrel{t}{\sum_{n}}\widehat{f}_{n}\times_{t}n\alpha)$. These, by construction, satisfy
-\begin{align}\label{eq:phit}
-f(\phi_{t}(\alpha)) = \phi_{t}(\trop_{t}f(\alpha))
-\end{align}
-and, using the fact that $\alpha\sumt{t}\beta\to\min\{\alpha,\beta\}$ and $\alpha\prodt{t}\beta\to \alpha+\beta$, we can deduce that $\trop f(\alpha)=\lim_{t\to 0}\trop_{t}f(\alpha)$ (for this, we use the fact that for all $m\in \BB N$,
-$\min_{i\leq m}\alpha_{i}\leq \stackrel{t}{\sum}_{i\leq m}\alpha_{i}\leq 
-(\min_{i\leq m}\alpha_{i})+t\log m$).
-
-Now, using the fact that $\trop f$ and $\trop_{t} f$ come arbitrarily close with $t$ small enough, together with \eqref{eq:phit}, the Lipschitz condition  
-$|\trop f(\phi_{t}(x))-\trop f(\phi_{t}(y))|\leq L\cdot |\phi_{t}(x)-\phi_{t}(y)|$ translates into the differential privacy condition 
-$f(x) \leq e^{L\cdot |\log x-\log y|} f(y)$ (supposing $y\geq x$ and $f(y)\neq 0$).
+\begin{proof}[Proof sketch]
+%We consider the case $X=Y=\B 1$ for simplicity. 
+%Express $f$ as a power series $f(x)=\sum_{n}\psi_{e}(\widehat f_{n})x^{n}$, so that $\trop f(\alpha)= \inf_{n}\left \{
+%n\alpha + \widehat f_{n}
+%\right\}$.
+%Let us define a family of intermediate functions $\trop_{t}f(\alpha)=\phi_{t}(\stackrel{t}{\sum_{n}}\widehat{f}_{n}\times_{t}n\alpha)$. These, by construction, satisfy
+%\begin{align}\label{eq:phit}
+%f(\phi_{t}(\alpha)) = \phi_{t}(\trop_{t}f(\alpha))
+%\end{align}
+%and, using the fact that $\alpha\sumt{t}\beta\to\min\{\alpha,\beta\}$ and $\alpha\prodt{t}\beta\to \alpha+\beta$, we can deduce that $\trop f(\alpha)=\lim_{t\to 0}\trop_{t}f(\alpha)$ (for this, we use the fact that for all $m\in \BB N$,
+%$\min_{i\leq m}\alpha_{i}\leq \stackrel{t}{\sum}_{i\leq m}\alpha_{i}\leq 
+%(\min_{i\leq m}\alpha_{i})+t\log m$).
+We consider $X=Y=\{\star\}$ for simplicity.
+With the notations from \S \ref{section3}.A, using the fact that $\lim_{t\to 0}\trop_{t}f= \trop f$, as well as the commutation $f(\phi_{t}(\alpha)) = \phi_{t}(\trop_{t}f(\alpha))$ (which follows from the definition of $\trop_{t}$), one can see that the Lipschitz condition
+$|\trop f(\phi_{t}(x))-\trop f(\phi_{t}(y))|\leq \epsilon |\phi_{t}(x)-\phi_{t}(y)|$ translates into the condition 
+$f(x) \leq e^{\epsilon |\log x-\log y|} f(y)$ (supposing $y\geq x$ and $f(y)\neq 0$).
 \end{proof}

### 2023-01-17 section4.tex
commit

@@
  Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\widehat f:\N^k\to\Lawv$.
- For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ s.t.:
- \begin{enumerate}
-  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
-  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
-  \item On all $[\epsilon,\infty]^k$, $f$ coincides with the tropical \emph{polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\widehat f(n)}$.
- \end{enumerate}
+ For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
+% 
+% \begin{enumerate}
+%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
+%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
+  %\item 
+$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical {polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\widehat f(n)}$.
+% \end{enumerate}
 \end{theorem}

### 2023-01-17 section5C.tex
commit

@@
 differing by $L$ items, for all $s\in X$, 
-$$
+$
 f(x)(s) \ = \ e^{\epsilon L} f(x')(s)
-$$
+$.
 \end{definition}

### 2023-01-17 section1.tex
commit

@@
 
-On the other hand there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected with the \emph{relational semantics} \cite{Manzo2010, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
+On the other hand there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected with the \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
 This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable to capture probabilistic properties \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.

### 2023-01-17 section2.tex
commit

@@
 The types of such systems are defined by combining linear constructors with a \emph{graded exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
-In the following sections we will sometimes make reference to a basic type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined via $A::= o \  \mid \  (!_{\leq n}A) \multimap A  $,
-where $n\in \BB N$. Intuitively, $!_{\leq n}A\multimap B$ is the type of functions from $A$ to $B$ that may use their input at most $n$ times. More details about $\BSTLC$ are provided in the Appendix.
+In the following sections we will sometimes make reference to a basic type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined via $A::= o  \mid   (!_{n}A) \multimap A  $,
+where $n\in \BB N$.
+Intuitively, $!_{ n}A\multimap B$ is the type of functions from $A$ to $B$ that may use their input \emph{at most} $n$ times.
+More details about $\BSTLC$ are provided in the Appendix.
 
@@
 cartesian \emph{closed} differential categories ($CC\partial C$), 
-which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential are both well-behaved with respect to the closed structure (see \cite{Manzo2010}).
+which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential are both well-behaved with respect to the closed structure (see \cite{Manzo2012}).
 
@@
 \section{Tropical mathematics%: a possible synthesis
-}
+}\label{sectionnew}
 
-\subsection{Tropical numbers}
+\subsection{Tropicalization of polynomials and power series}
 
@@
 \begin{align}\label{eq:polytrop}
-\varphi(\alpha)= \min_{i=1,\dots,k}\left\{ i\alpha + c_{i}\right\}
+\varphi(\alpha)= \min_{i_{1},\dots, i_{k}}\left\{ i_{j}\alpha + c_{i_{j}}\right\}
 \end{align}
-where the coefficients $c_{i}$ are taken from $\Lawv$. 
+where the $i_{j}$ are natural numbers and the coefficients $c_{i_{j}}$ are taken from $\Lawv$. For instance, the polynomial
+$\varphi_{3}(\alpha)=\min\{ 3\alpha+1/8,2\alpha+1/4, \alpha+1/2, \alpha\}$ will be discussed in Section \ref{section4}, and its graph is illustrated in Fig.~\ref{fig:plot1}.
 %A value $\alpha\in \Lawv$ is a \emph{root} of the polynomial $P$ when

### 2023-01-17 section3A.tex
commit

@@
 
-Its monoidal structure is given by a tensor product $\otimes$ acting on the objects as the Cartesian product of sets, and as the \emph{Kronecker product} {(\color{red} Ref?)} of matrices on morphisms.
+Its monoidal structure is given by a tensor product $\otimes$ acting on the objects as the Cartesian product of sets, and as the \emph{Kronecker product}
+ %{(\color{red} Ref?)} 
+ of matrices on morphisms.
 Its closed structures is also given by the Cartesian product on sets, with the usual evaluation and curry maps.

### 2023-01-17 section3B.tex
commit

@@
 In order to interpret the $\STLC$, we need a Cartesian closed category (CCC).
-It is well known ({\color{red}Ref?}) that it is always possible to construct a CCC by taking the \emph{co-Kleisli} $\C C_!$ of a so-called \emph{Lafont category} $\C C$.
+It is well known (\cite{MacLane}) that it is always possible to construct a CCC by taking the \emph{co-Kleisli} $\C C_!$ of a so-called \emph{Lafont category} $\C C$.
 %, a construction we now quickly recall.

### 2023-01-17 section3intro.tex
commit

--- 2023-01-17 section3intro.tex
@@
 For a fixed \emph{continuous} semi-ring $Q$ [Def.
-II.5, \cite{Manzo2013}], the category $\QREL$ has sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms (\cite{Manzo2013} would call it $Q^\Pi$).
-The identity morphism of $\QREL$ is the matrix $i:X\times X\to Q$ given by $i_{a,a}=1$ and $i_{a,b\neq a}=0$, and composition of morphisms $t:X\times Y\to Q$ and $s:Y\times Z\to Q$ is given by 
+II.5, \cite{Manzo2013}], the category $\QREL$ has sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms, i.e.~$\QREL(X,Y)=Q^{X\times Y}$ (\cite{Manzo2013} would call it $Q^\Pi$).
+The identity morphism of $\QREL$ is the identity matrix $I:X\times X\to Q$,
+% given by $i_{a,a}=1$ and $i_{a,b\neq a}=0$, 
+ and composition of morphisms $t:X\times Y\to Q$ and $s:Y\times Z\to Q$ is given by 
 $(st)_{a,c}:=\sum\limits_{b\in Y} s_{b,c}t_{a,b}$.
@@
 
- As it is expected, $Q^X$ is a $Q$-semimodule and the bijection $\hat{(\cdot)}$ identifies $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$ via
+ As it is expected, $Q^X$ is a $Q$-semimodule and the bijection $\hat{(\cdot)}$ identifies the set of linear maps from $Q^X$ to $Q^Y$ with 
+ $\HOM{\QREL}{X}{Y}$, any map $f:Q^{X}\to Q^{Y}$ being of the form
    \begin{equation}
- \hat t(x)_a:=\sum\limits_{b\in Y} t_{a,b}x_a
+  f(x)_b:=\sum\limits_{a\in X} \matr f_{a,b}x_a
  \end{equation}
+ for some matrix $\matr f:X\times Y\to Q$.
 
@@
 % fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum\limits_{b\in Y} s_{b,c}t_{a,b}$.
-In linear algebra, a map $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns.
-In particular, the product of such a matrix for a vector defines a map $Q^Y\to Q^X$, hence the opposite of what shown above.
+Notice that usual linear algebra conventions correspond to working in $\QREL^{\op}$: a matrix $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns, and the usual matrix-vector product defines a map $Q^Y\to Q^X$.
 \end{remark}
@@
 It is worth observing that the formula for composition in $\LREL$ is the tropicalisation of the one defining it in $\QREL$, i.e.\ $(st)_{a,c}:=\inf\limits_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
- similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$ induced by matrices, which we call \emph{tropical linear}, are exactly those of shape $f(x)=\inf\limits_{b\in Y} \set{\check f_{a,b}+x_a}$, for some matrix $\check f$ from $X$ to $Y$.
+ similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$ induced by matrices, which we call \emph{tropical linear}, are exactly those of shape $f(x)=\inf\limits_{b\in Y} \set{\matr f_{a,b}+x_a}$, for some matrix $\matr f$ from $X$ to $Y$.
 %\end{remark}

### 2023-01-17 section4.tex
commit

@@
 
-\subsection{From Tropical Polynomials to tLs}
+\subsection{From tropical polynomials to tLs}
 
@@
 A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable.
-In other words, the roots of $\varphi$ correspond with the points where the slope of $\varphi$ changes (i.e.~the points where the minimum defining $\varphi$ is attained at least twice).
-It can be shown that this corresponds to the usual factorization property of roots: for any root $x_{0}$, $f$ factorizes as
+In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes.
+For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
+With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
 $f(x)=\min\{x,x_{0}\}+ g(x)$.
@@
 
-As anticipated in Section \ref{sectionnew}, a tropical Laurent series (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function that can be expressed as $f(x)=\inf_{n}\{nx+\check f_{n}\}$, with $\check  f_{n}$ a sequence in $\Lawv$.
+As anticipated in Section \ref{sectionnew}, a tropical Laurent series (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function that can be expressed as $f(x)=\inf_{n}\{nx+\matr f_{n}\}$, with $\matr  f_{n}$ a sequence in $\Lawv$.
 In other words, a tLs is a ``limit'' of tropical polynomials of higher and higher degree.
-For instance, the function $\varphi(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ (illustrated in Fig.~\ref{fig:plot1}), that we will take as our running example,
+For instance, the function $\varphi(x):=\inf_{n\in\N}\set{nx+\frac{1}{2^n}}$ (illustrated in Fig.~\ref{fig:plot1}), that we will take as our running example,
 is the ``limit'' of the polynomials $\varphi_{n}$.
@@
  (a) Plot of the tropical polynomials $\varphi_{2}$ (in red), $\varphi_{3}$ (in blue), $\varphi_{4}$ (in orange) and of their limit tLs $\varphi$ (in violet).
-The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x$ where $ix+2^{-i}=(i+1)x+2^{-i+1}$, given by $x=2^{-i+1}$.
+The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x=2^{-(i+1)}$, satisfying $ix+2^{-i}=(i+1)x+2^{-(i+1)}$.
 %\ref{fig:plot2}
@@
 %\begin{remark}%[Tropical Laurent series]
- As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a linear map $\widehat t:\Lawv^{!X}\to\Lawv^Y$, but we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_!
-x$.
+ As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, but we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting 
+ $t^!(x):=t\circ_! x$.
 %  (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
@@
 %\begin{remark}
-Notice that, by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf\limits_{n\in\N}\set{nx+\widehat f(n)}$, for some $\widehat f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
+Notice that, by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf\limits_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
 % \end{remark}
@@
 $f:\Lawv^X\to\Lawv^Y$ 
- for which the support $\C F=\set{\mu\in!X\mid\check f_{\mu,b}\neq\infty}$ is \emph{finite}, which have thus shape $f(x)_b:=\min\limits_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$.
+ for which the support $\C F=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ is \emph{finite}, which have thus shape $f(x)_b:=\min\limits_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$.
  This is again the generalisation of usual tropical polynomials to the case of infinitely many variables.
@@
 Remark that $!\set{1,\dots,k}$ can be identified with $\N^k$.
-So the matrix of a tLs $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one variable $f(x)$ in output) can be given as a $\widehat f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf_{n\in \N^k}\set{nx+\widehat f(n)}$, where $nx$ is the scalar product.
+So the matrix of a tLs $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one variable $f(x)$ in output) can be given as a $\matr f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, where $nx$ is the scalar product.
 
 \begin{theorem}\label{theorem:fepsilon}
- Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\widehat f:\N^k\to\Lawv$.
+ Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\matr f:\N^k\to\Lawv$.
  For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
@@
   %\item 
-$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical {polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\widehat f(n)}$.
+$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical {polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
 % \end{enumerate}
@@
 
-\subsection{Lipschitz-Continuity of tLs}%$\Lawv^{X}$ as a metric space.}
+\subsection{Lipschitz-continuity of tLs}%$\Lawv^{X}$ as a metric space.}
 
-The norm $\norm{\cdot}_\infty$ naturally induces a metric $\|\B x-\B y\|_{\infty}$ over the spaces $\Lawv^{X}$. 
+The norm $\norm{\cdot}_\infty$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$. 
 A consequence of Theorem~\ref{theorem:fepsilon} is that tLs with finitely many variables are locally Lipschitz on all $\BB R_{>0}$.
@@
 %\begin{proof}[Proof sketch]
-%Using the fact that $f(\B x)_{b}= \inf_{a\in X}\check f_{a,b}+\B x_{a}$,
-%the problem reduces to checking that $|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
+%Using the fact that $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
+%the problem reduces to checking that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
 This result shows that, in analogy with that happens in usual metric semantics, linear programs are interpreted by non-expansive functions. 
 %\begin{proof}
-%Using $f(\B x)_{b}= \inf_{a\in X}\check f_{a,b}+\B x_{a}$,
-%first observe that $|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$; we now have
-%$|f(\B x)_{b}-f(\B y)_{b}| \leq |(\inf_{a\in X}\check f_{a,b}-\B x_{a})-(\inf_{a\in X}\check f_{a,b}-\B y_{a})| \leq
-%\sup_{a\in X}|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})|\leq 
+%Using $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
+%first observe that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$; we now have
+%$|f(\B x)_{b}-f(\B y)_{b}| \leq |(\inf_{a\in X}\matr f_{a,b}-\B x_{a})-(\inf_{a\in X}\matr f_{a,b}-\B y_{a})| \leq
+%\sup_{a\in X}|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})|\leq 
 % \| \B x-\B y\|_{\infty}$.
@@
 \begin{proposition}
-For all tropical linear functions $f,g:\Lawv^{X}\to \Lawv^{Y}$, $\| \check f-\check g\|_{\infty} =  \sup_{\B x\in \Lawv^{X}}
-\| f(\B x)-g(\B x)\|_{\infty}$.\end{proposition}
+For all tropical linear functions $f,g:\Lawv^{X}\to \Lawv^{Y}$, $\norm{ \matr f-\matr g}_{\infty} =  \sup_{\B x\in \Lawv^{X}}
+\norm{ f(\B x)-g(\B x)}_{\infty}$.\end{proposition}
 
@@
 \begin{proposition}\label{prop:boundedlip}
-If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\check f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
+If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
 \end{proposition}
 \begin{proof}[Proof sketch]
-This follows from Proposition \ref{prop:troplinear} and the remark that, for all $x\in \Lawv^{X}$, $\| (!_{n} x)-(!_{n} y)\|_{\infty}\leq n\cdot \| x- y\|_{\infty}$, where $!_{n} x$ is the restriction of $!
+This follows from Proposition \ref{prop:troplinear} and the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $!
 x$ to $\C M_{\leq n}(X)$.%
-%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \check f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
+%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
 %$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
@@
 \begin{proof}
-As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be a finite  and  its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Prop.~\ref{prop:polylip}.
+As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be finite  and  its corresponding linear map $\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Prop.~\ref{prop:polylip}.
 Claim 2) follows directly from Theorem~\ref{thmTLSlocLip}.
@@
 \begin{equation}
-D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\check f_{\mu*a}+x_{a}+\mu y\right\}
+D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu*a}+x_{a}+\mu y\right\}
 \end{equation}
@@
 the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz. Moreover 
-$\| \check f-\check g\|_{\infty}= \sup_{n} \| {\delta^{(n)}f}- {\delta^{(n)}g}\|_{\infty}$, 
+$\norm{ \matr f-\matr g}_{\infty}= \sup_{n} \norm{ {\delta^{(n)}f}- {\delta^{(n)}g}}_{\infty}$, 
 where $\delta^{(n)}h$ indicates the matrix of $D_{!}^{(n)}h$.

### 2023-01-17 section5.tex
Commit notturno

@@
 
-Let us consider in this subsection a probabilistic extension of $\lam$-calculus, call it $\STLC_\oplus$, adding as usual terms of shape $pM+qN$ and $M\oplus_p N$, for $p,q\in[0,1]$.
-These terms are typed via the rule:
+Let us consider in this subsection a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
+we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the rules:
 \[
  \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A} 
+ \qquad\qquad
+ \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
 \]
-and similar for $\Gamma\vdash pM+qN:A$.
-We add the reduction rule:
+We add reduction rules:
 \[
- M\oplus_p N \to pM+(1-p)N
+ M\oplus_p N \to pM \quad \textit{ and } \quad M\oplus_p N \to (1-p)N
 \]
-so that such terms play the role of probabilistic choices of parameter $p$, as well as the rule:
+so that $M\oplus_p N$ plays the role of a probabilistic sample with parameter $p$.
+We use this calculus as a toy example for our purposes, and it can be seen as a fragment of the PCF considered in \cite{Manzo2013}.
+
+Consider now the following term $M$ of type $\bool$:
 \[
- pM+qM\to (p+q)M.
+ (\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\true\oplus_p\false))
 \]
-
-Let us consider $M:=(I\oplus_p\Omega)\oplus_p((I\oplus_p\Omega)\oplus_p(I\oplus_p\Omega))$.
-Reducing to normal form, we have:
-\[
- M\twoheadrightarrow (p^2+(1-p)p^2+(1-p)^3)I+(p(1-p)+2(1-p)^2p)\Omega.
-\]
-Setting $q:=1-p$, we can consider the two weights of $I$ and $\Omega$ as polynomials in the variables $p$ and $q$.
-Let us call
-$Q_1(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ the weight of $I$ and
-$Q_2(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the weight of $\Omega$,
+Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, reading $0$ as ``left'' and $1$ as ``right''.
+The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
+For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$, starting from the left.
+Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
+$P_{00}(p,q)\true$, 
+$P_{100}(p,q)\true$, 
+$P_{111}(p,q)\true$, 
+$P_{01}(p,q)\false$, 
+$P_{110}(p,q)\false$,
+$P_{101}(p,q)\false$,
 where the $P$'s are the following monomials in $p,q$:
@@
 $P_{110}(p,q)=P_{101}(p,q):=q^2p$.
-The index $\omega\in\set{00,100,111,01,110,101}$ of each $P_\omega$ indicates the path of the reduction that led from $M$ to the respective occurrence $I_\omega$ of $I$ or $\Omega_\omega$ of $\Omega$ from $M$ to its normal form ($0$ means ``left'' and $1$ means ``right'').
-For instance, in order to reach $I_{100}$, i.e.\ the second occurrence of $I$ from the left in $M$, we have to take the right path in the outer $\oplus_p$ of $M$, then two times the left path in the new outer $\oplus_p$'s that we encounter during reduction.
-$P_{\omega}(p,q)$ gives then the probability (as a function of $p,q$) of obtaining the respective occurrence $I_{\omega}$ or $\Omega_\omega$ in the normal form, if we were to sample at each time we reduce a $\oplus_p$.
-It can thus also be read as the likelihood function of such event.
-The polynomials $Q_{1,2}(p,q)$ give instead the whole probability of obtaining respectively $I$ or $\Omega$, in the normal form after such samplings.
+They correspond to the respective reduction path from $M$ to the normal term of the same address.
+$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ in $M$ after reducing it to normal form.
+Thinking, as we will, to $p,q$ as parameters, it can thus also be read as the \emph{likelihood function} of such event.
+The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after such samplings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.
+%Let us consider in this subsection a probabilistic extension of $\lam$-calculus, call it $\STLC_\oplus$, adding as usual terms of shape $pM+qN$ and $M\oplus_p N$, for $p,q\in[0,1]$.These terms are typed via the rule:
+%\[
+%\dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
+%\]
+%and similar for $\Gamma\vdash pM+qN:A$.We add the reduction rule:
+%\[
+% M\oplus_p N \to pM+(1-p)N
+%\]
+%so that such terms play the role of probabilistic choices of parameter $p$, as well as the rule:
+%\[
+% pM+qM\to (p+q)M.
+%\]
+%Let us consider $M:=(I\oplus_p\Omega)\oplus_p((I\oplus_p\Omega)\oplus_p(I\oplus_p\Omega))$. Reducing to normal form, we have:
+%\[
+% M\twoheadrightarrow (p^2+(1-p)p^2+(1-p)^3)I+(p(1-p)+2(1-p)^2p)\Omega.
+%\]
+%The index $\omega\in\set{00,100,111,01,110,101}$ of each $P_\omega$ indicates the path of the reduction that led from $M$ to the respective occurrence $I_\omega$ of $I$ or $\Omega_\omega$ of $\Omega$ from $M$ to its normal form ($0$ means ``left'' and $1$ means ``right'').For instance, in order to reach $I_{100}$, i.e.\ the second occurrence of $I$ from the left in $M$, we have to take the right path in the outer $\oplus_p$ of $M$, then two times the left path in the new outer $\oplus_p$'s that we encounter during reduction.$P_{\omega}(p,q)$ gives then the probability (as a function of $p,q$) of obtaining the respective occurrence $I_{\omega}$ or $\Omega_\omega$ in the normal form, if we were to sample at each time we reduce a $\oplus_p$.It can thus also be read as the likelihood function of such event.The polynomials $Q_{1,2}(p,q)$ give instead the whole probability of obtaining respectively $I$ or $\Omega$, in the normal form after such samplings.
 
-The tropical point of view allows us to express two natural questions about this situation:
+The tropical point of view allows now to express two natural questions about this situation:
 \begin{enumerate}
- \item What is the maximum likelihood estimator for the event ``$M$ produces $I_\omega$'' (similarly for $\Omega$)?
-This means asking what is the choice of the parameters $p,q$ that maximizes the probability of getting $I_\omega$ by reducing $M$.
- \item Knowing that $M$ produced $I$ (similarly for $\Omega$), which is the choice of the parameters $p,q$ that maximizes the probability that the produced occurrence of $I$ was a fixed $I_{\omega_0}$ (i.e.
-that among the paths leading to $I$, the one taken was $\omega_0$)?
+ \item What is the \emph{maximum likelihood estimator} for the event ``$M$ produces $\true_\omega$'' (similarly for $\false_\omega$)?
+ This means asking which is the choice of the parameters $p,q$ that maximizes the probability of getting $\true_\omega$ by repeatedly sampling $M$.
+ \item Knowing that $M$ produced $\true$ (similarly for $\false$), which is the choice of the parameters $p,q$ that maximizes the probability that the produced occurrence of $\true$ was a fixed $\true_{\omega_0}$ (i.e.
+that among the paths leading to $\true$, the one taken was $\omega_0$)?
 \end{enumerate}
+
+These are questions much in the style of those asked in {\color{red}Ref?} in the framework of {\color{red}??}.
 
@@
 
-\begin{fact}
- By using \cite{Manzo2013}, it can be shown that $\LREL_!$ is a model of such calculus $\STLC_\oplus$. {\color{red}GIUSTIFICARLO BENE ?}
-\end{fact}
+\begin{remark}
+ By adapting [Section IV, Section VI, \cite{Manzo2013}], it can be seen that $\LREL_!$ is a model of $\STLC_\oplus$.
+ In particular, if we set $\model \bool:=\set{0,1}$, our running example $M$ is interpreted as $\model M\in\Lawv^{\set{0,1}}$ giving the tropicalised probabilities: $\model M_0=(\trop Q_0)(p,1-p)$, $\model M_1=(\trop Q_1)(p,1-p)$.
+\end{remark}
 
-Now, in $\LREL$, seen as a model of such probabilistic $\lam$-calculus, the interpretation of a term already computes the tropicalisation of the polynomials expressing the probabilities, because the underlying semiring of the model is tropical.
-For instance, for our running example $M$:
-\[\model M = \min\left\{\trop{Q_1}(p,1-p) \cdot \model I, \trop{Q_2}(p,1-p) \cdot \model \Omega\right\}.\]
-So, assuming the interpretation of a term known and assuming the tropical polynomials are extracted from it, one could study the optimisation problems \ref{eq:argmax}, \ref{eq:max} \emph{without} having to reduce the term.
+%Now, in $\LREL$, seen as a model of such probabilistic $\lam$-calculus, the interpretation of a term already computes the tropicalisation of the polynomials expressing the probabilities, because the underlying semiring of the model is tropical.
+%For instance, for our running example $M$:
+%\[\model M = \min\left\{\trop{Q_1}(p,1-p) \cdot \model I, \trop{Q_0}(p,1-p) \cdot \model \Omega\right\}.\]
+Therefore, assuming the interpretation of a term known and assuming the tropical polynomials (in $p$ and $q$, not after the substitution $q:=1-p$) are extracted from it, one could study the optimisation problems \ref{eq:argmax}, \ref{eq:max} \emph{without} having to reduce the term.
 
-\begin{example}\label{ex:study}
- For our running example $M$, we have $\trop Q_1(x,y)=\min\set{2x,y+2x,3y}$ and $\trop Q_2(x,y)=\min\set{x+y,2y+x}$.
- Studying $\trop Q_1$ we see that {\color{red}Vogliamo plottare $\trop Q_1$ ?}
- $\trop Q_1(x,y)=3y$ iff $y\leq \frac{2}{3}x$, and it coincides with $2x$ otherwise.
+%\begin{example}\label{ex:study}
+For our running example $M$, we have $\trop Q_1(x,y)=\min\set{2x,y+2x,3y}$ and $\trop Q_0(x,y)=\min\set{x+y,2y+x}$.
+Studying $\trop Q_1$, whose plot is in Fig.~\ref{fig:plot2}, we see that $\trop Q_1(x,y)=3y$ iff $y\leq \frac{2}{3}x$, and it coincides with $2x$ otherwise.
 Remembering that $3y=P_{111}(x,y)$, we can now solve the optimisation problem~\ref{eq:max} for $\omega_0=111$:
 via the substitution $x:=-\log_c p$, $y:=-\log_c (1-p)$, Equation~\ref{eq:max} is equivalent to $-\log_c (1-p)\leq -\frac{2}{3}\log_c p$, i.e.\ $1-p\geq p^{\frac{2}{3}}$.
- This means that, for $p\in[0,1]$ s.t.\ $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), the most likely occurrence of $I$ to obtain, knowing that $M$ produced $I$, is $I_{111}$.
- Remembering that $2x=P_{00}(x,y)$, for the other values of $p$ (for example, $p=\frac{1}{2}$), the most likely $I$ to be sampled is the occurrence $I_{00}$.
+This means that, for $p\in[0,1]$ s.t.\ $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), the most likely occurrence of $\true$ to obtain, knowing that $M$ sampled $\true$ in its normal form, is $\true_{111}$.
+Remembering that $2x=P_{00}(x,y)$, for the other values of $p$ (for example, $p=\frac{1}{2}$), the most likely $\true$ to be sampled is the occurrence $\true_{00}$.
 We have thus answered question 2) above.
-\end{example}
+%\end{example}
 
 \begin{remark}
- Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_1$ or of $\trop Q_2$ give, by definition of tropical root, the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable paths of samples during the reduction of $M$ to its normal form.
- Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log_c p,-\log_c(1-p))$ is a tropical root of, say, $\trop Q_1$ are the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable occurrences of $I$ that are sampled by $M$ during the reduction, knowing that some $I$ was sampled.
+ Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_1$ or of $\trop Q_0$ give, by definition of tropical root, the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable paths of samples from $M$ to its normal form.
+ Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log_c p,-\log_c(1-p))$ is a tropical root of, say, $\trop Q_1$ are the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are sampled by $M$ during its reduction to normal form, knowing that some $\true$ was sampled.
 \end{remark}

### 2023-01-18 section1.tex
commit

@@
 
-Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} (whose origin stands behind the adjective ``tropical'') as a way to do algebra and geometry by replacing the usual ring structure of numbers based on addition and multiplication by the semi-ring structure given by ``$\min$'' and ``$+$''.
+Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon}
+ (whose origin stands behind the adjective ``tropical'') 
+ as a combinatorial approach to algebra and geometry obtained by replacing the usual ring structure of numbers based on addition and multiplication by the semi-ring structure given by ``$\min$'' and ``$+$''.
 %
@@
 
-As working on the idempotent $\min,+$-semiring is often easier that working with addition and multiplication, the tropicalization of  algebraic concepts makes several intractable properties (e.g.~finding roots of polynomials, or studying geometric invariants like e.g.~the genus) become tractable.
-For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).
+Working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, one of the fundamental and most intractable problems in algebra, that of finding roots of polynomials, admits a linear time algorithm in the tropical case (and the roots of a tropicalized polynomial can be used to approximate the actual roots of the polynomial \cite{Noferini2015}).
+Such a computational nature of tropical concepts explains why these are widely applied in computer science, e.g.~for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).
 

### 2023-01-18 section5.tex
commit

@@
 $P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ in $M$ after reducing it to normal form.
-Thinking, as we will, to $p,q$ as parameters, it can thus also be read as the \emph{likelihood function} of such event.
+Thinking of $p,q$ as parameters, $P_{\omega}$ can thus be read as the \emph{likelihood function} of the event $\omega$.
 The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after such samplings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.
+
 %Let us consider in this subsection a probabilistic extension of $\lam$-calculus, call it $\STLC_\oplus$, adding as usual terms of shape $pM+qN$ and $M\oplus_p N$, for $p,q\in[0,1]$.These terms are typed via the rule:
@@
 
-The tropical point of view allows now to express two natural questions about this situation:
+This way the probabilistic evaluation of $M$ is presented in the form of a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}.
+Indeed, typical questions in this case would be:
+%
+%The tropical point of view allows now to express two natural questions about this situation:
 \begin{enumerate}
@@
 \end{enumerate}
+A similar argument could be done by replacing $\true$ and $\false$ by, respectively, a converging and a diverging term (e.g.~in a $\B{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.
 
-These are questions much in the style of those asked in {\color{red}Ref?} in the framework of {\color{red}??}.
-
+%These are questions much in the style of those asked in {\color{red}Ref?} in the framework of {\color{red}??}.
+Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega}$, which is more easily solved by 
+passing to the tropical polynomial $\trop P_{\omega}$. 
 For 1), we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)$ belong to:
@@
 \end{equation}
-In both cases, the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.
+In both cases, passing to $\trop P_{\omega}$ makes the problem easier, as this may be reduced to computing the tropical roots of the polynomial (which can be done in linear time \cite{Noferini2015}).
+%the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.
 
 \begin{remark}
- By adapting [Section IV, Section VI, \cite{Manzo2013}], it can be seen that $\LREL_!$ is a model of $\STLC_\oplus$.
+ By adapting [Section IV, \cite{Manzo2013}], it can be seen that $\LREL_!$ is a model of $\STLC_\oplus$.
  In particular, if we set $\model \bool:=\set{0,1}$, our running example $M$ is interpreted as $\model M\in\Lawv^{\set{0,1}}$ giving the tropicalised probabilities: $\model M_0=(\trop Q_0)(p,1-p)$, $\model M_1=(\trop Q_1)(p,1-p)$.
@@
 %\[\model M = \min\left\{\trop{Q_1}(p,1-p) \cdot \model I, \trop{Q_0}(p,1-p) \cdot \model \Omega\right\}.\]
-Therefore, assuming the interpretation of a term known and assuming the tropical polynomials (in $p$ and $q$, not after the substitution $q:=1-p$) are extracted from it, one could study the optimisation problems \ref{eq:argmax}, \ref{eq:max} \emph{without} having to reduce the term.
+Therefore, assuming the interpretation of a term is known, 
+so that one can extract the associated tropical polynomials from it,
+% (in $p$ and $q$, not after the substitution $q:=1-p$) 
+% are extracted from it, 
+one could study the optimisation problems \ref{eq:argmax}, \ref{eq:max} \emph{without} having to actually reduce the term.
 
@@
 
-\begin{remark}
- Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_1$ or of $\trop Q_0$ give, by definition of tropical root, the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable paths of samples from $M$ to its normal form.
+\begin{remark}\label{rem:troproots}
+ Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_1$ or of $\trop Q_0$ provide, by definition, the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable paths of samples from $M$ to its normal form.
  Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log_c p,-\log_c(1-p))$ is a tropical root of, say, $\trop Q_1$ are the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are sampled by $M$ during its reduction to normal form, knowing that some $\true$ was sampled.

### 2023-01-18 section8.tex
Mini mini commit
Ho levato la frase "In our opinion" dalla prima riga della conclusione. (Ma se la preferisci, rimettila pure)
commit

@@
-In our opinion, the main goals of this paper are two. Firstly,  to
+%In our opinion,
+The main goals of this paper are two. Firstly,  to
 demonstrate the existence of a conceptual bridge between two different well-studied quantitative approaches to higher-order programs, and to highlight the possibility of transferring results and techniques from one approach to the other.
@@
 
-While the first goal has been developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions (best-case analysis, log-probabilities, differential privacy).
-We believe that exploring in more depth these ideas could be a fruitful direction; moreover,
-since both generalized metrics and quantale-modules have been largely applied in computer science, 
-a natural question is whether the generalized approach of Section \ref{section6} could  lead to new applications of metric and tropical methods to the $\lambda$-calculus.
+While the first goal was here developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions (best-case analysis, log-probabilities, differential privacy).
+We believe that exploring these ideas in more depth could be a fruitful direction; moreover,
+since both generalized metrics and quantale-modules have been largely studied in computer science, 
+a natural question is if the generalized approach of Section \ref{section6} could  lead to new applications of metric and tropical methods to the $\lambda$-calculus.
 

### 2023-01-19 appIV.tex
commit

We here focus on the category $Q^{\Pi}_{!}$, that is, the matricial category associated with $\mathsf{Trop}_{!}$. 
Let us recall that a morphism of $Q^{\Pi}_{!}(X,Y)$ is a matrix $f\in Q^{\C M_{\mathsf f}(X)\times Y}$, and that composition of $f\in Q^{\Pi}_{!}(X,Y)$ and $g\in Q^{\Pi}_{!}(Y,Z)$ is given by 
$$
(g\circ f)_{\mu,z}= \inf\Big\{
\sum_{i=1}^{n}f_{\mu_{i},y_{i}}+
g_{[y_{1},\dots, y_{n}],z}
\mid 
n\in \BB N, 
y_{1},\dots, y_{n}\in Y,
\mu= \mu_{1}+\dots + \mu_{n}
\Big\}
$$

Notice that $\infty \circ f= f\circ \infty=\infty$. However, while $\min\{g,h\}\circ f= \min\{g\circ f, h\circ f \}$ holds, as it can be computed using the continuity of $\mu, \inf$ and $+$, it is not true in general that 
$h\circ \min\{f,g\}= \min\{h\circ f, h\circ g\}$ (as a consequence of the fact that 
$\sum_{i=1}^{n}\min\{ f_{\mu_{i},y_{i}}, g_{\mu_{i},y_{i}}\} \neq 
\min\{ \sum_{i=1}^{n} f_{\mu_{i},y_{i}}, \sum_{i=1}^{n} g_{\mu_{i},y_{i}}\}$).
In other words, morphisms in $Q^{\Pi}_{!}$ are only \emph{left-$Q$-additive}.
A morphism $h\in Q^{\Pi}_{!}(X,Y)$ that satisfies $h\circ \min\{f,g\}=\min\{h\circ f, h\circ g\}$ for all object $X'$ and $f,g\in Q^{\Pi}_{!}(X',X)$, is called \emph{additive}.
%We already know from Giulio's paper that $Q^{\Pi}_{!}$ is a post-linear continuous $Q$-ccc.

\begin{proposition}
$Q^{\Pi}_{!}$ is a cartesian closed left-$Q$-additive category.
\end{proposition}
\begin{proof} While left-$Q$-additivity was already proved, to show that $Q^{\Pi}_{!}$ is cartesian closed left-$Q$-additive we must also check that (1) products and projections of additive morphisms are additive, and that (2) $\Lambda(\min\{f,g\})=\min\{\Lambda(f),\Lambda(g)\}$, $\Lambda(\infty)=\infty$, where $\Lambda:Q^{\C M_{\mathsf f}(Z+X)\times Y}\to Q^{\C M_{\mathsf f}(Z)\times( \C M_{\mathsf f}(X)\times Y)}$ is the isomorphism given by
$(\Lambda(f))_{\mu,\nu,y}= f_{\mu\oplus\nu,y}$, where 
 $\mu\oplus\nu$ is defined by $(\mu\oplus\nu)(\langle 0,x\rangle)=\mu(x)$ and $(\mu\oplus \nu)(\langle 1,x\rangle)=\nu(x)$.
	\begin{enumerate}
	\item  Let $f\in Q^{\C M_{\mathsf f}(X)\times Y}$ and $g\in Q^{\C M_{\mathsf g}(X)\times Z}$ be additive; then $\langle f,g\rangle \in Q^{\C M_{\mathsf f}(X)\times (Y+Z)}$, which is defined by 
	$$
	\langle f,g\rangle_{\mu,\langle i,a\rangle}= \begin{cases}
	f_{\mu,a} & \text{ if }i=0\\
	g_{\mu,a} & \text{ if }i=1
	\end{cases}
	$$
	is also additive. Indeed, for all $h\in Q^{\Pi}_{!}(X',X)$, 
	in any cartesian category it holds that $\langle f,g\rangle \circ h=\langle f\circ h, g\circ h\rangle$. Now, 
		if $i=0$, then for all $h_{1},h_{2}\in Q^{\C M_{\mathsf f}(X')\times X}$, 
	\begin{align*}
	(\langle f,g\rangle\circ \min\{h_{1},h_{2}\})_{\rho,\langle i,z\rangle}&=
	(\langle f\circ  \min\{h_{1},h_{2}\},g\circ  \min\{h_{1},h_{2}\}\rangle)_{\rho,\langle i,z\rangle} 
	\\
&=	(f\circ \min\{h_{1},h_{2}\})_{\rho, z}\\ &
	= \min\{(f\circ h_{1})_{\rho,z},(f\circ h_{2})_{\rho,z}\} \\
	&=
	\min\{ \langle f\circ h_{1},g\circ h_{1}\rangle_{\rho,\langle i,z\rangle}, 
	\langle f\circ h_{2},g\circ h_{2}\rangle_{\rho,\langle i,z\rangle}
\}
	\\
	& = \min\{ (\langle f,g\rangle\circ h_{1})_{\rho,\langle i,z\rangle},(\langle f,g\rangle\circ h_{2})_{\rho,\langle i,z\rangle}\}	\end{align*}
	and similarly if $i=1$.

	Moreover, suppose $f\in Q^{\Pi}_{!}(X, Y+Z)$ is additive, and let us show that $\pi_{1}( f)\in Q^{\Pi}_{!}(X,Y)$, defined by 
	$
	(\pi_{1}( f))_{\mu,y}= f_{\mu, \langle 0,y\rangle}
	$, is also additive: first observe that $\pi_{1}(f)=\pi_{1}\circ f$, where $\pi_{1}\in Q^{\Pi}_{!}(Y+Z,Y)$ is given by $(\pi_{1})_{\mu,y}=0$ if $\mu=[y]\oplus \emptyset$ and is $\infty$ otherwise; moreover, 
	$\pi_{1}(\min\{g,h\})=\min\{\pi_{1}(g),\pi_{1}(h)\}$, since 
	$(\pi_{1}(\min\{g,h\}))_{\mu, y}=(\min\{g,h\})_{\mu, \langle 0,y\rangle}= \min\{g_{\mu, \langle 0,y\rangle},h_{\mu, \langle 0,y\rangle}\}=( \min\{\pi_{1}(g),\pi_{1}(h)\})_{\mu, y}$. Now,
		given $h_{1},h_{2}\in Q^{\Pi}_{!}(X',X)$, we have that 
	$\pi_{1}(f)\circ \min\{h_{1},h_{2}\}= (\pi_{1}\circ f)\circ \min\{h_{1},h_{2}\}= \pi_{1}\circ (f\circ \min\{h_{1},h_{2}\})= \pi_{1}\circ \min\{f\circ h_{1},f\circ h_{2}\}= \pi_{1}(\min\{f\circ h_{1},f\circ h_{2}\})= \min\{\pi_{1}(f\circ h_{1}), \pi_{1}(f\circ h_{2})\}= \min\{\pi_{1}(f)\circ h_{1},\pi_{1}(f)\circ h_{2})$.

	\item  It is clear then that $\Lambda(\infty)=\infty$, and moreover
	$$
	\Lambda (\min\{f,g\})_{\mu,\nu,y}= \min\{ f,g\}_{\mu\oplus \nu,y}= \min\{f_{\mu\oplus\nu,y},g_{\mu\oplus\nu,y}\}= \min(\Lambda(f),\Lambda(g))_{\mu,\nu,y}
	$$

	\end{enumerate}

\end{proof}

For any morphism $f\in Q^{\Pi}_{!}(X,Y)$, let us define a morphism $\mathsf D(f)\in Q^{\Pi}_{!}(X+X,Y)$, i.e.~$\mathsf D(f)\in Q^{\C M_{\mathsf f}(X+X),Y}$, by
$$
\mathsf D(f)_{\mu,y}=
\begin{cases}
f_{\mu'+x,y}
& \text{ if }\mu=[x]\oplus \mu'
\\
\infty & \text{ otherwise}
\end{cases}
$$

\begin{proposition}
The category $Q^{\Pi}_{!}$, endowed with the operator $\mathsf D$, is a cartesian closed differential category.
\end{proposition}
\begin{proof}
We must check axioms (1)-(7) of cartesian differential categories plus axiom (D-curry).
\begin{description}
\item[(1)] $\mathsf D(\min\{f,g\})=\min \{\mathsf D(f),\mathsf D(g)\}$ and $\mathsf D(\infty)=\infty$: 
while the latter is obvious, for the former we have 
$\mathsf D(\min\{f,g\})_{[x]\oplus\nu,y}= \min\{f,g\}_{\nu+x,y}= \min\{f_{\nu+x,y},g_{\nu+x,y}\}=\min\{\mathsf D(f),\mathsf D(g)\}_{[x]\oplus \nu,y}$, and if $\mu\neq [x]\oplus \nu$, 
$\mathsf D(\min\{f,g\})_{\mu,y}= \infty= \min\{\infty, \infty\}=\min\{ \mathsf D(f),\mathsf D(g)\}_{\mu,y}$. 
\item[(2)]
$\mathsf D(f)\circ \langle \min\{h,k\},v\rangle= \min\{ \mathsf D(f)\circ \langle h,v\rangle, \mathsf D(f)\circ \langle k,v\rangle\}$, and $\mathsf D(f)\circ \langle \infty,v\rangle=\infty$: we can compute
	\begin{align*}
	(\mathsf D(f)\circ \langle \min\{h,k\},v\rangle)_{\mu,y}&=
	\inf\Big\{ 
	\sum_{i=1}^{n}\min\{h,k\}_{\rho_{i},w_{i}}+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}\\
&	\qquad\qquad\mid
	\mu=\sum_{i=1}^{n}\rho_{i}+\sum_{j=1}^{m}\nu_{j},	[w]=[w_{1},\dots, w_{n}]
	\Big\}\\
	& 
=	\inf\Big\{ 
	\min\{h,k\}_{\rho,w}+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}\\
&	\qquad\qquad\mid
	\mu=\rho+\sum_{j=1}^{m}\nu_{j}
	\Big\}\\
	&=	\min\Big\{\inf\big\{ 
	h_{\rho,w}+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}	\mid
	\mu=\rho+\sum_{j=1}^{m}\nu_{j}\big\}, \\
	& \qquad\qquad \inf\big\{ 
	k_{\rho,w}+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}	\mid
	\mu=\rho+\sum_{j=1}^{m}\nu_{j}\big\}\Big\}\\
	& = \min\Big\{(\mathsf D(f)\circ \langle h,v\rangle)_{\mu,y}, (\mathsf D(f)\circ \langle k,v\rangle)_{\mu,y}\Big\}\\
		& = \Big(\min\big\{\mathsf D(f)\circ \langle h,v\rangle, \mathsf D(f)\circ \langle k,v\rangle\big\}\Big )_{\mu,y}
	\end{align*}
	where, in the first equation, the condition $[w_{1},\dots, w_{n}]=[w]$ (i.e.~$n=1$) is forced by the fact that, otherwise, the application of $\mathsf D(f)$ would give $\infty$. Moreover, we have
\begin{align*}
(\mathsf D(f) \circ \langle \infty, v\rangle)_{\mu,y}&=\inf\Big\{ 
	\infty+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}\mid
	\mu=\rho+\sum_{j=1}^{m}\nu_{j}
	\Big\} = \infty
\end{align*}

\item[(3)] $\mathsf D(\mathrm{id})=\pi_{1}$, $\mathsf D(\pi_{i})=\pi_{i}\circ \pi_{1}$: 
recall that $\mathrm{id}_{[x],x}=0$ and $\mathrm{id}_{\mu,x}=\infty$, if $\mu\neq [x]$. 
Moreover $(\pi_{1})_{\mu,x}=0$ if $\mu=[x]\oplus \emptyset$, and is $\infty$ otherwise, and $\pi_{2}$ is defined similarly.
Hence 
$\mathsf D(\mathrm{id})_{[x]\oplus\nu,y}=\mathrm{id}_{\nu+x,y}$ is $0$ precisely when $x=y$ and $\nu=\emptyset$, and in all other cases is $\infty$. This shows that $\mathsf D(\mathrm{id})=\pi_{1}$.

$\mathsf D(\pi_{1})\in Q^{\C M_{\mathsf f}((X+Y)+(X+Y))\times Y}$ is given by
$\mathsf D(\pi_{1})
_{
[x\oplus \emptyset] \oplus (\mu\oplus\nu),y
}= (\pi_{1})_{(\mu\oplus\nu)+\langle 0,x\rangle ,y }
$, 
which is 0 precisely when $(\mu\oplus\nu)+\langle 0,x\rangle= y\oplus\emptyset$, i.e.~when 
$x=y$ and $\mu=\nu=\emptyset$; in all other cases one can check that $\mathsf D(\pi_{1})_{\rho,y}=\infty$, so we conclude $\mathsf D(\pi_{1})=\pi_{1}\circ \pi_{1}$.
One can argue similarly for $\pi_{2}$.

\item[(4)] $\mathsf D(\langle f,g\rangle)=\langle \mathsf D(f),\mathsf D(g)\rangle$: 
we have
\begin{align*}
\mathsf D(\langle f,g\rangle)_{[x]\oplus\mu, \langle 0,y\rangle}& = 
(\langle f,g\rangle)_{\mu+x,\langle 0,y\rangle}= f_{\mu+x,y}= \mathsf D(f)_{[x]\oplus \mu,y}\\
\mathsf D(\langle f,g\rangle)_{[x]\oplus\mu, \langle 1,y\rangle}& = 
(\langle f,g\rangle)_{\mu+x,\langle 1,y\rangle}= g_{\mu+x,y}= \mathsf D(g)_{[x]\oplus \mu,y}
\end{align*}
from which we deduce
$\mathsf D(\langle f,g\rangle)_{[x]\oplus\mu, \langle i,y\rangle}=\langle \mathsf D(f),\mathsf D(g)\rangle_{[x]\oplus\mu, \langle i,y\rangle}$ 
 by the definition of $\langle \_,\_\rangle$.
 If $\rho\neq [x]\oplus\mu$, then
 $\mathsf D(\langle f,g\rangle)_{\rho, \langle i,y\rangle}=\infty=\langle \infty,\infty\rangle=\langle \mathsf D(f),\mathsf D(g)\rangle_{\rho, \langle i,y\rangle}$
 (where the equation $\infty=\langle \infty,\infty\rangle$ is to be read as an equality between the functions $ X+Y\longrightarrow Q$
 defined by $\langle i,y\rangle \mapsto \infty$ and by
 $\begin{matrix}
 \langle 0,x\rangle\mapsto\infty\\
  \langle 1,y\rangle\mapsto\infty
 \end{matrix}$, respectively).

\item[(5)] $\mathsf D(f\circ g)=\mathsf D(f)\circ \langle \mathsf D(g), g\circ \pi_{2}\rangle$: we can compute
\begin{align*}
\Big( \mathsf D(f)\circ \langle \mathsf D(g), g\circ \pi_{2}\rangle\Big)_{[x]\oplus\mu,z}
&=
\inf \Big\{
\mathsf D(g)_{[x]\oplus\mu',w}+
\sum_{i}g_{\mu_{i},w_{i}}+
\mathsf D(f)_{[w]\oplus [w_{1},\dots, w_{n}],z}\\
&\qquad\qquad \mid
w,w_{i}\in Y, 
 \mu=\mu'+ \sum_{i}\mu_{i},
\Big\}\\
&=
\inf \Big\{
g_{\mu'+x,w}+
\sum_{i}g_{\mu_{i},w_{i}}+
f_{ [w_{1},\dots, w_{n}]+w,z}\\
&\qquad\qquad \mid
w,w_{i}\in Y, 
 \mu= \mu'+\sum_{i}\mu_{i}
\Big\}
\\
&=\inf\Big\{
\sum_{i} g_{\mu_{i},w_{i}} + f_{[w_{1},\dots, w_{n}],z}
\mid
w_{1},\dots, w_{n}\in Y, 
\mu+x=\sum_{i}\mu_{i}
\Big\}
\\
&= (f\circ g)_{\mu+x,y} =\mathsf D(f\circ g)_{[x]\oplus\mu,z}
\end{align*}
if $\rho\neq [x]\oplus \mu$, then $\mathsf D(f\circ g)_{\rho,z}=\infty$ and 
from the first equation above it follows that 
also $( \mathsf D(f)\circ \langle \mathsf D(g), g\circ \pi_{2}\rangle)_{\rho,z}=\infty$.

\item[(6)] $\mathsf D(\mathsf D(f))\circ \langle \langle g,\infty\rangle,\langle h,k\rangle\rangle=\mathsf D(f)\circ  \langle g,k\rangle$:
observe that 
\begin{align*}
\Big(\mathsf D(\mathsf D(f))\Big)_{[\langle 1,x'\rangle]\oplus([x]\oplus \mu),z}&=
\big(\mathsf D(f) \big)_{[x]\oplus (\mu+x'),z }= f_{\mu+x'+x,z}\\
\Big(\mathsf D(\mathsf D(f))\Big)_{[\langle 0,x\rangle]\oplus (\emptyset \oplus \mu),z}&=
\big(\mathsf D(f) \big)_{[x]\oplus \mu,z }= f_{\mu+x,z}
\end{align*}
and in all other cases $(\mathsf D(\mathsf D(f)))_{\mu,z}=\infty$.
Using this fact we can compute:

{
\small
\begin{align*}
\Big(\mathsf D(\mathsf D(f))\circ \langle \langle g,\infty\rangle,\langle h,k\rangle\rangle\Big)_{\mu,z}&=
\min\left\{
\begin{matrix}
\inf\left\{
\begin{matrix}
\infty_{\rho_{1},x'} + h_{\rho_{2},x}+ \sum_{i}k_{\mu_{i},w_{i}}
+
f_{[w_{1},\dots, w_{n}]+x'+x,z}\\
\qquad \mid 
x,x',w_{i}\in Y, 
\mu=\rho_{1}+\rho_{2}+\sum_{i}\mu_{i}
\end{matrix}
\right\},\\
\inf\left\{
\begin{matrix}
g_{\rho,x}+ \sum_{i}k_{\mu_{i},w_{i}}
+f_{[w_{1},\dots, w_{n}]+x,z}\\
\qquad \mid 
x,w_{i}\in Y, 
\mu=\rho+\sum_{i}\mu_{i}
\end{matrix}
\right\}
\end{matrix}
\right\}\\
&=
\inf\left\{
g_{\rho,x}+ \sum_{i}k_{\mu_{i},w_{i}}
+f_{[w_{1},\dots, w_{n}]+x,z}
\mid 
x,w_{i}\in Y, 
\mu=\rho+\sum_{i}\mu_{i}
\right\}\\
&= \Big(\mathsf D(f)\circ \langle g,k\rangle\Big)_{\mu,z}
\end{align*}
}

\item[(7)] $\mathsf D(\mathsf D(f))\circ \langle\langle \infty,h\rangle, \langle g,k\rangle\rangle= \mathsf D(\mathsf D(f))\circ \langle \langle \infty,g\rangle, \langle h,k\rangle\rangle$:
by computations similar to the case above we obtain

{
\small
\begin{align*}
\Big(\mathsf D(\mathsf D(f))\circ \langle \langle & \infty,h\rangle,\langle g,k\rangle\rangle\Big)_{\mu,z}\\
%&=
%\inf\left\{
%\begin{matrix}
%h_{\mu',w} + g_{\mu'',x}+ \sum_{i}k_{\mu_{i},w_{i}}
%+
%f_{[w,w_{1},\dots, w_{n}]+x,z}\\
%\infty+ \sum_{i}k_{\mu_{i},w_{i}}
%+f_{[w,w_{1},\dots, w_{n}],z}
%\end{matrix}
%\mid 
%w,w_{i}\in Y, 
%\mu=\mu'+\mu''+\sum_{i}\mu_{i}
%\right\}\\
&=
\inf\left\{
h_{\rho',x'} + g_{\rho,x}+ \sum_{i}k_{\mu_{i},w_{i}}
+
f_{[w_{1},\dots, w_{n}]+x'+x,z}
\mid 
x,x',w_{i}\in Y, 
\mu=\rho'+\rho+\sum_{i}\mu_{i}
\right\}\\
&=
\inf\left\{
g_{\rho,x} + h_{\rho',x'}+ \sum_{i}k_{\mu_{i},w_{i}}
+
f_{[w_{1},\dots, w_{n}]+x+x',z}
\mid 
x,x',w_{i}\in Y, 
\mu=\rho+\rho'+\sum_{i}\mu_{i}
\right\}\\
&=
\Big(\mathsf D(\mathsf D(f))\circ \langle \langle\infty,g\rangle,\langle h,k\rangle\rangle\Big)_{\mu,z}
\end{align*}
}

\item[(D-curry)] $\mathsf D(\Lambda(f))=
\Lambda(\mathsf D(f)\circ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle)$:
by observing that both morphisms are 
in $Q^{\Pi}_{!}(X+X, Z^{Y})= Q^{\C M_{\mathsf f}(X+X)\times \C M_{\mathsf f}(Y)\times Z}$, 
and that $ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle
\in Q^{\Pi}_{!}( (X+X)+Y  , (X+Y)+(X+Y) )$,
we can compute:

{\small
\begin{align*}
\big(\Lambda(\mathsf D(f)\circ \langle&  \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle)\big)_{[x]\oplus\mu,\nu,z}\\
&=
\big(\mathsf D(f)\circ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle\big)_{([x]\oplus\mu)\oplus\nu, z}\\
&=
\inf\left\{
( \pi_{1})_{[x]\oplus \emptyset,x}+
\sum_{i}(\pi_{2})_{\emptyset\oplus [w_{i}],w_{i}}
+
\sum_{j}(\mathrm{id})_{[z_{j}],z_{j}}
 + \mathsf D(f)_{([x]\oplus \emptyset)\oplus(\mu\oplus \nu)}
\mid
\begin{matrix}
\mu=[w_{1},\dots, w_{n}],\\
\nu=[z_{1},\dots, z_{m}]
\end{matrix}
\right\}\\
&=
\inf\left\{
0+0
+
0 + \mathsf D(f)_{([x]\oplus \emptyset)\oplus(\mu\oplus \nu)}
\mid
\begin{matrix}
\mu=[w_{1},\dots, w_{n}],\\
\nu=[z_{1},\dots, z_{m}]
\end{matrix}
\right\}\\
&=
\big(\mathsf D(f)\big)_{ ([x]\oplus\emptyset)\oplus(\mu\oplus\nu),z}\\
&=
f_{(\mu+x)\oplus\nu, z}\\
& =
\big(\Lambda(f)\big)_{\mu+x,\nu,z}=\big(\mathsf D(\Lambda(f))\big)_{[x]\oplus\mu, \nu,z}
\end{align*}
}

If $\rho\neq [x]\oplus\mu$, then $(\mathsf D(\Lambda(f)))_{\rho,\nu,z}=\infty$ and $
(\Lambda(\mathsf D(f)\circ \langle  \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle))_{\rho,\nu,z}=
(\mathsf D(f)\circ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle)_{\rho\oplus\nu, z}
$, and one can check that also this is $\infty$, using the second equation above and the fact that $(\pi_{1})_{\rho\oplus\emptyset,x}=\infty$.\end{description}

\end{proof}

\paragraph{Taylor Expansion }

To check the validity of Taylor expansion we must further check the following equation, given $f\in Q^{\Pi}_{!}(C, B^{A})$ and $g\in Q^{\Pi}_{!}(C,A)$:
$$
\mathsf{ev}\circ \langle f,g\rangle= \inf_{n\in \BB N}
(( \cdots (\Lambda^{-}(f) \underbrace{\star g)\cdots )\star g}_{n\text{ times}})\circ \langle \mathrm{id}, \infty\rangle
$$
where:
\begin{enumerate}
\item $\mathsf{ev}\in Q^{\Pi}_{!}(B^{A}+A, B)$ is the canonical \emph{evaluation} morphism;

\item $\Lambda^{-}(\_):= \mathsf{ev}\circ (\_\times \mathrm{id})$ is the \emph{uncurry} operator;

\item given $f\in Q^{\Pi}_{!}(C+A,B)$ and $g\in Q^{\Pi}_{!}(C,A)$, 
$f\star g\in Q^{\Pi}_{!}(C+A,B)$ is the morphism obtained by differentiating $f$ in its second component and applying $g$ in that component, i.e.~
$$
f\star g =  \mathsf{D}(f)\circ \langle \langle \infty, g\circ \pi_{1}\rangle, \mathrm{id}_{C+A}\rangle.
$$

\end{enumerate}

Let us first compute the three morphisms $\mathsf{ev}, \Lambda^{-}$ and $\star$ explicitly:
\begin{enumerate}

\item $\mathsf{ev}\in Q^{ \C M_{\mathsf f}(  ( \C M_{\mathsf f}(A)\times B)       +  A    ) \times B  }$ is given by
$$\mathsf{ev}_{\mu,y}=
 \begin{cases}
 0 & \text{ if } \mu=[ \langle\rho, y\rangle]  \oplus \rho \\
 \infty & \text{ otherwise}
 \end{cases}
 $$
and observe that, given $f\in Q^{\Pi}_{!}(C, B^{A})$ and $g\in Q^{\Pi}_{!}(C,A)$, 
$$
\big(\mathsf{ev}\circ \langle f,g\rangle \big)_{\chi, y}= 
\inf\Big \{ 
\sum_{i=1}^{m}g_{\chi_{i},x_{i}}+
f_{\chi', \langle [x_{1},\dots, x_{m}],y\rangle}
\mid 
x_{1},\dots, x_{m}\in A,
\chi= \chi'+\sum_{i=1}^{m}\chi_{i}, 
\Big \}
$$

\item given $g\in Q^{\Pi}_{!}(C, B^{A})$, 
$\Lambda^{-}(g)\in Q^{\Pi}_{!}(C+A, B)$ is given by 
$$
\big(\Lambda^{-}(g)\big)_{\rho\oplus\mu,y}=g_{\rho, \langle \mu,y\rangle}
$$

\item $f\star g$ is given by 
$$
(f\star g)_{\rho\oplus\mu,y}=
\inf\Big\{
g_{\rho',x}+
f_{\rho''\oplus(\mu+x)}
\mid
x\in A,
\rho= \rho'+\rho''
\Big\}
$$

\end{enumerate}

Given the definition of $\mathsf{ev}\circ \langle f,g\rangle$, to check the Taylor equation it is enough to check that, for all $N\in \BB N$, 
$$
\left((( \cdots (\Lambda^{-}(f) \underbrace{\star g)\cdots )\star g}_{N\text{ times}})\circ \langle \mathrm{id}, \infty\rangle\right)_{\chi,y}=
\inf\Big \{ 
\sum_{i=1}^{N}g_{\chi_{i},x_{i}}+
f_{\chi', \langle [x_{1},\dots, x_{N}],y\rangle}
\mid 
\begin{matrix}
x_{1},\dots, x_{N}\in A,\\
\chi= \chi'+\sum_{i=1}^{N}\chi_{i}
\end{matrix}
\Big \}
$$

Let us show, by induction on $N$, the following equality, from which the desired equality easily descends:
$$
\big(( \cdots (\Lambda^{-}(f) \underbrace{\star g)\cdots )\star g}_{N\text{ times}}\big)_{\chi\oplus\mu,y}=
\inf\Big \{ 
\sum_{i=1}^{N}g_{\chi_{i},x_{i}}+
f_{\chi', \langle\mu+ [x_{1},\dots, x_{N}],y\rangle}
\mid 
\begin{matrix}
x_{1},\dots, x_{N}\in A,\\
\chi= \chi'+\sum_{i=1}^{N}\chi_{i}
\end{matrix}
\Big \}
$$
\begin{itemize}

\item if $N=0$, the right-hand term reduces to 
$f_{\chi, \langle \mu, y\rangle}=(\Lambda^{-}(f))_{\chi\oplus\mu,y}$;

\item otherwise, let $F=(( \cdots (\Lambda^{-}(f) \underbrace{\star g)\cdots )\star g}_{N-1\text{ times}})$, so that by I.H.~we have
$$
F_{\chi\oplus\mu,y}=
\inf\Big \{ 
\sum_{i=1}^{N-1}g_{\chi_{i},x_{i}}+
f_{\chi', \langle\mu+ [x_{1},\dots, x_{N-1}],y\rangle}
\mid 
\begin{matrix}
x_{1},\dots, x_{N-1}\in A,\\
\chi= \chi'+\sum_{i=1}^{N-1}\chi_{i}
\end{matrix}
\Big \}
$$
Then we have
{\small
\begin{align*}
\big( F\star g\big)_{\chi\oplus\mu,y}
&=
\inf \left \{
g_{\chi',x}+F_{\chi''\oplus(\mu+x)}
\mid
x\in A, \chi=\chi'+\chi''
\right\}\\
&=
\inf\left \{ 
g_{\chi',x}+
\inf\left\{
\sum_{i=1}^{N-1}g_{\chi_{i},x_{i}}+
f_{\chi'', \langle\mu+ [x_{1},\dots, x_{N-1}]+x,y\rangle}
\mid 
\begin{matrix}
x_{1},\dots, x_{N-1}\in A,\\
\chi^{*}= \chi''+\sum_{i=1}^{N-1}\chi_{i}
\end{matrix}
\right\}
\ 
\Big\vert \ 
\begin{matrix}
x\in A,\\
\chi=\chi'+\chi^{*}
\end{matrix}
\right \}\\
&=
\inf\Big \{ 
g_{\chi',x}+
\sum_{i=1}^{N-1}g_{\chi_{i},x_{i}}+
f_{\chi'', \langle\mu+ [x_{1},\dots, x_{N-1}]+x,y\rangle}
\mid 
\begin{matrix}
x,x_{1},\dots, x_{N-1}\in A,\\
\chi= \chi'+\chi''+\sum_{i=1}^{N-1}\chi_{i}
\end{matrix}
\Big \}\\
&=
\inf\Big \{ 
\sum_{i=1}^{N}g_{\chi_{i},x_{i}}+
f_{\chi', \langle\mu+ [x_{1},\dots, x_{N}],y\rangle}
\mid 
\begin{matrix}
x_{1},\dots, x_{N}\in A,\\
\chi= \chi'+\sum_{i=1}^{N}\chi_{i}
\end{matrix}
\Big \}.
\end{align*}
}
\end{itemize}

### 2023-01-19 appV.tex
commit

\subsubsection{Proof of Theorem \ref{theorem:fepsilon}}

Also proof of 
\begin{proposition}\label{prop:nondecr+conc}
 Any tLs $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
\end{proposition}

\subsubsection{Continuity of tLs}

Continuity of cones and metric

\subsubsection{Lipschitz continuity}

\begin{proposition}\label{prop:troplinear}
All tropical linear functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive.  
\end{proposition}

\begin{lemma}
Let $f: [0,\infty)^{X}\to Q$ be concave, monotone increasing, and continuous.  
Let $\B x\neq \B y\in [0,\infty)^{X}$, with $\|\B x-\B y\|_{\infty}<\infty$, and let $S(\B x, \B y)=\{ \alpha\B x + (1-\alpha)\B y\mid \alpha \in [0,1]\}$ be the segment generated by $\B x$ and $\B y$. 
Then $f$ is Lipschitz-continuous over $S(\B x, \B y)$.

\end{lemma}
\begin{proof}
Let us prove the lemma under the assumption that for all $a\in X$, $\B y_{a}-\B x_{a}\geq 1$. From the fact that the claim holds under the assumption, we can deduce the claim of the lemma:
indeed for $\alpha\in (0,1)$ large enough we have that $\B y':= \frac{\B y-\alpha\B x}{1-\alpha}$ is such that $\B y\in S(\B x, \B y')$ (and thus $S(\B x, \B y)\subseteq S(\B x, \B y')$) and 
$\B y'_{a}-\B x_{a}\geq 1$. Hence from our proof we deduce that $f$ is Lipschitz-continuous over $S(\B x, \B y')$, and thus a fortiori over $S(\B x, \B y)$ too.

Since $f$ is continuous over $[0,\infty)^{X}$ and $S(\B x, \B y)$ is compact, $f$ admits a maximum $\mathrm{MAX}$ over $S(\B x, \B y)$.
For all $\B z< \B z'\in S(\B x, \B y)$, let $M(\B z, \B z')\in Q^{X}$ be defined by
$$
M(\B z, \B z')_{a}= \frac{f(\B z')-f(\B z)}{\B z'_{a}-\B z_{a}}
$$
Observe that
\begin{align*}
M(\B x, \B y)_{a} & = \frac{f(\B y)-f(\B x)}{\B y_{a}-\B x_{a}} \leq 
f(\B y)-f(\B x) \leq \mathrm{MAX}
\end{align*}
using the fact that $\B y_{a}-\B x_{a}\geq 1$.

We now claim that $M(\B z, \B z')$ is contravariant in both $\B z$ and $\B z'$. 
Indeed suppose $\B z\leq \B z'' < \B z'$, so that $\B z= \lambda \B z' +(1-\lambda)\B z''$ for some $\lambda \in (0,1)$. Then, using the fact that $f$ is concave, we have 
\begin{align*}
M(\B z, \B z')_{a}&=\frac{f(\B z')-f(\lambda \B x +(1-\lambda)\B z'')}{\B z'_{a}-\lambda \B x_{a} -(1-\lambda)\B z''_{a}} \\
&
\geq 
\frac{f(\B z')-\lambda f(\B z') -(1-\lambda)f(\B z'')}{\B z'_{a}-\lambda \B z'_{a} -(1-\lambda)\B z''_{a}} \\
&
= 
\frac{(1-\lambda)(f(\B z')-f(\B z''))}{(1-\lambda )\B z'_{a}-\B z''_{a}} =M(\B z'', \B z')
\end{align*}
In a similar way it is shown that for $\B z < \B z''\leq \B z'$, $M(\B z,\B z')\leq M(\B z, \B z'')$.

Therefore, for all $\B z < \B z'\in S(\B x, \B y)$, we have that $M(\B z, \B z')_{a} \leq M(\B x, \B z')_{a} \leq M(\B x, \B y)_{a} \leq \mathrm{MAX}$. 
From this, using the fact that $f$ is monotone increasing, we deduce that 
$|f(\B z')-f(\B z)|=f(\B z')-f(\B z) \leq \mathrm{MAX}\cdot |\B z'_{a}-\B z_{a}|
$ and thus that 
$$|f(\B z')-f(\B z)|\leq \mathrm{MAX}\cdot \|\B z'-\B z\|_{\infty}$$
that is, that $f$ is $\mathrm{MAX}$-Lipschitz over $S(\B x, \B y)$. 
\end{proof}

\begin{proposition}
Let $f: [0,\infty)^{X}\to Q$ be concave, monotone increasing, and continuous.  
For all $\epsilon \in (0,\infty)$ and $\B x\in [0,\infty)^{X}$, $f$ is Lipschitz-continuous over the open ball $B(\B x, \epsilon)$.
\end{proposition}
\begin{proof}
Let $\mathrm{MAX}$ indicate the maximum of $f$ over $B(\B x, \epsilon)$. 
Let $\B y, \B z\in B(\B x, \epsilon)$; then $\|\B y-\B z\|_{\infty}\leq 2\epsilon < \infty$, so by the lemma above $f$ is $K$-Lipschitz over the segment $S(\B y, \B z)$ for some $K\leq \mathrm{MAX}$, so we deduce $|f(\B y)-f(\B z)|\leq \mathrm{MAX}\cdot \|\B y-\B z\|_{\infty}$.
\end{proof}

\begin{theorem}[local Lipschitz-continuity]
Let $f: [0,\infty)^{X}\to Q$ be concave, monotone increasing, and continuous.  
Then $f$ is locally Lipschitz-continuous.
\end{theorem}
\begin{proof}
For all $\B x\in [0,\infty)^{X}$, $f$ is Lipschitz-continuous over the open set $B(\B x, 1)$.
\end{proof}

### 2023-01-19 appVII.tex
commit

In the following we let $Q$ indicate the Lawvere quantale $([0,\infty], \geq, +,0)$. 
$Q$ can be seen as:
\begin{itemize}
\item a category $Q_{0}$ with a unique object and morphisms being the elements of $Q$, with identity 0 and composition $+$;

\item a degenerate category $Q_{1}$ with objects the elements of $Q$, and  
a morphism between $x$ and $y$ whenever $x\geq y$;

\item a 2-category $Q_{2}$ with 0-cells and 1-cells as in $Q_{0}$, and 2-cells the morphisms of $Q_{1}$.

\end{itemize}
These categorical formulations of $Q$ will help us here and there.

We let $x-y$ indicate truncated subtraction. Notice that 
$x \leq (x-y)+y$ and $(x+y)-y\leq x$. This corresponds to the fact that ``$\_-y$'', as a functor from $
Q_{1}$ to itself, 
 is \emph{right-adjoint} to the functor ``$\_+y$''. Equivalently, 
 $x+y \geq z $ iff $x\geq z-y$.

$Q\mathsf{Rel}$ is the category of sets and $Q$-relations, where a $Q$-relation 
$R:X\pfun Y$ is a function $R: X\times Y\to Q$, with identity $1_{X}:X\pfun X$ given by 
$$1_{X}(x,y)=\begin{cases} 0 & \text{ if }x=y\\ \infty & \text{ otherwise}\end{cases}$$
and composition of $R:X\pfun Y$ and $S:Y\pfun Z$ given by 
$$
(S\cdot R)(x,z)= \inf_{y\in Y}R(x,y)+S(y,z)
$$

\begin{remark}
$Q\mathsf{Rel}$ is also a (degenerate) 2-category, where for $R,S:X\pfun Y$, $R\To S$ iff for all $x\in X$, $y\in Y$, $R(x,y)\geq S(x,y)$.

\end{remark}

Composition with $R:X\pfun Y$ in Q$\mathsf{Rel}$ yields a functor
$$
R\cdot \_ : Q\mathsf{Rel}(A,X)   \longrightarrow Q\mathsf{Rel}(A,Y)
$$
which
has a right-adjoint
$$
\_ \multimapinv R : Q\mathsf{Rel}(A,Y)   \longrightarrow Q\mathsf{Rel}(A,X)
$$
given, for $S:A\pfun Y$ by 
\begin{align*}
(S\multimapinv R)(a,x)= \sup_{y\in Y}S(a,y) -R(x,y)
%(R\multimap S)(a,x)= \sup_{y\in Y}
\end{align*}

%
%\begin{lemma}
%\begin{itemize}
%\item[i.] $ (R\multimapinv T)\cdot
%(T\multimapinv S)\leq R \multimapinv S $.
%
%\item[ii.] $R\multimapinv T \leq (R\multimapinv S)\multimapinv (T\multimapinv S)$.
%\end{itemize}
%\end{lemma}
%\begin{proof}
%\begin{align*}
%(R\multimapinv S)(x,y)&= \sup_{z}R(x,z)-S(y,z) \\
%&=\sup_{z} (R(x,z)-T(z,z))+(T(z,z)-S(y,z))\\
%&= \inf_{w}\left(\sup_{z}R(x,z)-T(w,z)\right) +\left(\sup_{z'}T(w,z')-S(z',y)\right)
%\\
%&= \inf_{w}(R\multimapinv T)(x,w)+(T\multimapinv S)(w,y)  
%\\
%&= (R\multimapinv T)\cdot (T\multimapinv S)(x,y)
%\end{align*}
%
%\end{proof}

\subsubsection{$Q$-categories.}

A \emph{$Q$-category}, or \emph{generalized (quasi-)metric space} is a small $Q$-enriched category. Concretely, it is given by a pair $(X,q)$ made of a set $X$ and a $Q$-relation $q: X\pfun X$ satisfying:
\begin{align}
q(x,x) & = 0 \tag{Q1} \\
q(x,z)+q(z,y) & \geq q(x,y) \tag{Q2}
\end{align}
A $Q$-category is \emph{skeletal} if it further satisfies
\begin{align}
q(x,y) & = 0 \ \To \ x=y \tag{Q3}
 \end{align}
 and \emph{symmetric} if it satisfies
 \begin{align}
q(x,y) & = q(y,x) \tag{Q4}
 \end{align}
For simplicity, in the following we indicate the distance function of a $Q$-category $X$ simply as $X(x,y)$.

\begin{remark}
In more abstract terms, a $Q$-category can be identified with a \emph{monad} in $Q\mathsf{Rel}$, seen as a 2-category: it is given by a 1-cell $q:X\pfun X$ together with 2-cells $1_{X}\To  q$ and $q\cdot q \To q$. 
\end{remark}

We let $x\simeq y$ indicate the equivalence relation defined by  $X(x,y)=X(x',y)$ holds for all $y\in X$. 
Notice that $x\simeq y$ coincides with $x=y$ precisely when $X$ is skeletal.

Every $Q$-category $X$ is also a preorder $(X,\preceq_{X})$, with $x\preceq_{X}y$ iff $X(x,y)=0$. Moreover, if $X$ is skeletal, then $\preceq_{X}$ is an order.

For every $Q$-category $X$, the category $X^{\mathrm{op}}$ has same objects as $X$, and $X^{\mathrm{op}}(x,y)=X(y,x)$.

\begin{example}
$Q$ is a $Q$-category with the distance $Q(x,y)=y-x$. 
The category $Q^{\mathrm{op}}$ has distance $Q^{\mathrm{op}}(x,y)=Q(y,x)$.
\end{example}

\begin{example}
For every $Q$-category $X$, the distance $\C S(X)(x,y)=\max\{X(x,y),X^{\mathrm{op}}(x,y)\}$ defines a symmetric $Q$-category $\C S(X)$. In the case of $Q$, $\C S(Q)$ is just the Euclidean distance.
\end{example}

A \emph{functor} between $Q$-categories $X$ and $Y$ is just a non-expansive function $f:X\to Y$, that is, satisfying $Y(f(x),f(y))\leq X(x,y)$.

%
\begin{remark}\label{rem:functor}
Given $f:X\to Q$, $f$ is a functor precisely when for all $x,y\in X$ $f(x)\leq\inf_{x'\in X}f(x')+X(x',x)$. 
Indeed, if $f$ is a functor then $f(x)\leq f(x')+X(x',x)$, since $f(x)-f(x')=Q(f(x'),f(x))\leq X(x',x)$. 
Conversely, if $f(x)\leq f(x')+X(x',x)$ holds for all $x'$, then 
$Q(f(x),f(x'))=f(x')-f(x)\leq X(x',x)$. 
\end{remark}

%Indeed, being a functor means that any arrow $X(x,y)$ is mapped into an arrow $Y(f(x),f(y))$, that is, $Y(f(x),f(y))

\begin{example}
For every $Q$-category $X$, the $Q$-category $[X,Q]$ has objects all functors from $X$ to $Q$ (where the latter is seen as a $Q$-category), and $[X,Q](f,g)=\sup_{x\in X}Q(f(x),g(x))$.

When $X$ has the discrete metric, $[X,Q]=Q^{X}$ and $\C S([X,Q])(f,g)=\| f-g\|_{\infty}$.

\end{example}

\begin{example}
For every $Q$-categories $X,Y$ we can define:
\begin{itemize}
\item a $Q$-category $X\otimes Y$ with 
$(X\otimes Y)(\langle x,y\rangle, \langle x',y'\rangle)=X(x,x')+Y(y,y')$;
\item a $Q$-category $X\times Y$ with 
$(X\times Y)(\langle x,y\rangle, \langle x',y'\rangle)=\max\{X(x,x'),Y(y,y')\}$. 
\end{itemize}
\end{example}

\subsubsection{$Q$-distributors.}

%We let $Q\mathsf{Cont}$ indicate the category of $Q$-categories and continuous functors.

A \emph{$Q$-distributor} $\Phi: X\pfun Y$ is a $Q$-relation $\Phi: X\times Y\to Q$ satisfying
\begin{align*}
\Phi\cdot X   \leq\Phi \qquad \qquad
Y \cdot \Phi  \leq \Phi
\end{align*}

\begin{remark}
Using Remark \ref{rem:functor}, it follows that $\Phi$ is a $Q$-distributor precisely when it is a functor $\Phi: X^{\mathrm{op}}\otimes Y \to Q$:
the condition $\Phi(x,y) \leq X(x,x')+\Phi(x',y)$ implies that $\Phi$ is a functor over $X^{\mathrm{op}}$, while
the condition $\Phi(x,y) \leq \Phi(x,x')+Y(x',y)$ implies that $\Phi$ is a functor over $Y$.
But this implies then 
$\Phi(x,y) \leq X(x,x') +\Phi(x',y')+Y(y',y)$ and thus $Q(\Phi(x',y'),\Phi(x,y))\leq X(x,x')+Y(y',y)$.
\end{remark}

%where $(\Phi\cdot X)_{x,y}= \inf_{z\in X}\Phi_{x,z}+X(z,y)$ and 
%$(y\cdot \Phi)_{x,y}= \inf_{z\in X}Y(x,z)+\Phi_{z,y}$.

The identity distributor $1_{X}:X\pfun X$ is given by $(1_{X})_{x,y}=X(x,y)$. 
Distributors $\Phi:X\pfun Y$ and $\Psi:Y\pfun Z$ compose via $\Psi\cdot \Phi:X\pfun Z$. 
We let $Q\mathsf{Dist}$ indicate the category of $Q$-categories and $Q$-distributors.

\begin{remark}
Usually distributors $\Phi: X\pfun Y$ are presented as $Q$-relations $\Phi:Y\times X\to Q$ (notice the inversion of $X$ and $Y$), that is, as functors $\Phi: X\times Y^{\mathrm{op}}\to Q$. We chose to invert the presentation of distributors for uniformity with the usual notations in models of the differential $\lambda$-calculus. 
\end{remark}

Any functor $f:X\to Y$ induces two adjoint distributors $f^{\circ}:X\pfun Y$ and $f_{\circ}:Y\pfun X$ given by $f^{\circ}(x,y)=Y(f(x),y)$ and $f_{\circ}(y,x)=Y(f(y),x)$.
Here adjoint means that $1_{X}\leq f_{\circ}\cdot f^{\circ} $ and $f^{\circ}\cdot f_{\circ}\geq 1_{Y}$.

\begin{remark}[Yoneda embedding]
%$Q$ is a $Q$-category with $Q(x,y)=|x-y|$. 
%For any $Q$-category $X$ we can define a new $Q$-category $\mathsf{Dist}(X,Q)$ (that we note simply as $[X,Q]$) whose objects are the distributors $\B x:\B 1\pfun X$, or equivalently, the functors from $X$ to $Q$, i.e.~those $\B x\in Q^{X}$ such that $|\B x_{a}-\B x_{b}|\leq X(a,b)$, or equivalently $\B x_{a}+X(a,b) \geq \B x_{b}$.

The \emph{Yoneda embedding} is the faithful functor $\C Y: X\to [X,Q]$ given by $\C Y(x)(y)=X(y,x)$. The functoriality and faithfulness of $\C Y$ follow from
\begin{align}
[X,Q](\C Y(x),\C Y(x'))= X(x,x') \tag{Yoneda}
\end{align}
which is proved as follows: for all $y\in X$ we have 
\begin{align*}
Q( \C Y(x)(y),\C Y(x')(y))&= \C Y(x')(y)-\C Y(x)(y) \\
&= X(y,x')-X(y,x) \leq X(x,x')
\end{align*}
where the last step follows from $X(y,x)+X(x,x')\geq X(y,x')$. 
From this we deduce that $[X,Q](\C Y(x),\C Y(x'))=\sup_{y\in X}Q( \C Y(x)(y),\C Y(x')(y))\leq X(x,x')$. 
For the converse direction, we have  
\begin{align*}
X(x,x') &  = X(x,x') - 0 \\ &=
X(x,x')- X(x,x)
\\& =\C Y(x')(x)-\C Y(x)(x)= Q(\C Y(x)(x),\C Y(x')(x))\\
&\leq [X,Q](\C Y(x), \C Y(x'))
\end{align*}
\end{remark}

\begin{remark}
The \emph{opposite Yoneda embedding} is the faithful functor
$\C Y^{\mathrm{op}}: X\to [X,Q]^{\mathrm{op}}$ given by $\C Y^{\mathrm{op}}(x)(y)=X(x,y)$. The functoriality and faithfulness of $\C Y^{\mathrm{op}}$ follow from
\begin{align}\label{eq:yonedaop}
[X,Q](\C Y^{\mathrm{op}}(x),\C Y^{\mathrm{op}}(x'))= X(x',x) \tag{Yoneda$^{\mathrm{op}}$}
\end{align}
which is proved similarly to the case of $\C Y$.
\end{remark}

\subsubsection{cocompleteness}

Let $\Phi:Y\pfun Z$ be a distributor and $f:Y\to X$ be a functor.
A functor $ g:Z\to X$ is called the \emph{$\Phi$-weighted colimit of $f$} if it satisfies, for all
$z\in Z$ and 
 $x\in X$, 
$$
X(g(z), x) \ = \  \sup_{y\in Y}X(f(y),x) - \Phi(y,z)
$$
If this colimit exists, we write it as $\mathrm{colim}(\Phi,f)$.

A $Q$-category $X$ is said \emph{cocomplete} if it admits all weighted colimits, and a functor $f:X\to Y$ of $Q$-categories is said \emph{cocontinuous} if it commutes with all weighted colimits, meaning that $f \circ \mathrm{colim}(\Phi,g)= \mathrm{colim}(\Phi, f\circ g)$.

We let $Q\mathsf{CCat}$ indicate the category of skeletal and cocomplete $Q$-categories and cocontinuous functors.

\begin{proposition}\label{prop:yonedasup}
Let $X$ be a $Q$-category. Then $X$ is cocomplete iff the Yoneda embedding has a left-adjoint.
\end{proposition}
\begin{proof}
 For all $\B x\in [X,Q]$, let $\sup\B x$ be defined as a weighted colimit via
$$
X(\sup\B x, y)= \sup_{z\in X} X(z,y)-\B x_{z}
$$
that is, $\sup \B x= \mathrm{colim}( \B x, \mathrm{id}_{X})$, where $\B x$ is seen as a distributor $\B x:\B 1\pfun X$.

Let us check that $\sup: [X,Q]\to X$ is a functor. First, let us check the  inequality
\begin{align}\label{eq:in1}
(X\multimapinv\B y)\cdot (\B y\multimapinv\B x)\geq (X\multimapinv\B x)\end{align}
as follows:
\begin{align*}
\left((X\multimapinv \B y)\cdot (\B y\multimapinv\B x)\right)(a)&=
 \left (\sup_{b}X(b,a)-\B y_{b}\right)+\left(\sup_{b}\B y_{b}-\B x_{b}\right)\\
&\geq
\sup_{b}(X(b,a)-\B y_{b})+(\B y_{b}-\B x_{b})
\\
&=\sup_{b}(X(b,a)-\B y_{b}+\B y_{b})-\B x_{b}
\\
&=\sup_{b} X(b,a)-\B x_{a}\\
&= (X\multimapinv \B x)(a)
\end{align*}
From \eqref{eq:in1} we deduce immediately the inequality below:
\begin{align}\label{eq:in2}
\B y\multimapinv \B x \geq (X\multimapinv \B x)\multimapinv(X\multimapinv\B y)
\end{align}
and we can now compute:
\begin{align*}
[X,Q](\B x, \B y)&= \sup_{a\in X}\B y_{a}-\B x_{a}\\
 &\stackrel{\tiny\eqref{eq:in2}}{\geq}
 \sup_{a\in X} \left( \sup_{b\in X}X(b,a)-\B x_{a}\right) -\left(  \sup_{b\in X}X(b,a)-\B y_{a}   \right )\\
 &= \sup_{a\in X}X(\sup \B x, a)-X(\sup \B y, a)\\
 &= [X,Q](\C Y^{\mathrm{op}}(\sup\B y),\C Y^{\mathrm{op}}(\sup \B x))
 \tag{\ref{eq:yonedaop}}
 \\
&=X(\sup \B x, \sup \B y)
\end{align*}

Then for all $x\in X$, $\sup \C Y(x)\simeq x$. Indeed we have 
\begin{align*}
X(\sup \C Y(x), y) & = \sup_{z\in X}X(z,y) -\C Y(x)(z) \\
& = \sup_{z\in X}X(z,y)-X(z,x)\\
&= X(x,y)
\end{align*}
%where we use the fact that from $X(x,y)+X(z,x) \geq X(z,y)$ it follows $X(x,y) \geq X(z,y)-X(z,x)$ and thus $X(x,y) \geq \sup_{z}X(z,y)-X(x,z)$, and conversely, from $X(x,y)-X(x,x)=X(x,y)$ it follows
%%$X(x,y) \leq \sup_{z}X(z,y)-X(x,z)$.
%

Moreover, for all $\B x\in [X,Q]$, we have $\C Y(\sup \B x)\geq \B x$:  
\begin{align*}
\C Y(\sup \B x)(a) &= X(a,\sup \B x)\\
&= X(  \sup(\C Y(a)),\sup \B x) \\
& \geq [X,Q]( \C Y(a), \B x)\\
&\geq\B x_{a}-\C Y(a)(a) \\
&=
 \B x_{a}-X(a,a)  = \B x_{a}
\end{align*}
%where in the last step we use the fact that 
%$[X,Q](  \C Y(a),\B x)= \sup_{b}\B x_{b}-X(a,b) = \B x_{a}$.
%
%$\B y\in Q^{X}$, 
% $\| \C Y(\sup\B x)- \B y\|_{\infty}=\| \C Y(\sup\B x)- \B x\|_{\infty}$. Indeed we have 
% \begin{align*}
% \| \B x_{a}-\C Y(\sup\B x)\|_{\infty}&= \sup_{a\in X}|
%\B x_{a}-\C Y(\sup\B x)(a)|\\ 
%&=
%\sup_{a\in X}|
%\B x_{a}-X(\sup\B x,a)|\\
% &=
%\sup_{a\in X}\inf_{a'\in X}| \B x_{a}+X(a',a) -\B x_{a'}|\\
%&= \sup_{a\in X}|\B x_{a}+ X(a,a)- \B x_{a}| = 0
% \end{align*}
%
%

Conversely, if $\sup$ is well-defined and adjoint to $\C Y$, then 
given $\Phi: Y\pfun \B 1$ and $f:Y\to X$, 
we can define
$\mathrm{colim}(\Phi, f):= \sup\Psi$, where $\Psi=
 f^{\circ}\cdot \Phi: X\pfun \B 1$, since 
\begin{align*}
X(\sup \Psi, y)&=
\sup_{z\in X}X(z,y)-\Psi(z) \\
&=\sup_{z\in X}X(z,y)- \inf_{y\in Y}X(z,f(y))  +\Phi(y)\\
& = \sup_{z\in X}\sup_{y\in Y}X(z,y)-X(z,f(y))  -\Phi(y)\\
&=
\sup_{z\in X} X(f(z),y) - \Phi(z)
\end{align*}
\end{proof}

\begin{definition}[MacNeill Completion]
Let $X$ be a $Q$-category. For all $f: \B 1\pfun X$ and $g:X\pfun \B 1$, let $f \coh g$ iff 
$f  = X\multimapinv g $ and 
$g = f\multimap X$.

The \emph{MacNeill completion of $X$} is the $Q$-category $\B M(X)$ made of those 
$f:\B 1\pfun X$ such that $f\coh g$ for some $g:X\pfun \B 1$, with
$\B M(X)(f,f')=[X,Q](f,f')$. 
\end{definition}

Observe that if $f\coh g$, then $f= X\multimapinv (f\multimap X)$, i.e.:
\begin{align}
f(x)=  \sup_{y\in X}\inf_{z\in X}X(x,y)-X(y,z) +f(z)       
\tag{COH}
\end{align}

\begin{proposition}
Let $X$ be a $Q$-category. 
If $X$ is cocomplete, then $\C Y$ is an isomorphism between $X$ and $\B M(X)$. 
\end{proposition}
\begin{proof}
For all $x\in X$, one can check that $\C Y(x)\in \B M(X)$. Indeed, we can check that $\C Y(x) \coh \C Y^{\mathrm{op}}(x)$:
\begin{align*}
\C Y(x)(y)  &= X(y,x) \\
&=  \sup_{z\in X}X(y,z  )- X(x,z)\\
&= \sup_{z\in X}X(y,z  )- \C Y^{\mathrm{op}}(x)(z)\\
 &= (X\multimapinv \C Y^{\mathrm{op}}(x))(y)
\end{align*}
Since $\sup\C Y(x) \simeq x$ holds, it suffices to show that if $\B x \coh \B y$, then 
$\C Y(\sup \B x)=\B x$: 
\begin{align*}
\C Y(\sup \B x)(a) & \ \ = X(a, \sup\B x) \\
&\ \ = \sup_{b\in X}X(a,b)-X(\sup\B x, b) \\
&\ \ = \sup_{b\in X}\inf_{c\in X}X(a,b)- X(b,c) + \B x_{c} \\
 &\stackrel{{\tiny\text{(COH)}}}{=} \B x_{a}
\end{align*}
\end{proof}

%
%\begin{remark}[other notions of cocompleteness]
%The categorical notion of cocompleteness subsumes several other notions of cocompleteness, in the sense of being (strictly) stronger:
%\begin{itemize}
%\item \emph{order-completeness} is the case when the pre-order $\preceq_{X}$ is cocomplete.   
%Let $I\subseteq X$, so that $\mathrm{id}_{I}$ can be seen as a functor from the subcategory $I$ to $X$. Let $0_{I}: \B 1\pfun I$ be the distributor given by $(0_{I})_{i}=0$.
%Then $\mathrm{colim}(0_{I}, \mathrm{id}_{I})$ coincides with $\inf I$, via 
%$$
%X(\inf I, x) = \sup_{i\in I}X(i, x)
%$$
%Notice that in general an order-complete $Q$-category needs not be tensored, and thus it needs not be cocomplete.
%
%
%
%
% 
%
%\item \emph{Cauchy-completeness} is the case where $X$ contains the points $\sup \B x, \sup\B y$, for any 
% any two \emph{adjoint} presheaves $\B x,\B y\in [X,Q]$, i.e.~satisfying $0=\inf_{a}\B x_{a}+\B y_{a}$ and $\B x_{a}+ \B x_{b}\geq X(a,b)$. 
% Notice that given a pair $(\B x, \B y)$ we can define a Cauchy sequence by finding $a_{n}$ satisfying 
% $\B x_{a_{n}}+\B y_{a_{n}}\leq \frac{1}{n}$. Indeed this implies $X(a_{n},a_{n+1})\leq \B x_{a_{n}}+\B y_{a_{n+1}}\leq \B x_{a_{n}}+\B y_{a_{n}}+\B x_{a_{n+1}}+\B y_{a_{n+1}}=\frac{1}{n}+\frac{1}{n+1}$.
% Conversely, any (equivalence class of) Cauchy sequences $a_{n}$ yields an adjoint pair given by
% $\B x_{a}= \lim_{n}X(a_{n},a)$ and $\B y_{a}=\lim_{n}X(a,a_{n})$. 
% 
%% 
%% \item \emph{Isbell-completeness} (or \emph{MacNeill cocompleteness}) is the case where $X$ contains the points $\sup \B x, \sup \B y$, for any two presheaves $\B x, \B y\in [X,Q]$ satisfying 
%% $
%% \B x_{a}= \sup_{b\in X}X(b,a)-\B y_{b}
%% $.
%
%
%\end{itemize}
%
%\end{remark}

\subsubsection{Tensors and $Q$-Modules}

Among weighted colimits, one is of big importance for us. 
Any $\epsilon \in Q$ generates a constant distributor $(\epsilon):\B 1\pfun \B 1$, and any point $x\in X$ generates a constant functor $\Delta x: \B 1\to X$.
Given a $Q$-category $X$, a point $x\in X$ and  $\epsilon \in Q$, the \emph{tensor of $x$ and $\epsilon$}, if it exists, is defined as
$$
\epsilon \otimes x := \mathrm{colim}((\epsilon), \Delta x)
$$
A $Q$-category $X$ is \emph{tensored} if for all $x\in X$ and $\epsilon \in Q$, it admits the tensor $\epsilon \otimes x$.

\begin{proposition}
A tensored $Q$-category $X$ is a $Q$-module $(X, \preceq_{X}, \otimes)$.
A cocontinuous functor of cocomplete $Q$-categories is a $Q$-module morphism between the associated $Q$-modules.
\end{proposition}
\begin{proof}
We must show that tensors induce a continuous action. Observe that tensors are characterized by the equation 
\begin{align}\label{eq:tensor}
X(x\otimes \epsilon, x') = X(x,x') -\epsilon
\end{align}
If $\epsilon=0$, then \eqref{eq:tensor} forces $x\otimes \epsilon\simeq x$. 
If $\epsilon=\delta+\eta$, then using the fact that $\alpha-(\epsilon+\delta)=(\alpha-\epsilon)-\delta$ 
we deduce $X((x\otimes \epsilon)\otimes \delta, x')=X(x\otimes \epsilon, x')-\delta=(X(x,x')-\epsilon)-\delta=X(x,x')-(\epsilon+\delta)=X(x\otimes (\epsilon+\delta),x')$, which forces $x\otimes(\epsilon+\delta)\simeq (x\otimes \epsilon)\otimes \delta$.

A cocontinuous functor $f:X\to Y$ commutes with sups and with $\otimes$, and is thus a $Q$-module morphism.
\end{proof}

\begin{lemma}
\begin{itemize}
\item[i.] $\sup_{i\in I}a_{i}-\epsilon= (\sup_{i\in I}a_{i})-\epsilon$.
\item[ii.] $\sup_{i\in I}(a_{i}-\epsilon)-b_{i}= (\sup_{i\in I}a_{i}-b_{i})-\epsilon$.

\end{itemize}
\end{lemma}
\begin{proof}
Let $A= \sup_{i\in I}a_{i}-\epsilon$ and $B= (\sup_{i\in I}a_{i})-\epsilon$.
Let $J\subseteq I$ be the set of indexes $j$ such that $a_{j}>\epsilon$. 
If $J=\emptyset$ then $A=B=0$. Otherwise, 
$A= \sup_{j\in J}a_{j}-\epsilon$ (where ``$-$'' can be interpreted as subtraction on $\BB R$, and 
$B= (\sup_{j\in J}a_{j})-\epsilon$ (again with ``$-$'' being subtraction on $\BB R$), so $A=B$ follows from the continuity of ``$-$'' on $\BB R$.

Let now $A= \sup_{i\in I}(a_{i}-\epsilon)-b_{i}$ and $B= (\sup_{i\in I}a_{i}-b_{i})-\epsilon$.
Let $J\subseteq I$ be the set of indexes $j$ such that $a_{j}> b_{j}+\epsilon$.
If $J=\emptyset$, then $A=0$; suppose $B>0$, then $\sup_{i\in I}a_{i}-b_{i}>\epsilon$, but this implies that we can find $i\in I$ with $a_{i}>b_{i}+\epsilon$, against the assumption, so also $B=0$ holds. If $J$ is non-empty, then 
$A= \sup_{j\in J}(a_{j}-\epsilon)-b_{j}$, where ``$-$'' is not subtraction on $\BB R$ and 
$B= (\sup_{j\in J}a_{j}-b_{j})-\epsilon$, again with ``$-$'' usual subtraction, so $A=B$ follows from the continuity of ``$-$'' on $\BB R$.
 \end{proof}

\begin{lemma}
In any cocomplete $Q$-category, $x\otimes \epsilon \simeq \sup(  \C Y(x)+\epsilon  )$.
In the cocomplete $Q$-category $[X,Q]$, $\B x\otimes \epsilon= \B x+\epsilon$.
\end{lemma}
\begin{proof}
We have 
\begin{align*}
X(\sup(\C Y(x)+\epsilon), x')& =\sup_{y\in X}X(z,x')- (\C Y(x)(z)+\epsilon)\\
&= \sup_{y\in X}X(z,x')-(X(z,x)+\epsilon)\\
&= (\sup_{y\in X}X(z,x')-X(z,x))-\epsilon\\
&= X(x,x')-\epsilon
\end{align*}
which shows $x\otimes \epsilon=\sup(\C Y(x)+\epsilon)$. In $[X,Q]$ we have 
$[X,Q](\B x+\epsilon, \B x')=\sup_{a\in X}(\B x_{a}+\epsilon)-\B x'_{a}= (\sup_{a\in X}\B x_{a}-\B x'_{a})-\epsilon= [X,Q](\B x, \B x')-\epsilon$, which shows $\B x\otimes \epsilon \simeq \B x+\epsilon$, and since $[X,Q]$ is skeletal, $\B x\otimes \epsilon=\B x+\epsilon$.
\end{proof}

The dual notion of tensors is the \emph{cotensor} $x\multimapinv \epsilon$. Formally, it is defined as a \emph{weighted limit} (whose definition is dual to that of weighted colimit but we do not give details here), and characterized by the equation
\begin{align*}
X(x', x\multimapinv \epsilon)= X(x',x)-\epsilon
\end{align*}
In other words, in a tensored and cotensored $Q$-category we have $X(x\otimes \epsilon,y)= X(x,y\multimapinv \epsilon)$.

\begin{example}
The $Q$-category $[X,Q]$ is cotensored, with $\B x\multimapinv \epsilon:= \B x-\epsilon$. Indeed we have $[X,Q](\B x, \B y\multimapinv \epsilon)=\sup_{a\in X}(\B y_{a}-\epsilon)-\B x_{a}=(\sup_{a\in X}\B y_{a}-\B x_{a})-\epsilon= [X,Q](\B x, \B y)-\epsilon$.
\end{example}

\begin{definition}
A $Q$-category $X$ is \emph{order-complete} if it is a sup-lattice with respect to the order $\preceq_{X}$ (i.e.~all joins exist).
\end{definition}

\begin{lemma}\label{lemma:supinf}
Let $X$ be a $Q$-category. If $X$ is order-complete, then 
\begin{itemize}
\item if $X$ is co-tensored, $X(\bigvee_{i}x_{i},y)=  \sup_{i}X(x_{i},y) $;
\item if $X$ is tensored, $
X(x,\bigvee_{i}y_{i})=  \inf_{i}X(x,y_{i})$.

\end{itemize}
\end{lemma}
\begin{proof}
We only prove the second claim, the first being proved similarly.

 Let us show that $z\preceq_{X}z'$ iff for all $w\in X$, $X(w,z')\leq X(w,z)$: 
 on one direction we have $X(w,z')\leq X(w,z)+X(z,z')=X(w,z)$; on the other direction, 
 we have $X(z,z')\leq X(z,z)=0$.

 Using this, since $y_{i}\preceq_{X}y:=\bigvee_{i}y_{i}$ we deduce 
 $X(x,y_{i})\leq X(x,y)$, and thus $X(x,y)\geq \inf_{i}X(x,y_{i})$.

 For the converse direction, we argue as follows: let $X(x,y_{i})\leq \epsilon$ hold for all $i\in I$; then $0=X(x,y_{i})-\epsilon= X(x\otimes\epsilon,y_{i})$. Thus $x\otimes\epsilon\preceq_{X}y_{i}$, and thus
 $x\otimes\epsilon\preceq_{X}y$, that is $X(x\otimes\epsilon,y)=X(x,y)-\epsilon=0$, and consequently $X(x,y)\leq \epsilon$. 
 By letting $\epsilon:=X(x,y_{i})$ we conclude then $X(x,y)\leq X(x,y_{i})$, and thus $X(x,y)\leq \inf_{i}X(x,y_{i})$.

%
%First, if $
%By definition $x:=\bigvee_{i}x_{i}$ is characterized by (1) $X(x_{i},x)=0$ for all $i\in I$, and 
% (2) $X(x,y)=0$, for all $y$ such that $X(x_{i},y)=0$ holds for all $i\in I$.
%% 
%% Let us show that $z\preceq_{X}z'$ implies $X(z',y) \leq X(z,y)$: 
%% from $z\preceq_{X}z'$ we deduce $X(z,z')=0$, whence $
%% 
% Let now $y\in X$; 
% then $X(x_{i},y)\leq X(x_{i},x)+X(x,y)=X(x,y)$, which implies $\sup_{i}X(x_{i},y)\leq X(x,y)$.
% 
%
% 
% 
% Suppose now that, for some $i\in I$, $X(x_{i},y)>0$; then $X(x,y)\leq X(x,x_{i})+X(x_{i},y)$
% 
% 
%If $X(x_{i},y) \leq \epsilon$, for all $i\in I$, then 
%$0= X(x_{i},y)-\epsilon = X(x_{i},y\multimapinv \epsilon)$. Thus 
%$x_{i}\preceq_{X}y\multimapinv\epsilon$ and we deduce 
%$x\preceq_{X}y\multimapinv \epsilon$. Hence
%$0= X(x, y\multimapinv \epsilon)=X(x,y)-\epsilon$ and consequently
%$ X(x,y)\leq \epsilon$.

\end{proof}

\begin{proposition}\label{prop:tencoten}
If a $Q$-category $X$ is tensored, cotensored and order-complete, then it is cocomplete.
\end{proposition}
\begin{proof}
For all $\B x\in [X,Q]$, let $\sup \B x:= \bigvee_{a\in X}a\otimes\B x_{a}$. 
Let us check that $X(\sup \B x, b)= \sup_{a\in X}X(a,b)-\B x_{a}$, using Lemma \ref{lemma:supinf}:
\begin{align*}
X(\sup \B x, b) &= \sup_{a\in X}X(a\otimes \B x_{a},b)\\
&=\sup_{a\in X}X(a,b)-\B x_{a}
\end{align*}
We can thus conclude using Proposition \ref{prop:yonedasup}.
\end{proof}

\begin{proposition}\label{prop:tenfun}
Let $X,Y$ be two tensored $Q$-categories, and $f:X\to Y$ be a function.
\begin{itemize}
\item[i.] $f$ is a functor iff $f$ is order-preserving and for all $x\in X$ and $\epsilon\in Q$, $f(x)\otimes \epsilon \preceq_{Y} f(x\otimes \epsilon)$.

\item[ii.] $f$ is a cocontinuous functor iff $f$ commutes with joins and for all $x\in X$ and $\epsilon\in Q$, $f(x)\otimes \epsilon = f(x\otimes \epsilon)$.
\end{itemize}
\end{proposition}
\begin{proof}
\begin{itemize}
\item[i.] If $f$ is a functor then 
\begin{align*}
Y(f(x)\otimes \epsilon, f(x\otimes \epsilon))&= Y(f(x), f(x\otimes \epsilon)) -\epsilon \\
&\leq X(x, x\otimes \epsilon)-\epsilon \\
&= X(x\otimes \epsilon, x\otimes \epsilon)=0
\end{align*}
so $Y(f(x)\otimes \epsilon, f(x\otimes \epsilon))=0$, which implies
$f(x)\otimes \epsilon \preceq_{X}f(x\otimes \epsilon)$. 
Moreover, if $x\preceq_{X}x'$, then $0\geq X(x,x')\geq Y(f(x),f(x'))$, whence 
$f(x)\preceq_{Y}f(x')$, so $f$ is order-preserving.

Conversely, for all $x,x'\in X$, 
\begin{align*}
X(x\otimes X(x,x'), x') &=X(x,x')-X(x,x')=0 
\end{align*}
thus $x\otimes X(x,x') \preceq_{X}x'$. Since $f$ is order-preserving, it follows that
\begin{align*}
f(x)\otimes X(x,x') \preceq_{Y}f(x\otimes X(x,x'))\preceq_{Y}f(x') 
\end{align*}
which implies that 
\begin{align*}
Y(f(x),f(x')) - X(x,x') = Y(f(x)\otimes X(x,x'), f(x'))=0
\end{align*}
that is $Y(f(x),f(x'))\leq X(x,x')$, so $f$ is a functor.

\item[ii.]
Suppose $f$ is a cocontinuous functors, and let  $g:Y\to X$, be its right-adjoint, i.e.~satisfying $Y(f(x),y)=X(x,g(y))$. Then 
\begin{align*}
Y(f(x\otimes \epsilon), y)& = X(x\otimes \epsilon, g(y)) \\
&= X(x, g(y))-\epsilon \\
&= Y(f(x), y)-\epsilon
\end{align*}
which implies that $f(x\otimes \epsilon)$ coincides with the tensor $f(x)\otimes \epsilon$. 
Moreover, clearly also $f(x)\preceq_{Y}y$ iff $x\preceq_{X}g(y)$ holds, which means that $f$ is left-adjoint to $g$ also with respect to the order.

Conversely, suppose the function $f:X\to Y$ preserves joins and tensors. Since $f$ is order-preserving, by i.~it is a functor, so we must only prove that it is cocontinuous.
Since $f$ preserves joins there exists a function $g:Y\to X$ which is right-adjoint to $f$ with respect to orders, i.e.~$f(x)\preceq_{Y}y$ iff $x\preceq_{X}g(y)$. 
We need to prove then that $f$ is left-adjoint to $g$, i.e.~that $Y(f(x),y)=X(x,g(y))$.

On the one hand we have 
\begin{align*}
0 = X(x, g(y))-X(x,g(y))= X(x\otimes X(x,g(y)), g(y))
\end{align*}
from which it follows
\begin{align*}
0= Y(f(x\otimes X(x,g(y)), y)=Y(f(x)\otimes X(x,g(y)), y)=
Y(f(x),y)-X(x,g(y))
\end{align*}
where the first inequality follows from the fact that $f$ and $g$ are adjoint with respect to the order (so $Y(f(x),y)=0$ iff $X(x,g(y))=0$).
This implies then $Y(f(x),y)\leq X(x,g(y))$.

For the converse inequality, 
\begin{align*}
0=Y(f(x),y)-Y(f(x)-y)=Y(f(x)\otimes Y(f(x),y),y)=
Y(f(x\otimes Y(f(x),y)),y)
\end{align*}
and by a similar reasoning we deduce
\begin{align*}
0=X(x\otimes Y(f(x),y), g(y))=
X(x,g(y))-Y(f(x),y)
\end{align*}
whence $X(x,g(y))\leq Y(f(x),y)$.
\end{itemize}
\end{proof}

\begin{theorem}\label{thm:equivalence}
The category $Q\mathsf{Mod}$ of $Q$-modules and $Q$-module morphism coincides with the category $Q\mathsf{CCat}$ of cocomplete skeletal $Q$-categories and cocontinuous functors.
\end{theorem}
\begin{proof}
We have already seen that any cocomplete skeletal $Q$-category is a $Q$-module via tensors, 
and that cocontinuous functors are $Q$-module morphisms.
Let us now show that any $Q$-module is a cocomplete skeletal $Q$-category, and that a $Q$-module morphism is a cocontinuous functor.

Let then $M=(M,\preceq, \star)$ be a $Q$-module. Define $M(x,y)= \inf\{ \delta \mid x\star \delta \succeq y\}$.
 It is clear that $M(x,x)=0$. Let us prove $M(x,y)+M(y,z) \succeq M(x,z)$: 
from $x\star M(x,y)\succeq y$ and $y\star M(y,z)\succeq z$ we deduce 
$x\star(M(x,y)+M(y,z))= (x\star M(x,y))\star M(y,z) \succeq y\star M(y,z)\succeq z$, and thus 
$M(x,z)\preceq M(x,y)+M(y,z)$. 
Observe that $M(x,y)=0$ iff $x=x\star 0\geq y$, so the order $\preceq_{M}$ coincides with the order of $M$.

Let us check that the $Q$-category $M$ is tensored via $x\otimes \epsilon:=x\star\epsilon$.
Let $A_{x,y}= \{ \delta \mid (x\star\epsilon)\star \delta \geq y$ and 
$B_{x,y}=\{\delta-\epsilon\mid x\star \delta \geq y\}$.
Let us show that $A_{x,y}=B_{x,y}$: if $\delta \in A_{x,y}$, then 
$\delta=(\epsilon+\delta)-\epsilon$ satisfies 
$x\star(\epsilon+\delta)=(x\star\epsilon)\star \delta \geq y$, whence 
$\delta\in B_{x,y}$. Conversely, if $\eta=\delta-\epsilon\in B_{x,y}$, then 
$(x\star\epsilon)\star \eta \geq x\star \delta \geq y$, whence $\eta\in A_{x,y}$.
We conclude then that $M(x\star\epsilon,y)=\inf A_{x,y}=\inf B_{x,y}=
\inf\{\delta \mid x\star \delta \geq y\}-\epsilon=M(x,y)-\epsilon$.

Let us define the opposite action $x\multimapinv \epsilon= \bigwedge\{y \mid 
y\star \epsilon \geq x\}$. We must show that $M$ is cotensored via $\multimapinv$, for which it suffices to show $M(x\star \epsilon,y)=M(x,y\multimapinv \epsilon)$. Let $C_{x,y}=\{\delta \mid 
x\star \delta \geq y\multimapinv \epsilon\}$. We have that $\delta \in A_{x,y}$ iff  
$(x\star \delta)\star \epsilon=x\star(\delta+\epsilon)=x\star(\epsilon+\delta)=(x\star \epsilon)\star \delta \geq y$ which is equivalent to $x\star\delta \geq y\multimapinv \epsilon$. We conclude that $A_{x,y}=C_{x,y}$, from which $M(x\star \epsilon,y)=\inf A_{x,y}=\inf C_{x,y}=M(x,y\multimapinv \epsilon)$.

Since $M$, as a $Q$-category, is order-complete, tensored and cotensored, it is cocomplete by Proposition \ref{prop:tencoten}.

To conclude, notice that if $f:X\to Y$ is a cocontinuous functor, then it commutes with tensors and, by 
Proposition \ref{prop:tenfun} it commutes with joins, so it is a morphism of the respective $Q$-modules. Conversely, if $f:M\to N$ is a $Q$-module morphism, then, since $M$ and $N$ are both tensored $Q$-categories, the tensor coincides with the actions of $M$ and $N$, $f$ preserves the joins and the tensor, by Proposition \ref{prop:tenfun}, it is a cocontinuous functor of the respective $Q$-categories.
\end{proof}

\subsection{$Q\mathsf{Mod}$ is a $*$-Autonomous Category}

Let us first observe that:
\begin{itemize}
\item the hom-set $Hom(M,N)$ of two $Q$-modules is a $Q$-module with order and action defined pointwise;

\item for any $Q$-module $M=(M,\preceq, \star)$, there is a $Q$-module
$M^{\mathrm{op}}=(M,\succeq, \multimapinv)$, with $\multimapinv$ defined as in the proof of Theorem \ref{thm:equivalence}.

\end{itemize}

Let $M,N$ be two $Q$-modules. For all $A\in Q^{M\times N}$, we define the function
\begin{align*}
H_{A} & : Q^{M} \longrightarrow Q^{N}%\\
%K_{A} & : Q^{N} \longrightarrow Q^{M}
\end{align*}
via
\begin{align*}
H_{A}(\B x)(b)  &= \inf_{a\in M}\B x_{a}+A(a,b)\\
%K_{A}(\B x)(b)  &= \sup_{a\in M}\B x_{a}-A(a,b)
\end{align*}

\begin{lemma}
$H_{A}=H_{A'}$ iff $A=A'$. 
\end{lemma}
\begin{proof}
We only need to prove one direction, so suppose $A\neq A'$ and let $a,b$ be such that $A(a,b)\neq A'(a,b)$.
Let $\B x$ be defined by $\B x_{a}=1$ and $\B x_{a'}=\infty$ for all $a'\neq a$. Then $H_{A}(\B x)(b)=A(a,b)\neq A'(a,b)=H_{A'}(\B x)(b)$. 
\end{proof}

\begin{proposition}
$Q^{M\times N}$ and $Hom(Q^{M},Q^{N})$ are isomorphic $Q$-modules.
\end{proposition}
\begin{proof}
The map $A\mapsto H_{A}$ is injective, as shown above. We need to check that it commutes with joins:
\begin{align*}
H_{\bigvee_{i}A_{i}}(\B x)(b) & = \inf_{a}\B x_{a}+\bigvee_{i}A_{i}(a,b)\\
&=  \inf_{a}\bigvee_{i}\B x_{a}+A_{i}(a,b)\\
&=  \bigvee_{i}\inf_{a}\B x_{a}+A_{i}(a,b)\\
&=  \bigvee_{i}H_{A_{i}}(\B x)(b)\end{align*} 
(recall that $\inf$s are actually joins in $Q$!)

We must prove that $H$ is surjective: for all $f\in Hom(Q^{M},Q^{N})$, let 
$k_{f}\in Q^{M\times N}$ be given by $k_{f}(a,b)=f(e_{a})(b)$.

Then we have 
\begin{align*}
H_{k_{f}}(\B x)(b) & = \inf_{a}\B x_{a}+k_{f}(a,b) \\
&= \inf_{a}\B x_{a}+ f(e_{a})(b)\\
&= \left(\inf_{a}\B x_{a}+f(e_{a})\right)(b) \\
&= \left(\inf_{a}f(\B x_{a}+e_{a})\right)(b) \\
&= f(\inf_{a}\B x_{a}+e_{a})(b)\\
&=
f(\B x)(b)
\end{align*}
and conversely 
\begin{align*}
k_{H_{A}}(a,b)&= H_{A}(e_{a})(b)= \inf_{a'}(e_{a})_{a'}+A(a',b) =  A(a,b)
\end{align*}
\end{proof}

More generally, we have the following result:
\begin{proposition}
Let $X,Y$ be two $Q$-modules. For any morphism $f: X\to Y$ there is a matrix $k_{f}\in Q^{X\times Y}$ such that

\end{proposition}
\begin{proof}
By composing $f$ with the isomorphisms $\C Y^{\mathrm{op}}:X\to \B M(X)$, with inverse $\sup(\B x)=\bigvee_{a\in x}a\otimes \B x_{a}$, we obtain 
a morphism $\widehat f: \B M(X)\to \B M(Y)$  
\begin{align*}
\widehat f(\B x)(b)&:= \C Y^{\mathrm{op}}(f(\sup \B x))(b) \\
&= Y( f(\bigvee_{a\in X}a\otimes \B x_{a}),b) \\
&= Y( \bigvee_{a\in X}f(a)\otimes \B x_{a},b) \\
&= \sup_{a\in X}Y(f(a),b)-\B x_{a}
\end{align*}
Observe that $\widehat f$ can be extended to a function $f^{*}$ from $Q^{X}$ to $Q^{Y}$.
Now, $ f^{*}$ is generated by the matrix $k_{f}\in Q^{X\times Y}$ given by $k_{ f}(a,b)=f(e_{a})(b)$, so that for all $\B x\in Q^{X}$, 
$f^{*}(\B x)(b)=\bigvee_{a\in X}k_{f}(a,b)+\B x_{a}$, and thus
in particular, for all $\B x\in [X,Q]$, 
$\widehat f(\B x)(b)=f^{*}(\B x)(b)$.
\end{proof}

\subsubsection{The Tensor Product of $Q$-Modules}

The description of the tensor product of $Q$-modules requires some work. Let us first recall some important definitions:

\begin{definition}[congruence on a sup-lattice]
Let $(L, \leq)$ be a sup-lattice. An equivalence relation $R\subseteq L\times L$ is said a \emph{congruence} if it satisfies the following property:
\begin{align}
(\forall i\in I \ x_{i} Ry_{i})  \ \To  \ \left( \bigvee_{i}x_{i} \right ) R\left(\bigvee_{i}y_{i}\right)
\tag{congruence}
\end{align}
\end{definition}

\begin{lemma}
For all suplattices $(L,\leq)$, if $R$ is a congruence, then $(L/R, \leq_{R})$ is a sup-lattice, where $[x]\leq_{R}[y]$ iff $(x\vee y) R y$ (i.e.~$[x\vee y]=[y]$), and $\bigvee_{i}[x_{i}]=\left [\bigvee_{i}x_{i}\right]$.
\end{lemma}
\begin{proof}
Let us check that $\leq_{R}$ is an order. It is clear that $[x]\leq_{R}[x]$ holds. If $[x]\leq_{R}[y]$ and $[y]\leq_{R}[z]$ both hold, then 
$(x\vee y)Ry$ and $(y\vee z)Rz$ hold; 
then, since $R$ is a congruence $((x\vee y)\vee (y\vee z)) R (y\vee z)R z$, and moreover $x \vee (y\vee z) R (x \vee z)$, whence 
$(x\vee z) R(x \vee y\vee z)R z$, so $[x]\leq_{R}[z]$.  
If $[x]\leq_{R}[y]$ and $[y]\leq_{R}[x]$, then 
$xR(x\vee y)R y$, thus $[x]=[y]$.

Let us now check the definition of joins. 
From $[x_{i}]\vee [\bigvee_{i}x_{i}]=[x_{i}\vee \bigvee_{i}x_{i}]= [\bigvee_{i}x_{i}]$ we deduce $[x_{i}]\leq_{R}[\bigvee x_{i}]$.
Suppose now $[x_{i}]\leq [y]$ holds for all $i\in I$, that is, 
$(x_{i}\vee y)Ry$; then, since $R$ is a congruence, 
$(\bigvee_{i}(x_{i}\vee y))R y$, that is, 
$((\bigvee_{i}x_{i})\vee y)Ry$, which implies 
$[\bigvee_{i}x_{i}]\leq_{R}[y]$. We conclude then that $\bigvee_{i}[x_{i}]=[\bigvee_{i}x_{i}]$.
\end{proof}

\begin{corollary}\label{cor:bigvee}
Let $(L,\leq)$ be a suplattice and $R$ be a congruence on $L$.
Then, for any class $\beta\in L/R$, $\beta= [\bigvee\beta]$.
\end{corollary}
\begin{proof}
$[\bigvee \beta]=[\bigvee\{ x \mid x\in \beta\}]= \bigvee \{[x]\mid x\in \beta\}=\beta$.
\end{proof}

\begin{proposition}\label{prop:smallestcongruence}
Let $(L,\leq)$ be a sup-lattice. Let $R\subseteq L\times L$ be an equivalence relation, and for any ordinal $\alpha$, let the relation
$R^{(\alpha)}\subseteq L\times L$ be defined by:
\begin{itemize}
\item $xR^{(0)}y$ iff either $xRy$, $x=y$ or $yRx$ holds;
\item $xR^{(\alpha+1)}y$ iff one of the following holds:
	\begin{itemize}
	\item for some $z$, $xR^{(\alpha)}z$ and $zR^{(\alpha)}y$ holds;
	\item for some set $I$, and families $x_{i},y_{i}$, 
	$x=\bigvee x_{i}, y=\bigvee_{i}y_{i}$ and $x_{i} R^{(\alpha)}y_{i}$ holds for all $i\in I$.

	\end{itemize}
\item $xR^{(\gamma)}y$ iff $xR^{(\delta)}y$ holds for some $\delta <\gamma$, for $\gamma$ limit.
\end{itemize}
Then the relation $R^{*}\subseteq L\times L$ given by 
$$
xR^{*} y  \ \Leftrightarrow \ \exists \alpha . \mathrm{OR}(\alpha) \land xR^{(\alpha)}y
$$% defined as follows: $xR^{*} y$ iff
%\begin{enumerate}
%\item for any set $I$ and family $x_{i}$ such that $x=\bigvee_{i}x_{i}$, there exists a family $y_{i}$ such that $y=\bigvee_{i}y_{i}$ and $x_{i}R y_{i}$ holds for all $i\in I$;
%\item for any set $I$ and family $y_{i}$ such that $y=\bigvee_{i}y_{i}$, there exists a family $x_{i}$ such that $x=\bigvee_{i}x_{i}$ and $x_{i}R y_{i}$ holds for all $i\in I$.
%\end{enumerate}
%%\begin{align*}
%%xR^{*}y &\text{ iff }\  \forall I \ \forall x_{i} \  \text{ s.t. }
%%\left\{
%%\begin{matrix}
%%x=\bigvee_{i\in I}x_{i} \\
%%y=\bigvee_{i\in I}y_{i}\\
%%x_{i}Ry_{i} \ (\forall i\in I)
%%\end{matrix}
%%\right\}
%%\end{align*}
is a congruence, and is the smallest congruence containing $R$.
\end{proposition}
\begin{proof}
From $xR^{(0)}x$ it follows $xR^{*}x$.

Let us prove by induction that for any ordinal $\alpha$, $R^{(\alpha)}$ is symmetric:
for $\alpha=0$ this is immediate; suppose $xR^{(\alpha+1)}y$, then two cases are possible: either $xR^{(\alpha)}z$ and $zR^{(\alpha)}y$, then by IH 
$yR^{(\alpha)}z$ and $zR^{(\alpha)}x$, whence $yR^{(\alpha+1)}x$; oer
$x_{i}R^{(\alpha)}y_{i}$ for some decompositions $x=\bigvee_{i}x_{i}$ and $y=\bigvee_{i}y_{i}$; then by IH $y_{i}R^{(\alpha)}x_{i}$, so $yR^{(\alpha+1)}x$ holds.
Finally, if $\alpha$ is limit, then from $x R^{(\alpha)}y$ it follows
$xR^{(\beta)}y$ for some $\beta<\alpha$, whence 
$yR^{(\beta)}x$ by IH and we conclude $yR^{(\alpha)}x$.

Now, if $xR^{*}y$ then $xR^{(\alpha)}y$ holds for some ordinal $\alpha$, and thus $yR^{(\alpha)}x$ holds too, whence $yR^{*}x$.

Observe that $\alpha<\beta$ implies $R^{(\alpha)}\subseteq R^{(\beta)}$:
this is obvious if $\beta$ is limit, otherwise, if $\beta=\alpha+1$, from $xR^{(\alpha)}y$ and $x R^{(\alpha)}x$ we deduce
$xR^{(\alpha+1)}y$.

Suppose now $xR^{*}y$ and $yR^{*}z$. Then $xR^{(\alpha)}y$ and $yR^{(\beta)}z$ hold for some 
ordinals $\alpha$ and $\beta$; let $\gamma=\max\{\alpha,\beta\}$; then 
we have $xR^{(\gamma)}y$ and $yR^{(\gamma)}z$, whence $xR^{(\gamma+1)}z$ and thus 
$xR^{*}z$.

Suppose $x_{i}R^{*}y_{i}$ holds for all $i\in I$; then 
for all $i$ there is some ordinal $\alpha_{i}$ with 
$x_{i}R^{(\alpha_{i})}y_{i}$. 
Let $\gamma=\sup_{i}\alpha_{i}$, so that 
$x_{i}R^{(\gamma)}y_{i}$; then we have
$\bigvee_{i}x_{i} R^{(\gamma+1)}\bigvee y_{i}$, and thus
$\bigvee_{i}x_{i} R^{*}\bigvee_{i}y_{i}$.

%
%$R^{*}$ is symmetric, reflexive and transitive, given that $R$ is. 
%
%Let $I$ be a set and 
%suppose $x_{i}R^{*}y_{i}$ holds for all $i\in I$.
%Observe now that if $x=\bigvee_{i\in I} x_{i}=\bigvee_{j\in J}x'_{j}$, then
%\begin{enumerate}
%\item $x_{i}=\bigvee_{j\in J}x_{i}\land x_{j}$;
%\item $x= \bigvee_{(i,j)\in I\times J}x_{i}\land x'_{j}$. 
%\end{enumerate}
% $\bigvee_{j\in J}x_{i}\land x_{j}\leq x_{i} \leq x_{i}$ is clear.
% Conversely, we have $x_{i}= x \land x_{i} = \left (\bigvee_{j\in J}x_{j}\right) \land x_{i}= \bigvee_{j\in J}x_{j}\land x_{i}= \bigvee_{j\in J}x_{i}\land x_{j}$. {\color{red}NOT TRUE IN ANY LATTICE!}

%
% Then for each $i\in I$ there exists a set $J_{i}$ and sequences $x_{ij},y_{ij}$ such that 
%$\bigvee_{j\in J_{i}}x_{ij}=x_{i}$, $\bigvee_{j\in J_{i}}y_{ij}=y_{i}$ and 
%$x_{ij}Ry_{ij}$.
%
%Let then $K= \prod_{i\in I}\{i\}\times J_{i}$; then for all $(i,j)\in K$, 
%$x_{ij}R y_{ij}$, so we deduce that $\bigvee_{i\in I}x_{i}=\bigvee_{(i,j)\in K}x_{ij} R^{*} \bigvee_{(i,j)\in K}y_{ij}=\bigvee_{i\in I}y_{i}$, which proves that $R^{*}$ is a congruence.

Suppose now $S$ is a congruence containing $R$.
Let us show that for any ordinal $\alpha$, $R^{(\alpha)}\subseteq S$:
\begin{itemize}
\item $R^{(0)}\subseteq S$ holds since $S$ is an equivalence relation and contains $R$;

\item if $x R^{(\alpha+1)}y$ holds, then two cases occur:
	\begin{itemize}
	\item $xR^{(\alpha)}z$ and $zR^{(\alpha)}y$ hold, so by IH, 
	$xSz$ and $zSy$, and since $S$ is transitive, $xSy$ holds;
	\item $x=\bigvee_{i}x_{i}$, $y=\bigvee_{i}y_{i}$ and $x_{i}R^{(\alpha)}y_{i}$ holds; then by IH $x_{i}Sy_{i}$ holds, and since $S$ is a congruence, $xSy$ holds;

\item if $\alpha$ is limit and $xR^{(\alpha)}y$ holds, then $xR^{(\beta)}y$ holds for some $\beta<\alpha$, and by IH $xS y$ holds.

	\end{itemize}

\end{itemize}
Now, if $x R^{*} y$ holds, then $xR^{(\alpha)}y$ holds for some ordinal $\alpha$, whence $xS y$ holds.
This shows that $R^{*}\subseteq S$.
%
%Then there exists a set $I$ so that $x=\bigvee_{i\in I}x_{i}$, $y=\bigvee_{i\in I}y_{i}$ and $x_{i}Ry_{i}$; now, since $S$ contains $R$, we deduce $x_{i}Sy_{i}$ for all $i\in I$, and since $S$ is a congruence, $xSy$ holds, which proves that $R^{*}\subseteq S$.
\end{proof}

We can now introduce the tensor of $Q$-modules, that will be defined as a suitable quotient lattice.

\begin{definition}[tensor of $Q$-modules]
Let $M$ and $N$ be $Q$-modules. The \emph{tensor product} $M\otimes_{Q}N$ of $M$ and $N$ is the $Q$-module defined as $\C P(M\times N)/ R^{*}$, where $R^{*}$ is the smallest congruence containing the relation $R$ defined by:
\begin{align*}
R'= \left\{
\begin{matrix}
\left((\bigvee A, y), \bigcup_{a\in A}\{(a,y)\}\right)\\
\left((x,\bigvee B), \bigcup_{b\in B}\{(x,b)\}\right)\\
(\{(x\star \epsilon,y)\}, \{(x,y\star\epsilon)\})
\end{matrix}
\ \Bigg \vert\ 
\begin{matrix}
A\subseteq M, y\in N \\
B\subseteq N, x\in M \\
\epsilon \in Q
\end{matrix}
\right\}
\end{align*}
and the action is defined via $[A]\star \epsilon= \bigvee\{[ \{(x\star\epsilon,y)\}]\mid 
(x,y)\in A\}$.
\end{definition}

Let a \emph{$Q$-bimorphism} be a map $f:M\times N\to L$ such that $f$ preserves joins in each variable separately, and moreover $f(x,y\star \epsilon)=f(x\star\epsilon,y)$. A $Q$-bimorphism $f:M\times N\to L$ is \emph{universal} if for all $L'$ and bimorphism $g:M\times N\to L'$ there is a unique sup-lattice homomorphism $h:L\to L'$ such that $g=h\circ f$.

\begin{proposition}[universal property of the tensor product]
The tensor product $M\otimes_{Q}N$ is the codomain of the universal $Q$-bimorphism $M\times N \to M\otimes_{Q}N$.
\end{proposition}

\begin{remark}

For any $x\in M$ and $y\in N$, we indicate as $x\otimes_{Q}y$ the image of the pair $(x,y)$ under the universal $Q$-bimorphism $\tau:M\times N\to M\otimes_{Q}N$, or equivalently, as the $R^{*}$-equivalence class of $(x,y)$.
Since joins in $M\otimes_{Q}N$ are given by  
$\bigvee_{i}[A_{i}]= [\bigcup_{i}A_{i}]$, we have then 
that any element $[A]\in M\otimes_{Q}N$ can be written as 
$[A]= \bigvee\{x\otimes y\mid (x,y)\in A\}$.

%
% Any element of $M\otimes_{Q}N$ is a join of tensors, that is
%$$
%M\otimes_{Q}N= \left\{ \bigvee_{i\in I}x_{i}\otimes_{Q}y_{i}\ \Bigg \vert \  x_{i}\in M, y_{i}\in N \right\}
%$$
\end{remark}

\begin{lemma}
\begin{itemize}
\item $M\otimes_{Q}N\simeq N\otimes_{Q}M$;

\item $(M\otimes_{Q}N)\otimes_{Q}R \simeq M\otimes_{Q}(N\otimes_{Q}R)$.
\end{itemize}
\end{lemma}

\begin{proposition}
$Hom(M\otimes_{Q}N, R)   \simeq   Hom(M, Hom(N,R))$ (as an isomorphism of sup-lattices).
\end{proposition}
\begin{proof}
Given $h:M\otimes_{Q}N\to R$, for all $x\in M$ define $h_{x}:N\to R$ by 
$h_{x}(y)=h(x\otimes_{Q}y)$. We then have 
$h_{x}(\bigvee_{i}y_{i})=h(x\otimes_{Q}\bigvee_{i}y_{i})=
h(\bigvee_{i}x\otimes_{Q} y_{i})=
\bigvee_{i}h(x\otimes_{Q} y_{i})=\bigvee_{i}h_{x}(y_{i})$ and 
$h_{x}(y\star\epsilon)= h(x\otimes_{Q}(y\star \epsilon))=h((x\otimes_{Q}y)\star\epsilon)=h(x\otimes_{Q}y)\star \epsilon= h_{x}(y)\star\epsilon$, so $h_{x}\in Hom(N,R)$. Moreover, by a similar argument we have  $h_{\bigvee_{i}x_{i}}(y)=\bigvee_{i}h_{x_{i}}(y)$
and $h_{x\star\epsilon}(y)=h_{x}(y)\star\epsilon$, so the map $x\mapsto h_{x}$ is a $Q$-module morphism.

Finally, for any family $h_{i}:M\otimes_{Q}N\to R$, we have 
$(\bigvee_{i}h_{i})_{x}(y)=\bigvee_{i}h_{i}(x\otimes_{Q}y)=\bigvee_{i}(h_{i})_{x}(y)=(\bigvee_{i}h_{i})(x\otimes y)$, and thus we have a sup-lattice homomorphism $\zeta$ from $Hom(M\otimes_{Q}N,R)$ to $Hom(M,Hom(N,R))$ given by $\zeta(h)=h_{\_}$.

Let us show that $\zeta$ has an inverse:  for all $f\in Hom(M,Hom(N,R))$, define $f': M\times N\to R$ by $f'(x,y):=f(x)(y)$. This is clearly a bimorphism, so there is a unique homomorphism $h_{f'}:M\otimes_{Q}N\to M\times N$ such that $f'=h_{f'}\circ \tau$, i.e.~such that 
$h_{f'}(x\otimes y)=f'(x,y)=f(x)(y)$, and thus $\zeta(h_{f'})=f$.
On the other hand, if $f=\zeta(h)$, then the uniqueness of $h_{f'}$ ensures that $h_{f'}=h$.
\end{proof}

\begin{proposition}
\begin{itemize}
\item[i.] $Hom(Q,M)\simeq M$.
\item[ii.] $Hom(M,N)\simeq Hom(N^{\mathrm{op}},M^{\mathrm{op}})$.
\item[iii.] $Hom(M, Q^{\mathrm{op}})\simeq M^{\mathrm{op}}$.
\end{itemize}
(all isomorphisms of sup-lattices).
\end{proposition}
\begin{proof}
Define $\alpha:M\to Hom(Q,M)$ by $\alpha(x)(\epsilon)=x\star\epsilon$ and 
$\beta:Hom(Q,M)\to M$ by $\beta(f)=f(0)$. Then we have that 
$\alpha(\beta(f))(\epsilon)=\alpha(f(0))(\epsilon)=f(0)\star\epsilon=f(\epsilon)$, and 
$\beta(\alpha(x))=\beta(\lambda \epsilon.x\star\epsilon)=x\star0=x$.

If $f\in Hom(M,N)$, since it preserves joins, it has a right-adjoint $f^{*}:N^{\mathrm{op}}\to M^{\mathrm{op}}$, such that $f(x)\leq y$ iff $x\leq f^{*}(y)$.

By i.~$Hom(Q,M^{\mathrm{op}})\simeq M^{\mathrm{op}}$ and we conclude by ii.
\end{proof}

\begin{proposition}
\begin{itemize}
\item[i.] $Hom(M,N)\simeq (M\otimes_{Q}N^{\mathrm{op}})^{\mathrm{op}}$.
\item[ii.] $M\otimes_{Q}N\simeq Hom(M, N^{\mathrm{op}})^{\mathrm{op}}$.
\item[iii.] $Q\otimes_{Q} M \simeq M\otimes_{Q}Q\simeq Q$.
\end{itemize}
\end{proposition}
\begin{proof}
$Hom(M,N)\simeq Hom(M, Hom(N^{\mathrm{op}},Q^{\mathrm{op}}))
\simeq Hom(M\otimes N^{\mathrm{op}},Q^{\mathrm{op}})\simeq
(M\otimes N^{\mathrm{op}})^{\mathrm{op}}$.
Claim ii.~is proved similarly.

For claim iii.~$Q\otimes_{Q}M\simeq M\otimes_{Q}M\simeq Hom(M, Q^{\mathrm{op}})^{\mathrm{op}}\simeq (M^{\mathrm{op}})^{\mathrm{op}}=M$. 
\end{proof}

By putting all previous results together we obtain:
\begin{theorem}
$Q\mathsf{Mod}$ is a $^{*}$-autonomous category.
\end{theorem}

\begin{proposition}\label{prop:Qtensor}
\begin{itemize}
\item[i.] $Q^{X}\otimes_{Q}M \simeq M^{X}$;
\item[ii.] $Q^{X}\otimes_{Q}Q^{Y}\simeq Q^{X\times Y}$.
\end{itemize}
\end{proposition}
\begin{proof}
$M^{X}$ coincides with the product $\Pi_{x\in X}M$. We have then 
$Q^{X}\otimes_{Q}M \simeq (\Pi_{x}Q)\otimes_{Q}M \simeq \Pi_{x}(Q\otimes_{Q}M) \simeq \Pi_{x}M\simeq M^{X}$.

For ii., using i.~we have $Q^{X}\otimes_{Q}Q^{Y}\simeq( Q^{Y})^{X}\simeq Q^{X\times Y}$.
\end{proof}

\subsubsection{The Tensor Product of $Q$-Categories}

Thanks to Theorem \ref{thm:equivalence}, the $^{*}$-autonomous structure of $Q\mathsf{Mod}$ translates into a $^{*}$-autonomous structure for $Q\mathsf{CCat}$.

In $\mathsf{Met}$ the ``tensor product'' of two metric spaces $X$ and $Y$ is just their cartesian product, with the ``plus'' metric. What can we say about the tensor product in $Q\mathsf{CCat}$?

% 
%Let us start by discussing the construction of the tensor product $M\otimes_{Q}N$ of $Q$-modules $M$ and $N$ as a quotient of $\C P(M\times N)$. First, $\C P(M\times N)$ is the \emph{free} sup-lattice over $M\times N$: this means that any morphism 
%$f: M\times N \to L$, where $L$ is a sup-lattice, uniquely extends into a sup-lattice morphism $f^{*}:\C P(M\times N)\to L$, given by $f^{*}(A)=\bigvee_{x\in A}f(x)$. 
%
%
%\begin{lemma}
%For all $m\in M$ and $n\in N$, 
%\begin{align*}
%m\otimes n= &  \{A \mid \bigvee A= (m',n'+\epsilon+\delta), m=m'+\epsilon, n=n'+\delta\}\\
%& \cup
% \{A \mid \bigvee A= (m'+\epsilon+\delta, n'), m=m'+\epsilon, n=n'+\delta\}\end{align*}
%\end{lemma}
%\begin{proof}
%
%
%\end{proof}
The goal of this subsection is to describe the $Q$-categorical structure of the tensor product explicitly. The main intuition is that the elements of $X\otimes_{Q}Y$ can be seen as joins of pairs $x\otimes y$ of elements $x\in X$, $y\in Y$. We will then show that the metric coincides with the ``plus'' metric over pairs $x\otimes y$, and extends continuously to joins.

\begin{lemma}\label{lemma:tensorsum}
For all $m,m'\in M$ and $n,n'\in N$ and $\epsilon \in Q$, 
$(m\otimes n)\star \epsilon \succeq (m'\otimes n')$ iff there exists $\delta_{1},\delta_{2}$ such that $\delta_{1}+\delta_{2}=\epsilon$, 
$m+\delta_{1}\succeq m'$ and $n+\delta_{2}\succeq n'$.
\end{lemma}
\begin{proof}[Proof Sketch]
Notice that $(m\otimes n)\star \epsilon = [\{(m\star \epsilon,n)\}]=
[\{(m\star \delta_{1},n\star \delta_{2}\}]$ for all $\delta_{1}+\delta_{2}=\epsilon$. 
Hence, $m'\otimes n' \preceq (m\otimes n)\star \epsilon$ implies that 
for some $\delta_{1}+\delta_{2}=\epsilon$, 
$(m',n')\vee (m\star\delta_{1},n\star\delta_{2}) = (m'\vee (m\star\delta_{1}), n'\vee (n\star\delta_{2}))= (m\star \delta_{1}, n'\star\delta_{2})$, that is, that $m'\preceq_{M} m\star \delta_{1}$ and $n'\preceq_{n}n\star \delta_{2}$. 
\end{proof}

\begin{proposition}\label{prop:tensormetric}
For all $m,m'\in N$ and $n,n'\in N$, 
$$
(M\otimes_{Q} N)(m\otimes_{Q} n, m'\otimes_{Q}n')= M(m,m') +N(n,n')
$$
More generally, 
$$
(M\otimes_{Q}N)([A],[B])= \sup_{(x,y)\in A}\inf_{(x',y')\in B}M(x,x')+N(y,y')
$$
\end{proposition}
\begin{proof}
By definition, $(M\otimes_{Q} N)(m\otimes_{Q} n, m'\otimes_{Q}n')$ is given by $\inf A$, where
$$
A=\{ \epsilon \mid (m\otimes_{Q}n)\star\epsilon \geq m'\otimes_{Q}n'\}
$$
Let us show that $A$ coincides with 
$$
B=\{ \delta_{1}+\delta_{2} \mid m\star\delta_{1} \geq  m', n\star \delta_{2}\geq n'\}$$
On the one hand, if $\delta_{1}+\delta_{2}\in B$, it is clear that $\delta_{1}+\delta_{2}\in A$. Conversely, if $\epsilon\in A$, then by Lemma \ref{lemma:tensorsum} $\epsilon=\delta_{1}+\delta_{2}$ with $m\star\delta_{1}\geq m'$ and $n\star \delta_{2}\geq n'$, whence 
$\epsilon \in B$.

We can now conclude as follows:
\begin{align*}
(M\otimes_{Q} N)(m\otimes_{Q} n, m'\otimes_{Q}n')& =\inf A \\
&= \inf B \\
&= \inf \{ \delta_{1} \mid m\star\delta_{1} \geq  m'\}
+ \inf\{ \delta_{2} \mid  n\star \delta_{2}\geq n'\}\\
&= M(m,m')+N(n,n').
\end{align*}

For the second claim, using the fact that $M\otimes_{Q}N$, as a $Q$-category, is both tensored and cotensored, using the fact that 
$[A]=\bigvee_{(x,y)\in A}x\otimes y$ and $[B]=\bigvee_{(x',y')\in B}z\otimes w$, and Lemma \ref{lemma:supinf}:
\begin{align*}
(M\otimes_{Q} N)([A],[B])&=
\sup_{(x,y)\in A}(M\otimes_{Q}N)(x\otimes y, [B]) \\
&=\sup_{(x,y)\in A}\inf_{(x',y')\in B}(M\otimes_{Q}N)(x\otimes y, x'\otimes y') 
\\
&=
\sup_{(x,y)\in A} \inf_{(x',y')\in B}M(x,x')+N(y,y').
\end{align*}
\end{proof}

\subsection{Products and Coproducts in $Q\mathsf{Mod}$}

Products and coproducts exist in $Q\mathsf{Mod}$ and both coincide with the order product:
$$
\prod_{i\in I}X_{i} \simeq \coprod_{i\in I}X_{i}
$$
In particular, the projection and
 injection morphisms $\pi_{i}:\prod_{i}X_{i}\to X_{i}$ and $\iota_{i}:X_{i} \to \prod_{i\in I}X_{i}$ are defined by 
$$
\pi_{i}\big( (x_{j})_{j\in I}\big) = x_{i} \qquad \qquad
\iota_{i}(x)({j})= \begin{cases} x & \text{ if }i=j\\ \bot & \text{ otherwise}\end{cases}
$$

The following property is important
\begin{proposition}\label{prop:productvstensor}
$\prod_{i\in I} X\otimes Y_{i} \simeq X \otimes \prod_{i\in I}Y_{i}$.
\end{proposition}
\begin{proof}
For a complete proof, see \texttt{https://arxiv.org/pdf/0909.4493.pdf}. Here we just recall the isomorphism 
$h: \prod_{i\in I} X\otimes Y_{i} \simeq X \otimes \prod_{i\in I}Y_{i}$, defined as follows:
\begin{align*}
h\Big( i\mapsto \bigvee_{k\in J_{i}}x_{i,k}\otimes y_{i,k}\Big ) = 
\bigvee_{i\in I,k\in J_{i}}x_{i,k} \otimes \iota_{i}\big( (y_{j,k})_{j\in I}\big)
\end{align*}
and its inverse
\begin{align*}
k\Big( \bigvee_{k\in J}x_{k}\otimes \big(x_{k,i}\big)_{i\in I_{k}}\Big)(i)=
\bigvee_{k\in J,i\in I_{k}}x_{k}\otimes \pi_{i}\big((x_{k,j})_{j\in I_{k}}\big).
\end{align*}
\end{proof}

\subsection{Exponential and Differential Structure of $Q\mathsf{Mod}$}

\subsubsection{Symmetric Algebras in $Q\mathsf{Mod}$ }

We start by constructing the exponential co-modality in the equivalent category $Q\mathsf{Mod}$.

For all $n\in \BB N$ and $\sigma\in \F S_{n}$, let $\langle \sigma\rangle: X^{\otimes n}\to X^{\otimes n}$ be the morphism defined by 
$$
\langle \sigma \rangle (x_{1}\otimes \dots \otimes x_{n})=
x_{\sigma(1)}\otimes \dots \otimes x_{\sigma(n)}
$$
and extended to all other points by continuity.
%For any $Q$-module $X$, let $X_{\bullet}$ indicate the $Q$-module $X\times Q$.  
%
\begin{definition}Let $X$ be a $Q$-module and $n\in \BB N$. 
An element $x\in X^{\otimes_{n}}$ is said \emph{permutation-invariant} (in short, \emph{p-invariant}) if for all $\sigma\in \F S_{n}$, 
$\langle \sigma \rangle (x)=x$.

For all $n\in \BB N$, $x_{1},\dots, x_{n}\in X$, the \emph{$Q$-multiset} $[x_{1},\dots, x_{n}]$ is the element of $X^{\otimes_{n}}$ defined as $0$ if $n=0$, and otherwise as follows:
$$
[x_{1},\dots, x_{n}]:= \bigvee_{\sigma\in \F S_{n}}x_{\sigma(1)}\otimes \dots \otimes x_{\sigma(n)}
$$

Given multisets $A$ and $B$, we define the multiset $A\cup B$ as follows:
\begin{itemize}
\item if $A=0$, then $A\cup B=B$;
\item if $B=0$, then $A\cup B=A$;
\item if $A=[x_{1},\dots, x_{n}]$ and $B=[y_{1},\dots, y_{m}]$, then $A\cup B=[x_{1},\dots, x_{n},y_{1},\dots, y_{m}]$.

\end{itemize}

\end{definition}

\begin{proposition}
Let $X$ be a $Q$-module and $n\in \BB N$. Any $Q$-multiset $[x_{1},\dots, x_{n}]\in X^{\otimes_{n}}$ is p-invariant. Moreover, any p-invariant element $x\in X^{\otimes_{n}}$ can be written as 
a join of $Q$-multisets.
\end{proposition}
\begin{proof}
For the first claim we have, for all $\sigma\in \F S_{n}$, 
\begin{align*}
\langle \sigma \rangle ([x_{1},\dots, x_{n}]) & = 
\bigvee_{\tau\in \F S_{n}}\langle \sigma \rangle ([x_{\tau(1)},\dots, x_{\tau(n)}])\\
 & = 
\bigvee_{\tau\in \F S_{n}}[x_{\sigma\tau(1)},\dots, x_{\sigma\tau(n)}])\\
 & = 
\bigvee_{\tau\in \F S_{n}}[x_{\tau(1)},\dots, x_{\tau(n)}])\\
&= [x_{1},\dots, x_{n}].
\end{align*}
For the second claim, observe that $x$ can always be written as a join of tensors $x=\bigvee_{i}x_{1}^{i}\otimes \dots \otimes x_{n}^{i}$. Moreover, 
if $x_{1}^{i}\otimes \dots \otimes x_{n}^{i}\leq x$, since $x$ is p-invariant, for all $\sigma \in \F S_{n}$, also
$x_{\sigma(1)}^{i}\otimes \dots \otimes x_{n}^{i}\leq \langle \sigma\rangle(x)=x$, so we can conclude that 
$x=\bigvee_{i}[x_{1}^{i},\dots, x_{n}^{i}]$.
\end{proof}

\begin{proposition}
For any $Q$-module $X$, the set $!_{n}X\subseteq X$ of p-invariant elements of $X^{\otimes_{n}}$ is a sub-$Q$-module of $X$.
\end{proposition}
\begin{proof}
If $x_{i}\in X^{\otimes_{n}}$ is a family of p-invariant elements, then 
$x=\bigvee_{i}x_{i}$ is also p-invariant, since $\langle \sigma\rangle (x)=\bigvee_{i}\langle \sigma \rangle (x_{i})=\bigvee_{i}x_{i}=x$. Hence $!_{n}X$ is a sup-lattice.
Moreover, for all $x\in !_{n}X$ and $\epsilon \in Q$, 
$x\otimes \epsilon$ is also p-invariant, since $\langle \sigma \rangle (x\otimes \epsilon)= \langle \sigma \rangle (x)\otimes \epsilon=x\otimes \epsilon$. We conclude that $!_{n}X$ is a sup-lattice with a continuous action of $Q$, where both the order and the action are inherited from $X$, so it is a sub-$Q$-module of $X$.
\end{proof}

The fundamental property of $!_{n}X$ is the following:
\begin{proposition}
For any $Q$-module $X$ and $n\in \BB N$, the inclusion morphism 
$\iota:!_{n}X\longrightarrow X^{\otimes_{n}}$ is the equalizer of the diagram
$$
\begin{tikzcd}
!_{n}X \ar{r}{\iota} & X^{\otimes_{n}} \ar{r}{\langle \sigma\rangle} & X^{\otimes_{n}}
\end{tikzcd}
$$
generated by all actions $\langle \sigma\rangle$, for $\sigma\in \F S_{n}$.
\end{proposition}
\begin{proof}
It is clear that $\langle \sigma \rangle \circ \iota= \langle \tau\rangle \circ \iota$ holds for all $\sigma, \tau \in \F S_{n}$.
Suppose now $h: C\to X^{\otimes_{n}}$ is another morphism satisfying
$\langle \sigma \rangle \circ h= \langle \tau\rangle \circ h$ for all $\sigma, \tau \in \F S_{n}$.
Since $\langle \sigma \rangle \circ h= \langle \mathrm{id}\rangle \circ h=h$, we deduce that $h(c)$ is p-invariant, for all $c\in C$. Hence $h$ splits in a unique way as $C \stackrel{h}{\to} !_{n}X \stackrel{\iota}{\to} X^{\otimes_{n}}$.
\end{proof}

\begin{remark}[metric structure of $!_{n}X$]
As $!_{n}X$ is a sub-$Q$-module of $X^{\otimes_{n}}$, the distance function can be computed explicitly using Proposition \ref{prop:tensormetric}:
\begin{align*}
!_{n}X( [x_{1},\dots, x_{n}], [y_{1},\dots, y_{n}]) & = 
\sup_{\sigma\in \F S_{n}}\inf_{\tau\in \F S_{n}}
\sum_{i=1}^{n}
X(x_{\sigma(i)},y_{\tau(i)})
\end{align*}
\end{remark}

We now show that the $Q$-module $!_{n}X$ is isomorphic to the symmetric algebra, which is used in the construction of the exponential modality in the relational model.

\begin{definition}[symmetric algebra]
For any $Q$-module $X$ and $n\in \BB N$, we let $\mathrm{Sym}_{n}(X)$ indicate the $Q$-module defined as 
$\mathrm{Sym}_{n}(X):=\frac{X^{\otimes_{n}}}{\sim_{n}}
$, where $\sim_{n}$ is the least congruence generated by the action $\langle \sigma\rangle$ of permutations $\sigma\in \F S_{n}$.\end{definition}

\begin{proposition}
$!_{n}X\simeq \mathrm{Sym}_{n}(X) $.
\end{proposition}
\begin{proof}
First, observe that for any equivalence class $\alpha\in \mathrm{Sym}_{n}(X)$, the point $\bigvee\alpha$ is p-invariant: 
 since $x\in \alpha$ holds iff $\langle \sigma \rangle (x)\in \alpha$, for all $\sigma\in\F S_{n}$, 
it follows that $\langle \sigma \rangle (\bigvee \alpha)=\bigvee\{\langle \sigma \rangle (x)\mid x\in \alpha\}=\bigvee \{x\mid x\in \alpha\}=\bigvee \alpha$, and thus $\bigvee\alpha$ is p-invariant.

Now, let us show that for all $x\in X^{\otimes_{n}}$, $x \sim_{n} \bigvee[x]$: for all $y\in [x]$, by definition $x\sim_{n}y$ holds; hence, since $\sim_{n}$ is a congruence, we have that 
$x=\bigvee_{y\in[x]}x \sim_{n} \bigvee_{y\in [x]}y=\bigvee[x]$.
Observe that this implies that $[\bigvee[x]]=[x]$.

Let us now show that for all p-invariant point $x_{0}$, and for all $y,z\in X^{\otimes_{n}}$, if 
$y\leq x_{0}$ and 
$z\sim^{\alpha}y$ holds, then $z\leq x_{0}$.
\begin{itemize}
\item if $z\sim^{0} y$, then either $z=y$, in which case the claim follows from the hypothesis, or $z=z_{1}\otimes \dots \otimes z_{n}$ and $y=y_{\sigma(1)}\otimes \dots \otimes y_{\sigma(n)}$; then from $y\leq x_{0}$ we deduce $z=\langle \sigma^{-1} \rangle(y)\leq \langle \sigma^{-1}\rangle(x_{0})=x_{0}$.

\item if $z\sim^{\alpha+1}y$ two possibilities arise:
	\begin{enumerate}
	\item if $z\sim^{\alpha}z'\sim^{\alpha}y$, then by IH we have $z'\leq x_{0}$, and again by IH applied to $z'$ we deduce $z\leq x_{0}$;
	\item $z=\bigvee_{i}z_{i}$ and $y=\bigvee_{i}y_{i}$, with $z_{i}\sim^{\alpha}y_{i}$, then from $y_{i}\leq y\leq x_{0}$, we deduce, by IH, $z_{i}\leq x_{0}$, and thus $z\leq x_{0}$.

	\end{enumerate}

\item if $z\sim^{\gamma}y$ for $\gamma$ limit, then $z\sim^{\beta}y$ for some $\beta<\gamma$, so by IH we deduce $z\leq x_{0}$.

\end{itemize}
From the argument above we now deduce that for all p-invariant point $x_{0}$, and for all $y,z\in X^{\otimes_{n}}$, if 
$y\leq x_{0}$ and 
$z\sim_{n}y$ holds, then $z\leq x_{0}$.
From this we can deduce in turn that for all $x\in X^{\otimes_{n}}$, $\bigvee[x]$ is the smallest p-invariant over $x$: suppose $x_{0}$ is a p-invariant point and $x\leq x_{0}$; then for all $y\in [x]$, we deduce $y\leq x_{0}$ by the argument above, and we can thus conclude that $\bigvee[x]\leq x_{0}$.

Let now $x$ be p-invariant; as $x$ is the smallest p-invariant over $x$, we deduce that $x= \bigvee[x]$.

Using the previous facts we can now define an isomorphism $h:!_{n}X\to  \mathrm{Sym}_{n}(X)$ by letting  $h(x)=[x]$, with inverse $k([x])=\bigvee [x]$. Indeed, we have that 
$k(h(x))=\bigvee[x]=x$, and 
$h(k([x]))=[\bigvee[x]]=[x]$.
%
%First, let us show that for all $x\in !_{n}X\subseteq X^{\otimes_{n}}$, the corresponding equivalence class in $\mathrm{Sym}_{n}(X)$ is a singleton, i.e.~$[x]=\{x\}$. 
%To prove this, let us first show that for all ordinals $\alpha$ and p-invariant $x$, if $x \sim_{n}^{(\alpha)}y$ then $x=y$, where $\sim_{n}^{(\alpha)}$ is defined as in Proposition \ref{prop:smallestcongruence}. 
%\begin{itemize}
%\item $x\sim_{n}^{(0)}y$ holds iff either $x=y$, $x\sim_{n}y$ or $y\sim_{n}x$; if $x\sim_{n}y$, then it must be $x=x_{1}\otimes \dots \otimes x_{n}$ and $y=\langle \sigma\rangle(x)$, but since $x$ is permutation-closed, $x=\langle \mathrm{id}\rangle(x)=\langle \sigma\rangle(x)=y$. 
%
%\item $x\sim_{n}^{(\alpha+1)}y$ holds iff either $x\sim_{n}^{(\alpha)}z$ and $z\sim_{n}^{(\alpha)}y$ both hold, or $x=\bigvee_{i}x_{i}$, $y=\bigvee_{i}y_{i}$ and $x_{i}\sim_{n}^{(\alpha)}y_{i}$ all hold.
%In the first case, by IH we have $x=z$, so $z$ is p-invariant, and by applying again the IH, also $y=z$ holds, and thus $x=y$; 
%in the second case, {\color{red}by IH we have $x_{i}=y_{i}$ for all index $i$, whence 
%$x=\bigvee_{i}x_{i}=\bigvee_{i}y_{i}=y$.}
%
%\item if $x\sim_{n}^{(\gamma)}y$ for $\gamma$ limit, then
%$x\sim_{n}^{(\beta)}y$ holds for some $\beta<\gamma$, so by IH, $x=y$.
%
%
%\end{itemize}
%Now, if $x\sim_{n}y$ holds, then $x\sim_{n}^{(\alpha)}y$ holds for some ordinal $\alpha$, whence $x=y$. 
%
%
%
%
%Now, the main claim follows from the existence of the following two morphisms:
%a morphism $h:!_{n}X\to \mathrm{Sym}_{n}(X)$ given by $h(x)=[x]=\{x\}$ and a morphism $k: \mathrm{Sym}_{n}(X)\to !_{n}X$ given by $k(\alpha)=\bigvee\alpha$. Then $k\circ h(x)=x$ while $h\circ k(\alpha)=[\bigvee \alpha]=\{\bigvee\alpha\}$, so $k$ and $h$ define an isomorphism between $!_{n}X$ and the $\sim$-classes of $\mathrm{Sym}_{n}(X)$.
\end{proof}

The following lemma shows the compatibility of the construction of $!_{n}X$ with the usual construction of the exponential modality in weighted relational models.
\begin{lemma}
For any set $S$, there exists an isomorphism of $Q$-modules
$$!_{n}Q^{S} \simeq Q^{\C M_{n}(S)}$$
where $\C M_{n}(X)$ indicates the set of multisets of $X$ of cardinality $ n$.
\end{lemma}
\begin{proof}
Let us show that the morphism $h:Q^{\C M_{n}(S)}\to Q^{S\times \dots \times S}$ defined by 
$$
h(f)(\langle s_{1},\dots, s_{n}\rangle)=h([s_{1},\dots, s_{n}])
$$
is the equalizer of the diagram 
$$
\begin{tikzcd}
Q^{\C M_{n}(S)} \ar{r}{h} &Q^{S\times \dots \times S}\ar{r}{[\sigma]} &
Q^{S\times \dots \times S}
\end{tikzcd}
$$
where $[\sigma](\B x)(\langle s_{1},\dots, s_{n}\rangle)=\B x(\langle x_{\sigma(1)},\dots, x_{\sigma(n)}\rangle)$, with $\sigma$ varying over $\F S_{n}$.

It is immediate that $h\circ [\sigma]=h\circ [\tau]$, for all $\sigma,\tau\in \F S_{n}$. Let now $k: C\to Q^{S\times \dots \times S}$ satisfy $k\circ [\sigma]=k\circ [\tau]$: then for all $c\in C$, $k(c)(\langle s_{1},\dots, s_{n}\rangle)=k(c)(\langle s_{\sigma(1)},\dots, s_{\sigma(n)}\rangle)$, so $k(c)$ actually defines a unique element of $Q^{\C M_{n}(S)}$, and thus $k$ splits in a unique way as $C \stackrel{k'}{\to} Q^{\C M_{n}(S)} \stackrel{h}{\to}Q^{S\times \dots \times S}$.

Now, to conclude it suffices to observe that, by Proposition \ref{prop:Qtensor}, 
$Q^{S\times \dots \times S}\simeq (Q^{S})^{\otimes_{n}}$, and then, since equalizers are unique up to a unique isomorphism, we obtain an isomorphism $Q^{\C M_{n}(S)}\simeq !_{n}Q^{S}$.
%
%We define morphisms $h: !_{n}Q^{S}\to Q^{\C M_{n}(S)}$ and 
%$k: Q^{\C M_{n}(S)}\to !_{n}Q^{S}$ given by 
%\begin{align*}
%h([\B x_{1},\dots, \B x_{n}])(\{a_{1},\dots, a_{n}\})&=
%\sup_{\sigma\in \F S_{n}}\sum_{i=1}^{n}
%\B x_{i}(a_{\sigma(i)})\\
%k(f)& = 
%\bigvee_{a_{1},\dots, a_{n}\in S}[\C Y(a_{1}),\dots, \C Y(a_{n})]+ f(\{a_{1},\dots, a_{n}\})
%\end{align*}
%
\end{proof}

\subsubsection{Linear Differential Categories}

There exist many equivalent way to describe linear differential structure over symmetric monoidal categories with biproducts. Here we chose the approach via bialgebra modalities (see Ehrhard's stuff, but also Lemay and company's stuff for a clearer picture).

\begin{definition}[bialgebra modality]
Let $\BB C$ be an additive symmetric monoidal category. A \emph{bialgebra modality over $\BB C$} is a septuple $(!,\delta,\epsilon, \Delta, e,\nabla, u)$ consisting of:
\begin{enumerate}
\item a comonad $(!,\delta,\epsilon)$, that is, a functor $!$ together with natural transformations $\delta: !X\to !!X$ and $\epsilon:!X\to X$ satisfying
\begin{align}
\epsilon \circ \delta & = !\epsilon \circ \delta= 1 \\
!\delta \circ \delta  & = \delta\circ !\delta
\end{align}

\item two natural transformations $\Delta:!X\to !X\otimes !X$ and $e:!X\to \B 1$ such that $(!X, \Delta, e)$ is a cocommutative comonoid, that is the following equations hold:
\begin{align}
(\Delta \otimes  1)\circ \Delta & = ( 1\otimes \Delta )\circ \Delta \\
 ( 1\otimes e)\circ \Delta & = (e\otimes  1)\circ \Delta =1 \\
\sigma \circ\Delta& =  \Delta
\end{align}
and $\delta$ preserves the comultiplication, that is
\begin{align}
(\delta \otimes \delta)\circ \Delta = \Delta \circ \delta
\end{align}

\item two natural transformations $\nabla: !X\otimes !X \to !X$ and $u:\B 1\to !X$ such that $(!X,\nabla, u)$ is a commutative monoid, that is, the following equations hold:
\begin{align}
\nabla\circ (\nabla\otimes 1)  & = \nabla \circ (1\otimes \nabla) \\
 \nabla \circ(1\otimes u )& = \nabla \circ  (u\otimes 1) =1\\
\nabla\circ \sigma & = \nabla
\end{align}

\item $(!X,\nabla, u, \Delta, e)$ is a bialgebra, that is the following equations hold:
\begin{align}
e\circ\nabla  & =  e\otimes e\\
 \Delta \circ u& = u\otimes u \\
u\circ e & = 1\\
 \Delta \circ \nabla 
&=(\nabla\otimes \nabla)\circ (1\otimes \sigma\otimes 1)\circ (\Delta\otimes \Delta)
\end{align}

\item $\epsilon$ is compatible with $\nabla$, that is
\begin{align}
 \epsilon \circ \nabla& =  (\epsilon\otimes e)+(e\otimes \epsilon)
\end{align}

\end{enumerate}

\end{definition}

\begin{definition}
A \emph{codereliction} for a bialgebra modality $(!,\delta,\epsilon,\Delta, e,\nabla,u)$ is a natural transformation $\eta:X\to !X$ satisfying the following equations:
\begin{align}
e\circ \eta & = 0\\
 \Delta\circ \eta&  = (\eta\otimes u)+(u\otimes \eta) \\
 \epsilon\circ \eta & = 1 \\
\delta \circ \nabla\circ(1\otimes \eta) & =  \nabla \circ 
 (\delta\otimes \eta) \circ  (1\otimes \nabla)\circ
(\Delta \otimes \eta)
\end{align}
\end{definition}

\begin{definition}
Let $\BB C$ be an additive (i.e.~monoid-enriched) symmetric monoidal category with biproducts.
 $\BB C$ is a \emph{monoidal storage category} if it has a coalgebra modality satisfying the \emph{Seely isomorphisms}, that is, the maps
 $e:!\top \to \B 1$ and $\chi:!(X\times Y) \to !X\otimes !Y$, with $\chi=\Delta \circ  !(\pi_{0})\otimes!(\pi_{1})$, are isomorphisms (whence $!\top\simeq \B 1$ and $!(X\times Y)\simeq !X\otimes !Y$).
\end{definition}

\begin{definition}
Let $\BB C$ be an additive symmetric monoidal category. A bialgebra modality $(!,\delta,\epsilon,\Delta,e,\nabla,u)$ on $\BB C$ is \emph{additive} if the following equations hold:
\begin{align}
\nabla\circ (!f\otimes !g)\otimes \Delta  & = !(f+g)\\ 
u\circ e & = !0
\end{align}
\end{definition}

We use the following result:
\begin{theorem}[Lemay\&co]\label{theorem:lemay}
Every additive symmetric monoidal category with an additive bialgebra modality and finite products satisfies the Seely isomorphisms.
\end{theorem}

\begin{definition}
A \emph{linear differential category} is an additive symmetric monoidal category $\BB C$ with biproducts and a bialgebra modality with a codereliction and the Seely isomorphisms.
\end{definition}

\begin{theorem}
For any linear differential category $\BB C$, the co-Kleisli category $\BB C_{!}$ is a cartesian closed differential category, with deriving transformation $\mathsf{D}(f)$ defined as follows:
$$
\begin{tikzcd}
!(X\times X)\  \simeq\  !X\otimes !X 
\ar{r}{1\otimes \epsilon}  & !X\otimes X \ar{r}{1\otimes \eta}&
!X\otimes !X \ar{r}{\nabla}& !X \ar{r}{f} & Y
\end{tikzcd}
$$
\end{theorem}

\subsubsection{The Free Exponential Modality of $Q\mathsf{Mod}$}

Using the recipe from [Melli\'es, Tabareau, Tasson 2010], together with Proposition \ref{prop:productvstensor}, the free exponential modality of $Q\mathsf{Mod}$ can be defined as
$$
! X:= \prod_{n\in \BB N}!_{n}X
$$
The functorial action of $!$ is defined, for a morphism $f: X\to Y$, as follows:
\begin{align*}
!f(g)(0) & = g(0) \\
!f (g)(n+1)   & = \bigvee\Big \{
[f(x_{1}),\dots, f(x_{n+1})] \mid [x_{1}, \dots, x_{n+1}]\leq g(n+1)\Big \}
\end{align*}

%where $g(n)=\bigvee_{i}
%[x_{1}, \dots, x_{n}]$.

The bialgebra modality $(!, \delta, \epsilon, \Delta, e, \nabla,u)$ is defined as follows:
\begin{itemize}
\item the comonad $(!,\delta,\epsilon)$ is given by:
\begin{align*}
\epsilon(f)& =  f(1) \\
%\OV{\mathrm{der}}& = \iota_{1}\\
\delta(f)(n) & = \bigvee\left\{ 
\iota_{n}([\iota_{i_{1}}(a_{1}), \dots , \iota_{i_{n}}(a_{n})]) \ \Big \vert \ 
a_{j}\in !_{i_{j}}X,  
a_{1}\cup \dots \cup a_{n}\leq f(i_{1}+\dots+ i_{n})
\right\}
\end{align*}

We have $\epsilon(\delta(\alpha))=\delta(\alpha)(1)=
\bigvee\{\iota_{n}(\alpha(n))\mid n\in \BB N\}=\alpha$, and 
\begin{align*}
!\epsilon(\delta(\alpha))(n)&=
\bigvee\{ [\epsilon(\alpha_{1}),\dots, \epsilon(\alpha_{n})]
\mid
[\alpha_{1},\dots, \alpha_{n}] \leq \delta(\alpha)(n)\}
 \\
 &=
 \bigvee\left \{ [\epsilon(\alpha_{1}),\dots, \epsilon(\alpha_{n})]
\ \Bigg \vert \ 
\alpha_{i}=\iota_{j_{i}}(a_{i}),\bigcup_{i}a_{i}\leq 
%
%[ \iota_{j_{1}^{i}}(a_{1}^{i}),\dots, \iota_{j_{s_{i}}^{i}}(a_{s_{i}}^{i})], 
%\left [\bigcup_{j}a_{j}^{1},\dots, \bigcup_{j}a_{j}^{n}\right] \leq 
\alpha(n)
\right \} \\
&=
 \bigvee\left \{ [\epsilon(\alpha_{1}),\dots, \epsilon(\alpha_{n})]
\ \Bigg \vert \ 
\alpha_{i}=\iota_{1}(x_{i}),[x_{1},\dots, x_{n}]\leq 
\alpha(n)
\right \} \\
&= \bigvee\left \{ [
x_{1},\dots, x_{n}] \ \Bigg \vert  \ 
[x_{1},\dots, x_{n}]\leq \alpha(n)
\right \} = \alpha(n)
 \end{align*}
%But since $\epsilon(\beta)=\beta(1)$, this means that we are reduced to consider multisets $\alpha_{j}^{i} $ of cardinality 1, i.e.~of the form $[a_{j}^{i}]$, so we have  
% $!\epsilon(\delta(\alpha))(n)=
%\bigvee\{ [a_{1}^{i},\dots, a_{n}^{i}]\mid \bigvee_{i}[a_{1}^{i},\dots, a_{n}^{i}]= \alpha(n)\}= \alpha(n)$.

Let us now compute $!\delta(\delta (\alpha))$: 
{\small
\begin{align*}
!\delta(\delta(\alpha))(n)&=\bigvee_{i}\left \{[\delta(A_{1}),\dots, \delta(A_{n})]\ \Bigg\vert \   [A_{1},\dots, A_{n}]\leq \delta(\alpha)(n)\right\}\\
&=\bigvee\left \{[\delta(\iota_{m_{1}}(a_{1})),\dots, \delta(\iota_{m_{n}}(a_{n}))]\ \Bigg\vert \  
a_{j}\in !_{m_{j}}X, 
\bigcup_{j=1}^{n}a_{j}\leq \alpha(\sum_{l=1}^{n}m_{l})
\right\}\\
&= \bigvee \left\{
[ B_{1},\dots, B_{n}] \ \Bigg \vert \ 
B_{i}=\iota_{\sum_{j}r_{j}^{i}}\big([\iota_{r_{1}^{i}}(b_{1}^{i}),\dots, \iota_{r_{s_{i}}^{i}}(b_{m_{i}}^{i})]\big),\sum_{j}r_{j}^{i}=m_{i}, 
\left [\bigcup_{j}b_{j}^{1},\dots, \bigcup_{j}b_{j}^{n}\right ]\leq \alpha(\sum_{i}m_{i})
\right \}
\end{align*}
}

Let us compute $\delta(\delta(\alpha))$:
{\small
\begin{align*}
\delta(\delta(\alpha))(n) & = 
\bigvee\left \{
[\iota_{i_{1}}(A_{1}),\dots, \iota_{i_{n}}(A_{n})] \ \Bigg \vert \ 
A_{j}\in !_{i_{j}}X, \bigcup_{j}A_{j} \leq \delta(\alpha)(\sum_{j}i_{j})
\right\}\\
&=
\bigvee\left \{
[\iota_{i_{1}}(A_{1}),\dots, \iota_{i_{n}}(A_{n})] \ \Bigg \vert \ 
A_{j}= [\iota_{r_{1}^{j}}(b_{1}^{j}),\dots, \iota_{r_{s_{j}}^{j}}(b_{i_{j}}^{j})], 
\sum_{l} r_{l}^{j} = i_{j}, \left [\bigcup_{l}b_{l}^{1},\dots, \bigcup_{l}b_{l}^{n}\right ]\leq  \alpha(\sum_{j}i_{j})
\right\}
\end{align*}
}

From the two computations it is clear that $!\delta(\delta(\alpha)=\delta(\delta(\alpha))$.

\item the cocommutative comonoid structure $(!X, \Delta, e)$ is  defined as follows:
\begin{align*}
\Delta(f)& = \bigvee
\left\{ \iota_{n}(a)\otimes \iota_{m}(b) \ \Big \vert \  
n,m\in \BB N, a\cup b\leq f(n+m)
\right\}\\
e(f)&= f(0)
\end{align*}

Let us check the relevant equations:

\begin{align*}
(\Delta\otimes 1)\Big(\Delta(\alpha)\Big)&=
(\Delta\otimes 1)\left (\bigvee\left \{\iota_{n}(a)\otimes \iota_{m}(b)\ \Big \vert \ a\cup b \leq \alpha(n+m)\right\}\right)\\
&= 
\bigvee\left \{\iota_{n_{1}}(a_{1})\otimes \iota_{n_{2}}(a_{2})\otimes \iota_{m}(b)\ \Big \vert \ a_{1}\cup a_{2}\cup b \leq \alpha(n_{1}+n_{2}+m)\right\}\\
&= 
\bigvee\left \{\iota_{n}(a)\otimes \iota_{m_{1}}(b_{1})\otimes \iota_{m_{2}}(b_{2})\ \Big \vert \ a\cup b_{1}\cup b_{2} \leq \alpha(n+m_{1}+m_{2})\right\}
\\
&= (1\otimes \Delta)\Big(\Delta(\alpha)\Big)
\end{align*}

\begin{align*}
(e\otimes 1)\Big(\Delta(\alpha)\Big) & = 
( e\otimes 1)\left (\bigvee\left \{\iota_{n}(a)\otimes \iota_{m}(b)\ \Big \vert \ a\cup b \leq \alpha(n+m)\right\}\right)\\
&= 
\bigvee\left \{\iota_{m}(b)\ \Big \vert \ b \leq \alpha(m)\right\} =
\alpha
\end{align*}
and one can argue similarly for $(1\otimes e)(\Delta(\alpha))$.

\begin{align*}
\sigma(\Delta(\alpha)) & = 
\bigvee\left \{\iota_{m}(b)\otimes \iota_{n}(a)\ \Big \vert \ a\cup b \leq \alpha(n+m)\right\}
\\
& = 
\bigvee\left \{\iota_{n}(a)\otimes \iota_{m}(b)\ \Big \vert \ a\cup b \leq \alpha(n+m)\right\}
\\&= \Delta(\alpha)
\end{align*}

Finally, the commutation of $\Delta$ and $\delta$:
{\tiny
\begin{align*}
(\delta\otimes \delta)\Big(\Delta(\alpha)\Big) & =
(\delta\otimes\delta)\left(\bigvee
\left\{
\iota_{n}(a)\otimes \iota_{m}(b) \mid a\cup b \leq \alpha(n+m)
\right\}\right)\\
&=
\bigvee
\left \{\iota_{k_{1}}(
[\iota_{i_{1}}(a_{1}),\dots, \iota_{i_{k_{1}}}(a_{k_{1}})])\otimes
\iota_{k_{2}}(
[\iota_{j_{1}}(b_{1}),\dots, \iota_{j_{k_{2}}}(b_{k_{2}})]) \ \Bigg \vert \
\bigcup_{l}a_{l} \cup \bigcup_{l}b_{l} \leq \alpha\left (\sum_{l}i_{l}+\sum_{l}j_{l}\right)
\right\}
\\
&=
\Delta
\left(\bigvee\left \{
\iota_{n}(
[\iota_{i_{1}}(a_{1}),\dots, \iota_{i_{n}}(a_{n})]) \ \Bigg \vert \ 
\bigcup_{j}a_{j}\leq \alpha(\sum_{j}i_{j})
\right\}\right)
\\
& = \Delta(\delta(\alpha))
\end{align*}
}

\item the commutative monoid structure $(!X,\nabla,u)$ is given by 
\begin{align*}
\nabla( f \otimes g) (n)& = \bigvee_{k+h=n}f(k)\cup g(h) \\ 
u & =  \iota_{0}
\end{align*}
Observe that $\nabla(f\otimes g)=\bigvee_{n,k+h=n}\iota_{n}(f(k)\cup g(h))$.

Let us check the relevant equations:

\begin{align*}
\nabla\left((\nabla\otimes 1)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\otimes \gamma_{k}\right)\right) & =
\nabla \left(
\bigvee_{k,n}
\iota_{n}\left( \bigvee_{u+v=n}\alpha_{k}(u)\cup \beta_{k}(v)
\right)
\otimes \gamma_{k}
\right)\\
&=
\bigvee_{k,n}\iota_{n}\left( \bigvee_{u+v+w=n}
\alpha_{k}(u)\cup \beta_{k}(v)\cup \gamma_{k}(w)\right) \\
& =
\nabla \left(
\bigvee_{n,k}\alpha_{k}\otimes 
\iota_{n}\left( \bigvee_{v+w=n}\beta_{k}(v)\cup \gamma_{k}(w)\right)
\right)\\
&= 
\nabla\left((1\otimes\nabla)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\otimes \gamma_{k}\right)\right) 
\end{align*}

\begin{align*}
\nabla\left(
(1\otimes u)\left(\bigvee_{k}\alpha_{k}\otimes 0\right)\right) & =
\nabla\left( \bigvee_{k}\alpha_{k}\otimes\iota_{0}(\epsilon)\right)
=
\bigvee_{n,k}\iota_{n}( \alpha_{k}(n) )
 =  \bigvee_{k}\alpha_{k} \otimes 0
\end{align*}
and one can argue similarly for $\nabla(u\otimes 1)(\alpha)=\alpha$.

\begin{align*}
\nabla\left (\sigma\left (\bigvee_{k}\alpha_{k}\otimes \beta_{k}
\right) \right) & = 
\nabla \left (
\bigvee_{k}\beta_{k}\otimes \alpha_{k}
\right)\\
&= 
\bigvee_{k,n, p+q=n} \iota_{n}(\beta_{k}(p)\cup \alpha_{k}(q))
\\
&= 
\bigvee_{k,n, p+q=n}\iota_{n}( \alpha_{k}(p)\cup \beta_{k}(q))
\\
&= \nabla\left (\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)
\end{align*}

Let us check the bialgebra equations:

\begin{align*}
e\left (\nabla\left (\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)\right)  & =
e\left(
\bigvee_{k,n, p+q=n}\iota_{n}(\alpha_{k}(p)\cup \beta_{k}(q))
\right)
\\
&= \bigvee_{k}\alpha_{k}(0)\cup\beta_{k}(0) \\
&= \bigvee_{k} 0\cup 0\\
&= h_{Q}\left(\bigvee_{k}\alpha_{k}(0)\otimes\beta_{k}(0))\right) = h\left( (e\otimes e)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)\right)
\end{align*}
where $h_{Q}:Q\otimes Q \to Q$ indicates the isomorphism $h(\epsilon\otimes \delta)=\epsilon+\delta$.

\begin{align*}
\Delta(u(\epsilon))) = \Delta(\iota_{0}(\epsilon)) &=
 \left(
\bigvee\left\{\iota_{m}(a)\otimes\iota_{n}(b) \ \Bigg \vert \ 
a\cup b \leq \iota_{0}(\epsilon)(n+m)\right\}\right)\\
&=\bigvee\left\{\iota_{0}(\epsilon_{1})\otimes\iota_{0}(\epsilon_{2}) \ \Bigg \vert \ 
\epsilon_{1}+\epsilon_{2} \leq \epsilon\right\}\\
&= (u\otimes u)(\epsilon)
\end{align*}
where we are identifying $\epsilon \in q$ with the equivalence class
$\epsilon \otimes 0 = \{ \langle \epsilon_{1},\epsilon_{2}\rangle\mid \epsilon_{1}+\epsilon_{2}=\epsilon\}\in 
Q\otimes Q$ via the isomorphism $h_{Q}$ described above.

\begin{align*}
e(u(\epsilon))&= \iota_{0}(\epsilon)(0)= \epsilon
\end{align*}

{\tiny
\begin{align*}
&(\nabla\otimes \nabla)  (1\otimes \sigma\otimes 1)(\Delta\otimes \Delta)\left (\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right) \\
&=
(\nabla\otimes \nabla)(1\otimes \sigma\otimes 1)
\left(
\bigvee
\left\{
 \iota_{n}(a_{1})\otimes \iota_{m}(a_{2})\mid
a_{1}\cup a_{2}\leq \alpha_{k}(n+m)
\right\}\otimes
\left\{
 \iota_{n}(b_{1})\otimes \iota_{m}(b_{2})\mid
b_{1}\cup b_{2}\leq \beta_{k}(n+m)
\right\}
\right)\\
&=
(\nabla\otimes \nabla)(1\otimes \sigma\otimes 1)
\left(
\bigvee
\left\{
 \iota_{n}(a_{1})\otimes \iota_{m}(a_{2})
\otimes \iota_{n'}(b_{1})\otimes \iota_{m'}(b_{2})
\mid
a_{1}\cup a_{2}\leq \alpha_{k}(n+m),
b_{1}\cup b_{2}\leq \beta_{k}(n'+m')
\right\}
\right)\\
&=
(\nabla\otimes \nabla)\left(
\bigvee
\left\{
 \iota_{n}(a_{1})\otimes \iota_{n'}(b_{1})
\otimes \iota_{m}(a_{2})\otimes \iota_{m'}(b_{2})
\mid
a_{1}\cup a_{2}\leq \alpha_{k}(n+m),
b_{1}\cup b_{2}\leq \beta_{k}(n'+m')
\right\}
\right)\\
&=
\bigvee_{k,z,z'}\left\{
\left(
\iota_{z}\left( \bigvee_{p+q=z}
a_{1}\cup b_{1}\right)\right)\otimes
\left(\iota_{z'}\left( \bigvee_{p'+q'=z'}
a_{2}\cup b_{2}\right)\right)
\ \Bigg \vert \
a_{1}\cup a_{2} \leq \alpha_{k}(p+p'),
b_{1}\cup b_{2}\leq \beta_{k}(q+q')
\right\}
\\
&=\Delta\left(
\bigvee_{k,z}
\left\{
\iota_{z}\left( \bigvee_{p+q=z} \alpha_{k}(p)\cup \beta_{k}(q) \right)
\right\}
\right)
\\
&=
\Delta\left (\nabla\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right) \right)
\end{align*}
}

Finally, let us check the compatibility of $\epsilon$ and $\nabla$, which in $Q\mathsf{Mod}$ reads as
$ \epsilon\circ \nabla=(\epsilon \otimes e)\vee (e\otimes \epsilon)$:
\begin{align*}
\epsilon \left(\nabla\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)\right) & =
\epsilon\left (  \bigvee_{k,n,p+q=n}\iota_{n}(\alpha_{k}(p)\cup \beta_{k}(q))\right) \\
&=\left(\bigvee_{k}\alpha_{k}(1) \right) \vee \left(
\bigvee_{k}\beta_{k}(1)\right)
\\
&=
\left( (\epsilon\otimes e)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)
\right)
\vee
\left( (e\otimes \epsilon)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)
\right)
\\
&=
\Big((\epsilon\otimes e)\vee (e\otimes \epsilon)\Big)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)
 \end{align*}

\item the codereliction $\eta: X\to !X$ is defined by 
$\eta= \iota_{1}$.

Let us check the codereliction equations:
\begin{align*}
e(\eta(x)) & = \iota_{1}(x)(0)=0
\end{align*}
\begin{align*}
\Delta(\eta(x)) & = \Delta(\iota_{1}(x)) \\
&=
\bigvee\left\{\iota_{1}(x)\otimes \iota_{0}(r) \mid r\in Q\right\}\vee
\bigvee\left\{\iota_{0}(r)\otimes \iota_{1}(x) \mid r\in Q\right\}
\\
&=
(\iota_{1}(x)\otimes 0 )\vee
(0\otimes \iota_{1}(x) )
\\
&=
(\eta\otimes u)(x)\vee(u\otimes \eta) (x) \\
&=
\Big(\eta\otimes u)\vee(u\otimes \eta)\Big) (x) 
\end{align*}

\begin{align*}
\epsilon(\eta(x))&= \iota_{1}(x)(1) =x
\end{align*}

For the last equation, we only check it on basic tensors:
\begin{align*}
& \nabla((\delta\otimes \eta)((1\otimes \nabla))(\Delta\otimes \eta)(\alpha\otimes x)) \\
&=\nabla((\delta\otimes \eta)((1\otimes \nabla))\left(
\bigvee
\{\iota_{n}(a)\otimes \iota_{m}(b)\otimes \iota_{1}(x)\mid a\cup b \leq \alpha(n+m)\}
\right)\\ 
&=\nabla\left ((\delta\otimes \eta)\left(
\bigvee
\{\iota_{n}(a)\otimes \iota_{m+1}(b\cup \{x\})\mid a\cup b \leq \alpha(n+m)\}\right)\right)
\\ 
&=\nabla\left (
\bigvee\left 
\{[\iota_{i_{1}}(c_{1}),\dots,\iota_{i_{r}}(c_{r})]
\otimes \iota_{1}( \iota_{m+1}(b\cup \{x\}))\ \Bigg \vert \ 
\bigcup_{j}c_{j}\cup b \leq \alpha\left (\sum_{j}i_{j}+m\right )\right\}\right)
\\ 
&=
\bigvee\left\{
[\iota_{i_{1}}(c_{1}),\dots,\iota_{i_{r}}(c_{r}), \iota_{m+1}(b\cup \{x\})] \ \Bigg \vert \
\bigcup_{j}c_{j}\cup b \leq \alpha\left (\sum_{j}i_{j}+m\right )\right\}
\\
&=
\delta\left(
\bigvee_{n}\iota_{n+1}(
\alpha(n)\cup \{x\})
\right)
\\
&=
\delta(\nabla(\alpha\otimes \iota_{1}(x))
\\
&=
\delta(\nabla((1\otimes \eta)(\alpha\otimes x))) 
\end{align*}

\end{itemize}

It remains to check the Seely isomorphisms.
Using Theorem \ref{theorem:lemay} it suffices to check that the bialgebra modality defined above is additive (with respect to the ``tropical'' additive structure given by $\bot$ and $\vee$).

{\tiny
\begin{align*}
\nabla((!f\otimes !g)(\Delta(\alpha))) & =
\nabla\left((!f\otimes !g)\left(
\bigvee\{\iota_{n}([x_{1},\dots, x_{n}])\otimes \iota_{m}([y_{1},\dots, y_{m}])\mid [x_{1},\dots, x_{n},y_{1},\dots y_{m}]\leq \alpha(n+m)\}
\right)\right)\\
&=
\nabla\left(
\bigvee\{\iota_{n}([f(x_{1}),\dots, f(x_{n})])\otimes \iota_{m}([g(y_{1}),\dots, g(y_{m})])\mid [x_{1},\dots, x_{n},y_{1},\dots y_{m}]\leq \alpha(n+m)\}
\right)\\
&=
\bigvee\{\iota_{n+m}([f(x_{1}),\dots, f(x_{n}),g(y_{1}),\dots, g(y_{m})])\mid [x_{1},\dots, x_{n},y_{1},\dots y_{m}]\leq \alpha(n+m)\}\\
&=
\bigvee_{n}\{ \iota_{n}([f(x_{1})\vee g(x_{1}),\dots, f(x_{n})\vee g(x_{1})]
\mid [x_{1},\dots, x_{n}]\leq \alpha(n)\}\\
&=
!(f\vee g)(\alpha)
\end{align*}
}
\begin{align*}
u(e(\alpha)) = u(\alpha(0))& =\iota_{0}(\alpha({0})) =
\iota_{0}(\alpha(0))\vee
\bigvee\{\iota_{n+1}([\underbrace{\bot,\dots, \bot}_{n+1}])\mid n\in \BB N \}=
(!\bot) (\alpha)
\end{align*}

In particular, any $\alpha\in !(X\times Y)$ can be represented 
as an object $\alpha^{S}$ of $!X\otimes !Y$ defined as follows:
$$
\alpha^{S}=\bigvee \{ \iota_{n}([x_{1},\dots, x_{n}])\otimes \iota_{m}([y_{1},\dots, y_{m}]) \mid
\iota_{n+m}(\langle x_{1},\bot\rangle,\dots, 
\langle x_{n},\bot\rangle,
\langle \bot, y_{1}\rangle, \dots,
\langle \bot, y_{m}\rangle) \leq \alpha(n+m)
\}
$$

\begin{theorem}
$Q\mathsf{Mod}$ (equivalently, $Q\mathsf{CCat}$) is a linear differential category. Hence $Q\mathsf{Mod}_{!}$ (equivalently, $Q\mathsf{CCat}_{!}$) is a cartesian closed differential category. 
\end{theorem}

Let us compute the differential operator in $Q\mathsf{Mod}_{!}$: 
given $f:!X \to Y$, we have
\begin{align*}
\mathsf{D}[f](\alpha) & = 
\bigvee 
\left \{
f(\beta\cup[x])
\ \Big \vert \ 
\iota_{n}(\beta)\otimes \iota_{1}(x) \leq \alpha^{S}
\right\}
\end{align*}

%\begin{definition}
%A coalgebra modality $(!,\delta,\epsilon,\Delta,e)$ is a \emph{monoidal coalgebra modality} if there exist natural transformations $m_{\otimes}:!X\otimes !Y \to !(X\otimes Y)$ 
%and $m_{1}: \B 1 \to !\B 1$ such that 
%\begin{itemize}
%\item $m_{\otimes},m_{k}$ show $(!,\delta,\epsilon)$ as a symmetric monoidal comonad, that is, the following equations hold:
%
%
%\item the following further equations hold:
%\begin{align}
%\Delta \circ m_{\otimes} & = (m_{\otimes}\otimes m_{\otimes})\circ (1\otimes \sigma \otimes 1) \circ (\Delta \otimes \Delta) \\
%
%
%\end{align}
%
%\end{itemize}
%
%\end{definition}

%
%the morphism $\chi= (!\pi_{0}\otimes !\pi_{1})\circ \Delta$ is given by 
%%\begin{align*}
%%\chi(\alpha)&=(!\pi_{0}\otimes !\pi_{1})\left(
%\bigvee\{
%\iota_{n}(a)\otimes \iota_{m}(b) \mid
%a\cup b \leq \alpha(n+m)
%\}\right)\\
%&=
%\bigvee\{
%\iota_{n}(a)\otimes \iota_{m}(b) \mid
%a\cup b \leq \alpha(n+m)
%\}\right)\\
%\end{align*}

%$m_{\otimes}(\alpha\otimes \beta)(n)= \alpha(n)\otimes \beta(n)$
%
%
%
%$m_{Q}(\epsilon)=\iota_{1}(\epsilon)$.
%
%
%

\newpage

\subsection{$Q$-Rel is a model of Bounded Linear Logic}

Given SMCs $\BB C,\BB D$,  let $\B{SMC}_{l}(\BB C, \BB D)$ indicate the category of symmetric lax monoidal functors and monoidal natural transformations between them.
$\B{SMC}_{l}(\BB C, \BB D)$  is itself a SMC, with monoidal structure defined pointwise.

The set 
$\BB N$ can be seen as the category with objects the natural numbers and a morphism between $r$ and $r'$ precisely when $r\leq r'$. 
Moreover, $\BB N$ can be seen as a SMC in two ways:
\begin{itemize}

\item we indicate as $\BB N^{+}$ the SMC with monoidal product given by addition;
\item we indicate as $\BB N^{\times}$ the SMC with monoidal product given by multiplication.
\end{itemize}

%Below we let $\BB^{\min}$ indicate the monoid $(Q,\min,0)$ and $Q^{+}$ indicate the monoid $(Q,+,0)$.

\begin{definition}
A \emph{$\BB N$-graded linear exponential comonad} on a symmetric monoidal category $\BB C$ is a tuple
$(D, w,c,\epsilon,\delta)$ where:
\begin{itemize}

\item $D: \BB N\to \B{SMC}_{l}(\BB C, \BB C)$ is a functor. We write 
$m_{r}:\B 1 \to D(r)(\B 1)$ and $m_{r,A,B}: D(r)(A)\otimes D(r)(B) \to D(r)(A\otimes B)$ for the symmetric lax monoidal structure of $D(r)$;

\item $(D,w,c): \BB N^{+}\to \B{SMC}_{l}(\BB C, \BB C)$ is a symmetric colax monoidal functor;

\item $(D,\epsilon,\delta):\BB N^{\times}\to (\B{SMC}_{l},\mathrm{Id},\circ) $ is a colax monoidal functor.

\end{itemize}
further satisfying the axioms below:
\begin{align}
w_{A}& =  w_{D(s)(A)}\circ \delta_{0,s,A}\\
w_{A} & = D(s)(w_{A} )\circ \delta_{s,0,A} \\
(\delta_{r,s,A}\otimes \delta_{r',s,A})\circ c_{rs,r's,A}
&=
c_{r,r',D(s)(A)}\circ \delta_{r+r',s,A}\\
m_{s,D(r)(A),D(r')(A)}\circ (\delta_{r,s,A}\otimes \delta_{s,r',A})\circ c_{sr,sr',A}&=
D(s)(c_{r,r',A})\circ \delta_{s,r+r',A}
\end{align}
\end{definition}

Concretely, the definition above requires 6 natural transformations:
\begin{align*}
m_{r} & :\B 1\to  D(r)(\B 1)\\
m_{r,A,B}& :  D(r)(A)\otimes D(r)(B)\to  D(r)(A\otimes B)\\
w_{A}& : D(0)(A)\to \B 1 \\
c_{r,r',A}& : D(r+r')(A) \to D(r)(A)\otimes D(r')(A) \\
\epsilon_{A}& : D(1)(A)\to A \\
\delta_{r,r',A}&: D(r r')(A)\to D(r)(D(r')(A))
\end{align*}
subject to the following list of equations:
\begin{itemize}
\item $D(r)$ is a lax monoidal functor:
\begin{align}
m_{r,A\otimes B,C}\circ (m_{r,A,B}\otimes D(r)(C)) & = 
m_{r,A, B\otimes C}\circ (D(r)(A)\otimes m_{r,B,C})\\
m_{r,A,\B 1}\circ (D(r)(A)\otimes m_{r}) & = D(r)(A)\\
m_{r,\B 1, B}\circ (m_{r}\otimes D(r)(B))&= D(r)(B)
\end{align}

\item $(D,w,c)$ is a symmetric colax monoidal functor:
\begin{align}
(c_{r,s,-}\otimes D(t)(-))\circ c_{r+s,t} & =
(D(r)(-)\otimes c_{s,t,-})\circ c_{r,s+t}\\
(D(r)(-)\otimes w_{-})\circ c_{r,0,-} & = D(r)(-) \\
(w_{-}\otimes D(r)(-))\circ c_{0,r,-} & = D(r)(-)
\end{align}

\item $(D,\epsilon,\delta)$ is a colax monoidal functor:
\begin{align}
\delta_{r,s, D(t)(-)}\circ \delta_{(rs),t,-} & =
D(r)(\delta_{s,t,-})\circ \delta_{r,st,-}\\
D(r)(\epsilon_{-}) \circ \delta_{r,1,-} & = D(r)(-) \\
\epsilon_{D(r)(-)} \circ \delta_{1,r,-} & = D(r)(-)
\end{align}

\end{itemize}

\begin{definition}
We define the following structure $(M,w,c,\epsilon,\delta)$ over the category $Q\mathsf{Rel}$ as follows:
\begin{itemize}
\item for any set $X$ and $n\in \BB N$, let $M(n)(X)=\C M_{\leq n}(X)$;

\item for all $f: X\times Y\to Q$, let $M(n)(f): M(n)(X)\times M(n)(Y)\to Q$ be defined by 
\begin{align*}
M(n)(f)(\alpha,\beta)=
\begin{cases}
\min_{\sigma\in \F S_{k}}\sum_{i=1}^{k}f(x_{i},y_{\sigma(i)}) & 
\text{ if }\alpha=[x_{1},\dots, x_{k}], \beta=[y_{1},\dots, y_{k}]\\
\infty & \text{ otherwise}
\end{cases}
\end{align*}

\item $m_{r}(\star, \{\star\})=0$ and $m_{r}(\star, \emptyset)=\infty$;

\item $m_{r,A,B}: D(r)(A)\times D(r)(B)\times D(r)(A\times B)\to Q$ is defined by 
\begin{align*}
m_{r,A,B}((\alpha,\beta), \gamma)=
\begin{cases}
0 & \text{ if } \alpha=[x_{1},\dots, x_{k}], \beta=[y_{1},\dots, y_{k}], \gamma= [(x_{1},y_{1}),\dots, (x_{k},y_{k})]\\
\infty & \text{ otherwise}
\end{cases}
\end{align*}

\item $w_{A}:D(0)(A)\times \B 1\to Q$ is given by $w_{A}(\emptyset, \star)=0$ and is $\infty$ otherwise (observe that $D(0)(A)\simeq \B 1$);

\item $c_{r,s,A}: D(r+s)(A)\times D(r)(A)\times D(s)(A)\to Q$ is given by $c_{r,r',A}(\langle\alpha, \beta,\gamma\rangle)=0$ if $\alpha=\beta+\gamma$, and is $\infty$ otherwise;

\item $\epsilon_{A}(\emptyset, a)=\infty$, $\epsilon_{A}([a],a)=0$, $\epsilon_{A}([b],a)=\infty$ $(b\neq a)$,

\item $\delta_{r,r',A}(\alpha, B)=0$ if $\alpha= \sum B$ (where $\sum B$ indicates the multiset obtained by the sum of all multisets contained in $B$) and is $\infty$ otherwise.

\end{itemize}

\end{definition}

Let us check that this defines a $\BB N$-graded linear exponential comonad.

\begin{itemize}

\item $D(r)$ is a lax monoidal functor:
 $$ m_{r,A\times B,C}\circ (m_{r,A,B}\times D(r)(C))(\langle \alpha,\beta,\gamma,\delta\rangle)
 :
 D(r)(A)\times D(r)(B)\times D(r)(C) \times D(r)(A\times B\times C)\to Q
 $$
 is equal to $0$ 
precisely when $\alpha=[x_{1},\dots, x_{k}]$, $\beta=[y_{1},\dots, y_{k}]$, $\gamma=[z_{1},\dots, z_{k}]$ and 
$\delta= [(x_{1},y_{1},z_{1}),\dots, (x_{k},y_{k},z_{k})]$, and is $\infty $ in all other cases.

Observe that
$m_{r,A,B\times C}\circ (D(r)(A)\times m_{r,B,C})(\langle\alpha,\beta,\gamma, \delta\rangle)
  $ is equal to $0$ in the same situation, and is $\infty$ otherwise.

 We conclude that the two matrices coincide.

 Furthermore, we have that 
 $m_{r,A,\B 1}\circ (D(r)(A)\times m_{r})(\langle \alpha,\beta\rangle): D(r)(A)\times \B 1 \times D(r)(A)$ is equal to $0$ 
 precisely when $\alpha=\beta$ and is $\infty$ otherwise, that is, it coincides with $\mathrm{id}_{D(r)(A)}$.

\item $(D,w,c)$ is a symmetric colax monoidal functor.

$((c_{r,s,A}\times D(t)(A))\circ c_{r+s,t,A}) (\langle \alpha,\beta,\gamma,\delta\rangle)
: D(r+s+t)(A)\times D(r)(A)\times D(s)(A)\times D(t)(A)$
is equal to $0$ when $\alpha=\beta+\gamma+\delta$, and is $\infty$ otherwise, and the same holds for
$((D(r)(A)\times c_{s,t,A})\circ c_{r,s+t,A}) (\langle \alpha,\beta,\gamma,\delta\rangle)
$.

Furthermore, 
$((D(r)(A)\times w_{A})\circ c_{r,0,A})(\alpha,\beta)
: D(r)(A)\times D(r)(A)\to Q$ is equal to $0$ when 
$\alpha=\beta$, and is $\infty$ otherwise, so it coincides with 
$\mathrm{id}_{D(r)(A)}$.

\item $(D,\epsilon,\delta)$ is a colax monoidal functor:

$(\delta_{r,s,D(t)(A)}\circ \delta_{rs,t,A})
(\alpha, \Phi)
: D(rst)(A) \times D(r)(D(s)(D(t)(A))) \to Q
$
is $0$ precisely when $\alpha = \sum \sum \Phi$, and is $\infty$ otherwise, and similarly for 
$(D(r)(\delta_{s,t,A})\circ \delta_{r,st,A})(\alpha,\Phi)$.

Furthermore, 
$(D(r)(\epsilon_{A})\circ \delta_{r,1})( \alpha,\beta ):
D(r)(A) \times D(r)(A)\to Q
$ is equal to $0$ when $\alpha=\beta$ and is $\infty$ otherwise, so it coincides with $\mathrm{id}_{D(r)(A)}$.

\end{itemize}

Let us check the further equations:
\begin{itemize}

\item $(w_{D(s)(A)}\circ \delta_{0,s,A})(\langle \emptyset,\star\rangle: D(0)(A)\times \B 1\to Q$ is 0, precisely like $w_{A}$.

\item A similar argument holds for the second equation.

\item $((\delta_{r,s,A}\times \delta_{r',s,a})\circ c_{rs,r's,A})
(\langle \alpha, \Phi,\Psi  \rangle)
:
D(rs+r's)(A)\times  D(r)(s)(A)\times D(r')(s)(A)\to Q
$
is equal to $0$ when $\alpha=\sum \Phi + \sum \Psi$, and is $\infty$ otherwise.

Now, 
using the fact that $D(rs+r's)(A)=D((r+r')s)(A)$, we can check that the same holds for 
$c_{r,r',D(s)(A)}\circ \delta_{r+r',s,A})(\langle \alpha, \Phi,\Psi  \rangle)$: it is $0$ when 
$\alpha= \sum\Phi+\Psi= \sum \Phi+\sum \Psi$.

\item A similar argument holds for the fourth equation.

\end{itemize}

### 2023-01-19 appV.tex
commit

@@
+\subsubsection{Concavity}
+
+A function $f:Q^{X}\to Q^{Y}$ is \emph{concave} if for all $\alpha\in [0,1]$, $\B x, \B y\in Q^{X}$ and $b\in Y$ 
+$$
+f(\alpha\cdot \B x+(1-\alpha)\cdot \B y)_{b} \geq \alpha f(\B x)_{b} + (1-\alpha)f(\B y)_{b}
+$$
+
+\begin{proposition}
+All tropical functions $f: Q^{X}\to Q^{Y}$ are concave.
+\end{proposition}
+\begin{proof}
+Let us first show that all functions of the form $f(\B x)_{b}= \mu \B x+ c$ are concave:
+we have $f(\alpha\B x+ (1-\alpha)\B y)_{b}= \mu(\alpha\B x)+(1-\alpha)\B y)+c=
+ \mu(\alpha\B x)+(1-\alpha)\B y)+\alpha c+(1-\alpha)c=
+ \alpha(\mu \B x + c)+(1-\alpha)(\mu \B y+c)=\alpha f(\B x)_{b}+(1-\alpha) f(\B x)_{b}$.
+
+To conclude, let us show that if $(f_{i})_{i\in I}$ is a family of concave functions from $Q^{X}$ to $Q^{Y}$, the function $f=\inf_{i\in I}f_{i}$ is also concave: we have
+$f(\alpha\B x +(1-\alpha)\B y)_{b}=
+\inf_{i\in I}f_{i}(\alpha\B x+(1-\alpha)\B y)_{b} \geq 
+\inf_{i\in I}\alpha f_{i}(\B x)_{b}+(1-\alpha)f_{i}(\B y)_{b}
+\geq 
+\inf_{i\in I}\alpha f_{i}(\B x)_{b} + \inf_{j\in I}(1-\alpha)f_{j}(\B y)_{b}
+=
+\alpha \cdot (\inf_{i\in I}f_{i}(\B x)_{b})+ (1-\alpha)\cdot( \inf_{j\in I}f_{j}(\B y)_{b})=
+\alpha  f(\B x)_{b}+(1-\alpha)f(\B y)_{b}$, where we used the fact that given families $a_{i},b_{i}$ of reals,
+$\inf_{i}a_{i}+b_{i}\geq \inf_{i}a_{i}+\inf_{j}b_{j}$.
+This follows from the fact that for all $i\in I$, $a_{i}+b_{i}\geq \inf_{i}a_{i}+\inf_{i}b_{i}$.
+\end{proof}
+
+Concavity is a useful property, as it leads to establish the following:
+\begin{proposition}
+For any tropical function $f: Q^{X}\to Q^{Y}$, 
+\begin{enumerate}
+\item $ \alpha f(\B x)_{b}\leq  f(\alpha\B x)_{b}$;
+\item $f(\B x+\B y)_{b}\leq f(\B x)_{b}+f(\B y)_{b}$;
+\item $f(\B x+\B y)_{b} \leq f(\B x)_{b} + \phi_{\mu,b}(\B y)$ 
+for all $\mu\in \C M_{\mathrm{fin}}(X)$.
+
+\item if $\B x < \B y < \B z$, then 
+$$
+\frac{f(\B y)-f(\B x)}{|\B y-\B x|} \leq \frac{f(\B z)-f(\B x)}{|\B z-\B x|}
+\leq \frac{f(\B z)-f(\B y)}{|\B z-\B y|}
+$$
+
+\item the function $R_{b}(\B x, \B y)= \frac{f(\B x)_{b}-f(\B y)_{b}}{| \B x- \B y|}$ is monotonically non-decreasing in $\B x$ (for fixed $\B y$) and in $\B y$ (for fixed $\B x$).
+
+\end{enumerate}
+\end{proposition}
+\begin{proof}
+Since $f$ is concave, we have that for all $\B z$, 
+$f(\alpha\cdot\B z)=f(\alpha\cdot\B z+(1-\alpha)\cdot 0)\geq 
+\alpha f(\B z)+(1-\alpha)f(0)\geq \alpha f(\B z)$.
+
+For the second one, 
+if $f(\B x)=\mu \B x+c$, then $f(\B x+\B y)= \mu \B x+ \mu \B y + c  \leq
+\mu \B x + \mu \B y + 2c = f(\B x)+f(\B y)$.
+Moreover, if $(f_{i})_{i\in I}$ is a family of functions such that $f_{i}(\B x+\B y)\leq f_{i}(\B x)+f_{i}(\B y)$, then 
+$f=\inf _{i\in I}f_{i}$ also satisfies this property: 
+one has $f(\B x+\B y)=\inf_{i\in I}f_{i}(\B x+\B y) \geq \inf_{i} f_{i}(\B x)+f_{i}(\B y) \geq \inf_{i}f_{i}(\B x) + \inf_{i}f_{i}(\B y)=f(\B x)+f(\B y)$.
+%Now we can compute
+%\begin{align*}
+%f(\B x+\B y)_{b}& =
+%
+%\end{align*}
+
+The third one follows immediately from the second one, since $f(\B y)_{b}\leq \phi_{\mu,b}(\B y)$ for all $\mu\in \C M_{\mathrm{fin}}(X)$.
+
+\end{proof}
+
 \subsubsection{Proof of Theorem \ref{theorem:fepsilon}}
 
-Also proof of 
-\begin{proposition}\label{prop:nondecr+conc}
- Any tLs $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
-\end{proposition}
+DAVIDE's NOTES!
 
@@
 
-Continuity of cones and metric
+Continuity of cones and metric. DAVIDE'S NOTES
+!
 
@@
 \end{proposition}
+\begin{proof}[proof TO BE FIXED]
+All functions of the form $f(\B x)(b)= \mu_{b} \B x + c_{b}$ are Lipschitz with Lipschitz constant
+$\sum_{a\in X}\mu_{b}(a)$: 
+one has $| f(\B x)(b)- f(\B y)(b) |= |\mu_{b}\B x+c_{b} -\mu_{b}\B y+c_{b}|=
+|\mu_{b}\B x -\mu_{b} \B y  |\leq \left(\sum_{a\in X}\mu_{b}(a)\right)\cdot \sup_{a\in X}|\B x_{a}-\B y_{a}|$.
+
+\end{proof}
 
@@
 \end{proof}
+
+\subsubsection{Characterizations of the functional metric }
+
+  \begin{proposition}\label{prop:functionalmetric}
+For all maps $f,g: Q\langle X\rangle\to Q\langle Y\rangle$,  
+$$
+  \|\widehat f- \widehat g\|_{\infty}=
+  \sup\{\|f(\B x)- g(\B x)\|_{\infty} \mid \B x\in Q\langle X\rangle\}
+$$
+
+\end{proposition}
+
+Before proving the proposition we need one preliminary lemma:
+
+\begin{lemma}
+Let $u,v: I\to Q$ and suppose $|u(i)-v(i)|\leq \delta$, for all $i\in I$.
+Then $|\inf_{i\in I}u(i)- \inf_{i\in I}v(i)|\leq \delta$. 
+\end{lemma}
+\begin{proof}
+Let $A=\inf_{i\in I}u(i)$ and $B=\inf_{i\in I}v(i)$ and suppose $A\geq B$. 
+Suppose by way of contradiction $|A-B|> \delta$; then there exists $i\in I$ such that 
+$v(i)<A$ and $|A-v(i)|> \delta$. Indeed, otherwise we would have $|A-B|= \sup\{ |A-v(i)|\mid v(i)\leq A\}\leq \delta$. 
+Now, from $|A-v(i)|> \delta$ and $v(i)< A$ we deduce that $|u(j)-v(i)|> \delta$ for all $j\in I$, and thus in particular that $|u(i)-v(i)|>\delta$, against the assumption.
+We conclude then $|A-B| \leq \delta$.
+In case $B\geq A$, we can argue in a similar way. 
+\end{proof}
+
+%
+%\begin{lemma}
+%Let $f,g\in\mathsf{Trop}(1,1)$ (i.e.~$f,g:Q\to Q$), and suppose for all $n\in \BB N$, 
+%$|\widehat f_{n}-\widehat g_{n}|\leq \delta$. Then for all $x\in Q$, $|f(x)-g(x)|\leq \delta$.
+%
+%\end{lemma}
+%\begin{proof}
+%$|f(x)-g(x)|= |( \inf_{n}nx+\widehat f_{n})- (\inf_{n}nx+\widehat g_{n})|$. 
+%Since $|nx+\widehat f_{n} - nx -\widehat g_{n}|= |\widehat f_{n}-\widehat g_{n}|\leq \delta$, by the Lemma above we conclude $|f(x)-g(x)|\leq \delta$. 
+%\end{proof}
+
+  \begin{proof}[Proof of Proposition \ref{prop:functionalmetric}]
+For one side, suppose for all $\mu\in \C  M_{\mathrm{fin}}(X),b\in Y$, 
+$|\widehat f_{\mu,b}-\widehat g_{\mu,b}|\leq \delta$. Then for all $\B x\in Q\langle X\rangle$ and $b\in Y$, $|f(\B x)(b)-g(\B x)(b)|\leq \delta$.
+Indeed, $|f(\B x)(b)-g(\B x)(b)|= |( \inf_{\mu}\mu \B x+\widehat f_{\mu,b})- (\inf_{\mu}\mu \B x+\widehat g_{\mu,b})|$. 
+Since $|\mu \B x+\widehat f_{\mu,b} - \mu \B x -\widehat g_{\mu,b}|= |\widehat f_{\mu,b}-\widehat g_{\mu,b}|\leq \delta$, by the Lemma above we conclude $|f(\B x)(b)-g(\B x)(b)|\leq \delta$.
+
+For the other side, suppose for some $\mu\in \C M_{\mathrm f}(X)$ and $b\in Y$, $|\widehat f_{\mu,b}-\widehat g_{\mu,b}|> \epsilon$; 
+then, by letting $e_{\mu}\in Q\langle !X\rangle$ be the (tropical) characteristic function of $\mu$, we have
+ $|f(e_{\mu})({b})-g(e_{\mu})({b})| =
+|( \inf_{\mu'}\widehat f_{\mu',b}+ \mu'(e_{\mu}))-
+( \inf_{\mu'}\widehat g_{\mu',b}+\mu'(e_{\mu}))|=
+|\widehat f_{\mu,b}- \widehat g_{\mu,b}|> \epsilon$, so we deduce that 
+$\sup\{\| f(\B x)- g(\B x) \|_{\infty}\mid \B x\in Q\langle X\rangle\}> \epsilon$.
+\end{proof}
+
+The second characterization relates distances with the Taylor expansion. Let us briefly discuss the latter, first.
+For all $f: Q\langle X\rangle \to Q\langle Y\rangle$, let $\delta^{(n)}f:Q\langle X^{n}\rangle \to Q\langle Y\rangle$ indicate the $n$-linear function given by 
+$$
+\delta^{(n)}f(\B x^{n})=\mathsf D^{(n)}f(\B x^{n}, \infty)
+$$
+
+Notice that $\widehat{\delta^{(n)}f}\in Q^{X^{n}\times Y}$ satisfies 
+$\widehat{\delta^{(n)}f}_{a_{1},\dots, a_{n},b}= \widehat f_{[a_{1},\dots, a_{n}],b}$.
+In other words, $\delta^{(n)}f$ precisely captures the behavior of $f$ when applied to multisets of length $n$.
+Moreover, the full behavior of $f$ can be recovered from the functions $\delta^{(n)}f$ using the Taylor expansion which, in its tropical form, reads as:
+\begin{equation}
+f(\B x)= \inf_{n}\left \{\mathsf D^{(n)}f(\B x^{n},\infty)\right\}= \inf_{n}\left \{\delta^{(n)}f(\B x^{n})\right\}
+\tag{Tropical Taylor}
+\end{equation}
+
+\begin{proposition}\label{prop:taylormetric}
+For all $f,g:Q\langle X\rangle \to Q\langle Y\rangle$, 
+$$
+\| \widehat f-\widehat g\|_{\infty}= \sup_{n}\| \widehat{\delta^{(n)}f} -\widehat{\delta^{(n)}g}\|_{\infty}
+$$
+\end{proposition}
+\begin{proof}
+Let us first show that $\| \widehat f-\widehat g\|_{\infty}\leq \epsilon$ implies 
+$\| \delta^{(n)}f-\delta^{(n)}g\|_{\infty}\leq\epsilon$, for all $n\in \BB N$. 
+Notice that $\widehat{\delta^{(n)}f}\in Q^{X^{n}\times Y}$ satisfies 
+$\widehat{\delta^{(n)}f}_{a_{1},\dots, a_{n},b}= \widehat f_{[a_{1},\dots, a_{n}],b}$. Hence 
+from $\| \widehat f-\widehat g\|\leq\epsilon$ it follows that for all $\mu=[a_{1},\dots, a_{n}]$ and $b\in Y$, 
+$|\widehat f_{\mu,b}- \widehat g_{\mu,b}|\leq \epsilon$, so 
+we deduce that $\| \widehat{\delta^{(n)}f}- \widehat{\delta^{(n)}g}\|\leq \epsilon$.
+
+For the converse direction suppose that, for all $n\in \BB N$, $\| \widehat{\delta^{(n)}f}- \widehat{\delta^{(n)}g}\|\leq \epsilon_{n}$. Then, since the family of coefficients $
+F_{n,\mu,b}=
+( \widehat{\delta^{(n)}f})_{a_{1},\dots, a_{n},b}$ (where $\mu=[a_{1},\dots, a_{n}]$) is in bijection with the coefficients $\widehat f_{\mu,b}$, we deduce that $|\widehat f_{\mu,b}-\widehat g_{\mu,b}|\leq
+\epsilon_{\sharp \mu  }$, and thus that 
+$\| \widehat f-\widehat g\|_{\infty}\leq \sup_{n}\epsilon_{n}$. 
+\end{proof}

### 2023-01-19 appIII.tex
commit

--- 2023-01-19 appVII.tex
@@
-In the following we let $Q$ indicate the Lawvere quantale $([0,\infty], \geq, +,0)$. 
-$Q$ can be seen as:
-\begin{itemize}
-\item a category $Q_{0}$ with a unique object and morphisms being the elements of $Q$, with identity 0 and composition $+$;
+\subsubsection{The language $\BSTLC$}
+TO BE ADDED
 
-\item a degenerate category $Q_{1}$ with objects the elements of $Q$, and  
-a morphism between $x$ and $y$ whenever $x\geq y$;
+\subsubsection{$(\LREL, !_{n})$ is a model of Bounded Linear Logic}
 
-\item a 2-category $Q_{2}$ with 0-cells and 1-cells as in $Q_{0}$, and 2-cells the morphisms of $Q_{1}$.
-
-\end{itemize}
-These categorical formulations of $Q$ will help us here and there.
-
-We let $x-y$ indicate truncated subtraction. Notice that 
-$x \leq (x-y)+y$ and $(x+y)-y\leq x$. This corresponds to the fact that ``$\_-y$'', as a functor from $
-Q_{1}$ to itself, 
- is \emph{right-adjoint} to the functor ``$\_+y$''. Equivalently, 
- $x+y \geq z $ iff $x\geq z-y$.
-
-$Q\mathsf{Rel}$ is the category of sets and $Q$-relations, where a $Q$-relation 
-$R:X\pfun Y$ is a function $R: X\times Y\to Q$, with identity $1_{X}:X\pfun X$ given by 
-$$1_{X}(x,y)=\begin{cases} 0 & \text{ if }x=y\\ \infty & \text{ otherwise}\end{cases}$$
-and composition of $R:X\pfun Y$ and $S:Y\pfun Z$ given by 
-$$
-(S\cdot R)(x,z)= \inf_{y\in Y}R(x,y)+S(y,z)
-$$
-
-\begin{remark}
-$Q\mathsf{Rel}$ is also a (degenerate) 2-category, where for $R,S:X\pfun Y$, $R\To S$ iff for all $x\in X$, $y\in Y$, $R(x,y)\geq S(x,y)$.
-
-\end{remark}
-
-Composition with $R:X\pfun Y$ in Q$\mathsf{Rel}$ yields a functor
-$$
-R\cdot \_ : Q\mathsf{Rel}(A,X)   \longrightarrow Q\mathsf{Rel}(A,Y)
-$$
-which
-has a right-adjoint
-$$
-\_ \multimapinv R : Q\mathsf{Rel}(A,Y)   \longrightarrow Q\mathsf{Rel}(A,X)
-$$
-given, for $S:A\pfun Y$ by 
-\begin{align*}
-(S\multimapinv R)(a,x)= \sup_{y\in Y}S(a,y) -R(x,y)
-%(R\multimap S)(a,x)= \sup_{y\in Y}
-\end{align*}
-
-%
-%\begin{lemma}
-%\begin{itemize}
-%\item[i.] $ (R\multimapinv T)\cdot
-%(T\multimapinv S)\leq R \multimapinv S $.
-%
-%\item[ii.] $R\multimapinv T \leq (R\multimapinv S)\multimapinv (T\multimapinv S)$.
-%\end{itemize}
-%\end{lemma}
-%\begin{proof}
-%\begin{align*}
-%(R\multimapinv S)(x,y)&= \sup_{z}R(x,z)-S(y,z) \\
-%&=\sup_{z} (R(x,z)-T(z,z))+(T(z,z)-S(y,z))\\
-%&= \inf_{w}\left(\sup_{z}R(x,z)-T(w,z)\right) +\left(\sup_{z'}T(w,z')-S(z',y)\right)
-%\\
-%&= \inf_{w}(R\multimapinv T)(x,w)+(T\multimapinv S)(w,y)  
-%\\
-%&= (R\multimapinv T)\cdot (T\multimapinv S)(x,y)
-%\end{align*}
-%
-%\end{proof}
-
-\subsubsection{$Q$-categories.}
-
-A \emph{$Q$-category}, or \emph{generalized (quasi-)metric space} is a small $Q$-enriched category.
-Concretely, it is given by a pair $(X,q)$ made of a set $X$ and a $Q$-relation $q: X\pfun X$ satisfying:
-\begin{align}
-q(x,x) & = 0 \tag{Q1} \\
-q(x,z)+q(z,y) & \geq q(x,y) \tag{Q2}
-\end{align}
-A $Q$-category is \emph{skeletal} if it further satisfies
-\begin{align}
-q(x,y) & = 0 \ \To \ x=y \tag{Q3}
- \end{align}
- and \emph{symmetric} if it satisfies
- \begin{align}
-q(x,y) & = q(y,x) \tag{Q4}
- \end{align}
-For simplicity, in the following we indicate the distance function of a $Q$-category $X$ simply as $X(x,y)$.
-
-\begin{remark}
-In more abstract terms, a $Q$-category can be identified with a \emph{monad} in $Q\mathsf{Rel}$, seen as a 2-category: it is given by a 1-cell $q:X\pfun X$ together with 2-cells $1_{X}\To  q$ and $q\cdot q \To q$.
-\end{remark}
-
-We let $x\simeq y$ indicate the equivalence relation defined by  $X(x,y)=X(x',y)$ holds for all $y\in X$. 
-Notice that $x\simeq y$ coincides with $x=y$ precisely when $X$ is skeletal.
-
-Every $Q$-category $X$ is also a preorder $(X,\preceq_{X})$, with $x\preceq_{X}y$ iff $X(x,y)=0$. Moreover, if $X$ is skeletal, then $\preceq_{X}$ is an order.
-
-For every $Q$-category $X$, the category $X^{\mathrm{op}}$ has same objects as $X$, and $X^{\mathrm{op}}(x,y)=X(y,x)$.
-
-\begin{example}
-$Q$ is a $Q$-category with the distance $Q(x,y)=y-x$. 
-The category $Q^{\mathrm{op}}$ has distance $Q^{\mathrm{op}}(x,y)=Q(y,x)$.
-\end{example}
-
-\begin{example}
-For every $Q$-category $X$, the distance $\C S(X)(x,y)=\max\{X(x,y),X^{\mathrm{op}}(x,y)\}$ defines a symmetric $Q$-category $\C S(X)$. In the case of $Q$, $\C S(Q)$ is just the Euclidean distance.
-\end{example}
-
-A \emph{functor} between $Q$-categories $X$ and $Y$ is just a non-expansive function $f:X\to Y$, that is, satisfying $Y(f(x),f(y))\leq X(x,y)$.
-
-%
-\begin{remark}\label{rem:functor}
-Given $f:X\to Q$, $f$ is a functor precisely when for all $x,y\in X$ $f(x)\leq\inf_{x'\in X}f(x')+X(x',x)$. 
-Indeed, if $f$ is a functor then $f(x)\leq f(x')+X(x',x)$, since $f(x)-f(x')=Q(f(x'),f(x))\leq X(x',x)$. 
-Conversely, if $f(x)\leq f(x')+X(x',x)$ holds for all $x'$, then 
-$Q(f(x),f(x'))=f(x')-f(x)\leq X(x',x)$. 
-\end{remark}
-
-%Indeed, being a functor means that any arrow $X(x,y)$ is mapped into an arrow $Y(f(x),f(y))$, that is, $Y(f(x),f(y))
-
-\begin{example}
-For every $Q$-category $X$, the $Q$-category $[X,Q]$ has objects all functors from $X$ to $Q$ (where the latter is seen as a $Q$-category), and $[X,Q](f,g)=\sup_{x\in X}Q(f(x),g(x))$.
-
-When $X$ has the discrete metric, $[X,Q]=Q^{X}$ and $\C S([X,Q])(f,g)=\| f-g\|_{\infty}$.
-
-\end{example}
-
-\begin{example}
-For every $Q$-categories $X,Y$ we can define:
-\begin{itemize}
-\item a $Q$-category $X\otimes Y$ with 
-$(X\otimes Y)(\langle x,y\rangle, \langle x',y'\rangle)=X(x,x')+Y(y,y')$;
-\item a $Q$-category $X\times Y$ with 
-$(X\times Y)(\langle x,y\rangle, \langle x',y'\rangle)=\max\{X(x,x'),Y(y,y')\}$. 
-\end{itemize}
-\end{example}
-
-\subsubsection{$Q$-distributors.}
-
-%We let $Q\mathsf{Cont}$ indicate the category of $Q$-categories and continuous functors.
-
-A \emph{$Q$-distributor} $\Phi: X\pfun Y$ is a $Q$-relation $\Phi: X\times Y\to Q$ satisfying
-\begin{align*}
-\Phi\cdot X   \leq\Phi \qquad \qquad
-Y \cdot \Phi  \leq \Phi
-\end{align*}
-
-\begin{remark}
-Using Remark \ref{rem:functor}, it follows that $\Phi$ is a $Q$-distributor precisely when it is a functor $\Phi: X^{\mathrm{op}}\otimes Y \to Q$:
-the condition $\Phi(x,y) \leq X(x,x')+\Phi(x',y)$ implies that $\Phi$ is a functor over $X^{\mathrm{op}}$, while
-the condition $\Phi(x,y) \leq \Phi(x,x')+Y(x',y)$ implies that $\Phi$ is a functor over $Y$.
-But this implies then 
-$\Phi(x,y) \leq X(x,x') +\Phi(x',y')+Y(y',y)$ and thus $Q(\Phi(x',y'),\Phi(x,y))\leq X(x,x')+Y(y',y)$.
-\end{remark}
-
-%where $(\Phi\cdot X)_{x,y}= \inf_{z\in X}\Phi_{x,z}+X(z,y)$ and 
-%$(y\cdot \Phi)_{x,y}= \inf_{z\in X}Y(x,z)+\Phi_{z,y}$.
-
-The identity distributor $1_{X}:X\pfun X$ is given by $(1_{X})_{x,y}=X(x,y)$. 
-Distributors $\Phi:X\pfun Y$ and $\Psi:Y\pfun Z$ compose via $\Psi\cdot \Phi:X\pfun Z$. 
-We let $Q\mathsf{Dist}$ indicate the category of $Q$-categories and $Q$-distributors.
-
-\begin{remark}
-Usually distributors $\Phi: X\pfun Y$ are presented as $Q$-relations $\Phi:Y\times X\to Q$ (notice the inversion of $X$ and $Y$), that is, as functors $\Phi: X\times Y^{\mathrm{op}}\to Q$.
-We chose to invert the presentation of distributors for uniformity with the usual notations in models of the differential $\lambda$-calculus.
-\end{remark}
-
-Any functor $f:X\to Y$ induces two adjoint distributors $f^{\circ}:X\pfun Y$ and $f_{\circ}:Y\pfun X$ given by $f^{\circ}(x,y)=Y(f(x),y)$ and $f_{\circ}(y,x)=Y(f(y),x)$.
-Here adjoint means that $1_{X}\leq f_{\circ}\cdot f^{\circ} $ and $f^{\circ}\cdot f_{\circ}\geq 1_{Y}$.
-
-\begin{remark}[Yoneda embedding]
-%$Q$ is a $Q$-category with $Q(x,y)=|x-y|$. 
-%For any $Q$-category $X$ we can define a new $Q$-category $\mathsf{Dist}(X,Q)$ (that we note simply as $[X,Q]$) whose objects are the distributors $\B x:\B 1\pfun X$, or equivalently, the functors from $X$ to $Q$, i.e.~those $\B x\in Q^{X}$ such that $|\B x_{a}-\B x_{b}|\leq X(a,b)$, or equivalently $\B x_{a}+X(a,b) \geq \B x_{b}$.
-
-The \emph{Yoneda embedding} is the faithful functor $\C Y: X\to [X,Q]$ given by $\C Y(x)(y)=X(y,x)$. The functoriality and faithfulness of $\C Y$ follow from
-\begin{align}
-[X,Q](\C Y(x),\C Y(x'))= X(x,x') \tag{Yoneda}
-\end{align}
-which is proved as follows: for all $y\in X$ we have 
-\begin{align*}
-Q( \C Y(x)(y),\C Y(x')(y))&= \C Y(x')(y)-\C Y(x)(y) \\
-&= X(y,x')-X(y,x) \leq X(x,x')
-\end{align*}
-where the last step follows from $X(y,x)+X(x,x')\geq X(y,x')$. 
-From this we deduce that $[X,Q](\C Y(x),\C Y(x'))=\sup_{y\in X}Q( \C Y(x)(y),\C Y(x')(y))\leq X(x,x')$. 
-For the converse direction, we have  
-\begin{align*}
-X(x,x') &  = X(x,x') - 0 \\ &=
-X(x,x')- X(x,x)
-\\& =\C Y(x')(x)-\C Y(x)(x)= Q(\C Y(x)(x),\C Y(x')(x))\\
-&\leq [X,Q](\C Y(x), \C Y(x'))
-\end{align*}
-\end{remark}
-
-\begin{remark}
-The \emph{opposite Yoneda embedding} is the faithful functor
-$\C Y^{\mathrm{op}}: X\to [X,Q]^{\mathrm{op}}$ given by $\C Y^{\mathrm{op}}(x)(y)=X(x,y)$. The functoriality and faithfulness of $\C Y^{\mathrm{op}}$ follow from
-\begin{align}\label{eq:yonedaop}
-[X,Q](\C Y^{\mathrm{op}}(x),\C Y^{\mathrm{op}}(x'))= X(x',x) \tag{Yoneda$^{\mathrm{op}}$}
-\end{align}
-which is proved similarly to the case of $\C Y$.
-\end{remark}
-
-\subsubsection{cocompleteness}
-
-Let $\Phi:Y\pfun Z$ be a distributor and $f:Y\to X$ be a functor.
-A functor $ g:Z\to X$ is called the \emph{$\Phi$-weighted colimit of $f$} if it satisfies, for all
-$z\in Z$ and 
- $x\in X$, 
-$$
-X(g(z), x) \ = \  \sup_{y\in Y}X(f(y),x) - \Phi(y,z)
-$$
-If this colimit exists, we write it as $\mathrm{colim}(\Phi,f)$.
-
-A $Q$-category $X$ is said \emph{cocomplete} if it admits all weighted colimits, and a functor $f:X\to Y$ of $Q$-categories is said \emph{cocontinuous} if it commutes with all weighted colimits, meaning that $f \circ \mathrm{colim}(\Phi,g)= \mathrm{colim}(\Phi, f\circ g)$.
-
-We let $Q\mathsf{CCat}$ indicate the category of skeletal and cocomplete $Q$-categories and cocontinuous functors.
-
-\begin{proposition}\label{prop:yonedasup}
-Let $X$ be a $Q$-category. Then $X$ is cocomplete iff the Yoneda embedding has a left-adjoint.
-\end{proposition}
-\begin{proof}
- For all $\B x\in [X,Q]$, let $\sup\B x$ be defined as a weighted colimit via
-$$
-X(\sup\B x, y)= \sup_{z\in X} X(z,y)-\B x_{z}
-$$
-that is, $\sup \B x= \mathrm{colim}( \B x, \mathrm{id}_{X})$, where $\B x$ is seen as a distributor $\B x:\B 1\pfun X$.
-
-Let us check that $\sup: [X,Q]\to X$ is a functor. First, let us check the  inequality
-\begin{align}\label{eq:in1}
-(X\multimapinv\B y)\cdot (\B y\multimapinv\B x)\geq (X\multimapinv\B x)\end{align}
-as follows:
-\begin{align*}
-\left((X\multimapinv \B y)\cdot (\B y\multimapinv\B x)\right)(a)&=
- \left (\sup_{b}X(b,a)-\B y_{b}\right)+\left(\sup_{b}\B y_{b}-\B x_{b}\right)\\
-&\geq
-\sup_{b}(X(b,a)-\B y_{b})+(\B y_{b}-\B x_{b})
-\\
-&=\sup_{b}(X(b,a)-\B y_{b}+\B y_{b})-\B x_{b}
-\\
-&=\sup_{b} X(b,a)-\B x_{a}\\
-&= (X\multimapinv \B x)(a)
-\end{align*}
-From \eqref{eq:in1} we deduce immediately the inequality below:
-\begin{align}\label{eq:in2}
-\B y\multimapinv \B x \geq (X\multimapinv \B x)\multimapinv(X\multimapinv\B y)
-\end{align}
-and we can now compute:
-\begin{align*}
-[X,Q](\B x, \B y)&= \sup_{a\in X}\B y_{a}-\B x_{a}\\
- &\stackrel{\tiny\eqref{eq:in2}}{\geq}
- \sup_{a\in X} \left( \sup_{b\in X}X(b,a)-\B x_{a}\right) -\left(  \sup_{b\in X}X(b,a)-\B y_{a}   \right )\\
- &= \sup_{a\in X}X(\sup \B x, a)-X(\sup \B y, a)\\
- &= [X,Q](\C Y^{\mathrm{op}}(\sup\B y),\C Y^{\mathrm{op}}(\sup \B x))
- \tag{\ref{eq:yonedaop}}
- \\
-&=X(\sup \B x, \sup \B y)
-\end{align*}
-
-Then for all $x\in X$, $\sup \C Y(x)\simeq x$. Indeed we have 
-\begin{align*}
-X(\sup \C Y(x), y) & = \sup_{z\in X}X(z,y) -\C Y(x)(z) \\
-& = \sup_{z\in X}X(z,y)-X(z,x)\\
-&= X(x,y)
-\end{align*}
-%where we use the fact that from $X(x,y)+X(z,x) \geq X(z,y)$ it follows $X(x,y) \geq X(z,y)-X(z,x)$ and thus $X(x,y) \geq \sup_{z}X(z,y)-X(x,z)$, and conversely, from $X(x,y)-X(x,x)=X(x,y)$ it follows
-%%$X(x,y) \leq \sup_{z}X(z,y)-X(x,z)$.
-%
-
-Moreover, for all $\B x\in [X,Q]$, we have $\C Y(\sup \B x)\geq \B x$:  
-\begin{align*}
-\C Y(\sup \B x)(a) &= X(a,\sup \B x)\\
-&= X(  \sup(\C Y(a)),\sup \B x) \\
-& \geq [X,Q]( \C Y(a), \B x)\\
-&\geq\B x_{a}-\C Y(a)(a) \\
-&=
- \B x_{a}-X(a,a)  = \B x_{a}
-\end{align*}
-%where in the last step we use the fact that 
-%$[X,Q](  \C Y(a),\B x)= \sup_{b}\B x_{b}-X(a,b) = \B x_{a}$.
-%
-%$\B y\in Q^{X}$, 
-% $\| \C Y(\sup\B x)- \B y\|_{\infty}=\| \C Y(\sup\B x)- \B x\|_{\infty}$. Indeed we have 
-% \begin{align*}
-% \| \B x_{a}-\C Y(\sup\B x)\|_{\infty}&= \sup_{a\in X}|
-%\B x_{a}-\C Y(\sup\B x)(a)|\\ 
-%&=
-%\sup_{a\in X}|
-%\B x_{a}-X(\sup\B x,a)|\\
-% &=
-%\sup_{a\in X}\inf_{a'\in X}| \B x_{a}+X(a',a) -\B x_{a'}|\\
-%&= \sup_{a\in X}|\B x_{a}+ X(a,a)- \B x_{a}| = 0
-% \end{align*}
-%
-%
-
-Conversely, if $\sup$ is well-defined and adjoint to $\C Y$, then 
-given $\Phi: Y\pfun \B 1$ and $f:Y\to X$, 
-we can define
-$\mathrm{colim}(\Phi, f):= \sup\Psi$, where $\Psi=
- f^{\circ}\cdot \Phi: X\pfun \B 1$, since 
-\begin{align*}
-X(\sup \Psi, y)&=
-\sup_{z\in X}X(z,y)-\Psi(z) \\
-&=\sup_{z\in X}X(z,y)- \inf_{y\in Y}X(z,f(y))  +\Phi(y)\\
-& = \sup_{z\in X}\sup_{y\in Y}X(z,y)-X(z,f(y))  -\Phi(y)\\
-&=
-\sup_{z\in X} X(f(z),y) - \Phi(z)
-\end{align*}
-\end{proof}
-
-\begin{definition}[MacNeill Completion]
-Let $X$ be a $Q$-category. For all $f: \B 1\pfun X$ and $g:X\pfun \B 1$, let $f \coh g$ iff 
-$f  = X\multimapinv g $ and 
-$g = f\multimap X$.
-
-The \emph{MacNeill completion of $X$} is the $Q$-category $\B M(X)$ made of those 
-$f:\B 1\pfun X$ such that $f\coh g$ for some $g:X\pfun \B 1$, with
-$\B M(X)(f,f')=[X,Q](f,f')$. 
-\end{definition}
-
-Observe that if $f\coh g$, then $f= X\multimapinv (f\multimap X)$, i.e.:
-\begin{align}
-f(x)=  \sup_{y\in X}\inf_{z\in X}X(x,y)-X(y,z) +f(z)       
-\tag{COH}
-\end{align}
-
-\begin{proposition}
-Let $X$ be a $Q$-category. 
-If $X$ is cocomplete, then $\C Y$ is an isomorphism between $X$ and $\B M(X)$. 
-\end{proposition}
-\begin{proof}
-For all $x\in X$, one can check that $\C Y(x)\in \B M(X)$. Indeed, we can check that $\C Y(x) \coh \C Y^{\mathrm{op}}(x)$:
-\begin{align*}
-\C Y(x)(y)  &= X(y,x) \\
-&=  \sup_{z\in X}X(y,z  )- X(x,z)\\
-&= \sup_{z\in X}X(y,z  )- \C Y^{\mathrm{op}}(x)(z)\\
- &= (X\multimapinv \C Y^{\mathrm{op}}(x))(y)
-\end{align*}
-Since $\sup\C Y(x) \simeq x$ holds, it suffices to show that if $\B x \coh \B y$, then 
-$\C Y(\sup \B x)=\B x$: 
-\begin{align*}
-\C Y(\sup \B x)(a) & \ \ = X(a, \sup\B x) \\
-&\ \ = \sup_{b\in X}X(a,b)-X(\sup\B x, b) \\
-&\ \ = \sup_{b\in X}\inf_{c\in X}X(a,b)- X(b,c) + \B x_{c} \\
- &\stackrel{{\tiny\text{(COH)}}}{=} \B x_{a}
-\end{align*}
-\end{proof}
-
-%
-%\begin{remark}[other notions of cocompleteness]
-%The categorical notion of cocompleteness subsumes several other notions of cocompleteness, in the sense of being (strictly) stronger:
-%\begin{itemize}
-%\item \emph{order-completeness} is the case when the pre-order $\preceq_{X}$ is cocomplete.   
-%Let $I\subseteq X$, so that $\mathrm{id}_{I}$ can be seen as a functor from the subcategory $I$ to $X$. Let $0_{I}: \B 1\pfun I$ be the distributor given by $(0_{I})_{i}=0$.
-%Then $\mathrm{colim}(0_{I}, \mathrm{id}_{I})$ coincides with $\inf I$, via 
-%$$
-%X(\inf I, x) = \sup_{i\in I}X(i, x)
-%$$
-%Notice that in general an order-complete $Q$-category needs not be tensored, and thus it needs not be cocomplete.
-%
-%
-%
-%
-% 
-%
-%\item \emph{Cauchy-completeness} is the case where $X$ contains the points $\sup \B x, \sup\B y$, for any 
-% any two \emph{adjoint} presheaves $\B x,\B y\in [X,Q]$, i.e.~satisfying $0=\inf_{a}\B x_{a}+\B y_{a}$ and $\B x_{a}+ \B x_{b}\geq X(a,b)$. 
-% Notice that given a pair $(\B x, \B y)$ we can define a Cauchy sequence by finding $a_{n}$ satisfying 
-% $\B x_{a_{n}}+\B y_{a_{n}}\leq \frac{1}{n}$.
-Indeed this implies $X(a_{n},a_{n+1})\leq \B x_{a_{n}}+\B y_{a_{n+1}}\leq \B x_{a_{n}}+\B y_{a_{n}}+\B x_{a_{n+1}}+\B y_{a_{n+1}}=\frac{1}{n}+\frac{1}{n+1}$.
-% Conversely, any (equivalence class of) Cauchy sequences $a_{n}$ yields an adjoint pair given by
-% $\B x_{a}= \lim_{n}X(a_{n},a)$ and $\B y_{a}=\lim_{n}X(a,a_{n})$. 
-% 
-%% 
-%% \item \emph{Isbell-completeness} (or \emph{MacNeill cocompleteness}) is the case where $X$ contains the points $\sup \B x, \sup \B y$, for any two presheaves $\B x, \B y\in [X,Q]$ satisfying 
-%% $
-%% \B x_{a}= \sup_{b\in X}X(b,a)-\B y_{b}
-%% $.
-%
-%
-%\end{itemize}
-%
-%\end{remark}
-
-\subsubsection{Tensors and $Q$-Modules}
-
-Among weighted colimits, one is of big importance for us. 
-Any $\epsilon \in Q$ generates a constant distributor $(\epsilon):\B 1\pfun \B 1$, and any point $x\in X$ generates a constant functor $\Delta x: \B 1\to X$.
-Given a $Q$-category $X$, a point $x\in X$ and  $\epsilon \in Q$, the \emph{tensor of $x$ and $\epsilon$}, if it exists, is defined as
-$$
-\epsilon \otimes x := \mathrm{colim}((\epsilon), \Delta x)
-$$
-A $Q$-category $X$ is \emph{tensored} if for all $x\in X$ and $\epsilon \in Q$, it admits the tensor $\epsilon \otimes x$.
-
-\begin{proposition}
-A tensored $Q$-category $X$ is a $Q$-module $(X, \preceq_{X}, \otimes)$.
-A cocontinuous functor of cocomplete $Q$-categories is a $Q$-module morphism between the associated $Q$-modules.
-\end{proposition}
-\begin{proof}
-We must show that tensors induce a continuous action. Observe that tensors are characterized by the equation 
-\begin{align}\label{eq:tensor}
-X(x\otimes \epsilon, x') = X(x,x') -\epsilon
-\end{align}
-If $\epsilon=0$, then \eqref{eq:tensor} forces $x\otimes \epsilon\simeq x$. 
-If $\epsilon=\delta+\eta$, then using the fact that $\alpha-(\epsilon+\delta)=(\alpha-\epsilon)-\delta$ 
-we deduce $X((x\otimes \epsilon)\otimes \delta, x')=X(x\otimes \epsilon, x')-\delta=(X(x,x')-\epsilon)-\delta=X(x,x')-(\epsilon+\delta)=X(x\otimes (\epsilon+\delta),x')$, which forces $x\otimes(\epsilon+\delta)\simeq (x\otimes \epsilon)\otimes \delta$.
-
-A cocontinuous functor $f:X\to Y$ commutes with sups and with $\otimes$, and is thus a $Q$-module morphism.
-\end{proof}
-
-\begin{lemma}
-\begin{itemize}
-\item[i.] $\sup_{i\in I}a_{i}-\epsilon= (\sup_{i\in I}a_{i})-\epsilon$.
-\item[ii.] $\sup_{i\in I}(a_{i}-\epsilon)-b_{i}= (\sup_{i\in I}a_{i}-b_{i})-\epsilon$.
-
-\end{itemize}
-\end{lemma}
-\begin{proof}
-Let $A= \sup_{i\in I}a_{i}-\epsilon$ and $B= (\sup_{i\in I}a_{i})-\epsilon$.
-Let $J\subseteq I$ be the set of indexes $j$ such that $a_{j}>\epsilon$. 
-If $J=\emptyset$ then $A=B=0$. Otherwise, 
-$A= \sup_{j\in J}a_{j}-\epsilon$ (where ``$-$'' can be interpreted as subtraction on $\BB R$, and 
-$B= (\sup_{j\in J}a_{j})-\epsilon$ (again with ``$-$'' being subtraction on $\BB R$), so $A=B$ follows from the continuity of ``$-$'' on $\BB R$.
-
-Let now $A= \sup_{i\in I}(a_{i}-\epsilon)-b_{i}$ and $B= (\sup_{i\in I}a_{i}-b_{i})-\epsilon$.
-Let $J\subseteq I$ be the set of indexes $j$ such that $a_{j}> b_{j}+\epsilon$.
-If $J=\emptyset$, then $A=0$; suppose $B>0$, then $\sup_{i\in I}a_{i}-b_{i}>\epsilon$, but this implies that we can find $i\in I$ with $a_{i}>b_{i}+\epsilon$, against the assumption, so also $B=0$ holds.
-If $J$ is non-empty, then
-$A= \sup_{j\in J}(a_{j}-\epsilon)-b_{j}$, where ``$-$'' is not subtraction on $\BB R$ and 
-$B= (\sup_{j\in J}a_{j}-b_{j})-\epsilon$, again with ``$-$'' usual subtraction, so $A=B$ follows from the continuity of ``$-$'' on $\BB R$.
- \end{proof}
-
-\begin{lemma}
-In any cocomplete $Q$-category, $x\otimes \epsilon \simeq \sup(  \C Y(x)+\epsilon  )$.
-In the cocomplete $Q$-category $[X,Q]$, $\B x\otimes \epsilon= \B x+\epsilon$.
-\end{lemma}
-\begin{proof}
-We have 
-\begin{align*}
-X(\sup(\C Y(x)+\epsilon), x')& =\sup_{y\in X}X(z,x')- (\C Y(x)(z)+\epsilon)\\
-&= \sup_{y\in X}X(z,x')-(X(z,x)+\epsilon)\\
-&= (\sup_{y\in X}X(z,x')-X(z,x))-\epsilon\\
-&= X(x,x')-\epsilon
-\end{align*}
-which shows $x\otimes \epsilon=\sup(\C Y(x)+\epsilon)$. In $[X,Q]$ we have 
-$[X,Q](\B x+\epsilon, \B x')=\sup_{a\in X}(\B x_{a}+\epsilon)-\B x'_{a}= (\sup_{a\in X}\B x_{a}-\B x'_{a})-\epsilon= [X,Q](\B x, \B x')-\epsilon$, which shows $\B x\otimes \epsilon \simeq \B x+\epsilon$, and since $[X,Q]$ is skeletal, $\B x\otimes \epsilon=\B x+\epsilon$.
-\end{proof}
-
-The dual notion of tensors is the \emph{cotensor} $x\multimapinv \epsilon$.
-Formally, it is defined as a \emph{weighted limit} (whose definition is dual to that of weighted colimit but we do not give details here), and characterized by the equation
-\begin{align*}
-X(x', x\multimapinv \epsilon)= X(x',x)-\epsilon
-\end{align*}
-In other words, in a tensored and cotensored $Q$-category we have $X(x\otimes \epsilon,y)= X(x,y\multimapinv \epsilon)$.
-
-\begin{example}
-The $Q$-category $[X,Q]$ is cotensored, with $\B x\multimapinv \epsilon:= \B x-\epsilon$.
-Indeed we have $[X,Q](\B x, \B y\multimapinv \epsilon)=\sup_{a\in X}(\B y_{a}-\epsilon)-\B x_{a}=(\sup_{a\in X}\B y_{a}-\B x_{a})-\epsilon= [X,Q](\B x, \B y)-\epsilon$.
-\end{example}
-
-\begin{definition}
-A $Q$-category $X$ is \emph{order-complete} if it is a sup-lattice with respect to the order $\preceq_{X}$ (i.e.~all joins exist).
-\end{definition}
-
-\begin{lemma}\label{lemma:supinf}
-Let $X$ be a $Q$-category. If $X$ is order-complete, then 
-\begin{itemize}
-\item if $X$ is co-tensored, $X(\bigvee_{i}x_{i},y)=  \sup_{i}X(x_{i},y) $;
-\item if $X$ is tensored, $
-X(x,\bigvee_{i}y_{i})=  \inf_{i}X(x,y_{i})$.
-
-\end{itemize}
-\end{lemma}
-\begin{proof}
-We only prove the second claim, the first being proved similarly.
-
- Let us show that $z\preceq_{X}z'$ iff for all $w\in X$, $X(w,z')\leq X(w,z)$: 
- on one direction we have $X(w,z')\leq X(w,z)+X(z,z')=X(w,z)$; on the other direction, 
- we have $X(z,z')\leq X(z,z)=0$.
-
- Using this, since $y_{i}\preceq_{X}y:=\bigvee_{i}y_{i}$ we deduce 
- $X(x,y_{i})\leq X(x,y)$, and thus $X(x,y)\geq \inf_{i}X(x,y_{i})$.
-
- For the converse direction, we argue as follows: let $X(x,y_{i})\leq \epsilon$ hold for all $i\in I$; then $0=X(x,y_{i})-\epsilon= X(x\otimes\epsilon,y_{i})$.
-Thus $x\otimes\epsilon\preceq_{X}y_{i}$, and thus
- $x\otimes\epsilon\preceq_{X}y$, that is $X(x\otimes\epsilon,y)=X(x,y)-\epsilon=0$, and consequently $X(x,y)\leq \epsilon$. 
- By letting $\epsilon:=X(x,y_{i})$ we conclude then $X(x,y)\leq X(x,y_{i})$, and thus $X(x,y)\leq \inf_{i}X(x,y_{i})$.
-
-%
-%First, if $
-%By definition $x:=\bigvee_{i}x_{i}$ is characterized by (1) $X(x_{i},x)=0$ for all $i\in I$, and 
-% (2) $X(x,y)=0$, for all $y$ such that $X(x_{i},y)=0$ holds for all $i\in I$.
-%% 
-%% Let us show that $z\preceq_{X}z'$ implies $X(z',y) \leq X(z,y)$: 
-%% from $z\preceq_{X}z'$ we deduce $X(z,z')=0$, whence $
-%% 
-% Let now $y\in X$; 
-% then $X(x_{i},y)\leq X(x_{i},x)+X(x,y)=X(x,y)$, which implies $\sup_{i}X(x_{i},y)\leq X(x,y)$.
-% 
-%
-% 
-% 
-% Suppose now that, for some $i\in I$, $X(x_{i},y)>0$; then $X(x,y)\leq X(x,x_{i})+X(x_{i},y)$
-% 
-% 
-%If $X(x_{i},y) \leq \epsilon$, for all $i\in I$, then 
-%$0= X(x_{i},y)-\epsilon = X(x_{i},y\multimapinv \epsilon)$. Thus 
-%$x_{i}\preceq_{X}y\multimapinv\epsilon$ and we deduce 
-%$x\preceq_{X}y\multimapinv \epsilon$. Hence
-%$0= X(x, y\multimapinv \epsilon)=X(x,y)-\epsilon$ and consequently
-%$ X(x,y)\leq \epsilon$.
-
-\end{proof}
-
-\begin{proposition}\label{prop:tencoten}
-If a $Q$-category $X$ is tensored, cotensored and order-complete, then it is cocomplete.
-\end{proposition}
-\begin{proof}
-For all $\B x\in [X,Q]$, let $\sup \B x:= \bigvee_{a\in X}a\otimes\B x_{a}$. 
-Let us check that $X(\sup \B x, b)= \sup_{a\in X}X(a,b)-\B x_{a}$, using Lemma \ref{lemma:supinf}:
-\begin{align*}
-X(\sup \B x, b) &= \sup_{a\in X}X(a\otimes \B x_{a},b)\\
-&=\sup_{a\in X}X(a,b)-\B x_{a}
-\end{align*}
-We can thus conclude using Proposition \ref{prop:yonedasup}.
-\end{proof}
-
-\begin{proposition}\label{prop:tenfun}
-Let $X,Y$ be two tensored $Q$-categories, and $f:X\to Y$ be a function.
-\begin{itemize}
-\item[i.] $f$ is a functor iff $f$ is order-preserving and for all $x\in X$ and $\epsilon\in Q$, $f(x)\otimes \epsilon \preceq_{Y} f(x\otimes \epsilon)$.
-
-\item[ii.] $f$ is a cocontinuous functor iff $f$ commutes with joins and for all $x\in X$ and $\epsilon\in Q$, $f(x)\otimes \epsilon = f(x\otimes \epsilon)$.
-\end{itemize}
-\end{proposition}
-\begin{proof}
-\begin{itemize}
-\item[i.] If $f$ is a functor then 
-\begin{align*}
-Y(f(x)\otimes \epsilon, f(x\otimes \epsilon))&= Y(f(x), f(x\otimes \epsilon)) -\epsilon \\
-&\leq X(x, x\otimes \epsilon)-\epsilon \\
-&= X(x\otimes \epsilon, x\otimes \epsilon)=0
-\end{align*}
-so $Y(f(x)\otimes \epsilon, f(x\otimes \epsilon))=0$, which implies
-$f(x)\otimes \epsilon \preceq_{X}f(x\otimes \epsilon)$. 
-Moreover, if $x\preceq_{X}x'$, then $0\geq X(x,x')\geq Y(f(x),f(x'))$, whence 
-$f(x)\preceq_{Y}f(x')$, so $f$ is order-preserving.
-
-Conversely, for all $x,x'\in X$, 
-\begin{align*}
-X(x\otimes X(x,x'), x') &=X(x,x')-X(x,x')=0 
-\end{align*}
-thus $x\otimes X(x,x') \preceq_{X}x'$. Since $f$ is order-preserving, it follows that
-\begin{align*}
-f(x)\otimes X(x,x') \preceq_{Y}f(x\otimes X(x,x'))\preceq_{Y}f(x') 
-\end{align*}
-which implies that 
-\begin{align*}
-Y(f(x),f(x')) - X(x,x') = Y(f(x)\otimes X(x,x'), f(x'))=0
-\end{align*}
-that is $Y(f(x),f(x'))\leq X(x,x')$, so $f$ is a functor.
-
-\item[ii.]
-Suppose $f$ is a cocontinuous functors, and let  $g:Y\to X$, be its right-adjoint, i.e.~satisfying $Y(f(x),y)=X(x,g(y))$. Then 
-\begin{align*}
-Y(f(x\otimes \epsilon), y)& = X(x\otimes \epsilon, g(y)) \\
-&= X(x, g(y))-\epsilon \\
-&= Y(f(x), y)-\epsilon
-\end{align*}
-which implies that $f(x\otimes \epsilon)$ coincides with the tensor $f(x)\otimes \epsilon$. 
-Moreover, clearly also $f(x)\preceq_{Y}y$ iff $x\preceq_{X}g(y)$ holds, which means that $f$ is left-adjoint to $g$ also with respect to the order.
-
-Conversely, suppose the function $f:X\to Y$ preserves joins and tensors. Since $f$ is order-preserving, by i.~it is a functor, so we must only prove that it is cocontinuous.
-Since $f$ preserves joins there exists a function $g:Y\to X$ which is right-adjoint to $f$ with respect to orders, i.e.~$f(x)\preceq_{Y}y$ iff $x\preceq_{X}g(y)$. 
-We need to prove then that $f$ is left-adjoint to $g$, i.e.~that $Y(f(x),y)=X(x,g(y))$.
-
-On the one hand we have 
-\begin{align*}
-0 = X(x, g(y))-X(x,g(y))= X(x\otimes X(x,g(y)), g(y))
-\end{align*}
-from which it follows
-\begin{align*}
-0= Y(f(x\otimes X(x,g(y)), y)=Y(f(x)\otimes X(x,g(y)), y)=
-Y(f(x),y)-X(x,g(y))
-\end{align*}
-where the first inequality follows from the fact that $f$ and $g$ are adjoint with respect to the order (so $Y(f(x),y)=0$ iff $X(x,g(y))=0$).
-This implies then $Y(f(x),y)\leq X(x,g(y))$.
-
-For the converse inequality, 
-\begin{align*}
-0=Y(f(x),y)-Y(f(x)-y)=Y(f(x)\otimes Y(f(x),y),y)=
-Y(f(x\otimes Y(f(x),y)),y)
-\end{align*}
-and by a similar reasoning we deduce
-\begin{align*}
-0=X(x\otimes Y(f(x),y), g(y))=
-X(x,g(y))-Y(f(x),y)
-\end{align*}
-whence $X(x,g(y))\leq Y(f(x),y)$.
-\end{itemize}
-\end{proof}
-
-\begin{theorem}\label{thm:equivalence}
-The category $Q\mathsf{Mod}$ of $Q$-modules and $Q$-module morphism coincides with the category $Q\mathsf{CCat}$ of cocomplete skeletal $Q$-categories and cocontinuous functors.
-\end{theorem}
-\begin{proof}
-We have already seen that any cocomplete skeletal $Q$-category is a $Q$-module via tensors, 
-and that cocontinuous functors are $Q$-module morphisms.
-Let us now show that any $Q$-module is a cocomplete skeletal $Q$-category, and that a $Q$-module morphism is a cocontinuous functor.
-
-Let then $M=(M,\preceq, \star)$ be a $Q$-module. Define $M(x,y)= \inf\{ \delta \mid x\star \delta \succeq y\}$.
- It is clear that $M(x,x)=0$. Let us prove $M(x,y)+M(y,z) \succeq M(x,z)$: 
-from $x\star M(x,y)\succeq y$ and $y\star M(y,z)\succeq z$ we deduce 
-$x\star(M(x,y)+M(y,z))= (x\star M(x,y))\star M(y,z) \succeq y\star M(y,z)\succeq z$, and thus 
-$M(x,z)\preceq M(x,y)+M(y,z)$. 
-Observe that $M(x,y)=0$ iff $x=x\star 0\geq y$, so the order $\preceq_{M}$ coincides with the order of $M$.
-
-Let us check that the $Q$-category $M$ is tensored via $x\otimes \epsilon:=x\star\epsilon$.
-Let $A_{x,y}= \{ \delta \mid (x\star\epsilon)\star \delta \geq y$ and 
-$B_{x,y}=\{\delta-\epsilon\mid x\star \delta \geq y\}$.
-Let us show that $A_{x,y}=B_{x,y}$: if $\delta \in A_{x,y}$, then 
-$\delta=(\epsilon+\delta)-\epsilon$ satisfies 
-$x\star(\epsilon+\delta)=(x\star\epsilon)\star \delta \geq y$, whence 
-$\delta\in B_{x,y}$. Conversely, if $\eta=\delta-\epsilon\in B_{x,y}$, then 
-$(x\star\epsilon)\star \eta \geq x\star \delta \geq y$, whence $\eta\in A_{x,y}$.
-We conclude then that $M(x\star\epsilon,y)=\inf A_{x,y}=\inf B_{x,y}=
-\inf\{\delta \mid x\star \delta \geq y\}-\epsilon=M(x,y)-\epsilon$.
-
-Let us define the opposite action $x\multimapinv \epsilon= \bigwedge\{y \mid 
-y\star \epsilon \geq x\}$. We must show that $M$ is cotensored via $\multimapinv$, for which it suffices to show $M(x\star \epsilon,y)=M(x,y\multimapinv \epsilon)$. Let $C_{x,y}=\{\delta \mid 
-x\star \delta \geq y\multimapinv \epsilon\}$. We have that $\delta \in A_{x,y}$ iff  
-$(x\star \delta)\star \epsilon=x\star(\delta+\epsilon)=x\star(\epsilon+\delta)=(x\star \epsilon)\star \delta \geq y$ which is equivalent to $x\star\delta \geq y\multimapinv \epsilon$.
-We conclude that $A_{x,y}=C_{x,y}$, from which $M(x\star \epsilon,y)=\inf A_{x,y}=\inf C_{x,y}=M(x,y\multimapinv \epsilon)$.
-
-Since $M$, as a $Q$-category, is order-complete, tensored and cotensored, it is cocomplete by Proposition \ref{prop:tencoten}.
-
-To conclude, notice that if $f:X\to Y$ is a cocontinuous functor, then it commutes with tensors and, by 
-Proposition \ref{prop:tenfun} it commutes with joins, so it is a morphism of the respective $Q$-modules.
-Conversely, if $f:M\to N$ is a $Q$-module morphism, then, since $M$ and $N$ are both tensored $Q$-categories, the tensor coincides with the actions of $M$ and $N$, $f$ preserves the joins and the tensor, by Proposition \ref{prop:tenfun}, it is a cocontinuous functor of the respective $Q$-categories.
-\end{proof}
-
-\subsection{$Q\mathsf{Mod}$ is a $*$-Autonomous Category}
-
-Let us first observe that:
-\begin{itemize}
-\item the hom-set $Hom(M,N)$ of two $Q$-modules is a $Q$-module with order and action defined pointwise;
-
-\item for any $Q$-module $M=(M,\preceq, \star)$, there is a $Q$-module
-$M^{\mathrm{op}}=(M,\succeq, \multimapinv)$, with $\multimapinv$ defined as in the proof of Theorem \ref{thm:equivalence}.
-
-\end{itemize}
-
-Let $M,N$ be two $Q$-modules. For all $A\in Q^{M\times N}$, we define the function
-\begin{align*}
-H_{A} & : Q^{M} \longrightarrow Q^{N}%\\
-%K_{A} & : Q^{N} \longrightarrow Q^{M}
-\end{align*}
-via
-\begin{align*}
-H_{A}(\B x)(b)  &= \inf_{a\in M}\B x_{a}+A(a,b)\\
-%K_{A}(\B x)(b)  &= \sup_{a\in M}\B x_{a}-A(a,b)
-\end{align*}
-
-\begin{lemma}
-$H_{A}=H_{A'}$ iff $A=A'$. 
-\end{lemma}
-\begin{proof}
-We only need to prove one direction, so suppose $A\neq A'$ and let $a,b$ be such that $A(a,b)\neq A'(a,b)$.
-Let $\B x$ be defined by $\B x_{a}=1$ and $\B x_{a'}=\infty$ for all $a'\neq a$. Then $H_{A}(\B x)(b)=A(a,b)\neq A'(a,b)=H_{A'}(\B x)(b)$. 
-\end{proof}
-
-\begin{proposition}
-$Q^{M\times N}$ and $Hom(Q^{M},Q^{N})$ are isomorphic $Q$-modules.
-\end{proposition}
-\begin{proof}
-The map $A\mapsto H_{A}$ is injective, as shown above. We need to check that it commutes with joins:
-\begin{align*}
-H_{\bigvee_{i}A_{i}}(\B x)(b) & = \inf_{a}\B x_{a}+\bigvee_{i}A_{i}(a,b)\\
-&=  \inf_{a}\bigvee_{i}\B x_{a}+A_{i}(a,b)\\
-&=  \bigvee_{i}\inf_{a}\B x_{a}+A_{i}(a,b)\\
-&=  \bigvee_{i}H_{A_{i}}(\B x)(b)\end{align*} 
-(recall that $\inf$s are actually joins in $Q$!)
-
-We must prove that $H$ is surjective: for all $f\in Hom(Q^{M},Q^{N})$, let 
-$k_{f}\in Q^{M\times N}$ be given by $k_{f}(a,b)=f(e_{a})(b)$.
-
-Then we have 
-\begin{align*}
-H_{k_{f}}(\B x)(b) & = \inf_{a}\B x_{a}+k_{f}(a,b) \\
-&= \inf_{a}\B x_{a}+ f(e_{a})(b)\\
-&= \left(\inf_{a}\B x_{a}+f(e_{a})\right)(b) \\
-&= \left(\inf_{a}f(\B x_{a}+e_{a})\right)(b) \\
-&= f(\inf_{a}\B x_{a}+e_{a})(b)\\
-&=
-f(\B x)(b)
-\end{align*}
-and conversely 
-\begin{align*}
-k_{H_{A}}(a,b)&= H_{A}(e_{a})(b)= \inf_{a'}(e_{a})_{a'}+A(a',b) =  A(a,b)
-\end{align*}
-\end{proof}
-
-More generally, we have the following result:
-\begin{proposition}
-Let $X,Y$ be two $Q$-modules. For any morphism $f: X\to Y$ there is a matrix $k_{f}\in Q^{X\times Y}$ such that
-
-\end{proposition}
-\begin{proof}
-By composing $f$ with the isomorphisms $\C Y^{\mathrm{op}}:X\to \B M(X)$, with inverse $\sup(\B x)=\bigvee_{a\in x}a\otimes \B x_{a}$, we obtain 
-a morphism $\widehat f: \B M(X)\to \B M(Y)$  
-\begin{align*}
-\widehat f(\B x)(b)&:= \C Y^{\mathrm{op}}(f(\sup \B x))(b) \\
-&= Y( f(\bigvee_{a\in X}a\otimes \B x_{a}),b) \\
-&= Y( \bigvee_{a\in X}f(a)\otimes \B x_{a},b) \\
-&= \sup_{a\in X}Y(f(a),b)-\B x_{a}
-\end{align*}
-Observe that $\widehat f$ can be extended to a function $f^{*}$ from $Q^{X}$ to $Q^{Y}$.
-Now, $ f^{*}$ is generated by the matrix $k_{f}\in Q^{X\times Y}$ given by $k_{ f}(a,b)=f(e_{a})(b)$, so that for all $\B x\in Q^{X}$, 
-$f^{*}(\B x)(b)=\bigvee_{a\in X}k_{f}(a,b)+\B x_{a}$, and thus
-in particular, for all $\B x\in [X,Q]$, 
-$\widehat f(\B x)(b)=f^{*}(\B x)(b)$.
-\end{proof}
-
-\subsubsection{The Tensor Product of $Q$-Modules}
-
-The description of the tensor product of $Q$-modules requires some work. Let us first recall some important definitions:
-
-\begin{definition}[congruence on a sup-lattice]
-Let $(L, \leq)$ be a sup-lattice. An equivalence relation $R\subseteq L\times L$ is said a \emph{congruence} if it satisfies the following property:
-\begin{align}
-(\forall i\in I \ x_{i} Ry_{i})  \ \To  \ \left( \bigvee_{i}x_{i} \right ) R\left(\bigvee_{i}y_{i}\right)
-\tag{congruence}
-\end{align}
-\end{definition}
-
-\begin{lemma}
-For all suplattices $(L,\leq)$, if $R$ is a congruence, then $(L/R, \leq_{R})$ is a sup-lattice, where $[x]\leq_{R}[y]$ iff $(x\vee y) R y$ (i.e.~$[x\vee y]=[y]$), and $\bigvee_{i}[x_{i}]=\left [\bigvee_{i}x_{i}\right]$.
-\end{lemma}
-\begin{proof}
-Let us check that $\leq_{R}$ is an order. It is clear that $[x]\leq_{R}[x]$ holds. If $[x]\leq_{R}[y]$ and $[y]\leq_{R}[z]$ both hold, then 
-$(x\vee y)Ry$ and $(y\vee z)Rz$ hold; 
-then, since $R$ is a congruence $((x\vee y)\vee (y\vee z)) R (y\vee z)R z$, and moreover $x \vee (y\vee z) R (x \vee z)$, whence 
-$(x\vee z) R(x \vee y\vee z)R z$, so $[x]\leq_{R}[z]$.  
-If $[x]\leq_{R}[y]$ and $[y]\leq_{R}[x]$, then 
-$xR(x\vee y)R y$, thus $[x]=[y]$.
-
-Let us now check the definition of joins. 
-From $[x_{i}]\vee [\bigvee_{i}x_{i}]=[x_{i}\vee \bigvee_{i}x_{i}]= [\bigvee_{i}x_{i}]$ we deduce $[x_{i}]\leq_{R}[\bigvee x_{i}]$.
-Suppose now $[x_{i}]\leq [y]$ holds for all $i\in I$, that is, 
-$(x_{i}\vee y)Ry$; then, since $R$ is a congruence, 
-$(\bigvee_{i}(x_{i}\vee y))R y$, that is, 
-$((\bigvee_{i}x_{i})\vee y)Ry$, which implies 
-$[\bigvee_{i}x_{i}]\leq_{R}[y]$. We conclude then that $\bigvee_{i}[x_{i}]=[\bigvee_{i}x_{i}]$.
-\end{proof}
-
-\begin{corollary}\label{cor:bigvee}
-Let $(L,\leq)$ be a suplattice and $R$ be a congruence on $L$.
-Then, for any class $\beta\in L/R$, $\beta= [\bigvee\beta]$.
-\end{corollary}
-\begin{proof}
-$[\bigvee \beta]=[\bigvee\{ x \mid x\in \beta\}]= \bigvee \{[x]\mid x\in \beta\}=\beta$.
-\end{proof}
-
-\begin{proposition}\label{prop:smallestcongruence}
-Let $(L,\leq)$ be a sup-lattice. Let $R\subseteq L\times L$ be an equivalence relation, and for any ordinal $\alpha$, let the relation
-$R^{(\alpha)}\subseteq L\times L$ be defined by:
-\begin{itemize}
-\item $xR^{(0)}y$ iff either $xRy$, $x=y$ or $yRx$ holds;
-\item $xR^{(\alpha+1)}y$ iff one of the following holds:
-	\begin{itemize}
-	\item for some $z$, $xR^{(\alpha)}z$ and $zR^{(\alpha)}y$ holds;
-	\item for some set $I$, and families $x_{i},y_{i}$, 
-	$x=\bigvee x_{i}, y=\bigvee_{i}y_{i}$ and $x_{i} R^{(\alpha)}y_{i}$ holds for all $i\in I$.
-
-	\end{itemize}
-\item $xR^{(\gamma)}y$ iff $xR^{(\delta)}y$ holds for some $\delta <\gamma$, for $\gamma$ limit.
-\end{itemize}
-Then the relation $R^{*}\subseteq L\times L$ given by 
-$$
-xR^{*} y  \ \Leftrightarrow \ \exists \alpha . \mathrm{OR}(\alpha) \land xR^{(\alpha)}y
-$$% defined as follows: $xR^{*} y$ iff
-%\begin{enumerate}
-%\item for any set $I$ and family $x_{i}$ such that $x=\bigvee_{i}x_{i}$, there exists a family $y_{i}$ such that $y=\bigvee_{i}y_{i}$ and $x_{i}R y_{i}$ holds for all $i\in I$;
-%\item for any set $I$ and family $y_{i}$ such that $y=\bigvee_{i}y_{i}$, there exists a family $x_{i}$ such that $x=\bigvee_{i}x_{i}$ and $x_{i}R y_{i}$ holds for all $i\in I$.
-%\end{enumerate}
-%%\begin{align*}
-%%xR^{*}y &\text{ iff }\  \forall I \ \forall x_{i} \  \text{ s.t. }
-%%\left\{
-%%\begin{matrix}
-%%x=\bigvee_{i\in I}x_{i} \\
-%%y=\bigvee_{i\in I}y_{i}\\
-%%x_{i}Ry_{i} \ (\forall i\in I)
-%%\end{matrix}
-%%\right\}
-%%\end{align*}
-is a congruence, and is the smallest congruence containing $R$.
-\end{proposition}
-\begin{proof}
-From $xR^{(0)}x$ it follows $xR^{*}x$.
-
-Let us prove by induction that for any ordinal $\alpha$, $R^{(\alpha)}$ is symmetric:
-for $\alpha=0$ this is immediate; suppose $xR^{(\alpha+1)}y$, then two cases are possible: either $xR^{(\alpha)}z$ and $zR^{(\alpha)}y$, then by IH 
-$yR^{(\alpha)}z$ and $zR^{(\alpha)}x$, whence $yR^{(\alpha+1)}x$; oer
-$x_{i}R^{(\alpha)}y_{i}$ for some decompositions $x=\bigvee_{i}x_{i}$ and $y=\bigvee_{i}y_{i}$; then by IH $y_{i}R^{(\alpha)}x_{i}$, so $yR^{(\alpha+1)}x$ holds.
-Finally, if $\alpha$ is limit, then from $x R^{(\alpha)}y$ it follows
-$xR^{(\beta)}y$ for some $\beta<\alpha$, whence 
-$yR^{(\beta)}x$ by IH and we conclude $yR^{(\alpha)}x$.
-
-Now, if $xR^{*}y$ then $xR^{(\alpha)}y$ holds for some ordinal $\alpha$, and thus $yR^{(\alpha)}x$ holds too, whence $yR^{*}x$.
-
-Observe that $\alpha<\beta$ implies $R^{(\alpha)}\subseteq R^{(\beta)}$:
-this is obvious if $\beta$ is limit, otherwise, if $\beta=\alpha+1$, from $xR^{(\alpha)}y$ and $x R^{(\alpha)}x$ we deduce
-$xR^{(\alpha+1)}y$.
-
-Suppose now $xR^{*}y$ and $yR^{*}z$. Then $xR^{(\alpha)}y$ and $yR^{(\beta)}z$ hold for some 
-ordinals $\alpha$ and $\beta$; let $\gamma=\max\{\alpha,\beta\}$; then 
-we have $xR^{(\gamma)}y$ and $yR^{(\gamma)}z$, whence $xR^{(\gamma+1)}z$ and thus 
-$xR^{*}z$.
-
-Suppose $x_{i}R^{*}y_{i}$ holds for all $i\in I$; then 
-for all $i$ there is some ordinal $\alpha_{i}$ with 
-$x_{i}R^{(\alpha_{i})}y_{i}$. 
-Let $\gamma=\sup_{i}\alpha_{i}$, so that 
-$x_{i}R^{(\gamma)}y_{i}$; then we have
-$\bigvee_{i}x_{i} R^{(\gamma+1)}\bigvee y_{i}$, and thus
-$\bigvee_{i}x_{i} R^{*}\bigvee_{i}y_{i}$.
-
-%
-%$R^{*}$ is symmetric, reflexive and transitive, given that $R$ is. 
-%
-%Let $I$ be a set and 
-%suppose $x_{i}R^{*}y_{i}$ holds for all $i\in I$.
-%Observe now that if $x=\bigvee_{i\in I} x_{i}=\bigvee_{j\in J}x'_{j}$, then
-%\begin{enumerate}
-%\item $x_{i}=\bigvee_{j\in J}x_{i}\land x_{j}$;
-%\item $x= \bigvee_{(i,j)\in I\times J}x_{i}\land x'_{j}$. 
-%\end{enumerate}
-% $\bigvee_{j\in J}x_{i}\land x_{j}\leq x_{i} \leq x_{i}$ is clear.
-% Conversely, we have $x_{i}= x \land x_{i} = \left (\bigvee_{j\in J}x_{j}\right) \land x_{i}= \bigvee_{j\in J}x_{j}\land x_{i}= \bigvee_{j\in J}x_{i}\land x_{j}$.
-{\color{red}NOT TRUE IN ANY LATTICE!}
-
-%
-% Then for each $i\in I$ there exists a set $J_{i}$ and sequences $x_{ij},y_{ij}$ such that 
-%$\bigvee_{j\in J_{i}}x_{ij}=x_{i}$, $\bigvee_{j\in J_{i}}y_{ij}=y_{i}$ and 
-%$x_{ij}Ry_{ij}$.
-%
-%Let then $K= \prod_{i\in I}\{i\}\times J_{i}$; then for all $(i,j)\in K$, 
-%$x_{ij}R y_{ij}$, so we deduce that $\bigvee_{i\in I}x_{i}=\bigvee_{(i,j)\in K}x_{ij} R^{*} \bigvee_{(i,j)\in K}y_{ij}=\bigvee_{i\in I}y_{i}$, which proves that $R^{*}$ is a congruence.
-
-Suppose now $S$ is a congruence containing $R$.
-Let us show that for any ordinal $\alpha$, $R^{(\alpha)}\subseteq S$:
-\begin{itemize}
-\item $R^{(0)}\subseteq S$ holds since $S$ is an equivalence relation and contains $R$;
-
-\item if $x R^{(\alpha+1)}y$ holds, then two cases occur:
-	\begin{itemize}
-	\item $xR^{(\alpha)}z$ and $zR^{(\alpha)}y$ hold, so by IH, 
-	$xSz$ and $zSy$, and since $S$ is transitive, $xSy$ holds;
-	\item $x=\bigvee_{i}x_{i}$, $y=\bigvee_{i}y_{i}$ and $x_{i}R^{(\alpha)}y_{i}$ holds; then by IH $x_{i}Sy_{i}$ holds, and since $S$ is a congruence, $xSy$ holds;
-
-\item if $\alpha$ is limit and $xR^{(\alpha)}y$ holds, then $xR^{(\beta)}y$ holds for some $\beta<\alpha$, and by IH $xS y$ holds.
-
-	\end{itemize}
-
-\end{itemize}
-Now, if $x R^{*} y$ holds, then $xR^{(\alpha)}y$ holds for some ordinal $\alpha$, whence $xS y$ holds.
-This shows that $R^{*}\subseteq S$.
-%
-%Then there exists a set $I$ so that $x=\bigvee_{i\in I}x_{i}$, $y=\bigvee_{i\in I}y_{i}$ and $x_{i}Ry_{i}$; now, since $S$ contains $R$, we deduce $x_{i}Sy_{i}$ for all $i\in I$, and since $S$ is a congruence, $xSy$ holds, which proves that $R^{*}\subseteq S$.
-\end{proof}
-
-We can now introduce the tensor of $Q$-modules, that will be defined as a suitable quotient lattice.
-
-\begin{definition}[tensor of $Q$-modules]
-Let $M$ and $N$ be $Q$-modules.
-The \emph{tensor product} $M\otimes_{Q}N$ of $M$ and $N$ is the $Q$-module defined as $\C P(M\times N)/ R^{*}$, where $R^{*}$ is the smallest congruence containing the relation $R$ defined by:
-\begin{align*}
-R'= \left\{
-\begin{matrix}
-\left((\bigvee A, y), \bigcup_{a\in A}\{(a,y)\}\right)\\
-\left((x,\bigvee B), \bigcup_{b\in B}\{(x,b)\}\right)\\
-(\{(x\star \epsilon,y)\}, \{(x,y\star\epsilon)\})
-\end{matrix}
-\ \Bigg \vert\ 
-\begin{matrix}
-A\subseteq M, y\in N \\
-B\subseteq N, x\in M \\
-\epsilon \in Q
-\end{matrix}
-\right\}
-\end{align*}
-and the action is defined via $[A]\star \epsilon= \bigvee\{[ \{(x\star\epsilon,y)\}]\mid 
-(x,y)\in A\}$.
-\end{definition}
-
-Let a \emph{$Q$-bimorphism} be a map $f:M\times N\to L$ such that $f$ preserves joins in each variable separately, and moreover $f(x,y\star \epsilon)=f(x\star\epsilon,y)$.
-A $Q$-bimorphism $f:M\times N\to L$ is \emph{universal} if for all $L'$ and bimorphism $g:M\times N\to L'$ there is a unique sup-lattice homomorphism $h:L\to L'$ such that $g=h\circ f$.
-
-\begin{proposition}[universal property of the tensor product]
-The tensor product $M\otimes_{Q}N$ is the codomain of the universal $Q$-bimorphism $M\times N \to M\otimes_{Q}N$.
-\end{proposition}
-
-\begin{remark}
-
-For any $x\in M$ and $y\in N$, we indicate as $x\otimes_{Q}y$ the image of the pair $(x,y)$ under the universal $Q$-bimorphism $\tau:M\times N\to M\otimes_{Q}N$, or equivalently, as the $R^{*}$-equivalence class of $(x,y)$.
-Since joins in $M\otimes_{Q}N$ are given by  
-$\bigvee_{i}[A_{i}]= [\bigcup_{i}A_{i}]$, we have then 
-that any element $[A]\in M\otimes_{Q}N$ can be written as 
-$[A]= \bigvee\{x\otimes y\mid (x,y)\in A\}$.
-
-%
-% Any element of $M\otimes_{Q}N$ is a join of tensors, that is
-%$$
-%M\otimes_{Q}N= \left\{ \bigvee_{i\in I}x_{i}\otimes_{Q}y_{i}\ \Bigg \vert \  x_{i}\in M, y_{i}\in N \right\}
-%$$
-\end{remark}
-
-\begin{lemma}
-\begin{itemize}
-\item $M\otimes_{Q}N\simeq N\otimes_{Q}M$;
-
-\item $(M\otimes_{Q}N)\otimes_{Q}R \simeq M\otimes_{Q}(N\otimes_{Q}R)$.
-\end{itemize}
-\end{lemma}
-
-\begin{proposition}
-$Hom(M\otimes_{Q}N, R)   \simeq   Hom(M, Hom(N,R))$ (as an isomorphism of sup-lattices).
-\end{proposition}
-\begin{proof}
-Given $h:M\otimes_{Q}N\to R$, for all $x\in M$ define $h_{x}:N\to R$ by 
-$h_{x}(y)=h(x\otimes_{Q}y)$. We then have 
-$h_{x}(\bigvee_{i}y_{i})=h(x\otimes_{Q}\bigvee_{i}y_{i})=
-h(\bigvee_{i}x\otimes_{Q} y_{i})=
-\bigvee_{i}h(x\otimes_{Q} y_{i})=\bigvee_{i}h_{x}(y_{i})$ and 
-$h_{x}(y\star\epsilon)= h(x\otimes_{Q}(y\star \epsilon))=h((x\otimes_{Q}y)\star\epsilon)=h(x\otimes_{Q}y)\star \epsilon= h_{x}(y)\star\epsilon$, so $h_{x}\in Hom(N,R)$.
-Moreover, by a similar argument we have  $h_{\bigvee_{i}x_{i}}(y)=\bigvee_{i}h_{x_{i}}(y)$
-and $h_{x\star\epsilon}(y)=h_{x}(y)\star\epsilon$, so the map $x\mapsto h_{x}$ is a $Q$-module morphism.
-
-Finally, for any family $h_{i}:M\otimes_{Q}N\to R$, we have 
-$(\bigvee_{i}h_{i})_{x}(y)=\bigvee_{i}h_{i}(x\otimes_{Q}y)=\bigvee_{i}(h_{i})_{x}(y)=(\bigvee_{i}h_{i})(x\otimes y)$, and thus we have a sup-lattice homomorphism $\zeta$ from $Hom(M\otimes_{Q}N,R)$ to $Hom(M,Hom(N,R))$ given by $\zeta(h)=h_{\_}$.
-
-Let us show that $\zeta$ has an inverse:  for all $f\in Hom(M,Hom(N,R))$, define $f': M\times N\to R$ by $f'(x,y):=f(x)(y)$.
-This is clearly a bimorphism, so there is a unique homomorphism $h_{f'}:M\otimes_{Q}N\to M\times N$ such that $f'=h_{f'}\circ \tau$, i.e.~such that
-$h_{f'}(x\otimes y)=f'(x,y)=f(x)(y)$, and thus $\zeta(h_{f'})=f$.
-On the other hand, if $f=\zeta(h)$, then the uniqueness of $h_{f'}$ ensures that $h_{f'}=h$.
-\end{proof}
-
-\begin{proposition}
-\begin{itemize}
-\item[i.] $Hom(Q,M)\simeq M$.
-\item[ii.] $Hom(M,N)\simeq Hom(N^{\mathrm{op}},M^{\mathrm{op}})$.
-\item[iii.] $Hom(M, Q^{\mathrm{op}})\simeq M^{\mathrm{op}}$.
-\end{itemize}
-(all isomorphisms of sup-lattices).
-\end{proposition}
-\begin{proof}
-Define $\alpha:M\to Hom(Q,M)$ by $\alpha(x)(\epsilon)=x\star\epsilon$ and 
-$\beta:Hom(Q,M)\to M$ by $\beta(f)=f(0)$. Then we have that 
-$\alpha(\beta(f))(\epsilon)=\alpha(f(0))(\epsilon)=f(0)\star\epsilon=f(\epsilon)$, and 
-$\beta(\alpha(x))=\beta(\lambda \epsilon.x\star\epsilon)=x\star0=x$.
-
-If $f\in Hom(M,N)$, since it preserves joins, it has a right-adjoint $f^{*}:N^{\mathrm{op}}\to M^{\mathrm{op}}$, such that $f(x)\leq y$ iff $x\leq f^{*}(y)$.
-
-By i.~$Hom(Q,M^{\mathrm{op}})\simeq M^{\mathrm{op}}$ and we conclude by ii.
-\end{proof}
-
-\begin{proposition}
-\begin{itemize}
-\item[i.] $Hom(M,N)\simeq (M\otimes_{Q}N^{\mathrm{op}})^{\mathrm{op}}$.
-\item[ii.] $M\otimes_{Q}N\simeq Hom(M, N^{\mathrm{op}})^{\mathrm{op}}$.
-\item[iii.] $Q\otimes_{Q} M \simeq M\otimes_{Q}Q\simeq Q$.
-\end{itemize}
-\end{proposition}
-\begin{proof}
-$Hom(M,N)\simeq Hom(M, Hom(N^{\mathrm{op}},Q^{\mathrm{op}}))
-\simeq Hom(M\otimes N^{\mathrm{op}},Q^{\mathrm{op}})\simeq
-(M\otimes N^{\mathrm{op}})^{\mathrm{op}}$.
-Claim ii.~is proved similarly.
-
-For claim iii.~$Q\otimes_{Q}M\simeq M\otimes_{Q}M\simeq Hom(M, Q^{\mathrm{op}})^{\mathrm{op}}\simeq (M^{\mathrm{op}})^{\mathrm{op}}=M$. 
-\end{proof}
-
-By putting all previous results together we obtain:
-\begin{theorem}
-$Q\mathsf{Mod}$ is a $^{*}$-autonomous category.
-\end{theorem}
-
-\begin{proposition}\label{prop:Qtensor}
-\begin{itemize}
-\item[i.] $Q^{X}\otimes_{Q}M \simeq M^{X}$;
-\item[ii.] $Q^{X}\otimes_{Q}Q^{Y}\simeq Q^{X\times Y}$.
-\end{itemize}
-\end{proposition}
-\begin{proof}
-$M^{X}$ coincides with the product $\Pi_{x\in X}M$. We have then 
-$Q^{X}\otimes_{Q}M \simeq (\Pi_{x}Q)\otimes_{Q}M \simeq \Pi_{x}(Q\otimes_{Q}M) \simeq \Pi_{x}M\simeq M^{X}$.
-
-For ii., using i.~we have $Q^{X}\otimes_{Q}Q^{Y}\simeq( Q^{Y})^{X}\simeq Q^{X\times Y}$.
-\end{proof}
-
-\subsubsection{The Tensor Product of $Q$-Categories}
-
-Thanks to Theorem \ref{thm:equivalence}, the $^{*}$-autonomous structure of $Q\mathsf{Mod}$ translates into a $^{*}$-autonomous structure for $Q\mathsf{CCat}$.
-
-In $\mathsf{Met}$ the ``tensor product'' of two metric spaces $X$ and $Y$ is just their cartesian product, with the ``plus'' metric. What can we say about the tensor product in $Q\mathsf{CCat}$?
-
-% 
-%Let us start by discussing the construction of the tensor product $M\otimes_{Q}N$ of $Q$-modules $M$ and $N$ as a quotient of $\C P(M\times N)$.
-First, $\C P(M\times N)$ is the \emph{free} sup-lattice over $M\times N$: this means that any morphism
-%$f: M\times N \to L$, where $L$ is a sup-lattice, uniquely extends into a sup-lattice morphism $f^{*}:\C P(M\times N)\to L$, given by $f^{*}(A)=\bigvee_{x\in A}f(x)$. 
-%
-%
-%\begin{lemma}
-%For all $m\in M$ and $n\in N$, 
-%\begin{align*}
-%m\otimes n= &  \{A \mid \bigvee A= (m',n'+\epsilon+\delta), m=m'+\epsilon, n=n'+\delta\}\\
-%& \cup
-% \{A \mid \bigvee A= (m'+\epsilon+\delta, n'), m=m'+\epsilon, n=n'+\delta\}\end{align*}
-%\end{lemma}
-%\begin{proof}
-%
-%
-%\end{proof}
-The goal of this subsection is to describe the $Q$-categorical structure of the tensor product explicitly.
-The main intuition is that the elements of $X\otimes_{Q}Y$ can be seen as joins of pairs $x\otimes y$ of elements $x\in X$, $y\in Y$.
-We will then show that the metric coincides with the ``plus'' metric over pairs $x\otimes y$, and extends continuously to joins.
-
-\begin{lemma}\label{lemma:tensorsum}
-For all $m,m'\in M$ and $n,n'\in N$ and $\epsilon \in Q$, 
-$(m\otimes n)\star \epsilon \succeq (m'\otimes n')$ iff there exists $\delta_{1},\delta_{2}$ such that $\delta_{1}+\delta_{2}=\epsilon$, 
-$m+\delta_{1}\succeq m'$ and $n+\delta_{2}\succeq n'$.
-\end{lemma}
-\begin{proof}[Proof Sketch]
-Notice that $(m\otimes n)\star \epsilon = [\{(m\star \epsilon,n)\}]=
-[\{(m\star \delta_{1},n\star \delta_{2}\}]$ for all $\delta_{1}+\delta_{2}=\epsilon$. 
-Hence, $m'\otimes n' \preceq (m\otimes n)\star \epsilon$ implies that 
-for some $\delta_{1}+\delta_{2}=\epsilon$, 
-$(m',n')\vee (m\star\delta_{1},n\star\delta_{2}) = (m'\vee (m\star\delta_{1}), n'\vee (n\star\delta_{2}))= (m\star \delta_{1}, n'\star\delta_{2})$, that is, that $m'\preceq_{M} m\star \delta_{1}$ and $n'\preceq_{n}n\star \delta_{2}$.
-\end{proof}
-
-\begin{proposition}\label{prop:tensormetric}
-For all $m,m'\in N$ and $n,n'\in N$, 
-$$
-(M\otimes_{Q} N)(m\otimes_{Q} n, m'\otimes_{Q}n')= M(m,m') +N(n,n')
-$$
-More generally, 
-$$
-(M\otimes_{Q}N)([A],[B])= \sup_{(x,y)\in A}\inf_{(x',y')\in B}M(x,x')+N(y,y')
-$$
-\end{proposition}
-\begin{proof}
-By definition, $(M\otimes_{Q} N)(m\otimes_{Q} n, m'\otimes_{Q}n')$ is given by $\inf A$, where
-$$
-A=\{ \epsilon \mid (m\otimes_{Q}n)\star\epsilon \geq m'\otimes_{Q}n'\}
-$$
-Let us show that $A$ coincides with 
-$$
-B=\{ \delta_{1}+\delta_{2} \mid m\star\delta_{1} \geq  m', n\star \delta_{2}\geq n'\}$$
-On the one hand, if $\delta_{1}+\delta_{2}\in B$, it is clear that $\delta_{1}+\delta_{2}\in A$.
-Conversely, if $\epsilon\in A$, then by Lemma \ref{lemma:tensorsum} $\epsilon=\delta_{1}+\delta_{2}$ with $m\star\delta_{1}\geq m'$ and $n\star \delta_{2}\geq n'$, whence
-$\epsilon \in B$.
-
-We can now conclude as follows:
-\begin{align*}
-(M\otimes_{Q} N)(m\otimes_{Q} n, m'\otimes_{Q}n')& =\inf A \\
-&= \inf B \\
-&= \inf \{ \delta_{1} \mid m\star\delta_{1} \geq  m'\}
-+ \inf\{ \delta_{2} \mid  n\star \delta_{2}\geq n'\}\\
-&= M(m,m')+N(n,n').
-\end{align*}
-
-For the second claim, using the fact that $M\otimes_{Q}N$, as a $Q$-category, is both tensored and cotensored, using the fact that 
-$[A]=\bigvee_{(x,y)\in A}x\otimes y$ and $[B]=\bigvee_{(x',y')\in B}z\otimes w$, and Lemma \ref{lemma:supinf}:
-\begin{align*}
-(M\otimes_{Q} N)([A],[B])&=
-\sup_{(x,y)\in A}(M\otimes_{Q}N)(x\otimes y, [B]) \\
-&=\sup_{(x,y)\in A}\inf_{(x',y')\in B}(M\otimes_{Q}N)(x\otimes y, x'\otimes y') 
-\\
-&=
-\sup_{(x,y)\in A} \inf_{(x',y')\in B}M(x,x')+N(y,y').
-\end{align*}
-\end{proof}
-
-\subsection{Products and Coproducts in $Q\mathsf{Mod}$}
-
-Products and coproducts exist in $Q\mathsf{Mod}$ and both coincide with the order product:
-$$
-\prod_{i\in I}X_{i} \simeq \coprod_{i\in I}X_{i}
-$$
-In particular, the projection and
- injection morphisms $\pi_{i}:\prod_{i}X_{i}\to X_{i}$ and $\iota_{i}:X_{i} \to \prod_{i\in I}X_{i}$ are defined by 
-$$
-\pi_{i}\big( (x_{j})_{j\in I}\big) = x_{i} \qquad \qquad
-\iota_{i}(x)({j})= \begin{cases} x & \text{ if }i=j\\ \bot & \text{ otherwise}\end{cases}
-$$
-
-The following property is important
-\begin{proposition}\label{prop:productvstensor}
-$\prod_{i\in I} X\otimes Y_{i} \simeq X \otimes \prod_{i\in I}Y_{i}$.
-\end{proposition}
-\begin{proof}
-For a complete proof, see \texttt{https://arxiv.org/pdf/0909.4493.pdf}. Here we just recall the isomorphism 
-$h: \prod_{i\in I} X\otimes Y_{i} \simeq X \otimes \prod_{i\in I}Y_{i}$, defined as follows:
-\begin{align*}
-h\Big( i\mapsto \bigvee_{k\in J_{i}}x_{i,k}\otimes y_{i,k}\Big ) = 
-\bigvee_{i\in I,k\in J_{i}}x_{i,k} \otimes \iota_{i}\big( (y_{j,k})_{j\in I}\big)
-\end{align*}
-and its inverse
-\begin{align*}
-k\Big( \bigvee_{k\in J}x_{k}\otimes \big(x_{k,i}\big)_{i\in I_{k}}\Big)(i)=
-\bigvee_{k\in J,i\in I_{k}}x_{k}\otimes \pi_{i}\big((x_{k,j})_{j\in I_{k}}\big).
-\end{align*}
-\end{proof}
-
-\subsection{Exponential and Differential Structure of $Q\mathsf{Mod}$}
-
-\subsubsection{Symmetric Algebras in $Q\mathsf{Mod}$ }
-
-We start by constructing the exponential co-modality in the equivalent category $Q\mathsf{Mod}$.
-
-For all $n\in \BB N$ and $\sigma\in \F S_{n}$, let $\langle \sigma\rangle: X^{\otimes n}\to X^{\otimes n}$ be the morphism defined by 
-$$
-\langle \sigma \rangle (x_{1}\otimes \dots \otimes x_{n})=
-x_{\sigma(1)}\otimes \dots \otimes x_{\sigma(n)}
-$$
-and extended to all other points by continuity.
-%For any $Q$-module $X$, let $X_{\bullet}$ indicate the $Q$-module $X\times Q$.  
-%
-\begin{definition}Let $X$ be a $Q$-module and $n\in \BB N$. 
-An element $x\in X^{\otimes_{n}}$ is said \emph{permutation-invariant} (in short, \emph{p-invariant}) if for all $\sigma\in \F S_{n}$, 
-$\langle \sigma \rangle (x)=x$.
-
-For all $n\in \BB N$, $x_{1},\dots, x_{n}\in X$, the \emph{$Q$-multiset} $[x_{1},\dots, x_{n}]$ is the element of $X^{\otimes_{n}}$ defined as $0$ if $n=0$, and otherwise as follows:
-$$
-[x_{1},\dots, x_{n}]:= \bigvee_{\sigma\in \F S_{n}}x_{\sigma(1)}\otimes \dots \otimes x_{\sigma(n)}
-$$
-
-Given multisets $A$ and $B$, we define the multiset $A\cup B$ as follows:
-\begin{itemize}
-\item if $A=0$, then $A\cup B=B$;
-\item if $B=0$, then $A\cup B=A$;
-\item if $A=[x_{1},\dots, x_{n}]$ and $B=[y_{1},\dots, y_{m}]$, then $A\cup B=[x_{1},\dots, x_{n},y_{1},\dots, y_{m}]$.
-
-\end{itemize}
-
-\end{definition}
-
-\begin{proposition}
-Let $X$ be a $Q$-module and $n\in \BB N$. Any $Q$-multiset $[x_{1},\dots, x_{n}]\in X^{\otimes_{n}}$ is p-invariant. Moreover, any p-invariant element $x\in X^{\otimes_{n}}$ can be written as 
-a join of $Q$-multisets.
-\end{proposition}
-\begin{proof}
-For the first claim we have, for all $\sigma\in \F S_{n}$, 
-\begin{align*}
-\langle \sigma \rangle ([x_{1},\dots, x_{n}]) & = 
-\bigvee_{\tau\in \F S_{n}}\langle \sigma \rangle ([x_{\tau(1)},\dots, x_{\tau(n)}])\\
- & = 
-\bigvee_{\tau\in \F S_{n}}[x_{\sigma\tau(1)},\dots, x_{\sigma\tau(n)}])\\
- & = 
-\bigvee_{\tau\in \F S_{n}}[x_{\tau(1)},\dots, x_{\tau(n)}])\\
-&= [x_{1},\dots, x_{n}].
-\end{align*}
-For the second claim, observe that $x$ can always be written as a join of tensors $x=\bigvee_{i}x_{1}^{i}\otimes \dots \otimes x_{n}^{i}$. Moreover, 
-if $x_{1}^{i}\otimes \dots \otimes x_{n}^{i}\leq x$, since $x$ is p-invariant, for all $\sigma \in \F S_{n}$, also
-$x_{\sigma(1)}^{i}\otimes \dots \otimes x_{n}^{i}\leq \langle \sigma\rangle(x)=x$, so we can conclude that 
-$x=\bigvee_{i}[x_{1}^{i},\dots, x_{n}^{i}]$.
-\end{proof}
-
-\begin{proposition}
-For any $Q$-module $X$, the set $!_{n}X\subseteq X$ of p-invariant elements of $X^{\otimes_{n}}$ is a sub-$Q$-module of $X$.
-\end{proposition}
-\begin{proof}
-If $x_{i}\in X^{\otimes_{n}}$ is a family of p-invariant elements, then 
-$x=\bigvee_{i}x_{i}$ is also p-invariant, since $\langle \sigma\rangle (x)=\bigvee_{i}\langle \sigma \rangle (x_{i})=\bigvee_{i}x_{i}=x$. Hence $!_{n}X$ is a sup-lattice.
-Moreover, for all $x\in !_{n}X$ and $\epsilon \in Q$, 
-$x\otimes \epsilon$ is also p-invariant, since $\langle \sigma \rangle (x\otimes \epsilon)= \langle \sigma \rangle (x)\otimes \epsilon=x\otimes \epsilon$.
-We conclude that $!_{n}X$ is a sup-lattice with a continuous action of $Q$, where both the order and the action are inherited from $X$, so it is a sub-$Q$-module of $X$.
-\end{proof}
-
-The fundamental property of $!_{n}X$ is the following:
-\begin{proposition}
-For any $Q$-module $X$ and $n\in \BB N$, the inclusion morphism 
-$\iota:!_{n}X\longrightarrow X^{\otimes_{n}}$ is the equalizer of the diagram
-$$
-\begin{tikzcd}
-!_{n}X \ar{r}{\iota} & X^{\otimes_{n}} \ar{r}{\langle \sigma\rangle} & X^{\otimes_{n}}
-\end{tikzcd}
-$$
-generated by all actions $\langle \sigma\rangle$, for $\sigma\in \F S_{n}$.
-\end{proposition}
-\begin{proof}
-It is clear that $\langle \sigma \rangle \circ \iota= \langle \tau\rangle \circ \iota$ holds for all $\sigma, \tau \in \F S_{n}$.
-Suppose now $h: C\to X^{\otimes_{n}}$ is another morphism satisfying
-$\langle \sigma \rangle \circ h= \langle \tau\rangle \circ h$ for all $\sigma, \tau \in \F S_{n}$.
-Since $\langle \sigma \rangle \circ h= \langle \mathrm{id}\rangle \circ h=h$, we deduce that $h(c)$ is p-invariant, for all $c\in C$.
-Hence $h$ splits in a unique way as $C \stackrel{h}{\to} !_{n}X \stackrel{\iota}{\to} X^{\otimes_{n}}$.
-\end{proof}
-
-\begin{remark}[metric structure of $!_{n}X$]
-As $!_{n}X$ is a sub-$Q$-module of $X^{\otimes_{n}}$, the distance function can be computed explicitly using Proposition \ref{prop:tensormetric}:
-\begin{align*}
-!_{n}X( [x_{1},\dots, x_{n}], [y_{1},\dots, y_{n}]) & = 
-\sup_{\sigma\in \F S_{n}}\inf_{\tau\in \F S_{n}}
-\sum_{i=1}^{n}
-X(x_{\sigma(i)},y_{\tau(i)})
-\end{align*}
-\end{remark}
-
-We now show that the $Q$-module $!_{n}X$ is isomorphic to the symmetric algebra, which is used in the construction of the exponential modality in the relational model.
-
-\begin{definition}[symmetric algebra]
-For any $Q$-module $X$ and $n\in \BB N$, we let $\mathrm{Sym}_{n}(X)$ indicate the $Q$-module defined as 
-$\mathrm{Sym}_{n}(X):=\frac{X^{\otimes_{n}}}{\sim_{n}}
-$, where $\sim_{n}$ is the least congruence generated by the action $\langle \sigma\rangle$ of permutations $\sigma\in \F S_{n}$.\end{definition}
-
-\begin{proposition}
-$!_{n}X\simeq \mathrm{Sym}_{n}(X) $.
-\end{proposition}
-\begin{proof}
-First, observe that for any equivalence class $\alpha\in \mathrm{Sym}_{n}(X)$, the point $\bigvee\alpha$ is p-invariant: 
- since $x\in \alpha$ holds iff $\langle \sigma \rangle (x)\in \alpha$, for all $\sigma\in\F S_{n}$, 
-it follows that $\langle \sigma \rangle (\bigvee \alpha)=\bigvee\{\langle \sigma \rangle (x)\mid x\in \alpha\}=\bigvee \{x\mid x\in \alpha\}=\bigvee \alpha$, and thus $\bigvee\alpha$ is p-invariant.
-
-Now, let us show that for all $x\in X^{\otimes_{n}}$, $x \sim_{n} \bigvee[x]$: for all $y\in [x]$, by definition $x\sim_{n}y$ holds; hence, since $\sim_{n}$ is a congruence, we have that 
-$x=\bigvee_{y\in[x]}x \sim_{n} \bigvee_{y\in [x]}y=\bigvee[x]$.
-Observe that this implies that $[\bigvee[x]]=[x]$.
-
-Let us now show that for all p-invariant point $x_{0}$, and for all $y,z\in X^{\otimes_{n}}$, if 
-$y\leq x_{0}$ and 
-$z\sim^{\alpha}y$ holds, then $z\leq x_{0}$.
-\begin{itemize}
-\item if $z\sim^{0} y$, then either $z=y$, in which case the claim follows from the hypothesis, or $z=z_{1}\otimes \dots \otimes z_{n}$ and $y=y_{\sigma(1)}\otimes \dots \otimes y_{\sigma(n)}$; then from $y\leq x_{0}$ we deduce $z=\langle \sigma^{-1} \rangle(y)\leq \langle \sigma^{-1}\rangle(x_{0})=x_{0}$.
-
-\item if $z\sim^{\alpha+1}y$ two possibilities arise:
-	\begin{enumerate}
-	\item if $z\sim^{\alpha}z'\sim^{\alpha}y$, then by IH we have $z'\leq x_{0}$, and again by IH applied to $z'$ we deduce $z\leq x_{0}$;
-	\item $z=\bigvee_{i}z_{i}$ and $y=\bigvee_{i}y_{i}$, with $z_{i}\sim^{\alpha}y_{i}$, then from $y_{i}\leq y\leq x_{0}$, we deduce, by IH, $z_{i}\leq x_{0}$, and thus $z\leq x_{0}$.
-
-	\end{enumerate}
-
-\item if $z\sim^{\gamma}y$ for $\gamma$ limit, then $z\sim^{\beta}y$ for some $\beta<\gamma$, so by IH we deduce $z\leq x_{0}$.
-
-\end{itemize}
-From the argument above we now deduce that for all p-invariant point $x_{0}$, and for all $y,z\in X^{\otimes_{n}}$, if 
-$y\leq x_{0}$ and 
-$z\sim_{n}y$ holds, then $z\leq x_{0}$.
-From this we can deduce in turn that for all $x\in X^{\otimes_{n}}$, $\bigvee[x]$ is the smallest p-invariant over $x$: suppose $x_{0}$ is a p-invariant point and $x\leq x_{0}$; then for all $y\in [x]$, we deduce $y\leq x_{0}$ by the argument above, and we can thus conclude that $\bigvee[x]\leq x_{0}$.
-
-Let now $x$ be p-invariant; as $x$ is the smallest p-invariant over $x$, we deduce that $x= \bigvee[x]$.
-
-Using the previous facts we can now define an isomorphism $h:!_{n}X\to  \mathrm{Sym}_{n}(X)$ by letting  $h(x)=[x]$, with inverse $k([x])=\bigvee [x]$. Indeed, we have that 
-$k(h(x))=\bigvee[x]=x$, and 
-$h(k([x]))=[\bigvee[x]]=[x]$.
-%
-%First, let us show that for all $x\in !_{n}X\subseteq X^{\otimes_{n}}$, the corresponding equivalence class in $\mathrm{Sym}_{n}(X)$ is a singleton, i.e.~$[x]=\{x\}$. 
-%To prove this, let us first show that for all ordinals $\alpha$ and p-invariant $x$, if $x \sim_{n}^{(\alpha)}y$ then $x=y$, where $\sim_{n}^{(\alpha)}$ is defined as in Proposition \ref{prop:smallestcongruence}.
-%\begin{itemize}
-%\item $x\sim_{n}^{(0)}y$ holds iff either $x=y$, $x\sim_{n}y$ or $y\sim_{n}x$; if $x\sim_{n}y$, then it must be $x=x_{1}\otimes \dots \otimes x_{n}$ and $y=\langle \sigma\rangle(x)$, but since $x$ is permutation-closed, $x=\langle \mathrm{id}\rangle(x)=\langle \sigma\rangle(x)=y$.
-%
-%\item $x\sim_{n}^{(\alpha+1)}y$ holds iff either $x\sim_{n}^{(\alpha)}z$ and $z\sim_{n}^{(\alpha)}y$ both hold, or $x=\bigvee_{i}x_{i}$, $y=\bigvee_{i}y_{i}$ and $x_{i}\sim_{n}^{(\alpha)}y_{i}$ all hold.
-%In the first case, by IH we have $x=z$, so $z$ is p-invariant, and by applying again the IH, also $y=z$ holds, and thus $x=y$; 
-%in the second case, {\color{red}by IH we have $x_{i}=y_{i}$ for all index $i$, whence 
-%$x=\bigvee_{i}x_{i}=\bigvee_{i}y_{i}=y$.}
-%
-%\item if $x\sim_{n}^{(\gamma)}y$ for $\gamma$ limit, then
-%$x\sim_{n}^{(\beta)}y$ holds for some $\beta<\gamma$, so by IH, $x=y$.
-%
-%
-%\end{itemize}
-%Now, if $x\sim_{n}y$ holds, then $x\sim_{n}^{(\alpha)}y$ holds for some ordinal $\alpha$, whence $x=y$. 
-%
-%
-%
-%
-%Now, the main claim follows from the existence of the following two morphisms:
-%a morphism $h:!_{n}X\to \mathrm{Sym}_{n}(X)$ given by $h(x)=[x]=\{x\}$ and a morphism $k: \mathrm{Sym}_{n}(X)\to !_{n}X$ given by $k(\alpha)=\bigvee\alpha$.
-Then $k\circ h(x)=x$ while $h\circ k(\alpha)=[\bigvee \alpha]=\{\bigvee\alpha\}$, so $k$ and $h$ define an isomorphism between $!_{n}X$ and the $\sim$-classes of $\mathrm{Sym}_{n}(X)$.
-\end{proof}
-
-The following lemma shows the compatibility of the construction of $!_{n}X$ with the usual construction of the exponential modality in weighted relational models.
-\begin{lemma}
-For any set $S$, there exists an isomorphism of $Q$-modules
-$$!_{n}Q^{S} \simeq Q^{\C M_{n}(S)}$$
-where $\C M_{n}(X)$ indicates the set of multisets of $X$ of cardinality $ n$.
-\end{lemma}
-\begin{proof}
-Let us show that the morphism $h:Q^{\C M_{n}(S)}\to Q^{S\times \dots \times S}$ defined by 
-$$
-h(f)(\langle s_{1},\dots, s_{n}\rangle)=h([s_{1},\dots, s_{n}])
-$$
-is the equalizer of the diagram 
-$$
-\begin{tikzcd}
-Q^{\C M_{n}(S)} \ar{r}{h} &Q^{S\times \dots \times S}\ar{r}{[\sigma]} &
-Q^{S\times \dots \times S}
-\end{tikzcd}
-$$
-where $[\sigma](\B x)(\langle s_{1},\dots, s_{n}\rangle)=\B x(\langle x_{\sigma(1)},\dots, x_{\sigma(n)}\rangle)$, with $\sigma$ varying over $\F S_{n}$.
-
-It is immediate that $h\circ [\sigma]=h\circ [\tau]$, for all $\sigma,\tau\in \F S_{n}$.
-Let now $k: C\to Q^{S\times \dots \times S}$ satisfy $k\circ [\sigma]=k\circ [\tau]$: then for all $c\in C$, $k(c)(\langle s_{1},\dots, s_{n}\rangle)=k(c)(\langle s_{\sigma(1)},\dots, s_{\sigma(n)}\rangle)$, so $k(c)$ actually defines a unique element of $Q^{\C M_{n}(S)}$, and thus $k$ splits in a unique way as $C \stackrel{k'}{\to} Q^{\C M_{n}(S)} \stackrel{h}{\to}Q^{S\times \dots \times S}$.
-
-Now, to conclude it suffices to observe that, by Proposition \ref{prop:Qtensor}, 
-$Q^{S\times \dots \times S}\simeq (Q^{S})^{\otimes_{n}}$, and then, since equalizers are unique up to a unique isomorphism, we obtain an isomorphism $Q^{\C M_{n}(S)}\simeq !_{n}Q^{S}$.
-%
-%We define morphisms $h: !_{n}Q^{S}\to Q^{\C M_{n}(S)}$ and 
-%$k: Q^{\C M_{n}(S)}\to !_{n}Q^{S}$ given by 
-%\begin{align*}
-%h([\B x_{1},\dots, \B x_{n}])(\{a_{1},\dots, a_{n}\})&=
-%\sup_{\sigma\in \F S_{n}}\sum_{i=1}^{n}
-%\B x_{i}(a_{\sigma(i)})\\
-%k(f)& = 
-%\bigvee_{a_{1},\dots, a_{n}\in S}[\C Y(a_{1}),\dots, \C Y(a_{n})]+ f(\{a_{1},\dots, a_{n}\})
-%\end{align*}
-%
-\end{proof}
-
-\subsubsection{Linear Differential Categories}
-
-There exist many equivalent way to describe linear differential structure over symmetric monoidal categories with biproducts.
-Here we chose the approach via bialgebra modalities (see Ehrhard's stuff, but also Lemay and company's stuff for a clearer picture).
-
-\begin{definition}[bialgebra modality]
-Let $\BB C$ be an additive symmetric monoidal category. A \emph{bialgebra modality over $\BB C$} is a septuple $(!,\delta,\epsilon, \Delta, e,\nabla, u)$ consisting of:
-\begin{enumerate}
-\item a comonad $(!,\delta,\epsilon)$, that is, a functor $!$ together with natural transformations $\delta: !X\to !!X$ and $\epsilon:!X\to X$ satisfying
-\begin{align}
-\epsilon \circ \delta & = !\epsilon \circ \delta= 1 \\
-!\delta \circ \delta  & = \delta\circ !\delta
-\end{align}
-
-\item two natural transformations $\Delta:!X\to !X\otimes !X$ and $e:!X\to \B 1$ such that $(!X, \Delta, e)$ is a cocommutative comonoid, that is the following equations hold:
-\begin{align}
-(\Delta \otimes  1)\circ \Delta & = ( 1\otimes \Delta )\circ \Delta \\
- ( 1\otimes e)\circ \Delta & = (e\otimes  1)\circ \Delta =1 \\
-\sigma \circ\Delta& =  \Delta
-\end{align}
-and $\delta$ preserves the comultiplication, that is
-\begin{align}
-(\delta \otimes \delta)\circ \Delta = \Delta \circ \delta
-\end{align}
-
-\item two natural transformations $\nabla: !X\otimes !X \to !X$ and $u:\B 1\to !X$ such that $(!X,\nabla, u)$ is a commutative monoid, that is, the following equations hold:
-\begin{align}
-\nabla\circ (\nabla\otimes 1)  & = \nabla \circ (1\otimes \nabla) \\
- \nabla \circ(1\otimes u )& = \nabla \circ  (u\otimes 1) =1\\
-\nabla\circ \sigma & = \nabla
-\end{align}
-
-\item $(!X,\nabla, u, \Delta, e)$ is a bialgebra, that is the following equations hold:
-\begin{align}
-e\circ\nabla  & =  e\otimes e\\
- \Delta \circ u& = u\otimes u \\
-u\circ e & = 1\\
- \Delta \circ \nabla 
-&=(\nabla\otimes \nabla)\circ (1\otimes \sigma\otimes 1)\circ (\Delta\otimes \Delta)
-\end{align}
-
-\item $\epsilon$ is compatible with $\nabla$, that is
-\begin{align}
- \epsilon \circ \nabla& =  (\epsilon\otimes e)+(e\otimes \epsilon)
-\end{align}
-
-\end{enumerate}
-
-\end{definition}
-
-\begin{definition}
-A \emph{codereliction} for a bialgebra modality $(!,\delta,\epsilon,\Delta, e,\nabla,u)$ is a natural transformation $\eta:X\to !X$ satisfying the following equations:
-\begin{align}
-e\circ \eta & = 0\\
- \Delta\circ \eta&  = (\eta\otimes u)+(u\otimes \eta) \\
- \epsilon\circ \eta & = 1 \\
-\delta \circ \nabla\circ(1\otimes \eta) & =  \nabla \circ 
- (\delta\otimes \eta) \circ  (1\otimes \nabla)\circ
-(\Delta \otimes \eta)
-\end{align}
-\end{definition}
-
-\begin{definition}
-Let $\BB C$ be an additive (i.e.~monoid-enriched) symmetric monoidal category with biproducts.
- $\BB C$ is a \emph{monoidal storage category} if it has a coalgebra modality satisfying the \emph{Seely isomorphisms}, that is, the maps
- $e:!\top \to \B 1$ and $\chi:!(X\times Y) \to !X\otimes !Y$, with $\chi=\Delta \circ  !(\pi_{0})\otimes!(\pi_{1})$, are isomorphisms (whence $!\top\simeq \B 1$ and $!(X\times Y)\simeq !X\otimes !Y$).
-\end{definition}
-
-\begin{definition}
-Let $\BB C$ be an additive symmetric monoidal category. A bialgebra modality $(!,\delta,\epsilon,\Delta,e,\nabla,u)$ on $\BB C$ is \emph{additive} if the following equations hold:
-\begin{align}
-\nabla\circ (!f\otimes !g)\otimes \Delta  & = !(f+g)\\ 
-u\circ e & = !0
-\end{align}
-\end{definition}
-
-We use the following result:
-\begin{theorem}[Lemay\&co]\label{theorem:lemay}
-Every additive symmetric monoidal category with an additive bialgebra modality and finite products satisfies the Seely isomorphisms.
-\end{theorem}
-
-\begin{definition}
-A \emph{linear differential category} is an additive symmetric monoidal category $\BB C$ with biproducts and a bialgebra modality with a codereliction and the Seely isomorphisms.
-\end{definition}
-
-\begin{theorem}
-For any linear differential category $\BB C$, the co-Kleisli category $\BB C_{!}$ is a cartesian closed differential category, with deriving transformation $\mathsf{D}(f)$ defined as follows:
-$$
-\begin{tikzcd}
-!(X\times X)\  \simeq\  !X\otimes !X 
-\ar{r}{1\otimes \epsilon}  & !X\otimes X \ar{r}{1\otimes \eta}&
-!X\otimes !X \ar{r}{\nabla}& !X \ar{r}{f} & Y
-\end{tikzcd}
-$$
-\end{theorem}
-
-\subsubsection{The Free Exponential Modality of $Q\mathsf{Mod}$}
-
-Using the recipe from [Melli\'es, Tabareau, Tasson 2010], together with Proposition \ref{prop:productvstensor}, the free exponential modality of $Q\mathsf{Mod}$ can be defined as
-$$
-! X:= \prod_{n\in \BB N}!_{n}X
-$$
-The functorial action of $!$ is defined, for a morphism $f: X\to Y$, as follows:
-\begin{align*}
-!f(g)(0) & = g(0) \\
-!f (g)(n+1)   & = \bigvee\Big \{
-[f(x_{1}),\dots, f(x_{n+1})] \mid [x_{1}, \dots, x_{n+1}]\leq g(n+1)\Big \}
-\end{align*}
-
-%where $g(n)=\bigvee_{i}
-%[x_{1}, \dots, x_{n}]$.
-
-The bialgebra modality $(!, \delta, \epsilon, \Delta, e, \nabla,u)$ is defined as follows:
-\begin{itemize}
-\item the comonad $(!,\delta,\epsilon)$ is given by:
-\begin{align*}
-\epsilon(f)& =  f(1) \\
-%\OV{\mathrm{der}}& = \iota_{1}\\
-\delta(f)(n) & = \bigvee\left\{ 
-\iota_{n}([\iota_{i_{1}}(a_{1}), \dots , \iota_{i_{n}}(a_{n})]) \ \Big \vert \ 
-a_{j}\in !_{i_{j}}X,  
-a_{1}\cup \dots \cup a_{n}\leq f(i_{1}+\dots+ i_{n})
-\right\}
-\end{align*}
-
-We have $\epsilon(\delta(\alpha))=\delta(\alpha)(1)=
-\bigvee\{\iota_{n}(\alpha(n))\mid n\in \BB N\}=\alpha$, and 
-\begin{align*}
-!\epsilon(\delta(\alpha))(n)&=
-\bigvee\{ [\epsilon(\alpha_{1}),\dots, \epsilon(\alpha_{n})]
-\mid
-[\alpha_{1},\dots, \alpha_{n}] \leq \delta(\alpha)(n)\}
- \\
- &=
- \bigvee\left \{ [\epsilon(\alpha_{1}),\dots, \epsilon(\alpha_{n})]
-\ \Bigg \vert \ 
-\alpha_{i}=\iota_{j_{i}}(a_{i}),\bigcup_{i}a_{i}\leq 
-%
-%[ \iota_{j_{1}^{i}}(a_{1}^{i}),\dots, \iota_{j_{s_{i}}^{i}}(a_{s_{i}}^{i})], 
-%\left [\bigcup_{j}a_{j}^{1},\dots, \bigcup_{j}a_{j}^{n}\right] \leq 
-\alpha(n)
-\right \} \\
-&=
- \bigvee\left \{ [\epsilon(\alpha_{1}),\dots, \epsilon(\alpha_{n})]
-\ \Bigg \vert \ 
-\alpha_{i}=\iota_{1}(x_{i}),[x_{1},\dots, x_{n}]\leq 
-\alpha(n)
-\right \} \\
-&= \bigvee\left \{ [
-x_{1},\dots, x_{n}] \ \Bigg \vert  \ 
-[x_{1},\dots, x_{n}]\leq \alpha(n)
-\right \} = \alpha(n)
- \end{align*}
-%But since $\epsilon(\beta)=\beta(1)$, this means that we are reduced to consider multisets $\alpha_{j}^{i} $ of cardinality 1, i.e.~of the form $[a_{j}^{i}]$, so we have  
-% $!\epsilon(\delta(\alpha))(n)=
-%\bigvee\{ [a_{1}^{i},\dots, a_{n}^{i}]\mid \bigvee_{i}[a_{1}^{i},\dots, a_{n}^{i}]= \alpha(n)\}= \alpha(n)$.
-
-Let us now compute $!\delta(\delta (\alpha))$: 
-{\small
-\begin{align*}
-!\delta(\delta(\alpha))(n)&=\bigvee_{i}\left \{[\delta(A_{1}),\dots, \delta(A_{n})]\ \Bigg\vert \   [A_{1},\dots, A_{n}]\leq \delta(\alpha)(n)\right\}\\
-&=\bigvee\left \{[\delta(\iota_{m_{1}}(a_{1})),\dots, \delta(\iota_{m_{n}}(a_{n}))]\ \Bigg\vert \  
-a_{j}\in !_{m_{j}}X, 
-\bigcup_{j=1}^{n}a_{j}\leq \alpha(\sum_{l=1}^{n}m_{l})
-\right\}\\
-&= \bigvee \left\{
-[ B_{1},\dots, B_{n}] \ \Bigg \vert \ 
-B_{i}=\iota_{\sum_{j}r_{j}^{i}}\big([\iota_{r_{1}^{i}}(b_{1}^{i}),\dots, \iota_{r_{s_{i}}^{i}}(b_{m_{i}}^{i})]\big),\sum_{j}r_{j}^{i}=m_{i}, 
-\left [\bigcup_{j}b_{j}^{1},\dots, \bigcup_{j}b_{j}^{n}\right ]\leq \alpha(\sum_{i}m_{i})
-\right \}
-\end{align*}
-}
-
-Let us compute $\delta(\delta(\alpha))$:
-{\small
-\begin{align*}
-\delta(\delta(\alpha))(n) & = 
-\bigvee\left \{
-[\iota_{i_{1}}(A_{1}),\dots, \iota_{i_{n}}(A_{n})] \ \Bigg \vert \ 
-A_{j}\in !_{i_{j}}X, \bigcup_{j}A_{j} \leq \delta(\alpha)(\sum_{j}i_{j})
-\right\}\\
-&=
-\bigvee\left \{
-[\iota_{i_{1}}(A_{1}),\dots, \iota_{i_{n}}(A_{n})] \ \Bigg \vert \ 
-A_{j}= [\iota_{r_{1}^{j}}(b_{1}^{j}),\dots, \iota_{r_{s_{j}}^{j}}(b_{i_{j}}^{j})], 
-\sum_{l} r_{l}^{j} = i_{j}, \left [\bigcup_{l}b_{l}^{1},\dots, \bigcup_{l}b_{l}^{n}\right ]\leq  \alpha(\sum_{j}i_{j})
-\right\}
-\end{align*}
-}
-
-From the two computations it is clear that $!\delta(\delta(\alpha)=\delta(\delta(\alpha))$.
-
-\item the cocommutative comonoid structure $(!X, \Delta, e)$ is  defined as follows:
-\begin{align*}
-\Delta(f)& = \bigvee
-\left\{ \iota_{n}(a)\otimes \iota_{m}(b) \ \Big \vert \  
-n,m\in \BB N, a\cup b\leq f(n+m)
-\right\}\\
-e(f)&= f(0)
-\end{align*}
-
-Let us check the relevant equations:
-
-\begin{align*}
-(\Delta\otimes 1)\Big(\Delta(\alpha)\Big)&=
-(\Delta\otimes 1)\left (\bigvee\left \{\iota_{n}(a)\otimes \iota_{m}(b)\ \Big \vert \ a\cup b \leq \alpha(n+m)\right\}\right)\\
-&= 
-\bigvee\left \{\iota_{n_{1}}(a_{1})\otimes \iota_{n_{2}}(a_{2})\otimes \iota_{m}(b)\ \Big \vert \ a_{1}\cup a_{2}\cup b \leq \alpha(n_{1}+n_{2}+m)\right\}\\
-&= 
-\bigvee\left \{\iota_{n}(a)\otimes \iota_{m_{1}}(b_{1})\otimes \iota_{m_{2}}(b_{2})\ \Big \vert \ a\cup b_{1}\cup b_{2} \leq \alpha(n+m_{1}+m_{2})\right\}
-\\
-&= (1\otimes \Delta)\Big(\Delta(\alpha)\Big)
-\end{align*}
-
-\begin{align*}
-(e\otimes 1)\Big(\Delta(\alpha)\Big) & = 
-( e\otimes 1)\left (\bigvee\left \{\iota_{n}(a)\otimes \iota_{m}(b)\ \Big \vert \ a\cup b \leq \alpha(n+m)\right\}\right)\\
-&= 
-\bigvee\left \{\iota_{m}(b)\ \Big \vert \ b \leq \alpha(m)\right\} =
-\alpha
-\end{align*}
-and one can argue similarly for $(1\otimes e)(\Delta(\alpha))$.
-
-\begin{align*}
-\sigma(\Delta(\alpha)) & = 
-\bigvee\left \{\iota_{m}(b)\otimes \iota_{n}(a)\ \Big \vert \ a\cup b \leq \alpha(n+m)\right\}
-\\
-& = 
-\bigvee\left \{\iota_{n}(a)\otimes \iota_{m}(b)\ \Big \vert \ a\cup b \leq \alpha(n+m)\right\}
-\\&= \Delta(\alpha)
-\end{align*}
-
-Finally, the commutation of $\Delta$ and $\delta$:
-{\tiny
-\begin{align*}
-(\delta\otimes \delta)\Big(\Delta(\alpha)\Big) & =
-(\delta\otimes\delta)\left(\bigvee
-\left\{
-\iota_{n}(a)\otimes \iota_{m}(b) \mid a\cup b \leq \alpha(n+m)
-\right\}\right)\\
-&=
-\bigvee
-\left \{\iota_{k_{1}}(
-[\iota_{i_{1}}(a_{1}),\dots, \iota_{i_{k_{1}}}(a_{k_{1}})])\otimes
-\iota_{k_{2}}(
-[\iota_{j_{1}}(b_{1}),\dots, \iota_{j_{k_{2}}}(b_{k_{2}})]) \ \Bigg \vert \
-\bigcup_{l}a_{l} \cup \bigcup_{l}b_{l} \leq \alpha\left (\sum_{l}i_{l}+\sum_{l}j_{l}\right)
-\right\}
-\\
-&=
-\Delta
-\left(\bigvee\left \{
-\iota_{n}(
-[\iota_{i_{1}}(a_{1}),\dots, \iota_{i_{n}}(a_{n})]) \ \Bigg \vert \ 
-\bigcup_{j}a_{j}\leq \alpha(\sum_{j}i_{j})
-\right\}\right)
-\\
-& = \Delta(\delta(\alpha))
-\end{align*}
-}
-
-\item the commutative monoid structure $(!X,\nabla,u)$ is given by 
-\begin{align*}
-\nabla( f \otimes g) (n)& = \bigvee_{k+h=n}f(k)\cup g(h) \\ 
-u & =  \iota_{0}
-\end{align*}
-Observe that $\nabla(f\otimes g)=\bigvee_{n,k+h=n}\iota_{n}(f(k)\cup g(h))$.
-
-Let us check the relevant equations:
-
-\begin{align*}
-\nabla\left((\nabla\otimes 1)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\otimes \gamma_{k}\right)\right) & =
-\nabla \left(
-\bigvee_{k,n}
-\iota_{n}\left( \bigvee_{u+v=n}\alpha_{k}(u)\cup \beta_{k}(v)
-\right)
-\otimes \gamma_{k}
-\right)\\
-&=
-\bigvee_{k,n}\iota_{n}\left( \bigvee_{u+v+w=n}
-\alpha_{k}(u)\cup \beta_{k}(v)\cup \gamma_{k}(w)\right) \\
-& =
-\nabla \left(
-\bigvee_{n,k}\alpha_{k}\otimes 
-\iota_{n}\left( \bigvee_{v+w=n}\beta_{k}(v)\cup \gamma_{k}(w)\right)
-\right)\\
-&= 
-\nabla\left((1\otimes\nabla)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\otimes \gamma_{k}\right)\right) 
-\end{align*}
-
-\begin{align*}
-\nabla\left(
-(1\otimes u)\left(\bigvee_{k}\alpha_{k}\otimes 0\right)\right) & =
-\nabla\left( \bigvee_{k}\alpha_{k}\otimes\iota_{0}(\epsilon)\right)
-=
-\bigvee_{n,k}\iota_{n}( \alpha_{k}(n) )
- =  \bigvee_{k}\alpha_{k} \otimes 0
-\end{align*}
-and one can argue similarly for $\nabla(u\otimes 1)(\alpha)=\alpha$.
-
-\begin{align*}
-\nabla\left (\sigma\left (\bigvee_{k}\alpha_{k}\otimes \beta_{k}
-\right) \right) & = 
-\nabla \left (
-\bigvee_{k}\beta_{k}\otimes \alpha_{k}
-\right)\\
-&= 
-\bigvee_{k,n, p+q=n} \iota_{n}(\beta_{k}(p)\cup \alpha_{k}(q))
-\\
-&= 
-\bigvee_{k,n, p+q=n}\iota_{n}( \alpha_{k}(p)\cup \beta_{k}(q))
-\\
-&= \nabla\left (\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)
-\end{align*}
-
-Let us check the bialgebra equations:
-
-\begin{align*}
-e\left (\nabla\left (\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)\right)  & =
-e\left(
-\bigvee_{k,n, p+q=n}\iota_{n}(\alpha_{k}(p)\cup \beta_{k}(q))
-\right)
-\\
-&= \bigvee_{k}\alpha_{k}(0)\cup\beta_{k}(0) \\
-&= \bigvee_{k} 0\cup 0\\
-&= h_{Q}\left(\bigvee_{k}\alpha_{k}(0)\otimes\beta_{k}(0))\right) = h\left( (e\otimes e)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)\right)
-\end{align*}
-where $h_{Q}:Q\otimes Q \to Q$ indicates the isomorphism $h(\epsilon\otimes \delta)=\epsilon+\delta$.
-
-\begin{align*}
-\Delta(u(\epsilon))) = \Delta(\iota_{0}(\epsilon)) &=
- \left(
-\bigvee\left\{\iota_{m}(a)\otimes\iota_{n}(b) \ \Bigg \vert \ 
-a\cup b \leq \iota_{0}(\epsilon)(n+m)\right\}\right)\\
-&=\bigvee\left\{\iota_{0}(\epsilon_{1})\otimes\iota_{0}(\epsilon_{2}) \ \Bigg \vert \ 
-\epsilon_{1}+\epsilon_{2} \leq \epsilon\right\}\\
-&= (u\otimes u)(\epsilon)
-\end{align*}
-where we are identifying $\epsilon \in q$ with the equivalence class
-$\epsilon \otimes 0 = \{ \langle \epsilon_{1},\epsilon_{2}\rangle\mid \epsilon_{1}+\epsilon_{2}=\epsilon\}\in 
-Q\otimes Q$ via the isomorphism $h_{Q}$ described above.
-
-\begin{align*}
-e(u(\epsilon))&= \iota_{0}(\epsilon)(0)= \epsilon
-\end{align*}
-
-{\tiny
-\begin{align*}
-&(\nabla\otimes \nabla)  (1\otimes \sigma\otimes 1)(\Delta\otimes \Delta)\left (\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right) \\
-&=
-(\nabla\otimes \nabla)(1\otimes \sigma\otimes 1)
-\left(
-\bigvee
-\left\{
- \iota_{n}(a_{1})\otimes \iota_{m}(a_{2})\mid
-a_{1}\cup a_{2}\leq \alpha_{k}(n+m)
-\right\}\otimes
-\left\{
- \iota_{n}(b_{1})\otimes \iota_{m}(b_{2})\mid
-b_{1}\cup b_{2}\leq \beta_{k}(n+m)
-\right\}
-\right)\\
-&=
-(\nabla\otimes \nabla)(1\otimes \sigma\otimes 1)
-\left(
-\bigvee
-\left\{
- \iota_{n}(a_{1})\otimes \iota_{m}(a_{2})
-\otimes \iota_{n'}(b_{1})\otimes \iota_{m'}(b_{2})
-\mid
-a_{1}\cup a_{2}\leq \alpha_{k}(n+m),
-b_{1}\cup b_{2}\leq \beta_{k}(n'+m')
-\right\}
-\right)\\
-&=
-(\nabla\otimes \nabla)\left(
-\bigvee
-\left\{
- \iota_{n}(a_{1})\otimes \iota_{n'}(b_{1})
-\otimes \iota_{m}(a_{2})\otimes \iota_{m'}(b_{2})
-\mid
-a_{1}\cup a_{2}\leq \alpha_{k}(n+m),
-b_{1}\cup b_{2}\leq \beta_{k}(n'+m')
-\right\}
-\right)\\
-&=
-\bigvee_{k,z,z'}\left\{
-\left(
-\iota_{z}\left( \bigvee_{p+q=z}
-a_{1}\cup b_{1}\right)\right)\otimes
-\left(\iota_{z'}\left( \bigvee_{p'+q'=z'}
-a_{2}\cup b_{2}\right)\right)
-\ \Bigg \vert \
-a_{1}\cup a_{2} \leq \alpha_{k}(p+p'),
-b_{1}\cup b_{2}\leq \beta_{k}(q+q')
-\right\}
-\\
-&=\Delta\left(
-\bigvee_{k,z}
-\left\{
-\iota_{z}\left( \bigvee_{p+q=z} \alpha_{k}(p)\cup \beta_{k}(q) \right)
-\right\}
-\right)
-\\
-&=
-\Delta\left (\nabla\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right) \right)
-\end{align*}
-}
-
-Finally, let us check the compatibility of $\epsilon$ and $\nabla$, which in $Q\mathsf{Mod}$ reads as
-$ \epsilon\circ \nabla=(\epsilon \otimes e)\vee (e\otimes \epsilon)$:
-\begin{align*}
-\epsilon \left(\nabla\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)\right) & =
-\epsilon\left (  \bigvee_{k,n,p+q=n}\iota_{n}(\alpha_{k}(p)\cup \beta_{k}(q))\right) \\
-&=\left(\bigvee_{k}\alpha_{k}(1) \right) \vee \left(
-\bigvee_{k}\beta_{k}(1)\right)
-\\
-&=
-\left( (\epsilon\otimes e)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)
-\right)
-\vee
-\left( (e\otimes \epsilon)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)
-\right)
-\\
-&=
-\Big((\epsilon\otimes e)\vee (e\otimes \epsilon)\Big)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)
- \end{align*}
-
-\item the codereliction $\eta: X\to !X$ is defined by 
-$\eta= \iota_{1}$.
-
-Let us check the codereliction equations:
-\begin{align*}
-e(\eta(x)) & = \iota_{1}(x)(0)=0
-\end{align*}
-\begin{align*}
-\Delta(\eta(x)) & = \Delta(\iota_{1}(x)) \\
-&=
-\bigvee\left\{\iota_{1}(x)\otimes \iota_{0}(r) \mid r\in Q\right\}\vee
-\bigvee\left\{\iota_{0}(r)\otimes \iota_{1}(x) \mid r\in Q\right\}
-\\
-&=
-(\iota_{1}(x)\otimes 0 )\vee
-(0\otimes \iota_{1}(x) )
-\\
-&=
-(\eta\otimes u)(x)\vee(u\otimes \eta) (x) \\
-&=
-\Big(\eta\otimes u)\vee(u\otimes \eta)\Big) (x) 
-\end{align*}
-
-\begin{align*}
-\epsilon(\eta(x))&= \iota_{1}(x)(1) =x
-\end{align*}
-
-For the last equation, we only check it on basic tensors:
-\begin{align*}
-& \nabla((\delta\otimes \eta)((1\otimes \nabla))(\Delta\otimes \eta)(\alpha\otimes x)) \\
-&=\nabla((\delta\otimes \eta)((1\otimes \nabla))\left(
-\bigvee
-\{\iota_{n}(a)\otimes \iota_{m}(b)\otimes \iota_{1}(x)\mid a\cup b \leq \alpha(n+m)\}
-\right)\\ 
-&=\nabla\left ((\delta\otimes \eta)\left(
-\bigvee
-\{\iota_{n}(a)\otimes \iota_{m+1}(b\cup \{x\})\mid a\cup b \leq \alpha(n+m)\}\right)\right)
-\\ 
-&=\nabla\left (
-\bigvee\left 
-\{[\iota_{i_{1}}(c_{1}),\dots,\iota_{i_{r}}(c_{r})]
-\otimes \iota_{1}( \iota_{m+1}(b\cup \{x\}))\ \Bigg \vert \ 
-\bigcup_{j}c_{j}\cup b \leq \alpha\left (\sum_{j}i_{j}+m\right )\right\}\right)
-\\ 
-&=
-\bigvee\left\{
-[\iota_{i_{1}}(c_{1}),\dots,\iota_{i_{r}}(c_{r}), \iota_{m+1}(b\cup \{x\})] \ \Bigg \vert \
-\bigcup_{j}c_{j}\cup b \leq \alpha\left (\sum_{j}i_{j}+m\right )\right\}
-\\
-&=
-\delta\left(
-\bigvee_{n}\iota_{n+1}(
-\alpha(n)\cup \{x\})
-\right)
-\\
-&=
-\delta(\nabla(\alpha\otimes \iota_{1}(x))
-\\
-&=
-\delta(\nabla((1\otimes \eta)(\alpha\otimes x))) 
-\end{align*}
-
-\end{itemize}
-
-It remains to check the Seely isomorphisms.
-Using Theorem \ref{theorem:lemay} it suffices to check that the bialgebra modality defined above is additive (with respect to the ``tropical'' additive structure given by $\bot$ and $\vee$).
-
-{\tiny
-\begin{align*}
-\nabla((!f\otimes !g)(\Delta(\alpha))) & =
-\nabla\left((!f\otimes !g)\left(
-\bigvee\{\iota_{n}([x_{1},\dots, x_{n}])\otimes \iota_{m}([y_{1},\dots, y_{m}])\mid [x_{1},\dots, x_{n},y_{1},\dots y_{m}]\leq \alpha(n+m)\}
-\right)\right)\\
-&=
-\nabla\left(
-\bigvee\{\iota_{n}([f(x_{1}),\dots, f(x_{n})])\otimes \iota_{m}([g(y_{1}),\dots, g(y_{m})])\mid [x_{1},\dots, x_{n},y_{1},\dots y_{m}]\leq \alpha(n+m)\}
-\right)\\
-&=
-\bigvee\{\iota_{n+m}([f(x_{1}),\dots, f(x_{n}),g(y_{1}),\dots, g(y_{m})])\mid [x_{1},\dots, x_{n},y_{1},\dots y_{m}]\leq \alpha(n+m)\}\\
-&=
-\bigvee_{n}\{ \iota_{n}([f(x_{1})\vee g(x_{1}),\dots, f(x_{n})\vee g(x_{1})]
-\mid [x_{1},\dots, x_{n}]\leq \alpha(n)\}\\
-&=
-!(f\vee g)(\alpha)
-\end{align*}
-}
-\begin{align*}
-u(e(\alpha)) = u(\alpha(0))& =\iota_{0}(\alpha({0})) =
-\iota_{0}(\alpha(0))\vee
-\bigvee\{\iota_{n+1}([\underbrace{\bot,\dots, \bot}_{n+1}])\mid n\in \BB N \}=
-(!\bot) (\alpha)
-\end{align*}
-
-In particular, any $\alpha\in !(X\times Y)$ can be represented 
-as an object $\alpha^{S}$ of $!X\otimes !Y$ defined as follows:
-$$
-\alpha^{S}=\bigvee \{ \iota_{n}([x_{1},\dots, x_{n}])\otimes \iota_{m}([y_{1},\dots, y_{m}]) \mid
-\iota_{n+m}(\langle x_{1},\bot\rangle,\dots, 
-\langle x_{n},\bot\rangle,
-\langle \bot, y_{1}\rangle, \dots,
-\langle \bot, y_{m}\rangle) \leq \alpha(n+m)
-\}
-$$
-
-\begin{theorem}
-$Q\mathsf{Mod}$ (equivalently, $Q\mathsf{CCat}$) is a linear differential category. Hence $Q\mathsf{Mod}_{!}$ (equivalently, $Q\mathsf{CCat}_{!}$) is a cartesian closed differential category. 
-\end{theorem}
-
-Let us compute the differential operator in $Q\mathsf{Mod}_{!}$: 
-given $f:!X \to Y$, we have
-\begin{align*}
-\mathsf{D}[f](\alpha) & = 
-\bigvee 
-\left \{
-f(\beta\cup[x])
-\ \Big \vert \ 
-\iota_{n}(\beta)\otimes \iota_{1}(x) \leq \alpha^{S}
-\right\}
-\end{align*}
-
-%\begin{definition}
-%A coalgebra modality $(!,\delta,\epsilon,\Delta,e)$ is a \emph{monoidal coalgebra modality} if there exist natural transformations $m_{\otimes}:!X\otimes !Y \to !(X\otimes Y)$ 
-%and $m_{1}: \B 1 \to !\B 1$ such that 
-%\begin{itemize}
-%\item $m_{\otimes},m_{k}$ show $(!,\delta,\epsilon)$ as a symmetric monoidal comonad, that is, the following equations hold:
-%
-%
-%\item the following further equations hold:
-%\begin{align}
-%\Delta \circ m_{\otimes} & = (m_{\otimes}\otimes m_{\otimes})\circ (1\otimes \sigma \otimes 1) \circ (\Delta \otimes \Delta) \\
-%
-%
-%\end{align}
-%
-%\end{itemize}
-%
-%\end{definition}
-
-%
-%the morphism $\chi= (!\pi_{0}\otimes !\pi_{1})\circ \Delta$ is given by 
-%%\begin{align*}
-%%\chi(\alpha)&=(!\pi_{0}\otimes !\pi_{1})\left(
-%\bigvee\{
-%\iota_{n}(a)\otimes \iota_{m}(b) \mid
-%a\cup b \leq \alpha(n+m)
-%\}\right)\\
-%&=
-%\bigvee\{
-%\iota_{n}(a)\otimes \iota_{m}(b) \mid
-%a\cup b \leq \alpha(n+m)
-%\}\right)\\
-%\end{align*}
-
-%$m_{\otimes}(\alpha\otimes \beta)(n)= \alpha(n)\otimes \beta(n)$
-%
-%
-%
-%$m_{Q}(\epsilon)=\iota_{1}(\epsilon)$.
-%
-%
-%
-
-\newpage
-
-\subsection{$Q$-Rel is a model of Bounded Linear Logic}
-
-Given SMCs $\BB C,\BB D$,  let $\B{SMC}_{l}(\BB C, \BB D)$ indicate the category of symmetric lax monoidal functors and monoidal natural transformations between them.
-$\B{SMC}_{l}(\BB C, \BB D)$  is itself a SMC, with monoidal structure defined pointwise.
+Given SMCs $\C C,\C D$,  let $\B{SMC}_{l}(\C C, \C D)$ indicate the category of symmetric lax monoidal functors and monoidal natural transformations between them.
+$\B{SMC}_{l}(\C C, \C D)$  is itself a SMC, with monoidal structure defined pointwise.
 
@@
 
-%Below we let $\BB^{\min}$ indicate the monoid $(Q,\min,0)$ and $Q^{+}$ indicate the monoid $(Q,+,0)$.
+%Below we let $\BB^{\min}$ indicate the monoid $(Q,\min,0)$ and $\Lawv^{+}$ indicate the monoid $(Q,+,0)$.
 
-\begin{definition}
-A \emph{$\BB N$-graded linear exponential comonad} on a symmetric monoidal category $\BB C$ is a tuple
+\begin{definition}[cf.~\cite{Katsumata2018}]
+A \emph{$\BB N$-graded linear exponential comonad} on a symmetric monoidal category $\C C$ is a tuple
 $(D, w,c,\epsilon,\delta)$ where:
@@
 
-\item $D: \BB N\to \B{SMC}_{l}(\BB C, \BB C)$ is a functor. We write 
-$m_{r}:\B 1 \to D(r)(\B 1)$ and $m_{r,A,B}: D(r)(A)\otimes D(r)(B) \to D(r)(A\otimes B)$ for the symmetric lax monoidal structure of $D(r)$;
+\item $D: \BB N\to \B{SMC}_{l}(\C C, \C C)$ is a functor. We write 
+$m_{r}:\{\star\} \to D(r)(\{\star\})$ and $m_{r,A,B}: D(r)(A)\otimes D(r)(B) \to D(r)(A\otimes B)$ for the symmetric lax monoidal structure of $D(r)$;
 
-\item $(D,w,c): \BB N^{+}\to \B{SMC}_{l}(\BB C, \BB C)$ is a symmetric colax monoidal functor;
+\item $(D,w,c): \BB N^{+}\to \B{SMC}_{l}(\C C, \C C)$ is a symmetric colax monoidal functor;
 
@@
 \begin{align*}
-m_{r} & :\B 1\to  D(r)(\B 1)\\
+m_{r} & :\{\star\}\to  D(r)(\{\star\})\\
 m_{r,A,B}& :  D(r)(A)\otimes D(r)(B)\to  D(r)(A\otimes B)\\
-w_{A}& : D(0)(A)\to \B 1 \\
+w_{A}& : D(0)(A)\to \{\star\} \\
 c_{r,r',A}& : D(r+r')(A) \to D(r)(A)\otimes D(r')(A) \\
@@
 m_{r,A, B\otimes C}\circ (D(r)(A)\otimes m_{r,B,C})\\
-m_{r,A,\B 1}\circ (D(r)(A)\otimes m_{r}) & = D(r)(A)\\
-m_{r,\B 1, B}\circ (m_{r}\otimes D(r)(B))&= D(r)(B)
+m_{r,A,\{\star\}}\circ (D(r)(A)\otimes m_{r}) & = D(r)(A)\\
+m_{r,\{\star\}, B}\circ (m_{r}\otimes D(r)(B))&= D(r)(B)
 \end{align}
@@
 
-\item $w_{A}:D(0)(A)\times \B 1\to Q$ is given by $w_{A}(\emptyset, \star)=0$ and is $\infty$ otherwise (observe that $D(0)(A)\simeq \B 1$);
+\item $w_{A}:D(0)(A)\times \{\star\}\to Q$ is given by $w_{A}(\emptyset, \star)=0$ and is $\infty$ otherwise (observe that $D(0)(A)\simeq \{\star\}$);
 
@@
  Furthermore, we have that 
- $m_{r,A,\B 1}\circ (D(r)(A)\times m_{r})(\langle \alpha,\beta\rangle): D(r)(A)\times \B 1 \times D(r)(A)$ is equal to $0$ 
+ $m_{r,A,\{\star\}}\circ (D(r)(A)\times m_{r})(\langle \alpha,\beta\rangle): D(r)(A)\times \{\star\} \times D(r)(A)$ is equal to $0$ 
  precisely when $\alpha=\beta$ and is $\infty$ otherwise, that is, it coincides with $\mathrm{id}_{D(r)(A)}$.
@@
 
-\item $(w_{D(s)(A)}\circ \delta_{0,s,A})(\langle \emptyset,\star\rangle: D(0)(A)\times \B 1\to Q$ is 0, precisely like $w_{A}$.
+\item $(w_{D(s)(A)}\circ \delta_{0,s,A})(\langle \emptyset,\star\rangle: D(0)(A)\times \{\star\}\to Q$ is 0, precisely like $w_{A}$.
 

### 2023-01-19 appVII.tex
commit

@@
-In the following we let $Q$ indicate the Lawvere quantale $([0,\infty], \geq, +,0)$. 
-$Q$ can be seen as:
-\begin{itemize}
-\item a category $Q_{0}$ with a unique object and morphisms being the elements of $Q$, with identity 0 and composition $+$;
-
-\item a degenerate category $Q_{1}$ with objects the elements of $Q$, and  
-a morphism between $x$ and $y$ whenever $x\geq y$;
-
-\item a 2-category $Q_{2}$ with 0-cells and 1-cells as in $Q_{0}$, and 2-cells the morphisms of $Q_{1}$.
-
-\end{itemize}
-These categorical formulations of $Q$ will help us here and there.
-
-We let $x-y$ indicate truncated subtraction. Notice that 
-$x \leq (x-y)+y$ and $(x+y)-y\leq x$. This corresponds to the fact that ``$\_-y$'', as a functor from $
-Q_{1}$ to itself, 
- is \emph{right-adjoint} to the functor ``$\_+y$''. Equivalently, 
- $x+y \geq z $ iff $x\geq z-y$.
-
-$Q\mathsf{Rel}$ is the category of sets and $Q$-relations, where a $Q$-relation 
-$R:X\pfun Y$ is a function $R: X\times Y\to Q$, with identity $1_{X}:X\pfun X$ given by 
-$$1_{X}(x,y)=\begin{cases} 0 & \text{ if }x=y\\ \infty & \text{ otherwise}\end{cases}$$
-and composition of $R:X\pfun Y$ and $S:Y\pfun Z$ given by 
-$$
-(S\cdot R)(x,z)= \inf_{y\in Y}R(x,y)+S(y,z)
-$$
-
-\begin{remark}
-$Q\mathsf{Rel}$ is also a (degenerate) 2-category, where for $R,S:X\pfun Y$, $R\To S$ iff for all $x\in X$, $y\in Y$, $R(x,y)\geq S(x,y)$.
-
-\end{remark}
-
-Composition with $R:X\pfun Y$ in Q$\mathsf{Rel}$ yields a functor
-$$
-R\cdot \_ : Q\mathsf{Rel}(A,X)   \longrightarrow Q\mathsf{Rel}(A,Y)
-$$
-which
-has a right-adjoint
-$$
-\_ \multimapinv R : Q\mathsf{Rel}(A,Y)   \longrightarrow Q\mathsf{Rel}(A,X)
-$$
-given, for $S:A\pfun Y$ by 
-\begin{align*}
-(S\multimapinv R)(a,x)= \sup_{y\in Y}S(a,y) -R(x,y)
-%(R\multimap S)(a,x)= \sup_{y\in Y}
-\end{align*}
-
-%
-%\begin{lemma}
-%\begin{itemize}
-%\item[i.] $ (R\multimapinv T)\cdot
-%(T\multimapinv S)\leq R \multimapinv S $.
-%
-%\item[ii.] $R\multimapinv T \leq (R\multimapinv S)\multimapinv (T\multimapinv S)$.
-%\end{itemize}
-%\end{lemma}
-%\begin{proof}
-%\begin{align*}
-%(R\multimapinv S)(x,y)&= \sup_{z}R(x,z)-S(y,z) \\
-%&=\sup_{z} (R(x,z)-T(z,z))+(T(z,z)-S(y,z))\\
-%&= \inf_{w}\left(\sup_{z}R(x,z)-T(w,z)\right) +\left(\sup_{z'}T(w,z')-S(z',y)\right)
-%\\
-%&= \inf_{w}(R\multimapinv T)(x,w)+(T\multimapinv S)(w,y)  
-%\\
-%&= (R\multimapinv T)\cdot (T\multimapinv S)(x,y)
-%\end{align*}
-%
-%\end{proof}
-
-\subsubsection{$Q$-categories.}
-
-A \emph{$Q$-category}, or \emph{generalized (quasi-)metric space} is a small $Q$-enriched category.
-Concretely, it is given by a pair $(X,q)$ made of a set $X$ and a $Q$-relation $q: X\pfun X$ satisfying:
-\begin{align}
-q(x,x) & = 0 \tag{Q1} \\
-q(x,z)+q(z,y) & \geq q(x,y) \tag{Q2}
-\end{align}
-A $Q$-category is \emph{skeletal} if it further satisfies
-\begin{align}
-q(x,y) & = 0 \ \To \ x=y \tag{Q3}
- \end{align}
- and \emph{symmetric} if it satisfies
- \begin{align}
-q(x,y) & = q(y,x) \tag{Q4}
- \end{align}
-For simplicity, in the following we indicate the distance function of a $Q$-category $X$ simply as $X(x,y)$.
-
-\begin{remark}
-In more abstract terms, a $Q$-category can be identified with a \emph{monad} in $Q\mathsf{Rel}$, seen as a 2-category: it is given by a 1-cell $q:X\pfun X$ together with 2-cells $1_{X}\To  q$ and $q\cdot q \To q$.
-\end{remark}
-
-We let $x\simeq y$ indicate the equivalence relation defined by  $X(x,y)=X(x',y)$ holds for all $y\in X$. 
-Notice that $x\simeq y$ coincides with $x=y$ precisely when $X$ is skeletal.
-
-Every $Q$-category $X$ is also a preorder $(X,\preceq_{X})$, with $x\preceq_{X}y$ iff $X(x,y)=0$. Moreover, if $X$ is skeletal, then $\preceq_{X}$ is an order.
-
-For every $Q$-category $X$, the category $X^{\mathrm{op}}$ has same objects as $X$, and $X^{\mathrm{op}}(x,y)=X(y,x)$.
-
-\begin{example}
-$Q$ is a $Q$-category with the distance $Q(x,y)=y-x$. 
-The category $Q^{\mathrm{op}}$ has distance $Q^{\mathrm{op}}(x,y)=Q(y,x)$.
-\end{example}
-
-\begin{example}
-For every $Q$-category $X$, the distance $\C S(X)(x,y)=\max\{X(x,y),X^{\mathrm{op}}(x,y)\}$ defines a symmetric $Q$-category $\C S(X)$. In the case of $Q$, $\C S(Q)$ is just the Euclidean distance.
-\end{example}
-
-A \emph{functor} between $Q$-categories $X$ and $Y$ is just a non-expansive function $f:X\to Y$, that is, satisfying $Y(f(x),f(y))\leq X(x,y)$.
-
 %
@@
 
-%Indeed, being a functor means that any arrow $X(x,y)$ is mapped into an arrow $Y(f(x),f(y))$, that is, $Y(f(x),f(y))
-
-\begin{example}
-For every $Q$-category $X$, the $Q$-category $[X,Q]$ has objects all functors from $X$ to $Q$ (where the latter is seen as a $Q$-category), and $[X,Q](f,g)=\sup_{x\in X}Q(f(x),g(x))$.
-
-When $X$ has the discrete metric, $[X,Q]=Q^{X}$ and $\C S([X,Q])(f,g)=\| f-g\|_{\infty}$.
-
-\end{example}
-
-\begin{example}
-For every $Q$-categories $X,Y$ we can define:
-\begin{itemize}
-\item a $Q$-category $X\otimes Y$ with 
-$(X\otimes Y)(\langle x,y\rangle, \langle x',y'\rangle)=X(x,x')+Y(y,y')$;
-\item a $Q$-category $X\times Y$ with 
-$(X\times Y)(\langle x,y\rangle, \langle x',y'\rangle)=\max\{X(x,x'),Y(y,y')\}$. 
-\end{itemize}
-\end{example}
-
-\subsubsection{$Q$-distributors.}
-
-%We let $Q\mathsf{Cont}$ indicate the category of $Q$-categories and continuous functors.
-
-A \emph{$Q$-distributor} $\Phi: X\pfun Y$ is a $Q$-relation $\Phi: X\times Y\to Q$ satisfying
-\begin{align*}
-\Phi\cdot X   \leq\Phi \qquad \qquad
-Y \cdot \Phi  \leq \Phi
-\end{align*}
-
-\begin{remark}
-Using Remark \ref{rem:functor}, it follows that $\Phi$ is a $Q$-distributor precisely when it is a functor $\Phi: X^{\mathrm{op}}\otimes Y \to Q$:
-the condition $\Phi(x,y) \leq X(x,x')+\Phi(x',y)$ implies that $\Phi$ is a functor over $X^{\mathrm{op}}$, while
-the condition $\Phi(x,y) \leq \Phi(x,x')+Y(x',y)$ implies that $\Phi$ is a functor over $Y$.
-But this implies then 
-$\Phi(x,y) \leq X(x,x') +\Phi(x',y')+Y(y',y)$ and thus $Q(\Phi(x',y'),\Phi(x,y))\leq X(x,x')+Y(y',y)$.
-\end{remark}
-
-%where $(\Phi\cdot X)_{x,y}= \inf_{z\in X}\Phi_{x,z}+X(z,y)$ and 
-%$(y\cdot \Phi)_{x,y}= \inf_{z\in X}Y(x,z)+\Phi_{z,y}$.
-
-The identity distributor $1_{X}:X\pfun X$ is given by $(1_{X})_{x,y}=X(x,y)$. 
-Distributors $\Phi:X\pfun Y$ and $\Psi:Y\pfun Z$ compose via $\Psi\cdot \Phi:X\pfun Z$. 
-We let $Q\mathsf{Dist}$ indicate the category of $Q$-categories and $Q$-distributors.
-
-\begin{remark}
-Usually distributors $\Phi: X\pfun Y$ are presented as $Q$-relations $\Phi:Y\times X\to Q$ (notice the inversion of $X$ and $Y$), that is, as functors $\Phi: X\times Y^{\mathrm{op}}\to Q$.
-We chose to invert the presentation of distributors for uniformity with the usual notations in models of the differential $\lambda$-calculus.
-\end{remark}
-
-Any functor $f:X\to Y$ induces two adjoint distributors $f^{\circ}:X\pfun Y$ and $f_{\circ}:Y\pfun X$ given by $f^{\circ}(x,y)=Y(f(x),y)$ and $f_{\circ}(y,x)=Y(f(y),x)$.
-Here adjoint means that $1_{X}\leq f_{\circ}\cdot f^{\circ} $ and $f^{\circ}\cdot f_{\circ}\geq 1_{Y}$.
-
 \begin{remark}[Yoneda embedding]
-%$Q$ is a $Q$-category with $Q(x,y)=|x-y|$. 
-%For any $Q$-category $X$ we can define a new $Q$-category $\mathsf{Dist}(X,Q)$ (that we note simply as $[X,Q]$) whose objects are the distributors $\B x:\B 1\pfun X$, or equivalently, the functors from $X$ to $Q$, i.e.~those $\B x\in Q^{X}$ such that $|\B x_{a}-\B x_{b}|\leq X(a,b)$, or equivalently $\B x_{a}+X(a,b) \geq \B x_{b}$.
-
-The \emph{Yoneda embedding} is the faithful functor $\C Y: X\to [X,Q]$ given by $\C Y(x)(y)=X(y,x)$. The functoriality and faithfulness of $\C Y$ follow from
+%$\Lawv$ is a $\Lawv$-category with $Q(x,y)=|x-y|$. 
+%For any $\Lawv$-category $X$ we can define a new $\Lawv$-category $\mathsf{Dist}(X,Q)$ (that we note simply as $[X,Q]$) whose objects are the distributors $x:\{\star\}\pfun X$, or equivalently, the functors from $X$ to $\Lawv$, i.e.~those $x\in \Lawv^{X}$ such that $|x_{a}-x_{b}|\leq X(a,b)$, or equivalently $x_{a}+X(a,b) \geq x_{b}$.
+
+The \emph{Yoneda embedding} is the faithful functor $\B Y: X\to [X,Q]$ given by $\B Y(x)(y)=X(y,x)$. The functoriality and faithfulness of $\B Y$ follow from
 \begin{align}
-[X,Q](\C Y(x),\C Y(x'))= X(x,x') \tag{Yoneda}
+[X,Q](\B Y(x),\B Y(x'))= X(x,x') \tag{Yoneda}
 \end{align}
@@
 \begin{align*}
-Q( \C Y(x)(y),\C Y(x')(y))&= \C Y(x')(y)-\C Y(x)(y) \\
+Q( \B Y(x)(y),\B Y(x')(y))&= \B Y(x')(y)-\B Y(x)(y) \\
 &= X(y,x')-X(y,x) \leq X(x,x')
@@
 where the last step follows from $X(y,x)+X(x,x')\geq X(y,x')$. 
-From this we deduce that $[X,Q](\C Y(x),\C Y(x'))=\sup_{y\in X}Q( \C Y(x)(y),\C Y(x')(y))\leq X(x,x')$. 
+From this we deduce that $[X,Q](\B Y(x),\B Y(x'))=\sup_{y\in X}Q( \B Y(x)(y),\B Y(x')(y))\leq X(x,x')$. 
 For the converse direction, we have  
@@
 X(x,x')- X(x,x)
-\\& =\C Y(x')(x)-\C Y(x)(x)= Q(\C Y(x)(x),\C Y(x')(x))\\
-&\leq [X,Q](\C Y(x), \C Y(x'))
+\\& =\B Y(x')(x)-\B Y(x)(x)= Q(\B Y(x)(x),\B Y(x')(x))\\
+&\leq [X,Q](\B Y(x), \B Y(x'))
 \end{align*}
@@
 The \emph{opposite Yoneda embedding} is the faithful functor
-$\C Y^{\mathrm{op}}: X\to [X,Q]^{\mathrm{op}}$ given by $\C Y^{\mathrm{op}}(x)(y)=X(x,y)$. The functoriality and faithfulness of $\C Y^{\mathrm{op}}$ follow from
+$\B Y^{\mathrm{op}}: X\to [X,Q]^{\mathrm{op}}$ given by $\B Y^{\mathrm{op}}(x)(y)=X(x,y)$. The functoriality and faithfulness of $\B Y^{\mathrm{op}}$ follow from
 \begin{align}\label{eq:yonedaop}
-[X,Q](\C Y^{\mathrm{op}}(x),\C Y^{\mathrm{op}}(x'))= X(x',x) \tag{Yoneda$^{\mathrm{op}}$}
+[X,Q](\B Y^{\mathrm{op}}(x),\B Y^{\mathrm{op}}(x'))= X(x',x) \tag{Yoneda$^{\mathrm{op}}$}
 \end{align}
-which is proved similarly to the case of $\C Y$.
+which is proved similarly to the case of $\B Y$.
 \end{remark}
 
-\subsubsection{cocompleteness}
-
-Let $\Phi:Y\pfun Z$ be a distributor and $f:Y\to X$ be a functor.
-A functor $ g:Z\to X$ is called the \emph{$\Phi$-weighted colimit of $f$} if it satisfies, for all
-$z\in Z$ and 
- $x\in X$, 
-$$
-X(g(z), x) \ = \  \sup_{y\in Y}X(f(y),x) - \Phi(y,z)
-$$
-If this colimit exists, we write it as $\mathrm{colim}(\Phi,f)$.
-
-A $Q$-category $X$ is said \emph{cocomplete} if it admits all weighted colimits, and a functor $f:X\to Y$ of $Q$-categories is said \emph{cocontinuous} if it commutes with all weighted colimits, meaning that $f \circ \mathrm{colim}(\Phi,g)= \mathrm{colim}(\Phi, f\circ g)$.
-
-We let $Q\mathsf{CCat}$ indicate the category of skeletal and cocomplete $Q$-categories and cocontinuous functors.
+\subsubsection{completeness}
+
+Composition with $r:X\pfun Y$ yields a functor
+$$
+r\cdot \_ : \LREL(A,X)   \longrightarrow \LREL(A,Y)
+$$
+which
+has a right-adjoint
+$$
+\_ \multimapinv r : \LREL(A,Y)   \longrightarrow \LREL(A,X)
+$$
+given, for $s:A\pfun Y$ by 
+\begin{align*}
+(s\multimapinv r)(a,x)= \sup_{y\in Y}s(a,y) \dotdiv r(x,y)
+%(R\multimap S)(a,x)= \sup_{y\in Y}
+\end{align*}
+
+%
+%
+%Let $\Phi:Y\pfun Z$ be a distributor and $f:Y\to X$ be a functor.
+%A functor $ g:Z\to X$ is called the \emph{$\Phi$-weighted colimit of $f$} if it satisfies, for all
+%$z\in Z$ and 
+% $x\in X$, 
+%$$
+%X(g(z), x) \ = \  \sup_{y\in Y}X(f(y),x) - \Phi(y,z)
+%$$
+%If this colimit exists, we write it as $\mathrm{colim}(\Phi,f)$.
+%
+%
+%A $\Lawv$-category $X$ is said \emph{complete} if it admits all weighted colimits, and a functor $f:X\to Y$ of $\Lawv$-categories is said \emph{continuous} if it commutes with all weighted colimits, meaning that $f \circ \mathrm{colim}(\Phi,g)= \mathrm{colim}(\Phi, f\circ g)$.
+%
+%
+%
+%We let $\GMet$ indicate the category of skeletal and complete $\Lawv$-categories and continuous functors. 
+%
+%
+
+For a $\Lawv$-category $X$, we write $x\simeq y$ when $X(x,y)=X(y,x)=0$. $\simeq$ coincides with $=$ precisely when $x$ is skeletal.
 
 \begin{proposition}\label{prop:yonedasup}
-Let $X$ be a $Q$-category. Then $X$ is cocomplete iff the Yoneda embedding has a left-adjoint.
+Let $X$ be a skeletal $\Lawv$-category. Then $X$ is complete iff the Yoneda embedding has a left-adjoint.
 \end{proposition}
 \begin{proof}
- For all $\B x\in [X,Q]$, let $\sup\B x$ be defined as a weighted colimit via
-$$
-X(\sup\B x, y)= \sup_{z\in X} X(z,y)-\B x_{z}
-$$
-that is, $\sup \B x= \mathrm{colim}( \B x, \mathrm{id}_{X})$, where $\B x$ is seen as a distributor $\B x:\B 1\pfun X$.
+ For all $ x\in [X,Q]$, let $\sup x$ be defined as a weighted colimit via
+$$
+X(\sup x, b)= \sup_{a\in X} X(a,b)-x_{a}
+$$
+that is, $\sup  x= \mathrm{colim}(  x, \mathrm{id}_{X})$, where $ x$ is seen as a distributor $x:\{\star\}\pfun X$.
 
@@
 \begin{align}\label{eq:in1}
-(X\multimapinv\B y)\cdot (\B y\multimapinv\B x)\geq (X\multimapinv\B x)\end{align}
+(X\multimapinv y)\cdot ( y\multimapinv x)\geq (X\multimapinv x)\end{align}
 as follows:
 \begin{align*}
-\left((X\multimapinv \B y)\cdot (\B y\multimapinv\B x)\right)(a)&=
- \left (\sup_{b}X(b,a)-\B y_{b}\right)+\left(\sup_{b}\B y_{b}-\B x_{b}\right)\\
+\left((X\multimapinv  y)\cdot ( y\multimapinv x)\right)(a)&=
+ \left (\sup_{b}X(b,a)- y_{b}\right)+\left(\sup_{b} y_{b}-x_{b}\right)\\
 &\geq
-\sup_{b}(X(b,a)-\B y_{b})+(\B y_{b}-\B x_{b})
-\\
-&=\sup_{b}(X(b,a)-\B y_{b}+\B y_{b})-\B x_{b}
-\\
-&=\sup_{b} X(b,a)-\B x_{a}\\
-&= (X\multimapinv \B x)(a)
+\sup_{b}(X(b,a)- y_{b})+( y_{b}-x_{b})
+\\
+&=\sup_{b}(X(b,a)- y_{b}+ y_{b})-x_{b}
+\\
+&=\sup_{b} X(b,a)-x_{a}\\
+&= (X\multimapinv x)(a)
 \end{align*}
@@
 \begin{align}\label{eq:in2}
-\B y\multimapinv \B x \geq (X\multimapinv \B x)\multimapinv(X\multimapinv\B y)
+ y\multimapinv x \geq (X\multimapinv x)\multimapinv(X\multimapinv y)
 \end{align}
@@
 \begin{align*}
-[X,Q](\B x, \B y)&= \sup_{a\in X}\B y_{a}-\B x_{a}\\
+[X,Q](x,  y)&= \sup_{a\in X} y_{a}-x_{a}\\
  &\stackrel{\tiny\eqref{eq:in2}}{\geq}
- \sup_{a\in X} \left( \sup_{b\in X}X(b,a)-\B x_{a}\right) -\left(  \sup_{b\in X}X(b,a)-\B y_{a}   \right )\\
- &= \sup_{a\in X}X(\sup \B x, a)-X(\sup \B y, a)\\
- &= [X,Q](\C Y^{\mathrm{op}}(\sup\B y),\C Y^{\mathrm{op}}(\sup \B x))
+ \sup_{a\in X} \left( \sup_{b\in X}X(b,a)-x_{a}\right) -\left(  \sup_{b\in X}X(b,a)- y_{a}   \right )\\
+ &= \sup_{a\in X}X(\sup x, a)-X(\sup  y, a)\\
+ &= [X,Q](\B Y^{\mathrm{op}}(\sup y),\B Y^{\mathrm{op}}(\sup x))
  \tag{\ref{eq:yonedaop}}
  \\
-&=X(\sup \B x, \sup \B y)
-\end{align*}
-
-Then for all $x\in X$, $\sup \C Y(x)\simeq x$. Indeed we have 
-\begin{align*}
-X(\sup \C Y(x), y) & = \sup_{z\in X}X(z,y) -\C Y(x)(z) \\
+&=X(\sup x, \sup  y)
+\end{align*}
+
+Then for all $x\in X$, $\sup \B Y(x)\simeq x$. Indeed we have 
+\begin{align*}
+X(\sup \B Y(x), y) & = \sup_{z\in X}X(z,y) -\B Y(x)(z) \\
 & = \sup_{z\in X}X(z,y)-X(z,x)\\
@@
 
-Moreover, for all $\B x\in [X,Q]$, we have $\C Y(\sup \B x)\geq \B x$:  
-\begin{align*}
-\C Y(\sup \B x)(a) &= X(a,\sup \B x)\\
-&= X(  \sup(\C Y(a)),\sup \B x) \\
-& \geq [X,Q]( \C Y(a), \B x)\\
-&\geq\B x_{a}-\C Y(a)(a) \\
-&=
- \B x_{a}-X(a,a)  = \B x_{a}
+Moreover, for all $x\in [X,Q]$, we have $\B Y(\sup x)\geq x$:  
+\begin{align*}
+\B Y(\sup x)(a) &= X(a,\sup x)\\
+&= X(  \sup(\B Y(a)),\sup x) \\
+& \geq [X,Q]( \B Y(a), x)\\
+&\geq x_{a}-\B Y(a)(a) \\
+&=
+ x_{a}-X(a,a)  = x_{a}
 \end{align*}
 %where in the last step we use the fact that 
-%$[X,Q](  \C Y(a),\B x)= \sup_{b}\B x_{b}-X(a,b) = \B x_{a}$.
-%
-%$\B y\in Q^{X}$, 
-% $\| \C Y(\sup\B x)- \B y\|_{\infty}=\| \C Y(\sup\B x)- \B x\|_{\infty}$. Indeed we have 
+%$[X,Q](  \B Y(a),x)= \sup_{b}x_{b}-X(a,b) = x_{a}$.
+%
+%$ y\in \Lawv^{X}$, 
+% $\| \B Y(\sup x)-  y\|_{\infty}=\| \B Y(\sup x)- x\|_{\infty}$. Indeed we have 
 % \begin{align*}
-% \| \B x_{a}-\C Y(\sup\B x)\|_{\infty}&= \sup_{a\in X}|
-%\B x_{a}-\C Y(\sup\B x)(a)|\\ 
+% \| x_{a}-\B Y(\sup x)\|_{\infty}&= \sup_{a\in X}|
+%x_{a}-\B Y(\sup x)(a)|\\ 
 %&=
 %\sup_{a\in X}|
-%\B x_{a}-X(\sup\B x,a)|\\
+%x_{a}-X(\sup x,a)|\\
 % &=
-%\sup_{a\in X}\inf_{a'\in X}| \B x_{a}+X(a',a) -\B x_{a'}|\\
-%&= \sup_{a\in X}|\B x_{a}+ X(a,a)- \B x_{a}| = 0
+%\sup_{a\in X}\inf_{a'\in X}| x_{a}+X(a',a) -x_{a'}|\\
+%&= \sup_{a\in X}|x_{a}+ X(a,a)- x_{a}| = 0
 % \end{align*}
@@
 
-Conversely, if $\sup$ is well-defined and adjoint to $\C Y$, then 
-given $\Phi: Y\pfun \B 1$ and $f:Y\to X$, 
+Conversely, if $\sup$ is well-defined and adjoint to $\B Y$, then 
+given $\Phi: Y\pfun \{\star\}$ and $f:Y\to X$, 
 we can define
 $\mathrm{colim}(\Phi, f):= \sup\Psi$, where $\Psi=
- f^{\circ}\cdot \Phi: X\pfun \B 1$, since 
+ f^{\circ}\cdot \Phi: X\pfun \{\star\}$, since 
 \begin{align*}
@@
 \begin{definition}[MacNeill Completion]
-Let $X$ be a $Q$-category. For all $f: \B 1\pfun X$ and $g:X\pfun \B 1$, let $f \coh g$ iff 
+Let $X$ be a $\Lawv$-category. For all $f: \{\star\}\pfun X$ and $g:X\pfun \{\star\}$, let $f \coh g$ iff 
 $f  = X\multimapinv g $ and 
 $g = f\multimap X$. 
-
-The \emph{MacNeill completion of $X$} is the $Q$-category $\B M(X)$ made of those 
-$f:\B 1\pfun X$ such that $f\coh g$ for some $g:X\pfun \B 1$, with
+The \emph{MacNeill completion of $X$} is the $\Lawv$-category $\B M(X)$ made of those 
+$f:\{\star\}\pfun X$ such that $f\coh g$ for some $g:X\pfun \{\star\}$, with
 $\B M(X)(f,f')=[X,Q](f,f')$. 
@@
 \begin{proposition}
-Let $X$ be a $Q$-category. 
-If $X$ is cocomplete, then $\C Y$ is an isomorphism between $X$ and $\B M(X)$. 
+Let $X$ be a skeletal $\Lawv$-category. 
+If $X$ is complete, then $\B Y$ is an isomorphism between $X$ and $\B M(X)$. 
 \end{proposition}
 \begin{proof}
-For all $x\in X$, one can check that $\C Y(x)\in \B M(X)$. Indeed, we can check that $\C Y(x) \coh \C Y^{\mathrm{op}}(x)$:
-\begin{align*}
-\C Y(x)(y)  &= X(y,x) \\
-&=  \sup_{z\in X}X(y,z  )- X(x,z)\\
-&= \sup_{z\in X}X(y,z  )- \C Y^{\mathrm{op}}(x)(z)\\
- &= (X\multimapinv \C Y^{\mathrm{op}}(x))(y)
-\end{align*}
-Since $\sup\C Y(x) \simeq x$ holds, it suffices to show that if $\B x \coh \B y$, then 
-$\C Y(\sup \B x)=\B x$: 
-\begin{align*}
-\C Y(\sup \B x)(a) & \ \ = X(a, \sup\B x) \\
-&\ \ = \sup_{b\in X}X(a,b)-X(\sup\B x, b) \\
-&\ \ = \sup_{b\in X}\inf_{c\in X}X(a,b)- X(b,c) + \B x_{c} \\
- &\stackrel{{\tiny\text{(COH)}}}{=} \B x_{a}
-\end{align*}
-\end{proof}
-
-%
-%\begin{remark}[other notions of cocompleteness]
-%The categorical notion of cocompleteness subsumes several other notions of cocompleteness, in the sense of being (strictly) stronger:
+For all $x\in X$, one can check that $\B Y(x)\in \B M(X)$. Indeed, we can check that $\B Y(a) \coh \B Y^{\mathrm{op}}(a)$:
+\begin{align*}
+\B Y(a)(b)  &= X(b,a) \\
+&=  \sup_{c\in X}X(b,c  )- X(a,c)\\
+&= \sup_{c\in X}X(b,c  )- \B Y^{\mathrm{op}}(a)(c)\\
+ &= (X\multimapinv \B Y^{\mathrm{op}}(a))(b)
+\end{align*}
+Since $\sup\B Y(a) \simeq a$ holds, it suffices to show that if $x \coh  y$, then 
+$\B Y(\sup x)=x$: 
+\begin{align*}
+\B Y(\sup x)(a) & \ \ = X(a, \sup x) \\
+&\ \ = \sup_{b\in X}X(a,b)-X(\sup x, b) \\
+&\ \ = \sup_{b\in X}\inf_{c\in X}X(a,b)- X(b,c) + x_{c} \\
+ &\stackrel{{\tiny\text{(COH)}}}{=} x_{a}
+\end{align*}
+\end{proof}
+
+%
+%\begin{remark}[other notions of completeness]
+%The categorical notion of completeness subsumes several other notions of completeness, in the sense of being (strictly) stronger:
 %\begin{itemize}
-%\item \emph{order-completeness} is the case when the pre-order $\preceq_{X}$ is cocomplete.   
-%Let $I\subseteq X$, so that $\mathrm{id}_{I}$ can be seen as a functor from the subcategory $I$ to $X$. Let $0_{I}: \B 1\pfun I$ be the distributor given by $(0_{I})_{i}=0$.
+%\item \emph{order-completeness} is the case when the pre-order $\preceq_{X}$ is complete.   
+%Let $I\subseteq X$, so that $\mathrm{id}_{I}$ can be seen as a functor from the subcategory $I$ to $X$. Let $0_{I}: \{\star\}\pfun I$ be the distributor given by $(0_{I})_{i}=0$.
 %Then $\mathrm{colim}(0_{I}, \mathrm{id}_{I})$ coincides with $\inf I$, via 
@@
 %$$
-%Notice that in general an order-complete $Q$-category needs not be tensored, and thus it needs not be cocomplete.
-%
-%
-%
-%
-% 
-%
-%\item \emph{Cauchy-completeness} is the case where $X$ contains the points $\sup \B x, \sup\B y$, for any 
-% any two \emph{adjoint} presheaves $\B x,\B y\in [X,Q]$, i.e.~satisfying $0=\inf_{a}\B x_{a}+\B y_{a}$ and $\B x_{a}+ \B x_{b}\geq X(a,b)$. 
-% Notice that given a pair $(\B x, \B y)$ we can define a Cauchy sequence by finding $a_{n}$ satisfying 
-% $\B x_{a_{n}}+\B y_{a_{n}}\leq \frac{1}{n}$.
-Indeed this implies $X(a_{n},a_{n+1})\leq \B x_{a_{n}}+\B y_{a_{n+1}}\leq \B x_{a_{n}}+\B y_{a_{n}}+\B x_{a_{n+1}}+\B y_{a_{n+1}}=\frac{1}{n}+\frac{1}{n+1}$.
+%Notice that in general an order-complete $\Lawv$-category needs not be tensored, and thus it needs not be complete.
+%
+%
+%
+%
+% 
+%
+%\item \emph{Cauchy-completeness} is the case where $X$ contains the points $\sup x, \sup y$, for any 
+% any two \emph{adjoint} presheaves $x, y\in [X,Q]$, i.e.~satisfying $0=\inf_{a}x_{a}+ y_{a}$ and $x_{a}+ x_{b}\geq X(a,b)$. 
+% Notice that given a pair $(x,  y)$ we can define a Cauchy sequence by finding $a_{n}$ satisfying 
+% $x_{a_{n}}+ y_{a_{n}}\leq \frac{1}{n}$. Indeed this implies $X(a_{n},a_{n+1})\leq x_{a_{n}}+ y_{a_{n+1}}\leq x_{a_{n}}+ y_{a_{n}}+x_{a_{n+1}}+ y_{a_{n+1}}=\frac{1}{n}+\frac{1}{n+1}$.
 % Conversely, any (equivalence class of) Cauchy sequences $a_{n}$ yields an adjoint pair given by
-% $\B x_{a}= \lim_{n}X(a_{n},a)$ and $\B y_{a}=\lim_{n}X(a,a_{n})$. 
+% $x_{a}= \lim_{n}X(a_{n},a)$ and $ y_{a}=\lim_{n}X(a,a_{n})$. 
 % 
 %% 
-%% \item \emph{Isbell-completeness} (or \emph{MacNeill cocompleteness}) is the case where $X$ contains the points $\sup \B x, \sup \B y$, for any two presheaves $\B x, \B y\in [X,Q]$ satisfying 
+%% \item \emph{Isbell-completeness} (or \emph{MacNeill completeness}) is the case where $X$ contains the points $\sup x, \sup  y$, for any two presheaves $x,  y\in [X,Q]$ satisfying 
 %% $
-%% \B x_{a}= \sup_{b\in X}X(b,a)-\B y_{b}
+%% x_{a}= \sup_{b\in X}X(b,a)- y_{b}
 %% $.
@@
 
-\subsubsection{Tensors and $Q$-Modules}
-
-Among weighted colimits, one is of big importance for us. 
-Any $\epsilon \in Q$ generates a constant distributor $(\epsilon):\B 1\pfun \B 1$, and any point $x\in X$ generates a constant functor $\Delta x: \B 1\to X$.
-Given a $Q$-category $X$, a point $x\in X$ and  $\epsilon \in Q$, the \emph{tensor of $x$ and $\epsilon$}, if it exists, is defined as
-$$
-\epsilon \otimes x := \mathrm{colim}((\epsilon), \Delta x)
-$$
-A $Q$-category $X$ is \emph{tensored} if for all $x\in X$ and $\epsilon \in Q$, it admits the tensor $\epsilon \otimes x$.
+From now on, all $\Lawv$-categories will be tacitly assumed to be skeletal. As observed in Section \ref{section6}, any $\Lawv$-category $X$ can be made skeletal by a suitable quotient.
+
+\subsubsection{Tensors and $\Lawv$-Modules}
+
+%Among weighted colimits, one is of big importance for us. 
+%Any $\epsilon \in Q$ generates a constant distributor $(\epsilon):\{\star\}\pfun \{\star\}$, and any point $x\in X$ generates a constant functor $\Delta x: \{\star\}\to X$.
+%Given a $\Lawv$-category $X$, a point $x\in X$ and  $\epsilon \in Q$, the \emph{tensor of $x$ and $\epsilon$}, if it exists, is defined as
+%$$
+%\epsilon \otimes x := \mathrm{colim}((\epsilon), \Delta x)
+%$$
+\begin{definition}
+A $\Lawv$-category $X$ is said \emph{tensored} if for all $x\in X$ and $\epsilon \in Q$, it admits the tensor $\epsilon \otimes x$.
+\end{definition}
 
 \begin{proposition}
-A tensored $Q$-category $X$ is a $Q$-module $(X, \preceq_{X}, \otimes)$.
-A cocontinuous functor of cocomplete $Q$-categories is a $Q$-module morphism between the associated $Q$-modules.
+A tensored $\Lawv$-category $X$ is a $\Lawv$-module $(X, \preceq_{X}, \otimes)$.
+A continuous functor of complete $\Lawv$-categories is a $\Lawv$-module morphism between the associated $\Lawv$-modules.
 \end{proposition}
@@
 
-A cocontinuous functor $f:X\to Y$ commutes with sups and with $\otimes$, and is thus a $Q$-module morphism.
+A continuous functor $f:X\to Y$ commutes with sups and with $\otimes$, and is thus a $\Lawv$-module morphism.
 \end{proof}
@@
 \begin{lemma}
-In any cocomplete $Q$-category, $x\otimes \epsilon \simeq \sup(  \C Y(x)+\epsilon  )$.
-In the cocomplete $Q$-category $[X,Q]$, $\B x\otimes \epsilon= \B x+\epsilon$.
+In any complete $\Lawv$-category, $x\otimes \epsilon \simeq \sup(  \B Y(x)+\epsilon  )$.
+In the complete $\Lawv$-category $[X,Q]$, $x\otimes \epsilon= x+\epsilon$.
 \end{lemma}
@@
 \begin{align*}
-X(\sup(\C Y(x)+\epsilon), x')& =\sup_{y\in X}X(z,x')- (\C Y(x)(z)+\epsilon)\\
+X(\sup(\B Y(x)+\epsilon), x')& =\sup_{y\in X}X(z,x')- (\B Y(x)(z)+\epsilon)\\
 &= \sup_{y\in X}X(z,x')-(X(z,x)+\epsilon)\\
@@
 \end{align*}
-which shows $x\otimes \epsilon=\sup(\C Y(x)+\epsilon)$. In $[X,Q]$ we have 
-$[X,Q](\B x+\epsilon, \B x')=\sup_{a\in X}(\B x_{a}+\epsilon)-\B x'_{a}= (\sup_{a\in X}\B x_{a}-\B x'_{a})-\epsilon= [X,Q](\B x, \B x')-\epsilon$, which shows $\B x\otimes \epsilon \simeq \B x+\epsilon$, and since $[X,Q]$ is skeletal, $\B x\otimes \epsilon=\B x+\epsilon$.
+which shows $x\otimes \epsilon=\sup(\B Y(x)+\epsilon)$. In $[X,Q]$ we have 
+$[X,Q](x+\epsilon, x')=\sup_{a\in X}(x_{a}+\epsilon)-x'_{a}= (\sup_{a\in X}x_{a}-x'_{a})-\epsilon= [X,Q](x, x')-\epsilon$, which shows $x\otimes \epsilon \simeq x+\epsilon$, and since $[X,Q]$ is skeletal, $x\otimes \epsilon=x+\epsilon$.
 \end{proof}
@@
 \end{align*}
-In other words, in a tensored and cotensored $Q$-category we have $X(x\otimes \epsilon,y)= X(x,y\multimapinv \epsilon)$.
+In other words, in a tensored and cotensored $\Lawv$-category we have $X(x\otimes \epsilon,y)= X(x,y\multimapinv \epsilon)$.
 
 \begin{example}
-The $Q$-category $[X,Q]$ is cotensored, with $\B x\multimapinv \epsilon:= \B x-\epsilon$.
-Indeed we have $[X,Q](\B x, \B y\multimapinv \epsilon)=\sup_{a\in X}(\B y_{a}-\epsilon)-\B x_{a}=(\sup_{a\in X}\B y_{a}-\B x_{a})-\epsilon= [X,Q](\B x, \B y)-\epsilon$.
+The $\Lawv$-category $[X,Q]$ is cotensored, with $x\multimapinv \epsilon:= x-\epsilon$.
+Indeed we have $[X,Q](x,  y\multimapinv \epsilon)=\sup_{a\in X}( y_{a}-\epsilon)-x_{a}=(\sup_{a\in X} y_{a}-x_{a})-\epsilon= [X,Q](x,  y)-\epsilon$.
 \end{example}
@@
 \begin{definition}
-A $Q$-category $X$ is \emph{order-complete} if it is a sup-lattice with respect to the order $\preceq_{X}$ (i.e.~all joins exist).
+A $\Lawv$-category $X$ is \emph{order-complete} if it is a sup-lattice with respect to the order $\preceq_{X}$ (i.e.~all joins exist).
 \end{definition}
@@
 \begin{lemma}\label{lemma:supinf}
-Let $X$ be a $Q$-category. If $X$ is order-complete, then 
+Let $X$ be a $\Lawv$-category. If $X$ is order-complete, then 
 \begin{itemize}
@@
 \begin{proposition}\label{prop:tencoten}
-If a $Q$-category $X$ is tensored, cotensored and order-complete, then it is cocomplete.
+If a $\Lawv$-category $X$ is tensored, cotensored and order-complete, then it is complete.
 \end{proposition}
 \begin{proof}
-For all $\B x\in [X,Q]$, let $\sup \B x:= \bigvee_{a\in X}a\otimes\B x_{a}$. 
-Let us check that $X(\sup \B x, b)= \sup_{a\in X}X(a,b)-\B x_{a}$, using Lemma \ref{lemma:supinf}:
-\begin{align*}
-X(\sup \B x, b) &= \sup_{a\in X}X(a\otimes \B x_{a},b)\\
-&=\sup_{a\in X}X(a,b)-\B x_{a}
+For all $x\in [X,Q]$, let $\sup x:= \bigvee_{a\in X}a\otimes x_{a}$. 
+Let us check that $X(\sup x, b)= \sup_{a\in X}X(a,b)-x_{a}$, using Lemma \ref{lemma:supinf}:
+\begin{align*}
+X(\sup x, b) &= \sup_{a\in X}X(a\otimes x_{a},b)\\
+&=\sup_{a\in X}X(a,b)-x_{a}
 \end{align*}
@@
 \begin{proposition}\label{prop:tenfun}
-Let $X,Y$ be two tensored $Q$-categories, and $f:X\to Y$ be a function.
+Let $X,Y$ be two tensored $\Lawv$-categories, and $f:X\to Y$ be a function.
 \begin{itemize}
@@
 
-\item[ii.] $f$ is a cocontinuous functor iff $f$ commutes with joins and for all $x\in X$ and $\epsilon\in Q$, $f(x)\otimes \epsilon = f(x\otimes \epsilon)$.
+\item[ii.] $f$ is a continuous functor iff $f$ commutes with joins and for all $x\in X$ and $\epsilon\in Q$, $f(x)\otimes \epsilon = f(x\otimes \epsilon)$.
 \end{itemize}
@@
 \item[ii.]
-Suppose $f$ is a cocontinuous functors, and let  $g:Y\to X$, be its right-adjoint, i.e.~satisfying $Y(f(x),y)=X(x,g(y))$. Then 
+Suppose $f$ is a continuous functors, and let  $g:Y\to X$, be its right-adjoint, i.e.~satisfying $Y(f(x),y)=X(x,g(y))$. Then 
 \begin{align*}
@@
 
-Conversely, suppose the function $f:X\to Y$ preserves joins and tensors. Since $f$ is order-preserving, by i.~it is a functor, so we must only prove that it is cocontinuous.
+Conversely, suppose the function $f:X\to Y$ preserves joins and tensors. Since $f$ is order-preserving, by i.~it is a functor, so we must only prove that it is continuous.
 Since $f$ preserves joins there exists a function $g:Y\to X$ which is right-adjoint to $f$ with respect to orders, i.e.~$f(x)\preceq_{Y}y$ iff $x\preceq_{X}g(y)$. 
@@
 \begin{theorem}\label{thm:equivalence}
-The category $Q\mathsf{Mod}$ of $Q$-modules and $Q$-module morphism coincides with the category $Q\mathsf{CCat}$ of cocomplete skeletal $Q$-categories and cocontinuous functors.
+The category $\Mod$ of $\Lawv$-modules and $\Lawv$-module morphism coincides with the category $\GMet$ of complete skeletal $\Lawv$-categories and continuous functors.
 \end{theorem}
 \begin{proof}
-We have already seen that any cocomplete skeletal $Q$-category is a $Q$-module via tensors, 
-and that cocontinuous functors are $Q$-module morphisms.
-Let us now show that any $Q$-module is a cocomplete skeletal $Q$-category, and that a $Q$-module morphism is a cocontinuous functor.
-
-Let then $M=(M,\preceq, \star)$ be a $Q$-module. Define $M(x,y)= \inf\{ \delta \mid x\star \delta \succeq y\}$.
+We have already seen that any complete skeletal $\Lawv$-category is a $\Lawv$-module via tensors, 
+and that continuous functors are $\Lawv$-module morphisms.
+Let us now show that any $\Lawv$-module is a complete skeletal $\Lawv$-category, and that a $\Lawv$-module morphism is a continuous functor.
+
+Let then $M=(M,\preceq, \star)$ be a $\Lawv$-module. Define $M(x,y)= \inf\{ \delta \mid x\star \delta \succeq y\}$.
  It is clear that $M(x,x)=0$. Let us prove $M(x,y)+M(y,z) \succeq M(x,z)$: 
@@
 
-Let us check that the $Q$-category $M$ is tensored via $x\otimes \epsilon:=x\star\epsilon$.
+Let us check that the $\Lawv$-category $M$ is tensored via $x\otimes \epsilon:=x\star\epsilon$.
 Let $A_{x,y}= \{ \delta \mid (x\star\epsilon)\star \delta \geq y$ and 
@@
 
-Since $M$, as a $Q$-category, is order-complete, tensored and cotensored, it is cocomplete by Proposition \ref{prop:tencoten}.
-
-To conclude, notice that if $f:X\to Y$ is a cocontinuous functor, then it commutes with tensors and, by 
-Proposition \ref{prop:tenfun} it commutes with joins, so it is a morphism of the respective $Q$-modules.
-Conversely, if $f:M\to N$ is a $Q$-module morphism, then, since $M$ and $N$ are both tensored $Q$-categories, the tensor coincides with the actions of $M$ and $N$, $f$ preserves the joins and the tensor, by Proposition \ref{prop:tenfun}, it is a cocontinuous functor of the respective $Q$-categories.
-\end{proof}
-
-\subsection{$Q\mathsf{Mod}$ is a $*$-Autonomous Category}
-
-Let us first observe that:
-\begin{itemize}
-\item the hom-set $Hom(M,N)$ of two $Q$-modules is a $Q$-module with order and action defined pointwise;
-
-\item for any $Q$-module $M=(M,\preceq, \star)$, there is a $Q$-module
-$M^{\mathrm{op}}=(M,\succeq, \multimapinv)$, with $\multimapinv$ defined as in the proof of Theorem \ref{thm:equivalence}.
-
-\end{itemize}
-
-Let $M,N$ be two $Q$-modules. For all $A\in Q^{M\times N}$, we define the function
-\begin{align*}
-H_{A} & : Q^{M} \longrightarrow Q^{N}%\\
-%K_{A} & : Q^{N} \longrightarrow Q^{M}
-\end{align*}
-via
-\begin{align*}
-H_{A}(\B x)(b)  &= \inf_{a\in M}\B x_{a}+A(a,b)\\
-%K_{A}(\B x)(b)  &= \sup_{a\in M}\B x_{a}-A(a,b)
-\end{align*}
-
-\begin{lemma}
-$H_{A}=H_{A'}$ iff $A=A'$. 
-\end{lemma}
-\begin{proof}
-We only need to prove one direction, so suppose $A\neq A'$ and let $a,b$ be such that $A(a,b)\neq A'(a,b)$.
-Let $\B x$ be defined by $\B x_{a}=1$ and $\B x_{a'}=\infty$ for all $a'\neq a$. Then $H_{A}(\B x)(b)=A(a,b)\neq A'(a,b)=H_{A'}(\B x)(b)$. 
-\end{proof}
-
-\begin{proposition}
-$Q^{M\times N}$ and $Hom(Q^{M},Q^{N})$ are isomorphic $Q$-modules.
-\end{proposition}
-\begin{proof}
-The map $A\mapsto H_{A}$ is injective, as shown above. We need to check that it commutes with joins:
-\begin{align*}
-H_{\bigvee_{i}A_{i}}(\B x)(b) & = \inf_{a}\B x_{a}+\bigvee_{i}A_{i}(a,b)\\
-&=  \inf_{a}\bigvee_{i}\B x_{a}+A_{i}(a,b)\\
-&=  \bigvee_{i}\inf_{a}\B x_{a}+A_{i}(a,b)\\
-&=  \bigvee_{i}H_{A_{i}}(\B x)(b)\end{align*} 
-(recall that $\inf$s are actually joins in $Q$!)
-
-We must prove that $H$ is surjective: for all $f\in Hom(Q^{M},Q^{N})$, let 
-$k_{f}\in Q^{M\times N}$ be given by $k_{f}(a,b)=f(e_{a})(b)$.
-
-Then we have 
-\begin{align*}
-H_{k_{f}}(\B x)(b) & = \inf_{a}\B x_{a}+k_{f}(a,b) \\
-&= \inf_{a}\B x_{a}+ f(e_{a})(b)\\
-&= \left(\inf_{a}\B x_{a}+f(e_{a})\right)(b) \\
-&= \left(\inf_{a}f(\B x_{a}+e_{a})\right)(b) \\
-&= f(\inf_{a}\B x_{a}+e_{a})(b)\\
-&=
-f(\B x)(b)
-\end{align*}
-and conversely 
-\begin{align*}
-k_{H_{A}}(a,b)&= H_{A}(e_{a})(b)= \inf_{a'}(e_{a})_{a'}+A(a',b) =  A(a,b)
-\end{align*}
-\end{proof}
-
-More generally, we have the following result:
-\begin{proposition}
-Let $X,Y$ be two $Q$-modules. For any morphism $f: X\to Y$ there is a matrix $k_{f}\in Q^{X\times Y}$ such that
-
-\end{proposition}
-\begin{proof}
-By composing $f$ with the isomorphisms $\C Y^{\mathrm{op}}:X\to \B M(X)$, with inverse $\sup(\B x)=\bigvee_{a\in x}a\otimes \B x_{a}$, we obtain 
-a morphism $\widehat f: \B M(X)\to \B M(Y)$  
-\begin{align*}
-\widehat f(\B x)(b)&:= \C Y^{\mathrm{op}}(f(\sup \B x))(b) \\
-&= Y( f(\bigvee_{a\in X}a\otimes \B x_{a}),b) \\
-&= Y( \bigvee_{a\in X}f(a)\otimes \B x_{a},b) \\
-&= \sup_{a\in X}Y(f(a),b)-\B x_{a}
-\end{align*}
-Observe that $\widehat f$ can be extended to a function $f^{*}$ from $Q^{X}$ to $Q^{Y}$.
-Now, $ f^{*}$ is generated by the matrix $k_{f}\in Q^{X\times Y}$ given by $k_{ f}(a,b)=f(e_{a})(b)$, so that for all $\B x\in Q^{X}$, 
-$f^{*}(\B x)(b)=\bigvee_{a\in X}k_{f}(a,b)+\B x_{a}$, and thus
-in particular, for all $\B x\in [X,Q]$, 
-$\widehat f(\B x)(b)=f^{*}(\B x)(b)$.
-\end{proof}
-
-\subsubsection{The Tensor Product of $Q$-Modules}
-
-The description of the tensor product of $Q$-modules requires some work. Let us first recall some important definitions:
-
-\begin{definition}[congruence on a sup-lattice]
-Let $(L, \leq)$ be a sup-lattice. An equivalence relation $R\subseteq L\times L$ is said a \emph{congruence} if it satisfies the following property:
-\begin{align}
-(\forall i\in I \ x_{i} Ry_{i})  \ \To  \ \left( \bigvee_{i}x_{i} \right ) R\left(\bigvee_{i}y_{i}\right)
-\tag{congruence}
-\end{align}
-\end{definition}
-
-\begin{lemma}
-For all suplattices $(L,\leq)$, if $R$ is a congruence, then $(L/R, \leq_{R})$ is a sup-lattice, where $[x]\leq_{R}[y]$ iff $(x\vee y) R y$ (i.e.~$[x\vee y]=[y]$), and $\bigvee_{i}[x_{i}]=\left [\bigvee_{i}x_{i}\right]$.
-\end{lemma}
-\begin{proof}
-Let us check that $\leq_{R}$ is an order. It is clear that $[x]\leq_{R}[x]$ holds. If $[x]\leq_{R}[y]$ and $[y]\leq_{R}[z]$ both hold, then 
-$(x\vee y)Ry$ and $(y\vee z)Rz$ hold; 
-then, since $R$ is a congruence $((x\vee y)\vee (y\vee z)) R (y\vee z)R z$, and moreover $x \vee (y\vee z) R (x \vee z)$, whence 
-$(x\vee z) R(x \vee y\vee z)R z$, so $[x]\leq_{R}[z]$.  
-If $[x]\leq_{R}[y]$ and $[y]\leq_{R}[x]$, then 
-$xR(x\vee y)R y$, thus $[x]=[y]$.
-
-Let us now check the definition of joins. 
-From $[x_{i}]\vee [\bigvee_{i}x_{i}]=[x_{i}\vee \bigvee_{i}x_{i}]= [\bigvee_{i}x_{i}]$ we deduce $[x_{i}]\leq_{R}[\bigvee x_{i}]$.
-Suppose now $[x_{i}]\leq [y]$ holds for all $i\in I$, that is, 
-$(x_{i}\vee y)Ry$; then, since $R$ is a congruence, 
-$(\bigvee_{i}(x_{i}\vee y))R y$, that is, 
-$((\bigvee_{i}x_{i})\vee y)Ry$, which implies 
-$[\bigvee_{i}x_{i}]\leq_{R}[y]$. We conclude then that $\bigvee_{i}[x_{i}]=[\bigvee_{i}x_{i}]$.
-\end{proof}
-
-\begin{corollary}\label{cor:bigvee}
-Let $(L,\leq)$ be a suplattice and $R$ be a congruence on $L$.
-Then, for any class $\beta\in L/R$, $\beta= [\bigvee\beta]$.
-\end{corollary}
-\begin{proof}
-$[\bigvee \beta]=[\bigvee\{ x \mid x\in \beta\}]= \bigvee \{[x]\mid x\in \beta\}=\beta$.
-\end{proof}
-
-\begin{proposition}\label{prop:smallestcongruence}
-Let $(L,\leq)$ be a sup-lattice. Let $R\subseteq L\times L$ be an equivalence relation, and for any ordinal $\alpha$, let the relation
-$R^{(\alpha)}\subseteq L\times L$ be defined by:
-\begin{itemize}
-\item $xR^{(0)}y$ iff either $xRy$, $x=y$ or $yRx$ holds;
-\item $xR^{(\alpha+1)}y$ iff one of the following holds:
-	\begin{itemize}
-	\item for some $z$, $xR^{(\alpha)}z$ and $zR^{(\alpha)}y$ holds;
-	\item for some set $I$, and families $x_{i},y_{i}$, 
-	$x=\bigvee x_{i}, y=\bigvee_{i}y_{i}$ and $x_{i} R^{(\alpha)}y_{i}$ holds for all $i\in I$.
-
-	\end{itemize}
-\item $xR^{(\gamma)}y$ iff $xR^{(\delta)}y$ holds for some $\delta <\gamma$, for $\gamma$ limit.
-\end{itemize}
-Then the relation $R^{*}\subseteq L\times L$ given by 
-$$
-xR^{*} y  \ \Leftrightarrow \ \exists \alpha . \mathrm{OR}(\alpha) \land xR^{(\alpha)}y
-$$% defined as follows: $xR^{*} y$ iff
-%\begin{enumerate}
-%\item for any set $I$ and family $x_{i}$ such that $x=\bigvee_{i}x_{i}$, there exists a family $y_{i}$ such that $y=\bigvee_{i}y_{i}$ and $x_{i}R y_{i}$ holds for all $i\in I$;
-%\item for any set $I$ and family $y_{i}$ such that $y=\bigvee_{i}y_{i}$, there exists a family $x_{i}$ such that $x=\bigvee_{i}x_{i}$ and $x_{i}R y_{i}$ holds for all $i\in I$.
-%\end{enumerate}
-%%\begin{align*}
-%%xR^{*}y &\text{ iff }\  \forall I \ \forall x_{i} \  \text{ s.t. }
-%%\left\{
-%%\begin{matrix}
-%%x=\bigvee_{i\in I}x_{i} \\
-%%y=\bigvee_{i\in I}y_{i}\\
-%%x_{i}Ry_{i} \ (\forall i\in I)
-%%\end{matrix}
-%%\right\}
-%%\end{align*}
-is a congruence, and is the smallest congruence containing $R$.
-\end{proposition}
-\begin{proof}
-From $xR^{(0)}x$ it follows $xR^{*}x$.
-
-Let us prove by induction that for any ordinal $\alpha$, $R^{(\alpha)}$ is symmetric:
-for $\alpha=0$ this is immediate; suppose $xR^{(\alpha+1)}y$, then two cases are possible: either $xR^{(\alpha)}z$ and $zR^{(\alpha)}y$, then by IH 
-$yR^{(\alpha)}z$ and $zR^{(\alpha)}x$, whence $yR^{(\alpha+1)}x$; oer
-$x_{i}R^{(\alpha)}y_{i}$ for some decompositions $x=\bigvee_{i}x_{i}$ and $y=\bigvee_{i}y_{i}$; then by IH $y_{i}R^{(\alpha)}x_{i}$, so $yR^{(\alpha+1)}x$ holds.
-Finally, if $\alpha$ is limit, then from $x R^{(\alpha)}y$ it follows
-$xR^{(\beta)}y$ for some $\beta<\alpha$, whence 
-$yR^{(\beta)}x$ by IH and we conclude $yR^{(\alpha)}x$.
-
-Now, if $xR^{*}y$ then $xR^{(\alpha)}y$ holds for some ordinal $\alpha$, and thus $yR^{(\alpha)}x$ holds too, whence $yR^{*}x$.
-
-Observe that $\alpha<\beta$ implies $R^{(\alpha)}\subseteq R^{(\beta)}$:
-this is obvious if $\beta$ is limit, otherwise, if $\beta=\alpha+1$, from $xR^{(\alpha)}y$ and $x R^{(\alpha)}x$ we deduce
-$xR^{(\alpha+1)}y$.
-
-Suppose now $xR^{*}y$ and $yR^{*}z$. Then $xR^{(\alpha)}y$ and $yR^{(\beta)}z$ hold for some 
-ordinals $\alpha$ and $\beta$; let $\gamma=\max\{\alpha,\beta\}$; then 
-we have $xR^{(\gamma)}y$ and $yR^{(\gamma)}z$, whence $xR^{(\gamma+1)}z$ and thus 
-$xR^{*}z$.
-
-Suppose $x_{i}R^{*}y_{i}$ holds for all $i\in I$; then 
-for all $i$ there is some ordinal $\alpha_{i}$ with 
-$x_{i}R^{(\alpha_{i})}y_{i}$. 
-Let $\gamma=\sup_{i}\alpha_{i}$, so that 
-$x_{i}R^{(\gamma)}y_{i}$; then we have
-$\bigvee_{i}x_{i} R^{(\gamma+1)}\bigvee y_{i}$, and thus
-$\bigvee_{i}x_{i} R^{*}\bigvee_{i}y_{i}$.
-
-%
-%$R^{*}$ is symmetric, reflexive and transitive, given that $R$ is. 
-%
-%Let $I$ be a set and 
-%suppose $x_{i}R^{*}y_{i}$ holds for all $i\in I$.
-%Observe now that if $x=\bigvee_{i\in I} x_{i}=\bigvee_{j\in J}x'_{j}$, then
-%\begin{enumerate}
-%\item $x_{i}=\bigvee_{j\in J}x_{i}\land x_{j}$;
-%\item $x= \bigvee_{(i,j)\in I\times J}x_{i}\land x'_{j}$. 
-%\end{enumerate}
-% $\bigvee_{j\in J}x_{i}\land x_{j}\leq x_{i} \leq x_{i}$ is clear.
-% Conversely, we have $x_{i}= x \land x_{i} = \left (\bigvee_{j\in J}x_{j}\right) \land x_{i}= \bigvee_{j\in J}x_{j}\land x_{i}= \bigvee_{j\in J}x_{i}\land x_{j}$.
+Since $M$, as a $\Lawv$-category, is order-complete, tensored and cotensored, it is complete by Proposition \ref{prop:tencoten}.
+
+To conclude, notice that if $f:X\to Y$ is a continuous functor, then it commutes with tensors and, by 
+Proposition \ref{prop:tenfun} it commutes with joins, so it is a morphism of the respective $\Lawv$-modules.
+Conversely, if $f:M\to N$ is a $\Lawv$-module morphism, then, since $M$ and $N$ are both tensored $\Lawv$-categories, the tensor coincides with the actions of $M$ and $N$, $f$ preserves the joins and the tensor, by Proposition \ref{prop:tenfun}, it is a continuous functor of the respective $\Lawv$-categories.
+\end{proof}
+
+\subsection{$\Mod$ is a $*$-Autonomous Category}
+
+%
+%
+%Let us first observe that:
+%\begin{itemize}
+%\item the hom-set $\Mod(M,N)$ of two $\Lawv$-modules is a $\Lawv$-module with order and action defined pointwise;
+%
+%\item for any $\Lawv$-module $M=(M,\preceq, \star)$, there is a $\Lawv$-module
+%$M^{\mathrm{op}}=(M,\succeq, \multimapinv)$, with $\multimapinv$ defined as in the proof of Theorem \ref{thm:equivalence}.
+%
+%
+%\end{itemize}
+%
+%
+%
+%Let $M,N$ be two $\Lawv$-modules. For all $A\in \Lawv^{M\times N}$, we define the function
+%\begin{align*}
+%H_{A} & : \Lawv^{M} \longrightarrow \Lawv^{N}%\\
+%%K_{A} & : \Lawv^{N} \longrightarrow \Lawv^{M}
+%\end{align*}
+%via
+%\begin{align*}
+%H_{A}(x)(b)  &= \inf_{a\in M}x_{a}+A(a,b)\\
+%%K_{A}(x)(b)  &= \sup_{a\in M}x_{a}-A(a,b)
+%\end{align*}
+%
+%
+%\begin{lemma}
+%$H_{A}=H_{A'}$ iff $A=A'$. 
+%\end{lemma}
+%\begin{proof}
+%We only need to prove one direction, so suppose $A\neq A'$ and let $a,b$ be such that $A(a,b)\neq A'(a,b)$.
+%Let $x$ be defined by $x_{a}=1$ and $x_{a'}=\infty$ for all $a'\neq a$. Then $H_{A}(x)(b)=A(a,b)\neq A'(a,b)=H_{A'}(x)(b)$. 
+%\end{proof}
+%
+%\begin{proposition}
+%$\Lawv^{M\times N}$ and $\Mod(\Lawv^{M},\Lawv^{N})$ are isomorphic $\Lawv$-modules.
+%\end{proposition}
+%\begin{proof}
+%The map $A\mapsto H_{A}$ is injective, as shown above. We need to check that it commutes with joins:
+%\begin{align*}
+%H_{\bigvee_{i}A_{i}}(x)(b) & = \inf_{a}x_{a}+\bigvee_{i}A_{i}(a,b)\\
+%&=  \inf_{a}\bigvee_{i}x_{a}+A_{i}(a,b)\\
+%&=  \bigvee_{i}\inf_{a}x_{a}+A_{i}(a,b)\\
+%&=  \bigvee_{i}H_{A_{i}}(x)(b)\end{align*} 
+%(recall that $\inf$s are actually joins in $\Lawv$!)
+%
+%We must prove that $H$ is surjective: for all $f\in \Mod(\Lawv^{M},\Lawv^{N})$, let 
+%$k_{f}\in \Lawv^{M\times N}$ be given by $k_{f}(a,b)=f(e_{a})(b)$. 
+%
+%Then we have 
+%\begin{align*}
+%H_{k_{f}}(x)(b) & = \inf_{a}x_{a}+k_{f}(a,b) \\
+%&= \inf_{a}x_{a}+ f(e_{a})(b)\\
+%&= \left(\inf_{a}x_{a}+f(e_{a})\right)(b) \\
+%&= \left(\inf_{a}f(x_{a}+e_{a})\right)(b) \\
+%&= f(\inf_{a}x_{a}+e_{a})(b)\\
+%&=
+%f(x)(b)
+%\end{align*}
+%and conversely 
+%\begin{align*}
+%k_{H_{A}}(a,b)&= H_{A}(e_{a})(b)= \inf_{a'}(e_{a})_{a'}+A(a',b) =  A(a,b)
+%\end{align*}
+%\end{proof}
+%
+%
+%More generally, we have the following result:
+%\begin{proposition}
+%Let $X,Y$ be two $\Lawv$-modules. For any morphism $f: X\to Y$ there is a matrix $k_{f}\in \Lawv^{X\times Y}$ such that 
+%
+%\end{proposition}
+%\begin{proof}
+%By composing $f$ with the isomorphisms $\B Y^{\mathrm{op}}:X\to \B M(X)$, with inverse $\sup(x)=\bigvee_{a\in x}a\otimes x_{a}$, we obtain 
+%a morphism $\widehat f: \B M(X)\to \B M(Y)$  
+%\begin{align*}
+%\widehat f(x)(b)&:= \B Y^{\mathrm{op}}(f(\sup x))(b) \\
+%&= Y( f(\bigvee_{a\in X}a\otimes x_{a}),b) \\
+%&= Y( \bigvee_{a\in X}f(a)\otimes x_{a},b) \\
+%&= \sup_{a\in X}Y(f(a),b)-x_{a}
+%\end{align*}
+%Observe that $\widehat f$ can be extended to a function $f^{*}$ from $\Lawv^{X}$ to $\Lawv^{Y}$.
+%Now, $ f^{*}$ is generated by the matrix $k_{f}\in \Lawv^{X\times Y}$ given by $k_{ f}(a,b)=f(e_{a})(b)$, so that for all $x\in \Lawv^{X}$, 
+%$f^{*}(x)(b)=\bigvee_{a\in X}k_{f}(a,b)+x_{a}$, and thus
+%in particular, for all $x\in [X,Q]$, 
+%$\widehat f(x)(b)=f^{*}(x)(b)$.
+%\end{proof}
+
+\subsubsection{The Tensor Product of $\Lawv$-Modules}
+%
+% Let us first recall some important definitions:
+%
+%\begin{definition}[congruence on a sup-lattice]
+%Let $(L, \leq)$ be a sup-lattice. An equivalence relation $R\subseteq L\times L$ is said a \emph{congruence} if it satisfies the following property:
+%\begin{align}
+%(\forall i\in I \ x_{i} Ry_{i})  \ \To  \ \left( \bigvee_{i}x_{i} \right ) R\left(\bigvee_{i}y_{i}\right)
+%\tag{congruence}
+%\end{align}
+%\end{definition}
+%
+%\begin{lemma}
+%For all suplattices $(L,\leq)$, if $R$ is a congruence, then $(L/R, \leq_{R})$ is a sup-lattice, where $[x]\leq_{R}[y]$ iff $(x\vee y) R y$ (i.e.~$[x\vee y]=[y]$), and $\bigvee_{i}[x_{i}]=\left [\bigvee_{i}x_{i}\right]$.
+%\end{lemma}
+%\begin{proof}
+%Let us check that $\leq_{R}$ is an order. It is clear that $[x]\leq_{R}[x]$ holds. If $[x]\leq_{R}[y]$ and $[y]\leq_{R}[z]$ both hold, then 
+%$(x\vee y)Ry$ and $(y\vee z)Rz$ hold; 
+%then, since $R$ is a congruence $((x\vee y)\vee (y\vee z)) R (y\vee z)R z$, and moreover $x \vee (y\vee z) R (x \vee z)$, whence 
+%$(x\vee z) R(x \vee y\vee z)R z$, so $[x]\leq_{R}[z]$.  
+%If $[x]\leq_{R}[y]$ and $[y]\leq_{R}[x]$, then 
+%$xR(x\vee y)R y$, thus $[x]=[y]$.
+%
+%Let us now check the definition of joins. 
+%From $[x_{i}]\vee [\bigvee_{i}x_{i}]=[x_{i}\vee \bigvee_{i}x_{i}]= [\bigvee_{i}x_{i}]$ we deduce $[x_{i}]\leq_{R}[\bigvee x_{i}]$.
+%Suppose now $[x_{i}]\leq [y]$ holds for all $i\in I$, that is, 
+%$(x_{i}\vee y)Ry$; then, since $R$ is a congruence, 
+%$(\bigvee_{i}(x_{i}\vee y))R y$, that is, 
+%$((\bigvee_{i}x_{i})\vee y)Ry$, which implies 
+%$[\bigvee_{i}x_{i}]\leq_{R}[y]$. We conclude then that $\bigvee_{i}[x_{i}]=[\bigvee_{i}x_{i}]$.
+%\end{proof}
+%
+%
+%\begin{corollary}\label{cor:bigvee}
+%Let $(L,\leq)$ be a suplattice and $R$ be a congruence on $L$.
+%Then, for any class $\beta\in L/R$, $\beta= [\bigvee\beta]$.
+%\end{corollary}
+%\begin{proof}
+%$[\bigvee \beta]=[\bigvee\{ x \mid x\in \beta\}]= \bigvee \{[x]\mid x\in \beta\}=\beta$.
+%\end{proof}
+%
+%
+%\begin{proposition}\label{prop:smallestcongruence}
+%Let $(L,\leq)$ be a sup-lattice. Let $R\subseteq L\times L$ be an equivalence relation, and for any ordinal $\alpha$, let the relation
+%$R^{(\alpha)}\subseteq L\times L$ be defined by:
+%\begin{itemize}
+%\item $xR^{(0)}y$ iff either $xRy$, $x=y$ or $yRx$ holds;
+%\item $xR^{(\alpha+1)}y$ iff one of the following holds:
+%	\begin{itemize}
+%	\item for some $z$, $xR^{(\alpha)}z$ and $zR^{(\alpha)}y$ holds;
+%	\item for some set $I$, and families $x_{i},y_{i}$, 
+%	$x=\bigvee x_{i}, y=\bigvee_{i}y_{i}$ and $x_{i} R^{(\alpha)}y_{i}$ holds for all $i\in I$.
+%
+%	\end{itemize}
+%\item $xR^{(\gamma)}y$ iff $xR^{(\delta)}y$ holds for some $\delta <\gamma$, for $\gamma$ limit.
+%\end{itemize}
+%Then the relation $R^{*}\subseteq L\times L$ given by 
+%$$
+%xR^{*} y  \ \Leftrightarrow \ \exists \alpha . \mathrm{OR}(\alpha) \land xR^{(\alpha)}y
+%$$% defined as follows: $xR^{*} y$ iff
+%%\begin{enumerate}
+%%\item for any set $I$ and family $x_{i}$ such that $x=\bigvee_{i}x_{i}$, there exists a family $y_{i}$ such that $y=\bigvee_{i}y_{i}$ and $x_{i}R y_{i}$ holds for all $i\in I$;
+%%\item for any set $I$ and family $y_{i}$ such that $y=\bigvee_{i}y_{i}$, there exists a family $x_{i}$ such that $x=\bigvee_{i}x_{i}$ and $x_{i}R y_{i}$ holds for all $i\in I$.
+%%\end{enumerate}
+%%%\begin{align*}
+%%%xR^{*}y &\text{ iff }\  \forall I \ \forall x_{i} \  \text{ s.t. }
+%%%\left\{
+%%%\begin{matrix}
+%%%x=\bigvee_{i\in I}x_{i} \\
+%%%y=\bigvee_{i\in I}y_{i}\\
+%%%x_{i}Ry_{i} \ (\forall i\in I)
+%%%\end{matrix}
+%%%\right\}
+%%%\end{align*}
+%where $\mathrm{OR}(\alpha)$ is the property ``$\alpha$ is an ordinal'', 
+%is a congruence, and is the smallest congruence containing $R$.
+%\end{proposition}
+%\begin{proof}
+%From $xR^{(0)}x$ it follows $xR^{*}x$.
+%
+%Let us prove by induction that for any ordinal $\alpha$, $R^{(\alpha)}$ is symmetric:
+%for $\alpha=0$ this is immediate; suppose $xR^{(\alpha+1)}y$, then two cases are possible: either $xR^{(\alpha)}z$ and $zR^{(\alpha)}y$, then by IH 
+%$yR^{(\alpha)}z$ and $zR^{(\alpha)}x$, whence $yR^{(\alpha+1)}x$; oer
+%$x_{i}R^{(\alpha)}y_{i}$ for some decompositions $x=\bigvee_{i}x_{i}$ and $y=\bigvee_{i}y_{i}$; then by IH $y_{i}R^{(\alpha)}x_{i}$, so $yR^{(\alpha+1)}x$ holds.
+%Finally, if $\alpha$ is limit, then from $x R^{(\alpha)}y$ it follows
+%$xR^{(\beta)}y$ for some $\beta<\alpha$, whence 
+%$yR^{(\beta)}x$ by IH and we conclude $yR^{(\alpha)}x$.
+%
+%Now, if $xR^{*}y$ then $xR^{(\alpha)}y$ holds for some ordinal $\alpha$, and thus $yR^{(\alpha)}x$ holds too, whence $yR^{*}x$.
+%
+%
+%Observe that $\alpha<\beta$ implies $R^{(\alpha)}\subseteq R^{(\beta)}$:
+%this is obvious if $\beta$ is limit, otherwise, if $\beta=\alpha+1$, from $xR^{(\alpha)}y$ and $x R^{(\alpha)}x$ we deduce
+%$xR^{(\alpha+1)}y$.
+%
+%
+%Suppose now $xR^{*}y$ and $yR^{*}z$. Then $xR^{(\alpha)}y$ and $yR^{(\beta)}z$ hold for some 
+%ordinals $\alpha$ and $\beta$; let $\gamma=\max\{\alpha,\beta\}$; then 
+%we have $xR^{(\gamma)}y$ and $yR^{(\gamma)}z$, whence $xR^{(\gamma+1)}z$ and thus 
+%$xR^{*}z$.
+%
+%
+%Suppose $x_{i}R^{*}y_{i}$ holds for all $i\in I$; then 
+%for all $i$ there is some ordinal $\alpha_{i}$ with 
+%$x_{i}R^{(\alpha_{i})}y_{i}$. 
+%Let $\gamma=\sup_{i}\alpha_{i}$, so that 
+%$x_{i}R^{(\gamma)}y_{i}$; then we have
+%$\bigvee_{i}x_{i} R^{(\gamma+1)}\bigvee y_{i}$, and thus
+%$\bigvee_{i}x_{i} R^{*}\bigvee_{i}y_{i}$.
+%
+%%
+%%$R^{*}$ is symmetric, reflexive and transitive, given that $R$ is. 
+%%
+%%Let $I$ be a set and 
+%%suppose $x_{i}R^{*}y_{i}$ holds for all $i\in I$.
+%%Observe now that if $x=\bigvee_{i\in I} x_{i}=\bigvee_{j\in J}x'_{j}$, then
+%%\begin{enumerate}
+%%\item $x_{i}=\bigvee_{j\in J}x_{i}\land x_{j}$;
+%%\item $x= \bigvee_{(i,j)\in I\times J}x_{i}\land x'_{j}$. 
+%%\end{enumerate}
+%% $\bigvee_{j\in J}x_{i}\land x_{j}\leq x_{i} \leq x_{i}$ is clear.
+%% Conversely, we have $x_{i}= x \land x_{i} = \left (\bigvee_{j\in J}x_{j}\right) \land x_{i}= \bigvee_{j\in J}x_{j}\land x_{i}= \bigvee_{j\in J}x_{i}\land x_{j}$.
 {\color{red}NOT TRUE IN ANY LATTICE!}
-
-%
-% Then for each $i\in I$ there exists a set $J_{i}$ and sequences $x_{ij},y_{ij}$ such that 
-%$\bigvee_{j\in J_{i}}x_{ij}=x_{i}$, $\bigvee_{j\in J_{i}}y_{ij}=y_{i}$ and 
-%$x_{ij}Ry_{ij}$.
-%
-%Let then $K= \prod_{i\in I}\{i\}\times J_{i}$; then for all $(i,j)\in K$, 
-%$x_{ij}R y_{ij}$, so we deduce that $\bigvee_{i\in I}x_{i}=\bigvee_{(i,j)\in K}x_{ij} R^{*} \bigvee_{(i,j)\in K}y_{ij}=\bigvee_{i\in I}y_{i}$, which proves that $R^{*}$ is a congruence.
-
-Suppose now $S$ is a congruence containing $R$.
-Let us show that for any ordinal $\alpha$, $R^{(\alpha)}\subseteq S$:
-\begin{itemize}
-\item $R^{(0)}\subseteq S$ holds since $S$ is an equivalence relation and contains $R$;
-
-\item if $x R^{(\alpha+1)}y$ holds, then two cases occur:
-	\begin{itemize}
-	\item $xR^{(\alpha)}z$ and $zR^{(\alpha)}y$ hold, so by IH, 
-	$xSz$ and $zSy$, and since $S$ is transitive, $xSy$ holds;
-	\item $x=\bigvee_{i}x_{i}$, $y=\bigvee_{i}y_{i}$ and $x_{i}R^{(\alpha)}y_{i}$ holds; then by IH $x_{i}Sy_{i}$ holds, and since $S$ is a congruence, $xSy$ holds;
-
-\item if $\alpha$ is limit and $xR^{(\alpha)}y$ holds, then $xR^{(\beta)}y$ holds for some $\beta<\alpha$, and by IH $xS y$ holds.
-
-	\end{itemize}
-
-\end{itemize}
-Now, if $x R^{*} y$ holds, then $xR^{(\alpha)}y$ holds for some ordinal $\alpha$, whence $xS y$ holds.
-This shows that $R^{*}\subseteq S$.
-%
-%Then there exists a set $I$ so that $x=\bigvee_{i\in I}x_{i}$, $y=\bigvee_{i\in I}y_{i}$ and $x_{i}Ry_{i}$; now, since $S$ contains $R$, we deduce $x_{i}Sy_{i}$ for all $i\in I$, and since $S$ is a congruence, $xSy$ holds, which proves that $R^{*}\subseteq S$.
-\end{proof}
-
-We can now introduce the tensor of $Q$-modules, that will be defined as a suitable quotient lattice.
-
-\begin{definition}[tensor of $Q$-modules]
-Let $M$ and $N$ be $Q$-modules.
-The \emph{tensor product} $M\otimes_{Q}N$ of $M$ and $N$ is the $Q$-module defined as $\C P(M\times N)/ R^{*}$, where $R^{*}$ is the smallest congruence containing the relation $R$ defined by:
+% 
+%%
+%% Then for each $i\in I$ there exists a set $J_{i}$ and sequences $x_{ij},y_{ij}$ such that 
+%%$\bigvee_{j\in J_{i}}x_{ij}=x_{i}$, $\bigvee_{j\in J_{i}}y_{ij}=y_{i}$ and 
+%%$x_{ij}Ry_{ij}$.
+%%
+%%Let then $K= \prod_{i\in I}\{i\}\times J_{i}$; then for all $(i,j)\in K$, 
+%%$x_{ij}R y_{ij}$, so we deduce that $\bigvee_{i\in I}x_{i}=\bigvee_{(i,j)\in K}x_{ij} R^{*} \bigvee_{(i,j)\in K}y_{ij}=\bigvee_{i\in I}y_{i}$, which proves that $R^{*}$ is a congruence.
+%
+%Suppose now $S$ is a congruence containing $R$.
+%Let us show that for any ordinal $\alpha$, $R^{(\alpha)}\subseteq S$:
+%\begin{itemize}
+%\item $R^{(0)}\subseteq S$ holds since $S$ is an equivalence relation and contains $R$;
+%
+%\item if $x R^{(\alpha+1)}y$ holds, then two cases occur:
+%	\begin{itemize}
+%	\item $xR^{(\alpha)}z$ and $zR^{(\alpha)}y$ hold, so by IH, 
+%	$xSz$ and $zSy$, and since $S$ is transitive, $xSy$ holds;
+%	\item $x=\bigvee_{i}x_{i}$, $y=\bigvee_{i}y_{i}$ and $x_{i}R^{(\alpha)}y_{i}$ holds; then by IH $x_{i}Sy_{i}$ holds, and since $S$ is a congruence, $xSy$ holds;
+%	
+%\item if $\alpha$ is limit and $xR^{(\alpha)}y$ holds, then $xR^{(\beta)}y$ holds for some $\beta<\alpha$, and by IH $xS y$ holds.
+%	
+%	\end{itemize}
+%
+%\end{itemize}
+%Now, if $x R^{*} y$ holds, then $xR^{(\alpha)}y$ holds for some ordinal $\alpha$, whence $xS y$ holds.
+%This shows that $R^{*}\subseteq S$.
+%%
+%%Then there exists a set $I$ so that $x=\bigvee_{i\in I}x_{i}$, $y=\bigvee_{i\in I}y_{i}$ and $x_{i}Ry_{i}$; now, since $S$ contains $R$, we deduce $x_{i}Sy_{i}$ for all $i\in I$, and since $S$ is a congruence, $xSy$ holds, which proves that $R^{*}\subseteq S$.
+%\end{proof}
+
+The tensor of $\Lawv$-modules can be introduced as a suitable quotient lattice.
+
+\begin{definition}[tensor of $\Lawv$-modules]
+Let $M$ and $N$ be $\Lawv$-modules.
+The \emph{tensor product} $M\otimes N$ of $M$ and $N$ is the $\Lawv$-module defined as $\C P(M\times N)/ R^{*}$, where $R^{*}$ is the smallest congruence containing the relation $R$ defined by:
 \begin{align*}
@@
 
-Let a \emph{$Q$-bimorphism} be a map $f:M\times N\to L$ such that $f$ preserves joins in each variable separately, and moreover $f(x,y\star \epsilon)=f(x\star\epsilon,y)$.
-A $Q$-bimorphism $f:M\times N\to L$ is \emph{universal} if for all $L'$ and bimorphism $g:M\times N\to L'$ there is a unique sup-lattice homomorphism $h:L\to L'$ such that $g=h\circ f$.
-
-\begin{proposition}[universal property of the tensor product]
-The tensor product $M\otimes_{Q}N$ is the codomain of the universal $Q$-bimorphism $M\times N \to M\otimes_{Q}N$.
+Let a \emph{$\Lawv$-bimorphism} be a map $f:M\times N\to L$ such that $f$ preserves joins in each variable separately, and moreover $f(x,y\star \epsilon)=f(x\star\epsilon,y)$.
+A $\Lawv$-bimorphism $f:M\times N\to L$ is \emph{universal} if for all $L'$ and bimorphism $g:M\times N\to L'$ there is a unique sup-lattice homomorphism $h:L\to L'$ such that $g=h\circ f$.
+
+\begin{proposition}[universal property of the tensor product, cf.~\ref{Russo2007}]
+The tensor product $M\otimes N$ is the codomain of the universal $\Lawv$-bimorphism $M\times N \to M\otimes N$.
 \end{proposition}
@@
 
-For any $x\in M$ and $y\in N$, we indicate as $x\otimes_{Q}y$ the image of the pair $(x,y)$ under the universal $Q$-bimorphism $\tau:M\times N\to M\otimes_{Q}N$, or equivalently, as the $R^{*}$-equivalence class of $(x,y)$.
-Since joins in $M\otimes_{Q}N$ are given by  
+For any $x\in M$ and $y\in N$, we indicate as $x\otimes y$ the image of the pair $(x,y)$ under the universal $\Lawv$-bimorphism $\tau:M\times N\to M\otimes N$, or equivalently, as the $R^{*}$-equivalence class of $(x,y)$.
+Since joins in $M\otimes N$ are given by  
 $\bigvee_{i}[A_{i}]= [\bigcup_{i}A_{i}]$, we have then 
-that any element $[A]\in M\otimes_{Q}N$ can be written as 
+that any element $[A]\in M\otimes N$ can be written as 
 $[A]= \bigvee\{x\otimes y\mid (x,y)\in A\}$.
@@
 %
-% Any element of $M\otimes_{Q}N$ is a join of tensors, that is
+% Any element of $M\otimes N$ is a join of tensors, that is
 %$$
-%M\otimes_{Q}N= \left\{ \bigvee_{i\in I}x_{i}\otimes_{Q}y_{i}\ \Bigg \vert \  x_{i}\in M, y_{i}\in N \right\}
+%M\otimes N= \left\{ \bigvee_{i\in I}x_{i}\otimes y_{i}\ \Bigg \vert \  x_{i}\in M, y_{i}\in N \right\}
 %$$
@@
 
-\begin{lemma}
+\begin{lemma}[cf.~\ref{Russo2007}]
 \begin{itemize}
-\item $M\otimes_{Q}N\simeq N\otimes_{Q}M$;
-
-\item $(M\otimes_{Q}N)\otimes_{Q}R \simeq M\otimes_{Q}(N\otimes_{Q}R)$.
+\item $M\otimes N\simeq N\otimes M$;
+
+\item $(M\otimes N)\otimes R \simeq M\otimes (N\otimes R)$.
 \end{itemize}
@@
 \begin{proposition}
-$Hom(M\otimes_{Q}N, R)   \simeq   Hom(M, Hom(N,R))$ (as an isomorphism of sup-lattices).
+$\Mod(M\otimes N, R)   \simeq   \Mod(M, \Mod(N,R))$ (as an isomorphism of sup-lattices).
 \end{proposition}
 \begin{proof}
-Given $h:M\otimes_{Q}N\to R$, for all $x\in M$ define $h_{x}:N\to R$ by 
-$h_{x}(y)=h(x\otimes_{Q}y)$. We then have 
-$h_{x}(\bigvee_{i}y_{i})=h(x\otimes_{Q}\bigvee_{i}y_{i})=
-h(\bigvee_{i}x\otimes_{Q} y_{i})=
-\bigvee_{i}h(x\otimes_{Q} y_{i})=\bigvee_{i}h_{x}(y_{i})$ and 
-$h_{x}(y\star\epsilon)= h(x\otimes_{Q}(y\star \epsilon))=h((x\otimes_{Q}y)\star\epsilon)=h(x\otimes_{Q}y)\star \epsilon= h_{x}(y)\star\epsilon$, so $h_{x}\in Hom(N,R)$.
+Given $h:M\otimes N\to R$, for all $x\in M$ define $h_{x}:N\to R$ by 
+$h_{x}(y)=h(x\otimes y)$. We then have 
+$h_{x}(\bigvee_{i}y_{i})=h(x\otimes \bigvee_{i}y_{i})=
+h(\bigvee_{i}x\otimes  y_{i})=
+\bigvee_{i}h(x\otimes  y_{i})=\bigvee_{i}h_{x}(y_{i})$ and 
+$h_{x}(y\star\epsilon)= h(x\otimes (y\star \epsilon))=h((x\otimes y)\star\epsilon)=h(x\otimes y)\star \epsilon= h_{x}(y)\star\epsilon$, so $h_{x}\in \Mod(N,R)$.
 Moreover, by a similar argument we have  $h_{\bigvee_{i}x_{i}}(y)=\bigvee_{i}h_{x_{i}}(y)$
-and $h_{x\star\epsilon}(y)=h_{x}(y)\star\epsilon$, so the map $x\mapsto h_{x}$ is a $Q$-module morphism.
-
-Finally, for any family $h_{i}:M\otimes_{Q}N\to R$, we have 
-$(\bigvee_{i}h_{i})_{x}(y)=\bigvee_{i}h_{i}(x\otimes_{Q}y)=\bigvee_{i}(h_{i})_{x}(y)=(\bigvee_{i}h_{i})(x\otimes y)$, and thus we have a sup-lattice homomorphism $\zeta$ from $Hom(M\otimes_{Q}N,R)$ to $Hom(M,Hom(N,R))$ given by $\zeta(h)=h_{\_}$.
-
-Let us show that $\zeta$ has an inverse:  for all $f\in Hom(M,Hom(N,R))$, define $f': M\times N\to R$ by $f'(x,y):=f(x)(y)$.
-This is clearly a bimorphism, so there is a unique homomorphism $h_{f'}:M\otimes_{Q}N\to M\times N$ such that $f'=h_{f'}\circ \tau$, i.e.~such that
+and $h_{x\star\epsilon}(y)=h_{x}(y)\star\epsilon$, so the map $x\mapsto h_{x}$ is a $\Lawv$-module morphism.
+
+Finally, for any family $h_{i}:M\otimes N\to R$, we have 
+$(\bigvee_{i}h_{i})_{x}(y)=\bigvee_{i}h_{i}(x\otimes y)=\bigvee_{i}(h_{i})_{x}(y)=(\bigvee_{i}h_{i})(x\otimes y)$, and thus we have a sup-lattice homomorphism $\zeta$ from $\Mod(M\otimes N,R)$ to $\Mod(M,\Mod(N,R))$ given by $\zeta(h)=h_{\_}$.
+
+Let us show that $\zeta$ has an inverse:  for all $f\in \Mod(M,\Mod(N,R))$, define $f': M\times N\to R$ by $f'(x,y):=f(x)(y)$.
+This is clearly a bimorphism, so there is a unique homomorphism $h_{f'}:M\otimes N\to M\times N$ such that $f'=h_{f'}\circ \tau$, i.e.~such that
 $h_{f'}(x\otimes y)=f'(x,y)=f(x)(y)$, and thus $\zeta(h_{f'})=f$.
@@
 \begin{itemize}
-\item[i.] $Hom(Q,M)\simeq M$.
-\item[ii.] $Hom(M,N)\simeq Hom(N^{\mathrm{op}},M^{\mathrm{op}})$.
-\item[iii.] $Hom(M, Q^{\mathrm{op}})\simeq M^{\mathrm{op}}$.
+\item[i.] $\Mod(Q,M)\simeq M$.
+\item[ii.] $\Mod(M,N)\simeq \Mod(N^{\mathrm{op}},M^{\mathrm{op}})$.
+\item[iii.] $\Mod(M, \Lawv^{\mathrm{op}})\simeq M^{\mathrm{op}}$.
 \end{itemize}
@@
 \begin{proof}
-Define $\alpha:M\to Hom(Q,M)$ by $\alpha(x)(\epsilon)=x\star\epsilon$ and 
-$\beta:Hom(Q,M)\to M$ by $\beta(f)=f(0)$. Then we have that 
+Define $\alpha:M\to \Mod(Q,M)$ by $\alpha(x)(\epsilon)=x\star\epsilon$ and 
+$\beta:\Mod(Q,M)\to M$ by $\beta(f)=f(0)$. Then we have that 
 $\alpha(\beta(f))(\epsilon)=\alpha(f(0))(\epsilon)=f(0)\star\epsilon=f(\epsilon)$, and 
@@
 
-If $f\in Hom(M,N)$, since it preserves joins, it has a right-adjoint $f^{*}:N^{\mathrm{op}}\to M^{\mathrm{op}}$, such that $f(x)\leq y$ iff $x\leq f^{*}(y)$.
-
-By i.~$Hom(Q,M^{\mathrm{op}})\simeq M^{\mathrm{op}}$ and we conclude by ii.
+If $f\in \Mod(M,N)$, since it preserves joins, it has a right-adjoint $f^{*}:N^{\mathrm{op}}\to M^{\mathrm{op}}$, such that $f(x)\leq y$ iff $x\leq f^{*}(y)$.
+
+By i.~$\Mod(Q,M^{\mathrm{op}})\simeq M^{\mathrm{op}}$ and we conclude by ii.
 \end{proof}
@@
 \begin{itemize}
-\item[i.] $Hom(M,N)\simeq (M\otimes_{Q}N^{\mathrm{op}})^{\mathrm{op}}$.
-\item[ii.] $M\otimes_{Q}N\simeq Hom(M, N^{\mathrm{op}})^{\mathrm{op}}$.
-\item[iii.] $Q\otimes_{Q} M \simeq M\otimes_{Q}Q\simeq Q$.
+\item[i.] $\Mod(M,N)\simeq (M\otimes N^{\mathrm{op}})^{\mathrm{op}}$.
+\item[ii.] $M\otimes N\simeq \Mod(M, N^{\mathrm{op}})^{\mathrm{op}}$.
+\item[iii.] $Q\otimes  M \simeq M\otimes Q\simeq Q$.
 \end{itemize}
@@
 \begin{proof}
-$Hom(M,N)\simeq Hom(M, Hom(N^{\mathrm{op}},Q^{\mathrm{op}}))
-\simeq Hom(M\otimes N^{\mathrm{op}},Q^{\mathrm{op}})\simeq
+$\Mod(M,N)\simeq \Mod(M, \Mod(N^{\mathrm{op}},\Lawv^{\mathrm{op}}))
+\simeq \Mod(M\otimes N^{\mathrm{op}},\Lawv^{\mathrm{op}})\simeq
 (M\otimes N^{\mathrm{op}})^{\mathrm{op}}$.
@@
 
-For claim iii.~$Q\otimes_{Q}M\simeq M\otimes_{Q}M\simeq Hom(M, Q^{\mathrm{op}})^{\mathrm{op}}\simeq (M^{\mathrm{op}})^{\mathrm{op}}=M$. 
+For claim iii.~$Q\otimes M\simeq M\otimes M\simeq \Mod(M, \Lawv^{\mathrm{op}})^{\mathrm{op}}\simeq (M^{\mathrm{op}})^{\mathrm{op}}=M$. 
 \end{proof}
@@
 \begin{theorem}
-$Q\mathsf{Mod}$ is a $^{*}$-autonomous category.
+$\Mod$ is a $^{*}$-autonomous category.
 \end{theorem}
 
-\begin{proposition}\label{prop:Qtensor}
+Further useful properties are the following:
+
+\begin{proposition}\label{prop:Qtensor}[cf.~\cite{Russo2007}]
 \begin{itemize}
-\item[i.] $Q^{X}\otimes_{Q}M \simeq M^{X}$;
-\item[ii.] $Q^{X}\otimes_{Q}Q^{Y}\simeq Q^{X\times Y}$.
+\item[i.] $\Lawv^{X}\otimes M \simeq M^{X}$;
+\item[ii.] $\Lawv^{X}\otimes \Lawv^{Y}\simeq \Lawv^{X\times Y}$.
 \end{itemize}
@@
 $M^{X}$ coincides with the product $\Pi_{x\in X}M$. We have then 
-$Q^{X}\otimes_{Q}M \simeq (\Pi_{x}Q)\otimes_{Q}M \simeq \Pi_{x}(Q\otimes_{Q}M) \simeq \Pi_{x}M\simeq M^{X}$.
-
-For ii., using i.~we have $Q^{X}\otimes_{Q}Q^{Y}\simeq( Q^{Y})^{X}\simeq Q^{X\times Y}$.
-\end{proof}
-
-\subsubsection{The Tensor Product of $Q$-Categories}
-
-Thanks to Theorem \ref{thm:equivalence}, the $^{*}$-autonomous structure of $Q\mathsf{Mod}$ translates into a $^{*}$-autonomous structure for $Q\mathsf{CCat}$.
-
-In $\mathsf{Met}$ the ``tensor product'' of two metric spaces $X$ and $Y$ is just their cartesian product, with the ``plus'' metric. What can we say about the tensor product in $Q\mathsf{CCat}$?
-
-% 
-%Let us start by discussing the construction of the tensor product $M\otimes_{Q}N$ of $Q$-modules $M$ and $N$ as a quotient of $\C P(M\times N)$.
+$\Lawv^{X}\otimes M \simeq (\Pi_{x}Q)\otimes M \simeq \Pi_{x}(Q\otimes M) \simeq \Pi_{x}M\simeq M^{X}$.
+
+For ii., using i.~we have $\Lawv^{X}\otimes \Lawv^{Y}\simeq( \Lawv^{Y})^{X}\simeq \Lawv^{X\times Y}$.
+\end{proof}
+
+$\Mod$ admits biproducts, since products and coproducts exist in $\Mod$ and both coincide with the order product:
+$$
+\prod_{i\in I}X_{i} \simeq \coprod_{i\in I}X_{i}
+$$
+In particular, the projection and
+ injection morphisms $\pi_{i}:\prod_{i}X_{i}\to X_{i}$ and $\iota_{i}:X_{i} \to \prod_{i\in I}X_{i}$ are defined by 
+$$
+\pi_{i}\big( (x_{j})_{j\in I}\big) = x_{i} \qquad \qquad
+\iota_{i}(x)({j})= \begin{cases} x & \text{ if }i=j\\ \bot & \text{ otherwise}\end{cases}
+$$
+Moreover, biproducts commute with tensors as follows:
+\begin{proposition}\label{prop:productvstensor}[cf.~\cite{Russo2007}]
+$\prod_{i\in I} X\otimes Y_{i} \simeq X \otimes \prod_{i\in I}Y_{i}$.
+\end{proposition}
+\begin{proof}[Proof sketch]
+We just recall the isomorphism 
+$h: \prod_{i\in I} X\otimes Y_{i} \simeq X \otimes \prod_{i\in I}Y_{i}$, defined as follows:
+\begin{align*}
+h\Big( i\mapsto \bigvee_{k\in J_{i}}x_{i,k}\otimes y_{i,k}\Big ) = 
+\bigvee_{i\in I,k\in J_{i}}x_{i,k} \otimes \iota_{i}\big( (y_{j,k})_{j\in I}\big)
+\end{align*}
+and its inverse
+\begin{align*}
+k\Big( \bigvee_{k\in J}x_{k}\otimes \big(x_{k,i}\big)_{i\in I_{k}}\Big)(i)=
+\bigvee_{k\in J,i\in I_{k}}x_{k}\otimes \pi_{i}\big((x_{k,j})_{j\in I_{k}}\big).
+\end{align*}
+\end{proof}
+
+\subsubsection{The Tensor Product of $\Lawv$-Categories}
+
+Thanks to Theorem \ref{thm:equivalence}, the $^{*}$-autonomous structure of $\Mod$ translates into a $^{*}$-autonomous structure for $\GMet$.
+
+In $\mathsf{Met}$ the ``tensor product'' of two metric spaces $X$ and $Y$ is just their cartesian product, with the ``plus'' metric. What can we say about the tensor product in $\GMet$?
+
+% 
+%Let us start by discussing the construction of the tensor product $M\otimes N$ of $\Lawv$-modules $M$ and $N$ as a quotient of $\C P(M\times N)$.
 First, $\C P(M\times N)$ is the \emph{free} sup-lattice over $M\times N$: this means that any morphism
@@
 %\end{proof}
-The goal of this subsection is to describe the $Q$-categorical structure of the tensor product explicitly.
-The main intuition is that the elements of $X\otimes_{Q}Y$ can be seen as joins of pairs $x\otimes y$ of elements $x\in X$, $y\in Y$.
+The goal of this subsection is to describe the $\Lawv$-categorical structure of the tensor product explicitly.
+The main intuition is that the elements of $X\otimes Y$ can be seen as joins of pairs $x\otimes y$ of elements $x\in X$, $y\in Y$.
 We will then show that the metric coincides with the ``plus'' metric over pairs $x\otimes y$, and extends continuously to joins.
@@
 $$
-(M\otimes_{Q} N)(m\otimes_{Q} n, m'\otimes_{Q}n')= M(m,m') +N(n,n')
+(M\otimes  N)(m\otimes  n, m'\otimes n')= M(m,m') +N(n,n')
 $$
@@
 $$
-(M\otimes_{Q}N)([A],[B])= \sup_{(x,y)\in A}\inf_{(x',y')\in B}M(x,x')+N(y,y')
+(M\otimes N)([A],[B])= \sup_{(x,y)\in A}\inf_{(x',y')\in B}M(x,x')+N(y,y')
 $$
@@
 \begin{proof}
-By definition, $(M\otimes_{Q} N)(m\otimes_{Q} n, m'\otimes_{Q}n')$ is given by $\inf A$, where
-$$
-A=\{ \epsilon \mid (m\otimes_{Q}n)\star\epsilon \geq m'\otimes_{Q}n'\}
+By definition, $(M\otimes  N)(m\otimes  n, m'\otimes n')$ is given by $\inf A$, where
+$$
+A=\{ \epsilon \mid (m\otimes n)\star\epsilon \geq m'\otimes n'\}
 $$
@@
 \begin{align*}
-(M\otimes_{Q} N)(m\otimes_{Q} n, m'\otimes_{Q}n')& =\inf A \\
+(M\otimes  N)(m\otimes  n, m'\otimes n')& =\inf A \\
 &= \inf B \\
@@
 
-For the second claim, using the fact that $M\otimes_{Q}N$, as a $Q$-category, is both tensored and cotensored, using the fact that 
+For the second claim, using the fact that $M\otimes N$, as a $\Lawv$-category, is both tensored and cotensored, using the fact that 
 $[A]=\bigvee_{(x,y)\in A}x\otimes y$ and $[B]=\bigvee_{(x',y')\in B}z\otimes w$, and Lemma \ref{lemma:supinf}:
 \begin{align*}
-(M\otimes_{Q} N)([A],[B])&=
-\sup_{(x,y)\in A}(M\otimes_{Q}N)(x\otimes y, [B]) \\
-&=\sup_{(x,y)\in A}\inf_{(x',y')\in B}(M\otimes_{Q}N)(x\otimes y, x'\otimes y') 
+(M\otimes  N)([A],[B])&=
+\sup_{(x,y)\in A}(M\otimes N)(x\otimes y, [B]) \\
+&=\sup_{(x,y)\in A}\inf_{(x',y')\in B}(M\otimes N)(x\otimes y, x'\otimes y') 
 \\
@@
 
-\subsection{Products and Coproducts in $Q\mathsf{Mod}$}
-
-Products and coproducts exist in $Q\mathsf{Mod}$ and both coincide with the order product:
-$$
-\prod_{i\in I}X_{i} \simeq \coprod_{i\in I}X_{i}
-$$
-In particular, the projection and
- injection morphisms $\pi_{i}:\prod_{i}X_{i}\to X_{i}$ and $\iota_{i}:X_{i} \to \prod_{i\in I}X_{i}$ are defined by 
-$$
-\pi_{i}\big( (x_{j})_{j\in I}\big) = x_{i} \qquad \qquad
-\iota_{i}(x)({j})= \begin{cases} x & \text{ if }i=j\\ \bot & \text{ otherwise}\end{cases}
-$$
-
-The following property is important
-\begin{proposition}\label{prop:productvstensor}
-$\prod_{i\in I} X\otimes Y_{i} \simeq X \otimes \prod_{i\in I}Y_{i}$.
-\end{proposition}
-\begin{proof}
-For a complete proof, see \texttt{https://arxiv.org/pdf/0909.4493.pdf}. Here we just recall the isomorphism 
-$h: \prod_{i\in I} X\otimes Y_{i} \simeq X \otimes \prod_{i\in I}Y_{i}$, defined as follows:
-\begin{align*}
-h\Big( i\mapsto \bigvee_{k\in J_{i}}x_{i,k}\otimes y_{i,k}\Big ) = 
-\bigvee_{i\in I,k\in J_{i}}x_{i,k} \otimes \iota_{i}\big( (y_{j,k})_{j\in I}\big)
-\end{align*}
-and its inverse
-\begin{align*}
-k\Big( \bigvee_{k\in J}x_{k}\otimes \big(x_{k,i}\big)_{i\in I_{k}}\Big)(i)=
-\bigvee_{k\in J,i\in I_{k}}x_{k}\otimes \pi_{i}\big((x_{k,j})_{j\in I_{k}}\big).
-\end{align*}
-\end{proof}
-
-\subsection{Exponential and Differential Structure of $Q\mathsf{Mod}$}
-
-\subsubsection{Symmetric Algebras in $Q\mathsf{Mod}$ }
-
-We start by constructing the exponential co-modality in the equivalent category $Q\mathsf{Mod}$.
-
-For all $n\in \BB N$ and $\sigma\in \F S_{n}$, let $\langle \sigma\rangle: X^{\otimes n}\to X^{\otimes n}$ be the morphism defined by 
-$$
-\langle \sigma \rangle (x_{1}\otimes \dots \otimes x_{n})=
-x_{\sigma(1)}\otimes \dots \otimes x_{\sigma(n)}
-$$
-and extended to all other points by continuity.
-%For any $Q$-module $X$, let $X_{\bullet}$ indicate the $Q$-module $X\times Q$.  
-%
-\begin{definition}Let $X$ be a $Q$-module and $n\in \BB N$. 
-An element $x\in X^{\otimes_{n}}$ is said \emph{permutation-invariant} (in short, \emph{p-invariant}) if for all $\sigma\in \F S_{n}$, 
-$\langle \sigma \rangle (x)=x$.
-
-For all $n\in \BB N$, $x_{1},\dots, x_{n}\in X$, the \emph{$Q$-multiset} $[x_{1},\dots, x_{n}]$ is the element of $X^{\otimes_{n}}$ defined as $0$ if $n=0$, and otherwise as follows:
-$$
-[x_{1},\dots, x_{n}]:= \bigvee_{\sigma\in \F S_{n}}x_{\sigma(1)}\otimes \dots \otimes x_{\sigma(n)}
-$$
-
-Given multisets $A$ and $B$, we define the multiset $A\cup B$ as follows:
+\subsection{Exponential and Differential Structure of $\Mod$}
+
+\subsubsection{Symmetric Algebras in $\Mod$ }
+
+Given $\Lawv$-multisets $A$ and $B$, we define the multiset $A\cup B$ as follows:
 \begin{itemize}
@@
 
-\end{definition}
-
 \begin{proposition}
-Let $X$ be a $Q$-module and $n\in \BB N$. Any $Q$-multiset $[x_{1},\dots, x_{n}]\in X^{\otimes_{n}}$ is p-invariant. Moreover, any p-invariant element $x\in X^{\otimes_{n}}$ can be written as 
-a join of $Q$-multisets.
+Let $X$ be a $\Lawv$-module and $n\in \BB N$. Any $\Lawv$-multiset $[x_{1},\dots, x_{n}]\in X^{\otimes_{n}}$ is p-invariant. Moreover, any p-invariant element $x\in X^{\otimes_{n}}$ can be written as 
+a join of $\Lawv$-multisets.
 \end{proposition}
@@
 \begin{proposition}
-For any $Q$-module $X$, the set $!_{n}X\subseteq X$ of p-invariant elements of $X^{\otimes_{n}}$ is a sub-$Q$-module of $X$.
+For any $\Lawv$-module $X$, the set $!_{n}X\subseteq X$ of p-invariant elements of $X^{\otimes_{n}}$ is a sub-$\Lawv$-module of $X$.
 \end{proposition}
@@
 $x\otimes \epsilon$ is also p-invariant, since $\langle \sigma \rangle (x\otimes \epsilon)= \langle \sigma \rangle (x)\otimes \epsilon=x\otimes \epsilon$.
-We conclude that $!_{n}X$ is a sup-lattice with a continuous action of $Q$, where both the order and the action are inherited from $X$, so it is a sub-$Q$-module of $X$.
+We conclude that $!_{n}X$ is a sup-lattice with a continuous action of $\Lawv$, where both the order and the action are inherited from $X$, so it is a sub-$\Lawv$-module of $X$.
 \end{proof}
@@
 \begin{proposition}
-For any $Q$-module $X$ and $n\in \BB N$, the inclusion morphism 
+For any $\Lawv$-module $X$ and $n\in \BB N$, the inclusion morphism 
 $\iota:!_{n}X\longrightarrow X^{\otimes_{n}}$ is the equalizer of the diagram
@@
 \begin{remark}[metric structure of $!_{n}X$]
-As $!_{n}X$ is a sub-$Q$-module of $X^{\otimes_{n}}$, the distance function can be computed explicitly using Proposition \ref{prop:tensormetric}:
+As $!_{n}X$ is a sub-$\Lawv$-module of $X^{\otimes_{n}}$, the distance function can be computed explicitly using Proposition \ref{prop:tensormetric}:
 \begin{align*}
@@
 
-We now show that the $Q$-module $!_{n}X$ is isomorphic to the symmetric algebra, which is used in the construction of the exponential modality in the relational model.
+We now show that the $\Lawv$-module $!_{n}X$ is isomorphic to the symmetric algebra, which is used in the construction of the exponential modality in the relational model.
 
 \begin{definition}[symmetric algebra]
-For any $Q$-module $X$ and $n\in \BB N$, we let $\mathrm{Sym}_{n}(X)$ indicate the $Q$-module defined as 
+For any $\Lawv$-module $X$ and $n\in \BB N$, we let $\mathrm{Sym}_{n}(X)$ indicate the $\Lawv$-module defined as 
 $\mathrm{Sym}_{n}(X):=\frac{X^{\otimes_{n}}}{\sim_{n}}
@@
 Let us now show that for all p-invariant point $x_{0}$, and for all $y,z\in X^{\otimes_{n}}$, if 
-$y\leq x_{0}$ and 
-$z\sim^{\alpha}y$ holds, then $z\leq x_{0}$.
+$y\leq x_{0}$ and $z\sim y$ holds, then $z\leq x_{0}$.
+
+We will exploit the fact that $\sim$ is the least congruence containing the relation $\sim_{0}$ induced by the action of permutations. More precisely, $\sim$ can be defined explicitly as 
+$$
+x\sim y  \ \Leftrightarrow \ \exists \alpha . \mathrm{OR}(\alpha) \land x\sim^{(\alpha)}y
+$$
+where $\mathrm{OR}(\alpha)$ is the property ``$\alpha$ is an ordinal'', and the relations $\sim^{(\alpha)}$ are defined by induction as follows:
+\begin{itemize}
+\item $x\sim^{(0)}y$ iff either $x\sim_{0}y$, $x=y$ or $y\sim_{0}x$ holds;
+\item $x\sim^{(\alpha+1)}y$ iff one of the following holds:
+	\begin{itemize}
+	\item for some $z$, $x\sim^{(\alpha)}z$ and $z\sim^{(\alpha)}y$ holds;
+	\item for some set $I$, and families $x_{i},y_{i}$, 
+	$x=\bigvee x_{i}, y=\bigvee_{i}y_{i}$ and $x_{i} \sim^{(\alpha)}y_{i}$ holds for all $i\in I$.
+
+	\end{itemize}
+\item $x\sim^{(\gamma)}y$ iff $x\sim^{(\delta)}y$ holds for some $\delta <\gamma$, for $\gamma$ limit.
+\end{itemize}
+
+We will now prove, by induction on an ordinal $\alpha$, that for all p-invariant point $x_{0}$, and for all $y,z\in X^{\otimes_{n}}$, if 
+$y\leq x_{0}$ and $z\sim^{(\alpha)} y$ holds, then $z\leq x_{0}$.
+From this the claim will follow.
+
 \begin{itemize}
@@
 \begin{lemma}
-For any set $S$, there exists an isomorphism of $Q$-modules
-$$!_{n}Q^{S} \simeq Q^{\C M_{n}(S)}$$
+For any set $S$, there exists an isomorphism of $\Lawv$-modules
+$$!_{n}\Lawv^{S} \simeq \Lawv^{\C M_{n}(S)}$$
 where $\C M_{n}(X)$ indicates the set of multisets of $X$ of cardinality $ n$.
@@
 \begin{proof}
-Let us show that the morphism $h:Q^{\C M_{n}(S)}\to Q^{S\times \dots \times S}$ defined by 
+Let us show that the morphism $h:\Lawv^{\C M_{n}(S)}\to \Lawv^{S\times \dots \times S}$ defined by 
 $$
@@
 \begin{tikzcd}
-Q^{\C M_{n}(S)} \ar{r}{h} &Q^{S\times \dots \times S}\ar{r}{[\sigma]} &
-Q^{S\times \dots \times S}
+\Lawv^{\C M_{n}(S)} \ar{r}{h} &\Lawv^{S\times \dots \times S}\ar{r}{[\sigma]} &
+\Lawv^{S\times \dots \times S}
 \end{tikzcd}
 $$
-where $[\sigma](\B x)(\langle s_{1},\dots, s_{n}\rangle)=\B x(\langle x_{\sigma(1)},\dots, x_{\sigma(n)}\rangle)$, with $\sigma$ varying over $\F S_{n}$.
+where $[\sigma](x)(\langle s_{1},\dots, s_{n}\rangle)=x(\langle x_{\sigma(1)},\dots, x_{\sigma(n)}\rangle)$, with $\sigma$ varying over $\F S_{n}$.
 
 It is immediate that $h\circ [\sigma]=h\circ [\tau]$, for all $\sigma,\tau\in \F S_{n}$.
-Let now $k: C\to Q^{S\times \dots \times S}$ satisfy $k\circ [\sigma]=k\circ [\tau]$: then for all $c\in C$, $k(c)(\langle s_{1},\dots, s_{n}\rangle)=k(c)(\langle s_{\sigma(1)},\dots, s_{\sigma(n)}\rangle)$, so $k(c)$ actually defines a unique element of $Q^{\C M_{n}(S)}$, and thus $k$ splits in a unique way as $C \stackrel{k'}{\to} Q^{\C M_{n}(S)} \stackrel{h}{\to}Q^{S\times \dots \times S}$.
+Let now $k: C\to \Lawv^{S\times \dots \times S}$ satisfy $k\circ [\sigma]=k\circ [\tau]$: then for all $c\in C$, $k(c)(\langle s_{1},\dots, s_{n}\rangle)=k(c)(\langle s_{\sigma(1)},\dots, s_{\sigma(n)}\rangle)$, so $k(c)$ actually defines a unique element of $\Lawv^{\C M_{n}(S)}$, and thus $k$ splits in a unique way as $C \stackrel{k'}{\to} \Lawv^{\C M_{n}(S)} \stackrel{h}{\to}\Lawv^{S\times \dots \times S}$.
 
 Now, to conclude it suffices to observe that, by Proposition \ref{prop:Qtensor}, 
-$Q^{S\times \dots \times S}\simeq (Q^{S})^{\otimes_{n}}$, and then, since equalizers are unique up to a unique isomorphism, we obtain an isomorphism $Q^{\C M_{n}(S)}\simeq !_{n}Q^{S}$.
-%
-%We define morphisms $h: !_{n}Q^{S}\to Q^{\C M_{n}(S)}$ and 
-%$k: Q^{\C M_{n}(S)}\to !_{n}Q^{S}$ given by 
+$\Lawv^{S\times \dots \times S}\simeq (\Lawv^{S})^{\otimes_{n}}$, and then, since equalizers are unique up to a unique isomorphism, we obtain an isomorphism $\Lawv^{\C M_{n}(S)}\simeq !_{n}\Lawv^{S}$.
+%
+%We define morphisms $h: !_{n}\Lawv^{S}\to \Lawv^{\C M_{n}(S)}$ and 
+%$k: \Lawv^{\C M_{n}(S)}\to !_{n}\Lawv^{S}$ given by 
 %\begin{align*}
-%h([\B x_{1},\dots, \B x_{n}])(\{a_{1},\dots, a_{n}\})&=
+%h([x_{1},\dots, x_{n}])(\{a_{1},\dots, a_{n}\})&=
 %\sup_{\sigma\in \F S_{n}}\sum_{i=1}^{n}
-%\B x_{i}(a_{\sigma(i)})\\
+%x_{i}(a_{\sigma(i)})\\
 %k(f)& = 
-%\bigvee_{a_{1},\dots, a_{n}\in S}[\C Y(a_{1}),\dots, \C Y(a_{n})]+ f(\{a_{1},\dots, a_{n}\})
+%\bigvee_{a_{1},\dots, a_{n}\in S}[\B Y(a_{1}),\dots, \B Y(a_{n})]+ f(\{a_{1},\dots, a_{n}\})
 %\end{align*}
@@
 There exist many equivalent way to describe linear differential structure over symmetric monoidal categories with biproducts.
-Here we chose the approach via bialgebra modalities (see Ehrhard's stuff, but also Lemay and company's stuff for a clearer picture).
+Here we chose the approach via bialgebra modalities (see \cite{Lemay2017, Blute2019}).
 
@@
 
-\item two natural transformations $\Delta:!X\to !X\otimes !X$ and $e:!X\to \B 1$ such that $(!X, \Delta, e)$ is a cocommutative comonoid, that is the following equations hold:
+\item two natural transformations $\Delta:!X\to !X\otimes !X$ and $e:!X\to \{\star\}$ such that $(!X, \Delta, e)$ is a cocommutative comonoid, that is the following equations hold:
 \begin{align}
@@
 
-\item two natural transformations $\nabla: !X\otimes !X \to !X$ and $u:\B 1\to !X$ such that $(!X,\nabla, u)$ is a commutative monoid, that is, the following equations hold:
+\item two natural transformations $\nabla: !X\otimes !X \to !X$ and $u:\{\star\}\to !X$ such that $(!X,\nabla, u)$ is a commutative monoid, that is, the following equations hold:
 \begin{align}
@@
 
-\begin{definition}
+\begin{definition}[cf.~\cite{Lemay2017}]
 A \emph{codereliction} for a bialgebra modality $(!,\delta,\epsilon,\Delta, e,\nabla,u)$ is a natural transformation $\eta:X\to !X$ satisfying the following equations:
@@
  $\BB C$ is a \emph{monoidal storage category} if it has a coalgebra modality satisfying the \emph{Seely isomorphisms}, that is, the maps
- $e:!\top \to \B 1$ and $\chi:!(X\times Y) \to !X\otimes !Y$, with $\chi=\Delta \circ  !(\pi_{0})\otimes!(\pi_{1})$, are isomorphisms (whence $!\top\simeq \B 1$ and $!(X\times Y)\simeq !X\otimes !Y$).
+ $e:!\top \to \{\star\}$ and $\chi:!(X\times Y) \to !X\otimes !Y$, with $\chi=\Delta \circ  !(\pi_{0})\otimes!(\pi_{1})$, are isomorphisms (whence $!\top\simeq \{\star\}$ and $!(X\times Y)\simeq !X\otimes !Y$).
 \end{definition}
@@
 We use the following result:
-\begin{theorem}[Lemay\&co]\label{theorem:lemay}
+\begin{theorem}[cf.~\cite{Lemay2017}]\label{theorem:lemay}
 Every additive symmetric monoidal category with an additive bialgebra modality and finite products satisfies the Seely isomorphisms.
@@
 \begin{theorem}
-For any linear differential category $\BB C$, the co-Kleisli category $\BB C_{!}$ is a cartesian closed differential category, with deriving transformation $\mathsf{D}(f)$ defined as follows:
+For any linear differential category $\BB C$, the co-Kleisli category $\BB C_{!}$ is a cartesian closed differential category, with deriving transformation $Df$ defined as follows:
 $$
@@
 
-\subsubsection{The Free Exponential Modality of $Q\mathsf{Mod}$}
-
-Using the recipe from [Melli\'es, Tabareau, Tasson 2010], together with Proposition \ref{prop:productvstensor}, the free exponential modality of $Q\mathsf{Mod}$ can be defined as
+\subsubsection{The Free Exponential Modality of $\Mod$}
+
+Using the recipe from \cite{Mellies2018}, together with Proposition \ref{prop:productvstensor}, the free exponential modality of $\Mod$ can be defined as
 $$
@@
 
-Finally, let us check the compatibility of $\epsilon$ and $\nabla$, which in $Q\mathsf{Mod}$ reads as
+Finally, let us check the compatibility of $\epsilon$ and $\nabla$, which in $\Mod$ reads as
 $ \epsilon\circ \nabla=(\epsilon \otimes e)\vee (e\otimes \epsilon)$:
@@
 \begin{theorem}
-$Q\mathsf{Mod}$ (equivalently, $Q\mathsf{CCat}$) is a linear differential category. Hence $Q\mathsf{Mod}_{!}$ (equivalently, $Q\mathsf{CCat}_{!}$) is a cartesian closed differential category. 
+$\Mod$ (equivalently, $\GMet$) is a linear differential category. Hence $\Mod_{!}$ (equivalently, $\GMet_{!}$) is a cartesian closed differential category. 
 \end{theorem}
-
-Let us compute the differential operator in $Q\mathsf{Mod}_{!}$: 
-given $f:!X \to Y$, we have
-\begin{align*}
-\mathsf{D}[f](\alpha) & = 
-\bigvee 
-\left \{
-f(\beta\cup[x])
-\ \Big \vert \ 
-\iota_{n}(\beta)\otimes \iota_{1}(x) \leq \alpha^{S}
-\right\}
-\end{align*}
+%
+%Let us compute the differential operator in $\Mod_{!}$: 
+%given $f:!X \to Y$, we have
+%\begin{align*}
+%E[f](\alpha) & = 
+%\bigvee 
+%\left \{
+%f(\beta\cup[x])
+%\ \Big \vert \ 
+%\iota_{n}(\beta)\otimes \iota_{1}(x) \leq \alpha^{S}
+%\right\}
+%\end{align*}
 
@@
 %A coalgebra modality $(!,\delta,\epsilon,\Delta,e)$ is a \emph{monoidal coalgebra modality} if there exist natural transformations $m_{\otimes}:!X\otimes !Y \to !(X\otimes Y)$ 
-%and $m_{1}: \B 1 \to !\B 1$ such that 
+%and $m_{1}: \{\star\} \to !\{\star\}$ such that 
 %\begin{itemize}
@@
 %
-
-\newpage
-
-\subsection{$Q$-Rel is a model of Bounded Linear Logic}
-
-Given SMCs $\BB C,\BB D$,  let $\B{SMC}_{l}(\BB C, \BB D)$ indicate the category of symmetric lax monoidal functors and monoidal natural transformations between them.
-$\B{SMC}_{l}(\BB C, \BB D)$  is itself a SMC, with monoidal structure defined pointwise.
-
-The set 
-$\BB N$ can be seen as the category with objects the natural numbers and a morphism between $r$ and $r'$ precisely when $r\leq r'$. 
-Moreover, $\BB N$ can be seen as a SMC in two ways:
-\begin{itemize}
-
-\item we indicate as $\BB N^{+}$ the SMC with monoidal product given by addition;
-\item we indicate as $\BB N^{\times}$ the SMC with monoidal product given by multiplication.
-\end{itemize}
-
-%Below we let $\BB^{\min}$ indicate the monoid $(Q,\min,0)$ and $Q^{+}$ indicate the monoid $(Q,+,0)$.
-
-\begin{definition}
-A \emph{$\BB N$-graded linear exponential comonad} on a symmetric monoidal category $\BB C$ is a tuple
-$(D, w,c,\epsilon,\delta)$ where:
-\begin{itemize}
-
-\item $D: \BB N\to \B{SMC}_{l}(\BB C, \BB C)$ is a functor. We write 
-$m_{r}:\B 1 \to D(r)(\B 1)$ and $m_{r,A,B}: D(r)(A)\otimes D(r)(B) \to D(r)(A\otimes B)$ for the symmetric lax monoidal structure of $D(r)$;
-
-\item $(D,w,c): \BB N^{+}\to \B{SMC}_{l}(\BB C, \BB C)$ is a symmetric colax monoidal functor;
-
-\item $(D,\epsilon,\delta):\BB N^{\times}\to (\B{SMC}_{l},\mathrm{Id},\circ) $ is a colax monoidal functor.
-
-\end{itemize}
-further satisfying the axioms below:
-\begin{align}
-w_{A}& =  w_{D(s)(A)}\circ \delta_{0,s,A}\\
-w_{A} & = D(s)(w_{A} )\circ \delta_{s,0,A} \\
-(\delta_{r,s,A}\otimes \delta_{r',s,A})\circ c_{rs,r's,A}
-&=
-c_{r,r',D(s)(A)}\circ \delta_{r+r',s,A}\\
-m_{s,D(r)(A),D(r')(A)}\circ (\delta_{r,s,A}\otimes \delta_{s,r',A})\circ c_{sr,sr',A}&=
-D(s)(c_{r,r',A})\circ \delta_{s,r+r',A}
-\end{align}
-\end{definition}
-
-Concretely, the definition above requires 6 natural transformations:
-\begin{align*}
-m_{r} & :\B 1\to  D(r)(\B 1)\\
-m_{r,A,B}& :  D(r)(A)\otimes D(r)(B)\to  D(r)(A\otimes B)\\
-w_{A}& : D(0)(A)\to \B 1 \\
-c_{r,r',A}& : D(r+r')(A) \to D(r)(A)\otimes D(r')(A) \\
-\epsilon_{A}& : D(1)(A)\to A \\
-\delta_{r,r',A}&: D(r r')(A)\to D(r)(D(r')(A))
-\end{align*}
-subject to the following list of equations:
-\begin{itemize}
-\item $D(r)$ is a lax monoidal functor:
-\begin{align}
-m_{r,A\otimes B,C}\circ (m_{r,A,B}\otimes D(r)(C)) & = 
-m_{r,A, B\otimes C}\circ (D(r)(A)\otimes m_{r,B,C})\\
-m_{r,A,\B 1}\circ (D(r)(A)\otimes m_{r}) & = D(r)(A)\\
-m_{r,\B 1, B}\circ (m_{r}\otimes D(r)(B))&= D(r)(B)
-\end{align}
-
-\item $(D,w,c)$ is a symmetric colax monoidal functor:
-\begin{align}
-(c_{r,s,-}\otimes D(t)(-))\circ c_{r+s,t} & =
-(D(r)(-)\otimes c_{s,t,-})\circ c_{r,s+t}\\
-(D(r)(-)\otimes w_{-})\circ c_{r,0,-} & = D(r)(-) \\
-(w_{-}\otimes D(r)(-))\circ c_{0,r,-} & = D(r)(-)
-\end{align}
-
-\item $(D,\epsilon,\delta)$ is a colax monoidal functor:
-\begin{align}
-\delta_{r,s, D(t)(-)}\circ \delta_{(rs),t,-} & =
-D(r)(\delta_{s,t,-})\circ \delta_{r,st,-}\\
-D(r)(\epsilon_{-}) \circ \delta_{r,1,-} & = D(r)(-) \\
-\epsilon_{D(r)(-)} \circ \delta_{1,r,-} & = D(r)(-)
-\end{align}
-
-\end{itemize}
-
-\begin{definition}
-We define the following structure $(M,w,c,\epsilon,\delta)$ over the category $Q\mathsf{Rel}$ as follows:
-\begin{itemize}
-\item for any set $X$ and $n\in \BB N$, let $M(n)(X)=\C M_{\leq n}(X)$;
-
-\item for all $f: X\times Y\to Q$, let $M(n)(f): M(n)(X)\times M(n)(Y)\to Q$ be defined by 
-\begin{align*}
-M(n)(f)(\alpha,\beta)=
-\begin{cases}
-\min_{\sigma\in \F S_{k}}\sum_{i=1}^{k}f(x_{i},y_{\sigma(i)}) & 
-\text{ if }\alpha=[x_{1},\dots, x_{k}], \beta=[y_{1},\dots, y_{k}]\\
-\infty & \text{ otherwise}
-\end{cases}
-\end{align*}
-
-\item $m_{r}(\star, \{\star\})=0$ and $m_{r}(\star, \emptyset)=\infty$;
-
-\item $m_{r,A,B}: D(r)(A)\times D(r)(B)\times D(r)(A\times B)\to Q$ is defined by 
-\begin{align*}
-m_{r,A,B}((\alpha,\beta), \gamma)=
-\begin{cases}
-0 & \text{ if } \alpha=[x_{1},\dots, x_{k}], \beta=[y_{1},\dots, y_{k}], \gamma= [(x_{1},y_{1}),\dots, (x_{k},y_{k})]\\
-\infty & \text{ otherwise}
-\end{cases}
-\end{align*}
-
-\item $w_{A}:D(0)(A)\times \B 1\to Q$ is given by $w_{A}(\emptyset, \star)=0$ and is $\infty$ otherwise (observe that $D(0)(A)\simeq \B 1$);
-
-\item $c_{r,s,A}: D(r+s)(A)\times D(r)(A)\times D(s)(A)\to Q$ is given by $c_{r,r',A}(\langle\alpha, \beta,\gamma\rangle)=0$ if $\alpha=\beta+\gamma$, and is $\infty$ otherwise;
-
-\item $\epsilon_{A}(\emptyset, a)=\infty$, $\epsilon_{A}([a],a)=0$, $\epsilon_{A}([b],a)=\infty$ $(b\neq a)$,
-
-\item $\delta_{r,r',A}(\alpha, B)=0$ if $\alpha= \sum B$ (where $\sum B$ indicates the multiset obtained by the sum of all multisets contained in $B$) and is $\infty$ otherwise.
-
-\end{itemize}
-
-\end{definition}
-
-Let us check that this defines a $\BB N$-graded linear exponential comonad.
-
-\begin{itemize}
-
-\item $D(r)$ is a lax monoidal functor:
- $$ m_{r,A\times B,C}\circ (m_{r,A,B}\times D(r)(C))(\langle \alpha,\beta,\gamma,\delta\rangle)
- :
- D(r)(A)\times D(r)(B)\times D(r)(C) \times D(r)(A\times B\times C)\to Q
- $$
- is equal to $0$ 
-precisely when $\alpha=[x_{1},\dots, x_{k}]$, $\beta=[y_{1},\dots, y_{k}]$, $\gamma=[z_{1},\dots, z_{k}]$ and 
-$\delta= [(x_{1},y_{1},z_{1}),\dots, (x_{k},y_{k},z_{k})]$, and is $\infty $ in all other cases.
-
-Observe that
-$m_{r,A,B\times C}\circ (D(r)(A)\times m_{r,B,C})(\langle\alpha,\beta,\gamma, \delta\rangle)
-  $ is equal to $0$ in the same situation, and is $\infty$ otherwise.
-
- We conclude that the two matrices coincide.
-
- Furthermore, we have that 
- $m_{r,A,\B 1}\circ (D(r)(A)\times m_{r})(\langle \alpha,\beta\rangle): D(r)(A)\times \B 1 \times D(r)(A)$ is equal to $0$ 
- precisely when $\alpha=\beta$ and is $\infty$ otherwise, that is, it coincides with $\mathrm{id}_{D(r)(A)}$.
-
-\item $(D,w,c)$ is a symmetric colax monoidal functor.
-
-$((c_{r,s,A}\times D(t)(A))\circ c_{r+s,t,A}) (\langle \alpha,\beta,\gamma,\delta\rangle)
-: D(r+s+t)(A)\times D(r)(A)\times D(s)(A)\times D(t)(A)$
-is equal to $0$ when $\alpha=\beta+\gamma+\delta$, and is $\infty$ otherwise, and the same holds for
-$((D(r)(A)\times c_{s,t,A})\circ c_{r,s+t,A}) (\langle \alpha,\beta,\gamma,\delta\rangle)
-$.
-
-Furthermore, 
-$((D(r)(A)\times w_{A})\circ c_{r,0,A})(\alpha,\beta)
-: D(r)(A)\times D(r)(A)\to Q$ is equal to $0$ when 
-$\alpha=\beta$, and is $\infty$ otherwise, so it coincides with 
-$\mathrm{id}_{D(r)(A)}$.
-
-\item $(D,\epsilon,\delta)$ is a colax monoidal functor:
-
-$(\delta_{r,s,D(t)(A)}\circ \delta_{rs,t,A})
-(\alpha, \Phi)
-: D(rst)(A) \times D(r)(D(s)(D(t)(A))) \to Q
-$
-is $0$ precisely when $\alpha = \sum \sum \Phi$, and is $\infty$ otherwise, and similarly for 
-$(D(r)(\delta_{s,t,A})\circ \delta_{r,st,A})(\alpha,\Phi)$.
-
-Furthermore, 
-$(D(r)(\epsilon_{A})\circ \delta_{r,1})( \alpha,\beta ):
-D(r)(A) \times D(r)(A)\to Q
-$ is equal to $0$ when $\alpha=\beta$ and is $\infty$ otherwise, so it coincides with $\mathrm{id}_{D(r)(A)}$.
-
-\end{itemize}
-
-Let us check the further equations:
-\begin{itemize}
-
-\item $(w_{D(s)(A)}\circ \delta_{0,s,A})(\langle \emptyset,\star\rangle: D(0)(A)\times \B 1\to Q$ is 0, precisely like $w_{A}$.
-
-\item A similar argument holds for the second equation.
-
-\item $((\delta_{r,s,A}\times \delta_{r',s,a})\circ c_{rs,r's,A})
-(\langle \alpha, \Phi,\Psi  \rangle)
-:
-D(rs+r's)(A)\times  D(r)(s)(A)\times D(r')(s)(A)\to Q
-$
-is equal to $0$ when $\alpha=\sum \Phi + \sum \Psi$, and is $\infty$ otherwise.
-
-Now, 
-using the fact that $D(rs+r's)(A)=D((r+r')s)(A)$, we can check that the same holds for 
-$c_{r,r',D(s)(A)}\circ \delta_{r+r',s,A})(\langle \alpha, \Phi,\Psi  \rangle)$: it is $0$ when 
-$\alpha= \sum\Phi+\Psi= \sum \Phi+\sum \Psi$.
-
-\item A similar argument holds for the fourth equation.
-
-\end{itemize}

### 2023-01-19 appIII.tex
commit

@@
 \subsubsection{The language $\BSTLC$}
-TO BE ADDED
+In Section 4 we considered a variant $\BSTLC$ of $\STLC$ with affine simple types and an exponential $!_{r}A$, for all $r\in \BB R_{\geq 0}$, corresponding to a fragment of $\mathrm{Fuzz}$ \cite{Reed_2010}.
+We describe the rules of this fragment in Fig.~\ref{fig:fuzz}.
+The terms are generated by the grammar:
+
+{
+\begin{minipage}{\textwidth}
+\begin{align*}
+t,u:= x\mid \lambda x.t\mid tu \mid !t \mid \mathsf{let} !x=t\mathsf{in} u\mid (t,u)\mid  \mathsf{let} (x,y)=t\mathsf{in} u 
+\end{align*}\end{minipage}}\medskip\\
+For the purposes of this article we limited ourselves to a minimal fragment of this language. For a more practical language see \cite{Reed_2010,Gaboardi2017}.  
+Simple types are generated by the grammar below:
+
+{
+\begin{minipage}{\textwidth}
+\begin{align*}
+A,B:= X\mid !_{r}A \quad (r\in \BB R_{\geq 0}) \mid A\multimap B \mid A\otimes B
+\end{align*}\end{minipage}}\medskip\\
+Type judgements are of the form $\Gamma \vdash t:A$, where a context $\Gamma$ is a list of declarations of the form $x\in_{r}A$, for some $r\in \BB R_{\geq 0}$.
+We define the following operation $\Gamma+\Delta$  as follows:
+
+{
+\begin{minipage}{\textwidth}
+\begin{align*}
+() + () & =() \\
+(\Gamma, x\in_{r} A)+( \Delta, x\in_{s} A) & =  (\Gamma+\Delta), x\in_{r+s}A \\
+(\Gamma, x\in_{r}A)+\Delta & =(\Gamma+\Delta), x\in_{r}A \qquad (x\notin \Delta) \\
+\Gamma+ (\Delta, x\in_{r}A) &= (\Gamma+\Delta), x\in_{r} A \qquad (x\notin \Gamma)
+\end{align*}\end{minipage}}\medskip\\
+Moreover, we let $s\Gamma$ be the context made all judgmenets $x\in_{sr}A$, where $(x\in_{r}A)\in \Gamma$.
+
+Observe that one can always type an affine term like e.g.~$\lambda xy.x$ with a linear type $A\multimap B\multimap A$.
+Instead, a term like $\lambda xy.x(xy)$ containing two occurrences of $x$ cannot be given the linear type $(A\multimap A)\multimap (A\multimap A)$ but a type of the form
+$!_{2}(A\multimap A)\multimap (A\multimap A)$.
+
+\begin{figure}
+\fbox{
+\begin{minipage}{0.9\textwidth}
+\begin{center}
+
+  {
+\AXC{$x\in_{r} A\in \Gamma$}
+\UIC{$\Gamma\vdash x\in A$}
+\DP}
+
+\bigskip
+
+\begin{tabular}{c c }
+  {
+\AXC{$\Gamma, x\in_{1} A\vdash t\in B$}
+\UIC{$\Gamma\vdash \lambda x.t\in A\multimap B$}
+\DP}
+
+&
+
+  {
+\AXC{$\Gamma \vdash t\in A\multimap B$}
+\AXC{$\Gamma\vdash u\in A$}
+\BIC{$\Gamma \vdash tu\in B$}
+\DP}
+\\
+
+&
+
+\\
+
+  {
+\AXC{$\Gamma \vdash t\in A$}
+\AXC{$\Delta \vdash u\in B$}
+\BIC{$\Gamma+\Delta \vdash ( t,u) \in A\otimes B$}
+\DP}
+
+&
+
+  {
+\AXC{$\Gamma\vdash t\in A \otimes B$}
+\AXC{$\Delta,x\in_{r}A, y\in_{r} B\vdash u:C$}
+\BIC{$\Gamma+\Delta \vdash \mathsf{let} (x,y)=t \mathsf{in} u\in C$}
+\DP}
+\\
+&
+
+\\
+
+  {
+\AXC{$\Gamma \vdash t\in A$}
+\UIC{$s\Gamma \vdash !t \in \ !_{s}A$}
+\DP}
+
+&
+
+  {
+\AXC{$\Gamma\vdash t\in !_{s}A$}
+\AXC{$\Delta,x\in_{rs}A\vdash u:C$}
+\BIC{$r\Gamma+\Delta \vdash \mathsf{let} !x=t \mathsf{in} u\in C$}
+\DP}
+
+\end{tabular}
+\end{center}
+\end{minipage}
+}
+\caption{Typing rules for $\BSTLC$.}
+\label{fig:fuzz}
+\end{figure}
 
@@
 \begin{definition}
-We define the following structure $(M,w,c,\epsilon,\delta)$ over the category $Q\mathsf{Rel}$ as follows:
+We define the following structure $(M,w,c,\epsilon,\delta)$ over the category $\LREL$ as follows:
 \begin{itemize}
@@
 
-\item for all $f: X\times Y\to Q$, let $M(n)(f): M(n)(X)\times M(n)(Y)\to Q$ be defined by 
+\item for all $f: X\times Y\to \Lawv$, let $M(n)(f): M(n)(X)\times M(n)(Y)\to \Lawv$ be defined by 
 \begin{align*}
@@
 
-\item $m_{r,A,B}: D(r)(A)\times D(r)(B)\times D(r)(A\times B)\to Q$ is defined by 
+\item $m_{r,A,B}: D(r)(A)\times D(r)(B)\times D(r)(A\times B)\to \Lawv$ is defined by 
 \begin{align*}
@@
 
-\item $w_{A}:D(0)(A)\times \{\star\}\to Q$ is given by $w_{A}(\emptyset, \star)=0$ and is $\infty$ otherwise (observe that $D(0)(A)\simeq \{\star\}$);
+\item $w_{A}:D(0)(A)\times \{\star\}\to \Lawv$ is given by $w_{A}(\emptyset, \star)=0$ and is $\infty$ otherwise (observe that $D(0)(A)\simeq \{\star\}$);
 
-\item $c_{r,s,A}: D(r+s)(A)\times D(r)(A)\times D(s)(A)\to Q$ is given by $c_{r,r',A}(\langle\alpha, \beta,\gamma\rangle)=0$ if $\alpha=\beta+\gamma$, and is $\infty$ otherwise;
+\item $c_{r,s,A}: D(r+s)(A)\times D(r)(A)\times D(s)(A)\to \Lawv$ is given by $c_{r,r',A}(\langle\alpha, \beta,\gamma\rangle)=0$ if $\alpha=\beta+\gamma$, and is $\infty$ otherwise;
 
@@
  :
- D(r)(A)\times D(r)(B)\times D(r)(C) \times D(r)(A\times B\times C)\to Q
+ D(r)(A)\times D(r)(B)\times D(r)(C) \times D(r)(A\times B\times C)\to \Lawv
  $$
@@
 $((D(r)(A)\times w_{A})\circ c_{r,0,A})(\alpha,\beta)
-: D(r)(A)\times D(r)(A)\to Q$ is equal to $0$ when 
+: D(r)(A)\times D(r)(A)\to \Lawv$ is equal to $0$ when 
 $\alpha=\beta$, and is $\infty$ otherwise, so it coincides with 
@@
 $(\delta_{r,s,D(t)(A)}\circ \delta_{rs,t,A})
-(\alpha, \Phi)
-: D(rst)(A) \times D(r)(D(s)(D(t)(A))) \to Q
+(\alpha, \Gamma)
+: D(rst)(A) \times D(r)(D(s)(D(t)(A))) \to \Lawv
 $
-is $0$ precisely when $\alpha = \sum \sum \Phi$, and is $\infty$ otherwise, and similarly for 
-$(D(r)(\delta_{s,t,A})\circ \delta_{r,st,A})(\alpha,\Phi)$.
+is $0$ precisely when $\alpha = \sum \sum \Gamma$, and is $\infty$ otherwise, and similarly for 
+$(D(r)(\delta_{s,t,A})\circ \delta_{r,st,A})(\alpha,\Gamma)$.
 
@@
 $(D(r)(\epsilon_{A})\circ \delta_{r,1})( \alpha,\beta ):
-D(r)(A) \times D(r)(A)\to Q
+D(r)(A) \times D(r)(A)\to \Lawv
 $ is equal to $0$ when $\alpha=\beta$ and is $\infty$ otherwise, so it coincides with $\mathrm{id}_{D(r)(A)}$.
@@
 
-\item $(w_{D(s)(A)}\circ \delta_{0,s,A})(\langle \emptyset,\star\rangle: D(0)(A)\times \{\star\}\to Q$ is 0, precisely like $w_{A}$.
+\item $(w_{D(s)(A)}\circ \delta_{0,s,A})(\langle \emptyset,\star\rangle: D(0)(A)\times \{\star\}\to \Lawv$ is 0, precisely like $w_{A}$.
 
@@
 \item $((\delta_{r,s,A}\times \delta_{r',s,a})\circ c_{rs,r's,A})
-(\langle \alpha, \Phi,\Psi  \rangle)
+(\langle \alpha, \Gamma,\Delta  \rangle)
 :
-D(rs+r's)(A)\times  D(r)(s)(A)\times D(r')(s)(A)\to Q
+D(rs+r's)(A)\times  D(r)(s)(A)\times D(r')(s)(A)\to \Lawv
 $
-is equal to $0$ when $\alpha=\sum \Phi + \sum \Psi$, and is $\infty$ otherwise.
+is equal to $0$ when $\alpha=\sum \Gamma + \sum \Delta$, and is $\infty$ otherwise.
 
@@
 using the fact that $D(rs+r's)(A)=D((r+r')s)(A)$, we can check that the same holds for 
-$c_{r,r',D(s)(A)}\circ \delta_{r+r',s,A})(\langle \alpha, \Phi,\Psi  \rangle)$: it is $0$ when 
-$\alpha= \sum\Phi+\Psi= \sum \Phi+\sum \Psi$.
+$c_{r,r',D(s)(A)}\circ \delta_{r+r',s,A})(\langle \alpha, \Gamma,\Delta  \rangle)$: it is $0$ when 
+$\alpha= \sum\Gamma+\Delta= \sum \Gamma+\sum \Delta$.
 

### 2023-01-19 appIII.tex
commit

@@
 \subsubsection{The language $\BSTLC$}
-In Section 4 we considered a variant $\BSTLC$ of $\STLC$ with affine simple types and an exponential $!_{r}A$, for all $r\in \BB R_{\geq 0}$, corresponding to a fragment of $\mathrm{Fuzz}$ \cite{Reed_2010}.
+In Section 4 we considered a variant $\BSTLC$ of $\STLC$ with affine simple types and an exponential $!_{n}A$, for all $n\in \BB N$, corresponding to a fragment of $\mathrm{Fuzz}$ \cite{Reed2010}. 
 We describe the rules of this fragment in Fig.~\ref{fig:fuzz}.
@@
 \begin{align*}
-t,u:= x\mid \lambda x.t\mid tu \mid !t \mid \mathsf{let} !x=t\mathsf{in} u\mid (t,u)\mid  \mathsf{let} (x,y)=t\mathsf{in} u 
+t,u:= x\mid \lambda x.t\mid tu \mid !t \mid \mathsf{let}\  !x=t \ \mathsf{in}\  u\mid (t,u)\mid  \mathsf{let}\  (x,y)=t \ \mathsf{in} \  u 
 \end{align*}\end{minipage}}\medskip\\
-For the purposes of this article we limited ourselves to a minimal fragment of this language. For a more practical language see \cite{Reed_2010,Gaboardi2017}.  
+For the purposes of this article we limited ourselves to a minimal fragment of this language. For a more practical language see \cite{Reed2010,Gaboardi2017}.  
 Simple types are generated by the grammar below:
@@
 \begin{align*}
-A,B:= X\mid !_{r}A \quad (r\in \BB R_{\geq 0}) \mid A\multimap B \mid A\otimes B
+A,B:= X\mid  !_{n}A  \mid A\multimap B \mid A\otimes B
 \end{align*}\end{minipage}}\medskip\\
-Type judgements are of the form $\Gamma \vdash t:A$, where a context $\Gamma$ is a list of declarations of the form $x\in_{r}A$, for some $r\in \BB R_{\geq 0}$.
+Type judgements are of the form $\Gamma \vdash t:A$, where a context $\Gamma$ is a list of declarations of the form $x\in_{n}A$.
 We define the following operation $\Gamma+\Delta$  as follows:
@@
 () + () & =() \\
-(\Gamma, x\in_{r} A)+( \Delta, x\in_{s} A) & =  (\Gamma+\Delta), x\in_{r+s}A \\
-(\Gamma, x\in_{r}A)+\Delta & =(\Gamma+\Delta), x\in_{r}A \qquad (x\notin \Delta) \\
-\Gamma+ (\Delta, x\in_{r}A) &= (\Gamma+\Delta), x\in_{r} A \qquad (x\notin \Gamma)
+(\Gamma, x\in_{m} A)+( \Delta, x\in_{n} A) & =  (\Gamma+\Delta), x\in_{m+n}A \\
+(\Gamma, x\in_{m}A)+\Delta & =(\Gamma+\Delta), x\in_{n}A \qquad (x\notin \Delta) \\
+\Gamma+ (\Delta, x\in_{m}A) &= (\Gamma+\Delta), x\in_{n} A \qquad (x\notin \Gamma)
 \end{align*}\end{minipage}}\medskip\\
-Moreover, we let $s\Gamma$ be the context made all judgmenets $x\in_{sr}A$, where $(x\in_{r}A)\in \Gamma$.
+Moreover, we let $m\Gamma$ be the context made all judgmenets $x\in_{mn}A$, where $(x\in_{n}A)\in \Gamma$.
 
@@
   {
-\AXC{$x\in_{r} A\in \Gamma$}
+\AXC{$x\in_{n} A\in \Gamma$}
 \UIC{$\Gamma\vdash x\in A$}
@@
 \AXC{$\Gamma\vdash t\in A \otimes B$}
-\AXC{$\Delta,x\in_{r}A, y\in_{r} B\vdash u:C$}
-\BIC{$\Gamma+\Delta \vdash \mathsf{let} (x,y)=t \mathsf{in} u\in C$}
+\AXC{$\Delta,x\in_{n}A, y\in_{n} B\vdash u:C$}
+\BIC{$n\Gamma+\Delta \vdash \mathsf{let}\  (x,y)=t \ \mathsf{in}  \ u\in C$}
 \DP}
@@
 \AXC{$\Gamma \vdash t\in A$}
-\UIC{$s\Gamma \vdash !t \in \ !_{s}A$}
+\UIC{$s\Gamma \vdash !t \in \ !_{n}A$}
 \DP}
@@
   {
-\AXC{$\Gamma\vdash t\in !_{s}A$}
-\AXC{$\Delta,x\in_{rs}A\vdash u:C$}
-\BIC{$r\Gamma+\Delta \vdash \mathsf{let} !x=t \mathsf{in} u\in C$}
+\AXC{$\Gamma\vdash t\in !_{n}A$}
+\AXC{$\Delta,x\in_{mn}A\vdash u:C$}
+\BIC{$m\Gamma+\Delta \vdash \mathsf{let}\  !x=t \ \mathsf{in} \ u\in C$}
 \DP}
@@
 \begin{definition}
-We define the following structure $(M,w,c,\epsilon,\delta)$ over the category $\LREL$ as follows:
+We define the following structure $(!_{-}(-),w,c,\epsilon,\delta)$ over the category $\LREL$ as follows:
 \begin{itemize}
-\item for any set $X$ and $n\in \BB N$, let $M(n)(X)=\C M_{\leq n}(X)$;
+\item for any set $X$ and $n\in \BB N$, let $!_{n}(X)=\C M_{\leq n}(X)$;
 
-\item for all $f: X\times Y\to \Lawv$, let $M(n)(f): M(n)(X)\times M(n)(Y)\to \Lawv$ be defined by 
+\item for all $f: X\times Y\to \Lawv$, let $!_{n}(f): !_{n}(X)\times !_{n}(Y)\to \Lawv$ be defined by 
 \begin{align*}
-M(n)(f)(\alpha,\beta)=
+!_{n}(f)(\alpha,\beta)=
 \begin{cases}
@@
 
-Let us check that this defines a $\BB N$-graded linear exponential comonad.
-
+\begin{proposition}
+ $(!_{-}(-),w,c,\epsilon,\delta)$  is a $\BB N$-graded linear exponential comonad over $\LREL$.
+\end{proposition}
+\begin{proof}
 \begin{itemize}
@@
 \end{itemize}
+\end{proof}

### 2023-01-19 appIV.tex
commit

@@
-We here focus on the category $Q^{\Pi}_{!}$, that is, the matricial category associated with $\mathsf{Trop}_{!}$. 
-Let us recall that a morphism of $Q^{\Pi}_{!}(X,Y)$ is a matrix $f\in Q^{\C M_{\mathsf f}(X)\times Y}$, and that composition of $f\in Q^{\Pi}_{!}(X,Y)$ and $g\in Q^{\Pi}_{!}(Y,Z)$ is given by 
-$$
-(g\circ f)_{\mu,z}= \inf\Big\{
-\sum_{i=1}^{n}f_{\mu_{i},y_{i}}+
-g_{[y_{1},\dots, y_{n}],z}
-\mid 
-n\in \BB N, 
-y_{1},\dots, y_{n}\in Y,
-\mu= \mu_{1}+\dots + \mu_{n}
-\Big\}
-$$
-
-Notice that $\infty \circ f= f\circ \infty=\infty$.
-However, while $\min\{g,h\}\circ f= \min\{g\circ f, h\circ f \}$ holds, as it can be computed using the continuity of $\mu, \inf$ and $+$, it is not true in general that
-$h\circ \min\{f,g\}= \min\{h\circ f, h\circ g\}$ (as a consequence of the fact that 
-$\sum_{i=1}^{n}\min\{ f_{\mu_{i},y_{i}}, g_{\mu_{i},y_{i}}\} \neq 
-\min\{ \sum_{i=1}^{n} f_{\mu_{i},y_{i}}, \sum_{i=1}^{n} g_{\mu_{i},y_{i}}\}$).
-In other words, morphisms in $Q^{\Pi}_{!}$ are only \emph{left-$Q$-additive}.
-A morphism $h\in Q^{\Pi}_{!}(X,Y)$ that satisfies $h\circ \min\{f,g\}=\min\{h\circ f, h\circ g\}$ for all object $X'$ and $f,g\in Q^{\Pi}_{!}(X',X)$, is called \emph{additive}.
-%We already know from Giulio's paper that $Q^{\Pi}_{!}$ is a post-linear continuous $Q$-ccc.
+%We already know from Giulio's paper that $\LREL_!$ is a post-linear continuous $\Lawv$-ccc.
 
 \begin{proposition}
-$Q^{\Pi}_{!}$ is a cartesian closed left-$Q$-additive category.
+$\LREL_!$ is a cartesian closed left-$\Lawv$-additive category.
 \end{proposition}
-\begin{proof} While left-$Q$-additivity was already proved, to show that $Q^{\Pi}_{!}$ is cartesian closed left-$Q$-additive we must also check that (1) products and projections of additive morphisms are additive, and that (2) $\Lambda(\min\{f,g\})=\min\{\Lambda(f),\Lambda(g)\}$, $\Lambda(\infty)=\infty$, where $\Lambda:Q^{\C M_{\mathsf f}(Z+X)\times Y}\to Q^{\C M_{\mathsf f}(Z)\times( \C M_{\mathsf f}(X)\times Y)}$ is the isomorphism given by
+\begin{proof} 
+Since $\infty \circ f= f\circ \infty=\infty$ and $\min\{g,h\}\circ f= \min\{g\circ f, h\circ f \}$, $\LREL_!$ is {left-$\Lawv$-additive}.
+A morphism $h\in \LREL_!(X,Y)$ that further satisfies $h\circ \min\{f,g\}=\min\{h\circ f, h\circ g\}$ for all object $X'$ and $f,g\in \LREL_!(X',X)$, is called \emph{additive}.
+To show that $\LREL_!$ is cartesian closed left-$\Lawv$-additive we must also check that (1) products and projections of additive morphisms are additive, and that (2) $\Lambda(\min\{f,g\})=\min\{\Lambda(f),\Lambda(g)\}$, $\Lambda(\infty)=\infty$, where $\Lambda:\Lawv^{\C M_{\mathsf{fin}}(Z+X)\times Y}\to \Lawv^{\C M_{\mathsf{fin}}(Z)\times( \C M_{\mathsf{fin}}(X)\times Y)}$ is the isomorphism given by
 $(\Lambda(f))_{\mu,\nu,y}= f_{\mu\oplus\nu,y}$, where 
@@
 	\begin{enumerate}
-	\item  Let $f\in Q^{\C M_{\mathsf f}(X)\times Y}$ and $g\in Q^{\C M_{\mathsf g}(X)\times Z}$ be additive; then $\langle f,g\rangle \in Q^{\C M_{\mathsf f}(X)\times (Y+Z)}$, which is defined by 
+	\item  Let $f\in \Lawv^{\C M_{\mathsf{fin}}(X)\times Y}$ and $g\in \Lawv^{\C M_{\mathsf{fin}}(X)\times Z}$ be additive; then $\langle f,g\rangle \in \Lawv^{\C M_{\mathsf{fin}}(X)\times (Y+Z)}$, which is defined by
 	$$
@@
 	$$
-	is also additive. Indeed, for all $h\in Q^{\Pi}_{!}(X',X)$, 
+	is also additive. Indeed, for all $h\in \LREL_!(X',X)$, 
 	in any cartesian category it holds that $\langle f,g\rangle \circ h=\langle f\circ h, g\circ h\rangle$. Now, 
-		if $i=0$, then for all $h_{1},h_{2}\in Q^{\C M_{\mathsf f}(X')\times X}$, 
+		if $i=0$, then for all $h_{1},h_{2}\in \Lawv^{\C M_{\mathsf{fin}}(X')\times X}$, 
 	\begin{align*}
@@
 
-	Moreover, suppose $f\in Q^{\Pi}_{!}(X, Y+Z)$ is additive, and let us show that $\pi_{1}( f)\in Q^{\Pi}_{!}(X,Y)$, defined by 
+	Moreover, suppose $f\in \LREL_!(X, Y+Z)$ is additive, and let us show that $\pi_{1}( f)\in \LREL_!(X,Y)$, defined by 
 	$
 	(\pi_{1}( f))_{\mu,y}= f_{\mu, \langle 0,y\rangle}
-	$, is also additive: first observe that $\pi_{1}(f)=\pi_{1}\circ f$, where $\pi_{1}\in Q^{\Pi}_{!}(Y+Z,Y)$ is given by $(\pi_{1})_{\mu,y}=0$ if $\mu=[y]\oplus \emptyset$ and is $\infty$ otherwise; moreover,
+	$, is also additive: first observe that $\pi_{1}(f)=\pi_{1}\circ f$, where $\pi_{1}\in \LREL_!(Y+Z,Y)$ is given by $(\pi_{1})_{\mu,y}=0$ if $\mu=[y]\oplus \emptyset$ and is $\infty$ otherwise; moreover,
 	$\pi_{1}(\min\{g,h\})=\min\{\pi_{1}(g),\pi_{1}(h)\}$, since 
 	$(\pi_{1}(\min\{g,h\}))_{\mu, y}=(\min\{g,h\})_{\mu, \langle 0,y\rangle}= \min\{g_{\mu, \langle 0,y\rangle},h_{\mu, \langle 0,y\rangle}\}=( \min\{\pi_{1}(g),\pi_{1}(h)\})_{\mu, y}$. Now,
-		given $h_{1},h_{2}\in Q^{\Pi}_{!}(X',X)$, we have that 
+		given $h_{1},h_{2}\in \LREL_!(X',X)$, we have that 
 	$\pi_{1}(f)\circ \min\{h_{1},h_{2}\}= (\pi_{1}\circ f)\circ \min\{h_{1},h_{2}\}= \pi_{1}\circ (f\circ \min\{h_{1},h_{2}\})= \pi_{1}\circ \min\{f\circ h_{1},f\circ h_{2}\}= \pi_{1}(\min\{f\circ h_{1},f\circ h_{2}\})= \min\{\pi_{1}(f\circ h_{1}), \pi_{1}(f\circ h_{2})\}= \min\{\pi_{1}(f)\circ h_{1},\pi_{1}(f)\circ h_{2})$.
@@
 
-For any morphism $f\in Q^{\Pi}_{!}(X,Y)$, let us define a morphism $\mathsf D(f)\in Q^{\Pi}_{!}(X+X,Y)$, i.e.~$\mathsf D(f)\in Q^{\C M_{\mathsf f}(X+X),Y}$, by
+For any morphism $f\in \LREL_!(X,Y)$, let us define a morphism $D(f)\in \LREL_!(X+X,Y)$, i.e.~$D(f)\in \Lawv^{\C M_{\mathsf{fin}}(X+X),Y}$, by
 $$
-\mathsf D(f)_{\mu,y}=
+D(f)_{\mu,y}=
 \begin{cases}
@@
 \begin{proposition}
-The category $Q^{\Pi}_{!}$, endowed with the operator $\mathsf D$, is a cartesian closed differential category.
+The category $\LREL_!$, endowed with the operator $D$, is a cartesian closed differential category.
 \end{proposition}
 \begin{proof}
-We must check axioms (1)-(7) of cartesian differential categories plus axiom (D-curry).
+We must check axioms (D1)-(D7) of cartesian differential categories  plus axiom (D-curry) (cf.~\cite{Manzo2012}).
 \begin{description}
-\item[(1)] $\mathsf D(\min\{f,g\})=\min \{\mathsf D(f),\mathsf D(g)\}$ and $\mathsf D(\infty)=\infty$: 
+\item[(D1)]$D(\min\{f,g\})=\min \{D(f),D(g)\}$ and $D(\infty)=\infty$: 
 while the latter is obvious, for the former we have 
-$\mathsf D(\min\{f,g\})_{[x]\oplus\nu,y}= \min\{f,g\}_{\nu+x,y}= \min\{f_{\nu+x,y},g_{\nu+x,y}\}=\min\{\mathsf D(f),\mathsf D(g)\}_{[x]\oplus \nu,y}$, and if $\mu\neq [x]\oplus \nu$, 
-$\mathsf D(\min\{f,g\})_{\mu,y}= \infty= \min\{\infty, \infty\}=\min\{ \mathsf D(f),\mathsf D(g)\}_{\mu,y}$. 
-\item[(2)]
-$\mathsf D(f)\circ \langle \min\{h,k\},v\rangle= \min\{ \mathsf D(f)\circ \langle h,v\rangle, \mathsf D(f)\circ \langle k,v\rangle\}$, and $\mathsf D(f)\circ \langle \infty,v\rangle=\infty$: we can compute
+$D(\min\{f,g\})_{[x]\oplus\nu,y}= \min\{f,g\}_{\nu+x,y}= \min\{f_{\nu+x,y},g_{\nu+x,y}\}=\min\{D(f),D(g)\}_{[x]\oplus \nu,y}$, and if $\mu\neq [x]\oplus \nu$, 
+$D(\min\{f,g\})_{\mu,y}= \infty= \min\{\infty, \infty\}=\min\{ D(f),D(g)\}_{\mu,y}$. 
+\item[(D2)]
+$D(f)\circ \langle \min\{h,k\},v\rangle= \min\{ D(f)\circ \langle h,v\rangle, D(f)\circ \langle k,v\rangle\}$, and $D(f)\circ \langle \infty,v\rangle=\infty$: we can compute
 	\begin{align*}
-	(\mathsf D(f)\circ \langle \min\{h,k\},v\rangle)_{\mu,y}&=
+	(D(f)\circ \langle \min\{h,k\},v\rangle)_{\mu,y}&=
 	\inf\Big\{ 
@@
 	\mu=\rho+\sum_{j=1}^{m}\nu_{j}\big\}\Big\}\\
-	& = \min\Big\{(\mathsf D(f)\circ \langle h,v\rangle)_{\mu,y}, (\mathsf D(f)\circ \langle k,v\rangle)_{\mu,y}\Big\}\\
-		& = \Big(\min\big\{\mathsf D(f)\circ \langle h,v\rangle, \mathsf D(f)\circ \langle k,v\rangle\big\}\Big )_{\mu,y}
+	& = \min\Big\{(D(f)\circ \langle h,v\rangle)_{\mu,y}, (D(f)\circ \langle k,v\rangle)_{\mu,y}\Big\}\\
+		& = \Big(\min\big\{D(f)\circ \langle h,v\rangle, D(f)\circ \langle k,v\rangle\big\}\Big )_{\mu,y}
 	\end{align*}
-	where, in the first equation, the condition $[w_{1},\dots, w_{n}]=[w]$ (i.e.~$n=1$) is forced by the fact that, otherwise, the application of $\mathsf D(f)$ would give $\infty$. Moreover, we have
+	where, in the first equation, the condition $[w_{1},\dots, w_{n}]=[w]$ (i.e.~$n=1$) is forced by the fact that, otherwise, the application of $D(f)$ would give $\infty$. Moreover, we have
 \begin{align*}
-(\mathsf D(f) \circ \langle \infty, v\rangle)_{\mu,y}&=\inf\Big\{ 
+(D(f) \circ \langle \infty, v\rangle)_{\mu,y}&=\inf\Big\{ 
 	\infty+
@@
 
-\item[(3)] $\mathsf D(\mathrm{id})=\pi_{1}$, $\mathsf D(\pi_{i})=\pi_{i}\circ \pi_{1}$: 
+\item[(D3)] $D(\mathrm{id})=\pi_{1}$, $D(\pi_{i})=\pi_{i}\circ \pi_{1}$: 
 recall that $\mathrm{id}_{[x],x}=0$ and $\mathrm{id}_{\mu,x}=\infty$, if $\mu\neq [x]$. 
@@
 Hence 
-$\mathsf D(\mathrm{id})_{[x]\oplus\nu,y}=\mathrm{id}_{\nu+x,y}$ is $0$ precisely when $x=y$ and $\nu=\emptyset$, and in all other cases is $\infty$. This shows that $\mathsf D(\mathrm{id})=\pi_{1}$.
+$D(\mathrm{id})_{[x]\oplus\nu,y}=\mathrm{id}_{\nu+x,y}$ is $0$ precisely when $x=y$ and $\nu=\emptyset$, and in all other cases is $\infty$. This shows that $D(\mathrm{id})=\pi_{1}$.
 
-$\mathsf D(\pi_{1})\in Q^{\C M_{\mathsf f}((X+Y)+(X+Y))\times Y}$ is given by
-$\mathsf D(\pi_{1})
+$D(\pi_{1})\in \Lawv^{\C M_{\mathsf{fin}}((X+Y)+(X+Y))\times Y}$ is given by
+$D(\pi_{1})
 _{
@@
 which is 0 precisely when $(\mu\oplus\nu)+\langle 0,x\rangle= y\oplus\emptyset$, i.e.~when 
-$x=y$ and $\mu=\nu=\emptyset$; in all other cases one can check that $\mathsf D(\pi_{1})_{\rho,y}=\infty$, so we conclude $\mathsf D(\pi_{1})=\pi_{1}\circ \pi_{1}$.
+$x=y$ and $\mu=\nu=\emptyset$; in all other cases one can check that $D(\pi_{1})_{\rho,y}=\infty$, so we conclude $D(\pi_{1})=\pi_{1}\circ \pi_{1}$.
 One can argue similarly for $\pi_{2}$.
 
-\item[(4)] $\mathsf D(\langle f,g\rangle)=\langle \mathsf D(f),\mathsf D(g)\rangle$: 
+\item[(D4)] $D(\langle f,g\rangle)=\langle D(f),D(g)\rangle$: 
 we have
 \begin{align*}
-\mathsf D(\langle f,g\rangle)_{[x]\oplus\mu, \langle 0,y\rangle}& = 
-(\langle f,g\rangle)_{\mu+x,\langle 0,y\rangle}= f_{\mu+x,y}= \mathsf D(f)_{[x]\oplus \mu,y}\\
-\mathsf D(\langle f,g\rangle)_{[x]\oplus\mu, \langle 1,y\rangle}& = 
-(\langle f,g\rangle)_{\mu+x,\langle 1,y\rangle}= g_{\mu+x,y}= \mathsf D(g)_{[x]\oplus \mu,y}
+D(\langle f,g\rangle)_{[x]\oplus\mu, \langle 0,y\rangle}& = 
+(\langle f,g\rangle)_{\mu+x,\langle 0,y\rangle}= f_{\mu+x,y}= D(f)_{[x]\oplus \mu,y}\\
+D(\langle f,g\rangle)_{[x]\oplus\mu, \langle 1,y\rangle}& = 
+(\langle f,g\rangle)_{\mu+x,\langle 1,y\rangle}= g_{\mu+x,y}= D(g)_{[x]\oplus \mu,y}
 \end{align*}
 from which we deduce
-$\mathsf D(\langle f,g\rangle)_{[x]\oplus\mu, \langle i,y\rangle}=\langle \mathsf D(f),\mathsf D(g)\rangle_{[x]\oplus\mu, \langle i,y\rangle}$ 
+$D(\langle f,g\rangle)_{[x]\oplus\mu, \langle i,y\rangle}=\langle D(f),D(g)\rangle_{[x]\oplus\mu, \langle i,y\rangle}$ 
  by the definition of $\langle \_,\_\rangle$.
  If $\rho\neq [x]\oplus\mu$, then
- $\mathsf D(\langle f,g\rangle)_{\rho, \langle i,y\rangle}=\infty=\langle \infty,\infty\rangle=\langle \mathsf D(f),\mathsf D(g)\rangle_{\rho, \langle i,y\rangle}$
+ $D(\langle f,g\rangle)_{\rho, \langle i,y\rangle}=\infty=\langle \infty,\infty\rangle=\langle D(f),D(g)\rangle_{\rho, \langle i,y\rangle}$
  (where the equation $\infty=\langle \infty,\infty\rangle$ is to be read as an equality between the functions $ X+Y\longrightarrow Q$
@@
 
-\item[(5)] $\mathsf D(f\circ g)=\mathsf D(f)\circ \langle \mathsf D(g), g\circ \pi_{2}\rangle$: we can compute
+\item[(D5)] $D(f\circ g)=D(f)\circ \langle D(g), g\circ \pi_{2}\rangle$: we can compute
 \begin{align*}
-\Big( \mathsf D(f)\circ \langle \mathsf D(g), g\circ \pi_{2}\rangle\Big)_{[x]\oplus\mu,z}
+\Big( D(f)\circ \langle D(g), g\circ \pi_{2}\rangle\Big)_{[x]\oplus\mu,z}
 &=
 \inf \Big\{
-\mathsf D(g)_{[x]\oplus\mu',w}+
+D(g)_{[x]\oplus\mu',w}+
 \sum_{i}g_{\mu_{i},w_{i}}+
-\mathsf D(f)_{[w]\oplus [w_{1},\dots, w_{n}],z}\\
+D(f)_{[w]\oplus [w_{1},\dots, w_{n}],z}\\
 &\qquad\qquad \mid
@@
 \\
-&= (f\circ g)_{\mu+x,y} =\mathsf D(f\circ g)_{[x]\oplus\mu,z}
+&= (f\circ g)_{\mu+x,y} =D(f\circ g)_{[x]\oplus\mu,z}
 \end{align*}
-if $\rho\neq [x]\oplus \mu$, then $\mathsf D(f\circ g)_{\rho,z}=\infty$ and 
+if $\rho\neq [x]\oplus \mu$, then $D(f\circ g)_{\rho,z}=\infty$ and 
 from the first equation above it follows that 
-also $( \mathsf D(f)\circ \langle \mathsf D(g), g\circ \pi_{2}\rangle)_{\rho,z}=\infty$.
+also $( D(f)\circ \langle D(g), g\circ \pi_{2}\rangle)_{\rho,z}=\infty$.
 
-\item[(6)] $\mathsf D(\mathsf D(f))\circ \langle \langle g,\infty\rangle,\langle h,k\rangle\rangle=\mathsf D(f)\circ  \langle g,k\rangle$:
+\item[(D6)] $D(D(f))\circ \langle \langle g,\infty\rangle,\langle h,k\rangle\rangle=D(f)\circ  \langle g,k\rangle$:
 observe that 
 \begin{align*}
-\Big(\mathsf D(\mathsf D(f))\Big)_{[\langle 1,x'\rangle]\oplus([x]\oplus \mu),z}&=
-\big(\mathsf D(f) \big)_{[x]\oplus (\mu+x'),z }= f_{\mu+x'+x,z}\\
-\Big(\mathsf D(\mathsf D(f))\Big)_{[\langle 0,x\rangle]\oplus (\emptyset \oplus \mu),z}&=
-\big(\mathsf D(f) \big)_{[x]\oplus \mu,z }= f_{\mu+x,z}
+\Big(D(D(f))\Big)_{[\langle 1,x'\rangle]\oplus([x]\oplus \mu),z}&=
+\big(D(f) \big)_{[x]\oplus (\mu+x'),z }= f_{\mu+x'+x,z}\\
+\Big(D(D(f))\Big)_{[\langle 0,x\rangle]\oplus (\emptyset \oplus \mu),z}&=
+\big(D(f) \big)_{[x]\oplus \mu,z }= f_{\mu+x,z}
 \end{align*}
-and in all other cases $(\mathsf D(\mathsf D(f)))_{\mu,z}=\infty$.
+and in all other cases $(D(D(f)))_{\mu,z}=\infty$.
 Using this fact we can compute:
@@
 \begin{align*}
-\Big(\mathsf D(\mathsf D(f))\circ \langle \langle g,\infty\rangle,\langle h,k\rangle\rangle\Big)_{\mu,z}&=
+\Big(D(D(f))\circ \langle \langle g,\infty\rangle,\langle h,k\rangle\rangle\Big)_{\mu,z}&=
 \min\left\{
@@
 \right\}\\
-&= \Big(\mathsf D(f)\circ \langle g,k\rangle\Big)_{\mu,z}
+&= \Big(D(f)\circ \langle g,k\rangle\Big)_{\mu,z}
 \end{align*}
@@
 
-\item[(7)] $\mathsf D(\mathsf D(f))\circ \langle\langle \infty,h\rangle, \langle g,k\rangle\rangle= \mathsf D(\mathsf D(f))\circ \langle \langle \infty,g\rangle, \langle h,k\rangle\rangle$:
+\item[(D7)] $D(D(f))\circ \langle\langle \infty,h\rangle, \langle g,k\rangle\rangle= D(D(f))\circ \langle \langle \infty,g\rangle, \langle h,k\rangle\rangle$:
 by computations similar to the case above we obtain
@@
 \begin{align*}
-\Big(\mathsf D(\mathsf D(f))\circ \langle \langle & \infty,h\rangle,\langle g,k\rangle\rangle\Big)_{\mu,z}\\
+\Big(D(D(f))\circ \langle \langle & \infty,h\rangle,\langle g,k\rangle\rangle\Big)_{\mu,z}\\
 %&=
@@
 &=
-\Big(\mathsf D(\mathsf D(f))\circ \langle \langle\infty,g\rangle,\langle h,k\rangle\rangle\Big)_{\mu,z}
+\Big(D(D(f))\circ \langle \langle\infty,g\rangle,\langle h,k\rangle\rangle\Big)_{\mu,z}
 \end{align*}
@@
 
-\item[(D-curry)] $\mathsf D(\Lambda(f))=
-\Lambda(\mathsf D(f)\circ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle)$:
+\item[(D-curry)]  \ \ \ \ \  $D(\Lambda(f))=
+\Lambda(D(f)\circ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle)$:
 by observing that both morphisms are 
-in $Q^{\Pi}_{!}(X+X, Z^{Y})= Q^{\C M_{\mathsf f}(X+X)\times \C M_{\mathsf f}(Y)\times Z}$, 
+in $\LREL_!(X+X, Z^{Y})= \Lawv^{\C M_{\mathsf{fin}}(X+X)\times \C M_{\mathsf{fin}}(Y)\times Z}$, 
 and that $ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle
-\in Q^{\Pi}_{!}( (X+X)+Y  , (X+Y)+(X+Y) )$,
+\in \LREL_!( (X+X)+Y  , (X+Y)+(X+Y) )$,
 we can compute:
@@
 \begin{align*}
-\big(\Lambda(\mathsf D(f)\circ \langle&  \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle)\big)_{[x]\oplus\mu,\nu,z}\\
+\big(\Lambda(D(f)\circ \langle&  \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle)\big)_{[x]\oplus\mu,\nu,z}\\
 &=
-\big(\mathsf D(f)\circ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle\big)_{([x]\oplus\mu)\oplus\nu, z}\\
+\big(D(f)\circ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle\big)_{([x]\oplus\mu)\oplus\nu, z}\\
 &=
@@
 \sum_{j}(\mathrm{id})_{[z_{j}],z_{j}}
- + \mathsf D(f)_{([x]\oplus \emptyset)\oplus(\mu\oplus \nu)}
+ + D(f)_{([x]\oplus \emptyset)\oplus(\mu\oplus \nu)}
 \mid
@@
 +
-0 + \mathsf D(f)_{([x]\oplus \emptyset)\oplus(\mu\oplus \nu)}
+0 + D(f)_{([x]\oplus \emptyset)\oplus(\mu\oplus \nu)}
 \mid
@@
 &=
-\big(\mathsf D(f)\big)_{ ([x]\oplus\emptyset)\oplus(\mu\oplus\nu),z}\\
+\big(D(f)\big)_{ ([x]\oplus\emptyset)\oplus(\mu\oplus\nu),z}\\
 &=
@@
 & =
-\big(\Lambda(f)\big)_{\mu+x,\nu,z}=\big(\mathsf D(\Lambda(f))\big)_{[x]\oplus\mu, \nu,z}
+\big(\Lambda(f)\big)_{\mu+x,\nu,z}=\big(D(\Lambda(f))\big)_{[x]\oplus\mu, \nu,z}
 \end{align*}
@@
 
-If $\rho\neq [x]\oplus\mu$, then $(\mathsf D(\Lambda(f)))_{\rho,\nu,z}=\infty$ and $
-(\Lambda(\mathsf D(f)\circ \langle  \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle))_{\rho,\nu,z}=
-(\mathsf D(f)\circ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle)_{\rho\oplus\nu, z}
+If $\rho\neq [x]\oplus\mu$, then $(D(\Lambda(f)))_{\rho,\nu,z}=\infty$ and $
+(\Lambda(D(f)\circ \langle  \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle))_{\rho,\nu,z}=
+(D(f)\circ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle)_{\rho\oplus\nu, z}
 $, and one can check that also this is $\infty$, using the second equation above and the fact that $(\pi_{1})_{\rho\oplus\emptyset,x}=\infty$.\end{description}
@@
 
-To check the validity of Taylor expansion we must further check the following equation, given $f\in Q^{\Pi}_{!}(C, B^{A})$ and $g\in Q^{\Pi}_{!}(C,A)$:
+To check the validity of Taylor expansion we must further check the following equation, given $f\in \LREL_!(C, B^{A})$ and $g\in \LREL_!(C,A)$:
 $$
@@
 \begin{enumerate}
-\item $\mathsf{ev}\in Q^{\Pi}_{!}(B^{A}+A, B)$ is the canonical \emph{evaluation} morphism;
+\item $\mathsf{ev}\in \LREL_!(B^{A}+A, B)$ is the canonical \emph{evaluation} morphism;
 
@@
 
-\item given $f\in Q^{\Pi}_{!}(C+A,B)$ and $g\in Q^{\Pi}_{!}(C,A)$, 
-$f\star g\in Q^{\Pi}_{!}(C+A,B)$ is the morphism obtained by differentiating $f$ in its second component and applying $g$ in that component, i.e.~
+\item given $f\in \LREL_!(C+A,B)$ and $g\in \LREL_!(C,A)$, 
+$f\star g\in \LREL_!(C+A,B)$ is the morphism obtained by differentiating $f$ in its second component and applying $g$ in that component, i.e.~
 $$
@@
 
-\item $\mathsf{ev}\in Q^{ \C M_{\mathsf f}(  ( \C M_{\mathsf f}(A)\times B)       +  A    ) \times B  }$ is given by
+\item $\mathsf{ev}\in \Lawv^{ \C M_{\mathsf{fin}}(  ( \C M_{\mathsf{fin}}(A)\times B)       +  A    ) \times B  }$ is given by
 $$\mathsf{ev}_{\mu,y}=
@@
  $$
-and observe that, given $f\in Q^{\Pi}_{!}(C, B^{A})$ and $g\in Q^{\Pi}_{!}(C,A)$, 
+and observe that, given $f\in \LREL_!(C, B^{A})$ and $g\in \LREL_!(C,A)$, 
 $$
@@
 
-\item given $g\in Q^{\Pi}_{!}(C, B^{A})$, 
-$\Lambda^{-}(g)\in Q^{\Pi}_{!}(C+A, B)$ is given by 
+\item given $g\in \LREL_!(C, B^{A})$, 
+$\Lambda^{-}(g)\in \LREL_!(C+A, B)$ is given by 
 $$

### 2023-01-19 appV.tex
Commit

@@
 
-\subsubsection{Proof of Theorem \ref{theorem:fepsilon}}
+\newpage
 
-DAVIDE's NOTES!
+\subsubsection{Proof of Theorem~\ref{theorem:fepsilon}}
+
+We give below the complete statement of Theorem \ref{theorem:fepsilon} together with its proof.
+
+First, let us set the following:
+\begin{definition}
+ Let $\preceq$ be the pointwise order on $\N^k$ (i.e.\ for all $\mathbf m,\mathbf n\in \N^{K}$, $\mathbf m\preceq \mathbf n$ iff $m_{i}\leq n_{i}$ for all $1\leq i\leq K$).
+ Of course $\mathbf m\prec \mathbf n$ holds exactly when $\mathbf m\preceq \mathbf n$ and $m_{i}<n_{i}$ for at least one $1\leq i\leq K$.
+ Finally, we set $\mathbf m\prec_{1}\mathbf n$  iff
+$\mathbf m\prec \mathbf n$ and $\sum_{i=1}^{K}n_{i}-m_{i}=1$ (i.e.\ they differ on exactly one coordinate).
+\end{definition}
+
+We will exploit the following:
+
+\begin{remark}\label{rmk:AC}
+\text{If $U\subseteq \N^{K}$ is infinite, then $U$ contains an infinite ascending chain $\mathbf m_{0}\prec \mathbf m_{1} \prec \mathbf m_{2} \prec \dots$.}.
+
+This is a consequence of K\"onig Lemma (KL): consider the directed acyclic graph $(U,\prec_{1})$, indeed a $K$-branching tree; if there is no infinite ascending chain $ \mathbf m_{0}\prec \mathbf m_{1} \prec \mathbf m_{2} \prec \dots$, then in particular there is no infinite ascending chain $ \mathbf m_{0}\prec_{1} \mathbf m_{1} \prec_{1} \mathbf m_{2} \prec_{1} \dots$ so the tree $U$ has no infinite ascending chain; then by KL it is finite, contradicting the assumption.
+\end{remark}
+
+\begin{theorem}[Theorem \ref{theorem:fepsilon}]
+Let $f:\overline{\R}_{\geq 0}^{\{1,\dots, K\}} \to \overline{\R}_{\geq 0}$ tropical analytical, with matrix $\hat f : \N^{K} \to \overline{\R}_{\geq 0}$.
+Then, for all $0<\epsilon<+\infty$, there is $\mathcal{F}_\epsilon\subseteq\N^{K}$ s.t.\
+\begin{enumerate}
+ \item $\mathcal{F}_\epsilon$ is finite
+ \item If $\mathcal{F}_\epsilon= \emptyset$ then $f(\mathbf x) = +\infty$ for all $\mathbf x\in \overline{\R}_{\geq 0}^{\{1,\dots, K\}}$
+ \item If $f(\mathbf x_0) = +\infty$ for some $\mathbf x_0\in [\epsilon,\infty)^{K}$ then $\mathcal{F}_\epsilon= \emptyset$
+ \item The restriction of $f$ to $[\epsilon,+\infty]^{K}$ coincides with the  \emph{tropical polynomial}:
+      \[
+        P_\epsilon(\mathbf x)= \min\limits_{\mathbf n\in\mathcal{F}_\epsilon} \set{\hat f (\mathbf n) + \mathbf n\mathbf x}.
+      \]
+      where $\mathbf n\mathbf x= \sum_{i=1}^{K}n_{i}x_{i}$.
+\end{enumerate}
+\end{theorem}
+\begin{proof}
+We let $\mathcal F_\epsilon$ to be the complementary in $\N$ of the set:
+\[
+ \set{\mathbf n\in\N^{K} \mid \textit{either } \hat f (\mathbf n)=+\infty \textit{ or there is } \mathbf m\prec \mathbf n\textit{ s.t.\ } \hat f(\mathbf m)\leq\hat f(\mathbf n)+\epsilon}.
+\]
+In other words, $\mathbf n\in\mathcal F_\epsilon$ iff $\hat f(\mathbf n)<+\infty$ and for all $\mathbf m\prec \mathbf n$, one has $\hat f(\mathbf m)>\hat f(\mathbf n)+\epsilon$.
+
+1).
+Suppose that $\mathcal F_\epsilon$ is infinite; then, using Remark~\ref{rmk:AC}, it contains an infinite ascending chain
+\[\set{\mathbf m_0\prec \mathbf m_1\prec\cdots}.\]
+By definition of $\mathcal F_\epsilon$ we have then:
+\[+\infty>\hat f(\mathbf m_0)>\hat f(\mathbf m_1)+\epsilon>\hat f(\mathbf m_2)+2\epsilon>\cdots\]
+so that $+\infty>\hat f(\mathbf m_0)>\hat f(\mathbf m_{i})+i\epsilon\geq i\epsilon$ for all $i\in\N$.
+This contradicts the Archimedean property of $\R$.
+
+2).
+We show that if $\mathcal F_\epsilon=\emptyset$, then $\hat f(\mathbf n)=+\infty$ for all $\mathbf n\in\N^{K}$.
+This immediately entails the desired result.
+We go by induction on the well-founded order $\prec$ over $\mathbf n\in\N^{K}$:
+
+- if $\mathbf n=0^{K}\notin\mathcal F_\epsilon$, then $\hat f(\mathbf n)=+\infty$, because there is no $\mathbf m\prec\mathbf n$.
+
+- if $\mathbf n\notin\mathcal F_\epsilon$, with $\mathbf n\neq 0^{K}$ then either $\hat f(\mathbf n)=+\infty$ and we are done, or there is $\mathbf m\prec \mathbf n$ s.t.\ $\hat f(\mathbf m)\leq \hat f(\mathbf n)+\epsilon$.
+By induction $\hat f(\mathbf m)=+\infty$ and, since $\epsilon<+\infty$, this entails $\hat f(\mathbf n)=+\infty$.
+
+3).
+If $f(\mathbf x_0)=+\infty$ with $\mathbf x_0\in [\epsilon,\infty)^{K}$, then necessarily $\hat f(\mathbf n)=+\infty$ for all $\mathbf n\in\N^{K}$.
+Therefore, no $\mathbf n\in\N^{K}$ belongs to $\mathcal F_\epsilon$.
+
+4).
+We have to show that $f(\mathbf x)=P_\epsilon(\mathbf x)$ for all $\mathbf x\in [\epsilon,+\infty]^{K}$.
+By 1), it suffices to show that we can compute $f(\mathbf x)$ by taking the $\inf$, that is therefore a $\min$, only in $\mathcal F_\epsilon$ (instead of all $\N^{K}$).
+If $\mathcal F_\epsilon=\emptyset$ then by 2) we are done (remember that $\min\emptyset := +\infty$).
+If $\mathcal F_\epsilon\neq\emptyset$, we show that for all $\mathbf n\in\N^{K}$, if $\mathbf n \notin\mathcal F_\epsilon$, then there is $\mathbf m\in\mathcal F_\epsilon$ s.t.\ $\hat f(\mathbf m)+\mathbf m\mathbf x \leq \hat f(\mathbf n)+\mathbf n\mathbf x$.
+We do it again by induction on $\prec_{1}$:
+
+- if $\mathbf n=0^{K}$, then from $\mathbf  n\notin \mathcal F_{\epsilon}$, by definition of $\mathcal F_\epsilon$, we have $\hat f(\mathbf n)=+\infty$ (because there is no $\mathbf n'\prec\mathbf n$).
+So any element of $\mathcal F_\epsilon\neq\emptyset$ works.
+
+- if $\mathbf n\neq 0^{K}$, then we have two cases:
+either $\hat f(\mathbf n)=+\infty$, in which case we are done as before by taking any element of $\mathcal F_\epsilon\neq\emptyset$.
+Or $\hat f(\mathbf n)<+\infty$, in which case (again by definition of $\mathcal F_\epsilon$) there is $\mathbf n'\prec\mathbf n$ s.t.\ \begin{equation}\label{eq:n'neps} \hat f(\mathbf n')\leq \hat f(\mathbf n)+\epsilon.\end{equation}
+Therefore we have (remark that the following inequalities hold also for the case $x=+\infty$):
+\[\begin{array}{rclr}
+ \hat f(\mathbf n')+\mathbf n'\mathbf x & \leq & \hat f(\mathbf n) + \epsilon +\mathbf  n'\mathbf x & \textit{by \eqref{eq:n'neps}} \\
+ & \leq & \hat f(\mathbf n) + (\mathbf n-\mathbf n')\mathbf x + \mathbf n'\mathbf x & \textit{because $\epsilon\leq\min\mathbf x$ and $\min \mathbf x\leq(\mathbf n-\mathbf  n')\mathbf x$} \\
+ & = & \hat f(\mathbf n)+ \mathbf n\mathbf x. &
+\end{array}\]
+Now, if $\mathbf n'\in\mathcal F_\epsilon$ we are done.
+Otherwise $\mathbf n'\notin\mathcal F_\epsilon$ and we can apply the induction hypothesis on it, obtaining an $\mathbf m\in\mathcal F_\epsilon$ s.t.\ $\hat f(\mathbf m)+\mathbf m\mathbf x \leq \hat f(\mathbf n')+\mathbf n'\mathbf x$.
+Therefore this $\mathbf m$ works.
+\end{proof}
+
+\newpage
 
@@
 
-Continuity of cones and metric. DAVIDE'S NOTES
-!
+We prove here the two results Theorem~\ref{thm:ScottCont} and Theorem~\ref{thm:cont} from Section~\ref{subsec:cont}.
+
+In order to prove Theorem~\ref{thm:cont}, we place ourselves in the general setting of locally convex topological vector spaces (LCTVS), and prove a general result about convex functions in this setting, adapting the argument from [Theorem , \cite{}].
+
+\begin{definition}
+ A \emph{normed semiring} is the data of a semiring $R$ together with a \emph{norm} on it, that is, a function $\absv{.}\,:R\to\R$ satisfying the usual axioms of the absolute value function\footnote{I.e.: $\absv{x}\geq 0$, $\absv x =0 \textit{ iff } x=0$, $\absv{xy}=\absv x\absv y$ and $\absv{x+y}\leq \absv x+\absv y$.}.
+ One can immediately check that if $R$ is a ring, then any norm on it induces a metric $d(r,s):=\mid r-s\mid$ on it (and thus also a topology).
+ A \emph{topological normed ring} is a normed ring which is a topological ring w.r.t.\ the topology induced by the norm distance.
+\end{definition}
+
+\begin{definition}
+ Let $R$ be normed semiring, let $A$ be a topological $R$-module and let $C\subseteq A$.
+ We say that $C$ is \emph{convex} iff for all $x,y\in C$ and for all $r,s\in R$ s.t. $\mid r\mid +\mid s\mid =1$, we have $rx+sy\in C$.
+ We say that $C$ is \emph{balanced} iff for all $x\in C$ and for all $r,s\in R$ s.t. $\mid r\mid +\mid s\mid =1$, we have $rx,sx\in C$.
+ We say that $C$ is a \emph{disk at $x\in A$} iff it is of shape $C=x+D$ for some $D$ both convex and balanced.
+\end{definition}
+
+\begin{example}
+ An example of a balanced but not convex set in (the normed field) $\R^2$.
+ \[\begin{tikzpicture}
+ \draw[->] (0,-1) -- (0,1);
+ \draw[->] (-1,0) -- (1,0);
+ \draw[fill=gray, fill opacity=0.3] (0,0) -- (0.5,-0.5) -- (0.5,0.5) -- (0,0) -- (-0.5,0.5) -- (-0.5,-0.5) -- cycle;
+\end{tikzpicture}\]
+\end{example}
+
+\begin{definition}
+ Let $R$ be a normed field and let $\mathbb{V}$ be a topological $R$-vector space.
+We say that $\mathbb{V}$ is \emph{locally convex} iff every $x\in \mathbb{V}$ admits a local basis consisting of disks at $x$.
+ Since the topology of $\mathbb{V}$ is necessarily translation invariant,
+ it is equivalent to only ask that $0$ has such a local basis.
+\end{definition}
+
+The main result is the following:
+
+\begin{proposition}\label{prop:mainLCTVS}
+ Let $\mathbb{V}$ be a locally convex topological $\R$-vector space (where $\R$ is endowed with its usual absolute value and the induced topology by it).
+ Fix $x\in\mathbb{V}$, a neighbourhood $V\subseteq\mathbb{V}$ of $x$ and $f:V\to \R\cup\set{+\infty,-\infty}$.
+ If $V$ is convex, $f$ is concave and lower bounded by a finite constant $K$, then $f$ is continuous at $x$ (w.r.t.\ the subspace topology on $V$).
+\end{proposition}
+\begin{proof}
+ Since $V$ is a neighbourhood of $x$, $x$ admits an open neighbourhood $U\subseteq V$, and by local convexity of $\mathbb{V}$ there is a (non necessary open) disk $D\subseteq U$ at $x$.
+ Now fix $\eps\in(0,1)$ and $r,s\in \R$ with $0\leq r\leq\eps$, $s=1-r$.
+ Fix also $w\in D$.
+ By the convexity of $D$ we have $sx+rw\in D$, and thus the convexity of $f$ entails that:
+ \[
+  f(sx+rw)\geq s f(x) + r f(w) \geq s f(x) + r K = (1-r)f(x) + r K
+ \]
+that is,
+ \[
+  f(x)-f(sx+rw)\leq r (f(x)-K).
+ \]
+ Now remark that $x=\dfrac{1}{s+2r}(sx+rw)+\dfrac{r}{s+2r}(2x-w)$, where $\dfrac{1}{s+2r}+\dfrac{r}{s+2r}=1$, $\dfrac{1}{s+2r}<1$ and $2x-w\in D$ (because $D$ is a disk at $x$).
+ Therefore by the convexity of $f$ we have:
+ \[
+  f(x)\geq \dfrac{1}{s+2r}f(sx+rw)+\dfrac{r}{s+2r}f(2x-w)\geq \dfrac{1}{s+2r}(f(sx+rw)+rK)
+ \]
+ that is,
+ \[
+  f(sx+rw)-f(x)\leq r(f(x)-K).
+ \]
+ We have thus shown that $\mid f(sx+rw)-f(x)\mid\leq r(f(x)-K)$ for all $w\in D$.
+ Since this holds also for all $\eps\in(0,1)$, and due to the choice of $r,s$, the points of shape $sx+rw$ span $D$ when $w$ spans $D$ and $\eps$ spans $(0,1)$.
+ That is, we have shown that $\mid f(w)-f(x)\mid\leq r(f(x)-K)$ for all $w\in W$.
+ Since $f(x)-K\geq 0$ and $r\leq \eps$, we have that $\exists\lim\limits_{w\to x} f(w)=f(x)$.
+\end{proof}
+
+Now we show how to apply this argument to our tropical case $\Lawv^X$, which is \emph{not} a LCTVS (because it is not a vector space).
+
+First, we have:
+
+\begin{corollary}\label{cor:cont}
+ Let $f:({\R}_{> 0}^X,\supnorm{.}) \to (\overline\R_{\geq 0},\absv .)$ and $x\in {\R}_{> 0}^X$.
+ If there is a convex neighbourhood $V\subseteq \overline{\R}_{> 0}^X$ of $x$ s.t.\ $f_{\lVert V}$ is concave, then $f$ is continuous at $x$.
+\end{corollary}
+\begin{proof}
+ The $\R$-vector space $\R^X$ is topological w.r.t.\ the topology $\tau_\infty$ induced on it by the norm $\lVert .\lVert_\infty$, and it is clearly locally convex.
+ Call $\tau^+_\infty$ the topology induced by $\supnorm{.}$ on ${\R}_{> 0}^X$.
+ It is clear that it coincides with $\tau^+_\infty$.
+ Since moreover ${\R}_{> 0}^X$ is open in $(\R^X,\tau_\infty)$, the neighbourhood $V$ of $x$ in $({\R}_{> 0}^X,\supnorm{.})$ is also a neighbourhood of $x$ in $(\R^X,\tau_\infty)$.
+ We can therefore apply Proposition~\ref{prop:mainLCTVS} to $f_{\lVert V}$ (which is lower bounded by $0$ by definition of $f$) and obtain that $f_{\lVert V}$ is continuous at $x$ w.r.t.\ the subspace topology $\tau_V$ induced by $\tau_\infty$ on $V$.
+ But since $V$ is contained in ${\R}_{> 0}^X$, the topology $\tau_V$ coincides with the subspace topology induced on $V$ by $\tau^+_\infty$.
+ So $f$ is continuous at $x$ w.r.t.\ $\tau^+_\infty$.
+\end{proof}
+
+One may wonder if the same proof as above makes it possible to state the previous corollary replacing ${\R}_{> 0}^X$ with $\R_{\geq 0}^X$ (so, in particular, taking $x\in\R_{\geq 0}^X$).
+This is not possible because in the proof we crucially use that $\R_{> 0}^X$ is open in $(\R^X,\tau_\infty)$, which is not the case of $\R_{\geq 0}^X$.
+In fact, this allowed us to say that $V$ is a neighbourhood of $x$ in $(\R^X,\tau_\infty)$, and therefore to be able to apply Proposition~\ref{prop:mainLCTVS}.
+Taking $\R_{\geq 0}^X$ instead, this is in general not true: we could have a neighbourhood of $x$ w.r.t.\ the subspace topology on $\R_{\geq 0}^X$ induced by $\tau_\infty$, which does not contain any open neighbourhood of $x$ w.r.t.\ $\tau_\infty$ (i.e.
+it is not a neighbourhood w.r.t.\ $\tau_\infty$).
+
+\begin{example}
+ An example of a neighbourhood of $x$ w.r.t.\ the subspace topology on $\R_{\geq 0}^X$ induced by $\tau_\infty$, which is not a neighbourhood w.r.t.\ $\tau_\infty$.
+ \[\begin{tikzpicture}
+ \draw[->] (0,-0.5) -- (0,1.5);
+ \draw[->] (-0.5,0) -- (1.5,0);
+ \filldraw[black] (0.8,0) circle (0.7pt) node[anchor=north]{x};
+ \fill[fill=gray, fill opacity=0.3] (0.4,0) -- (1.2,0) -- (1.2,0.7) -- (0.4,0.7) -- (0.4,0) -- cycle;
+\end{tikzpicture}\]
+\end{example}
+
+Finally we have the desired result Theorem~\ref{thm:cont}:
+
+\begin{theorem}[Theorem~\ref{thm:cont}]
+ Tropical Laurent series $f:({\R}_{\geq 0}^X,\supnorm .) \to (\overline\R_{\geq 0},\absv .)$ are continuous on ${\R}_{> 0}^X$ w.r.t.\ the norm $\norm{\cdot}_\infty$.
+\end{theorem}
+\begin{proof}
+ We know that tLs's are concave on all their domain.
+ Therefore the result immediate follows by Corollary \ref{cor:cont}, since the topology induced by $\supnorm{.}$ on ${\R}_{> 0}^X$ coincides with the subspace topology induced by $({\R}_{\geq 0}^X,\supnorm .)$ on it.
+\end{proof}
+
+Due to the previous discussion about the impossibility of stating Corollary \ref{cor:cont} in the case where one of the coordinates of $x$ is $0$, the continuity of tropical functions on the hyperplanes $\mathcal H_a:=\set{x\in {\R}_{\geq 0}^X \mid x_a=0}$, for $a\in X$, must be treated separately.
+Similarly, the continuity at points with infinite coordinates must be treated separately as well.
+We left it for future investigations.
+
+\begin{remark}
+ We formulated the main argument (Proposition~\ref{prop:mainLCTVS}) in the general setting of LCTVS's.
+We did so just to show that it is actually a general argument, but we could have placed ourselves already in a more particular setting of our interest and carry on the proof.
+For instance, Theorem~\ref{thmTLSlocLip} will be proved by refining this kind of argument (Lemma~\ref{lm:mainLip}), but not in the setting of LCTVS.
+\end{remark}
+
+\newpage
+
+\begin{definition}
+ An \emph{$\overline{\R}_{\geq 0}$-cone} is a commutative $\overline{\R}_{\geq 0}$-semimodule with cancellative addition\footnote{I.e.: $x+y=x+y' \Rightarrow y=y'$.}.
+\end{definition}
+
+In [Selinger] cones are required to also have ``strict addition'', meaning that $x+y=0 \Rightarrow x=y=0$.
+We do not add this requirement since it will automatic hold when considering normed cones.
+
+\begin{remark}
+ The addition of a cone $P$ (which forms a commutative monoid) turns $P$ into a poset by setting:
+ \[
+  x \leq y \textit{ iff } y=x+z \textit{, for some }z\in P.
+ \]
+ By the cancellative property, when such $z$ exists it is unique, and we denote it by $y-x$.
+ Such order is called the \emph{cone-order} on $P$.
+\end{remark}
+
+\begin{definition}
+ A \emph{normed $\overline{\R}_{\geq 0}$-cone} is the data of a $\overline{\R}_{\geq 0}$-cone together with a $\leq$-monotone\footnote{I.e.: $x\leq y \Rightarrow \norm{x}\leq \norm y$.
+Remark that requiring this property (for all $x,y$) is equivalent to requiring that $\norm{x}\leq \norm{x+y}$ for all $x,y$.} norm\footnote{A \emph{norm} on a $\overline{\R}_{\geq 0}$-abstract cone $P$ is a map $\norm{.}:P\to \overline{\R}$ satisfying the usual axioms of norms:
+ $\norm x \geq 0$, $\norm x = 0 \Rightarrow x=0$, $\norm{rx}=r\norm x$ and $\norm{x+y}\leq \norm x + \norm y$.} on it.
+\end{definition}
+
+In [Ehrhard-Pagani-Tasson, Crubillie] a normed $\overline{\R}_{\geq 0}$-cone is simply called a cone.
+
+Remark that in a normed $\overline{\R}_{\geq 0}$-cone, by monotonicity of the norm, we have: $\norm{x+y}=0 \Rightarrow x=y=0$.
+Therefore, as already mentioned, in a normed cone we have: 
+$x+y=0 \Rightarrow \norm{x+y}=0 \Rightarrow x=y=0$, that is, addition is strict.
+
+\begin{example}
+ $\overline{\R}_{\geq 0}^X$ is a normed cone with the norm $\supnorm{x}:=\sup\limits_{a\in X} x_a\in \overline{\R}_{\geq 0}$.
+\end{example}
+
+\begin{remark}
+ The cone-order on $\overline{\R}_{\geq 0}^X$ is the pointwise usual order on $\overline{\R}_{\geq 0}$.
+
+ Remark also that tropical functions have no reason to be linear nor sublinear.
+\end{remark}
+
+\begin{remark}\label{rmk:tropMonot}
+ Tropical functions are monotone w.r.t.\ to the cone order on its domain and codomain.
+ This is clear, using the definition of tropical functions, because $\mu x\leq \mu y$ if $x\leq y$.
+\end{remark}
+
+\begin{remark}
+ The cone-order on $\overline{\R}_{\geq 0}^X$ makes it into a dcpo with least element $0$.
+\end{remark}
+
+The following is a usual domain theoretical lemma.
+
+\begin{lemma}\label{lm:sup=sup}
+ Let $P$ be a poset and $X, I$ sets. Fix the pointwise order on $P^X$.
+ \begin{enumerate}
+  \item Let $x^i\in P^X$ for $i\in I$.
+  If $\bigvee\limits_{i\in I} x^i$ exists $P^X$, then $\bigvee\limits_{i\in I} x^i_a$ exists in $P$ for all $a\in X$, and we have: $\bigvee\limits_{i\in I} x^i_a=\left(\bigvee\limits_{i\in I} x^i\right)_a$.
+ \item Let $x^i_a\in P$ for $i\in I,a\in X$.
+  If $\bigvee\limits_{i\in I} x^i_a$ exists $P$ for all $a\in X$, then $\bigvee\limits_{i\in I} x^i$ exists in $P^X$, and we have: $\left(\bigvee\limits_{i\in I} x^i\right)_a=\bigvee\limits_{i\in I} x^i_a$.
+ \end{enumerate}
+\end{lemma}
+
+A \emph{directed net} in a poset $P$ with indices in a set $I$ is a function $s:I\to P$, denoted by $(s_i)_{i\in I}$, s.t.\ its image is directed.
+We say that a directed net in $P$ \emph{admits a sup} iff its image admits a sup in $P$.
+We say that a directed net $s$ in a normed cone is \emph{bounded} iff the set $\set{\norm{s_i}\,\mid i\in I}$ is bounded in $\R_{\geq 0}$.
+
+Remember the definition of Scott-continuity:
+
+\begin{definition}
+ A function $f:P\to P'$ between posets is \emph{Scott-continuous} iff for all directed net $(s_i)_i$ in $P$ admitting a sup, we have $\exists \bigvee\limits_i f(s_i) = f(\bigvee\limits_i s_i)$ in $P'$.
+\end{definition}
+
+The fundamental result in order to prove Theorem~\ref{thm:ScottCont} is the following, taken from [Selinger].
+
+\begin{proposition}\label{prop:infsup}
+ Let $P$ be a normed $\overline{\R}_{\geq 0}$-cone s.t.\ every bounded directed net in $P$ admits a sup.
+ Let $(v_i)_{i\in I}$ be a directed net in $P$ with an upper bound $v\in P$.
+ Then $\exists\bigvee\limits_{i\in I} v_i \in P$ and, if $\inf\limits_{i\in I} \norm{v-v_i} =0$, one has: $\bigvee\limits_{i\in I} v_i = v$.
+\end{proposition}
+\begin{proof}
+ Remark that $v-v_i$ exists in $P$ by hypothesis and so does $\bigvee\limits_{i\in I} v_i$, thanks to the monotonicity of the norm.
+ Now, since $v\geq v_i$ for all $i$, we have that $v\geq \bigvee\limits_{i\in I} v_i$, and so $v-\bigvee\limits_{i\in I} v_i$ exists in $P$.
+ Fix $i\in I$.
+ Since $v_i\leq \bigvee\limits_{i\in I} v_i$, then $v-\bigvee\limits_{i\in I} v_i\leq v-v_i$ and, by monotonicity of the norm, $\norm{v-\bigvee\limits_{i\in I} v_i}\leq \norm{v-v_i}$.
+ Since this holds for all $i\in I$, we have:
+ $0\leq \norm{v-\bigvee\limits_{i\in I} v_i}\leq \inf\limits_{i\in I} \norm{v-v_i}=0$, where the last equality holds by hypothesis.
+ Thus $\norm{v-\bigvee\limits_{i\in I} v_i}=0$, i.e.\ $v=\bigvee\limits_{i\in I} v_i$.
+\end{proof}
+
+\begin{definition}
+ A normed $\overline{\R}_{\geq 0}$-cone $P$ is \emph{Scott-complete} iff its norm is Scott-continuous (where the codomain $\overline\R_{\geq 0}$ is endowed with its usual order) and every bounded directed net in $P$ admits a sup.
+ This is equivalent to asking that its closed unit ball is a dcpo.
+\end{definition}
+
+\begin{proposition}
+ The normed cone $\overline{\R}_{\geq 0}^X$ is Scott-complete.
+\end{proposition}
+\begin{proof}
+ $\overline{\R}_{\geq 0}^X$ being a dcpo, all the existences of sup's that we should check do automatically hold, so we only have to show that $\bigvee$ and $\supnorm{.}$ commute:
+ $\supnorm{\bigvee\limits_i x_i}=
+ \sup\limits_a\, \sup\limits_i\, (x_i)_a =
+ \sup\limits_i\, \sup\limits_a\, (x_i)_a =
+ \bigvee\limits_i \supnorm{x_i}$.
+\end{proof}
+
+%Let us set $\infty_b$ to be the hyper-semi-plane $\set{y\in\overline{\R}^Y_{\geq 0}\mid y_b=+\infty}$ and $\set{f<<+\infty}:=\bigcup\limits_{b\in Y} f^{-1}\infty_b$.
+
+%\begin{remark}\label{rmk:R-f<<oo}
+%Let $f:\overline{\R}^X_{\geq 0}\to\overline{\R}^Y_{\geq 0}$ be tropical.
+%Then, $\R^X_{> 0}-\set{f<<+\infty}$ is a dcpo (w.r.t.\ the usual order, that is, its cone order).
+%In order to see this, it is enough to show that, for any $b\in Y$ and for any directed net $(x^i)_i$ in $\R^X_{> 0}$ s.t.\ $f(x^i)_b<+\infty$ for all $i$, we have $f(\bigvee\limits_i x^i)_b<+\infty$ (where the sup is taken in $\R^X_{> 0}$).
+%Such property holds because otherwise it is easy to see that $f$ would be discontinuous at the point $\bigvee\limits_i x^i\in\R^X_{> 0}$ w.r.t.\ the topology\footnote{Remark that this argument would not be possible using Scott-continuity instead.
+Indeed, it could well be that $f(\bigvee\limits_i x^i)_b=\bigvee\limits_i f(x^i)_b=+\infty$ while $f(x^i)_b<+\infty$ for all $i$.} induced by $\supnorm{.}$.
+%This contradicts Corollary \ref{cor:tropCont}.
+%\end{remark}
+
+\begin{proposition}
+ All tropical functions $f:\overline{\R}^X_{\geq 0}\to\overline{\R}^Y_{\geq 0}$ are Scott-continuous on $\R^X_{> 0}$ w.r.t.\ the cone-orders on its domain and codomain.
+\end{proposition}
+\begin{proof}
+ Let $(x_i)_i$ a directed net in $\R^X_{> 0}$ s.t.\ $\bigvee\limits_i x^i$ exists in $\R^X_{> 0}$.
+ Then $\inf\limits_i \supnorm{\bigvee\limits_i x^i - x^i} =0$, where $\bigvee\limits_i x^i - x^i$ exists because $\bigvee\limits_i x^i \geq x^i$ for all $i$.
+ Since $f$ is $\supnorm{.}$-continuous on $\R^X_{> 0}$ (Corollary \ref{cor:tropCont}), then $\inf\limits_i \supnorm{f(\bigvee\limits_i x^i) - f(x^i)} =0$, where $f(\bigvee\limits_i x^i) - f(x^i)$ exists because $f(\bigvee\limits_i x^i) \geq f(x^i)$ for all $i$ being $f$ monotone (Remark \ref{rmk:tropMonot}).
+ We can therefore apply Proposition \ref{prop:infsup} to the directed net $(f(x^i))_i$ in $\overline{\R}^Y_{\geq 0}$, obtaining that $\bigvee\limits_i f(x^i)$ exists in $\overline{\R}^Y_{\geq 0}$ and it coincides with $f(\bigvee\limits_i x^i)$.
+\end{proof}
+
+%\begin{remark}
+%By looking at the definition of tropical functions, for $f$ tropical we have that: if $x,x'\in\R^X_{> 0}-\set{f<<+\infty}$ and $r\in\R_{> 0}$, then $x+x',rx\in\R^X_{> 0}-\set{f<<+\infty}$.
+%That is, $\R^X_{> 0}-\set{f<<+\infty}$ is a $\R_{>0}$-cone.
+%It is clearly still Scott-complete.
+%\end{remark}
+
+In the main paper we mention that $\R_{\geq0}^X$ is a Scott-complete dcpo.
+We did not state it as a theorem because it follows from standard and well-known domain theoretical considerations.
+Let us prove it here anyway.
+
+Recall the definition of Scott-complete dcpo:
+
+\begin{definition}
+ Let $P$ be a dcpo and define a relation by: $x<<y$ iff for all directed $D\subseteq P$, if $y\leq \bigvee D$, then $x\leq d$, for some $d\in D$.
+ Call $\twoheaddownarrow y:=\set{x\in P\mid x<<y}$.
+ A dcpo $P$ is \emph{Scott-continuous} iff $\twoheaddownarrow x$ is directed and $x=\bigvee\twoheaddownarrow x$ for all $x\in P$.
+\end{definition}
+
+\begin{remark}
+ In any dcpo we have: $\twoheaddownarrow x \subseteq \downarrow x$.
+ This immediately follows by considering the directed set $\set{x}$.
+\end{remark}
+
+\begin{lemma}\label{lm:<<dir}
+ In $\overline{\R}_{\geq 0}^X$, every set $\twoheaddownarrow x$ is directed.
+\end{lemma}
+\begin{proof}
+ It is immediate that $0\in\twoheaddownarrow x$, so it is non-empty.
+ Now let $y,y' << x$.
+ Since we are in $\overline{\R}_{\geq 0}^X$, there is $y\vee y'\in\overline{\R}_{\geq 0}^X$.
+ So we only have to show that $y\vee y'<<x$.
+ For that, let $D$ be a directed set in $\overline{\R}_{\geq 0}^X$ s.t.\ $x\leq\bigvee D$.
+ Since $y,y' << x$ we find $d,d'\in D$ s.t.\ $y\leq d$, $y'\leq d'$.
+ Since $D$ is directed, there is $\hat d\in D$ s.t.\ $y\leq d\leq \hat d\geq d'\geq y'$.
+ But then, by definition of sup, it must be $\hat d\geq y\vee y'$ and we are done.
+\end{proof}
+
+\begin{remark}
+ The following is a known property (that we will not use):
+ let $P$ be a complete normed cone which is a dcpo w.r.t.\ its cone-order. Then $P$ is Scott-continuous as a dcpo iff its closed unit ball is Scott-continuous as a dcpo.
+\end{remark}
+
+\begin{remark}\label{rmk:<<R}
+ Consider $\overline{\R}_{\geq 0}$ with its usual order (which coincides with its cone-order).
+ It is easily seen (and well known) that $y<<x$ iff either $y=0$ or $y<x$.
+ This immediately implies that $\overline{\R}_{\geq 0}$ is Scott-continuous, since $x=\sup\limits_{y<x} y$.
+\end{remark}
+
+\begin{lemma}\label{lm:<<RX}
+ Let $x,y\in \overline{\R}_{\geq 0}^X$ (considered as a dcpo with its cone order, which is the pointwise one). Fix $a\in X$.
+ If $y_a<<x_a$ and $y_c=0$ for all $c\neq a$, then $y<<x$.
+\end{lemma}
+\begin{proof}
+ Towards a contradiction, assume that there is a directed set $D$ in $\overline{\R}_{\geq 0}^X$ with $x\leq\bigvee D$ and s.t.\ $y\not\leq d$ for all $d\in D$.
+ Call $D_a:=\set{d_a\mid d\in D}$ and remark that it is directed, because $D$ is.
+ Also, by Lemma \ref{lm:sup=sup}, $\bigvee D_a = \left(\bigvee D\right)_a$.
+ Therefore from $y_a<<x_a\leq \bigvee D_a$ we obtain a $d\in D$ s.t.\ $y_a\leq d_a$.
+ By the absurd hypothesis we have $y\not\leq d$, so there must be $c\in X$ s.t.\ $y_c\not\leq d_c$, i.e. (because real numbers are totally ordered) $y_c>d_c$.
+ Therefore it must be $c\neq a$.
+ But then we have $0=y_c>d_c\geq 0$, contradiction.
+\end{proof}
+
+\begin{lemma}\label{lm:<<_a}
+ In $\overline{\R}_{\geq 0}^X$ we have:
+ if $y<<x$ then $y_a<< x_a$ for all $a\in X$.
+\end{lemma}
+\begin{proof}
+ Fix $a\in X$.
+ Let $D$ directed set in $\overline{\R}_{\geq 0}$ s.t.\ $x_a\leq\bigvee D$.
+ We look for a $d\in D$ s.t.\ $y_a\leq d$.
+ For $d\in D$, let $x^{a,d}\in\overline{\R}_{\geq 0}^X$ defined by $x^{a,d}_c:=x_c$ if $c\neq a$ and $x^{a,d}_a:=d$.
+ Let $D^x_a:=\set{x^{a,d}\mid d\in D}$.
+ By Lemma \ref{lm:sup=sup} $D^x_a$ admits sup in $\overline{\R}_{\geq 0}^X$ and it is $(\bigvee D^x_a)_c=x_c$ if $c\neq a$ and $(\bigvee D^x_a)_a=\bigvee D$.
+ Hence, $x\leq \bigvee D^x_a$.
+ If we prove that $D^x_a$ is directed, we are done: indeed, since $y<<x$, there is $d\in D$ s.t.\ $y\leq x^{a,d}$ and thus, in particular, $y_a\leq x^{a,d}_a=d$.
+ Let us finally prove that $D^x_a$ is directed: it is clearly non-empty, since $D$ is.
+ Let now $d,d'\in D$.
+ We want to show that there is $\hat d\in D$ s.t.\ $x^{a,d}\leq x^{a,\hat d}\geq x^{a,d'}$.
+ But since $D$ is directed, there is $\hat d\in D$ s.t.\ $d\leq \hat d\geq d'$, and therefore $x^{a,d}_c=x_c= x^{a,\hat d}_c =x_c=x^{a,d'}_c$ for all $c\neq a$, and $x^{a,d}_a=d\leq \hat d=x^{a,\hat d}_a=\hat d \geq d'=x^{a,d'}_a$.
+\end{proof}
+
+\begin{lemma}\label{rmk:sup_a=sup_a}
+In $\overline{\R}_{\geq 0}^X$ we have:
+$\bigvee\limits_{y<<x_a} y = \bigvee\limits_{y<<x} y_a$.
+\end{lemma}
+\begin{proof}
+ If we show that $\twoheaddownarrow x_a = \set{y_a\mid y\in\twoheaddownarrow x}$, then we are done because in the statement we are taking their sup's.
+ The inclusion ($\supseteq$) immediately follows from Lemma \ref{lm:<<_a}. For ($\subseteq$), let $d<< x_a$.
+ Then the $y\in \overline{\R}_{\geq 0}^X$ defined by $y_c:=0$ if $c\neq a$ and $y_a:=d$, is s.t.\ $y\in\twoheaddownarrow x$ by Lemma \ref{lm:<<RX}.
+ Thus, $d\in \set{y_a\mid y\in\twoheaddownarrow x}$.
+\end{proof}
+
+\begin{corollary}
+ The dcpo $\overline{\R}_{\geq 0}^X$ is Scott-continuous.
+\end{corollary}
+\begin{proof}
+ The fact that $\twoheaddownarrow x$ is directed is given by Lemma \ref{lm:<<dir}. The fact that $x=\bigvee \twoheaddownarrow x$ is given by the following equalities:
+ $x_a=\bigvee\limits_{y<<x_a} y = \bigvee\limits_{y<<x} y_a = \left(\bigvee\limits_{y<<x} y\right)_a=\left(\bigvee \twoheaddownarrow x\right)_a$.
+ The first equality follows from Remark \ref{rmk:<<R}, the second one from Lemma \ref{rmk:sup_a=sup_a}, the third one from Lemma \ref{lm:sup=sup}.
+\end{proof}
+
+\newpage
 
@@
 
-\begin{lemma}
+\begin{lemma}\label{lm:mainLip}
 Let $f: [0,\infty)^{X}\to Q$ be concave, monotone increasing, and continuous.  

### 2023-01-19 section4.tex
commit
Commit

@@
 With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
-$f(x)=\min\{x,x_{0}\}+ g(x)$.
+$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
 %
@@
 \end{theorem}
-%\begin{proof}
-%Call $\leq$ the well-founded pointwise order on $\N^k$.
-%Now we can let $\C F_\epsilon:=\set{n\in\N \mid
-%\widehat f(n)<\infty \textit{ and } \widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$.
-%\end{proof}
+\begin{proof}
+Call $\leq$ the (well-founded) pointwise order on $\N^k$.
+Now \mbox{$\C F_\epsilon:=\set{n\in\N \mid
+\widehat f(n)<\infty \textit{ and } 
+\widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$}
+does the job.
+\end{proof}
 
-\subsection{Continuity of tLs}%$\Lawv^{X}$ as a normed cone}
+\subsection{Continuity of tLs}\label{subsec:cont}%$\Lawv^{X}$ as a normed cone}
 
-$\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
-Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete normed cone.
-The normed cone structure induces an order on it, called its \emph{cone order}, by setting:
-$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
-This order makes it a Scott-continuous dcpo.
-Furthermore we have:
-
-\begin{proposition}
-  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders.
-  % on the domain and codomain.
-\end{proposition}
-The tLs $\varphi$ is continuous on $\BB R_{\geq0}=\Lawv-\set{\infty}$ (w.r.t.\ the usual norm of real numbers).
-By considering the usual norm $\norm{x}_\infty:=\sup\limits_{a\in X} \absv{x_a}$ on $\Lawv^X$, we could generalise this property by dropping the case of $x$ having some $0$ coordinate:
+The tLs $\varphi$ is continuous on $\BB R_{\geq0}$ (w.r.t.\ the usual norm of real numbers).
+By considering the usual norm $\norm{x}_\infty:=\sup\limits_{a\in X} \absv{x_a}$ on $\Lawv^X$, we can generalise this property by dropping the case of $x$ having some $0$ coordinate:
 
@@
 \begin{proof}
- The result follows after proving that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
+ It follows after adaptating [Theorem , \cite{}] in order to prove that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
 \end{proof}
+
+We conclude this subsection by noticing that $\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
+Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete \emph{normed cone} %(see~\cite{Selinger2004}, or the appendix, for such notions}.
+Its cone structure induces an order on it, called its \emph{cone order}:
+$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
+This makes it a Scott-continuous dcpo.
+Suitable categories of cones have been recently investigated as models of probabilistic computation %(\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
+We will just show that:
+
+\begin{theorem}\label{thm:ScottCont}
+  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders.
+  % on the domain and codomain.
+\end{theorem}
 
@@
 \begin{theorem}\label{thmTLSlocLip}
- All tropical Laurent series $\Lawv^X\to\Lawv$ are locally Lipschitz on $\BB R_{>0}$.
+ All tropical Laurent series $\Lawv^X\to\Lawv$ are locally Lipschitz on $\BB R_{>0}^X$.
 \end{theorem}

### 2023-01-19 section5.tex
Commit

@@
 \]
-so that $M\oplus_p N$ plays the role of a probabilistic sample with parameter $p$.
+so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
 We use this calculus as a toy example for our purposes, and it can be seen as a fragment of the PCF considered in \cite{Manzo2013}.
@@
 The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
-For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$, starting from the left.
+For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$ starting from the left.
 Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
@@
 They correspond to the respective reduction path from $M$ to the normal term of the same address.
-$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ in $M$ after reducing it to normal form.
+$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ after all the coin tossings in $M$ are performed.
 Thinking of $p,q$ as parameters, $P_{\omega}$ can thus be read as the \emph{likelihood function} of the event $\omega$.
-The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after such samplings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.
+The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after all the tossings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.
 
@@
 
-This way the probabilistic evaluation of $M$ is presented in the form of a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}.
-Indeed, typical questions in this case would be:
+This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}.
+Typical questions in this case would be:
 %
@@
  \item What is the \emph{maximum likelihood estimator} for the event ``$M$ produces $\true_\omega$'' (similarly for $\false_\omega$)?
- This means asking which is the choice of the parameters $p,q$ that maximizes the probability of getting $\true_\omega$ by repeatedly sampling $M$.
+ I.e., which is the choice of the parameters $p,q$ that maximizes the probability of getting $\true_\omega$ after all tossings in $M$?
  \item Knowing that $M$ produced $\true$ (similarly for $\false$), which is the choice of the parameters $p,q$ that maximizes the probability that the produced occurrence of $\true$ was a fixed $\true_{\omega_0}$ (i.e.
@@
 %These are questions much in the style of those asked in {\color{red}Ref?} in the framework of {\color{red}??}.
-Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega}$, which is more easily solved by 
-passing to the tropical polynomial $\trop P_{\omega}$. 
+Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega}, Q_\omega$, which is more easily solved by 
+passing to the tropical monomials/polynomial $\trop P_{\omega},\trop Q_{\omega}$. 
 For 1), we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)$ belong to:
@@
 \end{equation}
-In both cases, passing to $\trop P_{\omega}$ makes the problem easier, as this may be reduced to computing the tropical roots of the polynomial (which can be done in linear time \cite{Noferini2015}).
+In both cases, passing through $\trop$ %P_\omega, \trop Q_\omega$ 
+makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2015}).
 %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.
@@
 \begin{remark}\label{rem:troproots}
- Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_1$ or of $\trop Q_0$ provide, by definition, the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable paths of samples from $M$ to its normal form.
- Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log_c p,-\log_c(1-p))$ is a tropical root of, say, $\trop Q_1$ are the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are sampled by $M$ during its reduction to normal form, knowing that some $\true$ was sampled.
+ Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_1$ or of $\trop Q_0$ provide, by definition, the values of the bias of $\oplus_p$ for which there are at least two different equiprobable paths of tossings from $M$ to its normal form.
+ Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log_c p,-\log_c(1-p))$ is a tropical root of, say, $\trop Q_1$ are the values of the bias of $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are smapled by $M$ during its tossings to normal form, knowing that some $\true$ was sampled.
 \end{remark}

### 2023-01-19 appIV.tex
commit

@@
 $$
-f\star g =  \mathsf{D}(f)\circ \langle \langle \infty, g\circ \pi_{1}\rangle, \mathrm{id}_{C+A}\rangle.
+f\star g =  D(f)\circ \langle \langle \infty, g\circ \pi_{1}\rangle, \mathrm{id}_{C+A}\rangle.
 $$

### 2023-01-19 appVII.tex
commit

@@
 
-\begin{proposition}[universal property of the tensor product, cf.~\ref{Russo2007}]
+\begin{proposition}[universal property of the tensor product, cf.~\cite{Russo2007}]
 The tensor product $M\otimes N$ is the codomain of the universal $\Lawv$-bimorphism $M\times N \to M\otimes N$.
@@
 
-\begin{lemma}[cf.~\ref{Russo2007}]
+\begin{lemma}[cf.~\cite{Russo2007}]
 \begin{itemize}
@@
 \item the comonad $(!,\delta,\epsilon)$ is given by:
+{\small
 \begin{align*}
@@
 \end{align*}
+}
 
@@
 \bigvee\{\iota_{n}(\alpha(n))\mid n\in \BB N\}=\alpha$, and 
+{\small
 \begin{align*}
@@
  \end{align*}
+ }
 %But since $\epsilon(\beta)=\beta(1)$, this means that we are reduced to consider multisets $\alpha_{j}^{i} $ of cardinality 1, i.e.~of the form $[a_{j}^{i}]$, so we have  
@@
 \item the cocommutative comonoid structure $(!X, \Delta, e)$ is  defined as follows:
+{\small
 \begin{align*}
@@
 
+{\small
 \begin{align*}
@@
 \end{align*}
-
+}
+
+{\small
 \begin{align*}
@@
 \end{align*}
+}
 and one can argue similarly for $(1\otimes e)(\Delta(\alpha))$.
 
+{\small
 \begin{align*}
@@
 \end{align*}
+}
 
 Finally, the commutation of $\Delta$ and $\delta$:
-{\tiny
+{\small
 \begin{align*}
@@
 \end{align*}
+}
 and one can argue similarly for $\nabla(u\otimes 1)(\alpha)=\alpha$.
 
+{\small
 \begin{align*}
@@
 &= \bigvee_{k} 0\cup 0\\
-&= h_{Q}\left(\bigvee_{k}\alpha_{k}(0)\otimes\beta_{k}(0))\right) = h\left( (e\otimes e)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)\right)
-\end{align*}
-where $h_{Q}:Q\otimes Q \to Q$ indicates the isomorphism $h(\epsilon\otimes \delta)=\epsilon+\delta$.
-
+&= h\left(\bigvee_{k}\alpha_{k}(0)\otimes\beta_{k}(0))\right) = h\left( (e\otimes e)\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)\right)
+\end{align*}
+}
+where $h:\Lawv\otimes\Lawv\to \Lawv$ indicates the isomorphism $h(\epsilon\otimes \delta)=\epsilon+\delta$.
+
+{\small
 \begin{align*}
@@
 \end{align*}
+}
 where we are identifying $\epsilon \in q$ with the equivalence class
 $\epsilon \otimes 0 = \{ \langle \epsilon_{1},\epsilon_{2}\rangle\mid \epsilon_{1}+\epsilon_{2}=\epsilon\}\in 
-Q\otimes Q$ via the isomorphism $h_{Q}$ described above.
+\Lawv\otimes\Lawv$ via the isomorphism $h$ described above.
 
@@
 
-{\tiny
+{\small
 \begin{align*}
@@
 $ \epsilon\circ \nabla=(\epsilon \otimes e)\vee (e\otimes \epsilon)$:
-\begin{align*}
+{\small\begin{align*}
 \epsilon \left(\nabla\left(\bigvee_{k}\alpha_{k}\otimes \beta_{k}\right)\right) & =
@@
  \end{align*}
+}
 
@@
 For the last equation, we only check it on basic tensors:
+{\small
 \begin{align*}

### 2023-01-19 section4.tex
commit

@@
 \begin{proof}
- It follows after adaptating [Theorem , \cite{}] in order to prove that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
+ It follows after adapting [Theorem , \cite{}] in order to prove that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
 \end{proof}
@@
 We conclude this subsection by noticing that $\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
-Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete \emph{normed cone} %(see~\cite{Selinger2004}, or the appendix, for such notions}.
+Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete \emph{normed cone} (see~\cite{Selinger2004}, or the appendix, for such notions).
 Its cone structure induces an order on it, called its \emph{cone order}:
@@
 This makes it a Scott-continuous dcpo.
-Suitable categories of cones have been recently investigated as models of probabilistic computation %(\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
+Suitable categories of cones have been recently investigated as models of probabilistic computation (\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
 We will just show that:

### 2023-01-20 main.tex
commit
Commit
Commit correzioni Ugo

@@
 \usepackage{graphicx}
+\usepackage{bussproofs}
+\EnableBpAbbreviations
 \usepackage{hyperref}
+\usepackage{tikz,pgfplots}
+\pgfplotsset{compat=newest}
 \usepackage{tikz-cd}
@@
 \usepackage{amsthm}
+\usepackage{caption}
+\usepackage{subcaption}
 \usetikzlibrary{cd}
@@
 
-\title{Tropical Mathematics for the Lambda Calculus}
+\title{Tropical Mathematics and Lambda Calculus}
 
@@
 program metrics, as based on the analysis of program sensitivity via Lipschitz-conditions, and for resource analysis, as based on higher-order program differentiation.
-We discuss applications of this tropical semantics for ``best case'' resource analysis as well as for other quantitative properties like log-probabilities and differential privacy.
+We discuss applications of this tropical semantics to ``best case'' resource analysis as well as to other quantitative properties like log-probabilities and differential privacy.
 Finally, we show that a general foundation for this semantic approach is provided by an abstract correspondence between Lawvere's theory  
@@
 %\section{Tropical (non-)Linear Algebra and computation}
-\subsection{Tropical linear algebra via the relational model% and computation
+\subsection{Tropical linear algebra and the relational model% and computation
 }
 
-\input{section3intro}%
+\input{section3intro}
 
 %\subsection{As a model of the \emph{linear} $\STLC$}
-\section{Tropical models of several $\lam$-calculi}\label{section3}
+\section{Tropical Semantics of Several $\lam$-Calculi}\label{section3}
 \subsection{Prohibited duplication/erasure: the \emph{linear} $\STLC$}
@@
 \subsection{Allowed duplication/erasure: the $\STLC$}
-\input{section3B}%
+\input{section3B}
 
@@
 \subsection{Controlled duplication/erasure: the $\STDLC$}
-\input{section3C}%
+\input{section3C}
 
-\section{Tropical Laurent series}\label{section4}
-\input{section4}%
+\section{Tropical Laurent Series}\label{section4}
+\input{section4}
 
 \section{Applications}\label{sec:app}\label{section5}
-\input{section5}%
+\input{section5}
+\input{section5C}
 
@@
 
-\section{Conclusion and future work}\label{section8}
+\section{Conclusion and Future Work}\label{section8}
 \input{section8}
@@
 
+\onecolumn
+\appendix
+
+\subsection{Proofs from Section III}
+\input{appIII}
+\newpage
+
+\subsection{Proofs from Section IV}
+\input{appIV}
+
+\newpage
+\subsection{Proofs from Section V}
+\input{appV}
+
+\newpage
+\subsection{Proofs from Section VII}
+\input{appVII}
+
 \end{document}

### 2023-01-20 section1.tex
Commit correzioni Ugo

--- 2023-01-17 section1.tex
@@
-In recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability,
+In recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of programs, such as e.g.~number of computation steps, convergence probability,
 approximation bounds, 
-as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence. 
+as opposed to purely \emph{qualitative} properties like e.g.~termination or equivalence. 
 Notably, a significative effort has been made to extend, or adapt, well-established qualitative methods, like e.g.~type systems, relational logics or denotational semantics, to account for quantitative properties.
@@
 probabilistic \cite{Ehrhard2011, Staton2017} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
-The main reason to look for methods relying on (quantitative extensions of) type-theory or denotational semantics is that these approaches yield \emph{modular} and \emph{compositional} techniques, that is, allow one to deduce properties of complex algorithms from the properties of their component parts.
+The main reason to look for methods relying on (quantitative extensions of) type-theory or denotational semantics is that these approaches yield \emph{modular} and \emph{compositional} techniques, that is, allow one to deduce properties of complex programs from the properties of their constituent parts.
 
@@
 
-Among such quantitative approaches, two different directions have received a considerable attention.
+Among such quantitative approaches, two different directions have received considerable attention.
 
-On the one hand there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}.
+On the one hand one should certainly mention \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function (which is rarely the case), it is more useful to know whether they compute functions which do not differ \emph{too much}.
 This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
 
-On the other hand there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected with the \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
-This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable to capture probabilistic properties \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
+On the other hand there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
+This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for probabilistic settings \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
 
 In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using its input exactly once), plays a crucial role.
-In the metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
+In metric semantics, linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
 By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
 
-A natural question is thus whether these two, apparently unrelated, ways of interpreting linearity and duplication can be somehow connected.
+A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow connected.
 At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
-in metric models $n$ times duplication results in a linear (hence Lipschitz) function $n\cdot x$, while in differential models this results in a polynomial $x^{n}$, which is not Lipschitz.
+in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} $x^{n}$, hence not Lipschitz.
 The fundamental insight of this article is that a natural way to overcome
@@
 
-Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} (whose origin stands behind the adjective ``tropical'') as a way to do algebra and geometry by replacing the usual ring structure of numbers based on addition and multiplication by the semi-ring structure given by ``$\min$'' and ``$+$''.
+Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} as the study of algebra and geometry where one replaces the usual ring structure of numbers based on addition and multiplication by the semi-ring structure given by ``$\min$'' and ``$+$''.
 %
@@
 
-As working on the idempotent $\min,+$-semiring is often easier that working with addition and multiplication, the tropicalization of  algebraic concepts makes several intractable properties (e.g.~finding roots of polynomials, or studying geometric invariants like e.g.~the genus) become tractable.
-For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).
+This is not a \emph{ad-hoc} setting: tropical geometry turns out to be a rich and florid combinatorial counterpart of usual algebraic geometry, with important connections with optimisation theory.
+Furthermore, computationally speaking, working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, the important and generally intractable problem of finding roots of a polynomial, admits a linear time algorithm in the tropical case (and the roots of a tropicalized polynomial can be used to approximate the actual roots of the polynomial \cite{Noferini2015}).
+Such a computational nature of tropical concepts explains why these are also widely applied in computer science, e.g.~for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).
 
@@
 At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.
-Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}, have also been observed \cite{Fuji}.
+Furthermore, connections between tropical linear algebra and metric spaces (in the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}), have also been observed \cite{Fuji}.
 
@@
 This bridge does not only provide a conceptual unification of two different quantitative approaches to higher-order algorithms, but it also
-suggests ways in which techniques for resource-analysis could be used for sensitivity analysis and vice-versa, opening the way for new potential applications of tropical geometry to the study of quantitative properties of higher-order programs.
+suggests ways in which techniques from resource-analysis could be used in sensitivity analysis and \emph{vice-versa}, opening the way for new potential applications of tropical geometry to the study of quantitative properties of higher-order programs.
 
@@
 \item We study the relational model over the tropical semi-ring, showing that
-\emph{bounded} simply typed terms correspond to tropical polynomials, while general
+\emph{bounded} simply typed terms are interpreted as tropical polynomials, while general
  simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{Porzio2021}.
@@
 
-\item Using the relational model as our main reference, we suggest a few potential applications of tropical methods to the study of quantitative properties of non-deterministic/probabilistic functional programs like counting best-case computation steps,
+\item Using the relational model as our main source of intuition, we suggest a few potential applications of tropical methods to the study of quantitative properties of non-deterministic and probabilistic functional programs like counting best-case computation steps,
 measuring convergenge log-probabilities, and 
@@
 
-\item We conclude by putting the correspondence between differential and metric properties made possible by the tropical viewpoint at the right level of generality. 
+\item We conclude by considering a more general tropical setting between differential and metric properties.
 By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007}, we show that the category of \emph{complete}  GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.

### 2023-01-20 section2.tex
Commit correzioni Ugo

@@
 
-In this section we discuss in some more detail the two approaches in quantitative semantics recalled in the Introduction, at the same time providing an overview of how we aim at bridging them using tropical mathematics.
+In this section we discuss in some more detail the two approaches to quantitative semantics we mentioned in the Introduction, at the same time providing an overview of how we aim at bridging them using tropical mathematics.
 
@@
 
-As recalled in the Introduction, in many situations (typically, when dealing with computationally expensive function) one does not look for algorithms to compute some function \emph{exactly}, but rather to approximate it (e.g.~in an efficient way) within some error bound.
-In other common situations (typically, in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, that a small error in the input will produce a comparably small error in the output.
+As recalled in the Introduction, in many situations (e.g.~when dealing with computationally expensive functions) one does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound.
+In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, a small error in the input will produce a comparably small error in the output.
 
-A natural way to account for these situations is to consider denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which measures differences of behavior.
-A fundamental insight coming from this line of works is that \emph{affine} programs, that is, programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
+A natural way to account for these situations is to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which measures differences of behavior.
+A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
 $d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
@@
 
-As observed in \cite{Reed2010, Gaboardi2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications.
+As observed in \cite{Reed2010, Gaboardi2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplication.
 The fundamental intuition is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz} map.
@@
 
-These observations have led to the study of \emph{bounded} typed $\lambda$-calculi, like the system $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied in the study of higher-order differential privacy \cite{Gaboardi2013, Gaboardi2017}.
+These observations have led to the study of \emph{bounded} typed $\lambda$-calculi, like the system $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied in the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}.
 The types of such systems are defined by combining linear constructors with a \emph{graded exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
@@
 k\cdot x: \BB R\to \BB R$ are all Lipschitz-continuous, with Lipschitz constant $k$, the function $M=\lambda x.x^{2}$ obtained by ``duplicating'' $x$ is not Lipschitz anymore: $M$ is, so to say, \emph{too} sensitive to errors.
-More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of $\STLC$ (yet, $\Met$ contains several cartesian closed \emph{sub-}categories, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
+More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of $\STLC$ (yet, several cartesian closed \emph{sub-}categories of $\Met$ do exist, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
 Still, one might observe that the program $M$ above is actually Lipschitz-continuous, if not globally, at least \emph{locally} (i.e.~over any compact set).
@@
 \begin{itemize}
-\item $\C C$ is left-additive, i.e.~its hom-sets are commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
+\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
 \item $\C C$ is equipped with a differential operator $D:
-\C C(X,Y)\to \C C(X\times X,Y)$ satisfying 7 axioms $D1,\dots, D7$, which translate usual properties of the differential (e.g.~the linearity of $D$ in one of its two variables, the chain rule,  Schwarz's theorem etc.).
+\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule,  Schwarz's theorem etc.).
 \end{itemize}
@@
 
-Another suggestive similarity between program derivatives and 
+Another intriguing similarity between program derivatives and 
 actual derivatives is provided by the \emph{Taylor expansion}:
 in $\STDLC$ one can prove that any application $MN$ can be expanded as a sum of \emph{linear} applications
-$\Der^{(k)}[M]\cdot N^{k}$, i.e.~where $N$ is passed exactly $k$ times to $M$, as expressed below:
+$\Der^{(k)}[M]\cdot N^{k}$, i.e.~where $N$ is passed exactly $k$ times to $M$, expressed as:
 \begin{align}\label{eq:taylor}
@@
 \end{align}
-In other words, terms with unbounded duplications can be seen as some sort of limits of bounded, but arbitrarily large, duplications.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
+In other words, unbounded duplications are seen as a sort of limit of bounded, but arbitrarily large, ones.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
 
@@
 may somehow be seen as a decomposition of a $\lambda$-term 
-into a limit of Lipschitz maps, thus bridging the metric and differential approach.  
+into a limit of Lipschitz maps, thus bridging the metric and differential approaches.  
 While suggestive, an obstacle to realize this idea is that, as anticipated in the Introduction, 
@@
 Our question can thus be reformulated as follows: can we make the relational semantics \emph{Lipschitz}, hence amenable to metric and sensitivity analysis?
-The goal of this paper is to show that, by appealing to tropical mathematics, this is indeed possible and leads to a surprising
-bridge between the metric and differential study to higher-order programs.
+The goal of this paper is to show that, by appealing to tropical mathematics, this is indeed possible and leads to the somehow unespected discovery of a bridge between the metric and differential study to higher-order programs.
 %

### 2023-01-20 section1.tex
Super tiny committino

@@
 The goal of this paper is to study the tropical semantics of the differential $\lambda$-calculus and use it to connect it with metric semantics.
-This bridge does not only provide a conceptual unification of two different quantitative approaches to higher-order algorithms, but it also
+This bridge does not only provide a conceptual unification of two different quantitative approaches to higher-order programs, but it also
 suggests ways in which techniques from resource-analysis could be used in sensitivity analysis and \emph{vice-versa}, opening the way for new potential applications of tropical geometry to the study of quantitative properties of higher-order programs.

### 2023-01-20 section4.tex
Super tiny committino

@@
 \begin{proof}
-As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be finite  and  its corresponding linear map $\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Prop.~\ref{prop:polylip}.
+As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be a finite matrix  and  its corresponding linear map $\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Prop.~\ref{prop:polylip}.
 Claim 2) follows directly from Theorem~\ref{thmTLSlocLip}.

### 2023-01-21 section1.tex
Commit notturno sonnolento

@@
 
-\item We study the relational model over the tropical semi-ring, showing that
+\item In Section~\ref{section3}, we study the relational model over the tropical semi-ring% showing in particular that it is a model of $\STDLC$;
+and in Section~\ref{section4} we show that
 \emph{bounded} simply typed terms are interpreted as tropical polynomials, while general
- simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{Porzio2021}.
-Moreover, we show that such maps are locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
+ simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{Porzio2021} which, moreover, are locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
 %Moreover, we exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps.
 
-\item Using the relational model as our main source of intuition, we suggest a few potential applications of tropical methods to the study of quantitative properties of non-deterministic and probabilistic functional programs like counting best-case computation steps,
+\item Using the relational model as our main source of intuition, in Section~\ref{section5} we suggest a few potential applications of tropical methods to the study of quantitative properties of non-deterministic and probabilistic functional programs like counting best-case computation steps,
 measuring convergenge log-probabilities, and 
@@
 
-\item We conclude by considering a more general tropical setting between differential and metric properties.
+\item We conclude by considering, in Section~\ref{section6}, a more general tropical setting between differential and metric properties.
 By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007}, we show that the category of \emph{complete} GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.

### 2023-01-21 section2.tex
Commit notturno sonnolento

@@
 
-\subsection{Outline of the paper}
+\subsection{{\color{red}Spoiler section ?}}%Outline of the paper}
 
-In the next section we recall some basic ideas from tropical mathematics, and its connection with the study of the Lawvere quantale.
+Fino a qui niente a capo
+
+Fino a qui niente a capo
+
+Fino a qui niente a capo
+
+Fino a qui niente a capo
+
+Fino a qui niente a capo
+
+Limite! Alla prossima riga sforo oltre 12 pag.
+
+%In the next section we recall some basic ideas from tropical mathematics, and its connection with the study of the Lawvere quantale.
 %Since polynomials correspond to piecewise linear (hence Lipschitz) functions in tropical mathematics,
-The reconstruction of the relational semantics over the tropical semi-ring, presented in Section \ref{section3} and Section \ref{section4}, will provide a metric semantics of the differential $\lambda$-calculus, bridging sensitivity and resource analysis.
-In Section \ref{section5} we suggest potential applications of this approach, relating well-studied applications of program metrics, resource analysis with current uses of tropical mathematics in computer science.
-Finally, in Section \ref{section6} we show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between
-{generalized metric spaces} and modules over the tropical semiring.
+%The reconstruction of the relational semantics over the tropical semi-ring, presented in Section \ref{section3} and Section \ref{section4}, will provide a metric semantics of the differential $\lambda$-calculus, bridging sensitivity and resource analysis.
+%In Section \ref{section5} we suggest potential applications of this approach, relating well-studied applications of program metrics, resource analysis with current uses of tropical mathematics in computer science.
+%Finally, in Section \ref{section6} we show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between {generalized metric spaces} and modules over the tropical semiring.
 

### 2023-01-21 section3C.tex
Commit notturno sonnolento

@@
 
-Let us show the differential structure of $\LREL_!$ (remember that the Cartesian product of $\LREL_!$ is the disjoint union $+$):
+Let us show the differential structure of $\LREL_!$ (remember that the Cartesian product of $\LREL_!$ is the disjoint union $+$).
 
@@
 
-There is more: not only $\LREL_!$ is a model of the $\STDLC$, but this model is also well-behaved w.r.t.\ to the Taylor expansion of $\lam$-terms.
-This is expressed by the following property (see [Definition 4.22, \cite{Manzo2012}], where it is called the fact of ``modeling the Taylor expansion'').
+This ensures that one can define a sound interpretation of $\STDLC$-terms in the standard way (see [Section 4.3, \cite{Manzo2010}]).
+
+There is more: this model is also well-behaved w.r.t.\ to the Taylor expansion, as expressed by the following property (see [Definition 4.22, \cite{Manzo2012}]%, where it is called the fact of ``modeling the Taylor expansion''
+).
 
@@
  In $\LREL_!$ equipped with $D$, all morphisms can be Taylor expanded, i.e.\ in $\LREL_!$ we have that, for all
-$t\in\HOM{\LREL_!}{Z}{X\multimap Y}$ and $s\in\HOM{\LREL_!}{Z}{X}$, the evaluation of $t$ over $s$ yields 
+$t\in\HOM{\LREL_!}{Z}{X\multimap Y}$ and $s\in\HOM{\LREL_!}{Z}{X}$ we have:%, the evaluation of $t$ over $s$ yields 
  \begin{align}\label{eq:taylorcat}
@@
  \end{theorem}
-It is worth discussing the formula above a bit more. 
-Here, $\Lambda^-$
+%It is worth discussing the formula above a bit more. 
+Here,
 %:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
- is the uncurry operator and 
 %$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
-$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to linear application of $s$ to $u$.
-Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$,  i.e.~it coincides with the tropical interpretation of the Taylor formula \eqref{eq:taylor}.
-In other words, \eqref{eq:taylorcat} says that the series corresponding to the Taylor expansion of an ordinary $\lam$-term $M$ converges to $M$ [Theorem 4.23, \cite{Manzo2012}].
+$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to linear application of $s$ to $u$, and $\Lambda^-$ is the uncurry operator.
+Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$,  i.e.~it coincides with the tropical %interpretation of the
+version of the usual Taylor expansion.
+Since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
+
+As a consequence, we have:
+
+\begin{corollary}\label{thm:M=infLip}
+ If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is the $\inf$ of \emph{Lipschitz} functions.
+\end{corollary}
 

### 2023-01-21 section4.tex
Commit notturno sonnolento
Commit

@@
 The norm $\norm{\cdot}_\infty$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$.
-A consequence of Theorem~\ref{theorem:fepsilon} is that tLs with finitely many variables are locally Lipschitz on all $\BB R_{>0}$.
-In this subsection we will refine and generalise this result.
+
+As a first remark, let us mention that an immediate consequence of the lines after Theorem~\ref{thm:modelsTaylor} is:
+
+\begin{corollary}\label{thm:M=infLip}
+ If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is the $\inf$ of \emph{Lipschitz} functions.
+\end{corollary}
+
+Another remark is that, as consequence of Theorem~\ref{theorem:fepsilon}, the tLs with finitely many variables are locally Lipschitz on all $\BB R_{>0}$.
+In this subsection we refine and generalise this kind of results.
 %
@@
 \begin{enumerate}
-\item if $\Gamma \vdash_{\BSTLC} M:A$, then $\model M:\model\Gamma \to \model A$ is a Lipschitz map.
-\item if $\Gamma \vdash_{\STLC}M:A$, then $\model M: \model\Gamma \to \model A$ is a locally Lipschitz map.
+\item if $\Gamma \vdash_{\BSTLC} M:A$, then $\model M^!:\model\Gamma \to \model A$ is a Lipschitz map.
+\item if $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is a locally Lipschitz map.
 \end{enumerate}

### 2023-01-21 section1.tex
commit

@@
 
-\item In Section~\ref{section3}, we study the relational model over the tropical semi-ring% showing in particular that it is a model of $\STDLC$;
-and in Section~\ref{section4} we show that
+\item we study the relational model over the tropical semiring, in which
 \emph{bounded} simply typed terms are interpreted as tropical polynomials, while general
  simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{Porzio2021} which, moreover, are locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
+This is in Sections \ref{section3} and \ref{section4}.
 %Moreover, we exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps.
 
-\item Using the relational model as our main source of intuition, in Section~\ref{section5} we suggest a few potential applications of tropical methods to the study of quantitative properties of non-deterministic and probabilistic functional programs like counting best-case computation steps,
-measuring convergenge log-probabilities, and 
-differential privacy.
+\item Using the relational model as our main source of intuition,  we suggest a few potential applications of tropical methods to the study of quantitative properties of non-deterministic and probabilistic functional programs like counting best-case computation steps,
+measuring convergence log-probabilities, and 
+differential privacy. This is in Section~\ref{section5}
 
-\item We conclude by considering, in Section~\ref{section6}, a more general tropical setting between differential and metric properties.
+\item We conclude 
+by putting the connection between
+tropical, differential and metric properties at the right level of generality.
 By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007}, we show that the category of \emph{complete} GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
+This is in Section~\ref{section6}.
 \end{itemize}

### 2023-01-21 section2.tex
commit

@@
 
-A natural way to account for these situations is to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which measures differences of behavior.
+In all these cases, it is common to consider denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which measures differences of behavior. 
 A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
@@
 \item $\C C$ is equipped with a differential operator $D:
-\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule,  Schwarz's theorem etc.).
+\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
 \end{itemize}
@@
 
-At this point, given our previous discussion of bounded duplications and Lipschitz-continuity, it is natural to ask whether the Taylor formula
-may somehow be seen as a decomposition of a $\lambda$-term 
-into a limit of Lipschitz maps, thus bridging the metric and differential approaches.  
-While suggestive, an obstacle to realize this idea is that, as anticipated in the Introduction, 
-%in general, a $C\partial C$ needs not satisfy the Taylor formula
-%\eqref{eq:taylor}, and 
-in the canonical ``Taylor'' semantics for $\STDLC$, the relational semantics, the bounded terms appearing in the Taylor expansion correspond to \emph{polynomials}, and thus to non-Lipschitz functions.
-
-Our question can thus be reformulated as follows: can we make the relational semantics \emph{Lipschitz}, hence amenable to metric and sensitivity analysis?
-The goal of this paper is to show that, by appealing to tropical mathematics, this is indeed possible and leads to the somehow unespected discovery of a bridge between the metric and differential study to higher-order programs.
 %
@@
 
-\subsection{{\color{red}Spoiler section ?}}%Outline of the paper}
+\subsection{Tropical Mathematics: a possible synthesis}%Outline of the paper}
 
-Fino a qui niente a capo
+At this point, as the Taylor formula decomposes an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as decomposing  a $\lambda$-term
+into a limit of Lipschitz maps, thus bridging the metric and differential approaches.  
+A natural direction to look for here is the relational semantics,  the somehow canonical ``Taylor'' semantics for $\STDLC$. 
+However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.
 
-Fino a qui niente a capo
+Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions?
+This way, eq.~\eqref{eq:taylor} could really be interpreted as a decomposition of $\lambda$-terms via Lipschitz maps.
+In other words, unbounded term application could be seen
+as a limit of \emph{more and more sensitive} operations. 
+This viewpoint, that we develop in the following sections, leads to the somehow unexpected discovery of a bridge between the metric and differential study of higher-order programs.
+This connection not only suggest the application of optimization methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions, but it scales to a 
+more abstract level, leading to introduce a 
+differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}).
 
-Fino a qui niente a capo
-
-Fino a qui niente a capo
-
-Fino a qui niente a capo
-
-Limite! Alla prossima riga sforo oltre 12 pag.
+%
+%At this point, 
+%by interpreting such polynomials over the tropical semiring 
+%
+%Our question can thus be reformulated as follows: can we make the relational semantics \emph{Lipschitz}, hence amenable to metric and sensitivity analysis?
+The goal of this paper is to show that, by appealing to tropical mathematics, this is indeed possible and leads to the somehow unespected discovery of a bridge between the metric and differential study to higher-order programs.
+%
+%Fino a qui niente a capo
+%
+%Fino a qui niente a capo
+%
+%Fino a qui niente a capo
+%
+%Fino a qui niente a capo
+%
+%Fino a qui niente a capo
+%
+%Limite! Alla prossima riga sforo oltre 12 pag.
+%
 

### 2023-01-21 appIV.tex
Commit pre-pranzo

@@
 \end{itemize}
+
+Let us end this section by recalling the sound interpretation of $\STDLC$-terms in a $C\partial C$ category (which is taken from [Section 4.3, \cite{Manzo2010}]).
+
+\begin{definition}
+Let $\C C$ be a $C\partial C$ and $\model A$ fixed objects of $\C C$, for each ground type $A$ of the $\STDLC$.
+
+One lifts as usual the interpretation to types and contexts as $\model{A\to B}:=\model A\to \model B$ and $\model{x_1:A_1,\dots,x_n:A_n}:=\Pi_{i=1}^n\model{x_i:A_i}$.
+
+Then, one inductively defines an interpretation $\model{\Gamma \vdash M:A}\in\C C(\model \Gamma,\model A)$ of a $\STDLC$-term $\Gamma \vdash M:A$ (that can can be proved to be sound) by induction as:
+\begin{itemize}
+ \item $\model{\Gamma, x:A \vdash x:A}:=\pi_2$
+ \item $\model{\Gamma, x:A \vdash y:B}:=\model{\Gamma \vdash y:B}\circ \pi_1\quad\textit{ if } y\neq x$
+ \item $\model{\Gamma\vdash \lambda x.M:A\to B}:=\Lambda(\model{\Gamma,x:A\vdash M:B})$
+ \item $\model{\Gamma \vdash M\BB T:B}:=\mathsf{ev} \circ \langle \, \model{\Gamma \vdash M:A\to B}, \model{\Gamma \vdash \BB T:A} \, \rangle$
+ \item $\model{\Gamma \vdash \Der [M] \cdot N:A\to B}:=\Lambda\left(\Lambda^-(\model{\Gamma \vdash M:A\to B})\star \model{\Gamma \vdash N:A}\right)$
+ \item $\model{\Gamma \vdash 0:A}:=\infty$
+ \item $\model{\Gamma \vdash M+\BB T:A}:=\model{\Gamma \vdash M:A}+\model{\Gamma \vdash \BB T:A}$
+\end{itemize}
+\end{definition}
+
+In the main paper we write $\model M$, but of course we mean $\model{\Gamma \vdash M:A}$.

### 2023-01-21 appV.tex
commit

@@
 
-A function $f:Q^{X}\to Q^{Y}$ is \emph{concave} if for all $\alpha\in [0,1]$, $\B x, \B y\in Q^{X}$ and $b\in Y$ 
+A function $f:Q^{X}\to Q^{Y}$ is \emph{concave} if for all $\alpha\in [0,1]$, $ x ,  y  \in Q^{X}$ and $b\in Y$ 
 $$
-f(\alpha\cdot \B x+(1-\alpha)\cdot \B y)_{b} \geq \alpha f(\B x)_{b} + (1-\alpha)f(\B y)_{b}
+f(\alpha\cdot  x +(1-\alpha)\cdot  y  )_{b} \geq \alpha f( x )_{b} + (1-\alpha)f( y  )_{b}
 $$
@@
 \begin{proof}
-Let us first show that all functions of the form $f(\B x)_{b}= \mu \B x+ c$ are concave:
-we have $f(\alpha\B x+ (1-\alpha)\B y)_{b}= \mu(\alpha\B x)+(1-\alpha)\B y)+c=
- \mu(\alpha\B x)+(1-\alpha)\B y)+\alpha c+(1-\alpha)c=
- \alpha(\mu \B x + c)+(1-\alpha)(\mu \B y+c)=\alpha f(\B x)_{b}+(1-\alpha) f(\B x)_{b}$.
+Let us first show that all functions of the form $f( x )_{b}= \mu  x + c$ are concave:
+we have $f(\alpha x + (1-\alpha) y  )_{b}= \mu(\alpha x )+(1-\alpha) y  )+c=
+ \mu(\alpha x )+(1-\alpha) y  )+\alpha c+(1-\alpha)c=
+ \alpha(\mu  x  + c)+(1-\alpha)(\mu  y  +c)=\alpha f( x )_{b}+(1-\alpha) f( x )_{b}$.
 
 To conclude, let us show that if $(f_{i})_{i\in I}$ is a family of concave functions from $Q^{X}$ to $Q^{Y}$, the function $f=\inf_{i\in I}f_{i}$ is also concave: we have
-$f(\alpha\B x +(1-\alpha)\B y)_{b}=
-\inf_{i\in I}f_{i}(\alpha\B x+(1-\alpha)\B y)_{b} \geq 
-\inf_{i\in I}\alpha f_{i}(\B x)_{b}+(1-\alpha)f_{i}(\B y)_{b}
+$f(\alpha x  +(1-\alpha) y  )_{b}=
+\inf_{i\in I}f_{i}(\alpha x +(1-\alpha) y  )_{b} \geq 
+\inf_{i\in I}\alpha f_{i}( x )_{b}+(1-\alpha)f_{i}( y  )_{b}
 \geq 
-\inf_{i\in I}\alpha f_{i}(\B x)_{b} + \inf_{j\in I}(1-\alpha)f_{j}(\B y)_{b}
+\inf_{i\in I}\alpha f_{i}( x )_{b} + \inf_{j\in I}(1-\alpha)f_{j}( y  )_{b}
 =
-\alpha \cdot (\inf_{i\in I}f_{i}(\B x)_{b})+ (1-\alpha)\cdot( \inf_{j\in I}f_{j}(\B y)_{b})=
-\alpha  f(\B x)_{b}+(1-\alpha)f(\B y)_{b}$, where we used the fact that given families $a_{i},b_{i}$ of reals,
+\alpha \cdot (\inf_{i\in I}f_{i}( x )_{b})+ (1-\alpha)\cdot( \inf_{j\in I}f_{j}( y  )_{b})=
+\alpha  f( x )_{b}+(1-\alpha)f( y  )_{b}$, where we used the fact that given families $a_{i},b_{i}$ of reals,
 $\inf_{i}a_{i}+b_{i}\geq \inf_{i}a_{i}+\inf_{j}b_{j}$.
@@
 
-Concavity is a useful property, as it leads to establish the following:
-\begin{proposition}
-For any tropical function $f: Q^{X}\to Q^{Y}$, 
-\begin{enumerate}
-\item $ \alpha f(\B x)_{b}\leq  f(\alpha\B x)_{b}$;
-\item $f(\B x+\B y)_{b}\leq f(\B x)_{b}+f(\B y)_{b}$;
-\item $f(\B x+\B y)_{b} \leq f(\B x)_{b} + \phi_{\mu,b}(\B y)$ 
-for all $\mu\in \C M_{\mathrm{fin}}(X)$.
+%
+%
+%Concavity is a useful property, as it leads to establish the following:
+%\begin{proposition}
+%For any tropical function $f: Q^{X}\to Q^{Y}$, 
+%\begin{enumerate}
+%\item $ \alpha f( x )_{b}\leq  f(\alpha x )_{b}$;
+%\item $f( x + y  )_{b}\leq f( x )_{b}+f( y  )_{b}$;
+%\item $f( x + y  )_{b} \leq f( x )_{b} + \phi_{\mu,b}( y  )$ 
+%for all $\mu\in \C M_{\mathrm{fin}}(X)$.
+%
+%\item if $ x  <  y   <  z  $, then 
+%$$
+%\frac{f( y  )-f( x )}{| y  - x |} \leq \frac{f( z  )-f( x )}{| z  - x |}
+%\leq \frac{f( z  )-f( y  )}{| z  - y  |}
+%$$
+%
+%\item the function $R_{b}( x ,  y  )= \frac{f( x )_{b}-f( y  )_{b}}{|  x -  y  |}$ is monotonically non-decreasing in $ x $ (for fixed $ y  $) and in $ y  $ (for fixed $ x $).
+%
+%\end{enumerate}
+%\end{proposition}
+%\begin{proof}
+%Since $f$ is concave, we have that for all $ z  $, 
+%$f(\alpha\cdot z  )=f(\alpha\cdot z  +(1-\alpha)\cdot 0)\geq 
+%\alpha f( z  )+(1-\alpha)f(0)\geq \alpha f( z  )$. 
+%
+%For the second one, 
+%if $f( x )=\mu  x +c$, then $f( x + y  )= \mu  x + \mu  y   + c  \leq
+%\mu  x  + \mu  y   + 2c = f( x )+f( y  )$.
+%Moreover, if $(f_{i})_{i\in I}$ is a family of functions such that $f_{i}( x + y  )\leq f_{i}( x )+f_{i}( y  )$, then 
+%$f=\inf _{i\in I}f_{i}$ also satisfies this property: 
+%one has $f( x + y  )=\inf_{i\in I}f_{i}( x + y  ) \geq \inf_{i} f_{i}( x )+f_{i}( y  ) \geq \inf_{i}f_{i}( x ) + \inf_{i}f_{i}( y  )=f( x )+f( y  )$.
+%%Now we can compute
+%%\begin{align*}
+%%f( x + y  )_{b}& =
+%%
+%%\end{align*}
+%
+%The third one follows immediately from the second one, since $f( y  )_{b}\leq \phi_{\mu,b}( y  )$ for all $\mu\in \C M_{\mathrm{fin}}(X)$.
+%
 
-\item if $\B x < \B y < \B z$, then 
-$$
-\frac{f(\B y)-f(\B x)}{|\B y-\B x|} \leq \frac{f(\B z)-f(\B x)}{|\B z-\B x|}
-\leq \frac{f(\B z)-f(\B y)}{|\B z-\B y|}
-$$
+%\end{proof}
 
-\item the function $R_{b}(\B x, \B y)= \frac{f(\B x)_{b}-f(\B y)_{b}}{| \B x- \B y|}$ is monotonically non-decreasing in $\B x$ (for fixed $\B y$) and in $\B y$ (for fixed $\B x$).
-
-\end{enumerate}
-\end{proposition}
-\begin{proof}
-Since $f$ is concave, we have that for all $\B z$, 
-$f(\alpha\cdot\B z)=f(\alpha\cdot\B z+(1-\alpha)\cdot 0)\geq 
-\alpha f(\B z)+(1-\alpha)f(0)\geq \alpha f(\B z)$.
-
-For the second one, 
-if $f(\B x)=\mu \B x+c$, then $f(\B x+\B y)= \mu \B x+ \mu \B y + c  \leq
-\mu \B x + \mu \B y + 2c = f(\B x)+f(\B y)$.
-Moreover, if $(f_{i})_{i\in I}$ is a family of functions such that $f_{i}(\B x+\B y)\leq f_{i}(\B x)+f_{i}(\B y)$, then 
-$f=\inf _{i\in I}f_{i}$ also satisfies this property: 
-one has $f(\B x+\B y)=\inf_{i\in I}f_{i}(\B x+\B y) \geq \inf_{i} f_{i}(\B x)+f_{i}(\B y) \geq \inf_{i}f_{i}(\B x) + \inf_{i}f_{i}(\B y)=f(\B x)+f(\B y)$.
-%Now we can compute
-%\begin{align*}
-%f(\B x+\B y)_{b}& =
-%
-%\end{align*}
-
-The third one follows immediately from the second one, since $f(\B y)_{b}\leq \phi_{\mu,b}(\B y)$ for all $\mu\in \C M_{\mathrm{fin}}(X)$.
-
-\end{proof}
-
-\newpage
+%\newpage
 
@@
 \begin{definition}
- Let $\preceq$ be the pointwise order on $\N^k$ (i.e.\ for all $\mathbf m,\mathbf n\in \N^{K}$, $\mathbf m\preceq \mathbf n$ iff $m_{i}\leq n_{i}$ for all $1\leq i\leq K$).
- Of course $\mathbf m\prec \mathbf n$ holds exactly when $\mathbf m\preceq \mathbf n$ and $m_{i}<n_{i}$ for at least one $1\leq i\leq K$.
- Finally, we set $\mathbf m\prec_{1}\mathbf n$  iff
-$\mathbf m\prec \mathbf n$ and $\sum_{i=1}^{K}n_{i}-m_{i}=1$ (i.e.\ they differ on exactly one coordinate).
+ Let $\preceq$ be the pointwise order on $\N^k$ (i.e.\ for all $ m  , n  \in \N^{K}$, $ m  \preceq  n  $ iff $m_{i}\leq n_{i}$ for all $1\leq i\leq K$).
+ Of course $ m  \prec  n  $ holds exactly when $ m  \preceq  n  $ and $m_{i}<n_{i}$ for at least one $1\leq i\leq K$.
+ Finally, we set $ m  \prec_{1} n  $  iff
+$ m  \prec  n  $ and $\sum_{i=1}^{K}n_{i}-m_{i}=1$ (i.e.\ they differ on exactly one coordinate).
 \end{definition}
@@
 \begin{remark}\label{rmk:AC}
-\text{If $U\subseteq \N^{K}$ is infinite, then $U$ contains an infinite ascending chain $\mathbf m_{0}\prec \mathbf m_{1} \prec \mathbf m_{2} \prec \dots$.}.
+\text{If $U\subseteq \N^{K}$ is infinite, then $U$ contains an infinite ascending chain $ m  _{0}\prec  m  _{1} \prec  m  _{2} \prec \dots$.}.
 
-This is a consequence of K\"onig Lemma (KL): consider the directed acyclic graph $(U,\prec_{1})$, indeed a $K$-branching tree; if there is no infinite ascending chain $ \mathbf m_{0}\prec \mathbf m_{1} \prec \mathbf m_{2} \prec \dots$, then in particular there is no infinite ascending chain $ \mathbf m_{0}\prec_{1} \mathbf m_{1} \prec_{1} \mathbf m_{2} \prec_{1} \dots$ so the tree $U$ has no infinite ascending chain; then by KL it is finite, contradicting the assumption.
+This is a consequence of K\"onig Lemma (KL): consider the directed acyclic graph $(U,\prec_{1})$, indeed a $K$-branching tree; if there is no infinite ascending chain $  m  _{0}\prec  m  _{1} \prec  m  _{2} \prec \dots$, then in particular there is no infinite ascending chain $  m  _{0}\prec_{1}  m  _{1} \prec_{1}  m  _{2} \prec_{1} \dots$ so the tree $U$ has no infinite ascending chain; then by KL it is finite, contradicting the assumption.
 \end{remark}
@@
  \item $\mathcal{F}_\epsilon$ is finite
- \item If $\mathcal{F}_\epsilon= \emptyset$ then $f(\mathbf x) = +\infty$ for all $\mathbf x\in \overline{\R}_{\geq 0}^{\{1,\dots, K\}}$
- \item If $f(\mathbf x_0) = +\infty$ for some $\mathbf x_0\in [\epsilon,\infty)^{K}$ then $\mathcal{F}_\epsilon= \emptyset$
+ \item If $\mathcal{F}_\epsilon= \emptyset$ then $f( x ) = +\infty$ for all $ x \in \overline{\R}_{\geq 0}^{\{1,\dots, K\}}$
+ \item If $f( x _0) = +\infty$ for some $ x _0\in [\epsilon,\infty)^{K}$ then $\mathcal{F}_\epsilon= \emptyset$
  \item The restriction of $f$ to $[\epsilon,+\infty]^{K}$ coincides with the  \emph{tropical polynomial}:
       \[
-        P_\epsilon(\mathbf x)= \min\limits_{\mathbf n\in\mathcal{F}_\epsilon} \set{\hat f (\mathbf n) + \mathbf n\mathbf x}.
+        P_\epsilon( x )= \min\limits_{ n  \in\mathcal{F}_\epsilon} \set{\hat f ( n  ) +  n   x }.
       \]
-      where $\mathbf n\mathbf x= \sum_{i=1}^{K}n_{i}x_{i}$.
+      where $ n   x = \sum_{i=1}^{K}n_{i}x_{i}$.
 \end{enumerate}
@@
 \[
- \set{\mathbf n\in\N^{K} \mid \textit{either } \hat f (\mathbf n)=+\infty \textit{ or there is } \mathbf m\prec \mathbf n\textit{ s.t.\ } \hat f(\mathbf m)\leq\hat f(\mathbf n)+\epsilon}.
+ \set{ n  \in\N^{K} \mid \textit{either } \hat f ( n  )=+\infty \textit{ or there is }  m  \prec  n  \textit{ s.t.\ } \hat f( m  )\leq\hat f( n  )+\epsilon}.
 \]
-In other words, $\mathbf n\in\mathcal F_\epsilon$ iff $\hat f(\mathbf n)<+\infty$ and for all $\mathbf m\prec \mathbf n$, one has $\hat f(\mathbf m)>\hat f(\mathbf n)+\epsilon$.
+In other words, $ n  \in\mathcal F_\epsilon$ iff $\hat f( n  )<+\infty$ and for all $ m  \prec  n  $, one has $\hat f( m  )>\hat f( n  )+\epsilon$.
 
@@
 Suppose that $\mathcal F_\epsilon$ is infinite; then, using Remark~\ref{rmk:AC}, it contains an infinite ascending chain
-\[\set{\mathbf m_0\prec \mathbf m_1\prec\cdots}.\]
+\[\set{ m  _0\prec  m  _1\prec\cdots}.\]
 By definition of $\mathcal F_\epsilon$ we have then:
-\[+\infty>\hat f(\mathbf m_0)>\hat f(\mathbf m_1)+\epsilon>\hat f(\mathbf m_2)+2\epsilon>\cdots\]
-so that $+\infty>\hat f(\mathbf m_0)>\hat f(\mathbf m_{i})+i\epsilon\geq i\epsilon$ for all $i\in\N$.
+\[+\infty>\hat f( m  _0)>\hat f( m  _1)+\epsilon>\hat f( m  _2)+2\epsilon>\cdots\]
+so that $+\infty>\hat f( m  _0)>\hat f( m  _{i})+i\epsilon\geq i\epsilon$ for all $i\in\N$.
 This contradicts the Archimedean property of $\R$.
@@
 2).
-We show that if $\mathcal F_\epsilon=\emptyset$, then $\hat f(\mathbf n)=+\infty$ for all $\mathbf n\in\N^{K}$.
+We show that if $\mathcal F_\epsilon=\emptyset$, then $\hat f( n  )=+\infty$ for all $ n  \in\N^{K}$.
 This immediately entails the desired result.
-We go by induction on the well-founded order $\prec$ over $\mathbf n\in\N^{K}$:
+We go by induction on the well-founded order $\prec$ over $ n  \in\N^{K}$:
 
-- if $\mathbf n=0^{K}\notin\mathcal F_\epsilon$, then $\hat f(\mathbf n)=+\infty$, because there is no $\mathbf m\prec\mathbf n$.
+- if $ n  =0^{K}\notin\mathcal F_\epsilon$, then $\hat f( n  )=+\infty$, because there is no $ m  \prec n  $.
 
-- if $\mathbf n\notin\mathcal F_\epsilon$, with $\mathbf n\neq 0^{K}$ then either $\hat f(\mathbf n)=+\infty$ and we are done, or there is $\mathbf m\prec \mathbf n$ s.t.\ $\hat f(\mathbf m)\leq \hat f(\mathbf n)+\epsilon$.
-By induction $\hat f(\mathbf m)=+\infty$ and, since $\epsilon<+\infty$, this entails $\hat f(\mathbf n)=+\infty$.
+- if $ n  \notin\mathcal F_\epsilon$, with $ n  \neq 0^{K}$ then either $\hat f( n  )=+\infty$ and we are done, or there is $ m  \prec  n  $ s.t.\ $\hat f( m  )\leq \hat f( n  )+\epsilon$.
+By induction $\hat f( m  )=+\infty$ and, since $\epsilon<+\infty$, this entails $\hat f( n  )=+\infty$.
 
 3).
-If $f(\mathbf x_0)=+\infty$ with $\mathbf x_0\in [\epsilon,\infty)^{K}$, then necessarily $\hat f(\mathbf n)=+\infty$ for all $\mathbf n\in\N^{K}$.
-Therefore, no $\mathbf n\in\N^{K}$ belongs to $\mathcal F_\epsilon$.
+If $f( x _0)=+\infty$ with $ x _0\in [\epsilon,\infty)^{K}$, then necessarily $\hat f( n  )=+\infty$ for all $ n  \in\N^{K}$.
+Therefore, no $ n  \in\N^{K}$ belongs to $\mathcal F_\epsilon$.
 
 4).
-We have to show that $f(\mathbf x)=P_\epsilon(\mathbf x)$ for all $\mathbf x\in [\epsilon,+\infty]^{K}$.
-By 1), it suffices to show that we can compute $f(\mathbf x)$ by taking the $\inf$, that is therefore a $\min$, only in $\mathcal F_\epsilon$ (instead of all $\N^{K}$).
+We have to show that $f( x )=P_\epsilon( x )$ for all $ x \in [\epsilon,+\infty]^{K}$.
+By 1), it suffices to show that we can compute $f( x )$ by taking the $\inf$, that is therefore a $\min$, only in $\mathcal F_\epsilon$ (instead of all $\N^{K}$).
 If $\mathcal F_\epsilon=\emptyset$ then by 2) we are done (remember that $\min\emptyset := +\infty$).
-If $\mathcal F_\epsilon\neq\emptyset$, we show that for all $\mathbf n\in\N^{K}$, if $\mathbf n \notin\mathcal F_\epsilon$, then there is $\mathbf m\in\mathcal F_\epsilon$ s.t.\ $\hat f(\mathbf m)+\mathbf m\mathbf x \leq \hat f(\mathbf n)+\mathbf n\mathbf x$.
+If $\mathcal F_\epsilon\neq\emptyset$, we show that for all $ n  \in\N^{K}$, if $ n   \notin\mathcal F_\epsilon$, then there is $ m  \in\mathcal F_\epsilon$ s.t.\ $\hat f( m  )+ m   x  \leq \hat f( n  )+ n   x $.
 We do it again by induction on $\prec_{1}$:
 
-- if $\mathbf n=0^{K}$, then from $\mathbf  n\notin \mathcal F_{\epsilon}$, by definition of $\mathcal F_\epsilon$, we have $\hat f(\mathbf n)=+\infty$ (because there is no $\mathbf n'\prec\mathbf n$).
+- if $ n  =0^{K}$, then from $\mathbf  n\notin \mathcal F_{\epsilon}$, by definition of $\mathcal F_\epsilon$, we have $\hat f( n  )=+\infty$ (because there is no $ n  '\prec n  $).
 So any element of $\mathcal F_\epsilon\neq\emptyset$ works.
 
-- if $\mathbf n\neq 0^{K}$, then we have two cases:
-either $\hat f(\mathbf n)=+\infty$, in which case we are done as before by taking any element of $\mathcal F_\epsilon\neq\emptyset$.
-Or $\hat f(\mathbf n)<+\infty$, in which case (again by definition of $\mathcal F_\epsilon$) there is $\mathbf n'\prec\mathbf n$ s.t.\ \begin{equation}\label{eq:n'neps} \hat f(\mathbf n')\leq \hat f(\mathbf n)+\epsilon.\end{equation}
+- if $ n  \neq 0^{K}$, then we have two cases:
+either $\hat f( n  )=+\infty$, in which case we are done as before by taking any element of $\mathcal F_\epsilon\neq\emptyset$.
+Or $\hat f( n  )<+\infty$, in which case (again by definition of $\mathcal F_\epsilon$) there is $ n  '\prec n  $ s.t.\ \begin{equation}\label{eq:n'neps} \hat f( n  ')\leq \hat f( n  )+\epsilon.\end{equation}
 Therefore we have (remark that the following inequalities hold also for the case $x=+\infty$):
 \[\begin{array}{rclr}
- \hat f(\mathbf n')+\mathbf n'\mathbf x & \leq & \hat f(\mathbf n) + \epsilon +\mathbf  n'\mathbf x & \textit{by \eqref{eq:n'neps}} \\
- & \leq & \hat f(\mathbf n) + (\mathbf n-\mathbf n')\mathbf x + \mathbf n'\mathbf x & \textit{because $\epsilon\leq\min\mathbf x$ and $\min \mathbf x\leq(\mathbf n-\mathbf  n')\mathbf x$} \\
- & = & \hat f(\mathbf n)+ \mathbf n\mathbf x. &
+ \hat f( n  ')+ n  ' x  & \leq & \hat f( n  ) + \epsilon +\mathbf  n' x  & \textit{by \eqref{eq:n'neps}} \\
+ & \leq & \hat f( n  ) + ( n  - n  ') x  +  n  ' x  & \textit{because $\epsilon\leq\min x $ and $\min  x \leq( n  -\mathbf  n') x $} \\
+ & = & \hat f( n  )+  n   x . &
 \end{array}\]
-Now, if $\mathbf n'\in\mathcal F_\epsilon$ we are done.
-Otherwise $\mathbf n'\notin\mathcal F_\epsilon$ and we can apply the induction hypothesis on it, obtaining an $\mathbf m\in\mathcal F_\epsilon$ s.t.\ $\hat f(\mathbf m)+\mathbf m\mathbf x \leq \hat f(\mathbf n')+\mathbf n'\mathbf x$.
-Therefore this $\mathbf m$ works.
+Now, if $ n  '\in\mathcal F_\epsilon$ we are done.
+Otherwise $ n  '\notin\mathcal F_\epsilon$ and we can apply the induction hypothesis on it, obtaining an $ m  \in\mathcal F_\epsilon$ s.t.\ $\hat f( m  )+ m   x  \leq \hat f( n  ')+ n  ' x $.
+Therefore this $ m  $ works.
 \end{proof}
 
-\newpage
+%\newpage
 
@@
 
-\newpage
+%\newpage
 
@@
 
-\begin{proposition}\label{prop:troplinear}
-All tropical linear functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive.  
-\end{proposition}
-\begin{proof}[proof TO BE FIXED]
-All functions of the form $f(\B x)(b)= \mu_{b} \B x + c_{b}$ are Lipschitz with Lipschitz constant
-$\sum_{a\in X}\mu_{b}(a)$: 
-one has $| f(\B x)(b)- f(\B y)(b) |= |\mu_{b}\B x+c_{b} -\mu_{b}\B y+c_{b}|=
-|\mu_{b}\B x -\mu_{b} \B y  |\leq \left(\sum_{a\in X}\mu_{b}(a)\right)\cdot \sup_{a\in X}|\B x_{a}-\B y_{a}|$.
-
-\end{proof}
-
-\begin{lemma}\label{lm:mainLip}
-Let $f: [0,\infty)^{X}\to Q$ be concave, monotone increasing, and continuous.  
-Let $\B x\neq \B y\in [0,\infty)^{X}$, with $\|\B x-\B y\|_{\infty}<\infty$, and let $S(\B x, \B y)=\{ \alpha\B x + (1-\alpha)\B y\mid \alpha \in [0,1]\}$ be the segment generated by $\B x$ and $\B y$.
-Then $f$ is Lipschitz-continuous over $S(\B x, \B y)$.
-
-\end{lemma}
-\begin{proof}
-Let us prove the lemma under the assumption that for all $a\in X$, $\B y_{a}-\B x_{a}\geq 1$. From the fact that the claim holds under the assumption, we can deduce the claim of the lemma:
-indeed for $\alpha\in (0,1)$ large enough we have that $\B y':= \frac{\B y-\alpha\B x}{1-\alpha}$ is such that $\B y\in S(\B x, \B y')$ (and thus $S(\B x, \B y)\subseteq S(\B x, \B y')$) and 
-$\B y'_{a}-\B x_{a}\geq 1$. Hence from our proof we deduce that $f$ is Lipschitz-continuous over $S(\B x, \B y')$, and thus a fortiori over $S(\B x, \B y)$ too.
-
-Since $f$ is continuous over $[0,\infty)^{X}$ and $S(\B x, \B y)$ is compact, $f$ admits a maximum $\mathrm{MAX}$ over $S(\B x, \B y)$.
-For all $\B z< \B z'\in S(\B x, \B y)$, let $M(\B z, \B z')\in Q^{X}$ be defined by
-$$
-M(\B z, \B z')_{a}= \frac{f(\B z')-f(\B z)}{\B z'_{a}-\B z_{a}}
-$$
-Observe that
-\begin{align*}
-M(\B x, \B y)_{a} & = \frac{f(\B y)-f(\B x)}{\B y_{a}-\B x_{a}} \leq 
-f(\B y)-f(\B x) \leq \mathrm{MAX}
-\end{align*}
-using the fact that $\B y_{a}-\B x_{a}\geq 1$.
-
-We now claim that $M(\B z, \B z')$ is contravariant in both $\B z$ and $\B z'$. 
-Indeed suppose $\B z\leq \B z'' < \B z'$, so that $\B z= \lambda \B z' +(1-\lambda)\B z''$ for some $\lambda \in (0,1)$. Then, using the fact that $f$ is concave, we have 
-\begin{align*}
-M(\B z, \B z')_{a}&=\frac{f(\B z')-f(\lambda \B x +(1-\lambda)\B z'')}{\B z'_{a}-\lambda \B x_{a} -(1-\lambda)\B z''_{a}} \\
-&
-\geq 
-\frac{f(\B z')-\lambda f(\B z') -(1-\lambda)f(\B z'')}{\B z'_{a}-\lambda \B z'_{a} -(1-\lambda)\B z''_{a}} \\
-&
-= 
-\frac{(1-\lambda)(f(\B z')-f(\B z''))}{(1-\lambda )\B z'_{a}-\B z''_{a}} =M(\B z'', \B z')
-\end{align*}
-In a similar way it is shown that for $\B z < \B z''\leq \B z'$, $M(\B z,\B z')\leq M(\B z, \B z'')$.
-
-Therefore, for all $\B z < \B z'\in S(\B x, \B y)$, we have that $M(\B z, \B z')_{a} \leq M(\B x, \B z')_{a} \leq M(\B x, \B y)_{a} \leq \mathrm{MAX}$. 
-From this, using the fact that $f$ is monotone increasing, we deduce that 
-$|f(\B z')-f(\B z)|=f(\B z')-f(\B z) \leq \mathrm{MAX}\cdot |\B z'_{a}-\B z_{a}|
-$ and thus that 
-$$|f(\B z')-f(\B z)|\leq \mathrm{MAX}\cdot \|\B z'-\B z\|_{\infty}$$
-that is, that $f$ is $\mathrm{MAX}$-Lipschitz over $S(\B x, \B y)$. 
-\end{proof}
-
-\begin{proposition}
-Let $f: [0,\infty)^{X}\to Q$ be concave, monotone increasing, and continuous.  
-For all $\epsilon \in (0,\infty)$ and $\B x\in [0,\infty)^{X}$, $f$ is Lipschitz-continuous over the open ball $B(\B x, \epsilon)$.
-\end{proposition}
-\begin{proof}
-Let $\mathrm{MAX}$ indicate the maximum of $f$ over $B(\B x, \epsilon)$. 
-Let $\B y, \B z\in B(\B x, \epsilon)$; then $\|\B y-\B z\|_{\infty}\leq 2\epsilon < \infty$, so by the lemma above $f$ is $K$-Lipschitz over the segment $S(\B y, \B z)$ for some $K\leq \mathrm{MAX}$, so we deduce $|f(\B y)-f(\B z)|\leq \mathrm{MAX}\cdot \|\B y-\B z\|_{\infty}$.
-\end{proof}
-
-\begin{theorem}[local Lipschitz-continuity]
-Let $f: [0,\infty)^{X}\to Q$ be concave, monotone increasing, and continuous.  
-Then $f$ is locally Lipschitz-continuous.
-\end{theorem}
-\begin{proof}
-For all $\B x\in [0,\infty)^{X}$, $f$ is Lipschitz-continuous over the open set $B(\B x, 1)$.
-\end{proof}
-
-\subsubsection{Characterizations of the functional metric }
-
-  \begin{proposition}\label{prop:functionalmetric}
-For all maps $f,g: Q\langle X\rangle\to Q\langle Y\rangle$,  
-$$
-  \|\widehat f- \widehat g\|_{\infty}=
-  \sup\{\|f(\B x)- g(\B x)\|_{\infty} \mid \B x\in Q\langle X\rangle\}
-$$
-
-\end{proposition}
-
-Before proving the proposition we need one preliminary lemma:
-
-\begin{lemma}
+  We need a first, preliminary, lemma:
+\begin{lemma}\label{lemma:inf}
 Let $u,v: I\to Q$ and suppose $|u(i)-v(i)|\leq \delta$, for all $i\in I$.
@@
 \end{proof}
+
+\begin{proposition}\label{prop:troplinear}
+All tropical linear functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive.  
+\end{proposition}
+\begin{proof}
+%All functions of the form $f_{\theta,\eta}(x)_{b}= \theta(b)+x_{\eta(b)}$, where $\theta:Y \to \Lawv$ and $\eta:Y\to X$, are non-expansive: indeed $|f_{\theta,\eta}(x)_{b}-f_{\theta,\eta}(y)_{b}|= |\theta(b)+x_{\eta(b)}-\theta(b)-y_{\eta(b)}|=
+%|x_{\eta(b)}-y_{\eta(b)}|\leq \norm{x-y}_{\infty}$.
+Let $f(x)_{b}=\inf_{a}\{\widehat f_{a,b}+x_{a}\}$.
+For all $a\in X$ we have that 
+$|(\widehat f_{a,b}+x_{a})- (\widehat f_{a,b}+y_{a})|=
+|x_{a}-y_{a}|\leq \norm{x-y}_{\infty}$. Using Lemma \ref{lemma:inf}, we deduce then 
+$|f(x)_{b}-f(y)_{b}|=|\inf_{a}\{\widehat f_{a,b}+x_{a}\}-
+\inf_{a}\{\widehat f_{a,b}+y_{a}\}|\leq \norm{x-y}_{\infty}$.
+\end{proof}
+
+\begin{lemma}\label{lm:mainLip}
+Let $f: [0,\infty)^{X}\to Q$ be concave, monotone increasing, and continuous.  
+Let $ x \neq  y  \in [0,\infty)^{X}$, with $\| x - y  \|_{\infty}<\infty$, and let $S( x ,  y  )=\{ \alpha x  + (1-\alpha) y  \mid \alpha \in [0,1]\}$ be the segment generated by $ x $ and $ y  $. 
+Then $f$ is Lipschitz-continuous over $S( x ,  y  )$.
+
+\end{lemma}
+\begin{proof}
+Let us prove the lemma under the assumption that for all $a\in X$, $ y  _{a}- x _{a}\geq 1$. From the fact that the claim holds under the assumption, we can deduce the claim of the lemma:
+indeed for $\alpha\in (0,1)$ large enough we have that $ y  ':= \frac{ y  -\alpha x }{1-\alpha}$ is such that $ y  \in S( x ,  y  ')$ (and thus $S( x ,  y  )\subseteq S( x ,  y  ')$) and 
+$ y  '_{a}- x _{a}\geq 1$. Hence from our proof we deduce that $f$ is Lipschitz-continuous over $S( x ,  y  ')$, and thus a fortiori over $S( x ,  y  )$ too.
+
+Since $f$ is continuous over $[0,\infty)^{X}$ and $S( x ,  y  )$ is compact, $f$ admits a maximum $\mathrm{MAX}$ over $S( x ,  y  )$.
+For all $ z  <  z  '\in S( x ,  y  )$, let $M( z  ,  z  ')\in Q^{X}$ be defined by
+$$
+M( z  ,  z  ')_{a}= \frac{f( z  ')-f( z  )}{ z  '_{a}- z  _{a}}
+$$
+Observe that
+\begin{align*}
+M( x ,  y  )_{a} & = \frac{f( y  )-f( x )}{ y  _{a}- x _{a}} \leq 
+f( y  )-f( x ) \leq \mathrm{MAX}
+\end{align*}
+using the fact that $ y  _{a}- x _{a}\geq 1$.
+
+We now claim that $M( z  ,  z  ')$ is contravariant in both $ z  $ and $ z  '$. 
+Indeed suppose $ z  \leq  z  '' <  z  '$, so that $ z  = \lambda  z  ' +(1-\lambda) z  ''$ for some $\lambda \in (0,1)$. Then, using the fact that $f$ is concave, we have 
+\begin{align*}
+M( z  ,  z  ')_{a}&=\frac{f( z  ')-f(\lambda  x  +(1-\lambda) z  '')}{ z  '_{a}-\lambda  x _{a} -(1-\lambda) z  ''_{a}} \\
+&
+\geq 
+\frac{f( z  ')-\lambda f( z  ') -(1-\lambda)f( z  '')}{ z  '_{a}-\lambda  z  '_{a} -(1-\lambda) z  ''_{a}} \\
+&
+= 
+\frac{(1-\lambda)(f( z  ')-f( z  ''))}{(1-\lambda ) z  '_{a}- z  ''_{a}} =M( z  '',  z  ')
+\end{align*}
+In a similar way it is shown that for $ z   <  z  ''\leq  z  '$, $M( z  , z  ')\leq M( z  ,  z  '')$.
+
+Therefore, for all $ z   <  z  '\in S( x ,  y  )$, we have that $M( z  ,  z  ')_{a} \leq M( x ,  z  ')_{a} \leq M( x ,  y  )_{a} \leq \mathrm{MAX}$. 
+From this, using the fact that $f$ is monotone increasing, we deduce that 
+$|f( z  ')-f( z  )|=f( z  ')-f( z  ) \leq \mathrm{MAX}\cdot | z  '_{a}- z  _{a}|
+$ and thus that 
+$$|f( z  ')-f( z  )|\leq \mathrm{MAX}\cdot \| z  '- z  \|_{\infty}$$
+that is, that $f$ is $\mathrm{MAX}$-Lipschitz over $S( x ,  y  )$. 
+\end{proof}
+
+\begin{proposition}
+Let $f: [0,\infty)^{X}\to Q$ be concave, monotone increasing, and continuous.  
+For all $\epsilon \in (0,\infty)$ and $ x \in [0,\infty)^{X}$, $f$ is Lipschitz-continuous over the open ball $B( x , \epsilon)$.
+\end{proposition}
+\begin{proof}
+Let $\mathrm{MAX}$ indicate the maximum of $f$ over $B( x , \epsilon)$. 
+Let $ y  ,  z  \in B( x , \epsilon)$; then $\| y  - z  \|_{\infty}\leq 2\epsilon < \infty$, so by the lemma above $f$ is $K$-Lipschitz over the segment $S( y  ,  z  )$ for some $K\leq \mathrm{MAX}$, so we deduce $|f( y  )-f( z  )|\leq \mathrm{MAX}\cdot \| y  - z  \|_{\infty}$.
+\end{proof}
+
+\begin{theorem}[local Lipschitz-continuity]
+Let $f: [0,\infty)^{X}\to Q$ be concave, monotone increasing, and continuous.  
+Then $f$ is locally Lipschitz-continuous.
+\end{theorem}
+\begin{proof}
+For all $ x \in [0,\infty)^{X}$, $f$ is Lipschitz-continuous over the open set $B( x , 1)$.
+\end{proof}
+
+\subsubsection{Characterizations of the functional metric }
+
+  \begin{proposition}\label{prop:functionalmetric}
+For all maps $f,g: Q\langle X\rangle\to Q\langle Y\rangle$,  
+$$
+  \|\widehat f- \widehat g\|_{\infty}=
+  \sup\{\|f( x )- g( x )\|_{\infty} \mid  x \in Q\langle X\rangle\}
+$$
+
+\end{proposition}
 
@@
 For one side, suppose for all $\mu\in \C  M_{\mathrm{fin}}(X),b\in Y$, 
-$|\widehat f_{\mu,b}-\widehat g_{\mu,b}|\leq \delta$. Then for all $\B x\in Q\langle X\rangle$ and $b\in Y$, $|f(\B x)(b)-g(\B x)(b)|\leq \delta$.
-Indeed, $|f(\B x)(b)-g(\B x)(b)|= |( \inf_{\mu}\mu \B x+\widehat f_{\mu,b})- (\inf_{\mu}\mu \B x+\widehat g_{\mu,b})|$. 
-Since $|\mu \B x+\widehat f_{\mu,b} - \mu \B x -\widehat g_{\mu,b}|= |\widehat f_{\mu,b}-\widehat g_{\mu,b}|\leq \delta$, by the Lemma above we conclude $|f(\B x)(b)-g(\B x)(b)|\leq \delta$.
+$|\widehat f_{\mu,b}-\widehat g_{\mu,b}|\leq \delta$. Then for all $ x \in Q\langle X\rangle$ and $b\in Y$, $|f( x )(b)-g( x )(b)|\leq \delta$.
+Indeed, $|f( x )(b)-g( x )(b)|= |( \inf_{\mu}\mu  x +\widehat f_{\mu,b})- (\inf_{\mu}\mu  x +\widehat g_{\mu,b})|$. 
+Since $|\mu  x +\widehat f_{\mu,b} - \mu  x  -\widehat g_{\mu,b}|= |\widehat f_{\mu,b}-\widehat g_{\mu,b}|\leq \delta$, by Lemma \ref{lemma:inf} we conclude $|f( x )(b)-g( x )(b)|\leq \delta$.
 
@@
 |\widehat f_{\mu,b}- \widehat g_{\mu,b}|> \epsilon$, so we deduce that 
-$\sup\{\| f(\B x)- g(\B x) \|_{\infty}\mid \B x\in Q\langle X\rangle\}> \epsilon$.
+$\sup\{\| f( x )- g( x ) \|_{\infty}\mid  x \in Q\langle X\rangle\}> \epsilon$.
 \end{proof}
@@
 $$
-\delta^{(n)}f(\B x^{n})=\mathsf D^{(n)}f(\B x^{n}, \infty)
+\delta^{(n)}f( x ^{n})=\mathsf D^{(n)}f( x ^{n}, \infty)
 $$
@@
 Moreover, the full behavior of $f$ can be recovered from the functions $\delta^{(n)}f$ using the Taylor expansion which, in its tropical form, reads as:
-\begin{equation}
-f(\B x)= \inf_{n}\left \{\mathsf D^{(n)}f(\B x^{n},\infty)\right\}= \inf_{n}\left \{\delta^{(n)}f(\B x^{n})\right\}
-\tag{Tropical Taylor}
-\end{equation}
+\begin{equation*}
+f( x )= \inf_{n}\left \{\mathsf D^{(n)}f( x ^{n},\infty)\right\}= \inf_{n}\left \{\delta^{(n)}f( x ^{n})\right\}
+%\tag{Tropical Taylor}
+\end{equation*}
 

### 2023-01-21 section2.tex
commit

@@
 as a limit of \emph{more and more sensitive} operations.
+
 This viewpoint, that we develop in the following sections, leads to the somehow unexpected discovery of a bridge between the metric and differential study of higher-order programs.
-This connection not only suggest the application of optimization methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions, but it scales to a 
+This connection not only suggests the application of optimization methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions, but it scales to a 
 more abstract level, leading to introduce a 

### 2023-01-21 appIV.tex
Commit presentazione del STDLC

@@
 
-Let us end this section by recalling the sound interpretation of $\STDLC$-terms in a $C\partial C$ category (which is taken from [Section 4.3, \cite{Manzo2010}]).
+\subsubsection{The language and the semantics of $\STDLC$}
+
+The language we mainly consider in the main paper il the differential $\lambda$-calculus $\STDLC$, that we quickly define here, following [Section 3, \cite{Manzo2010}].
+
+The terms and the sums\footnote{Usually what we call here ``terms'' are called ``simple terms'', and what we call here ``sums'' are called ``differential terms''.} are mutually generated by the following grammars, then quotiented by usual $\alpha$-equivalence plus other equivalences handling the $+$ constructor.
+
+{
+\begin{minipage}{\textwidth}
+\begin{align*}
+\textit{Terms: } & M,N:= x\mid \lambda x.M \mid M\BB T \mid \Der [M] \cdot N
+\\
+\textit{Sums: } & \BB T:= 0 \mid M \mid M+\BB T
+\end{align*}\end{minipage}}\medskip\\
+Simple types are generated by the grammar:
+
+{
+\begin{minipage}{\textwidth}
+\begin{align*}
+A,B:= X\mid A\to B
+\end{align*}\end{minipage}}\medskip\\
+where $X$ varies in a fixed set of ground types.
+
+The typing system is given in Figure~\ref{fig:STDLC}.
+
+\begin{figure}
+\fbox{
+\begin{minipage}{0.9\textwidth}
+\begin{center}
+
+  {
+\AXC{$(x: A)\in \Gamma$}
+\UIC{$\Gamma\vdash x\in A$}
+\DP}
+
+\bigskip
+
+\begin{tabular}{c c }
+  {
+\AXC{$\Gamma, x: A\vdash M: B$}
+\UIC{$\Gamma\vdash \lambda x.M: A\to B$}
+\DP}
+
+&
+
+  {
+\AXC{$\Gamma \vdash M: A\to B$}
+\AXC{$\Gamma\vdash \BB T: A$}
+\BIC{$\Gamma \vdash M\BB T: B$}
+\DP}
+
+\end{tabular}
+
+\bigskip
+
+  {
+\AXC{$\Gamma \vdash M: A\to B$}
+\AXC{$\Gamma \vdash N: B$}
+\BIC{$\Gamma \vdash \Der [M] \cdot N: A\to B$}
+\DP}
+
+\bigskip
+
+\begin{tabular}{c c }
+
+  {
+\AXC{}
+\UIC{$\Gamma \vdash 0:A$}
+\DP}
+
+&
+
+  {
+\AXC{$\Gamma\vdash M_1: A \quad \cdots \quad \Gamma \vdash M_n:A \qquad (n\geq2)$}
+\UIC{$\Gamma \vdash M_1+\cdots +M_n : A$}
+\DP}
+
+\end{tabular}
+\end{center}
+\end{minipage}
+}
+\caption{Typing rules for $\STDLC$.}
+\label{fig:STDLC}
+\end{figure}
+
+We finally recall the sound interpretation of $\STDLC$-terms in a $C\partial C$ category, following [Section 4.3, \cite{Manzo2010}].
 
 \begin{definition}
-Let $\C C$ be a $C\partial C$ and $\model A$ fixed objects of $\C C$, for each ground type $A$ of the $\STDLC$.
+Let $\C C$ be a $C\partial C$ and $\model X$ fixed objects of $\C C$, one for each ground type $X$ of the $\STDLC$.
 
@@
 
-Then, one inductively defines an interpretation $\model{\Gamma \vdash M:A}\in\C C(\model \Gamma,\model A)$ of a $\STDLC$-term $\Gamma \vdash M:A$ (that can can be proved to be sound) by induction as:
+Then, one inductively defines an interpretation $\model{\Gamma \vdash M:A}\in\C C(\model \Gamma,\model A)$ of $\Gamma \vdash M:A$ and one of $\model{\Gamma \vdash \BB T:A}\in\C C(\model \Gamma,\model{A})$ of $\Gamma \vdash \BB T:A$ by induction as:
 \begin{itemize}
@@
  \item $\model{\Gamma \vdash \Der [M] \cdot N:A\to B}:=\Lambda\left(\Lambda^-(\model{\Gamma \vdash M:A\to B})\star \model{\Gamma \vdash N:A}\right)$
- \item $\model{\Gamma \vdash 0:A}:=\infty$
- \item $\model{\Gamma \vdash M+\BB T:A}:=\model{\Gamma \vdash M:A}+\model{\Gamma \vdash \BB T:A}$
+ \item $\model{\Gamma \vdash 0:A}:=:0 \qquad \textit{(the $0$ element of $\C C(\model \Gamma,\model A)$)}$
+ \item $\model{\Gamma \vdash M+\BB T:A}:=\model{\Gamma \vdash M:A}+\model{\Gamma \vdash \BB T:A}$ \qquad \textit{(the $+$ operation on $\C C(\model \Gamma,\model A)$)}
 \end{itemize}
@@
 
+One then defines a ``differential'' extension of usual $\beta$-reduction (see again [Section 3, \cite{Manzo2010}], or \cite{ER}), and thanks to the fact that $\C C$ is a $C\partial CC$, the interpretation above can can be proved to be sound for it ([Theorem 4.12, \cite{Manzo2010}]).
+
 In the main paper we write $\model M$, but of course we mean $\model{\Gamma \vdash M:A}$.

### 2023-01-22 main.tex
commit

@@
 
-\title{Tropical Mathematics and Lambda Calculus}
+\title{Tropical Mathematics and the Lambda Calculus}
 
@@
 \begin{abstract}
-In this paper we study a quantitative semantics for the $\lambda$-calculus based on tropical mathematics. Notably, we show that the relational model with values over the tropical semiring 
-provides a unified framework for both 
+In this paper we study an interpretation of the $\lambda$-calculus in the framework of tropical mathematics, and we show that it provides a unified framework for both 
 program metrics, as based on the analysis of program sensitivity via Lipschitz-conditions, and for resource analysis, as based on higher-order program differentiation.
-We discuss applications of this tropical semantics to ``best case'' resource analysis as well as to other quantitative properties like log-probabilities and differential privacy.
+We discuss applications of this tropical semantics to ``best case'' resource analysis as well as to other quantitative properties like convergence log-probabilities and differential privacy.
 Finally, we show that a general foundation for this semantic approach is provided by an abstract correspondence between Lawvere's theory  

### 2023-01-22 section1.tex
commit

@@
-In recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of programs, such as e.g.~number of computation steps, convergence probability,
-approximation bounds, 
-as opposed to purely \emph{qualitative} properties like e.g.~termination or equivalence. 
-Notably, a significative effort has been made to extend, or adapt, well-established qualitative methods, like e.g.~type systems, relational logics or denotational semantics, to account for quantitative properties.
+In recent years, more and more interest in the programming language community has been directed towards the study of \emph{quantitative} properties of programs like e.g.~computing the number of computation steps, or convergence probabilities,
+as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence. 
+Notably, a significant effort has been made to extend, or adapt, well-established qualitative methods, like e.g.~type systems, relational logics or denotational semantics, to account for quantitative properties.
 We can mention, for example,
-intersection type systems aimed at capturing time or space resources \cite{decarvalho2018, Accattoli2022}, or convergence probabilities \cite{Breuvart2018, PistoneLICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthe_2012} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for
+intersection type systems aimed at capturing time or space resources \cite{decarvalho2018, Accattoli2022} or convergence probabilities \cite{Breuvart2018, PistoneLICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthe_2012} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for
 probabilistic \cite{Ehrhard2011, Staton2017} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
@@
 
-On the one hand one should certainly mention \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function (which is rarely the case), it is more useful to know whether they compute functions which do not differ \emph{too much}.
-This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
+On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function (which is rarely the case), it makes more sense to ask   whether they compute functions which do not differ \emph{too much}.
+This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
 
-On the other hand there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
-This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for probabilistic settings \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
+On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
+This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
 
-In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using its input exactly once), plays a crucial role.
-In metric semantics, linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
+In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
+In metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
 By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
 
-A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow connected.
+A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled.
 At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
-in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} $x^{n}$, hence not Lipschitz.
-The fundamental insight of this article is that a natural way to overcome
-this obstacle and bridge the two viewpoints 
-is to develop differential semantics in the framework of 
-tropical mathematics.
+in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz.
+The fundamental motivation of this work is then the observation that
+this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ ``reads as'' the linear function $n\cdot x$.
+
+% from higher-order programs is based on  soon as one develops  differential semantics in the framework of 
+%tropical mathematics.
 %
@@
 
-Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} as the study of algebra and geometry where one replaces the usual ring structure of numbers based on addition and multiplication by the semi-ring structure given by ``$\min$'' and ``$+$''.
+Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} as an alternative approach to algebra and geometry where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given by ``$\min$'' and ``$+$''.
 %
@@
 %played by the idempotent operation $\min$.
-For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$, when interpreted over the tropical semi-ring, becomes the piecewise linear function
+For instance, a polynomial like e.g.~$p(x,y)=x^{2}+xy^{2}+y^{3}$, when interpreted over the tropical semiring, translates as the piecewise linear function
 $
-\varphi(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
+\varphi(x,y)=\min\{2x, x+2y, 3y\}
 $.
 
-This is not a \emph{ad-hoc} setting: tropical geometry turns out to be a rich and florid combinatorial counterpart of usual algebraic geometry, with important connections with optimisation theory.
-Furthermore, computationally speaking, working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, the important and generally intractable problem of finding roots of a polynomial, admits a linear time algorithm in the tropical case (and the roots of a tropicalized polynomial can be used to approximate the actual roots of the polynomial \cite{Noferini2015}).
-Such a computational nature of tropical concepts explains why these are also widely applied in computer science, e.g.~for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).
+%This is not a \emph{ad-hoc} setting: 
+In the last decades, tropical geometry evolved into a vast and rich research domain, providing a combinatorial counterpart of usual algebraic geometry, with important connections with optimisation theory \cite{Akian2011, Akian2012}.
+Computationally speaking, working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, the fundamental (and generally intractable) problem of finding the roots of a polynomial admits a \emph{linear time} algorithm in the tropical case (moreover, the tropical roots of a polynomial $p$ can be used to approximate the actual roots of $p$ \cite{Noferini2015}).
+The computational nature of tropical notions explains why these are so widely applied in computer science, notably for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).
 
-Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for, as the monomials $x^{n}$ turn precisely into the Lipschitz functions $nx$.
+Coming back to our previous discussion on program semantics, tropical geometry might seem to provide precisely what we are looking for, as the monomials $x^{n}$ turn precisely into the Lipschitz functions $nx$.
 At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.
@@
 
-The goal of this paper is to study the tropical semantics of the differential $\lambda$-calculus and use it to connect it with metric semantics.
-This bridge does not only provide a conceptual unification of two different quantitative approaches to higher-order programs, but it also
-suggests ways in which techniques from resource-analysis could be used in sensitivity analysis and \emph{vice-versa}, opening the way for new potential applications of tropical geometry to the study of quantitative properties of higher-order programs.
+In this paper we demonstrate that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provides the desired bridge between differential and metric semantics.
+Moreover, we show that the conceptual unification of these two approaches suggests ways in which techniques from resource-analysis could be used in sensitivity analysis and \emph{vice-versa}, paving the way for new  applications of tropical geometry to the  study of higher-order programs.
 
@@
 
-\item we study the relational model over the tropical semiring, in which
-\emph{bounded} simply typed terms are interpreted as tropical polynomials, while general
- simply typed lambda terms correspond to a generalization of \emph{tropical Laurent series} \cite{Porzio2021} which, moreover, are locally Lipschitz-continuous, yielding a full-scale metric semantics for the $\lambda$-calculus.
+\item we study the relational model over the tropical semiring  and we show that the functions interpreting simply-typed lambda terms, which correspond to a generalization of \emph{tropical Laurent series} \cite{Porzio2021}, are locally Lipschitz-continuous, thus yielding a full-scale metric semantics for the $\lambda$-calculus and its bounded fragments.
 This is in Sections \ref{section3} and \ref{section4}.
@@
 
-\item Using the relational model as our main source of intuition,  we suggest a few potential applications of tropical methods to the study of quantitative properties of non-deterministic and probabilistic functional programs like counting best-case computation steps,
+\item Using the relational model as our main source of inspiration,  we suggest a few potential applications of tropical methods to the study of quantitative properties of non-deterministic and probabilistic functional programs, like counting best-case computation steps,
 measuring convergence log-probabilities, and 
@@
 \item We conclude 
-by putting the connection between
-tropical, differential and metric properties at the right level of generality.
-By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} (GMS) \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007}, we show that the category of \emph{complete} GMSs provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
+by putting the connection between the 
+tropical, differential and metric viewpoints at the right level of generality.
+By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007}, we show that the category of \emph{complete} generalized metric spaces provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
 This is in Section~\ref{section6}.

### 2023-01-22 section2.tex
commit

@@
 
-In this section we discuss in some more detail the two approaches to quantitative semantics we mentioned in the Introduction, at the same time providing an overview of how we aim at bridging them using tropical mathematics.
+In this section, we discuss in some more detail the two approaches to quantitative semantics we mentioned in the Introduction, at the same time providing an overview of how we aim at bridging them using tropical mathematics.
 
@@
 
-As recalled in the Introduction, in many situations (e.g.~when dealing with computationally expensive functions) one does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound.
+In many situations (e.g.~when dealing with computationally difficult problems) one does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound.
 In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, a small error in the input will produce a comparably small error in the output.
 
-In all these cases, it is common to consider denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which measures differences of behavior. 
+In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior.
 A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
 $d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
-A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the {affine} simply typed $\lambda$-calculus, being both a \emph{symmetric monoidal closed} category (i.e.~a model of \emph{linear} $\STLC$) and a cartesian category in which cartesian and monoidal unit coincide.
+A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the \emph{linear} simply typed $\lambda$-calculus, being a \emph{symmetric monoidal closed} category, and in fact it also models {affine} terms (since the cartesian and monoidal units coincide).
 
-As observed in \cite{Reed2010, Gaboardi2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplication.
+As observed in \cite{Reed2010, Gaboardi2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications.
 The fundamental intuition is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz} map.
@@
 
-These observations have led to the study of \emph{bounded} typed $\lambda$-calculi, like the system $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied in the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}.
+These observations have led to the study of linear $\lambda$-calculi with \emph{graded} exponentials, like the system $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied to the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}.
 The types of such systems are defined by combining linear constructors with a \emph{graded exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
-In the following sections we will sometimes make reference to a basic type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined via $A::= o  \mid   (!_{n}A) \multimap A  $,
+In the following sections we will sometimes make reference to a basic graded type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined via $A::= o  \mid   !_{n}A\multimap A  $,
 where $n\in \BB N$.
@@
 Still, one might observe that the program $M$ above is actually Lipschitz-continuous, if not globally, at least \emph{locally} (i.e.~over any compact set).
-Indeed, some cartesian closed categories of locally Lipschitz maps have been produced in the literature (see \cite{Ehrhard2011, PistoneLICS}), and a new example will be exhibited in this paper.
+Indeed, some cartesian closed categories of locally Lipschitz maps have been produced in the literature \cite{Ehrhard2011, PistoneLICS}, and a new example will be exhibited in this paper.
 
@@
 The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus.
-The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M]\cdot N$, still of type $A\to B$, corresponds to the \emph{linear application} of $N$ to $M$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M]\cdot N$ has type $A\to B$, since $M$ might still need \emph{other} copies of an input of type $A$).
+The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M]\cdot N$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M]\cdot N$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).
 
-Interestingly, the categorical study of these differential operators has led to the introduction of \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, a class of categories providing an abstract axiomatization of differentiation, in the usual mathematical sense.
+Interestingly, the categorical study of the operator $\Der$ has led to the introduction of \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, a class of categories providing an abstract axiomatization of differentiation, in the usual mathematical sense.
 More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
@@
 
-Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with the differential operator $\Der[-]\cdot (-)$, together with a notion of \emph{linear substitution} (see \cite{difflambda} for details).
+Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
 The models of $\STDLC$ are the 
 cartesian \emph{closed} differential categories ($CC\partial C$), 
-which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential are both well-behaved with respect to the closed structure (see \cite{Manzo2012}).
+which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.
 
@@
 actual derivatives is provided by the \emph{Taylor expansion}:
-in $\STDLC$ one can prove that any application $MN$ can be expanded as a sum of \emph{linear} applications
-$\Der^{(k)}[M]\cdot N^{k}$, i.e.~where $N$ is passed exactly $k$ times to $M$, expressed as:
+in $\STDLC$ one can prove that any application $MN$ can be expanded as an infinite formal sum of \emph{linear} applications
+$\Der^{(k)}[M]\cdot N^{k}$, i.e.~where $N$ is passed exactly $k$ times to $M$:
 \begin{align}\label{eq:taylor}
@@
 \end{align}
-In other words, unbounded duplications are seen as a sort of limit of bounded, but arbitrarily large, ones.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
+In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
 
@@
 
-At this point, as the Taylor formula decomposes an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as decomposing  a $\lambda$-term
-into a limit of Lipschitz maps, thus bridging the metric and differential approaches.  
-A natural direction to look for here is the relational semantics,  the somehow canonical ``Taylor'' semantics for $\STDLC$. 
+At this point, as the Taylor formula decomposes an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as interpreting  a $\lambda$-term
+as a limit of Lipschitz maps, in some sense, thus bridging the metric and differential approaches.  
+Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
 However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.
@@
 Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions?
-This way, eq.~\eqref{eq:taylor} could really be interpreted as a decomposition of $\lambda$-terms via Lipschitz maps.
+This way, \eqref{eq:taylor} could really be interpreted as decomposition of $\lambda$-terms via limits (indeed, an $\inf$s) of Lipschitz maps.
 In other words, unbounded term application could be seen

### 2023-01-22 section2.tex
commit

@@
 %Finally, in Section \ref{section6} we show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between {generalized metric spaces} and modules over the tropical semiring.
-
-\section{Tropical mathematics%: a possible synthesis
-}\label{sectionnew}
-
-\subsection{Tropicalization of polynomials and power series}
-
-At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}.
-Let us recall that a quantale is a complete lattice endowed with a continuous monoid action.
-In the case of $\Lawv$ the lattice is defined by the reverse order $\geq$ on $\BB R$, and the monoid action is provided by addition.
-Notice that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$.
-A consequence of these observations is that, as we discussed below, the tropical approach to linear algebra coincides with the study of ``$\Lawv$-valued matrices'', i.e.~of maps of the form $s: X\times Y\to \Lawv$ .
-In particular, a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we explicitly take in Section \ref{section6}).
-
-Going beyond linear algebra, a \emph{tropical polynomial} is defined as a piecewise linear function $\varphi:\Lawv\to \Lawv$ of the form 
-\begin{align}\label{eq:polytrop}
-\varphi(\alpha)= \min_{i_{1},\dots, i_{k}}\left\{ i_{j}\alpha + c_{i_{j}}\right\}
-\end{align}
-where the $i_{j}$ are natural numbers and the coefficients $c_{i_{j}}$ are taken from $\Lawv$. For instance, the polynomial
-$\varphi_{3}(\alpha)=\min\{ 3\alpha+1/8,2\alpha+1/4, \alpha+1/2, \alpha\}$ will be discussed in Section \ref{section4}, and its graph is illustrated in Fig.~\ref{fig:plot1}.
-%A value $\alpha\in \Lawv$ is a \emph{root} of the polynomial $P$ when
-%the minimum at $\varphi(\alpha)$ is attained at least twice (equivalently, when 
-% $\varphi$ is not differentiable at $\alpha$). In other words, the tropical roots of $\varphi$ coincide with the points where the slope of $\varphi$ changes. 
-%
-Intuitively, tropical polynomials look much like standard polynomials, although with ``$+$'' replaced by ``$\min$'', and ``$\times$'' replaced by ``$+$''. 
-In fact, this intuition can be made precise as follows: for any positive real $t$, the tropical polynomials like $\varphi$ are in one-to-one correspondence with the functions $f:[0,1]\to [0,1]$ which can be written as a \emph{parameterized} polynomial
-$f_{t}(x)= \sum_{i=1}^{n}t^{c_{i}}x^{n}$, with the $c_{i}\in [0,\infty]$. 
-%Hence, for any polynomial $p(x)= 
-%For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
-%\begin{align}
-%\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
-%\end{align}
-
-More generally, for any function $f:[0,1]\to [0,1]$ which can be written as a (parameterized) \emph{power series} of the form $f_{t}(x)= \sum_{n}t^{c_{n}}x^{n}$ (as we'll see in Section \ref{section5}, such functions arise naturally from the interpretation of probabilistic programs), we can define its \emph{tropicalization} as the function $\trop f: \Lawv \to \Lawv$ defined as follows:
-\begin{align}\label{eq:defTLS}
-\trop f(\alpha) =\inf_{n}\left\{ n\alpha+c_{n}\right\}
-\end{align}
-Such functions, called \emph{tropical Laurent series} \cite{Porzio2021}, will be discussed in more detail in Section \ref{section5}.
-
-The functions $f$ and $\trop f$ can be related by a limit passage as follows: the functions $\phi_{t}(x)= -\log_{t}x$ and $\varphi_{t}(\alpha)= t^{-\alpha}$ define continuous bijections between $[0,1]$ and $[0,\infty]$ and, by letting
-$\trop_{t}f: [0,\infty]\to [0,\infty]$ be defined by 
-$\trop_{t}f(\alpha)= \phi_{t}\circ f \circ \psi_{t}$, one has that 
-$\trop f= \lim_{t\to 0}\trop_{t}f$. 
-Indeed, one can check that the ``parameterized'' sums and product $\alpha \sumt{t}\beta:= \phi_{t}(\psi_{t}(\alpha)+\psi_{t}(\beta))= \log_{t}(t^{-\alpha}+t^{-\beta})$ and 
-$\alpha \prodt{t}\beta:= \phi_{t}( \psi_{t}(\alpha)\psi_{t}(\beta))=
-\log_{t}(t^{-\alpha}t^{-\beta})$ converge respectively to $\min\{\alpha,\beta\}$ and $\alpha+\beta$, when $t\to 0$ (this remark is an example of the so-called \emph{Maslov dequantization}, relating the semi-ring $\BB R$ with the tropical semiring $((-\infty, \infty], \min, +)$, see e.g.~\cite{Litvinov2007}).
-
-%
-%Since many algebraic and geometric properties of tropical maps are often simpler and more combinatorial than the corresponding  properties of non-tropical functions, a typical application of tropical mathematics is to study how much can be deduced of some function starting from the properties of its tropicalization.
-%In Section \ref{section5} we will follow a similar direction, investigating what quantitative properties of a higher-order programs are revealed by the study of its tropical interpretation.
-%
-%
-
-%
-%
-%or more generally as a power series $f(x)=\sum_{n}\widehat f_{n}x^{n}$ with coefficient $\widehat f_{n}\in [0,1]$, we can define its \emph{tropicalization} $\trop f: \Lawv \to \Lawv$ as the function 
-%\begin{align}
-%\trop f(\alpha)= \inf_{n}\left\{ 
-%\end{align}
-%
-%This correspondence can be made precise through the the so-called
-% \emph{de Maslov dequantization} \cite{}.
-% For each positive real $t$, any polynomial in $\BB R[x]$ can be written under the \emph{$t$-parameterized} form:
-% \begin{align}
-% p_{t}(x)= \sum_{i=1}^{k}t^{c_{i}}x^{i}
-% \end{align}
-% with the coefficients $c_{i}$ taken from $\Lawv$. 
-% It is clear then that tropical polynomials and $t$-parameterized polynomials admit a one to one correspondence between their presentations.
-% 
-% Actually, the $\varphi$s and the $p_{t}$s can be related by passing through some intermediate functions $\varphi_{t}$ introduced by Maslov.
-%For any $t>1$, the functions $\phi_{t}(x)=-\log_{t}x$ and $\psi_{t}(\alpha)=t^{-\alpha}$ are inverse of each other and define thus continuous (w.r.t.\ the usual topologies) bijections between the space of probabilities $[0,1]$ and $\BB R_{\geq 0}\cup\set{\infty}$ (we write $\log$ for the natural logarithm).
-%Moreover, if we set $\alpha \widetilde+ \beta:= \frac{\alpha+\beta}{2}$, $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta:=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$, it is known that: $\lim\limits_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$.
-%In this sense, setting $\Lawv_t:=([0,\infty],\sumt{t},\prodt{t})$, one says that $\Lawv_t\to_{t\to 0^+}\Lawv$.
-%Moreover, setting $\widetilde\Lawv:=([0,\infty],\widetilde+,\cdot)$, it can be shown that $\Lawv_t\simeq\widetilde\Lawv$ for all $t>0$, so the $\Lawv_t$ are all isomorphic, whereas at the limit we have a discontinuity: it can be shown that $\Lawv_t\not\simeq\Lawv$.
-%
-
-%
-%{\color{red}Lista delle cose da dire:}
-%
-%1) Def di quantale, come lattice e come complete idempotent semiring.
-%Among the the so-called \emph{tropical semirings}, we consider the \emph{Lawvere quantale/semiring}.
-%
-%2) Def di $\Lawv$, the \emph{Lawvere quantale}: seen as the idempotent complete semiring, it is $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,\cdot,0)$.
-%Seen as lattice it is defined by the order $\preceq$, which is the reversed order $\geq$ of the usual order $\leq$ on $\BB R_{\geq0}\cup\set{\infty}$.
-%
-%3) Maslov dequantisation:
-%
-%First, let us recall that for any non-negative real $t$, the functions $\phi_{t}(x)=-t\log x$ and $\psi_{t}(\alpha)=e^{-\alpha/t}$ are inverse of each other and define thus continuous (w.r.t.\ the usual topologies) bijections between the space of probabilities $[0,1]$ and $\BB R_{\geq 0}\cup\set{\infty}$ (we write $\log$ for the natural logarithm).
-%Moreover, if we set $\alpha \widetilde+ \beta:= \frac{\alpha+\beta}{2}$, $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta:=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$, it is known that: $\lim\limits_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$.
-%In this sense, setting $\Lawv_t:=([0,\infty],\sumt{t},\prodt{t})$, one says that $\Lawv_t\to_{t\to 0^+}\Lawv$.
-%Moreover, setting $\widetilde\Lawv:=([0,\infty],\widetilde+,\cdot)$, it can be shown that $\Lawv_t\simeq\widetilde\Lawv$ for all $t>0$, so the $\Lawv_t$ are all isomorphic, whereas at the limit we have a discontinuity: it can be shown that $\Lawv_t\not\simeq\Lawv$.
-%
-%4) Def di $\trop$ di un polinomio/serie, (\`e la stessa formula, dipende solo se gli indici sono finiti/infiniti).
-%Come sta scritto sotto, giusto un po' pi\`u formale (per esempio, scriverlo come Definizione).
-%
-%The fundamental observation that led to the study of mathematics over the \emph{tropical semi-ring} $\Lawv=([0,\infty],\min,+)$ was that, by replacing everywhere the ``$+$'' by the ``$\min$'' and the ``$\times$'' by the ``$+$'', many algebraic and geometric objects becomes combinatorial and their computation simpler.
-%
-%For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
-%\begin{align}
-%\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
-%\end{align}
-%Notably, the \emph{tropical roots} (whose definition is recalled in Section \ref{section3}) of $\trop p(\alpha)$ can be found through a rather simple (indeed polytime \cite{}) algorithm, and can be used to \emph{approximate} the actual roots of $p(x)$ \cite{}.
-%More generally, the tropicalization of a power series $f(x)=\sum_{n}\widehat f_{n}x^{n}$ yields a \emph{tropical Laurent series} \cite{} 
-%\begin{align}
-%\trop f(\alpha)= \inf_{n}\left\{n\alpha+ \widehat f_{n}\right\}
-%\end{align}
-%a class of functions that we will study in detail in Section \ref{section4}.
-%
-%%- generalities about tropical maths (tropicalisation $\trop P$  of polynomials and of Laurent series, and their roots -- all that without $\LREL$)
-%
-%
-%

### 2023-01-22 section3new.tex
commit

--- 2023-01-17 section3intro.tex
@@
-The study of matrices with values over the tropical semi-ring can be seen as a special case of the
+\subsection{Tropical linear algebra% and computation
+}
+
+At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}.
+Let us recall that a quantale is a complete lattice endowed with a continuous monoid action.
+In the case of $\Lawv$ the lattice is defined by the reverse order $\geq$ on $\BB R$, and the monoid action is provided by addition.
+Notice that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$.
+A consequence of these observations is that, as we discussed below, the tropical approach to linear algebra coincides with the study of ``$\Lawv$-valued matrices'', i.e.~of maps of the form $s: X\times Y\to \Lawv$ .
+In particular, a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we explicitly take in Section \ref{section6}).
+
+The study of matrices with values over the tropical semiring can be seen as a special case of the
 \emph{quantitative relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus.
@@
 
-The tropical relational semantics is thus provided by the category $\LREL$ of matrices with values over $\Lawv$ (which, being a quantale, is indeed a continuous semi-ring).
+The tropical relational model is thus provided by the category $\LREL$ of matrices with values over $\Lawv$ (which, being a quantale, is indeed a continuous semi-ring).
 %, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
@@
 In the next sections we will see how $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus.
+
+%
+%
+%
+%
+
+%
+%
+%or more generally as a power series $f(x)=\sum_{n}\widehat f_{n}x^{n}$ with coefficient $\widehat f_{n}\in [0,1]$, we can define its \emph{tropicalization} $\trop f: \Lawv \to \Lawv$ as the function 
+%\begin{align}
+%\trop f(\alpha)= \inf_{n}\left\{ 
+%\end{align}
+%
+%This correspondence can be made precise through the the so-called
+% \emph{de Maslov dequantization} \cite{}.
+% For each positive real $t$, any polynomial in $\BB R[x]$ can be written under the \emph{$t$-parameterized} form:
+% \begin{align}
+% p_{t}(x)= \sum_{i=1}^{k}t^{c_{i}}x^{i}
+% \end{align}
+% with the coefficients $c_{i}$ taken from $\Lawv$. 
+% It is clear then that tropical polynomials and $t$-parameterized polynomials admit a one to one correspondence between their presentations.
+% 
+% Actually, the $\varphi$s and the $p_{t}$s can be related by passing through some intermediate functions $\varphi_{t}$ introduced by Maslov.
+%For any $t>1$, the functions $\phi_{t}(x)=-\log_{t}x$ and $\psi_{t}(\alpha)=t^{-\alpha}$ are inverse of each other and define thus continuous (w.r.t.\ the usual topologies) bijections between the space of probabilities $[0,1]$ and $\BB R_{\geq 0}\cup\set{\infty}$ (we write $\log$ for the natural logarithm).
+%Moreover, if we set $\alpha \widetilde+ \beta:= \frac{\alpha+\beta}{2}$, $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta:=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$, it is known that: $\lim\limits_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$.
+%In this sense, setting $\Lawv_t:=([0,\infty],\sumt{t},\prodt{t})$, one says that $\Lawv_t\to_{t\to 0^+}\Lawv$.
+%Moreover, setting $\widetilde\Lawv:=([0,\infty],\widetilde+,\cdot)$, it can be shown that $\Lawv_t\simeq\widetilde\Lawv$ for all $t>0$, so the $\Lawv_t$ are all isomorphic, whereas at the limit we have a discontinuity: it can be shown that $\Lawv_t\not\simeq\Lawv$.
+%
+
+%
+%{\color{red}Lista delle cose da dire:}
+%
+%1) Def di quantale, come lattice e come complete idempotent semiring.
+%Among the the so-called \emph{tropical semirings}, we consider the \emph{Lawvere quantale/semiring}.
+%
+%2) Def di $\Lawv$, the \emph{Lawvere quantale}: seen as the idempotent complete semiring, it is $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,\cdot,0)$.
+%Seen as lattice it is defined by the order $\preceq$, which is the reversed order $\geq$ of the usual order $\leq$ on $\BB R_{\geq0}\cup\set{\infty}$.
+%
+%3) Maslov dequantisation:
+%
+%First, let us recall that for any non-negative real $t$, the functions $\phi_{t}(x)=-t\log x$ and $\psi_{t}(\alpha)=e^{-\alpha/t}$ are inverse of each other and define thus continuous (w.r.t.\ the usual topologies) bijections between the space of probabilities $[0,1]$ and $\BB R_{\geq 0}\cup\set{\infty}$ (we write $\log$ for the natural logarithm).
+%Moreover, if we set $\alpha \widetilde+ \beta:= \frac{\alpha+\beta}{2}$, $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta:=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$, it is known that: $\lim\limits_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$.
+%In this sense, setting $\Lawv_t:=([0,\infty],\sumt{t},\prodt{t})$, one says that $\Lawv_t\to_{t\to 0^+}\Lawv$.
+%Moreover, setting $\widetilde\Lawv:=([0,\infty],\widetilde+,\cdot)$, it can be shown that $\Lawv_t\simeq\widetilde\Lawv$ for all $t>0$, so the $\Lawv_t$ are all isomorphic, whereas at the limit we have a discontinuity: it can be shown that $\Lawv_t\not\simeq\Lawv$.
+%
+%4) Def di $\trop$ di un polinomio/serie, (\`e la stessa formula, dipende solo se gli indici sono finiti/infiniti).
+%Come sta scritto sotto, giusto un po' pi\`u formale (per esempio, scriverlo come Definizione).
+%
+%The fundamental observation that led to the study of mathematics over the \emph{tropical semi-ring} $\Lawv=([0,\infty],\min,+)$ was that, by replacing everywhere the ``$+$'' by the ``$\min$'' and the ``$\times$'' by the ``$+$'', many algebraic and geometric objects becomes combinatorial and their computation simpler.
+%
+%For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
+%\begin{align}
+%\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
+%\end{align}
+%Notably, the \emph{tropical roots} (whose definition is recalled in Section \ref{section3}) of $\trop p(\alpha)$ can be found through a rather simple (indeed polytime \cite{}) algorithm, and can be used to \emph{approximate} the actual roots of $p(x)$ \cite{}.
+%More generally, the tropicalization of a power series $f(x)=\sum_{n}\widehat f_{n}x^{n}$ yields a \emph{tropical Laurent series} \cite{} 
+%\begin{align}
+%\trop f(\alpha)= \inf_{n}\left\{n\alpha+ \widehat f_{n}\right\}
+%\end{align}
+%a class of functions that we will study in detail in Section \ref{section4}.
+%
+%%- generalities about tropical maths (tropicalisation $\trop P$  of polynomials and of Laurent series, and their roots -- all that without $\LREL$)
+%
+%
+%

### 2023-01-22 section4.tex
commit

@@
 
-A tropical polynomial, as we saw, is a piece-wise linear function of the form $f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}$.
-For instance, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{in+2^{-i}\}$
+A tropical polynomial iss a piece-wise linear function of the form \begin{equation}
+f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}
+\end{equation}
+For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{in+2^{-i}\}$
 are plotted in Fig.~\ref{fig:plot1} for $n=2,3,4$.
@@
 
-As anticipated in Section \ref{sectionnew}, a tropical Laurent series (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function that can be expressed as $f(x)=\inf_{n}\{nx+\matr f_{n}\}$, with $\matr  f_{n}$ a sequence in $\Lawv$.
+A \emph{tropical Laurent series} (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function that can be expressed as $f(x)=\inf_{n}\{nx+\matr f_{n}\}$, with $\matr  f_{n}$ a sequence in $\Lawv$.
 In other words, a tLs is a ``limit'' of tropical polynomials of higher and higher degree.
@@
 \end{theorem}
-\begin{proof}
-Call $\leq$ the (well-founded) pointwise order on $\N^k$.
-Now \mbox{$\C F_\epsilon:=\set{n\in\N \mid
+\begin{proof}[Proof sketch]
+\mbox{$\C F_\epsilon:=\set{n\in\N \mid
 \widehat f(n)<\infty \textit{ and } 
-\widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$}
-does the job.
+\widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$}, where $\leq$ is the pointwise order on $\N^k$.
+The core of the proof is showing that this set is indeed finite and enough for computing $f$.
 \end{proof}
@@
 
-Before looking at what happens in the case of non-linear programs, let us make the metric structure of $\LREL$ explicit. 
+%Before looking at what happens in the case of non-linear programs, let us make the metric structure of $\LREL$ explicit. 
 The following proposition provides a useful characterization of the functional metrics in $\LREL$, relying on 

### 2023-01-22 section5.tex
commit

@@
 In this section we illustrate a few directions in which the tropical semantics just introduced could be used to analyze quantitative properties of higher-order programs.
-The general pattern is to consider a program $M$, which can be expressed in the form of a polynomial or a power series, and 
-investigate what quantitative properties of $M$ can be deduced from properties of its tropicalization $\trop M$ (which are often easier to study).
-By the way, this is the approach taken in several well-studied applications of tropical geometry (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}).
 
-\subsection{Best Case Analysis and Metric Reasoning}
+Since algebraic and geometric properties in tropical mathematics are usually more tractable from a computational point of view, in several well-known applications (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}) one starts from a given model, typically expressed by some polynomial function $f$, and studies
+what properties of the model can be deduced from the \emph{tropicalization} of $f$, noted $\trop f$, i.e.~the transformation of $f$ into a tropical polynomial.
+
+%
+%
+%several well-known applications of tropical mathematics is to study how much can be deduced of some function starting from the properties of its tropicalization.
+%In Section \ref{section5} we will follow a similar direction, investigating what quantitative properties of a higher-order programs are revealed by the study of its tropical interpretation.
+%
+
+Here we follow a similar patter: we consider a program $M$, which can be expressed in the form of a polynomial or a power series $f$, and we 
+investigate what quantitative properties of $M$ can be deduced from the properties of $\trop f$, that will indeed coincide with the interpretation of $M$ in $\LREL_{!}$.
+
+\input{tropicalization}
+
+\subsection{Best case analysis and metric reasoning}
 
@@
 
-\subsection{Tropical Polynomials and Log-Probabilities}
+\subsection{Convergence log-probabilities}
 

### 2023-01-22 tropicalization.tex
commit

\subsection{The tropicalization of polynomials and power series}

%Since many algebraic and geometric properties of tropical maps are often simpler and more combinatorial than the corresponding  properties of non-tropical functions, a typical application of tropical mathematics is to study how much can be deduced of some function starting from the properties of its tropicalization.
%In Section \ref{section5} we will follow a similar direction, investigating what quantitative properties of a higher-order programs are revealed by the study of its tropical interpretation.
%

Let us first recall how standard polynomials and power series over $[0,1]$ can be turned into tLs via the so-called \emph{Maslov dequantization} \cite{Litvinov2007}.
%
%Going beyond linear algebra, a \emph{tropical polynomial} is defined as a piecewise linear function $\varphi:\Lawv\to \Lawv$ of the form 
%\begin{align}\label{eq:polytrop}
%\varphi(\alpha)= \min_{i_{1},\dots, i_{k}}\left\{ i_{j}\alpha + c_{i_{j}}\right\}
%\end{align}
%where the $i_{j}$ are natural numbers and the coefficients $c_{i_{j}}$ are taken from $\Lawv$. For instance, the polynomial
%$\varphi_{3}(\alpha)=\min\{ 3\alpha+1/8,2\alpha+1/4, \alpha+1/2, \alpha\}$ will be discussed in Section \ref{section4}, and its graph is illustrated in Fig.~\ref{fig:plot1}.
%A value $\alpha\in \Lawv$ is a \emph{root} of the polynomial $P$ when
%the minimum at $\varphi(\alpha)$ is attained at least twice (equivalently, when 
% $\varphi$ is not differentiable at $\alpha$). In other words, the tropical roots of $\varphi$ coincide with the points where the slope of $\varphi$ changes. 
%%
%Intuitively, tropical polynomials look much like standard polynomials, although with ``$+$'' replaced by ``$\min$'', and ``$\times$'' replaced by ``$+$''. 
%In fact, this intuition can be made precise as follows:

%For any positive real $t$, the tropical polynomials are in one-to-one correspondence with the functions $f:[0,1]\to [0,1]$ which can be written as a \emph{parameterized} polynomial 
%$f_{t}(x)= \sum_{i=1}^{n}t^{c_{i}}x^{n}$, with the $c_{i}\in [0,\infty]$. 
%%Hence, for any polynomial $p(x)= 
%%For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
%%\begin{align}
%%\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
%%\end{align}
%More generally, 
Let us fix a positive real $t>0$. For any function $f:[0,1]\to [0,1]$ which can be written as a parameterized {power series} of the form $f_{t}(x)= \sum_{n}t^{c_{n}}x^{n}$, 
% (as we'll see in Section \ref{section5}, such functions arise naturally from the interpretation of probabilistic programs),
  we let its \emph{tropicalization} be the tLs $\trop f: \Lawv \to \Lawv$ defined as follows:
\begin{align}\label{eq:defTLS}
\trop f(\alpha) =\inf_{n}\left\{ n\alpha+c_{n}\right\}
\end{align}
%Such functions, called \emph{tropical Laurent series} \cite{Porzio2021}, will be discussed in more detail in Section \ref{section5}.
Clearly, for any $t>0$, there is a one-to-one correspondence between the representation of power series in parameterized form and the associated tLs. Moreover, 
$f$ and $\trop f$ can be related by a limit passage as follows: the functions $\phi_{t}(x)= -\log_{t}x$ and $\varphi_{t}(\alpha)= t^{-\alpha}$ define continuous bijections between $[0,1]$ and $[0,\infty]$ and, by letting
$\trop_{t}f: [0,\infty]\to [0,\infty]$ be defined by 
$\trop_{t}f(\alpha)= \phi_{t}\circ f \circ \psi_{t}$, one has that 
$\trop f= \lim_{t\to 0}\trop_{t}f$. 
Indeed, one can check that the ``parameterized'' sums and product $\alpha \sumt{t}\beta:= \phi_{t}(\psi_{t}(\alpha)+\psi_{t}(\beta))= \log_{t}(t^{-\alpha}+t^{-\beta})$ and 
$\alpha \prodt{t}\beta:= \phi_{t}( \psi_{t}(\alpha)\psi_{t}(\beta))=
\log_{t}(t^{-\alpha}t^{-\beta})$ converge respectively to $\min\{\alpha,\beta\}$ and $\alpha+\beta$, when $t\to 0$.

### 2023-01-22 section1.tex
commit

@@
 
-Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} as an alternative approach to algebra and geometry where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given by ``$\min$'' and ``$+$''.
+Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} as an alternative approach to algebra and geometry where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
 %
@@
 %played by the idempotent operation $\min$.
-For instance, a polynomial like e.g.~$p(x,y)=x^{2}+xy^{2}+y^{3}$, when interpreted over the tropical semiring, translates as the piecewise linear function
+For instance, the polynomial $p(x,y)=x^{2}+xy^{2}+y^{3}$, when interpreted over the tropical semiring, translates as the piecewise linear function
 $
@@
 %This is not a \emph{ad-hoc} setting: 
-In the last decades, tropical geometry evolved into a vast and rich research domain, providing a combinatorial counterpart of usual algebraic geometry, with important connections with optimisation theory \cite{Akian2011, Akian2012}.
+In the last decades, tropical geometry evolved into a vast and rich research domain, providing a combinatorial counterpart of usual algebraic geometry, with important connections with optimisation theory \cite{Sturmfelds}.
 Computationally speaking, working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, the fundamental (and generally intractable) problem of finding the roots of a polynomial admits a \emph{linear time} algorithm in the tropical case (moreover, the tropical roots of a polynomial $p$ can be used to approximate the actual roots of $p$ \cite{Noferini2015}).

### 2023-01-22 section3A.tex
commit

@@
+We now turn to discuss how different variants of the simply typed $\lambda$-calculus are interpreted in $\LREL$.
+
 The linear simply typed $\lam$-calculus is a restriction of the ordinary $\lam$-calculus in which programs can only use their arguments \emph{exactly once}.
@@
 
-Its monoidal structure is given by a tensor product $\otimes$ acting on the objects as the Cartesian product of sets, and as the \emph{Kronecker product}
+The monoidal structure of $\LREL$ is given by a tensor product $\otimes$ acting on the objects as the Cartesian product of sets, and as the \emph{Kronecker product}
  %{(\color{red} Ref?)} 

### 2023-01-22 section3B.tex
commit

@@
 
-In order to interpret the $\STLC$, we need a Cartesian closed category (CCC).
-It is well known (\cite{MacLane}) that it is always possible to construct a CCC by taking the \emph{co-Kleisli} $\C C_!$ of a so-called \emph{Lafont category} $\C C$.
+In order to interpret the full $\STLC$, we need a Cartesian closed category (CCC).
+It is well-known \cite{Mellies2009} that it is always possible to construct a CCC by taking the \emph{co-Kleisli} $\C C_!$ of a so-called \emph{Lafont category} $\C C$.
 %, a construction we now quickly recall.
@@
 A SMCC is Lafont when it has finite products and it is equipped with a comonad $!$ (its \emph{Lafont exponential}) which, at level of objects, sends $X$ to an object $!X$ being the free commutative comonoid on $X$.
-Such object $!X$ represents the \emph{bang} connective of linear logic, granting infinite duplications via the infinite product $X^0\otimes X\otimes X^2\otimes X^3\otimes\cdots$, each factor representing a possible number of duplications.
-It is well known that, under mild conditions satisfied by $\QREL$, one can explicit this idea via the fact that the map $X\mapsto \finMS{X}$ (where $\finMS{X}$ is the set of finite multi-sets on $X$) lifts to a functor $!:\QREL\to\QREL$ which is a Lafont-exponential comonad.
-Specialising [Corollary III.6, \cite{Manzo2013}] to our case, we have:
+Such objects $!X$ represent the \emph{bang} connective of linear logic, granting infinite duplications via the infinite product $X^0\otimes X\otimes X^2\otimes X^3\otimes\cdots$, each factor representing a possible number of duplications.
+It is well-known that, under mild conditions satisfied by $\QREL$, one can explicit this idea via the fact that the map $X\mapsto \finMS{X}$ (where $\finMS{X}$ is the set of finite multi-sets on $X$) lifts to a functor $!:\QREL\to\QREL$ which is a Lafont-exponential comonad.
+Specializing [Corollary III.6, \cite{Manzo2013}] to our case, we have:
 
@@
 
-As it is well known, the Cartesian closed structure %of a category $\C C$ 
-allows to define a sound interpretation $\model{\Gamma\vdash M:A}\in\HOM{}{\model \Gamma}{\model A}$ of terms as morphisms.
+As it is well-known, the Cartesian closed structure %of a category $\C C$ 
+allows to define a sound interpretation $\model{\Gamma\vdash M:A}\in\HOM{\LREL_{!}}{\model \Gamma}{\model A}$ of terms as morphisms.
 In our case, we have:
@@
 In particular, bounded arrow types are interpreted via
-%Concretely, the types of this calculus (Equation~\ref{eq:Btyped}) are interpreted as: $\model o=\B 1$ and 
+%Concretely, the types of this calculus (Equation~\ref{eq:Btyped}) are interpreted as: 
+$\model o=\{\star\}$ and 
 $\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$.

### 2023-01-22 section3C.tex
Commit
commit

@@
 \begin{definition}
- The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho*\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise.
+ The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho*\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
 \end{definition}
@@
  In $\LREL_!$ equipped with $D$, all morphisms can be Taylor expanded, i.e.\ in $\LREL_!$ we have that, for all
-$t\in\HOM{\LREL_!}{Z}{X\multimap Y}$ and $s\in\HOM{\LREL_!}{Z}{X}$ we have:%, the evaluation of $t$ over $s$ yields 
+$t\in\HOM{\LREL_!}{Z}{X\multimap Y}$ and $s\in\HOM{\LREL_!}{Z}{X}$:%, the evaluation of $t$ over $s$ yields 
  \begin{align}\label{eq:taylorcat}
@@
 version of the usual Taylor expansion.
-Since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
-
-As a consequence, we have:
-
-\begin{corollary}\label{thm:M=infLip}
- If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is the $\inf$ of \emph{Lipschitz} functions.
-\end{corollary}
-
+Indeed, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
 %

### 2023-01-22 section3new.tex
commit

@@
 %This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
-It is worth observing that the formula for composition in $\LREL$ is the tropicalisation of the one defining it in $\QREL$, i.e.\ $(st)_{a,c}:=\inf\limits_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
+It is worth observing that the formula for composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
+reads as \ $(st)_{a,c}:=\inf\limits_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
  similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$ induced by matrices, which we call \emph{tropical linear}, are exactly those of shape $f(x)=\inf\limits_{b\in Y} \set{\matr f_{a,b}+x_a}$, for some matrix $\matr f$ from $X$ to $Y$.

### 2023-01-22 section4.tex
commit

@@
 correspond to a generalization of tropical Laurent series, i.e.~the tropicalization of power series.
-The study of power series in tropical mathematics is very recent (see e.g.~\cite{Porzio2021}) and many properties of these functions still have to be well-understood.
 
-In this section we recall some known facts about tropical Laurent series, and we establish some new properties, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
+%many properties of these functions still have to be well-understood. 
+In this section we establish some tropological and metric properties of tropical Laurent series, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
+The literature on power series in tropical mathematics is very recent (see e.g.~\cite{Porzio2021}) and several results we prove in this section are, to our knowledge, new.
 
@@
 
-A tropical polynomial iss a piece-wise linear function of the form \begin{equation}
+A tropical polynomial is a piece-wise linear function of the form \begin{equation}
 f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}
@@
 $f:\Lawv^X\to\Lawv^Y$ 
- for which the support $\C F=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ is \emph{finite}, which have thus shape $f(x)_b:=\min\limits_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$.
+ for which the support $\C F=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ is \emph{finite}, and which have thus shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$.
  This is again the generalisation of usual tropical polynomials to the case of infinitely many variables.
@@
 In particular, for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$.
-Indeed, we can compute  by hand that $\varphi$ coincides with $\varphi_{n}$
-for $\epsilon \geq \log2/2^{n+1}$.
+More, precisely, $\varphi$ coincides with $\varphi_{n}$
+for $\epsilon \geq 2^{-(n+1)}$, the smallest tropical root of $\varphi_{n}$.
 However, at
@@
 \begin{proof}[Proof sketch]
-\mbox{$\C F_\epsilon:=\set{n\in\N \mid
-\widehat f(n)<\infty \textit{ and } 
-\widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$}, where $\leq$ is the pointwise order on $\N^k$.
+Let $\C F_\epsilon$ be the set of $n\in\N^{k}$ such that 
+$\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
 The core of the proof is showing that this set is indeed finite and enough for computing $f$.
@@
 The tLs $\varphi$ is continuous on $\BB R_{\geq0}$ (w.r.t.\ the usual norm of real numbers).
-By considering the usual norm $\norm{x}_\infty:=\sup\limits_{a\in X} \absv{x_a}$ on $\Lawv^X$, we can generalise this property by dropping the case of $x$ having some $0$ coordinate:
+By considering the usual norm $\norm{x}_\infty:=\sup_{a\in X} \absv{x_a}$ on $\Lawv^X$, we can generalise this property by dropping the case of $x$ having some $0$ coordinate:
 
@@
 Suitable categories of cones have been recently investigated as models of probabilistic computation (\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
-We will just show that:
+Here we will just show that:
 
@@
 The norm $\norm{\cdot}_\infty$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$.
-
-As a first remark, let us mention that an immediate consequence of the lines after Theorem~\ref{thm:modelsTaylor} is:
-
-\begin{corollary}\label{thm:M=infLip}
- If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is the $\inf$ of \emph{Lipschitz} functions.
-\end{corollary}
-
-Another remark is that, as consequence of Theorem~\ref{theorem:fepsilon}, the tLs with finitely many variables are locally Lipschitz on all $\BB R_{>0}$.
-In this subsection we refine and generalise this kind of results.
-%
-%We will show that tLs satisfy suitable Lipschitz properties  w.r.t.~these metrics.
+We will show that tLs satisfy suitable Lipschitz properties  w.r.t.~these metrics.
 
@@
 \begin{proposition}\label{prop:boundedlip}
-If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
+If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a bounded matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
 \end{proposition}
@@
 First, since a tropical polynomial can be represented as a tropical linear function from $\Lawv^{\C M_{\mathrm{deg}(\varphi)(X)}}$ to $\Lawv$, from Proposition \ref{prop:boundedlip} we deduce:
-\begin{proposition}\label{prop:polylip}
+\begin{corollary}\label{prop:polylip}
 Any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
-\end{proposition}
+\end{corollary}
 
-The proposition above, together with Theorem \ref{theorem:fepsilon}, can be used to show that tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz over $(0,\infty)^{X}$.
-Yet, we can prove a more general statement, covering also the infinitary case.
+Using Corollary \ref{prop:polylip}, an immediate consequence of the lines after Theorem~\ref{thm:modelsTaylor} is:
+
+\begin{corollary}\label{thm:M=infLip}
+ If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is an $\inf$ of \emph{Lipschitz} functions.
+\end{corollary}
+
+% 
+%
+%The proposition above, together with Theorem \ref{theorem:fepsilon}, can be used to show that tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz over $(0,\infty)^{X}$.
+
+Moreover, as consequence of Theorem~\ref{theorem:fepsilon}, the tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $\BB R_{>0}$.
+Actually, we can prove a more general statement, also covering the infinitary case.
 
@@
 \begin{proof}
-As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be a finite matrix  and  its corresponding linear map $\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Prop.~\ref{prop:polylip}.
+As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be a finite matrix  and  its corresponding linear map $\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Corollary~\ref{prop:polylip}.
 Claim 2) follows directly from Theorem~\ref{thmTLSlocLip}.
@@
 Finally, let us discuss the differential structure.
-The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tLs $f:\Lawv^{X}\to \Lawv^{Y}$ into a tLs $D_{!}(f):(\Lawv^{X})^{2}\to \Lawv^{Y}$, linear in its first variable, and given by
+The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tLs $f:\Lawv^{X}\to \Lawv^{Y}$ into a tLs $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by
 \begin{equation}
@@
 One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
-Moreover, the Taylor formula \eqref{eq:taylorcat} yields 
-a Taylor formula for tLs of the form 
-$f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}$.
+Moreover, the Taylor formula \eqref{eq:taylorcat} yields a ``tropical'' Taylor formula for tLs of the form 
+\begin{equation}
+f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}
+\end{equation}
 The following result shows that the distance between two tropical maps can be approximated using the terms appearing in their Taylor expansions:

### 2023-01-22 section5C.tex
commit

@@
-\subsection{Resource Analysis for Differential Privacy}
+\subsection{Resource analysis for differential privacy}
 
-The typical situation in differential privacy is where one considers a probabilistic query $f: \mathsf{db}\to [0,1]^{X}$ over some database, and one requires that $f$ should not be \emph{too sensitive} to small changes in the output, in other to prevent potential leaks of private information about individual items in $\mathsf{db}$ (for instance, an element $x\in\mathsf{db}$ could be the list of students of some university and $f(x)$ could indicate the percentage of female students$\mathsf{db}$).
+The typical situation in differential privacy is where one considers a probabilistic query $f: \mathsf{db}\to [0,1]^{X}$ over some database, and one requires that $f$ should not be \emph{too sensitive} to small changes in the output, in other to prevent potential leaks of private information about individual items in $\mathsf{db}$ (for instance, an element $x\in\mathsf{db}$ could be the list of students of some university and $f(x)$ could indicate the percentage of female students in $x:\mathsf{db}$).
 
@@
 
-A well-studied approach to ensure differential privacy for higher-order programs is to use bounded type systems like $\mathsf{Fuzz}$ \cite{Reed2010}.
-Indeed, such systems come with an \emph{a priori} warrant that well-typed programs are Lipschitz.
+A well-studied approach to ensure differential privacy for higher-order programs is via linear type systems with graded exponentials like $\mathsf{Fuzz}$ \cite{Reed2010}: these systems provide an \emph{a priori} warrant that well-typed programs are Lipschitz.
 
@@
 We consider $X=Y=\{\star\}$ for simplicity.
-With the notations from \S \ref{section3}.A, using the fact that $\lim_{t\to 0}\trop_{t}f= \trop f$, as well as the commutation $f(\phi_{t}(\alpha)) = \phi_{t}(\trop_{t}f(\alpha))$ (which follows from the definition of $\trop_{t}$), one can see that the Lipschitz condition
+Using the fact that $\lim_{t\to 0}\trop_{t}f= \trop f$, as well as the commutation $f(\phi_{t}(\alpha)) = \phi_{t}(\trop_{t}f(\alpha))$ (which follows from the definition of $\trop_{t}$), one can see that the Lipschitz condition
 $|\trop f(\phi_{t}(x))-\trop f(\phi_{t}(y))|\leq \epsilon |\phi_{t}(x)-\phi_{t}(y)|$ translates into the condition 
-$f(x) \leq e^{\epsilon |\log x-\log y|} f(y)$ (supposing $y\geq x$ and $f(y)\neq 0$).
+$f(x) \leq e^{\epsilon d(x,y)} f(y)$ (supposing $y\geq x$ and $f(y)\neq 0$).
 \end{proof}

### 2023-01-22 section6.tex
commit

@@
 In this section we first reconstruct this correspondence, by combining some {folklore} results with more abstract ones from recent literature in enriched category theory \cite{Fuji, Stubbe2006, Shen2014}.
-Then we show that this correspondence can be lifted to a model of the full differential $\lambda$-calculus, by suitably generalizing the construction of the exponential comonad of $\LREL$.
+Then we show that this correspondence can be lifted to a model of the full differential $\lambda$-calculus, by suitably generalizing the construction of the Lafont exponential of $\LREL$.
 
@@
 
-\subsection{$\Lawv$-Modules}
+\subsection{$\Lawv$-modules}
 
@@
 The hom-sets $\Mod(M,N)$ have a natural $\Lawv$-module structure, defined pointwise.
-Moreover, the tensor product of $\Lawv$-modules $M\otimes N$ can be defined as the quotient of the usual tensor product of sup-lattices under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$ (see e.g.~\cite{Russo2007}).
+The tensor product of $\Lawv$-modules $M\otimes N$ can be defined as the quotient of the usual tensor product of sup-lattices under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$ (see e.g.~\cite{Russo2007}).
 Notably, any element of $M\otimes N$ can be identified with a join of basic tensors $x\otimes y$, corresponding to the equivalence class of the pair $\langle x,y\rangle$.
@@
 $\Hom(M,N^{\op})\simeq (M\otimes N)^{\op}$.
-Finally, both products and coproducts in $\Mod$ are given by the Cartesian products of the underlying posets, with action defined pointwise.
+Finally, $\Mod$ has \emph{biproducts}: products and coproducts are both given by the Cartesian product of the underlying posets, with action defined pointwise.
 %
@@
 
-\subsection{$\Lawv$-Categories}
+\subsection{$\Lawv$-categories}
 
 Lawvere was the first to observe that a metric space can be described as a \emph{$\Lawv$-enriched} category.
-Indeed, spelling out the definition, a $\Lawv$-enriched category (in short, a $\Lawv$-category) is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$, satisfying
+Indeed, spelling out the definition, a $\Lawv$-enriched category (in short, a $\Lawv$-category) is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$ satisfying
 \begin{align}
@@
 \end{align}
-This structure is often referred as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}.%, or as a \emph{quasi-metric space}. 
-Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, i.e.~it must satisfy $Y(f(x),f(y))\leq X(x,y)$.
+This structure is often referred to as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}. %, or as a \emph{quasi-metric space}. 
+Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, as functoriality reads as $Y(f(x),f(y))\leq X(x,y)$.
 
@@
 For any $\Lawv$-category $X$, the presheafs $[X^{\op},\Lawv]$ on $X$ form another $\Lawv$-category, with metric $[X^{\op},\Lawv](f,g)= \sup_{x\in X}\Lawv(f(x),g(x))$.
-Notice that, when $X$ has the discrete metric, the metric space $[X,\Lawv]^{\sym}$ coincides with $\Lawv^{X}$ with the norm metric.
+Notice that, when $X$ has the discrete metric, the metric space $[X,\Lawv]^{\sym}$ coincides with $\Lawv^{X}$ with the $\infty$-norm metric.
 The \emph{Yoneda embedding} is the faithful functor $\Yon: X\to [X^{\op},\Lawv]$ given by $\Yon(x)(y)=X(y,x)$.
 
-Actually, a well-studied example of $\Lawv$-categories is provided by $\Lawv$-modules.
+Actually, an important example of $\Lawv$-categories are precisely the $\Lawv$-modules:
 
@@
 
-\subsection{Complete $\Lawv$-Categories Correspond to $\Lawv$-Modules}
+\subsection{Complete $\Lawv$-categories correspond to $\Lawv$-modules}
 
@@
 \end{align}
-and the distance on basic tensors 
-of $X\otimes Y$ is given by
+and the distance on $X\otimes Y$ is given by
 \begin{align}
-%(X\otimes Y)(\alpha, \beta)=
+(X\otimes Y)(\alpha, \beta)=
+\sup_{i}\inf_{j}\left\{X(x_{i},x'_{j})+Y(y_{j},y'_{j}\right\},
+\end{align}
+ where $\alpha=\bigvee_{i}x_{i}\otimes y_{i}$ and 
+$\beta=\bigvee_{j}x'_{j}\otimes y'_{j}$, 
+from which we deduce that, for basic tensors
+$x\otimes y, x'\otimes y'$, their distance is just $X(x,x')+Y(y,y')$.
+
+%
+%
+%and is extended to arbitrary joins by continuity, i.e.~
+%$(X\otimes Y)(\alpha, \beta)=
 %\sup_{i}\inf_{j}X(x_{i},x'_{j})+Y(y_{j},y'_{j}),
-(X\otimes Y)(x\otimes y, x'\otimes y')=
-X(x,x')+Y(y,y')
-\end{align}
-and is extended to arbitrary joins by continuity.
-%where $\alpha=\bigvee_{i}x_{i}\otimes y_{i}$ and 
-%$\beta=\bigvee_{j}x'_{j}\otimes y'_{j}$, 
-%and thus coincides with the sum-metric over basic tensors;
+%$, where $\alpha=\bigvee_{i}x_{i}\otimes y_{i}$ and 
+%$\beta=\bigvee_{j}x'_{j}\otimes y'_{j}$.
+%%and thus coincides with the sum-metric over basic tensors;
 %\item the distance on the bi-product $\Pi_{i\in i}X_{i}$ is given by
@@
 
-\subsection{Exponential and Differential Structure}
+\subsection{Exponential and differential structure}
 

### 2023-01-22 section8.tex
commit

@@
 
-Finally, as suggested in Section \ref{section5}, an interesting direction to look at is the semantics of finiteness spaces \cite{Ehrhard2005}, which allows to model typable terms by means of \emph{finitary} Taylor expansions.
-Indeed, knowing that the application of a program $M$ to $N$ expands as a finite sum of linear applications (hence by a Lipschitz function in our setting), may allow one to predict how sensitive $M$ will be ``around $N$''.
+Finally, since bounds on the Taylor expansion translate into Lipschitz conditions, 
+ two interesting directions to explore are provided by (non-idempotent) intersection types and finiteness spaces \cite{Ehrhard2005}, as both methods are in principle capable of capturing \emph{finitary} bounds on the Taylor expansion.
 %
+Notably, knowing that the application of a program $M$ to $N$ expands as a finite sum of linear applications may allow one to predict how sensitive $M$ will be ``around $N$''. 
 %We have shown that:

### 2023-01-22 tropicalization.tex
commit

@@
 % (as we'll see in Section \ref{section5}, such functions arise naturally from the interpretation of probabilistic programs),
-  we let its \emph{tropicalization} be the tLs $\trop f: \Lawv \to \Lawv$ defined as follows:
+  we let its \emph{tropicalization} $\trop f: \Lawv \to \Lawv$ be the tLs defined as follows:
 \begin{align}\label{eq:defTLS}
@@
 %Such functions, called \emph{tropical Laurent series} \cite{Porzio2021}, will be discussed in more detail in Section \ref{section5}.
-Clearly, for any $t>0$, there is a one-to-one correspondence between the representation of power series in parameterized form and the associated tLs. Moreover, 
+Clearly, for any $t>0$, there is a one-to-one correspondence between the representations of power series in parameterized form and the associated tLs. Moreover, 
 $f$ and $\trop f$ can be related by a limit passage as follows: the functions $\phi_{t}(x)= -\log_{t}x$ and $\varphi_{t}(\alpha)= t^{-\alpha}$ define continuous bijections between $[0,1]$ and $[0,\infty]$ and, by letting

### 2023-01-22 appIII.tex
commit

@@
 \subsubsection{The language $\BSTLC$}
-In Section 4 we considered a variant $\BSTLC$ of $\STLC$ with affine simple types and an exponential $!_{n}A$, for all $n\in \BB N$, corresponding to a fragment of $\mathrm{Fuzz}$ \cite{Reed2010}. 
-We describe the rules of this fragment in Fig.~\ref{fig:fuzz}.
-The terms are generated by the grammar:
+In Section 4 we considered a variant $\BSTLC$ of $\STLC$ with linear simple types and a graded exponential $!_{n}A$, for all $n\in \BB N$, corresponding to a somehow simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}.
+%The terms are generated by the grammar:
+%
+%{
+%\begin{minipage}{\textwidth}
+%\begin{align*}
+%M:= x\mid \lambda x.M\mid MM \mid !M \mid \mathsf{let}\  !x=M \ \mathsf{in}\  M%\mid (t,u)\mid  \mathsf{let}\  (x,y)=t \ \mathsf{in} \  u 
+%\end{align*}\end{minipage}}\medskip\\
+%For the purposes of this article we limited ourselves to a minimal fragment of this language. For a more practical language see \cite{Reed2010,Gaboardi2017}.
+
+The types are generated by the grammar below:
 
@@
 \begin{align*}
-t,u:= x\mid \lambda x.t\mid tu \mid !t \mid \mathsf{let}\  !x=t \ \mathsf{in}\  u\mid (t,u)\mid  \mathsf{let}\  (x,y)=t \ \mathsf{in} \  u 
+A:= X \ \mid  \ !_{n}A \multimap A
 \end{align*}\end{minipage}}\medskip\\
-For the purposes of this article we limited ourselves to a minimal fragment of this language. For a more practical language see \cite{Reed2010,Gaboardi2017}.  
-Simple types are generated by the grammar below:
-
-{
-\begin{minipage}{\textwidth}
-\begin{align*}
-A,B:= X\mid  !_{n}A  \mid A\multimap B \mid A\otimes B
-\end{align*}\end{minipage}}\medskip\\
-Type judgements are of the form $\Gamma \vdash t:A$, where a context $\Gamma$ is a list of declarations of the form $x\in_{n}A$.
+Type judgements are of the form $\Gamma \vdash t:A$, where a context $\Gamma$ is a list of declarations of the form $x :_{n}A$, with $n\in \BB N$.
 We define the following operation $\Gamma+\Delta$  as follows:
@@
 () + () & =() \\
-(\Gamma, x\in_{m} A)+( \Delta, x\in_{n} A) & =  (\Gamma+\Delta), x\in_{m+n}A \\
-(\Gamma, x\in_{m}A)+\Delta & =(\Gamma+\Delta), x\in_{n}A \qquad (x\notin \Delta) \\
-\Gamma+ (\Delta, x\in_{m}A) &= (\Gamma+\Delta), x\in_{n} A \qquad (x\notin \Gamma)
+(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) & =  (\Gamma+\Delta), x:_{m+n}A \\
+(\Gamma, x:_{n}A)+\Delta & =(\Gamma+\Delta), x:_{n}A \qquad (x\notin \Delta) \\
+\Gamma+ (\Delta, x:_{n}A) &= (\Gamma+\Delta), x:_{n} A \qquad (x\notin \Gamma)
 \end{align*}\end{minipage}}\medskip\\
-Moreover, we let $m\Gamma$ be the context made all judgmenets $x\in_{mn}A$, where $(x\in_{n}A)\in \Gamma$.
+Moreover, we let $m\Gamma$ be the context made all judgements $x:_{mn}A$, where $(x:_{n}A): \Gamma$.
 
-Observe that one can always type an affine term like e.g.~$\lambda xy.x$ with a linear type $A\multimap B\multimap A$.
+The typing rules of $\BSTLC$ are illustrated in Fig.~\ref{fig:fuzz}.
+
+%Observe that one can always type an affine term like e.g.~$\lambda xy.x$ with a linear type $A\multimap B\multimap A$.
 Instead, a term like $\lambda xy.x(xy)$ containing two occurrences of $x$ cannot be given the linear type $(A\multimap A)\multimap (A\multimap A)$ but a type of the form
-$!_{2}(A\multimap A)\multimap (A\multimap A)$.
+%$!_{2}(A\multimap A)\multimap (A\multimap A)$.
 
@@
   {
-\AXC{$x\in_{n} A\in \Gamma$}
-\UIC{$\Gamma\vdash x\in A$}
+\AXC{}
+\UIC{$ x:_{1}A\vdash x: A$}
 \DP}
@@
 \begin{tabular}{c c }
-  {
-\AXC{$\Gamma, x\in_{1} A\vdash t\in B$}
-\UIC{$\Gamma\vdash \lambda x.t\in A\multimap B$}
-\DP}
+\AXC{$\Gamma \vdash M:A$}
+\UIC{$\Gamma, x:_{0}B \vdash M:A$}
+\DP
 
@@
 
-  {
-\AXC{$\Gamma \vdash t\in A\multimap B$}
-\AXC{$\Gamma\vdash u\in A$}
-\BIC{$\Gamma \vdash tu\in B$}
-\DP}
+\AXC{$\Gamma, x:_{n}B, y_{m}:B\vdash M:A$}
+\UIC{$\Gamma, x:_{n+m}B\vdash M[x/y]:A$}
+\DP
+
 \\
@@
   {
-\AXC{$\Gamma \vdash t\in A$}
-\AXC{$\Delta \vdash u\in B$}
-\BIC{$\Gamma+\Delta \vdash ( t,u) \in A\otimes B$}
+\AXC{$\Gamma, x:_{n} A\vdash M: B$}
+\UIC{$\Gamma\vdash \lambda x.M: !_{n}A\multimap B$}
 \DP}
@@
   {
-\AXC{$\Gamma\vdash t\in A \otimes B$}
-\AXC{$\Delta,x\in_{n}A, y\in_{n} B\vdash u:C$}
-\BIC{$n\Gamma+\Delta \vdash \mathsf{let}\  (x,y)=t \ \mathsf{in}  \ u\in C$}
+\AXC{$\Gamma \vdash M: A\multimap B$}
+\AXC{$\Delta\vdash N: A$}
+\BIC{$\Gamma +\Delta\vdash MN: B$}
 \DP}
 \\
-&
 
-\\
+%&
+
+%\\
+%
+%  {
+%\AXC{$\Gamma \vdash t\in A$}
+%\AXC{$\Delta \vdash u\in B$}
+%\BIC{$\Gamma+\Delta \vdash ( t,u) \in A\otimes B$}
+%\DP}
+%
+%& 
+%
+%
+%  {
+%\AXC{$\Gamma\vdash t\in A \otimes B$}
+%\AXC{$\Delta,x\in_{n}A, y\in_{n} B\vdash u:C$}
+%\BIC{$n\Gamma+\Delta \vdash \mathsf{let}\  (x,y)=t \ \mathsf{in}  \ u\in C$}
+%\DP}
+%\\
+%&
+
+%\\
+%
+%
+%
+%
+%
+%& 
+%
+%  {
+%\AXC{$\Gamma\vdash M: !_{n}A$}
+%\AXC{$\Delta,x\in_{mn}A\vdash u:C$}
+%\BIC{$m\Gamma+\Delta \vdash \mathsf{let}\  !x=M \ \mathsf{in} \ u :  C$}
+%\DP}
+
+\end{tabular}
+
+\bigskip
 
   {
-\AXC{$\Gamma \vdash t\in A$}
-\UIC{$s\Gamma \vdash !t \in \ !_{n}A$}
+\AXC{$\Gamma \vdash M: A$}
+\UIC{$n\Gamma \vdash M : \ !_{n}A$}
 \DP}
 
-&
-
-  {
-\AXC{$\Gamma\vdash t\in !_{n}A$}
-\AXC{$\Delta,x\in_{mn}A\vdash u:C$}
-\BIC{$m\Gamma+\Delta \vdash \mathsf{let}\  !x=t \ \mathsf{in} \ u\in C$}
-\DP}
-
-\end{tabular}
 \end{center}
@@
 
-\subsubsection{$(\LREL, !_{n})$ is a model of Bounded Linear Logic}
+\subsubsection{$(\LREL, !_{n})$ is a model of $\BSTLC$}
 
@@
 \end{itemize}
+
+The following definition provides an interpretation of $\BSTLC$ in any symmetric monoidal closed category with a $\BB N$-graded linear exponential comonad.
+
+\begin{definition}[interpretation of $\BSTLC$]
+Let $\C C$ be a symmetric monoidal closed category and $(D, w,c,\epsilon,\delta)$ be a $\BB N$-graded linear exponential comonad. 
+Let $\model{X}$ be fixed objects of $\C C $, one for each ground type $X$ of $\BSTLC$.
+
+One lifts the interpretation to types as 
+$\model{!_{n}A\multimap B}=D(n)(\model{A})\multimap \model B$. Moreover, one extends the interpretation to contexts via $\model{x:_{n}A}:= D(n)(\model A)$ and 
+$\model{\{x_{1}:_{n_{1}}A_{1},\dots, x_{k}:_{n_{k}}A_{k}\}}=
+\bigotimes_{i=1}^{k}\model{x_{i}:_{n_{i}}A_{i}}$.
+
+Then, one inductively defines an interpretation $\model{\Gamma \vdash M:A}\in \C C(\model\Gamma, \model A)$ of $\Gamma \vdash M:A$ by induction as follows:
+ \begin{itemize}
+\item $\model{x\in_{1}A\vdash x\in A}=\epsilon_{A}$;
+\item $\model{\Gamma+x:_{0}B\vdash M:A} =
+\model{\Gamma \vdash M:A}
+\circ (\model{\Gamma}\otimes  w_{\model B})
+$;
+\item $\model{\Gamma, x:_{m+n}B \vdash M[x/y]:A}=
+\model{\Gamma, x:_{m}B, y:_{m}B \vdash M:A}
+\circ
+(\model\Gamma \otimes c_{m,n,\model B})
+$;
+\item $\model{\Gamma \vdash \lambda x.M: !_{n}A\multimap B}=
+\Lambda (\model{\Gamma, x:_{n}A \vdash M:B})
+$, where $\Lambda$ is the isomorphism $ \C C(\model \Gamma \otimes D(n)(\model A), \model B) \to \C C(\model \Gamma, D(n)(\model A)\multimap \model B)$;
+
+\item $\model{\Gamma+\Delta \vdash MN:B}=
+\mathsf{ev}\circ \big( \model{\Gamma \vdash M:A\multimap B}
+\otimes
+\model{\Delta \vdash N:A}\big)$, where $\mathsf{ev}\in \C C((\model{A}\multimap \model B)\otimes \model A, \model B)$ is the evaluation morphism of $\C C$;
+
+\item $\model{n\Gamma\vdash M:!_{n}A}=
+ !_{n}(\model{\Gamma \vdash M:A})\circ \big(\delta_{n,m_{1},\model{A_{1}}}\otimes \dots \otimes 
+\delta_{n,m_{k},\model{A_{k}}}\big)
+$, where $\Gamma=\{x_{1}:_{m_{1}}A_{1},\dots, x_{k}:_{m_{k}}A_{k}\}$.
+%\item $\model{m\Gamma+\Delta \vdash \mathsf{let}\ !x=M\ \mathsf{in} \ u:C}=
+%\model{\Delta,x:_{mn}A\vdash u:C}
+%\circ
+%!_{m}(\model{\Gamma \vdash M:!_{n}A})
+%$.
+
+\end{itemize}
+\end{definition}
+
+Let us now show that bounded multisets defined a $\BB N$-graded linear exponential comonad over $\LREL$.
 

### 2023-01-22 appIV.tex
commit

@@
 
-One then defines a ``differential'' extension of usual $\beta$-reduction (see again [Section 3, \cite{Manzo2010}], or \cite{ER}), and thanks to the fact that $\C C$ is a $C\partial CC$, the interpretation above can can be proved to be sound for it ([Theorem 4.12, \cite{Manzo2010}]).
+One then defines a ``differential'' extension of usual $\beta$-reduction (see again [Section 3, \cite{Manzo2010}], or \cite{difflambda}), and thanks to the fact that $\C C$ is a $C\partial CC$, the interpretation above can can be proved to be sound for it ([Theorem 4.12, \cite{Manzo2010}]).
 

### 2023-01-22 section4.tex
commit

@@
 
-A tropical polynomial is a piece-wise linear function of the form \begin{equation}
+A tropical polynomial is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form \begin{equation}
 f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}
 \end{equation}
-For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{in+2^{-i}\}$
-are plotted in Fig.~\ref{fig:plot1} for $n=2,3,4$.
+where the $i_{j}$ are natural numbers and the $c_{i_{j}}$ are in $\Lawv$.
+For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$
+are illustrated in Fig.~\ref{fig:plot1} for $n=2,3,4$.
 A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable.
-In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes.
+In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes).
 For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
@@
 %\begin{remark}
-Notice that, by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf\limits_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
+Notice that, by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
 % \end{remark}

### 2023-01-22 section6.tex
commit

@@
 Let $X$ be a $\Lawv$-category, $x\in X$ and $\epsilon \in \Lawv$. The \emph{tensor of $x$ and $\epsilon$}, if it exists, is the colimit $\epsilon \otimes x:= \colim( [\epsilon],\Delta x)$, where
-$[\epsilon]: \B 1\pfun \B 1$ is the constantly $\epsilon$ distributor
-and $\Delta x:\B 1\to X$ is the constant functor. 
+$[\epsilon]: \{\star\}\pfun \{\star\}$ is the constantly $\epsilon$ distributor
+and $\Delta x:\{\star\}\to X$ is the constant functor. 
 \end{definition}

### 2023-01-23 appIV.tex
Commit notturno (finita parte Taylor)

--- 2023-01-21 appIV.tex
@@
 %We already know from Giulio's paper that $\LREL_!$ is a post-linear continuous $\Lawv$-ccc.
+
+\subsubsection{Proof of Theorem~\ref{thm:LREL!CCDC}}
+
+Theorem~\ref{thm:LREL!CCDC} says that $(\LREL_!,D)$ is a $CC\partial C$.
+This is proved in Proposition~\ref{propApp:CCDC}.
+First we need the following:
 
@@
 
-\begin{proposition}
+\begin{proposition}\label{propApp:CCDC}
 The category $\LREL_!$, endowed with the operator $D$, is a cartesian closed differential category.
@@
 
-\paragraph{Taylor Expansion }
+%\paragraph{Taylor Expansion }
+\subsubsection{Taylor Expansion in $\LREL_!$}
 
-To check the validity of Taylor expansion we must further check the following equation, given $f\in \LREL_!(C, B^{A})$ and $g\in \LREL_!(C,A)$:
+To check the validity of Taylor expansion in $\LREL_!$ we must further check the following equation, given $f\in \LREL_!(C, B^{A})$ and $g\in \LREL_!(C,A)$:
 $$
 \mathsf{ev}\circ \langle f,g\rangle= \inf_{n\in \BB N}
-(( \cdots (\Lambda^{-}(f) \underbrace{\star g)\cdots )\star g}_{n\text{ times}})\circ \langle \mathrm{id}, \infty\rangle
+\left\{(( \cdots (\Lambda^{-}(f) \underbrace{\star g)\cdots )\star g}_{n\text{ times}})\circ \langle \mathrm{id}, \infty\rangle\right\}
 $$
@@
 
-\subsubsection{The language and the semantics of $\STDLC$}
+\subsubsection{The language and the semantics of the $\STDLC$}\label{secApp:STDLC}
 
-The language we mainly consider in the main paper il the differential $\lambda$-calculus $\STDLC$, that we quickly define here, following [Section 3, \cite{Manzo2010}].
+The language we mainly consider in the main paper is the differential $\lambda$-calculus $\STDLC$, that we briefly define here, following [Section 3, \cite{Manzo2010}].
 
-The terms and the sums\footnote{Usually what we call here ``terms'' are called ``simple terms'', and what we call here ``sums'' are called ``differential terms''.} are mutually generated by the following grammars, then quotiented by usual $\alpha$-equivalence plus other equivalences handling the $+$ constructor.
+The terms and the sums\footnote{Usually what we call here ``terms'' are called ``simple terms'', and what we call here ``sums'' are called ``differential terms''.} are mutually generated by the following grammars:
 
@@
 \begin{align*}
-\textit{Terms: } & M,N:= x\mid \lambda x.M \mid M\BB T \mid \Der [M] \cdot N
+\textit{Terms: } & M,N:= x\mid \lambda x.M \mid M\BB T \mid \Der [M] \cdot N \qquad \textit{(Call $\Lambda^\partial$ the set of such terms)}
 \\
@@
 \end{align*}\end{minipage}}\medskip\\
-Simple types are generated by the grammar:
+quotiented by usual $\alpha$-equivalence plus other equivalences.
+Among those, we ask that $+$ is commutative, associative, with neutral element $0$ (thus the name ``sums''), and the constructors $\lam x.(\cdot)$, $(\cdot)\BB T$ and $\Der [\cdot] \cdot (\cdot)$ commute with $+$.
+One says that those constructors are \emph{linear}.
+Remark that $M(\cdot)$ is \emph{not} set to be linear.
+So, while $\lambda x.0=0\BB T=\Der [0] \cdot N=\Der [M] \cdot 0=0$, we have $M0\neq0$ in general.
+This is crucial for the definition of the Taylor expansion.
+One also needs to quotient for an equation, called ``permutative equality'', which basically says that the order of \emph{consecutive} linear applications $\Der [\cdot] \cdot (\cdot)$ does not count.
+Finally, for simplicity we follow here the tradition of ``eliminating the coefficients'', that is, we quotient for the idempotency of the sum.
+Sums are, then, just (\emph{finite}) subsets of $\Lambda^\partial$.
+
+Simple types are as usual generated by the grammar:
 
@@
 
-We finally recall the sound interpretation of $\STDLC$-terms in a $C\partial C$ category, following [Section 4.3, \cite{Manzo2010}].
+We finally recall the sound interpretation of $\STDLC$-terms in a $CC\partial C$ category, following [Section 4.3, \cite{Manzo2010}].
 
 \begin{definition}
-Let $\C C$ be a $C\partial C$ and $\model X$ fixed objects of $\C C$, one for each ground type $X$ of the $\STDLC$.
+Let $\C C$ be a $CC\partial C$ and $\model X$ fixed objects of $\C C$, one for each ground type $X$ of the $\STDLC$.
 
@@
 
-One then defines a ``differential'' extension of usual $\beta$-reduction (see again [Section 3, \cite{Manzo2010}], or \cite{ER}), and thanks to the fact that $\C C$ is a $C\partial CC$, the interpretation above can can be proved to be sound for it ([Theorem 4.12, \cite{Manzo2010}]).
+One then defines a ``differential'' extension of usual $\beta$-reduction (see again [Section 3, \cite{Manzo2010}], or \cite{difflambda}), and thanks to the fact that $\C C$ is a $CC\partial C$, the interpretation above can can be proved to be sound for it ([Theorem 4.12, \cite{Manzo2010}]).
 
 In the main paper we write $\model M$, but of course we mean $\model{\Gamma \vdash M:A}$.
+
+\subsubsection{The Taylor expansion of the $\STLC$}
+
+In the last lines of Section~\ref{sec:3B} we state Corollary~\ref{cor:T(M)=M} which is about the Taylor expansion of a ordinary $\lambda$-terms $M$.
+In the main paper we do not define it, let us briefly recall it here.
+For details, the reader can look at the many papers on the subject pointed in the bibliography.
+
+If sums (as defined in the above lines of Section~\ref{secApp:STDLC}) are \emph{finite} subsets of $\Lambda^\partial$, the Taylor expansion considers in general infinite subsets of $\Lambda^\partial$.
+Since we decided to ``eliminate the coefficients'', what we will define is often called the ``qualitative'' Taylor expansion.
+
+\begin{definition}
+
+ The (qualitative) \emph{Taylor expansion} is the following inductively defined map $\Theta : \Lambda \to \C P\left(\Lambda^\partial\right)$.
+\[\begin{array}{rcl}
+ \Te x & := & x \\
+ \Te{\lambda x.M} & := & \lambda x.\Te M \\
+ \Te{M\BB T} & := & \sum\limits_{n\in\N} \left(\Der^n [\Te M] \cdot \Te{\BB T}^n\right)0
+\end{array}\]
+where $\Te{M+\BB T}:=\Te M+\Te{\BB T}$ and $\Der^{n+1} [\Te M] \cdot \Te N^{n+1}:=\Der[\Der^n [\Te M] \cdot \Te N^n] \cdot \Te N$.
+\end{definition}
+
+We can now prove Corollary~\ref{cor:T(M)=M}.
+
+\begin{theorem}[Corollary~\ref{cor:T(M)=M}]
+ In $(\LREL_!,D)$, we have:
+ \[ \model M = \model{\Te M} \]
+for all $M\in\Lambda$.
+\end{theorem}
+\begin{proof}
+ By induction on $M$.
+ We only give the crucial case, that is $M=P\left(\sum\limits_j N_j\right)$.
+Looking at the definition of $\model{\cdot}$ we have:
+ \[\begin{array}{rclr}
+	\model{P(\sum\limits_j N_j)} 
+	& = &
+	 \mathsf{ev} \circ \langle \, \model{P}, \model{\sum\limits_j N_j}  \, \rangle
+	&
+	\\
+	& = &
+	 \inf\limits_{k\in\N}\left\{
+	 ((\dots (\Lambda^-(\model{P})\star \model{\sum\limits_j N_j})\cdots ) \star \model{\sum\limits_j N_j} )
+	 \circ \langle id,\infty \rangle \right\}
+	&
+	\textit{(Theorem~\ref{thm:modelsTaylor})}
+	\\
+	& = &
+	 \inf\limits_{k\in\N}\left\{
+	 ((\dots (\Lambda^-(\model{\Te P})\star \model{\sum\limits_j \Te N_j})\cdots ) \star \model{\sum\limits_j \Te N_j} )
+	 \circ \langle id,\infty \rangle \right\}
+	&
+	\textit{(ind.\ hyp.)}
+	\\
+	& = &
+	 \inf\limits_{k\in\N}\left\{
+	 \mathsf{ev} \circ \langle\, \Lambda\left(
+	 (\dots (\Lambda^-(\model{\Te P})\star \model{\sum\limits_j \Te N_j})\cdots ) \star \model{\sum\limits_j \Te N_j}
+	 \right) , \infty \, \rangle \right\}
+	& \textit{(general property of CCC's)}
+	\\
+	& = &
+	 \inf\limits_{k\in\N}\left\{
+	 \mathsf{ev} \circ \langle\, \model{\Der^n [\Te P] \cdot \Te{\sum\limits_j N_j}^n}, \infty \, \rangle \right\}
+	&
+	\\
+	& = &
+	 \inf\limits_{k\in\N}\left\{\model{(\Der^n [\Te P] \cdot \Te{\sum\limits_j N_j}^n)0} \right\}
+	&
+	\\ \\
+	& = &
+	 \model{\Te{P(\sum\limits_j N_j)}}
+	&
+ \end{array}\]
+\end{proof}

### 2023-01-23 section2.tex
Commit notturno (finita parte Taylor)

@@
 Another intriguing similarity between program derivatives and 
-actual derivatives is provided by the \emph{Taylor expansion}:
-in $\STDLC$ one can prove that any application $MN$ can be expanded as an infinite formal sum of \emph{linear} applications
+actual derivatives is provided by the \emph{Taylor expansion} $\C T$:
+in $\STDLC$ one can expand any application $MN$ as an infinite formal sum of \emph{linear} applications
 $\Der^{(k)}[M]\cdot N^{k}$, i.e.~where $N$ is passed exactly $k$ times to $M$:
 \begin{align}\label{eq:taylor}
-MN  =  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot (\Der^{(k)}[M] \cdot N^{k})0
+\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot (\Der^{(k)}[M] \cdot N^{k})0
 \end{align}

### 2023-01-23 section3C.tex
Commit notturno (finita parte Taylor)

@@
 %$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
-$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to linear application of $s$ to $u$, and $\Lambda^-$ is the uncurry operator.
+$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
 Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$,  i.e.~it coincides with the tropical %interpretation of the
 version of the usual Taylor expansion.
-Indeed, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
-%
-% corresponding to the syntactical \emph{differential substitution} of $\STDLC$.
-%%\begin{proof}
-% {\color{red}(Che ci scriviamo ??}
-%\end{proof}
+%Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
+Now, since $\LREL_!$ has countable sums (i.e.~all $\inf$'s converge), an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] shows:
+
+\begin{corollary}\label{cor:T(M)=M}
+ The interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
+\end{corollary}

### 2023-01-23 section4.tex
Commit notturno (finita parte Taylor)
commit

@@
 \begin{proposition}
-For all tropical linear functions $f,g:\Lawv^{X}\to \Lawv^{Y}$, $\norm{ \matr f-\matr g}_{\infty} =  \sup_{\B x\in \Lawv^{X}}
-\norm{ f(\B x)-g(\B x)}_{\infty}$.\end{proposition}
+For all tropical linear functions $f,g:\Lawv^{X}\to \Lawv^{Y}$, $\norm{ \matr f-\matr g}_{\infty} =  \sup_{x\in \Lawv^{X}}
+\norm{ f( x)-g(x)}_{\infty}$.\end{proposition}
 
@@
 
-Using Corollary \ref{prop:polylip}, an immediate consequence of the lines after Theorem~\ref{thm:modelsTaylor} is:
+Using Corollary \ref{prop:polylip}, an immediate consequence of %the lines after Theorem~\ref{thm:modelsTaylor} is:
+Corollary~\ref{cor:T(M)=M} is:
 

### 2023-01-23 section1.tex
commit

@@
 In the last decades, tropical geometry evolved into a vast and rich research domain, providing a combinatorial counterpart of usual algebraic geometry, with important connections with optimisation theory \cite{Sturmfelds}.
-Computationally speaking, working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, the fundamental (and generally intractable) problem of finding the roots of a polynomial admits a \emph{linear time} algorithm in the tropical case (moreover, the tropical roots of a polynomial $p$ can be used to approximate the actual roots of $p$ \cite{Noferini2015}).
+Computationally speaking, working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, the fundamental (and generally intractable) problem of finding the roots of a polynomial admits a \emph{linear time} algorithm in the tropical case (and, moreover,  the tropical roots can be used to approximate the actual roots \cite{Noferini2015}).
 The computational nature of tropical notions explains why these are so widely applied in computer science, notably for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).
 
-Coming back to our previous discussion on program semantics, tropical geometry might seem to provide precisely what we are looking for, as the monomials $x^{n}$ turn precisely into the Lipschitz functions $nx$.
+Coming back to our previous discussion on program semantics, tropical geometry might seem to provide precisely what we are looking for, as it turns the monomials $x^{n}$ into the Lipschitz functions $n\cdot x$.
 At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.

### 2023-01-23 section4.tex
commit

@@
 For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$
-are illustrated in Fig.~\ref{fig:plot1} for $n=2,3,4$.
+are illustrated in Fig.~\ref{fig:plot1} for $n\leq 4$.
 A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable.
@@
 \begin{axis}[samples=250]
-\addplot[red,domain=0:0.8] {min(2*x+1/4, x+1/2, 1)};
+\addplot[yellow,domain=0:0.8] {1+0.02};
+
+\addplot[orange,domain=0:0.8] {min(x+1/2, 1)+0.01};
+
+\addplot[red,domain=0:0.8] {min(2*x+1/4, x+1/2, 1};
 \addplot[blue,domain=0:0.8] {min(3*x+1/8,2*x+1/4, x+1/2, 1)-0.01};
@@
 %\ref{fig:plot1}
- (a) Plot of the tropical polynomials $\varphi_{2}$ (in red), $\varphi_{3}$ (in blue), $\varphi_{4}$ (in orange) and of their limit tLs $\varphi$ (in violet).
+ (a) Plot of the tropical polynomials $\varphi_{n}$, for $n\leq 4$ (from top to bottom), and of their limit tLs $\varphi$ (in violet).
 The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x=2^{-(i+1)}$, satisfying $ix+2^{-i}=(i+1)x+2^{-(i+1)}$.

### 2023-01-23 appIII.tex
Commit serale

@@
-\subsubsection{The language $\BSTLC$}
+\subsection{Section~\ref{sec:3B}: the language $\BSTLC$}
+
 In Section 4 we considered a variant $\BSTLC$ of $\STLC$ with linear simple types and a graded exponential $!_{n}A$, for all $n\in \BB N$, corresponding to a somehow simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}.
@@
 
-\subsubsection{$(\LREL, !_{n})$ is a model of $\BSTLC$}
+\subsection{Proofs from Section~\ref{sec:3B}: $(\LREL, !_{n})$ is a model of $\BSTLC$}
 

### 2023-01-23 appIV.tex
Commit serale

@@
 
-\subsubsection{Proof of Theorem~\ref{thm:LREL!CCDC}}
+\subsection{Section~\ref{sec:3C}: Proof of Theorem~\ref{thm:LREL!CCDC}}
 
@@
 %\paragraph{Taylor Expansion }
-\subsubsection{Taylor Expansion in $\LREL_!$}
+\subsection{Section~\ref{sec:3C}: proof of  Theorem~\ref{thm:modelsTaylor}}
 
-To check the validity of Taylor expansion in $\LREL_!$ we must further check the following equation, given $f\in \LREL_!(C, B^{A})$ and $g\in \LREL_!(C,A)$:
+Theorem~\ref{thm:modelsTaylor} states the validity of Taylor expansion in $\LREL_!$.
+
+We must check for it the following equation, given $f\in \LREL_!(C, B^{A})$ and $g\in \LREL_!(C,A)$:
 $$
@@
 
-Let us first compute the three morphisms $\mathsf{ev}, \Lambda^{-}$ and $\star$ explicitly:
+We do it in the following $4$ steps.
+
 \begin{enumerate}
 
-\item $\mathsf{ev}\in \Lawv^{ \C M_{\mathsf{fin}}(  ( \C M_{\mathsf{fin}}(A)\times B)       +  A    ) \times B  }$ is given by
+\item Let us compute the morphism $\mathsf{ev}$ explicitly: $\mathsf{ev}\in \Lawv^{ \C M_{\mathsf{fin}}(  ( \C M_{\mathsf{fin}}(A)\times B)       +  A    ) \times B  }$ is given by
 $$\mathsf{ev}_{\mu,y}=
@@
 
-\item given $g\in \LREL_!(C, B^{A})$, 
+\item Let us compute the morphism $\Lambda^{-}$ explicitly: given $g\in \LREL_!(C, B^{A})$, 
 $\Lambda^{-}(g)\in \LREL_!(C+A, B)$ is given by 
@@
 
-\item $f\star g$ is given by 
+\item Let us compute the morphism $\star$ explicitly: $f\star g$ is given by 
 $$
@@
 
-\end{enumerate}
-
-Given the definition of $\mathsf{ev}\circ \langle f,g\rangle$, to check the Taylor equation it is enough to check that, for all $N\in \BB N$, 
+\item We can now conclude: given the definition of $\mathsf{ev}\circ \langle f,g\rangle$, to check the Taylor equation it is enough to check that, for all $N\in \BB N$, 
 $$
@@
 \end{itemize}
+\end{enumerate}
 
-\subsubsection{The language and the semantics of the $\STDLC$}\label{secApp:STDLC}
+\subsection{Section~\ref{sec:3C}: The language and the semantics of the $\STDLC$}\label{secApp:STDLC}
 
-The language we mainly consider in the main paper is the differential $\lambda$-calculus $\STDLC$, that we briefly define here, following [Section 3, \cite{Manzo2010}].
+The language we mainly consider in the paper is the differential $\lambda$-calculus $\STDLC$, that we briefly define here, following [Section 3, \cite{Manzo2010}].
 
-The terms and the sums\footnote{Usually what we call here ``terms'' are called ``simple terms'', and what we call here ``sums'' are called ``differential terms''.} are mutually generated by the following grammars:
+The \emph{terms} and the \emph{sums}\footnote{Usually what we call here ``terms'' are called ``simple terms'', and what we call here ``sums'' are called ``differential terms''.} are mutually generated by the following grammars:
 
@@
 
-One lifts as usual the interpretation to types and contexts as $\model{A\to B}:=\model A\to \model B$ and $\model{x_1:A_1,\dots,x_n:A_n}:=\Pi_{i=1}^n\model{x_i:A_i}$.
+One lifts as usual the interpretation to types and contexts as $\model{A\to B}:=\model A\to \model B$ (the exponential object in $\C C$) and $\model{x_1:A_1,\dots,x_n:A_n}:=\Pi_{i=1}^n\model{x_i:A_i}$.
 
-Then, one inductively defines an interpretation $\model{\Gamma \vdash M:A}\in\C C(\model \Gamma,\model A)$ of $\Gamma \vdash M:A$ and one of $\model{\Gamma \vdash \BB T:A}\in\C C(\model \Gamma,\model{A})$ of $\Gamma \vdash \BB T:A$ by induction as:
+One inductively defines an interpretation $\model{\Gamma \vdash \BB T:A}\in\C C(\model \Gamma,\model{A})$ of $\Gamma \vdash \BB T:A$ by induction as:
 \begin{itemize}
@@
 
-In the main paper we write $\model M$, but of course we mean $\model{\Gamma \vdash M:A}$.
+In the paper we write $\model M$, but of course we mean $\model{\Gamma \vdash M:A}$.
 
-\subsubsection{The Taylor expansion of the $\STLC$}
+\subsection{Section~\ref{sec:3C}: proof of Corollary~\ref{cor:T(M)=M}}
 
 In the last lines of Section~\ref{sec:3B} we state Corollary~\ref{cor:T(M)=M} which is about the Taylor expansion of a ordinary $\lambda$-terms $M$.
-In the main paper we do not define it, let us briefly recall it here.
+In the main paper we do not define it, let us briefly recall it here, and the we prove the Corollary.
 For details, the reader can look at the many papers on the subject pointed in the bibliography.

### 2023-01-23 appV.tex
commit
Commit serale

@@
-\subsubsection{Concavity}
+\subsection{Section~\ref{sec:4A}: Proof of Proposition~\ref{prop:nondecr+conc}}
+
+We just show the concavity, the rest being easy.
 
@@
 
-\subsubsection{Proof of Theorem~\ref{theorem:fepsilon}}
+\subsection{Section~\ref{sec:4A}: Proof of Theorem~\ref{theorem:fepsilon}}
 
@@
 
-\subsubsection{Continuity of tLs}
+\subsection{Section~\ref{subsec:cont}: proof of Theorem~\ref{thm:cont}}
 
-We prove here the two results Theorem~\ref{thm:ScottCont} and Theorem~\ref{thm:cont} from Section~\ref{subsec:cont}.
-
-In order to prove Theorem~\ref{thm:cont}, we place ourselves in the general setting of locally convex topological vector spaces (LCTVS), and prove a general result about convex functions in this setting, adapting the argument from [Theorem , \cite{}].
+In order to prove Theorem~\ref{thm:cont}, we place ourselves in the general setting of locally convex topological vector spaces (LCTVS), and prove a general result about convex functions in this setting, adapting, in the main Proposition~\ref{prop:mainLCTVS}, an argument taken from [Proposition 4.4, \cite{Cobzas2017}].
 
@@
 %\newpage
+\subsection{Section~\ref{subsec:cont}: proof of Theorem~\ref{thm:ScottCont}}
+
+In order to prove such result we must quickly recall the theory of normed cones, which we do below.
+
+Here we denote with $\overline{\R}_{\geq 0}$ the set $\R_{\geq 0}\cup\set{\infty}$ (the same as $\Lawv$, but we prefer to explicit it here).
+
+\subsubsection{Normed cones}
 
@@
 
-In [Selinger] cones are required to also have ``strict addition'', meaning that $x+y=0 \Rightarrow x=y=0$.
+In \cite{Selinger2004} cones are required to also have ``strict addition'', meaning that $x+y=0 \Rightarrow x=y=0$.
 We do not add this requirement since it will automatic hold when considering normed cones.
@@
 
-In [Ehrhard-Pagani-Tasson, Crubillie] a normed $\overline{\R}_{\geq 0}$-cone is simply called a cone.
+In, e.g.~\cite{EhrPagTas2018}, a normed $\overline{\R}_{\geq 0}$-cone is simply called a cone.
 
@@
 
-The fundamental result in order to prove Theorem~\ref{thm:ScottCont} is the following, taken from [Selinger].
+The fundamental result in order to prove Theorem~\ref{thm:ScottCont} is the following, taken from \cite{Selinger2004}.
 
@@
 
-\begin{proposition}
+We finally obtain the desired:
+
+\begin{theorem}[Theorem~\ref{thm:ScottCont}]
  All tropical functions $f:\overline{\R}^X_{\geq 0}\to\overline{\R}^Y_{\geq 0}$ are Scott-continuous on $\R^X_{> 0}$ w.r.t.\ the cone-orders on its domain and codomain.
-\end{proposition}
+\end{theorem}
 \begin{proof}
@@
  Then $\inf\limits_i \supnorm{\bigvee\limits_i x^i - x^i} =0$, where $\bigvee\limits_i x^i - x^i$ exists because $\bigvee\limits_i x^i \geq x^i$ for all $i$.
- Since $f$ is $\supnorm{.}$-continuous on $\R^X_{> 0}$ (Corollary \ref{cor:tropCont}), then $\inf\limits_i \supnorm{f(\bigvee\limits_i x^i) - f(x^i)} =0$, where $f(\bigvee\limits_i x^i) - f(x^i)$ exists because $f(\bigvee\limits_i x^i) \geq f(x^i)$ for all $i$ being $f$ monotone (Remark \ref{rmk:tropMonot}).
+ Since $f$ is $\supnorm{.}$-continuous on $\R^X_{> 0}$, then $\inf\limits_i \supnorm{f(\bigvee\limits_i x^i) - f(x^i)} =0$, where $f(\bigvee\limits_i x^i) - f(x^i)$ exists because $f(\bigvee\limits_i x^i) \geq f(x^i)$ for all $i$ being $f$ monotone (Remark \ref{rmk:tropMonot}).
  We can therefore apply Proposition \ref{prop:infsup} to the directed net $(f(x^i))_i$ in $\overline{\R}^Y_{\geq 0}$, obtaining that $\bigvee\limits_i f(x^i)$ exists in $\overline{\R}^Y_{\geq 0}$ and it coincides with $f(\bigvee\limits_i x^i)$.
@@
 
-\newpage
+%\newpage
+
+\subsection{Proofs from Section~\ref{sec:4C}}
 
@@
 $$
-\delta^{(n)}f( x ^{n})=\mathsf D^{(n)}f( x ^{n}, \infty)
+\delta^{(n)}f( x ^{n})= D^{(n)}f( x ^{n}, \infty)
 $$
@@
 \begin{equation*}
-f( x )= \inf_{n}\left \{\mathsf D^{(n)}f( x ^{n},\infty)\right\}= \inf_{n}\left \{\delta^{(n)}f( x ^{n})\right\}
+f( x )= \inf_{n}\left \{ D^{(n)}f( x ^{n},\infty)\right\}= \inf_{n}\left \{\delta^{(n)}f( x ^{n})\right\}
 %\tag{Tropical Taylor}

### 2023-01-23 section4.tex
Commit serale

@@
 
-\subsection{From tropical polynomials to tLs}
+\subsection{From tropical polynomials to tLs}\label{sec:4A}
 
@@
 \begin{proof}
- It follows after adapting [Theorem , \cite{}] in order to prove that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
+ It follows after adapting [Proposition 4.4, \cite{Cobzas2017}] in order to prove that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
 \end{proof}
@@
 
-\subsection{Lipschitz-continuity of tLs}%$\Lawv^{X}$ as a metric space.}
+\subsection{Lipschitz-continuity of tLs}\label{sec:4C}%$\Lawv^{X}$ as a metric space.}
 

### 2023-01-24 appV.tex
Commit notturno

@@
 \begin{proposition}
-All tropical functions $f: Q^{X}\to Q^{Y}$ are concave.
+All tLs $f: Q^{X}\to Q^{Y}$ are concave.
 \end{proposition}
@@
 \begin{theorem}[Theorem \ref{theorem:fepsilon}]
-Let $f:\overline{\R}_{\geq 0}^{\{1,\dots, K\}} \to \overline{\R}_{\geq 0}$ tropical analytical, with matrix $\hat f : \N^{K} \to \overline{\R}_{\geq 0}$.
-Then, for all $0<\epsilon<+\infty$, there is $\mathcal{F}_\epsilon\subseteq\N^{K}$ s.t.\
+Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\matr f:\N^k\to\Lawv$.
+ For all $0<\epsilon<\infty$, there is $\C F_\epsilon \subseteq \N^k$ such that:
 \begin{enumerate}
- \item $\mathcal{F}_\epsilon$ is finite
- \item If $\mathcal{F}_\epsilon= \emptyset$ then $f( x ) = +\infty$ for all $ x \in \overline{\R}_{\geq 0}^{\{1,\dots, K\}}$
+ \item $\C F_\epsilon$ is finite
+ \item If $\mathcal{F}_\epsilon= \emptyset$ then $f( x ) = +\infty$ for all $ x \in \Lawv^k$
  \item If $f( x _0) = +\infty$ for some $ x _0\in [\epsilon,\infty)^{K}$ then $\mathcal{F}_\epsilon= \emptyset$
- \item The restriction of $f$ to $[\epsilon,+\infty]^{K}$ coincides with the  \emph{tropical polynomial}:
-      \[
-        P_\epsilon( x )= \min\limits_{ n  \in\mathcal{F}_\epsilon} \set{\hat f ( n  ) +  n   x }.
-      \]
-      where $ n   x = \sum_{i=1}^{K}n_{i}x_{i}$.
+ \item The restriction of $f$ on $[\epsilon,\infty]^k$ coincides  with the tropical {polynomial} \[P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\matr f(n)}\]
+where $nx:=\sum_{i=1}^k n_ix_i$.
 \end{enumerate}
@@
 
-Due to the previous discussion about the impossibility of stating Corollary \ref{cor:cont} in the case where one of the coordinates of $x$ is $0$, the continuity of tropical functions on the hyperplanes $\mathcal H_a:=\set{x\in {\R}_{\geq 0}^X \mid x_a=0}$, for $a\in X$, must be treated separately.
+Due to the previous discussion about the impossibility of stating Corollary \ref{cor:cont} in the case where one of the coordinates of $x$ is $0$, the continuity of tLs on the hyperplanes $\mathcal H_a:=\set{x\in {\R}_{\geq 0}^X \mid x_a=0}$, for $a\in X$, must be treated separately.
 Similarly, the continuity at points with infinite coordinates must be treated separately as well.
@@
 
- Remark also that tropical functions have no reason to be linear nor sublinear.
+ Remark also that tLs have no reason to be linear nor sublinear.
 \end{remark}
@@
 \begin{remark}\label{rmk:tropMonot}
- Tropical functions are monotone w.r.t.\ to the cone order on its domain and codomain.
- This is clear, using the definition of tropical functions, because $\mu x\leq \mu y$ if $x\leq y$.
+ tLs are monotone w.r.t.\ to the cone order on its domain and codomain.
+ This is clear, using the definition of tLs, because $\mu x\leq \mu y$ if $x\leq y$.
 \end{remark}
@@
 \begin{theorem}[Theorem~\ref{thm:ScottCont}]
- All tropical functions $f:\overline{\R}^X_{\geq 0}\to\overline{\R}^Y_{\geq 0}$ are Scott-continuous on $\R^X_{> 0}$ w.r.t.\ the cone-orders on its domain and codomain.
+ All tLs $f:\overline{\R}^X_{\geq 0}\to\overline{\R}^Y_{\geq 0}$ are Scott-continuous on $\R^X_{> 0}$ w.r.t.\ the cone-orders on its domain and codomain.
 \end{theorem}

### 2023-01-24 section3C.tex
Commit notturno

@@
 \begin{corollary}\label{cor:T(M)=M}
- The interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
+ In $\LREL_!$, the interpretation of the Taylor expansion of a $\STLC$-term $M$, given in \eqref{eq:taylor}, converges to the one of $M$.
 \end{corollary}

### 2023-01-24 section3new.tex
Mini commit

@@
 reads as \ $(st)_{a,c}:=\inf\limits_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
- similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$ induced by matrices, which we call \emph{tropical linear}, are exactly those of shape $f(x)=\inf\limits_{b\in Y} \set{\matr f_{a,b}+x_a}$, for some matrix $\matr f$ from $X$ to $Y$.
+ similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$ induced by matrices, which we call \emph{tropical linear}, are exactly those of shape $f(x)_b=\inf\limits_{a\in X} \set{\matr f_{a,b}+x_a}$, for some matrix $\matr f$ from $X$ to $Y$.
 %\end{remark}
 
-Since $\Lawv$ is a continuous (commutative) semiring, [Proposition III.3, \cite{Manzo2013}] immediately applies and gives:
+Since $\Lawv$ is a continuous commutative semiring, [Proposition III.3, \cite{Manzo2013}] immediately applies and gives:
 

### 2023-01-24 appV.tex
commit

@@
 \[\begin{array}{rclr}
- \hat f( n  ')+ n  ' x  & \leq & \hat f( n  ) + \epsilon +\mathbf  n' x  & \textit{by \eqref{eq:n'neps}} \\
- & \leq & \hat f( n  ) + ( n  - n  ') x  +  n  ' x  & \textit{because $\epsilon\leq\min x $ and $\min  x \leq( n  -\mathbf  n') x $} \\
+ \hat f( n  ')+ n  ' x  & \leq & \hat f( n  ) + \epsilon +  n' x  & \textit{by \eqref{eq:n'neps}} \\
+ & \leq & \hat f( n  ) + ( n  - n  ') x  +  n  ' x  & \textit{because $\epsilon\leq\min x $ and $\min  x \leq( n  -  n') x $} \\
  & = & \hat f( n  )+  n   x . &
@@
 \begin{lemma}\label{lemma:inf}
-Let $u,v: I\to Q$ and suppose $|u(i)-v(i)|\leq \delta$, for all $i\in I$.
+Let $u,v: I\to \Lawv$ and suppose $|u(i)-v(i)|\leq \delta$, for all $i\in I$.
 Then $|\inf_{i\in I}u(i)- \inf_{i\in I}v(i)|\leq \delta$. 
@@
 \begin{theorem}[local Lipschitz-continuity]
-Let $f: [0,\infty)^{X}\to Q$ be concave, monotone increasing, and continuous.  
+Let $f: [0,\infty)^{X}\to \Lawv$ be concave, monotone increasing, and continuous.  
 Then $f$ is locally Lipschitz-continuous.
@@
   \begin{proposition}\label{prop:functionalmetric}
-For all maps $f,g: Q\langle X\rangle\to Q\langle Y\rangle$,  
-$$
+For all maps $f,g: \Lawv^{X}\to \Lawv^{Y}$,  
+$
   \|\widehat f- \widehat g\|_{\infty}=
   \sup\{\|f( x )- g( x )\|_{\infty} \mid  x \in Q\langle X\rangle\}
-$$
+$.
 
@@
 The second characterization relates distances with the Taylor expansion. Let us briefly discuss the latter, first.
-For all $f: Q\langle X\rangle \to Q\langle Y\rangle$, let $\delta^{(n)}f:Q\langle X^{n}\rangle \to Q\langle Y\rangle$ indicate the $n$-linear function given by 
+For all $f: \Lawv^{X}\to\Lawv^{Y}$, let $\delta^{(n)}f:(\Lawv^{X})^{n}\to \Lawv^{Y}$ indicate the $n$-linear function given by 
 $$
@@
 \begin{proposition}\label{prop:taylormetric}
-For all $f,g:Q\langle X\rangle \to Q\langle Y\rangle$, 
+For all $f,g:\Lawv^{X}\to\Lawv^{Y}$, 
 $$
@@
 $\| \delta^{(n)}f-\delta^{(n)}g\|_{\infty}\leq\epsilon$, for all $n\in \BB N$. 
-Notice that $\widehat{\delta^{(n)}f}\in Q^{X^{n}\times Y}$ satisfies 
+Notice that $\widehat{\delta^{(n)}f}\in \Lawv^{X^{n}\times Y}$ satisfies 
 $\widehat{\delta^{(n)}f}_{a_{1},\dots, a_{n},b}= \widehat f_{[a_{1},\dots, a_{n}],b}$. Hence 

### 2023-01-24 appVII.tex
Commit serale
commit

@@
 %
+\subsection{Proofs from Section \ref{section7}}
+
 \begin{remark}\label{rem:functor}
-Given $f:X\to Q$, $f$ is a functor precisely when for all $x,y\in X$ $f(x)\leq\inf_{x'\in X}f(x')+X(x',x)$. 
-Indeed, if $f$ is a functor then $f(x)\leq f(x')+X(x',x)$, since $f(x)-f(x')=Q(f(x'),f(x))\leq X(x',x)$. 
+Given $f:X\to \Lawv$, $f$ is a functor precisely when for all $x,y\in X$ $f(x)\leq\inf_{x'\in X}f(x')+X(x',x)$. 
+Indeed, if $f$ is a functor then $f(x)\leq f(x')+X(x',x)$, since $f(x)-f(x')=\Lawv(f(x'),f(x))\leq X(x',x)$. 
 Conversely, if $f(x)\leq f(x')+X(x',x)$ holds for all $x'$, then 
-$Q(f(x),f(x'))=f(x')-f(x)\leq X(x',x)$. 
+$\Lawv(f(x),f(x'))=f(x')-f(x)\leq X(x',x)$. 
 \end{remark}
@@
 \begin{remark}[Yoneda embedding]
-%$\Lawv$ is a $\Lawv$-category with $Q(x,y)=|x-y|$. 
-%For any $\Lawv$-category $X$ we can define a new $\Lawv$-category $\mathsf{Dist}(X,Q)$ (that we note simply as $[X,Q]$) whose objects are the distributors $x:\{\star\}\pfun X$, or equivalently, the functors from $X$ to $\Lawv$, i.e.~those $x\in \Lawv^{X}$ such that $|x_{a}-x_{b}|\leq X(a,b)$, or equivalently $x_{a}+X(a,b) \geq x_{b}$.
-
-The \emph{Yoneda embedding} is the faithful functor $\B Y: X\to [X,Q]$ given by $\B Y(x)(y)=X(y,x)$. The functoriality and faithfulness of $\B Y$ follow from
+%$\Lawv$ is a $\Lawv$-category with $\Lawv(x,y)=|x-y|$. 
+%For any $\Lawv$-category $X$ we can define a new $\Lawv$-category $\mathsf{Dist}(X,Q)$ (that we note simply as $[X,\Lawv]$) whose objects are the distributors $x:\{\star\}\pfun X$, or equivalently, the functors from $X$ to $\Lawv$, i.e.~those $x\in \Lawv^{X}$ such that $|x_{a}-x_{b}|\leq X(a,b)$, or equivalently $x_{a}+X(a,b) \geq x_{b}$.
+
+The \emph{Yoneda embedding} is the faithful functor $\B Y: X\to [X,\Lawv]$ given by $\B Y(x)(y)=X(y,x)$. The functoriality and faithfulness of $\B Y$ follow from
 \begin{align}
-[X,Q](\B Y(x),\B Y(x'))= X(x,x') \tag{Yoneda}
+[X,\Lawv](\B Y(x),\B Y(x'))= X(x,x') \tag{Yoneda}
 \end{align}
@@
 \begin{align*}
-Q( \B Y(x)(y),\B Y(x')(y))&= \B Y(x')(y)-\B Y(x)(y) \\
+\Lawv( \B Y(x)(y),\B Y(x')(y))&= \B Y(x')(y)-\B Y(x)(y) \\
 &= X(y,x')-X(y,x) \leq X(x,x')
@@
 where the last step follows from $X(y,x)+X(x,x')\geq X(y,x')$. 
-From this we deduce that $[X,Q](\B Y(x),\B Y(x'))=\sup_{y\in X}Q( \B Y(x)(y),\B Y(x')(y))\leq X(x,x')$. 
+From this we deduce that $[X,\Lawv](\B Y(x),\B Y(x'))=\sup_{y\in X}\Lawv( \B Y(x)(y),\B Y(x')(y))\leq X(x,x')$. 
 For the converse direction, we have  
@@
 X(x,x')- X(x,x)
-\\& =\B Y(x')(x)-\B Y(x)(x)= Q(\B Y(x)(x),\B Y(x')(x))\\
-&\leq [X,Q](\B Y(x), \B Y(x'))
+\\& =\B Y(x')(x)-\B Y(x)(x)= \Lawv(\B Y(x)(x),\B Y(x')(x))\\
+&\leq [X,\Lawv](\B Y(x), \B Y(x'))
 \end{align*}
@@
 The \emph{opposite Yoneda embedding} is the faithful functor
-$\B Y^{\mathrm{op}}: X\to [X,Q]^{\mathrm{op}}$ given by $\B Y^{\mathrm{op}}(x)(y)=X(x,y)$. The functoriality and faithfulness of $\B Y^{\mathrm{op}}$ follow from
+$\B Y^{\mathrm{op}}: X\to [X,\Lawv]^{\mathrm{op}}$ given by $\B Y^{\mathrm{op}}(x)(y)=X(x,y)$. The functoriality and faithfulness of $\B Y^{\mathrm{op}}$ follow from
 \begin{align}\label{eq:yonedaop}
-[X,Q](\B Y^{\mathrm{op}}(x),\B Y^{\mathrm{op}}(x'))= X(x',x) \tag{Yoneda$^{\mathrm{op}}$}
+[X,\Lawv](\B Y^{\mathrm{op}}(x),\B Y^{\mathrm{op}}(x'))= X(x',x) \tag{Yoneda$^{\mathrm{op}}$}
 \end{align}
@@
 \begin{proof}
- For all $ x\in [X,Q]$, let $\sup x$ be defined as a weighted colimit via
+ For all $ x\in [X,\Lawv]$, let $\sup x$ be defined as a weighted colimit via
 $$
@@
 
-Let us check that $\sup: [X,Q]\to X$ is a functor. First, let us check the  inequality
+Let us check that $\sup: [X,\Lawv]\to X$ is a functor. First, let us check the  inequality
 \begin{align}\label{eq:in1}
@@
 \begin{align*}
-[X,Q](x,  y)&= \sup_{a\in X} y_{a}-x_{a}\\
+[X,\Lawv](x,  y)&= \sup_{a\in X} y_{a}-x_{a}\\
  &\stackrel{\tiny\eqref{eq:in2}}{\geq}
@@
  &= \sup_{a\in X}X(\sup x, a)-X(\sup  y, a)\\
- &= [X,Q](\B Y^{\mathrm{op}}(\sup y),\B Y^{\mathrm{op}}(\sup x))
+ &= [X,\Lawv](\B Y^{\mathrm{op}}(\sup y),\B Y^{\mathrm{op}}(\sup x))
  \tag{\ref{eq:yonedaop}}
@@
 
-Moreover, for all $x\in [X,Q]$, we have $\B Y(\sup x)\geq x$:  
+Moreover, for all $x\in [X,\Lawv]$, we have $\B Y(\sup x)\geq x$:  
 \begin{align*}
@@
 &= X(  \sup(\B Y(a)),\sup x) \\
-& \geq [X,Q]( \B Y(a), x)\\
+& \geq [X,\Lawv]( \B Y(a), x)\\
 &\geq x_{a}-\B Y(a)(a) \\
@@
 %where in the last step we use the fact that 
-%$[X,Q](  \B Y(a),x)= \sup_{b}x_{b}-X(a,b) = x_{a}$.
+%$[X,\Lawv](  \B Y(a),x)= \sup_{b}x_{b}-X(a,b) = x_{a}$.
 %
@@
 $f:\{\star\}\pfun X$ such that $f\coh g$ for some $g:X\pfun \{\star\}$, with
-$\B M(X)(f,f')=[X,Q](f,f')$. 
+$\B M(X)(f,f')=[X,\Lawv](f,f')$. 
 \end{definition}
@@
 %\item \emph{Cauchy-completeness} is the case where $X$ contains the points $\sup x, \sup y$, for any 
-% any two \emph{adjoint} presheaves $x, y\in [X,Q]$, i.e.~satisfying $0=\inf_{a}x_{a}+ y_{a}$ and $x_{a}+ x_{b}\geq X(a,b)$. 
+% any two \emph{adjoint} presheaves $x, y\in [X,\Lawv]$, i.e.~satisfying $0=\inf_{a}x_{a}+ y_{a}$ and $x_{a}+ x_{b}\geq X(a,b)$. 
 % Notice that given a pair $(x,  y)$ we can define a Cauchy sequence by finding $a_{n}$ satisfying 
@@
 %% 
-%% \item \emph{Isbell-completeness} (or \emph{MacNeill completeness}) is the case where $X$ contains the points $\sup x, \sup  y$, for any two presheaves $x,  y\in [X,Q]$ satisfying 
+%% \item \emph{Isbell-completeness} (or \emph{MacNeill completeness}) is the case where $X$ contains the points $\sup x, \sup  y$, for any two presheaves $x,  y\in [X,\Lawv]$ satisfying 
 %% $
@@
 
-\subsubsection{Tensors and $\Lawv$-Modules}
+\subsection{Tensors and $\Lawv$-Modules}
 
@@
 In any complete $\Lawv$-category, $x\otimes \epsilon \simeq \sup(  \B Y(x)+\epsilon  )$.
-In the complete $\Lawv$-category $[X,Q]$, $x\otimes \epsilon= x+\epsilon$.
+In the complete $\Lawv$-category $[X,\Lawv]$, $x\otimes \epsilon= x+\epsilon$.
 \end{lemma}
@@
 \end{align*}
-which shows $x\otimes \epsilon=\sup(\B Y(x)+\epsilon)$. In $[X,Q]$ we have 
-$[X,Q](x+\epsilon, x')=\sup_{a\in X}(x_{a}+\epsilon)-x'_{a}= (\sup_{a\in X}x_{a}-x'_{a})-\epsilon= [X,Q](x, x')-\epsilon$, which shows $x\otimes \epsilon \simeq x+\epsilon$, and since $[X,Q]$ is skeletal, $x\otimes \epsilon=x+\epsilon$.
+which shows $x\otimes \epsilon=\sup(\B Y(x)+\epsilon)$. In $[X,\Lawv]$ we have 
+$[X,\Lawv](x+\epsilon, x')=\sup_{a\in X}(x_{a}+\epsilon)-x'_{a}= (\sup_{a\in X}x_{a}-x'_{a})-\epsilon= [X,\Lawv](x, x')-\epsilon$, which shows $x\otimes \epsilon \simeq x+\epsilon$, and since $[X,\Lawv]$ is skeletal, $x\otimes \epsilon=x+\epsilon$.
 \end{proof}
@@
 \begin{example}
-The $\Lawv$-category $[X,Q]$ is cotensored, with $x\multimapinv \epsilon:= x-\epsilon$.
-Indeed we have $[X,Q](x,  y\multimapinv \epsilon)=\sup_{a\in X}( y_{a}-\epsilon)-x_{a}=(\sup_{a\in X} y_{a}-x_{a})-\epsilon= [X,Q](x,  y)-\epsilon$.
+The $\Lawv$-category $[X,\Lawv]$ is cotensored, with $x\multimapinv \epsilon:= x-\epsilon$.
+Indeed we have $[X,\Lawv](x,  y\multimapinv \epsilon)=\sup_{a\in X}( y_{a}-\epsilon)-x_{a}=(\sup_{a\in X} y_{a}-x_{a})-\epsilon= [X,\Lawv](x,  y)-\epsilon$.
 \end{example}
@@
 \begin{proof}
-For all $x\in [X,Q]$, let $\sup x:= \bigvee_{a\in X}a\otimes x_{a}$. 
+For all $x\in [X,\Lawv]$, let $\sup x:= \bigvee_{a\in X}a\otimes x_{a}$. 
 Let us check that $X(\sup x, b)= \sup_{a\in X}X(a,b)-x_{a}$, using Lemma \ref{lemma:supinf}:
@@
 %$f^{*}(x)(b)=\bigvee_{a\in X}k_{f}(a,b)+x_{a}$, and thus
-%in particular, for all $x\in [X,Q]$, 
+%in particular, for all $x\in [X,\Lawv]$, 
 %$\widehat f(x)(b)=f^{*}(x)(b)$.
@@
 \begin{itemize}
-\item[i.] $\Mod(Q,M)\simeq M$.
+\item[i.] $\Mod(\Lawv,M)\simeq M$.
 \item[ii.] $\Mod(M,N)\simeq \Mod(N^{\mathrm{op}},M^{\mathrm{op}})$.
@@
 \begin{proof}
-Define $\alpha:M\to \Mod(Q,M)$ by $\alpha(x)(\epsilon)=x\star\epsilon$ and 
-$\beta:\Mod(Q,M)\to M$ by $\beta(f)=f(0)$. Then we have that 
+Define $\alpha:M\to \Mod(\Lawv,M)$ by $\alpha(x)(\epsilon)=x\star\epsilon$ and 
+$\beta:\Mod(\Lawv,M)\to M$ by $\beta(f)=f(0)$. Then we have that 
 $\alpha(\beta(f))(\epsilon)=\alpha(f(0))(\epsilon)=f(0)\star\epsilon=f(\epsilon)$, and 
@@
 
-By i.~$\Mod(Q,M^{\mathrm{op}})\simeq M^{\mathrm{op}}$ and we conclude by ii.
+By i.~$\Mod(\Lawv,M^{\mathrm{op}})\simeq M^{\mathrm{op}}$ and we conclude by ii.
 \end{proof}
@@
 \item[ii.] $M\otimes N\simeq \Mod(M, N^{\mathrm{op}})^{\mathrm{op}}$.
-\item[iii.] $Q\otimes  M \simeq M\otimes Q\simeq Q$.
+\item[iii.] $\Lawv\otimes  M \simeq M\otimes \Lawv\simeq \Lawv$.
 \end{itemize}
@@
 
-For claim iii.~$Q\otimes M\simeq M\otimes M\simeq \Mod(M, \Lawv^{\mathrm{op}})^{\mathrm{op}}\simeq (M^{\mathrm{op}})^{\mathrm{op}}=M$. 
+For claim iii.~$\Lawv\otimes M\simeq M\otimes M\simeq \Mod(M, \Lawv^{\mathrm{op}})^{\mathrm{op}}\simeq (M^{\mathrm{op}})^{\mathrm{op}}=M$. 
 \end{proof}
@@
 $M^{X}$ coincides with the product $\Pi_{x\in X}M$. We have then 
-$\Lawv^{X}\otimes M \simeq (\Pi_{x}Q)\otimes M \simeq \Pi_{x}(Q\otimes M) \simeq \Pi_{x}M\simeq M^{X}$.
+$\Lawv^{X}\otimes M \simeq (\Pi_{x}\Lawv)\otimes M \simeq \Pi_{x}(\Lawv\otimes M) \simeq \Pi_{x}M\simeq M^{X}$.
 

### 2023-01-24 section1.tex
commit

@@
 On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function (which is rarely the case), it makes more sense to ask   whether they compute functions which do not differ \emph{too much}.
-This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
+This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
 

### 2023-01-24 section4.tex
Commit notturno
commit

@@
 \begin{theorem}\label{thm:cont}
- Any tropical Laurent series $f:\Lawv^X\to\Lawv$ is continuous on $\BB R_{>0}^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
+ All tLs $f:\Lawv^X\to\Lawv$ are continuous on $\BB R_{>0}^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
 \end{theorem}
@@
 Suitable categories of cones have been recently investigated as models of probabilistic computation (\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
-Here we will just show that:
+Here we will just mention that:
 
 \begin{theorem}\label{thm:ScottCont}
-  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders.
+ All tLs $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders.
   % on the domain and codomain.
@@
 \begin{proposition}\label{prop:troplinear}
-All tropical linear functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive.  
+All tropical \emph{linear} functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive.  
 \end{proposition}
@@
 \begin{corollary}\label{prop:polylip}
-Any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
+Any tropical polynomials $\varphi:\Lawv^{X}\to\Lawv$ are $\mathrm{deg}(\varphi)$-Lipschitz continuous.
 \end{corollary}
@@
 \begin{theorem}\label{thmTLSlocLip}
- All tropical Laurent series $\Lawv^X\to\Lawv$ are locally Lipschitz on $\BB R_{>0}^X$.
+ All tLs $\Lawv^X\to\Lawv$ are locally Lipschitz on $\BB R_{>0}^X$.
 \end{theorem}

### 2023-01-24 section1.tex
commit

@@
-In recent years, more and more interest in the programming language community has been directed towards the study of \emph{quantitative} properties of programs like e.g.~computing the number of computation steps, or convergence probabilities,
-as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence. 
-Notably, a significant effort has been made to extend, or adapt, well-established qualitative methods, like e.g.~type systems, relational logics or denotational semantics, to account for quantitative properties.
+In recent years, more and more interest in the programming language community has been directed towards the study of \emph{quantitative} properties of programs like computing the number of computation steps or convergence probabilities,
+as opposed to purely \emph{qualitative} properties like termination or program equivalence. 
+Notably, a significant effort has been made to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties.
 We can mention, for example,
@@
 
-On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function (which is rarely the case), it makes more sense to ask   whether they compute functions which do not differ \emph{too much}.
+On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}.
 This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
@@
 The fundamental motivation of this work is then the observation that
-this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ ``reads as'' the linear function $n\cdot x$.
+this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
 
@@
 
-Coming back to our previous discussion on program semantics, tropical geometry might seem to provide precisely what we are looking for, as it turns the monomials $x^{n}$ into the Lipschitz functions $n\cdot x$.
+Coming back to our discussion on program semantics, tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
 At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.
-Furthermore, connections between tropical linear algebra and metric spaces (in the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}), have also been observed \cite{Fuji}.
+Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
+However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.
 

### 2023-01-24 section3C.tex
commit

@@
 \begin{definition}
- The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho*\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
+ The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
 \end{definition}
@@
 For $t\in\HOM{\LREL}{!X}{Y}$, we have:
- $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu*\nu'*\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho*\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.
+ $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu+\nu'+\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho+\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.
 \end{remark}

### 2023-01-24 section4.tex
commit

@@
 
-We can now look at what happens with tLs, i.e.~when considering the full exponential $!$.
-First, since a tropical polynomial can be represented as a tropical linear function from $\Lawv^{\C M_{\mathrm{deg}(\varphi)(X)}}$ to $\Lawv$, from Proposition \ref{prop:boundedlip} we deduce:
+Observe that, for any tropical polynomial $\varphi:\Lawv^{X}\to \Lawv^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \Lawv$ (as any monomial $ix+c_{i}$ yields a matrix entry on $!_{i}X\times Y$).
+Hence, using Proposition \ref{prop:boundedlip}, we have:
 \begin{corollary}\label{prop:polylip}
-Any tropical polynomials $\varphi:\Lawv^{X}\to\Lawv$ are $\mathrm{deg}(\varphi)$-Lipschitz continuous.
+For any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$, $\varphi$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
 \end{corollary}
 
-Using Corollary \ref{prop:polylip}, an immediate consequence of %the lines after Theorem~\ref{thm:modelsTaylor} is:
-Corollary~\ref{cor:T(M)=M} is:
+Moreover, from Proposition \ref{prop:boundedlip}, we can also deduce the desired Lipschitz decomposition of the Taylor expansion (using Corollary~\ref{cor:T(M)=M}):
 
 \begin{corollary}\label{thm:M=infLip}
- If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is an $\inf$ of \emph{Lipschitz} functions.
+ If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is an $\inf$ of \emph{Lipschitz} functions (via its Taylor expansion \eqref{eq:taylorcat}).
 \end{corollary}
 
-% 
-%
-%The proposition above, together with Theorem \ref{theorem:fepsilon}, can be used to show that tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz over $(0,\infty)^{X}$.
-
-Moreover, as consequence of Theorem~\ref{theorem:fepsilon}, the tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $\BB R_{>0}$.
+Let us now look at what happens with tLs, i.e.~when considering the full exponential $!$.
+As consequence of Theorem~\ref{theorem:fepsilon}, the tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $\BB R_{>0}$.
 Actually, we can prove a more general statement, also covering the infinitary case.
@@
 \begin{equation}
-D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu*a}+x_{a}+\mu y\right\}
+D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a}+x_{a}+\mu y\right\}
 \end{equation}

### 2023-01-24 section2.tex
commit

@@
 
-These observations have led to the study of linear $\lambda$-calculi with \emph{graded} exponentials, like the system $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied to the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}.
-The types of such systems are defined by combining linear constructors with a \emph{graded exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
+These observations have led to the study of linear $\lambda$-calculi with \emph{graded} exponentials, like $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied to the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}.
+The types of such systems are defined by combining linear constructors with a \emph{graded linear exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
 In the following sections we will sometimes make reference to a basic graded type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined via $A::= o  \mid   !_{n}A\multimap A  $,
@@
 Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions?
-This way, \eqref{eq:taylor} could really be interpreted as decomposition of $\lambda$-terms via limits (indeed, an $\inf$s) of Lipschitz maps.
+This way, \eqref{eq:taylor} could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps.
 In other words, unbounded term application could be seen

### 2023-01-24 section4.tex
commit

@@
 %many properties of these functions still have to be well-understood. 
-In this section we establish some tropological and metric properties of tropical Laurent series, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
+In this section we establish some topological and metric properties of tropical Laurent series, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
 The literature on power series in tropical mathematics is very recent (see e.g.~\cite{Porzio2021}) and several results we prove in this section are, to our knowledge, new.

### 2023-01-24 section4.tex
Commit finale

@@
 
-Observe that, for any tropical polynomial $\varphi:\Lawv^{X}\to \Lawv^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \Lawv$ (as any monomial $ix+c_{i}$ yields a matrix entry on $!_{i}X\times Y$).
+Observe that, for any tropical polynomial $\varphi:\Lawv^{X}\to \Lawv^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \Lawv$ (as a monomial $\mu_ix+c_{i}$ yields a matrix entry on $!_{\#\mu_i}X\times Y$).
 Hence, using Proposition \ref{prop:boundedlip}, we have:

### 2023-03-16 rebuttal.txt
commit

We thank the anonymous reviewers for their comments and suggestions.
We reply to a few of them below.

Reviewer 1:

> "which can be written as a parameterized power series of the form" what does this mean? $f$ is a single function, but $f_t$ is an infinite family of functions. does it mean $f$ can be written as $f_t$ for some fixed constant $t$?"

Rigirare la frittata su f, f_t ecc.

> Theorem 24 to bottom of column: I found this discussion hard to follow.

One of the core aspects of Theorem 24 is that on any L-module one can define a canonical metric; the goal of this paragraph is thus to describe in an explicit way the metrics associated with the SMCC structure of LMOD (or, equivalently, the metrics defining the SMCC structure of LCCAT). Yet we think that we should have noted X,Y rather as M,N, in order to avoid confusion.

Concerning tensors, in the previous page (after Prop. 19) we introduced the tensor product in LMOD and recalled that an element of it is a join of basic tensors x(x)y.

Reviewer 2:

> "Some of the calculations in the appendix regarding the differential story can be referenced away."

We thank the reviewer for his careful reading of the whole paper (including the long appendix), and for his numerous suggestions concerning results appearing in the recent literature. We were not aware of Lemay's "Coderelictions for Free Exponential Modalities" and we found its results very interesting. We will adjust our presentation following such remarks.

As a general remark, our results on LREL and LREL_! in Section III are based on relatively unsurprising checks (yet, it is always interesting to know that one can avoid long checks and use a theorem instead!). Since this paper is also addressed to the program metrics community, which is less familiar with linear logic and weighted relational models, this section is mainly thought to provide a relatively concrete presentation of (the tropical variant of) this model, studied in more detail in the following sections.

> "Theorem 81 is important and should have appeared in the main paper."

Theorem 29 is (essentially) Theorem 81 from the appendix, but without mention of differential linear categories. We chose not to mention the latter in the main text (and thus to limit our discussion of differential operators to the cartesian case) because our results concerning them are either folklore or based on well-known results/techniques.

> "In the introduction, you say "the somehow canonical "Taylor semantics for STdLC". Why is that? What is the justification for this statement?"

...

> "I suggest rewriting Section III.D based on the comments made above, and explain that the differential structure comes from the fact that LREL is a differential category."

OK? Connettere alla discussione sopra??

Reviewer 3:

> "It is good that the model can characterize
    differential privacy of programs, but its implication to the
    verification tasks is not discussed."

Of course, the applications in Section 5 are just sketched. Our main goal in this paper was to demonstrate that tropical methods, which are widely applied in computer science (but not in the linear logic/lambda-calculus community) can make sense also for studying the quantitative semantics of the lambda-calculus. Concrete developments of such applications are thus left for future work.

> "Another limitation of the
    paper's model is that it can not handle probabilistic programs
    involving continuous distributions"

This is indeed a well-known limitations of the semantics of probabilistic programs based on weighed relational models (e.g. probabilistic coherent spaces [33]). Still, such semantics are well-studied and led to strong results (e.g. full abstraction theorems). Still, we agree that this limitation should be mentioned.

> "In complex analysis, Laurent series include terms of negative
    degrees, while here I think we only consider terms with
    non-negative degrees."

This is true, yet in the literature on tropical mathematics (to which we refer, e.g. [59]) these are defined only with positive coefficients. We agree that this should be explicitly mentioned, in order to avoid confusion.

> "Is Theorem 24, which is one of the main claims in the paper,
    original, or subsumed by [71]?"

We think that the main contribution in Section 6 is to show that an abstract correspondence known in the categorical literature can be extended to account for the relevant structure for interpreting the differential lambda-calculus. Hence we see Theorems 27 and 29, rather than Theorem 24 (whose statement and proof can be deduced from the results of [71]) as the main claims of this section.

### 2023-03-17 rebuttal.txt
Commit pre-pranzo
Ho scritto le risposte al reviewer 1 ed aggiustato qualcosa al 3. Dopo pranzo vedo il 2.

@@
-We thank the anonymous reviewers for their comments and suggestions.
-We reply to a few of them below.
+We thank very much all the reviewers for the careful comments.
+We will of course take care in correcting all the typos and unclear sentences.
+We thank in particular reviewer 2 for his/her thorough reading, and we will perform the useful modifications he/her suggested (see below).
 
@@
 
-> "which can be written as a parameterized power series of the form" what does this mean?
+> p3 (1)
+
+One just takes the semimodule generated by differential terms with coefficients in Q (or another semiring).
+Equivalently, you can see them as formal linear combinations with coefficients in the choosen semiring.
+
+> p5 "setting $t^!(x) := t o_! x$"...
+
+We are tacitly making some standard identifications: one identifies Q^X with the morphisms in QRel_!(T,X)=QRel(!T,X) where T=\emptyset is the terminal object.
+This is possible by mapping x \in Q^X to x' \in QRel(!T,X) defined by x'_{\emptyset,a} := x_a (remark that !T={\emptyset}).
+In this way, given t \in QRel_!(X,Y), one defines the function t^!
+: Q^X --> Q^Y by t(x) := t °_!
+x', which is a morphism T -> Y in QRel_!, seen in the same way as an element of Q^Y.
+
+>  p7 what is $[[M]]^!$?
+
+It is the function just defined in the previous answer (i.e. in equation (6) at page 5).
+
+> p7 "which can be written as a parameterized power series of the form" what does this mean?
 $f$ is a single function, but $f_t$ is an infinite family of functions.
@@
 Rigirare la frittata su f, f_t ecc.
+
+> p9 Prop 17: why is the input of $f$ suddenly $[0, 1]^X$?
+
+In the discussion around Def2 we were informal on what we meant by "database".
+Prop17 is, instead, a rigorous statement, so we stated it with the correct types for f.
+We are supposing to see a database as a function X --> [0,1], which is indeed often the case (for instance, for a relational database -- take X as the key/attributes and [0,1] as the datas).
 
@@
 
-Of course, the applications in Section 5 are just sketched.
+You are right, the applications in Section 5 are just sketched.
 Our main goal in this paper was to demonstrate that tropical methods, which are widely applied in computer science (but not in the linear logic/lambda-calculus community) can make sense also for studying the quantitative semantics of the lambda-calculus.
@@
 
+> p.4 where is this hat operation defined?
+
+It is supposed to be defined by those line. We agree that the sentence is not clear, so we will rephrase it.
+
 > "In complex analysis, Laurent series include terms of negative
@@
 [59]) these are defined only with positive coefficients.
-We agree that this should be explicitly mentioned, in order to avoid confusion.
+We will explicitly mention it, in order to avoid confusion.
 
@@
 
-We think that the main contribution in Section 6 is to show that an abstract correspondence known in the categorical literature can be extended to account for the relevant structure for interpreting the differential lambda-calculus.
-Hence we see Theorems 27 and 29, rather than Theorem 24 (whose statement and proof can be deduced from the results of [71]) as the main claims of this section.
+We think that the main contribution in Section 6 is to show that an abstract correspondence (expressed indeed by Theorem 24) known in the categorical literature can be extended to account for the relevant structure for interpreting the differential lambda-calculus.
+Hence, we see Theorems 27 and 29, rather than Theorem 24 (whose statement and proof can be deduced from the results of [71]) as the main claims of this section.

### 2023-03-17 rebuttal.txt
Committ

@@
 We will of course take care in correcting all the typos and unclear sentences.
-We thank in particular reviewer 2 for his/her thorough reading, and we will perform the useful modifications he/her suggested (see below).
+We thank in particular reviewer 2 for his/her thorough reading (and numerous suggestions concerning results appearing in the recent literature), and we will perform the useful modifications he/her suggested (see below).
 
@@
 Yet we think that we should have noted X,Y rather as M,N, in order to avoid confusion.
-
 Concerning tensors, in the previous page (after Prop. 19) we introduced the tensor product in LMOD and recalled that an element of it is a join of basic tensors x(x)y.
@@
 
-> "Some of the calculations in the appendix regarding the differential story can be referenced away."
+> Comment 1 ("Some of the calculations in the appendix regarding the differential story can be referenced away"), comment 2, comment 7, comment 8.
 
-We thank the reviewer for his careful reading of the whole paper (including the long appendix), and for his numerous suggestions concerning results appearing in the recent literature.
 We were not aware of Lemay's "Coderelictions for Free Exponential Modalities" and we found its results very interesting.
-We will adjust our presentation following such remarks.
+We will surely adjust our presentation following such remarks!
+In particular, we will follow your suggestion to "carefully providing a proper reference for all the important math definitions".
 
 As a general remark, our results on LREL and LREL_!
-in Section III are based on relatively unsurprising checks (yet, it is always interesting to know that one can avoid long checks and use a theorem instead!).
-Since this paper is also addressed to the program metrics community, which is less familiar with linear logic and weighted relational models, this section is mainly thought to provide a relatively concrete presentation of (the tropical variant of) this model, studied in more detail in the following sections.
+in Section III are indeed based on relatively unsurprising checks; since this paper is also addressed to the program metrics community, which is less familiar with linear logic and weighted relational models, this section is mainly thought to provide a relatively concrete presentation of (the tropical variant of) this model, studied in more detail in the following sections.
+Therefore, our plan is to adjust Section III by making clearer what results are general (by also replacing the use of "Fact" with "Proposition/Lemma", toghether with the appropriate reference), and that the point of this section is to show how they look like in a tropical setting.
+We would also accordingly adjust the appendix in case it should appear in an eventual final version of the paper.
 
@@
 
-Theorem 29 is (essentially) Theorem 81 from the appendix, but without mention of differential linear categories.
+Theorem 29 is Theorem 81 from the appendix, except for the mention of differential linear categories.
 We chose not to mention the latter in the main text (and thus to limit our discussion of differential operators to the cartesian case) because our results concerning them are either folklore or based on well-known results/techniques.
@@
 
-...
-
-> "I suggest rewriting Section III.D based on the comments made above, and explain that the differential structure comes from the fact that LREL is a differential category."
-
-OK? Connettere alla discussione sopra??
+We did not mean it to be a formal statement, but just the fact that it is known that Taylor approximation and relational semantics are tightly related, so it is natural to look at what happends in that kind of models.
+We agree that the sentence is not clear, so will modify it.
 

### 2023-03-18 rebuttal.txt
Committ pre-cena

--- 2023-03-17 rebuttal.txt
@@
 We will of course take care in correcting all the typos and unclear sentences.
-We thank in particular reviewer 2 for his/her thorough reading, and we will perform the useful modifications he/her suggested (see below).
+Below we anwer to the main points.
 
@@
 
-> p3 (1)
-
-One just takes the semimodule generated by differential terms with coefficients in Q (or another semiring).
-Equivalently, you can see them as formal linear combinations with coefficients in the choosen semiring.
-
 > p5 "setting $t^!(x) := t o_! x$"...
 
-We are tacitly making some standard identifications: one identifies Q^X with the morphisms in QRel_!(T,X)=QRel(!T,X) where T=\emptyset is the terminal object.
-This is possible by mapping x \in Q^X to x' \in QRel(!T,X) defined by x'_{\emptyset,a} := x_a (remark that !T={\emptyset}).
-In this way, given t \in QRel_!(X,Y), one defines the function t^!
-: Q^X --> Q^Y by t(x) := t °_!
-x', which is a morphism T -> Y in QRel_!, seen in the same way as an element of Q^Y.
+We are tacitly using the standard indentification of Q^X with the morphisms in QRel_!(T,X)=QRel(!T,X) where T=\emptyset is the terminal object.
 
@@
 
-It is the function just defined in the previous answer (i.e. in equation (6) at page 5).
+It is the function discussed in the previous answer (i.e. in equation (6) at page 5).
 
@@
 
-Rigirare la frittata su f, f_t ecc.
+We mean that we consider f a power series of shape \sum_n a_n x^n, and for fixed t>0 we take c_n := log_t (a_n), so that f can be written in the shape given in the paper.
+We agree that the subsection is not clear, so we will rephrase it better.
 
@@
 
-In the discussion around Def2 we were informal on what we meant by "database".
-Prop17 is, instead, a rigorous statement, so we stated it with the correct types for f.
 We are supposing to see a database as a function X --> [0,1], which is indeed often the case (for instance, for a relational database -- take X as the key/attributes and [0,1] as the datas).
+We will mention that.
 
@@
 One of the core aspects of Theorem 24 is that on any L-module one can define a canonical metric; the goal of this paragraph is thus to describe in an explicit way the metrics associated with the SMCC structure of LMOD (or, equivalently, the metrics defining the SMCC structure of LCCAT).
-Yet we think that we should have noted X,Y rather as M,N, in order to avoid confusion.
-
+Maybe we should use M,N rather than X,Y in order to avoid confusion.
 Concerning tensors, in the previous page (after Prop. 19) we introduced the tensor product in LMOD and recalled that an element of it is a join of basic tensors x(x)y.
@@
 
-> "Some of the calculations in the appendix regarding the differential story can be referenced away."
+First of all, we thank you for your thorough reading. We will surely adjust our presentation following your useful remarks!
 
-We thank the reviewer for his careful reading of the whole paper (including the long appendix), and for his numerous suggestions concerning results appearing in the recent literature.
-We were not aware of Lemay's "Coderelictions for Free Exponential Modalities" and we found its results very interesting.
-We will adjust our presentation following such remarks.
+> Comment 1, comment 2, comment 7, comment 8.
 
+We were not aware of Lemay's "Coderelictions for Free Exponential Modalities" and we found its results very interesting/.
 As a general remark, our results on LREL and LREL_!
-in Section III are based on relatively unsurprising checks (yet, it is always interesting to know that one can avoid long checks and use a theorem instead!).
-Since this paper is also addressed to the program metrics community, which is less familiar with linear logic and weighted relational models, this section is mainly thought to provide a relatively concrete presentation of (the tropical variant of) this model, studied in more detail in the following sections.
+in Section III are indeed based on relatively unsurprising checks; but since this paper is also addressed to the program metrics community, which is less familiar with linear logic and weighted relational models, this section is mainly thought to provide a relatively concrete presentation of (the tropical variant of) this model, before studying it in more detail in the following sections.
+Therefore, our plan is to adjust Section III by making clearer what results are general (by also replacing the use of "Fact" with "Proposition/Lemma" toghether with the appropriate reference), and making clearer that the point of that section is to show how they look like in a tropical setting.
+We would also accordingly adjust the appendix in case it should appear in an eventual final version of the paper.
 
@@
 
-Theorem 29 is (essentially) Theorem 81 from the appendix, but without mention of differential linear categories.
-We chose not to mention the latter in the main text (and thus to limit our discussion of differential operators to the cartesian case) because our results concerning them are either folklore or based on well-known results/techniques.
+Theorem 81, from the appendix, appears in the main text as Theorem 29, except for the mention of differential linear categories.
+We chose not to mention the latter in the main text because our results concerning them are either folklore or based on well-known results/techniques.
 
@@
 
-...
-
-> "I suggest rewriting Section III.D based on the comments made above, and explain that the differential structure comes from the fact that LREL is a differential category."
-
-OK? Connettere alla discussione sopra??
+We agree that the sentence is not clear, so will modify it.
+We simply meant that since Taylor and relational semantics are tightly related, it is natural to look at what happens in that kind of models.
 
@@
 You are right, the applications in Section 5 are just sketched.
-Our main goal in this paper was to demonstrate that tropical methods, which are widely applied in computer science (but not in the linear logic/lambda-calculus community) can make sense also for studying the quantitative semantics of the lambda-calculus.
+However, one of the main goals of this paper is to demonstrate that tropical methods, which are widely applied in computer science (but not in the linear logic/lambda-calculus community) can make sense also for studying the quantitative semantics of the lambda-calculus.
 Concrete developments of such applications are thus left for future work.
+We will clarify that point.
 
@@
 
-This is indeed a well-known limitations of the semantics of probabilistic programs based on weighed relational models (e.g.
+This is indeed a well-known limitation of the semantics of probabilistic programs based on weighed relational models (e.g.
 probabilistic coherent spaces [33]).
@@
 full abstraction theorems).
-Still, we agree that this limitation should be mentioned.
+Anyway, we agree that this limitation should be mentioned.
 
@@
 
-It is supposed to be defined by those line. We agree that the sentence is not clear, so we will rephrase it.
+It is supposed to be defined by those line. But we agree that it is not clear, so we will rephrase it better.
 
@@
 
-This is true, yet in the literature on tropical mathematics (to which we refer, e.g.
-[59]) these are defined only with positive coefficients.
-We will explicitly mention it, in order to avoid confusion.
+This is true, yet often in the literature on tropical mathematics (to which we refer, e.g. [59]) these are defined only with positive coefficients. We will explicitly mention it.
 
@@
 We think that the main contribution in Section 6 is to show that an abstract correspondence (expressed indeed by Theorem 24) known in the categorical literature can be extended to account for the relevant structure for interpreting the differential lambda-calculus.
-Hence, we see Theorems 27 and 29, rather than Theorem 24 (whose statement and proof can be deduced from the results of [71]) as the main claims of this section.
+Hence, we see Theorems 27 and 29 as the main claims of this section, rather than Theorem 24 (whose statement and proof can be deduced from the results of [71]).
+We will clarify this.

### 2023-03-18 rebuttal.txt
Committ
Cambiato una parola :P

@@
 We think that the main contribution in Section 6 is to show that an abstract correspondence (expressed indeed by Theorem 24) known in the categorical literature can be extended to account for the relevant structure for interpreting the differential lambda-calculus.
-Hence, we see Theorems 27 and 29 as the main claims of this section, rather than Theorem 24 (whose statement and proof can be deduced from the results of [71]).
+Hence, we see Theorems 27 and 29 as the main claims of this section, rather than Theorem 24 (which can be deduced from the results of [71]).
 We will clarify this.

### 2023-03-18 rebuttal.txt
commit

@@
-We thank very much all the reviewers for the careful comments.
-We will of course take care in correcting all the typos and unclear sentences.
-Below we anwer to the main points.
+We thank very much all the reviewers for their careful reading and comments.
+We will take care of their suggestions in correcting typos and unclear sentences.
+Below we reply to a few questions and observations.
 
@@
 
-We are tacitly using the standard indentification of Q^X with the morphisms in QRel_!(T,X)=QRel(!T,X) where T=\emptyset is the terminal object.
+We are tacitly using the standard identification of Q^X with QRel_!(T,X)=QRel(!T,X) where T=\emptyset is the terminal object.
 
@@
 We mean that we consider f a power series of shape \sum_n a_n x^n, and for fixed t>0 we take c_n := log_t (a_n), so that f can be written in the shape given in the paper.
-We agree that the subsection is not clear, so we will rephrase it better.
+We agree that the subsection is not very clear, so we will rephrase it.
 
@@
 
-We are supposing to see a database as a function X --> [0,1], which is indeed often the case (for instance, for a relational database -- take X as the key/attributes and [0,1] as the datas).
-We will mention that.
+We are here supposing that a database can be encoded as a function X --> [0,1], which is indeed often the case (for instance, for a relational database -- take X as the key/attributes and [0,1] as the datas).
+We will make this tacit assumption explicit.
 
@@
 
-First of all, we thank you for your thorough reading. We will surely adjust our presentation following your useful remarks!
+First of all, we thank the reviewer for their thorough reading. We will adjust our presentation following their useful remarks!
 
@@
 
-We were not aware of Lemay's "Coderelictions for Free Exponential Modalities" and we found its results very interesting/.
+We were not aware of Lemay's "Coderelictions for Free Exponential Modalities" and we found its results very interesting.
 As a general remark, our results on LREL and LREL_!
-in Section III are indeed based on relatively unsurprising checks; but since this paper is also addressed to the program metrics community, which is less familiar with linear logic and weighted relational models, this section is mainly thought to provide a relatively concrete presentation of (the tropical variant of) this model, before studying it in more detail in the following sections.
-Therefore, our plan is to adjust Section III by making clearer what results are general (by also replacing the use of "Fact" with "Proposition/Lemma" toghether with the appropriate reference), and making clearer that the point of that section is to show how they look like in a tropical setting.
-We would also accordingly adjust the appendix in case it should appear in an eventual final version of the paper.
+in Section III are mostly based on relatively unsurprising checks; yet, as this paper is also addressed to the program metrics community, which is less familiar with linear logic and weighted relational models, this section is mainly thought to provide a relatively concrete presentation of (the tropical variant of) this model, before studying it in more detail in the following sections.
+We will adjust Section III with a clear distinction between general results about weighted relational models (also replacing the use of "Fact" with "Proposition/Lemma" together with the appropriate reference), and results and observations related to the tropical model.
+For the final version of the paper, our plan is to make a long version of the paper available online.
 
@@
 
-Theorem 81, from the appendix, appears in the main text as Theorem 29, except for the mention of differential linear categories.
+Theorem 81 from the appendix, appears in the main text as Theorem 29, except for the mention of differential linear categories.
 We chose not to mention the latter in the main text because our results concerning them are either folklore or based on well-known results/techniques.
@@
 
-You are right, the applications in Section 5 are just sketched.
-However, one of the main goals of this paper is to demonstrate that tropical methods, which are widely applied in computer science (but not in the linear logic/lambda-calculus community) can make sense also for studying the quantitative semantics of the lambda-calculus.
-Concrete developments of such applications are thus left for future work.
-We will clarify that point.
+As we see it, the main goal of this paper is to demonstrate that tropical methods, which are widely applied in computer science (yet not in the linear logic/lambda-calculus community) do make sense also for studying the lambda-calculus from a quantitative viewpoint, and potentially intersect active research directions (like the study of differential privacy for higher-order languages [64]).
+As we observe in the conclusions, the applications in Section 5 are indeed just sketched, and we left concrete developments for future work.
 
@@
 
-It is supposed to be defined by those line. But we agree that it is not clear, so we will rephrase it better.
+It is supposed to be defined by those lines. But we agree that it is not clear, so we will rephrase it better.
 
@@
 
-This is true, yet often in the literature on tropical mathematics (to which we refer, e.g. [59]) these are defined only with positive coefficients. We will explicitly mention it.
+This is true, yet in the literature on tropical mathematics (to which we refer, e.g.
+[59]) these are generally defined only with positive coefficients.
+In order to avoid confusion, wee will explicitly mention this remark.
 
@@
 
-We think that the main contribution in Section 6 is to show that an abstract correspondence (expressed indeed by Theorem 24) known in the categorical literature can be extended to account for the relevant structure for interpreting the differential lambda-calculus.
-Hence, we see Theorems 27 and 29 as the main claims of this section, rather than Theorem 24 (which can be deduced from the results of [71]).
-We will clarify this.
+In our opinion, the main contribution in Section 6 is not Theorem 24 per se.
+As recalled at the beginning of the section, all results up to Theorem 24 are either known from the literature or folklore.
+Indeed, Theorem 24 follows from much more general results in [71], and our work in this part, which is visible in the appendix, was to provide concrete calculations for the main claims (as the results in the literature are very abstract and cover a more general situation).
+Rather, we think that our main contribution is to show that the abstract correspondence expressed by Theorem 24 can be extended to account for the relevant structure necessary to interpret the differential lambda-calculus, and provides in this way a natural extension of the tropical relational model.
+Hence, we see Theorems 27 and 29 as our main claims of this section (these claims and their proofs are, to our knowledge, new).
+Anyway, we will clarify this.

### 2023-03-18 rebuttal.txt
commit

@@
 
-One of the core aspects of Theorem 24 is that on any L-module one can define a canonical metric; the goal of this paragraph is thus to describe in an explicit way the metrics associated with the SMCC structure of LMOD (or, equivalently, the metrics defining the SMCC structure of LCCAT).
-Maybe we should use M,N rather than X,Y in order to avoid confusion.
-Concerning tensors, in the previous page (after Prop. 19) we introduced the tensor product in LMOD and recalled that an element of it is a join of basic tensors x(x)y.
+One of the core aspects of Theorem 24 is that on any L-module one can define a canonical metric; the goal of this paragraph is to describe explicitly the metrics associated with the SMCC structure of LMOD (equivalently, the SMCC structure of LCCAT).
+We should use M,N rather than X,Y to avoid confusion.
+On tensors: in the previous page (after Prop. 19) we introduced the tensor product in LMOD and recalled that its elementsare joins of basic tensors x(x)y.
 

### 2023-03-18 rebuttal.txt
commit

@@
 As a general remark, our results on LREL and LREL_!
-in Section III are mostly based on relatively unsurprising checks; yet, as this paper is also addressed to the program metrics community, which is less familiar with linear logic and weighted relational models, this section is mainly thought to provide a relatively concrete presentation of (the tropical variant of) this model, before studying it in more detail in the following sections.
+in Section III are mostly based on relatively unsurprising checks; yet, as this paper is also addressed to the program metrics community, which is less familiar with linear logic and weighted relational models, this section is mainly thought to provide a relatively concrete presentation of (the tropical variant of) this model.
 We will adjust Section III with a clear distinction between general results about weighted relational models (also replacing the use of "Fact" with "Proposition/Lemma" together with the appropriate reference), and results and observations related to the tropical model.

### 2023-03-18 rebuttal.txt
commit

@@
 
-> p5 "setting $t^!(x) := t o_! x$"...
+> on $t^!(x) := t o_! x$:
 
@@
 
->  p7 what is $[[M]]^!$?
+> on $[[M]]^!$:
 
@@
 
-> p7 "which can be written as a parameterized power series of the form" what does this mean?
-$f$ is a single function, but $f_t$ is an infinite family of functions.
-does it mean $f$ can be written as $f_t$ for some fixed constant $t$?"
+> on the functions f and f_t:
 
@@
 
-> p9 Prop 17: why is the input of $f$ suddenly $[0, 1]^X$?
+> "Prop 17: why is the input of $f$ suddenly $[0, 1]^X$?"
 
@@
 
-> Theorem 24 to bottom of column: I found this discussion hard to follow.
+> On the discussion after Theorem 24:.
 
@@
 
-> Comment 1, comment 2, comment 7, comment 8.
+> Comments 1, 2, 7, 8.
 
-We were not aware of Lemay's "Coderelictions for Free Exponential Modalities" and we found its results very interesting.
-As a general remark, our results on LREL and LREL_!
-in Section III are mostly based on relatively unsurprising checks; yet, as this paper is also addressed to the program metrics community, which is less familiar with linear logic and weighted relational models, this section is mainly thought to provide a relatively concrete presentation of (the tropical variant of) this model.
-We will adjust Section III with a clear distinction between general results about weighted relational models (also replacing the use of "Fact" with "Proposition/Lemma" together with the appropriate reference), and results and observations related to the tropical model.
-For the final version of the paper, our plan is to make a long version of the paper available online.
+We thank the reviewer for his careful reading of the appendix.
+We were not aware of Lemay's "Coderelictions for Free Exponential Modalities" and we found it very interesting.
+As a general remark, our results on LREL_!
+in Section III are mostly unsurprising checks, yet, as this paper is also addressed to the program metrics community, which is less familiar with linear logic and its models, we think it is useful to provide a concrete presentation of the tropical relational model.
+We will adjust references and replace "Fact" with "Proposition/Lemma" together with the appropriate reference, as in Section IV.
+For the final version of the paper, instead of a long appendix we will make a long version of the paper available online.
 
-> "Theorem 81 is important and should have appeared in the main paper."
+> On Theorem 81.
 
-Theorem 81 from the appendix, appears in the main text as Theorem 29, except for the mention of differential linear categories.
-We chose not to mention the latter in the main text because our results concerning them are either folklore or based on well-known results/techniques.
+The main claim of this theorem appears in the main text as Theorem 29, but without mention of differential linear categories.
+We chose not to introduce the latter in the main text as we only use them, via well-known results, to prove results about non linear categories.
 
-> "In the introduction, you say "the somehow canonical "Taylor semantics for STdLC". Why is that? What is the justification for this statement?"
+> On "the somehow canonical "Taylor semantics for STdLC".
 
-We agree that the sentence is not clear, so will modify it.
+We agree that the sentence is not clear, so we'll modify it.
 We simply meant that since Taylor and relational semantics are tightly related, it is natural to look at what happens in that kind of models.
@@
 
-> p.4 where is this hat operation defined?
+> p.4 on the hat operation
 
@@
 
-> "In complex analysis, Laurent series include terms of negative
-    degrees, while here I think we only consider terms with
-    non-negative degrees."
+> on Laurent series and terms of negative degrees
 
-This is true, yet in the literature on tropical mathematics (to which we refer, e.g.
-[59]) these are generally defined only with positive coefficients.
-In order to avoid confusion, wee will explicitly mention this remark.
+We agree that this may cause confusion, yet in tropical mathematics literature, e.g. [59], these are generally defined only with positive coefficients. We will explicitly indicate this.
 

### 2023-03-19 rebuttal.txt
commit

@@
-We thank very much all the reviewers for their careful reading and comments.
+We thank very much the reviewers for their careful reading and comments.
 We will take care of their suggestions in correcting typos and unclear sentences.
@@
 
-First of all, we thank the reviewer for their thorough reading. We will adjust our presentation following their useful remarks!
-
 > Comments 1, 2, 7, 8.
 
-We thank the reviewer for his careful reading of the appendix.
+We thank the reviewer for their careful reading of the appendix.
 We were not aware of Lemay's "Coderelictions for Free Exponential Modalities" and we found it very interesting.
@@
 in Section III are mostly unsurprising checks, yet, as this paper is also addressed to the program metrics community, which is less familiar with linear logic and its models, we think it is useful to provide a concrete presentation of the tropical relational model.
-We will adjust references and replace "Fact" with "Proposition/Lemma" together with the appropriate reference, as in Section IV.
+We will adjust references and replace "Fact" with "Proposition/Lemma" together with the appropriate reference, as is done in Section VI.
 For the final version of the paper, instead of a long appendix we will make a long version of the paper available online.
@@
 
-We agree that this may cause confusion, yet in tropical mathematics literature, e.g. [59], these are generally defined only with positive coefficients. We will explicitly indicate this.
+We agree that this may cause confusion, yet in the tropical mathematics literature, e.g. [59], these are generally defined only with positive coefficients. We will explicitly indicate this.
 
@@
 As recalled at the beginning of the section, all results up to Theorem 24 are either known from the literature or folklore.
-Indeed, Theorem 24 follows from much more general results in [71], and our work in this part, which is visible in the appendix, was to provide concrete calculations for the main claims (as the results in the literature are very abstract and cover a more general situation).
-Rather, we think that our main contribution is to show that the abstract correspondence expressed by Theorem 24 can be extended to account for the relevant structure necessary to interpret the differential lambda-calculus, and provides in this way a natural extension of the tropical relational model.
-Hence, we see Theorems 27 and 29 as our main claims of this section (these claims and their proofs are, to our knowledge, new).
+Indeed, the statement of Theorem 24 follows from general results in [71], and our work in this part, which is visible in the appendix, was to provide concrete calculations for the case of LMOD (as the results in the literature are very abstract and cover a more general situation).
+Instead, we think that the main novelty in this section is to show that the abstract correspondence expressed by Theorem 24 can be extended to account for the relevant structure necessary to interpret the differential lambda-calculus, and that this yields a "coordinate-free" generalization of the tropical relational model.
+Hence, we see Theorems 27 and 29 as the main technical results of this section (these claims and their proofs are, to our knowledge, new).
 Anyway, we will clarify this.

### 2023-03-20 rebuttal.txt
Commit
Ho modificato tre parole. Ora la submitto.

@@
 
-We agree that the sentence is not clear, so we'll modify it.
+We agree that the sentence is not clear, so we will modify it.
 We simply meant that since Taylor and relational semantics are tightly related, it is natural to look at what happens in that kind of models.
@@
 
-It is supposed to be defined by those lines. But we agree that it is not clear, so we will rephrase it better.
+It is supposed to be defined by those lines. But we agree that it is not clear, so we will rephrase it.
 
@@
 
-We agree that this may cause confusion, yet in the tropical mathematics literature, e.g. [59], these are generally defined only with positive coefficients. We will explicitly indicate this.
+We agree that this may cause confusion, yet in the tropical mathematics literature, e.g. [59], these are often defined only with positive coefficients. We will explicitly indicate this.
 
@@
 In our opinion, the main contribution in Section 6 is not Theorem 24 per se.
-As recalled at the beginning of the section, all results up to Theorem 24 are either known from the literature or folklore.
+As recalled at the beginning of the section, all results up to Theorem 24 can be either derived from the literature or folklore.
 Indeed, the statement of Theorem 24 follows from general results in [71], and our work in this part, which is visible in the appendix, was to provide concrete calculations for the case of LMOD (as the results in the literature are very abstract and cover a more general situation).
-Instead, we think that the main novelty in this section is to show that the abstract correspondence expressed by Theorem 24 can be extended to account for the relevant structure necessary to interpret the differential lambda-calculus, and that this yields a "coordinate-free" generalization of the tropical relational model.
+Instead, we think that the main novelty in this section is to show that the abstract correspondence expressed by Theorem 24 can be actually extended to account for the relevant structure necessary to interpret the differential lambda-calculus, and that this yields a "coordinate-free" generalization of the tropical relational model.
 Hence, we see Theorems 27 and 29 as the main technical results of this section (these claims and their proofs are, to our knowledge, new).

### 2023-05-08 TLLA23/BarbaPisto.tex
Branch per TLLA
Spero di non aver fatto casini con Git

\documentclass[submission,%copyright,creativecommons
]{eptcs}
\providecommand{\event}{TLLA 2023} % Name of the event you are submitting to
%\usepackage{breakurl}             % Not needed if you use pdflatex only.
\usepackage{underscore}           % Only needed if you use pdflatex.

\usepackage{macros}
\usepackage{tikz}
  \usetikzlibrary{calc}
\usepackage{hyperref}
\usepackage{stmaryrd,amssymb,proof,cmll,mathrsfs,amsmath,amsthm}
\usepackage{todonotes}

\title{Tropical methods in Lambda-Calculus}
\author{Davide Barbarossa
\institute{DISI, Universit\`a di Bologna}
\email{davide.barbarossa@unibo.it}
\and
\qquad\qquad Paolo Pistone
\institute{\qquad\qquad\qquad DISI, Universit\`a di Bologna}
\email{\qquad\qquad\qquad paolo.pistone@unibo.it}
}
\def\titlerunning{Trends in Linear Logic and Applications}
\def\authorrunning{D.\ Barbarossa and P.\ Pistone}

\newcommand\eg{\textit{e.g.\ }}
\newcommand\etc{\textit{etc}}

\begin{document}
\maketitle

\begin{abstract}
We propose an interpretation of the $\lambda$-calculus in the framework of tropical mathematics, as a unified framework for both program metrics, based on the analysis of program sensitivity via Lipschitz-conditions, and for resource analysis, based on higher-order program differentiation.
We sketch the relation of this semantics to quantitative properties like differential privacy, convergence logprobabilities and Probabilistic Coherent spaces.
Finally, we study the abstract correspondence between this tropical semantics and Lawvere’s generalised metric spaces.
\end{abstract}

\section*{Motivation}

\section{Tropical Laurent Series}

\section{Relations to quantitative properties}

\section{Lawvere's generalised metric spaces}

%The optional arguments of {\tt $\backslash$documentclass$\{$eptcs$\}$} are
%\begin{itemize}
%\item at most one of
%{\tt adraft},
%{\tt submission} or
%{\tt preliminary},
%\item at most one of {\tt publicdomain} or {\tt copyright},
%\item and optionally {\tt creativecommons},
%  \begin{itemize}
%  \item possibly augmented with
%    \begin{itemize}
%    \item {\tt noderivs}
%    \item or {\tt sharealike},
%    \end{itemize}
%  \item and possibly augmented with {\tt noncommercial}.
% \end{itemize}
%\end{itemize}
%We use {\tt adraft} rather than {\tt draft} so as not to confuse hyperref.
%The style-file option {\tt submission} is for papers that are
%submitted to {\tt $\backslash$event}, where the value of the latter is
%to be filled in in line 2 of the tex-file. Use {\tt preliminary} only
%for papers that are accepted but not yet published. The final version
%of your paper that is to be uploaded at the EPTCS website should have
%none of these style-file options.

%By means of the style-file option
%\href{http://creativecommons.org/licenses/}{creativecommons}
%authors equip their paper with a Creative Commons license that allows
%everyone to copy, distribute, display, and perform their copyrighted
%work and derivative works based upon it, but only if they give credit
%the way you request. By invoking the additional style-file option {\tt
%noderivs} you let others copy, distribute, display, and perform only
%verbatim copies of your work, but not derivative works based upon
%it. Alternatively, the {\tt sharealike} option allows others to
%distribute derivative works only under a license identical to the
%license that governs your work. Finally, you can invoke the option
%{\tt noncommercial} that let others copy, distribute, display, and
%perform your work and derivative works based upon it for
%noncommercial purposes only.

%Authors' (multiple) affiliations and emails use the commands
%{\tt $\backslash$institute} and {\tt $\backslash$email}.
%Both are optional.
%Authors should moreover supply
%{\tt $\backslash$titlerunning} and {\tt $\backslash$authorrunning},
%and in case the copyrightholders are not the authors also
%{\tt $\backslash$copyrightholders}.
%As illustrated above, heuristic solutions may be called for to share
%affiliations. Authors may apply their own creativity here \cite{multipleauthors}.

\section*{Bibliography}

%We request that you use
%\href{http://eptcs.web.cse.unsw.edu.au/eptcs.bst}
%{\tt $\backslash$bibliographystyle$\{$eptcs$\}$}
%\cite{bibliographystylewebpage}, or one of its variants
%\href{http://eptcs.web.cse.unsw.edu.au/eptcsalpha.bst}{eptcsalpha},
%\href{http://eptcs.web.cse.unsw.edu.au/eptcsini.bst}{eptcsini} or
%\href{http://eptcs.web.cse.unsw.edu.au/eptcsalphaini.bst}{eptcsalphaini}
%\cite{bibliographystylewebpage}. Compared to the original {\LaTeX}
%{\tt $\backslash$biblio\-graphystyle$\{$plain$\}$},
%it ignores the field {\tt month}, and uses the extra bibtex fields {\tt eid}, {\tt doi}, {\tt ee} and {\tt url}. The first is for electronic identifiers (typically the number $n$ indicating the $n^{\rm th}$ paper in an issue) of papers in electronic journals that do not use page numbers. The other three are to refer, with life links, to electronic incarnations of the paper.

\nocite{*}
\bibliographystyle{eptcs}
%\bibliography{generic}
\end{document}

### 2023-05-08 Notes/Maslov.tex
Ho creato anche questo branch per le altre note

\documentclass[11pt]{article}
%\providecommand{\event}{TLLA 2022} % Name of the event you are submitting to
\usepackage{breakurl}             % Not needed if you use pdflatex only.
\usepackage{underscore}           % Only needed if you use pdflatex.

%myPackages
\usepackage[T1]{fontenc} 
\usepackage[utf8]{inputenc} 
\usepackage{lmodern} 
\usepackage[english]{babel} 
\usepackage{amssymb}  
\usepackage{stmaryrd} 
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{comment}
\usepackage{ebproof}
\usepackage{tkz-graph}
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary {positioning}
\usepackage{cmll}
\usepackage{mathrsfs}
\usepackage{marvosym}
\usepackage{cancel}
\usepackage{pst-solides3d}
\usepackage{MnSymbol}
\usetikzlibrary{matrix,arrows,decorations.pathmorphing,calc}
%\usepackage[top=1in, bottom=1.25in, left=1in, right=1in]{geometry}

\setlength{\parindent}{0.7em}
\setlength{\parskip}{1.3em}

%\bibliographystyle{alpha}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{Lemma}[theorem]{Lemma}
\newtheorem{Proposition}[theorem]{Proposition}
\newtheorem{Corollary}[theorem]{Corollary}
\newtheorem{Definition}[theorem]{Definition}
\newtheorem{Example}[theorem]{Example}
\newtheorem{Remark}[theorem]{Remark}
\newtheorem{Notation}[theorem]{Notation}
\newtheorem{Terminology}[theorem]{Terminology}

\newcommand\eg{\textit{e.g.\ }}
\newcommand\etc{\textit{etc}}

\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{claim}{Claim}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{assumption}{Assumption}

\newcommand{\iu}{û}
\newcommand{\e}{ê}
\newcommand{\A}{â}
\newcommand{\Xplus}{X^{+}}
\newcommand{\forces}{\ \digamma}
\newcommand{\cons}{\textbf{.}}
\newcommand{\brabra}[1]{\llbracket #1 \rrbracket}
\newcommand{\test}[1]{\Vert #1 \Vert}
\newcommand{\str}[1]{\vert #1 \vert}
\newcommand{\sub}[3]{#1\{#2/#3\}}
\newcommand{\oppose}[3]{#1 \not\varepsilon #2^{+}(#3)}
\newcommand{\prog}[1]{\texttt{\small{\textbf{#1}}}}
\newcommand{\callcc}{\prog{callcc}}
\newcommand{\parts}{\mathcal{P}}
\newcommand{\fparts}{\parts_{f}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\rbeta}{\rightarrow_{\beta}}
\newcommand{\infer}[2]{\dfrac{#2}{#1}}
\newcommand{\interpret}[1]{\llbracket #1\rrbracket}
\newcommand{\Env}[1]{\mathrm{Env}_{#1}}
\newcommand{\restr}[2]{#1_{\big|_{#2}}}
\newcommand{\linc}{c^\bullet}

\usepackage{setspace}

% General Symbols
\newcommand{\nat}{\mathbb{N}} % natural numbers
\newcommand\seq\vec % sequence
\newcommand{\st}{s.t.\ } % Such that
\newcommand{\set}[1]{\{#1\}}
\newcommand{\bnfeq}{::=}
\newcommand\pow[1]{\mathscr{P}(#1)}  % powerset
\newcommand{\void}{\varnothing} % empty set
\newcommand{\fmsets}[1]{\mathcal{M}_f(#1)} % set of finite multisets on #1
\newcommand{\emptymset}{1} % empty multyset
\newcommand\bpf{\begin{proof}}
\newcommand\epf{\end{proof}}
\newcommand{\cT}{\mathcal{T}}
\newcommand{\cH}{\mathcal{H}}
\newcommand\cX{\mathcal{X}}
\newcommand{\finsub}{\subseteq
\newcommand{\cL}{\mathcal{L}}_\mathrm{f}}
% Lambda calculus
\newcommand{\lam}{\ensuremath{\lambda}}
\newcommand{\lamu}{\lambda\mu}
\newcommand{\Lam}{\ensuremath{\Lambda}}
\newcommand{\Lamo}{\Lambda^o}
\newcommand{\lamb}{\ensuremath{\lambda\bot}}
\newcommand{\subst}[2]{\{#2/#1\}} % substitution
\newcommand{\cB}{\mathcal{B}} % B of Bohm tree theory
\newcommand{\msto}[1][]{\twoheadrightarrow_{#1}} % multistep reduction
\newcommand{\holes}[1]{\langle #1\rangle} % context holes
\newcommand{\LamContext}[1]{\Lam\holes{#1}} % Lam<\xi>
\newcommand{\FV}[1]{\mathrm{FV}(#1)} % free variables
\newcommand{\BT}[1]{\mathrm{BT}(#1)} % BT(M)
\newcommand{\App}[1]{\mathcal{A}(#1)} % Approximants(M)
\newcommand{\Appr}{\mathcal{A}pp} % Set of all the approximants
\newcommand{\BTle}{\le_\bot} % <=_\Bot
\newcommand{\BTge}{\ge_\bot}
\newcommand{\comb}[1]{\mathsf{#1}}
\newcommand{\bI}{\comb{I}}
\newcommand{\rel}[1]{\mathtt{#1}}
\newcommand{\Om}{\comb{\Omega}}
\newcommand{\blam}{\boldsymbol{\lambda}}
\newcommand{\sqle}{\sqsubseteq}
\newcommand{\TM}[2][]{\mathcal{M}_{#2}^{#1}}
\newcommand{\fresh}{\mathrel{\#}}
\newcommand{\namedapp}[3]{(#1)_#2 #3}
% Resource calculus
\newcommand{\bag}[1]{{[}#1{]}}
\newcommand{\sums}[1]{\Bool\langle #1 \rangle} % 2<Lam^r>
\newcommand{\Bool}{\mathsf{2}} % ring 0,1
\newcommand{\Lamr}{\Lambda^r} % set of resource terms
\newcommand{\Lamb}{\Lambda^b} % set of bags
\newcommand{\lsubst}[2]{\langle [#2]/#1\rangle} % linear substitution
\newcommand{\Sum}[1]{\mathbb{#1}} % Sum of terms
\newcommand{\Te}[1]{\mathcal{T}(#1)} % (support of) Taylor expansion
\newcommand{\NF}[2][]{\mathrm{NF}_{#1}(#2)} % Normal form
\newcommand{\ResContext}[1]{\Lamr\holes{#1}} % Lam^r<\xi>
\newcommand{\BagResContext}[1]{\Lamb\holes{#1}} % Lam^b<\xi>
\newcommand{\NFTeq}{=_\tau} % "=_\tau", meaning NF(T(M)) = NF(T(M))
\newcommand{\size}[1]{\mid#1\mid} % size of a resource term
\newcommand{\nf}[2][]{\mathrm{nf}_{#1}(#2)}
\newcommand{\NFT}[1]{\mathrm{NF}\mathcal{T}(#1)}
\newcommand{\lin}[1]{#1^\circ}
\newcommand{\nak}[1]{|#1|}
\newcommand{\tor}{\to_r}
\newcommand{\redto}[1][]{\to_{#1}}
\newcommand{\emptybag}{1}
\newcommand{\forg}[1]{\lfloor #1\rfloor}
\newcommand\hole[1]{\llparenthesis\hspace{1pt} #1\hspace{1pt} \rrparenthesis} %context hole
\newcommand\linsub[2]{\langle #2 / #1 \rangle} %t<b/x> (lin sub without writing "[]" fo the bags)
\newcommand\lnamedapp[3]{\langle #1 \rangle_{#2} #3}
\newcommand\bigstep{\rightrightarrows}

%MACROS GIULIO
%!TEX root = ../BarbaM.tex
%%%%%% Lambda
\newcommand{\bnf}{::=}
\newcommand{\Appof}[1]{\mathcal{A}(#1)}
\newcommand{\approf}[1]{P_{#1}}
\newcommand{\bool}{\mathsf{2}}
\newcommand\Mfin[1]{\mathcal{M}_\mathrm{f}(#1)}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\cI}{\mathcal{I}}
\newcommand{\PLL}{\textsf{PLL}}
\newcommand\dg[2]{\mathrm{deg}_{#1}(#2)}
\newcommand{\comp}[1]{#1\!\!\uparrow}

\newcommand{\Pfinstar}[1]{\mathscr{P}_\mathrm{fin}^*(#1)}
\newcommand{\web}[1]{\lvert #1\lvert}
\newcommand{\homology}[2]{\mathcal{H}_{#1}(#2)}
\newcommand{\cs}[1]{\mathrm{cs}(#1)}
\newcommand{\simplices}[1]{\mathcal{S}#1}
\newcommand{\orsimpl}[2]{\overrightarrow{\mathcal S_{#1}} #2}
\newcommand{\barIexp}[2]{\mathscr{I}^{#1}\!#2}
\newcommand{\barI}[1]{\mathscr{I}\!#1}
\newcommand{\barsubd}[1]{\mathbf{bs}(#1)}

\newcommand{\lolli}{\multimap}
\newcommand{\cutelim}{\rightsquigarrow}

\newcommand{\minimum}[2]{\emph{min}\{#1,#2\}}

\newcommand{\alarm}[1]{\color{red}#1\color{black}}

\newcommand{\qRel}[1]{\mathrm{Rel_{#1}}}

\newcommand{\eps}{\epsilon}

\newcommand{\norm}[1]{\lVert #1 \lVert}
\newcommand{\supnorm}[1]{\lVert #1 \lVert_\infty}
\newcommand{\absv}[1]{\mid #1\mid}

\title{Notes on the Maslov dequantisation}
\author{Davide Barbarossa\\
Dipartimento di Informatica: Scienza ed Ingegneria\\
Università di Bologna%\\ Bologna, Italia\thanks{A fine university.}
\\
Bologna, Italia
}
%\def\titlerunning{Maslov dequantisation}
%\def\authorrunning{Davide Barbarossa}
\begin{document}

\maketitle

%\begin{abstract}
%\end{abstract}

Fix $t>0$.
Define $\phi_t:[0,1]\to[0,+\infty]$, $\phi_t(0<x):=-t\ln x$, $\phi_t(0):=+\infty$.
It is bijective with inverse $\phi^{-1}_t:[0,+\infty]\to[0,1]$, $\phi^{-1}_t(0<\alpha<+\infty):=e^{-\alpha/t}$, $\phi^{-1}_t(0):=1$, $\phi^{-1}_t(+\infty):=0$.

Call $\widetilde{\mathbb L}:=([0,1],\tilde+,\cdot)$, where $x\tilde+y:=(x+y)/2$.
This is a commutative non-associative idempotent semiring\footnote{Here by \emph{semiring} I mean a set endowed with an associative multiplication, an associative and commutative addition, with distributivity of the multiplication over addition. By \emph{non-associative semiring} I mean as before but without requiring associative addition. By \emph{idempotent} I mean with $x+x=x$.} with unit $1$ and pseudo-zero\footnote{For a non-associative semiring $R$, a \emph{pseudo-zero} $0$ is an element s.t.\ $x+0=f(x)=0+x$ and $x\cdot0=0=0\cdot x$, where $f$ is a fixed function $R\to R$ not depending on $x$.} $0$.

Call $\mathbb L_t:=([0,+\infty],+^t,\cdot^t)$, where $\alpha+^t\beta:=\phi_t(\phi_t^{-1}(\alpha)\tilde+\phi_t^{-1}(\beta))=-t\ln (e^{-\alpha/t}+e^{-\beta/t})+t\ln 2$,
and $\alpha\cdot^t\beta:=\phi_t(\phi_t^{-1}(\alpha)\phi_t^{-1}(\beta))=\alpha+\beta$.
This is a commutative non-associative idempotent semiring with unit $0$ and pseudo-zero $+\infty$.
$\mathbb L_1$ is usually called the \emph{unit-log-semiring}.

Remark that the bijection $\phi_t:\widetilde{\mathbb L}\to\mathbb L_t$ is an isomorphism of semirings with unit and pseudo-zeros, as $\phi_t(x\tilde+y)=\phi_t(x)+^t\phi_t(y)$ and $\phi_t(x\cdot y)=\phi_t(x)\cdot^t\phi_t(y)$.
Therefore, all the $\mathbb L_t$, for $t>0$, are isomorphic.

Call $\mathbb L:=([0,+\infty],\inf,+)$ the \emph{Lawvere tropical semiring}.
This is a commutative idempotent semiring with unit $0$ and zero $+\infty$.

Since $\mathbb L$ is associative, $\mathbb L_t$ and $\mathbb L$ cannot be isomorphic, for no $t>0$ (this can also be proven, without considering the associativity, by showing that there cannot exist a bijective homomorphism of semiring with units and pseudo-zeros between them).

It can be proven that $\alpha+^t\beta \underset{t\to 0}{\longrightarrow} \min\set{\alpha,\beta}$ (and trivially $\alpha\cdot^t\beta=\alpha+\beta \underset{t\to 0}{\longrightarrow} \alpha+\beta$).
In this sense, one says that $\mathbb L_t \underset{t\to 0}{\longrightarrow} \mathbb L$.
Looking at $\mathbb L$ as this limit is usually referred to as seeing $\mathbb L$ as the \emph{Maslov dequantisation} of $\mathbb{\widetilde L}$ (i.e.\ of the $\mathbb L_t$).

Remark that this limit deforms a \emph{isomorphic family} of commutative \emph{non-associative} idempotent semirings with units and \emph{pseudo-zeros}, to a \emph{non-isomorphic} commutative idempotent semiring with unit and \emph{zero}.

\end{document}

### 2023-05-08 Notes/Tropical.tex
Ho creato anche questo branch per le altre note

\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{TLLA 2022} % Name of the event you are submitting to
\usepackage{breakurl}             % Not needed if you use pdflatex only.
\usepackage{underscore}           % Only needed if you use pdflatex.

%myPackages
\usepackage[T1]{fontenc} 
\usepackage[utf8]{inputenc} 
\usepackage{lmodern} 
\usepackage[english]{babel} 
\usepackage{amssymb}  
\usepackage{stmaryrd} 
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{comment}
\usepackage{ebproof}
\usepackage{tkz-graph}
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary {positioning}
\usepackage{cmll}
\usepackage{mathrsfs}
\usepackage{marvosym}
\usepackage{cancel}
\usepackage{pst-solides3d}
\usepackage{MnSymbol}
\usetikzlibrary{matrix,arrows,decorations.pathmorphing,calc}
\usepackage[top=1in, bottom=1.25in, left=1in, right=1in]{geometry}

\setlength{\parindent}{0.7em}
\setlength{\parskip}{1.3em}

%\bibliographystyle{alpha}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{Lemma}[theorem]{Lemma}
\newtheorem{Proposition}[theorem]{Proposition}
\newtheorem{Corollary}[theorem]{Corollary}
\newtheorem{Definition}[theorem]{Definition}
\newtheorem{Example}[theorem]{Example}
\newtheorem{Remark}[theorem]{Remark}
\newtheorem{Notation}[theorem]{Notation}
\newtheorem{Terminology}[theorem]{Terminology}

\newcommand\eg{\textit{e.g.\ }}
\newcommand\etc{\textit{etc}}

\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{claim}{Claim}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{assumption}{Assumption}

\newcommand{\iu}{û}
\newcommand{\e}{ê}
\newcommand{\A}{â}
\newcommand{\Xplus}{X^{+}}
\newcommand{\forces}{\ \digamma}
\newcommand{\cons}{\textbf{.}}
\newcommand{\brabra}[1]{\llbracket #1 \rrbracket}
\newcommand{\test}[1]{\Vert #1 \Vert}
\newcommand{\str}[1]{\vert #1 \vert}
\newcommand{\sub}[3]{#1\{#2/#3\}}
\newcommand{\oppose}[3]{#1 \not\varepsilon #2^{+}(#3)}
\newcommand{\prog}[1]{\texttt{\small{\textbf{#1}}}}
\newcommand{\callcc}{\prog{callcc}}
\newcommand{\parts}{\mathcal{P}}
\newcommand{\fparts}{\parts_{f}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\rbeta}{\rightarrow_{\beta}}
\newcommand{\infer}[2]{\dfrac{#2}{#1}}
\newcommand{\interpret}[1]{\llbracket #1\rrbracket}
\newcommand{\Env}[1]{\mathrm{Env}_{#1}}
\newcommand{\restr}[2]{#1_{\big|_{#2}}}
\newcommand{\linc}{c^\bullet}

\usepackage{setspace}

% General Symbols
\newcommand{\nat}{\mathbb{N}} % natural numbers
\newcommand\seq\vec % sequence
\newcommand{\st}{s.t.\ } % Such that
\newcommand{\set}[1]{\{#1\}}
\newcommand{\bnfeq}{::=}
\newcommand\pow[1]{\mathscr{P}(#1)}  % powerset
\newcommand{\void}{\varnothing} % empty set
\newcommand{\fmsets}[1]{\mathcal{M}_f(#1)} % set of finite multisets on #1
\newcommand{\emptymset}{1} % empty multyset
\newcommand\bpf{\begin{proof}}
\newcommand\epf{\end{proof}}
\newcommand{\cT}{\mathcal{T}}
\newcommand{\cH}{\mathcal{H}}
\newcommand\cX{\mathcal{X}}
\newcommand{\finsub}{\subseteq
\newcommand{\cL}{\mathcal{L}}_\mathrm{f}}
% Lambda calculus
\newcommand{\lam}{\ensuremath{\lambda}}
\newcommand{\lamu}{\lambda\mu}
\newcommand{\Lam}{\ensuremath{\Lambda}}
\newcommand{\Lamo}{\Lambda^o}
\newcommand{\lamb}{\ensuremath{\lambda\bot}}
\newcommand{\subst}[2]{\{#2/#1\}} % substitution
\newcommand{\cB}{\mathcal{B}} % B of Bohm tree theory
\newcommand{\msto}[1][]{\twoheadrightarrow_{#1}} % multistep reduction
\newcommand{\holes}[1]{\langle #1\rangle} % context holes
\newcommand{\LamContext}[1]{\Lam\holes{#1}} % Lam<\xi>
\newcommand{\FV}[1]{\mathrm{FV}(#1)} % free variables
\newcommand{\BT}[1]{\mathrm{BT}(#1)} % BT(M)
\newcommand{\App}[1]{\mathcal{A}(#1)} % Approximants(M)
\newcommand{\Appr}{\mathcal{A}pp} % Set of all the approximants
\newcommand{\BTle}{\le_\bot} % <=_\Bot
\newcommand{\BTge}{\ge_\bot}
\newcommand{\comb}[1]{\mathsf{#1}}
\newcommand{\bI}{\comb{I}}
\newcommand{\rel}[1]{\mathtt{#1}}
\newcommand{\Om}{\comb{\Omega}}
\newcommand{\blam}{\boldsymbol{\lambda}}
\newcommand{\sqle}{\sqsubseteq}
\newcommand{\TM}[2][]{\mathcal{M}_{#2}^{#1}}
\newcommand{\fresh}{\mathrel{\#}}
\newcommand{\namedapp}[3]{(#1)_#2 #3}
% Resource calculus
\newcommand{\bag}[1]{{[}#1{]}}
\newcommand{\sums}[1]{\Bool\langle #1 \rangle} % 2<Lam^r>
\newcommand{\Bool}{\mathsf{2}} % ring 0,1
\newcommand{\Lamr}{\Lambda^r} % set of resource terms
\newcommand{\Lamb}{\Lambda^b} % set of bags
\newcommand{\lsubst}[2]{\langle [#2]/#1\rangle} % linear substitution
\newcommand{\Sum}[1]{\mathbb{#1}} % Sum of terms
\newcommand{\Te}[1]{\mathcal{T}(#1)} % (support of) Taylor expansion
\newcommand{\NF}[2][]{\mathrm{NF}_{#1}(#2)} % Normal form
\newcommand{\ResContext}[1]{\Lamr\holes{#1}} % Lam^r<\xi>
\newcommand{\BagResContext}[1]{\Lamb\holes{#1}} % Lam^b<\xi>
\newcommand{\NFTeq}{=_\tau} % "=_\tau", meaning NF(T(M)) = NF(T(M))
\newcommand{\size}[1]{\mid#1\mid} % size of a resource term
\newcommand{\nf}[2][]{\mathrm{nf}_{#1}(#2)}
\newcommand{\NFT}[1]{\mathrm{NF}\mathcal{T}(#1)}
\newcommand{\lin}[1]{#1^\circ}
\newcommand{\nak}[1]{|#1|}
\newcommand{\tor}{\to_r}
\newcommand{\redto}[1][]{\to_{#1}}
\newcommand{\emptybag}{1}
\newcommand{\forg}[1]{\lfloor #1\rfloor}
\newcommand\hole[1]{\llparenthesis\hspace{1pt} #1\hspace{1pt} \rrparenthesis} %context hole
\newcommand\linsub[2]{\langle #2 / #1 \rangle} %t<b/x> (lin sub without writing "[]" fo the bags)
\newcommand\lnamedapp[3]{\langle #1 \rangle_{#2} #3}
\newcommand\bigstep{\rightrightarrows}

%MACROS GIULIO
%!TEX root = ../BarbaM.tex
%%%%%% Lambda
\newcommand{\bnf}{::=}
\newcommand{\Appof}[1]{\mathcal{A}(#1)}
\newcommand{\approf}[1]{P_{#1}}
\newcommand{\bool}{\mathsf{2}}
\newcommand\Mfin[1]{\mathcal{M}_\mathrm{f}(#1)}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\cI}{\mathcal{I}}
\newcommand{\PLL}{\textsf{PLL}}
\newcommand\dg[2]{\mathrm{deg}_{#1}(#2)}
\newcommand{\comp}[1]{#1\!\!\uparrow}

\newcommand{\Pfinstar}[1]{\mathscr{P}_\mathrm{fin}^*(#1)}
\newcommand{\web}[1]{\lvert #1\lvert}
\newcommand{\homology}[2]{\mathcal{H}_{#1}(#2)}
\newcommand{\cs}[1]{\mathrm{cs}(#1)}
\newcommand{\simplices}[1]{\mathcal{S}#1}
\newcommand{\orsimpl}[2]{\overrightarrow{\mathcal S_{#1}} #2}
\newcommand{\barIexp}[2]{\mathscr{I}^{#1}\!#2}
\newcommand{\barI}[1]{\mathscr{I}\!#1}
\newcommand{\barsubd}[1]{\mathbf{bs}(#1)}

\newcommand{\lolli}{\multimap}
\newcommand{\cutelim}{\rightsquigarrow}

\newcommand{\minimum}[2]{\emph{min}\{#1,#2\}}

\newcommand{\alarm}[1]{\color{red}#1\color{black}}

\newcommand{\qRel}[1]{\mathrm{Rel_{#1}}}

\newcommand{\eps}{\epsilon}

\newcommand{\norm}[1]{\lVert #1 \lVert}
\newcommand{\supnorm}[1]{\lVert #1 \lVert_\infty}
\newcommand{\absv}[1]{\mid #1\mid}

\title{Denotational semantics driven simplicial homology?}
\author{Davide Barbarossa\thanks{Thanks to Thomas Ehrhard who contributed to this work. Thanks also to Giulio Manzonetto and Lorenzo Tortora de Falco for instructive discussions.}
%\institute{Università di Bologna\\ Bologna, Italia}
\institute{Dipartimento di Informatica: Scienza ed Ingegneria\\
Università di Bologna%\\ Bologna, Italia\thanks{A fine university.}
\\
Bologna, Italia}
\email{davide.barbarossa@unibo.it}
%\and
%Co Author \qquad\qquad Yet S. Else
%\institute{Stanford Univeristy\\
%California, USA}
%\email{\quad is@gmail.com \quad\qquad somebody@else.org}
}
\def\titlerunning{Den.\ sem.\ and homology}
\def\authorrunning{Davide Barbarossa}
\begin{document}

%\maketitle

%\begin{abstract}
%\end{abstract}

\begin{Definition}
Let $R$ be a semiring.
The category $\mathrm{Rel}_R$ has sets as objects, and the homset $\mathrm{Rel}_R(X,Y)$ is the set $R^{X\times Y}$ of matrices with rows indexed by $X$, columns indexed by $Y$ and coefficients in $R$.
The identity $id_X$ is the usual diagonal matrix $id_{a,b}:=\delta_{a,b}$ ($\delta$ is the Kronecker symbol), and the composition of $t:Y\times Z\to R$ with $s:X\times Y\to R$ is the matrix $ts:X\times Z\to R$ given by $(ts)_{a,c}:=\sum\limits_{b\in Y} t_{b,c}s_{a,b}$.
We suppose for the moment that all the desired infinite series converge in $R$.
\end{Definition}

Our composition formula is slightly unconventional.
This is because, usually, a matrix $t$ with raws on $X$ and columns on $Y$ is seen as a function from $R^Y$ to $R^X$ ($R$ being here any set of coefficients);
however, we are using the opposite convention, which corresponds to taking the transpose.
In this way, the matrix $t$ applies to a vector $x\in R^X$ giving rise to a vector $tx\in R^Y$ (instead of the converse), and the formula takes the (slightly unconventional) form $(tx)_{b} := \sum\limits_{a\in X} t_{a,b}x_{a}$.
We use this convention because it simplifies notations, as it allows us to see matrices $t\in\mathrm{Rel}_R(X,Y)$ as functions $\hat{t}:R^X\to R^Y$.

\begin{Remark}
It is straightforward to see that each $R^X$ is a $R$-\emph{module} w.r.t.\ pointwise scalar multiplication and pointwise addition.
\alarm{If the desired infinite series behave well in $R$, then the functions $\hat{t}$ are all and only the linear ones, as in the finite case.}
\end{Remark}

Given $a\in X$, let us denote by $e_a\in Q^X$ the ``canonical base vector'' whose coordinates are $(e_a)_{a}:=1$ and $(e_a)_{a'}:=0$ if $a\neq a'$.

Let us set $\overline{\R}_{\geq 0}:=\R_{\geq 0}\cup\set{+\infty}$.

\begin{Terminology}
Consider the complete lattice $(\overline{\R}_{\geq 0},\geq)$, with joins $\bigvee A$ given by $\inf A$, and meets $\bigwedge A$ by $\sup A$.
Remark that we set $\inf \emptyset := +\infty$.
It becomes a semiring if we endow it with $\inf$ as addition and $+$ as multiplication\footnote{The neutral element for addition being $+\infty$ and the one for multiplication $0$.}.
It is called the \emph{non-negative (min-plus-)tropical (commutative and idempotent) semiring}.
Moreover, it is a (unital, commutative, idempotent) quantale with multiplication the one of its semiring structure (that is, $+$).
It is also a continuous semiring.
Let us call it $Q$.
\end{Terminology}

\begin{Remark}
The $Q$-module structure of $Q^X$ is given by scalar multiplication $(r,x)\in Q\times Q^X\to r+x\in Q^X$ and addition $(x,y)\in Q\times Q^X\to \minimum{x}{y}\in Q^X$.
\alarm{If $t\in\mathrm{Rel}_Q(X,Y)$, than the function $\hat{t}$ is \emph{linear}}.
\end{Remark}

\begin{Definition}
Let us call $\mathrm{Trop}$ the co-Kleisli category $(\mathrm{Rel}_Q)_!$ of $\mathrm{Rel}_Q$ w.r.t.\ the usual multiset comonad $!$.
That is, objects are sets and morphisms from $X$ to $Y$ are matrices $F:\fmsets{X}\times Y\to Q$, seen as maps $\hat{F}:Q^{\fmsets{X}}\to Q^Y$.
Moreover, the identity on $X$ is the function $id:\fmsets{X}\times X\to X$ s.t.\ $id_{[x],x}=0$ and $id_{\mu,x}=+\infty$ if $\mu\neq [x]$, and the composition $GF:\fmsets{X}\times Z\to Q$ of $G:\fmsets{Y}\times Z\to R$ with $F:\fmsets{X}\times Y\to R$ takes the form:
\[
 (GF)_{\mu,c} = \inf\limits_{\rho\in\fmsets{Y}} \left\{G_{\rho,c}+\sum\limits_{b\in Y} \rho_b F_{\mu,b}\right\}.
\]
The induced map $\hat{F}:Q^{\fmsets{X}}\to Q^Y$ by $F\in\mathrm{Trop}(X,Y)$ is given by:
\[
 \hat{F}(x)_b = \inf\limits_{\mu\in\fmsets{X}} \left\{F_{\mu,b}+\mu\cdot x\right\}
\]
where $\mu\cdot x := \sum\limits_{a\in X} \mu_a x_a \in Q$.
We call \emph{tropical analytical} the functions of the shape $\hat{F}$.
\end{Definition}

\begin{Remark}
Since $Q$ is a continuous semiring, [Weighted relational models] assures us that $\mathrm{Rel}_Q$ is:

- linear and continuous category

- $*$-autonomous with biproducts and bilinear and continuous tensor and monoidal currying

- Lafont category.

Furthermore, the same paper tells us that $(\mathrm{Rel}_Q)_!$ is a post-linear and continuous $Q$-CCC.
\end{Remark}

\begin{Example}
The terminal object of $\mathrm{Trop}$ is $1:=\set{*}$.
Using the trivial bijection $\fmsets{\set{*}}= \N$, the morphisms $F\in\mathrm{Trop}(1,1)$ are \emph{exactly} the functions $F:\N\to Q$, and the induced tropical analytical map $\hat F: Q=Q^{\set{*}}\to Q^{\set{*}}=Q$ has shape $\hat{F}(x)=\inf\limits_{n\N}\set{F(n)+nx}$.
A particular case is the one of tropical polynomials on $Q$, which coincide with the tropical analytical functions from $Q$ to itself for which the $\inf$ is always a $\min$.
In particular, all affine functions $\hat F(x)=r+mx$ \emph{with $m\in\N$}, $r\in Q$ from $Q$ to itself (and thus also all affine functions from $\R$ to itself), are of this shape: is suffices to take $F(n):=r$ if $n=m$ and $F(n)=+\infty$ otherwise.
\end{Example}

\begin{Remark}
We can endow $Q^X$ with two (potentially $+\infty$) metrics:
\[d^{\infty}(x,y):=\sup\limits_{a\in X}|x_a-y_a|\]
\[d^1(x,y):=\sum\limits_{a\in X}|x_a-y_a|. \alarm{(Forse \ questa \ la \ si \ dovrebbe \ prendere \ tropicale, \ ovvero \ inf_a ?)}\]
\end{Remark}

Remember the following easy property of infs and sequences.

\begin{Lemma}
If $\mid a_n-b_n\mid \,\leq K$ for all $n \in\N$, then $\mid \inf\limits_n a_n-\inf\limits_n b_n\mid \,\leq K$.
\end{Lemma}
\begin{proof}
We have $\mid \inf\limits_n b_n - \inf\limits_n a_n\mid \leq \mid \inf\limits_n b_n - b_{m}\mid + \mid b_{m} - a_m\mid + \mid a_m - \inf\limits_n a_n\mid\leq \mid \inf\limits_n b_n - b_{m}\mid + K + \mid a_m - \inf\limits_n a_n\mid$  for all $m\in\N$, so $\mid \inf\limits_n b_n - \inf\limits_n a_n\mid \leq \inf\limits_m \set{\mid \inf\limits_n b_n - b_{m}\mid + K + \mid a_m - \inf\limits_n a_n\mid} = K$.
\end{proof}

Let us call $\widetilde{F}_{\mu,b}(x):=F_{\mu,b}+\mu\cdot x$, so that $\hat{F}(x)_b = \inf\limits_{\mu\in\fmsets{X}} \widetilde{F}_{\mu,b}(x)$.

\begin{Remark}
Let $\hat F, \hat G : Q^X\to Q^Y$ tropical analytical, whose associated matrices are $F,G$ respectively.
We have:
\[
 d^{\infty}(\hat F(x),\hat{G}(x)) \leq  d^{\infty}(F,G)
\]
for all $x\in Q^X$.

Indeed, we have:
$|\widetilde{F}_{\mu,b}(x)-\widetilde{G}_{\mu,b}(x)|=|F_{\mu,b}-G_{\mu,b}|\leq d^{\infty}(F,G)$, so by the previous Lemma $|\hat F(x)_b-\hat{G}(x)_b|\leq d^{\infty}(F,G)$ and thus the desired result.
\end{Remark}

\begin{Remark}
The functions $\widetilde{F}_{\mu,b}$ are $\#\mu$-Lipschitz in $Q^X$ (w.r.t.\ both metrics).
Indeed we have the following straightforward inequalities:
\[
 |\widetilde{F}_{\mu,b}(x)-\widetilde{F}_{\mu,b}(y)| =
 | \mu\cdot (x-y) | \leq
 \mu\cdot| x-y | =
 \sum\limits_{a\in X} \mu_a | x_a - y_a | \leq
 d(x,y)\sum\limits_{a\in X} \mu_a = 
 d(x,y)\#\mu.
\]
\end{Remark}

\begin{Remark}
The functions $\widetilde{F}_{\mu,b}$ are always non-decreasing, in the sense that $\widetilde{F}_{\mu,b}(x)\leq \widetilde{F}_{\mu,b}(x+y)$.
This is clear because everything takes place in non-negative real numbers, so $\mu\cdot x\leq \mu\cdot (x+y)$.
Therefore, also $\hat{F}$ is non-decreasing (in the same sense: for all $b\in Y$, we have $\hat F (x)_b\leq \hat F (x+y)_b$).
\end{Remark}

\begin{Remark}
Fix $\epsilon>0$.
Then there is $\mu^\epsilon\in\fmsets{X}$ s.t.\ $|\hat{F}(x)_b-\widetilde{F}_{\mu^\epsilon,b}(x)|\leq\epsilon$ (this is by definition of $\inf$).
Furthermore, if $\hat F(x)_b < +\infty$ then $\sum\limits_{a\in X} \mu^\epsilon_a x_a=\mu^\epsilon\cdot x < +\infty$.
This means that $\mu^\epsilon_a x_a<+\infty$ for all $a\in X$.

Actually, for all $a\in X$, there must exist $K_a\in\R$ s.t.\ $\mu^\epsilon_a x_a \leq K_a$ for all $0<\epsilon<1$ \alarm{(limitare eps serve altrimenti l'argomento sotto non funziona!)}.
In fact, otherwise, there is an $a\in X$ s.t.\ for all $K_a$ there is $0<\epsilon<1$ s.t.\ $\widetilde{F}_{\mu^\epsilon,b}(x)\geq \mu^\epsilon_a x_a> K_a$.
But, taking $K_a$ large enough, we obtain a contradiction with he fact that, by definition of $\mu^\epsilon$, we must have $|\hat{F}(x)_b-\widetilde{F}_{\mu^\epsilon,b}(x)|\leq\epsilon<1$.

Therefore, for all $a\in X$ s.t.\ $x_a\neq 0$, there must still exist $L_a\in\R$ s.t.\ $\mu^\epsilon_a \leq L_a$ for all $0<\epsilon<1$.
\end{Remark}

\begin{Remark}
Let $q\in Q$. For all $0<\epsilon<1$, we have:
\[
 |\widetilde{F}_{\mu^\epsilon,b}(x)-\widetilde{F}_{\mu^\epsilon,b}(x+ qe_a)| =
 | \mu^\epsilon\cdot (-qe_a) |=
 q\mu^\epsilon_a\]
and, if $x_a\neq 0$, the inequality continues as $\leq qL_a$.
\end{Remark}

Putting all this together we get:

\begin{Proposition}
Let $\hat F$ tropical analytical.
Fix $x\in Q^X$ and let $a\in X$ s.t.\ $x_a\neq 0$.
For all $0<\epsilon<1$ we have:
\[
 |\hat{F}(x)_b - \hat{F}(x+ \epsilon e_a)_b| \leq (2+2L_a)\epsilon.
\]
This immediately implies that:
\[
 d^{\infty}(\hat{F}(x) , \hat{F}(x+ \epsilon e_a)) \leq (2+2L_a)\epsilon. \alarm{(ma \ non \ \`e \ vero \ per \ d^1)}
\]
\end{Proposition}
\begin{proof}
We have:
\[
 |\hat{F}(x)_b - \hat{F}(x+\epsilon e_a)_b| \leq
 |\hat{F}(x)_b-\widetilde{F}_{\mu^\epsilon,b}(x)| +
 |\widetilde{F}_{\mu^\epsilon,b}(x)-\widetilde{F}_{\mu^\epsilon,b}(x+\epsilon e_a)| +
 |\widetilde{F}_{\mu^\epsilon,b}(x+\epsilon e_a)-\hat{F}(x+\epsilon e_a)_b|
\]
where $|\hat{F}(x)_b-\widetilde{F}_{\mu^\epsilon,b}(x)|\leq \epsilon$ and $|\widetilde{F}_{\mu^\epsilon,b}(x)-\widetilde{F}_{\mu^\epsilon,b}(x+\epsilon e_a)|\leq \epsilon L_a$.
So if we show that: \[|\widetilde{F}_{\mu^\epsilon,b}(x+\epsilon e_a)-\hat{F}(x+\epsilon e_a)_b|\leq\epsilon + \epsilon L_a\] we are done.
But since $\widetilde{F}_{\mu^\epsilon,b}$ is non-decreasing, and by definition of $\inf$, we have:
\[
 \hat{F}(x)_b \leq \hat{F}(x+\epsilon e_a)_b \leq \widetilde{F}_{\mu^\epsilon,b}(x+\epsilon e_a),
\]
so that:
\[
 |\widetilde{F}_{\mu^\epsilon,b}(x+\epsilon e_a)-\hat{F}(x+\epsilon e_a)_b|\leq
 |\widetilde{F}_{\mu^\epsilon,b}(x+\epsilon e_a) - \hat{F}(x)_b | \leq
 |\widetilde{F}_{\mu^\epsilon,b}(x+\epsilon e_a) - \widetilde{F}_{\mu^\epsilon,b}(x) | + | \widetilde{F}_{\mu^\epsilon,b}(x) - \hat{F}(x)_b | \leq
 \epsilon L_a + \epsilon.
\]
\end{proof}

Remark that, in general, the previous results reads as follows:

Fix $x\in Q^X$, $I\subseteq\N$ an index set and let $a_i\in X$ s.t.\ $x_{a_i}\neq 0$ for all $i\in I$.
For all $0<\epsilon_i<\epsilon<1$ we have:
\[
 |\hat{F}(x)_b - \hat{F}(x+ \sum\limits_{i\in I} \epsilon_i e_{a_i})_b| \leq (2+2\sum\limits_{i\in I} L_{a_i})\epsilon.
\]
This is because we have:
\[
 |\widetilde{F}_{\mu^\epsilon,b}(x)-\widetilde{F}_{\mu^\epsilon,b}(x+ \sum\limits_{i\in I} \epsilon_i e_{a_i})| =
 | \mu^\epsilon\cdot (-\sum\limits_{i\in I}  \epsilon_i e_{a_i}) |=
 \sum\limits_{i\in I} \mu^\epsilon_{a_i}  \epsilon_i
\]
and, if $x_{a_i}\neq 0$ for all $i\in I$, the inequality follows as $\leq \sum\limits_{i\in I} \epsilon_i L_{a_i}\leq  \epsilon \sum\limits_{i\in I} L_{a_i}$.

\begin{Remark}
 From the previous Proposition it also immediately follows that, for $\hat F$ tropical analytical, $x\in Q^X$, $a\in X$ s.t.\ $x_a\neq 0$ and for all $0<\epsilon<1$, we have:
\[
 d^{\infty}(\hat{F}(x) , \widetilde{F}_{\mu^\epsilon}(x+\epsilon e_a)) \leq (3+2L_a)\epsilon,
\]
where we put $ \widetilde{F}_{\rho}(.):= (\widetilde{F}_{\rho,b}(.))_{b\in Y}:Q^X\to Q^X$, which is clearly still $\#\rho$-Lipschitz.
That is, we can approximate $\hat{F}(x)$, with arbitrary precision $\epsilon$, by using the Lipschitz functions $\widetilde{F}_{\mu^\epsilon}(x+\epsilon e_a)$.
\alarm{Significa veramente che la posso approssimare con robe lipschitz \emph{LOCALMENTE} ??}
\end{Remark}

We say that a function $f:\overline{\R}_{\geq 0} \to \overline{\R}_{\geq 0}$ is \emph{tropical analytical with matrix $\hat f : \N \to \overline{\R}_{\geq 0}$} iff $\hat f\in\mathrm{Trop}(1,1)$ and $f$ is the tropical analytical function associated with it, under the identification of $\overline{\R}_{\geq 0}$ and $Q^1$).

Remark that, since $0\cdot +\infty = 0$, we have: $f(+\infty)=\min\set{\hat f(0),+\infty}=\hat f(0) = \widetilde f(0,+\infty)=\min\limits_{n\in\N} \widetilde f(n,+\infty)$.
Here we set $\widetilde f(n,x):=nx+\hat f(n)$.

\begin{Example}
Let us study the tropical analytical function with matrix $\hat f(n)=\frac{1}{2^n}$, that is:
\[
 f(x)=\inf\limits_{n\in\N} \set{nx+ \frac{1}{2^n}},
\]
where $x\in [0,+\infty]$.

We immediately see that $f(+\infty)=1$.
We also immediately see that $f(0)=\inf\limits_{n\in\N} \frac{1}{2^n}=0$.
In particular, remark that $f(0)$ is \emph{not} a $\min$ but only an $\inf$.
In order to see what happens in $(0,+\infty)$, let us treat $n$ as a real variable and, for $0<x<+\infty$, let us compute:
$\frac{\partial}{\partial n}\widetilde f(n,x) = x-\frac{log_e 2}{2^n}\geq 0$ iff $n\geq \log_2 \left(\dfrac{log_e 2}{x}\right)$.
So $\widetilde f(n,x)$ (as a function of $n$) has exactly one minimum and it is reached at $n_x=\log_2 \left(\dfrac{log_e 2}{x}\right)$.
Remember now that we are only interested in $n\geq 0$, and $n_x\geq 0$ iff $x\leq log_e 2$.
Putting all this together we get:
\[f(x)=\begin{cases}
\widetilde f(0,x) = 1 \, \textit{ if } x\in [log_e 2,+\infty) \\
\widetilde f(\widetilde n_x,x) = (\widetilde n_x+\frac{1}{log_e 2})x \, \textit{ if } x\in (0,log_e 2]
\end{cases}\]
where $\widetilde n_x\in\N$ is either $\lceil n_x \rceil$ or $\lfloor n_x \rfloor$.
In particular, remark that thus $f(x)=\min\limits_{n\in\N} \set{nx+ \frac{1}{2^n}}$ on $(0,+\infty)$.
The only point in which $f(x)$ is \emph{not} a $\min$ is thus $x=0$.

We just proved that $f$ is almost always a $\min$, but in order to find such min we have, \`a priori, to consider all $n\in\N$.
But we can prove more: on intervals of shape $x\in[\frac{\log_e(2)}{2^{n+1}},\frac{\log_e(2)}{2^{n}})$ (we could also have chosen $(\frac{\log_e(2)}{2^{n+1}},\frac{\log_e(2)}{2^{n}}]$), $f$ coincides in fact with a unique tropical polynomial $P_n$.
To see this, fix $n\in\N$ and remark that if $\dfrac{\log_e(2)}{2^{n+1}}\leq x<\dfrac{\log_e(2)}{2^{n}}$ then $\underline f (n+1,m)\leq \widetilde f(m,x) < \underline f (n,m)$ for all $m\in\N$, where we set $\underline f (k,m):=\frac{m}{2^k}\log_e(2)+\frac{1}{2^{m}}$.
By considering $m$ as a real parameter and by computing the derivative $\frac{\partial}{\partial m}\underline f(k,m)=(\frac{1}{2^k}-\frac{1}{2^m})\log_e(2)$ we see that, as a function of $m$, $\underline f(k,m)$ has a minimum (at $m=k$).
Now, remark that $\lim\limits_{m\to+\infty} \underline f (k,m) = +\infty$ for all $k\in\N$, and therefore it must exist $N_n\in\N$ s.t. $\underline f (n+1,m) \geq \min\limits_{i\in\N} \underline f(n,i)$ for all $m\geq N_n$.
Putting all this together we have:
\[
 \widetilde f(m,x)\geq \underline f(n+1,m)\geq \min\limits_{i\in\N} \underline f(n,i) > \inf\limits_{i\in\N} \widetilde f(i,x) = f(x)
\]
for all $m\geq N_n$.
The last inequality\footnote{The crucial fact is that this inequality is strict. In fact we already knew, by definition of $f$, that $\widetilde f(m,x)\geq f(x)$.} holds because if a function admitting a minimum is always strictly greater than another function, then the same strict inequality holds for the respective min and inf.
This means that all the $m\geq N_n$ are useless in order to compute $f(x)$, i.e.\
$f(x)=\min\limits_{m\leq N_n} \widetilde f(m,x) =:P_n(x)$, and now this is really a tropical polynomial since the minimum is taken on a \emph{finite} number of indices.

The property that we just proved can be immediately strengthened.
In fact, for all $0<\epsilon<\log_e(2)$, we can consider the unique $n_\epsilon\in\N$ s.t.\ $\epsilon \in [\frac{\log_e(2)}{2^{n_\epsilon+1}},\frac{\log_e(2)}{2^{n_\epsilon}})$, and then we have a \emph{finite} number of intervals $\{[\frac{\log_e(2)}{2^{i+1}},\frac{\log_e(2)}{2^{i}})\}_{i=0,\dots,n_\epsilon}$ partitioning $[\frac{\log_e(2)}{2^{n_\epsilon+1}},\log_e(2))$.
Now, remembering the definition of $P_i$, for all $x\in [\epsilon,\log_e(2))$ we actually have that:
\[
 f(x)=\min\limits_{i=0,\dots,n_\epsilon} P_i(x) = \min\limits_{m\leq N_{n_\epsilon}} \widetilde f(m,x) =:P_\epsilon (x)
\]
where $P_\epsilon$ is still a tropical polynomial, because $N_{n_\epsilon}<+\infty$.
Actually, there is no reason to limit $x$ to be smaller than $\log_e(2)$.
In fact, remembering that $f$ is non-decreasing and that $f(x)=1$ for $\log_e(2)\leq x<+\infty$, with $f(+\infty)=0$, we immediately have that for all $0<\epsilon<+\infty$ and for all $x\in [\epsilon,+\infty]$:
\[
 f(x)=\min \set{1,P_\epsilon (x)} =:\overline P_\epsilon (x)
\]
and $\overline P_\epsilon(.)=\min\limits_{m\leq N_{n_\epsilon}} \set{1,\widetilde f(m,.)}$ is again a tropical polynomial.

Remark that this immediately implies that $f$  is continuous on its domain.
It is continuous on $(0,+\infty]$ because for all $\epsilon>0$, it coincides with a tropical polynomial (which is a continuous function) on $[\epsilon,+\infty]$.
And, since we know that $f$ is non-decreasing, in order to show that $\lim\limits_{x\rightarrow 0^+} f(x) = f(0) = 0$, it sufficies to show that for all $K>0$ there is $x>0$ s.t.$f(x)<K$.
This is true because we can certainly find $m,n\in\N$ s.t.\ $\underline f (n,m)=\frac{m}{2^n}\log_e(2)+\frac{1}{2^{m}}<K$, and then we have $f(x)\leq \widetilde f(m,x) < \underline f (n,m)<K$ for $x\in [\dfrac{\log_e(2)}{2^{n+1}},\dfrac{\log_e(2)}{2^{n}})$.

As a last remark, let us point out that we used that fact that $\lim\limits_{m\to+\infty} \underline f (k,m) = +\infty$ in order to show that an appropriate $N_n$ always exists.
But we can actually give a particular $N_n$ doing the job.
In fact we already said that, as a function of $m$, $\underline f(k,m)$ has a minimum at $m=k$, and its value for $k=n$ is $\frac{1+n\log_e(2)}{2^n}$.
Now a perfectly fine $N_n$ is given by any solution $m$ of the inequality:
$\underline f (n+1,m) \geq \min\limits_{i\in\N} \underline f(n,i)$, that is,
$\frac{m}{2}\log_e(2)\geq 1+n\log_e(2)-\frac{2^n}{2^m}$.
A plot in WolframAlpha (or do it by studying the function) shows that a possible solution, for $n\geq 4$, is to take $m=\frac{2n}{log_e(2)}$.
Since we require $m$ to be natural, we can take the ceiling of that value, which turns out to be lower bounded by $3n=:N_n$.
For $n=0,1,2,3$ we can find a particular solution by hand.
For $n=0$ the inequality becomes
$m\geq \frac{2}{\log_e(2)}-\frac{2\cdot 2^m}{\log_e(2)}$.
Now if $m\geq 1-\log_2(\log_e(2))$, i.e. $\frac{2\cdot 2^m}{\log_e(2)}\leq 1$, the desired inequality is satisfied if we take $m\geq \frac{2}{\log_e(2)}$.
Therefore we are done taking $m\geq \max\set{1-\log_2(\log_e(2)),\frac{2}{\log_e(2}}=\frac{2}{\log_e(2)}$, and we can take the ceiling in order to have a natural value, which turns out to be $3=:N_0$.
Analogously one can see that we can take $N_1=5$, $N_2=7$, $N_3=9$.
\end{Example}

The fact that approaching $0$ arbitrarily close (from the right) we can always find a tropical polynomial (which depends on how much we approached $0$) coinciding with $f$ from that point on, is not specific to the previous example.
As the following proposition shows, this is always the case.
In particular, remark that the point $x=0$ is always the only one in which $f(x)$ can fail to be a $\min$.

\begin{Proposition}
Let $f:\overline{\R}_{\geq 0} \to \overline{\R}_{\geq 0}$ tropical analytical, with matrix $\hat f : \N \to \overline{\R}_{\geq 0}$.
Then, for all $0<\epsilon<+\infty$, there is $\mathcal{F}_\epsilon\subseteq\N$ s.t.\
\begin{enumerate}
 \item $\mathcal{F}_\epsilon$ finite
 \item If $\mathcal{F}_\epsilon= \emptyset$ then $f(x) = +\infty$ for all $x\in \overline{\R}_{\geq 0}$
 \item If $f(x_0) = +\infty$ for some $x_0<+\infty$ then $\mathcal{F}_\epsilon= \emptyset$
 \item For all $x\in [\epsilon,+\infty]$ we have $f(x)=P_\epsilon(x)$, where $P_\epsilon$ is the \emph{tropical polynomial}:
      \[
        P_\epsilon(x)= \min\limits_{n\in\mathcal{F}_\epsilon} \set{\hat f (n) + nx}.
      \]
\end{enumerate}
\end{Proposition}
\begin{proof}
Let us set $\mathcal F_\epsilon$ to be the complementary in $\N$ of the set:
\[
 \set{n\in\N \mid \textit{either } \hat f (n)=+\infty \textit{ or there is } m<n\textit{ s.t.\ } \hat f(m)\leq\hat f(n)+\epsilon}.
\]
In other words, $n\in\mathcal F_\epsilon$ iff $\hat f(n)<+\infty$ and for all $m<n$, one has $\hat f(m)>\hat f(n)+\epsilon$.

1).
Suppose that $\mathcal F_\epsilon$ is infinite, and let us enumerate its elements as:
\[\mathcal F_\epsilon =:\set{m_0<m_1<\cdots}.\]
By definition of $\mathcal F_\epsilon$ we have then:
\[+\infty>\hat f(m_0)>\hat f(m_1)+\epsilon>\hat f(m_2)+2\epsilon>\cdots\]
so that $+\infty>\hat f(m_0)>\hat f(m_{i})+i\epsilon\geq i\epsilon$ for all $i\in\N$.
This contradicts the Archimedean property of $\R$.

2).
We show that if $\mathcal F_\epsilon=\emptyset$, then $\hat f(n)=+\infty$ for all $n\in\N$.
This immediately entails the desired result.
We go by strong induction on $n\in\N$:

- if $n=0\notin\mathcal F_\epsilon$, then $\hat f(n)=+\infty$, because there is no $m<n$.

- if $n\geq 1\notin\mathcal F_\epsilon$, then either $\hat f(n)=+\infty$ and we are done, or there is $m<n$ s.t.\ $\hat f(m)\leq \hat f(n)+\epsilon$.
By strong induction $\hat f(m)=+\infty$ and, since $\epsilon<+\infty$, this entails $\hat f(n)=+\infty$.

3).
If $f(x_0)=+\infty$ with $x_0<+\infty$, then necessary $\hat f(n)=+\infty$ for all $n\in\N$.
Therefore, no $n\in\N$ belongs to $\mathcal F_\epsilon$.

4).
By 1), it suffices to show that we can compute $f(x)$ by taking the $\inf$, that is therefore a $\min$, only in $\mathcal F_\epsilon$ (instead of all $\N$).
If $\mathcal F_\epsilon=\emptyset$ then by 2) we are done (remember that $\min\emptyset := +\infty$).
If $\mathcal F_\epsilon\neq\emptyset$, we show that for all $n\in\N-\mathcal F_\epsilon$, there is $m\in\mathcal F_\epsilon$ s.t.\ $\hat f(m)+mx \leq \hat f(n)+nx$.
We do it by induction on $n\in\N$:

- if $n=0$, then by definition of $\mathcal F_\epsilon$, we have $\hat f(n)=+\infty$ (because there is no $n'<n$).
So any element of $\mathcal F_\epsilon\neq\emptyset$ works.

- if $n\geq 1$, then we have two cases:
either $\hat f(n)=+\infty$, in which case we are done as before by taking any element of $\mathcal F_\epsilon\neq\emptyset$.
Or $\hat f(n)<+\infty$, in which case (again by definition of $\mathcal F_\epsilon$) there is $n'<n$ s.t.\ \begin{equation}\label{eq:n'neps} \hat f(n')\leq \hat f(n)+\epsilon.\end{equation}
Therefore we have (remark that the following inequalities hold also for the case $x=+\infty$):
\[\begin{array}{rclr}
 \hat f(n')+n'x & \leq & \hat f(n) + \epsilon + n'x & \textit{by \eqref{eq:n'neps}} \\
 & \leq & \hat f(n) + (n-n')x + n'x & \textit{because $\epsilon\leq x$ and $n\geq n'$} \\
 & = & \hat f(n)+ nx. &
\end{array}\]
Now, if $n'\in\mathcal F_\epsilon$ we are done.
Otherwise $n'\notin\mathcal F_\epsilon$ and we can apply the induction hypothesis on it, obtaining an $m\in\mathcal F_\epsilon$ s.t.\ $\hat f(m)+mx \leq \hat f(n')+n'x$.
Therefore this $m$ works.
\end{proof}

\alarm{Che fare quando y non \`e della forma x+qe_a ??

Continuit\`a a tratti ??

Derivate ??

Diff lambda cat ??
}

Remember that on the $\overline{\R}_{\geq 0}$-module $\overline{\R}_{\geq 0}^X$ (we mean $\overline\R_{\geq 0}$ with the usual $+$ and $\cdot$ operations, which is also called the \emph{probability semiring}), we can define the subtraction $(x_a)_{a\in X}-(y_a)_{a\in X}:=(x_a-y_a)_{a\in X}$ if $x_a-y_a\geq 0$ and $:=0$ otherwise.

\begin{Proposition}
Let $f:\overline{\R}_{\geq 0}^X \to \overline\R_{\geq 0}$ and $x\in \overline{\R}_{> 0}^X$ (so $x_a\neq 0$ for all $a\in X$).
If there is a convex \alarm{open } $V\subseteq \overline{\R}_{\geq 0}^X$ with $x\in V$ and s.t.\ $f_{\big\mid V}$ is concave and lower bounded by a constant $K\geq 0$, then $f$ is continuous at $x$.
\end{Proposition}
\begin{proof}
Fix now and for all the proof an $\eps\in(0,1)$.

If $f(x)=+\infty$ then for all $y\in V$ we have:
$f((1-\eps)x+\eps y)\geq (1-\eps)f(x)+\eps f(y)=+\infty$,
so $f=+\infty$ on an open neighbourhood of $x$ (actually, on all $V$), and in particular it is continuous at $x\in W$.

Suppose now $f(x)<+\infty$.
Since $x_a\neq 0$ for all $a\in X$, there certainly is a convex open neighbourhood $W\subseteq V$ of $x$ s.t.\ $W\subseteq\overline{\R}_{> 0}$ and $W$ is symmetrical w.r.t.\ $x$, meaning that if $w\in W$ then $2x-w\in W$.
Said differently, $W$ is of shape $x+ \hat W$, for some open $\hat W\subseteq \overline{\R}^X$ s.t.\ if $y\in \hat W$ then $-y\in\hat W$ (that is, symmetrical w.r.t.\ $0$).
Now fix $y\in \hat W$, that is, s.t.\ $x+y\in W$.

Since $W$ is convex, $x+\eps y=(1-\eps)x+\eps (x+y)\in W$, and therefore we have:
\[f(x+\eps y)\geq (1-\eps)f(x)+\eps f(x+y)\geq f(x)-\eps f(x)+\eps K=f(x)+\eps(K-f(x))\]
so:
\begin{equation}\label{eq:magg1}
\eps(f(x)-K)\geq f(x)-f(x+\eps y).
\end{equation}
%Now remark that if $x_a-y_a\neq 0$ for all $a\in X$, then also $x_a-\epsilon y_a\neq 0$ for all $a\in X$ and for all $\epsilon\in(0,1]$. Indeed, if $x_a=\epsilon y_a$ for some $a\in X,\epsilon\in(0,1]$, then $0\geq (\epsilon-1)y_a=x_a-\epsilon y_a +\epsilon y_a - y_a=x_a-y_a>0$, contradiction.

Now remark that $x=\dfrac{1}{1+\eps}(x+\eps y)+\dfrac{\eps}{1+\eps}(x-y)$ where $\dfrac{1}{1+\eps}=1-\dfrac{\eps}{1+\eps}$, $\dfrac{\eps}{1+\eps}\in(0,1)$ and $x-y=2x-(x+y)\in W$.
Therefore we have:
\[f(x)\geq \dfrac{1}{1+\eps}f(x+\eps y)+\dfrac{\eps}{1+\eps} f(x-y)\geq \dfrac{1}{1+\eps}(f(x+\eps y)+\eps K)\]
so:
\begin{equation}\label{eq:magg2}
f(x)- f(x+\eps y)\geq-\eps (f(x)-K).
\end{equation}
Putting \ref{eq:magg1} and \ref{eq:magg2} together and remembering that $y$ was arbitrary in $\hat W$, we have shown that for all points $w\in x+\eps \hat W\subset W$, $\mid f(x)-f(w)\mid\leq \eps(f(x)-K)$ (and $f(x)-K\geq 0$).
Since $\eps$ was arbitrary in $(0,1)$, we have shown that $\exists\lim\limits_{w\to x} f(w)=f(x)$.
\end{proof}

\newpage

\begin{Definition}
 A \emph{semifield} is a semiring $R$ s.t.\ $(R-\set{0},\cdot)$ is a group.
 A semifield is commutative iff it is commutative as a semiring.
 Remark that if the semiring $R$ is actually a ring (that is, if $(R,+)$ is a group), then the semifield is a division ring.
 \end{Definition}

 $\overline\R_{\geq 0}$ with the usual $+$ and $\cdot$ operations is a semiring, also called the \emph{probability semiring}.
 It is a semifield under the usual inverse operation.

\begin{Definition}
 A \emph{topological field} is the data of a field $R$ together with a topology on it s.t.\ $+:R\times R\to R$, $-(.):R\to R$, $\cdot:R\times R\to R$ and $(.)^{-1}:R\to R$ are continuous (where $R\times R$ is endowed with the product topology).
 Similarly for the definitions of topological division ring, semifield, ring and semiring.
\end{Definition}

\begin{Definition}
 Let $R$ be a topological semifield.
 A \emph{topological $R$-module} is the data of an $R$-module $A$ together with a topology on it s.t.\ $+:A\times A\to A$ and $\cdot:R\times A\to A$ are continuous (where $R$ is considered with its topology of topological semifield and the product spaces are endowed with the product topologies).
 As a particular case, we have the definition of topological $R$-module for $R$ a semiring.
 If $R$ is a field, a topological $R$-module is called a \emph{topological $R$-vector space}, and as a particular case we have the definition of topological module on a ring.
\end{Definition}

\begin{Remark}
 If $A$ is a topological module on a semiring $R$ then the translation maps $T_x:A\to A$, $z\mapsto z+x$, are continuous for all $x\in A$.
 Remark that the set $\set{T_x\mid x\in A}$ forms a monoid under composition.
 If $R$ is a ring, then the translation maps are homeomorphisms and they form a group under composition.
\end{Remark}

\begin{Definition} 
 Let $\Phi$ be a semigroup (under composition) of continuous maps from a topological module $A$ on a semiring $R$ to itself. The topology $\tau$ of $A$ is \emph{weakly invariant\footnote{Non penso che esista come terminologia nella letteratura, la uso qui giusto per capirci.} under $\Phi$} iff $\phi O$ is open for all $O$ open and $\phi\in\Phi$.
 We can write this as $\Phi\tau \subseteq \tau$.
 It is \emph{invariant under $\Phi$} iff it is weakly invariant and every open is of shape $\phi O$ for some open $O$ and $\phi\in\Phi$.
 We can write this as $\Phi\tau = \tau$.
\end{Definition}

\begin{Remark}
 The topology of a topological module $A$ on a semiring $R$ is weakly translation invariant.
 This means that for all $x\in A$ and open $O$, the set $\set{z+x\mid z\in O}=:O+x$ is open.
 If $R$ is a ring, then the topology is translation invariant (that is, $\tau+x= \tau$ for all $x\in A$).
\end{Remark}

\begin{Definition}
 A \emph{normed semiring} is the data of a semiring $R$ together with a \emph{norm} on it, that is, a function $\absv{.}\,:R\to\R$ satisfying the usual axioms of the absolute value function\footnote{I.e.: $\absv{x}\geq 0$, $\absv x =0 \textit{ iff } x=0$, $\absv{xy}=\absv x\absv y$ and $\absv{x+y}\leq \absv x+\absv y$.}.
 One can immediately check that if $R$ is a ring, then any norm on it induces a metric $d(r,s):=\mid r-s\mid$ on it (and thus also a topology).
 A \emph{topological normed ring} is a normed ring which is a topological ring w.r.t.\ the topology induced by the norm distance.
\end{Definition}

\begin{Definition}
 Let $R$ be normed semiring, let $A$ be a topological $R$-module and let $C\subseteq A$.
 We say that $C$ is \emph{convex} iff for all $x,y\in C$ and for all $r,s\in R$ s.t. $\mid r\mid +\mid s\mid =1$, we have $rx+sy\in C$.
 We say that $C$ is \emph{balanced} iff for all $x\in C$ and for all $r,s\in R$ s.t. $\mid r\mid +\mid s\mid =1$, we have $rx,sx\in C$.
 We say that $C$ is a \emph{disk at $x\in A$} iff it is of shape $C=x+D$ for some $D$ both convex and balanced.
\end{Definition}

\begin{Example}
 An example of a balanced but not convex set in (the normed field) $\R^2$.
 \[\begin{tikzpicture}
 \draw[->] (0,-1) -- (0,1);
 \draw[->] (-1,0) -- (1,0);
 \draw[fill=gray, fill opacity=0.3] (0,0) -- (0.5,-0.5) -- (0.5,0.5) -- (0,0) -- (-0.5,0.5) -- (-0.5,-0.5) -- cycle;
\end{tikzpicture}\]
\end{Example}

\begin{Definition}
 Let $R$ be a normed field and let $\mathbb{V}$ be a topological $R$-vector space. We say that $\mathbb{V}$ is \emph{locally convex} iff every $x\in \mathbb{V}$ admits a local basis consisting of disks at $x$.
 Since the topology of $\mathbb{V}$ is necessarily translation invariant,
 it is equivalent to only ask that $0$ has such a local basis.
\end{Definition}

\begin{Proposition}\label{prop:mainLCTVS}
 Let $\mathbb{V}$ be a locally convex topological $\R$-vector space (where $\R$ is endowed with its usual absolute value and the induced topology by it).
 Fix $x\in\mathbb{V}$, a neighbourhood $V\subseteq\mathbb{V}$ of $x$ and $f:V\to \R\cup\set{+\infty,-\infty}$.
 If $V$ is convex, $f$ is concave and lower bounded by a finite constant $K$, then $f$ is continuous at $x$ (w.r.t.\ the subspace topology on $V$).
\end{Proposition}
\begin{proof}
 Since $V$ is a neighbourhood of $x$, $x$ admits an open neighbourhood $U\subseteq V$, and by local convexity of $\mathbb{V}$ there is a (non necessary open) disk $D\subseteq U$ at $x$.
 Now fix $\eps\in(0,1)$ and $r,s\in \R$ with $0\leq r\leq\eps$, $s=1-r$.
 Fix also $w\in D$.
 By the convexity of $D$ we have $sx+rw\in D$, and thus the convexity of $f$ entails that:
 \[
  f(sx+rw)\geq s f(x) + r f(w) \geq s f(x) + r K = (1-r)f(x) + r K
 \]
that is,
 \[
  f(x)-f(sx+rw)\leq r (f(x)-K).
 \]
 Now remark that $x=\dfrac{1}{s+2r}(sx+rw)+\dfrac{r}{s+2r}(2x-w)$, where $\dfrac{1}{s+2r}+\dfrac{r}{s+2r}=1$, $\dfrac{1}{s+2r}<1$ and $2x-w\in D$ (because $D$ is a disk at $x$).
 Therefore by the convexity of $f$ we have:
 \[
  f(x)\geq \dfrac{1}{s+2r}f(sx+rw)+\dfrac{r}{s+2r}f(2x-w)\geq \dfrac{1}{s+2r}(f(sx+rw)+rK)
 \]
 that is,
 \[
  f(sx+rw)-f(x)\leq r(f(x)-K).
 \]
 We have thus shown that $\mid f(sx+rw)-f(x)\mid\leq r(f(x)-K)$ for all $w\in D$.
 Since this holds also for all $\eps\in(0,1)$, and due to the choice of $r,s$, the points of shape $sx+rw$ span $D$ when $w$ spans $D$ and $\eps$ spans $(0,1)$.
 That is, we have shown that $\mid f(w)-f(x)\mid\leq r(f(x)-K)$ for all $w\in W$.
 Since $f(x)-K\geq 0$ and $r\leq \eps$, we have that $\exists\lim\limits_{w\to x} f(w)=f(x)$.
\end{proof}

\begin{Corollary}\label{cor:cont}
 Let $f:({\R}_{> 0}^X,\supnorm{.}) \to (\overline\R_{\geq 0},\absv .)$ and $x\in {\R}_{> 0}^X$.
 If there is a convex neighbourhood $V\subseteq \overline{\R}_{> 0}^X$ of $x$ s.t.\ $f_{\big\mid V}$ is concave, then $f$ is continuous at $x$.
\end{Corollary}
\begin{proof}
 The $\R$-vector space $\R^X$ is topological w.r.t.\ the topology $\tau_\infty$ induced on it by the norm $\lVert .\lVert_\infty$, and it is clearly locally convex.
 Call $\tau^+_\infty$ the topology induced by $\supnorm{.}$ on ${\R}_{> 0}^X$.
 It is clear that it coincides with $\tau^+_\infty$.
 Since moreover ${\R}_{> 0}^X$ is open in $(\R^X,\tau_\infty)$, the neighbourhood $V$ of $x$ in $({\R}_{> 0}^X,\supnorm{.})$ is also a neighbourhood of $x$ in $(\R^X,\tau_\infty)$.
 We can therefore apply Proposition~\ref{prop:mainLCTVS} to $f_{\big\mid V}$ (which is lower bounded by $0$ by definition of $f$) and obtain that $f_{\big\mid V}$ is continuous at $x$ w.r.t.\ the subspace topology $\tau_V$ induced by $\tau_\infty$ on $V$.
 But since $V$ is contained in ${\R}_{> 0}^X$, the topology $\tau_V$ coincides with the subspace topology induced on $V$ by $\tau^+_\infty$.
 So $f$ is continuous at $x$ w.r.t.\ $\tau^+_\infty$.
\end{proof}

One may wonder if the same proof as above makes it possible to state the previous corollary replacing ${\R}_{> 0}^X$ with $\R_{\geq 0}^X$ (so, in particular, taking $x\in\R_{\geq 0}^X$).
This is not possible because in the proof we crucially use that $\R_{> 0}^X$ is open in $(\R^X,\tau_\infty)$, which is not the case of $\R_{\geq 0}^X$.
In fact, this allowed us to say that $V$ is a neighbourhood of $x$ in $(\R^X,\tau_\infty)$, and therefore to be able to apply Proposition~\ref{prop:mainLCTVS}.
Taking $\R_{\geq 0}^X$ instead, this is in general not true: we could have a neighbourhood of $x$ w.r.t.\ the subspace topology on $\R_{\geq 0}^X$ induced by $\tau_\infty$, which does not contain any open neighbourhood of $x$ w.r.t.\ $\tau_\infty$ (i.e. it is not a neighbourhood w.r.t.\ $\tau_\infty$).

\begin{Example}
 An example of a neighbourhood of $x$ w.r.t.\ the subspace topology on $\R_{\geq 0}^X$ induced by $\tau_\infty$, which is not a neighbourhood w.r.t.\ $\tau_\infty$.
 \[\begin{tikzpicture}
 \draw[->] (0,-0.5) -- (0,1.5);
 \draw[->] (-0.5,0) -- (1.5,0);
 \filldraw[black] (0.8,0) circle (0.7pt) node[anchor=north]{x};
 \fill[fill=gray, fill opacity=0.3] (0.4,0) -- (1.2,0) -- (1.2,0.7) -- (0.4,0.7) -- (0.4,0) -- cycle;
\end{tikzpicture}\]
\end{Example}

\begin{Corollary}\label{cor:tropCont}
 Tropical functions $f:({\R}_{\geq 0}^X,\supnorm .) \to (\overline\R_{\geq 0},\absv .)$ are continuous on ${\R}_{> 0}^X$.
\end{Corollary}
\begin{proof}
 We know that tropical functions are concave on all their domain.
 Therefore the result immediate follows by Corollary \ref{cor:cont}, since the topology induced by $\supnorm{.}$ on ${\R}_{> 0}^X$ coincides with the subspace topology induced by $({\R}_{\geq 0}^X,\supnorm .)$ on it.
\end{proof}

Due to the previous discussion about the impossibility of stating Corollary \ref{cor:cont} in the case where one of the coordinates of $x$ is $0$, the continuity of tropical functions on the hyperplanes $\mathcal H_a:=\set{x\in {\R}_{\geq 0}^X \mid x_a=0}$, for $a\in X$, must be treated separately.
Similarly, the continuity at points with infinite coordinates must be treated separately as well.

Discussione sui coni ? (Vedi altra pagina)

\newpage

\begin{Definition}
 An \emph{$\overline{\R}_{\geq 0}$-cone} is a commutative $\overline{\R}_{\geq 0}$-semimodule with cancellative addition\footnote{I.e.: $x+y=x+y' \Rightarrow y=y'$.}.
\end{Definition}

In [Selinger] cones are required to also have ``strict addition'', meaning that $x+y=0 \Rightarrow x=y=0$.
We do not add this requirement since it will automatic hold when considering normed cones.

\begin{Remark}
 The addition of a cone $P$ (which forms a commutative monoid) turns $P$ into a poset by setting:
 \[
  x \leq y \textit{ iff } y=x+z \textit{, for some }z\in P.
 \]
 By the cancellative property, when such $z$ exists it is unique, and we denote it by $y-x$.
 Such order is called the \emph{cone-order} on $P$.
\end{Remark}

\begin{Definition}
 A \emph{normed $\overline{\R}_{\geq 0}$-cone} is the data of a $\overline{\R}_{\geq 0}$-cone together with a $\leq$-monotone\footnote{I.e.: $x\leq y \Rightarrow \norm{x}\leq \norm y$. Remark that requiring this property (for all $x,y$) is equivalent to requiring that $\norm{x}\leq \norm{x+y}$ for all $x,y$.} norm\footnote{A \emph{norm} on a $\overline{\R}_{\geq 0}$-abstract cone $P$ is a map $\norm{.}:P\to \overline{\R}$ satisfying the usual axioms of norms:
 $\norm x \geq 0$, $\norm x = 0 \Rightarrow x=0$, $\norm{rx}=r\norm x$ and $\norm{x+y}\leq \norm x + \norm y$.} on it.
\end{Definition}

In [Ehrhard-Pagani-Tasson, Crubillie] a normed $\overline{\R}_{\geq 0}$-cone is simply called a cone.

Remark that in a normed $\overline{\R}_{\geq 0}$-cone, by monotonicity of the norm, we have: $\norm{x+y}=0 \Rightarrow x=y=0$.
Therefore, as already mentioned, in a normed cone we have: 
$x+y=0 \Rightarrow \norm{x+y}=0 \Rightarrow x=y=0$, that is, addition is strict.

\begin{Example}
 $\overline{\R}_{\geq 0}^X$ is a normed cone with the norm $\supnorm{x}:=\sup\limits_{a\in X} x_a\in \overline{\R}_{\geq 0}$.
\end{Example}

\begin{Notation}
 We call $\overline{\mathscr{B}}^{\norm{.}}_1(P)$ the closed unit ball $\set{x\in\overline{\R}_{\geq 0}\mid \norm{x}\leq 1}$ of a normed $\overline{\R}_{\geq 0}$-cone $P$.
 We will also simply call it $\overline{\mathscr B}$ when the other informations are clear from the context.
 Similarly, $\mathscr B$ denotes the open unit ball.
\end{Notation}

\begin{Remark}
 The cone-order on $\overline{\R}_{\geq 0}^X$ is the pointwise usual order on $\overline{\R}_{\geq 0}$.

 Remark also that tropical functions have no reason to be linear nor sublinear.
\end{Remark}

\begin{Remark}\label{rmk:tropMonot}
 Tropical functions are monotone w.r.t.\ to the cone order on its domain and codomain.
 This is clear, using the definition of tropical functions, because $\mu x\leq \mu y$ if $x\leq y$.
\end{Remark}

\begin{Remark}
 The image $fD$ of a directed subset $D$ of a poset $P$ under a monotone function $f:P\to P'$ is always directed.
 This is clear because for all $x,y\in P$, we have $f(x)\leq f(z)\geq f(y)$, where $z\in D$ is s.t.\ $x\leq z\geq y$ and it exists in $D$ since $D$ is directed.
 \end{Remark}

\begin{Remark}
 The cone-order on $\overline{\R}_{\geq 0}^X$ makes it into a dcpo with least element $0$.
\end{Remark}

\begin{Lemma}\label{lm:sup=sup}
 Let $P$ be a poset and $X, I$ sets. Fix the pointwise order on $P^X$.
 \begin{enumerate}
  \item Let $x^i\in P^X$ for $i\in I$.
  If $\bigvee\limits_{i\in I} x^i$ exists $P^X$, then $\bigvee\limits_{i\in I} x^i_a$ exists in $P$ for all $a\in X$, and we have: \[\bigvee\limits_{i\in I} x^i_a=\left(\bigvee\limits_{i\in I} x^i\right)_a.\]
 \item Let $x^i_a\in P$ for $i\in I,a\in X$.
  If $\bigvee\limits_{i\in I} x^i_a$ exists $P$ for all $a\in X$, then $\bigvee\limits_{i\in I} x^i$ exists in $P^X$, and we have: \[\left(\bigvee\limits_{i\in I} x^i\right)_a=\bigvee\limits_{i\in I} x^i_a.\]
 \end{enumerate}
\end{Lemma}
\begin{proof}
 (1). Fix $a\in X$. Let us show that $\left(\bigvee\limits_{i\in I} x^i\right)_a$ is indeed the sup of $\set{x^i_a\mid i\in I}$.
 It is an upper bound because $\bigvee\limits_{i\in I} x^i\geq x^i$.
 Now let $d\in P$ be an upper bound of $\set{x^i_a\mid i\in I}$. In order to show that $d\geq \left(\bigvee\limits_{i\in I} x^i\right)_a$, let us consider $h\in P^X$ defined by $h_c:=\left(\bigvee\limits_{i\in I} x^i\right)_c$ if $c\neq a$ and $h_a:=d$.
 By construction $h$ is an upper bound of $\set{x^i\mid i\in I}$.
 So, by definition of sup, $h\geq \bigvee\limits_{i\in I} x^i$.
 Hence $d=h_a\geq\left(\bigvee\limits_{i\in I} x^i\right)_a$.

 (2). Analogue.
\end{proof}

A \emph{directed net} in a poset $P$ with indices in a set $I$ is a function $s:I\to P$, denoted by $(s_i)_{i\in I}$, s.t.\ its image is directed.
We say that a directed net in $P$ \emph{admits a sup} iff its image admits a sup in $P$.
We say that a directed net $s$ in a normed cone is \emph{bounded} iff the set $\set{\norm{s_i}\,\mid i\in I}$ is bounded in $\R_{\geq 0}$.
In the study of dcpo's the role played by sequences in usual topology is often played by directed nets.

\begin{Remark}\label{rmk:trivial}
 Monotone functions between posets transport directed nets to directed nets\footnote{We simply mean that if $(s_i)_i$ is a directed net in the domain, then $(f(s_i))_i$ is a directed net in the codomain.}.
\end{Remark}

\begin{Definition}
 A function $f:P\to P'$ between posets is \emph{Scott-continuous} iff for all directed net $(s_i)_i$ in $P$ admitting a sup, we have $\exists \bigvee\limits_i f(s_i) = f(\bigvee\limits_i s_i)$ in $P'$. 
\end{Definition}

As it is well known, the order theoretic notion of Scott-continuity coincides with the topological continuity w.r.t.\ to the \emph{Scott-topology} on $P$, whose opens are exactly the upward closed $O\subseteq P$ s.t.\ for all directed $D\subseteq P$ with $\bigvee D\in O$, one has $D\cap O\neq\emptyset$.

\begin{Proposition}\label{prop:infsup}
 Let $P$ be a normed $\overline{\R}_{\geq 0}$-cone s.t.\ every bounded directed net in $P$ admits a sup.
 Let $(v_i)_{i\in I}$ be a directed net in $P$ with an upper bound $v\in P$.
 Then $\exists\bigvee\limits_{i\in I} v_i \in P$ and, if $\inf\limits_{i\in I} \norm{v-v_i} =0$, one has: $\bigvee\limits_{i\in I} v_i = v$.
\end{Proposition}
\begin{proof}
 Remark that $v-v_i$ exists in $P$ by hypothesis and so does $\bigvee\limits_{i\in I} v_i$, thanks to the monotonicity of the norm.
 Now, since $v\geq v_i$ for all $i$, we have that $v\geq \bigvee\limits_{i\in I} v_i$, and so $v-\bigvee\limits_{i\in I} v_i$ exists in $P$.
 Fix $i\in I$.
 Since $v_i\leq \bigvee\limits_{i\in I} v_i$, then $v-\bigvee\limits_{i\in I} v_i\leq v-v_i$ and, by monotonicity of the norm, $\norm{v-\bigvee\limits_{i\in I} v_i}\leq \norm{v-v_i}$.
 Since this holds for all $i\in I$, we have:
 $0\leq \norm{v-\bigvee\limits_{i\in I} v_i}\leq \inf\limits_{i\in I} \norm{v-v_i}=0$, where the last equality holds by hypothesis.
 Thus $\norm{v-\bigvee\limits_{i\in I} v_i}=0$, i.e.\ $v=\bigvee\limits_{i\in I} v_i$.
\end{proof}

\begin{Definition}
 A normed $\overline{\R}_{\geq 0}$-cone $P$ is \emph{Scott-complete} iff its norm is Scott-continuous (where the codomain $\overline\R_{\geq 0}$ is endowed with its usual order) and every bounded directed net in $P$ admits a sup.
 This is equivalent to asking that its closed unit ball is a dcpo.
\end{Definition}

\begin{Proposition}
 The normed cone $\overline{\R}_{\geq 0}^X$ is Scott-complete.
\end{Proposition}
\begin{proof}
 $\overline{\R}_{\geq 0}^X$ being a dcpo, all the existences of sup's that we should check do automatically hold, so we only have to show that $\bigvee$ and $\supnorm{.}$ commute:
 $\supnorm{\bigvee\limits_i x_i}=
 \sup\limits_a\, \sup\limits_i\, (x_i)_a =
 \sup\limits_i\, \sup\limits_a\, (x_i)_a =
 \bigvee\limits_i \supnorm{x_i}$.
\end{proof}

%Let us set $\infty_b$ to be the hyper-semi-plane $\set{y\in\overline{\R}^Y_{\geq 0}\mid y_b=+\infty}$ and $\set{f<<+\infty}:=\bigcup\limits_{b\in Y} f^{-1}\infty_b$.

%\begin{Remark}\label{rmk:R-f<<oo}
%Let $f:\overline{\R}^X_{\geq 0}\to\overline{\R}^Y_{\geq 0}$ be tropical.
%Then, $\R^X_{> 0}-\set{f<<+\infty}$ is a dcpo (w.r.t.\ the usual order, that is, its cone order).
%In order to see this, it is enough to show that, for any $b\in Y$ and for any directed net $(x^i)_i$ in $\R^X_{> 0}$ s.t.\ $f(x^i)_b<+\infty$ for all $i$, we have $f(\bigvee\limits_i x^i)_b<+\infty$ (where the sup is taken in $\R^X_{> 0}$).
%Such property holds because otherwise it is easy to see that $f$ would be discontinuous at the point $\bigvee\limits_i x^i\in\R^X_{> 0}$ w.r.t.\ the topology\footnote{Remark that this argument would not be possible using Scott-continuity instead. Indeed, it could well be that $f(\bigvee\limits_i x^i)_b=\bigvee\limits_i f(x^i)_b=+\infty$ while $f(x^i)_b<+\infty$ for all $i$.} induced by $\supnorm{.}$.
%This contradicts Corollary \ref{cor:tropCont}.
%\end{Remark}

\begin{Proposition}
 All tropical functions $f:\overline{\R}^X_{\geq 0}\to\overline{\R}^Y_{\geq 0}$ are Scott-continuous on $\R^X_{> 0}$ w.r.t.\ the cone-orders on its domain and codomain.
\end{Proposition}
\begin{proof}
 Let $(x_i)_i$ a directed net in $\R^X_{> 0}$ s.t.\ $\bigvee\limits_i x^i$ exists in $\R^X_{> 0}$.
 Then $\inf\limits_i \supnorm{\bigvee\limits_i x^i - x^i} =0$, where $\bigvee\limits_i x^i - x^i$ exists because $\bigvee\limits_i x^i \geq x^i$ for all $i$.
 Since $f$ is $\supnorm{.}$-continuous on $\R^X_{> 0}$ (Corollary \ref{cor:tropCont}), then $\inf\limits_i \supnorm{f(\bigvee\limits_i x^i) - f(x^i)} =0$, where $f(\bigvee\limits_i x^i) - f(x^i)$ exists because $f(\bigvee\limits_i x^i) \geq f(x^i)$ for all $i$ being $f$ monotone (Remark \ref{rmk:tropMonot}).
 We can therefore apply Proposition \ref{prop:infsup} to the directed net $(f(x^i))_i$ in $\overline{\R}^Y_{\geq 0}$, obtaining that $\bigvee\limits_i f(x^i)$ exists in $\overline{\R}^Y_{\geq 0}$ and it coincides with $f(\bigvee\limits_i x^i)$.
\end{proof}

%\begin{Remark}
%By looking at the definition of tropical functions, for $f$ tropical we have that: if $x,x'\in\R^X_{> 0}-\set{f<<+\infty}$ and $r\in\R_{> 0}$, then $x+x',rx\in\R^X_{> 0}-\set{f<<+\infty}$.
%That is, $\R^X_{> 0}-\set{f<<+\infty}$ is a $\R_{>0}$-cone.
%It is clearly still Scott-complete.
%\end{Remark}

\begin{Definition}
 Let $P$ be a dcpo.
 We define a relation by: $x<<y$ iff for all directed $D\subseteq P$, if $y\leq \bigvee D$, then $x\leq d$, for some $d\in D$.
 Call $\twoheaddownarrow y:=\set{x\in P\mid x<<y}$.
 A dcpo $P$ is \emph{Scott-continuous} iff $\twoheaddownarrow x$ is directed and $x=\bigvee\twoheaddownarrow x$ for all $x\in P$.
\end{Definition}

\begin{Remark}
 In any dcpo we have: $\twoheaddownarrow x \subseteq \downarrow x$.
 This immediately follows by considering the directed set $\set{x}$.
\end{Remark}

\begin{Lemma}\label{lm:<<dir}
 In $\overline{\R}_{\geq 0}^X$, every set $\twoheaddownarrow x$ is directed.
\end{Lemma}
\begin{proof}
 It is immediate that $0\in\twoheaddownarrow x$, so it is non-empty.
 Now let $y,y' << x$.
 Since we are in $\overline{\R}_{\geq 0}^X$, there is $y\vee y'\in\overline{\R}_{\geq 0}^X$.
 So we only have to show that $y\vee y'<<x$.
 For that, let $D$ be a directed set in $\overline{\R}_{\geq 0}^X$ s.t.\ $x\leq\bigvee D$.
 Since $y,y' << x$ we find $d,d'\in D$ s.t.\ $y\leq d$, $y'\leq d'$.
 Since $D$ is directed, there is $\hat d\in D$ s.t.\ $y\leq d\leq \hat d\geq d'\geq y'$.
 But then, by definition of sup, it must be $\hat d\geq y\vee y'$ and we are done.
\end{proof}

\begin{Remark}
 The following is a known property (that we will not use):
 let $P$ be a complete normed cone which is a dcpo w.r.t.\ its cone-order. Then $P$ is Scott-continuous as a dcpo iff its closed unit ball is Scott-continuous as a dcpo.
\end{Remark}

\begin{Remark}\label{rmk:<<R}
 Consider $\overline{\R}_{\geq 0}$ with its usual order (which coincides with its cone-order).
 It is easily seen (and well known) that $y<<x$ iff either $y=0$ or $y<x$.
 This immediately implies that $\overline{\R}_{\geq 0}$ is Scott-continuous, since $x=\sup\limits_{y<x} y$.
\end{Remark}

\begin{Lemma}\label{lm:<<RX}
 Let $x,y\in \overline{\R}_{\geq 0}^X$ (considered as a dcpo with its cone order, which is the pointwise one). Fix $a\in X$.
 If $y_a<<x_a$ and $y_c=0$ for all $c\neq a$, then $y<<x$.
\end{Lemma}
\begin{proof}
 Towards a contradiction, assume that there is a directed set $D$ in $\overline{\R}_{\geq 0}^X$ with $x\leq\bigvee D$ and s.t.\ $y\not\leq d$ for all $d\in D$.
 Call $D_a:=\set{d_a\mid d\in D}$ and remark that it is directed, because $D$ is.
 Also, by Lemma \ref{lm:sup=sup}, $\bigvee D_a = \left(\bigvee D\right)_a$.
 Therefore from $y_a<<x_a\leq \bigvee D_a$ we obtain a $d\in D$ s.t.\ $y_a\leq d_a$.
 By the absurd hypothesis we have $y\not\leq d$, so there must be $c\in X$ s.t.\ $y_c\not\leq d_c$, i.e. (because real numbers are totally ordered) $y_c>d_c$.
 Therefore it must be $c\neq a$.
 But then we have $0=y_c>d_c\geq 0$, contradiction.
\end{proof}

\begin{Lemma}\label{lm:<<_a}
 In $\overline{\R}_{\geq 0}^X$ we have:
 if $y<<x$ then $y_a<< x_a$ for all $a\in X$.
\end{Lemma}
\begin{proof}
 Fix $a\in X$.
 Let $D$ directed set in $\overline{\R}_{\geq 0}$ s.t.\ $x_a\leq\bigvee D$.
 We look for a $d\in D$ s.t.\ $y_a\leq d$.
 For $d\in D$, let $x^{a,d}\in\overline{\R}_{\geq 0}^X$ defined by $x^{a,d}_c:=x_c$ if $c\neq a$ and $x^{a,d}_a:=d$.
 Let $D^x_a:=\set{x^{a,d}\mid d\in D}$.
 By Lemma \ref{lm:sup=sup} $D^x_a$ admits sup in $\overline{\R}_{\geq 0}^X$ and it is $(\bigvee D^x_a)_c=x_c$ if $c\neq a$ and $(\bigvee D^x_a)_a=\bigvee D$.
 Hence, $x\leq \bigvee D^x_a$.
 If we prove that $D^x_a$ is directed, we are done: indeed, since $y<<x$, there is $d\in D$ s.t.\ $y\leq x^{a,d}$ and thus, in particular, $y_a\leq x^{a,d}_a=d$.
 Let us finally prove that $D^x_a$ is directed: it is clearly non-empty, since $D$ is.
 Let now $d,d'\in D$.
 We want to show that there is $\hat d\in D$ s.t.\ $x^{a,d}\leq x^{a,\hat d}\geq x^{a,d'}$.
 But since $D$ is directed, there is $\hat d\in D$ s.t.\ $d\leq \hat d\geq d'$, and therefore $x^{a,d}_c=x_c= x^{a,\hat d}_c =x_c=x^{a,d'}_c$ for all $c\neq a$, and $x^{a,d}_a=d\leq \hat d=x^{a,\hat d}_a=\hat d \geq d'=x^{a,d'}_a$.
\end{proof}

\begin{Lemma}\label{rmk:sup_a=sup_a}
In $\overline{\R}_{\geq 0}^X$ we have:
$\bigvee\limits_{y<<x_a} y = \bigvee\limits_{y<<x} y_a$.
\end{Lemma}
\begin{proof}
 If we show that $\twoheaddownarrow x_a = \set{y_a\mid y\in\twoheaddownarrow x}$, then we are done because in the statement we are taking their sup's.
 The inclusion ($\supseteq$) immediately follows from Lemma \ref{lm:<<_a}. For ($\subseteq$), let $d<< x_a$.
 Then the $y\in \overline{\R}_{\geq 0}^X$ defined by $y_c:=0$ if $c\neq a$ and $y_a:=d$, is s.t.\ $y\in\twoheaddownarrow x$ by Lemma \ref{lm:<<RX}.
 Thus, $d\in \set{y_a\mid y\in\twoheaddownarrow x}$.
\end{proof}

\begin{Corollary}
 The dcpo $\overline{\R}_{\geq 0}^X$ is Scott-continuous.
\end{Corollary}
\begin{proof}
 The fact that $\twoheaddownarrow x$ is directed is given by Lemma \ref{lm:<<dir}. The fact that $x=\bigvee \twoheaddownarrow x$ is given by the following equalities:
 $x_a=\bigvee\limits_{y<<x_a} y = \bigvee\limits_{y<<x} y_a = \left(\bigvee\limits_{y<<x} y\right)_a=\left(\bigvee \twoheaddownarrow x\right)_a$.
 The first equality follows from Remark \ref{rmk:<<R}, the second one from Lemma \ref{rmk:sup_a=sup_a}, the third one from Lemma \ref{lm:sup=sup}.
\end{proof}

%\section{Bibliography}

\end{document}

### 2023-05-12 TLLA23/BarbaPisto.tex
Commit

@@
 
-\usepackage{macros}
 \usepackage{tikz}
@@
 
-\title{Tropical methods in Lambda-Calculus}
+\input{macros}
+
+\title{Tropical methods in Lambda-Calculus?}
 \author{Davide Barbarossa
@@
 
-\begin{abstract}
-We propose an interpretation of the $\lambda$-calculus in the framework of tropical mathematics, as a unified framework for both program metrics, based on the analysis of program sensitivity via Lipschitz-conditions, and for resource analysis, based on higher-order program differentiation.
-We sketch the relation of this semantics to quantitative properties like differential privacy, convergence logprobabilities and Probabilistic Coherent spaces.
-Finally, we study the abstract correspondence between this tropical semantics and Lawvere’s generalised metric spaces.
-\end{abstract}
+%\begin{abstract}We propose to study the interpretation of the $\lambda$-calculus in the framework of tropical mathematics, as a unified framework for both program metrics -- based on the analysis of program sensitivity via Lipschitz-conditions -- and for resource analysis -- based on higher-order program differentiation.We sketch the relation of this semantics to quantitative properties like differential privacy, convergence logprobabilities and Probabilistic Coherent spaces.Finally, we study the abstract correspondence between this tropical semantics and Lawvere’s generalised metric spaces.\end{abstract}
 
-\section*{Motivation}
+%\section*{Motivation}
+
+Among the quantitative approaches towards which the programming language community has directed its interest in recent years%(e.g.\ \cite{decarvalho2018, Accattoli2022}, \cite{Breuvart2018, PistoneLICS2022}, \cite{Barthe_2012} \cite{Reed2010, dallago}, \cite{Ehrhard2011, Staton2017}, \cite{difflambda}),
+, two different directions have received considerable attention. 
+On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask  whether they compute functions which do not differ \emph{too much}, and thus study denotational frameworks in which types are endowed with a metric \cite{Reed2010},\cite{Bonchi2018}, \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
+On the other hand, there is the approach based on \emph{differential} \cite{difflambda}, \cite{difflambda}, \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
+or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
+
+In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
+In metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
+By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
+
+At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
+in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz.
+The fundamental motivation of this work is then the observation that 
+this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics where, for instance, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
+Tropical mathematics \cite{Simon} is a well established algebraic and geometrical framework, with tight connections with optimisation theory \cite{Sturmfelds}, where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
+For instance, the polynomial $p(x,y)=x^{2}+xy^{2}+y^{3}$, when interpreted over the tropical semiring, translates as the piecewise linear function
+$
+\varphi(x,y)=\min\{2x, x+2y, 3y\}
+$.
+
+Coming back to our discussion on program semantics, tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
+Actually, a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.
+Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
+However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.
+
+The viewpoint that we develop in the following sections, may suggests the application of optimization methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions, but it scales to a
+more abstract level, leading to introduce a 
+differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}).
+
+In this contribution we only sketch some of the main results of our still work in progress analysis.
+
+\section{Tropical mathematics in a nutshell}
+
+More rigorously, we let the \emph{tropical semiring}, the structure at the heart of tropical mathematics, be $([0,\infty], \min, +)$.
+Remark that this coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014} (the order is the reverse order $\geq$ on $\BB R$ and the monoid action is addition), the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}, and we will take this point of view in the last section of this contributions.
+A tropical polynomial is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form $f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}$
+where the $i_{j}$ are natural numbers and the $c_{i_{j}}$ are in $\Lawv$.
+For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$
+are illustrated in Fig.~\ref{fig:plot1} for $n\leq 4$.
+A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable.
+In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes).
+For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
+With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
+$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
+A \emph{tropical Laurent series} (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function that can be expressed as $f(x)=\inf_{n}\{nx+\matr f_{n}\}$, with $\matr  f_{n}$ a sequence in $\Lawv$.
+In other words, a tLs is a ``limit'' of tropical polynomials of higher and higher degree.
+For instance, the function $\varphi(x):=\inf_{n\in\N}\set{nx+\frac{1}{2^n}}$ (illustrated in Fig.~\ref{fig:plot1}), that we will take as our running example,
+is the ``limit'' of the polynomials $\varphi_{n}$.
+Since $\inf$s are not in general $\min$s, the behavior of tLS may be less predictable than that of tropical polynomials.
+For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points.
+
+\section{Tropical weighted semantics in a nutshell}
+
+The study of matrices with values over the tropical semiring can be seen as a special case of the
+\emph{quantitative relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus and linear logic:
+for a fixed \emph{continuous} semi-ring $Q$ [Def.
+II.5, \cite{Manzo2013}], the category $\QREL$ has sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms, i.e.~$\QREL(X,Y)=Q^{X\times Y}$ (\cite{Manzo2013} would call it $Q^\Pi$).
+As it is expected, $Q^X$ is a $Q$-semimodule and the bijection $\hat{(\cdot)}$ identifies the set of linear maps from $Q^X$ to $Q^Y$ with 
+ $\HOM{\QREL}{X}{Y}$.
+
+The tropical relational model is thus provided by the category $\LREL$ of matrices with values over $\Lawv$ (which, being a quantale, is indeed a continuous semi-ring).
+%, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
+%The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring (w.r.t.\ its quantale order $\preceq$.
+%This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
+It is worth observing that the formula for composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
+reads as \ $(st)_{a,c}:=\inf\limits_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
+ similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$ induced by matrices, which we call \emph{tropical linear}, are exactly those of shape $f(x)_b=\inf\limits_{a\in X} \set{\matr f_{a,b}+x_a}$, for some matrix $\matr f$ from $X$ to $Y$.
+%\end{remark}
+By applying more or less well-known results (taken from \cite{Manzo2013}, CITARE LEMAYYYY), one obtains that $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus:
+
+\begin{fact}
+ $\LREL$ is a SMCC, thus a model of the linear $\STLC$.
+
+ The coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.
+
+The exponential $!$ can be ``decomposed'' into a family of \emph{graded} exponentials $(!_n)_{n\in\N}$ turning $\LREL$, together with $(!_n)_{n\in\N}$, is a model for $\BSTLC$.
+
+ $\LREL_!$ equipped with $D$, is a CC$\partial$C and the interpretation of the Taylor expansion of a $\STLC$-term $M$, given in \eqref{eq:taylor}, converges to the one of $M$.
+\end{fact}
+
+In the previous result, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
+$
+ \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
+ \left\{s_{[b_1,\dots,b_n],c} + \sum\limits_{i=1}^n t_{\mu_i,b_i}\right\}$.
+and the \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
+
+In particular, the Taylor formula decomposes an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as interpreting  a $\lambda$-term 
+as a limit of Lipschitz maps, in some sense, thus bridging the metric and differential approaches.  
+Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
+However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.
+
+Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions?
+This way, \eqref{eq:taylor} could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps.
+In other words, unbounded term application could be seen
+as a limit of \emph{more and more sensitive} operations.
 
 \section{Tropical Laurent Series}
+
+ As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, but we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting 
+ $t^!(x):=t\circ_! x$.
+%  (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
+ This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
+ Concretely, we have
+ \begin{equation}
+ t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}
+ \end{equation}
+ where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
+ These functions correspond then to tLs with possibly infinitely many variables (in fact, as many as the elements of $X$).
+
+Therefore, we study in this section the properties of those functions, from the point of view of mathematical analysis.
+%In the following we will also refer to them as tLs.
+
+% 
+% We will call them simply \emph{tropical Laurent series (tLs)}.
+% %Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
+%\end{remark}
+%
+%We find the usual notion of tLs of one variable as follows:
+
+%\begin{remark}
+Notice that, by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
+In a similar way, the tropical polynomials can be identified with the tLs
+$f:\Lawv^X\to\Lawv^Y$ 
+ for which the support $\C F=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ is \emph{finite}, and which have thus shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$.
+ This is again the generalisation of usual tropical polynomials to the case of infinitely many variables.
 
@@
 \section{Lawvere's generalised metric spaces}
+
+We already remarked that the tropical semiring $\Lawv$ coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$.
+In particular, a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we explicitly take in Section \ref{section6}).
+
+\begin{remark}
+ The category $\QREL$ is (equivalent to) a subcategory of the category $Q\SF{Mod}$ of \emph{complete} $Q$-semimodules.
+ If $\QREL$ corresponds to considering semimodules (the $Q^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $Q\SF{Mod}$ corresponds to considering semimodules in abstract, without fixing a base.
+ We take this viewpoint in Section~\ref{section6}.
+\end{remark}
 
@@
 
-\section*{Bibliography}
+%\section*{Bibliography}
 
@@
 
-\nocite{*}
-\bibliographystyle{eptcs}
+%\nocite{*}
+%\bibliographystyle{eptcs}
 %\bibliography{generic}
+
+%\bibliographystyle{plain}
+%\bibliography{tropical.bib}
+
 \end{document}

### 2023-05-13 TLLA23/BarbaPisto.tex
Committ
Fine prima pagina. Mi piace molto come è venuta.

@@
 
-Among the quantitative approaches towards which the programming language community has directed its interest in recent years%(e.g.\ \cite{decarvalho2018, Accattoli2022}, \cite{Breuvart2018, PistoneLICS2022}, \cite{Barthe_2012} \cite{Reed2010, dallago}, \cite{Ehrhard2011, Staton2017}, \cite{difflambda}),
-, two different directions have received considerable attention. 
-On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask  whether they compute functions which do not differ \emph{too much}, and thus study denotational frameworks in which types are endowed with a metric \cite{Reed2010},\cite{Bonchi2018}, \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
-On the other hand, there is the approach based on \emph{differential} \cite{difflambda}, \cite{difflambda}, \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
-or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
+Two different quantitative approaches have received considerable attention from the programming language community in recent years: %(e.g.\ \cite{decarvalho2018, Accattoli2022}, \cite{Breuvart2018, PistoneLICS2022}, \cite{Barthe_2012} \cite{Reed2010, dallago}, \cite{Ehrhard2011, Staton2017}, \cite{difflambda}), on the one hand one there is
+the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk} is based on the fact that when considering e.g.\ probabilistic %or approximate
+computation, %rather than asking whether two programs compute \emph{the same} function,
+it is natural to ask whether two programs compute functions which %do not differ \emph{too much}
+approximate each other (instead of equality of functions).
+This led to denotational frameworks in which types are endowed with a metric \cite{Reed2010},\cite{Bonchi2018}, \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
+On the other hand, there is the approach based on \emph{differential} \cite{difflambda}, \cite{difflambda}, \cite{Manzo2013, Breuvart2018, PistoneLICS2022} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
+This led to syntactic or denotational frameworks in which one can define a \emph{Taylor expansion} of programs.
 
 In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
-In metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
-By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
+In metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret programs with a fixed duplication bound as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
+By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs gives rise to \emph{power series}.
 
 At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
-in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz.
-The fundamental motivation of this work is then the observation that 
+in metric models $n$ times duplication results in a  \emph{linear} $n$-Lipschitz function $n\cdot x$, while in differential models this results in a non-Lipschitz \emph{polynomial} function $x^{n}$.
+The fundamental motivation of this work is the observation that 
 this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics where, for instance, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
-Tropical mathematics \cite{Simon} is a well established algebraic and geometrical framework, with tight connections with optimisation theory \cite{Sturmfelds}, where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
-For instance, the polynomial $p(x,y)=x^{2}+xy^{2}+y^{3}$, when interpreted over the tropical semiring, translates as the piecewise linear function
-$
-\varphi(x,y)=\min\{2x, x+2y, 3y\}
-$.
-
-Coming back to our discussion on program semantics, tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
 Actually, a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.
@@
 
-The viewpoint that we develop in the following sections, may suggests the application of optimization methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions, but it scales to a
+Tropical mathematics \cite{Simon} is a well established algebraic and geometrical framework, with tight connections with optimisation theory \cite{Sturmfelds}, where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
+For instance, the polynomial $p(x,y)=x^{2}+xy^{2}+y^{3}$, when interpreted over the tropical semiring, translates as the piecewise linear function
+$
+\trop f(\alpha,\beta)=\min\{2\alpha, \alpha+2\beta, 3\beta\}
+$.
+
+The viewpoint that we develop in the following sections can be read as the proposition of a setting for bridging the two approaches mentioned at the beginning, and may suggests the application of methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions.
+It also scales to a 
 more abstract level, leading to introduce a 
 differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}).
-
-In this contribution we only sketch some of the main results of our still work in progress analysis.
+We will only sketch some of our main results.% of our our analysis.
 

### 2023-05-13 TLLA23/BarbaPisto.tex
Commit
Pagina 1 e 2 fatte. Mi stanno proprio piacendo, let me know che ne pensi !

@@
 
-\title{Tropical methods in Lambda-Calculus?}
+\title{Tropical methods in Lambda-Calculus}
 \author{Davide Barbarossa
@@
 At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
-in metric models $n$ times duplication results in a  \emph{linear} $n$-Lipschitz function $n\cdot x$, while in differential models this results in a non-Lipschitz \emph{polynomial} function $x^{n}$.
-The fundamental motivation of this work is the observation that 
-this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics where, for instance, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
-Actually, a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.
-Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
-However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.
-
+in metric models $n$ times duplication results in a $n$-Lipschitz \emph{linear} function $n\cdot x$, while in differential models this results in a non-Lipschitz \emph{polynomial} function $x^{n}$.
+Our fundamental observation is that 
+this gap is overcome once we interpret these functions in the framework of tropical mathematics where, for instance, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
 Tropical mathematics \cite{Simon} is a well established algebraic and geometrical framework, with tight connections with optimisation theory \cite{Sturmfelds}, where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
@@
 
-The viewpoint that we develop in the following sections can be read as the proposition of a setting for bridging the two approaches mentioned at the beginning, and may suggests the application of methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions.
+A tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties.
+Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
+However, a thorough investigation of the full power of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.
+We sketch here some first steps.
+The viewpoint that we develop can be read as the proposition of a setting able to bridge the two approaches mentioned above, and may suggests the application of methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions.
 It also scales to a 
-more abstract level, leading to introduce a 
-differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}).
-We will only sketch some of our main results.% of our our analysis.
+more abstract level, leading to introduce a differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}).
+%We will only sketch some of our main results.% of our our analysis.
 
-\section{Tropical mathematics in a nutshell}
+\paragraph{Tropical mathematics in a nutshell}
 
-More rigorously, we let the \emph{tropical semiring}, the structure at the heart of tropical mathematics, be $([0,\infty], \min, +)$.
-Remark that this coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014} (the order is the reverse order $\geq$ on $\BB R$ and the monoid action is addition), the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}, and we will take this point of view in the last section of this contributions.
-A tropical polynomial is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form $f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}$
-where the $i_{j}$ are natural numbers and the $c_{i_{j}}$ are in $\Lawv$.
-For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$
-are illustrated in Fig.~\ref{fig:plot1} for $n\leq 4$.
-A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable.
-In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes).
-For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
-With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
-$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
-A \emph{tropical Laurent series} (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function that can be expressed as $f(x)=\inf_{n}\{nx+\matr f_{n}\}$, with $\matr  f_{n}$ a sequence in $\Lawv$.
-In other words, a tLs is a ``limit'' of tropical polynomials of higher and higher degree.
-For instance, the function $\varphi(x):=\inf_{n\in\N}\set{nx+\frac{1}{2^n}}$ (illustrated in Fig.~\ref{fig:plot1}), that we will take as our running example,
-is the ``limit'' of the polynomials $\varphi_{n}$.
-Since $\inf$s are not in general $\min$s, the behavior of tLS may be less predictable than that of tropical polynomials.
-For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points.
+We let the \emph{tropical semiring}  $\Lawv$, the structure at the heart of tropical mathematics, be $[0,\infty]$ with addition $\min$ and multiplication $+$.
+This coincides with the \emph{Lawvere quantale} $\Lawv$ \cite{Hofmann2014, Stubbe2014}, i.e.\ $[0,\infty]$ with order the reversed order from $\BB R$ and usual addition as the monoid action. 
+This is the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}, and we will take this point of view in the last section of this contribution.
+A \emph{tropical polynomial} is a piece-wise linear function $\varphi:\Lawv\to \Lawv$ of the form $\varphi(x)=\min_{j=1,\dots,k}\{i_{j}x+\matr \varphi_{i_{j}}\}$, with $i_{j}\in\N$, $\matr \varphi_{i_{j}}\in\Lawv$ and $k$ finite.
+Those are always Lipschitz functions.
+For example, $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$.
+A \emph{tropical root} of a tropical polynomial $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable, i.e.\ where the minimum defining $\varphi$ is attained at least twice, i.e.~where the slope of $\varphi$ changes.
+E.g., the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, $i \leq n$.
+%With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as $f(x)=\min\{x,x_{0}\}+ g(x)$.
+Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
+A \emph{tropical Laurent series} (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function $\varphi:\Lawv\to \Lawv$ of the form $\varphi(x)=\inf_{n\in\N}\{nx+\matr \varphi_{n}\}$, with $\matr \varphi_{n}\in\Lawv$.
+That is, a tLs is a ``limit'' of tropical polynomials of higher and higher degree.
+For example $\varphi(x):=\inf_{i\in\N}\set{ix+2^{-i}}$ %that we will take as our running example, 
+is the ``limit'' of the polynomials $\varphi_{n}$. %Since $\inf$s are not in general $\min$s, the behavior of tLS may be less predictable than that of tropical polynomials.
+%For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points.
+TLs may be non Lipschitz, and their study is less developed than that of tropical polynomials.
 
-\section{Tropical weighted semantics in a nutshell}
+\paragraph{Tropical weighted relational semantics in a nutshell}
 
 The study of matrices with values over the tropical semiring can be seen as a special case of the
-\emph{quantitative relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus and linear logic:
+\emph{weighted relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus and linear logic:
 for a fixed \emph{continuous} semi-ring $Q$ [Def.
-II.5, \cite{Manzo2013}], the category $\QREL$ has sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms, i.e.~$\QREL(X,Y)=Q^{X\times Y}$ (\cite{Manzo2013} would call it $Q^\Pi$).
-As it is expected, $Q^X$ is a $Q$-semimodule and the bijection $\hat{(\cdot)}$ identifies the set of linear maps from $Q^X$ to $Q^Y$ with 
- $\HOM{\QREL}{X}{Y}$.
-
-The tropical relational model is thus provided by the category $\LREL$ of matrices with values over $\Lawv$ (which, being a quantale, is indeed a continuous semi-ring).
+II.5, \cite{Manzo2013}], the category $\QREL$ has sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms, i.e.~$\HOM \QREL X Y=Q^{X\times Y}$.
+%(\cite{Manzo2013} would call it $Q^\Pi$).
+As expected, $Q^X$ is a $Q$-semimodule and we can identify $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.
+Taking $Q:=\Lawv$ we obtain the \emph{tropical weighted relational model} $\LREL$.
 %, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
@@
 %This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
-It is worth observing that the formula for composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
-reads as \ $(st)_{a,c}:=\inf\limits_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
- similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$ induced by matrices, which we call \emph{tropical linear}, are exactly those of shape $f(x)_b=\inf\limits_{a\in X} \set{\matr f_{a,b}+x_a}$, for some matrix $\matr f$ from $X$ to $Y$.
+It is worth observing that the composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
+reads as \ $(s\circ t)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
+similarly, linear maps $f:\Lawv^X\to \Lawv^Y$ are of shape $f(x)_b=\inf_{a\in X} \set{x_a+\matr f_{a,b}}$, for some matrix $\matr f\in Q^{X\times Y}$, are precisely those induced by $\matr f\in\HOM \LREL X Y$.
+We call them \emph{tropical linear}.
 %\end{remark}
 By applying more or less well-known results (taken from \cite{Manzo2013}, CITARE LEMAYYYY), one obtains that $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus:
-
-\begin{fact}
- $\LREL$ is a SMCC, thus a model of the linear $\STLC$.
-
- The coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.
-
-The exponential $!$ can be ``decomposed'' into a family of \emph{graded} exponentials $(!_n)_{n\in\N}$ turning $\LREL$, together with $(!_n)_{n\in\N}$, is a model for $\BSTLC$.
-
- $\LREL_!$ equipped with $D$, is a CC$\partial$C and the interpretation of the Taylor expansion of a $\STLC$-term $M$, given in \eqref{eq:taylor}, converges to the one of $M$.
-\end{fact}
-
-In the previous result, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
+$\LREL$ is a SMCC, i.e.\ a model of the linear $\STLC$.
+The coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$, where $!$ is the usual multiset comonad (so $!X$ is the set of finite multisets on $X$).
+Here, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ given by:
 $
@@
  \left\{s_{[b_1,\dots,b_n],c} + \sum\limits_{i=1}^n t_{\mu_i,b_i}\right\}$.
-and the \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
-
-In particular, the Taylor formula decomposes an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as interpreting  a $\lambda$-term 
-as a limit of Lipschitz maps, in some sense, thus bridging the metric and differential approaches.  
-Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
-However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.
-
-Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions?
-This way, \eqref{eq:taylor} could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps.
-In other words, unbounded term application could be seen
+Furthermore, $!$ can be decomposed into a family of \emph{graded} exponentials $(!_n)_{n\in\N}$ turning $(\LREL,(!_n)_{n\in\N})$ in a model for $\BSTLC$. 
+Finally, $\LREL_!$ can be equipped with a \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined by: $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
+This turns $(\LREL_!,D)$ into a CC$\partial$C, i.e.\ a model of the differential $\lambda$-calculus.
+Remember that the (quantitative) Taylor expansion of an ordinary $\lambda$-term $M$ is an inductively defined series $\Te{M}$ of differential $\lambda$-terms, the only non-trivial case being $\Te{PQ} :=  \sum_{k=0}^{\infty} (\Der^{k}[P] \cdot Q^{k})0$.
+It can be seen that the series interpreting in $\LREL_!$ the Taylor expansion of a $\STLC$-term $M$, converges to the interpretation of $M$.
+Now, the Taylor formula decomposes an unbounded application as a limit of bounded ones.
+So one might well ask whether this formula interprets a $\lambda$-term 
+as a limit of Lipschitz maps and thus bridging, in a sense, the metric and differential approaches previously mentioned.  
+%Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
+But terms with bounded applications correspond semantically to \emph{polynomials}, usually non-Lipschitz functions. 
+Yet, if these polynomials were tropical ones, the Taylor expansion would really decompose $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps: just like in mathematical analysis, unbounded application could be seen
 as a limit of \emph{more and more sensitive} operations.
+This is precisely the content of our Theorem~\ref{}.
 
@@
  This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
- Concretely, we have
- \begin{equation}
- t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}
- \end{equation}
+ Concretely, we have $t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$
  where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.

### 2023-05-13 TLLA23/BarbaPisto.tex
Mini-commit serale

@@
 
- As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, but we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting 
+ As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a \emph{linear} map $\Lawv^{!X}\to\Lawv^Y$, but we can also ``express it in the base $X$'' and see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting
  $t^!(x):=t\circ_! x$.
 %  (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
- This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
- Concretely, we have $t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$
- where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
- These functions correspond then to tLs with possibly infinitely many variables (in fact, as many as the elements of $X$).
-
-Therefore, we study in this section the properties of those functions, from the point of view of mathematical analysis.
+% This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
+ Concretely, we have $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$
+ where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
+ These functions correspond then to generalised tLs, i.e.\ with possibly infinitely many variables (as many as the elements of $X$).
+For $X=Y=\set{*}$, we get usual tLs of one variable.
+Instead, if the support $\C F:=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ of $f$ is \emph{finite}, we get maps of shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, i.e.\ tropical polynomials in possibly infinitely many variables.
+In this section we study the properties of generalised tLs, from the point of view of mathematical analysis.
 %In the following we will also refer to them as tLs.
@@
 %\begin{remark}
-Notice that, by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
-In a similar way, the tropical polynomials can be identified with the tLs
-$f:\Lawv^X\to\Lawv^Y$ 
- for which the support $\C F=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ is \emph{finite}, and which have thus shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$.
- This is again the generalisation of usual tropical polynomials to the case of infinitely many variables.
 

### 2023-05-14 TLLA23/BarbaPisto.tex
Committ pomeridiano (Sezione sulle tLs fatta!)
Non ho lavorato quanto avrei voluto, ma ora faccio una pausa e poi continuo !

@@
 \usepackage{underscore}           % Only needed if you use pdflatex.
-
-\usepackage{tikz}
-  \usetikzlibrary{calc}
+\usepackage{cite}
+\usepackage{amsmath,amssymb,amsfonts}
+\usepackage{algorithmic}
+\usepackage{graphicx}
+\usepackage{bussproofs}
+\EnableBpAbbreviations
 \usepackage{hyperref}
-\usepackage{stmaryrd,amssymb,proof,cmll,mathrsfs,amsmath,amsthm}
+\usepackage{tikz,pgfplots}
+\pgfplotsset{compat=newest}
+\usepackage{tikz-cd}
+\usepackage{textcomp}
+\usepackage{xcolor}
+\usepackage{amsthm}
+\usepackage{caption}
+\usepackage{subcaption}
+\usepackage{wrapfig}
+\usetikzlibrary{cd}
+%\usepackage{mathabx}
+\usepackage{cmll}
+\usepackage{stmaryrd}
 \usepackage{todonotes}
@@
 Those are always Lipschitz functions.
-For example, $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$.
-A \emph{tropical root} of a tropical polynomial $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable, i.e.\ where the minimum defining $\varphi$ is attained at least twice, i.e.~where the slope of $\varphi$ changes.
+For example, $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$, plotted in Fig~\ref{fig:plot1}.
+A \emph{tropical root} of $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable, i.e.\ where the minimum defining $\varphi$ is attained at least twice, i.e.~where the slope of $\varphi$ changes.
 E.g., the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, $i \leq n$.
@@
 For example $\varphi(x):=\inf_{i\in\N}\set{ix+2^{-i}}$ %that we will take as our running example, 
-is the ``limit'' of the polynomials $\varphi_{n}$. %Since $\inf$s are not in general $\min$s, the behavior of tLS may be less predictable than that of tropical polynomials.
+is the ``limit'' of the $\varphi_{n}$, see Fig~\ref{fig:plot1}. %Since $\inf$s are not in general $\min$s, the behavior of tLS may be less predictable than that of tropical polynomials.
 %For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points.
@@
 
-The study of matrices with values over the tropical semiring can be seen as a special case of the
+The study of matrices with values over the tropical semiring can is a special case of the
 \emph{weighted relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus and linear logic:
@@
 Remember that the (quantitative) Taylor expansion of an ordinary $\lambda$-term $M$ is an inductively defined series $\Te{M}$ of differential $\lambda$-terms, the only non-trivial case being $\Te{PQ} :=  \sum_{k=0}^{\infty} (\Der^{k}[P] \cdot Q^{k})0$.
-It can be seen that the series interpreting in $\LREL_!$ the Taylor expansion of a $\STLC$-term $M$, converges to the interpretation of $M$.
+It can be seen that the morphisms of $\LREL_!$ can always be Taylor expanded, and the series interpreting in $\LREL_!$ the Taylor expansion of a $\STLC$-term $M$, converges to the interpretation of $M$.
 Now, the Taylor formula decomposes an unbounded application as a limit of bounded ones.
@@
 as a limit of \emph{more and more sensitive} operations.
-This is precisely the content of our Theorem~\ref{}.
+This is precisely the content of our Theorem~\ref{thm:main}(3).
 
@@
 %\begin{remark}
+
+\begin{wrapfigure}{r}{0.3\textwidth}%
+%\begin{figure}
+\begin{tikzpicture}[scale=0.55]
+\begin{axis}[samples=250]
+\addplot[yellow,domain=0:0.8] {1+0.02};
+
+\addplot[orange,domain=0:0.8] {min(x+1/2, 1)+0.01};
+
+\addplot[red,domain=0:0.8] {min(2*x+1/4, x+1/2, 1};
+\addplot[blue,domain=0:0.8] {min(3*x+1/8,2*x+1/4, x+1/2, 1)-0.01};
+\addplot[orange,domain=0:0.8] {min(4*x+1/16,3*x+1/8,2*x+1/4, x+1/2, 1)-0.02};
+
+\addplot[violet,domain=0:0.8] {min(
+10*x+1/1424,
+9*x+1/712,
+8*x+1/356,
+7*x+1/128,
+6*x+1/64,
+5*x+1/32,
+4*x+1/16,3*x+1/8,2*x+1/4, x+1/2, 1)-.03};
+
+\end{axis}
+
+\end{tikzpicture}
+\caption{Plot of the tropical polynomials $\varphi_1,\varphi_2,\varphi_3,\varphi_4$ (from top to bottom), and of their limit tLs $\varphi$ (in violet).}
+\label{fig:plot1}
+%\end{figure} 
+\end{wrapfigure} %%
+
+\begin{proposition}\label{prop:nondecr+conc}
+ Any tLs $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
+\end{proposition}
+
+%In Example~\ref{ex:famous_ex}, 
+Look at Fig~\ref{fig:plot1}: it appears that $\varphi$ behaves \emph{locally} like the polynomials $\varphi_{n}$. 
+%In particular, for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$.
+More, precisely, $\varphi$ coincides with $\varphi_{n}$ for $\epsilon \geq 2^{-(n+1)}$, the smallest tropical root of $\varphi_{n}$.
+However, at
+%
+%$x\in [\epsilon,\infty]$, with $\epsilon>0$
+%It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
+ $x=0$ we have that $\varphi(x=0)=\inf_{i\in\N} 2^{-i}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
+Also, while the derivative of $\varphi$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
+This phenomenon is reminiscent of [Example 7, \cite{Ehrhard2005}],
+%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
+which actually motivated our first investigations.
+In fact, these properties are shared by all tLs with \emph{finitely} many variables, as shown by the following result.
+
+Remark that $!\set{1,\dots,k}$ can be identified with $\N^k$.
+So the matrix of a tLs $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one variable $f(x)$ in output) can be given as a $\matr f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, where $nx$ is the scalar product.
+
+\begin{theorem}\label{theorem:fepsilon}
+ Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\matr f:\N^k\to\Lawv$.
+ For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
+% 
+% \begin{enumerate}
+%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
+%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
+  %\item 
+$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical {polynomial} $P_\epsilon(x):=\min_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
+% \end{enumerate}
+\end{theorem}
+\begin{proof}[Proof sketch]
+Let $\C F_\epsilon$ be the set of $n\in\N^{k}$ such that 
+$\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
+We can show that it is finite and does the job.
+\end{proof}
+
+The norm $\norm{\cdot}_\infty$ on $\BB R$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$.
+We will show that tLs satisfy suitable Lipschitz properties  w.r.t.~these metrics.
+\emph{Wlog}, we will often take $Y=\set{*}$.
+
+First, it can be seen that all tropical \emph{linear} functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive.  
+%\begin{proof}[Proof sketch]
+%Using the fact that $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
+%the problem reduces to checking that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
+This result shows that, in analogy with that happens in usual metric semantics, linear programs are interpreted by non-expansive functions. 
+%\begin{proof}
+%Using $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
+%first observe that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$; we now have
+%$|f(\B x)_{b}-f(\B y)_{b}| \leq |(\inf_{a\in X}\matr f_{a,b}-\B x_{a})-(\inf_{a\in X}\matr f_{a,b}-\B y_{a})| \leq
+%\sup_{a\in X}|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})|\leq 
+% \| \B x-\B y\|_{\infty}$.
+%\end{proof}
+Let us now consider the case of bounded exponentials:
+\begin{proposition}\label{prop:boundedlip}
+If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a bounded matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
+\end{proposition}
+\begin{proof}[Proof sketch]
+It follows from the above lines about tropical linear functions, plus the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $!
+x$ to finite multisets of degree $\leq n$.%
+%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
+%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
+%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
+\end{proof}
+This result is perfectly analogous to what happens in the metric models discussed in Section \ref{section2}, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
+It also entails that for any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$, $\varphi$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
+
+Let us now look at what happens with tLs, i.e.~when considering the full exponential $!$.
+As consequence of Theorem~\ref{theorem:fepsilon}, the tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $\BB R_{>0}$.
+Actually, we can prove a more general statement, also covering the infinitary case.
+
+\begin{theorem}\label{thmTLSlocLip}
+ All tLs $\Lawv^X\to\Lawv$ are locally Lipschitz on $\BB R_{>0}^X$.
+\end{theorem}
+\begin{proof}[Proof sketch]
+The core of the proof is a convex analysis argument showing that an arbitrary function $f:\Lawv^X\to\Lawv$ which is non-decreasing, concave and continuous, must be locally Lipschitz. 
+\end{proof}
+
+Finally, the differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tLs $f:\Lawv^{X}\to \Lawv^{Y}$ into a tLs $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by $D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a}+x_{a}+\mu y\right\}$.
+One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
+Moreover, the Taylor formula enjoyed by $\LREL_!$ morphisms, yields a ``tropical'' Taylor formula for tLs of the form $f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}$.
+
+The above results translate into the following facts about the interpretation of higher-order programs:
+\begin{corollary}\label{thm:main}
+For any $\lambda$-term $M$:
+\begin{enumerate}
+\item if $\Gamma \vdash_{\BSTLC} M:A$, then $\model M^!:\model\Gamma \to \model A$ is a \emph{Lipschitz} map.
+\item if $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is a \emph{locally Lipschitz} map.
+\item  if $\Gamma \vdash_{\STLC}M:A$, then its Taylor exp.\ decomposes $\model M^!: \model\Gamma \to \model A$ into an $\inf$ of \emph{Lipschitz}~maps.
+\end{enumerate}
+\end{corollary}
 

### 2023-05-14 TLLA23/BarbaPisto.tex
Commit pre-cena
Parte su Lawver fatta. Manca solo la parte delle "applicazioni".

@@
 as a limit of \emph{more and more sensitive} operations.
-This is precisely the content of our Theorem~\ref{thm:main}(3).
+This is precisely the content of our Corollary~\ref{cor:main}(3).
 
@@
 The above results translate into the following facts about the interpretation of higher-order programs:
-\begin{corollary}\label{thm:main}
+\begin{corollary}\label{cor:main}
 For any $\lambda$-term $M$:
@@
 
-\section{Lawvere's generalised metric spaces}
+\section{Getting rid of bases}
 
-We already remarked that the tropical semiring $\Lawv$ coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$.
-In particular, a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we explicitly take in Section \ref{section6}).
+Remember that a $\Lawv$-module is a triple $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$.% (notice that this is indeed the usual notion of module over the Lawvere quantale $\Lawv$).
+$\Lawv$ is the most basic example of $\Lawv$-module.
+ Other basic examples of $\Lawv$-modules are the sets $\Lawv^{X}$, with order and action defined pointwise. 
+A $\Lawv$-module homomorphism is a map $f:M\to N$ commuting with both joins and the $\Lawv$-action. We let $\Mod$ indicate the category of $\Lawv$-modules and their homomorphisms.
 
-\begin{remark}
- The category $\QREL$ is (equivalent to) a subcategory of the category $Q\SF{Mod}$ of \emph{complete} $Q$-semimodules.
- If $\QREL$ corresponds to considering semimodules (the $Q^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $Q\SF{Mod}$ corresponds to considering semimodules in abstract, without fixing a base.
- We take this viewpoint in Section~\ref{section6}.
-\end{remark}
+Observe that the category $\LREL$ is (equivalent to) a subcategory of $\Lawv\SF{Mod}$.
+While $\LREL$ corresponds to considering modules (the $\LREL^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $\Lawv\SF{Mod}$ corresponds to considering semimodules in abstract, without fixing a base.
+
+We already remarked that the tropical semiring coincides with the \emph{Lawvere quantale} $\Lawv$.
+In particular, Lawvere was the first to observe that a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law.
+Indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories} (in short, a $\Lawv$-category) \cite{Lawvere1973, Hofmann2014, Stubbe2014}.
+%Indeed, spelling out the definition, a $\Lawv$-enriched category is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$ satisfying 
+%\begin{align}
+%0  & \geq X(x,x) \tag{$\Lawv$-cat 1}\\
+%X(y,z)+X(x,y)&\geq  X(x,z) \tag{$\Lawv$-cat 2}
+%\end{align}
+%This structure is often referred to as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}. %, or as a \emph{quasi-metric space}. 
+%A $\Lawv$-category is \emph{skeletal} \cite{Stubbe2014} when $X(x,y)=0$ implies $x=y$, and \emph{symmetric} when it coincides with its opposite category $X^{\op}(x,y):=X(y,x)$, i.e.~when $X(x,y)=X(y,x)$.
+A metric space, in the usual sense, is just a particular case of $\Lawv$-category.
+%Notice that any $\Lawv$-category $X$ induces a skeletal category $X^{\sk}$ by quotienting points under $X(x,y)=0$, and a symmetric one by letting $X^{\sym}(x,y)=\max\{X(x,y),X^{\op}(x,y)\}$.
+Remark that $\Lawv$ has a canonical $\Lawv$-enriched structure given by $\Lawv(r,s)=s \overset{\cdot}{-} r$ (where ``$ \overset{\cdot}{-}$'' indicates truncated subtraction), 
+and the Euclidean distance coincides with its symmetrization $\Lawv^{\sym}(x,y):=\max\{\Lawv(x,y),\Lawv^{\op}(x,y)\}$.
+
+%A $\Lawv$-enriched \emph{functor} between $\Lawv$-categories $X,Y$ is just a non-expansive map $f:X\to Y$, as functoriality reads as $Y(f(x),f(y))\leq X(x,y)$.
+Now, by considering distributors, one defines the notion of \emph{continuous} functor $f:X\to Y$ as one that commutes with all existing weighted colimits in $X$.
+A $\Lawv$-enriched category $X$ is said \emph{categorically complete} (or just \emph{complete}) if all weighted colimits over $X$ exist. 
+We let $\GMet$ indicate the category of complete and skeletal $\Lawv$-categories and continuous functors.
+We mention this notions because it can be shown (~\cite{Stubbe2006}) that, actually, $\Mod$ and $\GMet$ are isomorphic categories.
+This is interesting for us because it suggests to transport the metric point of view of $\GMet$ on our modules in $\Mod$.
+Actually, we can even show that:
+\begin{theorem}
+$\Mod_{!}$ or, equivalently, $\GMet_{!}$, can be equipped with a differential $D$ making it a $CC\partial C$.
+\end{theorem}
+That is, one does not need to stick to a base in order to interpret types/programs.
+This differential extends the one for $\LREL_{!}$;
+in order to define the exponential $!$ over $\Mod$ we use the well-known recipe from \cite{Mellies2018, Manzo2013}, consisting in defining a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$, and then define $!$ as the product of the symmetric algebras.
 

### 2023-05-14 TLLA23/BarbaPisto.tex
Committ notturno

@@
 it is natural to ask whether two programs compute functions which %do not differ \emph{too much}
-approximate each other (instead of equality of functions).
+approximate each other, instead of equality of functions.
 This led to denotational frameworks in which types are endowed with a metric \cite{Reed2010},\cite{Bonchi2018}, \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
-On the other hand, there is the approach based on \emph{differential} \cite{difflambda}, \cite{difflambda}, \cite{Manzo2013, Breuvart2018, PistoneLICS2022} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
+On the other hand, there is the approach based on \emph{differential} \cite{difflambda}, \cite{difflambda}, \cite{Manzo2013, Breuvart2018, PistoneLICS2022} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
 This led to syntactic or denotational frameworks in which one can define a \emph{Taylor expansion} of programs.
-
 In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
-In metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret programs with a fixed duplication bound as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
+In metric semantics, linear programs correspond to \emph{non-expansive} maps, i.e.\ that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret programs with a fixed duplication bound as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
 By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs gives rise to \emph{power series}.
@@
 in metric models $n$ times duplication results in a $n$-Lipschitz \emph{linear} function $n\cdot x$, while in differential models this results in a non-Lipschitz \emph{polynomial} function $x^{n}$.
-Our fundamental observation is that 
-this gap is overcome once we interpret these functions in the framework of tropical mathematics where, for instance, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
+Our main observation is that 
+this gap is overcome once we interpret these functions in the framework of tropical mathematics where, for instance, $x^{n}$ precisely reads as $n\cdot x$.
 Tropical mathematics \cite{Simon} is a well established algebraic and geometrical framework, with tight connections with optimisation theory \cite{Sturmfelds}, where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
@@
 A tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties.
-Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
+Connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
 However, a thorough investigation of the full power of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.
-We sketch here some first steps.
-The viewpoint that we develop can be read as the proposition of a setting able to bridge the two approaches mentioned above, and may suggests the application of methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions.
-It also scales to a 
+We sketch here some first steps, with the idea of bridging the two approaches mentioned above, and suggesting the application of tropical methods to the study of the $\lambda$-calculus and its quantitative extensions.
+This also scales to a 
 more abstract level, leading to introduce a differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}).
@@
 We let the \emph{tropical semiring}  $\Lawv$, the structure at the heart of tropical mathematics, be $[0,\infty]$ with addition $\min$ and multiplication $+$.
-This coincides with the \emph{Lawvere quantale} $\Lawv$ \cite{Hofmann2014, Stubbe2014}, i.e.\ $[0,\infty]$ with order the reversed order from $\BB R$ and usual addition as the monoid action. 
-This is the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}, and we will take this point of view in the last section of this contribution.
+This coincides with the \emph{Lawvere quantale} $\Lawv$ \cite{Hofmann2014, Stubbe2014}, i.e.\ $[0,\infty]$ with order $\geq$ and usual $+$ as the monoid action. 
+$\Lawv$ is at the heart of the categorical study of metric spaces initiated by Lawvere \cite{Lawvere1973}, and we will take this point of view in the last section.
 A \emph{tropical polynomial} is a piece-wise linear function $\varphi:\Lawv\to \Lawv$ of the form $\varphi(x)=\min_{j=1,\dots,k}\{i_{j}x+\matr \varphi_{i_{j}}\}$, with $i_{j}\in\N$, $\matr \varphi_{i_{j}}\in\Lawv$ and $k$ finite.
@@
 For example, $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$, plotted in Fig~\ref{fig:plot1}.
-A \emph{tropical root} of $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable, i.e.\ where the minimum defining $\varphi$ is attained at least twice, i.e.~where the slope of $\varphi$ changes.
+A \emph{tropical root} of $\varphi$ is a point $x\in \Lawv$ where the minimum defining $\varphi$ is attained at least twice.%, i.e.\ where $\varphi$ is not differentiable, i.e.~where the slope of $\varphi$ changes.
 E.g., the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, $i \leq n$.
@@
 %For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points.
+Finally, for a polynomial/power series $f(x)=\sum_n a_nx^n$, one defines its \emph{tropicalization} $\trop f(\alpha):=\inf_n\set{-\log a_n+n\alpha}$.
 TLs may be non Lipschitz, and their study is less developed than that of tropical polynomials.
@@
 
-The study of matrices with values over the tropical semiring can is a special case of the
+The study of matrices with values over the tropical semiring is a special case of the
 \emph{weighted relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus and linear logic:
-for a fixed \emph{continuous} semi-ring $Q$ [Def.
-II.5, \cite{Manzo2013}], the category $\QREL$ has sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms, i.e.~$\HOM \QREL X Y=Q^{X\times Y}$.
+for a fixed \emph{continuous} semi-ring $Q$, take the category $\QREL$ whose objects are sets and $\HOM \QREL X Y=Q^{X\times Y}$ (set-indexed matrices with coefficients in $Q$).
 %(\cite{Manzo2013} would call it $Q^\Pi$).
@@
 %This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
-It is worth observing that the composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
+Remark that the composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
 reads as \ $(s\circ t)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
-similarly, linear maps $f:\Lawv^X\to \Lawv^Y$ are of shape $f(x)_b=\inf_{a\in X} \set{x_a+\matr f_{a,b}}$, for some matrix $\matr f\in Q^{X\times Y}$, are precisely those induced by $\matr f\in\HOM \LREL X Y$.
+similarly, linear maps $f:\Lawv^X\to \Lawv^Y$ are of shape $f(x)_b=\inf_{a\in X} \set{x_a+\matr f_{a,b}}$, for some matrix $\matr f\in Q^{X\times Y}$ and are precisely those induced by $\matr f\in\HOM \LREL X Y$.
 We call them \emph{tropical linear}.
 %\end{remark}
-By applying more or less well-known results (taken from \cite{Manzo2013}, CITARE LEMAYYYY), one obtains that $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus:
-$\LREL$ is a SMCC, i.e.\ a model of the linear $\STLC$.
-The coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$, where $!$ is the usual multiset comonad (so $!X$ is the set of finite multisets on $X$).
+By applying known results (taken from \cite{Manzo2013},  {\color{red}CITARE LEMAYYYY}), one obtains that $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus:
+$\LREL$ is a SMCC, i.e.\ a model of the linear $\STLC$;
+the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$, where $!$ is the usual multiset comonad (so $!X$ is the set of finite multisets on $X$).
 Here, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ given by:
@@
  \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
- \left\{s_{[b_1,\dots,b_n],c} + \sum\limits_{i=1}^n t_{\mu_i,b_i}\right\}$.
+ \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
 Furthermore, $!$ can be decomposed into a family of \emph{graded} exponentials $(!_n)_{n\in\N}$ turning $(\LREL,(!_n)_{n\in\N})$ in a model for $\BSTLC$. 
-Finally, $\LREL_!$ can be equipped with a \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined by: $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
+Finally, $\LREL_!$ can be equipped with a \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined by: $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise.
+%(where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
 This turns $(\LREL_!,D)$ into a CC$\partial$C, i.e.\ a model of the differential $\lambda$-calculus.
@@
 It can be seen that the morphisms of $\LREL_!$ can always be Taylor expanded, and the series interpreting in $\LREL_!$ the Taylor expansion of a $\STLC$-term $M$, converges to the interpretation of $M$.
-Now, the Taylor formula decomposes an unbounded application as a limit of bounded ones.
+The Taylor formula decomposes an unbounded application as a limit of bounded ones.
 So one might well ask whether this formula interprets a $\lambda$-term 
@@
 %Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
-But terms with bounded applications correspond semantically to \emph{polynomials}, usually non-Lipschitz functions. 
-Yet, if these polynomials were tropical ones, the Taylor expansion would really decompose $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps: just like in mathematical analysis, unbounded application could be seen
+But terms with bounded applications correspond to \emph{polynomials}, usually non-Lipschitz functions. 
+Yet, if these polynomials were tropical, the Taylor expansion would really decompose $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps: just like in mathematical analysis, unbounded application could be seen
 as a limit of \emph{more and more sensitive} operations.
@@
 For $X=Y=\set{*}$, we get usual tLs of one variable.
-Instead, if the support $\C F:=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ of $f$ is \emph{finite}, we get maps of shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, i.e.\ tropical polynomials in possibly infinitely many variables.
+Instead, if the support $\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ of $f$ is \emph{finite}, we get %maps of shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, i.e.\ 
+tropical polynomials in possibly infinitely many variables.
 In this section we study the properties of generalised tLs, from the point of view of mathematical analysis.
@@
 Also, while the derivative of $\varphi$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
-This phenomenon is reminiscent of [Example 7, \cite{Ehrhard2005}],
-%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
-which actually motivated our first investigations.
-In fact, these properties are shared by all tLs with \emph{finitely} many variables, as shown by the following result.
-
-Remark that $!\set{1,\dots,k}$ can be identified with $\N^k$.
-So the matrix of a tLs $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one variable $f(x)$ in output) can be given as a $\matr f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, where $nx$ is the scalar product.
+This phenomenon is reminiscent of [Example 7, \cite{Ehrhard2005}].
+%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019 which actually motivated our first investigations.
+In fact, these properties are shared by all tLs with \emph{finitely} many variables, as shown by the following result (we identify $!\set{1,\dots,k}$ with $\N^k$, so the matrix of a tLs $f$ with variables $x=x_1,\dots,x_k$ can be given as a $\matr f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, where $nx$ is the scalar product).
 
@@
 \begin{proof}[Proof sketch]
-Let $\C F_\epsilon$ be the set of $n\in\N^{k}$ such that 
+Take $\C F_\epsilon$ the set of $n\in\N^{k}$ such that 
 $\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
-We can show that it is finite and does the job.
 \end{proof}
@@
 \end{proposition}
-\begin{proof}[Proof sketch]
-It follows from the above lines about tropical linear functions, plus the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $!
+%\begin{proof}[Proof sketch]
+%It follows from the above lines about tropical linear functions, plus the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $!
 x$ to finite multisets of degree $\leq n$.%
-%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
-%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
-%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
-\end{proof}
+%%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
+%%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
+%%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
+%\end{proof}
 This result is perfectly analogous to what happens in the metric models discussed in Section \ref{section2}, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
@@
 \end{theorem}
-\begin{proof}[Proof sketch]
 The core of the proof is a convex analysis argument showing that an arbitrary function $f:\Lawv^X\to\Lawv$ which is non-decreasing, concave and continuous, must be locally Lipschitz.
-\end{proof}
 
@@
 
+Since algebraic and geometric properties in tropical mathematics are usually more tractable from a computational point of view, in several well-known applications (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}) one starts from a given modelised phenomenon, typically expressed by some polynomial function $f$, and studies
+what of its properties can be deduced from the \emph{tropicalization} $\trop f$ of $f$.
+Let us illustrate a few possible direction following a similar pattern, where of course our modelled phenomenon are programs.%: for a $\lambda$-term $M$, which can be expressed in $\LREL_!$ as a tropical polynomial or power series $\model M$, and we investigate what quantitative properties of $M$ can be deduced from it.
+In particular, it is useful to think of $\model M = \trop f$ for some polynomial or power series $f$.
+We sketch here only one example, regarding probabilities, but we are currently working on other setting including differential privacy, finiteness spaces and best case analysis.
+
+As a toy example, consider a probabilistic extension $\STLC_\oplus$ of $\STLC$, with a new ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the usual rules.
+%\[ \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}  \qquad\qquad \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}\]
+We add reduction rules:
+$M\oplus_p N \to pM$ and $M\oplus_p N \to (1-p)N$,
+so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
+$\STLC_\oplus$ can be seen as a fragment of the PCF in \cite{Manzo2013}.
+
+Consider now the following term
+$
+ M:=(\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true))
+$
+of type $\bool$.
+Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, reading $0$ as ``left'' and $1$ as ``right'' in the tree structure of $M$.
+%The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
+Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
+$P_{00}(p,q)\true$, 
+$P_{100}(p,q)\true$, 
+$P_{111}(p,q)\true$, 
+$P_{01}(p,q)\false$, 
+$P_{110}(p,q)\false$,
+$P_{101}(p,q)\false$,
+where the $P$'s are the following monomials in $p,q$:
+$P_{00}(p,q):=p^2$,
+$P_{100}(p,q):=qp^2$,
+$P_{111}(p,q):=q^3$,
+$P_{01}(p,q):=pq$,
+$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
+They correspond to the respective reduction path from $M$ to the normal term of the same address.
+$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$.%after all the coin tossings in $M$ are performed.
+Thinking of $p,q$ as parameters, $P_{\omega}$ is the \emph{likelihood function} of the event $\omega$.
+The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after all the tossings (similarly a $Q_{0}(p,q)$ for $\false$).
+This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}.
+A typical question in this case would be: \emph{knowing that $M$ produced $\true$, which is the choice of the parameters $p,q$ that maximizes the probability that %the produced occurrence of $\true$ was a fixed $\true_{\omega_0}$ (i.e.
+that
+among the paths leading to $\true$, the one taken was $\omega_0$?}
+%These are questions much in the style of those asked in {\color{red}Ref?} in the framework of {\color{red}??}.
+Answering it, amounts at solving a optimisation problem related to $P_{\omega}, Q_\omega$, which is more easily solved by 
+passing to the tropicalizations $\trop P_{\omega},\trop Q_{\omega}$. 
+We are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
+$\max_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$.
+This last condition is equivalent to ask that
+$\min\limits_{\omega\in\set{00,100,111}} \, -\log P_\omega(x,y) = -\log P_{\omega_0}(x,y)$,
+i.e.\
+$(\trop Q_1)( -\log x, -\log y) = (\trop P_{\omega_0})( -\log x, -\log y)$.
+%Passing through $\trop $ %P_\omega, \trop Q_\omega$ makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2015}).
+%the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.
+
+\begin{remark}
+ By adapting [Section IV, \cite{Manzo2013}], it can be seen that $\LREL_!$ is a model of $\STLC_\oplus$.
+ In particular, if we set $\model \bool:=\set{0,1}$, our running example $M$ is interpreted as $\model M\in\Lawv^{\set{0,1}}$ giving the tropicalised probabilities: $\model M_0=(\trop Q_0)(p,1-p)$, $\model M_1=(\trop Q_1)(p,1-p)$.
+\end{remark}
+
+%\begin{example}\label{ex:study}
+For our $M$, we have $\trop Q_1(x,y)=\min\set{2x,y+2x,3y}$ and $\trop Q_0(x,y)=\min\set{x+y,2y+x}$.
+Studying $\trop Q_1$, we see that $\trop Q_1(x,y)=3y$ iff $y\leq \frac{2}{3}x$, and it coincides with $2x$ otherwise.
+Remembering that $3y=P_{111}(x,y)$, we can now solve our optimisation problem above for $\omega_0=111$:
+via the substitution $x:=-\log p$, $y:=-\log (1-p)$, it is equivalent to $-\log (1-p)\leq -\frac{2}{3}\log p$, i.e.\ $1-p\geq p^{\frac{2}{3}}$.
+This means that, for $p\in[0,1]$ s.t.\ $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), the most likely occurrence of $\true$ to obtain, knowing that $M$ sampled $\true$ in its normal form, is $\true_{111}$.
+Remembering that $2x=P_{00}(x,y)$, for the other values of $p$ (for example, $p=\frac{1}{2}$), the most likely $\true$ to be sampled is the occurrence $\true_{00}$.
+We have thus answered our question.
+%\end{example}
+Also, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a root of $\trop Q_1$ or of $\trop Q_0$ provide the values of the bias of $\oplus_p$ for which there are at least two different \emph{equiprobable} paths from $M$ to its normal form.
+Yet, we do not have a full understanding of the role of roots in this setting.
+% Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log p,-\log(1-p))$ is a tropical root of, say, $\trop Q_1$ are the values of the bias of $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are smapled by $M$ during its tossings to normal form, knowing that some $\true$ was sampled.
+
 \section{Getting rid of bases}
 
-Remember that a $\Lawv$-module is a triple $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$.% (notice that this is indeed the usual notion of module over the Lawvere quantale $\Lawv$).
+We take $\Lawv$-modules to be triples $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$.% (notice that this is indeed the usual notion of module over the Lawvere quantale $\Lawv$).
 $\Lawv$ is the most basic example of $\Lawv$-module.
@@
 \end{theorem}
-That is, one does not need to stick to a base in order to interpret types/programs.
-This differential extends the one for $\LREL_{!}$;
-in order to define the exponential $!$ over $\Mod$ we use the well-known recipe from \cite{Mellies2018, Manzo2013}, consisting in defining a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$, and then define $!$ as the product of the symmetric algebras.
+In order to define the exponential $!$ over $\Mod$ we use the well-known recipe from \cite{Mellies2018, Manzo2013}, consisting in defining a symmetric algebra %$\Sym_{n}(M)$
+as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$, and then define $!$ as the product of the symmetric algebras.
 
@@
 
-%\bibliographystyle{plain}
-%\bibliography{tropical.bib}
+\bibliographystyle{plain}
+\bibliography{tropical.bib}
 

### 2023-05-14 section5.tex
Committ
Ho corretto un minierrore nella versione LICS, non si sa mai la usiamo per le future versioni e ci scordavamo di correggerlo

@@
 
-Here we follow a similar patter: we consider a program $M$, which can be expressed in the form of a polynomial or a power series $f$, and we 
+Here we follow a similar pattern: we consider a program $M$, which can be expressed in the form of a polynomial or a power series $f$, and we 
 investigate what quantitative properties of $M$ can be deduced from the properties of $\trop f$, that will indeed coincide with the interpretation of $M$ in $\LREL_{!}$.
@@
 
-The possibility of using the relational model over the tropical semi-ring for ``best case'' resource analysis has already been explored in \cite{Manzo2013}.
+The possibility of using the relational model over the tropical semiring for ``best case'' resource analysis has already been explored in \cite{Manzo2013}.
 Notably, they considered an interpretation of a language for $\B{PCF}$ with non-deterministic choice in which each $\lambda$-abstraction and each occurrence of the fixpoint operator $Y$ is assigned a ``weight'' 1, and showed that for any program $M$ of type $\B{nat}$,
@@
 and similarly for $\TT C[N]$.
-It is here worth recalling that, for $\STDLC$, a well-known result \cite{difflambda} is that the Taylor expansion of a closed application $MN$ is always finite, although its non-zero coefficients may be arbitrarily high.
+It is here worth recalling that, for $\STDLC$, a well-known result \cite{difflambda} is that the Taylor expansion of a closed application $MN$ is always \emph{finite}, although its non-zero coefficients may be arbitrarily high.
 Notably, these observations suggest to study tropical versions of \emph{finiteness spaces} \cite{Ehrhard2005}, 
 a variant of the relational semantics modeling strongly normalizing programs via \emph{finite} power series.
-We mention this point in Section~\ref{section8}
+%We mention this point in Section~\ref{section8}.
 
@@
 
-Let us consider in this subsection a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
+Let us consider a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
 we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the rules:
@@
 \[
- (\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\true\oplus_p\false))
+ (\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true))
 \]
@@
 \end{equation}
-In both cases, passing through $\trop$ %P_\omega, \trop Q_\omega$ 
+In both cases, passing through $\trop $ %P_\omega, \trop Q_\omega$ 
 makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2015}).

### 2023-05-15 TLLA23/BarbaPisto.tex
commit

@@
 
-\title{Tropical methods in Lambda-Calculus}
+\title{Tropical Mathematics and the Lambda-Calculus}
 \author{Davide Barbarossa
@@
 \institute{\qquad\qquad\qquad DISI, Universit\`a di Bologna}
-\email{\qquad\qquad\qquad paolo.pistone@unibo.it}
+\email{\qquad\qquad\qquad paolo.pistone2@unibo.it}
 }
@@
 
-Two different quantitative approaches have received considerable attention from the programming language community in recent years: %(e.g.\ \cite{decarvalho2018, Accattoli2022}, \cite{Breuvart2018, PistoneLICS2022}, \cite{Barthe_2012} \cite{Reed2010, dallago}, \cite{Ehrhard2011, Staton2017}, \cite{difflambda}), on the one hand one there is
-the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk} is based on the fact that when considering e.g.\ probabilistic %or approximate
-computation, %rather than asking whether two programs compute \emph{the same} function,
-it is natural to ask whether two programs compute functions which %do not differ \emph{too much}
-approximate each other, instead of equality of functions.
-This led to denotational frameworks in which types are endowed with a metric \cite{Reed2010},\cite{Bonchi2018}, \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
-On the other hand, there is the approach based on \emph{differential} \cite{difflambda}, \cite{difflambda}, \cite{Manzo2013, Breuvart2018, PistoneLICS2022} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
-This led to syntactic or denotational frameworks in which one can define a \emph{Taylor expansion} of programs.
-In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
-In metric semantics, linear programs correspond to \emph{non-expansive} maps, i.e.\ that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret programs with a fixed duplication bound as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
+In recent years, more and more interest in the programming language community has been directed towards the study of quantitative properties of programs like computing the number of computation steps or convergence probabilities, as opposed to purely qualitative properties like termination or program equivalence.
+In particular, two different quantitative approaches have received considerable attention from the programming language community.
+On the one hand, the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk} is based on the observation that probabilistic or numerical algorithms are not though to compute a target function $f$ \emph{exactly}, but only in an approximate way; hence, one is usually interested to know whether the computed function is \emph{not too far} from $f$, rather than equal to $f$.
+This led to study denotational frameworks in which types are endowed with metrics measuring similarities in program behavior \cite{Reed2010},\cite{Bonchi2018}, \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
+On the other hand, there is the approach based on \emph{differential} \cite{difflambda}, \cite{difflambda}, \cite{Manzo2013, Breuvart2018, PistoneLICS2022} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the \emph{relational semantics}\cite{Manzo2012, Manzo2013, dill} and \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
+This led to study syntactic or denotational frameworks in which one can define a \emph{Taylor expansion} of programs.
+
+In both approaches a crucial role is played by the notion of \emph{linearity}, in the sense of linear logic, i.e.~of using inputs exactly once.
+In metric semantics, linear programs correspond to \emph{non-expansive} functions, i.e.\ maps that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret programs with a fixed duplication bound as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
 By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs gives rise to \emph{power series}.
 
-At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
+The starting observation of this work is that, at a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
 in metric models $n$ times duplication results in a $n$-Lipschitz \emph{linear} function $n\cdot x$, while in differential models this results in a non-Lipschitz \emph{polynomial} function $x^{n}$.
-Our main observation is that 
-this gap is overcome once we interpret these functions in the framework of tropical mathematics where, for instance, $x^{n}$ precisely reads as $n\cdot x$.
+ At the same time, 
+this gap may be overcome once we interpret these functions in the framework of tropical mathematics where, for instance, $x^{n}$ precisely reads as $n\cdot x$.
+
 Tropical mathematics \cite{Simon} is a well established algebraic and geometrical framework, with tight connections with optimisation theory \cite{Sturmfelds}, where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
@@
 
+\section{The Tropical Semantics of Linear Logic }
+
 \paragraph{Tropical mathematics in a nutshell}
@@
 For example, $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$, plotted in Fig~\ref{fig:plot1}.
-A \emph{tropical root} of $\varphi$ is a point $x\in \Lawv$ where the minimum defining $\varphi$ is attained at least twice.%, i.e.\ where $\varphi$ is not differentiable, i.e.~where the slope of $\varphi$ changes.
+A \emph{tropical root} of $\varphi$ is a point $x\in \Lawv$ where the minimum defining $\varphi$ is attained at least twice.
+%, i.e.\ where $\varphi$ is not differentiable, i.e.~where the slope of $\varphi$ changes.
 E.g., the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, $i \leq n$.
@@
 Finally, for a polynomial/power series $f(x)=\sum_n a_nx^n$, one defines its \emph{tropicalization} $\trop f(\alpha):=\inf_n\set{-\log a_n+n\alpha}$.
-TLs may be non Lipschitz, and their study is less developed than that of tropical polynomials.
+TLs are in general not Lipschitz, and their study is less developed than that of tropical polynomials.
 
@@
  \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
-Furthermore, $!$ can be decomposed into a family of \emph{graded} exponentials $(!_n)_{n\in\N}$ turning $(\LREL,(!_n)_{n\in\N})$ in a model for $\BSTLC$. 
-Finally, $\LREL_!$ can be equipped with a \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined by: $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise.
-%(where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
-This turns $(\LREL_!,D)$ into a CC$\partial$C, i.e.\ a model of the differential $\lambda$-calculus.
-Remember that the (quantitative) Taylor expansion of an ordinary $\lambda$-term $M$ is an inductively defined series $\Te{M}$ of differential $\lambda$-terms, the only non-trivial case being $\Te{PQ} :=  \sum_{k=0}^{\infty} (\Der^{k}[P] \cdot Q^{k})0$.
-It can be seen that the morphisms of $\LREL_!$ can always be Taylor expanded, and the series interpreting in $\LREL_!$ the Taylor expansion of a $\STLC$-term $M$, converges to the interpretation of $M$.
-The Taylor formula decomposes an unbounded application as a limit of bounded ones.
-So one might well ask whether this formula interprets a $\lambda$-term 
-as a limit of Lipschitz maps and thus bridging, in a sense, the metric and differential approaches previously mentioned.  
-%Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
-But terms with bounded applications correspond to \emph{polynomials}, usually non-Lipschitz functions. 
-Yet, if these polynomials were tropical, the Taylor expansion would really decompose $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps: just like in mathematical analysis, unbounded application could be seen
-as a limit of \emph{more and more sensitive} operations.
-This is precisely the content of our Corollary~\ref{cor:main}(3).
-
-\section{Tropical Laurent Series}
-
  As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a \emph{linear} map $\Lawv^{!X}\to\Lawv^Y$, but we can also ``express it in the base $X$'' and see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting
@@
 tropical polynomials in possibly infinitely many variables.
-In this section we study the properties of generalised tLs, from the point of view of mathematical analysis.
+Furthermore, $!$ can be decomposed into a family of \emph{graded} exponentials $(!_n)_{n\in\N}$ turning $(\LREL,(!_n)_{n\in\N})$ in a model for a $\mathit{BLL}$-style simply typed $\lambda$-calculus, that we call $\BSTLC$ (see e.g.~\cite{Brunel2014}).
+
+Finally, $\LREL_!$ can be equipped with a \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined by: $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise.
+%(where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
+This turns $(\LREL_!,D)$ into a CC$\partial$C, i.e.\ a model of the differential $\lambda$-calculus.
+Remember that the (quantitative) Taylor expansion of an ordinary $\lambda$-term $M$ is an inductively defined series $\Te{M}$ of differential $\lambda$-terms, the only non-trivial case being $\Te{PQ} :=  \sum_{k=0}^{\infty} (\Der^{k}[P] \cdot Q^{k})0$.
+It can be seen that the morphisms of $\LREL_!$ can always be Taylor expanded, and the series interpreting in $\LREL_!$ the Taylor expansion of a $\STLC$-term $M$, converges to the interpretation of $M$.
+
+%
+%
+%The Taylor formula decomposes an unbounded application as a limit of bounded ones.
+%So one might well ask whether this formula interprets a $\lambda$-term 
+%as a limit of Lipschitz maps and thus bridging, in a sense, the metric and differential approaches previously mentioned.  
+%%Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
+%But terms with bounded applications correspond to \emph{polynomials}, usually non-Lipschitz functions. 
+%Yet, if these polynomials were tropical, the Taylor expansion would really decompose $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps: just like in mathematical analysis, unbounded application could be seen
+%as a limit of \emph{more and more sensitive} operations.
+%This is precisely the content of our Corollary~\ref{cor:main}(3).
+
+\section{Tropical Laurent Series Yield a Metric Semantics of $\STLC$}
+
+The main goal of this section is to show that the interpretation of $\STLC$ in the tropical relational model yields a metric semantics, where the spaces $\Lawv^{X}$ are endowed with the $\|\cdot\|_{\infty}$-norm metric, and programs are interpreted as \emph{locally Lipschitz} maps:
+\begin{theorem}\label{cor:main}
+For any $\lambda$-term $M$:
+\begin{varenumerate}
+\item if $\Gamma \vdash_{\BSTLC} M:A$, then $\model M^!:\model\Gamma \to \model A$ is a \emph{Lipschitz} map.
+\item if $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is a \emph{locally Lipschitz} map.
+Moreover, the Taylor expansion decomposes $\model M^!$ into an $\inf$ of \emph{Lipschitz}~maps.
+\end{varenumerate}
+\end{theorem}
+Recall that the Taylor formula decomposes an unbounded application as a limit of bounded ones; the result above presents this as a decomposition of an higher-order program as a limit of Lipschitz maps, and provides thus a bridge between the metric and the differential approaches.
+
+The proof of the result above passes through the study of topological and metric properties of tLs. 
+We first studied tLs with respect to the topology induced by the $\|\cdot\|_{\infty}$-norm.
+%
+%In this section we study the properties of generalised tLs, from the point of view of mathematical analysis.
 %In the following we will also refer to them as tLs.
@@
 %\begin{remark}
+
+\begin{proposition}\label{prop:nondecr+conc}
+ Any tLs $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order, and continuous w.r.t.~the norm $\|\cdot\|_{\infty}$.
+\end{proposition}
+
+%
+%The norm $\norm{\cdot}_\infty$ on $\BB R$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$.
+%We will show that tLs satisfy suitable Lipschitz properties  w.r.t.~these metrics.
+%\emph{Wlog}, we will often take $Y=\set{*}$.
+
+Let us now consider metric properties. First, it can be seen that all tropical \emph{linear} functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive. 
+%\begin{proof}[Proof sketch]
+%Using the fact that $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
+%the problem reduces to checking that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
+This result shows that, in analogy with that happens in usual metric semantics, linear programs are interpreted by non-expansive functions. 
+%\begin{proof}
+%Using $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
+%first observe that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$; we now have
+%$|f(\B x)_{b}-f(\B y)_{b}| \leq |(\inf_{a\in X}\matr f_{a,b}-\B x_{a})-(\inf_{a\in X}\matr f_{a,b}-\B y_{a})| \leq
+%\sup_{a\in X}|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})|\leq 
+% \| \B x-\B y\|_{\infty}$.
+%\end{proof}
+More generally, linear maps with bounded exponentials yield Lipschitz maps:
+\begin{proposition}\label{prop:boundedlip}
+If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a bounded matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
+\end{proposition}
+%\begin{proof}[Proof sketch]
+%It follows from the above lines about tropical linear functions, plus the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $!
+x$ to finite multisets of degree $\leq n$.%
+%%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
+%%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
+%%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
+%\end{proof}
+This result is perfectly analogous to what happens in the metric models recalled in the introduction, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
+It also entails that for any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$, $\varphi$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
 
@@
 \end{tikzpicture}
-\caption{Plot of the tropical polynomials $\varphi_1,\varphi_2,\varphi_3,\varphi_4$ (from top to bottom), and of their limit tLs $\varphi$ (in violet).}
+\caption{\small Plot of the tropical polynomials $\varphi_1,\varphi_2,\varphi_3,\varphi_4$ (from top to bottom), and of their limit tLs $\varphi$ (in violet).}
 \label{fig:plot1}
@@
 
-\begin{proposition}\label{prop:nondecr+conc}
- Any tLs $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
-\end{proposition}
-
-%In Example~\ref{ex:famous_ex}, 
-Look at Fig~\ref{fig:plot1}: it appears that $\varphi$ behaves \emph{locally} like the polynomials $\varphi_{n}$. 
+Let us now look at what happens with tLs, i.e.~when considering the full exponential $!$.
+First, we consider the interesting case of tLs with \emph{finitely many variables}. 
+Let us start by an example, shown in Fig~\ref{fig:plot1}: the 1-variable tLs $\varphi(x)=\inf_{i}\{x+2^{-i}\}$ behaves \emph{locally} like the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{x+2^{-i}\}$. 
 %In particular, for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$.
@@
 %Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019 which actually motivated our first investigations.
-In fact, these properties are shared by all tLs with \emph{finitely} many variables, as shown by the following result (we identify $!\set{1,\dots,k}$ with $\N^k$, so the matrix of a tLs $f$ with variables $x=x_1,\dots,x_k$ can be given as a $\matr f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, where $nx$ is the scalar product).
+In fact, these properties are shared by all tLs with {finitely} many variables, as shown by the following result (we identify $!\set{1,\dots,k}$ with $\N^k$, so the matrix of a tLs $f$ with variables $x=x_1,\dots,x_k$ can be given as a $\matr f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, where $nx$ is the scalar product).
 
@@
 \end{theorem}
-\begin{proof}[Proof sketch]
-Take $\C F_\epsilon$ the set of $n\in\N^{k}$ such that 
-$\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
-\end{proof}
-
-The norm $\norm{\cdot}_\infty$ on $\BB R$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$.
-We will show that tLs satisfy suitable Lipschitz properties  w.r.t.~these metrics.
-\emph{Wlog}, we will often take $Y=\set{*}$.
-
-First, it can be seen that all tropical \emph{linear} functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive.  
 %\begin{proof}[Proof sketch]
-%Using the fact that $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
-%the problem reduces to checking that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
-This result shows that, in analogy with that happens in usual metric semantics, linear programs are interpreted by non-expansive functions. 
-%\begin{proof}
-%Using $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
-%first observe that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$; we now have
-%$|f(\B x)_{b}-f(\B y)_{b}| \leq |(\inf_{a\in X}\matr f_{a,b}-\B x_{a})-(\inf_{a\in X}\matr f_{a,b}-\B y_{a})| \leq
-%\sup_{a\in X}|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})|\leq 
-% \| \B x-\B y\|_{\infty}$.
+%Take $\C F_\epsilon$ the set of $n\in\N^{k}$ such that 
+%$\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
 %\end{proof}
-Let us now consider the case of bounded exponentials:
-\begin{proposition}\label{prop:boundedlip}
-If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a bounded matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
-\end{proposition}
-%\begin{proof}[Proof sketch]
-%It follows from the above lines about tropical linear functions, plus the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $!
-x$ to finite multisets of degree $\leq n$.%
-%%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
-%%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
-%%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
-%\end{proof}
-This result is perfectly analogous to what happens in the metric models discussed in Section \ref{section2}, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
-It also entails that for any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$, $\varphi$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
-
-Let us now look at what happens with tLs, i.e.~when considering the full exponential $!$.
-As consequence of Theorem~\ref{theorem:fepsilon}, the tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $\BB R_{>0}$.
-Actually, we can prove a more general statement, also covering the infinitary case.
+The result above suggests that the local behavior of tLs with finitely many variables can be studied with the tools of tropical geometry (e.g.~tropical roots, Newton polygones), as suggested in the next section.
+Moreover, a consequence of Theorem~\ref{theorem:fepsilon} is that all tLs with {finitely many} variables are always \emph{locally} Lipschitz on $\BB R_{>0}$.
+The following result extends this property to all tLs, also covering the case with infinitely many variables.
 
@@
 One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
-Moreover, the Taylor formula enjoyed by $\LREL_!$ morphisms, yields a ``tropical'' Taylor formula for tLs of the form $f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}$.
+Finally, the Taylor formula enjoyed by $\LREL_!$ morphisms, yields a ``tropical'' Taylor formula for tLs of the form $f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}$.
 
-The above results translate into the following facts about the interpretation of higher-order programs:
-\begin{corollary}\label{cor:main}
-For any $\lambda$-term $M$:
-\begin{enumerate}
-\item if $\Gamma \vdash_{\BSTLC} M:A$, then $\model M^!:\model\Gamma \to \model A$ is a \emph{Lipschitz} map.
-\item if $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is a \emph{locally Lipschitz} map.
-\item  if $\Gamma \vdash_{\STLC}M:A$, then its Taylor exp.\ decomposes $\model M^!: \model\Gamma \to \model A$ into an $\inf$ of \emph{Lipschitz}~maps.
-\end{enumerate}
-\end{corollary}
-
-\section{Relations to quantitative properties}
+\section{Tropical Semantics as Quantitative Semantics: Computing Likelihoods of Reduction Paths}
 
@@
 what of its properties can be deduced from the \emph{tropicalization} $\trop f$ of $f$.
-Let us illustrate a few possible direction following a similar pattern, where of course our modelled phenomenon are programs.%: for a $\lambda$-term $M$, which can be expressed in $\LREL_!$ as a tropical polynomial or power series $\model M$, and we investigate what quantitative properties of $M$ can be deduced from it.
+This idea suggests several natural directions in which the tropical semantics of a higher-order program could be used to deduce properties which can be expressed as an optimization problem.%: for a $\lambda$-term $M$, which can be expressed in $\LREL_!$ as a tropical polynomial or power series $\model M$, and we investigate what quantitative properties of $M$ can be deduced from it.
 In particular, it is useful to think of $\model M = \trop f$ for some polynomial or power series $f$.
-We sketch here only one example, regarding probabilities, but we are currently working on other setting including differential privacy, finiteness spaces and best case analysis.
+We sketch here only one example, regarding probabilities, but we are currently working on other directions including \emph{differential privacy} and \emph{best case analysis}.
 
@@
 
-\section{Getting rid of bases}
+\section{Getting rid of bases: $\Lawv$-modules as generalized metric spaces}
 
-We take $\Lawv$-modules to be triples $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$.% (notice that this is indeed the usual notion of module over the Lawvere quantale $\Lawv$).
-$\Lawv$ is the most basic example of $\Lawv$-module.
- Other basic examples of $\Lawv$-modules are the sets $\Lawv^{X}$, with order and action defined pointwise. 
-A $\Lawv$-module homomorphism is a map $f:M\to N$ commuting with both joins and the $\Lawv$-action. We let $\Mod$ indicate the category of $\Lawv$-modules and their homomorphisms.
+As we have seen, tropical semantics provides a viewpoint in which metric and differential properties are somehow unified.
+This approach can be made more abstract, thanks to a fundamental categorical correspondence between tropical linear algebra (i.e.~the study of quantale modules over $\Lawv$) and the theory of Lawvere's \emph{generalized metric spaces} (see \cite{Fuji, Stubbe2006}).
 
-Observe that the category $\LREL$ is (equivalent to) a subcategory of $\Lawv\SF{Mod}$.
-While $\LREL$ corresponds to considering modules (the $\LREL^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $\Lawv\SF{Mod}$ corresponds to considering semimodules in abstract, without fixing a base.
+A quantale module over $\Lawv$, shortly a \emph{$\Lawv$-module} (not to be confused with a module on $\Lawv$ \emph{as a semiring}), is a triple  $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$.
+%Notice that this defines the usual notion of module over the Lawvere quantale $\Lawv$, and this does \emph{not} coincide with the modules on $\Lawv$ \emph{as a semiring} (which are indeed well-studied and well-known to provide a $CC\partial C$ {\color{red}LEMAYYYY}).
+The most basic example of $\Lawv$-modules are given by the spaces $\Lawv^{X}$, with order and action defined pointwise.
 
@@
 Indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories} (in short, a $\Lawv$-category) \cite{Lawvere1973, Hofmann2014, Stubbe2014}.
+Many topological properties of metric spaces translate in this way into purely categorical ones.
+In particular, $\Lawv$-enriched categories which are \emph{cocomplete} in the (enriched) categorical sense (i.e.~all weighted colimits exist) satisfy usual metric completeness properties (e.g.~Cauchy-completeness or Isbell-completeness).
+
+Now, any $\Lawv$-module $M$ can be endowed with a metric 
+$M(x,y)=\inf\{\epsilon \mid \epsilon \star x \geq y\}$, yielding a cocomplete $\Lawv$-enriched category and, conversely, any cocomplete $\Lawv$-enriched category $X$ has a $\Lawv$-module structure with order $x\leq y$ when $X(y,x)=0$, and monoidal operation defined via a suitable weighted colimit.
+This induces an equivalence between the SMCC $\Mod$ of $\Lawv$-modules and their homomorphisms and the SMCC $\GMet$ of cocomplete $\Lawv$-enriched categories and cocontinuous functors (notice that functoriality in $\GMet$ is just non-expansiveness).
+
+%
+%A $\Lawv$-module homomorphism is a map $f:M\to N$ commuting with both joins and the $\Lawv$-action. We let $\Mod$ indicate the category of $\Lawv$-modules and their homomorphisms. 
+%
+%Observe that the category $\LREL$ is (equivalent to) a subcategory of $\Lawv\SF{Mod}$.
+%While $\LREL$ corresponds to considering modules (the $\LREL^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $\Lawv\SF{Mod}$ corresponds to considering semimodules in abstract, without fixing a base.
+
 %Indeed, spelling out the definition, a $\Lawv$-enriched category is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$ satisfying 
@@
 %A $\Lawv$-category is \emph{skeletal} \cite{Stubbe2014} when $X(x,y)=0$ implies $x=y$, and \emph{symmetric} when it coincides with its opposite category $X^{\op}(x,y):=X(y,x)$, i.e.~when $X(x,y)=X(y,x)$.
-A metric space, in the usual sense, is just a particular case of $\Lawv$-category.
-%Notice that any $\Lawv$-category $X$ induces a skeletal category $X^{\sk}$ by quotienting points under $X(x,y)=0$, and a symmetric one by letting $X^{\sym}(x,y)=\max\{X(x,y),X^{\op}(x,y)\}$.
-Remark that $\Lawv$ has a canonical $\Lawv$-enriched structure given by $\Lawv(r,s)=s \overset{\cdot}{-} r$ (where ``$ \overset{\cdot}{-}$'' indicates truncated subtraction), 
-and the Euclidean distance coincides with its symmetrization $\Lawv^{\sym}(x,y):=\max\{\Lawv(x,y),\Lawv^{\op}(x,y)\}$.
+%A metric space, in the usual sense, is just a particular case of $\Lawv$-category.
+%%Notice that any $\Lawv$-category $X$ induces a skeletal category $X^{\sk}$ by quotienting points under $X(x,y)=0$, and a symmetric one by letting $X^{\sym}(x,y)=\max\{X(x,y),X^{\op}(x,y)\}$.
+%Remark that $\Lawv$ has a canonical $\Lawv$-enriched structure given by $\Lawv(r,s)=s \overset{\cdot}{-} r$ (where ``$ \overset{\cdot}{-}$'' indicates truncated subtraction), 
+%and the Euclidean distance coincides with its symmetrization $\Lawv^{\sym}(x,y):=\max\{\Lawv(x,y),\Lawv^{\op}(x,y)\}$.
 
 %A $\Lawv$-enriched \emph{functor} between $\Lawv$-categories $X,Y$ is just a non-expansive map $f:X\to Y$, as functoriality reads as $Y(f(x),f(y))\leq X(x,y)$.
-Now, by considering distributors, one defines the notion of \emph{continuous} functor $f:X\to Y$ as one that commutes with all existing weighted colimits in $X$.
-A $\Lawv$-enriched category $X$ is said \emph{categorically complete} (or just \emph{complete}) if all weighted colimits over $X$ exist. 
-We let $\GMet$ indicate the category of complete and skeletal $\Lawv$-categories and continuous functors.
-We mention this notions because it can be shown (~\cite{Stubbe2006}) that, actually, $\Mod$ and $\GMet$ are isomorphic categories.
-This is interesting for us because it suggests to transport the metric point of view of $\GMet$ on our modules in $\Mod$.
-Actually, we can even show that:
+%Now, by considering distributors, one defines the notion of \emph{continuous} functor $f:X\to Y$ as one that commutes with all existing weighted colimits in $X$.
+%A $\Lawv$-enriched category $X$ is said \emph{categorically complete} (or just \emph{complete}) if all weighted colimits over $X$ exist. 
+%We let $\GMet$ indicate the category of complete and skeletal $\Lawv$-categories and continuous functors.
+%We mention this notions because it can be shown (~\cite{Stubbe2006}) that, actually, $\Mod$ and $\GMet$ are isomorphic categories.
+
+Our contribution was, first, to observe that the SMCC structure of $\LREL$ lifts to that of the more general category $\Mod\equiv\GMet$ and, secondly, to define an exponential $!$ in $\Mod\equiv\GMet$, using a well-known recipe from \cite{Mellies2018, Manzo2013} based on the construction of symmetric algebras, and show that in this way also the exponential and differential structure of $\LREL_{!}$ lifts to this more general setting:
+%
+%This is interesting for us because it suggests to transport the metric point of view of $\GMet$ onto our modules in $\Mod$.
+%Actually, we can even show that:
 \begin{theorem}
@@
 \end{theorem}
-In order to define the exponential $!$ over $\Mod$ we use the well-known recipe from \cite{Mellies2018, Manzo2013}, consisting in defining a symmetric algebra %$\Sym_{n}(M)$
-as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$, and then define $!$ as the product of the symmetric algebras.
+%In order to ,, consisting in defining a symmetric algebra %$\Sym_{n}(M)$
+%as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$, and then define $!$ as the product of the symmetric algebras.
 

### 2023-05-15 TLLA23/BarbaPisto.tex
Committ

@@
 In particular, two different quantitative approaches have received considerable attention from the programming language community.
-On the one hand, the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk} is based on the observation that probabilistic or numerical algorithms are not though to compute a target function $f$ \emph{exactly}, but only in an approximate way; hence, one is usually interested to know whether the computed function is \emph{not too far} from $f$, rather than equal to $f$.
+On the one hand, the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk} is based on the observation that probabilistic or numerical algorithms are not thought to compute a target function $f$ \emph{exactly}, but only in an approximate way.
+%hence, one is usually interested to know whether the computed function is \emph{not too far} from $f$, rather than equal to $f$.
 This led to study denotational frameworks in which types are endowed with metrics measuring similarities in program behavior \cite{Reed2010},\cite{Bonchi2018}, \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
@@
 $
-\trop f(\alpha,\beta)=\min\{2\alpha, \alpha+2\beta, 3\beta\}
+\trop f(x,y)=\min\{2x, x+2y, 3y\}
 $.
@@
 This coincides with the \emph{Lawvere quantale} $\Lawv$ \cite{Hofmann2014, Stubbe2014}, i.e.\ $[0,\infty]$ with order $\geq$ and usual $+$ as the monoid action. 
-$\Lawv$ is at the heart of the categorical study of metric spaces initiated by Lawvere \cite{Lawvere1973}, and we will take this point of view in the last section.
+$\Lawv$ is at the heart of the categorical study of metric spaces initiated by Lawvere \cite{Lawvere1973}, a viewpoint we will take in the last section.
 A \emph{tropical polynomial} is a piece-wise linear function $\varphi:\Lawv\to \Lawv$ of the form $\varphi(x)=\min_{j=1,\dots,k}\{i_{j}x+\matr \varphi_{i_{j}}\}$, with $i_{j}\in\N$, $\matr \varphi_{i_{j}}\in\Lawv$ and $k$ finite.
@@
 %(\cite{Manzo2013} would call it $Q^\Pi$).
-As expected, $Q^X$ is a $Q$-semimodule and we can identify $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.
+As expected, $Q^X$ is a $Q$-module and we can identify $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.
 Taking $Q:=\Lawv$ we obtain the \emph{tropical weighted relational model} $\LREL$.
@@
 By applying known results (taken from \cite{Manzo2013},  {\color{red}CITARE LEMAYYYY}), one obtains that $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus:
-$\LREL$ is a SMCC, i.e.\ a model of the linear $\STLC$;
-the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$, where $!$ is the usual multiset comonad (so $!X$ is the set of finite multisets on $X$).
-Here, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ given by:
-$
+first, $\LREL$ is a SMCC, i.e.\ a model of the linear $\STLC$.
+Also, the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$, where $!$ is the usual multiset comonad (so $!X$ is the set of finite multisets on $X$).
+Here, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ given by
+$(s\circ_! t)_{\mu,c}:=
  \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
  \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
- As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a \emph{linear} map $\Lawv^{!X}\to\Lawv^Y$, but we can also ``express it in the base $X$'' and see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting
+ As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a \emph{linear} map $\Lawv^{!X}\to\Lawv^Y$.
+However, we can also ``express it in the base $X$'' and see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting
  $t^!(x):=t\circ_! x$.
@@
  where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
- These functions correspond then to generalised tLs, i.e.\ with possibly infinitely many variables (as many as the elements of $X$).
-For $X=Y=\set{*}$, we get usual tLs of one variable.
+ These functions correspond then to generalised tLs, i.e.\ with possibly infinitely many variables (as many as the elements of $X$) and for $X=Y=\set{*}$, we get usual tLs of one variable.
 Instead, if the support $\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ of $f$ is \emph{finite}, we get %maps of shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, i.e.\ 
@@
 
-Finally, $\LREL_!$ can be equipped with a \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined by: $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise.
+Finally, $\LREL_!$ can be equipped with a \emph{differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined by: $(Dt)_{\mu\oplus\rho,b}:=t_{\rho+\mu,b}$ if $\#\mu=1$ and $:=\infty$ otherwise.
 %(where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
@@
 
-\section{Tropical Laurent Series Yield a Metric Semantics of $\STLC$}
+\section{Tropical Laurent Series and Metric Semantics for $\STLC$}
 
@@
 \item if $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is a \emph{locally Lipschitz} map.
-Moreover, the Taylor expansion decomposes $\model M^!$ into an $\inf$ of \emph{Lipschitz}~maps.
+
+Moreover, the Taylor expansion $\Te M$ decomposes $\model M^!$ into an $\inf$ of \emph{Lipschitz}~maps.
 \end{varenumerate}
 \end{theorem}
-Recall that the Taylor formula decomposes an unbounded application as a limit of bounded ones; the result above presents this as a decomposition of an higher-order program as a limit of Lipschitz maps, and provides thus a bridge between the metric and the differential approaches.
+Recall that the syntactic Taylor expansion decomposes an unbounded application as a limit of bounded ones;
+the result above lifts this decomposition to a semantic level, presenting a higher-order program as limits of Lipschitz maps: it provides thus a bridge between the metric and the differential approaches.
 
@@
 %\end{proof}
-This result is perfectly analogous to what happens in the metric models recalled in the introduction, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
-It also entails that for any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$, $\varphi$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
+This result is perfectly analogous to what happens in the metric models recalled in the introduction.%, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
+It also entails that any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
 
@@
 
-Let us now look at what happens with tLs, i.e.~when considering the full exponential $!$.
-First, we consider the interesting case of tLs with \emph{finitely many variables}. 
+Let us now %look at what happens with tLs, i.e.~when considering the full exponential $!$.
+%First, we
+first consider the interesting case of tLs with \emph{finitely many variables}. 
 Let us start by an example, shown in Fig~\ref{fig:plot1}: the 1-variable tLs $\varphi(x)=\inf_{i}\{x+2^{-i}\}$ behaves \emph{locally} like the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{x+2^{-i}\}$. 
@@
 %It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
- $x=0$ we have that $\varphi(x=0)=\inf_{i\in\N} 2^{-i}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
+ $x=0$ we have that $\varphi(0)=\inf_{i\in\N} 2^{-i}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
 Also, while the derivative of $\varphi$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
-This phenomenon is reminiscent of [Example 7, \cite{Ehrhard2005}].
+This phenomenon is reminiscent of \cite[Example 7]{Ehrhard2005}.
 %Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019 which actually motivated our first investigations.
@@
 %\end{proof}
-The result above suggests that the local behavior of tLs with finitely many variables can be studied with the tools of tropical geometry (e.g.~tropical roots, Newton polygones), as suggested in the next section.
+The result above suggests that the local behavior of tLs with finitely many variables can be studied with the tools of tropical geometry (e.g.~tropical roots, Newton polygones).
+%as suggested in the next section.
 Moreover, a consequence of Theorem~\ref{theorem:fepsilon} is that all tLs with {finitely many} variables are always \emph{locally} Lipschitz on $\BB R_{>0}$.
@@
 
-Finally, the differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tLs $f:\Lawv^{X}\to \Lawv^{Y}$ into a tLs $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by $D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a}+x_{a}+\mu y\right\}$.
-One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
+Finally, let us look at the differential operator $D$ of $\LREL_{!}$.
+It translates into a differential operator $D_{!}$ turing a tLs $f:\Lawv^{X}\to \Lawv^{Y}$ into a tLs $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by $D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a}+x_{a}+\mu y\right\}$.
+One can check that, when $f$ is a tropical polynomial, $D_{!}f$ gives the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
 Finally, the Taylor formula enjoyed by $\LREL_!$ morphisms, yields a ``tropical'' Taylor formula for tLs of the form $f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}$.
 
-\section{Tropical Semantics as Quantitative Semantics: Computing Likelihoods of Reduction Paths}
+\section{Tropical Semantics and Quantitative Properties: Likelihoods of Reduction Paths}
 
-Since algebraic and geometric properties in tropical mathematics are usually more tractable from a computational point of view, in several well-known applications (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}) one starts from a given modelised phenomenon, typically expressed by some polynomial function $f$, and studies
+Since algebraic and geometric properties in tropical mathematics are usually more tractable from a computational point of view, in several well-known applications (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}) one starts from a given modelled phenomenon, typically expressed by some polynomial function $f$, and studies
 what of its properties can be deduced from the \emph{tropicalization} $\trop f$ of $f$.
@@
 $P_{110}(p,q)=P_{101}(p,q):=q^2p$.
-They correspond to the respective reduction path from $M$ to the normal term of the same address.
+They correspond to the respective reduction path from $M$ to the normal term of the given address.
 $P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$.%after all the coin tossings in $M$ are performed.
@@
 that
-among the paths leading to $\true$, the one taken was $\omega_0$?}
+among the paths leading to $\true$, the one taken was a fixed $\omega_0$?}
 %These are questions much in the style of those asked in {\color{red}Ref?} in the framework of {\color{red}??}.
-Answering it, amounts at solving a optimisation problem related to $P_{\omega}, Q_\omega$, which is more easily solved by 
-passing to the tropicalizations $\trop P_{\omega},\trop Q_{\omega}$. 
+Answering it, amounts at solving an optimisation problem related to $P_{\omega}, Q_\omega$, which is more easily solved via the tropicalizations $\trop P_{\omega},\trop Q_{\omega}$. 
 We are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
@@
 \begin{remark}
- By adapting [Section IV, \cite{Manzo2013}], it can be seen that $\LREL_!$ is a model of $\STLC_\oplus$.
+ By adapting \cite[Section IV]{Manzo2013}, it can be seen that $\LREL_!$ is a model of $\STLC_\oplus$.
  In particular, if we set $\model \bool:=\set{0,1}$, our running example $M$ is interpreted as $\model M\in\Lawv^{\set{0,1}}$ giving the tropicalised probabilities: $\model M_0=(\trop Q_0)(p,1-p)$, $\model M_1=(\trop Q_1)(p,1-p)$.
@@
 
-A quantale module over $\Lawv$, shortly a \emph{$\Lawv$-module} (not to be confused with a module on $\Lawv$ \emph{as a semiring}), is a triple  $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$.
+A quantale module over $\Lawv$, shortly a \emph{$\Lawv$-module}%(not to be confused with a module on $\Lawv$ \emph{as a semiring})
+, is a triple  $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$.
 %Notice that this defines the usual notion of module over the Lawvere quantale $\Lawv$, and this does \emph{not} coincide with the modules on $\Lawv$ \emph{as a semiring} (which are indeed well-studied and well-known to provide a $CC\partial C$ {\color{red}LEMAYYYY}).
 The most basic example of $\Lawv$-modules are given by the spaces $\Lawv^{X}$, with order and action defined pointwise.
+We see them as modules given together with a fixed base, $X$.
+Our Theorem~\ref{thm:no-base} says that one can actually give a semantic of high order programs without need to fix a base.
 
@@
 $M(x,y)=\inf\{\epsilon \mid \epsilon \star x \geq y\}$, yielding a cocomplete $\Lawv$-enriched category and, conversely, any cocomplete $\Lawv$-enriched category $X$ has a $\Lawv$-module structure with order $x\leq y$ when $X(y,x)=0$, and monoidal operation defined via a suitable weighted colimit.
-This induces an equivalence between the SMCC $\Mod$ of $\Lawv$-modules and their homomorphisms and the SMCC $\GMet$ of cocomplete $\Lawv$-enriched categories and cocontinuous functors (notice that functoriality in $\GMet$ is just non-expansiveness).
+This induces an isomorphism between the SMCC $\Mod$ of $\Lawv$-modules and their homomorphisms and the SMCC $\GMet$ of cocomplete $\Lawv$-enriched categories and cocontinuous functors (notice that functoriality in $\GMet$ is just non-expansiveness).
 
@@
 %Observe that the category $\LREL$ is (equivalent to) a subcategory of $\Lawv\SF{Mod}$.
-%While $\LREL$ corresponds to considering modules (the $\LREL^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $\Lawv\SF{Mod}$ corresponds to considering semimodules in abstract, without fixing a base.
+%While $\LREL$ corresponds to considering modules (the $\LREL^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $\Lawv\SF{Mod}$ corresponds to considering modules in abstract, without fixing a base.
 
@@
 
-Our contribution was, first, to observe that the SMCC structure of $\LREL$ lifts to that of the more general category $\Mod\equiv\GMet$ and, secondly, to define an exponential $!$ in $\Mod\equiv\GMet$, using a well-known recipe from \cite{Mellies2018, Manzo2013} based on the construction of symmetric algebras, and show that in this way also the exponential and differential structure of $\LREL_{!}$ lifts to this more general setting:
+Our contribution was, first, to observe that the SMCC structure of $\LREL$ lifts to that of the more general category $\Mod\equiv\GMet$; secondly, to define an exponential $!$ in $\Mod\equiv\GMet$, using a well-known recipe from \cite{Mellies2018, Manzo2013} based on the construction of symmetric algebras, and show that in this way also the exponential and differential structure of $\LREL_{!}$ lift to this more general setting:
 %
@@
 %Actually, we can even show that:
-\begin{theorem}
+\begin{theorem}\label{thm:no-base}
 $\Mod_{!}$ or, equivalently, $\GMet_{!}$, can be equipped with a differential $D$ making it a $CC\partial C$.

### 2023-05-18 TLLA23/BarbaPisto.tex
commit

@@
 %\end{remark}
-By applying known results (taken from \cite{Manzo2013},  {\color{red}CITARE LEMAYYYY}), one obtains that $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus:
+By applying known results (taken from \cite{Manzo2013}, \cite{Laird2020}, \cite{LemayCALCO2021}), one obtains that $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus:
 first, $\LREL$ is a SMCC, i.e.\ a model of the linear $\STLC$.
@@
 %\end{proof}
-This result is perfectly analogous to what happens in the metric models recalled in the introduction.%, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
+This result is perfectly analogous to what happens in the metric models recalled in the introduction. %, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
 It also entails that any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
@@
 
-Our contribution was, first, to observe that the SMCC structure of $\LREL$ lifts to that of the more general category $\Mod\equiv\GMet$; secondly, to define an exponential $!$ in $\Mod\equiv\GMet$, using a well-known recipe from \cite{Mellies2018, Manzo2013} based on the construction of symmetric algebras, and show that in this way also the exponential and differential structure of $\LREL_{!}$ lift to this more general setting:
+At this point, it can be shown that the SMCC structure of $\LREL$ lifts in a natural way to that of the more general category $\Mod\equiv\GMet$; secondly, an exponential $!$ in $\Mod\equiv\GMet$, can be defined via a well-known recipe  based on the construction of symmetric algebras \cite{Mellies2018, Manzo2013, Laird2020} .
+Since $!$ makes $\GMet$ a Lafont category, one can either do a few calculations or apply general theorems from the literature (e.g.~\cite{LemayCALCO2021}) to obtain the following result, which lifts also the exponential and differential structure of $\LREL_{!}$ to this more general setting:
 %

### 2023-05-18 TLLA23/BarbaPisto.tex
Committ

@@
 Connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
+
 However, a thorough investigation of the full power of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.
-We sketch here some first steps, with the idea of bridging the two approaches mentioned above, and suggesting the application of tropical methods to the study of the $\lambda$-calculus and its quantitative extensions.
+We sketch here some first steps.
+The aim is to bride the two approaches mentioned above by making them \emph{coexist}, and suggesting the application of tropical methods to the study of the $\lambda$-calculus and its quantitative extensions.
 This also scales to a 
@@
 %\end{remark}
-By applying known results (taken from \cite{Manzo2013}, \cite{Laird2020}, \cite{LemayCALCO2021}), one obtains that $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus:
+By applying known results (taken from \cite{Manzo2013}, \cite{Laird2020}, \cite{LemayCALCO2021}), one obtains that $\LREL$ gives rise to denotational models of several variants of the simply typed $\lam$-calculus ($\STLC$):
 first, $\LREL$ is a SMCC, i.e.\ a model of the linear $\STLC$.
@@
 tropical polynomials in possibly infinitely many variables.
-Furthermore, $!$ can be decomposed into a family of \emph{graded} exponentials $(!_n)_{n\in\N}$ turning $(\LREL,(!_n)_{n\in\N})$ in a model for a $\mathit{BLL}$-style simply typed $\lambda$-calculus, that we call $\BSTLC$ (see e.g.~\cite{Brunel2014}).
+Furthermore, $!$ can be decomposed into a family of \emph{graded} exponentials $(!_n)_{n\in\N}$ turning $(\LREL,(!_n)_{n\in\N})$ in a model for a %$\mathit{BLL}$-style
+simply typed $\lambda$-calculus with \emph{bounded} duplications, on the style of~\cite{Brunel2014}, call it $\BSTLC$.%(see e.g.~\cite{Brunel2014}).
 
@@
 This turns $(\LREL_!,D)$ into a CC$\partial$C, i.e.\ a model of the differential $\lambda$-calculus.
-Remember that the (quantitative) Taylor expansion of an ordinary $\lambda$-term $M$ is an inductively defined series $\Te{M}$ of differential $\lambda$-terms, the only non-trivial case being $\Te{PQ} :=  \sum_{k=0}^{\infty} (\Der^{k}[P] \cdot Q^{k})0$.
+Remember that the (qualitative) Taylor expansion of an ordinary $\lambda$-term $M$ is an inductively defined series $\Te{M}$ of differential $\lambda$-terms, the only non-trivial case being $\Te{PQ} :=  \sum_{k=0}^{\infty} (\Der^{k}[P] \cdot Q^{k})0$.
 It can be seen that the morphisms of $\LREL_!$ can always be Taylor expanded, and the series interpreting in $\LREL_!$ the Taylor expansion of a $\STLC$-term $M$, converges to the interpretation of $M$.
@@
 
-The main goal of this section is to show that the interpretation of $\STLC$ in the tropical relational model yields a metric semantics, where the spaces $\Lawv^{X}$ are endowed with the $\|\cdot\|_{\infty}$-norm metric, and programs are interpreted as \emph{locally Lipschitz} maps:
+The main goal of this section is to show that the interpretation of the above mentioned variants of the $\STLC$ in the tropical relational model yields a metric semantics, where the spaces $\Lawv^{X}$ are endowed with the $\|\cdot\|_{\infty}$-norm metric, and programs are interpreted as \emph{locally Lipschitz} maps:
 \begin{theorem}\label{cor:main}
@@
 
-The proof of the result above passes through the study of topological and metric properties of tLs. 
+The proof of the result above requires the study of %topological and metric properties of
+tLs from the point of view of mathematical analysis. 
 We first studied tLs with respect to the topology induced by the $\|\cdot\|_{\infty}$-norm.
+
 %
@@
 \begin{proposition}\label{prop:boundedlip}
-If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a bounded matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
+If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a %bounded
+matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is a $n$-Lipschitz map.
 \end{proposition}
@@
 This result is perfectly analogous to what happens in the metric models recalled in the introduction. %, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
-It also entails that any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
+It also entails that any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$ is $\mathrm{deg}(\varphi)$-Lipschitz.
 
@@
   %\item 
-$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical {polynomial} $P_\epsilon(x):=\min_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
+$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical \emph{polynomial} $P_\epsilon(x):=\min_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
 % \end{enumerate}
@@
 %\end{proof}
-The result above suggests that the local behavior of tLs with finitely many variables can be studied with the tools of tropical geometry (e.g.~tropical roots, Newton polygones).
+The result above suggests that, far from $0$, the behavior of tLs with finitely many variables can be studied with the tools of tropical geometry (e.g.~tropical roots, Newton polygones).
 %as suggested in the next section.
 Moreover, a consequence of Theorem~\ref{theorem:fepsilon} is that all tLs with {finitely many} variables are always \emph{locally} Lipschitz on $\BB R_{>0}$.
-The following result extends this property to all tLs, also covering the case with infinitely many variables.
+The following result extends this property to all tLs, including the case with infinitely many variables.
 
@@
 \end{theorem}
-The core of the proof is a convex analysis argument showing that an arbitrary function $f:\Lawv^X\to\Lawv$ which is non-decreasing, concave and continuous, must be locally Lipschitz.
+The core of the proof is a simple convex analysis argument showing that an arbitrary function $f:\Lawv^X\to\Lawv$ which is non-decreasing, concave and continuous, must be locally Lipschitz.
 
@@
 
-Since algebraic and geometric properties in tropical mathematics are usually more tractable from a computational point of view, in several well-known applications (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}) one starts from a given modelled phenomenon, typically expressed by some polynomial function $f$, and studies
-what of its properties can be deduced from the \emph{tropicalization} $\trop f$ of $f$.
-This idea suggests several natural directions in which the tropical semantics of a higher-order program could be used to deduce properties which can be expressed as an optimization problem.%: for a $\lambda$-term $M$, which can be expressed in $\LREL_!$ as a tropical polynomial or power series $\model M$, and we investigate what quantitative properties of $M$ can be deduced from it.
+Since algebraic and geometric properties in tropical mathematics are usually more tractable from a computational point of view, in several well-known applications (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}) one starts from a given modelled phenomenon, typically expressed by some polynomial function $f$, and studies which of its properties can be deduced from the \emph{tropicalization} $\trop f$ of $f$.
+This idea suggests several natural directions in which the tropical semantics of a higher-order programs could be used to deduce properties which can be expressed as an optimization problem.%: for a $\lambda$-term $M$, which can be expressed in $\LREL_!$ as a tropical polynomial or power series $\model M$, and we investigate what quantitative properties of $M$ can be deduced from it.
 In particular, it is useful to think of $\model M = \trop f$ for some polynomial or power series $f$.
@@
 They correspond to the respective reduction path from $M$ to the normal term of the given address.
-$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$.%after all the coin tossings in $M$ are performed.
-Thinking of $p,q$ as parameters, $P_{\omega}$ is the \emph{likelihood function} of the event $\omega$.
+$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$,%after all the coin tossings in $M$ are performed.
+i.e.\ the \emph{likelihood function} of the reduction given by $\omega$.
 The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after all the tossings (similarly a $Q_{0}(p,q)$ for $\false$).
@@
  By adapting \cite[Section IV]{Manzo2013}, it can be seen that $\LREL_!$ is a model of $\STLC_\oplus$.
- In particular, if we set $\model \bool:=\set{0,1}$, our running example $M$ is interpreted as $\model M\in\Lawv^{\set{0,1}}$ giving the tropicalised probabilities: $\model M_0=(\trop Q_0)(p,1-p)$, $\model M_1=(\trop Q_1)(p,1-p)$.
+ In particular, if we set $\model \bool:=\set{0,1}$, our running example $M$ is interpreted as $\model M\in\Lawv^{\set{0,1}}$ giving the following tropicalised probabilities: $\model M_0=(\trop Q_0)(p,1-p)$, $\model M_1=(\trop Q_1)(p,1-p)$.
 \end{remark}
@@
 %\end{example}
-Also, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a root of $\trop Q_1$ or of $\trop Q_0$ provide the values of the bias of $\oplus_p$ for which there are at least two different \emph{equiprobable} paths from $M$ to its normal form.
+Also, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a root of $\trop Q_1$ or of $\trop Q_0$, provide the values of the bias of $\oplus_p$ for which there are at least two different \emph{equiprobable} paths from $M$ to its normal form.
 Yet, we do not have a full understanding of the role of roots in this setting.
@@
 
-As we have seen, tropical semantics provides a viewpoint in which metric and differential properties are somehow unified.
+As we have seen, tropical semantics provides a viewpoint in which metric and differential properties coexist.
 This approach can be made more abstract, thanks to a fundamental categorical correspondence between tropical linear algebra (i.e.~the study of quantale modules over $\Lawv$) and the theory of Lawvere's \emph{generalized metric spaces} (see \cite{Fuji, Stubbe2006}).
@@
 %Notice that this defines the usual notion of module over the Lawvere quantale $\Lawv$, and this does \emph{not} coincide with the modules on $\Lawv$ \emph{as a semiring} (which are indeed well-studied and well-known to provide a $CC\partial C$ {\color{red}LEMAYYYY}).
-The most basic example of $\Lawv$-modules are given by the spaces $\Lawv^{X}$, with order and action defined pointwise.
+The most basic examples of $\Lawv$-modules are given by the spaces $\Lawv^{X}$, with order and action defined pointwise.
 We see them as modules given together with a fixed base, $X$.
-Our Theorem~\ref{thm:no-base} says that one can actually give a semantic of high order programs without need to fix a base.
+Theorem~\ref{thm:no-base} says that one can actually give a semantic of high order programs without need to fix a base.
 
@@
 
-At this point, it can be shown that the SMCC structure of $\LREL$ lifts in a natural way to that of the more general category $\Mod\equiv\GMet$; secondly, an exponential $!$ in $\Mod\equiv\GMet$, can be defined via a well-known recipe  based on the construction of symmetric algebras \cite{Mellies2018, Manzo2013, Laird2020} .
+At this point, it can be shown that the SMCC structure of $\LREL$ lifts in a natural way to that of the more general category $\Mod\simeq\GMet$; secondly, an exponential $!$ in $\Mod\simeq\GMet$, can be defined via a well-known recipe  based on the construction of symmetric algebras \cite{Mellies2018, Manzo2013, Laird2020} .
 Since $!$ makes $\GMet$ a Lafont category, one can either do a few calculations or apply general theorems from the literature (e.g.~\cite{LemayCALCO2021}) to obtain the following result, which lifts also the exponential and differential structure of $\LREL_{!}$ to this more general setting:

### 2023-07-21 CSL24/main.tex
Committ
Per ora è ancora molto copia ed incolla da LICS. Fai tutte le modifiche che vuoi, se vuoi. Io domani non posso lavorare, quindi prendi il lock quando vuoi. Da domenica, dimmelo così ci organizziamo. Notte !

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{bussproofs}
\EnableBpAbbreviations
\usepackage{hyperref}
\usepackage{tikz,pgfplots}
\pgfplotsset{compat=newest}
\usepackage{tikz-cd}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{amsthm}
\usepackage{caption}
\usepackage{subcaption}
\usetikzlibrary{cd}
\usepackage{mathabx}
\usepackage{cmll}
\usepackage{stmaryrd}
\usepackage{wrapfig}

\input{prooftree}

\usepackage{proof}

\bibliographystyle{plainurl}% the mandatory bibstyle

\input{macros}

\title{Linearity and tropical mathematics in the Lambda-calculus} %TODO Please add

%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

\author{Davide {Barbarossa}}{Dipartimento di Informatica, Universit\`a di Bologna, %[optional: Address], 
Italy% \and My second affiliation, Country 
\and \url{https://lipn.univ-paris13.fr/\~barbarossa/index.html} }{davide.barbarossa@unibo.it}{https://orcid.org/0000-0003-4608-8282}{%(Optional)Funding each author
}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Paolo {Pistone}}{Dipartimento di Informatica, Universit\`a di Bologna, %[optional: Address], 
Italy% \and LIP, Universit\'e de Lyon, France 
\and \url{%WebPage
} }{paolo.pistone2@unibo.it}{%Orchid
}{%(Optional)Funding each author
}

\authorrunning{D.\,Barbarossa and P.\,Pistone} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Davide Barbarossa and Paolo Pistone} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[500]{Theory of computation~Lambda calculus}
\ccsdesc[300]{Theory of computation~Categorical semantics}
\ccsdesc[100]{Theory of computation~Linear logic}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Differential lambda-calculus, Tropical semiring, Relational semantics, Lawvere quantale, Program metrics} %TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

\funding{This work has been supported by the DIAPASoN project}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{%I want to thank \dots
}%optional

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

%TODO mandatory: add short abstract of the document
\begin{abstract}
In this paper we study an interpretation of the $\lambda$-calculus in the framework of tropical mathematics, by considering the relational semantics weighted on the tropical semiring.
We consider it as providing a unified framework where both program metrics, as based on the analysis of program sensitivity via Lipschitz-conditions, and resource analysis, as based on higher-order program differentiation, coexist. This is possible thanks to the fact that this is also a model of both a graded lambda-calculus ensuring bounded duplications, and the differential lambda-calculus. We only show the starting result of our line of research: thanks to the tropical operations we consider, in such model programs with bounded duplication are Lipschitz maps, simply typed terms are locally Lipschitz maps and the differential terms appearing in the Taylor expansion of a term (also called resource terms) are Lipschitz maps. Therefore, the lambda-calculus Taylor expansion decomposes semantically the simply typed term as an inf of more and more sensitive functions.
\end{abstract}

\section{Introduction}
\input{section1}

%\section{The Two Quantitative Approaches to the $\lambda$-calculus}\label{section2}
\input{section2}

\section{The category $\LREL$}\label{section3}

\input{section3new}

\subsection{... As (weighted relational) model of several $\lambda$-calculi}

The category $\LREL$ satisfies many nice properties which make it able to model all the variants of the $\lambda$-calculi to which we are interested in here.
Although the majority of the properties mentioned in the present section are shared with the general class of models of $\lambda$-calculi called \emph{weighted relational models}, thoroughly studied in \cite{Manzo2013}, we mention them for our particular case in order to highlight how these look like in the tropical setting.
Moreover, we end the section by stating some property of $\LREL$ not necessarily shared with all the models in the general class.
The real interest in our particular model will come in the next \autoref{sec:tls}.

\paragraph{Prohibited duplication/erasure: the \emph{linear} $\STLC$}\label{sec:3A}\input{section3A}

\paragraph{Allowed duplication/erasure: the $\STLC$}\label{sec:3B}\input{section3B}

\paragraph{Controlled duplication/erasure: the $\BSTLC$ and the $\STDLC$}\label{sec:3C}\input{section3C}

\section{Tropical Laurent Series}\label{sec:tls}\input{section4}

\section{Future lines of research}\label{sec:future}

{\color{red}2 pagine in totale}

\subsection{Quantitaive analysis: probabilistic, differential privacy}\label{sec:app}
%\input{section5}
%\input{section5C}

\newpage

\subsection{Generalized Metric Spaces and Modules over the Lawvere Quantale}\label{section6}
%At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}.
%Let us recall that a quantale is a complete lattice endowed with a continuous monoid action. In the case of $\Lawv$ the lattice is defined by the reverse order $\geq$ on $\BB R$, and the monoid action is provided by addition. Notice that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$. 
%A consequence of these observations is that, as we discussed below, the tropical approach to linear algebra coincides with the study of ``$\Lawv$-valued matrices'', i.e.~of maps of the form $s: X\times Y\to \Lawv$ .
%In particular, a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we explicitly take in Section \ref{section6}). 
%\input{section6}%

\newpage

\section{Related Work}\label{section7}
\input{section7}

\section{Conclusion}\label{section8}
\input{section8}

%%
%% Bibliography
%%

%% Please use bibtex,

\bibliography{tropical}

\appendix

\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}

\section{Theorem-like environments}\label{sec:theorem-environments}

\end{document}

### 2023-07-21 CSL24/section2.tex
Committ
Per ora è ancora molto copia ed incolla da LICS. Fai tutte le modifiche che vuoi, se vuoi. Io domani non posso lavorare, quindi prendi il lock quando vuoi. Da domenica, dimmelo così ci organizziamo. Notte !

--- 2023-01-24 section2.tex
@@
 
-In this section, we discuss in some more detail the two approaches to quantitative semantics we mentioned in the Introduction, at the same time providing an overview of how we aim at bridging them using tropical mathematics.
+%In this section, we discuss in some more detail the two approaches to quantitative semantics we mentioned in the Introduction, at the same time providing an overview of how we aim at bridging them using tropical mathematics.
 
-\subsection{Metric approach: bounded $\lambda$-terms as Lipschitz functions}
+\subsection{Metric approach: bounded $\lambda$-calculus in a nutshell}
 
@@
 
-In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior.
+%In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior.
 A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
@@
 The types of such systems are defined by combining linear constructors with a \emph{graded linear exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
-In the following sections we will sometimes make reference to a basic graded type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined via $A::= o  \mid   !_{n}A\multimap A  $,
-where $n\in \BB N$.
-Intuitively, $!_{ n}A\multimap B$ is the type of functions from $A$ to $B$ that may use their input \emph{at most} $n$ times.
-More details about $\BSTLC$ are provided in the Appendix.
+The language $\BSTLC$ we consider is a simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}. 
+The terms are as for the $\STLC$, the types are $A::= o \ \mid  \ !_{n}A \multimap A$ and type judgements are of the form $\Gamma \vdash M:A$, where a context $\Gamma$ is a set of declarations of the form $x :_{n}A$, with $n\in \mathbb N$.
+The typing rules are given in Fig.~\ref{fig:rules}, where $\Gamma+\Delta$ is defined letting $(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) =  (\Gamma+\Delta), x:_{m+n}A$, and $m\Gamma$ is made all $x:_{mn}A$ for $(x:_{n}A) \in \Gamma$.
+The axiom is $x:_{1}A\vdash x: A$.
+The main feature of this language is that if $\vdash \lambda x.M:\,!_nA\multimap B$, then $x$ will be duplicated exactly $n$ times in the reduction to the normal form.
+%, i.e.\ $\vdash \lambda x.M:\,!_nA\multimap B$ is a program which performs $n$ duplications.
+E.g., $\vdash_{\BSTLC} \lambda {\color{red}z}.\left( \lambda x{\color{green}y}.
+yxx \right)z : \, !_{\color{red}2} o\multimap !_{\color{green}1}(!_{\color{violet}1} o \multimap !_{\color{blue}1} o \multimap o) \multimap o$.
 
-Now, what about the good old, ``unbounded'', simply typed $\lambda$-calculus?
-Actually, by using unbounded duplications, one might lose the Lipschitz property.
-For instance, while the functions $M_{k}=\lambda x.
-k\cdot x: \BB R\to \BB R$ are all Lipschitz-continuous, with Lipschitz constant $k$, the function $M=\lambda x.x^{2}$ obtained by ``duplicating'' $x$ is not Lipschitz anymore: $M$ is, so to say, \emph{too} sensitive to errors.
-More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of $\STLC$ (yet, several cartesian closed \emph{sub-}categories of $\Met$ do exist, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
-Still, one might observe that the program $M$ above is actually Lipschitz-continuous, if not globally, at least \emph{locally} (i.e.~over any compact set).
-Indeed, some cartesian closed categories of locally Lipschitz maps have been produced in the literature \cite{Ehrhard2011, PistoneLICS}, and a new example will be exhibited in this paper.
+The $\BSTLC$ can be interpreted in a symmetric monoidal closed category (SMCC) equipped with a \emph{$\mathbb N$-graded linear exponential comonad} \cite{Katsumata2018}.
+%Observe that one can always type an affine term like e.g.~$\lambda xy.x$ with a linear type $A\multimap B\multimap A$.
+Instead, a term like $\lambda xy.x(xy)$ containing two occurrences of $x$ cannot be given the linear type $(A\multimap A)\multimap (A\multimap A)$ but a type of the form
+%$!_{2}(A\multimap A)\multimap (A\multimap A)$.
 
-\subsection{Resource approach: differential $\lambda$-terms as polynomials}
+\begin{figure*}
+
+	\scriptsize
+
+	\[ \arraycolsep=5pt\def\arraystretch{2.8}
+	\begin{array}{cccc}
+		\prooftree
+		\Gamma \vdash M:A
+		\justifies
+		\Gamma, x:_{0}B \vdash M:A
+		\endprooftree 
+		&
+		\prooftree
+		\Gamma, x:_{n}B, y:_{m} B\vdash M:A
+		\justifies
+		\Gamma, x:_{n+m}B\vdash M[x/y]:A
+		\endprooftree 
+		&
+		\prooftree
+		\Gamma, x:_{n} A\vdash M: B
+		\justifies
+		\Gamma\vdash \lambda x.M: !_{n}A\multimap B
+		\endprooftree
+		&
+		\prooftree
+		\Gamma \vdash M: !_nA\multimap B
+		\quad
+		\Delta\vdash N: A
+		\justifies
+		\Gamma +n\Delta\vdash MN: B
+		\endprooftree
+		\\
+		\\
+		\hline
+		\\
+		\prooftree
+		\Gamma, x: A\vdash M: B
+		\justifies
+		\Gamma\vdash \lambda x.M: A\to B
+		\endprooftree 
+		&
+		\prooftree
+		\Gamma \vdash M: A\to B
+		\quad
+		\Gamma\vdash \mathbb T: A
+		\justifies
+		\Gamma \vdash M\mathbb T: B
+		\endprooftree 
+		&
+		\prooftree
+		\Gamma \vdash M: A\to B
+		\quad
+		\Gamma \vdash N: A
+		\justifies
+		\Gamma \vdash \Diff{M}{N}: A\to B
+		\endprooftree
+		&
+		\prooftree
+		\Gamma\vdash M_1: A
+		\,\cdots\,
+		\Gamma \vdash M_n:A
+		\justifies
+		\Gamma \vdash M_1+\cdots +M_n : A
+		\using (n\geq 2)
+		\endprooftree
+	\end{array}
+	\]
+	\caption{Typing rules (axiom rules are given in the text) for $\BSTLC$ (top) and $\STDLC$ (bottom).}\label{fig:rules}
+\end{figure*}
+
+\subsection{Resource approach: differential $\lambda$-calculus in a nutshell}
 
@@
 The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus.
-The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M]\cdot N$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M]\cdot N$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).
+The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Diff{M}{N}$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Diff{M}{N}$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).
 
-Interestingly, the categorical study of the operator $\Der$ has led to the introduction of \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, a class of categories providing an abstract axiomatization of differentiation, in the usual mathematical sense.
-More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
-\begin{itemize}
-\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
-\item $\C C$ is equipped with a differential operator $D:
-\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
-\end{itemize}
+The syntax of the differential $\lambda$-calculus $\STDLC$ (e.g.\ \cite[Section 3]{Manzo2010}), is given by the \emph{terms} $M$ and the \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
+quotiented by: $\alpha$-equivalence; $+,0$ form a commutative monoid on the set of sums; %, i.e.\ commutativity and associativity of $+$ and neutrality of $0$ w.r.t.\ $+$;
+linearity of $\lam x.(\_)$, $\Diff{\_}{\_}$ and $(\_)\mathbb T$ (but \emph{not} of $M(\_)$); irrelevance of the order of consecutive $\Diff{\_}{\_}$.
+%Remark that $M(\_)$ is \emph{not} set to be linear: $\lambda x.0=0\mathbb T=\Diff{0}{N}=\Diff{M}{0}=0$ but $M0\neq0$ in general.
+%This is crucial for the definition of the Taylor expansion.
+We follow the tradition of quotienting also for the idempotency of $+$.
+%Sums are, then, just \emph{finite} sets of terms.
+Simple types are $A::= o\mid A\to A$.
+The typing system is in Figure~\ref{fig:rules}, where a context $\Gamma$ is a list of typed variable declarations.
+The axioms are $\Gamma, x:A \vdash x: A$ and $\Gamma\vdash 0:A$.
 
-Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
-The models of $\STDLC$ are the 
-cartesian \emph{closed} differential categories ($CC\partial C$), 
-which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.
+%In $\BSTLC$ the typing system handles duplications; in $\STDLC$ the syntax with its operational semantics (that we do not give) does it.
+Writing $\Der^2[\_,(\_)^2]$ as a shortcut for $\Der[\Der[\_,\_],\_]$ and $\Der^1[\_,(\_)^1]$ for $\Diff{\_}{\_}$, the analogue of the previous $\BSTLC$-term is $\vdash_{\STDLC} \lambda {\color{red}z}.
+\Der^{\color{red}2}[
+	\lambda x{\color{green}y}.
+		\Der^{\color{violet}1} [
+				\Der^{\color{blue}1} [y, x^{\color{blue}1}]
+        0, x^{\color{violet}1}
+	]0
+, z^{\color{red}2}]0
+: {\color{red}o}\to ({\color{green}o \to o \to o}) \to o$.
+%Here we wrote $\Der^2[\_]\cdot (\_)^2$ as a shortcut for $\Der[\Der[\_]\cdot (\_)]\cdot (\_)$ and $\Der^1[\_]\cdot (\_)^1$ for $\Der[\_]\cdot (\_)$.
+In particular, if the \emph{multiplicities} of the arguments (the colored exponents) do not exactly match the number of duplications, e.g.\ in $\vdash_{\STDLC} \lambda z. \Der^{\color{red}3}[
+	\lambda xy.
+		\Diff{
+				\Diff{y}{x}
+		0}{x}
+	0
+, z^{\color{red}3}]0
+: o\to (o \to o \to o) \to o$, then the term reduces to the empty sum $0$ (representing an \emph{error}).
+
+%Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
+%The models of $\STDLC$ are the cartesian \emph{closed} differential categories ($CC\partial C$), which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.
+
+The $\STDLC$ can be interpreted in Cartesian closed differential $\lambda$-categories (CC$\partial\lambda$C)\cite{Manzo2010,Blute2009, Blute2019}.
+In them, homsets are equipped with a structure of commutative monoid %(i.e.\ we can add morphisms and there is a $0$ morphism)
+%, it is Cartesian closed%(with the closed structure compatible with the additive structure)
+and with a certain \emph{differential operator} $D$, generalising the usual notion of differentiation, see e.g.\ \cite{BluteEhrhTass10}.
+%An example is the CC$\partial\lambda$C of convenient vector spaces with smooth maps, where $D$ is the ``real'' differential $Df:\mathbb{V}\times\mathbb{V}\rightarrow \mathbb{W}$, $Df(x,u):=\dfrac{d}{dt}{\!\Big|_{t=0}} f(x+tu)$, of smooth maps $f:\mathbb{V}\rightarrow \mathbb{W}$.
+%More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
+%\begin{itemize}
+%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
+%\item $\C C$ is equipped with a differential operator $D:
+%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
+%\end{itemize}
 
@@
 actual derivatives is provided by the \emph{Taylor expansion} $\C T$:
-in $\STDLC$ one can expand any application $MN$ as an infinite formal sum of \emph{linear} applications
-$\Der^{(k)}[M]\cdot N^{k}$, i.e.~where $N$ is passed exactly $k$ times to $M$:
-\begin{align}\label{eq:taylor}
-\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot (\Der^{(k)}[M] \cdot N^{k})0
-\end{align}
-In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
+%In analysis, a sufficiently regular function can be seen (locally around a point) as its Taylor expansion, i.e.\ a series of polynomials weighted via some rational coefficient.
+%Now, a polynomial in $x$ of degree $k$ can be thought of as a function which uses its argument $x$ exactly $k$ times.
+in $\STDLC$ (even untyped), we can perform a syntactic Taylor expansion of an ordinary $\lambda$-term via an inductively defined map $\Theta$ giving rise to an infinite series of terms, whose crucial case of the definition is
+$
+\Theta(MN)=\sum_{n\in\mathbb{N}}\frac{1}{n!}\Der^n\left[\Theta(M),\Theta(N)^n\right]0.
+$
+As in analysis, it decomposes an application as a series of $k$-linear functions, which can be seen as its approximants.
+In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.
+Since we only consider idempotent sum, the factorial coefficient disappears, and the resulting Taylor map is called the \emph{qualitative} Taylor expansion, indicated by $\Te{\_}$.
+% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
 
@@
 % $\Der[\sum_{n}a_{n}x^{n}]=\sum_{n}\Der[a_{n}x^{n}]$ and $\Der[a_{0}x^{0}]=0$ and $\Der[a_{n+1}x^{n+1}]= (n+1)a_{n+1}x^{n}$, so that power series can be Taylor expanded.
-
-\subsection{Tropical Mathematics: a possible synthesis}%Outline of the paper}
-
-At this point, as the Taylor formula decomposes an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as interpreting  a $\lambda$-term
-as a limit of Lipschitz maps, in some sense, thus bridging the metric and differential approaches.  
-Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
-However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.
-
-Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions?
-This way, \eqref{eq:taylor} could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps.
-In other words, unbounded term application could be seen
-as a limit of \emph{more and more sensitive} operations.
-
-This viewpoint, that we develop in the following sections, leads to the somehow unexpected discovery of a bridge between the metric and differential study of higher-order programs.
-This connection not only suggests the application of optimization methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions, but it scales to a 
-more abstract level, leading to introduce a 
-differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}).
-
-%
-%At this point, 
-%by interpreting such polynomials over the tropical semiring 
-%
-%Our question can thus be reformulated as follows: can we make the relational semantics \emph{Lipschitz}, hence amenable to metric and sensitivity analysis?
-The goal of this paper is to show that, by appealing to tropical mathematics, this is indeed possible and leads to the somehow unespected discovery of a bridge between the metric and differential study to higher-order programs.
-%
-%Fino a qui niente a capo
-%
-%Fino a qui niente a capo
-%
-%Fino a qui niente a capo
-%
-%Fino a qui niente a capo
-%
-%Fino a qui niente a capo
-%
-%Limite! Alla prossima riga sforo oltre 12 pag.
-%
-
-%In the next section we recall some basic ideas from tropical mathematics, and its connection with the study of the Lawvere quantale.
-%Since polynomials correspond to piecewise linear (hence Lipschitz) functions in tropical mathematics,
-%The reconstruction of the relational semantics over the tropical semi-ring, presented in Section \ref{section3} and Section \ref{section4}, will provide a metric semantics of the differential $\lambda$-calculus, bridging sensitivity and resource analysis.
-%In Section \ref{section5} we suggest potential applications of this approach, relating well-studied applications of program metrics, resource analysis with current uses of tropical mathematics in computer science.
-%Finally, in Section \ref{section6} we show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between {generalized metric spaces} and modules over the tropical semiring.

### 2023-07-21 CSL24/section3A.tex
Committ
Per ora è ancora molto copia ed incolla da LICS. Fai tutte le modifiche che vuoi, se vuoi. Io domani non posso lavorare, quindi prendi il lock quando vuoi. Da domenica, dimmelo così ci organizziamo. Notte !

The linear simply typed $\lam$-calculus is a restriction of the ordinary $\lam$-calculus in which programs, following the linear logic mantra, can only use their arguments \emph{exactly once} (and indeed this calculus can be embedded in its \emph{intuitionistic multiplicative} fragment \emph{IMLL}).
%More precisely, the linear $\lam$-calculus is obtained from the ordinary one by adding the constraint that each $\lam$-abstracted variable appears exactly once in the scope of the $\lam$-abstraction.
%In order to frame it in a category $\C C$, one needs a symmetric tensor product $\otimes$ together with internal hom-set objects $X\multimap Y$ s.t.\ the evaluation and curry maps yield a \emph{symmetric monoidal adjunction}: $\C C(Z\otimes X, Y) \simeq \C C (Z, X\multimap Y)$.
It can be interpreted in any Symmetric Monoidal Closed Category (SMCC).
Now, \cite[Section III.A]{Manzo2013} immediately gives that $\LREL$ is a SMCC (actually, even a model of IMLL), where its monoidal product $\otimes$ acts on the objects as the Cartesian product of sets, and as the \emph{Kronecker product}
 %{(\color{red} Ref?)} 
 of matrices on morphisms.
Its internal hom $X\multimap Y$ is also given by the Cartesian product $X\times Y$ of sets, with the usual evaluation and curry maps.

### 2023-07-21 CSL24/section3B.tex
Committ
Per ora è ancora molto copia ed incolla da LICS. Fai tutte le modifiche che vuoi, se vuoi. Io domani non posso lavorare, quindi prendi il lock quando vuoi. Da domenica, dimmelo così ci organizziamo. Notte !

--- 2023-01-22 section3B.tex
@@
 %\subsubsection{Unbounded duplications}
+
+\begin{comment}
 
@@
 
-\begin{fact}
+\begin{proposition}
  $\LREL$ is Lafont.
-\end{fact}
+\end{proposition}
+
+\end{comment}
 
@@
 %We will explicit this constructions in our tropical setting in the next lines.
-As already mentioned, one obtains a CCC from a Lafont category by defining the exponential objects as $X\to Y:=!X \multimap Y$.
-In our case, specialising [Theorem III.7, \cite{Manzo2013}] we have indeed:
-\begin{fact}
- The coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.
-\end{fact}
-
+As well-known, if a SMCC is \emph{Lafont}, then one obtains a CCC from it by defining the exponential objects as $X\to Y:=!X \multimap Y$.
+With no surprise $\LREL$ is Lafont (\cite[Corollary III.6]{Manzo2013}) and specialising \cite[Theorem III.7]{Manzo2013} we have indeed that the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$, where as usual $!$ acts on objects by taking the set of finite multisets.
 %It is instructive what its CCC-structure looks like in our tropical world.
-In our tropical setting, 
-the exponential object $X\multimap Y$ is $!X\times Y$, and 
-the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
-\begin{align}
- \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1*\cdots *\mu_n}
- \left\{s_{[b_1,\dots,b_n],c} + \sum\limits_{i=1}^n t_{\mu_i,b_i}\right\}.
-\end{align}
-% (where $+$ is the disjoint union).
+Remark that the cartesian product of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_!
+t\in\Lawv^{!X\times Z}$ where $(s\circ_!
+t)_{\mu,c}:=
+%\begin{align}
+\inf_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
+ \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
+%\end{align}
+%where $+$ is the multiset union.
 
@@
 
-As it is well-known, the Cartesian closed structure %of a category $\C C$ 
-allows to define a sound interpretation $\model{\Gamma\vdash M:A}\in\HOM{\LREL_{!}}{\model \Gamma}{\model A}$ of terms as morphisms.
-In our case, we have:
-
-\begin{proposition}\label{prop:descrete}
- The interpretation $\model M$ of a $\lam$-term $M$ in $\LREL$ is a \emph{discrete} matrix, i.e.\ its coefficients are either $0$ or $\infty$.
-\end{proposition}
-\begin{proof}
- First easily prove that composition preserves discreteness.
- Then go by straightforward induction on $M$, using that the projections and evaluation of  $\LREL$ are discrete.
-\end{proof}
-
-While ordinary $\lam$-terms yield discrete matrices, 
-as discussed in Section~\ref{sec:app}, the other coefficients from $[0,\infty]$ will appear when considering quantitative effects like computation steps or probabilities.
-%, give rise to  we will however discuss how 
-%we can use this model in order to make it keep track of quantitative effects of programs.
-
-\begin{remark}\label{rmk:ModelsOfBSTLC}
-%Similarly to It is known {\color{red}(([reference??] e dire meglio)} 
-The exponential $!$ can be ``decomposed'' into a family of \emph{graded} exponentials $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined by functors $X\mapsto \C M_{\leq n}(X)$. %  lift to functors 
-The sequence $(!_n)_{n\in\N}$ can be then turned into a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, satisfying the adjunction:
-$\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
-Therefore, $\LREL$, together with $(!_n)_{n\in\N}$, is a model for $\BSTLC$. 
-In particular, bounded arrow types are interpreted via
-%Concretely, the types of this calculus (Equation~\ref{eq:Btyped}) are interpreted as: 
-$\model o=\{\star\}$ and 
-$\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$.
-Notice that, for any type $A$ of $\BSTLC$, its interpretation $\model{A}$ is a finite set.
-\end{remark}
-%Bounded types are interesting because of the following proposition:
-%\begin{proposition}
-%For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) correspond to polynomials.
-%\end{proposition}
-%\begin{proof}
-%It suffices to check that $\model A$ is finite for all bounded types $A$.
-Indeed this implies that a morphism $t:\model A\to \model B$ is a finite matrix $t: \model A \times \model B \to \Lawv$.Hence, its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is a polynomial.
-%\end{proof}
-
-%For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
-%\begin{itemize}
-%\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
-%\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
-%\end{itemize}
+%As it is well-known, the Cartesian closed structure %of a category $\C C$  allows to define a sound interpretation $\model{\Gamma\vdash M:A}\in\HOM{\LREL_{!}}{\model \Gamma}{\model A}$ of terms as morphisms.
+%In our case, we have:

### 2023-07-21 CSL24/section3C.tex
Committ
Per ora è ancora molto copia ed incolla da LICS. Fai tutte le modifiche che vuoi, se vuoi. Io domani non posso lavorare, quindi prendi il lock quando vuoi. Da domenica, dimmelo così ci organizziamo. Notte !

%\begin{remark}\label{rmk:ModelsOfBSTLC}
%Similarly to It is known {\color{red}(([reference??] e dire meglio)} 
For $\BSTLC$, one can see that the exponential $!$ can be ``decomposed'' into a family of \emph{graded} exponentials functors $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined by taking multisets of cardinality \emph{at most} $n$. %  lift to functors 
The sequence $(!_n)_{n\in\N}$ gives rise to a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, %satisfying the adjunction: $\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
i.e.\ $(\LREL,(!_n)_{n\in\N})$ is a model of $\BSTLC$. 
In particular, bounded arrow types are interpreted via $\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$ and so, 
if e.g.\ $\model o=\{\star\}$, then $\model{A}$ is a finite set for any type $A$ of $\BSTLC$.
%\end{remark}
%Bounded types are interesting because of the following proposition:
%\begin{proposition}
%For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) correspond to polynomials.
%\end{proposition}
%\begin{proof}
%It suffices to check that $\model A$ is finite for all bounded types $A$. Indeed this implies that a morphism $t:\model A\to \model B$ is a finite matrix $t: \model A \times \model B \to \Lawv$.Hence, its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is a polynomial.
%\end{proof}

%For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
%\begin{itemize}
%\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
%\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
%\end{itemize}

%Until now we simply specialised well-known results in our tropical case, with the intent of showing how things read in this particular case.
%Now we go further, by showing that $\LREL_!$ actually admits a \emph{differential structure}, turning it into a model of the $\STDLC$, i.e.~a $CC\partial C$.
%This viewpoint
%, is where the \emph{metric} and the \emph{differential} viewpoints converge, as explained in the Introduction and Section II, and it % will be further generalised in Section \ref{section6}.

%A model of the $\STDLC$ is usually understood as so-called \emph{Cartesian closed differential categories} (CC$\partial$C), see \cite{Manzo2012} for details.
%In order to treat the $+$ and the constructor $D[\_]\cdot (\_)$ of $\STDLC$, the main features of a CC$\partial$C $\C C$ are that:
%
%1) $\C C$ is a left-additive-CCC, i.e.\ its Homsets are commutative monoids and its Cartesian closed structure is well behaved w.r.t.\ this monoid structure;
%
%2) $\C C$ is equipped with a differential operator map $D:\HOM{\C C}{X}{Y}\to \HOM{\C C}{X\times X}{Y}$ (here $\times$ is the Cartesian product of $\C C$) satisfying $8$ axioms, called D1, ..., D7, D-curry.

%Let us show the differential structure of $\LREL_!$ (remember that the Cartesian product of $\LREL_!$ is the disjoint union $+$).

%\begin{definition} The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).\end{definition}

For what concerns $\STDLC$, by patiently checking all the needed axioms {\color{red}or by using  LEMAY??}, one sees that
%\begin{proposition}[{\color{red}LEMAY??}]\label{thm:LREL!CCDC}
 $\LREL_!$ becomes a CC$\partial\lambda$C, i.e.\ a model of the $\STDLC$, when equipped with the \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
%\end{proposition}
%\begin{remark}For $t\in\HOM{\LREL}{!X}{Y}$, we have: $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu+\nu'+\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho+\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.\end{remark}
%This ensures that one can define a sound interpretation of $\STDLC$-terms in the standard way (see [Section 4.3, \cite{Manzo2010}]).
This model is also well-behaved w.r.t.\ the Taylor expansion, as expressed by the following two results.
As it can be patientely checked, in $(\LREL_!,D)$ all morphisms can be Taylor expanded  (see \cite[Definition 4.22]{Manzo2012}):

\begin{theorem}\label{thm:modelsTaylor}
 For all $t\in\HOM{\LREL_!}{Z}{X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$ we have:%, the evaluation of $t$ over $s$ yields 
 \begin{align}\label{eq:taylorcat}
  \RM{ev}\circ_!\langle t,s\rangle =
  \inf\limits_{n\in\N}
  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
 \end{align} 
\end{theorem}
%It is worth discussing the formula above a bit more. 
Here,
%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$,  i.e.~it coincides with the tropical %interpretation of the
version of the usual Taylor expansion.
%Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
Finally, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] shows:

\begin{corollary}\label{cor:T(M)=M}
 If $\Gamma\vdash_{\STLC} M:A$, then $\model{\Gamma\vdash_{\STDLC} \Te M:A}:=\inf_{t\in\Te{M}} \model{\Gamma\vdash_{\STDLC} t:A}$ converges to $\model{\Gamma\vdash_{\STLC} M:A}$. %the interpretation of the Taylor expansion of a $\STLC$-term $M$, given in \eqref{eq:taylor}, converges to the one of $M$.
\end{corollary}

By straightforward induction on $M$ (using that composition and $D$ easily preserve ``booleaness'' and projections and evaluation of  $\LREL$ are boolean) one can prove that:

\begin{proposition}\label{prop:descrete}
 For all the previous four languages, $\model{\Gamma\vdash M:A}\in\Lawv^{!\model{\Gamma}\times\model{A}}$ is a \emph{boolean} matrix, i.e.\ actually $\model{\Gamma\vdash M:A}\in\{0,\infty\}^{!\model{\Gamma}\times\model{A}}$.
\end{proposition}

The fact that the coefficients are either $0$ or $\infty$ can be seen as the fact that $\STLC$ is relatively trivial, from this point of view.
As we will discuss in \autoref{sec:app}, non-trivial coefficients appear when interpreting more interesting languages, like probabilistic ones.
However, in this paper we shall still focus on $\STLC$, in order to set the basis of the further studies and in order to see some already interesting properties of the interpretation.

### 2023-07-21 CSL24/section4.tex
Committ
Per ora è ancora molto copia ed incolla da LICS. Fai tutte le modifiche che vuoi, se vuoi. Io domani non posso lavorare, quindi prendi il lock quando vuoi. Da domenica, dimmelo così ci organizziamo. Notte !

--- 2023-01-24 section4.tex
@@
-As morphisms in $\LREL$ correspond to tropical linear functions, morphisms in the coKleisli category $\LREL_{!}$ 
-correspond to a generalization of tropical Laurent series, i.e.~the tropicalization of power series.
-
-%many properties of these functions still have to be well-understood. 
-In this section we establish some topological and metric properties of tropical Laurent series, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
-The literature on power series in tropical mathematics is very recent (see e.g.~\cite{Porzio2021}) and several results we prove in this section are, to our knowledge, new.
+For starting, remark that although a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, by exploiting the coKleisli structure we can also ``express it in the base $X$'', i.e.\ see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_!
+x$.
+%(we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
+Concretely, we have $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
+In this section we establish some topological and metric properties of such functions, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
+As we explain in a moment, those are generalised \emph{tropical Laurent series} (\emph{tLs} for short), and as the literature in tropical mathematics is often recent (e.g.~\cite{Porzio2021}), several results we prove in this section are, to our knowledge, new.
 
@@
 
-A tropical polynomial is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form \begin{equation}
-f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}
-\end{equation}
-where the $i_{j}$ are natural numbers and the $c_{i_{j}}$ are in $\Lawv$.
-For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$
-are illustrated in Fig.~\ref{fig:plot1} for $n\leq 4$.
+A \emph{tropical polynomial} is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$, where $i_{j}\in\N$ and $c_{i_{j}}\in\Lawv$.
+For example, the polynomials $\varphi_{n}(x)=\min_{0\leq j\leq n}\{jx+2^{-j}\}$
+are illustrated in Fig.~\ref{fig:plot1} for $0\leq n \leq 4$.
 A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable.
 In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes).
-For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
-With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
+For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
+Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
 $f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
+%Tropical polynomials and their roots are the main object of study of tropical geometry.
 %
@@
 
-A \emph{tropical Laurent series} (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function that can be expressed as $f(x)=\inf_{n}\{nx+\matr f_{n}\}$, with $\matr  f_{n}$ a sequence in $\Lawv$.
-In other words, a tLs is a ``limit'' of tropical polynomials of higher and higher degree.
-For instance, the function $\varphi(x):=\inf_{n\in\N}\set{nx+\frac{1}{2^n}}$ (illustrated in Fig.~\ref{fig:plot1}), that we will take as our running example,
+A \emph{tropical Laurent series} of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\{nx+\matr f_{n}\}$, with $\matr  f_{n}\in\Lawv$.
+In other words, it is a ``limit'' of tropical polynomials of higher and higher degree.
+E.g., the tLs $\varphi(x):=\inf_{n\in\N}\set{nx+2^{-n}}$ (see Fig.~\ref{fig:plot1}), that we will take as our running example, 
 is the ``limit'' of the polynomials $\varphi_{n}$.
-Since $\inf$s are not in general $\min$s, the behavior of tLS may be less predictable than that of tropical polynomials.
-For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points.
+Since $\inf$s are not in general $\min$s, the behaviour of tLS may be less predictable than that of tropical polynomials~\cite{Porzio2021}.
+%For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points of their domain.
 
-\begin{figure}
-\begin{subfigure}{0.24\textwidth}
-\begin{tikzpicture}[scale=0.55]
+%\begin{figure}
+%\begin{subfigure}{0.4\textwidth}
+%\begin{tikzpicture}[scale=0.6]
+%\begin{axis}[samples=250]
+%\addplot[yellow,domain=0:0.8] {1+0.02};
+%
+%\addplot[orange,domain=0:0.8] {min(x+1/2, 1)+0.01};
+%
+%
+%\addplot[red,domain=0:0.8] {min(2*x+1/4, x+1/2, 1};
+%\addplot[blue,domain=0:0.8] {min(3*x+1/8,2*x+1/4, x+1/2, 1)-0.01};
+%\addplot[orange,domain=0:0.8] {min(4*x+1/16,3*x+1/8,2*x+1/4, x+1/2, 1)-0.02};
+%
+%\addplot[violet,domain=0:0.8] {min(
+%10*x+1/1424,
+%9*x+1/712,
+%8*x+1/356,
+%7*x+1/128,
+%6*x+1/64,
+%5*x+1/32,
+%4*x+1/16,3*x+1/8,2*x+1/4, x+1/2, 1)-.03};
+%
+%
+%\end{axis}
+%
+%\end{tikzpicture}
+%\caption{}
+%\label{fig:plot1}
+%\end{subfigure}
+%\begin{subfigure}{0.4\textwidth}
+%\begin{tikzpicture}[scale=0.6]
+%\begin{axis}[samples=50, view={15}{45}]
+%\addplot3[orange,domain=0:2] {min(2*x, 2*x+y, 3*y)};
+%
+%\end{axis}
+%\end{tikzpicture}
+%\caption{}
+%\label{fig:plot2}
+%\end{subfigure}
+%\label{fig:plot12}
+%\caption{
+%\ref{fig:plot1}
+% (a) Plot of the tropical polynomials $\varphi_{n}$, for $0\leq n\leq 4$ (from top to bottom), and of their limit tLs $\varphi$ (in violet).
+The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x=2^{-(i+1)}$, satisfying $ix+2^{-i}=(i+1)x+2^{-(i+1)}$.
+%\ref{fig:plot2}
+%(b) Plot of $\varphi(x,y)=\min\{2x, 2x+y,3y\}$.
+%}
+%\end{figure}
+
+\begin{wrapfigure}{r}{0.5\textwidth}%\begin{figure}
+\begin{tikzpicture}[scale=0.9]
 \begin{axis}[samples=250]
@@
 \end{tikzpicture}
-\caption{}
-\label{fig:plot1}
-\end{subfigure}
-\begin{subfigure}{0.24\textwidth}
-\begin{tikzpicture}[scale=0.55]
-\begin{axis}[samples=50, view={15}{45}]
-\addplot3[orange,domain=0:2] {min(2*x, 2*x+y, 3*y)};
-
-\end{axis}
-\end{tikzpicture}
-\caption{}
-\label{fig:plot2}
-\end{subfigure}
-\label{fig:plot12}
-\caption{
-%\ref{fig:plot1}
- (a) Plot of the tropical polynomials $\varphi_{n}$, for $n\leq 4$ (from top to bottom), and of their limit tLs $\varphi$ (in violet).
-The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x=2^{-(i+1)}$, satisfying $ix+2^{-i}=(i+1)x+2^{-(i+1)}$.
-%\ref{fig:plot2}
-(b) Plot of
-$\varphi(x,y)=\min\{2x, 2x+y,3y\}$.}
-\end{figure}
+\caption{\small Tropical polynomials $\varphi_0,\dots,\varphi_4$ (top to bottom), and the limit tLs $\varphi$ (in violet).
+The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x=2^{-(i+1)}$, satisfying $ix+2^{-i}=(i+1)x+2^{-(i+1)}$.}
+\label{fig:plot1}%\end{figure}%
+\end{wrapfigure} %%
 
@@
 %\begin{remark}%[Tropical Laurent series]
- As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, but we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting 
- $t^!(x):=t\circ_! x$.
-%  (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
- This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
- Concretely, we have
- \begin{equation}
- t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}
- \end{equation}
- where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
- These functions correspond then to tLs with possibly infinitely many variables (in fact, as many as the elements of $X$). %In the following we will also refer to them as tLs.
-
-% 
+ Remark that the map $t^!:\Lawv^X\to\Lawv^Y$ introduced above from a matrix $t\in\HOM{\LREL_!}{X}{Y}$, corresponds to a tLs with \emph{possibly infinitely} many variables (in fact, as many as the elements of $X$).
+%In the following we will also refer to them as tLs.
 % We will call them simply \emph{tropical Laurent series (tLs)}.
@@
 %We find the usual notion of tLs of one variable as follows:
-
 %\begin{remark}
-Notice that, by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
+By identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the %functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ we recover the
+usual tLs's of one variable.
 % \end{remark}
@@
  Our running example is indeed of shape $\varphi=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
-% By Proposition~\ref{prop:descrete}, $f$ is not the interpretation of a $\lam$-term, because its matrix is not discrete.
+ However, it is not the interpretation of a $\lam$-term, because its matrix $t$ is not discrete.
 % Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
@@
 
- In a similar way, the tropical polynomials can be identified with the tLs
+ In a similar way, tLs
 $f:\Lawv^X\to\Lawv^Y$ 
- for which the support $\C F=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ is \emph{finite}, and which have thus shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$.
- This is again the generalisation of usual tropical polynomials to the case of infinitely many variables.
+ with \emph{finite} supports $\C F_b=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$, and which have thus shape $f(x)_b=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, are generalisation of usual tropical polynomials to the case of possibly infinitely many variables.
 
@@
 
-Looking at Fig~\ref{fig:plot1}, we see that the function $\varphi$, just like the polynomials $\varphi_{1}$, is non-decreasing and concave.
-It can be shown that this is indeed always the case:
+Looking at Fig~\ref{fig:plot1}, we see that $\varphi$, just like the polynomials $\varphi_{n}$, is non-decreasing and concave.
+This is indeed always the case:
 
@@
 %In Example~\ref{ex:famous_ex}, 
-Again, by looking at Fig~\ref{fig:plot1}, it appears that $\varphi$ behaves \emph{locally} like the polynomials $\varphi_{n}$.
-In particular, for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$.
-More, precisely, $\varphi$ coincides with $\varphi_{n}$
-for $\epsilon \geq 2^{-(n+1)}$, the smallest tropical root of $\varphi_{n}$.
+Again by looking at Fig~\ref{fig:plot1} it appears that, \emph{far from $0$}, $\varphi$ behaves like some polynomials $\varphi_{n}$.
+In particular, %for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$. More, precisely, 
+$\varphi$ coincides on $[\epsilon,\infty]$ with $\varphi_{n}$,
+for $\epsilon \geq 2^{-(n+1)}$ (the smallest tropical root of $\varphi_{n}$).
 However, at
@@
 %It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
- $x=0$ we have that $\varphi(x=0)=\inf_{n\in\N} \frac{1}{2^n}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
-Also, while the derivative of $f$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
-This phenomenon is reminiscent of [Example 7, \cite{Ehrhard2005}],
+ $x=0$ we have that $\varphi(x=0)=\inf_{n\in\N} 2^{-n}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
+Also, while the derivative of $f$ is bounded on all $(0,\infty)$, for $x\to 0^+$ it tends to $\infty$.
+This phenomenon is reminiscent of \cite[Example 7]{Ehrhard2005},
 %Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
 which actually motivated our first investigations.
-In fact, these properties are shared by all tLs with \emph{finitely} many variables, as shown by the following result.
-
-Remark that $!\set{1,\dots,k}$ can be identified with $\N^k$.
-So the matrix of a tLs $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one variable $f(x)$ in output) can be given as a $\matr f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, where $nx$ is the scalar product.
+In fact, this behaviour is shared by all tLs with \emph{finitely} many variables, as shown by the following result (we identify $!\set{1,\dots,k}\simeq \N^k$, so the matrix of a tLs $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one output variable) is given as a $\matr f:\N^k\to\Lawv$, $f$ having shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, with $nx$ the scalar product).
 
@@
   %\item 
-$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical {polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
+$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical \emph{polynomial} $P_\epsilon(x):=\min_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
 % \end{enumerate}
@@
 $x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
+Such order actually coincides with the pointwise order on $\Lawv^X$.
 This makes it a Scott-continuous dcpo.
@@
 \begin{theorem}\label{thm:ScottCont}
- All tLs $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders.
-  % on the domain and codomain.
+ All monotone (w.r.t.\ pointwise order) and $\norm{\cdot}_{\infty}$-continuous functions $f:(0,\infty)^X\to (0,\infty)$ are Scott-continuous.
+ In particular, all tLs $f:\Lawv^X\to\Lawv$ are Scott-continuous on $(0,\infty)^X$ w.r.t.\ the pointwise orders.
 \end{theorem}
+\begin{proof}
+ Use the fact, taken from \cite{Selinger2004}, that in a normed $\R_{\geq 0}$-cone $P$, considered with its cone-order, if every bounded directed net in $P$ admits a sup, and if $(v_i)_{i\in I}$ is a directed net in $P$ with an upper bound $v\in P$, then $\exists\bigvee_{i\in I} v_i \in P$ and, if $\inf_{i\in I} \norm{v-v_i} =0$, one has $\bigvee_{i\in I} v_i = v$.
+\end{proof}
 
@@
 
-The norm $\norm{\cdot}_\infty$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$.
-We will show that tLs satisfy suitable Lipschitz properties  w.r.t.~these metrics.
+%The norm $\norm{\cdot}_\infty$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$.
+We will show that tLs satisfy suitable Lipschitz properties w.r.t.\ $\norm{\cdot}_\infty$.
 

### 2023-07-21 CSL24/section7.tex
Committ
Per ora è ancora molto copia ed incolla da LICS. Fai tutte le modifiche che vuoi, se vuoi. Io domani non posso lavorare, quindi prendi il lock quando vuoi. Da domenica, dimmelo così ci organizziamo. Notte !

--- 2023-01-16 section7.tex
@@
-The connections between the differential $\lambda$-calculus (and differential linear logic), the relational semantics, and the theory of non-idempotent intersection types is very well-studied (see \cite{decarvalho2018}, and more recently, \cite{Mazza2016} for a more abstract perspective, and \cite{Olimpieri2021, Galal2021} for a 2-categorical, or proof-relevant, extension).
+The connections between differential $\lambda$-calculus (and differential linear logic), relational semantics, and non-idempotent intersection types are very well-studied (see \cite{decarvalho2018}, and more recently, \cite{Mazza2016} for a more abstract perspective, and \cite{Olimpieri2021, Galal2021} for a 2-categorical, or proof-relevant, extension).
 As we said, the relational semantics over the tropical semi-ring was quickly explored in \cite{Manzo2013}, to provide a ``best case'' resource analysis of a $\B{PCF}$-like language with non-deterministic choice.
@@
 The study of linear or bounded type systems for sensitivity analysis was initiated in \cite{Girard92tcs} and later developed \cite{Schopp, SchoppDalLago, Reed2010}.
-As recalled in the paper, the use bounded exponentials ensures that well-typed programs satisfy a Lipschitz condition.
+%As recalled in the paper, the use bounded exponentials ensures that well-typed programs satisfy a Lipschitz condition.
 Related approaches, although not based on metrics, are provided by \emph{differential logical relations} \cite{dallago} and \emph{change action} models \cite{Picallo2019}.

### 2023-07-21 CSL24/tropicalization.tex
Committ
Per ora è ancora molto copia ed incolla da LICS. Fai tutte le modifiche che vuoi, se vuoi. Io domani non posso lavorare, quindi prendi il lock quando vuoi. Da domenica, dimmelo così ci organizziamo. Notte !

--- 2023-01-22 tropicalization.tex
@@
 $\trop f= \lim_{t\to 0}\trop_{t}f$. 
-Indeed, one can check that the ``parameterized'' sums and product $\alpha \sumt{t}\beta:= \phi_{t}(\psi_{t}(\alpha)+\psi_{t}(\beta))= \log_{t}(t^{-\alpha}+t^{-\beta})$ and 
+Indeed, one can check that the ``parameterized'' sums and product $\alpha \sumt{t}\beta:= \phi_{t}(\psi_{t}(\alpha)+\psi_{t}(\beta))= -\log_{t}(t^{-\alpha}+t^{-\beta})$ and 
 $\alpha \prodt{t}\beta:= \phi_{t}( \psi_{t}(\alpha)\psi_{t}(\beta))=
-\log_{t}(t^{-\alpha}t^{-\beta})$ converge respectively to $\min\{\alpha,\beta\}$ and $\alpha+\beta$, when $t\to 0$.
+-\log_{t}(t^{-\alpha}t^{-\beta})$ converge respectively to $\min\{\alpha,\beta\}$ and $\alpha+\beta$, when $t\to 0$.

### 2023-07-23 CSL24/section1.tex
Commit

--- 2023-01-24 section1.tex
@@
 The main reason to look for methods relying on (quantitative extensions of) type-theory or denotational semantics is that these approaches yield \emph{modular} and \emph{compositional} techniques, that is, allow one to deduce properties of complex programs from the properties of their constituent parts.
+Among such quantitative approaches, two different directions have received considerable attention, which we recall below.
 
-\subsection{Two kinds of quantitative approaches}
+%\subsection{Two kinds of quantitative approaches}\label{section2}
 
-Among such quantitative approaches, two different directions have received considerable attention.
+\input{Section2}
 
-On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}.
+%On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}.
 This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
 
-On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
+%On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
 This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
-
-In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
-In metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
-By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
-
-A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled.
-At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
-in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz.
-The fundamental motivation of this work is then the observation that
-this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
 
@@
 
+In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
+In metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
+By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
+
+A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled.
+At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
+in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz.
+The fundamental motivation of this work is then the observation that
+this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
+
 Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} as an alternative approach to algebra and geometry where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
@@
 $.
-
 %This is not a \emph{ad-hoc} setting: 
@@
 
-Coming back to our discussion on program semantics, tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
+\subsection{This paper and our contributions}
+
+Coming back to our discussion on program semantics %tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
+the metric approach usually yields Lipschitz conditions for affine programs, but what about ``unbounded'', simply typed $\lambda$-calculus (the ordinary one)?
+In such case, one might lose the Lipschitz property, so the programs might be \emph{too} sensitive to errors. 
+%For instance, while the functions $M_{k}=\lambda x.
+k\cdot x: \BB R\to \BB R$ are all Lipschitz-continuous, with Lipschitz constant $k$, the program $M=\lambda x.x^{2}$ obtained by ``duplicating'' $x$ is not Lipschitz anymore: $M$ is, so to say, \emph{too} sensitive to errors.
+It is indeed well-known that the category $\Met$ is \emph{not} Cartesian closed, so it is not a model of $\STLC$ (yet, several Cartesian closed \emph{sub-}categories of $\Met$ do exist, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
+%Still, one might observe that the program $M$ above is actually \emph{locally} Lipschitz-continuous (i.e.~Lipschitz over any open ball).
+However, one can still consider Cartesian closed categories of \emph{locally} Lipschitz maps  (i.e.~Lipschitz over any open ball), see e.g.\ \cite{Ehrhard2011, PistoneLICS}, and a new example will be considered in this paper.
+
+On the other hand, since the resource approach decomposes, through the Taylor formula, an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as a decomposing the associated locally Lipschitz map by a limit of the associated Lipschitz ones, thus bridging the metric and differential approaches.
+Here, a natural direction to look for is the \emph{relational semantics}, since strictly related with the Taylor expansion $\STDLC$ {\color{red}CITE DeCarvalho ecc}. 
+However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.
+
+Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions?
+This way, the Taylor expansion could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps: unbounded term application could be seen as a limit of \emph{more and more sensitive} operations.
+
+%This viewpoint, that we develop in the following sections, not only suggests the application of optimization methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions, but it scales to a more abstract level, leading to introduce a differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}).
+%
+%At this point, 
+%by interpreting such polynomials over the tropical semiring 
+%
+%Our question can thus be reformulated as follows: can we make the relational semantics \emph{Lipschitz}, hence amenable to metric and sensitivity analysis?
+The goal of this paper is to show that, by appealing to tropical mathematics, this is indeed possible and leads to the somehow unespected discovery of a bridge between the metric and differential study to higher-order programs.
+
 At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.
@@
 
+%\subsection{Contributions}
+
 In this paper we demonstrate that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provides the desired bridge between differential and metric semantics.
 Moreover, we show that the conceptual unification of these two approaches suggests ways in which techniques from resource-analysis could be used in sensitivity analysis and \emph{vice-versa}, paving the way for new  applications of tropical geometry to the  study of higher-order programs.
-
-\subsection{Contributions}
 

### 2023-07-23 CSL24/section2.tex
Commit

@@
 
-\subsection{Metric approach: bounded $\lambda$-calculus in a nutshell}
+\paragraph*{Metric approach: bounded $\lambda$-calculus in a nutshell}
 
-In many situations (e.g.~when dealing with computationally difficult problems) one does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound.
+On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask whether they compute functions which do not differ \emph{too much} by controlling the propagation of an error during the computation.
+This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
+%As already mentioned, in many situations (e.g.~when dealing with computationally difficult problems, or in differential privacy \cite{Alvim2011, Reed2010}), one %does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound.
 In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, a small error in the input will produce a comparably small error in the output.
-
 %In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior.
@@
 $d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
-A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the \emph{linear} simply typed $\lambda$-calculus, being a \emph{symmetric monoidal closed} category, and in fact it also models {affine} terms (since the cartesian and monoidal units coincide).
-
-As observed in \cite{Reed2010, Gaboardi2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications.
-The fundamental intuition is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz} map.
-For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the functional input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.
-By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $r\cdot X$ (i.e.~with distance $d_{r}(x,y)=r\cdot d(x,y)$) and $Y$, the program $M$ above
-can thus be interpreted as a non-expansive map from $2\cdot(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.
+This is due to the fact that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the %\emph{linear} simply typed $\lambda$-calculus, being a \emph{symmetric monoidal closed} category, and in fact it also models
+\emph{affine} $\lambda$-calculus (it is symmetric monoidal closed, cartesian and monoidal units coincide). 
+The metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications \cite{Reed2010, Gaboardi2017}: the idea is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz} map.
+For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the function input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself.
+%: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.
+By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $rX:=(X,%d_{r}(x,y)=
+rd(x,y))$ and $Y$, the program $M$ above 
+can thus be interpreted as a non-expansive map from $2(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.
 
@@
 The language $\BSTLC$ we consider is a simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}. 
-The terms are as for the $\STLC$, the types are $A::= o \ \mid  \ !_{n}A \multimap A$ and type judgements are of the form $\Gamma \vdash M:A$, where a context $\Gamma$ is a set of declarations of the form $x :_{n}A$, with $n\in \mathbb N$.
+The terms are as for the $\STLC$, the types are $A::= * \ \mid  \ !_{n}A \multimap A$ and type judgements are of the form $\Gamma \vdash M:A$, where a context $\Gamma$ is a set of declarations of the form $x :_{n}A$, with $n\in \mathbb N$.
 The typing rules are given in Fig.~\ref{fig:rules}, where $\Gamma+\Delta$ is defined letting $(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) =  (\Gamma+\Delta), x:_{m+n}A$, and $m\Gamma$ is made all $x:_{mn}A$ for $(x:_{n}A) \in \Gamma$.
@@
 E.g., $\vdash_{\BSTLC} \lambda {\color{red}z}.\left( \lambda x{\color{green}y}.
-yxx \right)z : \, !_{\color{red}2} o\multimap !_{\color{green}1}(!_{\color{violet}1} o \multimap !_{\color{blue}1} o \multimap o) \multimap o$.
-
+yxx \right)z : \, !_{\color{red}2} *\multimap !_{\color{green}1}(!_{\color{violet}1} * \multimap !_{\color{blue}1} * \multimap *) \multimap *$.
 The $\BSTLC$ can be interpreted in a symmetric monoidal closed category (SMCC) equipped with a \emph{$\mathbb N$-graded linear exponential comonad} \cite{Katsumata2018}.
@@
 
-\subsection{Resource approach: differential $\lambda$-calculus in a nutshell}
+\paragraph*{Resource approach: differential $\lambda$-calculus in a nutshell}
 
-A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models.
+On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
+This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
+
+%A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models.
 The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus.
@@
 %Sums are, then, just \emph{finite} sets of terms.
-Simple types are $A::= o\mid A\to A$.
+Simple types are $A::= *\mid A\to A$.
 The typing system is in Figure~\ref{fig:rules}, where a context $\Gamma$ is a list of typed variable declarations.
@@
 , z^{\color{red}2}]0
-: {\color{red}o}\to ({\color{green}o \to o \to o}) \to o$.
+: {\color{red}*}\to ({\color{green}* \to * \to *}) \to *$.
 %Here we wrote $\Der^2[\_]\cdot (\_)^2$ as a shortcut for $\Der[\Der[\_]\cdot (\_)]\cdot (\_)$ and $\Der^1[\_]\cdot (\_)^1$ for $\Der[\_]\cdot (\_)$.
@@
 , z^{\color{red}3}]0
-: o\to (o \to o \to o) \to o$, then the term reduces to the empty sum $0$ (representing an \emph{error}).
+: *\to (* \to * \to *) \to *$, then the term reduces to the empty sum $0$ (representing an \emph{error}).
 

### 2023-07-23 CSL24/section3A.tex
Commit

@@
 It can be interpreted in any Symmetric Monoidal Closed Category (SMCC).
-Now, \cite[Section III.A]{Manzo2013} immediately gives that $\LREL$ is a SMCC (actually, even a model of IMLL), where its monoidal product $\otimes$ acts on the objects as the Cartesian product of sets, and as the \emph{Kronecker product}
+Now, \cite[Section III.A]{Manzo2013} immediately gives that $\LREL$ is a SMCC (actually, even a model of IMLL), where its monoidal product $\otimes$ acts on the objects as the Cartesian product of sets, %and as the \emph{Kronecker product} of matrices on morphisms.
  %{(\color{red} Ref?)} 
- of matrices on morphisms.
-Its internal hom $X\multimap Y$ is also given by the Cartesian product $X\times Y$ of sets, with the usual evaluation and curry maps.
+ and its internal hom $X\multimap Y$ is also given by the Cartesian product $X\times Y$ of sets, with the usual evaluation and curry maps.

### 2023-07-23 CSL24/section3B.tex
Commit

@@
 %It is instructive what its CCC-structure looks like in our tropical world.
-Remark that the cartesian product of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_!
+The cartesian product of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_!
 t\in\Lawv^{!X\times Z}$ where $(s\circ_!

### 2023-07-23 CSL24/section4.tex
Commit

@@
 % simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.
-
-A \emph{tropical Laurent series} of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\{nx+\matr f_{n}\}$, with $\matr  f_{n}\in\Lawv$.
-In other words, it is a ``limit'' of tropical polynomials of higher and higher degree.
-E.g., the tLs $\varphi(x):=\inf_{n\in\N}\set{nx+2^{-n}}$ (see Fig.~\ref{fig:plot1}), that we will take as our running example, 
-is the ``limit'' of the polynomials $\varphi_{n}$.
-Since $\inf$s are not in general $\min$s, the behaviour of tLS may be less predictable than that of tropical polynomials~\cite{Porzio2021}.
-%For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points of their domain.
 
@@
 \end{wrapfigure} %%
+
+A \emph{tropical Laurent series} of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\{nx+\matr f_{n}\}$, with $\matr  f_{n}\in\Lawv$.
+In other words, it is a ``limit'' of tropical polynomials of higher and higher degree.
+E.g., the tLs $\varphi(x):=\inf_{n\in\N}\set{nx+2^{-n}}$ (see Fig.~\ref{fig:plot1}), that we will take as our running example, 
+is the ``limit'' of the polynomials $\varphi_{n}$.
+Since $\inf$s are not in general $\min$s, the behaviour of tLS may be less predictable than that of tropical polynomials~\cite{Porzio2021}.
+%For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points of their domain.
 
@@
 the bijection between $\HOM{\LREL}{X}{Y}$ and set of tropical linear functions from $\Lawv^{X}$ to $\Lawv^{Y}$.
+
 \begin{proposition}
-For all tropical linear functions $f,g:\Lawv^{X}\to \Lawv^{Y}$, $\norm{ \matr f-\matr g}_{\infty} =  \sup_{x\in \Lawv^{X}}
-\norm{ f( x)-g(x)}_{\infty}$.\end{proposition}
+For all tropical linear functions $f,g:\Lawv^{X}\to \Lawv^{Y}$, $d_{\infty}(\matr f,\matr g)=d_\infty(f,g)$.% $\norm{ \matr f-\matr g}_{\infty} =  \sup_{x\in \Lawv^{X}} \norm{ f( x)-g(x)}_{\infty}$.
+\end{proposition}
 
@@
 \begin{proposition}\label{prop:boundedlip}
-If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a bounded matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
+If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
 \end{proposition}
@@
 
-Moreover, from Proposition \ref{prop:boundedlip}, we can also deduce the desired Lipschitz decomposition of the Taylor expansion (using Corollary~\ref{cor:T(M)=M}):
-
-\begin{corollary}\label{thm:M=infLip}
- If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is an $\inf$ of \emph{Lipschitz} functions (via its Taylor expansion \eqref{eq:taylorcat}).
-\end{corollary}
-
 Let us now look at what happens with tLs, i.e.~when considering the full exponential $!$.
@@
 The core of the proof is a convex analysis argument (see the Appendix) showing that an arbitrary function $f:\Lawv^X\to\Lawv$ which is non-decreasing, concave and continuous, must be locally Lipschitz.
-\end{proof}
-
-The results just presented translate into the following facts about the interpretation of higher-order programs:
-\begin{corollary}
-For any $\lambda$-term $M$:
-\begin{enumerate}
-\item if $\Gamma \vdash_{\BSTLC} M:A$, then $\model M^!:\model\Gamma \to \model A$ is a Lipschitz map.
-\item if $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is a locally Lipschitz map.
-\end{enumerate}
-\end{corollary} 
-\begin{proof}
-As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be a finite matrix  and  its corresponding linear map $\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Corollary~\ref{prop:polylip}.
-Claim 2) follows directly from Theorem~\ref{thmTLSlocLip}.
 \end{proof}
@@
 
+The results just presented translate into the following facts about the interpretation of higher-order programs:
+
+\begin{corollary}
+Let $\model A$ be a finite set.
+\begin{enumerate}
+\item $\model{\Gamma \vdash_{\BSTLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{Lipschitz} map.
+\item $\model{\Gamma \vdash_{\STLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{locally} Lipschitz map.
+\item $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!$ of \emph{tropical polynomials}, thus (as $\model A$ is finite), \emph{Lipschitz} functions.
+\end{enumerate}
+\end{corollary} 
+\begin{proof}
+1).
+As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be a finite matrix  and  its corresponding linear map $\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Corollary~\ref{prop:polylip}.
+2). It follows directly from Theorem~\ref{thmTLSlocLip}.
+3).
+It follows from the easily checked fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have $\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with $\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
+\end{proof}
+
+Remark that the restriction to the ground type $*$ in (2) is without loss of generality, thanks Currying.
+
+%Moreover, from Proposition \ref{prop:boundedlip}, we can also deduce the desired Lipschitz decomposition of the Taylor expansion (using Corollary~\ref{cor:T(M)=M}):
+
+%\begin{corollary}\label{thm:M=infLip}
+% If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is an $\inf$ of \emph{Lipschitz} functions (via its Taylor expansion \eqref{eq:taylorcat}).
+%\end{corollary}
+
 %A consequence of this result is that the distance between two differential programs can always be approximated via bounded programs:
@@
 %For all $\lambda$-terms $M,N: A\to B$ of $\STDLC$, for all $\epsilon >0$, there exists $K\in \BB N$ such that 
-%$|\|\model{M}-\model{N}\|_{\infty}- | \| \model{
+%$|\|\model{M}-\model{N}\|_{\infty}- |\| \model{
 %\end{corollary}

### 2023-07-24 CSL24/section4.tex
Commit giornaliero

@@
 are illustrated in Fig.~\ref{fig:plot1} for $0\leq n \leq 4$.
-A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable.
-In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes).
-For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
-Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
-$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
-%Tropical polynomials and their roots are the main object of study of tropical geometry.
-%
-%A
-% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.
 
@@
 \end{wrapfigure} %%
+
+A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable.
+In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes).
+For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
+Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
+$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
+%Tropical polynomials and their roots are the main object of study of tropical geometry.
+%
+%A
+% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.
 
@@
 \begin{enumerate}
-\item $\model{\Gamma \vdash_{\BSTLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{Lipschitz} map.
+\item $\model{\Gamma \vdash_{\BSTLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{tropical polynomial}, thus (as $\model A$ is finite), a \emph{Lipschitz} function.
 \item $\model{\Gamma \vdash_{\STLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{locally} Lipschitz map.
@@
 
-Remark that the restriction to the ground type $*$ in (2) is without loss of generality, thanks Currying.
+Remark that the restriction $\model A$ finite is without loss of generality, since by Currying all programs can be seen having type $*$, which is natural to interpret as a singleton.
 
-%Moreover, from Proposition \ref{prop:boundedlip}, we can also deduce the desired Lipschitz decomposition of the Taylor expansion (using Corollary~\ref{cor:T(M)=M}):
+A consequence of (3) is that the distance between two programs can always be approximated via Lipschitz tropical polynomial approximants of the initial two programs.
+\begin{corollary}
+ Let $\Gamma \vdash_{\STLC} M:A$ and $\Delta\vdash_{\STLC} N:B$.
+ For all $\epsilon>0$, $x\in\Lawv^{\model \Gamma}$, $b\in\model A$, there exist $t\in\Te{M}$, $u\in\Te{N}$ s.t.\ $\big| \model{\Gamma \vdash_{\STLC} M:A}^!(x)_b - \model{\Delta \vdash_{\STLC} N:B}^!(x)_b \big| \leq 2\epsilon + \big| \model{\Gamma \vdash_{\STDLC} t:A}^!(x)_b - \model{\Delta \vdash_{\STDLC} u:B}^!(x)_b \big|$.
+\end{corollary}
 
-%\begin{corollary}\label{thm:M=infLip}
-% If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is an $\inf$ of \emph{Lipschitz} functions (via its Taylor expansion \eqref{eq:taylorcat}).
-%\end{corollary}
-
-%A consequence of this result is that the distance between two differential programs can always be approximated via bounded programs:
-%\begin{corollary}
-%For all $\lambda$-terms $M,N: A\to B$ of $\STDLC$, for all $\epsilon >0$, there exists $K\in \BB N$ such that 
-%$|\|\model{M}-\model{N}\|_{\infty}- |\| \model{
-%\end{corollary}
+\begin{remark}
+ Let $X$, $Y$ be sets and let $\langle \_,\_\rangle:X\times Y \to \mathbb{R}$.
+ For $f:X\to \mathbb R$, define $f^*:Y\to \mathbb R$ by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
+ Then for $X=!A$, $Y=\Lawv^A$, where $A$ is a set, and $\langle \mu, y \rangle:= \mu y$, we have that $f^!=(-f)^*$ for all $f\in\Lawv^{!A}$.
+ This is precisely the same formal construction yielding the well-known convex conjugate $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
+ This construction is in turn a generalisation of the Legendre transformation.
+ Despite the formal constructions being the same, we ignore for the moment if these could be connected to the study of high-order programs in our setting.
+\end{remark}

### 2023-07-24 CSL24/section5.tex
Committ
Per ora è ancora molto copia ed incolla da LICS. Fai tutte le modifiche che vuoi, se vuoi. Io domani non posso lavorare, quindi prendi il lock quando vuoi. Da domenica, dimmelo così ci organizziamo. Notte !
Commit giornaliero

--- 2023-05-14 section5.tex
@@
-In this section we illustrate a few directions in which the tropical semantics just introduced could be used to analyze quantitative properties of higher-order programs.
+%In this section we illustrate a few directions in which the tropical semantics just introduced could be used to analyze quantitative properties of higher-order programs.
 
@@
 what properties of the model can be deduced from the \emph{tropicalization} of $f$, noted $\trop f$, i.e.~the transformation of $f$ into a tropical polynomial.
-
 %
@@
 %
-
 Here we follow a similar pattern: we consider a program $M$, which can be expressed in the form of a polynomial or a power series $f$, and we 
@@
 \input{tropicalization}
+
+\begin{comment}
 
@@
 
-\subsection{Convergence log-probabilities}
+\end{comment}
+
+\subsection{Log-probabilities with a worked out example}
 
 Let us consider a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
-we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the rules:
-\[
+we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the usual rules and whose operational semantics adds the usual rules:
+{\small{\[\begin{array}{cccc}
            \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
- \qquad\qquad
+           &
            \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
-\]
-We add reduction rules:
-\[
- M\oplus_p N \to pM \quad \textit{ and } \quad M\oplus_p N \to (1-p)N
-\]
+           &
+           M\oplus_p N \to pM
+           &
+           M\oplus_p N \to (1-p)N
+          \end{array}\]}}
+%We add reduction rules:$ M\oplus_p N \to pM \quad$ and $\quad M\oplus_p N \to (1-p)N,$
 so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
-We use this calculus as a toy example for our purposes, and it can be seen as a fragment of the PCF considered in \cite{Manzo2013}.
+We use this calculus as a toy example for our purposes.
+It can be compiled into $\mathrm{PCF}^{\Lawv}$, an instance of the $\mathrm{PCF}^{\mathcal R}$-languages studied in \cite{Manzo2013}, via the translation $(\_)^\circ$ of terms generated by $(M\oplus_p N)^\circ:=pM^\circ\prog{ or } (1-p)N^\circ$.
+It is easy to see that if $\Gamma \vdash_{\STLC_\oplus} M:A$, then $\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A$.
+This is slightly imprecise because in \cite{Manzo2013}, their $\mathrm{PCF}^{\mathcal R}$-languages only have one ground type $\mathrm{int}$, used for modeling $\N$, and has no term $\prog{True}$, $\prog{False}$.
+So we are actually compiling $\STLC_\oplus$ into the immediate extension of $\mathrm{PCF}^{\mathcal R}$ which adds two new ground types, $*$ and $\bool$ (and, since we will not consider the integers, we can also drop $\mathrm{int}$), plus the two boolean terms.
+This does not invalidate the results (not concerning $\mathrm{int}$) of \cite{Manzo2013}.
+Remark also that the operational semantics of $\mathrm{PCF}^{\Lawv}$ \cite[Fig.\ 1]{Manzo2013} simulates the one of $\STLC_\oplus$, potentially using multiple steps, e.g.\ $(M\oplus_p N)^\circ \overset{0}{\rightarrow} pM^\circ \overset{p}{\rightarrow} M^\circ$.
+Since $\LREL_!$ is an instance of models for $\mathrm{PCF}^{\mathcal{R}}$, similarly to \cite[Section VI]{Manzo2013}, we use it in order to get a model of $\STLC_\oplus$, by setting $\model{\Gamma \vdash_{\STLC_\oplus} M:A}:=\model{\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A}\in\HOM{\LREL_!}{\model{\Gamma}}{\model{A}}$.
+Here we set $\model{\bool}:=\{0,1\}$ and $\model{\vdash_{\mathrm{PCF}^{\Lawv}} \true:\bool}\in\Lawv^{\{0,1\}}$ gives value $0$ on $1$ and value $\infty$ on $0$, similarly for $\false$.
 
-Consider now the following term $M$ of type $\bool$:
-\[
- (\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true))
-\]
+\begin{remark}
+ Now the matrix interpreting a term need no longer be boolean.
+ For instance, $\model{\vdash_{\STLC_\oplus} \true:\bool}_i$ is $p$ if $i=1$ and it is $1-p$ if $i=0$.
+ In general, one can show that $\model{\Gamma \vdash_{\STLC_\oplus} M:A}\in[0,1]^{!\model \Gamma \times \model A}$.
+ Coefficients $> 1$ may be considered when dealing with other kinds of effects.
+\end{remark}
+
+Consider %the following closed term $M$ of type $\bool$:
+$
+ \vdash_{\STLC_\oplus} M:=(\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true)):\bool.
+ $
 Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, reading $0$ as ``left'' and $1$ as ``right''.
 The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
-For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$ starting from the left.
+%For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$ starting from the left.
 Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
@@
 passing to the tropical monomials/polynomial $\trop P_{\omega},\trop Q_{\omega}$. 
-For 1), we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)$ belong to:
-\begin{equation}
-  \begin{array}{r@{}c@{}l@{}}
+For 1), we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)\in
+%\begin{equation}
+%  \begin{array}{ccccc}
    \underset{(x,y)}{\RM{arg max}} P_\omega (x,y)
-   & = &
+   %& 
+   = %&
    \underset{(x,y)}{\RM{arg min}}\set{-\log_c P_\omega (x,y)}
-   \\
-   & = &
-   \underset{(x,y)}{\RM{arg max}}\set{(\trop P_\omega) (-\log_c x,-\log_c y)} \label{eq:argmax}
-  \end{array}
-\end{equation}
+   %&
+   =%&
+   \underset{(x,y)}{\RM{arg max}}\set{(\trop P_\omega) (-\log_c x,-\log_c y)} \label{eq:argmax}$
+%  \end{array}
+%\end{equation}
 where the equalities hold for all $0<c\neq1$.
-
 For 2), %the $\omega_0$ is s.t.\ $\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$. So
@@
 $\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$.
-This last condition is equivalent to ask that:
-\begin{equation}
-  \begin{array}{l@{}l@{}}
-    \min\limits_{\omega\in\set{00,100,111}} \, -\log_c P_\omega(x,y) = -\log_c P_{\omega_0}(x,y)
-   & \textit{ i.e.: }
-   \\
-   (\trop Q_1)( -\log_c x, -\log_c y) = (\trop P_{\omega_0})( -\log_c x, -\log_c y). &  \label{eq:max}
-  \end{array}
-\end{equation}
+This last condition is equivalent to ask that $\min\limits_{\omega\in\set{00,100,111}} \, -\log_c P_\omega(x,y) = -\log_c P_{\omega_0}(x,y)$, i.e.\ $(\trop Q_1)( -\log_c x, -\log_c y) = (\trop P_{\omega_0})( -\log_c x, -\log_c y)\label{eq:max}$.
 In both cases, passing through $\trop $ %P_\omega, \trop Q_\omega$ 
@@
 \begin{remark}
- By adapting [Section IV, \cite{Manzo2013}], it can be seen that $\LREL_!$ is a model of $\STLC_\oplus$.
- In particular, if we set $\model \bool:=\set{0,1}$, our running example $M$ is interpreted as $\model M\in\Lawv^{\set{0,1}}$ giving the tropicalised probabilities: $\model M_0=(\trop Q_0)(p,1-p)$, $\model M_1=(\trop Q_1)(p,1-p)$.
+ The interpretation $\model{\vdash_{\STLC_\oplus} M:\bool}\in\Lawv^{\set{0,1}}$ of our running example gives the tropicalised probabilities: $\model M_0=(\trop Q_0)(p,1-p)$, $\model M_1=(\trop Q_1)(p,1-p)$.
+ Therefore, the model $\LREL_!$ expresses such optimisation problems 
+% (in $p$ and $q$, not after the substitution $q:=1-p$) 
+% are extracted from it,
+ \ref{eq:argmax}, \ref{eq:max} of $\STLC_\oplus$-programs. 
 \end{remark}
@@
 %\[\model M = \min\left\{\trop{Q_1}(p,1-p) \cdot \model I, \trop{Q_0}(p,1-p) \cdot \model \Omega\right\}.\]
-Therefore, assuming the interpretation of a term is known, 
-so that one can extract the associated tropical polynomials from it,
-% (in $p$ and $q$, not after the substitution $q:=1-p$) 
-% are extracted from it, 
-one could study the optimisation problems \ref{eq:argmax}, \ref{eq:max} \emph{without} having to actually reduce the term.
 
@@
 For our running example $M$, we have $\trop Q_1(x,y)=\min\set{2x,y+2x,3y}$ and $\trop Q_0(x,y)=\min\set{x+y,2y+x}$.
-Studying $\trop Q_1$, whose plot is in Fig.~\ref{fig:plot2}, we see that $\trop Q_1(x,y)=3y$ iff $y\leq \frac{2}{3}x$, and it coincides with $2x$ otherwise.
+Studying $\trop Q_1$ %, whose plot is in Fig.~\ref{fig:plot2}, 
+we see that $\trop Q_1(x,y)=3y$ iff $y\leq \frac{2}{3}x$, and it coincides with $2x$ otherwise.
 Remembering that $3y=P_{111}(x,y)$, we can now solve the optimisation problem~\ref{eq:max} for $\omega_0=111$:

### 2023-07-25 CSL24/main.tex
Commit
Commit giornaliero
Committino notturno

@@
 
-\title{Linearity and tropical mathematics in the Lambda-calculus} %TODO Please add
+\title{Tropical mathematics for the Lambda-calculus} %TODO Please add
 
@@
 
-\section{Introduction}
+\section{Two kinds of quantitative approaches}
 \input{section1}
@@
 %\section{The Two Quantitative Approaches to the $\lambda$-calculus}\label{section2}
-\input{section2}
+%\input{section2}
 
@@
 
-{\color{red}2 pagine in totale}
+{\color{red}3 pagine in totale}
 
-\subsection{Quantitaive analysis: probabilistic, differential privacy}\label{sec:app}
-%\input{section5}
+\subsection{Quantitaive analysis%: probabilistic, differential privacy
+}\label{sec:app}
+\input{section5}
 %\input{section5C}
-
-\newpage
 

### 2023-07-25 CSL24/section4.tex
Committino notturno

@@
  with \emph{finite} supports $\C F_b=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$, and which have thus shape $f(x)_b=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, are generalisation of usual tropical polynomials to the case of possibly infinitely many variables.
-
+For instance, for a $t\in\HOM{\LREL}{!_n X}{Y}$ (this is in particular the case for the interpretations of $\BSTLC$-terms), we will consider the (generalised) tropical polynomial $t^!:\Lawv^X\to\Lawv^Y$ defined by $t^!(x)_b:=\inf_{\mu\in !_nX}\{\mu x+t_{\mu,b}\}$.
 % 
@@
 \begin{proof}
-1).
-As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be a finite matrix  and  its corresponding linear map $\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Corollary~\ref{prop:polylip}.
-2). It follows directly from Theorem~\ref{thmTLSlocLip}.
+1). We already observed that the interpretation of a $\BSTLC$-type $B$ is alway a finite set, so thus is $!_n \model B$.
+So the $\inf$ defining $(\_)^!$ is actually a $\min$.
+So the interpretation of a bounded term is a tropical polynomial, so we apply Corollary~\ref{prop:polylip} to each coordinate of the image, and since $\model A$ is finite (by taking the maximum Lipschitz constant among the finite number of them), we obtain the thesis.
+2). It follows immediately from Theorem~\ref{thmTLSlocLip} and the fact that $\model A$ is finite.
 3).
-It follows from the easily checked fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have $\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with $\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
+It follows from \autoref{cor:T(M)=M} plus the easily checked fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have $\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with $\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
 \end{proof}

### 2023-07-25 CSL24/section5.tex
Committino notturno

@@
  Now the matrix interpreting a term need no longer be boolean.
- For instance, $\model{\vdash_{\STLC_\oplus} \true:\bool}_i$ is $p$ if $i=1$ and it is $1-p$ if $i=0$.
- In general, one can show that $\model{\Gamma \vdash_{\STLC_\oplus} M:A}\in[0,1]^{!\model \Gamma \times \model A}$.
+ For instance, $\model{\vdash_{\STLC_\oplus} \true\oplus_p\false :\bool}_i$ is $p$ if $i=1$ and it is $1-p$ if $i=0$.
+ %In general, one can show that $\model{\Gamma \vdash_{\STLC_\oplus} M:A}\in[0,1]\cup\{\infty\}^{!\model \Gamma \times \model A}$.
 Coefficients $> 1$ may be considered when dealing with other kinds of effects.

### 2023-07-25 CSL24/section4.tex
Committino pomeridiano

@@
 In this section we establish some topological and metric properties of such functions, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
-As we explain in a moment, those are generalised \emph{tropical Laurent series} (\emph{tLs} for short), and as the literature in tropical mathematics is often recent (e.g.~\cite{Porzio2021}), several results we prove in this section are, to our knowledge, new.
+In analogy with \cite{DanEhrh2011}, we could call them \emph{tropical entire functions}, but since, as we explain in a moment, those are generalised \emph{tropical Laurent series}, we will use the latter terminology, or \emph{tLs} for short.
+The literature in tropical mathematics is often recent (e.g.~\cite{Porzio2021}), and several results we prove in this section are, to our knowledge, new.
+
+Remark that the map $(\_)^!$ is a surjective homomorphism (w.r.t.\ coKleisli composition and identity and function composition and identiy) on its image.
+{\color{red}In Ehrhard-Danos fanno vedere (Lemma 14) che \`e anche iniettivo.
+Non riesco a capire se anche a noi lo \`e, ma non vedo perch\'e non dovrebbe.}
 
@@
 The tLs $\varphi$ is continuous on $\BB R_{\geq0}$ (w.r.t.\ the usual norm of real numbers).
-By considering the usual norm $\norm{x}_\infty:=\sup_{a\in X} \absv{x_a}$ on $\Lawv^X$, we can generalise this property by dropping the case of $x$ having some $0$ coordinate:
+By considering the usual norm $\norm{x}_\infty:=\sup_{a\in X} x_a$ on $\Lawv^X$, we can generalise this property, dropping the case of $x$ having some $0$ coordinate:
 
 \begin{theorem}\label{thm:cont}
- All tLs $f:\Lawv^X\to\Lawv$ are continuous on $\BB R_{>0}^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
+ All tLs $f:\Lawv^X\to\Lawv$ are continuous on $(0,\infty)^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
 \end{theorem}
 \begin{proof}
- It follows after adapting [Proposition 4.4, \cite{Cobzas2017}] in order to prove that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
+%It follows after adapting \cite[Proposition 4.4]{Cobzas2017} in order to prove that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
+\cite[Proposition 4.4.(3)]{Cobzas2017} shows that if a real-valued convex function with domain a convex open subset of a locally convex topological $\BB R$-vector space is, locally around any point, bounded from above by a finite non-zero constant, then it is continuous on all its domain.
+Now, our $f$ is concave, so $-f$ is convex.
+Also, since $f\geq 0$ on all $\Lawv^X$, we have e.g.\ $-f\leq 1$ on $\Lawv^X$.
+Remark that we cannot immediately apply \cite[Proposition 4.4.(3)]{Cobzas2017}, since $\Lawv^X$ is not even a $\BB R$-vector space.
+But $(0,\infty)^X\subseteq \BB R^X$ is open and convex, and $\BB R^X$ is a locally convex topological $\BB R$-vector space.
+Thus \cite[Proposition 4.4.(3)]{Cobzas2017} entails the continuity of $-f\big|_{(0,\infty)^X}$, hence that of $f\big|_{(0,\infty)^X}$.
 \end{proof}
@@
 Suitable categories of cones have been recently investigated as models of probabilistic computation (\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
-Here we will just mention that:
+In analogy with \cite[Proposition 17]{DanEhrh2011}, we have:
 

### 2023-07-26 CSL24/main.tex
Commit notturno

@@
 \usepackage{wrapfig}
+\usepackage{relsize}
 
@@
 \begin{abstract}
-In this paper we study an interpretation of the $\lambda$-calculus in the framework of tropical mathematics, by considering the relational semantics weighted on the tropical semiring.
-We consider it as providing a unified framework where both program metrics, as based on the analysis of program sensitivity via Lipschitz-conditions, and resource analysis, as based on higher-order program differentiation, coexist.
-This is possible thanks to the fact that this is also a model of both a graded lambda-calculus ensuring bounded duplications, and the differential lambda-calculus.
-We only show the starting result of our line of research: thanks to the tropical operations we consider, in such model programs with bounded duplication are Lipschitz maps, simply typed terms are locally Lipschitz maps and the differential terms appearing in the Taylor expansion of a term (also called resource terms) are Lipschitz maps.
-Therefore, the lambda-calculus Taylor expansion decomposes semantically the simply typed term as an inf of more and more sensitive functions.
+We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, by considering the relational semantics weighted on the tropical semiring.
+The aim is to provide a framework where both program metrics, based on the analysis of program sensitivity via Lipschitz conditions, and resource analysis, based on higher-order program differentiation, coexist.
+In order to do that, we consider a graded-types lambda-calculus and the differential lambda-calculus, both languages ensuring bounded duplications.
+A typical result of such line of research is that, thanks to the tropical operations, a simply typed term is interpreted as a locally Lipschitz map, which is in turn decomposed as an inf of more and more sensitive Lipschitz ones, corresponding to the interpretations of the approximants given by the lambda-calculus Taylor expansion.
+We finally describe some promising lines of research, showing in particular how such tropical model expresses natural optimisation problems of probabilistic calculi.
 \end{abstract}
@@
 \input{section1}
-
-%\section{The Two Quantitative Approaches to the $\lambda$-calculus}\label{section2}
-%\input{section2}
 
@@
 \subsection{... As (weighted relational) model of several $\lambda$-calculi}
-
 The category $\LREL$ satisfies many nice properties which make it able to model all the variants of the $\lambda$-calculi to which we are interested in here.
-Although the majority of the properties mentioned in the present section are shared with the general class of models of $\lambda$-calculi called \emph{weighted relational models}, thoroughly studied in \cite{Manzo2013}, we mention them for our particular case in order to highlight how these look like in the tropical setting.
-Moreover, we end the section by stating some property of $\LREL$ not necessarily shared with all the models in the general class.
+Although the majority of such properties are shared with the general class of models of $\lambda$-calculi called \emph{weighted relational models}, thoroughly studied in \cite{Manzo2013}, we mention them for our particular case in order to highlight how these look like in the tropical setting, but also to make the non-expert in such topics reader able to follow.
 The real interest in our particular model will come in the next \autoref{sec:tls}.
@@
 
-\subsection{Quantitaive analysis%: probabilistic, differential privacy
-}\label{sec:app}
+\subsection{Quantitaive analysis}\label{sec:app}
 \input{section5}
@@
 \subsection{Generalized Metric Spaces and Modules over the Lawvere Quantale}\label{section6}
-%At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}.
+%%At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}.
 %Let us recall that a quantale is a complete lattice endowed with a continuous monoid action.
@@
 
-\section{Related Work}\label{section7}
-\input{section7}
+\section{Related Work}\label{section7} \input{section7}
 
-\section{Conclusion}\label{section8}
-\input{section8}
+\section{Conclusion}\label{section8} \input{section8}
 
@@
 
-\bibliography{tropical}
+%\bibliography{tropical}
 
@@
 
-\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}
-
-\section{Theorem-like environments}\label{sec:theorem-environments}
+\section{\autoref{section3}}\label{app-sec:section3}
 

### 2023-07-26 CSL24/section2.tex
Committino notturno
Commit notturno

@@
 The main feature of this language is that if $\vdash \lambda x.M:\,!_nA\multimap B$, then $x$ will be duplicated exactly $n$ times in the reduction to the normal form.
-%, i.e.\ $\vdash \lambda x.M:\,!_nA\multimap B$ is a program which performs $n$ duplications.
-E.g., $\vdash_{\BSTLC} \lambda {\color{red}z}.\left( \lambda x{\color{green}y}.
+%E.g., $\vdash_{\BSTLC} \lambda {\color{red}z}.\left( \lambda x{\color{green}y}.
 yxx \right)z : \, !_{\color{red}2} *\multimap !_{\color{green}1}(!_{\color{violet}1} * \multimap !_{\color{blue}1} * \multimap *) \multimap *$.
@@
 		\justifies
-		\Gamma, x:_{n+m}B\vdash M[x/y]:A
+		\Gamma, x:_{n+m}B\vdash M\{x/y\}:A
 		\endprooftree 
@@
 This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
-
 %A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models.
@@
 %In $\BSTLC$ the typing system handles duplications; in $\STDLC$ the syntax with its operational semantics (that we do not give) does it.
+\begin{comment}
 Writing $\Der^2[\_,(\_)^2]$ as a shortcut for $\Der[\Der[\_,\_],\_]$ and $\Der^1[\_,(\_)^1]$ for $\Diff{\_}{\_}$, the analogue of the previous $\BSTLC$-term is $\vdash_{\STDLC} \lambda {\color{red}z}.
@@
 : *\to (* \to * \to *) \to *$, then the term reduces to the empty sum $0$ (representing an \emph{error}).
-
+\end{comment}
 %Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
@@
 %, it is Cartesian closed%(with the closed structure compatible with the additive structure)
-and with a certain \emph{differential operator} $D$, generalising the usual notion of differentiation, see e.g.\ \cite{BluteEhrhTass10}.
+and with a certain \emph{differential operator} $D$, generalising the usual notion of differential, see e.g.\ \cite{BluteEhrhTass10}.
 %An example is the CC$\partial\lambda$C of convenient vector spaces with smooth maps, where $D$ is the ``real'' differential $Df:\mathbb{V}\times\mathbb{V}\rightarrow \mathbb{W}$, $Df(x,u):=\dfrac{d}{dt}{\!\Big|_{t=0}} f(x+tu)$, of smooth maps $f:\mathbb{V}\rightarrow \mathbb{W}$.

### 2023-07-26 CSL24/section3A.tex
Commit notturno

@@
-The linear simply typed $\lam$-calculus is a restriction of the ordinary $\lam$-calculus in which programs, following the linear logic mantra, can only use their arguments \emph{exactly once} (and indeed this calculus can be embedded in its \emph{intuitionistic multiplicative} fragment \emph{IMLL}).
+The linear simply typed $\lam$-calculus is a restriction of the ordinary $\lam$-calculus in which programs, following the linear logic mantra, can only use their arguments \emph{exactly once}.
+%(and indeed this calculus can be embedded in its \emph{intuitionistic multiplicative} fragment \emph{IMLL}).
 %More precisely, the linear $\lam$-calculus is obtained from the ordinary one by adding the constraint that each $\lam$-abstracted variable appears exactly once in the scope of the $\lam$-abstraction.
@@
 It can be interpreted in any Symmetric Monoidal Closed Category (SMCC).
-Now, \cite[Section III.A]{Manzo2013} immediately gives that $\LREL$ is a SMCC (actually, even a model of IMLL), where its monoidal product $\otimes$ acts on the objects as the Cartesian product of sets, %and as the \emph{Kronecker product} of matrices on morphisms.
+Now, $\LREL$ is a SMCC (\cite[Section III.A]{Manzo2013}), %(actually, even a model of IMLL), 
+whose monoidal product $\otimes$ acts on objects as the Cartesian product of sets, %and as the \emph{Kronecker product} of matrices on morphisms.
  %{(\color{red} Ref?)} 

### 2023-07-26 CSL24/section3B.tex
Commit notturno

@@
 
-Let us recall that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
+Remember that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
 %We will explicit this constructions in our tropical setting in the next lines.
 As well-known, if a SMCC is \emph{Lafont}, then one obtains a CCC from it by defining the exponential objects as $X\to Y:=!X \multimap Y$.
-With no surprise $\LREL$ is Lafont (\cite[Corollary III.6]{Manzo2013}) and specialising \cite[Theorem III.7]{Manzo2013} we have indeed that the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$, where as usual $!$ acts on objects by taking the set of finite multisets.
+With no surprise $\LREL$ is Lafont (~\cite[Corollary III.6]{Manzo2013}) %and specialising \cite[Theorem III.7]{Manzo2013} 
+w.r.t.\ the usual $!$ acting on objects by taking the set of finite multisets.
+So the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.
 %It is instructive what its CCC-structure looks like in our tropical world.
-The cartesian product of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_!
-t\in\Lawv^{!X\times Z}$ where $(s\circ_!
+The cartesian product $\&$ of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms $\RM{ev}$ are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_!
+t\in\Lawv^{!X\times Z}$, $(s\circ_!
 t)_{\mu,c}:=

### 2023-07-26 CSL24/section3C.tex
Commit giornaliero
Commit notturno

@@
 %Similarly to It is known {\color{red}(([reference??] e dire meglio)} 
-For $\BSTLC$, one can see that the exponential $!$ can be ``decomposed'' into a family of \emph{graded} exponentials functors $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined by taking multisets of cardinality \emph{at most} $n$.
+For $\BSTLC$, one can see that the exponential $!$ can be ``decomposed'' into a family of \emph{graded} exponentials functors $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined on objects by taking multisets of cardinality \emph{at most} $n$.
 %  lift to functors
@@
 i.e.\ $(\LREL,(!_n)_{n\in\N})$ is a model of $\BSTLC$. 
-In particular, bounded arrow types are interpreted via $\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$ and so, 
-if e.g.\ $\model o=\{\star\}$, then $\model{A}$ is a finite set for any type $A$ of $\BSTLC$.
+In particular, bounded arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$ and so, 
+if $\model *$ is finite, then $\model{A}$ is a finite set for any type $A$ of $\BSTLC$.
 %\end{remark}
@@
 
-For what concerns $\STDLC$, by patiently checking all the needed axioms {\color{red}or by using  LEMAY??}, one sees that
+For what concerns $\STDLC$, applying \cite[Theorem 6.1]{lemay2020} one can check that
 %\begin{proposition}[{\color{red}LEMAY??}]\label{thm:LREL!CCDC}
- $\LREL_!$ becomes a CC$\partial\lambda$C, i.e.\ a model of the $\STDLC$, when equipped with the \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
+ $\LREL_!$ becomes a CC$\partial\lambda$C, i.e.\ a model of the $\STDLC$, when equipped with the \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
 %\end{proposition}
@@
 %Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
-Finally, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] shows:
+Finally, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of \cite[Theorem 4.23]{Manzo2012} shows:
 
@@
 As we will discuss in \autoref{sec:app}, non-trivial coefficients appear when interpreting more interesting languages, like probabilistic ones.
-However, in this paper we shall still focus on $\STLC$, in order to set the basis of the further studies and in order to see some already interesting properties of the interpretation.
+However, in this paper we shall still focus on $\STLC$, in order to set the basis of the further studies and in order to see some interesting properties already of its interpretation.

### 2023-07-26 CSL24/section4.tex
Commit notturno

@@
 Concretely, we have $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
+
 In this section we establish some topological and metric properties of such functions, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
@@
 Non riesco a capire se anche a noi lo \`e, ma non vedo perch\'e non dovrebbe.}
+
+\begin{remark}
+ Since $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}\in\HOM{\LREL_!}{\mathlarger{\&}_{i=1}^n \model{A_i}}{\model B}\simeq\Lawv^{\prod_{i=1}^n !
+\model {A_i}\times \model B}$, the tLs $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ is defined by $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !
+\model{A_i}}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
+Same for $\STDLC$.
+Finally, the interpretations of $\BSTLC$-terms are matrices $t\in\HOM{\LREL}{\bigotimes_{i=1}^n !_{n_i}X_i}{Y}=\Lawv^{\prod_{i=1}^n !_{n_i}X_i\times Y}$.
+In such situation, we define $t^!:\Lawv^{\mathlarger{+}_{i=1}^n X_{i}}\to\Lawv^Y \simeq \prod_{i=1}^n\Lawv^{X_{i}}\to\Lawv^Y$ as $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !_{n_i} X_i}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
+Clearly $t^!={\widetilde t}^!$, where $\widetilde t\in\Lawv^{\prod_{i=1}^n !
+{X_i}\times Y}$ is the matrix $\widetilde t_{(\mu_1,\dots,\mu_n),b}:=t_{(\mu_1,\dots,\mu_n),b}$ if $\mu_i\in !_{n_i} X_i$ for all $i$, and $:=\infty$ otherwise, which has the same support as $t$.
+So, such $t^!$'s are special cases of tLs and thus, \emph{Wlog}, in the following we only consider tLs.
+\end{remark}
 
@@
  with \emph{finite} supports $\C F_b=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$, and which have thus shape $f(x)_b=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, are generalisation of usual tropical polynomials to the case of possibly infinitely many variables.
-For instance, for a $t\in\HOM{\LREL}{!_n X}{Y}$ (this is in particular the case for the interpretations of $\BSTLC$-terms), we will consider the (generalised) tropical polynomial $t^!:\Lawv^X\to\Lawv^Y$ defined by $t^!(x)_b:=\inf_{\mu\in !_nX}\{\mu x+t_{\mu,b}\}$.
+If $\model *$ is finite, this is in particular the case for the interpretations of $\BSTLC$-terms, since then the interpretation of any $\BSTLC$-type $B$ is a finite set, and thus also $!_n \model B$ is.
 % 
@@
 Also, since $f\geq 0$ on all $\Lawv^X$, we have e.g.\ $-f\leq 1$ on $\Lawv^X$.
-Remark that we cannot immediately apply \cite[Proposition 4.4.(3)]{Cobzas2017}, since $\Lawv^X$ is not even a $\BB R$-vector space.
-But $(0,\infty)^X\subseteq \BB R^X$ is open and convex, and $\BB R^X$ is a locally convex topological $\BB R$-vector space.
-Thus \cite[Proposition 4.4.(3)]{Cobzas2017} entails the continuity of $-f\big|_{(0,\infty)^X}$, hence that of $f\big|_{(0,\infty)^X}$.
+%Remark that we cannot immediately apply \cite[Proposition 4.4.(3)]{Cobzas2017}, since $\Lawv^X$ is not even a $\BB R$-vector space.
+Now $(0,\infty)^X\subseteq \BB R^X$ is open and convex, and {\color{red}$\BB R^X$ is a locally convex topological $\BB R$-vector space}, so \cite[Proposition 4.4.(3)]{Cobzas2017} entails the continuity of $-f\big|_{(0,\infty)^X}$, hence that of $f\big|_{(0,\infty)^X}$.
 \end{proof}
@@
 \begin{proof}
-1). We already observed that the interpretation of a $\BSTLC$-type $B$ is alway a finite set, so thus is $!_n \model B$.
-So the $\inf$ defining $(\_)^!$ is actually a $\min$.
-So the interpretation of a bounded term is a tropical polynomial, so we apply Corollary~\ref{prop:polylip} to each coordinate of the image, and since $\model A$ is finite (by taking the maximum Lipschitz constant among the finite number of them), we obtain the thesis.
+1). Since $\model A$ is finite, also $\model *$ is.
+Thus, as we already observed, the interpretation of a bounded term is a tropical polynomial.
+Now we apply Corollary~\ref{prop:polylip} to each coordinate of the image, and by taking the maximum Lipschitz constant among the finite number $\mathrm{Card}(\model A)$ of them, we obtain the thesis.
 2). It follows immediately from Theorem~\ref{thmTLSlocLip} and the fact that $\model A$ is finite.
@@
 
-A consequence of (3) is that the distance between two programs can always be approximated via Lipschitz tropical polynomial approximants of the initial two programs.
+A consequence of (3) is that the pointwise distance between two interpretations of programs can always be bounded via Lipschitz tropical polynomial approximants of the initial two programs.
 \begin{corollary}

### 2023-07-26 CSL24/section6.tex
Commit notturno

--- 2023-01-22 section6.tex
@@
+\begin{remark}
+ $\QREL$ is (equivalent to) a subcategory of the category $Q\SF{Mod}$ of \emph{complete} $Q$-semimodules.
+ If $\QREL$ corresponds to considering semimodules (the $Q^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $Q\SF{Mod}$ corresponds to considering semimodules in abstract, without fixing a base.
+\end{remark}
+
 As we have seen, the morphisms of $\LREL$ can be seen as continuous functions between the $\Lawv$-modules $\Lawv^{X}$, when the latter are taken with the metric induced by the $\infty$-norm.

### 2023-07-26 CSL24/section1.tex
Commit
Intanto committo, ovviamente è ancora tutto da aggiustare, ma almeno se vuoi vedere la nuova organizzazione ce l'hai disponibile.

@@
 Among such quantitative approaches, two different directions have received considerable attention, which we recall below.
+In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
 
-%\subsection{Two kinds of quantitative approaches}\label{section2}
+\paragraph*{Two kind of quantitative approaches}
 
-\input{Section2}
+On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask whether they compute functions which do not differ \emph{too much} by controlling the propagation of an error during the computation.
+This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
+%As already mentioned, in many situations (e.g.~when dealing with computationally difficult problems, or in differential privacy \cite{Alvim2011, Reed2010}), one %does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound.
+In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, a small error in the input will produce a comparably small error in the output.
+%In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior.
+A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
+$d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
+This is due to the fact that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the %\emph{linear} simply typed $\lambda$-calculus, being a \emph{symmetric monoidal closed} category, and in fact it also models
+\emph{affine} $\lambda$-calculus (it is symmetric monoidal closed, cartesian and monoidal units coincide). 
+The metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications \cite{Reed2010, Gaboardi2017}: the idea is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz} map.
+For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the function input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself.
+%: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.
+By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $rX:=(X,%d_{r}(x,y)=
+rd(x,y))$ and $Y$, the program $M$ above 
+can thus be interpreted as a non-expansive map from $2(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.
+
+On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
+This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
+%A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models.
+The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus.
+The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Diff{M}{N}$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Diff{M}{N}$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).
 
@@
 
-\subsection{Tropical mathematics and program semantics }
-
-In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
-In metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
-By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
-
-A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled.
-At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
-in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz.
-The fundamental motivation of this work is then the observation that
-this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
+\paragraph*{Tropical mathematics and program semantics }
 
@@
 
-\subsection{This paper and our contributions}
+Coming back to our discussion on program linearity, in metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
+By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
 
-Coming back to our discussion on program semantics %tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
-the metric approach usually yields Lipschitz conditions for affine programs, but what about ``unbounded'', simply typed $\lambda$-calculus (the ordinary one)?
+A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled.
+At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
+in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz.
+The fundamental motivation of this work is then the observation that
+this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
+
+Now, the metric approach usually yields Lipschitz conditions for affine programs, but what about ``unbounded'', simply typed $\lambda$-calculus (the ordinary one)?
 In such case, one might lose the Lipschitz property, so the programs might be \emph{too} sensitive to errors. 
@@
 
+{\color{red}Dire proba! Non solo metric vs differential}
+
+\paragraph*{This paper and our contributions}
+
+{\color{red}Dire proba!}
 %\subsection{Contributions}

### 2023-07-26 CSL24/section2.tex
Commit
Intanto committo, ovviamente è ancora tutto da aggiustare, ma almeno se vuoi vedere la nuova organizzazione ce l'hai disponibile.

@@
 
-\paragraph*{Metric approach: bounded $\lambda$-calculus in a nutshell}
-
-On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask whether they compute functions which do not differ \emph{too much} by controlling the propagation of an error during the computation.
-This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
-%As already mentioned, in many situations (e.g.~when dealing with computationally difficult problems, or in differential privacy \cite{Alvim2011, Reed2010}), one %does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound.
-In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, a small error in the input will produce a comparably small error in the output.
-%In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior.
-A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
-$d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
-This is due to the fact that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the %\emph{linear} simply typed $\lambda$-calculus, being a \emph{symmetric monoidal closed} category, and in fact it also models
-\emph{affine} $\lambda$-calculus (it is symmetric monoidal closed, cartesian and monoidal units coincide). 
-The metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications \cite{Reed2010, Gaboardi2017}: the idea is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz} map.
-For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the function input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself.
-%: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.
-By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $rX:=(X,%d_{r}(x,y)=
-rd(x,y))$ and $Y$, the program $M$ above 
-can thus be interpreted as a non-expansive map from $2(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.
+\paragraph*{Controlled duplication/erasure via graded types: bounded $\lambda$-calculus $\BSTLC$}\label{sec:BSTLC}
 
@@
 The terms are as for the $\STLC$, the types are $A::= * \ \mid  \ !_{n}A \multimap A$ and type judgements are of the form $\Gamma \vdash M:A$, where a context $\Gamma$ is a set of declarations of the form $x :_{n}A$, with $n\in \mathbb N$.
-The typing rules are given in Fig.~\ref{fig:rules}, where $\Gamma+\Delta$ is defined letting $(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) =  (\Gamma+\Delta), x:_{m+n}A$, and $m\Gamma$ is made all $x:_{mn}A$ for $(x:_{n}A) \in \Gamma$.
+The typing rules are: %given in Fig.~\ref{fig:rules}
+	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
+	\begin{array}{cccc}
+		\prooftree
+		\Gamma \vdash M:A
+		\justifies
+		\Gamma, x:_{0}B \vdash M:A
+		\endprooftree 
+		&
+		\prooftree
+		\Gamma, x:_{n}B, y:_{m} B\vdash M:A
+		\justifies
+		\Gamma, x:_{n+m}B\vdash M\{x/y\}:A
+		\endprooftree 
+		&
+		\prooftree
+		\Gamma, x:_{n} A\vdash M: B
+		\justifies
+		\Gamma\vdash \lambda x.M: !_{n}A\multimap B
+		\endprooftree
+		&
+		\prooftree
+		\Gamma \vdash M: !_nA\multimap B
+		\quad
+		\Delta\vdash N: A
+		\justifies
+		\Gamma +n\Delta\vdash MN: B
+		\endprooftree
+	\end{array}
+	\]
+where $\Gamma+\Delta$ is defined letting $(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) =  (\Gamma+\Delta), x:_{m+n}A$, and $m\Gamma$ is made all $x:_{mn}A$ for $(x:_{n}A) \in \Gamma$.  
 The axiom is $x:_{1}A\vdash x: A$.
@@
 yxx \right)z : \, !_{\color{red}2} *\multimap !_{\color{green}1}(!_{\color{violet}1} * \multimap !_{\color{blue}1} * \multimap *) \multimap *$.
-The $\BSTLC$ can be interpreted in a symmetric monoidal closed category (SMCC) equipped with a \emph{$\mathbb N$-graded linear exponential comonad} \cite{Katsumata2018}.
 %Observe that one can always type an affine term like e.g.~$\lambda xy.x$ with a linear type $A\multimap B\multimap A$.
@@
 %$!_{2}(A\multimap A)\multimap (A\multimap A)$.
+
+%\begin{remark}\label{rmk:ModelsOfBSTLC}
+%Similarly to It is known {\color{red}(([reference??] e dire meglio)} 
+The $\BSTLC$ can be interpreted in a symmetric monoidal closed category (SMCC) equipped with a \emph{$\mathbb N$-graded linear exponential comonad} \cite{Katsumata2018}.
+Now, one can see that the exponential $!$ can be ``decomposed'' into a family of \emph{graded} exponentials functors $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined on objects by taking multisets of cardinality \emph{at most} $n$.
+%  lift to functors
+The sequence $(!_n)_{n\in\N}$ gives rise to a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, %satisfying the adjunction: $\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
+i.e.\ $(\LREL,(!_n)_{n\in\N})$ is a model of $\BSTLC$. 
+In particular, bounded arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$ and so, 
+if $\model *$ is finite, then $\model{A}$ is a finite set for any type $A$ of $\BSTLC$.
+%\end{remark}
+%Bounded types are interesting because of the following proposition:
+%\begin{proposition}
+%For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) correspond to polynomials.
+%\end{proposition}
+%\begin{proof}
+%It suffices to check that $\model A$ is finite for all bounded types $A$.
+Indeed this implies that a morphism $t:\model A\to \model B$ is a finite matrix $t: \model A \times \model B \to \Lawv$.Hence, its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is a polynomial.
+%\end{proof}
+
+%For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
+%\begin{itemize}
+%\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
+%\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
+%\end{itemize}
+
+\begin{comment}
 
@@
 
-\paragraph*{Resource approach: differential $\lambda$-calculus in a nutshell}
+\end{comment}
 
-On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
-This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
-%A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models.
-The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus.
-The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Diff{M}{N}$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Diff{M}{N}$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).
+\paragraph*{Controlled duplication/erasure via resources: the differential $\lambda$-calculus $\STDLC$}\label{sec:STDLC}
 
@@
 Simple types are $A::= *\mid A\to A$.
-The typing system is in Figure~\ref{fig:rules}, where a context $\Gamma$ is a list of typed variable declarations.
+The typing rules are: %in Figure~\ref{fig:rules}
+	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
+	\begin{array}{cccc}
+		\prooftree
+		\Gamma, x: A\vdash M: B
+		\justifies
+		\Gamma\vdash \lambda x.M: A\to B
+		\endprooftree 
+		&
+		\prooftree
+		\Gamma \vdash M: A\to B
+		\quad
+		\Gamma\vdash \mathbb T: A
+		\justifies
+		\Gamma \vdash M\mathbb T: B
+		\endprooftree 
+		&
+		\prooftree
+		\Gamma \vdash M: A\to B
+		\quad
+		\Gamma \vdash N: A
+		\justifies
+		\Gamma \vdash \Diff{M}{N}: A\to B
+		\endprooftree
+		&
+		\prooftree
+		\Gamma\vdash M_1: A
+		\,\cdots\,
+		\Gamma \vdash M_n:A
+		\justifies
+		\Gamma \vdash M_1+\cdots +M_n : A
+		\using (n\geq 2)
+		\endprooftree
+	\end{array}
+	\]
+where a context $\Gamma$ is a list of typed variable declarations.
 The axioms are $\Gamma, x:A \vdash x: A$ and $\Gamma\vdash 0:A$.
@@
 %Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
-%The models of $\STDLC$ are the cartesian \emph{closed} differential categories ($CC\partial C$), which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.
-
-The $\STDLC$ can be interpreted in Cartesian closed differential $\lambda$-categories (CC$\partial\lambda$C)\cite{Manzo2010,Blute2009, Blute2019}.
-In them, homsets are equipped with a structure of commutative monoid %(i.e.\ we can add morphisms and there is a $0$ morphism)
-%, it is Cartesian closed%(with the closed structure compatible with the additive structure)
-and with a certain \emph{differential operator} $D$, generalising the usual notion of differential, see e.g.\ \cite{BluteEhrhTass10}.
-%An example is the CC$\partial\lambda$C of convenient vector spaces with smooth maps, where $D$ is the ``real'' differential $Df:\mathbb{V}\times\mathbb{V}\rightarrow \mathbb{W}$, $Df(x,u):=\dfrac{d}{dt}{\!\Big|_{t=0}} f(x+tu)$, of smooth maps $f:\mathbb{V}\rightarrow \mathbb{W}$.
-%More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
-%\begin{itemize}
-%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
-%\item $\C C$ is equipped with a differential operator $D:
-%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
-%\end{itemize}
 
@@
 % (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
+
+%Until now we simply specialised well-known results in our tropical case, with the intent of showing how things read in this particular case.
+%Now we go further, by showing that $\LREL_!$ actually admits a \emph{differential structure}, turning it into a model of the $\STDLC$, i.e.~a $CC\partial C$.
+%This viewpoint
+%, is where the \emph{metric} and the \emph{differential} viewpoints converge, as explained in the Introduction and Section II, and it % will be further generalised in Section \ref{section6}.
+
+%A model of the $\STDLC$ is usually understood as so-called \emph{Cartesian closed differential categories} (CC$\partial$C), see \cite{Manzo2012} for details.
+%In order to treat the $+$ and the constructor $D[\_]\cdot (\_)$ of $\STDLC$, the main features of a CC$\partial$C $\C C$ are that:
+%
+%1) $\C C$ is a left-additive-CCC, i.e.\ its Homsets are commutative monoids and its Cartesian closed structure is well behaved w.r.t.\ this monoid structure;
+%
+%2) $\C C$ is equipped with a differential operator map $D:\HOM{\C C}{X}{Y}\to \HOM{\C C}{X\times X}{Y}$ (here $\times$ is the Cartesian product of $\C C$) satisfying $8$ axioms, called D1, ..., D7, D-curry.
+
+%Let us show the differential structure of $\LREL_!$ (remember that the Cartesian product of $\LREL_!$ is the disjoint union $+$).
+
+%\begin{definition} The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).\end{definition}
+
+%The models of $\STDLC$ are the cartesian \emph{closed} differential categories ($CC\partial C$), which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.
+The $\STDLC$ can be interpreted in Cartesian closed differential $\lambda$-categories (CC$\partial\lambda$C)\cite{Manzo2010,Blute2009, Blute2019}.
+In them, homsets are equipped with a structure of commutative monoid %(i.e.\ we can add morphisms and there is a $0$ morphism)
+%, it is Cartesian closed%(with the closed structure compatible with the additive structure)
+and with a certain \emph{differential operator} $D$, generalising the usual notion of differential, see e.g.\ \cite{BluteEhrhTass10}.
+%An example is the CC$\partial\lambda$C of convenient vector spaces with smooth maps, where $D$ is the ``real'' differential $Df:\mathbb{V}\times\mathbb{V}\rightarrow \mathbb{W}$, $Df(x,u):=\dfrac{d}{dt}{\!\Big|_{t=0}} f(x+tu)$, of smooth maps $f:\mathbb{V}\rightarrow \mathbb{W}$.
+%More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
+%\begin{itemize}
+%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
+%\item $\C C$ is equipped with a differential operator $D:
+%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
+%\end{itemize}
+Now, applying \cite[Theorem 6.1]{lemay2020} one can check that
+%\begin{proposition}[{\color{red}LEMAY??}]\label{thm:LREL!CCDC}
+ $\LREL_!$ becomes a CC$\partial\lambda$C, i.e.\ a model of the $\STDLC$, when equipped with the \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
+%\end{proposition}
+%\begin{remark}For $t\in\HOM{\LREL}{!X}{Y}$, we have: $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu+\nu'+\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho+\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.\end{remark}
+%This ensures that one can define a sound interpretation of $\STDLC$-terms in the standard way (see [Section 4.3, \cite{Manzo2010}]).
+This model is also well-behaved w.r.t.\ the Taylor expansion, as expressed by the following two results.
+As it can be patientely checked, in $(\LREL_!,D)$ all morphisms can be Taylor expanded  (see \cite[Definition 4.22]{Manzo2012}):
+
+\begin{theorem}\label{thm:modelsTaylor}
+ For all $t\in\HOM{\LREL_!}{Z}{X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$ we have:%, the evaluation of $t$ over $s$ yields 
+ \begin{align}\label{eq:taylorcat}
+  \RM{ev}\circ_!\langle t,s\rangle =
+  \inf\limits_{n\in\N}
+  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
+ \end{align} 
+\end{theorem}
+%It is worth discussing the formula above a bit more. 
+Here,
+%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
+%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
+$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
+Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$,  i.e.~it coincides with the tropical %interpretation of the
+version of the usual Taylor expansion.
+%Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
+Finally, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of \cite[Theorem 4.23]{Manzo2012} shows:
+
+\begin{corollary}\label{cor:T(M)=M}
+ If $\Gamma\vdash_{\STLC} M:A$, then $\model{\Gamma\vdash_{\STDLC} \Te M:A}:=\inf_{t\in\Te{M}} \model{\Gamma\vdash_{\STDLC} t:A}$ converges to $\model{\Gamma\vdash_{\STLC} M:A}$.
+%the interpretation of the Taylor expansion of a $\STLC$-term $M$, given in \eqref{eq:taylor}, converges to the one of $M$.
+\end{corollary}
+
+As already mentioned, by straightforward induction on $M$ (using that composition and $D$ easily preserve ``booleaness'' and projections and evaluation of  $\LREL$ are boolean) one can prove that:
+
+\begin{proposition}\label{prop:descrete}
+ For all the previous four languages, except $\STLC_\oplus$, we have that $\model{\Gamma\vdash M:A}\in\Lawv^{!\model{\Gamma}\times\model{A}}$ is a \emph{boolean} matrix, i.e.\ actually $\model{\Gamma\vdash M:A}\in\{0,\infty\}^{!\model{\Gamma}\times\model{A}}$.
+\end{proposition}
+
+Even if the interpretation of those languages is trivia, we shall still focus on them in order to set the basis of the further studies and in order to see some already interesting properties of their interpretation.{\color{red}SPIEGARE MEGLIO}
 

### 2023-07-26 CSL24/section2bis.tex
Commit
Intanto committo, ovviamente è ancora tutto da aggiustare, ma almeno se vuoi vedere la nuova organizzazione ce l'hai disponibile.

--- 2023-07-25 CSL24/section4.tex
@@
-For starting, remark that although a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, by exploiting the coKleisli structure we can also ``express it in the base $X$'', i.e.\ see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_!
-x$.
-%(we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
-Concretely, we have $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
-In this section we establish some topological and metric properties of such functions, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
-As we explain in a moment, those are generalised \emph{tropical Laurent series} (\emph{tLs} for short), and as the literature in tropical mathematics is often recent (e.g.~\cite{Porzio2021}), several results we prove in this section are, to our knowledge, new.
+\subsection{Tropical linear algebra}
+\input{section3new}
 
-%
-%In this section we study the tropical Laurent series $\Lawv^X\to \Lawv ^Y$ from the viewpoint of analysis.
-
-\subsection{From tropical polynomials to tLs}\label{sec:4A}
+\subsection{Tropical non-linear algebra}\label{sec:4A}
 
@@
 
-%\begin{remark}%[Tropical Laurent series]
- Remark that the map $t^!:\Lawv^X\to\Lawv^Y$ introduced above from a matrix $t\in\HOM{\LREL_!}{X}{Y}$, corresponds to a tLs with \emph{possibly infinitely} many variables (in fact, as many as the elements of $X$).
-%In the following we will also refer to them as tLs.
-% We will call them simply \emph{tropical Laurent series (tLs)}.
-% %Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
-%\end{remark}
-%
-%We find the usual notion of tLs of one variable as follows:
-%\begin{remark}
-By identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the %functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ we recover the
-usual tLs's of one variable.
-% \end{remark}
- \begin{remark}
- Our running example is indeed of shape $\varphi=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
- However, it is not the interpretation of a $\lam$-term, because its matrix $t$ is not discrete.
-% Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
-\end{remark}
-
  In a similar way, tLs
@@
  with \emph{finite} supports $\C F_b=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$, and which have thus shape $f(x)_b=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, are generalisation of usual tropical polynomials to the case of possibly infinitely many variables.
-For instance, for a $t\in\HOM{\LREL}{!_n X}{Y}$ (this is in particular the case for the interpretations of $\BSTLC$-terms), we will consider the (generalised) tropical polynomial $t^!:\Lawv^X\to\Lawv^Y$ defined by $t^!(x)_b:=\inf_{\mu\in !_nX}\{\mu x+t_{\mu,b}\}$.
-% 
-% , i.e.~for a \emph{finite} set $\C F\subseteq \, !X$.
-%%Remark that we also find usual \emph{tropical polynomials} of tropical geometry as a particular case: they 
-%correspond to the tLs for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}. 
-%Actually, for us \emph{tropical polynomial} will mean a function 
-%\end{remark}
 
-Looking at Fig~\ref{fig:plot1}, we see that $\varphi$, just like the polynomials $\varphi_{n}$, is non-decreasing and concave.
-This is indeed always the case:
+Remember that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
+%We will explicit this constructions in our tropical setting in the next lines.
 
-\begin{proposition}\label{prop:nondecr+conc}
- Any tLs $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
-\end{proposition}
+The cartesian product $\&$ of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms $\RM{ev}$ are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_!
+t\in\Lawv^{!X\times Z}$, $(s\circ_!
+t)_{\mu,c}:=
+%\begin{align}
+\inf_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
+ \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
+%\end{align}
+%where $+$ is the multiset union.
 
-%In Example~\ref{ex:famous_ex}, 
-Again by looking at Fig~\ref{fig:plot1} it appears that, \emph{far from $0$}, $\varphi$ behaves like some polynomials $\varphi_{n}$.
-In particular, %for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$. More, precisely, 
-$\varphi$ coincides on $[\epsilon,\infty]$ with $\varphi_{n}$,
-for $\epsilon \geq 2^{-(n+1)}$ (the smallest tropical root of $\varphi_{n}$).
-However, at
-%
-%$x\in [\epsilon,\infty]$, with $\epsilon>0$
-%It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
- $x=0$ we have that $\varphi(x=0)=\inf_{n\in\N} 2^{-n}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
-Also, while the derivative of $f$ is bounded on all $(0,\infty)$, for $x\to 0^+$ it tends to $\infty$.
-This phenomenon is reminiscent of \cite[Example 7]{Ehrhard2005},
-%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
-which actually motivated our first investigations.
-In fact, this behaviour is shared by all tLs with \emph{finitely} many variables, as shown by the following result (we identify $!\set{1,\dots,k}\simeq \N^k$, so the matrix of a tLs $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one output variable) is given as a $\matr f:\N^k\to\Lawv$, $f$ having shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, with $nx$ the scalar product).
+%Remember that in $\Lawv$ the neutral element for addition is $\infty$ and the neutral for multiplication is $0$, so for instance the evaluation map is the matrix $\RM{eval}^{X,Y}\in\Lawv^{!((X\multimap Y) \times X)\times Y}\simeq\Lawv^{(!!X\times !Y\times !X)\times Y}$ given by $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=0$ if $\rho_1=[\mu]$ and $\rho_2=[b]$, and $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=\infty$ otherwise.
 
-\begin{theorem}\label{theorem:fepsilon}
- Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\matr f:\N^k\to\Lawv$.
- For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
-% 
-% \begin{enumerate}
-%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
-%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
-  %\item 
-$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical \emph{polynomial} $P_\epsilon(x):=\min_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
-% \end{enumerate}
-\end{theorem}
-\begin{proof}[Proof sketch]
-Let $\C F_\epsilon$ be the set of $n\in\N^{k}$ such that 
-$\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
-The core of the proof is showing that this set is indeed finite and enough for computing $f$.
-\end{proof}
-
-\subsection{Continuity of tLs}\label{subsec:cont}%$\Lawv^{X}$ as a normed cone}
-
-The tLs $\varphi$ is continuous on $\BB R_{\geq0}$ (w.r.t.\ the usual norm of real numbers).
-By considering the usual norm $\norm{x}_\infty:=\sup_{a\in X} \absv{x_a}$ on $\Lawv^X$, we can generalise this property by dropping the case of $x$ having some $0$ coordinate:
-
-\begin{theorem}\label{thm:cont}
- All tLs $f:\Lawv^X\to\Lawv$ are continuous on $\BB R_{>0}^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
-\end{theorem}
-\begin{proof}
- It follows after adapting [Proposition 4.4, \cite{Cobzas2017}] in order to prove that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
-\end{proof}
-
-We conclude this subsection by noticing that $\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
-Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete \emph{normed cone} (see~\cite{Selinger2004}, or the appendix, for such notions).
-Its cone structure induces an order on it, called its \emph{cone order}:
-$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
-Such order actually coincides with the pointwise order on $\Lawv^X$.
-This makes it a Scott-continuous dcpo.
-Suitable categories of cones have been recently investigated as models of probabilistic computation (\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
-Here we will just mention that:
-
-\begin{theorem}\label{thm:ScottCont}
- All monotone (w.r.t.\ pointwise order) and $\norm{\cdot}_{\infty}$-continuous functions $f:(0,\infty)^X\to (0,\infty)$ are Scott-continuous.
- In particular, all tLs $f:\Lawv^X\to\Lawv$ are Scott-continuous on $(0,\infty)^X$ w.r.t.\ the pointwise orders.
-\end{theorem}
-\begin{proof}
- Use the fact, taken from \cite{Selinger2004}, that in a normed $\R_{\geq 0}$-cone $P$, considered with its cone-order, if every bounded directed net in $P$ admits a sup, and if $(v_i)_{i\in I}$ is a directed net in $P$ with an upper bound $v\in P$, then $\exists\bigvee_{i\in I} v_i \in P$ and, if $\inf_{i\in I} \norm{v-v_i} =0$, one has $\bigvee_{i\in I} v_i = v$.
-\end{proof}
-
-\subsection{Lipschitz-continuity of tLs}\label{sec:4C}%$\Lawv^{X}$ as a metric space.}
-
-%The norm $\norm{\cdot}_\infty$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$.
-We will show that tLs satisfy suitable Lipschitz properties w.r.t.\ $\norm{\cdot}_\infty$.
-
-Let us first look at tropical linear functions:
-
-\begin{proposition}\label{prop:troplinear}
-All tropical \emph{linear} functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive.  
-\end{proposition}
-%\begin{proof}[Proof sketch]
-%Using the fact that $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
-%the problem reduces to checking that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
-This result shows that, in analogy with that happens in usual metric semantics, linear programs are interpreted by non-expansive functions. 
-%\begin{proof}
-%Using $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
-%first observe that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$; we now have
-%$|f(\B x)_{b}-f(\B y)_{b}| \leq |(\inf_{a\in X}\matr f_{a,b}-\B x_{a})-(\inf_{a\in X}\matr f_{a,b}-\B y_{a})| \leq
-%\sup_{a\in X}|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})|\leq 
-% \| \B x-\B y\|_{\infty}$.
-%\end{proof}
-
-%Before looking at what happens in the case of non-linear programs, let us make the metric structure of $\LREL$ explicit. 
-The following proposition provides a useful characterization of the functional metrics in $\LREL$, relying on 
-the bijection between $\HOM{\LREL}{X}{Y}$ and set of tropical linear functions from $\Lawv^{X}$ to $\Lawv^{Y}$.
-
-\begin{proposition}
-For all tropical linear functions $f,g:\Lawv^{X}\to \Lawv^{Y}$, $d_{\infty}(\matr f,\matr g)=d_\infty(f,g)$.% $\norm{ \matr f-\matr g}_{\infty} =  \sup_{x\in \Lawv^{X}} \norm{ f( x)-g(x)}_{\infty}$.
-\end{proposition}
-
-Let us now consider the case of bounded exponentials:
-\begin{proposition}\label{prop:boundedlip}
-If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
-\end{proposition}
-\begin{proof}[Proof sketch]
-This follows from Proposition \ref{prop:troplinear} and the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $!
-x$ to $\C M_{\leq n}(X)$.%
-%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
-%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
-%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
-\end{proof}
-This result is perfectly analogous to what happens in the metric models discussed in Section \ref{section2}, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
-
-Observe that, for any tropical polynomial $\varphi:\Lawv^{X}\to \Lawv^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \Lawv$ (as a monomial $\mu_ix+c_{i}$ yields a matrix entry on $!_{\#\mu_i}X\times Y$).
-Hence, using Proposition \ref{prop:boundedlip}, we have:
-\begin{corollary}\label{prop:polylip}
-For any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$, $\varphi$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
-\end{corollary}
-
-Let us now look at what happens with tLs, i.e.~when considering the full exponential $!$.
-As consequence of Theorem~\ref{theorem:fepsilon}, the tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $\BB R_{>0}$.
-Actually, we can prove a more general statement, also covering the infinitary case.
-
-\begin{theorem}\label{thmTLSlocLip}
- All tLs $\Lawv^X\to\Lawv$ are locally Lipschitz on $\BB R_{>0}^X$.
-\end{theorem}
-\begin{proof}[Proof sketch]
-The core of the proof is a convex analysis argument (see the Appendix) showing that an arbitrary function $f:\Lawv^X\to\Lawv$ which is non-decreasing, concave and continuous, must be locally Lipschitz.
-\end{proof}
-
-Finally, let us discuss the differential structure.
-The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tLs $f:\Lawv^{X}\to \Lawv^{Y}$ into a tLs $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by
-\begin{equation}
-D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a}+x_{a}+\mu y\right\}
-\end{equation}
-One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
-Moreover, the Taylor formula \eqref{eq:taylorcat} yields a ``tropical'' Taylor formula for tLs of the form 
-\begin{equation}
-f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}
-\end{equation}
-The following result shows that the distance between two tropical maps can be approximated using the terms appearing in their Taylor expansions:
-\begin{proposition}
-For all tLs $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
-the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz. Moreover 
-$\norm{ \matr f-\matr g}_{\infty}= \sup_{n} \norm{ {\delta^{(n)}f}- {\delta^{(n)}g}}_{\infty}$, 
-where $\delta^{(n)}h$ indicates the matrix of $D_{!}^{(n)}h$.
-%where $\delta^{(n)}f:( \Lawv^{X})^{n}\to \Lawv^{X}$ is the tropical linear function $\delta^{(n)}f(\B x_{1},\dots, \B x_{n})=
-%(\Der^{(n)}f)(\B x_{1},\dots, \B x_{n}, \infty)$. 
-\end{proposition}
-
-The results just presented translate into the following facts about the interpretation of higher-order programs:
-
-\begin{corollary}
-Let $\model A$ be a finite set.
-\begin{enumerate}
-\item $\model{\Gamma \vdash_{\BSTLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{tropical polynomial}, thus (as $\model A$ is finite), a \emph{Lipschitz} function.
-\item $\model{\Gamma \vdash_{\STLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{locally} Lipschitz map.
-\item $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!$ of \emph{tropical polynomials}, thus (as $\model A$ is finite), \emph{Lipschitz} functions.
-\end{enumerate}
-\end{corollary} 
-\begin{proof}
-1). We already observed that the interpretation of a $\BSTLC$-type $B$ is alway a finite set, so thus is $!_n \model B$.
-So the $\inf$ defining $(\_)^!$ is actually a $\min$.
-So the interpretation of a bounded term is a tropical polynomial, so we apply Corollary~\ref{prop:polylip} to each coordinate of the image, and since $\model A$ is finite (by taking the maximum Lipschitz constant among the finite number of them), we obtain the thesis.
-2). It follows immediately from Theorem~\ref{thmTLSlocLip} and the fact that $\model A$ is finite.
-3).
-It follows from \autoref{cor:T(M)=M} plus the easily checked fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have $\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with $\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
-\end{proof}
-
-Remark that the restriction $\model A$ finite is without loss of generality, since by Currying all programs can be seen having type $*$, which is natural to interpret as a singleton.
-
-A consequence of (3) is that the distance between two programs can always be approximated via Lipschitz tropical polynomial approximants of the initial two programs.
-\begin{corollary}
- Let $\Gamma \vdash_{\STLC} M:A$ and $\Delta\vdash_{\STLC} N:B$.
- For all $\epsilon>0$, $x\in\Lawv^{\model \Gamma}$, $b\in\model A$, there exist $t\in\Te{M}$, $u\in\Te{N}$ s.t.\ $\big| \model{\Gamma \vdash_{\STLC} M:A}^!(x)_b - \model{\Delta \vdash_{\STLC} N:B}^!(x)_b \big| \leq 2\epsilon + \big| \model{\Gamma \vdash_{\STDLC} t:A}^!(x)_b - \model{\Delta \vdash_{\STDLC} u:B}^!(x)_b \big|$.
-\end{corollary}
-
-\begin{remark}
- Let $X$, $Y$ be sets and let $\langle \_,\_\rangle:X\times Y \to \mathbb{R}$.
- For $f:X\to \mathbb R$, define $f^*:Y\to \mathbb R$ by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
- Then for $X=!A$, $Y=\Lawv^A$, where $A$ is a set, and $\langle \mu, y \rangle:= \mu y$, we have that $f^!=(-f)^*$ for all $f\in\Lawv^{!A}$.
- This is precisely the same formal construction yielding the well-known convex conjugate $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
- This construction is in turn a generalisation of the Legendre transformation.
- Despite the formal constructions being the same, we ignore for the moment if these could be connected to the study of high-order programs in our setting.
-\end{remark}
+\subsection{Probabilities and a worked out motivational example}\label{sec:proba}
+\input{section5}

### 2023-07-26 CSL24/section3B.tex
Commit
Intanto committo, ovviamente è ancora tutto da aggiustare, ma almeno se vuoi vedere la nuova organizzazione ce l'hai disponibile.

@@
 
-Remember that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
-%We will explicit this constructions in our tropical setting in the next lines.
 As well-known, if a SMCC is \emph{Lafont}, then one obtains a CCC from it by defining the exponential objects as $X\to Y:=!X \multimap Y$.
@@
 So the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.
+
+One can see that the matrix interpreting a $\STLC$-term is always boolean (i.e.\ its coefficients are either $0$ or $\infty$) can be seen as the fact that $\STLC$ is relatively trivial, from this point of view.
+This is also the case for the differential $\lambda$-calculus that we will introduce in a moment, so we state and discuss this property at that point (\autoref{prop:descrete}).
 %It is instructive what its CCC-structure looks like in our tropical world.
-The cartesian product $\&$ of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms $\RM{ev}$ are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_!
-t\in\Lawv^{!X\times Z}$, $(s\circ_!
-t)_{\mu,c}:=
-%\begin{align}
-\inf_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
- \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
-%\end{align}
-%where $+$ is the multiset union.
-
-%Remember that in $\Lawv$ the neutral element for addition is $\infty$ and the neutral for multiplication is $0$, so for instance the evaluation map is the matrix $\RM{eval}^{X,Y}\in\Lawv^{!((X\multimap Y) \times X)\times Y}\simeq\Lawv^{(!!X\times !Y\times !X)\times Y}$ given by $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=0$ if $\rho_1=[\mu]$ and $\rho_2=[b]$, and $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=\infty$ otherwise.
 
@@
 %In our case, we have:
+
+For what concerns the probabilistic calculus, let us consider a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
+we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the usual rules and whose operational semantics adds the usual rules:
+{\small{\[\begin{array}{cccc}
+           \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
+           &
+           \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
+           &
+           M\oplus_p N \to pM
+           &
+           M\oplus_p N \to (1-p)N
+          \end{array}\]}}
+%We add reduction rules:$ M\oplus_p N \to pM \quad$ and $\quad M\oplus_p N \to (1-p)N,$
+so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
+We use this calculus as a toy example for our purposes.
+It can be compiled into $\mathrm{PCF}^{\Lawv}$, an instance of the $\mathrm{PCF}^{\mathcal R}$-languages studied in \cite{Manzo2013}, via the translation $(\_)^\circ$ of terms generated by $(M\oplus_p N)^\circ:=pM^\circ\prog{ or } (1-p)N^\circ$.
+It is easy to see that if $\Gamma \vdash_{\STLC_\oplus} M:A$, then $\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A$.
+This is slightly imprecise because in \cite{Manzo2013}, their $\mathrm{PCF}^{\mathcal R}$-languages only have one ground type $\mathrm{int}$, used for modeling $\N$, and has no term $\prog{True}$, $\prog{False}$.
+So we are actually compiling $\STLC_\oplus$ into the immediate extension of $\mathrm{PCF}^{\mathcal R}$ which adds two new ground types, $*$ and $\bool$ (and, since we will not consider the integers, we can also drop $\mathrm{int}$), plus the two boolean terms.
+This does not invalidate the results (not concerning $\mathrm{int}$) of \cite{Manzo2013}.
+Remark also that the operational semantics of $\mathrm{PCF}^{\Lawv}$ \cite[Fig.\ 1]{Manzo2013} simulates the one of $\STLC_\oplus$, potentially using multiple steps, e.g.\ $(M\oplus_p N)^\circ \overset{0}{\rightarrow} pM^\circ \overset{p}{\rightarrow} M^\circ$.
+Since $\LREL_!$ is an instance of models for $\mathrm{PCF}^{\mathcal{R}}$, similarly to \cite[Section VI]{Manzo2013}, we use it in order to get a model of $\STLC_\oplus$, by setting $\model{\Gamma \vdash_{\STLC_\oplus} M:A}:=\model{\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A}\in\HOM{\LREL_!}{\model{\Gamma}}{\model{A}}$.
+Here we set $\model{\bool}:=\{0,1\}$ and $\model{\vdash_{\mathrm{PCF}^{\Lawv}} \true:\bool}\in\Lawv^{\{0,1\}}$ gives value $0$ on $1$ and value $\infty$ on $0$, similarly for $\false$.
+
+Since $\LREL_!$ is an instance of models for $\mathrm{PCF}^{\mathcal{R}}$, similarly to \cite[Section VI]{Manzo2013}, we use it in order to get a model of $\STLC_\oplus$, by setting $\model{\Gamma \vdash_{\STLC_\oplus} M:A}:=\model{\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A}\in\HOM{\LREL_!}{\model{\Gamma}}{\model{A}}$.
+Here we set $\model{\bool}:=\{0,1\}$ and $\model{\vdash_{\mathrm{PCF}^{\Lawv}} \true:\bool}\in\Lawv^{\{0,1\}}$ gives value $0$ on $1$ and value $\infty$ on $0$, similarly for $\false$.
+
+\begin{remark}
+ In the probabilistic case, the matrix interpreting a term need no longer be boolean.
+ For instance, $\model{\vdash_{\STLC_\oplus} \true\oplus_p\false :\bool}_i=p$ if $i=1$ and $=1-p$ if $i=0$.
+ %In general, one can show that $\model{\Gamma \vdash_{\STLC_\oplus} M:A}\in[0,1]\cup\{\infty\}^{!\model \Gamma \times \model A}$.
+Coefficients $> 1$ may be considered when dealing with other kinds of effects.
+\end{remark}

### 2023-07-26 CSL24/section4.tex
Commit
Intanto committo, ovviamente è ancora tutto da aggiustare, ma almeno se vuoi vedere la nuova organizzazione ce l'hai disponibile.

@@
 
-\subsection{From tropical polynomials to tLs}\label{sec:4A}
-
-A \emph{tropical polynomial} is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$, where $i_{j}\in\N$ and $c_{i_{j}}\in\Lawv$.
-For example, the polynomials $\varphi_{n}(x)=\min_{0\leq j\leq n}\{jx+2^{-j}\}$
-are illustrated in Fig.~\ref{fig:plot1} for $0\leq n \leq 4$.
-
-%\begin{figure}
-%\begin{subfigure}{0.4\textwidth}
-%\begin{tikzpicture}[scale=0.6]
-%\begin{axis}[samples=250]
-%\addplot[yellow,domain=0:0.8] {1+0.02};
-%
-%\addplot[orange,domain=0:0.8] {min(x+1/2, 1)+0.01};
-%
-%
-%\addplot[red,domain=0:0.8] {min(2*x+1/4, x+1/2, 1};
-%\addplot[blue,domain=0:0.8] {min(3*x+1/8,2*x+1/4, x+1/2, 1)-0.01};
-%\addplot[orange,domain=0:0.8] {min(4*x+1/16,3*x+1/8,2*x+1/4, x+1/2, 1)-0.02};
-%
-%\addplot[violet,domain=0:0.8] {min(
-%10*x+1/1424,
-%9*x+1/712,
-%8*x+1/356,
-%7*x+1/128,
-%6*x+1/64,
-%5*x+1/32,
-%4*x+1/16,3*x+1/8,2*x+1/4, x+1/2, 1)-.03};
-%
-%
-%\end{axis}
-%
-%\end{tikzpicture}
-%\caption{}
-%\label{fig:plot1}
-%\end{subfigure}
-%\begin{subfigure}{0.4\textwidth}
-%\begin{tikzpicture}[scale=0.6]
-%\begin{axis}[samples=50, view={15}{45}]
-%\addplot3[orange,domain=0:2] {min(2*x, 2*x+y, 3*y)};
-%
-%\end{axis}
-%\end{tikzpicture}
-%\caption{}
-%\label{fig:plot2}
-%\end{subfigure}
-%\label{fig:plot12}
-%\caption{
-%\ref{fig:plot1}
-% (a) Plot of the tropical polynomials $\varphi_{n}$, for $0\leq n\leq 4$ (from top to bottom), and of their limit tLs $\varphi$ (in violet).
-The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x=2^{-(i+1)}$, satisfying $ix+2^{-i}=(i+1)x+2^{-(i+1)}$.
-%\ref{fig:plot2}
-%(b) Plot of $\varphi(x,y)=\min\{2x, 2x+y,3y\}$.
-%}
-%\end{figure}
-
-\begin{wrapfigure}{r}{0.5\textwidth}%\begin{figure}
-\begin{tikzpicture}[scale=0.9]
-\begin{axis}[samples=250]
-\addplot[yellow,domain=0:0.8] {1+0.02};
-
-\addplot[orange,domain=0:0.8] {min(x+1/2, 1)+0.01};
-
-\addplot[red,domain=0:0.8] {min(2*x+1/4, x+1/2, 1};
-\addplot[blue,domain=0:0.8] {min(3*x+1/8,2*x+1/4, x+1/2, 1)-0.01};
-\addplot[orange,domain=0:0.8] {min(4*x+1/16,3*x+1/8,2*x+1/4, x+1/2, 1)-0.02};
-
-\addplot[violet,domain=0:0.8] {min(
-10*x+1/1424,
-9*x+1/712,
-8*x+1/356,
-7*x+1/128,
-6*x+1/64,
-5*x+1/32,
-4*x+1/16,3*x+1/8,2*x+1/4, x+1/2, 1)-.03};
-
-\end{axis}
-
-\end{tikzpicture}
-\caption{\small Tropical polynomials $\varphi_0,\dots,\varphi_4$ (top to bottom), and the limit tLs $\varphi$ (in violet).
-The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x=2^{-(i+1)}$, satisfying $ix+2^{-i}=(i+1)x+2^{-(i+1)}$.}
-\label{fig:plot1}%\end{figure}%
-\end{wrapfigure} %%
-
-A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable.
-In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes).
-For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
-Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
-$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
-%Tropical polynomials and their roots are the main object of study of tropical geometry.
-%
-%A
-% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.
-
-A \emph{tropical Laurent series} of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\{nx+\matr f_{n}\}$, with $\matr  f_{n}\in\Lawv$.
-In other words, it is a ``limit'' of tropical polynomials of higher and higher degree.
-E.g., the tLs $\varphi(x):=\inf_{n\in\N}\set{nx+2^{-n}}$ (see Fig.~\ref{fig:plot1}), that we will take as our running example, 
-is the ``limit'' of the polynomials $\varphi_{n}$.
-Since $\inf$s are not in general $\min$s, the behaviour of tLS may be less predictable than that of tropical polynomials~\cite{Porzio2021}.
-%For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points of their domain.
-
-%
-%
-%We will take as our running example
-%%given respectively by Equation~\ref{eq:polytrop} and Equation~\ref{eq:defTLS}.
-%A positive $x\in[0,\infty)$ is said to be a (finite) \emph{tropical root} of a tropical Laurent series $f:\Lawv\to\Lawv$ iff $f$ is not differentiable at $x$ (w.r.t.\ the usual topology on $\BB R_{\geq0}$.This is equivalent to ask that the $\inf$ defining $f$ at $x$ is obtained \emph{at least} twice.{\color{red} \`e vero anche per TLS o solo per poly questo?
-Inoltre, Robol parla anche di finite end-points of the domain: nel nostro caso sarebbe $x=0$}
-
-%\begin{example}\label{ex:famous_ex}
-% The function $f:\Lawv\to\Lawv$ defined by $f(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ is a tropical Laurent series.
-%\end{example}
-
-%By plotting its graph {\color{red}vogliamo plottarlo?}, we observe several properties that we will lift to the more general case that we will consider in the next lines, and this example will serve as running one.
-
 %\begin{remark}%[Tropical Laurent series]
@@
 
- In a similar way, tLs
-$f:\Lawv^X\to\Lawv^Y$ 
- with \emph{finite} supports $\C F_b=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$, and which have thus shape $f(x)_b=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, are generalisation of usual tropical polynomials to the case of possibly infinitely many variables.
-If $\model *$ is finite, this is in particular the case for the interpretations of $\BSTLC$-terms, since then the interpretation of any $\BSTLC$-type $B$ is a finite set, and thus also $!_n \model B$ is.
+If $\model *$ is finite, then the interpretations of $\BSTLC$-terms is a (generalised) tropical polynomial, since then the interpretation of any $\BSTLC$-type $B$ is a finite set, and thus also $!_n \model B$ is.
 % 

### 2023-07-26 CSL24/section5.tex
Commit
Intanto committo, ovviamente è ancora tutto da aggiustare, ma almeno se vuoi vedere la nuova organizzazione ce l'hai disponibile.

--- 2023-07-24 CSL24/section5.tex
@@
-%In this section we illustrate a few directions in which the tropical semantics just introduced could be used to analyze quantitative properties of higher-order programs.
+%%In this section we illustrate a few directions in which the tropical semantics just introduced could be used to analyze quantitative properties of higher-order programs.
 
-Since algebraic and geometric properties in tropical mathematics are usually more tractable from a computational point of view, in several well-known applications (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}) one starts from a given model, typically expressed by some polynomial function $f$, and studies
-what properties of the model can be deduced from the \emph{tropicalization} of $f$, noted $\trop f$, i.e.~the transformation of $f$ into a tropical polynomial.
+%Since algebraic and geometric properties in tropical mathematics are usually more tractable from a computational point of view, in several well-known applications (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}) one starts from a given model, typically expressed by some polynomial function $f$, and studies  what properties of the model can be deduced from the \emph{tropicalization} of $f$, noted $\trop f$, i.e.~the transformation of $f$ into a tropical polynomial.
+Here we follow a similar pattern: we consider a program $M$, which can be expressed in the form of a polynomial or a power series $f$, and we  investigate what quantitative properties of $M$ can be deduced from the properties of $\trop f$, that will indeed coincide with the interpretation of $M$ in $\LREL_{!}$.
+
 %
 %
-%several well-known applications of tropical mathematics is to study how much can be deduced of some function starting from the properties of its tropicalization.
-%In Section \ref{section5} we will follow a similar direction, investigating what quantitative properties of a higher-order programs are revealed by the study of its tropical interpretation.
+%%several well-known applications of tropical mathematics is to study how much can be deduced of some function starting from the properties of its tropicalization.
+%%In Section \ref{section5} we will follow a similar direction, investigating what quantitative properties of a higher-order programs are revealed by the study of its tropical interpretation.
 %
-Here we follow a similar pattern: we consider a program $M$, which can be expressed in the form of a polynomial or a power series $f$, and we 
-investigate what quantitative properties of $M$ can be deduced from the properties of $\trop f$, that will indeed coincide with the interpretation of $M$ in $\LREL_{!}$.
 
-\input{tropicalization}
+%\input{tropicalization}
 
@@
 \end{comment}
-
-\subsection{Log-probabilities with a worked out example}
-
-Let us consider a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
-we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the usual rules and whose operational semantics adds the usual rules:
-{\small{\[\begin{array}{cccc}
-           \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
-           &
-           \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
-           &
-           M\oplus_p N \to pM
-           &
-           M\oplus_p N \to (1-p)N
-          \end{array}\]}}
-%We add reduction rules:$ M\oplus_p N \to pM \quad$ and $\quad M\oplus_p N \to (1-p)N,$
-so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
-We use this calculus as a toy example for our purposes.
-It can be compiled into $\mathrm{PCF}^{\Lawv}$, an instance of the $\mathrm{PCF}^{\mathcal R}$-languages studied in \cite{Manzo2013}, via the translation $(\_)^\circ$ of terms generated by $(M\oplus_p N)^\circ:=pM^\circ\prog{ or } (1-p)N^\circ$.
-It is easy to see that if $\Gamma \vdash_{\STLC_\oplus} M:A$, then $\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A$.
-This is slightly imprecise because in \cite{Manzo2013}, their $\mathrm{PCF}^{\mathcal R}$-languages only have one ground type $\mathrm{int}$, used for modeling $\N$, and has no term $\prog{True}$, $\prog{False}$.
-So we are actually compiling $\STLC_\oplus$ into the immediate extension of $\mathrm{PCF}^{\mathcal R}$ which adds two new ground types, $*$ and $\bool$ (and, since we will not consider the integers, we can also drop $\mathrm{int}$), plus the two boolean terms.
-This does not invalidate the results (not concerning $\mathrm{int}$) of \cite{Manzo2013}.
-Remark also that the operational semantics of $\mathrm{PCF}^{\Lawv}$ \cite[Fig.\ 1]{Manzo2013} simulates the one of $\STLC_\oplus$, potentially using multiple steps, e.g.\ $(M\oplus_p N)^\circ \overset{0}{\rightarrow} pM^\circ \overset{p}{\rightarrow} M^\circ$.
-Since $\LREL_!$ is an instance of models for $\mathrm{PCF}^{\mathcal{R}}$, similarly to \cite[Section VI]{Manzo2013}, we use it in order to get a model of $\STLC_\oplus$, by setting $\model{\Gamma \vdash_{\STLC_\oplus} M:A}:=\model{\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A}\in\HOM{\LREL_!}{\model{\Gamma}}{\model{A}}$.
-Here we set $\model{\bool}:=\{0,1\}$ and $\model{\vdash_{\mathrm{PCF}^{\Lawv}} \true:\bool}\in\Lawv^{\{0,1\}}$ gives value $0$ on $1$ and value $\infty$ on $0$, similarly for $\false$.
-
-\begin{remark}
- Now the matrix interpreting a term need no longer be boolean.
- For instance, $\model{\vdash_{\STLC_\oplus} \true:\bool}_i$ is $p$ if $i=1$ and it is $1-p$ if $i=0$.
- In general, one can show that $\model{\Gamma \vdash_{\STLC_\oplus} M:A}\in[0,1]^{!\model \Gamma \times \model A}$.
- Coefficients $> 1$ may be considered when dealing with other kinds of effects.
-\end{remark}
 

### 2023-07-26 CSL24/section2bis.tex
Committino pre-cena

@@
 
-A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable.
-In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes).
-For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
+A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable ((i.e.~where the slope of $\varphi$ changes), or equivalently, the points where the minimum defining $\varphi$ is attained at least twice (see \autoref{fig:plot1}).
+%For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
 Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
-$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
+$f(x)=k\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
 %Tropical polynomials and their roots are the main object of study of tropical geometry.

### 2023-07-26 CSL24/section2bis.tex
Altro commit pre-cena

@@
 
-A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable ((i.e.~where the slope of $\varphi$ changes), or equivalently, the points where the minimum defining $\varphi$ is attained at least twice (see \autoref{fig:plot1}).
+A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable (i.e.~where the slope of $\varphi$ changes), or equivalently, where the minimum defining $\varphi$ is attained at least twice (see \autoref{fig:plot1}).
 %For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
@@
 In other words, it is a ``limit'' of tropical polynomials of higher and higher degree.
-E.g., the tLs $\varphi(x):=\inf_{n\in\N}\set{nx+2^{-n}}$ (see Fig.~\ref{fig:plot1}), that we will take as our running example, 
+E.g., the tLs $\varphi(x):=\inf_{n\in\N}\set{nx+2^{-n}}$ (see Fig.~\ref{fig:plot1}) %that we will take as our running example, 
 is the ``limit'' of the polynomials $\varphi_{n}$.
@@
 
+Tropical Laurent series arise from a general categorical construction:
+it is well known that $\LREL$ admits a comonad $!$ which acts, on objects, by taking the finite multisets, and remember that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
+%We will explicit this constructions in our tropical setting in the next lines.
+Now, although a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, by exploiting the coKleisli structure we can also ``express it in the base $X$'', i.e.\ see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_!
+x$.
+%(we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
+Concretely, we have $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
+%\begin{remark}%[Tropical Laurent series]
+These functions corresponds then to tLs with \emph{possibly infinitely} many variables (in fact, as many as the elements of $X$), %In the following we will also refer to them as tLs. 
+% We will call them simply \emph{tropical Laurent series (tLs)}.
+% %Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
+%\end{remark}
+%
+%We find the usual notion of tLs of one variable as follows:
+%\begin{remark}
+and by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the %functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ we recover the
+usual tLs's of one variable.
+% \end{remark}
+
 In a similar way, tLs
@@
 
-Remember that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
-%We will explicit this constructions in our tropical setting in the next lines.
-
-The cartesian product $\&$ of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms $\RM{ev}$ are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_!
-t\in\Lawv^{!X\times Z}$, $(s\circ_!
-t)_{\mu,c}:=
-%\begin{align}
-\inf_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
- \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
 %\end{align}

### 2023-07-26 CSL24/section4.tex
Altro commit pre-cena

@@
-For starting, remark that although a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, by exploiting the coKleisli structure we can also ``express it in the base $X$'', i.e.\ see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_!
-x$.
-%(we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
-Concretely, we have $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
-
 In this section we establish some topological and metric properties of such functions, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
@@
 
-%\begin{remark}%[Tropical Laurent series]
- Remark that the map $t^!:\Lawv^X\to\Lawv^Y$ introduced above from a matrix $t\in\HOM{\LREL_!}{X}{Y}$, corresponds to a tLs with \emph{possibly infinitely} many variables (in fact, as many as the elements of $X$).
-%In the following we will also refer to them as tLs.
-% We will call them simply \emph{tropical Laurent series (tLs)}.
-% %Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
-%\end{remark}
-%
-%We find the usual notion of tLs of one variable as follows:
-%\begin{remark}
-By identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the %functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ we recover the
-usual tLs's of one variable.
-% \end{remark}
  \begin{remark}

### 2023-07-27 CSL24/section1.tex
Commit notturno

@@
 can thus be interpreted as a non-expansive map from $2(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.
+These observations have led to the study of linear $\lambda$-calculi with \emph{graded} exponentials, like $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied to the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}.
+The types of such systems are defined by combining linear constructors with a \emph{graded linear exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
 
@@
 The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Diff{M}{N}$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Diff{M}{N}$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).
+Another intriguing similarity between program derivatives and 
+actual derivatives is provided by the \emph{Taylor expansion} $\C T$:
+%In analysis, a sufficiently regular function can be seen (locally around a point) as its Taylor expansion, i.e.\ a series of polynomials weighted via some rational coefficient.
+%Now, a polynomial in $x$ of degree $k$ can be thought of as a function which uses its argument $x$ exactly $k$ times.
+in $\STDLC$ (even untyped), we can perform a syntactic Taylor expansion of an ordinary $\lambda$-term via an inductively defined map $\Theta$ giving rise to an infinite series of terms, whose crucial case of the definition is
+$
+\Theta(MN)=\sum_{n\in\mathbb{N}}\frac{1}{n!}\Der^n\left[\Theta(M),\Theta(N)^n\right]0.
+$
+As in analysis, it decomposes an application as a series of $k$-linear functions, which can be seen as its approximants.
+In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.
+Since we only consider idempotent sum, the factorial coefficient disappears, and the resulting Taylor map is called the \emph{qualitative} Taylor expansion, indicated by $\Te{\_}$.
 

### 2023-07-27 CSL24/section2.tex
Commit notturno

@@
 
-These observations have led to the study of linear $\lambda$-calculi with \emph{graded} exponentials, like $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied to the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}.
-The types of such systems are defined by combining linear constructors with a \emph{graded linear exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
 The language $\BSTLC$ we consider is a simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}. 
-The terms are as for the $\STLC$, the types are $A::= * \ \mid  \ !_{n}A \multimap A$ and type judgements are of the form $\Gamma \vdash M:A$, where a context $\Gamma$ is a set of declarations of the form $x :_{n}A$, with $n\in \mathbb N$.
-The typing rules are: %given in Fig.~\ref{fig:rules}
+The terms are as for the $\STLC$, the types are $A::= * \ \mid  \ !_{n}A \multimap A$, the contexts of the typing judgements are sets of declarations of the form $x :_{n}A$, with $n\in \mathbb N$, and the typing rules are: %given in Fig.~\ref{fig:rules}
 	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
@@
 The axiom is $x:_{1}A\vdash x: A$.
-The main feature of this language is that if $\vdash \lambda x.M:\,!_nA\multimap B$, then $x$ will be duplicated exactly $n$ times in the reduction to the normal form.
+The main feature of this language is that if $\vdash \lambda x.M:\,!_nA\multimap B$, then $x$ is duplicated at most $n$ times in the reduction of $\vdash (\lambda x.M)N :B$ to the normal form.
 %E.g., $\vdash_{\BSTLC} \lambda {\color{red}z}.\left( \lambda x{\color{green}y}.
@@
 The syntax of the differential $\lambda$-calculus $\STDLC$ (e.g.\ \cite[Section 3]{Manzo2010}), is given by the \emph{terms} $M$ and the \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
-quotiented by: $\alpha$-equivalence; $+,0$ form a commutative monoid on the set of sums; %, i.e.\ commutativity and associativity of $+$ and neutrality of $0$ w.r.t.\ $+$;
-linearity of $\lam x.(\_)$, $\Diff{\_}{\_}$ and $(\_)\mathbb T$ (but \emph{not} of $M(\_)$); irrelevance of the order of consecutive $\Diff{\_}{\_}$.
+quotiented by $\alpha$-equivalence, by equations that make $+,0$ form a commutative monoid on the set of sums, %, i.e.\ commutativity and associativity of $+$ and neutrality of $0$ w.r.t.\ $+$;
+by linearity of $\lam x.(\_)$, $\Diff{\_}{\_}$ and $(\_)\mathbb T$ (but \emph{not} of $M(\_)$) and by irrelevance of the order of consecutive $\Diff{\_}{\_}$.
 %Remark that $M(\_)$ is \emph{not} set to be linear: $\lambda x.0=0\mathbb T=\Diff{0}{N}=\Diff{M}{0}=0$ but $M0\neq0$ in general.
@@
 %Sums are, then, just \emph{finite} sets of terms.
-Simple types are $A::= *\mid A\to A$.
-The typing rules are: %in Figure~\ref{fig:rules}
+The types are $A::= *\mid A\to A$, the typing rules: %in Figure~\ref{fig:rules}
 	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
@@
 The axioms are $\Gamma, x:A \vdash x: A$ and $\Gamma\vdash 0:A$.
-
+The main feature of this language is that $\Der^n[\lambda x.M,N^n]0$ has a non-zero normal form iff $x$ is duplicated exactly $n$ times during the reduction to normal form.
 %In $\BSTLC$ the typing system handles duplications; in $\STDLC$ the syntax with its operational semantics (that we do not give) does it.
@@
 %Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
-
-Another intriguing similarity between program derivatives and 
-actual derivatives is provided by the \emph{Taylor expansion} $\C T$:
-%In analysis, a sufficiently regular function can be seen (locally around a point) as its Taylor expansion, i.e.\ a series of polynomials weighted via some rational coefficient.
-%Now, a polynomial in $x$ of degree $k$ can be thought of as a function which uses its argument $x$ exactly $k$ times.
-in $\STDLC$ (even untyped), we can perform a syntactic Taylor expansion of an ordinary $\lambda$-term via an inductively defined map $\Theta$ giving rise to an infinite series of terms, whose crucial case of the definition is
-$
-\Theta(MN)=\sum_{n\in\mathbb{N}}\frac{1}{n!}\Der^n\left[\Theta(M),\Theta(N)^n\right]0.
-$
-As in analysis, it decomposes an application as a series of $k$-linear functions, which can be seen as its approximants.
-In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.
-Since we only consider idempotent sum, the factorial coefficient disappears, and the resulting Taylor map is called the \emph{qualitative} Taylor expansion, indicated by $\Te{\_}$.
-% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
 
@@
 %This ensures that one can define a sound interpretation of $\STDLC$-terms in the standard way (see [Section 4.3, \cite{Manzo2010}]).
-This model is also well-behaved w.r.t.\ the Taylor expansion, as expressed by the following two results.
-As it can be patientely checked, in $(\LREL_!,D)$ all morphisms can be Taylor expanded  (see \cite[Definition 4.22]{Manzo2012}):
+
+As already mentioned, by induction on $M$ %(using that composition and $D$ easily preserve ``booleaness'' and projections and evaluation of  $\LREL$ are boolean)
+one can prove that:
+
+\begin{proposition}\label{prop:descrete}
+ For $\BSTLC$ and $\STDLC$ (but not $\STLC_\oplus$), we have that $\model{\Gamma\vdash M:A}\in\Lawv^{!\model{\Gamma}\times\model{A}}$ is a \emph{boolean} matrix, i.e.\ actually $\model{\Gamma\vdash M:A}\in\{0,\infty\}^{!\model{\Gamma}\times\model{A}}$.
+\end{proposition}
+
+Even if the interpretation of those languages is trivia, we shall still focus on them in order to set the basis of the further studies and in order to see some already interesting properties of their interpretation.{\color{red}SPIEGARE MEGLIO}
+
+Let us conclude this subsection by recalling that the Taylor expansion $\Te{MN}$ of $MN$ is the set $\set{\Der^{n}[t,u_1,\dots,u_n]0 \mid n\in\N,t\in\Te{M},u_i\in\Te{N}}$.
+%One defines its interpretation as $\model{\Te{MN}}:=\inf_{n\in\N,t\in\Te{M},u_i\in\Te{N}} \model{\Der^{n}[t,u_1,\dots,u_n]0}$.
+$\LREL_!$ is also well-behaved w.r.t.\ the Taylor expansion, as expressed by the following two results.
+First, it can be patientely checked that in $(\LREL_!,D)$ all morphisms can be Taylor expanded  (see \cite[Definition 4.22]{Manzo2012}):
 
@@
 %Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
-Finally, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of \cite[Theorem 4.23]{Manzo2012} shows:
+Second, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of \cite[Theorem 4.23]{Manzo2012} shows:
 
 \begin{corollary}\label{cor:T(M)=M}
- If $\Gamma\vdash_{\STLC} M:A$, then $\model{\Gamma\vdash_{\STDLC} \Te M:A}:=\inf_{t\in\Te{M}} \model{\Gamma\vdash_{\STDLC} t:A}$ converges to $\model{\Gamma\vdash_{\STLC} M:A}$.
+ For $\Gamma\vdash_{\STLC} M:A$, we have %$\model{\Gamma\vdash_{\STDLC} \Te M:A}:=
+$\model{\Gamma\vdash_{\STLC} M:A}=\inf_{t\in\Te{M}} \model{\Gamma\vdash_{\STDLC} t:A}$.
 %the interpretation of the Taylor expansion of a $\STLC$-term $M$, given in \eqref{eq:taylor}, converges to the one of $M$.
 \end{corollary}
-
-As already mentioned, by straightforward induction on $M$ (using that composition and $D$ easily preserve ``booleaness'' and projections and evaluation of  $\LREL$ are boolean) one can prove that:
-
-\begin{proposition}\label{prop:descrete}
- For all the previous four languages, except $\STLC_\oplus$, we have that $\model{\Gamma\vdash M:A}\in\Lawv^{!\model{\Gamma}\times\model{A}}$ is a \emph{boolean} matrix, i.e.\ actually $\model{\Gamma\vdash M:A}\in\{0,\infty\}^{!\model{\Gamma}\times\model{A}}$.
-\end{proposition}
-
-Even if the interpretation of those languages is trivia, we shall still focus on them in order to set the basis of the further studies and in order to see some already interesting properties of their interpretation.{\color{red}SPIEGARE MEGLIO}
 

### 2023-07-27 CSL24/section2bis.tex
Commit notturno

@@
 
+\begin{remark}
+ There is a clear and well-known relation between tropical polynomial/power series and usual polynomials/power series:
+fix a (formal) power series $f=\sum_{i\in I\subseteq\N} a_i x^i$ (polynomials being the case for $I$ finite) with coefficitents $a_i$ in a semiring $Q$;
+fix a \emph{valuation} $\mathrm{val}:Q\to \Lawv$.
+Then one defines the \emph{tropicalisation} $\trop f:\Lawv \to \Lawv$ of $f$ as $\trop f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
+Often, if $Q=[0,1]$, one takes as valuation $\mathrm{val}(a)=-\log a$.
+\end{remark}
+
 \subsection{Probabilities and a worked out motivational example}\label{sec:proba}

### 2023-07-27 CSL24/section3B.tex
Altro commit pre-cena
Commit notturno

@@
 So the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.
+The cartesian product $\&$ of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms $\RM{ev}$ are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_!
+t\in\Lawv^{!X\times Z}$, $(s\circ_!
+t)_{\mu,c}:=
+%\begin{align}
+\inf_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
+ \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
 
@@
 
-For what concerns the probabilistic calculus, let us consider a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
-we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the usual rules and whose operational semantics adds the usual rules:
-{\small{\[\begin{array}{cccc}
-           \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
-           &
-           \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
-           &
-           M\oplus_p N \to pM
-           &
-           M\oplus_p N \to (1-p)N
-          \end{array}\]}}
-%We add reduction rules:$ M\oplus_p N \to pM \quad$ and $\quad M\oplus_p N \to (1-p)N,$
-so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
-We use this calculus as a toy example for our purposes.
+Remember the probabilistic calculus $\STLC_\oplus$.
 It can be compiled into $\mathrm{PCF}^{\Lawv}$, an instance of the $\mathrm{PCF}^{\mathcal R}$-languages studied in \cite{Manzo2013}, via the translation $(\_)^\circ$ of terms generated by $(M\oplus_p N)^\circ:=pM^\circ\prog{ or } (1-p)N^\circ$.
@@
 
-Since $\LREL_!$ is an instance of models for $\mathrm{PCF}^{\mathcal{R}}$, similarly to \cite[Section VI]{Manzo2013}, we use it in order to get a model of $\STLC_\oplus$, by setting $\model{\Gamma \vdash_{\STLC_\oplus} M:A}:=\model{\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A}\in\HOM{\LREL_!}{\model{\Gamma}}{\model{A}}$.
-Here we set $\model{\bool}:=\{0,1\}$ and $\model{\vdash_{\mathrm{PCF}^{\Lawv}} \true:\bool}\in\Lawv^{\{0,1\}}$ gives value $0$ on $1$ and value $\infty$ on $0$, similarly for $\false$.
-
 \begin{remark}

### 2023-07-27 CSL24/section4.tex
Commit notturno

@@
 %It follows after adapting \cite[Proposition 4.4]{Cobzas2017} in order to prove that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
-\cite[Proposition 4.4.(3)]{Cobzas2017} shows that if a real-valued convex function with domain a convex open subset of a locally convex topological $\BB R$-vector space is, locally around any point, bounded from above by a finite non-zero constant, then it is continuous on all its domain.
+\cite[Proposition 4.4.(3)]{Cobzas2017} shows that if a real-valued convex function with domain a convex open subset of a locally convex topological $\BB R$-vector space (LCTVS) is, locally around any point, bounded from above by a finite non-zero constant, then it is continuous on all its domain.
 Now, our $f$ is concave, so $-f$ is convex.
@@
 %Remark that we cannot immediately apply \cite[Proposition 4.4.(3)]{Cobzas2017}, since $\Lawv^X$ is not even a $\BB R$-vector space.
-Now $(0,\infty)^X\subseteq \BB R^X$ is open and convex, and {\color{red}$\BB R^X$ is a locally convex topological $\BB R$-vector space}, so \cite[Proposition 4.4.(3)]{Cobzas2017} entails the continuity of $-f\big|_{(0,\infty)^X}$, hence that of $f\big|_{(0,\infty)^X}$.
+Now $(0,\infty)^X\subseteq \BB R^X$ is open and convex, and {\color{red}$\BB R^X$ is a LCTVS}, so \cite[Proposition 4.4.(3)]{Cobzas2017} entails the continuity of $-f\big|_{(0,\infty)^X}$, hence that of $f\big|_{(0,\infty)^X}$.
 \end{proof}

### 2023-07-27 CSL24/section5.tex
Commit notturno

@@
 
+For what concerns the probabilistic calculus, let us consider a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
+we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the usual rules and whose operational semantics adds the usual rules:
+{\small{\[\begin{array}{cccc}
+           \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
+           &
+           \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
+           &
+           M\oplus_p N \to pM
+           &
+           M\oplus_p N \to (1-p)N
+          \end{array}\]}}
+%We add reduction rules:$ M\oplus_p N \to pM \quad$ and $\quad M\oplus_p N \to (1-p)N,$
+so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
+We use this calculus as a toy example for our purposes.
 Consider %the following closed term $M$ of type $\bool$:

### 2023-07-27 CSL24/main.tex
Commit pre-cena
Guarda in particolare la sezione 2.3 ed il suo remark 3. In generale, io sto scrivendo un po' tutte le cose che vanno dette, poi vediamo come scriverle bene

@@
 
-\title{Tropical mathematics for the Lambda-calculus} %TODO Please add
+\title{Tropical mathematics for the Lambda-calculus 1}
+\subtitle{The potential of tropical wheighted relational semantics%for probabilities and Taylor expansion
+} %TODO Please add
 
-%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line
+\titlerunning{Tropical mathematics for the Lambda-calculus 1} %TODO optional, please use if title is longer than one line
 
@@
 
-\section{Two kinds of quantitative approaches}
+\section{Introduction}
 \input{section1}
 
-\section{The category $\LREL$}\label{section3}
+\section{Preliminaries and motivation}
+\input{section2bis}
 
-\input{section3new}
+\section{Tropical weighted relational semantics for high-order languagesi}\label{section3}
 
-\subsection{... As (weighted relational) model of several $\lambda$-calculi}
 The category $\LREL$ satisfies many nice properties which make it able to model all the variants of the $\lambda$-calculi to which we are interested in here.
@@
 
-\paragraph{Prohibited duplication/erasure: the \emph{linear} $\STLC$}\label{sec:3A}\input{section3A}
+\paragraph*{Prohibited duplication/erasure: the \emph{linear} $\STLC$}\label{sec:3A}
+\input{section3A}
 
-\paragraph{Allowed duplication/erasure: the $\STLC$}\label{sec:3B}\input{section3B}
+\paragraph*{Allowed duplication/erasure: the $\STLC$ and the probabilistic $\lambda$-calculus $\STLC_\oplus$}\label{sec:STLC+STLCO}
 
-\paragraph{Controlled duplication/erasure: the $\BSTLC$ and the $\STDLC$}\label{sec:3C}\input{section3C}
+\input{section3B}
 
-\section{Tropical Laurent Series}\label{sec:tls}\input{section4}
+\input{section2}
 
-\section{Future lines of research}\label{sec:future}
+\section{Tropical Laurent Series}\label{sec:tls}
+\input{section4}
 
-{\color{red}3 pagine in totale}
+%\section{Future lines of research}\label{sec:future}
 
-\subsection{Quantitaive analysis}\label{sec:app}
-\input{section5}
-%\input{section5C}
+%\subsection{Quantitaive analysis}\label{sec:app}
+%%\input{section5C}
 
-\subsection{Generalized Metric Spaces and Modules over the Lawvere Quantale}\label{section6}
+%\subsection{Generalized Metric Spaces and Modules over the Lawvere Quantale}\label{section6}
 %%At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}.
@@
 
-\newpage
-
 \section{Related Work}\label{section7} \input{section7}
 
-\section{Conclusion}\label{section8} \input{section8}
+\section{Conclusion and future work}\label{section8} \input{section8}
 
@@
 
-\appendix
+%\appendix
 
-\section{\autoref{section3}}\label{app-sec:section3}
+%\section{\autoref{section3}}\label{app-sec:section3}
 

### 2023-07-27 CSL24/section2bis.tex
Commit pre-cena
Guarda in particolare la sezione 2.3 ed il suo remark 3. In generale, io sto scrivendo un po' tutte le cose che vanno dette, poi vediamo come scriverle bene

@@
 fix a \emph{valuation} $\mathrm{val}:Q\to \Lawv$.
-Then one defines the \emph{tropicalisation} $\trop f:\Lawv \to \Lawv$ of $f$ as $\trop f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
-Often, if $Q=[0,1]$, one takes as valuation $\mathrm{val}(a)=-\log a$.
+Then one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
+The trivial valutation is the constantly equal to $0$ one, called the \emph{null valuation}.
+A common one, for $Q=[0,1]$, is $\mathrm{val}(a)=-\log a$.
 \end{remark}
 
-\subsection{Probabilities and a worked out motivational example}\label{sec:proba}
+\subsection{Log-Probabilities and Tropicalisation through an exemple}\label{sec:proba}
 \input{section5}

### 2023-07-27 CSL24/section3B.tex
Commit pre-cena
Guarda in particolare la sezione 2.3 ed il suo remark 3. In generale, io sto scrivendo un po' tutte le cose che vanno dette, poi vediamo come scriverle bene

@@
 
+\begin{remark}
 One can see that the matrix interpreting a $\STLC$-term is always boolean (i.e.\ its coefficients are either $0$ or $\infty$) can be seen as the fact that $\STLC$ is relatively trivial, from this point of view.
@@
 %It is instructive what its CCC-structure looks like in our tropical world.
+\end{remark}
 
@@
 
-Remember the probabilistic calculus $\STLC_\oplus$.
+For what concerns the probabilistic calculus, let us consider a probabilistic extension of $\STLC$, call it $\STLC_\oplus$, which also extends the first-order language we used in~\autoref{sec:proba}:
+we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the usual rules and whose operational semantics adds the usual rules:
+{\small{\[\begin{array}{cccc}
+           \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
+           &
+           \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
+           &
+           M\oplus_p N \to pM
+           &
+           M\oplus_p N \to (1-p)N.
+          \end{array}\]}}
+We use this calculus as a toy example for our purposes.
 It can be compiled into $\mathrm{PCF}^{\Lawv}$, an instance of the $\mathrm{PCF}^{\mathcal R}$-languages studied in \cite{Manzo2013}, via the translation $(\_)^\circ$ of terms generated by $(M\oplus_p N)^\circ:=pM^\circ\prog{ or } (1-p)N^\circ$.

### 2023-07-27 CSL24/section5.tex
Commit pre-cena
Guarda in particolare la sezione 2.3 ed il suo remark 3. In generale, io sto scrivendo un po' tutte le cose che vanno dette, poi vediamo come scriverle bene

--- 2023-07-26 CSL24/section5.tex
@@
 
+As a toy exemple, let us consider a first-order probabilistic calculus on booleans:
+the terms are $M::= \true \mid \false \mid M\oplus_p M \mid pM$, for $p\in[0,1]$, and the operational semantics is $M\oplus_p N\to pM$ and $M\oplus_p N \to (1-p)N$, so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
+
 Consider %the following closed term $M$ of type $\bool$:
 $
- \vdash_{\STLC_\oplus} M:=(\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true)):\bool.
+ M:=(\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true)).
  $
-Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, reading $0$ as ``left'' and $1$ as ``right''.
+Let us give addresses $\omega\in\set{ll,lr,rll,rlr,rrl,rrr}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, ($l$ is ``left'' and $r$ is ``right'').
 The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
-%For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$ starting from the left.
+%For instance, $rll$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{rll}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$ starting from the left.
 Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
-$P_{00}(p,q)\true$, 
-$P_{100}(p,q)\true$, 
-$P_{111}(p,q)\true$, 
-$P_{01}(p,q)\false$, 
-$P_{110}(p,q)\false$,
-$P_{101}(p,q)\false$,
-where the $P$'s are the following monomials in $p,q$:
-$P_{00}(p,q):=p^2$,
-$P_{100}(p,q):=qp^2$,
-$P_{111}(p,q):=q^3$,
-$P_{01}(p,q):=pq$,
-$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
-They correspond to the respective reduction path from $M$ to the normal term of the same address.
-$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ after all the coin tossings in $M$ are performed.
-Thinking of $p,q$ as parameters, $P_{\omega}$ can thus be read as the \emph{likelihood function} of the event $\omega$.
-The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after all the tossings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.
-
-%Let us consider in this subsection a probabilistic extension of $\lam$-calculus, call it $\STLC_\oplus$, adding as usual terms of shape $pM+qN$ and $M\oplus_p N$, for $p,q\in[0,1]$.These terms are typed via the rule:
+$P_{ll}(p,q)\true$, 
+$P_{rll}(p,q)\true$, 
+$P_{rrr}(p,q)\true$, 
+$P_{lr}(p,q)\false$, 
+$P_{rrl}(p,q)\false$,
+$P_{rlr}(p,q)\false$,
+where the $P$'s are the following monomial functions in $p,q$:
+$P_{ll}(p,q):=p^2$,
+$P_{rll}(p,q):=qp^2$,
+$P_{rrr}(p,q):=q^3$,
+$P_{lr}(p,q):=pq$,
+$P_{rrl}(p,q)=P_{rlr}(p,q):=q^2p$.
+%They correspond to the respective reduction path from $M$ to the normal term of the same address.
+$P_{\omega}(p,1-p)$ is then the probability of the event ``$M\twoheadrightarrow \true_\omega/\false_\omega$'' (depending on $\omega$).
+Thinking of $p,q$ as parameters, $P_{\omega}(p,q)$ can thus be read as the \emph{likelihood function} of the event ``$M\twoheadrightarrow \true_\omega/\false_\omega$''.
+The polynomial function $Q_{\true}(p,q):=P_{ll}(p,q)+P_{rll}(p,q)+P_{rrr}(p,q)=p^2+p^2q+q^3$ gives instead the probability of the event ``$M\twoheadrightarrow \true$'', and analogously for $Q_{\false}(p,q):=P_{lr}(p,q)+P_{rrl}(p,q)+P_{rlr}(p,q)=pq+2pq^2$.
+%Let us consider in this subsection a probabilistic extension of $\lam$-calculus, call it $\STLC_\oplus$, adding as usual terms of shape $pM+qN$ and $M\oplus_p N$, for $p,q\in[l,r]$.These terms are typed via the rule:
 %\[
@@
 %\[
-% M\oplus_p N \to pM+(1-p)N
+% M\oplus_p N \to pM+(r-p)N
 %\]
@@
 %\[
-% M\twoheadrightarrow (p^2+(1-p)p^2+(1-p)^3)I+(p(1-p)+2(1-p)^2p)\Omega.
+% M\twoheadrightarrow (p^2+(r-p)p^2+(r-p)^3)I+(p(r-p)+2(r-p)^2p)\Omega.
 %\]
-%The index $\omega\in\set{00,100,111,01,110,101}$ of each $P_\omega$ indicates the path of the reduction that led from $M$ to the respective occurrence $I_\omega$ of $I$ or $\Omega_\omega$ of $\Omega$ from $M$ to its normal form ($0$ means ``left'' and $1$ means ``right'').For instance, in order to reach $I_{100}$, i.e.\ the second occurrence of $I$ from the left in $M$, we have to take the right path in the outer $\oplus_p$ of $M$, then two times the left path in the new outer $\oplus_p$'s that we encounter during reduction.$P_{\omega}(p,q)$ gives then the probability (as a function of $p,q$) of obtaining the respective occurrence $I_{\omega}$ or $\Omega_\omega$ in the normal form, if we were to sample at each time we reduce a $\oplus_p$.It can thus also be read as the likelihood function of such event.The polynomials $Q_{1,2}(p,q)$ give instead the whole probability of obtaining respectively $I$ or $\Omega$, in the normal form after such samplings.
+%The index $\omega\in\set{ll,rll,rrr,lr,rrl,rlr}$ of each $P_\omega$ indicates the path of the reduction that led from $M$ to the respective occurrence $I_\omega$ of $I$ or $\Omega_\omega$ of $\Omega$ from $M$ to its normal form ($l$ means ``left'' and $r$ means ``right'').For instance, in order to reach $I_{rll}$, i.e.\ the second occurrence of $I$ from the left in $M$, we have to take the right path in the outer $\oplus_p$ of $M$, then two times the left path in the new outer $\oplus_p$'s that we encounter during reduction.$P_{\omega}(p,q)$ gives then the probability (as a function of $p,q$) of obtaining the respective occurrence $I_{\omega}$ or $\Omega_\omega$ in the normal form, if we were to sample at each time we reduce a $\oplus_p$.It can thus also be read as the likelihood function of such event.The polynomials $Q_{r,2}(p,q)$ give instead the whole probability of obtaining respectively $I$ or $\Omega$, in the normal form after such samplings.
+This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baumr966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2ll4}.
 
-This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}.
-Typical questions in this case would be:
+Typical questions in this case would be, for a fixed $\omega_0$:
 %
@@
 \begin{enumerate}
- \item What is the \emph{maximum likelihood estimator} for the event ``$M$ produces $\true_\omega$'' (similarly for $\false_\omega$)?
- I.e., which is the choice of the parameters $p,q$ that maximizes the probability of getting $\true_\omega$ after all tossings in $M$?
- \item Knowing that $M$ produced $\true$ (similarly for $\false$), which is the choice of the parameters $p,q$ that maximizes the probability that the produced occurrence of $\true$ was a fixed $\true_{\omega_0}$ (i.e.
-that among the paths leading to $\true$, the one taken was $\omega_0$)?
+ \item Which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow \false_{\omega_0}$''?
+ I.e., which is the choice of $p,q$ that maximizes the probability $P_{\omega_0}$ of the event ``$M\twoheadrightarrow \false_{\omega_0}$''  ?
+ \item Which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow \false_{\omega_0}$'', knowing that ``$M\twoheadrightarrow \false$''?
+I.e., which is the choice of $p,q$ that makes $\omega_0$ the most likely path among those leading to $\false$ (i.e.\ that maximizes the conditional probability $\BB P(``M\twoheadrightarrow \false_{\omega_0}'' \mid ``M\twoheadrightarrow \false'')$)?
 \end{enumerate}
-A similar argument could be done by replacing $\true$ and $\false$ by, respectively, a converging and a diverging term (e.g.~in a $\B{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.
+%A similar argument could be done by replacing $\false$ and $\true$ by, respectively, a converging and a diverging term (e.g.~in a $\B{PCF}$-style language), so r) would be about finding maximum likelihood estimators for the event ``$M$ converges''.
 
-%These are questions much in the style of those asked in {\color{red}Ref?} in the framework of {\color{red}??}.
-Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega}, Q_\omega$, which is more easily solved by 
-passing to the tropical monomials/polynomial $\trop P_{\omega},\trop Q_{\omega}$. 
-For 1), we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)\in
+Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega_0}, Q_{\omega_0}$, which is more easily solved by 
+passing to the tropical monomial/polynomial functions $\trop^{\mathrm{val}} P_{\omega_0},\trop^{\mathrm{val}} Q_{\omega_0}$. 
+For 1), by definition of $\RM{arg max}$ and because $-\log$ is stricly decreasing, we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)\in
 %\begin{equation}
 %  \begin{array}{ccccc}
-   \underset{(x,y)}{\RM{arg max}} P_\omega (x,y)
+   \RM{arg max}_{(x,y)} P_{\omega_0} (x,y)
    %& 
    = %&
-   \underset{(x,y)}{\RM{arg min}}\set{-\log_c P_\omega (x,y)}
+   \RM{arg min}_{(x,y)}\set{-\log P_{\omega_0} (x,y)}
    %&
    =%&
-   \underset{(x,y)}{\RM{arg max}}\set{(\trop P_\omega) (-\log_c x,-\log_c y)} \label{eq:argmax}$
+   \RM{arg min}_{(x,y)}\set{(\trop P_{\omega_0}) (-\log x,-\log y)} \label{eq:argmax}$
+where this holds for any valuation.
+Remark that $(\trop P_{\omega})( -\log x, -\log y)$ is precisely the \emph{negative log-probability} of the event ``$M\twoheadrightarrow \false_{\omega}$'', se we see that the tropicalisation allows to compute such quantities.
 %  \end{array}
 %\end{equation}
-where the equalities hold for all $0<c\neq1$.
-For 2), %the $\omega_0$ is s.t.\ $\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$. So
+For 2), %the $\omega_l$ is s.t.\ $\max\limits_{\omega\in\set{ll,rll,rrr}} \, P_\omega(x,y) = P_{\omega_l}(x,y)$. So
 we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
-$\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$.
-This last condition is equivalent to ask that $\min\limits_{\omega\in\set{00,100,111}} \, -\log_c P_\omega(x,y) = -\log_c P_{\omega_0}(x,y)$, i.e.\ $(\trop Q_1)( -\log_c x, -\log_c y) = (\trop P_{\omega_0})( -\log_c x, -\log_c y)\label{eq:max}$.
-In both cases, passing through $\trop $ %P_\omega, \trop Q_\omega$ 
-makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2015}).
+$\max_{\omega\in\set{lr,rrl,rlr}} \, P_\omega(p,q) = P_{\omega_0}(p,q)$, i.e.\ $\min_{\omega\in\set{lr,rrl,rlr}} \, -\log P_\omega(p,q) = -\log P_{\omega_0}(p,q)$.
+Remembering that $\trop^{\mathrm{val}} Q_{\false}(p,q)=\min\set{p+q, \mathrm{val}(2)+p+2q}$, we see that our minimization problem is equivalent to the equality $(\trop^0 Q_{\false})( -\log p, -\log q) = (\trop^0 P_{\omega_0})( -\log x, -\log y)\label{eq:max}$
+and this holds only for the null valuation.
+Remark that, in both cases, passing through $\trop^0 $ %P_\omega, \trop Q_\omega$ 
+makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2lr5}), and this basically corresponds to study negative log-probabilities.
 %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.
 
-\begin{remark}
- The interpretation $\model{\vdash_{\STLC_\oplus} M:\bool}\in\Lawv^{\set{0,1}}$ of our running example gives the tropicalised probabilities: $\model M_0=(\trop Q_0)(p,1-p)$, $\model M_1=(\trop Q_1)(p,1-p)$.
- Therefore, the model $\LREL_!$ expresses such optimisation problems 
-% (in $p$ and $q$, not after the substitution $q:=1-p$) 
-% are extracted from it,
- \ref{eq:argmax}, \ref{eq:max} of $\STLC_\oplus$-programs. 
+%For our running example $M$, we have $\trop Q_{\true}(x,y)=\min\set{2x,y+2x,3y}$ and $\trop Q_{\false}(x,y)=\min\set{x+y,2y+x}$.
+Studying $\trop Q_{\true}$ %, whose plot is in Fig.~\ref{fig:plot2}, we see that $\trop Q_{\true}(x,y)=3y$ iff $y\leq \frac{2}{3}x$, and it coincides with $2x$ otherwise.
+Remembering that $3y=P_{rrr}(x,y)$, we can now solve the optimisation problem~\ref{eq:max} for $\omega_l=rrr$: via the substitution $x:=-\log p$, $y:=-\log (r-p)$, Equation~\ref{eq:max} is equivalent to $-\log (r-p)\leq -\frac{2}{3}\log_c p$, i.e.\ $r-p\geq p^{\frac{2}{3}}$.
+This means that, for $p\in[0,1]$ s.t.\ $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), the most likely occurrence of $\true$ to obtain, knowing that $M$ sampled $\true$ in its normal form, is $\true_{rrr}$.
+Remembering that $2x=P_{ll}(x,y)$, for the other values of $p$ (for example, $p=\frac{1}{2}$), the most likely $\true$ to be sampled is the occurrence $\true_{ll}$.
+We have thus answered question 2) above for $\true$.
+
+From this situation we notice the following important:
+
+\begin{remark}\label{rmk:tropof01Rel}
+This toy first-order language can be interpreted in the already mentioned $\overline{R_{\geq 0}}\mathrm{Rel}$ and in $\LREL$.
+We do not give details now since, from the following section, we will introduce the interpretation of interesting high-order calculi, including a probabilistic one containing the one of this section.
+But it is important to mention already at this point that the probabilities are already captured by the $[0,\infty]\mathrm{Rel}$ model: the $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}\in[0,\infty]^{\set{0,1}}$ of our running example $M$ is $\model M_0=Q_{\false}(p,1-p)$, $\model M_1=Q_{\true}(p,1-p)$.
+Therefore, this optimisation problems are already expressible by taking $\trop^0\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$. 
+Now, the model $\LREL$ is precisely null-valuation tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$, i.e.\ $\model{M}^{\LREL}=\trop^0\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$.
+This corresponds to quotienting the polynomial interpreting $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$ w.r.t.\ idempotent sum, as the null valuation eliminates all the coefficients (different from $1$).
+A precise study of the ``tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$'' is left for future investigations, as it is related with power series arising from more sophisticated calculi with paramethers (in the style of \cite{} {\color{red}Ehrhard!}).
+We therefore have that $\model{M}^{\LREL}_0(-\log p,-\log (1-p)$ is the negative log-probability of \emph{any} of the (equiprobable) most likely paths of reduction to normal form.
+We take these observations as a motivation for considering such model, and the point of this work is to dig into such model, concentrating for this first paper on the relations between the metric and differential tools that we will consider.
+% (in $p$ and $q$, not after the substitution $q:=r-p$) 
+% are extracted from it, \ref{eq:argmax}, \ref{eq:max} of $\STLC_\oplus$-programs. 
 \end{remark}
@@
 %For instance, for our running example $M$:
-%\[\model M = \min\left\{\trop{Q_1}(p,1-p) \cdot \model I, \trop{Q_0}(p,1-p) \cdot \model \Omega\right\}.\]
-
-%\begin{example}\label{ex:study}
-For our running example $M$, we have $\trop Q_1(x,y)=\min\set{2x,y+2x,3y}$ and $\trop Q_0(x,y)=\min\set{x+y,2y+x}$.
-Studying $\trop Q_1$ %, whose plot is in Fig.~\ref{fig:plot2}, 
-we see that $\trop Q_1(x,y)=3y$ iff $y\leq \frac{2}{3}x$, and it coincides with $2x$ otherwise.
-Remembering that $3y=P_{111}(x,y)$, we can now solve the optimisation problem~\ref{eq:max} for $\omega_0=111$:
-via the substitution $x:=-\log_c p$, $y:=-\log_c (1-p)$, Equation~\ref{eq:max} is equivalent to $-\log_c (1-p)\leq -\frac{2}{3}\log_c p$, i.e.\ $1-p\geq p^{\frac{2}{3}}$.
-This means that, for $p\in[0,1]$ s.t.\ $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), the most likely occurrence of $\true$ to obtain, knowing that $M$ sampled $\true$ in its normal form, is $\true_{111}$.
-Remembering that $2x=P_{00}(x,y)$, for the other values of $p$ (for example, $p=\frac{1}{2}$), the most likely $\true$ to be sampled is the occurrence $\true_{00}$.
-We have thus answered question 2) above.
-%\end{example}
-
-\begin{remark}\label{rem:troproots}
- Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_1$ or of $\trop Q_0$ provide, by definition, the values of the bias of $\oplus_p$ for which there are at least two different equiprobable paths of tossings from $M$ to its normal form.
- Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log_c p,-\log_c(1-p))$ is a tropical root of, say, $\trop Q_1$ are the values of the bias of $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are smapled by $M$ during its tossings to normal form, knowing that some $\true$ was sampled.
-\end{remark}
+%\[\model M = \min\left\{\trop{Q_r}(p,r-p) \cdot \model I, \trop{Q_l}(p,r-p) \cdot \model \Omega\right\}.\]

### 2023-07-27 CSL24/section8.tex
Commit pre-cena
Guarda in particolare la sezione 2.3 ed il suo remark 3. In generale, io sto scrivendo un po' tutte le cose che vanno dette, poi vediamo come scriverle bene

--- 2023-01-22 section8.tex
@@
+{\color{red}due parole sul fatto che non sappiamo se si posso usare strumenti propri della tropical geo per lo studio dei programmi. Per esempio:}
+\begin{remark}\label{rem:troproots}
+ Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_{\true}$ or of $\trop Q_{\false}$ provide, by definition, the values of the bias of $\oplus_p$ for which there are at least two different equiprobable paths of tossings from $M$ to its normal form.
+ Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log p,-\log(r-p))$ is a tropical root of, say, $\trop Q_{\true}$ are the values of the bias of $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are smapled by $M$ during its tossings to normal form, knowing that some $\true$ was sampled.
+\end{remark}
+
+{\color{red}due parole generalised metric spaces and modules}
+
+{\color{red}due parole su PCoh, $\overline{R_{\geq 0}}$Rel e le loro tropicalizzazioni (gia accennato in \autoref{rmk:tropof01Rel})}
+
 %In our opinion,
@@
 Notably, knowing that the application of a program $M$ to $N$ expands as a finite sum of linear applications may allow one to predict how sensitive $M$ will be ``around $N$''.
-%We have shown that:
-%
-%- intuitively, the interplay between tropical mathematics and $\lam$-calculus \emph{could} relate the metric and differential approaches on approximations of $\lam$-calc (introduction+section 2)
-%
-%- this relation \emph{can} take place, because the natural category $\LREL$ and its generalised versions are metric models of the differential $\lam$-calculus (section 3 and 6)
-%
-%- this relation \emph{seems} to provide applications in different fields (section 5).
-%
-%Therefore, we mainly set the basis for future interplays between all these areas, hopefully motivating the interest in such an interplay.
-%For instance, the general questions are:
-%
-%- Can we improve the results of section 4 ?
-%
-%- Can we develop and make the applications of section 5 useful ?
-%
-%- What does the general setting of section 6 give in terms of theoretical and applied results ?
-%
-%- Do tools from tropical \emph{geometry} provide something for $\lam$-calculus ? (for instance, the role of tropical roots, tropical varieties,...)
-%
-%- Finally, there is a last point which we think of interest and we did not mention through the paper: the inclusion of finiteness spaces in the picture.
-%Say what could they do and why.
+
+\newpage

### 2023-07-28 CSL24/section2bis.tex
Committino notturno

@@
 fix a \emph{valuation} $\mathrm{val}:Q\to \Lawv$.
+Remember that a valuation by definition always sends $1\in Q$ to $0$.
 Then one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
-The trivial valutation is the constantly equal to $0$ one, called the \emph{null valuation}.
-A common one, for $Q=[0,1]$, is $\mathrm{val}(a)=-\log a$.
+The constantly equal to $0$ valuation is called the \emph{null valuation}.
+A common valuation on $Q=[0,1]$ is $\mathrm{val}(a)=-\log a$.
+\end{remark}
+
+\begin{remark}
+ Let $X$, $Y$ be sets and let $\langle \_,\_\rangle:X\times Y \to \mathbb{R}$.
+ For $f:X\to \mathbb R$, define $f^*:Y\to \mathbb R$ by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
+ Then for $X=!A$, $Y=\Lawv^A$, where $A$ is a set, and $\langle \mu, y \rangle:= \mu y$, we have that $f^!(y)=(-f)^*(-y)$ for all $f\in\Lawv^{!A}$.
+ This is precisely the same formal construction yielding the well-known convex conjugate $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
+ This construction is in turn a generalisation of the Legendre transformation.
+ Despite the formal constructions being the same, we ignore for the moment if these could be connected to the study of high-order programs in our setting.
 \end{remark}

### 2023-07-28 CSL24/section4.tex
Committino notturno

@@
 
-The results just presented translate into the following facts about the interpretation of higher-order programs:
+The results just presented translate into the following facts about the interpretation of higher-order programs.
+Remark that the metric spaces $(\Lawv^{\mathlarger{+}_{i=1}^n X_i},d_{\infty})$ and $(\prod_{i=1}^n \Lawv^{X_i},\max_{i=1}^n d^{X_i}_\infty)$ are trivially isometric, so we identify them.
+
+% Since $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}\in\HOM{\LREL_!}{\mathlarger{\&}_{i=1}^n \model{A_i}}{\model B}\simeq\Lawv^{\prod_{i=1}^n !
+\model {A_i}\times \model B}$, the tLs $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ is defined by $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !
+\model{A_i}}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
+%Finally, the interpretations of $\BSTLC$-terms are matrices $t\in\HOM{\LREL}{\bigotimes_{i=1}^n !_{n_i}X_i}{Y}=\Lawv^{\prod_{i=1}^n !_{n_i}X_i\times Y}$.In such situation, we define $t^!:\Lawv^{\mathlarger{+}_{i=1}^n X_{i}}\to\Lawv^Y \simeq \prod_{i=1}^n\Lawv^{X_{i}}\to\Lawv^Y$ as $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !_{n_i} X_i}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
+Clearly $t^!={\widetilde t}^!$, where $\widetilde t\in\Lawv^{\prod_{i=1}^n !
+{X_i}\times Y}$ is the matrix $\widetilde t_{(\mu_1,\dots,\mu_n),b}:=t_{(\mu_1,\dots,\mu_n),b}$ if $\mu_i\in !_{n_i} X_i$ for all $i$, and $:=\infty$ otherwise, which has the same support as $t$.
 
@@
 \item $\model{\Gamma \vdash_{\BSTLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{tropical polynomial}, thus (as $\model A$ is finite), a \emph{Lipschitz} function.
-\item $\model{\Gamma \vdash_{\STLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{locally} Lipschitz map.
+\item $\model{\Gamma \vdash_{\STLC} M:A}^!, \model{\Gamma \vdash_{\STLC_\oplus} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ are \emph{locally} Lipschitz maps.
 \item $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!$ of \emph{tropical polynomials}, thus (as $\model A$ is finite), \emph{Lipschitz} functions.
@@
 
+Remark that, if $Y$ is finite and $f:\Lawv^{\mathlarger{+}_{i=1}^n X_i}\to\Lawv^Y$ is $K$-Lipschitz, then $f:\prod_{i=1}^n \Lawv^{X_i}\to \Lawv^Y$ is $K$-Lipschitz on each of the $n$ variable separately.
+
 A consequence of (3) is that the pointwise distance between two interpretations of programs can always be bounded via Lipschitz tropical polynomial approximants of the initial two programs.
@@
 \end{corollary}
-
-\begin{remark}
- Let $X$, $Y$ be sets and let $\langle \_,\_\rangle:X\times Y \to \mathbb{R}$.
- For $f:X\to \mathbb R$, define $f^*:Y\to \mathbb R$ by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
- Then for $X=!A$, $Y=\Lawv^A$, where $A$ is a set, and $\langle \mu, y \rangle:= \mu y$, we have that $f^!=(-f)^*$ for all $f\in\Lawv^{!A}$.
- This is precisely the same formal construction yielding the well-known convex conjugate $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
- This construction is in turn a generalisation of the Legendre transformation.
- Despite the formal constructions being the same, we ignore for the moment if these could be connected to the study of high-order programs in our setting.
-\end{remark}

### 2023-07-28 CSL24/section5.tex
Committino notturno

@@
 A precise study of the ``tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$'' is left for future investigations, as it is related with power series arising from more sophisticated calculi with paramethers (in the style of \cite{} {\color{red}Ehrhard!}).
-We therefore have that $\model{M}^{\LREL}_0(-\log p,-\log (1-p)$ is the negative log-probability of \emph{any} of the (equiprobable) most likely paths of reduction to normal form.
+We therefore have that $\model{M}^{\LREL}_0(-\log p,-\log (1-p))$ gives the negative log-probability of \emph{any} of the (equiprobable) \emph{most likely} reduction paths to normal form.
 We take these observations as a motivation for considering such model, and the point of this work is to dig into such model, concentrating for this first paper on the relations between the metric and differential tools that we will consider.

### 2023-07-28 CSL24/section2.tex
Commit pre-cena

@@
 Even if the interpretation of those languages is trivia, we shall still focus on them in order to set the basis of the further studies and in order to see some already interesting properties of their interpretation.{\color{red}SPIEGARE MEGLIO}
-
-Let us conclude this subsection by recalling that the Taylor expansion $\Te{MN}$ of $MN$ is the set $\set{\Der^{n}[t,u_1,\dots,u_n]0 \mid n\in\N,t\in\Te{M},u_i\in\Te{N}}$.
-%One defines its interpretation as $\model{\Te{MN}}:=\inf_{n\in\N,t\in\Te{M},u_i\in\Te{N}} \model{\Der^{n}[t,u_1,\dots,u_n]0}$.
-$\LREL_!$ is also well-behaved w.r.t.\ the Taylor expansion, as expressed by the following two results.
-First, it can be patientely checked that in $(\LREL_!,D)$ all morphisms can be Taylor expanded  (see \cite[Definition 4.22]{Manzo2012}):
-
-\begin{theorem}\label{thm:modelsTaylor}
- For all $t\in\HOM{\LREL_!}{Z}{X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$ we have:%, the evaluation of $t$ over $s$ yields 
- \begin{align}\label{eq:taylorcat}
-  \RM{ev}\circ_!\langle t,s\rangle =
-  \inf\limits_{n\in\N}
-  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
- \end{align} 
-\end{theorem}
-%It is worth discussing the formula above a bit more. 
-Here,
-%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
-%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
-$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
-Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$,  i.e.~it coincides with the tropical %interpretation of the
-version of the usual Taylor expansion.
-%Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
-Second, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of \cite[Theorem 4.23]{Manzo2012} shows:
-
-\begin{corollary}\label{cor:T(M)=M}
- For $\Gamma\vdash_{\STLC} M:A$, we have %$\model{\Gamma\vdash_{\STDLC} \Te M:A}:=
-$\model{\Gamma\vdash_{\STLC} M:A}=\inf_{t\in\Te{M}} \model{\Gamma\vdash_{\STDLC} t:A}$.
-%the interpretation of the Taylor expansion of a $\STLC$-term $M$, given in \eqref{eq:taylor}, converges to the one of $M$.
-\end{corollary}
-
-%More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument. 
-%Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
-%For example, if $M$ is the term $\lambda fx.f(fx)$ considered before, $\Der[M]$ takes a first input $N$ and passes it linearly to $M$.
-Notice that there are two ways of doing so, corresponding to the two bound occurrences of $f$ in $M$: either by applying $N$ to $fx$, or by
-%applying $f$ linearly to $Nx$ (indeed, if $f$ were applied in an unrestricted way, it might duplicate $Nx$, so that $N$ would not be used linearly).
-This justifies the equation below, in which $\Der[M]$ is identified with the non-deterministic sum of the two possible linear choices:
-%\begin{align}
-%\Der\left[\lambda f x.f(fx)\right]\cdot N = 
-%\lambda fx. N(fx) + \left(\Der[f]\cdot (Nx)\right)(fx)
-%\end{align}
-%More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
-%
-%
-%The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy.
-\frac{\mathsf df(y)}{\mathsf dy}\cdot x$.
-Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
-%Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:
-
-%
-%More generally, the relational semantics interprets unbounded programs as \emph{analytic functions}, that is, as functions admitting a representation as power series.
-For instance, observing that an analytic map $f: \BB R\to \BB R$, where $f(x)=\sum_{n}\widehat f_{n}\cdot x^{n}$ is uniquely determined by the sequence $\widehat f_{n}$, the program $M_{\infty}:=\lambda fx.fx: (\BB R\To \BB R)\To (\BB R\To \BB R)$ is represented by the power series below:
-%\begin{align}
-%F_{\infty}(f,x)= \sum_{n=0}^{\infty} \widehat f_{n} x^{n}
-%\end{align}
-%By restricting ourselves to bounded applications, the terms in the power series become finite, that is, the interpretation becomes a \emph{polynomial}: for instance, the program $M_{2}:=\lambda fx.
-\sum_{i=0}^{2}\Der^{(i)}[f]\cdot x^{i}$, corresponding to passing $x$ \emph{at most twice} to $f$, is represented by the polynomial
-%\begin{align}
-%F_{\leq 2}(f,x)=\widehat f_{2} x^{2}  + \widehat f_{1}x +  \widehat f_{0} 
-%\end{align}
-% In this framework the differential operator is naturally represented by formal differentiation of polynomials, where, as one would expect, 
-% $\Der[\sum_{n}a_{n}x^{n}]=\sum_{n}\Der[a_{n}x^{n}]$ and $\Der[a_{0}x^{0}]=0$ and $\Der[a_{n+1}x^{n+1}]= (n+1)a_{n+1}x^{n}$, so that power series can be Taylor expanded.

### 2023-07-28 CSL24/section2bis.tex
Commit pre-cena

@@
 
-\begin{remark}
- There is a clear and well-known relation between tropical polynomial/power series and usual polynomials/power series:
-fix a (formal) power series $f=\sum_{i\in I\subseteq\N} a_i x^i$ (polynomials being the case for $I$ finite) with coefficitents $a_i$ in a semiring $Q$;
-fix a \emph{valuation} $\mathrm{val}:Q\to \Lawv$.
-Remember that a valuation by definition always sends $1\in Q$ to $0$.
-Then one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
-The constantly equal to $0$ valuation is called the \emph{null valuation}.
-A common valuation on $Q=[0,1]$ is $\mathrm{val}(a)=-\log a$.
+\begin{remark}\label{rmk:val_trop}
+ There is a clear and well-known relation between tropical polynomial/power series and usual polynomials/power series. 
+Given $Q,L$ semirings with units and zeros, $L$ commutative idempotent, if $L$ is totally ordered by the partial order: $\alpha \preceq \beta$ iff $\alpha +  \beta = \beta$, then one defines a \emph{valuation} \cite{Izhakian2015} to be a map $\mathrm{val}:Q\to L$ s.t.\ $\mathrm{val}(0)=0$, $\mathrm{val}(1)=1$, $\mathrm{val}(ab)=\mathrm{val}(a)\mathrm{val}(b)$, $\mathrm{val}(a+b)\preceq \max\set{\mathrm{val}(a),\mathrm{val}(b)}$.
+The valuation which is constantly equal to $1\in L$ on all the invertibles of $Q$ is called the \emph{trivial valuation} $\mathrm{val}_1$.
+
+Now, for a (formal) power series $f=\sum_{i\in I\subseteq\N} a_i x^i$ (polynomials being the case for $I$ finite) with coefficitents $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$ (notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$), one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
+
+Remark that tropicalising w.r.t.\ the trivial valuation gives $\trop^{\mathrm{val}_1} \sum_{i\in I} a_i x^i = \inf_{i\in I} \set{i\alpha} = \trop^{\mathrm{val}} \sum_{i\in I\subseteq\N} x^i$ for all valuation $\mathrm{val}$, because in $\Lawv$ the unit is $0$.
+%and the zero is $\infty$.
+%Remember that a_i \neq 0 by def of formal power series
+That is, it coincides to consider the same power series but with coefficients in $Q$ quotiented for the idempotency of $+$.
+
 \end{remark}

### 2023-07-28 CSL24/section4.tex
Commit pre-cena

@@
 
-The results just presented translate into the following facts about the interpretation of higher-order programs.
+\subsection{The tropical interpretation of high-order programs}\label{subsec:tro_int}
+
+Rcall that the Taylor expansion $\Te{MN}$ of $MN$ is the set $\set{\Der^{n}[t,u_1,\dots,u_n]0 \mid n\in\N,t\in\Te{M},u_i\in\Te{N}}$.
+%One defines its interpretation as $\model{\Te{MN}}:=\inf_{n\in\N,t\in\Te{M},u_i\in\Te{N}} \model{\Der^{n}[t,u_1,\dots,u_n]0}$.
+$\LREL_!$ is also well-behaved w.r.t.\ the Taylor expansion, as expressed by the following two results.
+First, it can be patientely checked that in $(\LREL_!,D)$ all morphisms can be Taylor expanded  (see \cite[Definition 4.22]{Manzo2012}):
+
+\begin{theorem}\label{thm:modelsTaylor}
+ For all $t\in\HOM{\LREL_!}{Z}{X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$ we have:%, the evaluation of $t$ over $s$ yields 
+ \begin{align}\label{eq:taylorcat}
+  \RM{ev}\circ_!\langle t,s\rangle =
+  \inf\limits_{n\in\N}
+  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
+ \end{align} 
+\end{theorem}
+%It is worth discussing the formula above a bit more. 
+Here,
+%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
+%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
+$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
+Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$,  i.e.~it coincides with the tropical %interpretation of the
+version of the usual Taylor expansion.
+%Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
+Second, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of \cite[Theorem 4.23]{Manzo2012} shows:
+
+\begin{corollary}\label{cor:T(M)=M}
+ For $\Gamma\vdash_{\STLC} M:A$, we have %$\model{\Gamma\vdash_{\STDLC} \Te M:A}:=
+$\model{\Gamma\vdash_{\STLC} M:A}=\inf_{t\in\Te{M}} \model{\Gamma\vdash_{\STDLC} t:A}$.
+%the interpretation of the Taylor expansion of a $\STLC$-term $M$, given in \eqref{eq:taylor}, converges to the one of $M$.
+\end{corollary}
+
+%More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument. 
+%Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
+%For example, if $M$ is the term $\lambda fx.f(fx)$ considered before, $\Der[M]$ takes a first input $N$ and passes it linearly to $M$.
+Notice that there are two ways of doing so, corresponding to the two bound occurrences of $f$ in $M$: either by applying $N$ to $fx$, or by
+%applying $f$ linearly to $Nx$ (indeed, if $f$ were applied in an unrestricted way, it might duplicate $Nx$, so that $N$ would not be used linearly).
+This justifies the equation below, in which $\Der[M]$ is identified with the non-deterministic sum of the two possible linear choices:
+%\begin{align}
+%\Der\left[\lambda f x.f(fx)\right]\cdot N = 
+%\lambda fx. N(fx) + \left(\Der[f]\cdot (Nx)\right)(fx)
+%\end{align}
+%More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
+%
+%
+%The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy.
+\frac{\mathsf df(y)}{\mathsf dy}\cdot x$.
+Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
+%Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:
+
+%
+%More generally, the relational semantics interprets unbounded programs as \emph{analytic functions}, that is, as functions admitting a representation as power series.
+For instance, observing that an analytic map $f: \BB R\to \BB R$, where $f(x)=\sum_{n}\widehat f_{n}\cdot x^{n}$ is uniquely determined by the sequence $\widehat f_{n}$, the program $M_{\infty}:=\lambda fx.fx: (\BB R\To \BB R)\To (\BB R\To \BB R)$ is represented by the power series below:
+%\begin{align}
+%F_{\infty}(f,x)= \sum_{n=0}^{\infty} \widehat f_{n} x^{n}
+%\end{align}
+%By restricting ourselves to bounded applications, the terms in the power series become finite, that is, the interpretation becomes a \emph{polynomial}: for instance, the program $M_{2}:=\lambda fx.
+\sum_{i=0}^{2}\Der^{(i)}[f]\cdot x^{i}$, corresponding to passing $x$ \emph{at most twice} to $f$, is represented by the polynomial
+%\begin{align}
+%F_{\leq 2}(f,x)=\widehat f_{2} x^{2}  + \widehat f_{1}x +  \widehat f_{0} 
+%\end{align}
+% In this framework the differential operator is naturally represented by formal differentiation of polynomials, where, as one would expect, 
+% $\Der[\sum_{n}a_{n}x^{n}]=\sum_{n}\Der[a_{n}x^{n}]$ and $\Der[a_{0}x^{0}]=0$ and $\Der[a_{n+1}x^{n+1}]= (n+1)a_{n+1}x^{n}$, so that power series can be Taylor expanded.
+
+The results presented in all this paper translate now into the following facts about the interpretation of higher-order programs.
 Remark that the metric spaces $(\Lawv^{\mathlarger{+}_{i=1}^n X_i},d_{\infty})$ and $(\prod_{i=1}^n \Lawv^{X_i},\max_{i=1}^n d^{X_i}_\infty)$ are trivially isometric, so we identify them.

### 2023-07-28 CSL24/section5.tex
Commit pre-cena

--- 2023-07-27 CSL24/section5.tex
@@
 Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega_0}, Q_{\omega_0}$, which is more easily solved by 
-passing to the tropical monomial/polynomial functions $\trop^{\mathrm{val}} P_{\omega_0},\trop^{\mathrm{val}} Q_{\omega_0}$. 
+passing to the tropical monomial/polynomial functions $\trop P_{\omega_0},\trop^{\mathrm{val}} Q_{\omega_0}$, for any fixed valuation.
 For 1), by definition of $\RM{arg max}$ and because $-\log$ is stricly decreasing, we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)\in
@@
    \RM{arg min}_{(x,y)}\set{(\trop P_{\omega_0}) (-\log x,-\log y)} \label{eq:argmax}$
-where this holds for any valuation.
-Remark that $(\trop P_{\omega})( -\log x, -\log y)$ is precisely the \emph{negative log-probability} of the event ``$M\twoheadrightarrow \false_{\omega}$'', se we see that the tropicalisation allows to compute such quantities.
+where this holds for any valuation (because monomials do not carry any coefficient).
+Remark that $(\trop P_{\omega})( -\log x, -\log y)$ is precisely the \emph{negative log-probability} of the event ``$M\twoheadrightarrow \false_{\omega}$'', so we see that the tropicalisation allows to compute such quantities.
 %  \end{array}
@@
 $\max_{\omega\in\set{lr,rrl,rlr}} \, P_\omega(p,q) = P_{\omega_0}(p,q)$, i.e.\ $\min_{\omega\in\set{lr,rrl,rlr}} \, -\log P_\omega(p,q) = -\log P_{\omega_0}(p,q)$.
-Remembering that $\trop^{\mathrm{val}} Q_{\false}(p,q)=\min\set{p+q, \mathrm{val}(2)+p+2q}$, we see that our minimization problem is equivalent to the equality $(\trop^0 Q_{\false})( -\log p, -\log q) = (\trop^0 P_{\omega_0})( -\log x, -\log y)\label{eq:max}$
-and this holds only for the null valuation.
+Remembering that $\trop^{\mathrm{val}} Q_{\false}(p,q)=\min\set{p+q, \mathrm{val}(2)+p+2q}$, we see that our minimization problem is equivalent to asking $(\trop^{\mathrm{val}_1} Q_{\false})( -\log p, -\log q) = (\trop^{\mathrm{val}_1} P_{\omega_0})( -\log x, -\log y)\label{eq:max}$,
+and this time this holds, in general, only for the trivial valuation.
 Remark that, in both cases, passing through $\trop^0 $ %P_\omega, \trop Q_\omega$ 
-makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2lr5}), and this basically corresponds to study negative log-probabilities.
+makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2lr5}), and this essentially corresponds to study negative log-probabilities.
 %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.
@@
 \begin{remark}\label{rmk:tropof01Rel}
-This toy first-order language can be interpreted in the already mentioned $\overline{R_{\geq 0}}\mathrm{Rel}$ and in $\LREL$.
-We do not give details now since, from the following section, we will introduce the interpretation of interesting high-order calculi, including a probabilistic one containing the one of this section.
-But it is important to mention already at this point that the probabilities are already captured by the $[0,\infty]\mathrm{Rel}$ model: the $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}\in[0,\infty]^{\set{0,1}}$ of our running example $M$ is $\model M_0=Q_{\false}(p,1-p)$, $\model M_1=Q_{\true}(p,1-p)$.
-Therefore, this optimisation problems are already expressible by taking $\trop^0\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$. 
-Now, the model $\LREL$ is precisely null-valuation tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$, i.e.\ $\model{M}^{\LREL}=\trop^0\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$.
-This corresponds to quotienting the polynomial interpreting $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$ w.r.t.\ idempotent sum, as the null valuation eliminates all the coefficients (different from $1$).
+Our toy first-order language can be interpreted both in the already mentioned $\overline{R_{\geq 0}}\mathrm{Rel}$ and in $\LREL$.
+We do not give details now since in the rest of the paper we will interpret interesting \emph{high-order} calculi, including a probabilistic one containing our toy language.
+But it is important to mention already at this point that the probabilities we just discussed above, are already captured by the $[0,\infty]\mathrm{Rel}$ model: the $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}\in[0,\infty]^{\set{0,1}}$ of our running example $M$ is $\model M_0=Q_{\false}(p,1-p)$, $\model M_1=Q_{\true}(p,1-p)$.
+Therefore, the optimisation problems concerning the negative log-probabilities and likelihoods are already expressible by taking $\trop^{\mathrm{val}_1}\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$. 
+Now, the model $\LREL$ is precisely trivial-valuation tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$, i.e.\ $\model{M}^{\LREL}=\trop^{\mathrm{val}_1}\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$.
+As we remarked in \autoref{rmk:val_trop}, this corresponds to quotienting the polynomial interpreting $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$ w.r.t.\ idempotent sum.
 A precise study of the ``tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$'' is left for future investigations, as it is related with power series arising from more sophisticated calculi with paramethers (in the style of \cite{} {\color{red}Ehrhard!}).
-We therefore have that $\model{M}^{\LREL}_0(-\log p,-\log (1-p)$ is the negative log-probability of \emph{any} of the (equiprobable) most likely paths of reduction to normal form.
-We take these observations as a motivation for considering such model, and the point of this work is to dig into such model, concentrating for this first paper on the relations between the metric and differential tools that we will consider.
+For now, just have in mind that $\model{M}^{\LREL}_0(-\log p,-\log (1-p))$ gives the negative log-probability of \emph{any} of the (equiprobable) \emph{most likely} reduction paths to normal form, so $\LREL$ directly expresses such optimisations problems (no need to pass through a tropicalisation).
+We take these observations as incouraging, and the point of this work is to dig into such model, concentrating for this first paper on the relations between the metric and differential properties that coexist in it.
 % (in $p$ and $q$, not after the substitution $q:=r-p$) 

### 2023-07-28 CSL24/main.tex
Altro commit pre-cena

@@
 
-\section{Tropical weighted relational semantics for high-order languagesi}\label{section3}
+\section{The category $\LREL$}\label{section3}
+
+\subsection{... As tropical linear/non-linear algebra}
+
+For $Q$ a continuous semiring, one lets the category $\QREL$ (\cite{Manzo2013} calls it $Q^\Pi$) have sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms, i.e.~$\QREL(X,Y):=Q^{X\times Y}$.
+%The identity morphism of $\QREL$ is the identity matrix $I\in Q^{X\times X}$,
+% given by $i_{a,a}=1$ and $i_{a,b\neq a}=0$, and
+The composition $st\in Q^{X\times Z}$ of $t\in Q^{X\times Y}$ and $s\in Q^{Y\times Z}$ is given by $(st)_{a,c}:=\sum_{b\in Y} s_{b,c}t_{a,b}$ (such series converges because $Q$ is continuous).
+As expected, $Q^X$ is a $Q$-semimodule and we identify $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$, which have shape $f(x)_b=:\sum_{a\in X} \matr f_{a,b}x_a$, for some matrix $\matr f\in Q^{X\times Y}$.
+% \begin{remark}
+ %Following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we chose to see a matrix $t$ from $X$ to $Y$ as a map $t:X\times Y\to Q$.
+% 
+% fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum_{b\in Y} s_{b,c}t_{a,b}$.
+Notice that usual linear algebra conventions correspond to work in $\QREL^{\op}$, %a matrix $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns
+e.g.\ the usual matrix-vector product defines a map $Q^Y\to Q^X$.
+Following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we are instead working with transpose matrices.
+%\end{remark}
+%
+%In $\QREL^{op}$ (which corresponds to systematically taking transpose matrices), composition coincides with the product matrix/matrix and $\hat{(\cdot)}$ with the product matrix/vector.
+%In order to avoid confusion, we will refer to a $t\in Q^{X\times Y}$ just as a \emph{matrix from $X$ to $Y$}.
+%
+%
+%
+%We must fix a convention for matrices: following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum_{b\in Y} s_{b,c}t_{a,b}$.
+%In linear algebra, a map $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns.
+%In particular, the product of such a matrix for a vector defines a map $Q^Y\to Q^X$.
+%Instead, we prefer to see a $t\in\HOM{\QREL}{X}{Y}$ as giving rise to a map $\hat t:Q^X\to Q^Y$ defined by $\hat t(x)_a:=\sum_{b\in Y} t_{a,b}x_a$.
+%In $\QREL^{op}$ (which corresponds to systematically taking transpose matrices), composition coincides with the product matrix/matrix and $\hat{(\cdot)}$ with the product matrix/vector.
+%In order to avoid confusion, we will refer to a $t\in Q^{X\times Y}$ just as a \emph{matrix from $X$ to $Y$}.
+%As it is expected, $Q^X$ is a $Q$-semimodule and the bijection $\hat{(\cdot)}$ identifies $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.
+
+The main actor of this paper is the category $\LREL$. % of sets and matrices with values over $\Lawv$ (which is a continuous semi-ring).
+%, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
+%The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring (w.r.t.\ its quantale order $\preceq$.
+%This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
+It is worth observing that the composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
+reads as \ $(st)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
+similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$, %induced by matrices
+which we call \emph{tropical linear}, are exactly those of shape $f(x)_b=\inf_{a\in X} \set{\matr f_{a,b}+x_a}$, for some $\matr f\in\Lawv^{X\times Y}$.
+%\end{remark}
+
+Tropical Laurent series arise from a general categorical construction:
+it is well known that $\LREL$ admits a comonad $!$ which acts, on objects, by taking the finite multisets, and remember that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
+%We will explicit this constructions in our tropical setting in the next lines.
+Now, although a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, by exploiting the coKleisli structure we can also ``express it in the base $X$'', i.e.\ see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_!
+x$.
+%(we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
+Concretely, we have $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
+%\begin{remark}%[Tropical Laurent series]
+These functions corresponds then to tLs with \emph{possibly infinitely} many variables (in fact, as many as the elements of $X$), %In the following we will also refer to them as tLs. 
+% We will call them simply \emph{tropical Laurent series (tLs)}.
+% %Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
+%\end{remark}
+%
+%We find the usual notion of tLs of one variable as follows:
+%\begin{remark}
+and by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the %functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ we recover the
+usual tLs's of one variable.
+% \end{remark}
+
+In a similar way, tLs
+$f:\Lawv^X\to\Lawv^Y$ 
+ with \emph{finite} supports $\C F_b=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$, and which have thus shape $f(x)_b=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, are generalisation of usual tropical polynomials to the case of possibly infinitely many variables.
+
+%\end{align}
+%where $+$ is the multiset union.
+
+%Remember that in $\Lawv$ the neutral element for addition is $\infty$ and the neutral for multiplication is $0$, so for instance the evaluation map is the matrix $\RM{eval}^{X,Y}\in\Lawv^{!((X\multimap Y) \times X)\times Y}\simeq\Lawv^{(!!X\times !Y\times !X)\times Y}$ given by $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=0$ if $\rho_1=[\mu]$ and $\rho_2=[b]$, and $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=\infty$ otherwise.
+
+\begin{remark}
+ Let $X$, $Y$ be sets and let $\langle \_,\_\rangle:X\times Y \to \mathbb{R}$.
+ For $f:X\to \mathbb R$, define $f^*:Y\to \mathbb R$ by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
+ Then for $X=!A$, $Y=\Lawv^A$, where $A$ is a set, and $\langle \mu, y \rangle:= \mu y$, we have that $f^!(y)=(-f)^*(-y)$ for all $f\in\Lawv^{!A}$.
+ This is precisely the same formal construction yielding the well-known convex conjugate $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
+ This construction is in turn a generalisation of the Legendre transformation.
+ Despite the formal constructions being the same, we ignore for the moment if these could be connected to the study of high-order programs in our setting.
+\end{remark}
+
+\subsection{... As tropical weighted relational semantics for high-order languages}
 

### 2023-07-28 CSL24/section2bis.tex
Altro commit pre-cena

@@
-\subsection{Tropical linear algebra}
+\subsection{Elements of tropical mathematics}
 \input{section3new}
 
-\subsection{Tropical non-linear algebra}\label{sec:4A}
+%\subsection{Tropical non-linear algebra}\label{sec:4A}
 
@@
 
-Tropical Laurent series arise from a general categorical construction:
-it is well known that $\LREL$ admits a comonad $!$ which acts, on objects, by taking the finite multisets, and remember that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
-%We will explicit this constructions in our tropical setting in the next lines.
-Now, although a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, by exploiting the coKleisli structure we can also ``express it in the base $X$'', i.e.\ see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_!
-x$.
-%(we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
-Concretely, we have $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
-%\begin{remark}%[Tropical Laurent series]
-These functions corresponds then to tLs with \emph{possibly infinitely} many variables (in fact, as many as the elements of $X$), %In the following we will also refer to them as tLs. 
-% We will call them simply \emph{tropical Laurent series (tLs)}.
-% %Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
-%\end{remark}
-%
-%We find the usual notion of tLs of one variable as follows:
-%\begin{remark}
-and by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the %functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ we recover the
-usual tLs's of one variable.
-% \end{remark}
-
-In a similar way, tLs
-$f:\Lawv^X\to\Lawv^Y$ 
- with \emph{finite} supports $\C F_b=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$, and which have thus shape $f(x)_b=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, are generalisation of usual tropical polynomials to the case of possibly infinitely many variables.
-
-%\end{align}
-%where $+$ is the multiset union.
-
-%Remember that in $\Lawv$ the neutral element for addition is $\infty$ and the neutral for multiplication is $0$, so for instance the evaluation map is the matrix $\RM{eval}^{X,Y}\in\Lawv^{!((X\multimap Y) \times X)\times Y}\simeq\Lawv^{(!!X\times !Y\times !X)\times Y}$ given by $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=0$ if $\rho_1=[\mu]$ and $\rho_2=[b]$, and $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=\infty$ otherwise.
-
 \begin{remark}\label{rmk:val_trop}
@@
 That is, it coincides to consider the same power series but with coefficients in $Q$ quotiented for the idempotency of $+$.
-
-\end{remark}
-
-\begin{remark}
- Let $X$, $Y$ be sets and let $\langle \_,\_\rangle:X\times Y \to \mathbb{R}$.
- For $f:X\to \mathbb R$, define $f^*:Y\to \mathbb R$ by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
- Then for $X=!A$, $Y=\Lawv^A$, where $A$ is a set, and $\langle \mu, y \rangle:= \mu y$, we have that $f^!(y)=(-f)^*(-y)$ for all $f\in\Lawv^{!A}$.
- This is precisely the same formal construction yielding the well-known convex conjugate $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
- This construction is in turn a generalisation of the Legendre transformation.
- Despite the formal constructions being the same, we ignore for the moment if these could be connected to the study of high-order programs in our setting.
 \end{remark}

### 2023-07-28 CSL24/section2.tex
Commit serale

@@
 
-The language $\BSTLC$ we consider is a simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}. 
-The terms are as for the $\STLC$, the types are $A::= * \ \mid  \ !_{n}A \multimap A$, the contexts of the typing judgements are sets of declarations of the form $x :_{n}A$, with $n\in \mathbb N$, and the typing rules are: %given in Fig.~\ref{fig:rules}
+One can see that the comonad $!$ of $\LREL$ can be ``decomposed'' {\color{red}REF} into a family of ``graded exponentials functors'' $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined on objects by taking multisets of cardinality \emph{at most} $n$.
+%  lift to functors
+The sequence $(!_n)_{n\in\N}$ gives rise to a so-called \emph{$\N$-graded linear exponential comonad} on (the SMC) $\LREL$.
+%satisfying the adjunction: $\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
+
+As such, $(\LREL,(!_n)_{n\in\N})$ is a model of graded calculi ensuring bounded duplications via graded typing systems, for instance it is a model  \cite{Katsumata2018} of the language $\BSTLC$, a simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}, defined as follows:
+the terms are as for the $\STLC$, the types are $A::= * \ \mid  \ !_{n}A \multimap A$, the contexts of the typing judgements are sets of declarations of the form $x :_{n}A$, with $n\in \mathbb N$, and the typing rules are: %given in Fig.~\ref{fig:rules}
 	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
@@
 %Similarly to It is known {\color{red}(([reference??] e dire meglio)}
-The $\BSTLC$ can be interpreted in a symmetric monoidal closed category (SMCC) equipped with a \emph{$\mathbb N$-graded linear exponential comonad} \cite{Katsumata2018}.
-Now, one can see that the exponential $!$ can be ``decomposed'' into a family of \emph{graded} exponentials functors $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined on objects by taking multisets of cardinality \emph{at most} $n$.
-%  lift to functors
-The sequence $(!_n)_{n\in\N}$ gives rise to a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, %satisfying the adjunction: $\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
-i.e.\ $(\LREL,(!_n)_{n\in\N})$ is a model of $\BSTLC$. 
-In particular, bounded arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$ and so, 
+
+Remark that since arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$, 
 if $\model *$ is finite, then $\model{A}$ is a finite set for any type $A$ of $\BSTLC$.
@@
 
-The syntax of the differential $\lambda$-calculus $\STDLC$ (e.g.\ \cite[Section 3]{Manzo2010}), is given by the \emph{terms} $M$ and the \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
+A Cartesian closed differential $\lambda$-category (CC$\partial\lambda$C)\cite{Manzo2010,Blute2009, Blute2019} is a CCC enriched over commutative monoids %(i.e.\ we can add morphisms and there is a $0$ morphism)
+%, it is Cartesian closed%(with the closed structure compatible with the additive structure)
+and equipped with a certain \emph{differential operator} $D$, generalising the usual notion of differential, see e.g.\ \cite{BluteEhrhTass10}.
+%An example is the CC$\partial\lambda$C of convenient vector spaces with smooth maps, where $D$ is the ``real'' differential $Df:\mathbb{V}\times\mathbb{V}\rightarrow \mathbb{W}$, $Df(x,u):=\dfrac{d}{dt}{\!\Big|_{t=0}} f(x+tu)$, of smooth maps $f:\mathbb{V}\rightarrow \mathbb{W}$.
+%More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
+%\begin{itemize}
+%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
+%\item $\C C$ is equipped with a differential operator $D:
+%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
+%\end{itemize}
+Now, applying \cite[Theorem 6.1]{lemay2020} one can check that
+%\begin{proposition}[{\color{red}LEMAY??}]\label{thm:LREL!CCDC}
+ $\LREL_!$ becomes a CC$\partial\lambda$C when equipped with the \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
+%\end{proposition}
+%\begin{remark}For $t\in\HOM{\LREL}{!X}{Y}$, we have: $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu+\nu'+\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho+\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.\end{remark}
+%This ensures that one can define a sound interpretation of $\STDLC$-terms in the standard way (see [Section 4.3, \cite{Manzo2010}]).
+
+As such, $(\LREL_!,D)$ is a model of the differential $\lambda$-calculus $\STDLC$ \cite{ER}, a language ensuring exact control of duplications via a notion of linear substitution.
+Its syntax (e.g.\ \cite[Section 3]{Manzo2010}) is given by the \emph{terms} $M$ and the \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
 quotiented by $\alpha$-equivalence, by equations that make $+,0$ form a commutative monoid on the set of sums, %, i.e.\ commutativity and associativity of $+$ and neutrality of $0$ w.r.t.\ $+$;
@@
 %The models of $\STDLC$ are the cartesian \emph{closed} differential categories ($CC\partial C$), which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.
-The $\STDLC$ can be interpreted in Cartesian closed differential $\lambda$-categories (CC$\partial\lambda$C)\cite{Manzo2010,Blute2009, Blute2019}.
-In them, homsets are equipped with a structure of commutative monoid %(i.e.\ we can add morphisms and there is a $0$ morphism)
-%, it is Cartesian closed%(with the closed structure compatible with the additive structure)
-and with a certain \emph{differential operator} $D$, generalising the usual notion of differential, see e.g.\ \cite{BluteEhrhTass10}.
-%An example is the CC$\partial\lambda$C of convenient vector spaces with smooth maps, where $D$ is the ``real'' differential $Df:\mathbb{V}\times\mathbb{V}\rightarrow \mathbb{W}$, $Df(x,u):=\dfrac{d}{dt}{\!\Big|_{t=0}} f(x+tu)$, of smooth maps $f:\mathbb{V}\rightarrow \mathbb{W}$.
-%More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
-%\begin{itemize}
-%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
-%\item $\C C$ is equipped with a differential operator $D:
-%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
-%\end{itemize}
-Now, applying \cite[Theorem 6.1]{lemay2020} one can check that
-%\begin{proposition}[{\color{red}LEMAY??}]\label{thm:LREL!CCDC}
- $\LREL_!$ becomes a CC$\partial\lambda$C, i.e.\ a model of the $\STDLC$, when equipped with the \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
-%\end{proposition}
-%\begin{remark}For $t\in\HOM{\LREL}{!X}{Y}$, we have: $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu+\nu'+\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho+\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.\end{remark}
-%This ensures that one can define a sound interpretation of $\STDLC$-terms in the standard way (see [Section 4.3, \cite{Manzo2010}]).
-
-As already mentioned, by induction on $M$ %(using that composition and $D$ easily preserve ``booleaness'' and projections and evaluation of  $\LREL$ are boolean)
-one can prove that:
-
-\begin{proposition}\label{prop:descrete}
- For $\BSTLC$ and $\STDLC$ (but not $\STLC_\oplus$), we have that $\model{\Gamma\vdash M:A}\in\Lawv^{!\model{\Gamma}\times\model{A}}$ is a \emph{boolean} matrix, i.e.\ actually $\model{\Gamma\vdash M:A}\in\{0,\infty\}^{!\model{\Gamma}\times\model{A}}$.
-\end{proposition}
-
-Even if the interpretation of those languages is trivia, we shall still focus on them in order to set the basis of the further studies and in order to see some already interesting properties of their interpretation.{\color{red}SPIEGARE MEGLIO}

### 2023-07-28 CSL24/section2bis.tex
Commit serale

@@
 
-A \emph{tropical polynomial} is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$, where $i_{j}\in\N$ and $c_{i_{j}}\in\Lawv$.
+\begin{definition}
+A \emph{tropical power series} (tPs) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$ and $I\subseteq \N$.
+If $I$ is finite, $f$ is said to be a \emph{tropical polynomial}. 
+\end{definition}
+
+A tropical polynomial is always a piece-wise linear function since it has shape $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$
 For example, the polynomials $\varphi_{n}(x)=\min_{0\leq j\leq n}\{jx+2^{-j}\}$
@@
 
-A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable (i.e.~where the slope of $\varphi$ changes), or equivalently, where the minimum defining $\varphi$ is attained at least twice (see \autoref{fig:plot1}).
+A \emph{tropical root} of a tropical power series $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable (i.e.~where the slope of $\varphi$ changes), or equivalently, where the minimum defining $\varphi$ is attained at least twice (see \autoref{fig:plot1}).
 %For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
-Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
-$f(x)=k\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
+%Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as $f(x)=k\min\{x,x_{0}\}+ g(x)$.
+Unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
 %Tropical polynomials and their roots are the main object of study of tropical geometry.
@@
 
-A \emph{tropical Laurent series} of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\{nx+\matr f_{n}\}$, with $\matr  f_{n}\in\Lawv$.
-In other words, it is a ``limit'' of tropical polynomials of higher and higher degree.
+Since $\inf$s are not in general $\min$s, a tPs is a ``limit'' of tropical polynomials of higher and higher degree, and its behaviour may be less predictable than that of tropical polynomials~\cite{Porzio2021}.
+%For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points of their domain.
 E.g., the tLs $\varphi(x):=\inf_{n\in\N}\set{nx+2^{-n}}$ (see Fig.~\ref{fig:plot1}) %that we will take as our running example, 
 is the ``limit'' of the polynomials $\varphi_{n}$.
-Since $\inf$s are not in general $\min$s, the behaviour of tLS may be less predictable than that of tropical polynomials~\cite{Porzio2021}.
-%For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points of their domain.
 
@@
 
-Now, for a (formal) power series $f=\sum_{i\in I\subseteq\N} a_i x^i$ (polynomials being the case for $I$ finite) with coefficitents $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$ (notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$), one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
+Now, for a (formal) power series $f=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) with coefficitents $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$ (notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$), one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
 

### 2023-07-28 CSL24/section3A.tex
Commit serale

@@
-The linear simply typed $\lam$-calculus is a restriction of the ordinary $\lam$-calculus in which programs, following the linear logic mantra, can only use their arguments \emph{exactly once}.
-%(and indeed this calculus can be embedded in its \emph{intuitionistic multiplicative} fragment \emph{IMLL}).
-%More precisely, the linear $\lam$-calculus is obtained from the ordinary one by adding the constraint that each $\lam$-abstracted variable appears exactly once in the scope of the $\lam$-abstraction.
-%In order to frame it in a category $\C C$, one needs a symmetric tensor product $\otimes$ together with internal hom-set objects $X\multimap Y$ s.t.\ the evaluation and curry maps yield a \emph{symmetric monoidal adjunction}: $\C C(Z\otimes X, Y) \simeq \C C (Z, X\multimap Y)$.
-It can be interpreted in any Symmetric Monoidal Closed Category (SMCC).
-Now, $\LREL$ is a SMCC (\cite[Section III.A]{Manzo2013}), %(actually, even a model of IMLL), 
+$\LREL$ is a Symmetric Monoidal Closed Category (SMCC) \cite[Section III.A]{Manzo2013}, %(actually, even a model of IMLL), 
 whose monoidal product $\otimes$ acts on objects as the Cartesian product of sets, %and as the \emph{Kronecker product} of matrices on morphisms.
@@
  and its internal hom $X\multimap Y$ is also given by the Cartesian product $X\times Y$ of sets, with the usual evaluation and curry maps.
+As such, it is able to interpret the \emph{linear simply typed $\lam$-calculus}, the restriction of the ordinary $\lam$-calculus in which programs ,% following the linear logic mantra, 
+can only use their arguments \emph{exactly once}. %(and indeed this calculus can be embedded in its \emph{intuitionistic multiplicative} fragment \emph{IMLL}).
+%More precisely, the linear $\lam$-calculus is obtained from the ordinary one by adding the constraint that each $\lam$-abstracted variable appears exactly once in the scope of the $\lam$-abstraction.
+%In order to frame it in a category $\C C$, one needs a symmetric tensor product $\otimes$ together with internal hom-set objects $X\multimap Y$ s.t.\ the evaluation and curry maps yield a \emph{symmetric monoidal adjunction}: $\C C(Z\otimes X, Y) \simeq \C C (Z, X\multimap Y)$.

### 2023-07-28 CSL24/section3B.tex
Commit serale

@@
 
-\begin{remark}
-One can see that the matrix interpreting a $\STLC$-term is always boolean (i.e.\ its coefficients are either $0$ or $\infty$) can be seen as the fact that $\STLC$ is relatively trivial, from this point of view.
-This is also the case for the differential $\lambda$-calculus that we will introduce in a moment, so we state and discuss this property at that point (\autoref{prop:descrete}).
+By induction on $M$ %(using that composition and $D$ easily preserve ``booleaness'' and projections and evaluation of  $\LREL$ are boolean)
+one can prove that:
+
+\begin{proposition}\label{prop:descrete}
+ For $\BSTLC$ and $\STDLC$ (and thus also for its fragment $\STLC$), we have that $\model{\Gamma\vdash M:A}\in\Lawv^{!\model{\Gamma}\times\model{A}}$ is a \emph{boolean} matrix, i.e.\ actually $\model{\Gamma\vdash M:A}\in\{0,\infty\}^{!\model{\Gamma}\times\model{A}}$.
+\end{proposition}
+
+This can be seen as the fact that such calculi are relatively trivial, from this point of view.
+Even if the interpretation of those languages is trivia, we shall still focus on them in order to set the basis of the further studies and in order to see some already interesting properties of their interpretation.{\color{red}SPIEGARE MEGLIO}
 %It is instructive what its CCC-structure looks like in our tropical world.
-\end{remark}
 

### 2023-07-28 CSL24/section3C.tex
Commit serale

@@
-%\begin{remark}\label{rmk:ModelsOfBSTLC}
-%Similarly to It is known {\color{red}(([reference??] e dire meglio)} 
-For $\BSTLC$, one can see that the exponential $!$ can be ``decomposed'' into a family of \emph{graded} exponentials functors $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined on objects by taking multisets of cardinality \emph{at most} $n$.
-%  lift to functors
-The sequence $(!_n)_{n\in\N}$ gives rise to a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, %satisfying the adjunction: $\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
-i.e.\ $(\LREL,(!_n)_{n\in\N})$ is a model of $\BSTLC$. 
-In particular, bounded arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$ and so, 
-if $\model *$ is finite, then $\model{A}$ is a finite set for any type $A$ of $\BSTLC$.
+For $Q$ a continuous semiring, one lets the category $\QREL$ (\cite{Manzo2013} calls it $Q^\Pi$) have sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms, i.e.~$\QREL(X,Y):=Q^{X\times Y}$.
+%The identity morphism of $\QREL$ is the identity matrix $I\in Q^{X\times X}$,
+% given by $i_{a,a}=1$ and $i_{a,b\neq a}=0$, and
+The composition $st\in Q^{X\times Z}$ of $t\in Q^{X\times Y}$ and $s\in Q^{Y\times Z}$ is given by $(st)_{a,c}:=\sum_{b\in Y} s_{b,c}t_{a,b}$ (such series converges because $Q$ is continuous).
+As expected, $Q^X$ is a $Q$-semimodule and we identify $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$, which have shape $f(x)_b=:\sum_{a\in X} \matr f_{a,b}x_a$, for some matrix $\matr f\in Q^{X\times Y}$.
+% \begin{remark}
+ %Following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we chose to see a matrix $t$ from $X$ to $Y$ as a map $t:X\times Y\to Q$.
+% 
+% fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum_{b\in Y} s_{b,c}t_{a,b}$.
+Notice that usual linear algebra conventions correspond to work in $\QREL^{\op}$, %a matrix $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns
+e.g.\ the usual matrix-vector product defines a map $Q^Y\to Q^X$.
+Following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we are instead working with transpose matrices.
 %\end{remark}
-%Bounded types are interesting because of the following proposition:
-%\begin{proposition}
-%For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) correspond to polynomials.
-%\end{proposition}
-%\begin{proof}
-%It suffices to check that $\model A$ is finite for all bounded types $A$.
-Indeed this implies that a morphism $t:\model A\to \model B$ is a finite matrix $t: \model A \times \model B \to \Lawv$.Hence, its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is a polynomial.
-%\end{proof}
+%
+%In $\QREL^{op}$ (which corresponds to systematically taking transpose matrices), composition coincides with the product matrix/matrix and $\hat{(\cdot)}$ with the product matrix/vector.
+%In order to avoid confusion, we will refer to a $t\in Q^{X\times Y}$ just as a \emph{matrix from $X$ to $Y$}.
+%
+%
+%
+%We must fix a convention for matrices: following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum_{b\in Y} s_{b,c}t_{a,b}$.
+%In linear algebra, a map $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns.
+%In particular, the product of such a matrix for a vector defines a map $Q^Y\to Q^X$.
+%Instead, we prefer to see a $t\in\HOM{\QREL}{X}{Y}$ as giving rise to a map $\hat t:Q^X\to Q^Y$ defined by $\hat t(x)_a:=\sum_{b\in Y} t_{a,b}x_a$.
+%In $\QREL^{op}$ (which corresponds to systematically taking transpose matrices), composition coincides with the product matrix/matrix and $\hat{(\cdot)}$ with the product matrix/vector.
+%In order to avoid confusion, we will refer to a $t\in Q^{X\times Y}$ just as a \emph{matrix from $X$ to $Y$}.
+%As it is expected, $Q^X$ is a $Q$-semimodule and the bijection $\hat{(\cdot)}$ identifies $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.
+It is well known that $\QREL$ admits a comonad $!$ which acts, on objects, by taking the finite multisets.
+Remember that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
+%We will explicit this constructions in our tropical setting in the next lines.
 
-%For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
-%\begin{itemize}
-%\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
-%\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
-%\end{itemize}
+The main actor of this paper is the category $\LREL$. % of sets and matrices with values over $\Lawv$ (which is a continuous semi-ring).
+%, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
+%The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring (w.r.t.\ its quantale order $\preceq$.
+%This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
+It is worth observing that the composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
+reads as \ $(st)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
+similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$, %induced by matrices
+which we call \emph{tropical linear}, are exactly those of shape $f(x)_b=\inf_{a\in X} \set{\matr f_{a,b}+x_a}$, for some $\matr f\in\Lawv^{X\times Y}$.
+%\end{remark}
 
-%Until now we simply specialised well-known results in our tropical case, with the intent of showing how things read in this particular case.
-%Now we go further, by showing that $\LREL_!$ actually admits a \emph{differential structure}, turning it into a model of the $\STDLC$, i.e.~a $CC\partial C$.
-%This viewpoint
-%, is where the \emph{metric} and the \emph{differential} viewpoints converge, as explained in the Introduction and Section II, and it % will be further generalised in Section \ref{section6}.
+Now, although a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a tropical linear map $\Lawv^{!X}\to\Lawv^Y$, by exploiting the coKleisli structure we can also ``express it in the base $X$'', i.e.\ see it as a \emph{non-linear} map, that we call $t^!$.
+Tropical Laurent series precisely arise from such general construction:
 
-%A model of the $\STDLC$ is usually understood as so-called \emph{Cartesian closed differential categories} (CC$\partial$C), see \cite{Manzo2012} for details.
-%In order to treat the $+$ and the constructor $D[\_]\cdot (\_)$ of $\STDLC$, the main features of a CC$\partial$C $\C C$ are that:
+\begin{definition}
+Let $t\in\HOM{\LREL_!}{X}{Y}$.
+We define $t^!:\Lawv^X\to\Lawv^Y$ by setting $t^!(x):=t\circ_! x$. %(we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
+Concretely, $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
+\end{definition}
+%\begin{remark}%[Tropical Laurent series]
+The functions $t^!$ correspond then to the generalisation of tLs with \emph{possibly infinitely} many variables (in fact, as many as the elements of $X$), %In the following we will also refer to them as tLs.
+% We will call them simply \emph{tropical Laurent series (tLs)}.
+% %Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
+%\end{remark}
 %
-%1) $\C C$ is a left-additive-CCC, i.e.\ its Homsets are commutative monoids and its Cartesian closed structure is well behaved w.r.t.\ this monoid structure;
-%
-%2) $\C C$ is equipped with a differential operator map $D:\HOM{\C C}{X}{Y}\to \HOM{\C C}{X\times X}{Y}$ (here $\times$ is the Cartesian product of $\C C$) satisfying $8$ axioms, called D1, ..., D7, D-curry.
+%We find the usual notion of tLs of one variable as follows:
+%\begin{remark}
+and by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the %functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ we recover the
+usual tLs's of one variable.
+% \end{remark}
+In a similar way, tLs
+$f:\Lawv^X\to\Lawv^Y$ 
+ with \emph{finite} supports $\C F_b=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$, and which have thus shape $f(x)_b=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, are generalisation of usual tropical polynomials to the case of possibly infinitely many variables.
 
-%Let us show the differential structure of $\LREL_!$ (remember that the Cartesian product of $\LREL_!$ is the disjoint union $+$).
+%\end{align}
+%where $+$ is the multiset union.
 
-%\begin{definition} The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).\end{definition}
+%Remember that in $\Lawv$ the neutral element for addition is $\infty$ and the neutral for multiplication is $0$, so for instance the evaluation map is the matrix $\RM{eval}^{X,Y}\in\Lawv^{!((X\multimap Y) \times X)\times Y}\simeq\Lawv^{(!!X\times !Y\times !X)\times Y}$ given by $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=0$ if $\rho_1=[\mu]$ and $\rho_2=[b]$, and $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=\infty$ otherwise.
 
-For what concerns $\STDLC$, applying \cite[Theorem 6.1]{lemay2020} one can check that
-%\begin{proposition}[{\color{red}LEMAY??}]\label{thm:LREL!CCDC}
- $\LREL_!$ becomes a CC$\partial\lambda$C, i.e.\ a model of the $\STDLC$, when equipped with the \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
-%\end{proposition}
-%\begin{remark}For $t\in\HOM{\LREL}{!X}{Y}$, we have: $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu+\nu'+\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho+\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.\end{remark}
-%This ensures that one can define a sound interpretation of $\STDLC$-terms in the standard way (see [Section 4.3, \cite{Manzo2010}]).
-This model is also well-behaved w.r.t.\ the Taylor expansion, as expressed by the following two results.
-As it can be patientely checked, in $(\LREL_!,D)$ all morphisms can be Taylor expanded  (see \cite[Definition 4.22]{Manzo2012}):
-
-\begin{theorem}\label{thm:modelsTaylor}
- For all $t\in\HOM{\LREL_!}{Z}{X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$ we have:%, the evaluation of $t$ over $s$ yields 
- \begin{align}\label{eq:taylorcat}
-  \RM{ev}\circ_!\langle t,s\rangle =
-  \inf\limits_{n\in\N}
-  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
- \end{align} 
-\end{theorem}
-%It is worth discussing the formula above a bit more. 
-Here,
-%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
-%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
-$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
-Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$,  i.e.~it coincides with the tropical %interpretation of the
-version of the usual Taylor expansion.
-%Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
-Finally, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of \cite[Theorem 4.23]{Manzo2012} shows:
-
-\begin{corollary}\label{cor:T(M)=M}
- If $\Gamma\vdash_{\STLC} M:A$, then $\model{\Gamma\vdash_{\STDLC} \Te M:A}:=\inf_{t\in\Te{M}} \model{\Gamma\vdash_{\STDLC} t:A}$ converges to $\model{\Gamma\vdash_{\STLC} M:A}$.
-%the interpretation of the Taylor expansion of a $\STLC$-term $M$, given in \eqref{eq:taylor}, converges to the one of $M$.
-\end{corollary}
-
-By straightforward induction on $M$ (using that composition and $D$ easily preserve ``booleaness'' and projections and evaluation of  $\LREL$ are boolean) one can prove that:
-
-\begin{proposition}\label{prop:descrete}
- For all the previous four languages, $\model{\Gamma\vdash M:A}\in\Lawv^{!\model{\Gamma}\times\model{A}}$ is a \emph{boolean} matrix, i.e.\ actually $\model{\Gamma\vdash M:A}\in\{0,\infty\}^{!\model{\Gamma}\times\model{A}}$.
-\end{proposition}
-
-The fact that the coefficients are either $0$ or $\infty$ can be seen as the fact that $\STLC$ is relatively trivial, from this point of view.
-As we will discuss in \autoref{sec:app}, non-trivial coefficients appear when interpreting more interesting languages, like probabilistic ones.
-However, in this paper we shall still focus on $\STLC$, in order to set the basis of the further studies and in order to see some interesting properties already of its interpretation.
+\begin{remark}
+ Let $X$, $Y$ be sets and let $\langle \_,\_\rangle:X\times Y \to \mathbb{R}$.
+ For $f:X\to \mathbb R$, define $f^*:Y\to \mathbb R$ by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
+ Then for $X=!A$, $Y=\Lawv^A$, where $A$ is a set, and $\langle \mu, y \rangle:= \mu y$, we have that $f^!(y)=(-f)^*(-y)$ for all $f\in\Lawv^{!A}$.
+ This is precisely the same formal construction yielding the well-known convex conjugate $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
+ This construction is in turn a generalisation of the Legendre transformation.
+ Despite the formal constructions being the same, we ignore for the moment if these could be connected to the study of high-order programs in our setting.
+\end{remark}

### 2023-07-28 CSL24/section4.tex
Commit serale

@@
 \begin{enumerate}
-\item $\model{\Gamma \vdash_{\BSTLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{tropical polynomial}, thus (as $\model A$ is finite), a \emph{Lipschitz} function.
-\item $\model{\Gamma \vdash_{\STLC} M:A}^!, \model{\Gamma \vdash_{\STLC_\oplus} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ are \emph{locally} Lipschitz maps.
+\item $\model{\Gamma \vdash_{\BSTLC} M:B}^!:\prod\limits_{(x_i:_{n_i} A_i)\in\Gamma} \!\!\!\!\!\Lawv^{\model{A_i}} \to \Lawv^{\model B}$ is a \emph{tropical polynomial}, thus (as $\model A$ is finite), a \emph{Lipschitz} function.
+\item $\model{\Gamma \vdash_{\STLC} M:B}^!, \model{\Gamma \vdash_{\STLC_\oplus} M:A}^!:\prod\limits_{(x_i: A_i)\in\Gamma} \!\!\!\!\!\Lawv^{\model{A_i}} \to \Lawv^{\model B}$ are \emph{locally} Lipschitz maps.
 \item $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!$ of \emph{tropical polynomials}, thus (as $\model A$ is finite), \emph{Lipschitz} functions.

### 2023-07-29 CSL24/main.tex
Commit serale
Commit sonnolento
La parte sulla probability ancora mi manca qualche dettaglio da chiarire, ma ci siamo quasi...

--- 2023-07-27 CSL24/main.tex
@@
 
-\section{Tropical weighted relational semantics for high-order languagesi}\label{section3}
+\section{Topical linear/non-linear algebra}\label{section3}
+\input{section3C}
+
+\section{Tropical weighted relational semantics for high-order languages}
 
@@
 
-\paragraph*{Prohibited duplication/erasure: the \emph{linear} $\STLC$}\label{sec:3A}
 \input{section3A}
 
-\paragraph*{Allowed duplication/erasure: the $\STLC$ and the probabilistic $\lambda$-calculus $\STLC_\oplus$}\label{sec:STLC+STLCO}
+\input{section2}
+
+\paragraph*{Allowed duplication/erasure: the probabilistic $\lambda$-calculus $\STLC_\oplus$}\label{sec:STLCO}
 
 \input{section3B}
-
-\input{section2}
 

### 2023-07-29 CSL24/section2bis.tex
Commit sonnolento
La parte sulla probability ancora mi manca qualche dettaglio da chiarire, ma ci siamo quasi...

@@
 \begin{definition}
-A \emph{tropical power series} (tPs) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$ and $I\subseteq \N$.
+A \emph{tropical power series}\footnote{In analogy with \cite{DanEhrh2011}, we could also call them \emph{tropical entire functions}.} (tPs) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
 If $I$ is finite, $f$ is said to be a \emph{tropical polynomial}. 

### 2023-07-29 CSL24/section3A.tex
Commit sonnolento
La parte sulla probability ancora mi manca qualche dettaglio da chiarire, ma ci siamo quasi...

@@
+\paragraph*{Prohibited duplication/erasure: the \emph{linear} $\STLC$}\label{sec:3A}
+
 $\LREL$ is a Symmetric Monoidal Closed Category (SMCC) \cite[Section III.A]{Manzo2013}, %(actually, even a model of IMLL), 
@@
 %In order to frame it in a category $\C C$, one needs a symmetric tensor product $\otimes$ together with internal hom-set objects $X\multimap Y$ s.t.\ the evaluation and curry maps yield a \emph{symmetric monoidal adjunction}: $\C C(Z\otimes X, Y) \simeq \C C (Z, X\multimap Y)$.
+
+%\subsubsection{Unbounded duplications}
+
+\paragraph*{Allowed duplication/erasure: the $\STLC$}\label{sec:STLC}
+
+\begin{comment}
+
+In order to interpret the full $\STLC$, we need a Cartesian closed category (CCC).
+It is well-known \cite{Mellies2009} that it is always possible to construct a CCC by taking the \emph{co-Kleisli} $\C C_!$ of a so-called \emph{Lafont category} $\C C$.
+%, a construction we now quickly recall.
+%defined via a \emph{Lafont exponential} comonad $!$.
+%Let us quickly recall the ideas behind these notions.
+A SMCC is Lafont when it has finite products and it is equipped with a comonad $!$ (its \emph{Lafont exponential}) which, at level of objects, sends $X$ to an object $!X$ being the free commutative comonoid on $X$.
+Such objects $!X$ represent the \emph{bang} connective of linear logic, granting infinite duplications via the infinite product $X^0\otimes X\otimes X^2\otimes X^3\otimes\cdots$, each factor representing a possible number of duplications.
+It is well-known that, under mild conditions satisfied by $\QREL$, one can explicit this idea via the fact that the map $X\mapsto \finMS{X}$ (where $\finMS{X}$ is the set of finite multi-sets on $X$) lifts to a functor $!:\QREL\to\QREL$ which is a Lafont-exponential comonad.
+Specializing [Corollary III.6, \cite{Manzo2013}] to our case, we have:
+
+\begin{proposition}
+ $\LREL$ is Lafont.
+\end{proposition}
+
+\end{comment}
+
+As well-known, if a SMCC is \emph{Lafont}, then one obtains a CCC from it by defining the exponential objects as $X\to Y:=!X \multimap Y$.
+With no surprise $\LREL$ is Lafont~\cite[Corollary III.6]{Manzo2013} %and specialising \cite[Theorem III.7]{Manzo2013} 
+w.r.t.\ the usual $!$ acting on objects by taking the set of finite multisets.
+So the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.
+The cartesian product $\&$ of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms $\RM{ev}$ are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_!
+t\in\Lawv^{!X\times Z}$, $(s\circ_!
+t)_{\mu,c}:=
+%\begin{align}
+\inf_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
+ \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.

### 2023-07-29 CSL24/section3B.tex
Commit sonnolento
La parte sulla probability ancora mi manca qualche dettaglio da chiarire, ma ci siamo quasi...

--- 2023-07-27 CSL24/section3B.tex
@@
-%\subsubsection{Unbounded duplications}
-
-\begin{comment}
-
-In order to interpret the full $\STLC$, we need a Cartesian closed category (CCC).
-It is well-known \cite{Mellies2009} that it is always possible to construct a CCC by taking the \emph{co-Kleisli} $\C C_!$ of a so-called \emph{Lafont category} $\C C$.
-%, a construction we now quickly recall.
-%defined via a \emph{Lafont exponential} comonad $!$.
-%Let us quickly recall the ideas behind these notions.
-A SMCC is Lafont when it has finite products and it is equipped with a comonad $!$ (its \emph{Lafont exponential}) which, at level of objects, sends $X$ to an object $!X$ being the free commutative comonoid on $X$.
-Such objects $!X$ represent the \emph{bang} connective of linear logic, granting infinite duplications via the infinite product $X^0\otimes X\otimes X^2\otimes X^3\otimes\cdots$, each factor representing a possible number of duplications.
-It is well-known that, under mild conditions satisfied by $\QREL$, one can explicit this idea via the fact that the map $X\mapsto \finMS{X}$ (where $\finMS{X}$ is the set of finite multi-sets on $X$) lifts to a functor $!:\QREL\to\QREL$ which is a Lafont-exponential comonad.
-Specializing [Corollary III.6, \cite{Manzo2013}] to our case, we have:
-
-\begin{proposition}
- $\LREL$ is Lafont.
-\end{proposition}
-
-\end{comment}
-
-As well-known, if a SMCC is \emph{Lafont}, then one obtains a CCC from it by defining the exponential objects as $X\to Y:=!X \multimap Y$.
-With no surprise $\LREL$ is Lafont (~\cite[Corollary III.6]{Manzo2013}) %and specialising \cite[Theorem III.7]{Manzo2013} 
-w.r.t.\ the usual $!$ acting on objects by taking the set of finite multisets.
-So the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.
-The cartesian product $\&$ of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms $\RM{ev}$ are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_!
-t\in\Lawv^{!X\times Z}$, $(s\circ_!
-t)_{\mu,c}:=
-%\begin{align}
-\inf_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
- \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
-
-\begin{remark}
-One can see that the matrix interpreting a $\STLC$-term is always boolean (i.e.\ its coefficients are either $0$ or $\infty$) can be seen as the fact that $\STLC$ is relatively trivial, from this point of view.
-This is also the case for the differential $\lambda$-calculus that we will introduce in a moment, so we state and discuss this property at that point (\autoref{prop:descrete}).
-%It is instructive what its CCC-structure looks like in our tropical world.
-\end{remark}
-
-%As it is well-known, the Cartesian closed structure %of a category $\C C$  allows to define a sound interpretation $\model{\Gamma\vdash M:A}\in\HOM{\LREL_{!}}{\model \Gamma}{\model A}$ of terms as morphisms.
-%In our case, we have:
-
 For what concerns the probabilistic calculus, let us consider a probabilistic extension of $\STLC$, call it $\STLC_\oplus$, which also extends the first-order language we used in~\autoref{sec:proba}:
@@
 \end{remark}
+
+A precise investigation of the relation between probabilistic programs and their $\LREL_!$ interpretation is left for further studies.

### 2023-07-29 CSL24/section4.tex
Commit sonnolento
La parte sulla probability ancora mi manca qualche dettaglio da chiarire, ma ci siamo quasi...

@@
 In this section we establish some topological and metric properties of such functions, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
-In analogy with \cite{DanEhrh2011}, we could call them \emph{tropical entire functions}, but since, as we explain in a moment, those are generalised \emph{tropical Laurent series}, we will use the latter terminology, or \emph{tLs} for short.
 The literature in tropical mathematics is often recent (e.g.~\cite{Porzio2021}), and several results we prove in this section are, to our knowledge, new.
@@
 Non riesco a capire se anche a noi lo \`e, ma non vedo perch\'e non dovrebbe.}
+Remark that, since $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}\in\HOM{\LREL_!}{\mathlarger{\&}_{i=1}^n \model{A_i}}{\model B}\simeq\Lawv^{\prod_{i=1}^n !
+\model {A_i}\times \model B}$, the tLs $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ is $t^!(x^1,\dots,x^n)_b=\inf_{\mu_i\in !
+\model{A_i}}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
+Same for $\STDLC$.
 
 \begin{remark}
- Since $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}\in\HOM{\LREL_!}{\mathlarger{\&}_{i=1}^n \model{A_i}}{\model B}\simeq\Lawv^{\prod_{i=1}^n !
-\model {A_i}\times \model B}$, the tLs $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ is defined by $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !
-\model{A_i}}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
-Same for $\STDLC$.
-Finally, the interpretations of $\BSTLC$-terms are matrices $t\in\HOM{\LREL}{\bigotimes_{i=1}^n !_{n_i}X_i}{Y}=\Lawv^{\prod_{i=1}^n !_{n_i}X_i\times Y}$.
+The interpretations of $\BSTLC$-terms are matrices $t\in\HOM{\LREL}{\bigotimes_{i=1}^n !_{n_i}X_i}{Y}=\Lawv^{\prod_{i=1}^n !_{n_i}X_i\times Y}$.
 In such situation, we define $t^!:\Lawv^{\mathlarger{+}_{i=1}^n X_{i}}\to\Lawv^Y \simeq \prod_{i=1}^n\Lawv^{X_{i}}\to\Lawv^Y$ as $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !_{n_i} X_i}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
@@
 {X_i}\times Y}$ is the matrix $\widetilde t_{(\mu_1,\dots,\mu_n),b}:=t_{(\mu_1,\dots,\mu_n),b}$ if $\mu_i\in !_{n_i} X_i$ for all $i$, and $:=\infty$ otherwise, which has the same support as $t$.
-So, such $t^!$'s are special cases of tLs and thus, \emph{Wlog}, in the following we only consider tLs.
+So, such $t^!$'s are special cases of tLs and thus we can \emph{Wlog} only consider tPs's.
 \end{remark}
@@
 
- \begin{remark}
- Our running example is indeed of shape $\varphi=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
- However, it is not the interpretation of a $\lam$-term, because its matrix $t$ is not discrete.
-% Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
-\end{remark}
-
-If $\model *$ is finite, then the interpretations of $\BSTLC$-terms is a (generalised) tropical polynomial, since then the interpretation of any $\BSTLC$-type $B$ is a finite set, and thus also $!_n \model B$ is.
 % 
@@
 %\end{remark}
+
+\subsection{Continuity of tLs}\label{subsec:cont}%$\Lawv^{X}$ as a normed cone}
 
@@
 
-\subsection{Continuity of tLs}\label{subsec:cont}%$\Lawv^{X}$ as a normed cone}
-
 The tLs $\varphi$ is continuous on $\BB R_{\geq0}$ (w.r.t.\ the usual norm of real numbers).
@@
 \begin{proof}
- Use the fact, taken from \cite{Selinger2004}, that in a normed $\R_{\geq 0}$-cone $P$, considered with its cone-order, if every bounded directed net in $P$ admits a sup, and if $(v_i)_{i\in I}$ is a directed net in $P$ with an upper bound $v\in P$, then $\exists\bigvee_{i\in I} v_i \in P$ and, if $\inf_{i\in I} \norm{v-v_i} =0$, one has $\bigvee_{i\in I} v_i = v$.
+ Uses \autoref{thm:cont} plus the fact, taken from \cite{Selinger2004}, that in a normed $\R_{\geq 0}$-cone $P$, considered with its cone-order, if every bounded directed net in $P$ admits a sup, and if $(v_i)_{i\in I}$ is a directed net in $P$ with an upper bound $v\in P$, then $\exists\bigvee_{i\in I} v_i \in P$ and, if $\inf_{i\in I} \norm{v-v_i} =0$, one has $\bigvee_{i\in I} v_i = v$.
 \end{proof}
@@
 
-The results presented in all this paper translate now into the following facts about the interpretation of higher-order programs.
+Let us now see what the results proved in the previous section translate into, when referred to the interpretation of higher-order programs.
 Remark that the metric spaces $(\Lawv^{\mathlarger{+}_{i=1}^n X_i},d_{\infty})$ and $(\prod_{i=1}^n \Lawv^{X_i},\max_{i=1}^n d^{X_i}_\infty)$ are trivially isometric, so we identify them.

### 2023-07-29 CSL24/section5.tex
Altro commit pre-cena
Commit sonnolento
La parte sulla probability ancora mi manca qualche dettaglio da chiarire, ma ci siamo quasi...

@@
 $\max_{\omega\in\set{lr,rrl,rlr}} \, P_\omega(p,q) = P_{\omega_0}(p,q)$, i.e.\ $\min_{\omega\in\set{lr,rrl,rlr}} \, -\log P_\omega(p,q) = -\log P_{\omega_0}(p,q)$.
-Remembering that $\trop^{\mathrm{val}} Q_{\false}(p,q)=\min\set{p+q, \mathrm{val}(2)+p+2q}$, we see that our minimization problem is equivalent to asking $(\trop^{\mathrm{val}_1} Q_{\false})( -\log p, -\log q) = (\trop^{\mathrm{val}_1} P_{\omega_0})( -\log x, -\log y)\label{eq:max}$,
+Remembering that $\trop^{\mathrm{val}} Q_{\false}(p,q)=\min\set{p+q, \mathrm{val}(2)+p+2q}$, we see that our minimization problem is equivalent to asking $(\trop^{\mathrm{val}_1} Q_{\false})( -\log p, -\log q) = (\trop^{\mathrm{val}_1} P_{\omega_0})( -\log p, -\log q)\label{eq:max}$,
 and this time this holds, in general, only for the trivial valuation.
@@
 
-From this situation we notice the following important:
+Our toy first-order language can be interpreted in $\LREL$.
+We do not give details now since in the rest of the paper we will interpret interesting \emph{high-order} calculi, including a probabilistic one containing our toy language.
+We notice the following important:
 
 \begin{remark}\label{rmk:tropof01Rel}
-Our toy first-order language can be interpreted both in the already mentioned $\overline{R_{\geq 0}}\mathrm{Rel}$ and in $\LREL$.
-We do not give details now since in the rest of the paper we will interpret interesting \emph{high-order} calculi, including a probabilistic one containing our toy language.
+$\model{M}^{\LREL}_0(-\log p,-\log (1-p))$ gives the minimum negative log-probability of all the reductions from $M$ to normal form, i.e.\ it coincides with the  negative log-probability of any of the (equiprobable) \emph{most} likely reductions from $M$ to normal form.
+Equivalently, given $\omega_0$ a fixed reduction path from $M$ to normal form $P_{\omega_0}$, for all solution $p\in[0,1]$ to the equation $%\trop P_\omega(-\log p, -\log(1-p)) 
+-\log  P_{\omega_0} = \model{M}^{\LREL}_0(-\log p,-\log (1-p))$, the path $\omega_0$ is one of the \emph{most} likely one among all paths from $M$ to normal form.
+Similarly, $\model{M}^{\LREL}_0(\log p,\log (1-p))$ gives the minimum log-probability of all the reductions from $M$ to normal form, i.e.\ it coincides with the log-probability of any of the (equiprobable) \emph{less} likely reductions from $M$ to normal form.
+Equivalently, given $\omega_0$ a fixed reduction path from $M$ to normal form $P_{\omega_0}$, for all solution $p\in[0,1]$ to the equation $%\trop P_\omega(\log p, \log(1-p)) 
+\log  P_{\omega_0} = \model{M}^{\LREL}_0(\log p,\log (1-p))$, the path $\omega_0$ is one of the \emph{less} likely one among all paths from $M$ to normal form.
+% (in $p$ and $q$, not after the substitution $q:=r-p$) 
+% are extracted from it, \ref{eq:argmax}, \ref{eq:max} of $\STLC_\oplus$-programs. 
+\end{remark}
+
+\begin{remark}
+Our toy first-order language can be also interpreted both in the already mentioned $\overline{R_{\geq 0}}\mathrm{Rel}$.
 But it is important to mention already at this point that the probabilities we just discussed above, are already captured by the $[0,\infty]\mathrm{Rel}$ model: the $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}\in[0,\infty]^{\set{0,1}}$ of our running example $M$ is $\model M_0=Q_{\false}(p,1-p)$, $\model M_1=Q_{\true}(p,1-p)$.
@@
 A precise study of the ``tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$'' is left for future investigations, as it is related with power series arising from more sophisticated calculi with paramethers (in the style of \cite{} {\color{red}Ehrhard!}).
-For now, just have in mind that $\model{M}^{\LREL}_0(-\log p,-\log (1-p))$ gives the negative log-probability of \emph{any} of the (equiprobable) \emph{most likely} reduction paths to normal form, so $\LREL$ directly expresses such optimisations problems (no need to pass through a tropicalisation).
-We take these observations as incouraging, and the point of this work is to dig into such model, concentrating for this first paper on the relations between the metric and differential properties that coexist in it.
-% (in $p$ and $q$, not after the substitution $q:=r-p$) 
-% are extracted from it, \ref{eq:argmax}, \ref{eq:max} of $\STLC_\oplus$-programs. 
+However, it is important to keep in mind that those properties are expressed by $\LREL$ directly with no need to pass through a tropicalsation.
 \end{remark}

### 2023-07-29 CSL24/section5.tex
Commit ancora più sonnolento
Chiarito

@@
 \begin{remark}\label{rmk:tropof01Rel}
-$\model{M}^{\LREL}_0(-\log p,-\log (1-p))$ gives the minimum negative log-probability of all the reductions from $M$ to normal form, i.e.\ it coincides with the  negative log-probability of any of the (equiprobable) \emph{most} likely reductions from $M$ to normal form.
-Equivalently, given $\omega_0$ a fixed reduction path from $M$ to normal form $P_{\omega_0}$, for all solution $p\in[0,1]$ to the equation $%\trop P_\omega(-\log p, -\log(1-p)) 
--\log  P_{\omega_0} = \model{M}^{\LREL}_0(-\log p,-\log (1-p))$, the path $\omega_0$ is one of the \emph{most} likely one among all paths from $M$ to normal form.
-Similarly, $\model{M}^{\LREL}_0(\log p,\log (1-p))$ gives the minimum log-probability of all the reductions from $M$ to normal form, i.e.\ it coincides with the log-probability of any of the (equiprobable) \emph{less} likely reductions from $M$ to normal form.
-Equivalently, given $\omega_0$ a fixed reduction path from $M$ to normal form $P_{\omega_0}$, for all solution $p\in[0,1]$ to the equation $%\trop P_\omega(\log p, \log(1-p)) 
-\log  P_{\omega_0} = \model{M}^{\LREL}_0(\log p,\log (1-p))$, the path $\omega_0$ is one of the \emph{less} likely one among all paths from $M$ to normal form.
+$\model{M}^{\LREL}_0(-\log p,-\log (1-p))$ gives the minimum negative log-probability of all the reductions from $M$ to $\false$, i.e.\ it coincides with the  negative log-probability of any of the (equiprobable) \emph{most} likely reductions from $M$ to $\false$.
+Equivalently, given $\omega_0$ a fixed reduction path from $M$ to $\false_{\omega_0}$, for all solution $p\in[0,1]$ to the equation $%\trop P_\omega(-\log p, -\log(1-p)) 
+-\log  P_{\omega_0} = \model{M}^{\LREL}_0(-\log p,-\log (1-p))$, the path $\omega_0$ is one of the \emph{most} likely one among all paths from $M$ to $\false$.
+Similarly, $\model{M}^{\LREL}_0(\log p,\log (1-p))$ gives the minimum log-probability of all the reductions from $M$ to $\false$, i.e.\ it coincides with the log-probability of any of the (equiprobable) \emph{less} likely reductions from $M$ to $\false$.
+Equivalently, given $\omega_0$ a fixed reduction path from $M$ to $\false_{\omega_0}$, for all solution $p\in[0,1]$ to the equation $%\trop P_\omega(\log p, \log(1-p)) 
+\log  P_{\omega_0} = \model{M}^{\LREL}_0(\log p,\log (1-p))$, the path $\omega_0$ is one of the \emph{less} likely one among all paths from $M$ to $\false$.
 % (in $p$ and $q$, not after the substitution $q:=r-p$) 

### 2023-07-29 CSL24/main.tex
commit

@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
+\documentclass[a4paper,english, UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
 %This is a template for producing LIPIcs articles. 
@@
 
-\title{Tropical mathematics for the Lambda-calculus 1}
-\subtitle{The potential of tropical wheighted relational semantics%for probabilities and Taylor expansion
+\title{Tropical mathematics and the Lambda-calculus I}
+\subtitle{Metric and Differential Analysis of Effectful Programs
 } %TODO Please add
@@
 
-\section{Preliminaries and motivation}
+\section{A Bridge between Metric and Differential Aspects}\label{section5bis}
+\input{section5bis}
+
+\section{Tropical Power Series and Effectful Programs}\label{section2}
 \input{section2bis}
+\input{section2bisbis}
 
-\section{Topical linear/non-linear algebra}\label{section3}
-\input{section3C}
+\section{The Tropical Relational Model}\label{section3}
+\input{section3fancy}
 
-\section{Tropical weighted relational semantics for high-order languages}
-
-The category $\LREL$ satisfies many nice properties which make it able to model all the variants of the $\lambda$-calculi to which we are interested in here.
+%%
+%%\section{Topical linear/non-linear algebra}\label{section3}
+%%\input{section3C}
+%%
+%%\section{Tropical weighted relational semantics for high-order languages}
+%%
+%%The category $\LREL$ satisfies many nice properties which make it able to model all the variants of the $\lambda$-calculi to which we are interested in here.
 Although the majority of such properties are shared with the general class of models of $\lambda$-calculi called \emph{weighted relational models}, thoroughly studied in \cite{Manzo2013}, we mention them for our particular case in order to highlight how these look like in the tropical setting, but also to make the non-expert in such topics reader able to follow.
-The real interest in our particular model will come in the next \autoref{sec:tls}.
-
-\input{section3A}
-
-\input{section2}
-
-\paragraph*{Allowed duplication/erasure: the probabilistic $\lambda$-calculus $\STLC_\oplus$}\label{sec:STLCO}
-
-\input{section3B}
+%%The real interest in our particular model will come in the next \autoref{sec:tls}.
+%%
+%%
+%%\input{section3A}
+%%
+%%\input{section2}
+%
+%\paragraph*{Allowed duplication/erasure: the probabilistic $\lambda$-calculus $\STLC_\oplus$}\label{sec:STLCO}
+%
+%\input{section3B}
 
@@
 
-%\bibliography{tropical}
+\bibliography{tropical}
 

### 2023-07-29 CSL24/section1.tex
commit

--- 2023-07-23 CSL24/section1.tex
@@
 The main reason to look for methods relying on (quantitative extensions of) type-theory or denotational semantics is that these approaches yield \emph{modular} and \emph{compositional} techniques, that is, allow one to deduce properties of complex programs from the properties of their constituent parts.
-Among such quantitative approaches, two different directions have received considerable attention, which we recall below.
 
-%\subsection{Two kinds of quantitative approaches}\label{section2}
+\subparagraph*{Two kinds of quantitative approaches}
 
-\input{Section2}
+Among such quantitative approaches, two different directions have received considerable attention.
+
+On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}.
+This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
+
+On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
+This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
+
+In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
+In metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
+By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
+
+A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled.
+At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
+in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz.
+The fundamental motivation of this work is then the observation that
+this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
+
+%%
+%%In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
+%%
+%%\paragraph*{Two kind of quantitative approaches}
+%On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask whether they compute functions which do not differ \emph{too much} by controlling the propagation of an error during the computation.
+%This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
+%%As already mentioned, in many situations (e.g.~when dealing with computationally difficult problems, or in differential privacy \cite{Alvim2011, Reed2010}), one %does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound.
+In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, a small error in the input will produce a comparably small error in the output.
+%%In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior.
+%A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
+%$d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
+%This is due to the fact that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the %\emph{linear} simply typed $\lambda$-calculus, being a \emph{symmetric monoidal closed} category, and in fact it also models
+%\emph{affine} $\lambda$-calculus (it is symmetric monoidal closed, cartesian and monoidal units coincide). 
+%The metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications \cite{Reed2010, Gaboardi2017}: the idea is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz} map.
+%For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the function input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself.
+%: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.
+%By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $rX:=(X,%d_{r}(x,y)=
+%rd(x,y))$ and $Y$, the program $M$ above 
+%can thus be interpreted as a non-expansive map from $2(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.
+%These observations have led to the study of linear $\lambda$-calculi with \emph{graded} exponentials, like $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied to the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}.
+The types of such systems are defined by combining linear constructors with a \emph{graded linear exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
+
+%On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
+This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
+%%A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models.
+%The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus.
+The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Diff{M}{N}$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Diff{M}{N}$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).
+%Another intriguing similarity between program derivatives and 
+%actual derivatives is provided by the \emph{Taylor expansion} $\C T$:
+%%In analysis, a sufficiently regular function can be seen (locally around a point) as its Taylor expansion, i.e.\ a series of polynomials weighted via some rational coefficient.
+%%Now, a polynomial in $x$ of degree $k$ can be thought of as a function which uses its argument $x$ exactly $k$ times.
+%in $\STDLC$ (even untyped), we can perform a syntactic Taylor expansion of an ordinary $\lambda$-term via an inductively defined map $\Theta$ giving rise to an infinite series of terms, whose crucial case of the definition is
+%$
+%\Theta(MN)=\sum_{n\in\mathbb{N}}\frac{1}{n!}\Der^n\left[\Theta(M),\Theta(N)^n\right]0.
+%$
+%As in analysis, it decomposes an application as a series of $k$-linear functions, which can be seen as its approximants.
+%In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.
+%Since we only consider idempotent sum, the factorial coefficient disappears, and the resulting Taylor map is called the \emph{qualitative} Taylor expansion, indicated by $\Te{\_}$.
 
@@
 
-\subsection{Tropical mathematics and program semantics }
-
-In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
-In metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
-By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
-
-A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled.
-At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
-in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz.
-The fundamental motivation of this work is then the observation that
-this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
+\subparagraph*{Tropical mathematics and program semantics }
 
@@
 
-\subsection{This paper and our contributions}
+Coming back to our discussion on program semantics, tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
+At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.
+Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
+However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.
 
-Coming back to our discussion on program semantics %tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
-the metric approach usually yields Lipschitz conditions for affine programs, but what about ``unbounded'', simply typed $\lambda$-calculus (the ordinary one)?
-In such case, one might lose the Lipschitz property, so the programs might be \emph{too} sensitive to errors. 
-%For instance, while the functions $M_{k}=\lambda x.
+In this paper we demonstrate that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provides the desired bridge between differential and metric semantics, and we suggest how this approach might be used to develop a best-case analysis of probabilistic and non-deterministic programs (an aspect that will be developed in more detail in a future paper of this series).
+
+%Moreover, we show that the conceptual unification of these two approaches suggests ways in which techniques from resource-analysis could be used in sensitivity analysis and \emph{vice-versa}, paving the way for new  applications of tropical geometry to the  study of higher-order programs.
+
+%
+%Coming back to our discussion on program linearity, in metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
+%By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
+%
+%A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled.
+At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
+%in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz.
+The fundamental motivation of this work is then the observation that
+%this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
+%
+%Now, the metric approach usually yields Lipschitz conditions for affine programs, but what about ``unbounded'', simply typed $\lambda$-calculus (the ordinary one)?
+%In such case, one might lose the Lipschitz property, so the programs might be \emph{too} sensitive to errors. 
+%%For instance, while the functions $M_{k}=\lambda x.
 k\cdot x: \BB R\to \BB R$ are all Lipschitz-continuous, with Lipschitz constant $k$, the program $M=\lambda x.x^{2}$ obtained by ``duplicating'' $x$ is not Lipschitz anymore: $M$ is, so to say, \emph{too} sensitive to errors.
-It is indeed well-known that the category $\Met$ is \emph{not} Cartesian closed, so it is not a model of $\STLC$ (yet, several Cartesian closed \emph{sub-}categories of $\Met$ do exist, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
-%Still, one might observe that the program $M$ above is actually \emph{locally} Lipschitz-continuous (i.e.~Lipschitz over any open ball).
-However, one can still consider Cartesian closed categories of \emph{locally} Lipschitz maps  (i.e.~Lipschitz over any open ball), see e.g.\ \cite{Ehrhard2011, PistoneLICS}, and a new example will be considered in this paper.
-
-On the other hand, since the resource approach decomposes, through the Taylor formula, an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as a decomposing the associated locally Lipschitz map by a limit of the associated Lipschitz ones, thus bridging the metric and differential approaches.
-Here, a natural direction to look for is the \emph{relational semantics}, since strictly related with the Taylor expansion $\STDLC$ {\color{red}CITE DeCarvalho ecc}. 
-However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.
-
-Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions?
+%It is indeed well-known that the category $\Met$ is \emph{not} Cartesian closed, so it is not a model of $\STLC$ (yet, several Cartesian closed \emph{sub-}categories of $\Met$ do exist, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
+%%Still, one might observe that the program $M$ above is actually \emph{locally} Lipschitz-continuous (i.e.~Lipschitz over any open ball).
+%However, one can still consider Cartesian closed categories of \emph{locally} Lipschitz maps  (i.e.~Lipschitz over any open ball), see e.g.\ \cite{Ehrhard2011, PistoneLICS}, and a new example will be considered in this paper.
+%
+%On the other hand, since the resource approach decomposes, through the Taylor formula, an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as a decomposing the associated locally Lipschitz map by a limit of the associated Lipschitz ones, thus bridging the metric and differential approaches.
+%Here, a natural direction to look for is the \emph{relational semantics}, since strictly related with the Taylor expansion $\STDLC$ {\color{red}CITE DeCarvalho ecc}. 
+%However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.
+%
+%Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions?
 This way, the Taylor expansion could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps: unbounded term application could be seen as a limit of \emph{more and more sensitive} operations.
@@
 
-At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.
+%At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.
 Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
-However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.
+%However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.
+
+\subparagraph*{This paper and our contributions}
 
@@
 
-In this paper we demonstrate that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provides the desired bridge between differential and metric semantics.
+%In this paper we demonstrate that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provides the desired bridge between differential and metric semantics.
 Moreover, we show that the conceptual unification of these two approaches suggests ways in which techniques from resource-analysis could be used in sensitivity analysis and \emph{vice-versa}, paving the way for new  applications of tropical geometry to the  study of higher-order programs.
@@
 
-\item we study the relational model over the tropical semiring  and we show that the functions interpreting simply-typed lambda terms, which correspond to a generalization of \emph{tropical Laurent series} \cite{Porzio2021}, are locally Lipschitz-continuous, thus yielding a full-scale metric semantics for the $\lambda$-calculus and its bounded fragments.
+\item we show how tropical methods naturally arise in the \emph{best-case} analysis of probabilistic and non-deterministic programs, by studying a few very simple examples.
+This is in Section \ref{section2}.
+
+\item we study the relational model over the tropical semiring, which provides a semantics of effectful extensions of $\STLC$ and $\mathrm{PCF}$. Notably, 
+ we show that the functions interpreting higher-order programs, which correspond to a generalization of \emph{tropical Laurent series} \cite{Porzio2021}, are locally Lipschitz-continuous, thus yielding a full-scale metric semantics for the $\lambda$-calculus and its bounded fragments.
 This is in Sections \ref{section3} and \ref{section4}.
-%Moreover, we exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps.
 
-\item Using the relational model as our main source of inspiration,  we suggest a few potential applications of tropical methods to the study of quantitative properties of non-deterministic and probabilistic functional programs, like counting best-case computation steps,
-measuring convergence log-probabilities, and 
-differential privacy. This is in Section~\ref{section5}
+ \item We exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps, and we show that it can be used to deduce approximated Lipschitz-constants for higher-order programs.
+
+%\item Using the relational model as our main source of inspiration,  we suggest a few potential applications of tropical methods to the study of quantitative properties of non-deterministic and probabilistic functional programs, like counting best-case computation steps,
+%measuring convergence log-probabilities. This is in Section~\ref{section5}
 

### 2023-07-29 CSL24/section2bis.tex
commit

--- 2023-07-28 CSL24/section2bis.tex
@@
-\subsection{Elements of tropical mathematics}
-\input{section3new}
+%\subsection{Elements of tropical mathematics}
+%\input{section3new}
+
+In this section we recall tropical polynomials and power series and we discuss how these naturally arise in the study of \emph{optimal} behavior of probabilistic and non-deterministic programs.
+
+At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}.
+Let us recall that a quantale is a complete lattice endowed with a continuous monoid action.
+In the case of $\Lawv$ the lattice is defined by the reverse order $\geq$ on $\BB R$, and the monoid action is provided by addition.
+Notice that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$.
+
+%%A consequence of these observations is that, as we discussed below, the tropical approach to linear algebra coincides with the study of ``$\Lawv$-valued matrices'', i.e.~of maps of the form $s: X\times Y\to \Lawv$ .
+%%In particular, a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we explicitly take in Section \ref{section6}).
+%%
 
@@
 
+Power series and polynomials over the tropical semiring are defined as follows:
+
 \begin{definition}
-A \emph{tropical power series} (tPs) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$ and $I\subseteq \N$.
+A \emph{tropical power series}
+%\footnote{In analogy with \cite{DanEhrh2011}, we could also call them \emph{tropical entire functions}.} 
+(tPs) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
 If $I$ is finite, $f$ is said to be a \emph{tropical polynomial}. 
@@
 
-A \emph{tropical root} of a tropical power series $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable (i.e.~where the slope of $\varphi$ changes), or equivalently, where the minimum defining $\varphi$ is attained at least twice (see \autoref{fig:plot1}).
+A \emph{tropical root} of a tropical power series $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable (i.e.~where the slope of $\varphi$ changes).
+When $\varphi$ is a polynomial, the roots of $\varphi$ coincide with the points where the minimum defining $\varphi$ is attained at least twice (see \autoref{fig:plot1}).
 %For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
 %Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as $f(x)=k\min\{x,x_{0}\}+ g(x)$.
-Unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
+Unlike in standard algebra, tropical roots of tropical polynomials can be computed in linear time \cite{Noferini2015}. 
 %Tropical polynomials and their roots are the main object of study of tropical geometry.
@@
 
-Since $\inf$s are not in general $\min$s, a tPs is a ``limit'' of tropical polynomials of higher and higher degree, and its behaviour may be less predictable than that of tropical polynomials~\cite{Porzio2021}.
+While tropical polynomials are essentially combinatorial objects, this cannot be said for tPs: since $\inf$s are not in general $\min$s, a tPs is a ``limit'' of tropical polynomials of higher and higher degree, and its behavior is in general way more difficult to study than that of tropical polynomials~\cite{Porzio2021}.
 %For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points of their domain.
@@
 The valuation which is constantly equal to $1\in L$ on all the invertibles of $Q$ is called the \emph{trivial valuation} $\mathrm{val}_1$.
+Another common valuation is $\mathrm{logval}: \overline{\mathbb R}_{\geq 0} \to [-\infty, \infty]$ given by $\mathrm{logval}(x)=-\log x$.
+Notice that $\mathrm{logval}$ restricts to a bijection between $[0,1]$ and $\Lawv$.
 
-Now, for a (formal) power series $f=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) with coefficitents $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$ (notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$), one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
+Now, for a (formal) power series $f=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) with coefficients $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$ (notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$), one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
 
@@
 
-\subsection{Log-Probabilities and Tropicalisation through an exemple}\label{sec:proba}
-\input{section5}
+%\subsection{Log-Probabilities and Tropicalisation through an exemple}\label{sec:proba}
+%\input{section5}

### 2023-07-29 CSL24/section2bisbis.tex
commit

% !TEX root = /Users/paolopistone/Documents/GitHub/tropicalnew/CSL24/main.tex

Before discussing how full-scale higher-order programming languages can be interpreted in terms of tropical power series, we highlight how such functions may naturally arise in the study of effectful programming languages.
We will see that, when considering probabilistic and non-deterministic programs, tropical tools can be used to describe \emph{optimal} properties of programs, that is, the behavior of programs in \emph{the best case}. 
%Moreover, since, as we'll see, tropical semantics is also a \emph{metric} semantics, it can be used to study how much program behavior is \emph{sensitive} to errors, that is, how a small error in input may be increased in output.

\subparagraph*{Maximum Likelihood Estimators for Probabilistic Languages}

%
%Tropical methods have been largely applied as a means to solve optimization problems. Typically, to solve a problem of the form 
%$$
%\mathrm{maximize} \ \ p(x_{1},\dots, x_{n})
%$$
%where $p(x_{1},\dots, x_{n})$ is some polynomial function, one can instead try to solve the (generally much simpler) problem of maximizing the associated tropical polynomial $\mathsf tp$ instead.

Let us start with a very basic probabilistic language:
the terms are $M::= \true \mid \false \mid M\oplus_p M$, for $p\in[0,1]$, and the operational semantics is $M\oplus_p N\to pM$ and $M\oplus_p N \to (1-p)N$, so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
Consider the program
$$
 M:=(\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true)).
 $$
 Calling $q=1-p$, to each occurrence of $\true$ or $\false$ in $M$, univocally determined by an address
$\omega\in \{l,r\}^{*}$, is associated a monomial $P_{\omega}(p,q)$ which determines the probability of the event ``$M\twoheadrightarrow_{\omega} \true/\false$'', that is, that $M$ reduces to $\true/\false$ according to the choices in $\omega$.
 For instance, we have
$P_{rll}(p,q):=qp^2$,
$P_{rrr}(p,q):=q^3$, and 
$P_{rrl}(p,q)=P_{rlr}(p,q):=q^2p$.
The polynomial function $Q_{\true}(p,q):=P_{ll}(p,q)+P_{rll}(p,q)+P_{rrr}(p,q)=p^2+p^2q+q^3$ gives instead the probability of the event ``$M\twoheadrightarrow \true$'', and analogously for $Q_{\false}(p,q):=P_{lr}(p,q)+P_{rrl}(p,q)+P_{rlr}(p,q)=pq+2pq^2$.

Two natural questions at this point concern the \emph{likelihood} of a given reduction address $\omega_{0}$:
%
%The tropical point of view allows now to express two natural questions about this situation:
\begin{enumerate}
 \item Which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow \false_{\omega_0}$''?
 I.e., which is the choice of $p,q$ that maximizes the probability $P_{\omega_0}$?
 \item 
Knowing that $M$ produced $\true$ (similarly for $\false$) , which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow \false_{\omega_0}$'', knowing that ``$M\twoheadrightarrow \false$''?
I.e., which is the choice of $p,q$ that makes $\omega_0$ the most likely path among those leading to $\false$ (i.e.\ that maximizes the conditional probability $\BB P(``M\twoheadrightarrow \false_{\omega_0}'' \mid ``M\twoheadrightarrow \false'')$)?
\end{enumerate}

Answering 1) and 2) amounts then at solving a maximization problem related to $P_{\omega_{0}}(p,q)$ or $Q_{\true}(p,q)$. These are more easily solved by passing to the associated tropical polynomials.
For 1), the maximum values $x,y$ of $P_{rll}(p,q)$ can be computed by finding the \emph{minimum} values of $\mathsf tP_{rll}(-\log p, -\log q)= -2\log p- \log q$. Notice that $\mathsf tP_{rll}$ is precisely the \emph{negative log-probability} of the event ``$M\twoheadrightarrow_{rll} \false$''. For 2), the maximum values of $Q_{\true}(p,q): [0,1]^{2}\to [0,1]$ can be computed as $e^{-\alpha},e^{-\beta}$, where $\alpha,\beta\in[0,\infty]$ are the \emph{minimum} values of the tropical polynomial 
$\mathsf t Q_{\true}(\alpha,\beta) = \min \{ 2\alpha, 2\alpha+\beta, 3\beta\}$.

While we here consider a very basic language, similar arguments could be done by replacing $\true$ and $\false$ by, respectively, a converging and diverging term (e.g.~in a $\B{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.

\subparagraph*{Best Case Analysis for Non-Deterministic Languages}

This example is inspired from \cite{Manzo2013}. Again, we start from a very basic non-deterministic language with terms $M::= \true \mid  \mathsf{gen} \mid  M + M$, with an operation semantics comprising a non-deterministic reduction rule 
$M_{1}+M_{2} \stackrel{\alpha}{\to} M_{i}$ and a generation rule
$\mathsf{gen}\stackrel{\beta}{\to} \true +\mathsf{gen}$, 
where in each case the value $\alpha,\beta\in \Lawv$ indicates a \emph{cost} associated with the reduction (e.g.~the estimated clock value for the simulation of each reduction on a given machine model). 
Then, any reduction $\omega: M \to^{*} N$ of a term to (one of its) normal form is  associated with a tropical monomial $P_{\omega}( \alpha,\beta)$ consisting of the sum of the costs of all reductions in $\omega$. For a given normal form $N$, the reductions $\omega_{i}: M \to^{*}N$ give rise to a tLs $\inf_{i\in I}P_{\omega_{i}}(\vec \alpha,\beta)$. 
For example, consider the non-deterministic term below
$$
M :=\mathsf{gen}+  ((\true + \true) + \mathsf{gen})
$$
The (infinitely many) reduction paths leading to $\true$ can be grouped as follows:
\begin{itemize}
\item left, then reduce gen $n+1$-times, then left;
\item right, then left and then either left or right;
\item right twice, then reduce gen $n+1$-times and then left.
\end{itemize}
This leads to the tLs 
$$\varphi_{M\to^{*}\true}(\alpha)=\inf_{n\in \mathbb N}\big\{2\alpha+(n+1)\beta, 3\alpha, 3\alpha+(n+1)\beta\big \}= \min\{ 2\alpha+\beta, 3\alpha\}$$which describes all possible behaviors of $M$. Notice that, since $\alpha$ and $\beta$ are always positive, the power series $\varphi_{M\to^{*}\true}(\alpha)$ is indeed equivalent to the tropical polynomial $\min\{ 2\alpha+\beta, 3\alpha\}$. In other words, of the infinitely many behaviors of $M$, only finitely many have chances to be \emph{optimal}: either left + gen + left, or right + left + (left,right). This seems to be a very interesting feature of tropical semantics: reducing to best-case analysis leads, in many cases, to collapse the infinitary description of \emph{all} behaviors to a purely combinatorial description of the finitely many optimal ones.

Once reduced $\varphi_{M\to^{*}\true}$ to a polynomial, the best behavior among these will depend on the values of $\alpha$ and $\beta$, and by studying the tropical polynomial $\varphi_{M\to^{*}\true}$ one can thus answer questions analogous to 2) above, that is, what are the best choices of costs $\alpha,\beta$ making a \emph{chosen} reduction of $M$ to $\true$ the cheapest one?

### 2023-07-29 CSL24/section3fancy.tex
commit

% !TEX root = main.tex

What we here call the tropical relational model is indeed an instance of a general and well-known construction that yields, for any \emph{continuous} semiring $Q$, a model $\QREL_{!}$ of effectful extensions of the simply typed $\lambda$-calculus and $\mathrm{PCF}$. In this section we shortly recall this construction, and also discuss how it adapts to model \emph{graded} and \emph{differential} variants of $\STLC$. In the next sections we will instead focus on the peculiar features of the tropical case.

\subparagraph*{Weighted Relational Models}
For any \emph{continuous} semiring $Q$ (i.e.~a cpo equipped with an order-compatible semiring structure), one can define a category $\QREL$ of ``$Q$-valued matrices'' which is well-known to yield a model of both probabilistic and non-deterministic versions of $\mathrm{PCF}$ (see e.g.~\cite{Manzo2013, Pagani2018}).
More precisely, $\QREL$ (\cite{Manzo2013} calls it $Q^\Pi$) has sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms, i.e.~$\QREL(X,Y):=Q^{X\times Y}$.
%The identity morphism of $\QREL$ is the identity matrix $I\in Q^{X\times X}$,
% given by $i_{a,a}=1$ and $i_{a,b\neq a}=0$, and
The composition $st\in Q^{X\times Z}$ of $t\in Q^{X\times Y}$ and $s\in Q^{Y\times Z}$ is given by $(st)_{a,c}:=\sum_{b\in Y} s_{b,c}t_{a,b}$ (observe that this series always converges because $Q$ is continuous).
For any set $X$, $Q^X$ is a $Q$-semimodule and we can identify $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$, which have shape $f(x)_b=:\sum_{a\in X} \matr f_{a,b}x_a$, for some matrix $\matr f\in Q^{X\times Y}$.
% \begin{remark}
 %Following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we chose to see a matrix $t$ from $X$ to $Y$ as a map $t:X\times Y\to Q$.
% 
% fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum_{b\in Y} s_{b,c}t_{a,b}$.
Notice that usual linear algebra conventions correspond to work in $\QREL^{\op}$, %a matrix $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns
e.g.\ the usual matrix-vector product defines a map $Q^Y\to Q^X$.
Following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we are instead working with transpose matrices.

$\QREL$ admits a comonad $!$ which acts, on objects, by taking the finite multisets.
Remember that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
Now, although a matrix $t\in\HOM{\QREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, by exploiting the coKleisli structure we can also ``express it in the base $X$'', which leads to the \emph{non-linear} map $t^{!}.Q^{X}\to Q^{Y}$ defined by the power series
$
t^{!}(x)=t\circ_{!}x : b \mapsto \sum_{\mu\in !X}t_{\mu,b}\cdot x^{\mu}
$, 
where $x^{\mu}= \prod_{a\in x}x_{a}^{\mu(a)}$.

%\end{remark}

When we instantiate $Q=\Lawv$, we obtain a category $\LREL$ of $\Lawv$-values matrices. As one might expect, this category is tightly related to Lawvere's theory of (generalized) metric spaces. For the moment, let us just observe that a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law.
We will come back to this viewpoint in Section \ref{section6}.

Composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
reads as \ $(st)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$, moreover, the non-linear functions $t^{!}: \Lawv^{X}\to \Lawv^{Y}$ give rise to functions of shape 
$t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$, which correspond to the generalisation of tLs with \emph{possibly infinitely} many variables (in fact, as many as the elements of $X$).
By identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the %functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ we recover the 
usual tLs's of one variable. %
%similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$, %induced by matrices
%which we call \emph{tropical linear}, are exactly those of shape $f(x)_b=\inf_{a\in X} \set{\matr f_{a,b}+x_a}$, for some $\matr f\in\Lawv^{X\times Y}$.
%%\end{remark}
For example, the $\varphi$ of \autoref{fig:plot1} is indeed of shape $\varphi=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
% Therefore $\LREL_!$ is not a full-complete model of $\STLC$.

\begin{remark}
 Let $X$, $Y$ be sets and let $\langle \_,\_\rangle:X\times Y \to \mathbb{R}$.
 For $f:X\to \mathbb R$, define $f^*:Y\to \mathbb R$ by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
 Then for $X=!A$, $Y=\Lawv^A$, where $A$ is a set, and $\langle \mu, y \rangle:= \mu y$, we have that $f^!(y)=(-f)^*(-y)$ for all $f\in\Lawv^{!A}$.
 This is precisely the same formal construction yielding the well-known convex conjugate $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
 This construction is in turn a generalisation of the Legendre transformation.
% Despite the formal constructions being the same, we ignore for the moment if these could be connected to the study of high-order programs in our setting.
\end{remark}

\subparagraph*{Interpreting effectful higher-order programs}
The interpretation of the simply typed $\lambda$-calculus $\STLC$ in $\LREL$ relies on the fact that all categories $\QREL_{!}$ are cartesian closed \cite{Manzo2013}, with cartesian product and exponential objects acting on objects $X,Y$ as, respectively, $X+Y$ and $!X\times Y$. 
Hence, given an environment $\eta$ associating ground types with sets, any 
typable term $\Gamma \vdash M:A$ gives rise to a morphism 
$\model{M}_{\eta}\in \LREL_{!}(\model{\Gamma}_{\eta}, \model{A}_{\eta})$, and thus to a generalized tLs $\model{M}_{\eta}^{!}:\Lawv^{\model{\Gamma}_{\eta}}\to \Lawv^{\model{A}_{\eta}}$. 
This interpretation extends to $\mathrm{PCF}$ by interpreting the fixpoint combinator $\mathbf Y$ via the matrices $\mathrm{fix}^{X}= \inf_{n}\big\{\mathrm{fix}^{X}_{n}\big\} \in \Lawv^{  !(!X\times X) \times X  }$, where 
 $\mathrm{fix}^{X}_{0}=0$ and $\mathrm{fix}^{X}_{n+1}= \RM{ev}\circ_{!}\langle \mathrm{fix}_{n}^{X}, \RM{id} \rangle $.

One can easily check, by induction on a typing derivation, that for any program of $\STLC$ or $\mathrm{PCF}$, the associated matrix is \emph{discrete}, that is, its values are included in $\{0,\infty\}$. Indeed, as suggested in Section \ref{section2}, the actual interest of tropical semantics lies in the interpretation of effectful programs. 
%
%In particular, the \emph{evaluation} morphisms $\RM{ev}$ in $\LREL$ are discrete matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$ given by 
%$\RM{ev}_{[ (\mu,b) ,\mu ] , b }=0$, and being $\infty$ in all other cases, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$, $(s\circ_! t)_{\mu,c}:=
%%\begin{align}
%\inf_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
% \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
% For example, setting $\model{A}=\{\star\}$, the interpretation of the $\lambda$-term $\lambda x.\lambda y.xy: (A\to A)\to (A\to A)$ yields the tLs
% $\model{M}^{!}: \Lawv^{\mathbb N}\to \Lawv \to \Lawv$ (recalling that we can identify $\mathcal M_{\mathrm{fin}}(\{\star\})$ with $\mathbb N$) given by 
% $$
% \model{M}^{!}(x)(y)= \RM{ev}^{!}(\langle x,y\rangle)= \inf_{n\in \mathbb N}\big\{ x_{n} + ny\big\}.  
% $$
% Moreover,  
As the homsets $\LREL_{!}(X,Y)$ are $\Lawv$-modules, it is possible to interpret in extensions of $\STLC$ and $\mathrm{PCF}$ comprising $\Lawv$-module operations $\alpha \cdot M$ and $M+N$ \cite{Manzo2013}, by 
letting $\model{\alpha\cdot M}_{\eta}=\model{M}_{\eta}+\alpha$ and 
$\model{M+N}_{\eta}=\min\{\model{M}_{\eta},\model{N}_{\eta}\}$. 
More precisely, \cite{Manzo2013} considers a language $\mathrm{PCF}^{Q}$ corresponding to $\mathrm{PCF}$ extended with $Q$-module operations, with an operational semantics given by rules $M\stackrel{1}{\to}M'$ for each rule $M\to M'$ of $\mathrm{PCF}$ (here $1$ is the monoidal unit of $Q$) as well as 
$M_{1}+M_{2} \stackrel{1}{\to}M_{i}$ and $\alpha\cdot M \stackrel{\alpha}{\to} M $.
Hence, any reduction $\omega: M\to^{*} N$ is naturally associated with a weight $\mathsf w(\omega)\in Q$.
In particular, from [\cite{Manzo2013}, Theorem V.6] we deduce the following result, which relates the tropical semantics of a program with its best-case operational behavior.  
\begin{proposition}
For every closed term $M$ of type $\mathrm{Nat}$ in $\mathrm{PCF}^{\Lawv}$ and $n\in \mathbb N$, $\model{M}_{n}=  \inf\big\{ \mathsf w(\omega) \ \big \vert \ \omega : M \to \underline n\big \}.$
\end{proposition}

Observe that the two examples shown in Section \ref{section2} can easily be rephrased in the language $\mathrm{PCF}^{\Lawv}$. 
For instance, for the probabilistic example, for a given valuation $\RM{val}$, one can use a translation $(M\oplus_{p}N)^{\RM{val}}= \min\{(M^{\RM{val}}+\RM{val}(p)), N^{\RM{val}}+\RM{val}(1-p)\}$, so that the reductions 
$:M\oplus_{p}N\stackrel{p}{\to} M$ and $M\oplus_{p}N\stackrel{p-1}{\to} N$ translate into a sequence of two reductions
$(M\oplus_{p}N)^{\RM{val}} \stackrel{0}{\to} M^{\RM{val}}\stackrel{\RM{val}(p)}{\to} M$ 
and $(M\oplus_{p}N)^{\RM{val}} \stackrel{0}{\to} N^{\RM{val}}\stackrel{\RM{val}(1-p)}{\to} M$, with the weight of the latter corresponding to the valuation of the weight of the former.

Let $\mathrm{PPCF}$ (for \emph{probabilistic $\mathrm{PCF}$} \cite{Pagani2018}) be standard $\mathrm{PCF}$ extended with the constructor $M\oplus_{\frac{1}{2}}N$ and its associated reduction rules. From the above discussion we deduce the following:

\begin{corollary}
For every closed probabilistic term $M$ of type $\mathrm{Nat}$ in $\mathrm{PPCF}$ and $n\in \mathbb N$, 
$\model{M}^{\mathrm{logval}}_{n}\in \Lawv$ (cf.~Remark \ref{rmk:val_trop}) is the minimum negative log-probability of any
reduction from $M$ to $\underline n$.
\end{corollary}

\cite{Manzo2013} obtained a similar result for a non-deterministic version of PCF, by translating each term into $\mathrm{PCF}^{\Lawv}$via $(\lambda x.M)^{\circ}=\lambda x.M^{\circ}+1$ and $(\mathbf YM)^{\circ}= \mathbf Y(M^{\circ}+1)$
(\cite{Manzo2013} considers the discrete tropical semiring $\mathbb N\cup\{\infty\}$, but the result obviously transports to $\Lawv$).

\begin{corollary}[cf.~\cite{Manzo2013}, Corollary VI.10]
For every closed probabilistic term $M$ of type $\mathrm{Nat}$ in non-deterministic $\mathrm{PCF}$ and $n\in \mathbb N$, 
$\model{M^{\circ}}_{n}$ is the minimum number of $\beta$- and $\mathsf{fix}$- redexes reduced in a reduction sequence from $M $ to $\underline n$. 
\end{corollary}

%, an instance of the language $\mathrm{PCF}^{\mathcal R}$ of \cite{Manzo2013}, via the translations 
%$(M\oplus_{p}N)^{\circ} =pM^{\circ}\mathtt{or}(1-p)N^{\circ}$ and $(M+N)^{\circ}=M^{\circ}\mathtt{or}N^{\circ}$.
%In particular,

\subparagraph*{Interpreting $\mathbb N$-Graded Types}\label{sec:BSTLC}

We now show how to interpret in $\LREL_{!}$ a graded version of $\STLC$, that we call $\BSTLC$, indeed a simplified version of the well-studied language $\mathrm{Fuzz}$ \cite{Reed2010}. 
The terms of $\BSTLC$ as for the $\STLC$, the types are $A::= * \ \mid  \ !_{n}A \multimap A$, the contexts of the typing judgements are sets of declarations of the form $x :_{n}A$, with $n\in \mathbb N$, and the typing rules are: %given in Fig.~\ref{fig:rules}
	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
	\begin{array}{cccc}
		\prooftree
		\Gamma \vdash M:A
		\justifies
		\Gamma, x:_{0}B \vdash M:A
		\endprooftree 
		&
		\prooftree
		\Gamma, x:_{n}B, y:_{m} B\vdash M:A
		\justifies
		\Gamma, x:_{n+m}B\vdash M\{x/y\}:A
		\endprooftree 
		&
		\prooftree
		\Gamma, x:_{n} A\vdash M: B
		\justifies
		\Gamma\vdash \lambda x.M: !_{n}A\multimap B
		\endprooftree
		&
		\prooftree
		\Gamma \vdash M: !_nA\multimap B
		\quad
		\Delta\vdash N: A
		\justifies
		\Gamma +n\Delta\vdash MN: B
		\endprooftree
	\end{array}
	\]
where $\Gamma+\Delta$ is defined letting $(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) =  (\Gamma+\Delta), x:_{m+n}A$, and $m\Gamma$ is made all $x:_{mn}A$ for $(x:_{n}A) \in \Gamma$.  
The axiom is $x:_{1}A\vdash x: A$.
The main feature of this language is that if $\vdash \lambda x.M:\,!_nA\multimap B$, then $x$ is duplicated at most $n$ times in the reduction of $\vdash (\lambda x.M)N :B$ to the normal form.

Now, one can see that the comonad $!$ of $\LREL$ can be ``decomposed'' {\color{red}REF} into a family of ``graded exponentials functors'' $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined on objects by taking multisets of cardinality \emph{at most} $n$. %  lift to functors 
The sequence $(!_n)_{n\in\N}$ gives rise to a so-called \emph{$\N$-graded linear exponential comonad} on (the SMC) $\LREL$. %satisfying the adjunction: $\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
As such, $(\LREL,(!_n)_{n\in\N})$ yields then a model of $\BSTLC$. Remark that since arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$. Notice that, whenever $\model *$ is finite, the set $\model{A}$ is finite for any type $A$ of $\BSTLC$.

\subparagraph*{Interpreting the Differential $\lambda$-Calculus}\label{sec:STDLC}

%We show that $\LREL_{!}$ also interprets the $\STDLC$,  
A Cartesian closed differential $\lambda$-category (CC$\partial\lambda$C)\cite{Manzo2010,Blute2009, Blute2019} is a CCC enriched over commutative monoids (i.e.\ we can add morphisms and there is a $0$ morphism), with the cartesian closed structure compatible with the additive structure, 
and equipped with a certain \emph{differential operator} $D$, generalising the usual notion of differential, see e.g.\ \cite{BluteEhrhTass10}.
An example is the CC$\partial\lambda$C of convenient vector spaces with smooth maps, where $D$ is the ``real'' differential of smooth maps.
%$Df:\mathbb{V}\times\mathbb{V}\rightarrow \mathbb{W}$, 
%$Df(x,u):=\dfrac{d}{dt}{\!\Big|_{t=0}} f(x+tu)$, of smooth maps $f:\mathbb{V}\rightarrow \mathbb{W}$.
%More precisely, a cartesian closed category $\C C$ is a $C\partial C$ when:
%\begin{itemize}
%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
%\item $\C C$ is equipped with a differential operator $D:
%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
%\end{itemize}

Applying \cite[Theorem 6.1]{lemay2020} one can check that
%\begin{proposition}[{\color{red}LEMAY??}]\label{thm:LREL!CCDC}
 $\LREL_!$ is a CC$\partial\lambda$C (see Section \ref{section5bis}) when equipped with the \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
%\end{proposition}
%\begin{remark}For $t\in\HOM{\LREL}{!X}{Y}$, we have: $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu+\nu'+\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho+\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.\end{remark}
%This ensures that one can define a sound interpretation of $\STDLC$-terms in the standard way (see [Section 4.3, \cite{Manzo2010}]).
As such, $(\LREL_!,D)$ is a model of the differential $\lambda$-calculus $\STDLC$ \cite{ER}.

% an extension of $\STLC$ ensuring exact control of duplications via a notion of linear substitution in its operational semantics.
%We will not deal with its operational properties, so let us only give its syntax (see \cite[Section 3]{Manzo2010}): there are \emph{terms} $M$ and \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
%quotiented by $\alpha$-equivalence, by equations that make $+,0$ form a commutative monoid on the set of sums, %, i.e.\ commutativity and associativity of $+$ and neutrality of $0$ w.r.t.\ $+$;
%by linearity of $\lam x.(\_)$, $\Diff{\_}{\_}$ and $(\_)\mathbb T$ (but \emph{not} of $M(\_)$) and by irrelevance of the order of consecutive $\Diff{\_}{\_}$.
%%Remark that $M(\_)$ is \emph{not} set to be linear: $\lambda x.0=0\mathbb T=\Diff{0}{N}=\Diff{M}{0}=0$ but $M0\neq0$ in general.
%%This is crucial for the definition of the Taylor expansion.
%We follow the tradition of quotienting also for the idempotency of $+$.
%%Sums are, then, just \emph{finite} sets of terms.
%The types are $A::= *\mid A\to A$, the typing rules: %in Figure~\ref{fig:rules}
%	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
%	\begin{array}{cccc}
%		\prooftree
%		\Gamma, x: A\vdash M: B
%		\justifies
%		\Gamma\vdash \lambda x.M: A\to B
%		\endprooftree 
%		&
%		\prooftree
%		\Gamma \vdash M: A\to B
%		\quad
%		\Gamma\vdash \mathbb T: A
%		\justifies
%		\Gamma \vdash M\mathbb T: B
%		\endprooftree 
%		&
%		\prooftree
%		\Gamma \vdash M: A\to B
%		\quad
%		\Gamma \vdash N: A
%		\justifies
%		\Gamma \vdash \Diff{M}{N}: A\to B
%		\endprooftree
%		&
%		\prooftree
%		\Gamma\vdash M_1: A
%		\,\cdots\,
%		\Gamma \vdash M_n:A
%		\justifies
%		\Gamma \vdash M_1+\cdots +M_n : A
%		\using (n\geq 2)
%		\endprooftree
%	\end{array}
%	\]
%where a context $\Gamma$ is a list of typed variable declarations.
%The axioms are $\Gamma, x:A \vdash x: A$ and $\Gamma\vdash 0:A$.
%The main feature of this language is that $\Der^n[\lambda x.M,N^n]0$ has a non-zero normal form iff $x$ is duplicated exactly $n$ times during the reduction to normal form.
%In $\BSTLC$ the typing system handles duplications; in $\STDLC$ the syntax with its operational semantics (that we do not give) does it.
\begin{comment}
Writing $\Der^2[\_,(\_)^2]$ as a shortcut for $\Der[\Der[\_,\_],\_]$ and $\Der^1[\_,(\_)^1]$ for $\Diff{\_}{\_}$, the analogue of the previous $\BSTLC$-term is $\vdash_{\STDLC} \lambda {\color{red}z}. \Der^{\color{red}2}[
	\lambda x{\color{green}y}.
		\Der^{\color{violet}1} [
				\Der^{\color{blue}1} [y, x^{\color{blue}1}]
        0, x^{\color{violet}1}
	]0
, z^{\color{red}2}]0
: {\color{red}*}\to ({\color{green}* \to * \to *}) \to *$.
%Here we wrote $\Der^2[\_]\cdot (\_)^2$ as a shortcut for $\Der[\Der[\_]\cdot (\_)]\cdot (\_)$ and $\Der^1[\_]\cdot (\_)^1$ for $\Der[\_]\cdot (\_)$.
In particular, if the \emph{multiplicities} of the arguments (the colored exponents) do not exactly match the number of duplications, e.g.\ in $\vdash_{\STDLC} \lambda z. \Der^{\color{red}3}[
	\lambda xy.
		\Diff{
				\Diff{y}{x}
		0}{x}
	0
, z^{\color{red}3}]0
: *\to (* \to * \to *) \to *$, then the term reduces to the empty sum $0$ (representing an \emph{error}).
\end{comment}
%Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).

%Until now we simply specialised well-known results in our tropical case, with the intent of showing how things read in this particular case.
%Now we go further, by showing that $\LREL_!$ actually admits a \emph{differential structure}, turning it into a model of the $\STDLC$, i.e.~a $CC\partial C$.
%This viewpoint
%, is where the \emph{metric} and the \emph{differential} viewpoints converge, as explained in the Introduction and Section II, and it % will be further generalised in Section \ref{section6}.

%A model of the $\STDLC$ is usually understood as so-called \emph{Cartesian closed differential categories} (CC$\partial$C), see \cite{Manzo2012} for details.
%In order to treat the $+$ and the constructor $D[\_]\cdot (\_)$ of $\STDLC$, the main features of a CC$\partial$C $\C C$ are that:
%
%1) $\C C$ is a left-additive-CCC, i.e.\ its Homsets are commutative monoids and its Cartesian closed structure is well behaved w.r.t.\ this monoid structure;
%
%2) $\C C$ is equipped with a differential operator map $D:\HOM{\C C}{X}{Y}\to \HOM{\C C}{X\times X}{Y}$ (here $\times$ is the Cartesian product of $\C C$) satisfying $8$ axioms, called D1, ..., D7, D-curry.

%Let us show the differential structure of $\LREL_!$ (remember that the Cartesian product of $\LREL_!$ is the disjoint union $+$).

%\begin{definition} The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).\end{definition}

%The models of $\STDLC$ are the cartesian \emph{closed} differential categories ($CC\partial C$), which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}. 
%
%By induction on $M$ %(using that composition and $D$ easily preserve ``booleaness'' and projections and evaluation of  $\LREL$ are boolean)
%one can prove that:
%
%\begin{proposition}\label{prop:descrete}
% For $\BSTLC$ and $\STDLC$ (and thus also for its fragment $\STLC$), we have that $\model{\Gamma\vdash M:A}\in\Lawv^{!\model{\Gamma}\times\model{A}}$ is a \emph{boolean} matrix, i.e.\ actually $\model{\Gamma\vdash M:A}\in\{0,\infty\}^{!\model{\Gamma}\times\model{A}}$.
%\end{proposition}
%
%For example, the $\varphi$ of \autoref{fig:plot1} is not the interpretation of a $\lam$-term, because its matrix is not discrete.
%This can be seen as the fact that such calculi are relatively trivial, from this point of view.
%Even if the interpretation of those languages is trivia, we shall still focus on them in order to set the basis of the further studies and in order to see some already interesting properties of their interpretation.{\color{red}SPIEGARE MEGLIO}
%
%

### 2023-07-29 CSL24/section5bis.tex
commit

--- 2023-01-24 section2.tex
@@
-%Recall the two approaches with more details on lambda-calculus and on existing challenges.
+% !TEX root = /Users/paolopistone/Documents/GitHub/tropicalnew/CSL24/main.tex
 
@@
 
-\subsection{Metric approach: bounded $\lambda$-terms as Lipschitz functions}
+\subparagraph*{Metric approach: bounded $\lambda$-terms as Lipschitz functions}
 
@@
 
-In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior.
-A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
-$d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
-A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the \emph{linear} simply typed $\lambda$-calculus, being a \emph{symmetric monoidal closed} category, and in fact it also models {affine} terms (since the cartesian and monoidal units coincide).
+In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which accounts for differences in behavior.
+A fundamental insight coming from this line of work is that, if one can somehow   \emph{bound} the number of times that a program may duplicate its input, the resulting program will be \emph{Lipschitz-continuous}:
+if $M$ may duplicate at most $L$ times, then an error $\epsilon$ between two inputs will result in an error less or equal to $L\cdot \epsilon$ in the corresponding outputs \cite{Reed2010, Gaboardi2017}.
+For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the functional input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.
 
-As observed in \cite{Reed2010, Gaboardi2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications.
-The fundamental intuition is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz} map.
-For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the functional input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.
-By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $r\cdot X$ (i.e.~with distance $d_{r}(x,y)=r\cdot d(x,y)$) and $Y$, the program $M$ above
-can thus be interpreted as a non-expansive map from $2\cdot(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.
+%
+%By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $r\cdot X$ (i.e.~with distance $d_{r}(x,y)=r\cdot d(x,y)$) and $Y$, the program $M$ above
+%can thus be interpreted as a non-expansive map from $2\cdot(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.
 
-These observations have led to the study of linear $\lambda$-calculi with \emph{graded} exponentials, like $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied to the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}.
+These observations have led to the study of $\lambda$-calculi with \emph{graded} exponentials, like $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied to the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}.
 The types of such systems are defined by combining linear constructors with a \emph{graded linear exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
-In the following sections we will sometimes make reference to a basic graded type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined via $A::= o  \mid   !_{n}A\multimap A  $,
-where $n\in \BB N$.
+%In the following sections we will sometimes make reference to a basic graded type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined via $A::= o  \mid   !_{n}A\multimap A  $,
+%where $n\in \BB N$.
 Intuitively, $!_{ n}A\multimap B$ is the type of functions from $A$ to $B$ that may use their input \emph{at most} $n$ times.
@@
 
-Now, what about the good old, ``unbounded'', simply typed $\lambda$-calculus?
+What about the good old, ``unbounded'', simply typed $\lambda$-calculus?
 Actually, by using unbounded duplications, one might lose the Lipschitz property.
@@
 
-\subsection{Resource approach: differential $\lambda$-terms as polynomials}
+\subparagraph*{Resource approach: differential $\lambda$-terms as polynomials}
 
@@
 
-Interestingly, the categorical study of the operator $\Der$ has led to the introduction of \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, a class of categories providing an abstract axiomatization of differentiation, in the usual mathematical sense.
+%Interestingly, the categorical study of the operator $\Der$ has led to the introduction of \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, a class of categories providing an abstract axiomatization of differentiation, in the usual mathematical sense.
 More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
-\begin{itemize}
-\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
-\item $\C C$ is equipped with a differential operator $D:
-\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
-\end{itemize}
+%\begin{itemize}
+%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
+%\item $\C C$ is equipped with a differential operator $D:
+%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
+%\end{itemize}
+%
+%Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
+%The models of $\STDLC$ are the 
+%cartesian \emph{closed} differential categories ($CC\partial C$), 
+%which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.
 
-Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
-The models of $\STDLC$ are the 
-cartesian \emph{closed} differential categories ($CC\partial C$), 
-which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.
-
-Another intriguing similarity between program derivatives and 
+An intriguing similarity between program derivatives and 
 actual derivatives is provided by the \emph{Taylor expansion} $\C T$:
@@
 
-\subsection{Tropical Mathematics: a possible synthesis}%Outline of the paper}
+\subparagraph*{Tropical Mathematics: a possible synthesis}%Outline of the paper}
 
@@
 differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}).
-
-%
-%At this point, 
-%by interpreting such polynomials over the tropical semiring 
-%
-%Our question can thus be reformulated as follows: can we make the relational semantics \emph{Lipschitz}, hence amenable to metric and sensitivity analysis?
-The goal of this paper is to show that, by appealing to tropical mathematics, this is indeed possible and leads to the somehow unespected discovery of a bridge between the metric and differential study to higher-order programs.
-%
-%Fino a qui niente a capo
-%
-%Fino a qui niente a capo
-%
-%Fino a qui niente a capo
-%
-%Fino a qui niente a capo
-%
-%Fino a qui niente a capo
-%
-%Limite! Alla prossima riga sforo oltre 12 pag.
-%
-
-%In the next section we recall some basic ideas from tropical mathematics, and its connection with the study of the Lawvere quantale.
-%Since polynomials correspond to piecewise linear (hence Lipschitz) functions in tropical mathematics,
-%The reconstruction of the relational semantics over the tropical semi-ring, presented in Section \ref{section3} and Section \ref{section4}, will provide a metric semantics of the differential $\lambda$-calculus, bridging sensitivity and resource analysis.
-%In Section \ref{section5} we suggest potential applications of this approach, relating well-studied applications of program metrics, resource analysis with current uses of tropical mathematics in computer science.
-%Finally, in Section \ref{section6} we show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between {generalized metric spaces} and modules over the tropical semiring.

### 2023-07-29 CSL24/section2bisbis.tex
commit

@@
 $\omega\in \{l,r\}^{*}$, is associated a monomial $P_{\omega}(p,q)$ which determines the probability of the event ``$M\twoheadrightarrow_{\omega} \true/\false$'', that is, that $M$ reduces to $\true/\false$ according to the choices in $\omega$.
+Thinking of $p,q$ as parameters, $P_{\omega}(p,q)$ can thus be read as the \emph{likelihood function} of the event ``$M\twoheadrightarrow \true_\omega/\false_\omega$''.
  For instance, we have
@@
 The polynomial function $Q_{\true}(p,q):=P_{ll}(p,q)+P_{rll}(p,q)+P_{rrr}(p,q)=p^2+p^2q+q^3$ gives instead the probability of the event ``$M\twoheadrightarrow \true$'', and analogously for $Q_{\false}(p,q):=P_{lr}(p,q)+P_{rrl}(p,q)+P_{rlr}(p,q)=pq+2pq^2$.
-
-Two natural questions at this point concern the \emph{likelihood} of a given reduction address $\omega_{0}$:
+This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baumr966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2ll4}.
+Typical questions in this case would be, for a fixed $\omega_0$:
 %

### 2023-07-29 CSL24/section7.tex
commit

@@
+The applications of tropical mathematics in computer science abound, e.g.~in automata theory \cite{Chua1992, Simon}, machine learning \cite{Maragos2021, Pachter2004, Zhang2018}, optimization \cite{Akian2011, Akian2012}, and convex analysis \cite{Lucet2009}.
+
 The connections between differential $\lambda$-calculus (and differential linear logic), relational semantics, and non-idempotent intersection types are very well-studied (see \cite{decarvalho2018}, and more recently, \cite{Mazza2016} for a more abstract perspective, and \cite{Olimpieri2021, Galal2021} for a 2-categorical, or proof-relevant, extension).
-As we said, the relational semantics over the tropical semi-ring was quickly explored in \cite{Manzo2013}, to provide a ``best case'' resource analysis of a $\B{PCF}$-like language with non-deterministic choice.
+As we said, the relational semantics over the tropical semiring was quickly explored in \cite{Manzo2013}, to provide a ``best case'' resource analysis of a $\mathrm{PCF}$-like language with non-deterministic choice.
 \emph{Probabilistic coherent spaces} \cite{Ehrhard2011}, a variant of  the relational semantics, provide an interpretation of higher-order probabilistic programs
-as analytic functions. In \cite{Ehrhard2022} it was observed that such functions satisfy a local Lipschitz condition somehow reminiscent of our examples in Section \ref{section4}.
+as analytic functions. In \cite{Ehrhard2022} it was observed that such functions satisfy a local Lipschitz condition somehow reminiscent of our examples in Section \ref{section2}.
 
@@
 More generally, the literature on program metrics in denotational semantics is vast.
-Since at least \cite{VANBREUGEL20011} metric spaces, also in Lawvere's generalized sense \cite{Lawvere1973}, have been exploited as an alternative framework to standard, domain-theoretic, denotational semantics.
-While standard categories of metric spaces are not models of the full simply typed $\lambda$-calculus, several constructions of cartesian closed categories of metric spaces can be found in the literature.
-For instance,
-\emph{ultra}-metric spaces form a CCC, and have been shown to model $\B{PCF}$ \cite{Escardo1999}.
-Also \emph{partial} metrics, introduced in \cite{matthews}, have been shown to provide models of $\STLC$, under suitable generalizations \cite{Geoffroy2020}.
-More generally, \cite{Clementino2006} provides a general characterization of exponentiable objects in categories of (generalized) metric spaces, and \cite{PistoneLICS, PistoneFSCD2022} provide other ways to construct CCCs of
-(generalized) metrics, including one based on locally Lipschitz maps, 
-using ideas from {differential logical relations} and quantitative equational theories \cite{Plotk}.
+Since at least \cite{VANBREUGEL20011} metric spaces, also in Lawvere's generalized sense \cite{Lawvere1973}, have been exploited as an alternative framework to standard, domain-theoretic, denotational semantics, notably
+\emph{ultra}-metrics and \emph{partial} metrics have been shown to form a CCC \cite{Escardo1999,PistoneLICS, PistoneFSCD2022}.
 
@@
 
-Finally, applications of tropical mathematics in computer science abound, originally for automata theory \cite{Chua1992, Simon}, and more recently
- in optimization methods for machine learning and other statistical models (see e.g.~\cite{Maragos2021, Pachter2004, Zhang2018}), optimization \cite{Akian2011, Akian2012}, and convex analysis \cite{Lucet2009}.
-While our discussion in Section \ref{section5} is inspired by a well-known application of tropical polynomials to Hidden Markov Models \cite{Pachter2004},
-the vast literature in this domain lets us think that other ways to 
-apply tropical semantics to the analysis of higher-order programs might be studied.
-
+%While our discussion in Section \ref{section5} is inspired by a well-known application of tropical polynomials to Hidden Markov Models \cite{Pachter2004},  
+%the vast literature in this domain lets us think that other ways to 
+%apply tropical semantics to the analysis of higher-order programs might be studied.
+%
 %

### 2023-07-29 CSL24/section8.tex
commit

@@
-{\color{red}due parole sul fatto che non sappiamo se si posso usare strumenti propri della tropical geo per lo studio dei programmi. Per esempio:}
-\begin{remark}\label{rem:troproots}
- Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_{\true}$ or of $\trop Q_{\false}$ provide, by definition, the values of the bias of $\oplus_p$ for which there are at least two different equiprobable paths of tossings from $M$ to its normal form.
- Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log p,-\log(r-p))$ is a tropical root of, say, $\trop Q_{\true}$ are the values of the bias of $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are smapled by $M$ during its tossings to normal form, knowing that some $\true$ was sampled.
-\end{remark}
-
-{\color{red}due parole generalised metric spaces and modules}
-
-{\color{red}due parole su PCoh, $\overline{R_{\geq 0}}$Rel e le loro tropicalizzazioni (gia accennato in \autoref{rmk:tropof01Rel})}
+%{\color{red}due parole sul fatto che non sappiamo se si posso usare strumenti propri della tropical geo per lo studio dei programmi. Per esempio:}
+%\begin{remark}\label{rem:troproots}
+% Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_{\true}$ or of $\trop Q_{\false}$ provide, by definition, the values of the bias of $\oplus_p$ for which there are at least two different equiprobable paths of tossings from $M$ to its normal form.
+% Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log p,-\log(r-p))$ is a tropical root of, say, $\trop Q_{\true}$ are the values of the bias of $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are smapled by $M$ during its tossings to normal form, knowing that some $\true$ was sampled.
+%\end{remark}
+%
+%{\color{red}due parole generalised metric spaces and modules}
+%
+%
+%{\color{red}due parole su PCoh, $\overline{R_{\geq 0}}$Rel e le loro tropicalizzazioni (gia accennato in \autoref{rmk:tropof01Rel})}
 
@@
 
-While the first goal was here developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions (best-case analysis, log-probabilities, differential privacy).
-We believe that exploring these ideas in more depth could be a fruitful direction; moreover,
-since both generalized metrics and quantale-modules have been largely studied in computer science, 
-a natural question is if the generalized approach of Section \ref{section6} could  lead to new applications of metric and tropical methods to the $\lambda$-calculus.
+While the first goal was here developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions (best-case analysis, log-probabilities), and we leave their development to a second paper of this series.
+To conclude, while the main ideas of this article only use basic concepts from the toolbox of tropical mathematics, it looks like an exciting direction to look for applications of more advanced tools from tropical algebraic and differential geometry (e.g.~Newton polytopes, tropical varieties, tropical differential equations).
+%We believe that exploring these ideas in more depth could be a fruitful direction; moreover, 
+%since both generalized metrics and quantale-modules have been largely studied in computer science, 
+%a natural question is if the generalized approach of Section \ref{section6} could  lead to new applications of metric and tropical methods to the $\lambda$-calculus.
 
-Finally, since bounds on the Taylor expansion translate into Lipschitz conditions, 
- two interesting directions to explore are provided by (non-idempotent) intersection types and finiteness spaces \cite{Ehrhard2005}, as both methods are in principle capable of capturing \emph{finitary} bounds on the Taylor expansion.
+%Finally, since bounds on the Taylor expansion translate into Lipschitz conditions, 
+% two interesting directions to explore are provided by (non-idempotent) intersection types and finiteness spaces \cite{Ehrhard2005}, as both methods are in principle capable of capturing \emph{finitary} bounds on the Taylor expansion.
 %
-Notably, knowing that the application of a program $M$ to $N$ expands as a finite sum of linear applications may allow one to predict how sensitive $M$ will be ``around $N$''.
-
-\newpage
+%Notably, knowing that the application of a program $M$ to $N$ expands as a finite sum of linear applications may allow one to predict how sensitive $M$ will be ``around $N$''.

### 2023-07-29 CSL24/sectionLaw.tex
commit

As we have seen, the morphisms of $\LREL$ can be seen as continuous functions between the $\Lawv$-modules $\Lawv^{X}$, when the latter are taken with the metric induced by the $\infty$-norm. This viewpoint gives a metric flavor to $\LREL$, and allowed us to relate differential and metric structure. Yet, how far can this correspondence be pushed?
In particular, is this correspondence restricted to $\Lawv$-modules of the form $\Lawv^{X}$ (i.e.~with a fixed base), or does it hold in some sense for arbitrary $\Lawv$-modules? Is this correspondence restricted to the $\infty$-norm metric, or does it hold for other metrics too?

An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}, that it's worth recalling.

%In this section we first reconstruct this correspondence, by combining some {folklore} results with more abstract ones from recent literature in enriched category theory \cite{Fuji, Stubbe2006, Shen2014}. Then we show that this correspondence can be lifted to a model of the full differential $\lambda$-calculus, by suitably generalizing the construction of the Lafont exponential of $\LREL$.
%
%
%also extends exploit this correspondence to prove that arbitrary $\Lawv$-modules (equivalently, arbitrary \emph{complete} generalized metric spaces) provide a model of the differential $\lambda$-calculus, hence generalizing the tropical relational model.
%

%{$\Lawv$-modules}

On the one hand we have $\Lawv$-modules: these are triples $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$.% (notice that this is indeed the usual notion of module over the Lawvere quantale $\Lawv$).
A $\Lawv$-module homomorphism is a map $f:M\to N$ commuting with both joins and the $\Lawv$-action. We let $\Mod$ indicate the category of $\Lawv$-modules and their homomorphisms.

% 
% 
%
%$\Lawv$ is the most basic example of $\Lawv$-module.
%Any $\Lawv$-module $M$ has a dual $M^{\op}$, with reversed order and (right-)action $x\multimapinv \epsilon= \bigvee\{y\mid \epsilon \star y\succeq x\}$.
% Other basic examples of $\Lawv$-modules are the sets $\Lawv^{X}$, with order and action defined pointwise. 
% 
% 
% While the $\Lawv^{X}$ have a fixed base, for an arbitrary $\Lawv$-module one can retrieve a base via the \emph{Yoneda embedding}
%$\Yon: M \to \Mod(M^{\op},\Lawv)$, where $\Yon(x)(y)=\inf\{\epsilon\mid \epsilon\star y\succeq x\}$. 
%
%
%\begin{proposition}[cf.~\cite{Stubbe2014, Shen2014}]\label{prop:yonemod}
%For any $\Lawv$-module $M$, the Yoneda embedding has a left-adjoint $\sup(f)=\bigvee_{x\in M}f(x)\star x$.
%%\end{proposition}
%
%
%Like $\LREL$, the category $\Mod$ has the relevant structure to interpret the linear $\lambda$-calculus:
%\begin{proposition}
%$\Mod$ is a SMCC.
%\end{proposition}
%%\begin{proof}
%The hom-sets $\Mod(M,N)$ have a natural $\Lawv$-module structure, defined pointwise. The tensor product of $\Lawv$-modules $M\otimes N$ can be defined as the quotient of the usual tensor product of sup-lattices under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$ (see e.g.~\cite{Russo2007}).
%Notably, any element of $M\otimes N$ can be identified with a join of basic tensors $x\otimes y$, corresponding to the equivalence class of the pair $\langle x,y\rangle$.
%Beyond the required adjointness of the internal hom and the tensor, one can check that $\Mod$ is actually \emph{$^{*}$-autonomous}, since it satisfies $(M^{\op})^{\op}\simeq M$ and 
%$\Hom(M,N^{\op})\simeq (M\otimes N)^{\op}$.
%Finally, $\Mod$ has \emph{biproducts}: products and coproducts are both given by the Cartesian product of the underlying posets, with action defined pointwise.
%
%, similarly to the case of sup-lattices \cite{}, as the quotient of the free sup-lattice $\C P(M\times N)$ under the smallest congruence containing all pairs $((\bigvee A,y),\bigcup_{a\in A}\{(a,y)\})$, for $A\subseteq M, y\in N$, 
%$((x,\bigvee B),\bigcup_{b\in B}\{(x,b)\})$, for all $x\in A$, $B\subseteq N$, and all pairws $\{
%\end{proof}
%
%
%\begin{remark}
%The SMCC structure of $\Mod$ coincides with that of $\LREL$ for the modules $\Lawv^{X}$: we already know that $\Mod(\Lawv^{X},\Lawv^{Y})\simeq \Lawv^{X\times Y} $, and 
%one can prove $\Lawv^{X}\otimes \Lawv^{Y}\simeq \Lawv^{X\times Y} $ (cf.~\cite{Russo2007}).% and 
%%$\Pi_{i\in I}\Lawv^{X_{i}}\simeq \Lawv^{\coprod_{i\in I}X_{i}}$.
%\end{remark}

%\subsection{$\Lawv$-categories}
On the other hand we have Lawvere's \emph{generalized metric spaces}:
Lawvere was the first to observe that a metric space can be described as a \emph{$\Lawv$-enriched} category. Indeed, spelling out the definition, a $\Lawv$-enriched category (in short, a $\Lawv$-category) is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$ satisfying 
$0  \geq X(x,x)$ and $X(y,z)+X(x,y)&\geq  X(x,z)$. This structure is often referred to as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}. %, or as a \emph{quasi-metric space}. 
 A metric space, in the usual sense, is thus the same as a skeletal (i.e.~$X(x,y)=0$ implies $x=y$) and symmetric (i.e.~$X(x,y)=X(y,x)$) $\Lawv$-category.
Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, as functoriality reads as $Y(f(x),f(y))\leq X(x,y)$.

A $\Lawv$-category is \emph{complete} when all weighted colimits (cf.~Appendix for a precise definition) exist in $X$. This categorical notion of completeness can be seen as a generalization of the usual notion of Cauchy-completeness.
We let $\GMet$ indicate the category of complete $\Lawv$-categories and continuous (i.e.~colimit-preserving) $\Lawv$-enriched functors.

Now, it turns out that the categories $\Mod$ and $\GMet$ are isomorphic \cite{Stubbe2006}. First,  any $\Lawv$-module $(M,\preceq, \star)$ is a $\Lawv$-category via
$M(x,y) = \inf\left\{ \epsilon \mid \epsilon \star x\geq y\right\}$. Moreover, a homomorphism of $\Lawv$-modules is a functor of the associated $\Lawv$-categories. Conversely, complete $\Lawv$-categories are \emph{tensored}, that is, admit a $\Lawv$-module structure $\epsilon \otimes x$ (see the Appendix).
Moreover, a continuous functor between complete $\Lawv$-categories is the same as a homomorphism of the associated $\Lawv$-modules. 
%
%
%To conclude our correspondence between $\Lawv$-modules and complete $\Lawv$-categories, it remains to observe that the 
%two constructions leading from one structure to the other are one the inverse of the other: for any $\Lawv$-module $(M,\preceq,\star)$,
%$x\preceq_{M}y$ iff $M(y,x)=0$ iff $x\preceq y=0\star y$, and, from  
%$M(\epsilon \star x, y)= M(x,y)\dotdiv \epsilon$, we deduce $\epsilon\otimes x=\epsilon \star x$. 
%Conversely, 
%for any complete $\Lawv$-category $X$ and $x,y\in X$, one can check that 
%$X(y,x)=\inf\{ \epsilon \mid X(\epsilon\otimes y,x )=0\}$.
%
%This leads to the following:
%
%
%\begin{theorem}[cf.~\cite{Stubbe2006}]
%$\Mod$ and $\GMet$ are isomorphic categories.
%\end{theorem}

%
%Since $\Mod$ (and thus $\GMet$ too) is a SMCC, it is worth making its metric structure explicit. Given complete $\Lawv$-categories $X,Y$, we have that the distance on the hom-set $\Mod(X,Y)$ is given by 
%\begin{align}
%\Mod(X,Y)(f,g)= \sup_{x\in X}Y(f(x),g(x));
%\end{align}
%and the distance on $X\otimes Y$ is given by
%\begin{align}
%(X\otimes Y)(\alpha, \beta)=
%\sup_{i}\inf_{j}\left\{X(x_{i},x'_{j})+Y(y_{j},y'_{j}\right\},
%\end{align}
% where $\alpha=\bigvee_{i}x_{i}\otimes y_{i}$ and 
%$\beta=\bigvee_{j}x'_{j}\otimes y'_{j}$, 
%from which we deduce that, for basic tensors
%$x\otimes y, x'\otimes y'$, their distance is just $X(x,x')+Y(y,y')$.

%
%
%and is extended to arbitrary joins by continuity, i.e.~
%$(X\otimes Y)(\alpha, \beta)=
%\sup_{i}\inf_{j}X(x_{i},x'_{j})+Y(y_{j},y'_{j}),
%$, where $\alpha=\bigvee_{i}x_{i}\otimes y_{i}$ and 
%$\beta=\bigvee_{j}x'_{j}\otimes y'_{j}$.
%%and thus coincides with the sum-metric over basic tensors;
%\item the distance on the bi-product $\Pi_{i\in i}X_{i}$ is given by
%\begin{align}
%(\Pi_{i\in I}X_{i})(f,g)= \sup_{i\in I}X_{i}(f(x),g(x));
%\end{align}

%\end{itemize}

To conclude, we show how the correspondence $\Mod\simeq \GMet$ lifts to a model of the differential $\lambda$-calculus, extending the co-Kleisli category $\LREL_{!}$.

First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies2018, Manzo2013}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting suitable properties of $\Mod$, we may define $!$ as the product of the symmetric algebras.  
This ensures that the coKleisli category $\Mod_{!}$ is a CCC.
Moreover, the constructions for $\Mod$ generalize those of $\LREL$, in the sense that $!(\Lawv^{X})\simeq \Lawv^{\multiset(X)}$ and that $\Mod_{!}(\Lawv^{X},\Lawv^{Y})\simeq \LREL_{!}(X,Y)$.

Finally, general results {\color{red}CITARE} allow us to conclude 
\begin{theorem}
$\Mod_{!}\simeq\GMet_{!}$ is a CC$\partial$C, when equipped with 
the differential operator $E$ defined, for $f:!M\to N$, by 
$Ef(\alpha)=
\bigvee\left\{
f(\beta\cup [x]) \ \Big \vert  \ 
\iota_{n}(\beta)\otimes \iota_{1}(x) \leq S(\alpha)
\right\}
$, where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
and $S: !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{Mellies2018}.
\end{theorem}

### 2023-07-29 CSL24/sectionTaylip.tex
commit

% !TEX root = /Users/paolopistone/Documents/GitHub/tropicalnew/CSL24/main.tex

We now have all elements necessary to relate the metric and differential analysis of higher-order programs in the tropical relational model.

The key ingredient is the notion of Taylor expansion of $\lambda$-terms, that we now quickly recall. 
Let $M\cdot \langle N_{1},\dots, N_{k}\rangle$ be an abbreviation for
$\mathsf D^{n}[M]\cdot \langle N_{1},\dots, N_{k}\rangle$. A \emph{resource $\lambda$-term} is any term belonging to the grammar $t::= x \mid \lambda x.t \mid t\cdot \langle t,\dots, t\rangle$.
Observe that in a resource term applications are always \emph{bounded}: they may use an exact number of copies of their input.

The Taylor expansion of a $\lambda$-term is a set of resource terms $\Te{M}$ defined inductively as $\Te{x}=\{x\}$, $\Te{\lambda x.M}=\{\lambda x.t\mid t\in \Te{M}\}$ and 
$\Te{MN}=\{ t\cdot \langle u_{1},\dots, u_{k}\rangle \mid k\in \mathbb N, t\in \Te{M}, u_{i}\in \Te{N} \}$. 
Considering the term $M=zxx$ from Example \ref{ex:zxx}, all resource terms
 $t_{n,m}=(z\cdot x^{n})\cdot x^{m}$, for $n,m\in \mathbb N$, are in $\Te{M}$. 
Observe that the interpretation of $t_{n,m}$ yields a tropical polynomial
$\model{t_{n,m}}^{!}(x)(z)= y_{[n,m]}+(n+m)x$, rather than a tLs. 
However, resource term need not always yield a tropical polynomial. For instance, consider $y:(*\to*)\to (*\to*), x:(*\to *) \vdash t:  (*\to  *)$
with $t=y\cdot (y\cdot x)$. Then 
$\model{t}^{!}: \Lawv^{  \mathcal M_{\mathrm{fin}}(\mathbb N)\times \mathbb N}
\times \Lawv^{\mathbb N} \to \Lawv^{\mathbb N}$ is given by
$
\model{t}^{!}(y,x)_{i}= \inf_{m,n\in \mathbb N}\big\{    y_{[m],i}  +  y_{[n],m}+  x_{n}\}
$, 
which is not a polynomial. Yet, $\model{t}^{!}$ is Lipschitz, more precisely, 1-Lipschitz in $x$ and 2-Lipschitz in $y$. This is a general fact, as shown below.

%
%$\Te{MN}$ of $MN$ is the set $\set{\Der^{n}[t,u_1,\dots,u_n]0 \mid n\in\N,t\in\Te{M},u_i\in\Te{N}}$.
We have already shown that the tropical differential makes $\LREL_{!}$ a model of the differential $\lambda$-calculus. We now show that it also models the Taylor expansion (this needs not be true for \emph{any} CC$\partial$C).
First, it can be patiently checked that in $(\LREL_!,D)$ all morphisms can be Taylor expanded  (see \cite[Definition 4.22]{Manzo2012}):

\begin{theorem}\label{thm:modelsTaylor}
 For all $t\in\HOM{\LREL_!}{Z}{!X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$ we have:%, the evaluation of $t$ over $s$ yields 
 \begin{align}\label{eq:taylorcat}
  \RM{ev}\circ_!\langle t,s\rangle =
  \inf\limits_{n\in\N}
  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
 \end{align} 
\end{theorem}
%It is worth discussing the formula above a bit more. 
The equation above is a nothing more than a tropical version of the Taylor formula \eqref{eq:taylor}:
%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$.
%,  i.e.~it coincides with the tropical %interpretation of the
%version of the usual Taylor expansion.
%Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
Second, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of \cite[Theorem 4.23]{Manzo2012} shows:

\begin{corollary}\label{cor:T(M)=M}
 %For $\Gamma\vdash_{\STLC} M:A$, we have %$\model{\Gamma\vdash_{\STDLC} \Te M:A}:=
$\model{\Gamma\vdash_{\STLC} M:A}=\inf_{t\in\Te{M}} \model{\Gamma\vdash_{\STDLC} t:A}$. %the interpretation of the Taylor expansion of a $\STLC$-term $M$, given in \eqref{eq:taylor}, converges to the one of $M$.
\end{corollary}

Using the results of the previous section, as well as the results above, we can deduce the following properties:

\begin{theorem}
Let $\mathcal S\in \{\RM{PCF}^{\Lawv}, \STLC,\BSTLC,\STDLC\}$, $\Gamma\vdash_{\mathcal S}M:A$; for all $a\in \model{A}$,
\begin{enumerate}
\item if $\mathcal S=\BSTLC$, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is a tropical polynomial, and thus Lipschitz;

\item if $\mathcal S=\STDLC$ and $M$ is a resource term, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is Lipschitz;

\item if $\mathcal S=\STLC,\RM{PCF}^{\Lawv}$, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is locally Lipschitz;

\item if $\mathcal S=\STLC$, then $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!_{a}$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!_{a}$ of {Lipschitz} functions.
\end{enumerate}

\end{theorem}

%
%From this we deduce, inductively, the following fact:
%\begin{proposition}
%For any resource $\lambda$-term $t$, if $\Gamma\vdash_{\STDLC}t:A$, then $\model{\Gamma\vdash_{\STDLC}t:A}^{!}$ is a Lipschitz function.
%\end{proposition}
%
%
%This result shows that the interpretation of a program can be approximated via a family of tropical polynomials, and leads to.
% \begin{theorem}\label{thm:taylor}
% $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!_{a}$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!_{a}$ of \emph{tropical polynomials}, that is,of \emph{Lipschitz} functions.
%\end{theorem}

We conclude our discussion by presenting an interesting application of the Taylor expansion in $\LREL_{!}$: as proved in the previous section, all tLs are locally Lipschitz; now, Theorem \ref{thm:taylor} can be used to compute approximations of the Lipschitz constants of an actual higher-order program.

\begin{corollary}
Suppose $x: A \vdash_{\STLC}M:B$ and $\vdash_{\STLC} N:A$. 
Then for all $t\in \Te{M}$, with $t\neq \infty$, and $\delta>0$, the tLs $\model{ x:A\vdash_{\STLC}M:B}^{!}$ is $\frac{\model{t}(\model{N}+2\delta)}{\delta}$-Lipschitz over the open ball
 $B_{\delta}(\model{N})$.
\end{corollary}
{\color{red}ADD PROOF SKETCH}

\begin{example}
Consider again the term $M=zxx$ from Example \ref{ex:zxx}-
%, satisfying $x:*, z:(*\to*\to*) \vdash M:  *$. Since $\mathcal M_{\mathrm{fin}}(\{*\})\simeq \mathbb N$, we have that 
%
% $\model{M}^{!}: \Lawv \times \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}\to \Lawv$ is given by
 The (generalized) tLs $\model{M}^{!}(x)(y)= \inf_{n,n'\in \mathbb N}\{y_{[(n,n')]}+(n+n')x\}$ is not (globally) Lipschitz: for any 
$L>0$, choose a natural number $N>L$, let $Y\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ be such that $Y_{\mu}<\infty$ only if $\mu=[(n,n')]$ with $n+n'\geq N$; then $|\model{M}^{!}(x)(Y)- \model{M}^{!}(x+\epsilon)(Y)|\geq N\epsilon > L\epsilon$. 
Now take the approximant $t= (z\cdot \langle x^{N-1}\rangle) \cdot \langle x\rangle \in \Te{M}$ (chosen so that $\model t^{!}(x)(Y)<\infty$). Its interpretation is the monomial 
$\model{t}^{!}(x)(Y) = Y_{[(N-1,1)]}+Nx$. We can then compute a Lipschitz-constant for $\model{M}^{!}$ around $\langle x,Y\rangle$
as $\frac{1}{\delta}\model{t}^{!}(\langle x,Y\rangle+\delta)= N+2 + \frac{Y_{[(N-1,1)]}+Nx}{\delta}$. 
\end{example}
%    
%    %
%  Consider now $F\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ given by $F_{\mu}=0$ if $\mu=[(1,1)]$ and $F_{\mu}=\infty$ otherwise. 
%  Then $\model{M}^{!}(\langle x,F\rangle)=2x$
%%We show that we can compute a local 
%More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument. 
%Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
%For example, if $M$ is the term $\lambda fx.f(fx)$ considered before, $\Der[M]$ takes a first input $N$ and passes it linearly to $M$. Notice that there are two ways of doing so, corresponding to the two bound occurrences of $f$ in $M$: either by applying $N$ to $fx$, or by 
%applying $f$ linearly to $Nx$ (indeed, if $f$ were applied in an unrestricted way, it might duplicate $Nx$, so that $N$ would not be used linearly). This justifies the equation below, in which $\Der[M]$ is identified with the non-deterministic sum of the two possible linear choices:
%\begin{align}
%\Der\left[\lambda f x.f(fx)\right]\cdot N = 
%\lambda fx. N(fx) + \left(\Der[f]\cdot (Nx)\right)(fx)
%\end{align}
%More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
%
%
%The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy. \frac{\mathsf df(y)}{\mathsf dy}\cdot x$. Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
%Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:

%
%More generally, the relational semantics interprets unbounded programs as \emph{analytic functions}, that is, as functions admitting a representation as power series. For instance, observing that an analytic map $f: \BB R\to \BB R$, where $f(x)=\sum_{n}\widehat f_{n}\cdot x^{n}$ is uniquely determined by the sequence $\widehat f_{n}$, the program $M_{\infty}:=\lambda fx.fx: (\BB R\To \BB R)\To (\BB R\To \BB R)$ is represented by the power series below:
%\begin{align}
%F_{\infty}(f,x)= \sum_{n=0}^{\infty} \widehat f_{n} x^{n}
%\end{align}
%By restricting ourselves to bounded applications, the terms in the power series become finite, that is, the interpretation becomes a \emph{polynomial}: for instance, the program $M_{2}:=\lambda fx. \sum_{i=0}^{2}\Der^{(i)}[f]\cdot x^{i}$, corresponding to passing $x$ \emph{at most twice} to $f$, is represented by the polynomial
%\begin{align}
%F_{\leq 2}(f,x)=\widehat f_{2} x^{2}  + \widehat f_{1}x +  \widehat f_{0} 
%\end{align}
% In this framework the differential operator is naturally represented by formal differentiation of polynomials, where, as one would expect, 
% $\Der[\sum_{n}a_{n}x^{n}]=\sum_{n}\Der[a_{n}x^{n}]$ and $\Der[a_{0}x^{0}]=0$ and $\Der[a_{n+1}x^{n+1}]= (n+1)a_{n+1}x^{n}$, so that power series can be Taylor expanded. 
%
%Let us now see what the results proved in the previous section translate into, when referred to the interpretation of higher-order programs.
%Remark that the metric spaces $(\Lawv^{\mathlarger{+}_{i=1}^n X_i},d_{\infty})$ and $(\prod_{i=1}^n \Lawv^{X_i},\max_{i=1}^n d^{X_i}_\infty)$ are trivially isometric, so we identify them.

% Since $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}\in\HOM{\LREL_!}{\mathlarger{\&}_{i=1}^n \model{A_i}}{\model B}\simeq\Lawv^{\prod_{i=1}^n ! \model {A_i}\times \model B}$, the tLs $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ is defined by $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in ! \model{A_i}}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
%Finally, the interpretations of $\BSTLC$-terms are matrices $t\in\HOM{\LREL}{\bigotimes_{i=1}^n !_{n_i}X_i}{Y}=\Lawv^{\prod_{i=1}^n !_{n_i}X_i\times Y}$.In such situation, we define $t^!:\Lawv^{\mathlarger{+}_{i=1}^n X_{i}}\to\Lawv^Y \simeq \prod_{i=1}^n\Lawv^{X_{i}}\to\Lawv^Y$ as $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !_{n_i} X_i}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$. Clearly $t^!={\widetilde t}^!$, where $\widetilde t\in\Lawv^{\prod_{i=1}^n ! {X_i}\times Y}$ is the matrix $\widetilde t_{(\mu_1,\dots,\mu_n),b}:=t_{(\mu_1,\dots,\mu_n),b}$ if $\mu_i\in !_{n_i} X_i$ for all $i$, and $:=\infty$ otherwise, which has the same support as $t$.

%\begin{corollary}
%Let $\model A$ be a finite set.
%\begin{enumerate}
%\item $\model{\Gamma \vdash_{\BSTLC} M:B}^!:\prod\limits_{(x_i:_{n_i} A_i)\in\Gamma} \!\!\!\!\!\Lawv^{\model{A_i}} \to \Lawv^{\model B}$ is a \emph{tropical polynomial}, thus (as $\model A$ is finite), a \emph{Lipschitz} function.
%\item $\model{\Gamma \vdash_{\STLC} M:B}^!, \model{\Gamma \vdash_{\STLC_\oplus} M:A}^!:\prod\limits_{(x_i: A_i)\in\Gamma} \!\!\!\!\!\Lawv^{\model{A_i}} \to \Lawv^{\model B}$ are \emph{locally} Lipschitz maps.
%\item $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!$ of \emph{tropical polynomials}, thus (as $\model A$ is finite), \emph{Lipschitz} functions.
%\end{enumerate}
%\end{corollary} 
%\begin{proof}
%1). Since $\model A$ is finite, also $\model *$ is.
%Thus, as we already observed, the interpretation of a bounded term is a tropical polynomial.
%Now we apply Corollary~\ref{prop:polylip} to each coordinate of the image, and by taking the maximum Lipschitz constant among the finite number $\mathrm{Card}(\model A)$ of them, we obtain the thesis.
%2). It follows immediately from Theorem~\ref{thmTLSlocLip} and the fact that $\model A$ is finite.
%3). It follows from \autoref{cor:T(M)=M} plus the easily checked fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have $\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with $\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
%\end{proof}
%
%Remark that the restriction $\model A$ finite is without loss of generality, since by Currying all programs can be seen having type $*$, which is natural to interpret as a singleton.
%
%Remark that, if $Y$ is finite and $f:\Lawv^{\mathlarger{+}_{i=1}^n X_i}\to\Lawv^Y$ is $K$-Lipschitz, then $f:\prod_{i=1}^n \Lawv^{X_i}\to \Lawv^Y$ is $K$-Lipschitz on each of the $n$ variable separately.

%A consequence of (3) is that the pointwise distance between two interpretations of programs can always be bounded via Lipschitz tropical polynomial approximants of the initial two programs.
%\begin{corollary}
% Let $\Gamma \vdash_{\STLC} M:A$ and $\Delta\vdash_{\STLC} N:B$.
% For all $\epsilon>0$, $x\in\Lawv^{\model \Gamma}$, $b\in\model A$, there exist $t\in\Te{M}$, $u\in\Te{N}$ s.t.\ $\big| \model{\Gamma \vdash_{\STLC} M:A}^!(x)_b - \model{\Delta \vdash_{\STLC} N:B}^!(x)_b \big| \leq 2\epsilon + \big| \model{\Gamma \vdash_{\STDLC} t:A}^!(x)_b - \model{\Delta \vdash_{\STDLC} u:B}^!(x)_b \big|$.
%\end{corollary}
%
%

### 2023-07-29 CSL24/sectionLaw.tex
commit

@@
 Indeed, spelling out the definition, a $\Lawv$-enriched category (in short, a $\Lawv$-category) is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$ satisfying
-$0  \geq X(x,x)$ and $X(y,z)+X(x,y)&\geq  X(x,z)$.
+$0  \geq X(x,x)$ and $X(y,z)+X(x,y)\geq  X(x,z)$.
 This structure is often referred to as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}.

### 2023-07-29 CSL24/sectionTaylip.tex
commit

@@
 Suppose $x: A \vdash_{\STLC}M:B$ and $\vdash_{\STLC} N:A$. 
-Then for all $t\in \Te{M}$, with $t\neq \infty$, and $\delta>0$, the tLs $\model{ x:A\vdash_{\STLC}M:B}^{!}$ is $\frac{\model{t}(\model{N}+2\delta)}{\delta}$-Lipschitz over the open ball
+Then for all $t\in \Te{M}$ such that $\model{t}^{!}(\model{N})\neq \infty$, and $\delta>0$, the tLs $\model{ x:A\vdash_{\STLC}M:B}^{!}$ is $\frac{\model{t}^{!}(\model{N}+2\delta)}{\delta}$-Lipschitz over the open ball
  $B_{\delta}(\model{N})$.

### 2023-07-30 CSL24/section2bis.tex
Commit
Dettagli (più qualcosa da discutere quando ci sentiamo)

@@
 \begin{remark}\label{rmk:val_trop}
- There is a clear and well-known relation between tropical polynomial/power series and usual polynomials/power series. 
+ There is a well-known relation between tropical polynomial/power series and usual polynomials/power series. 
 Given $Q,L$ semirings with units and zeros, $L$ commutative idempotent, if $L$ is totally ordered by the partial order: $\alpha \preceq \beta$ iff $\alpha +  \beta = \beta$, then one defines a \emph{valuation} \cite{Izhakian2015} to be a map $\mathrm{val}:Q\to L$ s.t.\ $\mathrm{val}(0)=0$, $\mathrm{val}(1)=1$, $\mathrm{val}(ab)=\mathrm{val}(a)\mathrm{val}(b)$, $\mathrm{val}(a+b)\preceq \max\set{\mathrm{val}(a),\mathrm{val}(b)}$.
 The valuation which is constantly equal to $1\in L$ on all the invertibles of $Q$ is called the \emph{trivial valuation} $\mathrm{val}_1$.
-Another common valuation is $\mathrm{logval}: \overline{\mathbb R}_{\geq 0} \to [-\infty, \infty]$ given by $\mathrm{logval}(x)=-\log x$.
-Notice that $\mathrm{logval}$ restricts to a bijection between $[0,1]$ and $\Lawv$.
+%Notice that if $\mathrm{val}:Q\to ([0,+\infty],+,\cdot,0,1)$ is a valuation, then $\mathrm{logval}(x):Q\to ([-\infty,+\infty],\min,+,+\infty,0)$, $\mathrm{logval}(x):=-\log \mathrm{val}(x)$ is a valuation.
 

### 2023-07-30 CSL24/section2bisbis.tex
Commit
Dettagli (più qualcosa da discutere quando ci sentiamo)

@@
 Before discussing how full-scale higher-order programming languages can be interpreted in terms of tropical power series, we highlight how such functions may naturally arise in the study of effectful programming languages.
-We will see that, when considering probabilistic and non-deterministic programs, tropical tools can be used to describe \emph{optimal} properties of programs, that is, the behavior of programs in \emph{the best case}.
+We will see that, when considering probabilistic and non-deterministic programs, tropical tools can be used to describe \emph{optimal} properties of programs, that is, the behavior of programs in \emph{the best/worst case}.
 %Moreover, since, as we'll see, tropical semantics is also a \emph{metric} semantics, it can be used to study how much program behavior is \emph{sensitive} to errors, that is, how a small error in input may be increased in output.
@@
 For 1), the maximum values $x,y$ of $P_{rll}(p,q)$ can be computed by finding the \emph{minimum} values of $\mathsf tP_{rll}(-\log p, -\log q)= -2\log p- \log q$.
-Notice that $\mathsf tP_{rll}$ is precisely the \emph{negative log-probability} of the event ``$M\twoheadrightarrow_{rll} \false$''.
+Notice that the latter is precisely the \emph{negative log-probability} of the event ``$M\twoheadrightarrow_{rll} \false$''.
 For 2), the maximum values of $Q_{\true}(p,q): [0,1]^{2}\to [0,1]$ can be computed as $e^{-\alpha},e^{-\beta}$, where $\alpha,\beta\in[0,\infty]$ are the \emph{minimum} values of the tropical polynomial

### 2023-07-30 CSL24/section3fancy.tex
Commit
Dettagli (più qualcosa da discutere quando ci sentiamo)

@@
 
-What we here call the tropical relational model is indeed an instance of a general and well-known construction that yields, for any \emph{continuous} semiring $Q$, a model $\QREL_{!}$ of effectful extensions of the simply typed $\lambda$-calculus and $\mathrm{PCF}$.
-In this section we shortly recall this construction, and also discuss how it adapts to model \emph{graded} and \emph{differential} variants of $\STLC$.
+What we will call in this paper the tropical relational model is indeed an instance of a general and well-known construction that yields, for any \emph{continuous} semiring $Q$, a model $\QREL_{!}$ of effectful extensions of the simply typed $\lambda$-calculus and $\mathrm{PCF}$.
+Let us shortly recall this construction and then discuss how it adapts to model \emph{graded} and \emph{differential} variants of $\STLC$.
 In the next sections we will instead focus on the peculiar features of the tropical case.
 
-\subparagraph*{Weighted Relational Models}
-For any \emph{continuous} semiring $Q$ (i.e.~a cpo equipped with an order-compatible semiring structure), one can define a category $\QREL$ of ``$Q$-valued matrices'' which is well-known to yield a model of both probabilistic and non-deterministic versions of $\mathrm{PCF}$ (see e.g.~\cite{Manzo2013, Pagani2018}).
+\subparagraph*{As Linear/Non-Linear Algebra on Moduli}
+For any \emph{continuous} semiring $Q$ (i.e.~a cpo equipped with an order-compatible semiring structure), one can define a category $\QREL$ of ``$Q$-valued matrices''.
 More precisely, $\QREL$ (\cite{Manzo2013} calls it $Q^\Pi$) has sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms, i.e.~$\QREL(X,Y):=Q^{X\times Y}$.
@@
 Following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we are instead working with transpose matrices.
-
 $\QREL$ admits a comonad $!$ which acts, on objects, by taking the finite multisets.
@@
 For the moment, let us just observe that a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law.
-We will come back to this viewpoint in Section \ref{section6}.
+We will come back to this viewpoint in\autoref{sec:GMS}.
 
@@
  For $f:X\to \mathbb R$, define $f^*:Y\to \mathbb R$ by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
- Then for $X=!A$, $Y=\Lawv^A$, where $A$ is a set, and $\langle \mu, y \rangle:= \mu y$, we have that $f^!(y)=(-f)^*(-y)$ for all $f\in\Lawv^{!A}$.
+ Then for $X=!A$, $Y=\Lawv^A$, where $A$ is a set, and $\langle \mu, y \rangle:= \mu y$, we have $f^!(y)=(-f)^*(-y)$ for all $f\in\Lawv^{!A}$.
  This is precisely the same formal construction yielding the well-known convex conjugate $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
@@
 
-\subparagraph*{Interpreting effectful higher-order programs}
+\subparagraph*{As Tropical Weighted Relational Models for effectful higher-order programs}
+
+The categories $\QREL$ are well-known to yield a model of both probabilistic and non-deterministic versions of $\mathrm{PCF}$ (see e.g.~\cite{Manzo2013, Pagani2018}), which are called \emph{Weighted Relational Models}.
 The interpretation of the simply typed $\lambda$-calculus $\STLC$ in $\LREL$ relies on the fact that all categories $\QREL_{!}$ are cartesian closed \cite{Manzo2013}, with cartesian product and exponential objects acting on objects $X,Y$ as, respectively, $X+Y$ and $!X\times Y$.
-Hence, given an environment $\eta$ associating ground types with sets, any 
-typable term $\Gamma \vdash M:A$ gives rise to a morphism 
-$\model{M}_{\eta}\in \LREL_{!}(\model{\Gamma}_{\eta}, \model{A}_{\eta})$, and thus to a generalized tLs $\model{M}_{\eta}^{!}:\Lawv^{\model{\Gamma}_{\eta}}\to \Lawv^{\model{A}_{\eta}}$. 
+Hence, %given an environment $\eta$ associating ground types with sets, 
+any typable term $\Gamma \vdash M:A$ gives rise to a morphism 
+$\model{\Gamma \vdash M:A}\in \LREL_{!}(\model{\Gamma}, \model{A})$, and thus to a generalized tLs $\model{\Gamma \vdash M:A}^{!}:\Lawv^{\model{\Gamma}}\to \Lawv^{\model{A}}$. 
 This interpretation extends to $\mathrm{PCF}$ by interpreting the fixpoint combinator $\mathbf Y$ via the matrices $\mathrm{fix}^{X}= \inf_{n}\big\{\mathrm{fix}^{X}_{n}\big\} \in \Lawv^{  !(!X\times X) \times X  }$, where
@@
 As the homsets $\LREL_{!}(X,Y)$ are $\Lawv$-modules, it is possible to interpret in extensions of $\STLC$ and $\mathrm{PCF}$ comprising $\Lawv$-module operations $\alpha \cdot M$ and $M+N$ \cite{Manzo2013}, by
-letting $\model{\alpha\cdot M}_{\eta}=\model{M}_{\eta}+\alpha$ and 
-$\model{M+N}_{\eta}=\min\{\model{M}_{\eta},\model{N}_{\eta}\}$. 
+letting $\model{\Gamma \vdash \alpha\cdot M:A}=\model{\Gamma \vdash M:A}+\alpha$ and 
+$\model{\Gamma \vdash M+N:A}=\min\{\model{\Gamma \vdash M:A},\model{\Gamma \vdash N:A}\}$. 
 More precisely, \cite{Manzo2013} considers a language $\mathrm{PCF}^{Q}$ corresponding to $\mathrm{PCF}$ extended with $Q$-module operations, with an operational semantics given by rules $M\stackrel{1}{\to}M'$ for each rule $M\to M'$ of $\mathrm{PCF}$ (here $1$ is the monoidal unit of $Q$) as well as
@@
 \begin{proposition}
-For every closed term $M$ of type $\mathrm{Nat}$ in $\mathrm{PCF}^{\Lawv}$ and $n\in \mathbb N$, $\model{M}_{n}=  \inf\big\{ \mathsf w(\omega) \ \big \vert \ \omega : M \to \underline n\big \}.$
+If $\vdash_{\mathrm{PCF}^{\Lawv}} M:\mathrm{Nat}$, then $\model{\vdash_{\mathrm{PCF}^{\Lawv}} M:\mathrm{Nat}}_{n}=  \inf\big\{ \mathsf w(\omega) \ \big \vert \ \omega : M \to \underline n\big \}$ for all $n\in \mathbb N$.
 \end{proposition}
@@
 Observe that the two examples shown in Section \ref{section2} can easily be rephrased in the language $\mathrm{PCF}^{\Lawv}$. 
-For instance, for the probabilistic example, for a given valuation $\RM{val}$, one can use a translation $(M\oplus_{p}N)^{\RM{val}}= \min\{(M^{\RM{val}}+\RM{val}(p)), N^{\RM{val}}+\RM{val}(1-p)\}$, so that the reductions
-$:M\oplus_{p}N\stackrel{p}{\to} M$ and $M\oplus_{p}N\stackrel{p-1}{\to} N$ translate into a sequence of two reductions
-$(M\oplus_{p}N)^{\RM{val}} \stackrel{0}{\to} M^{\RM{val}}\stackrel{\RM{val}(p)}{\to} M$ 
-and $(M\oplus_{p}N)^{\RM{val}} \stackrel{0}{\to} N^{\RM{val}}\stackrel{\RM{val}(1-p)}{\to} M$, with the weight of the latter corresponding to the valuation of the weight of the former.
+For instance, for the probabilistic example, %for a given valuation $\RM{val}$, one can use a translation %$(M\oplus_{p}N)^{\RM{val}}= \min\{(M^{\RM{val}}+\RM{val}(p)), N^{\RM{val}}+\RM{val}(1-p)\}$, so that the reductions $:M\oplus_{p}N\stackrel{p}{\to} M$ and $M\oplus_{p}N\stackrel{p-1}{\to} N$ translate into a sequence of two reductions $(M\oplus_{p}N)^{\RM{val}} \stackrel{0}{\to} M^{\RM{val}}\stackrel{\RM{val}(p)}{\to} M$ and $(M\oplus_{p}N)^{\RM{val}} \stackrel{0}{\to} N^{\RM{val}}\stackrel{\RM{val}(1-p)}{\to} M$, with the weight of the latter corresponding to the valuation of the weight of the former.
+one can use the translation $(M\oplus_{p}N)^{\circ}= \min\{M^{\circ}+p, N^{\circ}+(1-p)\}$, so that the reductions $:M\oplus_{p}N\stackrel{p}{\to} M$ and $M\oplus_{p}N\stackrel{1-p}{\to} N$ translate into a sequence of two reductions $(M\oplus_{p}N)^{\circ} \stackrel{0}{\to} M^{\circ}\stackrel{p}{\to} M$ and $(M\oplus_{p}N)^{\circ} \stackrel{0}{\to} N^{\circ}\stackrel{1-p}{\to} N^{\circ}$.%, with the weight of the latter corresponding to the valuation of the weight of the former.
 
-Let $\mathrm{PPCF}$ (for \emph{probabilistic $\mathrm{PCF}$} \cite{Pagani2018}) be standard $\mathrm{PCF}$ extended with the constructor $M\oplus_{\frac{1}{2}}N$ and its associated reduction rules.
+Let $\mathrm{PPCF}$ (for \emph{probabilistic $\mathrm{PCF}$} \cite{Pagani2018}) be standard $\mathrm{PCF}$ extended with the constructor $M\oplus_{p}N$ ($p\in[0,1]$) and its associated reduction rules.
 From the above discussion we deduce the following:
@@
 \begin{corollary}
-For every closed probabilistic term $M$ of type $\mathrm{Nat}$ in $\mathrm{PPCF}$ and $n\in \mathbb N$, 
-$\model{M}^{\mathrm{logval}}_{n}\in \Lawv$ (cf.~Remark \ref{rmk:val_trop}) is the minimum negative log-probability of any
-reduction from $M$ to $\underline n$.
+%For every closed probabilistic term $M$ of type $\mathrm{Nat}$ in $\mathrm{PPCF}$ and $n\in \mathbb N$, 
+Let $\vdash_{\mathrm{PPCF}^{\Lawv}} M:\mathrm{Nat}$ and $n\in\N$.
+Considering its interpretation as a function of $p,1-p$, we have that $\model{\vdash_{\mathrm{PPCF}^{\Lawv}} M^{\circ}:\mathrm{Nat}}_{n}(-\log(p),-\log(1-p))$%\in \Lawv$ (cf.~Remark \ref{rmk:val_trop})
+is the minimum negative log-probability of any reduction from $M$ to $\underline n$, i.e.\ the negative log-probability of (any of) the (equiprobable) most likely reduction path from $M$.
 \end{corollary}
 
-\cite{Manzo2013} obtained a similar result for a non-deterministic version of PCF, by translating each term into $\mathrm{PCF}^{\Lawv}$via $(\lambda x.M)^{\circ}=\lambda x.M^{\circ}+1$ and $(\mathbf YM)^{\circ}= \mathbf Y(M^{\circ}+1)$
-(\cite{Manzo2013} considers the discrete tropical semiring $\mathbb N\cup\{\infty\}$, but the result obviously transports to $\Lawv$).
+Remark that this implies that all solution $p\in[0,1]$ to the equation $-\log w(\omega)= \model{\vdash_{\mathrm{PPCF}^{\Lawv}} M^{\circ}:\mathrm{Nat}}_{n}(-\log(p),-\log(1-p))$ are the values of the probabilistic parameter which make the reduction $\omega$ the most likely.
 
-\begin{corollary}[cf.~\cite{Manzo2013}, Corollary VI.10]
-For every closed probabilistic term $M$ of type $\mathrm{Nat}$ in non-deterministic $\mathrm{PCF}$ and $n\in \mathbb N$, 
-$\model{M^{\circ}}_{n}$ is the minimum number of $\beta$- and $\mathsf{fix}$- redexes reduced in a reduction sequence from $M $ to $\underline n$. 
-\end{corollary}
+Finally, \cite{Manzo2013} obtained a similar result for a non-deterministic version of PCF, by translating each term into $\mathrm{PCF}^{\Lawv}$via $(\lambda x.M)^{\circ}=\lambda x.M^{\circ}+1$ and $(\mathbf YM)^{\circ}= \mathbf Y(M^{\circ}+1)$
+(\cite{Manzo2013} considers the discrete tropical semiring $\mathbb N\cup\{\infty\}$, but the result obviously transports to $\Lawv$), and in that case \cite[Corollary VI.10]{Manzo2013} gives that $\model{\vdash M^{\circ}:\mathrm{Nat}}_{n}$ computes the minimum number of $\beta$- and $\mathsf{fix}$- redexes reduced in a reduction sequence from $M $ to $\underline n$.
+
+%\begin{corollary}[cf.~\cite{Manzo2013}, Corollary VI.10]
+%For every closed probabilistic term $M$ of type $\mathrm{Nat}$ in non-deterministic $\mathrm{PCF}$ and $n\in \mathbb N$, $\model{M^{\circ}}_{n}$ is the minimum number of $\beta$- and $\mathsf{fix}$- redexes reduced in a reduction sequence from $M $ to $\underline n$.
+%\end{corollary}
 

### 2023-07-30 CSL24/main.tex
commit
Commit
Dettagli (più qualcosa da discutere quando ci sentiamo)
Commit
Locally Lipschitz aggiustata.

@@
 
-\keywords{Differential lambda-calculus, Tropical semiring, Relational semantics, Lawvere quantale, Program metrics} %TODO mandatory; please add comma-separated list of keywords
+\keywords{Relational semantics, Differential lambda-calculus, Tropical semiring,  Program metrics, Lawvere quantale} %TODO mandatory; please add comma-separated list of keywords
 
@@
 
-\section{The Tropical Relational Model}\label{section3}
+\section{The category $\LREL$%The Tropical Relational Model
+}\label{section3}
 \input{section3fancy}
@@
 \input{section4}
+
+\section{Taylor Expansion and Lipschitz Approximation}
+\input{sectionTaylip}
+
+\section{Generalized Metric Spaces and $\Lawv$-Modules}\label{sec:GMS}
+\input{sectionLaw}
 

### 2023-07-30 CSL24/section4.tex
commit
Commit
Dettagli (più qualcosa da discutere quando ci sentiamo)
Commit
Locally Lipschitz aggiustata.

@@
 Remark that, since $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}\in\HOM{\LREL_!}{\mathlarger{\&}_{i=1}^n \model{A_i}}{\model B}\simeq\Lawv^{\prod_{i=1}^n !
-\model {A_i}\times \model B}$, the tLs $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ is $t^!(x^1,\dots,x^n)_b=\inf_{\mu_i\in !
+\model {A_i}\times \model B}$, the tLs $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ has shape $t^!(x^1,\dots,x^n)_b=\inf_{\mu_i\in !
 \model{A_i}}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
@@
 
+In all this section we consider the sets $\Lawv^X$ endowed with the usual sup-norm $\norm{x}_\infty:=\sup_{a\in X} x_a$.
+
 %
@@
 
-\subsection{Continuity of tLs}\label{subsec:cont}%$\Lawv^{X}$ as a normed cone}
+\subparagraph*{Continuity of tLs}\label{subsec:cont}%$\Lawv^{X}$ as a normed cone}
 
@@
 The tLs $\varphi$ is continuous on $\BB R_{\geq0}$ (w.r.t.\ the usual norm of real numbers).
-By considering the usual norm $\norm{x}_\infty:=\sup_{a\in X} x_a$ on $\Lawv^X$, we can generalise this property, dropping the case of $x$ having some $0$ coordinate:
+We can generalise this property, dropping the case of $x$ having some $0$ coordinate.
+But we have to be careful, because while in the finite dimensional $\BB R^n$, every real convex function is continuous because it is necessarily locally bounded from above (the sup-norm and the euclidean one are equivalent) \cite[Proposition 4.7]{Cobzas2017}, in infinite dimensions the former condition is no longer true \cite[Example 4.8]{Cobzas2017}.
+However, \cite[Proposition 4.4.(3)]{Cobzas2017} shows that it is the only requirement to ask: if a real-valued convex function with domain a convex open subset of a locally convex topological $\BB R$-vector space (LCTVS) is, locally around any point, bounded from above by a finite non-zero constant, then it is continuous on all its domain.
 
@@
 %It follows after adapting \cite[Proposition 4.4]{Cobzas2017} in order to prove that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
-\cite[Proposition 4.4.(3)]{Cobzas2017} shows that if a real-valued convex function with domain a convex open subset of a locally convex topological $\BB R$-vector space (LCTVS) is, locally around any point, bounded from above by a finite non-zero constant, then it is continuous on all its domain.
-Now, our $f$ is concave, so $-f$ is convex.
-Also, since $f\geq 0$ on all $\Lawv^X$, we have e.g.\ $-f\leq 1$ on $\Lawv^X$.
+%Now, our $f$ is concave, so 
+By \autoref{prop:nondecr+conc}, $-f$ is convex.
+Since $f\geq 0$ on all $\Lawv^X$, we have e.g.\ $-f\leq 1$ on $\Lawv^X$.
 %Remark that we cannot immediately apply \cite[Proposition 4.4.(3)]{Cobzas2017}, since $\Lawv^X$ is not even a $\BB R$-vector space.
@@
 
-\subsection{Lipschitz-continuity of tLs}\label{sec:4C}%$\Lawv^{X}$ as a metric space.}
+\subparagraph*{Lipschitz-continuity of tLs}\label{sec:4C}%$\Lawv^{X}$ as a metric space.}
 
@@
 As consequence of Theorem~\ref{theorem:fepsilon}, the tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $\BB R_{>0}$.
-Actually, we can prove a more general statement, also covering the infinitary case.
+%Actually, we can prove a more general statement, also covering the infinitary case.
+
+One could now obtain the locally Lipschitz property of also infinitely many variables tLs from the results in \cite{Cobzas2017}.
+In \cite[Theorem 6.4]{Cobzas2017} the locally Lipschitz property is obtained in general for LCTVS under the hypothesis of continuity.
+Instead, we replace the continuity hypothesis by the locally bounded condition in our next theorem, which corresponds to the one considered in \cite[Property 6.5]{Cobzas2017}.
+Our next theorem is basically a particular case of \cite[Theorem 6.9, Lemma 6.10]{Cobzas2017}, but we prefer instead to state it for a less general case and not talk about more complex argument involving LCTVS.
+
+\begin{theorem}\label{thm:locLip}
+ Let $f:V\subseteq (\BB R^X,d) \to (\BB R,\absv \cdot)$, with $V$ open and convex and $d$ any metric.
+ If $f$ is concave and locally bounded, then $f$ is locally Lipschitz.
+ Moreover, the Lipschitz constant of $f_{\big|_{\overline{B_{\delta}(x)}}}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} \absv f$.
+\end{theorem}
+\begin{proof}[Proof]
+ Call $\overline{B_{\delta}(x)}:=B_1$, $\overline{B_{3\delta}(x)}:=B_3$.
+ It suffices to show that for all $x\in V$, there is $\delta>0$ s.t.\ $B_3\subseteq \mathrm{interior}(V)$, $K:=\max_{B_3}  \absv f$ exists and $f_{\big|_{B_1}}$ is $(\frac{1}{\delta}\max_{B_3} \absv f)$-Lipschitz.
+A $\delta$ satisfying the first two conditions exists since $V$ is open and bcause $f$ is locally bounded and $B_3$ is compact.
+Now fix $y,z\in B_1$.
+Call $r:=\frac{d(y,z)}{2\delta}\in[0,1]$.
+We have to show that $\absv{f(y)-f(z)}\leq \frac{K}{\delta}d(y,z)=2Kr$.
+Wlog $y\neq z$.
+So $r\neq 0$ and we can consider $u:=\frac{1+r}{r}z-\frac{1}{r}y$, $v:=\frac{1}{r}y-\frac{r-1}{r}z$.
+We have $u,v\in \overline{B_{2\delta}(z)}=:B_2$ (geometrically, those are the intersections between $B_2$ and the segment between $y$ and $z$).
+Now we have the convex combinations $z=\frac{1}{1+r}y+\frac{r}{1+r}u$ and $y=(1-r)z+rv$, so the concavity of $f$ entails on one hand:
+$f(z)\geq \frac{1}{1+r}f(y)+\frac{r}{1+r}f(z)\geq \frac{f(y)}{1+r} - \frac{rK}{1+r}$, i.e.\ $f(y)-f(z)\leq r(K+f(z))\leq 2rK$, and on the other hand:
+$f(y)\geq (1-r)f(z)+rf(v)\geq f(z)-r(f(z)+K)$, i.e.\ $f(z)-f(y)\leq r(f(z)+K)\leq 2rK$.
+The last inequalities follow because $B_2\subseteq B_3$.
+Putting those together, we have the thesis.
+\end{proof}
+
+Now since $(0,\infty)^X$ is open and convex in $\BB R^X$, %and all tLs are non-negative, 
+we immediately have:
 
 \begin{theorem}\label{thmTLSlocLip}
- All tLs $\Lawv^X\to\Lawv$ are locally Lipschitz on $\BB R_{>0}^X$.
+ All tLs $f:\Lawv^X\to\Lawv$ are locally Lipschitz on $(0,\infty)^X$.
+ Moreover, the Lipschitz constant of $f_{\big|_{\overline{B_{\delta}(x)}}}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} f$.
 \end{theorem}
-\begin{proof}[Proof sketch]
-The core of the proof is a convex analysis argument (see the Appendix) showing that an arbitrary function $f:\Lawv^X\to\Lawv$ which is non-decreasing, concave and continuous, must be locally Lipschitz.
-\end{proof}
 
-Finally, let us discuss the differential structure.
+\subparagraph*{Tropical differential of tLs}
+
 The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tLs $f:\Lawv^{X}\to \Lawv^{Y}$ into a tLs $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by
@@
 \end{proposition}
+
+\begin{comment}
 
@@
 \end{corollary}
+
+\end{comment}

### 2023-07-30 CSL24/section2bisbis.tex
Commit
Dettagli vari

@@
 Consider the program
-$$
+$
  M:=(\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true)).
- $$
+ $
  Calling $q=1-p$, to each occurrence of $\true$ or $\false$ in $M$, univocally determined by an address
@@
 Then, any reduction $\omega: M \to^{*} N$ of a term to (one of its) normal form is  associated with a tropical monomial $P_{\omega}( \alpha,\beta)$ consisting of the sum of the costs of all reductions in $\omega$.
-For a given normal form $N$, the reductions $\omega_{i}: M \to^{*}N$ give rise to a tLs $\inf_{i\in I}P_{\omega_{i}}(\vec \alpha,\beta)$.
-For example, consider the non-deterministic term below
-$$
+For a given normal form $N$, the reductions $\omega_{i}: M \twoheadrightarrow N$ give rise to a tLs $\inf_{i\in I}P_{\omega_{i}}(\vec \alpha,\beta)$.
+For example, consider the non-deterministic term
+$
 M :=\mathsf{gen}+  ((\true + \true) + \mathsf{gen})
-$$
+$
 The (infinitely many) reduction paths leading to $\true$ can be grouped as follows:
@@
 This leads to the tLs 
-$$\varphi_{M\to^{*}\true}(\alpha)=\inf_{n\in \mathbb N}\big\{2\alpha+(n+1)\beta, 3\alpha, 3\alpha+(n+1)\beta\big \}= \min\{ 2\alpha+\beta, 3\alpha\}$$which describes all possible behaviors of $M$.
-Notice that, since $\alpha$ and $\beta$ are always positive, the power series $\varphi_{M\to^{*}\true}(\alpha)$ is indeed equivalent to the tropical polynomial $\min\{ 2\alpha+\beta, 3\alpha\}$.
+$\varphi_{M\twoheadrightarrow \true}(\alpha)=\inf_{n\in \mathbb N}\big\{2\alpha+(n+1)\beta, 3\alpha, 3\alpha+(n+1)\beta\big \}= \min\{ 2\alpha+\beta, 3\alpha\}$, which describes all possible behaviors of $M$.
+Notice that, since $\alpha$ and $\beta$ are always positive, the power series $\varphi_{M\twoheadrightarrow\true}(\alpha)$ is indeed equivalent to the tropical polynomial $\min\{ 2\alpha+\beta, 3\alpha\}$.
 In other words, of the infinitely many behaviors of $M$, only finitely many have chances to be \emph{optimal}: either left + gen + left, or right + left + (left,right).
@@
 
-Once reduced $\varphi_{M\to^{*}\true}$ to a polynomial, the best behavior among these will depend on the values of $\alpha$ and $\beta$, and by studying the tropical polynomial $\varphi_{M\to^{*}\true}$ one can thus answer questions analogous to 2) above, that is, what are the best choices of costs $\alpha,\beta$ making a \emph{chosen} reduction of $M$ to $\true$ the cheapest one?
+Once reduced $\varphi_{M\twoheadrightarrow\true}$ to a polynomial, the best behavior among these will depend on the values of $\alpha$ and $\beta$, and by studying the tropical polynomial $\varphi_{M\twoheadrightarrow\true}$ one can thus answer questions analogous to 2) above, that is, what are the best choices of costs $\alpha,\beta$ making a \emph{chosen} reduction of $M$ to $\true$ the cheapest one?

### 2023-07-30 CSL24/section3fancy.tex
Commit
Dettagli vari

@@
 Hence, any reduction $\omega: M\to^{*} N$ is naturally associated with a weight $\mathsf w(\omega)\in Q$.
-In particular, from [\cite{Manzo2013}, Theorem V.6] we deduce the following result, which relates the tropical semantics of a program with its best-case operational behavior.  
+In particular, from [\cite{Manzo2013}, Theorem V.6] we deduce the following adequation result:
 \begin{proposition}
@@
 
+The previous result allows to relate the tropical semantics of a program with its best-case operational behavior.  
 Observe that the two examples shown in Section \ref{section2} can easily be rephrased in the language $\mathrm{PCF}^{\Lawv}$. 

### 2023-07-30 CSL24/section4.tex
Commit
Dettagli vari

@@
 
-In all this section we consider the sets $\Lawv^X$ endowed with the usual sup-norm $\norm{x}_\infty:=\sup_{a\in X} x_a$.
-
 %
@@
 %\end{remark}
-
-\subparagraph*{Continuity of tLs}\label{subsec:cont}%$\Lawv^{X}$ as a normed cone}
 
@@
 
+In all the remainder of this section we consider the sets $\Lawv^X$ endowed with the usual sup-norm $\norm{x}_\infty:=\sup_{a\in X} x_a$.
+
+\subparagraph*{Continuity of tLs}\label{subsec:cont}%$\Lawv^{X}$ as a normed cone}
+
 The tLs $\varphi$ is continuous on $\BB R_{\geq0}$ (w.r.t.\ the usual norm of real numbers).
@@
 %The norm $\norm{\cdot}_\infty$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$.
-We will show that tLs satisfy suitable Lipschitz properties w.r.t.\ $\norm{\cdot}_\infty$.
+%We will show that tLs satisfy suitable Lipschitz properties w.r.t.\ $\norm{\cdot}_\infty$.
 
 Let us first look at tropical linear functions:
+in analogy with that happens in usual metric semantics, we have:
 
 \begin{proposition}\label{prop:troplinear}
-All tropical \emph{linear} functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive.  
+All tropical \emph{linear} functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive (i.e.\ $1$-Lipschitz).  
 \end{proposition}
@@
 %the problem reduces to checking that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof} 
-This result shows that, in analogy with that happens in usual metric semantics, linear programs are interpreted by non-expansive functions. 
 %\begin{proof}
@@
 the bijection between $\HOM{\LREL}{X}{Y}$ and set of tropical linear functions from $\Lawv^{X}$ to $\Lawv^{Y}$.
+Remember that $d_\infty(f,g):=\sup_{a\in X} d_\infty(f(x),f(y))$.
 
@@
 Let us now look at what happens with tLs, i.e.~when considering the full exponential $!$.
-As consequence of Theorem~\ref{theorem:fepsilon}, the tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $\BB R_{>0}$.
+As consequence of Theorem~\ref{theorem:fepsilon}, the tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $(0,\infty)^X$.
 %Actually, we can prove a more general statement, also covering the infinitary case.
 
-One could now obtain the locally Lipschitz property of also infinitely many variables tLs from the results in \cite{Cobzas2017}.
-In \cite[Theorem 6.4]{Cobzas2017} the locally Lipschitz property is obtained in general for LCTVS under the hypothesis of continuity.
-Instead, we replace the continuity hypothesis by the locally bounded condition in our next theorem, which corresponds to the one considered in \cite[Property 6.5]{Cobzas2017}.
-Our next theorem is basically a particular case of \cite[Theorem 6.9, Lemma 6.10]{Cobzas2017}, but we prefer instead to state it for a less general case and not talk about more complex argument involving LCTVS.
+We wish now to move to the case of infinitely many variables.
+In \cite[Theorem 6.4]{Cobzas2017} a locally Lipschitz property is obtained for locally convex topological vector spaces, under the hypothesis of continuity.
+\cite[Proposition 6.5]{Cobzas2017} shows that continuity is used in order to have a locally bounded condition, the true main ingredient of the proof.
+Instead of showing how our case fits into such theorems, we prefer to state a theorem, which is basically a particular case of \cite[Theorem 6.9, Lemma 6.10]{Cobzas2017}, already fitting in the topics under consideration in this paper.
 
-\begin{theorem}\label{thm:locLip}
+\begin{theorem}\label{thmTLSlocLip}
  Let $f:V\subseteq (\BB R^X,d) \to (\BB R,\absv \cdot)$, with $V$ open and convex and $d$ any metric.
  If $f$ is concave and locally bounded, then $f$ is locally Lipschitz.
- Moreover, the Lipschitz constant of $f_{\big|_{\overline{B_{\delta}(x)}}}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} \absv f$.
+ Moreover, the Lipschitz constant of $f$ on $\overline{B_{\delta}(x)}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} \absv f$.
+
+Therefore, since $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$, %and all tLs are non-negative, 
+we immediately have that all tLs $f:\Lawv^X\to\Lawv$ are locally Lipschitz on $(0,\infty)^X$.
+Moreover, the Lipschitz constant of $f$ on $\overline{B_{\delta}(x)}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} f$.
 \end{theorem}
-\begin{proof}[Proof]
- Call $\overline{B_{\delta}(x)}:=B_1$, $\overline{B_{3\delta}(x)}:=B_3$.
- It suffices to show that for all $x\in V$, there is $\delta>0$ s.t.\ $B_3\subseteq \mathrm{interior}(V)$, $K:=\max_{B_3}  \absv f$ exists and $f_{\big|_{B_1}}$ is $(\frac{1}{\delta}\max_{B_3} \absv f)$-Lipschitz.
-A $\delta$ satisfying the first two conditions exists since $V$ is open and bcause $f$ is locally bounded and $B_3$ is compact.
-Now fix $y,z\in B_1$.
-Call $r:=\frac{d(y,z)}{2\delta}\in[0,1]$.
-We have to show that $\absv{f(y)-f(z)}\leq \frac{K}{\delta}d(y,z)=2Kr$.
-Wlog $y\neq z$.
-So $r\neq 0$ and we can consider $u:=\frac{1+r}{r}z-\frac{1}{r}y$, $v:=\frac{1}{r}y-\frac{r-1}{r}z$.
-We have $u,v\in \overline{B_{2\delta}(z)}=:B_2$ (geometrically, those are the intersections between $B_2$ and the segment between $y$ and $z$).
-Now we have the convex combinations $z=\frac{1}{1+r}y+\frac{r}{1+r}u$ and $y=(1-r)z+rv$, so the concavity of $f$ entails on one hand:
-$f(z)\geq \frac{1}{1+r}f(y)+\frac{r}{1+r}f(z)\geq \frac{f(y)}{1+r} - \frac{rK}{1+r}$, i.e.\ $f(y)-f(z)\leq r(K+f(z))\leq 2rK$, and on the other hand:
-$f(y)\geq (1-r)f(z)+rf(v)\geq f(z)-r(f(z)+K)$, i.e.\ $f(z)-f(y)\leq r(f(z)+K)\leq 2rK$.
-The last inequalities follow because $B_2\subseteq B_3$.
-Putting those together, we have the thesis.
-\end{proof}
-
-Now since $(0,\infty)^X$ is open and convex in $\BB R^X$, %and all tLs are non-negative, 
-we immediately have:
-
-\begin{theorem}\label{thmTLSlocLip}
- All tLs $f:\Lawv^X\to\Lawv$ are locally Lipschitz on $(0,\infty)^X$.
- Moreover, the Lipschitz constant of $f_{\big|_{\overline{B_{\delta}(x)}}}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} f$.
-\end{theorem}
+%\begin{proof}[Proof]
+% Call $\overline{B_{\delta}(x)}:=B_1$, $\overline{B_{3\delta}(x)}:=B_3$.
+% It suffices to show that for all $x\in V$, there is $\delta>0$ s.t.\ $B_3\subseteq \mathrm{interior}(V)$, $K:=\max_{B_3}  \absv f$ exists and $f_{\big|_{B_1}}$ is $(\frac{1}{\delta}\max_{B_3} \absv f)$-Lipschitz.
+%A $\delta$ satisfying the first two conditions exists since $V$ is open and bcause $f$ is locally bounded and $B_3$ is compact.
+%Now fix $y,z\in B_1$.
+%Call $r:=\frac{d(y,z)}{2\delta}\in[0,1]$.
+%We have to show that $\absv{f(y)-f(z)}\leq \frac{K}{\delta}d(y,z)=2Kr$.
+%Wlog $y\neq z$.
+%So $r\neq 0$ and we can consider $u:=\frac{1+r}{r}z-\frac{1}{r}y$, $v:=\frac{1}{r}y-\frac{r-1}{r}z$.
+%We have $u,v\in \overline{B_{2\delta}(z)}=:B_2$ (geometrically, those are the intersections between $B_2$ and the segment between $y$ and $z$).
+%Now we have the convex combinations $z=\frac{1}{1+r}y+\frac{r}{1+r}u$ and $y=(1-r)z+rv$, so the concavity of $f$ entails on one hand:
+%$f(z)\geq \frac{1}{1+r}f(y)+\frac{r}{1+r}f(z)\geq \frac{f(y)}{1+r} - \frac{rK}{1+r}$, i.e.\ $f(y)-f(z)\leq r(K+f(z))\leq 2rK$, and on the other hand:
+%$f(y)\geq (1-r)f(z)+rf(v)\geq f(z)-r(f(z)+K)$, i.e.\ $f(z)-f(y)\leq r(f(z)+K)\leq 2rK$.
+%The last inequalities follow because $B_2\subseteq B_3$.
+%Putting those together, we have the thesis.
+%\end{proof}
 

### 2023-07-30 CSL24/section5bis.tex
Commit
Dettagli vari

@@
 
-\subparagraph*{Metric approach: bounded $\lambda$-terms as Lipschitz functions}
+\subparagraph*{Metric approach: bounded $\lambda$-calculus% as Lipschitz functions
+}
 
@@
 
-\subparagraph*{Resource approach: differential $\lambda$-terms as polynomials}
+\subparagraph*{Resource approach: the differential $\lambda$-calculus% as polynomials
+}
 
 A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models.
-The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus.
-The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M]\cdot N$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M]\cdot N$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).
-
+The key ingredient is a \emph{differential constructor} $\Der[\_,\_]$,  added to the usual syntax of the $\lambda$-calculus.
+The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M,N]$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M,N]$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).
 %Interestingly, the categorical study of the operator $\Der$ has led to the introduction of \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, a class of categories providing an abstract axiomatization of differentiation, in the usual mathematical sense.
@@
 %which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.
-
-An intriguing similarity between program derivatives and 
-actual derivatives is provided by the \emph{Taylor expansion} $\C T$:
-in $\STDLC$ one can expand any application $MN$ as an infinite formal sum of \emph{linear} applications
-$\Der^{(k)}[M]\cdot N^{k}$, i.e.~where $N$ is passed exactly $k$ times to $M$:
-\begin{align}\label{eq:taylor}
-\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot (\Der^{(k)}[M] \cdot N^{k})0
-\end{align}
-In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
+The reason why $\Der$ is called a ``differential'', is twofold: semantically, its interpretation is a generalisation of the usual differential (see \autoref{section3}); syntactically, it allows to define the so-called \emph{Taylor expansion} $\C T$ of programs:
+the idea is that, in $\STDLC$, one can expand any application $MN$ as an infinite formal sum of \emph{linear} applications
+$\Der^{k}[M,N^k]$, i.e.~where $N$ is linearly passed exactly $k$ times to $M$, and doing this recursively gives rise to the series $\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot \Der^{k}[\Te{M},\Te{N}^k]0$.
+In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.
+Actually, we will work with the so-called \emph{qualitative} Taylor expansion, where one quotients for idempotency of the $+$ operator in the series, and thus the Taylor expansion is just a set (see \autoref{sec:TayLip}).% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
 

### 2023-07-30 CSL24/sectionTaylip.tex
Commit
Dettagli vari

@@
 The key ingredient is the notion of Taylor expansion of $\lambda$-terms, that we now quickly recall. 
-Let $M\cdot \langle N_{1},\dots, N_{k}\rangle$ be an abbreviation for
-$\mathsf D^{n}[M]\cdot \langle N_{1},\dots, N_{k}\rangle$. A \emph{resource $\lambda$-term} is any term belonging to the grammar $t::= x \mid \lambda x.t \mid t\cdot \langle t,\dots, t\rangle$.
+Let $M\langle N_{1},\dots, N_{k}\rangle$ be an abbreviation for
+$\mathsf D^{k}[M, N_{1},\dots, N_{k}]0$. A \emph{resource $\lambda$-term} is any term belonging to the grammar $t::= x \mid \lambda x.t \mid t \langle t,\dots, t\rangle$.
 Observe that in a resource term applications are always \emph{bounded}: they may use an exact number of copies of their input.
@@
 The Taylor expansion of a $\lambda$-term is a set of resource terms $\Te{M}$ defined inductively as $\Te{x}=\{x\}$, $\Te{\lambda x.M}=\{\lambda x.t\mid t\in \Te{M}\}$ and 
-$\Te{MN}=\{ t\cdot \langle u_{1},\dots, u_{k}\rangle \mid k\in \mathbb N, t\in \Te{M}, u_{i}\in \Te{N} \}$. 
+$\Te{MN}=\{ t \langle u_{1},\dots, u_{k}\rangle \mid k\in \mathbb N, t\in \Te{M}, u_{i}\in \Te{N} \}$. 
 Considering the term $M=zxx$ from Example \ref{ex:zxx}, all resource terms
- $t_{n,m}=(z\cdot x^{n})\cdot x^{m}$, for $n,m\in \mathbb N$, are in $\Te{M}$. 
+ $t_{n,m}=z \,\langle x^{n}\rangle\,\langle x^{m}\rangle$, for $n,m\in \mathbb N$, are in $\Te{M}$. 
 Observe that the interpretation of $t_{n,m}$ yields a tropical polynomial
@@
 \begin{theorem}
-Let $\mathcal S\in \{\RM{PCF}^{\Lawv}, \STLC,\BSTLC,\STDLC\}$, $\Gamma\vdash_{\mathcal S}M:A$; for all $a\in \model{A}$,
+Let $\mathcal S$ be one of $\RM{PCF}^{\Lawv},\STLC,\BSTLC,\STDLC$. Let $\Gamma\vdash_{\mathcal S}M:A$ and $a\in \model{A}$.
 \begin{enumerate}
-\item if $\mathcal S=\BSTLC$, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is a tropical polynomial, and thus Lipschitz;
+\item For $\mathcal S=\BSTLC$, $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is a tropical polynomial, and thus Lipschitz;
 
-\item if $\mathcal S=\STDLC$ and $M$ is a resource term, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is Lipschitz;
+\item For $\mathcal S=\STDLC$, if $t$ is a resource term, then $\model{\Gamma\vdash_{\mathcal S}t:A}^{!}_{a}$ is Lipschitz;
 
-\item if $\mathcal S=\STLC,\RM{PCF}^{\Lawv}$, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is locally Lipschitz;
+\item For $\mathcal S=\STLC,\RM{PCF}^{\Lawv}$, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is locally Lipschitz;
 
-\item if $\mathcal S=\STLC$, then $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!_{a}$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!_{a}$ of {Lipschitz} functions.
+\item For $\mathcal S=\STLC$, $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!_{a}$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!_{a}$ of {Lipschitz} functions.
 \end{enumerate}
@@
 $L>0$, choose a natural number $N>L$, let $Y\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ be such that $Y_{\mu}<\infty$ only if $\mu=[(n,n')]$ with $n+n'\geq N$; then $|\model{M}^{!}(x)(Y)- \model{M}^{!}(x+\epsilon)(Y)|\geq N\epsilon > L\epsilon$.
-Now take the approximant $t= (z\cdot \langle x^{N-1}\rangle) \cdot \langle x\rangle \in \Te{M}$ (chosen so that $\model t^{!}(x)(Y)<\infty$). Its interpretation is the monomial 
+Now take the approximant $t= z \langle x^{N-1}\rangle \langle x\rangle \in \Te{M}$ (chosen so that $\model t^{!}(x)(Y)<\infty$). Its interpretation is the monomial 
 $\model{t}^{!}(x)(Y) = Y_{[(N-1,1)]}+Nx$. We can then compute a Lipschitz-constant for $\model{M}^{!}$ around $\langle x,Y\rangle$

### 2023-07-31 CSL24/main.tex
Commit
Dettagli vari
commit

@@
 
-\section{Tropical Power Series and Effectful Programs}\label{section2}
+\section{Tropical Polynomials and Power Series}\label{section2}
 \input{section2bis}
+
+\section{Tropical Semantics of Effectful Programs}\label{section22}
 \input{section2bisbis}
 
-\section{The category $\LREL$%The Tropical Relational Model
-}\label{section3}
+\section{Tropical Semantics of Higher-Order Programs}\label{section3}
 \input{section3fancy}
@@
 
-\section{Tropical Laurent Series}\label{sec:tls}
+\section{On Tropical Power Series}\label{sec:tls}
 \input{section4}
 
-\section{Taylor Expansion and Lipschitz Approximation}
+\section{Taylor Expansion and Lipschitz Approximation}\label{sec:TayLip}
 \input{sectionTaylip}

### 2023-07-31 CSL24/section1.tex
commit

@@
 
-In this paper we demonstrate that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provides the desired bridge between differential and metric semantics, and we suggest how this approach might be used to develop a best-case analysis of probabilistic and non-deterministic programs (an aspect that will be developed in more detail in a future paper of this series).
+In this paper we demonstrate that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provide the desired bridge between differential and metric semantics, and we suggest how this approach might be used to study probabilistic and non-deterministic programs (an aspect that will be developed in more detail in a future paper of this series).
 
@@
 
-Our contributions in this paper are threefold:
+Our contributions in this paper are the following:
 \begin{itemize}
 
-\item we show how tropical methods naturally arise in the \emph{best-case} analysis of probabilistic and non-deterministic programs, by studying a few very simple examples.
-This is in Section \ref{section2}.
+\item after recalling the notions of tropical polynomials and power series, we show how these objects naturally arise in the \emph{best-case} analysis of probabilistic and non-deterministic programs, by studying a few very simple examples.
+This is in Sections \ref{section2} and \ref{section22}.
 
-\item we study the relational model over the tropical semiring, which provides a semantics of effectful extensions of $\STLC$ and $\mathrm{PCF}$. Notably, 
- we show that the functions interpreting higher-order programs, which correspond to a generalization of \emph{tropical Laurent series} \cite{Porzio2021}, are locally Lipschitz-continuous, thus yielding a full-scale metric semantics for the $\lambda$-calculus and its bounded fragments.
-This is in Sections \ref{section3} and \ref{section4}.
+\item We study the relational model over the tropical semiring, which provides a semantics of effectful extensions of $\STLC$ and $\mathrm{PCF}$. Notably, 
+ we show that the functions interpreting higher-order programs, which correspond to a generalization of \emph{tropical power series} \cite{Porzio2021}, are locally Lipschitz-continuous and thus yield a full-scale metric semantics for the $\lambda$-calculus and its bounded fragments.
+This is in Sections \ref{section3} and \ref{sec:tls}.
 
  \item We exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps, and we show that it can be used to deduce approximated Lipschitz-constants for higher-order programs.
+This is in Section \ref{sec:TayLip}.
 
@@
 tropical, differential and metric viewpoints at the right level of generality.
-By recalling and suitably extending a well-known correspondence between Lawvere's \emph{generalized metric spaces} \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007}, we show that the category of \emph{complete} generalized metric spaces provides a model of the differential $\lambda$-calculus which extends the tropical relational model.
-This is in Section~\ref{section6}.
+We recall a well-known correspondence between Lawvere's \emph{generalized metric spaces} \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007} and we show that it yields a model of the differential $\lambda$-calculus which extends the tropical relational model.
+This is in Section~\ref{sec:GMS}.
 \end{itemize}

### 2023-07-31 CSL24/section2bis.tex
commit

@@
 
-In this section we recall tropical polynomials and power series and we discuss how these naturally arise in the study of \emph{optimal} behavior of probabilistic and non-deterministic programs.
+%In this section we recall tropical polynomials and power series and we discuss how these naturally arise in the study of \emph{optimal} behavior of probabilistic and non-deterministic programs.
 
@@
 %\footnote{In analogy with \cite{DanEhrh2011}, we could also call them \emph{tropical entire functions}.} 
-(tPs) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
-If $I$ is finite, $f$ is said to be a \emph{tropical polynomial}. 
+(tps) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
+If $I$ is finite, $f$ is called a \emph{tropical polynomial}. 
 \end{definition}
@@
 Unlike in standard algebra, tropical roots of tropical polynomials can be computed in linear time \cite{Noferini2015}.
+
 %Tropical polynomials and their roots are the main object of study of tropical geometry.
@@
 % simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root. 
+tps with more than one variable are defined as $f(x)=\inf_{n\in \mathbb N^{k}}\{nx + \matr f(n) \}$, where 
+$x\in \Lawv^{k}$, $n\in \mathbb N^{k}$, $nx$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients.
+In Section \ref{section3} we also consider tps with \emph{infinitely many} variables.
 
-While tropical polynomials are essentially combinatorial objects, this cannot be said for tPs: since $\inf$s are not in general $\min$s, a tPs is a ``limit'' of tropical polynomials of higher and higher degree, and its behavior is in general way more difficult to study than that of tropical polynomials~\cite{Porzio2021}.
+While tropical polynomials are essentially combinatorial objects, this cannot be said for tps: since $\inf$s are not in general $\min$s, a tps is a ``limit'' of tropical polynomials of higher and higher degree, and its behavior is in general way more difficult to study than that of tropical polynomials~\cite{Porzio2021}.
 %For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points of their domain.
@@
 %Remember that a_i \neq 0 by def of formal power series
-That is, it coincides to consider the same power series but with coefficients in $Q$ quotiented for the idempotency of $+$.
+That is, it corresponds to quotienting the coefficients in $Q$ for the idempotency of $+$.
 \end{remark}
+
+Observe that tropicalisation is \emph{not} in general an injective operation. 
+This is because tps have a tendency to collapse, if not globally at least locally, onto tropical polynomials.
+For instance, the tps $\varphi(x,y)=\inf_{n}\{2x+(n+1)y,3x\}$ is  equivalent to the polynomial $\min\{2x+y,3x\}$, since $2x+(n+1)y\geq 2x+y$ holds for all $n$.
+So the trivial valuation sends both the infinite series $\sum_{n}(x^{2}y^{n+1}+x^{3})$ and the polynomial $x^{2}y+x^{3}$ onto the same map.
+
+Similarly, by looking at Fig~\ref{fig:plot1} it appears that, \emph{far from $0$}, $\varphi$ behaves like some of the polynomials $\varphi_{n}$.
+In particular, %for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$. More, precisely, 
+$\varphi$ coincides on $[\epsilon,\infty]$ with $\varphi_{n}$,
+for $\epsilon \geq 2^{-(n+1)}$ (the smallest tropical root of $\varphi_{n}$).
+However, at
+%
+%$x\in [\epsilon,\infty]$, with $\epsilon>0$
+%It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
+ $x=0$ we have that $\varphi(x=0)=\inf_{n\in\N} 2^{-n}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
+Also, while the derivative of $f$ is bounded on all $(0,\infty)$, for $x\to 0^+$ it tends to $\infty$.
+In fact, this is a general phenomenon, as showed below:
+%shared by all tps with a finite number of variables, i.e.~of the form $f(x)=\inf_{n\in \mathbb N^{k}}\{nx + \matr f(n) \}$, where 
+%$x\in \Lawv^{k}$, $n\in \mathbb N^{k}$, $nx$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients. 
+% 
+%%This phenomenon is reminiscent of \cite[Example 7]{Ehrhard2005},
+%%%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
+%%which actually motivated our first investigations.
+%In fact, this behavior is shared by all tps with \emph{finitely} many variables, as shown by the following result (we identify $!\set{1,\dots,k}\simeq \N^k$, so the matrix of a tps $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one output variable) is given as a $\matr f:\N^k\to\Lawv$, $f$ having shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, with $nx$ the scalar product).
+
+\begin{theorem}\label{theorem:fepsilon}
+For all tLs $f(x_{1},\dots, x_{k})=\inf_{n\in \mathbb N^{k}}\{n x+\matr f(n)\}$, for all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
+% 
+% \begin{enumerate}
+%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
+%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
+  %\item 
+$f$ coincides on all $[\epsilon,\infty]^k$ with $P_\epsilon( x):=\min_{n\in \C F_\epsilon}\set{n x+\matr f(n)}$.
+% \end{enumerate}
+\end{theorem}
+\begin{proof}[Proof sketch]
+Let $\C F_\epsilon$ be the set of $n\in\N^{k}$ such that 
+$\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
+The core of the proof is showing that this set is indeed finite and sufficient for computing $f$.
+\end{proof}
+
+As we'll see, the potential of transforming infinitary objects (i.e.~tps) into combinatorial ones (i.e.~tropical polynomials), is one of the most interesting features of tropical semantics.
 

### 2023-07-31 CSL24/section2bisbis.tex
commit

@@
 $\omega\in \{l,r\}^{*}$, is associated a monomial $P_{\omega}(p,q)$ which determines the probability of the event ``$M\twoheadrightarrow_{\omega} \true/\false$'', that is, that $M$ reduces to $\true/\false$ according to the choices in $\omega$.
-Thinking of $p,q$ as parameters, $P_{\omega}(p,q)$ can thus be read as the \emph{likelihood function} of the event ``$M\twoheadrightarrow \true_\omega/\false_\omega$''.
+Thinking of $p,q$ as parameters, $P_{\omega}(p,q)$ can thus be read as the \emph{likelihood function} of the event ``$M\twoheadrightarrow_{\omega} \true/\false$''.
  For instance, we have
@@
 The polynomial function $Q_{\true}(p,q):=P_{ll}(p,q)+P_{rll}(p,q)+P_{rrr}(p,q)=p^2+p^2q+q^3$ gives instead the probability of the event ``$M\twoheadrightarrow \true$'', and analogously for $Q_{\false}(p,q):=P_{lr}(p,q)+P_{rrl}(p,q)+P_{rlr}(p,q)=pq+2pq^2$.
-This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baumr966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2ll4}.
+This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}.
 Typical questions in this case would be, for a fixed $\omega_0$:
@@
 \begin{enumerate}
- \item Which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow \false_{\omega_0}$''?
+ \item What is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0} \false$''?
  I.e., which is the choice of $p,q$ that maximizes the probability $P_{\omega_0}$?
  \item 
-Knowing that $M$ produced $\true$ (similarly for $\false$) , which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow \false_{\omega_0}$'', knowing that ``$M\twoheadrightarrow \false$''?
-I.e., which is the choice of $p,q$ that makes $\omega_0$ the most likely path among those leading to $\false$ (i.e.\ that maximizes the conditional probability $\BB P(``M\twoheadrightarrow \false_{\omega_0}'' \mid ``M\twoheadrightarrow \false'')$)?
+Knowing that $M$ produced $\true$ (similarly for $\false$) , which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0}\false$'', knowing that ``$M\twoheadrightarrow \false$''?
+I.e., which is the choice of $p,q$ that makes $\omega_0$ the most likely path among those leading to $\false$ (i.e.\ that maximizes the conditional probability $\BB P(``M\twoheadrightarrow_{\omega_0} \false'' \mid ``M\twoheadrightarrow \false'')$)?
 \end{enumerate}
@@
 
-While we here consider a very basic language, similar arguments could be done by replacing $\true$ and $\false$ by, respectively, a converging and diverging term (e.g.~in a $\B{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.
+Similar arguments could be done by replacing $\true$ and $\false$ by, respectively, a converging and diverging term (e.g.~in a $\mathrm{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.
 
@@
 This example is inspired from \cite{Manzo2013}.
-Again, we start from a very basic non-deterministic language with terms $M::= \true \mid  \mathsf{gen} \mid  M + M$, with an operation semantics comprising a non-deterministic reduction rule
+We consider now a basic non-deterministic language with terms $M::= \true \mid  \mathsf{gen} \mid  M + M$, with an operation semantics comprising a non-deterministic reduction rule
 $M_{1}+M_{2} \stackrel{\alpha}{\to} M_{i}$ and a generation rule
@@
 where in each case the value $\alpha,\beta\in \Lawv$ indicates a \emph{cost} associated with the reduction (e.g.~the estimated clock value for the simulation of each reduction on a given machine model).
-Then, any reduction $\omega: M \to^{*} N$ of a term to (one of its) normal form is  associated with a tropical monomial $P_{\omega}( \alpha,\beta)$ consisting of the sum of the costs of all reductions in $\omega$.
-For a given normal form $N$, the reductions $\omega_{i}: M \twoheadrightarrow N$ give rise to a tLs $\inf_{i\in I}P_{\omega_{i}}(\vec \alpha,\beta)$.
+Then, any reduction $\omega: M \twoheadrightarrow N$ of a term to (one of its) normal form is  associated with a tropical monomial $P_{\omega}( \alpha,\beta)$ consisting of the sum of the costs of all reductions in $\omega$.
+For a given normal form $N$, the reductions $\omega_{i}: M \twoheadrightarrow N$ give rise to a tps $\inf_{i\in I}P_{\omega_{i}}(\vec \alpha,\beta)$.
 For example, consider the non-deterministic term
@@
 \end{itemize}
-This leads to the tLs 
+This leads to the tps 
 $\varphi_{M\twoheadrightarrow \true}(\alpha)=\inf_{n\in \mathbb N}\big\{2\alpha+(n+1)\beta, 3\alpha, 3\alpha+(n+1)\beta\big \}= \min\{ 2\alpha+\beta, 3\alpha\}$, which describes all possible behaviors of $M$.
@@
 In other words, of the infinitely many behaviors of $M$, only finitely many have chances to be \emph{optimal}: either left + gen + left, or right + left + (left,right).
-This seems to be a very interesting feature of tropical semantics: reducing to best-case analysis leads, in many cases, to collapse the infinitary description of \emph{all} behaviors to a purely combinatorial description of the finitely many optimal ones.
+Hence, reducing to best-case analysis may lead, in several cases, to collapse the infinitary description of \emph{all} behaviors to a purely combinatorial description of the finitely many optimal ones.
 

### 2023-07-31 CSL24/section3fancy.tex
commit

@@
 
-What we will call in this paper the tropical relational model is indeed an instance of a general and well-known construction that yields, for any \emph{continuous} semiring $Q$, a model $\QREL_{!}$ of effectful extensions of the simply typed $\lambda$-calculus and $\mathrm{PCF}$.
-Let us shortly recall this construction and then discuss how it adapts to model \emph{graded} and \emph{differential} variants of $\STLC$.
-In the next sections we will instead focus on the peculiar features of the tropical case.
+There exists a general and well-known construction that yields, for any \emph{continuous} semiring $Q$, a model $\QREL_{!}$ of effectful extensions of the simply typed $\lambda$-calculus and $\mathrm{PCF}$.
+In this section we recall this construction and show how, in the case $Q=\Lawv$, it captures optimal program behavior; moreover, we discuss how it adapts to model \emph{graded} and \emph{differential} variants of $\STLC$.
+%In the next sections we will instead focus on the peculiar features of the tropical case.
 
-\subparagraph*{As Linear/Non-Linear Algebra on Moduli}
+\subparagraph*{Linear/Non-Linear Algebra on $Q$-Modules}
 For any \emph{continuous} semiring $Q$ (i.e.~a cpo equipped with an order-compatible semiring structure), one can define a category $\QREL$ of ``$Q$-valued matrices''.
@@
 
-When we instantiate $Q=\Lawv$, we obtain a category $\LREL$ of $\Lawv$-values matrices.
+When we instantiate $Q=\Lawv$, we obtain a category $\LREL$ of $\Lawv$-valued matrices.
 As one might expect, this category is tightly related to Lawvere's theory of (generalized) metric spaces.
 For the moment, let us just observe that a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law.
-We will come back to this viewpoint in\autoref{sec:GMS}.
+We will come back to this viewpoint in \autoref{sec:GMS}.
 
 Composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
-reads as \ $(st)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$, moreover, the non-linear functions $t^{!}: \Lawv^{X}\to \Lawv^{Y}$ give rise to functions of shape 
-$t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$, which correspond to the generalisation of tLs with \emph{possibly infinitely} many variables (in fact, as many as the elements of $X$).
-By identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the %functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ we recover the
-usual tLs's of one variable. %
+reads as \ $(st)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$, and the non-linear functions $t^{!}: \Lawv^{X}\to \Lawv^{Y}$ give rise to functions of shape 
+$t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$, which correspond to the generalisation of tps with \emph{possibly infinitely} many variables (in fact, as many as the elements of $X$).
+By identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tps generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the %functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ we recover the
+usual tps's of one variable. %
 %similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$, %induced by matrices
@@
  Let $X$, $Y$ be sets and let $\langle \_,\_\rangle:X\times Y \to \mathbb{R}$.
- For $f:X\to \mathbb R$, define $f^*:Y\to \mathbb R$ by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
+ For $f:X\to \mathbb R$, let $f^*:Y\to \mathbb R$ by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
  Then for $X=!A$, $Y=\Lawv^A$, where $A$ is a set, and $\langle \mu, y \rangle:= \mu y$, we have $f^!(y)=(-f)^*(-y)$ for all $f\in\Lawv^{!A}$.
- This is precisely the same formal construction yielding the well-known convex conjugate $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
+ This is the same formal construction yielding the \emph{convex conjugate} $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
  This construction is in turn a generalisation of the Legendre transformation.
@@
 
-\subparagraph*{As Tropical Weighted Relational Models for effectful higher-order programs}
+\subparagraph*{The $\Lawv$-Weighted Relational Model}
 
-The categories $\QREL$ are well-known to yield a model of both probabilistic and non-deterministic versions of $\mathrm{PCF}$ (see e.g.~\cite{Manzo2013, Pagani2018}), which are called \emph{Weighted Relational Models}.
+The categories $\QREL$ are well-known to yield a model of both probabilistic and non-deterministic versions of $\mathrm{PCF}$ (see e.g.~\cite{Manzo2013, Pagani2018}), which are called \emph{weighted relational models}.
 The interpretation of the simply typed $\lambda$-calculus $\STLC$ in $\LREL$ relies on the fact that all categories $\QREL_{!}$ are cartesian closed \cite{Manzo2013}, with cartesian product and exponential objects acting on objects $X,Y$ as, respectively, $X+Y$ and $!X\times Y$.
@@
 any typable term $\Gamma \vdash M:A$ gives rise to a morphism 
-$\model{\Gamma \vdash M:A}\in \LREL_{!}(\model{\Gamma}, \model{A})$, and thus to a generalized tLs $\model{\Gamma \vdash M:A}^{!}:\Lawv^{\model{\Gamma}}\to \Lawv^{\model{A}}$. 
+$\model{\Gamma \vdash M:A}\in \LREL_{!}(\model{\Gamma}, \model{A})$, and thus to a generalized tps $\model{\Gamma \vdash M:A}^{!}:\Lawv^{\model{\Gamma}}\to \Lawv^{\model{A}}$. 
+\begin{example}\label{ex:zxx}
+The evaluation morphism $\mathsf{ev}\in\LREL_{!}((!X\times Y) + X, Y)$ yields the tps  
+$\mathsf{ev}^{!}: \Lawv^{!X\times Y}\times \Lawv^{X}\to \Lawv^{Y}$ given by 
+$b\in Y \mapsto \mathsf{ev}^{!}(F,x)_{b}= \inf_{(\mu,b)}\{ F_{\mu,b}+ \mu x\}$.
+So, for instance, supposing the ground type $*$ of $\STLC$ is interpreted as the singleton set $\{*\}$, and recalling the identification $!\{*\}\simeq \mathbb N$, the interpretation of the term $x:*, z:(*\to*\to *) \vdash zxx$, involving two consecutive evaluations, yields the tps $\varphi:\Lawv\times \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}\to \Lawv$ given by $\varphi(x,z)=\inf_{n,n'}\{z_{[(n,n')]}+(n+n')x\}$.
+\end{example}
 This interpretation extends to $\mathrm{PCF}$ by interpreting the fixpoint combinator $\mathbf Y$ via the matrices $\mathrm{fix}^{X}= \inf_{n}\big\{\mathrm{fix}^{X}_{n}\big\} \in \Lawv^{  !(!X\times X) \times X  }$, where
@@
 % \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
-% For example, setting $\model{A}=\{\star\}$, the interpretation of the $\lambda$-term $\lambda x.\lambda y.xy: (A\to A)\to (A\to A)$ yields the tLs
+% For example, setting $\model{A}=\{\star\}$, the interpretation of the $\lambda$-term $\lambda x.\lambda y.xy: (A\to A)\to (A\to A)$ yields the tps
 % $\model{M}^{!}: \Lawv^{\mathbb N}\to \Lawv \to \Lawv$ (recalling that we can identify $\mathcal M_{\mathrm{fin}}(\{\star\})$ with $\mathbb N$) given by 
@@
 % Moreover,  
-As the homsets $\LREL_{!}(X,Y)$ are $\Lawv$-modules, it is possible to interpret in extensions of $\STLC$ and $\mathrm{PCF}$ comprising $\Lawv$-module operations $\alpha \cdot M$ and $M+N$ \cite{Manzo2013}, by
+As the homsets $\LREL_{!}(X,Y)$ are $\Lawv$-modules, it is possible to interpret in it  extensions of $\STLC$ and $\mathrm{PCF}$ comprising $\Lawv$-module operations $\alpha \cdot M$ and $M+N$ \cite{Manzo2013}, by
 letting $\model{\Gamma \vdash \alpha\cdot M:A}=\model{\Gamma \vdash M:A}+\alpha$ and 
@@
 $M_{1}+M_{2} \stackrel{1}{\to}M_{i}$ and $\alpha\cdot M \stackrel{\alpha}{\to} M $.
-Hence, any reduction $\omega: M\to^{*} N$ is naturally associated with a weight $\mathsf w(\omega)\in Q$.
+Hence, any reduction $\omega=\rho_{1}\dots \rho_{k}: M \twoheadrightarrow N$ is naturally associated with a weight $\mathsf w(\omega)=\sum_{i=1}^{k}\mathsf w(\rho_{i})\in Q$.
 In particular, from [\cite{Manzo2013}, Theorem V.6] we deduce the following adequation result:
 \begin{proposition}
-If $\vdash_{\mathrm{PCF}^{\Lawv}} M:\mathrm{Nat}$, then $\model{\vdash_{\mathrm{PCF}^{\Lawv}} M:\mathrm{Nat}}_{n}=  \inf\big\{ \mathsf w(\omega) \ \big \vert \ \omega : M \to \underline n\big \}$ for all $n\in \mathbb N$.
+$\model{\vdash_{\mathrm{PCF}^{\Lawv}} M:\mathrm{Nat}}_{n}=  \inf\big\{ \mathsf w(\omega) \ \big \vert \ \omega : M \to \underline n\big \}$ for all $n\in \mathbb N$.
 \end{proposition}
@@
 For instance, for the probabilistic example, %for a given valuation $\RM{val}$, one can use a translation %$(M\oplus_{p}N)^{\RM{val}}= \min\{(M^{\RM{val}}+\RM{val}(p)), N^{\RM{val}}+\RM{val}(1-p)\}$, so that the reductions $:M\oplus_{p}N\stackrel{p}{\to} M$ and $M\oplus_{p}N\stackrel{p-1}{\to} N$ translate into a sequence of two reductions $(M\oplus_{p}N)^{\RM{val}} \stackrel{0}{\to} M^{\RM{val}}\stackrel{\RM{val}(p)}{\to} M$ and $(M\oplus_{p}N)^{\RM{val}} \stackrel{0}{\to} N^{\RM{val}}\stackrel{\RM{val}(1-p)}{\to} M$, with the weight of the latter corresponding to the valuation of the weight of the former.
-one can use the translation $(M\oplus_{p}N)^{\circ}= \min\{M^{\circ}+p, N^{\circ}+(1-p)\}$, so that the reductions $:M\oplus_{p}N\stackrel{p}{\to} M$ and $M\oplus_{p}N\stackrel{1-p}{\to} N$ translate into a sequence of two reductions $(M\oplus_{p}N)^{\circ} \stackrel{0}{\to} M^{\circ}\stackrel{p}{\to} M$ and $(M\oplus_{p}N)^{\circ} \stackrel{0}{\to} N^{\circ}\stackrel{1-p}{\to} N^{\circ}$.%, with the weight of the latter corresponding to the valuation of the weight of the former.
-
+one can use the translation $(M\oplus_{p}N)^{\circ}= \min\{M^{\circ}+p, N^{\circ}+(1-p)\}$, so that the reductions $:M\oplus_{p}N\stackrel{p}{\to} M$ and $M\oplus_{p}N\stackrel{1-p}{\to} N$ translate into a sequence of two reductions $(M\oplus_{p}N)^{\circ} \stackrel{0}{\to} M^{\circ}\stackrel{p}{\to} M$ and $(M\oplus_{p}N)^{\circ} \stackrel{0}{\to} N^{\circ}\stackrel{1-p}{\to} N^{\circ}$.
+%, with the weight of the latter corresponding to the valuation of the weight of the former.
 Let $\mathrm{PPCF}$ (for \emph{probabilistic $\mathrm{PCF}$} \cite{Pagani2018}) be standard $\mathrm{PCF}$ extended with the constructor $M\oplus_{p}N$ ($p\in[0,1]$) and its associated reduction rules.
@@
 %For every closed probabilistic term $M$ of type $\mathrm{Nat}$ in $\mathrm{PPCF}$ and $n\in \mathbb N$, 
-Let $\vdash_{\mathrm{PPCF}^{\Lawv}} M:\mathrm{Nat}$ and $n\in\N$.
-Considering its interpretation as a function of $p,1-p$, we have that $\model{\vdash_{\mathrm{PPCF}^{\Lawv}} M^{\circ}:\mathrm{Nat}}_{n}(-\log(p),-\log(1-p))$%\in \Lawv$ (cf.~Remark \ref{rmk:val_trop})
+Let $\vdash_{\mathrm{PPCF}} M:\mathrm{Nat}$ and $n\in\N$.
+Considering its interpretation as a function of $p,1-p$, we have that $\model{\vdash_{\mathrm{PPCF}} M^{\circ}:\mathrm{Nat}}_{n}(-\log(p),-\log(1-p))$%\in \Lawv$ (cf.~Remark \ref{rmk:val_trop})
 is the minimum negative log-probability of any reduction from $M$ to $\underline n$, i.e.\ the negative log-probability of (any of) the (equiprobable) most likely reduction path from $M$.
@@
 
-Remark that this implies that all solution $p\in[0,1]$ to the equation $-\log w(\omega)= \model{\vdash_{\mathrm{PPCF}^{\Lawv}} M^{\circ}:\mathrm{Nat}}_{n}(-\log(p),-\log(1-p))$ are the values of the probabilistic parameter which make the reduction $\omega$ the most likely.
+Remark that this implies that all solution $p\in[0,1]$ to the equation $-\log \mathsf  w(\omega)= \model{\vdash_{\mathrm{PPCF}^{\Lawv}} M^{\circ}:\mathrm{Nat}}_{n}(-\log(p),-\log(1-p))$ are the values of the probabilistic parameter which make the reduction $\omega$ the most likely.
 
-Finally, \cite{Manzo2013} obtained a similar result for a non-deterministic version of PCF, by translating each term into $\mathrm{PCF}^{\Lawv}$via $(\lambda x.M)^{\circ}=\lambda x.M^{\circ}+1$ and $(\mathbf YM)^{\circ}= \mathbf Y(M^{\circ}+1)$
+\begin{remark}
+The function $\model{\vdash_{\mathrm{PPCF}} M^{\circ}:\mathrm{Nat}}_{n}(\alpha,\beta)$ is a tps, and Theorem \ref{theorem:fepsilon} ensures that this function coincides \emph{locally} with a tropical polynomial.
+This means that, for any choice of $p,1-p$, the most likely reduction path of $M$ can be searched for within a \emph{finite} space.
+\end{remark}
+
+Finally, \cite{Manzo2013} obtained a similar result for a non-deterministic version of PCF, by translating each term into $\mathrm{PCF}^{\Lawv}$ via $(\lambda x.M)^{\circ}=\lambda x.M^{\circ}+1$ and $(\mathbf YM)^{\circ}= \mathbf Y(M^{\circ}+1)$
 (\cite{Manzo2013} considers the discrete tropical semiring $\mathbb N\cup\{\infty\}$, but the result obviously transports to $\Lawv$), and in that case \cite[Corollary VI.10]{Manzo2013} gives that $\model{\vdash M^{\circ}:\mathrm{Nat}}_{n}$ computes the minimum number of $\beta$- and $\mathsf{fix}$- redexes reduced in a reduction sequence from $M $ to $\underline n$.
@@
 We now show how to interpret in $\LREL_{!}$ a graded version of $\STLC$, that we call $\BSTLC$, indeed a simplified version of the well-studied language $\mathrm{Fuzz}$ \cite{Reed2010}.
-The terms of $\BSTLC$ as for the $\STLC$, the types are $A::= * \ \mid  \ !_{n}A \multimap A$, the contexts of the typing judgements are sets of declarations of the form $x :_{n}A$, with $n\in \mathbb N$, and the typing rules are: %given in Fig.~\ref{fig:rules}
+The main feature of this language is that if $\vdash \lambda x.M:\,!_nA\multimap B$, then $x$ is duplicated \emph{at most} $n$ times in any reduction of $\vdash (\lambda x.M)N :B$ to the normal form.
+
+The types of $\BSTLC$ are $A::= * \ \mid  \ !_{n}A \multimap A$, the contexts of the typing judgements are sets of declarations of the form $x :_{n}A$, with $n\in \mathbb N$, and the typing rules are: %given in Fig.~\ref{fig:rules}
 	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
@@
 The axiom is $x:_{1}A\vdash x: A$.
-The main feature of this language is that if $\vdash \lambda x.M:\,!_nA\multimap B$, then $x$ is duplicated at most $n$ times in the reduction of $\vdash (\lambda x.M)N :B$ to the normal form.
 
@@
 As such, $(\LREL,(!_n)_{n\in\N})$ yields then a model of $\BSTLC$.
-Remark that since arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$.
+Remark that arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$.
 Notice that, whenever $\model *$ is finite, the set $\model{A}$ is finite for any type $A$ of $\BSTLC$.

### 2023-07-31 CSL24/section4.tex
commit

@@
-In this section we establish some topological and metric properties of such functions, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
+% !TEX root = /Users/paolopistone/Documents/GitHub/tropicalnew/CSL24/main.tex
+
+In this section we establish some topological and metric properties of tps, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
 The literature in tropical mathematics is often recent (e.g.~\cite{Porzio2021}), and several results we prove in this section are, to our knowledge, new.
-
-Remark that the map $(\_)^!$ is a surjective homomorphism (w.r.t.\ coKleisli composition and identity and function composition and identiy) on its image.
+%
+%Remark that the map $(\_)^!$ is a surjective homomorphism (w.r.t.\ coKleisli composition and identity and function composition and identiy) on its image.
 {\color{red}In Ehrhard-Danos fanno vedere (Lemma 14) che \`e anche iniettivo.
 Non riesco a capire se anche a noi lo \`e, ma non vedo perch\'e non dovrebbe.}
-Remark that, since $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}\in\HOM{\LREL_!}{\mathlarger{\&}_{i=1}^n \model{A_i}}{\model B}\simeq\Lawv^{\prod_{i=1}^n !
-\model {A_i}\times \model B}$, the tLs $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ has shape $t^!(x^1,\dots,x^n)_b=\inf_{\mu_i\in !
+%Remark that, since $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}\in\HOM{\LREL_!}{\mathlarger{\&}_{i=1}^n \model{A_i}}{\model B}\simeq\Lawv^{\prod_{i=1}^n !
+\model {A_i}\times \model B}$, the tps $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ has shape $t^!(x^1,\dots,x^n)_b=\inf_{\mu_i\in !
 \model{A_i}}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
-Same for $\STDLC$.
-
-\begin{remark}
-The interpretations of $\BSTLC$-terms are matrices $t\in\HOM{\LREL}{\bigotimes_{i=1}^n !_{n_i}X_i}{Y}=\Lawv^{\prod_{i=1}^n !_{n_i}X_i\times Y}$.
-In such situation, we define $t^!:\Lawv^{\mathlarger{+}_{i=1}^n X_{i}}\to\Lawv^Y \simeq \prod_{i=1}^n\Lawv^{X_{i}}\to\Lawv^Y$ as $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !_{n_i} X_i}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
-Clearly $t^!={\widetilde t}^!$, where $\widetilde t\in\Lawv^{\prod_{i=1}^n !
+%Same for $\STDLC$.
+%
+%\begin{remark}
+%The interpretations of $\BSTLC$-terms are matrices $t\in\HOM{\LREL}{\bigotimes_{i=1}^n !_{n_i}X_i}{Y}=\Lawv^{\prod_{i=1}^n !_{n_i}X_i\times Y}$.
+%In such situation, we define $t^!:\Lawv^{\mathlarger{+}_{i=1}^n X_{i}}\to\Lawv^Y \simeq \prod_{i=1}^n\Lawv^{X_{i}}\to\Lawv^Y$ as $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !_{n_i} X_i}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
+%Clearly $t^!={\widetilde t}^!$, where $\widetilde t\in\Lawv^{\prod_{i=1}^n !
 {X_i}\times Y}$ is the matrix $\widetilde t_{(\mu_1,\dots,\mu_n),b}:=t_{(\mu_1,\dots,\mu_n),b}$ if $\mu_i\in !_{n_i} X_i$ for all $i$, and $:=\infty$ otherwise, which has the same support as $t$.
-So, such $t^!$'s are special cases of tLs and thus we can \emph{Wlog} only consider tPs's.
-\end{remark}
+%So, such $t^!$'s are special cases of tps and thus we can \emph{Wlog} only consider tps's.
+%\end{remark}
 
@@
 %%Remark that we also find usual \emph{tropical polynomials} of tropical geometry as a particular case: they 
-%correspond to the tLs for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}. 
+%correspond to the tps for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}. 
 %Actually, for us \emph{tropical polynomial} will mean a function 
 %\end{remark}
+
+%
+%%In Example~\ref{ex:famous_ex}, 
+%Again by looking at Fig~\ref{fig:plot1} it appears that, \emph{far from $0$}, $\varphi$ behaves like some polynomials $\varphi_{n}$.
+%In particular, %for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$. More, precisely, 
+%$\varphi$ coincides on $[\epsilon,\infty]$ with $\varphi_{n}$,
+%for $\epsilon \geq 2^{-(n+1)}$ (the smallest tropical root of $\varphi_{n}$).
+%However, at
+%%
+%%$x\in [\epsilon,\infty]$, with $\epsilon>0$
+%%It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
+% $x=0$ we have that $\varphi(x=0)=\inf_{n\in\N} 2^{-n}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
+%Also, while the derivative of $f$ is bounded on all $(0,\infty)$, for $x\to 0^+$ it tends to $\infty$.
+%This phenomenon is reminiscent of \cite[Example 7]{Ehrhard2005},
+%%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
+%which actually motivated our first investigations.
+%In fact, this behaviour is shared by all tps with \emph{finitely} many variables, as shown by the following result (we identify $!\set{1,\dots,k}\simeq \N^k$, so the matrix of a tps $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one output variable) is given as a $\matr f:\N^k\to\Lawv$, $f$ having shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, with $nx$ the scalar product).
+%
+%\begin{theorem}\label{theorem:fepsilon}
+% Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tps with matrix $\matr f:\N^k\to\Lawv$.
+% For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
+%% 
+%% \begin{enumerate}
+%%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
+%%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
+%  %\item 
+%$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical \emph{polynomial} $P_\epsilon(x):=\min_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
+%% \end{enumerate}
+%\end{theorem}
+%\begin{proof}[Proof sketch]
+%Let $\C F_\epsilon$ be the set of $n\in\N^{k}$ such that 
+%$\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
+%The core of the proof is showing that this set is indeed finite and enough for computing $f$.
+%\end{proof}
+%
+In all the remainder of this section we consider the sets $\Lawv^X$ endowed with the usual sup-norm $\norm{x}_\infty:=\sup_{a\in X} x_a$.
+
+\subparagraph*{Continuity of tps}\label{subsec:cont}%$\Lawv^{X}$ as a normed cone}
 
@@
 \begin{proposition}\label{prop:nondecr+conc}
- Any tLs $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
+ All tps 
+ %$f:\Lawv^X\to\Lawv^Y$
+are non-decreasing and concave, w.r.t.\ the pointwise order on $\Lawv^{X}$.
 \end{proposition}
 
-%In Example~\ref{ex:famous_ex}, 
-Again by looking at Fig~\ref{fig:plot1} it appears that, \emph{far from $0$}, $\varphi$ behaves like some polynomials $\varphi_{n}$.
-In particular, %for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$. More, precisely, 
-$\varphi$ coincides on $[\epsilon,\infty]$ with $\varphi_{n}$,
-for $\epsilon \geq 2^{-(n+1)}$ (the smallest tropical root of $\varphi_{n}$).
-However, at
-%
-%$x\in [\epsilon,\infty]$, with $\epsilon>0$
-%It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
- $x=0$ we have that $\varphi(x=0)=\inf_{n\in\N} 2^{-n}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
-Also, while the derivative of $f$ is bounded on all $(0,\infty)$, for $x\to 0^+$ it tends to $\infty$.
-This phenomenon is reminiscent of \cite[Example 7]{Ehrhard2005},
-%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
-which actually motivated our first investigations.
-In fact, this behaviour is shared by all tLs with \emph{finitely} many variables, as shown by the following result (we identify $!\set{1,\dots,k}\simeq \N^k$, so the matrix of a tLs $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one output variable) is given as a $\matr f:\N^k\to\Lawv$, $f$ having shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, with $nx$ the scalar product).
-
-\begin{theorem}\label{theorem:fepsilon}
- Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\matr f:\N^k\to\Lawv$.
- For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
-% 
-% \begin{enumerate}
-%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
-%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
-  %\item 
-$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical \emph{polynomial} $P_\epsilon(x):=\min_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
-% \end{enumerate}
-\end{theorem}
-\begin{proof}[Proof sketch]
-Let $\C F_\epsilon$ be the set of $n\in\N^{k}$ such that 
-$\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
-The core of the proof is showing that this set is indeed finite and enough for computing $f$.
-\end{proof}
-
-In all the remainder of this section we consider the sets $\Lawv^X$ endowed with the usual sup-norm $\norm{x}_\infty:=\sup_{a\in X} x_a$.
-
-\subparagraph*{Continuity of tLs}\label{subsec:cont}%$\Lawv^{X}$ as a normed cone}
-
-The tLs $\varphi$ is continuous on $\BB R_{\geq0}$ (w.r.t.\ the usual norm of real numbers).
+The tps $\varphi$ is continuous on $\BB R_{\geq0}$ (w.r.t.\ the usual norm of real numbers).
 We can generalise this property, dropping the case of $x$ having some $0$ coordinate.
@@
 \begin{theorem}\label{thm:cont}
- All tLs $f:\Lawv^X\to\Lawv$ are continuous on $(0,\infty)^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
+ All tps $f:\Lawv^X\to\Lawv$ are continuous on $(0,\infty)^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
 \end{theorem}
@@
  All monotone (w.r.t.\ pointwise order) and $\norm{\cdot}_{\infty}$-continuous functions $f:(0,\infty)^X\to (0,\infty)$ are Scott-continuous.
- In particular, all tLs $f:\Lawv^X\to\Lawv$ are Scott-continuous on $(0,\infty)^X$ w.r.t.\ the pointwise orders.
+ In particular, all tps $f:\Lawv^X\to\Lawv$ are Scott-continuous on $(0,\infty)^X$ w.r.t.\ the pointwise orders.
 \end{theorem}
@@
 
-\subparagraph*{Lipschitz-continuity of tLs}\label{sec:4C}%$\Lawv^{X}$ as a metric space.}
+\subparagraph*{Lipschitz-continuity of tps}\label{sec:4C}%$\Lawv^{X}$ as a metric space.}
 
 %The norm $\norm{\cdot}_\infty$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$.
-%We will show that tLs satisfy suitable Lipschitz properties w.r.t.\ $\norm{\cdot}_\infty$.
+%We will show that tps satisfy suitable Lipschitz properties w.r.t.\ $\norm{\cdot}_\infty$.
 
-Let us first look at tropical linear functions:
-in analogy with that happens in usual metric semantics, we have:
+Let us first look at what happens with those tps which are either \emph{linear} or obtained via bounded exponentials.
+The result below is in analogy with that happens in the usual metric semantics of Fuzz, where 
+linear functions are non-expansive and $n$-bounded functions are $n$-Lipschitz \cite{Reed2010}.
 
 \begin{proposition}\label{prop:troplinear}
-All tropical \emph{linear} functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive (i.e.\ $1$-Lipschitz).  
+\begin{enumerate}
+\item If a tps $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:X\times Y\to \Lawv$ (i.e.~it is linear), then $f$ is non-expansive (i.e.\ $1$-Lipschitz).  
+\item If  $f \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f: \ !_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
+\end{enumerate}
 \end{proposition}
-%\begin{proof}[Proof sketch]
-%Using the fact that $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
-%the problem reduces to checking that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof} 
+\begin{proof}[Proof sketch]
+1). Using the fact that $f( x)_{b}= \inf_{a\in X}\{\matr f_{a,b}+ x_{a}\}$,
+the problem reduces to checking that $|(\matr f_{a,b}- x_{a})- (\matr f_{a,b}- y_{a})| = |x_{a}- y_{a}|\leq \|  x- y\|_{\infty}$.
+
+2.) Follows from 1.~and the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $!
+x$ to $\C M_{\leq n}(X)$.%
+\end{proof} 
 %\begin{proof}
@@
 
-Let us now consider the case of bounded exponentials:
-\begin{proposition}\label{prop:boundedlip}
-If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
-\end{proposition}
-\begin{proof}[Proof sketch]
-This follows from Proposition \ref{prop:troplinear} and the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $!
+%Let us now consider the case of bounded exponentials:
+%\begin{proposition}\label{prop:boundedlip}
+%If a tps $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
+%\end{proposition}
+%\begin{proof}[Proof sketch]
+%This follows from Proposition \ref{prop:troplinear} and the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $!
 x$ to $\C M_{\leq n}(X)$.%
-%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
-%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
-%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
-\end{proof}
-This result is perfectly analogous to what happens in the metric models discussed in Section \ref{section2}, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
+%%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
+%%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
+%%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
+%\end{proof}
+%This result is perfectly analogous to what happens in the metric models discussed in Section \ref{section2}, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
 
 Observe that, for any tropical polynomial $\varphi:\Lawv^{X}\to \Lawv^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \Lawv$ (as a monomial $\mu_ix+c_{i}$ yields a matrix entry on $!_{\#\mu_i}X\times Y$).
-Hence, using Proposition \ref{prop:boundedlip}, we have:
+Hence, using Proposition \ref{prop:troplinear} 2., we have:
 \begin{corollary}\label{prop:polylip}
@@
 
-Let us now look at what happens with tLs, i.e.~when considering the full exponential $!$.
-As consequence of Theorem~\ref{theorem:fepsilon}, the tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $(0,\infty)^X$.
-%Actually, we can prove a more general statement, also covering the infinitary case.
-
-We wish now to move to the case of infinitely many variables.
+Let us now look at what happens with tps, i.e.~when considering the full exponential $!$.
+First observe that, as consequence of Theorem~\ref{theorem:fepsilon}, the tps with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $(0,\infty)^X$.
+Yet, for the case of infinitely many variables we need a more general argument.
 In \cite[Theorem 6.4]{Cobzas2017} a locally Lipschitz property is obtained for locally convex topological vector spaces, under the hypothesis of continuity.
 \cite[Proposition 6.5]{Cobzas2017} shows that continuity is used in order to have a locally bounded condition, the true main ingredient of the proof.
-Instead of showing how our case fits into such theorems, we prefer to state a theorem, which is basically a particular case of \cite[Theorem 6.9, Lemma 6.10]{Cobzas2017}, already fitting in the topics under consideration in this paper.
+Instead of showing how our case fits into such theorems, we prefer to state a theorem, which is basically a particular case of \cite[Theorem 6.9, Lemma 6.10]{Cobzas2017}:
+%, already fitting in the topics under consideration in this paper.
 
 \begin{theorem}\label{thmTLSlocLip}
- Let $f:V\subseteq (\BB R^X,d) \to (\BB R,\absv \cdot)$, with $V$ open and convex and $d$ any metric.
- If $f$ is concave and locally bounded, then $f$ is locally Lipschitz.
- Moreover, the Lipschitz constant of $f$ on $\overline{B_{\delta}(x)}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} \absv f$.
-
-Therefore, since $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$, %and all tLs are non-negative, 
-we immediately have that all tLs $f:\Lawv^X\to\Lawv$ are locally Lipschitz on $(0,\infty)^X$.
+%Let $f:V\subseteq (\BB R^X,d) \to (\BB R,\absv \cdot)$, with $V$ open and convex and $d$ any metric.
+%If $f$ is concave and locally bounded, then $f$ is locally Lipschitz.
+%Moreover, the Lipschitz constant of $f$ on $\overline{B_{\delta}(x)}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} \absv f$.
+%Therefore, since $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$, %and all tps are non-negative, 
+%we immediately have that 
+All tps $f:\Lawv^X\to\Lawv$ are locally Lipschitz on $(0,\infty)^X$.
 Moreover, the Lipschitz constant of $f$ on $\overline{B_{\delta}(x)}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} f$.
 \end{theorem}
+\begin{proof}
+By observing that $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$, we apply the result that for all $f:V\subseteq (\BB R^X,d) \to (\BB R,\absv \cdot)$ concave and locally bounded, with $V$ open and convex and $d$ any metric, $f$ is locally Lipschitz, with the stated Lipschitz constant on $\overline{B_{\delta}(x)}$.
+%
+% Moreover, the Lipschitz constant of $f$ on $\overline{B_{\delta}(x)}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} \absv f$.
+\end{proof}
 %\begin{proof}[Proof]
@@
 
-\subparagraph*{Tropical differential of tLs}
+\subparagraph*{Tropical differential of tps}
+We conclude this section with a few observation on the differential operator over tps.
+The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tps $f:\Lawv^{X}\to \Lawv^{Y}$ into a tps $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by
+$D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a}+x_{a}+\mu y\right\}$.
+One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
+Moreover, the Taylor formula \eqref{eq:taylorcat} yields a ``tropical'' Taylor formula for tps of the form 
+$f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}$.
 
-The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tLs $f:\Lawv^{X}\to \Lawv^{Y}$ into a tLs $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by
-\begin{equation}
-D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a}+x_{a}+\mu y\right\}
-\end{equation}
-One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
-Moreover, the Taylor formula \eqref{eq:taylorcat} yields a ``tropical'' Taylor formula for tLs of the form 
-\begin{equation}
-f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}
-\end{equation}
 The following result shows that the distance between two tropical maps can be approximated using the terms appearing in their Taylor expansions:
 \begin{proposition}
-For all tLs $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
+For all tps $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
 the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz. Moreover 
@@
 % Since $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}\in\HOM{\LREL_!}{\mathlarger{\&}_{i=1}^n \model{A_i}}{\model B}\simeq\Lawv^{\prod_{i=1}^n !
-\model {A_i}\times \model B}$, the tLs $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ is defined by $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !
+\model {A_i}\times \model B}$, the tps $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ is defined by $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !
 \model{A_i}}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.

### 2023-07-31 CSL24/section5bis.tex
commit

@@
 In many situations (e.g.~when dealing with computationally difficult problems) one does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound.
-In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, a small error in the input will produce a comparably small error in the output.
-
+In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, that a small error in the input will produce a comparably small error in the output.
 In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which accounts for differences in behavior.
@@
 as a limit of Lipschitz maps, in some sense, thus bridging the metric and differential approaches.  
-Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
+Here, a natural direction to look for is the \emph{weighted relational semantics} \cite{Manzo2013}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
 However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.
@@
 Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions?
-This way, \eqref{eq:taylor} could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps.
+This way, the Taylor formula could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps.
 In other words, unbounded term application could be seen

### 2023-07-31 CSL24/sectionLaw.tex
commit

@@
 Moreover, a homomorphism of $\Lawv$-modules is a functor of the associated $\Lawv$-categories.
-Conversely, complete $\Lawv$-categories are \emph{tensored}, that is, admit a $\Lawv$-module structure $\epsilon \otimes x$ (see the Appendix).
+Conversely, complete $\Lawv$-categories are \emph{tensored} (cf.~\cite{Stubbe2014}), and thus admit a $\Lawv$-module structure $\epsilon \otimes x$ (see the Appendix).
 Moreover, a continuous functor between complete $\Lawv$-categories is the same as a homomorphism of the associated $\Lawv$-modules. 

### 2023-07-31 CSL24/sectionTaylip.tex
commit

@@
 
-The key ingredient is the notion of Taylor expansion of $\lambda$-terms, that we now quickly recall. 
-Let $M\langle N_{1},\dots, N_{k}\rangle$ be an abbreviation for
-$\mathsf D^{k}[M, N_{1},\dots, N_{k}]0$. A \emph{resource $\lambda$-term} is any term belonging to the grammar $t::= x \mid \lambda x.t \mid t \langle t,\dots, t\rangle$.
-Observe that in a resource term applications are always \emph{bounded}: they may use an exact number of copies of their input.
-
-The Taylor expansion of a $\lambda$-term is a set of resource terms $\Te{M}$ defined inductively as $\Te{x}=\{x\}$, $\Te{\lambda x.M}=\{\lambda x.t\mid t\in \Te{M}\}$ and 
-$\Te{MN}=\{ t \langle u_{1},\dots, u_{k}\rangle \mid k\in \mathbb N, t\in \Te{M}, u_{i}\in \Te{N} \}$. 
-Considering the term $M=zxx$ from Example \ref{ex:zxx}, all resource terms
+The key ingredient is the notion of Taylor expansion $\Te{M}$ of a $\lambda$-term $M$, which is a set of terms of the differential $\lambda$-calculus defined as follows:
+%
+%
+%
+% which is defined as follows: 
+%
+%that we now quickly recall. 
+%
+%
+%Let $M\langle N_{1},\dots, N_{k}\rangle$ be an abbreviation for
+%$\mathsf D^{k}[M, N_{1},\dots, N_{k}]0$. A \emph{resource $\lambda$-term} is any term belonging to the grammar $t::= x \mid \lambda x.t \mid t \langle t,\dots, t\rangle$.
+%%
+%
+%The Taylor expansion of a $\lambda$-term is a set of resource terms $\Te{M}$ defined inductively as 
+%
+$\Te{x}=\{x\}$, $\Te{\lambda x.M}=\{\lambda x.t\mid t\in \Te{M}\}$ and 
+$\Te{MN}=\{t\cdot \langle u_{1},\dots, u_{k}\rangle  \mid k\in \mathbb N, t\in \Te{M}, u_{i}\in \Te{N} \}$, where $t\cdot \langle u_{1},\dots, u_{k}\rangle$ is an abbreviation for $\mathsf D^{k}[t, u_{1},\dots, u_{k}] 0$.
+Observe that in the terms appearing in $\Te{M}$ all applications are always \emph{bounded}: they may use an exact number of copies of their input. 
+Such terms are usually called \emph{resource $\lambda$-terms} \cite{Pagani2009, Manzo2012}.
+Considering the term $M=zxx$ from Example \ref{ex:zxx}, all terms
  $t_{n,m}=z \,\langle x^{n}\rangle\,\langle x^{m}\rangle$, for $n,m\in \mathbb N$, are in $\Te{M}$. 
 Observe that the interpretation of $t_{n,m}$ yields a tropical polynomial
-$\model{t_{n,m}}^{!}(x)(z)= y_{[n,m]}+(n+m)x$, rather than a tLs. 
-However, resource term need not always yield a tropical polynomial. For instance, consider $y:(*\to*)\to (*\to*), x:(*\to *) \vdash t:  (*\to  *)$
-with $t=y\cdot (y\cdot x)$. Then 
+$\model{t_{n,m}}^{!}(x)(z)= y_{[n,m]}+(n+m)x$, rather than a tps. 
+However, resource $\lambda$-terms need not always yield a tropical polynomial. For instance, consider $y:(*\to*)\to (*\to*), x:(*\to *) \vdash t:  (*\to  *)$
+with $t=y\cdot \langle y\cdot \langle x\rangle \rangle\in \Te{y(yx)}$. Then 
 $\model{t}^{!}: \Lawv^{  \mathcal M_{\mathrm{fin}}(\mathbb N)\times \mathbb N}
@@
 %It is worth discussing the formula above a bit more. 
-The equation above is a nothing more than a tropical version of the Taylor formula \eqref{eq:taylor}:
+The equation above is a nothing more than a tropical version of the Taylor formula from the Introduction:
 %:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
@@
 %Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
+
 Second, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of \cite[Theorem 4.23]{Manzo2012} shows:
@@
 
-\begin{theorem}
+\begin{theorem}\label{thm:taylor}
 Let $\mathcal S$ be one of $\RM{PCF}^{\Lawv},\STLC,\BSTLC,\STDLC$. Let $\Gamma\vdash_{\mathcal S}M:A$ and $a\in \model{A}$.
@@
 
-\item For $\mathcal S=\STDLC$, if $t$ is a resource term, then $\model{\Gamma\vdash_{\mathcal S}t:A}^{!}_{a}$ is Lipschitz;
+\item For $\mathcal S=\STDLC$, if $M\in \Te{M'}$, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is Lipschitz;
 
@@
 \end{theorem}
+\begin{proof}
+1). As we already observed, the interpretation of a bounded term is a tropical polynomial.
+2.) From Proposition \ref{prop:troplinear} 2.~observing that a resource term $t(x)$ may use a variable $x$ a fixed number $n$ times, so that its matrix lies in $\Lawv^{!_{n}X\times Y}$. 
+%Now we apply Corollary~\ref{prop:polylip} to each coordinate of the image, and by taking the maximum Lipschitz constant among the finite number $\mathrm{Card}(\model A)$ of them, we obtain the thesis.
+3). From Theorem~\ref{thmTLSlocLip}.
+4).
+It follows from \autoref{cor:T(M)=M} plus the fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have $\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with $\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
+\end{proof}
 
@@
 
-We conclude our discussion by presenting an interesting application of the Taylor expansion in $\LREL_{!}$: as proved in the previous section, all tLs are locally Lipschitz; now, Theorem \ref{thm:taylor} can be used to compute approximations of the Lipschitz constants of an actual higher-order program.
+We conclude our discussion with an application of the Taylor expansion in $\LREL_{!}$: as proved in the previous section, all tps are locally Lipschitz; now, Theorem \ref{thm:taylor} can be used to compute approximations of the Lipschitz constants of an actual higher-order program.
 
@@
 Suppose $x: A \vdash_{\STLC}M:B$ and $\vdash_{\STLC} N:A$. 
-Then for all $t\in \Te{M}$ such that $\model{t}^{!}(\model{N})\neq \infty$, and $\delta>0$, the tLs $\model{ x:A\vdash_{\STLC}M:B}^{!}$ is $\frac{\model{t}^{!}(\model{N}+2\delta)}{\delta}$-Lipschitz over the open ball
+Then for all $t\in \Te{M}$ such that $\model{t}^{!}(\model{N})\neq \infty$, and $\delta>0$, the tps $\model{ x:A\vdash_{\STLC}M:B}^{!}$ is $\frac{\model{t}^{!}(\model{N}+2\delta)}{\delta}$-Lipschitz over the open ball
  $B_{\delta}(\model{N})$.
 \end{corollary}
-{\color{red}ADD PROOF SKETCH}
+\begin{proof}
+We apply the Lipschitz estimate from Thm.~\ref{thmTLSlocLip} for $\model{M}^{!}$,  and use the fact that, from Thm.~\ref{thm:taylor} 4.~it follows that $\model{t}^{!}\geq \model{M}^{!}$.
+To conclude, since $\model{t}^{!}$ is concave and non-decreasing, $\max_{\overline{B_{3\delta}(x)}}\model{t}^{!}= \model{t}^{!}(x+3\delta)$, which yields the stated Lipschitz estimate. 
+\end{proof}
+%{\color{red}ADD PROOF SKETCH}
 
@@
 % $\model{M}^{!}: \Lawv \times \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}\to \Lawv$ is given by
- The (generalized) tLs $\model{M}^{!}(x)(y)= \inf_{n,n'\in \mathbb N}\{y_{[(n,n')]}+(n+n')x\}$ is not (globally) Lipschitz: for any 
+ The (generalized) tps $\model{M}^{!}(x)(y)= \inf_{n,n'\in \mathbb N}\{y_{[(n,n')]}+(n+n')x\}$ is not (globally) Lipschitz: for any 
 $L>0$, choose a natural number $N>L$, let $Y\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ be such that $Y_{\mu}<\infty$ only if $\mu=[(n,n')]$ with $n+n'\geq N$; then $|\model{M}^{!}(x)(Y)- \model{M}^{!}(x+\epsilon)(Y)|\geq N\epsilon > L\epsilon$.
@@
 $\model{t}^{!}(x)(Y) = Y_{[(N-1,1)]}+Nx$. We can then compute a Lipschitz-constant for $\model{M}^{!}$ around $\langle x,Y\rangle$
-as $\frac{1}{\delta}\model{t}^{!}(\langle x,Y\rangle+\delta)= N+2 + \frac{Y_{[(N-1,1)]}+Nx}{\delta}$. 
+as $\frac{1}{\delta}\model{t}^{!}(\langle x,Y\rangle+\delta)= 3N+3 + \frac{Y_{[(N-1,1)]}+Nx}{\delta}$. 
 \end{example}

### 2023-07-31 CSL24/section4.tex
commit

@@
 1). Using the fact that $f( x)_{b}= \inf_{a\in X}\{\matr f_{a,b}+ x_{a}\}$,
-the problem reduces to checking that $|(\matr f_{a,b}- x_{a})- (\matr f_{a,b}- y_{a})| = |x_{a}- y_{a}|\leq \|  x- y\|_{\infty}$.
+the problem reduces to checking that $|(\matr f_{a,b}- x_{a})- (\matr f_{a,b}- y_{a})| = |x_{a}- y_{a}|\leq \|  x- y\|_{\infty}$ (see the Appendix).
 

### 2023-07-31 CSL24/section7.tex
commit

@@
 More generally, the literature on program metrics in denotational semantics is vast.
-Since at least \cite{VANBREUGEL20011} metric spaces, also in Lawvere's generalized sense \cite{Lawvere1973}, have been exploited as an alternative framework to standard, domain-theoretic, denotational semantics, notably
-\emph{ultra}-metrics and \emph{partial} metrics have been shown to form a CCC \cite{Escardo1999,PistoneLICS, PistoneFSCD2022}.
+Since at least \cite{VANBREUGEL20011} metric spaces, also in Lawvere's generalized sense \cite{Lawvere1973}, have been exploited as an alternative framework to standard, domain-theoretic, denotational semantics; notably models of $\STLC$ and $\mathrm{PCF}$ based on
+\emph{ultra}-metrics and \emph{partial} metrics have been proposed  \cite{Escardo1999,PistoneLICS, PistoneFSCD2022}.
 

### 2023-07-31 CSL24/section3fancy.tex
Commit dettagli

@@
 Hence, any reduction $\omega=\rho_{1}\dots \rho_{k}: M \twoheadrightarrow N$ is naturally associated with a weight $\mathsf w(\omega)=\sum_{i=1}^{k}\mathsf w(\rho_{i})\in Q$.
-In particular, from [\cite{Manzo2013}, Theorem V.6] we deduce the following adequation result:
+In particular, from [Theorem V.6]\cite{Manzo2013} we deduce the following adequation result:
 \begin{proposition}
@@
 Considering its interpretation as a function of $p,1-p$, we have that $\model{\vdash_{\mathrm{PPCF}} M^{\circ}:\mathrm{Nat}}_{n}(-\log(p),-\log(1-p))$%\in \Lawv$ (cf.~Remark \ref{rmk:val_trop})
-is the minimum negative log-probability of any reduction from $M$ to $\underline n$, i.e.\ the negative log-probability of (any of) the (equiprobable) most likely reduction path from $M$.
+is the minimum negative log-probability of any reduction from $M$ to $\underline n$, i.e.\ the negative log-probability of (any of) the (equiprobable) most likely reduction path from $M$ to $n$.
 \end{corollary}

### 2023-07-31 CSL24/section4.tex
Commit dettagli

@@
 %Remark that we cannot immediately apply \cite[Proposition 4.4.(3)]{Cobzas2017}, since $\Lawv^X$ is not even a $\BB R$-vector space.
-Now $(0,\infty)^X\subseteq \BB R^X$ is open and convex, and {\color{red}$\BB R^X$ is a LCTVS}, so \cite[Proposition 4.4.(3)]{Cobzas2017} entails the continuity of $-f\big|_{(0,\infty)^X}$, hence that of $f\big|_{(0,\infty)^X}$.
+Now $(0,\infty)^X\subseteq \BB R^X$ is open and convex, %and {\color{red}$\BB R^X$ is a LCTVS}, 
+so \cite[Proposition 4.4.(3)]{Cobzas2017} entails the continuity of $-f$ on $(0,\infty)^X$, hence that of $f$ on it.
 \end{proof}
@@
 \item If a tps $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:X\times Y\to \Lawv$ (i.e.~it is linear), then $f$ is non-expansive (i.e.\ $1$-Lipschitz).  
-\item If  $f \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f: \ !_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
+\item If  $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f: \ !_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
 \end{enumerate}
@@
 In \cite[Theorem 6.4]{Cobzas2017} a locally Lipschitz property is obtained for locally convex topological vector spaces, under the hypothesis of continuity.
-\cite[Proposition 6.5]{Cobzas2017} shows that continuity is used in order to have a locally bounded condition, the true main ingredient of the proof.
-Instead of showing how our case fits into such theorems, we prefer to state a theorem, which is basically a particular case of \cite[Theorem 6.9, Lemma 6.10]{Cobzas2017}:
+\cite[Proposition 6.5]{Cobzas2017} shows that continuity is used in order to have a locally bounded condition, the crucial ingredient of the proof.
+Instead of showing how our case fits into such theorems, we prefer to state the following theorem, basically a particular case of \cite[Theorem 6.9, Lemma 6.10]{Cobzas2017}:
 %, already fitting in the topics under consideration in this paper.

### 2023-07-31 CSL24/section1.tex
Commit

@@
 in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz.
-The fundamental motivation of this work is then the observation that
+The fundamental intuition of this work is then the observation that
 this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
@@
 
-In this paper we demonstrate that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provide the desired bridge between differential and metric semantics, and we suggest how this approach might be used to study probabilistic and non-deterministic programs (an aspect that will be developed in more detail in a future paper of this series).
+In this paper we make a first step in such direction, by demonstrating that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provide the desired bridge between differential and metric semantics, and we suggest how this approach might be used to study probabilistic and non-deterministic programs (this last aspect will be developed in more detail in a future paper of this series).
 
@@
 
-\item after recalling the notions of tropical polynomials and power series, we show how these objects naturally arise in the \emph{best-case} analysis of probabilistic and non-deterministic programs, by studying a few very simple examples.
+\item after recalling the notions of tropical polynomials and power series, we show how these objects naturally arise in the \emph{best-case} analysis of probabilistic and non-deterministic programs, by studying a few simple examples.
 This is in Sections \ref{section2} and \ref{section22}.
 
-\item We study the relational model over the tropical semiring, which provides a semantics of effectful extensions of $\STLC$ and $\mathrm{PCF}$. Notably, 
- we show that the functions interpreting higher-order programs, which correspond to a generalization of \emph{tropical power series} \cite{Porzio2021}, are locally Lipschitz-continuous and thus yield a full-scale metric semantics for the $\lambda$-calculus and its bounded fragments.
-This is in Sections \ref{section3} and \ref{sec:tls}.
+\item We study the relational model over the tropical semiring, which provides a semantics of effectful extensions of $\STLC$ (simply typed $\lambda$-calculus) and $\mathrm{PCF}$. Notably, 
+ we show that the functions interpreting higher-order programs, %which correspond to a 
+generalizations of \emph{tropical power series} \cite{Porzio2021}, are locally Lipschitz-continuous and thus yield a full-scale metric semantics for the $\lambda$-calculus and its bounded fragments.
+This is in Sections \ref{section3} and~\ref{sec:tls}.
 

### 2023-07-31 CSL24/section2bis.tex
Commit

@@
 %\footnote{In analogy with \cite{DanEhrh2011}, we could also call them \emph{tropical entire functions}.} 
-(tps) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
-If $I$ is finite, $f$ is called a \emph{tropical polynomial}. 
+%(tps) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
+in $k$-variables is a function $f:\Lawv^k\to\Lawv$ of shape $f(x)=\inf_{i\in I}\{ix + \matr f(i) \}$, where 
+$I\subseteq\mathbb N^{k}$, $i x$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients.
+When $I$ is finite, $f$ is called a \emph{tropical polynomial}. 
 \end{definition}
 
-A tropical polynomial is always a piece-wise linear function since it has shape $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$
+In Section \ref{section3} we also consider tps in \emph{infinitely many} variables.
+
+A tropical polynomial is always a piece-wise linear function since, e.g.\ in one variable, it has shape $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$
 For example, the polynomials $\varphi_{n}(x)=\min_{0\leq j\leq n}\{jx+2^{-j}\}$
@@
 % simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root. 
-tps with more than one variable are defined as $f(x)=\inf_{n\in \mathbb N^{k}}\{nx + \matr f(n) \}$, where 
-$x\in \Lawv^{k}$, $n\in \mathbb N^{k}$, $nx$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients.
+%Tps with more than one variable are defined as $f(x)=\inf_{n\in \mathbb N^{k}}\{nx + \matr f(n) \}$, where 
+%$x\in \Lawv^{k}$, $n\in \mathbb N^{k}$, $nx$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients.
 In Section \ref{section3} we also consider tps with \emph{infinitely many} variables.
@@
 Given $Q,L$ semirings with units and zeros, $L$ commutative idempotent, if $L$ is totally ordered by the partial order: $\alpha \preceq \beta$ iff $\alpha +  \beta = \beta$, then one defines a \emph{valuation} \cite{Izhakian2015} to be a map $\mathrm{val}:Q\to L$ s.t.\ $\mathrm{val}(0)=0$, $\mathrm{val}(1)=1$, $\mathrm{val}(ab)=\mathrm{val}(a)\mathrm{val}(b)$, $\mathrm{val}(a+b)\preceq \max\set{\mathrm{val}(a),\mathrm{val}(b)}$.
-The valuation which is constantly equal to $1\in L$ on all the invertibles of $Q$ is called the \emph{trivial valuation} $\mathrm{val}_1$.
+For example, the valuation which gives $1\in L$ on all the invertibles of $Q$ and $0$ otherwise, is called the \emph{trivial valuation} $\mathrm{val}_1$.
 %Notice that if $\mathrm{val}:Q\to ([0,+\infty],+,\cdot,0,1)$ is a valuation, then $\mathrm{logval}(x):Q\to ([-\infty,+\infty],\min,+,+\infty,0)$, $\mathrm{logval}(x):=-\log \mathrm{val}(x)$ is a valuation.
 
-Now, for a (formal) power series $f=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) with coefficients $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$ (notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$), one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
-
-Remark that tropicalising w.r.t.\ the trivial valuation gives $\trop^{\mathrm{val}_1} \sum_{i\in I} a_i x^i = \inf_{i\in I} \set{i\alpha} = \trop^{\mathrm{val}} \sum_{i\in I\subseteq\N} x^i$ for all valuation $\mathrm{val}$, because in $\Lawv$ the unit is $0$.
+Now, for a power series $f=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) with coefficients $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$, one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
+%Remark that tropicalising w.r.t.\ the trivial valuation gives $\trop^{\mathrm{val}_1} \sum_{i\in I} a_i x^i = \inf_{i\in I} \set{i\alpha} = \trop^{\mathrm{val}} \sum_{i\in I\subseteq\N} x^i$ for all valuation $\mathrm{val}$, because in $\Lawv$ the unit is $0$.
 %and the zero is $\infty$.
 %Remember that a_i \neq 0 by def of formal power series
-That is, it corresponds to quotienting the coefficients in $Q$ for the idempotency of $+$.
+%That is, it corresponds to quotienting the coefficients in $Q$ for the idempotency of $+$.
 \end{remark}
 
-Observe that tropicalisation is \emph{not} in general an injective operation. 
-This is because tps have a tendency to collapse, if not globally at least locally, onto tropical polynomials.
-For instance, the tps $\varphi(x,y)=\inf_{n}\{2x+(n+1)y,3x\}$ is  equivalent to the polynomial $\min\{2x+y,3x\}$, since $2x+(n+1)y\geq 2x+y$ holds for all $n$.
+%For instance, the tps $\varphi(x,y)=\inf_{n}\{2x+(n+1)y,3x\}$ coincides with the polynomial $\min\{2x+y,3x\}$, since $2x+(n+1)y\geq 2x+y$ holds for all $n$.
 So the trivial valuation sends both the infinite series $\sum_{n}(x^{2}y^{n+1}+x^{3})$ and the polynomial $x^{2}y+x^{3}$ onto the same map.
+Notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$ and the zero is $\infty$.
+Therefore, for instance, the tropicalisation (w.r.t.\ any valuation) sends the infinite power series $\sum_{n\geq 2} x^n$ to the tps $\varphi(x)=\inf_n (n+2)x$.
+But the latter always coincides with the tropical polynomial $\varphi(x)=2x$, since $nx\geq 0$ for all $n\in\N$, which is in turn the tropicalisation of the polynomial $x^2$.
+This shows that tropicalisation is \emph{not} in general an injective operation and in fact, as we show in \autoref{theorem:fepsilon}, tps (in finitely many variables) have a tendency to collapse, if not globally at least locally, onto tropical polynomials.
 

### 2023-07-31 CSL24/section5bis.tex
Commit

@@
 The key ingredient is a \emph{differential constructor} $\Der[\_,\_]$,  added to the usual syntax of the $\lambda$-calculus.
-The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M,N]$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M,N]$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).
+The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M,N]$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once.
+This is also why $\Der[M,N]$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$. 
+In particular, the application of $\Der[M,N]$ to an ``error term'' $0$ ensures that $M$ will use $N$ exactly once (we say \emph{linearly}).
 %Interestingly, the categorical study of the operator $\Der$ has led to the introduction of \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, a class of categories providing an abstract axiomatization of differentiation, in the usual mathematical sense.
@@
 %which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.
+
 The reason why $\Der$ is called a ``differential'', is twofold: semantically, its interpretation is a generalisation of the usual differential (see \autoref{section3}); syntactically, it allows to define the so-called \emph{Taylor expansion} $\C T$ of programs:
-the idea is that, in $\STDLC$, one can expand any application $MN$ as an infinite formal sum of \emph{linear} applications
-$\Der^{k}[M,N^k]$, i.e.~where $N$ is linearly passed exactly $k$ times to $M$, and doing this recursively gives rise to the series $\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot \Der^{k}[\Te{M},\Te{N}^k]0$.
+the idea is that one can expand any application $MN$ as an infinite formal sum of \emph{linear} applications
+$\Der^{k}[M,N^k]0$, i.e.~where $N$ is linearly passed exactly $k$ times to $M$; 
+doing this recursively gives rise to the series $\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot \Der^{k}[\Te{M},\Te{N}^k]0$.
 In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.
-Actually, we will work with the so-called \emph{qualitative} Taylor expansion, where one quotients for idempotency of the $+$ operator in the series, and thus the Taylor expansion is just a set (see \autoref{sec:TayLip}).% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
+Actually, we will work with the so-called \emph{qualitative} Taylor expansion, where one quotients for idempotency of the $+$ operator in the series, and thus the Taylor expansion becomes just a set (see \autoref{sec:TayLip}).% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
 
@@
 as a limit of Lipschitz maps, in some sense, thus bridging the metric and differential approaches.  
-Here, a natural direction to look for is the \emph{weighted relational semantics} \cite{Manzo2013}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
+Here, a natural direction to look for is the \emph{weighted relational semantics} \cite{Manzo2013}, %i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
+due to its strict relations with the Taylor expansion of programs.
 However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.

### 2023-07-31 CSL24/main.tex
Commit
commit

@@
 
-\author{Davide {Barbarossa}}{Dipartimento di Informatica, Universit\`a di Bologna, %[optional: Address], 
+\author{Davide {Barbarossa}}{Dipartimento di Informatica - Scienza e Ingegneria, Universit\`a di Bologna, %[optional: Address], 
 Italy% \and My second affiliation, Country 
@@
 
-\author{Paolo {Pistone}}{Dipartimento di Informatica, Universit\`a di Bologna, %[optional: Address], 
+\author{Paolo {Pistone}}{Dipartimento di Informatica - Scienza e Ingegneria, Universit\`a di Bologna, %[optional: Address], 
 Italy% \and LIP, Universit\'e de Lyon, France 
-\and \url{%WebPage
-} }{paolo.pistone2@unibo.it}{%Orchid
-}{%(Optional)Funding each author
+\and \url{http://logica.uniroma3.it/pistone/
+} }{paolo.pistone2@unibo.it}{https://orcid.org/0000-0003-4250-9051}{%(Optional)Funding each author
 }
@@
 
-\funding{This work has been supported by the DIAPASoN project}%optional, to capture a funding statement, which applies to all authors.
+\funding{This work has been supported by the ERC CoG 818616 DIAPASoN.}%optional, to capture a funding statement, which applies to all authors.
 Please enter author specific funding statements as fifth argument of the \author macro.
@@
 \begin{abstract}
-We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, by considering the relational semantics weighted on the tropical semiring.
-The aim is to provide a framework where both program metrics, based on the analysis of program sensitivity via Lipschitz conditions, and resource analysis, based on higher-order program differentiation, coexist.
-In order to do that, we consider a graded-types lambda-calculus and the differential lambda-calculus, both languages ensuring bounded duplications.
-A typical result of such line of research is that, thanks to the tropical operations, a simply typed term is interpreted as a locally Lipschitz map, which is in turn decomposed as an inf of more and more sensitive Lipschitz ones, corresponding to the interpretations of the approximants given by the lambda-calculus Taylor expansion.
-We finally describe some promising lines of research, showing in particular how such tropical model expresses natural optimisation problems of probabilistic calculi.
+We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, and we show that it provides a unified framework for both program  metrics, based on the analysis of program sensitivity via Lipschitz conditions, and resource analysis, based on higher-order program differentiation.
+To do that we focus on the semantics arising from the relational model weighted over the tropical semiring, and we discuss its application to the study of “best case” program behavior for languages with probabilistic and non-deterministic effects.
+Finally, we show that a general foundation for this approach is provided by an abstract correspondence between tropical algebra and Lawvere’s theory of generalized metric spaces.
+%A typical result of such line of research is that, thanks to the tropical operations, a simply typed term is interpreted as a locally Lipschitz map, which is in turn decomposed as an inf of more and more sensitive Lipschitz ones, corresponding to the interpretations of the approximants given by the lambda-calculus Taylor expansion.
+%We finally describe some promising lines of research, showing in particular how such tropical model expresses natural optimisation problems of probabilistic calculi.
 \end{abstract}
@@
 
-\section{Tropical Semantics of Effectful Programs}\label{section22}
+\section{Tropical Semantics and First Order Effectful Programs}\label{section22}
 \input{section2bisbis}
@@
 
-\section{Taylor Expansion and Lipschitz Approximation}\label{sec:TayLip}
+\section{Taylor Meets Lipschitz}\label{sec:TayLip}
 \input{sectionTaylip}

### 2023-07-31 CSL24/section1.tex
commit

@@
 
-\subparagraph*{Two kinds of quantitative approaches}
+\subparagraph*{Two approaches to quantitative semantics}
 
-Among such quantitative approaches, two different directions have received considerable attention.
+Among such quantitative approaches, two have received considerable attention.
 
@@
 On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
-This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
+This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable to the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.
 
 In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
-In metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
+In metric semantics, linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances, and the possibility of duplicating inputs leads to interpret programs with a fixed duplication bound as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
 By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.
 
-A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled.
+A natural question, at this point, is whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled.
 At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
-in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz.
+in metric models $n$ times duplication results in a {linear} (hence \emph{Lipschitz}) function $n x$, while in differential models this results in a {polynomial} function $x^{n}$, thus not Lipschitz.
 The fundamental intuition of this work is then the observation that
-this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
+this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n x$.
 
@@
 
-Coming back to our discussion on program semantics, tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
-At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.
+Coming back to our discussion on program semantics, tropical mathematics seems to be just  what we look for, as it turns polynomial functions like $x^{n}$ into Lipschitz maps like $n x$.
+At this point, it is worth mentioning that a tropical variant of the usual relational semantics of linear logic and the $\lambda$-calculus has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail.
 Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
@@
 
-In this paper we make a first step in such direction, by demonstrating that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provide the desired bridge between differential and metric semantics, and we suggest how this approach might be used to study probabilistic and non-deterministic programs (this last aspect will be developed in more detail in a future paper of this series).
+In this paper we make a first step in such direction, by demonstrating that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provide the desired bridge between differential and metric semantics, and we suggest how this approach might be used to study probabilistic and non-deterministic programs.% (an aspect that will be developed in more detail in a future paper of this series).
 
@@
 
-\subparagraph*{This paper and our contributions}
+\subparagraph*{Contributions and outline of the paper}
 
@@
 Moreover, we show that the conceptual unification of these two approaches suggests ways in which techniques from resource-analysis could be used in sensitivity analysis and \emph{vice-versa}, paving the way for new  applications of tropical geometry to the  study of higher-order programs.
-
 Our contributions in this paper are the following:
@@
 
-\item after recalling the notions of tropical polynomials and power series, we show how these objects naturally arise in the \emph{best-case} analysis of probabilistic and non-deterministic programs, by studying a few simple examples.
+\item We first show that tropical polynomials naturally arise in the {best-case} analysis of probabilistic and non-deterministic programs, turning the study of quantitative program behavior into a purely combinatorial problem.
 This is in Sections \ref{section2} and \ref{section22}.
@@
 \item We study the relational model over the tropical semiring, which provides a semantics of effectful extensions of $\STLC$ (simply typed $\lambda$-calculus) and $\mathrm{PCF}$. Notably, 
- we show that the functions interpreting higher-order programs, %which correspond to a 
-generalizations of \emph{tropical power series} \cite{Porzio2021}, are locally Lipschitz-continuous and thus yield a full-scale metric semantics for the $\lambda$-calculus and its bounded fragments.
+ we show that higher-order programs are interpreted by a  
+generalizations of {tropical power series} \cite{Porzio2021}, and we show that these functions are locally Lipschitz-continuous, thus yielding a full-scale metric semantics.
 This is in Sections \ref{section3} and~\ref{sec:tls}.
 
- \item We exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps, and we show that it can be used to deduce approximated Lipschitz-constants for higher-order programs.
+ \item We exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps, and we show that it can be used to compute approximated Lipschitz-constants for higher-order programs.
 This is in Section \ref{sec:TayLip}.
@@
 by putting the connection between the 
-tropical, differential and metric viewpoints at the right level of generality.
+tropical, differential and metric viewpoints at the right level of abstraction and generality.
 We recall a well-known correspondence between Lawvere's \emph{generalized metric spaces} \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007} and we show that it yields a model of the differential $\lambda$-calculus which extends the tropical relational model.

### 2023-07-31 CSL24/section2bis.tex
commit

@@
 \begin{definition}
-A \emph{tropical power series}
+A \emph{tropical power series} (tps) 
 %\footnote{In analogy with \cite{DanEhrh2011}, we could also call them \emph{tropical entire functions}.} 
@@
 \end{definition}
-
+Hence, a unary tps is a function $f:\Lawv\to \Lawv$ of the form $f(x)=\inf_{i\in I}\{ix+a_{i}\}$, with $I\subseteq\mathbb N$ and the $a_{i}\in \Lawv$.
 In Section \ref{section3} we also consider tps in \emph{infinitely many} variables.
 
-A tropical polynomial is always a piece-wise linear function since, e.g.\ in one variable, it has shape $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$
+A tropical polynomial is always a piece-wise linear function since, e.g.\ in one variable, it has shape $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$.
 For example, the polynomials $\varphi_{n}(x)=\min_{0\leq j\leq n}\{jx+2^{-j}\}$
@@
 
-A \emph{tropical root} of a tropical power series $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable (i.e.~where the slope of $\varphi$ changes).
+A \emph{tropical root} of a tps $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable (i.e.~where the slope of $\varphi$ changes).
 When $\varphi$ is a polynomial, the roots of $\varphi$ coincide with the points where the minimum defining $\varphi$ is attained at least twice (see \autoref{fig:plot1}).
@@
 
-Now, for a power series $f=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) with coefficients $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$, one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
+Now, for a power series $f(x)=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) with coefficients $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$, one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
 %Remark that tropicalising w.r.t.\ the trivial valuation gives $\trop^{\mathrm{val}_1} \sum_{i\in I} a_i x^i = \inf_{i\in I} \set{i\alpha} = \trop^{\mathrm{val}} \sum_{i\in I\subseteq\N} x^i$ for all valuation $\mathrm{val}$, because in $\Lawv$ the unit is $0$.
@@
 But the latter always coincides with the tropical polynomial $\varphi(x)=2x$, since $nx\geq 0$ for all $n\in\N$, which is in turn the tropicalisation of the polynomial $x^2$.
-This shows that tropicalisation is \emph{not} in general an injective operation and in fact, as we show in \autoref{theorem:fepsilon}, tps (in finitely many variables) have a tendency to collapse, if not globally at least locally, onto tropical polynomials.
+This shows that tropicalisation is \emph{not} in general an injective operation and in fact, as we show in \autoref{theorem:fepsilon} below, tps (in finitely many variables) have a tendency to collapse, if not globally at least locally, onto tropical polynomials.
 
-Similarly, by looking at Fig~\ref{fig:plot1} it appears that, \emph{far from $0$}, $\varphi$ behaves like some of the polynomials $\varphi_{n}$.
+For instance, by looking at Fig~\ref{fig:plot1} it appears that, \emph{far from $0$}, $\varphi$ behaves like some of the polynomials $\varphi_{n}$.
 In particular, %for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$. More, precisely, 
@@
 \begin{theorem}\label{theorem:fepsilon}
-For all tLs $f(x_{1},\dots, x_{k})=\inf_{n\in \mathbb N^{k}}\{n x+\matr f(n)\}$, for all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
+For all tps $f(x)=\inf_{n\in \mathbb N^{k}}\{n x+\matr f(n)\}$, for all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
 % 
@@
 
-As we'll see, the potential of transforming infinitary objects (i.e.~tps) into combinatorial ones (i.e.~tropical polynomials), is one of the most interesting features of tropical semantics.
+As we'll see, the potential of collapsing infinitary objects (i.e.~tps) into combinatorial ones (i.e.~tropical polynomials), is one of the most intriguing features of tropical semantics.
 

### 2023-07-31 CSL24/section3fancy.tex
commit

@@
 % !TEX root = main.tex
-
-There exists a general and well-known construction that yields, for any \emph{continuous} semiring $Q$, a model $\QREL_{!}$ of effectful extensions of the simply typed $\lambda$-calculus and $\mathrm{PCF}$.
-In this section we recall this construction and show how, in the case $Q=\Lawv$, it captures optimal program behavior; moreover, we discuss how it adapts to model \emph{graded} and \emph{differential} variants of $\STLC$.
+In this section we first recall a general and well-known construction that yields, for any \emph{continuous} semiring $Q$, a model $\QREL_{!}$ of effectful extensions of the simply typed $\lambda$-calculus and $\mathrm{PCF}$, and we show how, when $Q=\Lawv$, it captures optimal program behavior; moreover, we discuss how this model adapts to \emph{graded} and \emph{differential} variants of $\STLC$.
 %In the next sections we will instead focus on the peculiar features of the tropical case.
@@
 \subparagraph*{Linear/Non-Linear Algebra on $Q$-Modules}
-For any \emph{continuous} semiring $Q$ (i.e.~a cpo equipped with an order-compatible semiring structure), one can define a category $\QREL$ of ``$Q$-valued matrices''.
-More precisely, $\QREL$ (\cite{Manzo2013} calls it $Q^\Pi$) has sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms, i.e.~$\QREL(X,Y):=Q^{X\times Y}$.
+For any \emph{continuous} semiring $Q$ (i.e.~a cpo equipped with an order-compatible semiring structure), one can define a category $\QREL$  (\cite{Manzo2013} calls it $Q^\Pi$) of ``$Q$-valued matrices'' as follows: $\QREL$ has sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms, i.e.~$\QREL(X,Y):=Q^{X\times Y}$.
 %The identity morphism of $\QREL$ is the identity matrix $I\in Q^{X\times X}$,
@@
 The composition $st\in Q^{X\times Z}$ of $t\in Q^{X\times Y}$ and $s\in Q^{Y\times Z}$ is given by $(st)_{a,c}:=\sum_{b\in Y} s_{b,c}t_{a,b}$ (observe that this series always converges because $Q$ is continuous).
-For any set $X$, $Q^X$ is a $Q$-semimodule and we can identify $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$, which have shape $f(x)_b=:\sum_{a\in X} \matr f_{a,b}x_a$, for some matrix $\matr f\in Q^{X\times Y}$.
+For any set $X$, $Q^X$ is a $Q$-semimodule and we can identify $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$, which have shape $f(x)_b:=\sum_{a\in X} \matr f_{a,b}x_a$, for some matrix $\matr f\in Q^{X\times Y}$.
 % \begin{remark}
@@
 Following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we are instead working with transpose matrices.
-$\QREL$ admits a comonad $!$ which acts, on objects, by taking the finite multisets.
-Remember that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
-Now, although a matrix $t\in\HOM{\QREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, by exploiting the coKleisli structure we can also ``express it in the base $X$'', which leads to the \emph{non-linear} map $t^{!}.Q^{X}\to Q^{Y}$ defined by the power series
+
+$\QREL$ admits a comonad $!$ which acts on objects by taking the finite multisets.
+Remember that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose objects are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined via the co-multiplication of $!$.
+Now, although a matrix $t\in\HOM{\QREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, by exploiting the coKleisli structure we can also ``express it in the base $X$'', which leads to the \emph{non-linear} map $t^{!}:Q^{X}\to Q^{Y}$ defined by the power series
 $
@@
 reads as \ $(st)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$, and the non-linear functions $t^{!}: \Lawv^{X}\to \Lawv^{Y}$ give rise to functions of shape 
-$t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$, which correspond to the generalisation of tps with \emph{possibly infinitely} many variables (in fact, as many as the elements of $X$).
+$t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
+These correspond to the generalisation of tps with \emph{possibly infinitely} many variables (in fact, as many as the elements of $X$).
 By identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tps generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the %functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ we recover the
@@
 \begin{remark}
- Let $X$, $Y$ be sets and let $\langle \_,\_\rangle:X\times Y \to \mathbb{R}$.
- For $f:X\to \mathbb R$, let $f^*:Y\to \mathbb R$ by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
+The operation $f\mapsto f^{!}$ turning a matrix into a function is reminiscent of the well-known operation of taking the 
+ \emph{convex conjugate} $f^*$ of a function $f$ defined over a vector space (itself a generalization of the Legendre transformation).
+ Indeed, let $X$, $Y$ be sets and let $\langle \_,\_\rangle:X\times Y \to \mathbb{R}$.
+ For $f:X\to \mathbb R$, let $f^*:Y\to \mathbb R$ be defined by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
  Then for $X=!A$, $Y=\Lawv^A$, where $A$ is a set, and $\langle \mu, y \rangle:= \mu y$, we have $f^!(y)=(-f)^*(-y)$ for all $f\in\Lawv^{!A}$.
- This is the same formal construction yielding the \emph{convex conjugate} $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
- This construction is in turn a generalisation of the Legendre transformation.
+%Hence the definition  is the same formal construction yielding the \emph{convex conjugate} $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
+ %This construction is in turn a generalisation of the Legendre transformation.
 % Despite the formal constructions being the same, we ignore for the moment if these could be connected to the study of high-order programs in our setting.
@@
 $\mathsf{ev}^{!}: \Lawv^{!X\times Y}\times \Lawv^{X}\to \Lawv^{Y}$ given by 
-$b\in Y \mapsto \mathsf{ev}^{!}(F,x)_{b}= \inf_{(\mu,b)}\{ F_{\mu,b}+ \mu x\}$.
-So, for instance, supposing the ground type $*$ of $\STLC$ is interpreted as the singleton set $\{*\}$, and recalling the identification $!\{*\}\simeq \mathbb N$, the interpretation of the term $x:*, z:(*\to*\to *) \vdash zxx$, involving two consecutive evaluations, yields the tps $\varphi:\Lawv\times \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}\to \Lawv$ given by $\varphi(x,z)=\inf_{n,n'}\{z_{[(n,n')]}+(n+n')x\}$.
+$b\in Y \mapsto \mathsf{ev}^{!}(F,x)_{b}= \inf_{\mu,b}\{ F_{\mu,b}+ \mu x\}$.
+So, for instance, supposing the ground type $o$ of $\STLC$ is interpreted as the singleton set $\{*\}$, and recalling the identification $!\{*\}\simeq \mathbb N$, the interpretation of the term $x:o, z:(o\to o\to o) \vdash zxx:o$, involving two consecutive evaluations, yields the tps $\varphi:\Lawv\times \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}\to \Lawv$ given by $\varphi(x,z)=\inf_{n,n'}\{z_{[(n,n')]}+(n+n')x\}$.
 \end{example}
@@
 Let $\vdash_{\mathrm{PPCF}} M:\mathrm{Nat}$ and $n\in\N$.
-Considering its interpretation as a function of $p,1-p$, we have that $\model{\vdash_{\mathrm{PPCF}} M^{\circ}:\mathrm{Nat}}_{n}(-\log(p),-\log(1-p))$%\in \Lawv$ (cf.~Remark \ref{rmk:val_trop})
+Considering its interpretation as a function of $p,1-p$, we have that $\model{\vdash_{\mathrm{PPCF}} M^{\circ}:\mathrm{Nat}}_{n}(-\log(p),-\log(1-p))$ %\in \Lawv$ (cf.~Remark \ref{rmk:val_trop})
 is the minimum negative log-probability of any reduction from $M$ to $\underline n$, i.e.\ the negative log-probability of (any of) the (equiprobable) most likely reduction path from $M$ to $n$.
@@
 
-\subparagraph*{Interpreting $\mathbb N$-Graded Types}\label{sec:BSTLC}
+\subparagraph*{$\mathbb N$-Graded Types}\label{sec:BSTLC}
 
@@
 
-The types of $\BSTLC$ are $A::= * \ \mid  \ !_{n}A \multimap A$, the contexts of the typing judgements are sets of declarations of the form $x :_{n}A$, with $n\in \mathbb N$, and the typing rules are: %given in Fig.~\ref{fig:rules}
+The types of $\BSTLC$ are $A::= o \ \mid  \ !_{n}A \multimap A$, the contexts of the typing judgements are sets of declarations of the form $x :_{n}A$, with $n\in \mathbb N$, and the typing rules are: %given in Fig.~\ref{fig:rules}
 	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
@@
 
-Now, one can see that the comonad $!$ of $\LREL$ can be ``decomposed'' {\color{red}REF} into a family of ``graded exponentials functors'' $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined on objects by taking multisets of cardinality \emph{at most} $n$.
+Now, one can see that the comonad $!$ of $\LREL$ can be ``decomposed'' into a family of ``graded exponentials functors'' $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined on objects by taking multisets of cardinality \emph{at most} $n$.
 %  lift to functors
@@
 
-\subparagraph*{Interpreting the Differential $\lambda$-Calculus}\label{sec:STDLC}
+\subparagraph*{The Differential $\lambda$-Calculus}\label{sec:STDLC}
 
-%We show that $\LREL_{!}$ also interprets the $\STDLC$,  
-A Cartesian closed differential $\lambda$-category (CC$\partial\lambda$C)\cite{Manzo2010,Blute2009, Blute2019} is a CCC enriched over commutative monoids (i.e.\ we can add morphisms and there is a $0$ morphism), with the cartesian closed structure compatible with the additive structure,
+We recall the interpretation in $\LREL_{!}$ of $\STDLC$,  
+ an extension of $\STLC$ ensuring exact control of duplications.
+The syntax of $\STDLC$ (see \cite[Section 3]{Manzo2010}) is made of \emph{terms} $M$ and \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
+quotiented by equations that make $+,0$ form a commutative monoid on the set of sums, %, i.e.\ commutativity and associativity of $+$ and neutrality of $0$ w.r.t.\ $+$;
+by linearity of $\lam x.(\_)$, $\Diff{\_}{\_}$ and $(\_)\mathbb T$ (but \emph{not} of $M(\_)$) and by irrelevance of the order of consecutive $\Diff{\_}{\_}$.
+%Remark that $M(\_)$ is \emph{not} set to be linear: $\lambda x.0=0\mathbb T=\Diff{0}{N}=\Diff{M}{0}=0$ but $M0\neq0$ in general.
+%This is crucial for the definition of the Taylor expansion.
+We follow the tradition of quotienting also for the idempotency of $+$.
+%Sums are, then, just \emph{finite} sets of terms.
+The typing rules are as follows: %in Figure~\ref{fig:rules}
+	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
+	\begin{array}{cccc}
+		\prooftree
+		\Gamma, x: A\vdash M: B
+		\justifies
+		\Gamma\vdash \lambda x.M: A\to B
+		\endprooftree 
+		&
+		\prooftree
+		\Gamma \vdash M: A\to B
+		\quad
+		\Gamma\vdash \mathbb T: A
+		\justifies
+		\Gamma \vdash M\mathbb T: B
+		\endprooftree 
+		&
+		\prooftree
+		\Gamma \vdash M: A\to B
+		\quad
+		\Gamma \vdash N: A
+		\justifies
+		\Gamma \vdash \Diff{M}{N}: A\to B
+		\endprooftree
+		&
+		\prooftree
+		\Gamma\vdash M_1: A
+		\,\cdots\,
+		\Gamma \vdash M_n:A
+		\justifies
+		\Gamma \vdash M_1+\cdots +M_n : A
+		\using (n\geq 2)
+		\endprooftree
+	\end{array}
+	\]
+where a context $\Gamma$ is a list of typed variable declarations.
+The axioms are $\Gamma, x:A \vdash x: A$ and $\Gamma\vdash 0:A$.
+The main feature of this language is that $\Der^n[\lambda x.M,N^n]0$ has a non-zero normal form iff $x$ is duplicated exactly $n$ times during reduction.
+
+The categorical models of $\STDLC$ are called \emph{cartesian closed
+ differential $\lambda$-categories} (CC$\partial\lambda$C)\cite{Manzo2010,Blute2009, Blute2019}.
+These are CCCs enriched over commutative monoids (i.e.\ morphisms are summable and there is a $0$ morphism), with the cartesian closed structure compatible with the additive structure,
 and equipped with a certain \emph{differential operator} $D$, generalising the usual notion of differential, see e.g.\ \cite{BluteEhrhTass10}.
@@
 %\begin{proposition}[{\color{red}LEMAY??}]\label{thm:LREL!CCDC}
- $\LREL_!$ is a CC$\partial\lambda$C (see Section \ref{section5bis}) when equipped with the \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
+ $\LREL_!$ is a CC$\partial\lambda$C (see Section \ref{section5bis}) when equipped with $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
+ The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tps $f:\Lawv^{X}\to \Lawv^{Y}$ into a tps $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by
+$D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a}+x_{a}+\mu y\right\}$.
+One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
+%Moreover, the Taylor formula \eqref{eq:taylorcat} yields a ``tropical'' Taylor formula for tps of the form 
+%$f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}$. 
 %\end{proposition}
@@
 %This ensures that one can define a sound interpretation of $\STDLC$-terms in the standard way (see [Section 4.3, \cite{Manzo2010}]).
-As such, $(\LREL_!,D)$ is a model of the differential $\lambda$-calculus $\STDLC$ \cite{ER}.
+%As such, $(\LREL_!,D)$ is a model of the differential $\lambda$-calculus $\STDLC$ \cite{ER}.
 
-% an extension of $\STLC$ ensuring exact control of duplications via a notion of linear substitution in its operational semantics.
-%We will not deal with its operational properties, so let us only give its syntax (see \cite[Section 3]{Manzo2010}): there are \emph{terms} $M$ and \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
-%quotiented by $\alpha$-equivalence, by equations that make $+,0$ form a commutative monoid on the set of sums, %, i.e.\ commutativity and associativity of $+$ and neutrality of $0$ w.r.t.\ $+$;
-%by linearity of $\lam x.(\_)$, $\Diff{\_}{\_}$ and $(\_)\mathbb T$ (but \emph{not} of $M(\_)$) and by irrelevance of the order of consecutive $\Diff{\_}{\_}$.
-%%Remark that $M(\_)$ is \emph{not} set to be linear: $\lambda x.0=0\mathbb T=\Diff{0}{N}=\Diff{M}{0}=0$ but $M0\neq0$ in general.
-%%This is crucial for the definition of the Taylor expansion.
-%We follow the tradition of quotienting also for the idempotency of $+$.
-%%Sums are, then, just \emph{finite} sets of terms.
-%The types are $A::= *\mid A\to A$, the typing rules: %in Figure~\ref{fig:rules}
-%	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
-%	\begin{array}{cccc}
-%		\prooftree
-%		\Gamma, x: A\vdash M: B
-%		\justifies
-%		\Gamma\vdash \lambda x.M: A\to B
-%		\endprooftree 
-%		&
-%		\prooftree
-%		\Gamma \vdash M: A\to B
-%		\quad
-%		\Gamma\vdash \mathbb T: A
-%		\justifies
-%		\Gamma \vdash M\mathbb T: B
-%		\endprooftree 
-%		&
-%		\prooftree
-%		\Gamma \vdash M: A\to B
-%		\quad
-%		\Gamma \vdash N: A
-%		\justifies
-%		\Gamma \vdash \Diff{M}{N}: A\to B
-%		\endprooftree
-%		&
-%		\prooftree
-%		\Gamma\vdash M_1: A
-%		\,\cdots\,
-%		\Gamma \vdash M_n:A
-%		\justifies
-%		\Gamma \vdash M_1+\cdots +M_n : A
-%		\using (n\geq 2)
-%		\endprooftree
-%	\end{array}
-%	\]
-%where a context $\Gamma$ is a list of typed variable declarations.
-%The axioms are $\Gamma, x:A \vdash x: A$ and $\Gamma\vdash 0:A$.
-%The main feature of this language is that $\Der^n[\lambda x.M,N^n]0$ has a non-zero normal form iff $x$ is duplicated exactly $n$ times during the reduction to normal form.
 %In $\BSTLC$ the typing system handles duplications; in $\STDLC$ the syntax with its operational semantics (that we do not give) does it.

### 2023-07-31 CSL24/section4.tex
commit

@@
 
-In this section we establish some topological and metric properties of tps, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
-The literature in tropical mathematics is often recent (e.g.~\cite{Porzio2021}), and several results we prove in this section are, to our knowledge, new.
+As seen in Section \ref{section2}, tropical polynomials are piecewise-linear functions, hence concave and Lipschitz-continuous.
+Moreover, tps in finitely many variables are locally equivalent to tropical polynomials (except at some singular points), and are thus also concave and locally Lipschitz-continuous.
+In this section we show that much of these properties extend also to tps with infinitely many variables, as those arising from the tropical relational model.
+%establish some topological and metric properties of tps, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
+The literature on tropical power series is often recent (e.g.~\cite{Porzio2021}), and several results we prove in this section are, to our knowledge, new.
 %
@@
 %
-In all the remainder of this section we consider the sets $\Lawv^X$ endowed with the usual sup-norm $\norm{x}_\infty:=\sup_{a\in X} x_a$.
+In the remainder of this section, by a tropical polynomial or power series we mean one with possibly infintely many variables.
+Moreover, we consider the sets $\Lawv^X$ endowed with the usual sup-norm $\norm{x}_\infty:=\sup_{a\in X} x_a$.
 
@@
 Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete \emph{normed cone} (see~\cite{Selinger2004}, or the appendix, for such notions).
-Its cone structure induces an order on it, called its \emph{cone order}:
+Its cone structure induces an order on it, called the \emph{cone order}:
 $x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
@@
 \end{theorem}
-\begin{proof}
- Uses \autoref{thm:cont} plus the fact, taken from \cite{Selinger2004}, that in a normed $\R_{\geq 0}$-cone $P$, considered with its cone-order, if every bounded directed net in $P$ admits a sup, and if $(v_i)_{i\in I}$ is a directed net in $P$ with an upper bound $v\in P$, then $\exists\bigvee_{i\in I} v_i \in P$ and, if $\inf_{i\in I} \norm{v-v_i} =0$, one has $\bigvee_{i\in I} v_i = v$.
-\end{proof}
+%\begin{proof}
+% From \autoref{thm:cont} plus the fact, taken from \cite{Selinger2004}, that in a normed $\R_{\geq 0}$-cone $P$, considered with its cone-order, if every bounded directed net in $P$ admits a sup, and if $(v_i)_{i\in I}$ is a directed net in $P$ with an upper bound $v\in P$, then $\exists\bigvee_{i\in I} v_i \in P$ and, if $\inf_{i\in I} \norm{v-v_i} =0$, one has $\bigvee_{i\in I} v_i = v$.
+%\end{proof}
 
@@
 Let us first look at what happens with those tps which are either \emph{linear} or obtained via bounded exponentials.
-The result below is in analogy with that happens in the usual metric semantics of Fuzz, where 
+The result below is in analogy with what happens in the usual metric semantics of Fuzz, where 
 linear functions are non-expansive and $n$-bounded functions are $n$-Lipschitz \cite{Reed2010}.
@@
 \begin{proposition}
-For all tropical linear functions $f,g:\Lawv^{X}\to \Lawv^{Y}$, $d_{\infty}(\matr f,\matr g)=d_\infty(f,g)$.% $\norm{ \matr f-\matr g}_{\infty} =  \sup_{x\in \Lawv^{X}} \norm{ f( x)-g(x)}_{\infty}$.
+For all $f,g\in \LREL_{!}(X,Y)$, $d_{\infty}( f,g)=d_\infty(f^{!},g^{!})$.% $\norm{ \matr f-\matr g}_{\infty} =  \sup_{x\in \Lawv^{X}} \norm{ f( x)-g(x)}_{\infty}$.
 \end{proposition}
@@
 
-Let us now look at what happens with tps, i.e.~when considering the full exponential $!$.
-First observe that, as consequence of Theorem~\ref{theorem:fepsilon}, the tps with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $(0,\infty)^X$.
+We now show that, if we consider the \emph{full} exponential $!$, i.e.~arbitrary tps, we can still prove that a local Lipschitz condition holds.
+%First observe that, as a consequence of Theorem~\ref{theorem:fepsilon}, the tps with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $(0,\infty)^X$.
 Yet, for the case of infinitely many variables we need a more general argument.
@@
 
-\subparagraph*{Tropical differential of tps}
-We conclude this section with a few observation on the differential operator over tps.
-The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tps $f:\Lawv^{X}\to \Lawv^{Y}$ into a tps $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by
-$D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a}+x_{a}+\mu y\right\}$.
+%\subparagraph*{Tropical differential of tps}
+%We conclude this section with a few observations on the differential operator over tps.
+%The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tps $f:\Lawv^{X}\to \Lawv^{Y}$ into a tps $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by
+%$D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a}+x_{a}+\mu y\right\}$.
 One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
-Moreover, the Taylor formula \eqref{eq:taylorcat} yields a ``tropical'' Taylor formula for tps of the form 
-$f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}$.
-
-The following result shows that the distance between two tropical maps can be approximated using the terms appearing in their Taylor expansions:
-\begin{proposition}
-For all tps $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
-the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz. Moreover 
-$\norm{ \matr f-\matr g}_{\infty}= \sup_{n} \norm{ {\delta^{(n)}f}- {\delta^{(n)}g}}_{\infty}$, 
-where $\delta^{(n)}h$ indicates the matrix of $D_{!}^{(n)}h$.
-%where $\delta^{(n)}f:( \Lawv^{X})^{n}\to \Lawv^{X}$ is the tropical linear function $\delta^{(n)}f(\B x_{1},\dots, \B x_{n})=
-%(\Der^{(n)}f)(\B x_{1},\dots, \B x_{n}, \infty)$. 
-\end{proposition}
+%Moreover, the Taylor formula \eqref{eq:taylorcat} yields a ``tropical'' Taylor formula for tps of the form 
+%$f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}$. 
+%
+%The following result shows that the distance between two tropical maps can be approximated using the terms appearing in their Taylor expansions:
+%\begin{proposition}
+%For all tps $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
+%the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz. Moreover 
+%$\norm{ \matr f-\matr g}_{\infty}= \sup_{n} \norm{ {\delta^{(n)}f}- {\delta^{(n)}g}}_{\infty}$, 
+%where $\delta^{(n)}h$ indicates the matrix of $D_{!}^{(n)}h$.
+%%where $\delta^{(n)}f:( \Lawv^{X})^{n}\to \Lawv^{X}$ is the tropical linear function $\delta^{(n)}f(\B x_{1},\dots, \B x_{n})=
+%%(\Der^{(n)}f)(\B x_{1},\dots, \B x_{n}, \infty)$. 
+%\end{proposition}
 

### 2023-07-31 CSL24/section5bis.tex
commit

@@
 
-What about the good old, ``unbounded'', simply typed $\lambda$-calculus?
+Yet, what about the good old, ``unbounded'', simply typed $\lambda$-calculus?
 Actually, by using unbounded duplications, one might lose the Lipschitz property.
@@
 
-The reason why $\Der$ is called a ``differential'', is twofold: semantically, its interpretation is a generalisation of the usual differential (see \autoref{section3}); syntactically, it allows to define the so-called \emph{Taylor expansion} $\C T$ of programs:
+The reason why $\Der$ is called a ``differential'', is twofold: semantically, its interpretation is a generalisation of the usual differential form analysis (see \autoref{section3}); syntactically, it allows to define the so-called \emph{Taylor expansion} $\C T$ of programs:
 the idea is that one can expand any application $MN$ as an infinite formal sum of \emph{linear} applications
 $\Der^{k}[M,N^k]0$, i.e.~where $N$ is linearly passed exactly $k$ times to $M$; 
-doing this recursively gives rise to the series $\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot \Der^{k}[\Te{M},\Te{N}^k]0$.
+doing this recursively gives rise to the suggestive {Taylor formula} $\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot \Der^{k}[\Te{M},\Te{N}^k]0$.
 In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.
-Actually, we will work with the so-called \emph{qualitative} Taylor expansion, where one quotients for idempotency of the $+$ operator in the series, and thus the Taylor expansion becomes just a set (see \autoref{sec:TayLip}).% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
+%Actually, we will work with the so-called \emph{qualitative} Taylor expansion, where one quotients for idempotency of the $+$ operator in the series, and thus the Taylor expansion becomes just a set (see \autoref{sec:TayLip}).% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).
 
@@
 due to its strict relations with the Taylor expansion of programs.
-However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.
+However, in this semantics, arbitrary terms correspond to power series, and terms with bounded applications correspond to {polynomials}, hence in any case to functions which are \emph{not} Lipschitz.
 
-Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions?
+Yet, what if such polynomials were tropical ones, i.e.~piecewise linear functions?
 This way, the Taylor formula could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps.
@@
 more abstract level, leading to introduce a 
-differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}).
+differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}), as shown in Section \ref{sec:GMS}.

### 2023-07-31 CSL24/sectionLaw.tex
commit

--- 2023-07-29 CSL24/sectionLaw.tex
@@
 
-An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}, that it's worth recalling.
+An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}, that it's certainly worth recalling here, albeit very briefly (more details in the Appendix).
 
@@
 
-On the one hand we have $\Lawv$-modules: these are triples $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$.% (notice that this is indeed the usual notion of module over the Lawvere quantale $\Lawv$).
+On the one hand we have $\Lawv$-modules: these are triples $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$.
+% (notice that this is indeed the usual notion of module over the Lawvere quantale $\Lawv$).
 A $\Lawv$-module homomorphism is a map $f:M\to N$ commuting with both joins and the $\Lawv$-action. We let $\Mod$ indicate the category of $\Lawv$-modules and their homomorphisms.
@@
 %\subsection{$\Lawv$-categories}
-On the other hand we have Lawvere's \emph{generalized metric spaces}:
+On the other hand we have Lawvere's \emph{generalized metric spaces}  \cite{Lawvere1973, Hofmann2014, Stubbe2014}:
 Lawvere was the first to observe that a metric space can be described as a \emph{$\Lawv$-enriched} category.
@@
 $0  \geq X(x,x)$ and $X(y,z)+X(x,y)\geq  X(x,z)$. 
-This structure is often referred to as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}.
-%, or as a \emph{quasi-metric space}.
- A metric space, in the usual sense, is thus the same as a skeletal (i.e.~$X(x,y)=0$ implies $x=y$) and symmetric (i.e.~$X(x,y)=X(y,x)$) $\Lawv$-category.
-Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, as functoriality reads as $Y(f(x),f(y))\leq X(x,y)$.
+This structure clearly generalizes the usual definition of metric spaces, which indeed are the same as skeletal (i.e.~$X(x,y)=0$ implies $x=y$) and symmetric (i.e.~$X(x,y)=X(y,x)$) $\Lawv$-categories.
+Moreover, a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is nothing but a non-expansive map $f:X\to Y$, since functoriality reads as $Y(f(x),f(y))\leq X(x,y)$.
 
-A $\Lawv$-category is \emph{complete} when all weighted colimits (cf.~Appendix for a precise definition) exist in $X$.
-This categorical notion of completeness can be seen as a generalization of the usual notion of Cauchy-completeness.
-We let $\GMet$ indicate the category of complete $\Lawv$-categories and continuous (i.e.~colimit-preserving) $\Lawv$-enriched functors.
+Lawvere also observed that the usual notion of Cauchy-completeness can be formulated, in this framework, as the existence of suitable colimits \cite{Lawvere1973}.
+A $\Lawv$-category is said \emph{complete} when \emph{all} {weighted colimits} exist in $X$.
+Let $\GMet$ indicate the category of complete $\Lawv$-categories and continuous (i.e.~colimit-preserving) $\Lawv$-enriched functors.
 
-Now, it turns out that the categories $\Mod$ and $\GMet$ are isomorphic \cite{Stubbe2006}. First,  any $\Lawv$-module $(M,\preceq, \star)$ is a $\Lawv$-category via
+It turns out that the categories $\Mod$ and $\GMet$ are isomorphic \cite{Stubbe2006}. First,  any $\Lawv$-module $(M,\preceq, \star)$ is a $\Lawv$-category via
 $M(x,y) = \inf\left\{ \epsilon \mid \epsilon \star x\geq y\right\}$.
 Moreover, a homomorphism of $\Lawv$-modules is a functor of the associated $\Lawv$-categories.
-Conversely, complete $\Lawv$-categories are \emph{tensored}, that is, admit a $\Lawv$-module structure $\epsilon \otimes x$ (see the Appendix).
+Conversely, complete $\Lawv$-categories are \emph{tensored} (cf.~\cite{Stubbe2014}), and thus admit a suitable $\Lawv$-module structure $\epsilon \otimes x$.
 Moreover, a continuous functor between complete $\Lawv$-categories is the same as a homomorphism of the associated $\Lawv$-modules. 
@@
 
-To conclude, we show how the correspondence $\Mod\simeq \GMet$ lifts to a model of the differential $\lambda$-calculus, extending the co-Kleisli category $\LREL_{!}$.
-
+To relate this elegant correspondence with our work on $\LREL_{!}$, we show that it lifts to a model of the differential $\lambda$-calculus, extending the co-Kleisli category $\LREL_{!}$. 
 First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies2018, Manzo2013}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting suitable properties of $\Mod$, we may define $!$ as the product of the symmetric algebras.
@@
 Moreover, the constructions for $\Mod$ generalize those of $\LREL$, in the sense that $!(\Lawv^{X})\simeq \Lawv^{\multiset(X)}$ and that $\Mod_{!}(\Lawv^{X},\Lawv^{Y})\simeq \LREL_{!}(X,Y)$.
-
-Finally, general results {\color{red}CITARE} allow us to conclude 
+Finally, as the coKleisli category of a Lafont category with biproducts is always a CC$\partial$C \cite{LemayCALCO2021}, we can endow $\Mod_{!}$ with a suitable differential operator $E$ generalizing $D^{!}$ (see the Appendix) so that:
 \begin{theorem}
-$\Mod_{!}\simeq\GMet_{!}$ is a CC$\partial$C, when equipped with 
-the differential operator $E$ defined, for $f:!M\to N$, by 
-$Ef(\alpha)=
-\bigvee\left\{
-f(\beta\cup [x]) \ \Big \vert  \ 
-\iota_{n}(\beta)\otimes \iota_{1}(x) \leq S(\alpha)
-\right\}
-$, where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
-and $S: !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{Mellies2018}.
+($\Mod_{!}/\GMet_{!},E$) is a CC$\partial$C.
+%, when equipped with 
+%the differential operator $E$ defined, for $f:\ !M\to N$, by 
+%$Ef(\alpha)=
+%\bigvee\left\{
+%f(\beta\cup [x]) \ \Big \vert  \ 
+%\iota_{n}(\beta)\otimes \iota_{1}(x) \leq S(\alpha)
+%\right\}
+%$, where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
+%and $S: \ !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{Mellies2018}.
 \end{theorem}

### 2023-07-31 CSL24/sectionTaylip.tex
Commit dettagli
commit

@@
 
-We now have all elements necessary to relate the metric and differential analysis of higher-order programs in the tropical relational model.
+In this section we finally relate the metric and differential analysis of higher-order programs in the tropical relational model.
 
-The key ingredient is the notion of Taylor expansion $\Te{M}$ of a $\lambda$-term $M$, which is a set of terms of the differential $\lambda$-calculus defined as follows:
+The key ingredient is the notion of Taylor expansion $\Te{M}$ of a $\lambda$-term $M$. This is a set of terms of the differential $\lambda$-calculus defined inductively as: 
 %
@@
 %The Taylor expansion of a $\lambda$-term is a set of resource terms $\Te{M}$ defined inductively as 
-%
 $\Te{x}=\{x\}$, $\Te{\lambda x.M}=\{\lambda x.t\mid t\in \Te{M}\}$ and 
 $\Te{MN}=\{t\cdot \langle u_{1},\dots, u_{k}\rangle  \mid k\in \mathbb N, t\in \Te{M}, u_{i}\in \Te{N} \}$, where $t\cdot \langle u_{1},\dots, u_{k}\rangle$ is an abbreviation for $\mathsf D^{k}[t, u_{1},\dots, u_{k}] 0$.
-Observe that in the terms appearing in $\Te{M}$ all applications are always \emph{bounded}: they may use an exact number of copies of their input. 
-Such terms are usually called \emph{resource $\lambda$-terms} \cite{Pagani2009, Manzo2012}.
+Observe that in the terms appearing in $\Te{M}$ all applications are \emph{bounded}: they may use an exact number of copies of their input. 
+Such terms are usually called \emph{resource $\lambda$-terms} \cite{Pagani2009, Manzo2012}. One can easily check that for all terms $\Gamma \vdash_{\STLC}M:A$ and  $t\in \Te{M}$, also 
+$\Gamma \vdash_{\STDLC}t:A$ holds. 
+\begin{example}
 Considering the term $M=zxx$ from Example \ref{ex:zxx}, all terms
  $t_{n,m}=z \,\langle x^{n}\rangle\,\langle x^{m}\rangle$, for $n,m\in \mathbb N$, are in $\Te{M}$. 
-Observe that the interpretation of $t_{n,m}$ yields a tropical polynomial
+Notice that the interpretation of $t_{n,m}$ yields a tropical polynomial
 $\model{t_{n,m}}^{!}(x)(z)= y_{[n,m]}+(n+m)x$, rather than a tps. 
-However, resource $\lambda$-terms need not always yield a tropical polynomial. For instance, consider $y:(*\to*)\to (*\to*), x:(*\to *) \vdash t:  (*\to  *)$
+However, this is not a general fact: consider $y:(o\to o )\to (o\to o), x:(o\to o) \vdash t:  (o\to  o)$
 with $t=y\cdot \langle y\cdot \langle x\rangle \rangle\in \Te{y(yx)}$. Then 
-$\model{t}^{!}: \Lawv^{  \mathcal M_{\mathrm{fin}}(\mathbb N)\times \mathbb N}
+$\model{t}^{!}: \Lawv^{  !\mathbb N\times \mathbb N}
 \times \Lawv^{\mathbb N} \to \Lawv^{\mathbb N}$ is given by
@@
 $, 
-which is not a polynomial. Yet, $\model{t}^{!}$ is Lipschitz, more precisely, 1-Lipschitz in $x$ and 2-Lipschitz in $y$. This is a general fact, as shown below.
+which is not a polynomial. Yet, $\model{t}^{!}$ is Lipschitz, more precisely, 1-Lipschitz in $x$ and 2-Lipschitz in $y$. This is a general fact, as shown by Theorem \ref{thm:taylor} below.
+\end{example}
 
@@
 We have already shown that the tropical differential makes $\LREL_{!}$ a model of the differential $\lambda$-calculus.
-We now show that it also models the Taylor expansion (this needs not be true for \emph{any} CC$\partial$C).
-First, it can be patiently checked that in $(\LREL_!,D)$ all morphisms can be Taylor expanded  (see \cite[Definition 4.22]{Manzo2012}):
-
+We now show that it also models the Taylor expansion (this needs not be true for \emph{any} CC$\partial\lambda$C).
+First, it can be patiently checked that (see \cite[Definition 4.22]{Manzo2012}):
+%
+% in $(\LREL_!,D)$ all morphisms can be Taylor expanded  (see \cite[Definition 4.22]{Manzo2012}):
 \begin{theorem}\label{thm:modelsTaylor}
- For all $t\in\HOM{\LREL_!}{Z}{!X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$ we have:%, the evaluation of $t$ over $s$ yields 
- \begin{align}\label{eq:taylorcat}
-  \RM{ev}\circ_!\langle t,s\rangle =
+Morphisms in $(\LREL_{!},D)$ can be Taylor-expanded: for all $t\in\HOM{\LREL_!}{Z}{!X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$ we have %, the evaluation of $t$ over $s$ yields 
+% \begin{align}\label{eq:taylorcat}
+  $\RM{ev}\circ_!\langle t,s\rangle =
   \inf\limits_{n\in\N}
-  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
- \end{align} 
+  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.$
+% \end{align} 
 \end{theorem}
 %It is worth discussing the formula above a bit more. 
-The equation above is a nothing more than a tropical version of the Taylor formula from the Introduction:
+The equation above is a tropical reformulation of the Taylor formula from the Introduction:
 %:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
@@
 $u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
-Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$.
+Hence the right-hand term corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$.
 %,  i.e.~it coincides with the tropical %interpretation of the
@@
 %Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
+%Moreover, one can show that the metrics $d_{\infty}$ are compatible with the Taylor formula in the following sense:
+%\begin{proposition}
+%For all tps $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
+%the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz. Moreover 
+%$\norm{ \matr f-\matr g}_{\infty}= \sup_{n} \norm{ {\delta^{(n)}f}- {\delta^{(n)}g}}_{\infty}$, 
+%where $\delta^{(n)}h$ indicates the matrix of $D_{!}^{(n)}h$.
+%%where $\delta^{(n)}f:( \Lawv^{X})^{n}\to \Lawv^{X}$ is the tropical linear function $\delta^{(n)}f(\B x_{1},\dots, \B x_{n})=
+%%(\Der^{(n)}f)(\B x_{1},\dots, \B x_{n}, \infty)$. 
+%\end{proposition}
 
@@
 
-Using the results of the previous section, as well as the results above, we can deduce the following properties:
+Using the results of the previous section, as well as the results above, we now deduce the following properties:
 
@@
 
-\item For $\mathcal S=\STDLC$, if $M\in \Te{M'}$, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is Lipschitz;
+\item For $\mathcal S=\STDLC$, if $t\in \Te{M}$, then $\model{\Gamma\vdash_{\mathcal S}t:A}^{!}_{a}$ is Lipschitz;
 
@@
 \begin{proof}
-1). As we already observed, the interpretation of a bounded term is a tropical polynomial.
-2.) From Proposition \ref{prop:troplinear} 2.~observing that a resource term $t(x)$ may use a variable $x$ a fixed number $n$ times, so that its matrix lies in $\Lawv^{!_{n}X\times Y}$. 
+1). From Proposition~\ref{prop:troplinear} 2.~and the remark that for any type of $\BSTLC$, $\model{A}$ is finite.
+2.) From Proposition~\ref{prop:troplinear} 2.~observing that a resource term $t(x)$ may use a variable $x$ a fixed number $n$ times, so that its matrix lies in $\Lawv^{!_{n}X\times Y}$. 
 %Now we apply Corollary~\ref{prop:polylip} to each coordinate of the image, and by taking the maximum Lipschitz constant among the finite number $\mathrm{Card}(\model A)$ of them, we obtain the thesis.
 3). From Theorem~\ref{thmTLSlocLip}.
-4).
-It follows from \autoref{cor:T(M)=M} plus the fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have $\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with $\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
+4). It follows from \autoref{cor:T(M)=M} plus the fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have 
+%$\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with 
+$\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
 \end{proof}
@@
 \begin{example}
-Consider again the term $M=zxx$ from Example \ref{ex:zxx}-
+Consider again the term $M=zxx$ from Example \ref{ex:zxx}. 
 %, satisfying $x:*, z:(*\to*\to*) \vdash M:  *$. Since $\mathcal M_{\mathrm{fin}}(\{*\})\simeq \mathbb N$, we have that 

### 2023-08-02 CSL24/appSection3.tex
commit

--- 2023-01-24 appV.tex
@@
-\subsection{Section~\ref{sec:4A}: Proof of Proposition~\ref{prop:nondecr+conc}}
-
-We just show the concavity, the rest being easy.
-
-A function $f:Q^{X}\to Q^{Y}$ is \emph{concave} if for all $\alpha\in [0,1]$, $ x ,  y  \in Q^{X}$ and $b\in Y$ 
-$$
-f(\alpha\cdot  x +(1-\alpha)\cdot  y  )_{b} \geq \alpha f( x )_{b} + (1-\alpha)f( y  )_{b}
-$$
-
-\begin{proposition}
-All tLs $f: Q^{X}\to Q^{Y}$ are concave.
-\end{proposition}
-\begin{proof}
-Let us first show that all functions of the form $f( x )_{b}= \mu  x + c$ are concave:
-we have $f(\alpha x + (1-\alpha) y  )_{b}= \mu(\alpha x )+(1-\alpha) y  )+c=
- \mu(\alpha x )+(1-\alpha) y  )+\alpha c+(1-\alpha)c=
- \alpha(\mu  x  + c)+(1-\alpha)(\mu  y  +c)=\alpha f( x )_{b}+(1-\alpha) f( x )_{b}$.
-
-To conclude, let us show that if $(f_{i})_{i\in I}$ is a family of concave functions from $Q^{X}$ to $Q^{Y}$, the function $f=\inf_{i\in I}f_{i}$ is also concave: we have
-$f(\alpha x  +(1-\alpha) y  )_{b}=
-\inf_{i\in I}f_{i}(\alpha x +(1-\alpha) y  )_{b} \geq 
-\inf_{i\in I}\alpha f_{i}( x )_{b}+(1-\alpha)f_{i}( y  )_{b}
-\geq 
-\inf_{i\in I}\alpha f_{i}( x )_{b} + \inf_{j\in I}(1-\alpha)f_{j}( y  )_{b}
-=
-\alpha \cdot (\inf_{i\in I}f_{i}( x )_{b})+ (1-\alpha)\cdot( \inf_{j\in I}f_{j}( y  )_{b})=
-\alpha  f( x )_{b}+(1-\alpha)f( y  )_{b}$, where we used the fact that given families $a_{i},b_{i}$ of reals,
-$\inf_{i}a_{i}+b_{i}\geq \inf_{i}a_{i}+\inf_{j}b_{j}$.
-This follows from the fact that for all $i\in I$, $a_{i}+b_{i}\geq \inf_{i}a_{i}+\inf_{i}b_{i}$.
-\end{proof}
-
-%
-%
-%Concavity is a useful property, as it leads to establish the following:
-%\begin{proposition}
-%For any tropical function $f: Q^{X}\to Q^{Y}$, 
-%\begin{enumerate}
-%\item $ \alpha f( x )_{b}\leq  f(\alpha x )_{b}$;
-%\item $f( x + y  )_{b}\leq f( x )_{b}+f( y  )_{b}$;
-%\item $f( x + y  )_{b} \leq f( x )_{b} + \phi_{\mu,b}( y  )$ 
-%for all $\mu\in \C M_{\mathrm{fin}}(X)$.
-%
-%\item if $ x  <  y   <  z  $, then 
-%$$
-%\frac{f( y  )-f( x )}{| y  - x |} \leq \frac{f( z  )-f( x )}{| z  - x |}
-%\leq \frac{f( z  )-f( y  )}{| z  - y  |}
-%$$
-%
-%\item the function $R_{b}( x ,  y  )= \frac{f( x )_{b}-f( y  )_{b}}{|  x -  y  |}$ is monotonically non-decreasing in $ x $ (for fixed $ y  $) and in $ y  $ (for fixed $ x $).
-%
-%\end{enumerate}
-%\end{proposition}
-%\begin{proof}
-%Since $f$ is concave, we have that for all $ z  $, 
-%$f(\alpha\cdot z  )=f(\alpha\cdot z  +(1-\alpha)\cdot 0)\geq 
-%\alpha f( z  )+(1-\alpha)f(0)\geq \alpha f( z  )$. 
-%
-%For the second one, 
-%if $f( x )=\mu  x +c$, then $f( x + y  )= \mu  x + \mu  y   + c  \leq
-%\mu  x  + \mu  y   + 2c = f( x )+f( y  )$.
-%Moreover, if $(f_{i})_{i\in I}$ is a family of functions such that $f_{i}( x + y  )\leq f_{i}( x )+f_{i}( y  )$, then 
-%$f=\inf _{i\in I}f_{i}$ also satisfies this property: 
-%one has $f( x + y  )=\inf_{i\in I}f_{i}( x + y  ) \geq \inf_{i} f_{i}( x )+f_{i}( y  ) \geq \inf_{i}f_{i}( x ) + \inf_{i}f_{i}( y  )=f( x )+f( y  )$.
-%%Now we can compute
-%%\begin{align*}
-%%f( x + y  )_{b}& =
-%%
-%%\end{align*}
-%
-%The third one follows immediately from the second one, since $f( y  )_{b}\leq \phi_{\mu,b}( y  )$ for all $\mu\in \C M_{\mathrm{fin}}(X)$.
-%
-
-%\end{proof}
-
-%\newpage
-
-\subsection{Section~\ref{sec:4A}: Proof of Theorem~\ref{theorem:fepsilon}}
+%\subsection{Section~\ref{sec:4A}: Proof of Theorem~\ref{theorem:fepsilon}}
 
@@
 \begin{theorem}[Theorem \ref{theorem:fepsilon}]
-Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\matr f:\N^k\to\Lawv$.
+Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tps with matrix $\matr f:\N^k\to\Lawv$.
  For all $0<\epsilon<\infty$, there is $\C F_\epsilon \subseteq \N^k$ such that:
@@
 \end{proof}
-
-%\newpage
-
-\subsection{Section~\ref{subsec:cont}: proof of Theorem~\ref{thm:cont}}
-
-In order to prove Theorem~\ref{thm:cont}, we place ourselves in the general setting of locally convex topological vector spaces (LCTVS), and prove a general result about convex functions in this setting, adapting, in the main Proposition~\ref{prop:mainLCTVS}, an argument taken from [Proposition 4.4, \cite{Cobzas2017}].
-
-\begin{definition}
- A \emph{normed semiring} is the data of a semiring $R$ together with a \emph{norm} on it, that is, a function $\absv{.}\,:R\to\R$ satisfying the usual axioms of the absolute value function\footnote{I.e.: $\absv{x}\geq 0$, $\absv x =0 \textit{ iff } x=0$, $\absv{xy}=\absv x\absv y$ and $\absv{x+y}\leq \absv x+\absv y$.}.
- One can immediately check that if $R$ is a ring, then any norm on it induces a metric $d(r,s):=\mid r-s\mid$ on it (and thus also a topology).
- A \emph{topological normed ring} is a normed ring which is a topological ring w.r.t.\ the topology induced by the norm distance.
-\end{definition}
-
-\begin{definition}
- Let $R$ be normed semiring, let $A$ be a topological $R$-module and let $C\subseteq A$.
- We say that $C$ is \emph{convex} iff for all $x,y\in C$ and for all $r,s\in R$ s.t. $\mid r\mid +\mid s\mid =1$, we have $rx+sy\in C$.
- We say that $C$ is \emph{balanced} iff for all $x\in C$ and for all $r,s\in R$ s.t. $\mid r\mid +\mid s\mid =1$, we have $rx,sx\in C$.
- We say that $C$ is a \emph{disk at $x\in A$} iff it is of shape $C=x+D$ for some $D$ both convex and balanced.
-\end{definition}
-
-\begin{example}
- An example of a balanced but not convex set in (the normed field) $\R^2$.
- \[\begin{tikzpicture}
- \draw[->] (0,-1) -- (0,1);
- \draw[->] (-1,0) -- (1,0);
- \draw[fill=gray, fill opacity=0.3] (0,0) -- (0.5,-0.5) -- (0.5,0.5) -- (0,0) -- (-0.5,0.5) -- (-0.5,-0.5) -- cycle;
-\end{tikzpicture}\]
-\end{example}
-
-\begin{definition}
- Let $R$ be a normed field and let $\mathbb{V}$ be a topological $R$-vector space.
-We say that $\mathbb{V}$ is \emph{locally convex} iff every $x\in \mathbb{V}$ admits a local basis consisting of disks at $x$.
- Since the topology of $\mathbb{V}$ is necessarily translation invariant,
- it is equivalent to only ask that $0$ has such a local basis.
-\end{definition}
-
-The main result is the following:
-
-\begin{proposition}\label{prop:mainLCTVS}
- Let $\mathbb{V}$ be a locally convex topological $\R$-vector space (where $\R$ is endowed with its usual absolute value and the induced topology by it).
- Fix $x\in\mathbb{V}$, a neighbourhood $V\subseteq\mathbb{V}$ of $x$ and $f:V\to \R\cup\set{+\infty,-\infty}$.
- If $V$ is convex, $f$ is concave and lower bounded by a finite constant $K$, then $f$ is continuous at $x$ (w.r.t.\ the subspace topology on $V$).
-\end{proposition}
-\begin{proof}
- Since $V$ is a neighbourhood of $x$, $x$ admits an open neighbourhood $U\subseteq V$, and by local convexity of $\mathbb{V}$ there is a (non necessary open) disk $D\subseteq U$ at $x$.
- Now fix $\eps\in(0,1)$ and $r,s\in \R$ with $0\leq r\leq\eps$, $s=1-r$.
- Fix also $w\in D$.
- By the convexity of $D$ we have $sx+rw\in D$, and thus the convexity of $f$ entails that:
- \[
-  f(sx+rw)\geq s f(x) + r f(w) \geq s f(x) + r K = (1-r)f(x) + r K
- \]
-that is,
- \[
-  f(x)-f(sx+rw)\leq r (f(x)-K).
- \]
- Now remark that $x=\dfrac{1}{s+2r}(sx+rw)+\dfrac{r}{s+2r}(2x-w)$, where $\dfrac{1}{s+2r}+\dfrac{r}{s+2r}=1$, $\dfrac{1}{s+2r}<1$ and $2x-w\in D$ (because $D$ is a disk at $x$).
- Therefore by the convexity of $f$ we have:
- \[
-  f(x)\geq \dfrac{1}{s+2r}f(sx+rw)+\dfrac{r}{s+2r}f(2x-w)\geq \dfrac{1}{s+2r}(f(sx+rw)+rK)
- \]
- that is,
- \[
-  f(sx+rw)-f(x)\leq r(f(x)-K).
- \]
- We have thus shown that $\mid f(sx+rw)-f(x)\mid\leq r(f(x)-K)$ for all $w\in D$.
- Since this holds also for all $\eps\in(0,1)$, and due to the choice of $r,s$, the points of shape $sx+rw$ span $D$ when $w$ spans $D$ and $\eps$ spans $(0,1)$.
- That is, we have shown that $\mid f(w)-f(x)\mid\leq r(f(x)-K)$ for all $w\in W$.
- Since $f(x)-K\geq 0$ and $r\leq \eps$, we have that $\exists\lim\limits_{w\to x} f(w)=f(x)$.
-\end{proof}
-
-Now we show how to apply this argument to our tropical case $\Lawv^X$, which is \emph{not} a LCTVS (because it is not a vector space).
-
-First, we have:
-
-\begin{corollary}\label{cor:cont}
- Let $f:({\R}_{> 0}^X,\supnorm{.}) \to (\overline\R_{\geq 0},\absv .)$ and $x\in {\R}_{> 0}^X$.
- If there is a convex neighbourhood $V\subseteq \overline{\R}_{> 0}^X$ of $x$ s.t.\ $f_{\lVert V}$ is concave, then $f$ is continuous at $x$.
-\end{corollary}
-\begin{proof}
- The $\R$-vector space $\R^X$ is topological w.r.t.\ the topology $\tau_\infty$ induced on it by the norm $\lVert .\lVert_\infty$, and it is clearly locally convex.
- Call $\tau^+_\infty$ the topology induced by $\supnorm{.}$ on ${\R}_{> 0}^X$.
- It is clear that it coincides with $\tau^+_\infty$.
- Since moreover ${\R}_{> 0}^X$ is open in $(\R^X,\tau_\infty)$, the neighbourhood $V$ of $x$ in $({\R}_{> 0}^X,\supnorm{.})$ is also a neighbourhood of $x$ in $(\R^X,\tau_\infty)$.
- We can therefore apply Proposition~\ref{prop:mainLCTVS} to $f_{\lVert V}$ (which is lower bounded by $0$ by definition of $f$) and obtain that $f_{\lVert V}$ is continuous at $x$ w.r.t.\ the subspace topology $\tau_V$ induced by $\tau_\infty$ on $V$.
- But since $V$ is contained in ${\R}_{> 0}^X$, the topology $\tau_V$ coincides with the subspace topology induced on $V$ by $\tau^+_\infty$.
- So $f$ is continuous at $x$ w.r.t.\ $\tau^+_\infty$.
-\end{proof}
-
-One may wonder if the same proof as above makes it possible to state the previous corollary replacing ${\R}_{> 0}^X$ with $\R_{\geq 0}^X$ (so, in particular, taking $x\in\R_{\geq 0}^X$).
-This is not possible because in the proof we crucially use that $\R_{> 0}^X$ is open in $(\R^X,\tau_\infty)$, which is not the case of $\R_{\geq 0}^X$.
-In fact, this allowed us to say that $V$ is a neighbourhood of $x$ in $(\R^X,\tau_\infty)$, and therefore to be able to apply Proposition~\ref{prop:mainLCTVS}.
-Taking $\R_{\geq 0}^X$ instead, this is in general not true: we could have a neighbourhood of $x$ w.r.t.\ the subspace topology on $\R_{\geq 0}^X$ induced by $\tau_\infty$, which does not contain any open neighbourhood of $x$ w.r.t.\ $\tau_\infty$ (i.e.
-it is not a neighbourhood w.r.t.\ $\tau_\infty$).
-
-\begin{example}
- An example of a neighbourhood of $x$ w.r.t.\ the subspace topology on $\R_{\geq 0}^X$ induced by $\tau_\infty$, which is not a neighbourhood w.r.t.\ $\tau_\infty$.
- \[\begin{tikzpicture}
- \draw[->] (0,-0.5) -- (0,1.5);
- \draw[->] (-0.5,0) -- (1.5,0);
- \filldraw[black] (0.8,0) circle (0.7pt) node[anchor=north]{x};
- \fill[fill=gray, fill opacity=0.3] (0.4,0) -- (1.2,0) -- (1.2,0.7) -- (0.4,0.7) -- (0.4,0) -- cycle;
-\end{tikzpicture}\]
-\end{example}
-
-Finally we have the desired result Theorem~\ref{thm:cont}:
-
-\begin{theorem}[Theorem~\ref{thm:cont}]
- Tropical Laurent series $f:({\R}_{\geq 0}^X,\supnorm .) \to (\overline\R_{\geq 0},\absv .)$ are continuous on ${\R}_{> 0}^X$ w.r.t.\ the norm $\norm{\cdot}_\infty$.
-\end{theorem}
-\begin{proof}
- We know that tLs's are concave on all their domain.
- Therefore the result immediate follows by Corollary \ref{cor:cont}, since the topology induced by $\supnorm{.}$ on ${\R}_{> 0}^X$ coincides with the subspace topology induced by $({\R}_{\geq 0}^X,\supnorm .)$ on it.
-\end{proof}
-
-Due to the previous discussion about the impossibility of stating Corollary \ref{cor:cont} in the case where one of the coordinates of $x$ is $0$, the continuity of tLs on the hyperplanes $\mathcal H_a:=\set{x\in {\R}_{\geq 0}^X \mid x_a=0}$, for $a\in X$, must be treated separately.
-Similarly, the continuity at points with infinite coordinates must be treated separately as well.
-We left it for future investigations.
-
-\begin{remark}
- We formulated the main argument (Proposition~\ref{prop:mainLCTVS}) in the general setting of LCTVS's.
-We did so just to show that it is actually a general argument, but we could have placed ourselves already in a more particular setting of our interest and carry on the proof.
-For instance, Theorem~\ref{thmTLSlocLip} will be proved by refining this kind of argument (Lemma~\ref{lm:mainLip}), but not in the setting of LCTVS.
-\end{remark}
-
-%\newpage
-\subsection{Section~\ref{subsec:cont}: proof of Theorem~\ref{thm:ScottCont}}
-
-In order to prove such result we must quickly recall the theory of normed cones, which we do below.
-
-Here we denote with $\overline{\R}_{\geq 0}$ the set $\R_{\geq 0}\cup\set{\infty}$ (the same as $\Lawv$, but we prefer to explicit it here).
-
-\subsubsection{Normed cones}
-
-\begin{definition}
- An \emph{$\overline{\R}_{\geq 0}$-cone} is a commutative $\overline{\R}_{\geq 0}$-semimodule with cancellative addition\footnote{I.e.: $x+y=x+y' \Rightarrow y=y'$.}.
-\end{definition}
-
-In \cite{Selinger2004} cones are required to also have ``strict addition'', meaning that $x+y=0 \Rightarrow x=y=0$.
-We do not add this requirement since it will automatic hold when considering normed cones.
-
-\begin{remark}
- The addition of a cone $P$ (which forms a commutative monoid) turns $P$ into a poset by setting:
- \[
-  x \leq y \textit{ iff } y=x+z \textit{, for some }z\in P.
- \]
- By the cancellative property, when such $z$ exists it is unique, and we denote it by $y-x$.
- Such order is called the \emph{cone-order} on $P$.
-\end{remark}
-
-\begin{definition}
- A \emph{normed $\overline{\R}_{\geq 0}$-cone} is the data of a $\overline{\R}_{\geq 0}$-cone together with a $\leq$-monotone\footnote{I.e.: $x\leq y \Rightarrow \norm{x}\leq \norm y$.
-Remark that requiring this property (for all $x,y$) is equivalent to requiring that $\norm{x}\leq \norm{x+y}$ for all $x,y$.} norm\footnote{A \emph{norm} on a $\overline{\R}_{\geq 0}$-abstract cone $P$ is a map $\norm{.}:P\to \overline{\R}$ satisfying the usual axioms of norms:
- $\norm x \geq 0$, $\norm x = 0 \Rightarrow x=0$, $\norm{rx}=r\norm x$ and $\norm{x+y}\leq \norm x + \norm y$.} on it.
-\end{definition}
-
-In, e.g.~\cite{EhrPagTas2018}, a normed $\overline{\R}_{\geq 0}$-cone is simply called a cone.
-
-Remark that in a normed $\overline{\R}_{\geq 0}$-cone, by monotonicity of the norm, we have: $\norm{x+y}=0 \Rightarrow x=y=0$.
-Therefore, as already mentioned, in a normed cone we have: 
-$x+y=0 \Rightarrow \norm{x+y}=0 \Rightarrow x=y=0$, that is, addition is strict.
-
-\begin{example}
- $\overline{\R}_{\geq 0}^X$ is a normed cone with the norm $\supnorm{x}:=\sup\limits_{a\in X} x_a\in \overline{\R}_{\geq 0}$.
-\end{example}
-
-\begin{remark}
- The cone-order on $\overline{\R}_{\geq 0}^X$ is the pointwise usual order on $\overline{\R}_{\geq 0}$.
-
- Remark also that tLs have no reason to be linear nor sublinear.
-\end{remark}
-
-\begin{remark}\label{rmk:tropMonot}
- tLs are monotone w.r.t.\ to the cone order on its domain and codomain.
- This is clear, using the definition of tLs, because $\mu x\leq \mu y$ if $x\leq y$.
-\end{remark}
-
-\begin{remark}
- The cone-order on $\overline{\R}_{\geq 0}^X$ makes it into a dcpo with least element $0$.
-\end{remark}
-
-The following is a usual domain theoretical lemma.
-
-\begin{lemma}\label{lm:sup=sup}
- Let $P$ be a poset and $X, I$ sets. Fix the pointwise order on $P^X$.
- \begin{enumerate}
-  \item Let $x^i\in P^X$ for $i\in I$.
-  If $\bigvee\limits_{i\in I} x^i$ exists $P^X$, then $\bigvee\limits_{i\in I} x^i_a$ exists in $P$ for all $a\in X$, and we have: $\bigvee\limits_{i\in I} x^i_a=\left(\bigvee\limits_{i\in I} x^i\right)_a$.
- \item Let $x^i_a\in P$ for $i\in I,a\in X$.
-  If $\bigvee\limits_{i\in I} x^i_a$ exists $P$ for all $a\in X$, then $\bigvee\limits_{i\in I} x^i$ exists in $P^X$, and we have: $\left(\bigvee\limits_{i\in I} x^i\right)_a=\bigvee\limits_{i\in I} x^i_a$.
- \end{enumerate}
-\end{lemma}
-
-A \emph{directed net} in a poset $P$ with indices in a set $I$ is a function $s:I\to P$, denoted by $(s_i)_{i\in I}$, s.t.\ its image is directed.
-We say that a directed net in $P$ \emph{admits a sup} iff its image admits a sup in $P$.
-We say that a directed net $s$ in a normed cone is \emph{bounded} iff the set $\set{\norm{s_i}\,\mid i\in I}$ is bounded in $\R_{\geq 0}$.
-
-Remember the definition of Scott-continuity:
-
-\begin{definition}
- A function $f:P\to P'$ between posets is \emph{Scott-continuous} iff for all directed net $(s_i)_i$ in $P$ admitting a sup, we have $\exists \bigvee\limits_i f(s_i) = f(\bigvee\limits_i s_i)$ in $P'$.
-\end{definition}
-
-The fundamental result in order to prove Theorem~\ref{thm:ScottCont} is the following, taken from \cite{Selinger2004}.
-
-\begin{proposition}\label{prop:infsup}
- Let $P$ be a normed $\overline{\R}_{\geq 0}$-cone s.t.\ every bounded directed net in $P$ admits a sup.
- Let $(v_i)_{i\in I}$ be a directed net in $P$ with an upper bound $v\in P$.
- Then $\exists\bigvee\limits_{i\in I} v_i \in P$ and, if $\inf\limits_{i\in I} \norm{v-v_i} =0$, one has: $\bigvee\limits_{i\in I} v_i = v$.
-\end{proposition}
-\begin{proof}
- Remark that $v-v_i$ exists in $P$ by hypothesis and so does $\bigvee\limits_{i\in I} v_i$, thanks to the monotonicity of the norm.
- Now, since $v\geq v_i$ for all $i$, we have that $v\geq \bigvee\limits_{i\in I} v_i$, and so $v-\bigvee\limits_{i\in I} v_i$ exists in $P$.
- Fix $i\in I$.
- Since $v_i\leq \bigvee\limits_{i\in I} v_i$, then $v-\bigvee\limits_{i\in I} v_i\leq v-v_i$ and, by monotonicity of the norm, $\norm{v-\bigvee\limits_{i\in I} v_i}\leq \norm{v-v_i}$.
- Since this holds for all $i\in I$, we have:
- $0\leq \norm{v-\bigvee\limits_{i\in I} v_i}\leq \inf\limits_{i\in I} \norm{v-v_i}=0$, where the last equality holds by hypothesis.
- Thus $\norm{v-\bigvee\limits_{i\in I} v_i}=0$, i.e.\ $v=\bigvee\limits_{i\in I} v_i$.
-\end{proof}
-
-\begin{definition}
- A normed $\overline{\R}_{\geq 0}$-cone $P$ is \emph{Scott-complete} iff its norm is Scott-continuous (where the codomain $\overline\R_{\geq 0}$ is endowed with its usual order) and every bounded directed net in $P$ admits a sup.
- This is equivalent to asking that its closed unit ball is a dcpo.
-\end{definition}
-
-\begin{proposition}
- The normed cone $\overline{\R}_{\geq 0}^X$ is Scott-complete.
-\end{proposition}
-\begin{proof}
- $\overline{\R}_{\geq 0}^X$ being a dcpo, all the existences of sup's that we should check do automatically hold, so we only have to show that $\bigvee$ and $\supnorm{.}$ commute:
- $\supnorm{\bigvee\limits_i x_i}=
- \sup\limits_a\, \sup\limits_i\, (x_i)_a =
- \sup\limits_i\, \sup\limits_a\, (x_i)_a =
- \bigvee\limits_i \supnorm{x_i}$.
-\end{proof}
-
-%Let us set $\infty_b$ to be the hyper-semi-plane $\set{y\in\overline{\R}^Y_{\geq 0}\mid y_b=+\infty}$ and $\set{f<<+\infty}:=\bigcup\limits_{b\in Y} f^{-1}\infty_b$.
-
-%\begin{remark}\label{rmk:R-f<<oo}
-%Let $f:\overline{\R}^X_{\geq 0}\to\overline{\R}^Y_{\geq 0}$ be tropical.
-%Then, $\R^X_{> 0}-\set{f<<+\infty}$ is a dcpo (w.r.t.\ the usual order, that is, its cone order).
-%In order to see this, it is enough to show that, for any $b\in Y$ and for any directed net $(x^i)_i$ in $\R^X_{> 0}$ s.t.\ $f(x^i)_b<+\infty$ for all $i$, we have $f(\bigvee\limits_i x^i)_b<+\infty$ (where the sup is taken in $\R^X_{> 0}$).
-%Such property holds because otherwise it is easy to see that $f$ would be discontinuous at the point $\bigvee\limits_i x^i\in\R^X_{> 0}$ w.r.t.\ the topology\footnote{Remark that this argument would not be possible using Scott-continuity instead.
-Indeed, it could well be that $f(\bigvee\limits_i x^i)_b=\bigvee\limits_i f(x^i)_b=+\infty$ while $f(x^i)_b<+\infty$ for all $i$.} induced by $\supnorm{.}$.
-%This contradicts Corollary \ref{cor:tropCont}.
-%\end{remark}
-
-We finally obtain the desired:
-
-\begin{theorem}[Theorem~\ref{thm:ScottCont}]
- All tLs $f:\overline{\R}^X_{\geq 0}\to\overline{\R}^Y_{\geq 0}$ are Scott-continuous on $\R^X_{> 0}$ w.r.t.\ the cone-orders on its domain and codomain.
-\end{theorem}
-\begin{proof}
- Let $(x_i)_i$ a directed net in $\R^X_{> 0}$ s.t.\ $\bigvee\limits_i x^i$ exists in $\R^X_{> 0}$.
- Then $\inf\limits_i \supnorm{\bigvee\limits_i x^i - x^i} =0$, where $\bigvee\limits_i x^i - x^i$ exists because $\bigvee\limits_i x^i \geq x^i$ for all $i$.
- Since $f$ is $\supnorm{.}$-continuous on $\R^X_{> 0}$, then $\inf\limits_i \supnorm{f(\bigvee\limits_i x^i) - f(x^i)} =0$, where $f(\bigvee\limits_i x^i) - f(x^i)$ exists because $f(\bigvee\limits_i x^i) \geq f(x^i)$ for all $i$ being $f$ monotone (Remark \ref{rmk:tropMonot}).
- We can therefore apply Proposition \ref{prop:infsup} to the directed net $(f(x^i))_i$ in $\overline{\R}^Y_{\geq 0}$, obtaining that $\bigvee\limits_i f(x^i)$ exists in $\overline{\R}^Y_{\geq 0}$ and it coincides with $f(\bigvee\limits_i x^i)$.
-\end{proof}
-
-%\begin{remark}
-%By looking at the definition of tropical functions, for $f$ tropical we have that: if $x,x'\in\R^X_{> 0}-\set{f<<+\infty}$ and $r\in\R_{> 0}$, then $x+x',rx\in\R^X_{> 0}-\set{f<<+\infty}$.
-%That is, $\R^X_{> 0}-\set{f<<+\infty}$ is a $\R_{>0}$-cone.
-%It is clearly still Scott-complete.
-%\end{remark}
-
-In the main paper we mention that $\R_{\geq0}^X$ is a Scott-complete dcpo.
-We did not state it as a theorem because it follows from standard and well-known domain theoretical considerations.
-Let us prove it here anyway.
-
-Recall the definition of Scott-complete dcpo:
-
-\begin{definition}
- Let $P$ be a dcpo and define a relation by: $x<<y$ iff for all directed $D\subseteq P$, if $y\leq \bigvee D$, then $x\leq d$, for some $d\in D$.
- Call $\twoheaddownarrow y:=\set{x\in P\mid x<<y}$.
- A dcpo $P$ is \emph{Scott-continuous} iff $\twoheaddownarrow x$ is directed and $x=\bigvee\twoheaddownarrow x$ for all $x\in P$.
-\end{definition}
-
-\begin{remark}
- In any dcpo we have: $\twoheaddownarrow x \subseteq \downarrow x$.
- This immediately follows by considering the directed set $\set{x}$.
-\end{remark}
-
-\begin{lemma}\label{lm:<<dir}
- In $\overline{\R}_{\geq 0}^X$, every set $\twoheaddownarrow x$ is directed.
-\end{lemma}
-\begin{proof}
- It is immediate that $0\in\twoheaddownarrow x$, so it is non-empty.
- Now let $y,y' << x$.
- Since we are in $\overline{\R}_{\geq 0}^X$, there is $y\vee y'\in\overline{\R}_{\geq 0}^X$.
- So we only have to show that $y\vee y'<<x$.
- For that, let $D$ be a directed set in $\overline{\R}_{\geq 0}^X$ s.t.\ $x\leq\bigvee D$.
- Since $y,y' << x$ we find $d,d'\in D$ s.t.\ $y\leq d$, $y'\leq d'$.
- Since $D$ is directed, there is $\hat d\in D$ s.t.\ $y\leq d\leq \hat d\geq d'\geq y'$.
- But then, by definition of sup, it must be $\hat d\geq y\vee y'$ and we are done.
-\end{proof}
-
-\begin{remark}
- The following is a known property (that we will not use):
- let $P$ be a complete normed cone which is a dcpo w.r.t.\ its cone-order. Then $P$ is Scott-continuous as a dcpo iff its closed unit ball is Scott-continuous as a dcpo.
-\end{remark}
-
-\begin{remark}\label{rmk:<<R}
- Consider $\overline{\R}_{\geq 0}$ with its usual order (which coincides with its cone-order).
- It is easily seen (and well known) that $y<<x$ iff either $y=0$ or $y<x$.
- This immediately implies that $\overline{\R}_{\geq 0}$ is Scott-continuous, since $x=\sup\limits_{y<x} y$.
-\end{remark}
-
-\begin{lemma}\label{lm:<<RX}
- Let $x,y\in \overline{\R}_{\geq 0}^X$ (considered as a dcpo with its cone order, which is the pointwise one). Fix $a\in X$.
- If $y_a<<x_a$ and $y_c=0$ for all $c\neq a$, then $y<<x$.
-\end{lemma}
-\begin{proof}
- Towards a contradiction, assume that there is a directed set $D$ in $\overline{\R}_{\geq 0}^X$ with $x\leq\bigvee D$ and s.t.\ $y\not\leq d$ for all $d\in D$.
- Call $D_a:=\set{d_a\mid d\in D}$ and remark that it is directed, because $D$ is.
- Also, by Lemma \ref{lm:sup=sup}, $\bigvee D_a = \left(\bigvee D\right)_a$.
- Therefore from $y_a<<x_a\leq \bigvee D_a$ we obtain a $d\in D$ s.t.\ $y_a\leq d_a$.
- By the absurd hypothesis we have $y\not\leq d$, so there must be $c\in X$ s.t.\ $y_c\not\leq d_c$, i.e. (because real numbers are totally ordered) $y_c>d_c$.
- Therefore it must be $c\neq a$.
- But then we have $0=y_c>d_c\geq 0$, contradiction.
-\end{proof}
-
-\begin{lemma}\label{lm:<<_a}
- In $\overline{\R}_{\geq 0}^X$ we have:
- if $y<<x$ then $y_a<< x_a$ for all $a\in X$.
-\end{lemma}
-\begin{proof}
- Fix $a\in X$.
- Let $D$ directed set in $\overline{\R}_{\geq 0}$ s.t.\ $x_a\leq\bigvee D$.
- We look for a $d\in D$ s.t.\ $y_a\leq d$.
- For $d\in D$, let $x^{a,d}\in\overline{\R}_{\geq 0}^X$ defined by $x^{a,d}_c:=x_c$ if $c\neq a$ and $x^{a,d}_a:=d$.
- Let $D^x_a:=\set{x^{a,d}\mid d\in D}$.
- By Lemma \ref{lm:sup=sup} $D^x_a$ admits sup in $\overline{\R}_{\geq 0}^X$ and it is $(\bigvee D^x_a)_c=x_c$ if $c\neq a$ and $(\bigvee D^x_a)_a=\bigvee D$.
- Hence, $x\leq \bigvee D^x_a$.
- If we prove that $D^x_a$ is directed, we are done: indeed, since $y<<x$, there is $d\in D$ s.t.\ $y\leq x^{a,d}$ and thus, in particular, $y_a\leq x^{a,d}_a=d$.
- Let us finally prove that $D^x_a$ is directed: it is clearly non-empty, since $D$ is.
- Let now $d,d'\in D$.
- We want to show that there is $\hat d\in D$ s.t.\ $x^{a,d}\leq x^{a,\hat d}\geq x^{a,d'}$.
- But since $D$ is directed, there is $\hat d\in D$ s.t.\ $d\leq \hat d\geq d'$, and therefore $x^{a,d}_c=x_c= x^{a,\hat d}_c =x_c=x^{a,d'}_c$ for all $c\neq a$, and $x^{a,d}_a=d\leq \hat d=x^{a,\hat d}_a=\hat d \geq d'=x^{a,d'}_a$.
-\end{proof}
-
-\begin{lemma}\label{rmk:sup_a=sup_a}
-In $\overline{\R}_{\geq 0}^X$ we have:
-$\bigvee\limits_{y<<x_a} y = \bigvee\limits_{y<<x} y_a$.
-\end{lemma}
-\begin{proof}
- If we show that $\twoheaddownarrow x_a = \set{y_a\mid y\in\twoheaddownarrow x}$, then we are done because in the statement we are taking their sup's.
- The inclusion ($\supseteq$) immediately follows from Lemma \ref{lm:<<_a}. For ($\subseteq$), let $d<< x_a$.
- Then the $y\in \overline{\R}_{\geq 0}^X$ defined by $y_c:=0$ if $c\neq a$ and $y_a:=d$, is s.t.\ $y\in\twoheaddownarrow x$ by Lemma \ref{lm:<<RX}.
- Thus, $d\in \set{y_a\mid y\in\twoheaddownarrow x}$.
-\end{proof}
-
-\begin{corollary}
- The dcpo $\overline{\R}_{\geq 0}^X$ is Scott-continuous.
-\end{corollary}
-\begin{proof}
- The fact that $\twoheaddownarrow x$ is directed is given by Lemma \ref{lm:<<dir}. The fact that $x=\bigvee \twoheaddownarrow x$ is given by the following equalities:
- $x_a=\bigvee\limits_{y<<x_a} y = \bigvee\limits_{y<<x} y_a = \left(\bigvee\limits_{y<<x} y\right)_a=\left(\bigvee \twoheaddownarrow x\right)_a$.
- The first equality follows from Remark \ref{rmk:<<R}, the second one from Lemma \ref{rmk:sup_a=sup_a}, the third one from Lemma \ref{lm:sup=sup}.
-\end{proof}
-
-%\newpage
-
-\subsection{Proofs from Section~\ref{sec:4C}}
-
-\subsubsection{Lipschitz continuity}
-
-  We need a first, preliminary, lemma:
-\begin{lemma}\label{lemma:inf}
-Let $u,v: I\to \Lawv$ and suppose $|u(i)-v(i)|\leq \delta$, for all $i\in I$.
-Then $|\inf_{i\in I}u(i)- \inf_{i\in I}v(i)|\leq \delta$. 
-\end{lemma}
-\begin{proof}
-Let $A=\inf_{i\in I}u(i)$ and $B=\inf_{i\in I}v(i)$ and suppose $A\geq B$. 
-Suppose by way of contradiction $|A-B|> \delta$; then there exists $i\in I$ such that 
-$v(i)<A$ and $|A-v(i)|> \delta$. Indeed, otherwise we would have $|A-B|= \sup\{ |A-v(i)|\mid v(i)\leq A\}\leq \delta$. 
-Now, from $|A-v(i)|> \delta$ and $v(i)< A$ we deduce that $|u(j)-v(i)|> \delta$ for all $j\in I$, and thus in particular that $|u(i)-v(i)|>\delta$, against the assumption.
-We conclude then $|A-B| \leq \delta$.
-In case $B\geq A$, we can argue in a similar way. 
-\end{proof}
-
-\begin{proposition}\label{prop:troplinear}
-All tropical linear functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive.  
-\end{proposition}
-\begin{proof}
-%All functions of the form $f_{\theta,\eta}(x)_{b}= \theta(b)+x_{\eta(b)}$, where $\theta:Y \to \Lawv$ and $\eta:Y\to X$, are non-expansive: indeed $|f_{\theta,\eta}(x)_{b}-f_{\theta,\eta}(y)_{b}|= |\theta(b)+x_{\eta(b)}-\theta(b)-y_{\eta(b)}|=
-%|x_{\eta(b)}-y_{\eta(b)}|\leq \norm{x-y}_{\infty}$.
-Let $f(x)_{b}=\inf_{a}\{\widehat f_{a,b}+x_{a}\}$.
-For all $a\in X$ we have that 
-$|(\widehat f_{a,b}+x_{a})- (\widehat f_{a,b}+y_{a})|=
-|x_{a}-y_{a}|\leq \norm{x-y}_{\infty}$. Using Lemma \ref{lemma:inf}, we deduce then 
-$|f(x)_{b}-f(y)_{b}|=|\inf_{a}\{\widehat f_{a,b}+x_{a}\}-
-\inf_{a}\{\widehat f_{a,b}+y_{a}\}|\leq \norm{x-y}_{\infty}$.
-\end{proof}
-
-\begin{lemma}\label{lm:mainLip}
-Let $f: [0,\infty)^{X}\to Q$ be concave, monotone increasing, and continuous.  
-Let $ x \neq  y  \in [0,\infty)^{X}$, with $\| x - y  \|_{\infty}<\infty$, and let $S( x ,  y  )=\{ \alpha x  + (1-\alpha) y  \mid \alpha \in [0,1]\}$ be the segment generated by $ x $ and $ y  $. 
-Then $f$ is Lipschitz-continuous over $S( x ,  y  )$.
-
-\end{lemma}
-\begin{proof}
-Let us prove the lemma under the assumption that for all $a\in X$, $ y  _{a}- x _{a}\geq 1$. From the fact that the claim holds under the assumption, we can deduce the claim of the lemma:
-indeed for $\alpha\in (0,1)$ large enough we have that $ y  ':= \frac{ y  -\alpha x }{1-\alpha}$ is such that $ y  \in S( x ,  y  ')$ (and thus $S( x ,  y  )\subseteq S( x ,  y  ')$) and 
-$ y  '_{a}- x _{a}\geq 1$. Hence from our proof we deduce that $f$ is Lipschitz-continuous over $S( x ,  y  ')$, and thus a fortiori over $S( x ,  y  )$ too.
-
-Since $f$ is continuous over $[0,\infty)^{X}$ and $S( x ,  y  )$ is compact, $f$ admits a maximum $\mathrm{MAX}$ over $S( x ,  y  )$.
-For all $ z  <  z  '\in S( x ,  y  )$, let $M( z  ,  z  ')\in Q^{X}$ be defined by
-$$
-M( z  ,  z  ')_{a}= \frac{f( z  ')-f( z  )}{ z  '_{a}- z  _{a}}
-$$
-Observe that
-\begin{align*}
-M( x ,  y  )_{a} & = \frac{f( y  )-f( x )}{ y  _{a}- x _{a}} \leq 
-f( y  )-f( x ) \leq \mathrm{MAX}
-\end{align*}
-using the fact that $ y  _{a}- x _{a}\geq 1$.
-
-We now claim that $M( z  ,  z  ')$ is contravariant in both $ z  $ and $ z  '$. 
-Indeed suppose $ z  \leq  z  '' <  z  '$, so that $ z  = \lambda  z  ' +(1-\lambda) z  ''$ for some $\lambda \in (0,1)$. Then, using the fact that $f$ is concave, we have 
-\begin{align*}
-M( z  ,  z  ')_{a}&=\frac{f( z  ')-f(\lambda  x  +(1-\lambda) z  '')}{ z  '_{a}-\lambda  x _{a} -(1-\lambda) z  ''_{a}} \\
-&
-\geq 
-\frac{f( z  ')-\lambda f( z  ') -(1-\lambda)f( z  '')}{ z  '_{a}-\lambda  z  '_{a} -(1-\lambda) z  ''_{a}} \\
-&
-= 
-\frac{(1-\lambda)(f( z  ')-f( z  ''))}{(1-\lambda ) z  '_{a}- z  ''_{a}} =M( z  '',  z  ')
-\end{align*}
-In a similar way it is shown that for $ z   <  z  ''\leq  z  '$, $M( z  , z  ')\leq M( z  ,  z  '')$.
-
-Therefore, for all $ z   <  z  '\in S( x ,  y  )$, we have that $M( z  ,  z  ')_{a} \leq M( x ,  z  ')_{a} \leq M( x ,  y  )_{a} \leq \mathrm{MAX}$. 
-From this, using the fact that $f$ is monotone increasing, we deduce that 
-$|f( z  ')-f( z  )|=f( z  ')-f( z  ) \leq \mathrm{MAX}\cdot | z  '_{a}- z  _{a}|
-$ and thus that 
-$$|f( z  ')-f( z  )|\leq \mathrm{MAX}\cdot \| z  '- z  \|_{\infty}$$
-that is, that $f$ is $\mathrm{MAX}$-Lipschitz over $S( x ,  y  )$. 
-\end{proof}
-
-\begin{proposition}
-Let $f: [0,\infty)^{X}\to Q$ be concave, monotone increasing, and continuous.  
-For all $\epsilon \in (0,\infty)$ and $ x \in [0,\infty)^{X}$, $f$ is Lipschitz-continuous over the open ball $B( x , \epsilon)$.
-\end{proposition}
-\begin{proof}
-Let $\mathrm{MAX}$ indicate the maximum of $f$ over $B( x , \epsilon)$. 
-Let $ y  ,  z  \in B( x , \epsilon)$; then $\| y  - z  \|_{\infty}\leq 2\epsilon < \infty$, so by the lemma above $f$ is $K$-Lipschitz over the segment $S( y  ,  z  )$ for some $K\leq \mathrm{MAX}$, so we deduce $|f( y  )-f( z  )|\leq \mathrm{MAX}\cdot \| y  - z  \|_{\infty}$.
-\end{proof}
-
-\begin{theorem}[local Lipschitz-continuity]
-Let $f: [0,\infty)^{X}\to \Lawv$ be concave, monotone increasing, and continuous.  
-Then $f$ is locally Lipschitz-continuous.
-\end{theorem}
-\begin{proof}
-For all $ x \in [0,\infty)^{X}$, $f$ is Lipschitz-continuous over the open set $B( x , 1)$.
-\end{proof}
-
-\subsubsection{Characterizations of the functional metric }
-
-  \begin{proposition}\label{prop:functionalmetric}
-For all maps $f,g: \Lawv^{X}\to \Lawv^{Y}$,  
-$
-  \|\widehat f- \widehat g\|_{\infty}=
-  \sup\{\|f( x )- g( x )\|_{\infty} \mid  x \in Q\langle X\rangle\}
-$.
-
-\end{proposition}
-
-%
-%\begin{lemma}
-%Let $f,g\in\mathsf{Trop}(1,1)$ (i.e.~$f,g:Q\to Q$), and suppose for all $n\in \BB N$, 
-%$|\widehat f_{n}-\widehat g_{n}|\leq \delta$. Then for all $x\in Q$, $|f(x)-g(x)|\leq \delta$.
-%
-%\end{lemma}
-%\begin{proof}
-%$|f(x)-g(x)|= |( \inf_{n}nx+\widehat f_{n})- (\inf_{n}nx+\widehat g_{n})|$. 
-%Since $|nx+\widehat f_{n} - nx -\widehat g_{n}|= |\widehat f_{n}-\widehat g_{n}|\leq \delta$, by the Lemma above we conclude $|f(x)-g(x)|\leq \delta$. 
-%\end{proof}
-
-  \begin{proof}[Proof of Proposition \ref{prop:functionalmetric}]
-For one side, suppose for all $\mu\in \C  M_{\mathrm{fin}}(X),b\in Y$, 
-$|\widehat f_{\mu,b}-\widehat g_{\mu,b}|\leq \delta$. Then for all $ x \in Q\langle X\rangle$ and $b\in Y$, $|f( x )(b)-g( x )(b)|\leq \delta$.
-Indeed, $|f( x )(b)-g( x )(b)|= |( \inf_{\mu}\mu  x +\widehat f_{\mu,b})- (\inf_{\mu}\mu  x +\widehat g_{\mu,b})|$. 
-Since $|\mu  x +\widehat f_{\mu,b} - \mu  x  -\widehat g_{\mu,b}|= |\widehat f_{\mu,b}-\widehat g_{\mu,b}|\leq \delta$, by Lemma \ref{lemma:inf} we conclude $|f( x )(b)-g( x )(b)|\leq \delta$.
-
-For the other side, suppose for some $\mu\in \C M_{\mathrm f}(X)$ and $b\in Y$, $|\widehat f_{\mu,b}-\widehat g_{\mu,b}|> \epsilon$; 
-then, by letting $e_{\mu}\in Q\langle !X\rangle$ be the (tropical) characteristic function of $\mu$, we have
- $|f(e_{\mu})({b})-g(e_{\mu})({b})| =
-|( \inf_{\mu'}\widehat f_{\mu',b}+ \mu'(e_{\mu}))-
-( \inf_{\mu'}\widehat g_{\mu',b}+\mu'(e_{\mu}))|=
-|\widehat f_{\mu,b}- \widehat g_{\mu,b}|> \epsilon$, so we deduce that 
-$\sup\{\| f( x )- g( x ) \|_{\infty}\mid  x \in Q\langle X\rangle\}> \epsilon$.
-\end{proof}
-
-The second characterization relates distances with the Taylor expansion. Let us briefly discuss the latter, first.
-For all $f: \Lawv^{X}\to\Lawv^{Y}$, let $\delta^{(n)}f:(\Lawv^{X})^{n}\to \Lawv^{Y}$ indicate the $n$-linear function given by 
-$$
-\delta^{(n)}f( x ^{n})= D^{(n)}f( x ^{n}, \infty)
-$$
-
-Notice that $\widehat{\delta^{(n)}f}\in Q^{X^{n}\times Y}$ satisfies 
-$\widehat{\delta^{(n)}f}_{a_{1},\dots, a_{n},b}= \widehat f_{[a_{1},\dots, a_{n}],b}$.
-In other words, $\delta^{(n)}f$ precisely captures the behavior of $f$ when applied to multisets of length $n$.
-Moreover, the full behavior of $f$ can be recovered from the functions $\delta^{(n)}f$ using the Taylor expansion which, in its tropical form, reads as:
-\begin{equation*}
-f( x )= \inf_{n}\left \{ D^{(n)}f( x ^{n},\infty)\right\}= \inf_{n}\left \{\delta^{(n)}f( x ^{n})\right\}
-%\tag{Tropical Taylor}
-\end{equation*}
-
-\begin{proposition}\label{prop:taylormetric}
-For all $f,g:\Lawv^{X}\to\Lawv^{Y}$, 
-$$
-\| \widehat f-\widehat g\|_{\infty}= \sup_{n}\| \widehat{\delta^{(n)}f} -\widehat{\delta^{(n)}g}\|_{\infty}
-$$
-\end{proposition}
-\begin{proof}
-Let us first show that $\| \widehat f-\widehat g\|_{\infty}\leq \epsilon$ implies 
-$\| \delta^{(n)}f-\delta^{(n)}g\|_{\infty}\leq\epsilon$, for all $n\in \BB N$. 
-Notice that $\widehat{\delta^{(n)}f}\in \Lawv^{X^{n}\times Y}$ satisfies 
-$\widehat{\delta^{(n)}f}_{a_{1},\dots, a_{n},b}= \widehat f_{[a_{1},\dots, a_{n}],b}$. Hence 
-from $\| \widehat f-\widehat g\|\leq\epsilon$ it follows that for all $\mu=[a_{1},\dots, a_{n}]$ and $b\in Y$, 
-$|\widehat f_{\mu,b}- \widehat g_{\mu,b}|\leq \epsilon$, so 
-we deduce that $\| \widehat{\delta^{(n)}f}- \widehat{\delta^{(n)}g}\|\leq \epsilon$.
-
-For the converse direction suppose that, for all $n\in \BB N$, $\| \widehat{\delta^{(n)}f}- \widehat{\delta^{(n)}g}\|\leq \epsilon_{n}$. Then, since the family of coefficients $
-F_{n,\mu,b}=
-( \widehat{\delta^{(n)}f})_{a_{1},\dots, a_{n},b}$ (where $\mu=[a_{1},\dots, a_{n}]$) is in bijection with the coefficients $\widehat f_{\mu,b}$, we deduce that $|\widehat f_{\mu,b}-\widehat g_{\mu,b}|\leq
-\epsilon_{\sharp \mu  }$, and thus that 
-$\| \widehat f-\widehat g\|_{\infty}\leq \sup_{n}\epsilon_{n}$. 
-\end{proof}

### 2023-08-02 CSL24/appSection5.tex
commit

--- 2023-01-22 appIII.tex
@@
-\subsubsection{The language $\BSTLC$}
-In Section 4 we considered a variant $\BSTLC$ of $\STLC$ with linear simple types and a graded exponential $!_{n}A$, for all $n\in \BB N$, corresponding to a somehow simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}.
-%The terms are generated by the grammar:
-%
-%{
-%\begin{minipage}{\textwidth}
-%\begin{align*}
-%M:= x\mid \lambda x.M\mid MM \mid !M \mid \mathsf{let}\  !x=M \ \mathsf{in}\  M%\mid (t,u)\mid  \mathsf{let}\  (x,y)=t \ \mathsf{in} \  u 
-%\end{align*}\end{minipage}}\medskip\\
-%For the purposes of this article we limited ourselves to a minimal fragment of this language. For a more practical language see \cite{Reed2010,Gaboardi2017}.
-
-The types are generated by the grammar below:
-
-{
-\begin{minipage}{\textwidth}
-\begin{align*}
-A:= X \ \mid  \ !_{n}A \multimap A
-\end{align*}\end{minipage}}\medskip\\
-Type judgements are of the form $\Gamma \vdash t:A$, where a context $\Gamma$ is a list of declarations of the form $x :_{n}A$, with $n\in \BB N$.
-We define the following operation $\Gamma+\Delta$  as follows:
-
-{
-\begin{minipage}{\textwidth}
-\begin{align*}
-() + () & =() \\
-(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) & =  (\Gamma+\Delta), x:_{m+n}A \\
-(\Gamma, x:_{n}A)+\Delta & =(\Gamma+\Delta), x:_{n}A \qquad (x\notin \Delta) \\
-\Gamma+ (\Delta, x:_{n}A) &= (\Gamma+\Delta), x:_{n} A \qquad (x\notin \Gamma)
-\end{align*}\end{minipage}}\medskip\\
-Moreover, we let $m\Gamma$ be the context made all judgements $x:_{mn}A$, where $(x:_{n}A): \Gamma$.
-
-The typing rules of $\BSTLC$ are illustrated in Fig.~\ref{fig:fuzz}.
-
-%Observe that one can always type an affine term like e.g.~$\lambda xy.x$ with a linear type $A\multimap B\multimap A$.
-Instead, a term like $\lambda xy.x(xy)$ containing two occurrences of $x$ cannot be given the linear type $(A\multimap A)\multimap (A\multimap A)$ but a type of the form
-%$!_{2}(A\multimap A)\multimap (A\multimap A)$.
-
-\begin{figure}
-\fbox{
-\begin{minipage}{0.9\textwidth}
-\begin{center}
-
-  {
-\AXC{}
-\UIC{$ x:_{1}A\vdash x: A$}
-\DP}
-
-\bigskip
-
-\begin{tabular}{c c }
-\AXC{$\Gamma \vdash M:A$}
-\UIC{$\Gamma, x:_{0}B \vdash M:A$}
-\DP
-
-&
-
-\AXC{$\Gamma, x:_{n}B, y_{m}:B\vdash M:A$}
-\UIC{$\Gamma, x:_{n+m}B\vdash M[x/y]:A$}
-\DP
-
-\\
-
-&
-
-\\
-
-  {
-\AXC{$\Gamma, x:_{n} A\vdash M: B$}
-\UIC{$\Gamma\vdash \lambda x.M: !_{n}A\multimap B$}
-\DP}
-
-&
-
-  {
-\AXC{$\Gamma \vdash M: A\multimap B$}
-\AXC{$\Delta\vdash N: A$}
-\BIC{$\Gamma +\Delta\vdash MN: B$}
-\DP}
-\\
-
-%&
-
-%\\
-%
-%  {
-%\AXC{$\Gamma \vdash t\in A$}
-%\AXC{$\Delta \vdash u\in B$}
-%\BIC{$\Gamma+\Delta \vdash ( t,u) \in A\otimes B$}
-%\DP}
-%
-%& 
-%
-%
-%  {
-%\AXC{$\Gamma\vdash t\in A \otimes B$}
-%\AXC{$\Delta,x\in_{n}A, y\in_{n} B\vdash u:C$}
-%\BIC{$n\Gamma+\Delta \vdash \mathsf{let}\  (x,y)=t \ \mathsf{in}  \ u\in C$}
-%\DP}
-%\\
-%&
-
-%\\
-%
-%
-%
-%
-%
-%& 
-%
-%  {
-%\AXC{$\Gamma\vdash M: !_{n}A$}
-%\AXC{$\Delta,x\in_{mn}A\vdash u:C$}
-%\BIC{$m\Gamma+\Delta \vdash \mathsf{let}\  !x=M \ \mathsf{in} \ u :  C$}
-%\DP}
-
-\end{tabular}
-
-\bigskip
-
-  {
-\AXC{$\Gamma \vdash M: A$}
-\UIC{$n\Gamma \vdash M : \ !_{n}A$}
-\DP}
-
-\end{center}
-\end{minipage}
-}
-\caption{Typing rules for $\BSTLC$.}
-\label{fig:fuzz}
-\end{figure}
-
-\subsubsection{$(\LREL, !_{n})$ is a model of $\BSTLC$}
+%\subsection{Proofs from Section~\ref{sec:3B}: $(\LREL, !_{n})$ is a model of $\BSTLC$}
 
@@
 \item $\model{x\in_{1}A\vdash x\in A}=\epsilon_{A}$;
-\item $\model{\Gamma+x:_{0}B\vdash M:A} =
+\item $\model{\Gamma,x:_{0}B\vdash M:A} =
 \model{\Gamma \vdash M:A}

### 2023-08-02 CSL24/appSection7.tex
commit

--- 2023-01-22 appIV.tex
@@
-%We already know from Giulio's paper that $\LREL_!$ is a post-linear continuous $\Lawv$-ccc.
+%\paragraph{Taylor Expansion }
+\subsection{Proof of  Theorem~\ref{thm:modelsTaylor}}
 
-\begin{proposition}
-$\LREL_!$ is a cartesian closed left-$\Lawv$-additive category.
-\end{proposition}
-\begin{proof} 
-Since $\infty \circ f= f\circ \infty=\infty$ and $\min\{g,h\}\circ f= \min\{g\circ f, h\circ f \}$, $\LREL_!$ is {left-$\Lawv$-additive}.
-A morphism $h\in \LREL_!(X,Y)$ that further satisfies $h\circ \min\{f,g\}=\min\{h\circ f, h\circ g\}$ for all object $X'$ and $f,g\in \LREL_!(X',X)$, is called \emph{additive}.
-To show that $\LREL_!$ is cartesian closed left-$\Lawv$-additive we must also check that (1) products and projections of additive morphisms are additive, and that (2) $\Lambda(\min\{f,g\})=\min\{\Lambda(f),\Lambda(g)\}$, $\Lambda(\infty)=\infty$, where $\Lambda:\Lawv^{\C M_{\mathsf{fin}}(Z+X)\times Y}\to \Lawv^{\C M_{\mathsf{fin}}(Z)\times( \C M_{\mathsf{fin}}(X)\times Y)}$ is the isomorphism given by
-$(\Lambda(f))_{\mu,\nu,y}= f_{\mu\oplus\nu,y}$, where 
- $\mu\oplus\nu$ is defined by $(\mu\oplus\nu)(\langle 0,x\rangle)=\mu(x)$ and $(\mu\oplus \nu)(\langle 1,x\rangle)=\nu(x)$.
-	\begin{enumerate}
-	\item  Let $f\in \Lawv^{\C M_{\mathsf{fin}}(X)\times Y}$ and $g\in \Lawv^{\C M_{\mathsf{fin}}(X)\times Z}$ be additive; then $\langle f,g\rangle \in \Lawv^{\C M_{\mathsf{fin}}(X)\times (Y+Z)}$, which is defined by
-	$$
-	\langle f,g\rangle_{\mu,\langle i,a\rangle}= \begin{cases}
-	f_{\mu,a} & \text{ if }i=0\\
-	g_{\mu,a} & \text{ if }i=1
-	\end{cases}
-	$$
-	is also additive. Indeed, for all $h\in \LREL_!(X',X)$, 
-	in any cartesian category it holds that $\langle f,g\rangle \circ h=\langle f\circ h, g\circ h\rangle$. Now, 
-		if $i=0$, then for all $h_{1},h_{2}\in \Lawv^{\C M_{\mathsf{fin}}(X')\times X}$, 
-	\begin{align*}
-	(\langle f,g\rangle\circ \min\{h_{1},h_{2}\})_{\rho,\langle i,z\rangle}&=
-	(\langle f\circ  \min\{h_{1},h_{2}\},g\circ  \min\{h_{1},h_{2}\}\rangle)_{\rho,\langle i,z\rangle} 
-	\\
-&=	(f\circ \min\{h_{1},h_{2}\})_{\rho, z}\\ &
-	= \min\{(f\circ h_{1})_{\rho,z},(f\circ h_{2})_{\rho,z}\} \\
-	&=
-	\min\{ \langle f\circ h_{1},g\circ h_{1}\rangle_{\rho,\langle i,z\rangle}, 
-	\langle f\circ h_{2},g\circ h_{2}\rangle_{\rho,\langle i,z\rangle}
-\}
-	\\
-	& = \min\{ (\langle f,g\rangle\circ h_{1})_{\rho,\langle i,z\rangle},(\langle f,g\rangle\circ h_{2})_{\rho,\langle i,z\rangle}\}	\end{align*}
-	and similarly if $i=1$.
-
-	Moreover, suppose $f\in \LREL_!(X, Y+Z)$ is additive, and let us show that $\pi_{1}( f)\in \LREL_!(X,Y)$, defined by 
-	$
-	(\pi_{1}( f))_{\mu,y}= f_{\mu, \langle 0,y\rangle}
-	$, is also additive: first observe that $\pi_{1}(f)=\pi_{1}\circ f$, where $\pi_{1}\in \LREL_!(Y+Z,Y)$ is given by $(\pi_{1})_{\mu,y}=0$ if $\mu=[y]\oplus \emptyset$ and is $\infty$ otherwise; moreover,
-	$\pi_{1}(\min\{g,h\})=\min\{\pi_{1}(g),\pi_{1}(h)\}$, since 
-	$(\pi_{1}(\min\{g,h\}))_{\mu, y}=(\min\{g,h\})_{\mu, \langle 0,y\rangle}= \min\{g_{\mu, \langle 0,y\rangle},h_{\mu, \langle 0,y\rangle}\}=( \min\{\pi_{1}(g),\pi_{1}(h)\})_{\mu, y}$. Now,
-		given $h_{1},h_{2}\in \LREL_!(X',X)$, we have that 
-	$\pi_{1}(f)\circ \min\{h_{1},h_{2}\}= (\pi_{1}\circ f)\circ \min\{h_{1},h_{2}\}= \pi_{1}\circ (f\circ \min\{h_{1},h_{2}\})= \pi_{1}\circ \min\{f\circ h_{1},f\circ h_{2}\}= \pi_{1}(\min\{f\circ h_{1},f\circ h_{2}\})= \min\{\pi_{1}(f\circ h_{1}), \pi_{1}(f\circ h_{2})\}= \min\{\pi_{1}(f)\circ h_{1},\pi_{1}(f)\circ h_{2})$.
-
-	\item  It is clear then that $\Lambda(\infty)=\infty$, and moreover
-	$$
-	\Lambda (\min\{f,g\})_{\mu,\nu,y}= \min\{ f,g\}_{\mu\oplus \nu,y}= \min\{f_{\mu\oplus\nu,y},g_{\mu\oplus\nu,y}\}= \min(\Lambda(f),\Lambda(g))_{\mu,\nu,y}
-	$$
-
-	\end{enumerate}
-
-\end{proof}
-
-For any morphism $f\in \LREL_!(X,Y)$, let us define a morphism $D(f)\in \LREL_!(X+X,Y)$, i.e.~$D(f)\in \Lawv^{\C M_{\mathsf{fin}}(X+X),Y}$, by
-$$
-D(f)_{\mu,y}=
-\begin{cases}
-f_{\mu'+x,y}
-& \text{ if }\mu=[x]\oplus \mu'
-\\
-\infty & \text{ otherwise}
-\end{cases}
-$$
-
-\begin{proposition}
-The category $\LREL_!$, endowed with the operator $D$, is a cartesian closed differential category.
-\end{proposition}
-\begin{proof}
-We must check axioms (D1)-(D7) of cartesian differential categories  plus axiom (D-curry) (cf.~\cite{Manzo2012}).
-\begin{description}
-\item[(D1)]$D(\min\{f,g\})=\min \{D(f),D(g)\}$ and $D(\infty)=\infty$: 
-while the latter is obvious, for the former we have 
-$D(\min\{f,g\})_{[x]\oplus\nu,y}= \min\{f,g\}_{\nu+x,y}= \min\{f_{\nu+x,y},g_{\nu+x,y}\}=\min\{D(f),D(g)\}_{[x]\oplus \nu,y}$, and if $\mu\neq [x]\oplus \nu$, 
-$D(\min\{f,g\})_{\mu,y}= \infty= \min\{\infty, \infty\}=\min\{ D(f),D(g)\}_{\mu,y}$. 
-\item[(D2)]
-$D(f)\circ \langle \min\{h,k\},v\rangle= \min\{ D(f)\circ \langle h,v\rangle, D(f)\circ \langle k,v\rangle\}$, and $D(f)\circ \langle \infty,v\rangle=\infty$: we can compute
-	\begin{align*}
-	(D(f)\circ \langle \min\{h,k\},v\rangle)_{\mu,y}&=
-	\inf\Big\{ 
-	\sum_{i=1}^{n}\min\{h,k\}_{\rho_{i},w_{i}}+
-	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
-	+
-	f_{[z_{1},\dots, z_{m}]+w,y}\\
-&	\qquad\qquad\mid
-	\mu=\sum_{i=1}^{n}\rho_{i}+\sum_{j=1}^{m}\nu_{j},	[w]=[w_{1},\dots, w_{n}]
-	\Big\}\\
-	& 
-=	\inf\Big\{ 
-	\min\{h,k\}_{\rho,w}+
-	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
-	+
-	f_{[z_{1},\dots, z_{m}]+w,y}\\
-&	\qquad\qquad\mid
-	\mu=\rho+\sum_{j=1}^{m}\nu_{j}
-	\Big\}\\
-	&=	\min\Big\{\inf\big\{ 
-	h_{\rho,w}+
-	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
-	+
-	f_{[z_{1},\dots, z_{m}]+w,y}	\mid
-	\mu=\rho+\sum_{j=1}^{m}\nu_{j}\big\}, \\
-	& \qquad\qquad \inf\big\{ 
-	k_{\rho,w}+
-	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
-	+
-	f_{[z_{1},\dots, z_{m}]+w,y}	\mid
-	\mu=\rho+\sum_{j=1}^{m}\nu_{j}\big\}\Big\}\\
-	& = \min\Big\{(D(f)\circ \langle h,v\rangle)_{\mu,y}, (D(f)\circ \langle k,v\rangle)_{\mu,y}\Big\}\\
-		& = \Big(\min\big\{D(f)\circ \langle h,v\rangle, D(f)\circ \langle k,v\rangle\big\}\Big )_{\mu,y}
-	\end{align*}
-	where, in the first equation, the condition $[w_{1},\dots, w_{n}]=[w]$ (i.e.~$n=1$) is forced by the fact that, otherwise, the application of $D(f)$ would give $\infty$. Moreover, we have
-\begin{align*}
-(D(f) \circ \langle \infty, v\rangle)_{\mu,y}&=\inf\Big\{ 
-	\infty+
-	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
-	+
-	f_{[z_{1},\dots, z_{m}]+w,y}\mid
-	\mu=\rho+\sum_{j=1}^{m}\nu_{j}
-	\Big\} = \infty
-\end{align*}
-
-\item[(D3)] $D(\mathrm{id})=\pi_{1}$, $D(\pi_{i})=\pi_{i}\circ \pi_{1}$: 
-recall that $\mathrm{id}_{[x],x}=0$ and $\mathrm{id}_{\mu,x}=\infty$, if $\mu\neq [x]$. 
-Moreover $(\pi_{1})_{\mu,x}=0$ if $\mu=[x]\oplus \emptyset$, and is $\infty$ otherwise, and $\pi_{2}$ is defined similarly.
-Hence 
-$D(\mathrm{id})_{[x]\oplus\nu,y}=\mathrm{id}_{\nu+x,y}$ is $0$ precisely when $x=y$ and $\nu=\emptyset$, and in all other cases is $\infty$. This shows that $D(\mathrm{id})=\pi_{1}$.
-
-$D(\pi_{1})\in \Lawv^{\C M_{\mathsf{fin}}((X+Y)+(X+Y))\times Y}$ is given by
-$D(\pi_{1})
-_{
-[x\oplus \emptyset] \oplus (\mu\oplus\nu),y
-}= (\pi_{1})_{(\mu\oplus\nu)+\langle 0,x\rangle ,y }
-$, 
-which is 0 precisely when $(\mu\oplus\nu)+\langle 0,x\rangle= y\oplus\emptyset$, i.e.~when 
-$x=y$ and $\mu=\nu=\emptyset$; in all other cases one can check that $D(\pi_{1})_{\rho,y}=\infty$, so we conclude $D(\pi_{1})=\pi_{1}\circ \pi_{1}$.
-One can argue similarly for $\pi_{2}$.
-
-\item[(D4)] $D(\langle f,g\rangle)=\langle D(f),D(g)\rangle$: 
-we have
-\begin{align*}
-D(\langle f,g\rangle)_{[x]\oplus\mu, \langle 0,y\rangle}& = 
-(\langle f,g\rangle)_{\mu+x,\langle 0,y\rangle}= f_{\mu+x,y}= D(f)_{[x]\oplus \mu,y}\\
-D(\langle f,g\rangle)_{[x]\oplus\mu, \langle 1,y\rangle}& = 
-(\langle f,g\rangle)_{\mu+x,\langle 1,y\rangle}= g_{\mu+x,y}= D(g)_{[x]\oplus \mu,y}
-\end{align*}
-from which we deduce
-$D(\langle f,g\rangle)_{[x]\oplus\mu, \langle i,y\rangle}=\langle D(f),D(g)\rangle_{[x]\oplus\mu, \langle i,y\rangle}$ 
- by the definition of $\langle \_,\_\rangle$.
- If $\rho\neq [x]\oplus\mu$, then
- $D(\langle f,g\rangle)_{\rho, \langle i,y\rangle}=\infty=\langle \infty,\infty\rangle=\langle D(f),D(g)\rangle_{\rho, \langle i,y\rangle}$
- (where the equation $\infty=\langle \infty,\infty\rangle$ is to be read as an equality between the functions $ X+Y\longrightarrow Q$
- defined by $\langle i,y\rangle \mapsto \infty$ and by
- $\begin{matrix}
- \langle 0,x\rangle\mapsto\infty\\
-  \langle 1,y\rangle\mapsto\infty
- \end{matrix}$, respectively).
-
-\item[(D5)] $D(f\circ g)=D(f)\circ \langle D(g), g\circ \pi_{2}\rangle$: we can compute
-\begin{align*}
-\Big( D(f)\circ \langle D(g), g\circ \pi_{2}\rangle\Big)_{[x]\oplus\mu,z}
-&=
-\inf \Big\{
-D(g)_{[x]\oplus\mu',w}+
-\sum_{i}g_{\mu_{i},w_{i}}+
-D(f)_{[w]\oplus [w_{1},\dots, w_{n}],z}\\
-&\qquad\qquad \mid
-w,w_{i}\in Y, 
- \mu=\mu'+ \sum_{i}\mu_{i},
-\Big\}\\
-&=
-\inf \Big\{
-g_{\mu'+x,w}+
-\sum_{i}g_{\mu_{i},w_{i}}+
-f_{ [w_{1},\dots, w_{n}]+w,z}\\
-&\qquad\qquad \mid
-w,w_{i}\in Y, 
- \mu= \mu'+\sum_{i}\mu_{i}
-\Big\}
-\\
-&=\inf\Big\{
-\sum_{i} g_{\mu_{i},w_{i}} + f_{[w_{1},\dots, w_{n}],z}
-\mid
-w_{1},\dots, w_{n}\in Y, 
-\mu+x=\sum_{i}\mu_{i}
-\Big\}
-\\
-&= (f\circ g)_{\mu+x,y} =D(f\circ g)_{[x]\oplus\mu,z}
-\end{align*}
-if $\rho\neq [x]\oplus \mu$, then $D(f\circ g)_{\rho,z}=\infty$ and 
-from the first equation above it follows that 
-also $( D(f)\circ \langle D(g), g\circ \pi_{2}\rangle)_{\rho,z}=\infty$.
-
-\item[(D6)] $D(D(f))\circ \langle \langle g,\infty\rangle,\langle h,k\rangle\rangle=D(f)\circ  \langle g,k\rangle$:
-observe that 
-\begin{align*}
-\Big(D(D(f))\Big)_{[\langle 1,x'\rangle]\oplus([x]\oplus \mu),z}&=
-\big(D(f) \big)_{[x]\oplus (\mu+x'),z }= f_{\mu+x'+x,z}\\
-\Big(D(D(f))\Big)_{[\langle 0,x\rangle]\oplus (\emptyset \oplus \mu),z}&=
-\big(D(f) \big)_{[x]\oplus \mu,z }= f_{\mu+x,z}
-\end{align*}
-and in all other cases $(D(D(f)))_{\mu,z}=\infty$.
-Using this fact we can compute:
-
-{
-\small
-\begin{align*}
-\Big(D(D(f))\circ \langle \langle g,\infty\rangle,\langle h,k\rangle\rangle\Big)_{\mu,z}&=
-\min\left\{
-\begin{matrix}
-\inf\left\{
-\begin{matrix}
-\infty_{\rho_{1},x'} + h_{\rho_{2},x}+ \sum_{i}k_{\mu_{i},w_{i}}
-+
-f_{[w_{1},\dots, w_{n}]+x'+x,z}\\
-\qquad \mid 
-x,x',w_{i}\in Y, 
-\mu=\rho_{1}+\rho_{2}+\sum_{i}\mu_{i}
-\end{matrix}
-\right\},\\
-\inf\left\{
-\begin{matrix}
-g_{\rho,x}+ \sum_{i}k_{\mu_{i},w_{i}}
-+f_{[w_{1},\dots, w_{n}]+x,z}\\
-\qquad \mid 
-x,w_{i}\in Y, 
-\mu=\rho+\sum_{i}\mu_{i}
-\end{matrix}
-\right\}
-\end{matrix}
-\right\}\\
-&=
-\inf\left\{
-g_{\rho,x}+ \sum_{i}k_{\mu_{i},w_{i}}
-+f_{[w_{1},\dots, w_{n}]+x,z}
-\mid 
-x,w_{i}\in Y, 
-\mu=\rho+\sum_{i}\mu_{i}
-\right\}\\
-&= \Big(D(f)\circ \langle g,k\rangle\Big)_{\mu,z}
-\end{align*}
-}
-
-\item[(D7)] $D(D(f))\circ \langle\langle \infty,h\rangle, \langle g,k\rangle\rangle= D(D(f))\circ \langle \langle \infty,g\rangle, \langle h,k\rangle\rangle$:
-by computations similar to the case above we obtain
-
-{
-\small
-\begin{align*}
-\Big(D(D(f))\circ \langle \langle & \infty,h\rangle,\langle g,k\rangle\rangle\Big)_{\mu,z}\\
-%&=
-%\inf\left\{
-%\begin{matrix}
-%h_{\mu',w} + g_{\mu'',x}+ \sum_{i}k_{\mu_{i},w_{i}}
-%+
-%f_{[w,w_{1},\dots, w_{n}]+x,z}\\
-%\infty+ \sum_{i}k_{\mu_{i},w_{i}}
-%+f_{[w,w_{1},\dots, w_{n}],z}
-%\end{matrix}
-%\mid 
-%w,w_{i}\in Y, 
-%\mu=\mu'+\mu''+\sum_{i}\mu_{i}
-%\right\}\\
-&=
-\inf\left\{
-h_{\rho',x'} + g_{\rho,x}+ \sum_{i}k_{\mu_{i},w_{i}}
-+
-f_{[w_{1},\dots, w_{n}]+x'+x,z}
-\mid 
-x,x',w_{i}\in Y, 
-\mu=\rho'+\rho+\sum_{i}\mu_{i}
-\right\}\\
-&=
-\inf\left\{
-g_{\rho,x} + h_{\rho',x'}+ \sum_{i}k_{\mu_{i},w_{i}}
-+
-f_{[w_{1},\dots, w_{n}]+x+x',z}
-\mid 
-x,x',w_{i}\in Y, 
-\mu=\rho+\rho'+\sum_{i}\mu_{i}
-\right\}\\
-&=
-\Big(D(D(f))\circ \langle \langle\infty,g\rangle,\langle h,k\rangle\rangle\Big)_{\mu,z}
-\end{align*}
-}
-
-\item[(D-curry)]  \ \ \ \ \  $D(\Lambda(f))=
-\Lambda(D(f)\circ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle)$:
-by observing that both morphisms are 
-in $\LREL_!(X+X, Z^{Y})= \Lawv^{\C M_{\mathsf{fin}}(X+X)\times \C M_{\mathsf{fin}}(Y)\times Z}$, 
-and that $ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle
-\in \LREL_!( (X+X)+Y  , (X+Y)+(X+Y) )$,
-we can compute:
-
-{\small
-\begin{align*}
-\big(\Lambda(D(f)\circ \langle&  \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle)\big)_{[x]\oplus\mu,\nu,z}\\
-&=
-\big(D(f)\circ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle\big)_{([x]\oplus\mu)\oplus\nu, z}\\
-&=
-\inf\left\{
-( \pi_{1})_{[x]\oplus \emptyset,x}+
-\sum_{i}(\pi_{2})_{\emptyset\oplus [w_{i}],w_{i}}
-+
-\sum_{j}(\mathrm{id})_{[z_{j}],z_{j}}
- + D(f)_{([x]\oplus \emptyset)\oplus(\mu\oplus \nu)}
-\mid
-\begin{matrix}
-\mu=[w_{1},\dots, w_{n}],\\
-\nu=[z_{1},\dots, z_{m}]
-\end{matrix}
-\right\}\\
-&=
-\inf\left\{
-0+0
-+
-0 + D(f)_{([x]\oplus \emptyset)\oplus(\mu\oplus \nu)}
-\mid
-\begin{matrix}
-\mu=[w_{1},\dots, w_{n}],\\
-\nu=[z_{1},\dots, z_{m}]
-\end{matrix}
-\right\}\\
-&=
-\big(D(f)\big)_{ ([x]\oplus\emptyset)\oplus(\mu\oplus\nu),z}\\
-&=
-f_{(\mu+x)\oplus\nu, z}\\
-& =
-\big(\Lambda(f)\big)_{\mu+x,\nu,z}=\big(D(\Lambda(f))\big)_{[x]\oplus\mu, \nu,z}
-\end{align*}
-}
-
-If $\rho\neq [x]\oplus\mu$, then $(D(\Lambda(f)))_{\rho,\nu,z}=\infty$ and $
-(\Lambda(D(f)\circ \langle  \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle))_{\rho,\nu,z}=
-(D(f)\circ \langle \pi_{1}\times \infty, \pi_{2}\times \mathrm{id}\rangle)_{\rho\oplus\nu, z}
-$, and one can check that also this is $\infty$, using the second equation above and the fact that $(\pi_{1})_{\rho\oplus\emptyset,x}=\infty$.\end{description}
-
-\end{proof}
-
-\paragraph{Taylor Expansion }
-
-To check the validity of Taylor expansion we must further check the following equation, given $f\in \LREL_!(C, B^{A})$ and $g\in \LREL_!(C,A)$:
+Theorem~\ref{thm:modelsTaylor} states the validity of Taylor expansion in $\LREL_!$.
+We must check for it the following equation, given $f\in \LREL_!(C, B^{A})$ and $g\in \LREL_!(C,A)$:
 $$
 \mathsf{ev}\circ \langle f,g\rangle= \inf_{n\in \BB N}
-(( \cdots (\Lambda^{-}(f) \underbrace{\star g)\cdots )\star g}_{n\text{ times}})\circ \langle \mathrm{id}, \infty\rangle
+\left\{(( \cdots (\Lambda^{-}(f) \underbrace{\star g)\cdots )\star g}_{n\text{ times}})\circ \langle \mathrm{id}, \infty\rangle\right\}
 $$
@@
 \end{itemize}
+\end{enumerate}
 
-\subsubsection{The language and the semantics of $\STDLC$}
+\subsection{Proof of Corollary~\ref{cor:taylor}}
 
-The language we mainly consider in the main paper il the differential $\lambda$-calculus $\STDLC$, that we quickly define here, following [Section 3, \cite{Manzo2010}].
-
-The terms and the sums\footnote{Usually what we call here ``terms'' are called ``simple terms'', and what we call here ``sums'' are called ``differential terms''.} are mutually generated by the following grammars, then quotiented by usual $\alpha$-equivalence plus other equivalences handling the $+$ constructor.
-
-{
-\begin{minipage}{\textwidth}
-\begin{align*}
-\textit{Terms: } & M,N:= x\mid \lambda x.M \mid M\BB T \mid \Der [M] \cdot N
-\\
-\textit{Sums: } & \BB T:= 0 \mid M \mid M+\BB T
-\end{align*}\end{minipage}}\medskip\\
-Simple types are generated by the grammar:
-
-{
-\begin{minipage}{\textwidth}
-\begin{align*}
-A,B:= X\mid A\to B
-\end{align*}\end{minipage}}\medskip\\
-where $X$ varies in a fixed set of ground types.
-
-The typing system is given in Figure~\ref{fig:STDLC}.
-
-\begin{figure}
-\fbox{
-\begin{minipage}{0.9\textwidth}
-\begin{center}
-
-  {
-\AXC{$(x: A)\in \Gamma$}
-\UIC{$\Gamma\vdash x\in A$}
-\DP}
-
-\bigskip
-
-\begin{tabular}{c c }
-  {
-\AXC{$\Gamma, x: A\vdash M: B$}
-\UIC{$\Gamma\vdash \lambda x.M: A\to B$}
-\DP}
-
-&
-
-  {
-\AXC{$\Gamma \vdash M: A\to B$}
-\AXC{$\Gamma\vdash \BB T: A$}
-\BIC{$\Gamma \vdash M\BB T: B$}
-\DP}
-
-\end{tabular}
-
-\bigskip
-
-  {
-\AXC{$\Gamma \vdash M: A\to B$}
-\AXC{$\Gamma \vdash N: B$}
-\BIC{$\Gamma \vdash \Der [M] \cdot N: A\to B$}
-\DP}
-
-\bigskip
-
-\begin{tabular}{c c }
-
-  {
-\AXC{}
-\UIC{$\Gamma \vdash 0:A$}
-\DP}
-
-&
-
-  {
-\AXC{$\Gamma\vdash M_1: A \quad \cdots \quad \Gamma \vdash M_n:A \qquad (n\geq2)$}
-\UIC{$\Gamma \vdash M_1+\cdots +M_n : A$}
-\DP}
-
-\end{tabular}
-\end{center}
-\end{minipage}
-}
-\caption{Typing rules for $\STDLC$.}
-\label{fig:STDLC}
-\end{figure}
-
-We finally recall the sound interpretation of $\STDLC$-terms in a $C\partial C$ category, following [Section 4.3, \cite{Manzo2010}].
-
-\begin{definition}
-Let $\C C$ be a $C\partial C$ and $\model X$ fixed objects of $\C C$, one for each ground type $X$ of the $\STDLC$.
-
-One lifts as usual the interpretation to types and contexts as $\model{A\to B}:=\model A\to \model B$ and $\model{x_1:A_1,\dots,x_n:A_n}:=\Pi_{i=1}^n\model{x_i:A_i}$.
-
-Then, one inductively defines an interpretation $\model{\Gamma \vdash M:A}\in\C C(\model \Gamma,\model A)$ of $\Gamma \vdash M:A$ and one of $\model{\Gamma \vdash \BB T:A}\in\C C(\model \Gamma,\model{A})$ of $\Gamma \vdash \BB T:A$ by induction as:
-\begin{itemize}
- \item $\model{\Gamma, x:A \vdash x:A}:=\pi_2$
- \item $\model{\Gamma, x:A \vdash y:B}:=\model{\Gamma \vdash y:B}\circ \pi_1\quad\textit{ if } y\neq x$
- \item $\model{\Gamma\vdash \lambda x.M:A\to B}:=\Lambda(\model{\Gamma,x:A\vdash M:B})$
- \item $\model{\Gamma \vdash M\BB T:B}:=\mathsf{ev} \circ \langle \, \model{\Gamma \vdash M:A\to B}, \model{\Gamma \vdash \BB T:A} \, \rangle$
- \item $\model{\Gamma \vdash \Der [M] \cdot N:A\to B}:=\Lambda\left(\Lambda^-(\model{\Gamma \vdash M:A\to B})\star \model{\Gamma \vdash N:A}\right)$
- \item $\model{\Gamma \vdash 0:A}:=:0 \qquad \textit{(the $0$ element of $\C C(\model \Gamma,\model A)$)}$
- \item $\model{\Gamma \vdash M+\BB T:A}:=\model{\Gamma \vdash M:A}+\model{\Gamma \vdash \BB T:A}$ \qquad \textit{(the $+$ operation on $\C C(\model \Gamma,\model A)$)}
-\end{itemize}
-\end{definition}
-
-One then defines a ``differential'' extension of usual $\beta$-reduction (see again [Section 3, \cite{Manzo2010}], or \cite{difflambda}), and thanks to the fact that $\C C$ is a $C\partial CC$, the interpretation above can can be proved to be sound for it ([Theorem 4.12, \cite{Manzo2010}]).
-
-In the main paper we write $\model M$, but of course we mean $\model{\Gamma \vdash M:A}$.
+\begin{proof}[Proof of Corollary \ref{cor:taylor}]
+Thm.~\ref{thmTLSlocLip} yields the estimate $\max_{\overline{B_{3\delta}(\model{N})}}\model{M}^{!}$.
+As from Thm.~\ref{thm:taylor} 4.~it follows that $\model{t}^{!}\geq \model{M}^{!}$, we deduce that $K=\max_{\overline{B_{3\delta}(\model{N})}}\model{t}^{!}\geq \max_{\overline{B_{3\delta}(\model{N})}}\model{M}^{!}$ is also a local Lipschitz constant for $\model{M}^{!}$.
+Moreover, since $\model{t}^{!}$ is concave and non-decreasing, the $\max$ of $\model{t}^{!}$ is attained at the maximum point of $\overline{B_{3\delta}(\model{N})}$, that is,
+$K= \model{t}^{!}(x+3\delta)$. Finally, from $\model{t}^{!}(\model{N})<\infty$ and the continuity of $\model{t}^{!}$ we deduce $K<\infty$.
+\end{proof}

### 2023-08-02 CSL24/main.tex
commit
Commit

@@
 \usepackage{relsize}
+
+\usetikzlibrary{intersections,through}
 
@@
 
-\section{Taylor Meets Lipschitz}\label{sec:TayLip}
+\section{Lipschitz Meets Taylor}\label{sec:TayLip}
 \input{sectionTaylip}
@@
 
-\section{Conclusion and future work}\label{section8} \input{section8}
+\section{Conclusion and Future Work}\label{section8} \input{section8}
 
@@
 
-%\appendix
+\appendix
 
-%\section{\autoref{section3}}\label{app-sec:section3}
+\section{Proofs from \autoref{section2}: Theorem \ref{theorem:fepsilon}}
+\input{appSection3}
+
+\section{Proofs from \autoref{section22}: $(\LREL, !_{n})$ is a model of $\BSTLC$}
+\input{appSection5}
+
+\section{Proofs from \autoref{sec:tls}}\label{appsec:tls}
+%\input{app_section4}
+
+\section{Proofs from \autoref{sec:TayLip}: $\Lawv$-modules and Generalized Metric Spaces}
+\input{appSection7}
+
+\section{Proofs of \autoref{sec:GMS}}
+\input{appSection8}
 

### 2023-08-02 CSL24/sectionTaylip.tex
commit

@@
 
-\begin{corollary}
+\begin{corollary}\label{cor:taylor}
 Suppose $x: A \vdash_{\STLC}M:B$ and $\vdash_{\STLC} N:A$. 
@@
 \end{corollary}
-\begin{proof}
+\begin{proof}[Proof sketch.]
 We apply the Lipschitz estimate from Thm.~\ref{thmTLSlocLip} for $\model{M}^{!}$,  and use the fact that, from Thm.~\ref{thm:taylor} 4.~it follows that $\model{t}^{!}\geq \model{M}^{!}$.

### 2023-08-02 CSL24/app_section4.tex

\begin{figure}[h]
\begin{tikzpicture}[thick]
    \coordinate (u) at (-2,0);
    \coordinate (v) at (2,0);
    \coordinate (y) at (-1.2,0);
    \coordinate (z) at (0,0);

    \draw[name path={u--v}] (u) -- (v);
    \node [draw,name path=z] at (z) [circle through={(u)}] {};

    \path[name intersections={of=u--v and z,by={int_u,int_v}}]
      foreach \X in {u,v}{(int_\X) node[below left]{$\X$}};

    \fill   (int_u) circle[radius=2pt]  node [below left] {$u$};
    \fill   (int_v) circle[radius=2pt]  node [below left] {$v$};

    \fill   (z) circle[radius=0pt]  node [xshift=-1cm, yshift=2.2cm] {$\overline{B_{2\delta}(z)}$};

    \fill   (z) circle[radius=2pt]  node [above] {$z$};
    \fill   (y) circle[radius=2pt]  node [above] {$y$};

    \coordinate (x) at (-0.7,0.4);

    \draw   (x) circle[radius=1cm] node [above] {$x$};
    \fill   (x) circle[radius=2pt]  node [xshift=10mm, yshift=10mm] {$\overline{B_{\delta}(x)}$};

    \draw   (x) circle[radius=3cm] node {};
    \fill   (x) circle[radius=0pt]  node [xshift=-1.7cm, yshift=3cm] {$\overline{B_{3\delta}(x)}$};
\end{tikzpicture}
\caption{Drawing of the proof of \autoref{th:locLip}.}
\label{fig:proof_loc_lip}
\end{figure}

\begin{theorem}\label{th:locLip}
Let $f:V\subseteq (\BB R^X,\norm\cdot) \to (\BB R,\absv \cdot)$, with $V$ open and convex and $\norm\cdot$ any norm.
If $f$ is concave and locally bounded, then $f$ is locally Lipschitz.
Moreover, the Lipschitz constant of $f$ on $\overline{B_{\delta}(x)}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} \absv f$.
\end{theorem}
\begin{proof}
 Call $\overline{B_{\delta}(x)}:=B_1$, $\overline{B_{3\delta}(x)}:=B_3$.
 It suffices to show that for all $x\in V$, there is $\delta>0$ s.t.\ $B_3\subseteq \mathrm{interior}(V)$, $K:=\max_{B_3}  \absv f$ exists and $f$ is $(\frac{1}{\delta}\max_{B_3} \absv f)$-Lipschitz on $B_1$.
 A $\delta$ satisfying the first two conditions exists since $V$ is open and bcause $f$ is locally bounded and $B_3$ is compact.
 We will show that for all such $\delta$, the third condition already holds.

 For that, fix $y,z\in B_1$ and call $r:=\frac{d(y,z)}{2\delta}\in[0,1]$.
 We want to show that $\absv{f(y)-f(z)}\leq \frac{K}{\delta}d(y,z)=2Kr$.
 Wlog $y\neq z$, otherwise there is nothing to show.

 So $r\neq 0$ and we can consider $u:=\frac{1+r}{r}z-\frac{1}{r}y$, $v:=\frac{1}{r}y-\frac{r-1}{r}z$.
 We have $u,v\in \overline{B_{2\delta}(z)}=:B_2$.
 Indeed, $d(u,z)=\norm{u-z}=\norm{\frac{z}{r}+z-\frac{y}{r}-z}=\frac{\norm{z-y}}{r}=2\delta$ and similarly $d(v,z)=2\delta$.
 Geometrically, those are actually the intersections between $B_2$ and the line generated by $y$ and $z$, see \autoref{fig:proof_loc_lip}.
 Now we have the convex combinations $z=\frac{1}{1+r}y+\frac{r}{1+r}u$ and $y=(1-r)z+rv$, so the concavity of $f$ entails on one hand:
 $f(z)\geq \frac{1}{1+r}f(y)+\frac{r}{1+r}f(u)\geq \frac{f(y)}{1+r} - \frac{rK}{1+r}$, i.e.\ $f(y)-f(z)\leq r(K+f(z))\leq 2rK$, and on the other hand:
 $f(y)\geq (1-r)f(z)+rf(v)\geq f(z)-r(f(z)+K)$, i.e.\ $f(z)-f(y)\leq r(f(z)+K)\leq 2rK$.
 In the previous inequalities we have used that $f(u),f(v)\geq -K$.
 This follows because $u,v\in B_2\subseteq B_3$, as it can be immediately checked, thus $\absv{f(u)},\absv{f(v)}\leq K$.
 Putting the final inequalities together, we have $\absv{f(y)-f(z)}\leq 2rK$, i.e.\ the thesis.
\end{proof}

Therefore, since $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$ and all tps are non-negative, we immediately have:

\begin{theorem}[\autoref{thmTLSlocLip}]
All tps $f:\Lawv^X\to\Lawv$ are locally Lipschitz on $(0,\infty)^X$.
Moreover, the Lipschitz constant of $f$ on $\overline{B_{\delta}(x)}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} f$.
\end{theorem}

### 2023-08-02 CSL24/main.tex
Commit

@@
 
-\section{Proofs from \autoref{section22}: $(\LREL, !_{n})$ is a model of $\BSTLC$}
+\section{Proofs from \autoref{section3}: $(\LREL, !_{n})$ is a model of $\BSTLC$}
 \input{appSection5}
 
-\section{Proofs from \autoref{sec:tls}}\label{appsec:tls}
-%\input{app_section4}
+\section{Proofs from \autoref{sec:tls}: the Analysis of Tropical Power Series}\label{appsec:tls}
+\input{app_section4}
 
-\section{Proofs from \autoref{sec:TayLip}: $\Lawv$-modules and Generalized Metric Spaces}
+\section{Proofs from \autoref{sec:TayLip}: Lipschitz meets Taylor}
 \input{appSection7}
 
-\section{Proofs of \autoref{sec:GMS}}
+\section{Proofs of \autoref{sec:GMS}: $\Lawv$-modules and Generalized Metric Spaces}
 \input{appSection8}

### 2023-08-02 CSL24/app_section4.tex
Commit

@@
+\subsection{Proof of \autoref{prop:nondecr+conc}}
+
+Remember that a function $f:Q^{X}\to Q^{Y}$ is \emph{concave} if for all $\alpha\in [0,1]$, $ x ,  y  \in Q^{X}$ and $b\in Y$ 
+\[
+f(\alpha\cdot  x +(1-\alpha)\cdot  y  )_{b} \geq \alpha f( x )_{b} + (1-\alpha)f( y  )_{b}
+\]
+
+We now prove \autoref{prop:nondecr+conc}, i.e.\ that $f: Q^{X}\to Q^{Y}$ are non-decreasing and concave.
+
+The fact that $f$ is non-decreasing is clear, since the multiplicities of the multisets and all coordinates of the points are non-negative.
+Let us show the concavity.
+Let us first show that all functions of the form $f( x )_{b}= \mu  x + c$ are concave:
+we have $f(\alpha x + (1-\alpha) y  )_{b}= \mu(\alpha x )+(1-\alpha) y  )+c=
+ \mu(\alpha x )+(1-\alpha) y  )+\alpha c+(1-\alpha)c=
+ \alpha(\mu  x  + c)+(1-\alpha)(\mu  y  +c)=\alpha f( x )_{b}+(1-\alpha) f( x )_{b}$.
+To conclude, let us show that if $(f_{i})_{i\in I}$ is a family of concave functions from $Q^{X}$ to $Q^{Y}$, the function $f=\inf_{i\in I}f_{i}$ is also concave: we have
+$f(\alpha x  +(1-\alpha) y  )_{b}=
+\inf_{i\in I}f_{i}(\alpha x +(1-\alpha) y  )_{b} \geq 
+\inf_{i\in I}\alpha f_{i}( x )_{b}+(1-\alpha)f_{i}( y  )_{b}
+\geq 
+\inf_{i\in I}\alpha f_{i}( x )_{b} + \inf_{j\in I}(1-\alpha)f_{j}( y  )_{b}
+=
+\alpha \cdot (\inf_{i\in I}f_{i}( x )_{b})+ (1-\alpha)\cdot( \inf_{j\in I}f_{j}( y  )_{b})=
+\alpha  f( x )_{b}+(1-\alpha)f( y  )_{b}$, where we used the fact that given families $a_{i},b_{i}$ of reals,
+$\inf_{i}a_{i}+b_{i}\geq \inf_{i}a_{i}+\inf_{j}b_{j}$.
+This follows from the fact that for all $i\in I$, $a_{i}+b_{i}\geq \inf_{i}a_{i}+\inf_{i}b_{i}$.
+
+\subsection{Proof of \autoref{thm:ScottCont}}
+
+The part of \autoref{thm:ScottCont} about tPs immedley follows from the first part of the same theorem.
+Let us quickly recall the basic definitions about cones that we need in order to prove it.
+
+\begin{definition}
+ An \emph{$\overline{\R}_{\geq 0}$-cone} is a commutative $\overline{\R}_{\geq 0}$-semimodule with cancellative addition (i.e.\ $x+y=x+y' \Rightarrow y=y'$).
+\end{definition}
+
+In \cite{Selinger2004} cones are required to also have ``strict addition'', meaning that $x+y=0 \Rightarrow x=y=0$.
+We do not add this requirement since it will automatic hold when considering normed cones.
+
+\begin{remark}
+ The addition of a cone $P$ (which forms a commutative monoid) turns $P$ into a poset by setting:
+ $
+  x \leq y \textit{ iff } y=x+z \textit{, for some }z\in P.
+ $
+ This is called the \emph{cone-order} on $P$
+ By the cancellative property, when such $z$ exists it is unique, and we denote it by $y-x$.
+\end{remark}
+
+\begin{definition}
+ A \emph{normed $\overline{\R}_{\geq 0}$-cone} $P$ is the data of a $\overline{\R}_{\geq 0}$-cone together with a $\leq$-monotone\footnote{I.e.: $x\leq y \Rightarrow \norm{x}\leq \norm y$.
+Remark that requiring this property (for all $x,y$) is equivalent to requiring that $\norm{x}\leq \norm{x+y}$ for all $x,y$.} norm on it, where a \emph{norm} on $P$ is a map $\norm{.}:P\to \overline{\R}$ satisfying the usual axioms of norms:
+ $\norm x \geq 0$, $\norm x = 0 \Rightarrow x=0$, $\norm{rx}=r\norm x$ and $\norm{x+y}\leq \norm x + \norm y$.
+\end{definition}
+
+In, e.g.~\cite{EhrPagTas2018}, a normed $\overline{\R}_{\geq 0}$-cone is simply called a cone.
+
+Remark that in a normed $\overline{\R}_{\geq 0}$-cone, by monotonicity of the norm, we have: $\norm{x+y}=0 \Rightarrow x=y=0$.
+Therefore, as already mentioned, in a normed cone we have: 
+$x+y=0 \Rightarrow \norm{x+y}=0 \Rightarrow x=y=0$, that is, addition is strict.
+
+\begin{example}
+ $\overline{\R}_{\geq 0}^X$ is a normed cone with the norm $\supnorm{x}:=\sup\limits_{a\in X} x_a\in \overline{\R}_{\geq 0}$.
+\end{example}
+
+\begin{remark}
+ The cone-order on $\overline{\R}_{\geq 0}^X$ is the pointwise usual order on $\overline{\R}_{\geq 0}$.
+% Remark also that tLs have no reason to be linear nor sublinear.
+\end{remark}
+
+A \emph{directed net} in a poset $P$ with indices in a set $I$ is a function $s:I\to P$, denoted by $(s_i)_{i\in I}$, s.t.\ its image is directed.
+We say that a directed net in $P$ \emph{admits a sup} iff its image admits a sup in $P$.
+We say that a directed net $s$ in a normed cone is \emph{bounded} iff the set $\set{\norm{s_i}\,\mid i\in I}$ is bounded in $\R_{\geq 0}$.
+
+Remember the definition of Scott-continuity:
+
+\begin{definition}
+ A function $f:P\to P'$ between posets is \emph{Scott-continuous} iff for all directed net $(s_i)_i$ in $P$ admitting a sup, we have $\exists \bigvee\limits_i f(s_i) = f(\bigvee\limits_i s_i)$ in $P'$.
+\end{definition}
+
+The fundamental result in order to prove Theorem~\ref{thm:ScottCont} is the following, taken from \cite{Selinger2004}.
+
+\begin{proposition}\label{prop:infsup}
+ Let $P$ be a normed $\overline{\R}_{\geq 0}$-cone s.t.\ every bounded directed net in $P$ admits a sup.
+ Let $(v_i)_{i\in I}$ be a directed net in $P$ with an upper bound $v\in P$.
+ Then $\exists\bigvee\limits_{i\in I} v_i \in P$ and, if $\inf\limits_{i\in I} \norm{v-v_i} =0$, one has: $\bigvee\limits_{i\in I} v_i = v$.
+\end{proposition}
+\begin{proof}
+ Remark that $v-v_i$ exists in $P$ by hypothesis and so does $\bigvee\limits_{i\in I} v_i$. %, thanks to the monotonicity of the norm.
+ Now, since $v\geq v_i$ for all $i$, we have that $v\geq \bigvee\limits_{i\in I} v_i$, and so $v-\bigvee\limits_{i\in I} v_i$ exists in $P$.
+ Fix $i\in I$.
+ Since $v_i\leq \bigvee\limits_{i\in I} v_i$, then $v-\bigvee\limits_{i\in I} v_i\leq v-v_i$ and, by monotonicity of the norm, $\norm{v-\bigvee\limits_{i\in I} v_i}\leq \norm{v-v_i}$.
+ Since this holds for all $i\in I$, we have:
+ $0\leq \norm{v-\bigvee\limits_{i\in I} v_i}\leq \inf\limits_{i\in I} \norm{v-v_i}=0$, where the last equality holds by hypothesis.
+ Thus $\norm{v-\bigvee\limits_{i\in I} v_i}=0$, i.e.\ $v=\bigvee\limits_{i\in I} v_i$.
+\end{proof}
+
+We finally obtain the desired:
+
+\begin{theorem}[Theorem~\ref{thm:ScottCont}]
+  All monotone (w.r.t.\ pointwise order) and $\norm{\cdot}_{\infty}$-continuous functions $f:(0,\infty)^X\to (0,\infty)$ are Scott-continuous.
+\end{theorem}
+\begin{proof}
+ Let $(x_i)_i$ a directed net in $(0,\infty)^X$ s.t.\ $\bigvee\limits_i x^i$ exists in $(0,\infty)^X$.
+ Then $\inf\limits_i \supnorm{\bigvee\limits_i x^i - x^i} =0$, where $\bigvee\limits_i x^i - x^i$ exists because $\bigvee\limits_i x^i \geq x^i$ for all $i$.
+ Since $f$ is $\supnorm{.}$-continuous on $(0,\infty)^X$, then $\inf\limits_i \supnorm{f(\bigvee\limits_i x^i) - f(x^i)} =0$, where $f(\bigvee\limits_i x^i) - f(x^i)$ exists because $f(\bigvee\limits_i x^i) \geq f(x^i)$ for all $i$ being $f$ monotone.
+ We can therefore apply Proposition \ref{prop:infsup} to the directed net $(f(x^i))_i$ in $(0,\infty)$, obtaining that $\bigvee\limits_i f(x^i)$ exists in $(0,\infty)$ and it coincides with $f(\bigvee\limits_i x^i)$.
+\end{proof}
+
+\subsection{Proof of \autoref{thmTLSlocLip}}
+
 \begin{figure}[h]
@@
 
+The main ingredient of the proof, that we mention in the proof sketch of \autoref{thmTLSlocLip}, is the following:
+
 \begin{theorem}\label{th:locLip}
@@
 
-Therefore, since $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$ and all tps are non-negative, we immediately have:
-
-\begin{theorem}[\autoref{thmTLSlocLip}]
-All tps $f:\Lawv^X\to\Lawv$ are locally Lipschitz on $(0,\infty)^X$.
-Moreover, the Lipschitz constant of $f$ on $\overline{B_{\delta}(x)}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} f$.
-\end{theorem}
+Therefore, since $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$ and all tps are non-negative, we immediately have \autoref{thmTLSlocLip}.

### 2023-08-02 CSL24/section4.tex
Commit

@@
 
-We conclude this subsection by noticing that $\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
-Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete \emph{normed cone} (see~\cite{Selinger2004}, or the appendix, for such notions).
+We conclude this subsection by noticing that, as a set, $\Lawv=[0,\infty]$, and  $[0,\infty]^X$ with the usual $+$ and $\cdot$ is a $\BB R_{\geq0}$-semimodule, let us call it $\overline{\BB R}_{\geq 0}^X$.
+Together with the norm $\norm{\cdot}_\infty$, it can be showed to be a Scott-complete \emph{normed cone} (see~\cite{Selinger2004} or the appendix).
 Its cone structure induces an order on it, called the \emph{cone order}:
-$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
-Such order actually coincides with the pointwise order on $\Lawv^X$.
+$x\leq y$ iff $y=x+z$ for some (unique) $z\in\overline{\BB R}_{\geq 0}^X$.
+Such order actually coincides with the pointwise order on $\overline{\BB R}_{\geq 0}$.
 This makes it a Scott-continuous dcpo.
@@
  All monotone (w.r.t.\ pointwise order) and $\norm{\cdot}_{\infty}$-continuous functions $f:(0,\infty)^X\to (0,\infty)$ are Scott-continuous.
- In particular, all tps $f:\Lawv^X\to\Lawv$ are Scott-continuous on $(0,\infty)^X$ w.r.t.\ the pointwise orders.
+ In particular, all tps $f:\overline{\BB R}_{\geq 0}^X\to\overline{\BB R}_{\geq 0}$ are Scott-continuous on $(0,\infty)^X$ w.r.t.\ the pointwise orders.
 \end{theorem}
@@
 \begin{enumerate}
-\item If a tps $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:X\times Y\to \Lawv$ (i.e.~it is linear), then $f$ is non-expansive (i.e.\ $1$-Lipschitz).  
+\item If a tps $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:X\times Y\to \Lawv$ (i.e.~it is tropical linear), then $f$ is non-expansive (i.e.\ $1$-Lipschitz).  
 \item If  $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f: \ !_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.

### 2023-08-02 CSL24/section4.tex
Commit

@@
 Moreover, tps in finitely many variables are locally equivalent to tropical polynomials (except at some singular points), and are thus also concave and locally Lipschitz-continuous.
-In this section we show that much of these properties extend also to tps with infinitely many variables, as those arising from the tropical relational model.
-%establish some topological and metric properties of tps, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
+In this section we show that much of these properties extend also to tps with infinitely many variables, as those arising from the tropical relational model, %establish some topological and metric properties of tps, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
+and by a tropical polynomial or power series we mean one with possibly infintely many variables.
 The literature on tropical power series is often recent (e.g.~\cite{Porzio2021}), and several results we prove in this section are, to our knowledge, new.
@@
 %
-In the remainder of this section, by a tropical polynomial or power series we mean one with possibly infintely many variables.
-Moreover, we consider the sets $\Lawv^X$ endowed with the usual sup-norm $\norm{x}_\infty:=\sup_{a\in X} x_a$.
+Notice that, as a set, $\Lawv^X=[0,\infty]^X$, and with the usual $+$ and $\cdot$ it is a $\BB R_{\geq0}$-semimodule, let us call it $\overline{\BB R}_{\geq 0}^X$.
+Together with the usual sup-norm $\norm{x}_\infty:=\sup_{a\in X} x_a$, it can be showed to be a Scott-complete \emph{normed cone} (see~\cite{Selinger2004} or the appendix).
+Suitable categories of cones have been recently investigated as models of probabilistic computation (\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
+The cone structure of $\overline{\BB R}_{\geq 0}^X$ also induces a partial order on it, its \emph{cone partial-order}:
+$x\leq y$ iff $y=x+z$ for some (unique) $z\in\overline{\BB R}_{\geq 0}^X$.
+It actually coincides with the pointwise order on $\overline{\BB R}_{\geq 0}$ (and makes it a Scott-continuous dcpo).
+In this section we consider tps w.r.t.\ this structure.
+%Moreover, we consider the sets $\Lawv^X$ endowed with the usual sup-norm $\norm{x}_\infty:=\sup_{a\in X} x_a$.
 
@@
  %$f:\Lawv^X\to\Lawv^Y$
-are non-decreasing and concave, w.r.t.\ the pointwise order on $\Lawv^{X}$.
+are non-decreasing and concave, w.r.t.\ the pointwise order on $\overline{\BB R}_{\geq 0}^{X}$.
 \end{proposition}
@@
 \begin{theorem}\label{thm:cont}
- All tps $f:\Lawv^X\to\Lawv$ are continuous on $(0,\infty)^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
+ All tps $f:\overline{\BB R}_{\geq 0}^X\to\overline{\BB R}_{\geq 0}$ are continuous on $(0,\infty)^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
 \end{theorem}
@@
 By \autoref{prop:nondecr+conc}, $-f$ is convex.
-Since $f\geq 0$ on all $\Lawv^X$, we have e.g.\ $-f\leq 1$ on $\Lawv^X$.
+Since $f\geq 0$ on all $\overline{\BB R}_{\geq 0}^X$, we have e.g.\ $-f\leq 1$ on $\overline{\BB R}_{\geq 0}^X$.
 %Remark that we cannot immediately apply \cite[Proposition 4.4.(3)]{Cobzas2017}, since $\Lawv^X$ is not even a $\BB R$-vector space.
@@
 
-We conclude this subsection by noticing that, as a set, $\Lawv=[0,\infty]$, and  $[0,\infty]^X$ with the usual $+$ and $\cdot$ is a $\BB R_{\geq0}$-semimodule, let us call it $\overline{\BB R}_{\geq 0}^X$.
-Together with the norm $\norm{\cdot}_\infty$, it can be showed to be a Scott-complete \emph{normed cone} (see~\cite{Selinger2004} or the appendix).
-Its cone structure induces an order on it, called the \emph{cone order}:
-$x\leq y$ iff $y=x+z$ for some (unique) $z\in\overline{\BB R}_{\geq 0}^X$.
-Such order actually coincides with the pointwise order on $\overline{\BB R}_{\geq 0}$.
-This makes it a Scott-continuous dcpo.
-Suitable categories of cones have been recently investigated as models of probabilistic computation (\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
-In analogy with \cite[Proposition 17]{DanEhrh2011}, we have:
+In analogy with \cite[Proposition 17]{DanEhrh2011}, we also have:
 
@@
 \begin{enumerate}
-\item If a tps $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:X\times Y\to \Lawv$ (i.e.~it is tropical linear), then $f$ is non-expansive (i.e.\ $1$-Lipschitz).  
-\item If  $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f: \ !_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
+\item If a tps $f: \overline{\BB R}_{\geq 0}^{X}\to \overline{\BB R}_{\geq 0}^{Y}$ arises from a matrix $\matr f:X\times Y\to \overline{\BB R}_{\geq 0}$ (i.e.~it is tropical linear), then $f$ is non-expansive (i.e.\ $1$-Lipschitz).
+\item If  $f: \overline{\BB R}_{\geq 0}^{X}\to \overline{\BB R}_{\geq 0}^{Y}$ arises from a matrix $\matr f: \ !_{n}X\times Y\to \overline{\BB R}_{\geq 0}$, then $f$ is $n$-Lipschitz-continuous.
 \end{enumerate}
@@
 
-Observe that, for any tropical polynomial $\varphi:\Lawv^{X}\to \Lawv^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \Lawv$ (as a monomial $\mu_ix+c_{i}$ yields a matrix entry on $!_{\#\mu_i}X\times Y$).
+Observe that, for any tropical polynomial $\varphi:\overline{\BB R}_{\geq 0}^{X}\to \overline{\BB R}_{\geq 0}^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \overline{\BB R}_{\geq 0}$ (as a monomial $\mu_ix+c_{i}$ yields a matrix entry on $!_{\#\mu_i}X\times Y$).
 Hence, using Proposition \ref{prop:troplinear} 2., we have:
 \begin{corollary}\label{prop:polylip}
-For any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$, $\varphi$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
+Any tropical polynomial $\varphi:\overline{\BB R}_{\geq 0}^{X}\to\overline{\BB R}_{\geq 0}$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
 \end{corollary}
@@
 %we immediately have that 
-All tps $f:\Lawv^X\to\Lawv$ are locally Lipschitz on $(0,\infty)^X$.
+All tps $f:\overline{\BB R}_{\geq 0}^X\to\overline{\BB R}_{\geq 0}$ are locally Lipschitz on $(0,\infty)^X$.
 Moreover, the Lipschitz constant of $f$ on $\overline{B_{\delta}(x)}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} f$.

### 2023-08-02 CSL24/sectionLaw.tex
commit

@@
 Moreover, the constructions for $\Mod$ generalize those of $\LREL$, in the sense that $!(\Lawv^{X})\simeq \Lawv^{\multiset(X)}$ and that $\Mod_{!}(\Lawv^{X},\Lawv^{Y})\simeq \LREL_{!}(X,Y)$.
-Finally, as the coKleisli category of a Lafont category with biproducts is always a CC$\partial$C \cite{LemayCALCO2021}, we can endow $\Mod_{!}$ with a suitable differential operator $E$ generalizing $D^{!}$ (see the Appendix) so that:
-\begin{theorem}
+Finally, as the coKleisli category of a Lafont category with biproducts is always a CC$\partial$C \cite[Theorem 21]{LemayCALCO2021}, we can endow $\Mod_{!}$ with a  differential operator $E$ generalizing $D^{!}$ (see the Appendix) so that:
+\begin{theorem}\label{thm:lemay}
 ($\Mod_{!}/\GMet_{!},E$) is a CC$\partial$C.

### 2023-08-02 CSL24/section3fancy.tex
Commit dettaglini

@@
 
-When we instantiate $Q=\Lawv$, we obtain a category $\LREL$ of $\Lawv$-valued matrices.
+When we instantiate $Q=\Lawv$, we obtain the category $\LREL$ of $\Lawv$-valued matrices.
 As one might expect, this category is tightly related to Lawvere's theory of (generalized) metric spaces.
@@
 Composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
-reads as \ $(st)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$, and the non-linear functions $t^{!}: \Lawv^{X}\to \Lawv^{Y}$ give rise to functions of shape 
+reads as \ $(st)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$, and the non-linear maps $t^{!}: \Lawv^{X}\to \Lawv^{Y}$ have shape 
 $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum_{a\in X} \mu(a)x_a$.

### 2023-08-02 CSL24/section4.tex
Commit dettaglini

@@
 1). Using the fact that $f( x)_{b}= \inf_{a\in X}\{\matr f_{a,b}+ x_{a}\}$,
-the problem reduces to checking that $|(\matr f_{a,b}- x_{a})- (\matr f_{a,b}- y_{a})| = |x_{a}- y_{a}|\leq \|  x- y\|_{\infty}$ (see the Appendix).
+the problem reduces to: $|(\matr f_{a,b}- x_{a})- (\matr f_{a,b}- y_{a})| = |x_{a}- y_{a}|\leq \norm{x- y}_{\infty}$. %(see the Appendix).
 

### 2023-08-02 CSL24/main.tex
commit

@@
 \begin{abstract}
-We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, and we show that it provides a unified framework for both program  metrics, based on the analysis of program sensitivity via Lipschitz conditions, and resource analysis, based on higher-order program differentiation.
+We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, and we show that it provides a unifying framework for two well-developed quantitative approaches to program semantics: on the one hand program metrics, based on the analysis of program sensitivity via Lipschitz conditions, on the other hand resource analysis, based on linear logic and higher-order program differentiation.
 To do that we focus on the semantics arising from the relational model weighted over the tropical semiring, and we discuss its application to the study of “best case” program behavior for languages with probabilistic and non-deterministic effects.

### 2023-08-02 CSL24/section1.tex
commit

@@
-In recent years, more and more interest in the programming language community has been directed towards the study of \emph{quantitative} properties of programs like computing the number of computation steps or convergence probabilities,
+In recent years, more and more interest in the programming language community has been directed towards the study of \emph{quantitative} properties of programs like e.g.~the number of computation steps or the probability of convergence,
 as opposed to purely \emph{qualitative} properties like termination or program equivalence. 
@@
 
-On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}.
+On the one hand one one could mention the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}.
 This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
@@
 At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
-in metric models $n$ times duplication results in a {linear} (hence \emph{Lipschitz}) function $n x$, while in differential models this results in a {polynomial} function $x^{n}$, thus not Lipschitz.
-The fundamental intuition of this work is then the observation that
-this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n x$.
+in metric models a program duplicating an input $n$ times yields a {linear} (hence \emph{Lipschitz}) function $n x$, whereas in differential models it would lead to a {polynomial} function $x^{n}$, thus not Lipschitz.
+The fundamental idea behind this work is the observation that
+this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we will see, the monomial $x^{n}$ precisely reads as the linear function $n x$.
 
@@
 Computationally speaking, working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, the fundamental (and generally intractable) problem of finding the roots of a polynomial admits a \emph{linear time} algorithm in the tropical case (and, moreover,  the tropical roots can be used to approximate the actual roots \cite{Noferini2015}).
-The computational nature of tropical notions explains why these are so widely applied in computer science, notably for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).
+The combinatorial nature of several methods in tropical mathematics explains why these are so widely applied in computer science, notably for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).
 
@@
 Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
-However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.
+However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics and of the potentialities of its applications has not yet been undertaken.
 
-In this paper we make a first step in such direction, by demonstrating that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provide the desired bridge between differential and metric semantics, and we suggest how this approach might be used to study probabilistic and non-deterministic programs.% (an aspect that will be developed in more detail in a future paper of this series).
+In this paper we make a first step in such direction, by demonstrating that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provide the desired bridge between differential and metric semantics, and suggests new combinatorial methods to study probabilistic and non-deterministic programs.% (an aspect that will be developed in more detail in a future paper of this series).
 
@@
 
-\item We study the relational model over the tropical semiring, which provides a semantics of effectful extensions of $\STLC$ (simply typed $\lambda$-calculus) and $\mathrm{PCF}$. Notably, 
+\item We study the relational model over the tropical semiring, which provides a semantics of effectful extensions of the simply typed $\lambda$-calculus ($\STLC$ in the following) and $\mathrm{PCF}$ \cite{Plotkin1977}.
+Notably,
  we show that higher-order programs are interpreted by a  
@@
 \item We conclude 
-by putting the connection between the 
-tropical, differential and metric viewpoints at the right level of abstraction and generality.
+by framing the connection between the 
+tropical, differential and metric viewpoints at a more abstract level.
 We recall a well-known correspondence between Lawvere's \emph{generalized metric spaces} \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007} and we show that it yields a model of the differential $\lambda$-calculus which extends the tropical relational model.

### 2023-08-02 CSL24/section4.tex
commit

@@
 %Before looking at what happens in the case of non-linear programs, let us make the metric structure of $\LREL$ explicit. 
-The following proposition provides a useful characterization of the functional metrics in $\LREL$, relying on 
-the bijection between $\HOM{\LREL}{X}{Y}$ and set of tropical linear functions from $\Lawv^{X}$ to $\Lawv^{Y}$.
-Remember that $d_\infty(f,g):=\sup_{a\in X} d_\infty(f(x),f(y))$.
+Observe that on the hom-sets $\HOM{\LREL_{!}}{X}{Y}$ there are two natural notions of distance: the metric $\| f-g\|_{\infty}$ arising from the norm and the one arising from the usual $\sup$-metric $d_\infty(f,g):=\sup_{a\in X} d_\infty(f^{!}(x),f^{!}(y))$.
+In general one has $\| f-g\|_{\infty}\geq d_\infty(f,g)$, the equality holding when $f,g$ are linear.
 
-\begin{proposition}
-For all $f,g\in \LREL_{!}(X,Y)$, $d_{\infty}( f,g)=d_\infty(f^{!},g^{!})$.% $\norm{ \matr f-\matr g}_{\infty} =  \sup_{x\in \Lawv^{X}} \norm{ f( x)-g(x)}_{\infty}$.
-\end{proposition}
+%The following proposition provides a useful characterization of the functional metrics in $\LREL$, relying on 
+%the bijection between $\HOM{\LREL}{X}{Y}$ and set of tropical linear functions from $\Lawv^{X}$ to $\Lawv^{Y}$.
+%Remember that $d_\infty(f,g):=\sup_{a\in X} d_\infty(f(x),f(y))$.
+%
+%\begin{proposition}
+%For all $f,g\in \LREL_{!}(X,Y)$, $d_{\infty}( f,g)=d_\infty(f^{!},g^{!})$.% $\norm{ \matr f-\matr g}_{\infty} =  \sup_{x\in \Lawv^{X}} \norm{ f( x)-g(x)}_{\infty}$.
+%\end{proposition}
 
@@
 
-Observe that, for any tropical polynomial $\varphi:\overline{\BB R}_{\geq 0}^{X}\to \overline{\BB R}_{\geq 0}^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \overline{\BB R}_{\geq 0}$ (as a monomial $\mu_ix+c_{i}$ yields a matrix entry on $!_{\#\mu_i}X\times Y$).
+For any tropical polynomial $\varphi:\overline{\BB R}_{\geq 0}^{X}\to \overline{\BB R}_{\geq 0}^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \overline{\BB R}_{\geq 0}$ (as a monomial $\mu_ix+c_{i}$ yields a matrix entry on $!_{\#\mu_i}X\times Y$).
 Hence, using Proposition \ref{prop:troplinear} 2., we have:

### 2023-08-02 CSL24/section5bis.tex
commit

@@
 
-\subparagraph*{Metric approach: bounded $\lambda$-calculus% as Lipschitz functions
+\subparagraph*{Metric Approach: Bounded $\lambda$-Calculus% as Lipschitz functions
 }
@@
 
-\subparagraph*{Resource approach: the differential $\lambda$-calculus% as polynomials
+\subparagraph*{Resource Approach: the Differential $\lambda$-Calculus% as polynomials
 }
@@
 
-\subparagraph*{Tropical Mathematics: a possible synthesis}%Outline of the paper}
+\subparagraph*{Tropical Mathematics: Lipschitz Meets Taylor}%Outline of the paper}
 

### 2023-08-02 CSL24/sectionTaylip.tex
commit

@@
 \end{corollary}
-\begin{proof}[Proof sketch.]
-We apply the Lipschitz estimate from Thm.~\ref{thmTLSlocLip} for $\model{M}^{!}$,  and use the fact that, from Thm.~\ref{thm:taylor} 4.~it follows that $\model{t}^{!}\geq \model{M}^{!}$.
-To conclude, since $\model{t}^{!}$ is concave and non-decreasing, $\max_{\overline{B_{3\delta}(x)}}\model{t}^{!}= \model{t}^{!}(x+3\delta)$, which yields the stated Lipschitz estimate. 
+\begin{proof}
+Thm.~\ref{thmTLSlocLip} yields the estimate $\max_{\overline{B_{3\delta}(\model{N})}}\model{M}^{!}$.
+As from Thm.~\ref{thm:taylor} 4.~it follows that $\model{t}^{!}\geq \model{M}^{!}$, we deduce that $K=\max_{\overline{B_{3\delta}(\model{N})}}\model{t}^{!}\geq \max_{\overline{B_{3\delta}(\model{N})}}\model{M}^{!}$ is also a local Lipschitz constant for $\model{M}^{!}$.
+Moreover, since $\model{t}^{!}$ is concave and non-decreasing, the $\max$ of $\model{t}^{!}$ is attained at the maximum point of $\overline{B_{3\delta}(\model{N})}$, that is,
+$K= \model{t}^{!}(x+3\delta)$. Finally, from $\model{t}^{!}(\model{N})<\infty$ and the continuity of $\model{t}^{!}$ we deduce $K<\infty$.
 \end{proof}
+%\begin{proof}[Proof sketch.]
+%We apply the Lipschitz estimate from Thm.~\ref{thmTLSlocLip} for $\model{M}^{!}$,  and use the fact that, from Thm.~\ref{thm:taylor} 4.~it follows that $\model{t}^{!}\geq \model{M}^{!}$.
+%To conclude, since $\model{t}^{!}$ is concave and non-decreasing, $\max_{\overline{B_{3\delta}(x)}}\model{t}^{!}= \model{t}^{!}(x+3\delta)$, which yields the stated Lipschitz estimate. 
+%\end{proof}
 %{\color{red}ADD PROOF SKETCH}

### 2023-08-02 CSL24/section4.tex
commit

@@
 %Before looking at what happens in the case of non-linear programs, let us make the metric structure of $\LREL$ explicit. 
-Observe that on the hom-sets $\HOM{\LREL_{!}}{X}{Y}$ there are two natural notions of distance: the metric $\| f-g\|_{\infty}$ arising from the norm and the one arising from the usual $\sup$-metric $d_\infty(f,g):=\sup_{a\in X} d_\infty(f^{!}(x),f^{!}(y))$.
-In general one has $\| f-g\|_{\infty}\geq d_\infty(f,g)$, the equality holding when $f,g$ are linear.
+Observe that on the hom-sets $\HOM{\LREL_{!}}{X}{Y}$ there are two natural notions of distance: the metric $\| f-g\|_{\infty}$ arising from the norm and the one arising from the usual $\sup$-metric $d_\infty(f,g):=\sup_{a\in X} d_\infty(f^{!}(a),g^{!}(a))$.
+In general one has $\| f-g\|_{\infty}\geq d_\infty(f,g)$, the equality holding when $f^{!},g^{!}$ are linear (i.e.~when they arise from morphisms of $ \LREL(X,Y)$).
 

### 2023-11-26 CSL24/main.tex
commit

@@
 
-\title{Tropical mathematics and the Lambda-calculus I}
+\title{Tropical Mathematics and the Lambda-calculus I}
 \subtitle{Metric and Differential Analysis of Effectful Programs
@@
 
-\titlerunning{Tropical mathematics for the Lambda-calculus 1} %TODO optional, please use if title is longer than one line
+\titlerunning{Tropical mathematics for the Lambda-calculus I} %TODO optional, please use if title is longer than one line
 
@@
 
-\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
+\relatedversion{{\color{red}ADD URL HERE.}} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
 %\relatedversiondetails[linktext={opt.
@@
 
-\funding{This work has been supported by the ERC CoG 818616 DIAPASoN.}%optional, to capture a funding statement, which applies to all authors.
+\funding{This work has been supported by the European Research Council through the project ERC CoG 818616 DIAPASoN.}%optional, to capture a funding statement, which applies to all authors.
 Please enter author specific funding statements as fifth argument of the \author macro.

### 2023-11-26 CSL24/section2bis.tex
commit

@@
 \end{theorem}
-\begin{proof}[Proof sketch]
-Let $\C F_\epsilon$ be the set of $n\in\N^{k}$ such that 
-$\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
-The core of the proof is showing that this set is indeed finite and sufficient for computing $f$.
+%\begin{proof}[Proof sketch]
+%Let $\C F_\epsilon$ be the set of $n\in\N^{k}$ such that 
+%$\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
+%The core of the proof is showing that this set is indeed finite and sufficient for computing $f$.
+%\end{proof}
+
+In order to prove Theorem \ref{theorem:fepsilon}, let us set the following:
+\begin{definition}
+ Let $\preceq$ be the product order on $\N^k$ (i.e.\ for all $ m  , n  \in \N^{K}$, $ m  \preceq  n  $ iff $m_{i}\leq n_{i}$ for all $1\leq i\leq K$).
+ Of course $ m  \prec  n  $ holds exactly when $ m  \preceq  n  $ and $m_{i}<n_{i}$ for at least one $1\leq i\leq K$.
+ Finally, we set $ m  \prec_{1} n  $  iff
+$ m  \prec  n  $ and $\sum_{i=1}^{K}n_{i}-m_{i}=1$ (i.e.\ they differ on exactly one coordinate).
+\end{definition}
+
+\begin{remark}\label{rmk:AC}
+If $U\subseteq \N^{K}$ is infinite, then $U$ contains an infinite ascending chain $ m  _{0}\prec  m  _{1} \prec  m  _{2} \prec \dots$.
+This is a consequence of K\"onig Lemma (KL): consider the directed acyclic graph $(U,\prec_{1})$, indeed a $K$-branching tree; if there is no infinite ascending chain $  m  _{0}\prec  m  _{1} \prec  m  _{2} \prec \dots$, then in particular there is no infinite ascending chain $  m  _{0}\prec_{1}  m  _{1} \prec_{1}  m  _{2} \prec_{1} \dots$ so the tree $U$ has no infinite ascending chain; then by KL it is finite, contradicting the assumption.
+\end{remark}
+
+%Theorem \ref{theorem:fepsilon} now follows from the following result:
+%\begin{proposition}%[Theorem \ref{theorem:fepsilon}]
+%Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tps with matrix $\matr f:\N^k\to\Lawv$.
+% For all $0<\epsilon<\infty$, there is $\C F_\epsilon \subseteq_{\mathrm{fin}} \N^k$ such that:
+%\begin{enumerate}
+%% \item $\C F_\epsilon$ is finite
+% \item if $\mathcal{F}_\epsilon= \emptyset$ then $f( x ) = +\infty$ for all $ x \in \Lawv^k$;
+% \item if $f( x _0) = +\infty$ for some $ x _0\in [\epsilon,\infty)^{K}$ then $\mathcal{F}_\epsilon= \emptyset$;
+% \item the restriction of $f$ on $[\epsilon,\infty]^k$ coincides  with  $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{n x+\matr f(n)}$.
+%%where $nx:=\sum_{i=1}^k n_ix_i$.
+%\end{enumerate}
+%\end{proposition}
+\begin{proof}[Proof of Theorem \ref{theorem:fepsilon}]
+We will actually show the existence of $\C F_\epsilon \subseteq_{\mathrm{fin}} \N^k$ such that:
+\begin{enumerate}
+% \item $\C F_\epsilon$ is finite
+ \item if $\mathcal{F}_\epsilon= \emptyset$ then $f( x ) = +\infty$ for all $ x \in \Lawv^k$;
+ \item if $f( x _0) = +\infty$ for some $ x _0\in [\epsilon,\infty)^{K}$ then $\mathcal{F}_\epsilon= \emptyset$;
+ \item the restriction of $f$ on $[\epsilon,\infty]^k$ coincides  with  $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{n x+\matr f(n)}$.
+%where $nx:=\sum_{i=1}^k n_ix_i$.
+\end{enumerate}
+Let $\mathcal F_\epsilon$ be the complementary in $\N$ of the set:
+\[
+ \set{ n  \in\N^{K} \mid \textit{either } \hat f ( n  )=+\infty \textit{ or there is }  m  \prec  n  \textit{ s.t.\ } \hat f( m  )\leq\hat f( n  )+\epsilon}.
+\]
+In other words, $ n  \in\mathcal F_\epsilon$ iff $\hat f( n  )<+\infty$ and for all $ m  \prec  n  $, one has $\hat f( m  )>\hat f( n  )+\epsilon$.
+
+Suppose that $\mathcal F_\epsilon$ is infinite; then, using Remark~\ref{rmk:AC}, it contains an infinite ascending chain $\set{ m  _0\prec  m  _1\prec\cdots}$.  
+By definition of $\mathcal F_\epsilon$ we have then 
+$+\infty>\hat f( m  _0)>\hat f( m  _1)+\epsilon>\hat f( m  _2)+2\epsilon>\cdots$, 
+so that $+\infty>\hat f( m  _0)>\hat f( m  _{i})+i\epsilon\geq i\epsilon$ for all $i\in\N$.
+This contradicts the Archimedean property of $\R$.
+
+1).
+We show that if $\mathcal F_\epsilon=\emptyset$, then $\hat f( n  )=+\infty$ for all $ n  \in\N^{K}$.
+This immediately entails the desired result.
+We go by induction on the well-founded order $\prec$ over $ n  \in\N^{K}$:
+
+- if $ n  =0^{K}\notin\mathcal F_\epsilon$, then $\hat f( n  )=+\infty$, because there is no $ m  \prec n  $.
+
+- if $ n  \notin\mathcal F_\epsilon$, with $ n  \neq 0^{K}$ then either $\hat f( n  )=+\infty$ and we are done, or there is $ m  \prec  n  $ s.t.\ $\hat f( m  )\leq \hat f( n  )+\epsilon$.
+By induction $\hat f( m  )=+\infty$ and, since $\epsilon<+\infty$, this entails $\hat f( n  )=+\infty$.
+
+2).
+If $f( x _0)=+\infty$ with $ x _0\in [\epsilon,\infty)^{K}$, then necessarily $\hat f( n  )=+\infty$ for all $ n  \in\N^{K}$.
+Therefore, no $ n  \in\N^{K}$ belongs to $\mathcal F_\epsilon$.
+
+3).
+We have to show that $f( x )=P_\epsilon( x )$ for all $ x \in [\epsilon,+\infty]^{K}$.
+By 1), it suffices to show that we can compute $f( x )$ by taking the $\inf$, that is therefore a $\min$, only in $\mathcal F_\epsilon$ (instead of all $\N^{K}$).
+If $\mathcal F_\epsilon=\emptyset$ then by 2) we are done (remember that $\min\emptyset := +\infty$).
+If $\mathcal F_\epsilon\neq\emptyset$, we show that for all $ n  \in\N^{K}$, if $ n   \notin\mathcal F_\epsilon$, then there is $ m  \in\mathcal F_\epsilon$ s.t.\ $\hat f( m  )+ m   x  \leq \hat f( n  )+ n   x $.
+We do it again by induction on $\prec_{1}$:
+
+- if $ n  =0^{K}$, then from $\mathbf  n\notin \mathcal F_{\epsilon}$, by definition of $\mathcal F_\epsilon$, we have $\hat f( n  )=+\infty$ (because there is no $ n  '\prec n  $).
+So any element of $\mathcal F_\epsilon\neq\emptyset$ works.
+
+- if $ n  \neq 0^{K}$, then we have two cases:
+either $\hat f( n  )=+\infty$, in which case we are done as before by taking any element of $\mathcal F_\epsilon\neq\emptyset$.
+Or $\hat f( n  )<+\infty$, in which case (again by definition of $\mathcal F_\epsilon$) there is $ n  '\prec n  $ such that $ \hat f( n  ')\leq \hat f( n  )+\epsilon$  ($\star$).
+Therefore we have (remark that the following inequalities hold also for the case $x=+\infty$):
+\[\begin{array}{rclr}
+ \hat f( n  ')+ n  ' x  & \leq & \hat f( n  ) + \epsilon +  n' x  & \textit{by }(\star) \\
+ & \leq & \hat f( n  ) + ( n  - n  ') x  +  n  ' x  & \textit{because $\epsilon\leq\min x $ and $\min  x \leq( n  -  n') x $} \\
+ & = & \hat f( n  )+  n   x . &
+\end{array}\]
+Now, if $ n  '\in\mathcal F_\epsilon$ we are done.
+Otherwise $ n  '\notin\mathcal F_\epsilon$ and we can apply the induction hypothesis on it, obtaining an $ m  \in\mathcal F_\epsilon$ s.t.\ $\hat f( m  )+ m   x  \leq \hat f( n  ')+ n  ' x $.
+Therefore this $ m  $ works.
 \end{proof}

### 2023-11-26 CSL24/sectionLaw.tex
commit

@@
+% !TEX root = main.tex
+
 As we have seen, the morphisms of $\LREL$ can be seen as continuous functions between the $\Lawv$-modules $\Lawv^{X}$, when the latter are taken with the metric induced by the $\infty$-norm.
@@
 
-An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}, that it's certainly worth recalling here, albeit very briefly (more details in the Appendix).
+\subsection{$\Lawv$-Modules and Complete $\Lawv$-Categories}
+
+An answer to the questions above comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}.
 
@@
 $0  \geq X(x,x)$ and $X(y,z)+X(x,y)\geq  X(x,z)$. 
-This structure clearly generalizes the usual definition of metric spaces, which indeed are the same as skeletal (i.e.~$X(x,y)=0$ implies $x=y$) and symmetric (i.e.~$X(x,y)=X(y,x)$) $\Lawv$-categories.
+This structure clearly generalizes the usual definition of metric spaces, which are indeed precisely the  
+$\Lawv$-categories which are \emph{skeletal} (i.e.~$X(x,y)=0$ implies $x=y$) and \emph{symmetric} (i.e.~$X(x,y)=X^{\mathrm{op}}(x,y)$, where $X^{\mathrm{op}}(x,y)=X(y,x)$).
+A basic example of $\Lawv$-category is $\Lawv$ itself, with the distance $\Lawv(x,y)=y \menus x$.
+
 Moreover, a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is nothing but a non-expansive map $f:X\to Y$, since functoriality reads as $Y(f(x),f(y))\leq X(x,y)$.
+Functors of shape $\Phi: X\times Y^{\op}\to \Lawv$ are called \emph{distributors} and usually noted $\Phi: Y \pfun X$.
+Notice that distributors $\Phi: Y\pfun X$ and $\Psi: Z\pfun Y$ can be composed just like
+ordinary matrices in $\LREL$: $\Psi\diamond \Phi : Z\pfun X$ is given by
+$(\Psi\diamond \Phi)_{z,x}=\inf_{y\in Y}\Psi(z,y)+\Phi(y,x)$.
 
 Lawvere also observed that the usual notion of Cauchy-completeness can be formulated, in this framework, as the existence of suitable colimits \cite{Lawvere1973}.
-A $\Lawv$-category is said \emph{complete} when \emph{all} {weighted colimits} exist in $X$.
-Let $\GMet$ indicate the category of complete $\Lawv$-categories and continuous (i.e.~colimit-preserving) $\Lawv$-enriched functors.
+Let us recall the notion of weighted colimit in this context:
 
-It turns out that the categories $\Mod$ and $\GMet$ are isomorphic \cite{Stubbe2006}. First,  any $\Lawv$-module $(M,\preceq, \star)$ is a $\Lawv$-category via
-$M(x,y) = \inf\left\{ \epsilon \mid \epsilon \star x\geq y\right\}$.
-Moreover, a homomorphism of $\Lawv$-modules is a functor of the associated $\Lawv$-categories.
-Conversely, complete $\Lawv$-categories are \emph{tensored} (cf.~\cite{Stubbe2014}), and thus admit a suitable $\Lawv$-module structure $\epsilon \otimes x$.
-Moreover, a continuous functor between complete $\Lawv$-categories is the same as a homomorphism of the associated $\Lawv$-modules. 
+\begin{definition}[weighted colimits]
+Let $X,Y,Z$ be $\Lawv$-categories,
+$\Phi: Z\pfun Y$ be a distributor and  $f:Y\to X$ be a functor.
+A functor $g:Z\to X$ is the \emph{$\Phi$-weighted colimit of $f$ over $X$}, noted $\colim(\Phi,f)$, if for all $z\in Z$ and $x\in X$
+\begin{align}
+X(g(z),x)= \sup_{y\in Y}\left\{X(f(y),x)\menus \Phi(y,z)\right\}
+\end{align}
+
+\end{definition}
+
+A functor $f:X\to Y$ is said \emph{cocontinuous} if it commutes with all existing weighted colimits in $X$, i.e.~$f(\colim(\Phi,g))=\colim(\Phi,f\circ g)$. A $\Lawv$-enriched category.
+A $\Lawv$-category $X$ is said \emph{cocomplete} if all weighted colimits over $X$ exist. 
+We let $\GMet$ indicate the category of cocomplete $\Lawv$-categories and cocontinuous $\Lawv$-enriched functors as morphisms.
+
+Observe that the usual Cauchy completeness for a $\Lawv$-category $X$ follows from its cocompleteness.
+Indeed, a Cauchy sequence $(x_{n})_{n\in \mathbb N}$ in $X$ yields two adjoint distributors $\phi^{*}:1\pfun X$ and $\phi_{*}:X\pfun 1$, where
+$\phi^{*}(x')=\lim_{n\to \infty}X(x_{n},x')$ and
+$\phi_{*}(x')=\lim_{n\to \infty}X(x',x_{n})$. Hence,  
+$\colim(\phi^{*},1_{X}):1\to X$ must be a point $x$ satisfying $0=X(x,x)=\sup_{y\in X}\lim_{n\to \infty}(X(y,x)\menus X(x_{n},y))$, which implies $\lim_{n\to \infty}X(x_{n},x)=0$.
+
+% to the existence 
+%of weighted colimits of the form $\colim(\phi^{*}, 1_{X}):1\to X$, where $\phi^{*}: 1\pfun X$ is a \emph{left-adjoint} distributor, and $1_{X}:X\to X$ is the identity functor. 
+%That $\phi^{*}$ is left-adjoint means that there exists a distributor $\phi_{*}:X\pfun 1$ such that 
+%$0=  \phi^{*}\diamond \phi_{*}$ and $\phi_{*}\diamond \phi^{*} \geq X$, i.e.~$0=\inf_{x\in X}\phi^{*}(x)+\phi_{*}(x)$ and 
+%$\phi_{*}(x)+\phi^{*}(x')\geq X(x,x')$, for all $x\in X$.
+
+It turns out that the notions of $\Lawv$-module and cocomplete $\Lawv$-category are indeed equivalent. More precisely, the categories $\Mod$ and $\GMet$ are isomorphic \cite{Stubbe2006}. 
+First, any $\Lawv$-module $(M,\preceq, \star)$ can be endowed with the structure of a $\Lawv$-category by letting
+$M(x,y) = \inf\left\{ \epsilon \mid \epsilon \star x\geq y\right\}$. Moreover, a homomorphism of $\Lawv$-modules induces a cocontinuous functor of the associated $\Lawv$-categories. 
+Conversely, in cocomplete $\Lawv$-categories it is possible to introduce a continuous $\Lawv$-action
+via suitable weighted colimits called \emph{tensors} (cf.~\cite{Stubbe2014}):
+
+\begin{definition}[tensors]
+Let $X$ be a $\Lawv$-category, $x\in X$ and $\epsilon \in \Lawv$. The \emph{tensor of $x$ and $\epsilon$}, if it exists, is the colimit $\epsilon \otimes x:= \colim( [\epsilon],\Delta x)$, where
+$[\epsilon]: \{\star\}\pfun \{\star\}$ is the constantly $\epsilon$ distributor
+and $\Delta x:\{\star\}\to X$ is the constant functor. 
+\end{definition}
+
+A cocomplete $\Lawv$-category can thus be endowed with a $\Lawv$-module structure with order given by  $x\preceq_{X}y $ iff $X(y,x)=0$, and 
+action given by tensors $\epsilon \otimes x$. 
+Moreover, a cocontinuous functor between cocomplete $\Lawv$-categories is the same as a homomorphism of the associated $\Lawv$-modules. 
 %
@@
 
-To relate this elegant correspondence with our work on $\LREL_{!}$, we show that it lifts to a model of the differential $\lambda$-calculus, extending the co-Kleisli category $\LREL_{!}$. 
-First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies2018, Manzo2013}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting suitable properties of $\Mod$, we may define $!$ as the product of the symmetric algebras.
-This ensures that the coKleisli category $\Mod_{!}$ is a CCC.
+\subsection{Exponential and Differential Structure of $\Mod\simeq\GMet$}
+
+We now show that the correspondence between $\Lawv$-modules and cocomplete $\Lawv$-categories lifts to a model of the differential $\lambda$-calculus, generalizing the co-Kleisli category $\LREL_{!}$.
+
+In order to define a Lafont exponential $!$ over $\Mod$, we exploit a well-known recipe from \cite{Mellies2018, Manzo2013}.
+The first step is to define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$.
+Notice that each element of $!_{n}M$ can be described as a join of ``multisets''
+$[x_{1},\dots, x_{n}]$, where the latter is the equivalence class of the tensors 
+$x_{1}\otimes \dots \otimes x_{n}\in M^{\otimes_{n}}$ under the action of permutations $\sigma\in \F S_{n}$.
+The $\Lawv$-module $!_{n}M$ is a cocomplete $\Lawv$-category with distance function defined on basic ``multisets'' as follows:
+\begin{align}
+(!_{n}M)(\alpha,\beta)=
+\sup_{\sigma\in \F S_{n}}\inf_{\tau\in \F S_{n}}\sum_{i=1}^{n}
+X(x_{\sigma(i)},y_{\tau(i)})
+\end{align}
+where $\alpha=[x_{1},\dots, x_{n}]$ and $\beta= [y_{1},\dots, y_{n}]$, and extended to arbitrary elements $\alpha=\bigvee_{i}\alpha_{i}$ and $\beta=\bigvee_{j}\beta_{j}$
+by $(!_{n}M)(\alpha,\beta)=\sup_{i}\inf_{j}(!_{n}M)(\alpha_{i},\beta_{j})$.
+
+Next, we define $!M$ as the infinite biproduct $\prod_{n}!_{n}M$, yielding the cofree commutative comonoid over $M$ (cf.~\cite[Proposition 1]{Mellies2018}).
+Using the fact that biproducts commute with tensors in $\LREL$, by standard results \cite{Mellies2018}, we obtain that the coKleisli category $\Mod_{!}$ is a CCC.
 Moreover, the constructions for $\Mod$ generalize those of $\LREL$, in the sense that $!(\Lawv^{X})\simeq \Lawv^{\multiset(X)}$ and that $\Mod_{!}(\Lawv^{X},\Lawv^{Y})\simeq \LREL_{!}(X,Y)$.
-Finally, as the coKleisli category of a Lafont category with biproducts is always a CC$\partial$C \cite[Theorem 21]{LemayCALCO2021}, we can endow $\Mod_{!}$ with a  differential operator $E$ generalizing $D^{!}$ (see the Appendix) so that:
+
+Finally, since the coKleisli category of a Lafont category with biproducts is always a CC$\partial$C \cite[Theorem 21]{LemayCALCO2021}, we can endow $\Mod_{!}$ with a  differential operator $E$,  generalizing $D^{!}$, given by
+\begin{align}\label{eq:dermod}
+Ef(\alpha)=
+\bigvee\left\{
+f(\beta\cup [x]) \ \Big \vert  \ 
+\iota_{n}(\beta)\otimes \iota_{1}(x) \leq S(\alpha)
+\right\}
+\end{align}
+where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
+and $S: !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{Mellies2018}, and conclude that:
 \begin{theorem}\label{thm:lemay}

### 2023-11-27 CSL24/ArXiv/appSection3.tex
commit

--- 2023-08-02 CSL24/appSection3.tex
@@
 \begin{definition}
- Let $\preceq$ be the pointwise order on $\N^k$ (i.e.\ for all $ m  , n  \in \N^{K}$, $ m  \preceq  n  $ iff $m_{i}\leq n_{i}$ for all $1\leq i\leq K$).
+ Let $\preceq$ be the product order on $\N^k$ (i.e.\ for all $ m  , n  \in \N^{K}$, $ m  \preceq  n  $ iff $m_{i}\leq n_{i}$ for all $1\leq i\leq K$).
  Of course $ m  \prec  n  $ holds exactly when $ m  \preceq  n  $ and $m_{i}<n_{i}$ for at least one $1\leq i\leq K$.

### 2023-11-27 CSL24/ArXiv/appSection7.tex
commit

--- 2023-08-02 CSL24/appSection7.tex
@@
 \end{enumerate}
-
-\subsection{Proof of Corollary~\ref{cor:taylor}}
-
-\begin{proof}[Proof of Corollary \ref{cor:taylor}]
-Thm.~\ref{thmTLSlocLip} yields the estimate $\max_{\overline{B_{3\delta}(\model{N})}}\model{M}^{!}$.
-As from Thm.~\ref{thm:taylor} 4.~it follows that $\model{t}^{!}\geq \model{M}^{!}$, we deduce that $K=\max_{\overline{B_{3\delta}(\model{N})}}\model{t}^{!}\geq \max_{\overline{B_{3\delta}(\model{N})}}\model{M}^{!}$ is also a local Lipschitz constant for $\model{M}^{!}$.
-Moreover, since $\model{t}^{!}$ is concave and non-decreasing, the $\max$ of $\model{t}^{!}$ is attained at the maximum point of $\overline{B_{3\delta}(\model{N})}$, that is,
-$K= \model{t}^{!}(x+3\delta)$. Finally, from $\model{t}^{!}(\model{N})<\infty$ and the continuity of $\model{t}^{!}$ we deduce $K<\infty$.
-\end{proof}

### 2023-11-27 CSL24/ArXiv/main.tex
commit

--- 2023-08-02 CSL24/main.tex
@@
-\documentclass[a4paper,english, UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
-%This is a template for producing LIPIcs articles. 
-%See lipics-v2021-authors-guidelines.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
-%for producing a PDF according the PDF/A standard, add "pdfa"
+\documentclass[english]{article}
 
-%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
+\usepackage{eurosym}
+\usepackage{xspace}
+\usepackage{ifthen}
+\setlength\voffset{-1in}
+\setlength\topmargin{1cm}
+\setlength\headheight{1cm}
+\setlength\headsep{1cm}
+\setlength\textheight{22.7cm}
+\setlength\footskip{0.5cm}
+\setlength\hoffset{-1in}
+\setlength\oddsidemargin{3.3cm}
+\setlength\textwidth{15cm}
+\usepackage{titlesec}
+\titlespacing*{\section}{0ex}{1ex}{1.5ex}
+\titlespacing*{\paragraph}{0ex}{1ex}{1.3ex}
+\titlespacing*{\optionalsubsection}{0ex}{1ex}{1ex}
+\usepackage{enumitem}
+\setitemize{itemsep=0pt,topsep=0pt,parsep=0pt,partopsep=0pt}
+\setenumerate{itemsep=0pt,topsep=0pt,parsep=0pt,partopsep=0pt}
+\setdescription{itemsep=0pt,topsep=0pt,parsep=0pt,partopsep=0pt}
 
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
-
+\usepackage[english]{babel}  
 \usepackage{cite}
@@
 
-\title{Tropical mathematics and the Lambda-calculus I}
-\subtitle{Metric and Differential Analysis of Effectful Programs
+\addto\extrasenglish{
+  \def\sectionautorefname{Section}
+}
+
+\title{Tropical Mathematics and the Lambda-calculus I:\\
+{\Large Metric and Differential Analysis of Effectful Programs}
 } %TODO Please add
 
-\titlerunning{Tropical mathematics for the Lambda-calculus 1} %TODO optional, please use if title is longer than one line
-
-\author{Davide {Barbarossa}}{Dipartimento di Informatica - Scienza e Ingegneria, Universit\`a di Bologna, %[optional: Address], 
-Italy% \and My second affiliation, Country 
-\and \url{https://lipn.univ-paris13.fr/\~barbarossa/index.html} }{davide.barbarossa@unibo.it}{https://orcid.org/0000-0003-4608-8282}{%(Optional)Funding each author
-}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
-Please provide at least the name of the affiliation and the country.
-The full address is optional.
-Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.
-
-\author{Paolo {Pistone}}{Dipartimento di Informatica - Scienza e Ingegneria, Universit\`a di Bologna, %[optional: Address], 
-Italy% \and LIP, Universit\'e de Lyon, France 
-\and \url{http://logica.uniroma3.it/pistone/
-} }{paolo.pistone2@unibo.it}{https://orcid.org/0000-0003-4250-9051}{%(Optional)Funding each author
+\author{
+Davide {Barbarossa}\\
+{\small Dipartimento di Informatica - Scienza e Ingegneria}, \\
+{\small Universit\`a di Bologna, Italy}\\
+{\small \texttt{davide.barbarossa@unibo.it}}\\
+\ \\
+Paolo {Pistone}\\
+{\small Dipartimento di Informatica - Scienza e Ingegneria}, \\
+{\small Universit\`a di Bologna, %[optional: Address], 
+Italy}\\
+{\small \texttt{paolo.pistone2@unibo.it}}
 }
-
-\authorrunning{D.\,Barbarossa and P.\,Pistone} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{Davide Barbarossa and Paolo Pistone} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
-
-\ccsdesc[500]{Theory of computation~Lambda calculus}
-\ccsdesc[300]{Theory of computation~Categorical semantics}
-\ccsdesc[100]{Theory of computation~Linear logic}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
-
-\keywords{Relational semantics, Differential lambda-calculus, Tropical semiring,  Program metrics, Lawvere quantale} %TODO mandatory; please add comma-separated list of keywords
-
-\category{} %optional, e.g. invited paper
-
-\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
-%\relatedversiondetails[linktext={opt.
-text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g.
-Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional
-
-%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-%\supplementdetails[linktext={opt.
-text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g.
-Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
-
-\funding{This work has been supported by the ERC CoG 818616 DIAPASoN.}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-\acknowledgements{%I want to thank \dots
-}%optional
-
-%\nolinenumbers %uncomment to disable line numbering
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\date{}
 
@@
 
-%TODO mandatory: add short abstract of the document
 \begin{abstract}
@@
 Finally, we show that a general foundation for this approach is provided by an abstract correspondence between tropical algebra and Lawvere’s theory of generalized metric spaces.
-%A typical result of such line of research is that, thanks to the tropical operations, a simply typed term is interpreted as a locally Lipschitz map, which is in turn decomposed as an inf of more and more sensitive Lipschitz ones, corresponding to the interpretations of the approximants given by the lambda-calculus Taylor expansion.
-%We finally describe some promising lines of research, showing in particular how such tropical model expresses natural optimisation problems of probabilistic calculi.
 \end{abstract}
@@
 
-%%
-%%\section{Topical linear/non-linear algebra}\label{section3}
-%%\input{section3C}
-%%
-%%\section{Tropical weighted relational semantics for high-order languages}
-%%
-%%The category $\LREL$ satisfies many nice properties which make it able to model all the variants of the $\lambda$-calculi to which we are interested in here.
-Although the majority of such properties are shared with the general class of models of $\lambda$-calculi called \emph{weighted relational models}, thoroughly studied in \cite{Manzo2013}, we mention them for our particular case in order to highlight how these look like in the tropical setting, but also to make the non-expert in such topics reader able to follow.
-%%The real interest in our particular model will come in the next \autoref{sec:tls}.
-%%
-%%
-%%\input{section3A}
-%%
-%%\input{section2}
-%
-%\paragraph*{Allowed duplication/erasure: the probabilistic $\lambda$-calculus $\STLC_\oplus$}\label{sec:STLCO}
-%
-%\input{section3B}
-
 \section{On Tropical Power Series}\label{sec:tls}
@@
 
-%\section{Future lines of research}\label{sec:future}
-
-%\subsection{Quantitaive analysis}\label{sec:app}
-%%\input{section5C}
-
-%\subsection{Generalized Metric Spaces and Modules over the Lawvere Quantale}\label{section6}
-%%At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}.
-%Let us recall that a quantale is a complete lattice endowed with a continuous monoid action.
-In the case of $\Lawv$ the lattice is defined by the reverse order $\geq$ on $\BB R$, and the monoid action is provided by addition.
-Notice that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$.
-%A consequence of these observations is that, as we discussed below, the tropical approach to linear algebra coincides with the study of ``$\Lawv$-valued matrices'', i.e.~of maps of the form $s: X\times Y\to \Lawv$ .
-%In particular, a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we explicitly take in Section \ref{section6}).
-%\input{section6}%
-
 \section{Related Work}\label{section7} \input{section7}
@@
 
-%%
-%% Bibliography
-%%
-
-%% Please use bibtex,
-
-\bibliography{tropical}
+\bibliography{main}
 

### 2023-11-27 CSL24/ArXiv/section2.tex
commit

--- 2023-07-28 CSL24/section2.tex
@@
 
-As such, $(\LREL_!,D)$ is a model of the differential $\lambda$-calculus $\STDLC$ \cite{ER}, a language ensuring exact control of duplications via a notion of linear substitution.
-Its syntax (e.g.\ \cite[Section 3]{Manzo2010}) is given by the \emph{terms} $M$ and the \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
+As such, $(\LREL_!,D)$ is a model of the differential $\lambda$-calculus $\STDLC$ \cite{ER}, an extension of $\STLC$ ensuring exact control of duplications via a notion of linear substitution in its operational semantics.
+We will not deal with its operational properties, so let us only give its syntax (see \cite[Section 3]{Manzo2010}): there are \emph{terms} $M$ and \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
 quotiented by $\alpha$-equivalence, by equations that make $+,0$ form a commutative monoid on the set of sums, %, i.e.\ commutativity and associativity of $+$ and neutrality of $0$ w.r.t.\ $+$;
@@
 %The models of $\STDLC$ are the cartesian \emph{closed} differential categories ($CC\partial C$), which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.
+
+By induction on $M$ %(using that composition and $D$ easily preserve ``booleaness'' and projections and evaluation of  $\LREL$ are boolean)
+one can prove that:
+
+\begin{proposition}\label{prop:descrete}
+ For $\BSTLC$ and $\STDLC$ (and thus also for its fragment $\STLC$), we have that $\model{\Gamma\vdash M:A}\in\Lawv^{!\model{\Gamma}\times\model{A}}$ is a \emph{boolean} matrix, i.e.\ actually $\model{\Gamma\vdash M:A}\in\{0,\infty\}^{!\model{\Gamma}\times\model{A}}$.
+\end{proposition}
+
+For example, the $\varphi$ of \autoref{fig:plot1} is not the interpretation of a $\lam$-term, because its matrix is not discrete.
+This can be seen as the fact that such calculi are relatively trivial, from this point of view.
+Even if the interpretation of those languages is trivia, we shall still focus on them in order to set the basis of the further studies and in order to see some already interesting properties of their interpretation.{\color{red}SPIEGARE MEGLIO}

### 2023-11-27 CSL24/ArXiv/section2bisbis.tex
commit

--- 2023-07-31 CSL24/section2bisbis.tex
@@
 Before discussing how full-scale higher-order programming languages can be interpreted in terms of tropical power series, we highlight how such functions may naturally arise in the study of effectful programming languages.
-We will see that, when considering probabilistic and non-deterministic programs, tropical tools can be used to describe \emph{optimal} properties of programs, that is, the behavior of programs in \emph{the best/worst case}.
+We will see that, when considering probabilistic and non-deterministic programs, tropical tools can be used to describe the behavior of programs in \emph{the best/worst case}, and may lead to collapse the description of infinitely many possible behaviors into a combinatorial account of the optimal ones.
 %Moreover, since, as we'll see, tropical semantics is also a \emph{metric} semantics, it can be used to study how much program behavior is \emph{sensitive} to errors, that is, how a small error in input may be increased in output.
@@
 The polynomial function $Q_{\true}(p,q):=P_{ll}(p,q)+P_{rll}(p,q)+P_{rrr}(p,q)=p^2+p^2q+q^3$ gives instead the probability of the event ``$M\twoheadrightarrow \true$'', and analogously for $Q_{\false}(p,q):=P_{lr}(p,q)+P_{rrl}(p,q)+P_{rlr}(p,q)=pq+2pq^2$.
+
 This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}.
@@
 
-Answering 1) and 2) amounts then at solving a maximization problem related to $P_{\omega_{0}}(p,q)$ or $Q_{\true}(p,q)$. These are more easily solved by passing to the associated tropical polynomials.
+Answering 1) and 2) amounts then at solving a maximization problem related to $P_{\omega_{0}}(p,q)$ or $Q_{\true}(p,q)$.
+In fact, these problems are more easily solved by passing to the associated \emph{tropical} polynomials.
 For 1), the maximum values $x,y$ of $P_{rll}(p,q)$ can be computed by finding the \emph{minimum} values of $\mathsf tP_{rll}(-\log p, -\log q)= -2\log p- \log q$.
@@
 
-Similar arguments could be done by replacing $\true$ and $\false$ by, respectively, a converging and diverging term (e.g.~in a $\mathrm{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.
+As we'll see in Section \ref{section3}, this analysis extends to PCF-style programs.
+For example, the program $M=\mathbf Y(\lambda x.\true \oplus_{p} x)$ yields the power series $Q_{\true}(p,q)=\sum_{n=0}^{\infty}pq^{n}=\frac{p}{1-q}$ that sums all \emph{infinitely many} ways in which $M$ may reduce to $\true$.
+Notice that the tropicalised series $\mathsf tQ_{\true}(-\log p,-\log q)=\inf_{n\in \mathbb N}\{-\log p -n\log q\}=-\log p$ collapses onto a single monomial describing the \emph{unique} most likely reduction path of $M$, namely the one that passes through a coin toss only once.
+%
+%Similar arguments could be done by replacing $\true$ and $\false$ by, respectively, a converging and diverging term (e.g.~in a $\mathrm{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.
 
@@
 This example is inspired from \cite{Manzo2013}.
-We consider now a basic non-deterministic language with terms $M::= \true \mid  \mathsf{gen} \mid  M + M$, with an operation semantics comprising a non-deterministic reduction rule
+We consider now a basic non-deterministic language with terms $M::= \true \mid  \mathtt{Gen} \mid  M + M$, with an operation semantics comprising a non-deterministic reduction rule
 $M_{1}+M_{2} \stackrel{\alpha}{\to} M_{i}$ and a generation rule
-$\mathsf{gen}\stackrel{\beta}{\to} \true +\mathsf{gen}$, 
+$\mathtt{Gen}\stackrel{\beta}{\to} \true +\mathtt{Gen}$, 
 where in each case the value $\alpha,\beta\in \Lawv$ indicates a \emph{cost} associated with the reduction (e.g.~the estimated clock value for the simulation of each reduction on a given machine model).
 Then, any reduction $\omega: M \twoheadrightarrow N$ of a term to (one of its) normal form is  associated with a tropical monomial $P_{\omega}( \alpha,\beta)$ consisting of the sum of the costs of all reductions in $\omega$.
-For a given normal form $N$, the reductions $\omega_{i}: M \twoheadrightarrow N$ give rise to a tps $\inf_{i\in I}P_{\omega_{i}}(\vec \alpha,\beta)$.
+For a given normal form $N$, the reductions $\omega_{i}: M \twoheadrightarrow N$ give rise to a tps $\inf_{i\in I}P_{\omega_{i}}( \alpha,\beta)$.
 For example, consider the non-deterministic term
 $
-M :=\mathsf{gen}+  ((\true + \true) + \mathsf{gen})
-$
+M :=\mathtt{Gen}+  ((\true + \true) + \mathtt{Gen})
+$. 
 The (infinitely many) reduction paths leading to $\true$ can be grouped as follows:
 \begin{itemize}
-\item left, then reduce gen $n+1$-times, then left;
+\item left, then reduce $\mathtt{Gen}$ $n+1$-times, then left;
 \item right, then left and then either left or right;
-\item right twice, then reduce gen $n+1$-times and then left.
+\item right twice, then reduce $\mathtt{Gen}$ $n+1$-times and then left.
 \end{itemize}
@@
 Notice that, since $\alpha$ and $\beta$ are always positive, the power series $\varphi_{M\twoheadrightarrow\true}(\alpha)$ is indeed equivalent to the tropical polynomial $\min\{ 2\alpha+\beta, 3\alpha\}$.
-In other words, of the infinitely many behaviors of $M$, only finitely many have chances to be \emph{optimal}: either left + gen + left, or right + left + (left,right).
-Hence, reducing to best-case analysis may lead, in several cases, to collapse the infinitary description of \emph{all} behaviors to a purely combinatorial description of the finitely many optimal ones.
+In other words, of the infinitely many behaviors of $M$, only finitely many have chances to be \emph{optimal}: either left + $\mathtt{Gen}$ + left, or right + left + (left,right).
+Also in this case, reducing to best-case analysis leads to collapse the infinitary description of \emph{all} behaviors to a purely combinatorial description of the finitely many optimal ones.
 

### 2023-11-27 CSL24/ArXiv/section3fancy.tex
commit

--- 2023-08-02 CSL24/section3fancy.tex
@@
 We now show how to interpret in $\LREL_{!}$ a graded version of $\STLC$, that we call $\BSTLC$, indeed a simplified version of the well-studied language $\mathrm{Fuzz}$ \cite{Reed2010}.
-The main feature of this language is that if $\vdash \lambda x.M:\,!_nA\multimap B$, then $x$ is duplicated \emph{at most} $n$ times in any reduction of $\vdash (\lambda x.M)N :B$ to the normal form.
+This language is based on a graded exponential $!_{n}A$, corresponding to the possibility of using an element of type $A$ \emph{at most} $n$ times.
+In particular, if a function $ \lambda x.M$ of type $!_nA\multimap B$, then, for any $N$ of type $A$, $x$ is duplicated \emph{at most} $n$ times in any reduction of $(\lambda x.M)N $ to the normal form.
 
-The types of $\BSTLC$ are $A::= o \ \mid  \ !_{n}A \multimap A$, the contexts of the typing judgements are sets of declarations of the form $x :_{n}A$, with $n\in \mathbb N$, and the typing rules are: %given in Fig.~\ref{fig:rules}
-	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
-	\begin{array}{cccc}
-		\prooftree
+Graded simple types are defined by $A::= o \ \mid  \ !_{n}A \multimap A$; the contexts of the typing judgements are sets of declarations of the form $x :_{n}A$, with $n\in \mathbb N$;
+given two contexts $\Gamma,\Delta$, we define $\Gamma+\Delta$ recursively as follows: if $\Gamma$ and $\Delta$ have no variable in common, then $\Gamma+\Delta=\Gamma\cup \Delta$; otherwise, we let $(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) =  (\Gamma+\Delta), x:_{m+n}A$.
+Moreover, for any context $\Gamma$ and $m\in \mathbb N$, we let $m\Gamma$ be made all $x:_{mn}A$ for $(x:_{n}A) \in \Gamma$.  
+The  typing rules of $\BSTLC$ are illustrated in Fig.~\ref{fig:rules},
+
+%	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
+\begin{figure}
+\fbox{
+	\begin{minipage}{0.9\textwidth}
+	\begin{center}
+	$\prooftree
+	\justifies
+	x:_{1}A\vdash x: A
+	\endprooftree$
+
+	\bigskip
+
+	\begin{minipage}{0.42\textwidth}
+	\begin{center}
+	$\prooftree
 		\Gamma \vdash M:A
@@
 		\endprooftree 
-		&
-		\prooftree
+		$
+
+	\bigskip
+
+	$\prooftree
+		\Gamma, x:_{n} A\vdash M: B
+		\justifies
+		\Gamma\vdash \lambda x.M: !_{n}A\multimap B
+		\endprooftree$
+
+		\medskip
+
+	\end{center}
+	\end{minipage} \ \ \ \ 
+	\begin{minipage}{0.42\textwidth}
+	\begin{center}
+
+		$\prooftree
 		\Gamma, x:_{n}B, y:_{m} B\vdash M:A
@@
 		\Gamma, x:_{n+m}B\vdash M\{x/y\}:A
-		\endprooftree 
-		&
-		\prooftree
-		\Gamma, x:_{n} A\vdash M: B
-		\justifies
-		\Gamma\vdash \lambda x.M: !_{n}A\multimap B
-		\endprooftree
-		&
-		\prooftree
+		\endprooftree $
+
+		\bigskip
+
+		$\prooftree
 		\Gamma \vdash M: !_nA\multimap B
@@
 		\endprooftree
-	\end{array}
-	\]
-where $\Gamma+\Delta$ is defined letting $(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) =  (\Gamma+\Delta), x:_{m+n}A$, and $m\Gamma$ is made all $x:_{mn}A$ for $(x:_{n}A) \in \Gamma$.  
-The axiom is $x:_{1}A\vdash x: A$.
+		 $
 
-Now, one can see that the comonad $!$ of $\LREL$ can be ``decomposed'' into a family of ``graded exponentials functors'' $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined on objects by taking multisets of cardinality \emph{at most} $n$.
+		 \medskip
+
+	\end{center}
+	\end{minipage}
+	\end{center}
+	\end{minipage}
+	}
+	\caption{Typing rules for $\BSTLC$.}
+	\label{fig:rules}
+	\end{figure}
+
+Now, one can see that the comonad $!$ of $\LREL$ can be ``decomposed'' into a family of ``graded exponentials functors'' $!_n:\LREL\to\LREL$ ($n\in\BB N$), where $!_{n}X$ is the set of multisets on $X$ of cardinality \emph{at most} $n$.
 %  lift to functors
-The sequence $(!_n)_{n\in\N}$ gives rise to a so-called \emph{$\N$-graded linear exponential comonad} on (the SMC) $\LREL$.
+The sequence $(!_n)_{n\in\N}$ gives rise to a so-called \emph{$\N$-graded linear exponential comonad} on (the SMC) $\LREL$ \cite{Katsumata2018}.
 %satisfying the adjunction: $\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
@@
 
-We recall the interpretation in $\LREL_{!}$ of $\STDLC$,  
+We recall the interpretation in $\LREL_{!}$ of the simply typed \emph{differential} $\lambda$-calculus $\STDLC$,  
  an extension of $\STLC$ ensuring exact control of duplications.
@@
 %Sums are, then, just \emph{finite} sets of terms.
-The typing rules are as follows: %in Figure~\ref{fig:rules}
-	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
-	\begin{array}{cccc}
-		\prooftree
+The typing rules are illustrated in Figure~\ref{fig:rules2}, where a context $\Gamma$ is a list of typed variable declarations.
+%The axioms are $\Gamma, x:A \vdash x: A$ and $\Gamma\vdash 0:A$.
+The main feature of this language is that $\Der^n[\lambda x.M,N^n]0$ has a non-zero normal form iff $x$ is duplicated exactly $n$ times during reduction.
+
+\begin{figure}
+\fbox{
+	\begin{minipage}{0.9\textwidth}
+	\begin{center}
+	\begin{minipage}{0.42\textwidth}
+	\begin{center}
+	$\prooftree
+	\justifies
+	\Gamma\vdash 0:A
+	\endprooftree$
+
+	\bigskip
+
+	$\prooftree
 		\Gamma, x: A\vdash M: B
@@
 		\Gamma\vdash \lambda x.M: A\to B
-		\endprooftree 
-		&
-		\prooftree
+		\endprooftree $
+
+		\bigskip
+
+		$ \prooftree
+		\Gamma \vdash M: A\to B
+		\quad
+		\Gamma \vdash N: A
+		\justifies
+		\Gamma \vdash \Diff{M}{N}: A\to B
+		\endprooftree$
+
+		\medskip
+
+	\end{center}
+	\end{minipage} \ \ \ \ 
+	\begin{minipage}{0.42\textwidth}
+	\begin{center}
+	$\prooftree 
+	\justifies
+	\Gamma, x:A \vdash x: A
+	\endprooftree $
+
+	\bigskip
+
+	$\prooftree
 		\Gamma \vdash M: A\to B
@@
 		\Gamma \vdash M\mathbb T: B
-		\endprooftree 
-		&
-		\prooftree
-		\Gamma \vdash M: A\to B
-		\quad
-		\Gamma \vdash N: A
-		\justifies
-		\Gamma \vdash \Diff{M}{N}: A\to B
-		\endprooftree
-		&
-		\prooftree
+		\endprooftree $
+
+		\bigskip
+
+		$\prooftree
 		\Gamma\vdash M_1: A
@@
 		\endprooftree
-	\end{array}
-	\]
-where a context $\Gamma$ is a list of typed variable declarations.
-The axioms are $\Gamma, x:A \vdash x: A$ and $\Gamma\vdash 0:A$.
-The main feature of this language is that $\Der^n[\lambda x.M,N^n]0$ has a non-zero normal form iff $x$ is duplicated exactly $n$ times during reduction.
+		$
+
+		\medskip
+
+	\end{center}
+	\end{minipage}
+	\end{center}
+	\end{minipage}
+	}
+	\caption{Typing rules of $\STDLC$.}
+	\label{fig:rules2}
+	\end{figure}
 
@@
 These are CCCs enriched over commutative monoids (i.e.\ morphisms are summable and there is a $0$ morphism), with the cartesian closed structure compatible with the additive structure,
-and equipped with a certain \emph{differential operator} $D$, generalising the usual notion of differential, see e.g.\ \cite{BluteEhrhTass10}.
+and equipped with a certain \emph{differential operator} $D$, turning a morphism $f:A\to B$ into a morphism $Df: A\times A\to B$, and 
+generalising the usual notion of differential, see e.g.\ \cite{BluteEhrhTass10}.
 An example is the CC$\partial\lambda$C of convenient vector spaces with smooth maps, where $D$ is the ``real'' differential of smooth maps.

### 2023-11-27 CSL24/ArXiv/section4.tex
commit

--- 2023-08-02 CSL24/section4.tex
@@
 %Before looking at what happens in the case of non-linear programs, let us make the metric structure of $\LREL$ explicit. 
-Observe that on the hom-sets $\HOM{\LREL_{!}}{X}{Y}$ there are two natural notions of distance: the metric $\| f-g\|_{\infty}$ arising from the norm and the one arising from the usual $\sup$-metric $d_\infty(f,g):=\sup_{a\in X} d_\infty(f^{!}(a),g^{!}(a))$.
+Observe that on the hom-sets $\HOM{\LREL_{!}}{X}{Y}$ there are two natural notions of distance: the metric $\| f-g\|_{\infty}$ arising from the norm and the one arising from the usual $\sup$-metric $d_\infty(f,g):=\sup_{x\in \Lawv^{X}} \|f^{!}(x)-g^{!}(x)\|_{\infty}$.
 In general one has $\| f-g\|_{\infty}\geq d_\infty(f,g)$, the equality holding when $f^{!},g^{!}$ are linear (i.e.~when they arise from morphisms of $ \LREL(X,Y)$).

### 2023-11-27 CSL24/ArXiv/section7.tex
commit

--- 2023-07-31 CSL24/section7.tex
@@
 
+As we said, the relational semantics over the tropical semiring was quickly explored in \cite{Manzo2013}, to provide a ``best case'' resource analysis of a $\mathrm{PCF}$-like language with non-deterministic choice.
 The connections between differential $\lambda$-calculus (and differential linear logic), relational semantics, and non-idempotent intersection types are very well-studied (see \cite{decarvalho2018}, and more recently, \cite{Mazza2016} for a more abstract perspective, and \cite{Olimpieri2021, Galal2021} for a 2-categorical, or proof-relevant, extension).
-As we said, the relational semantics over the tropical semiring was quickly explored in \cite{Manzo2013}, to provide a ``best case'' resource analysis of a $\mathrm{PCF}$-like language with non-deterministic choice.
 \emph{Probabilistic coherent spaces} \cite{Ehrhard2011}, a variant of  the relational semantics, provide an interpretation of higher-order probabilistic programs
@@
 Related approaches, although not based on metrics, are provided by \emph{differential logical relations} \cite{dallago} and \emph{change action} models \cite{Picallo2019}.
-
 More generally, the literature on program metrics in denotational semantics is vast.
@@
 Motivated by connections with computer science and fuzzy set-theory, 
-the abstract study of generalized metric spaces in the framework of \emph{quantale}- or even \emph{quantaloid}-enriched categories has led to a vast literature in recent years \cite{Hofmann2014, Stubbe2014},
-and its connections with tropical mathematics have been explored e.g.~in \cite{Fuji, Willerton2013}.
+the abstract study of generalized metric spaces in the framework of \emph{quantale}- or even \emph{quantaloid}-enriched categories has led to a significant literature in recent years (e.g.~\cite{Hofmann2014, Stubbe2014}),
+and connections with tropical mathematics also have been explored e.g.~in \cite{Fuji, Willerton2013}.
 Moreover, applications of quantale-modules to both logic and computer science have also been studied \cite{Abramsky1993b, Russo2007}.
 
-Connections between program metrics and the differential $\lambda$-calculus have been already suggested in \cite{PistoneLICS}; moreover, \emph{cartesian difference categories} \cite{Picallo2020} have been proposed as a way to relate derivatives in differential categories with those found in change action models.
+Finally, connections between program metrics and the differential $\lambda$-calculus have been already suggested in \cite{PistoneLICS}; moreover, \emph{cartesian difference categories} \cite{Picallo2020} have been proposed as a way to relate derivatives in differential categories with those found in change action models.
 

### 2023-11-27 CSL24/ArXiv/section8.tex
commit

--- 2023-07-29 CSL24/section8.tex
@@
 The main goals of this paper are two. Firstly,  to
-demonstrate the existence of a conceptual bridge between two different well-studied quantitative approaches to higher-order programs, and to highlight the possibility of transferring results and techniques from one approach to the other.
+demonstrate the existence of a conceptual bridge between two well-studied quantitative approaches to higher-order programs, and to highlight the possibility of transferring results and techniques from one approach to the other.
 Secondly, to suggest that tropical mathematics, a
-field which has been largely and successfully applied in computer science, could be used to study quantitative properties of higher-order programs.
+field which has been largely and successfully applied in computer science, could be used for the quantitative analysis of functional programming languages. 
+While the first goal was here developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions, and we leave their development to a second paper of this series.
 
-While the first goal was here developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions (best-case analysis, log-probabilities), and we leave their development to a second paper of this series.
-To conclude, while the main ideas of this article only use basic concepts from the toolbox of tropical mathematics, it looks like an exciting direction to look for applications of more advanced tools from tropical algebraic and differential geometry (e.g.~Newton polytopes, tropical varieties, tropical differential equations).
+While the main ideas of this article only use basic concepts from the toolbox of tropical mathematics, an exciting direction is that of looking at potential applications of more advanced tools from tropical algebraic and differential geometry (e.g.~Newton polytopes, tropical varieties, tropical differential equations).
+Another interesting question is how much of our results on tps and their tropical Taylor expansion can be extended to the abstract setting of generalized metric spaces and continuous functors.
 %We believe that exploring these ideas in more depth could be a fruitful direction; moreover, 

### 2023-11-27 CSL24bis/main.tex
commit

--- 2023-08-02 CSL24/main.tex
@@
 
-\usetikzlibrary{intersections,through}
-
 \input{prooftree}
@@
 
-\title{Tropical mathematics and the Lambda-calculus I}
-\subtitle{Metric and Differential Analysis of Effectful Programs
+\title{Tropical mathematics for the Lambda-calculus 1}
+\subtitle{The potential of tropical wheighted relational semantics%for probabilities and Taylor expansion
 } %TODO Please add
@@
 
-\author{Davide {Barbarossa}}{Dipartimento di Informatica - Scienza e Ingegneria, Universit\`a di Bologna, %[optional: Address], 
+\author{Davide {Barbarossa}}{Dipartimento di Informatica, Universit\`a di Bologna, %[optional: Address], 
 Italy% \and My second affiliation, Country 
@@
 
-\author{Paolo {Pistone}}{Dipartimento di Informatica - Scienza e Ingegneria, Universit\`a di Bologna, %[optional: Address], 
+\author{Paolo {Pistone}}{Dipartimento di Informatica, Universit\`a di Bologna, %[optional: Address], 
 Italy% \and LIP, Universit\'e de Lyon, France 
-\and \url{http://logica.uniroma3.it/pistone/
-} }{paolo.pistone2@unibo.it}{https://orcid.org/0000-0003-4250-9051}{%(Optional)Funding each author
+\and \url{%WebPage
+} }{paolo.pistone2@unibo.it}{%Orchid
+}{%(Optional)Funding each author
 }
@@
 
-\keywords{Relational semantics, Differential lambda-calculus, Tropical semiring,  Program metrics, Lawvere quantale} %TODO mandatory; please add comma-separated list of keywords
+\keywords{Differential lambda-calculus, Tropical semiring, Relational semantics, Lawvere quantale, Program metrics} %TODO mandatory; please add comma-separated list of keywords
 
@@
 
-\funding{This work has been supported by the ERC CoG 818616 DIAPASoN.}%optional, to capture a funding statement, which applies to all authors.
+\funding{This work has been supported by the DIAPASoN project}%optional, to capture a funding statement, which applies to all authors.
 Please enter author specific funding statements as fifth argument of the \author macro.
@@
 \begin{abstract}
-We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, and we show that it provides a unified framework for both program  metrics, based on the analysis of program sensitivity via Lipschitz conditions, and resource analysis, based on higher-order program differentiation.
-To do that we focus on the semantics arising from the relational model weighted over the tropical semiring, and we discuss its application to the study of “best case” program behavior for languages with probabilistic and non-deterministic effects.
-Finally, we show that a general foundation for this approach is provided by an abstract correspondence between tropical algebra and Lawvere’s theory of generalized metric spaces.
-%A typical result of such line of research is that, thanks to the tropical operations, a simply typed term is interpreted as a locally Lipschitz map, which is in turn decomposed as an inf of more and more sensitive Lipschitz ones, corresponding to the interpretations of the approximants given by the lambda-calculus Taylor expansion.
-%We finally describe some promising lines of research, showing in particular how such tropical model expresses natural optimisation problems of probabilistic calculi.
+We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, by considering the relational semantics weighted on the tropical semiring.
+The aim is to provide a framework where both program metrics, based on the analysis of program sensitivity via Lipschitz conditions, and resource analysis, based on higher-order program differentiation, coexist.
+In order to do that, we consider a graded-types lambda-calculus and the differential lambda-calculus, both languages ensuring bounded duplications.
+A typical result of such line of research is that, thanks to the tropical operations, a simply typed term is interpreted as a locally Lipschitz map, which is in turn decomposed as an inf of more and more sensitive Lipschitz ones, corresponding to the interpretations of the approximants given by the lambda-calculus Taylor expansion.
+We finally describe some promising lines of research, showing in particular how such tropical model expresses natural optimisation problems of probabilistic calculi.
 \end{abstract}
@@
 
-\section{A Bridge between Metric and Differential Aspects}\label{section5bis}
-\input{section5bis}
-
-\section{Tropical Polynomials and Power Series}\label{section2}
+\section{Preliminaries on Tropical (Linear and Non-Linear) Algebra}
 \input{section2bis}
 
-\section{Tropical Semantics and First Order Effectful Programs}\label{section22}
+\section{Tropical Semantics of Effectful Programs: Two Examples}
 \input{section2bisbis}
@@
 \section{Tropical Semantics of Higher-Order Programs}\label{section3}
-\input{section3fancy}
 
-%%
-%%\section{Topical linear/non-linear algebra}\label{section3}
-%%\input{section3C}
-%%
-%%\section{Tropical weighted relational semantics for high-order languages}
-%%
-%%The category $\LREL$ satisfies many nice properties which make it able to model all the variants of the $\lambda$-calculi to which we are interested in here.
+The category $\LREL$ satisfies many nice properties which make it able to model all the variants of the $\lambda$-calculi to which we are interested in here.
 Although the majority of such properties are shared with the general class of models of $\lambda$-calculi called \emph{weighted relational models}, thoroughly studied in \cite{Manzo2013}, we mention them for our particular case in order to highlight how these look like in the tropical setting, but also to make the non-expert in such topics reader able to follow.
-%%The real interest in our particular model will come in the next \autoref{sec:tls}.
-%%
-%%
-%%\input{section3A}
-%%
-%%\input{section2}
-%
-%\paragraph*{Allowed duplication/erasure: the probabilistic $\lambda$-calculus $\STLC_\oplus$}\label{sec:STLCO}
-%
-%\input{section3B}
+The real interest in our particular model will come in the next \autoref{sec:tls}.
 
-\section{On Tropical Power Series}\label{sec:tls}
+\paragraph*{Prohibited duplication/erasure: the \emph{linear} $\STLC$}\label{sec:3A}
+\input{section3A}
+
+\paragraph*{Allowed duplication/erasure: the $\STLC$ and a toy probabilistic $\lambda$-calculus $\STLC_\oplus$}\label{sec:STLC+STLCO}
+
+\input{section3B}
+
+\input{section2}
+
+\section{Tropical Laurent Series}\label{sec:tls}
 \input{section4}
-
-\section{Lipschitz Meets Taylor}\label{sec:TayLip}
-\input{sectionTaylip}
-
-\section{Generalized Metric Spaces and $\Lawv$-Modules}\label{sec:GMS}
-\input{sectionLaw}
 
@@
 
-\section{Conclusion and Future Work}\label{section8} \input{section8}
+\section{Conclusion}\label{section8} \input{section8}
 
@@
 
-\bibliography{tropical}
+%\bibliography{tropical}
 
-\appendix
+%\appendix
 
-\section{Proofs from \autoref{section2}: Theorem \ref{theorem:fepsilon}}
-\input{appSection3}
-
-\section{Proofs from \autoref{section3}: $(\LREL, !_{n})$ is a model of $\BSTLC$}
-\input{appSection5}
-
-\section{Proofs from \autoref{sec:tls}: the Analysis of Tropical Power Series}\label{appsec:tls}
-\input{app_section4}
-
-\section{Proofs from \autoref{sec:TayLip}: Lipschitz meets Taylor}
-\input{appSection7}
-
-\section{Proofs of \autoref{sec:GMS}: $\Lawv$-modules and Generalized Metric Spaces}
-\input{appSection8}
+%\section{\autoref{section3}}\label{app-sec:section3}
 

### 2023-11-27 CSL24/section2bis.tex
commit

@@
 %\ref{fig:plot1}
-% (a) Plot of the tropical polynomials $\varphi_{n}$, for $0\leq n\leq 4$ (from top to bottom), and of their limit tLs $\varphi$ (in violet).
+% (a) Plot of the tropical polynomials $\varphi_{n}$, for $0\leq n\leq 4$ (from top to bottom), and of their limit tps $\varphi$ (in violet).
 The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x=2^{-(i+1)}$, satisfying $ix+2^{-i}=(i+1)x+2^{-(i+1)}$.
@@
 \end{tikzpicture}
-\caption{\small Tropical polynomials $\varphi_0,\dots,\varphi_4$ (top to bottom), and the limit tLs $\varphi$ (in violet).
+\caption{\small Tropical polynomials $\varphi_0,\dots,\varphi_4$ (top to bottom), and their limit tps $\varphi$ (in violet).
 The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x=2^{-(i+1)}$, satisfying $ix+2^{-i}=(i+1)x+2^{-(i+1)}$.}
@@
 While tropical polynomials are essentially combinatorial objects, this cannot be said for tps: since $\inf$s are not in general $\min$s, a tps is a ``limit'' of tropical polynomials of higher and higher degree, and its behavior is in general way more difficult to study than that of tropical polynomials~\cite{Porzio2021}.
-%For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points of their domain.
-E.g., the tLs $\varphi(x):=\inf_{n\in\N}\set{nx+2^{-n}}$ (see Fig.~\ref{fig:plot1}) %that we will take as our running example, 
+%For instance, tropical roots for tps (see \cite{Porzio2021}) may also include limit points of their domain.
+E.g., the tps $\varphi(x):=\inf_{n\in\N}\set{nx+2^{-n}}$ (see Fig.~\ref{fig:plot1}) %that we will take as our running example, 
 is the ``limit'' of the polynomials $\varphi_{n}$.
@@
 %\end{proof}
+As we'll see, the potential of collapsing infinitary objects (i.e.~tps) into combinatorial ones (i.e.~tropical polynomials), is one of the most intriguing features of tropical semantics.
 
-In order to prove Theorem \ref{theorem:fepsilon}, let us set the following:
+For the interested reader, we provide the proof of Theorem \ref{theorem:fepsilon} below. Let us first set the following:
 \begin{definition}
@@
 
-As we'll see, the potential of collapsing infinitary objects (i.e.~tps) into combinatorial ones (i.e.~tropical polynomials), is one of the most intriguing features of tropical semantics.
-
 %\subsection{Log-Probabilities and Tropicalisation through an exemple}\label{sec:proba}

### 2023-11-30 CSL24/main.tex
Commit notturno

@@
 
+\addto\extrasenglish{
+  \def\sectionautorefname{Section}
+}
+
 \title{Tropical Mathematics and the Lambda-calculus I}
@@
 
-\relatedversion{{\color{red}ADD URL HERE.}} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
+\relatedversion{\url{https://arxiv.org/abs/2311.15704}} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
 %\relatedversiondetails[linktext={opt.
@@
 We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, and we show that it provides a unifying framework for two well-developed quantitative approaches to program semantics: on the one hand program metrics, based on the analysis of program sensitivity via Lipschitz conditions, on the other hand resource analysis, based on linear logic and higher-order program differentiation.
-To do that we focus on the semantics arising from the relational model weighted over the tropical semiring, and we discuss its application to the study of “best case” program behavior for languages with probabilistic and non-deterministic effects.
+To do that, we focus on the semantics arising from the relational model weighted over the tropical semiring, and we discuss its application to the study of “best case” program behavior for languages with probabilistic and non-deterministic effects.
 Finally, we show that a general foundation for this approach is provided by an abstract correspondence between tropical algebra and Lawvere’s theory of generalized metric spaces.

### 2023-11-30 CSL24/section1.tex
Commit notturno

@@
-In recent years, more and more interest in the programming language community has been directed towards the study of \emph{quantitative} properties of programs like e.g.~the number of computation steps or the probability of convergence,
+In recent years, more and more interest in the programming language community has been directed towards the study of \emph{quantitative} properties of programs like, e.g., the number of computation steps or the probability of convergence,
 as opposed to purely \emph{qualitative} properties like termination or program equivalence. 
@@
 We can mention, for example,
-intersection type systems aimed at capturing time or space resources \cite{decarvalho2018, Accattoli2022} or convergence probabilities \cite{Breuvart2018, PistoneLICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthe_2012} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for
+intersection type systems aimed at capturing time or space resources \cite{decarvalho2018, Accattoli2022} or convergence probabilities \cite{Breuvart2018, PistoneLICS2022},  relational logics to account for probabilistic properties like, e.g., differential privacy \cite{Barthe_2012} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for
 probabilistic \cite{Ehrhard2011, Staton2017} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
@@
 
-On the one hand one one could mention the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}.
-This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
+On the one hand, one one could mention the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}.
+This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in, e.g., differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
 

### 2023-11-30 CSL24/section2bisbis.tex
Commit notturno

@@
 \begin{enumerate}
- \item What is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0} \false$''?
+ \item What is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0} \false$''  (similarly for $\true$)?
  I.e., which is the choice of $p,q$ that maximizes the probability $P_{\omega_0}$?
  \item 
-Knowing that $M$ produced $\true$ (similarly for $\false$) , which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0}\false$'', knowing that ``$M\twoheadrightarrow \false$''?
-I.e., which is the choice of $p,q$ that makes $\omega_0$ the most likely path among those leading to $\false$ (i.e.\ that maximizes the conditional probability $\BB P(``M\twoheadrightarrow_{\omega_0} \false'' \mid ``M\twoheadrightarrow \false'')$)?
+Knowing that $M$ reached a normal form, %produced $\false$ (similarly for $\true$),
+ which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0}\false$''? %, knowing that ``$M\twoheadrightarrow \false$''?
+I.e., which is the choice of $p,q$ that makes $\omega_0$ the most likely path %among those leading to $\false$ 
+(i.e.\ that maximizes the conditional probability $\BB P(``M\twoheadrightarrow_{\omega_0} \false'')$?% \mid ``M\twoheadrightarrow \false'')$)?
 \end{enumerate}
 
-Answering 1) and 2) amounts then at solving a maximization problem related to $P_{\omega_{0}}(p,q)$ or $Q_{\true}(p,q)$. These are more easily solved by passing to the associated tropical polynomials.
-For 1), the maximum values $x,y$ of $P_{rll}(p,q)$ can be computed by finding the \emph{minimum} values of $\mathsf tP_{rll}(-\log p, -\log q)= -2\log p- \log q$.
+Answering 1) amounts at solving a contrained maximization problem:
+\[(p,q)\in\arg\max_{y=1-x} P_{\omega_{0}}(x,y)\] or, equivalently: $p\in\arg\max_{x\in[0,1]} P_{\omega_0}(x,1-x)$,
+related to the monomial $P_{\omega_{0}}(p,q)$ and for which a tropical transformation does not, in general, help factoring the contraint $q=1-p$ out of the maximization.
+We will, instead, mainly concentrate on 2).
+This asks for finding a solution to the following contrained maximization problem in the unknown $p\in[0,1]$:
+\[
+P_{\omega_0}(p,1-p) = \max_w P_{\omega}(p,1-p)
+\]
+which is related to the polynomial, say, $Q_{\true}(p,q)$.
+Now, the difficulty raised by the constraint between the two variables of the monomials might be overcome by passing to the associated \emph{tropical} polynomial.
+In fact, since $-\log P_\omega(p,q)=(\trop P)(-\log p,-\log(1-p))$, this is equivalent to the following minimization problem in the unknown $p\in[0,1]$:
+\[
+(\trop P_{\omega_0})(-\log p,-\log(1-p)) = (\trop Q)(-\log p,-\log(1-p))
+\]
+and with a change of variable we can now move the difficult constraint outside the minimization problem, as this is equivalent to finding a solution of the following system of equations in the unknowns $(p,x,y)\in[0,1]\times[0,\infty]\times[0,\infty]$:
+\[
+\trop P_{\omega_0}(x,y) = \trop Q (x,y), \qquad x= -\log p,\qquad y=-\log(1-p).
+\]
+Since studying a tropical polynomial could be a feasable task, the first equation might be easily trasformed into an explicit relation between $x$ and $y$, and this might make it possible to solve the whole system, finally finding our $p$, as the next example shows.
+%Problem 1) the maximum values $x,y$ of $P_{rll}(p,q)$ can be computed by finding the \emph{minimum} values of $\mathsf tP_{rll}(-\log p, -\log q)= -2\log p- \log q$ under the contraint $q=1-p$.
 Notice that the latter is precisely the \emph{negative log-probability} of the event ``$M\twoheadrightarrow_{rll} \false$''.
-For 2), the maximum values of $Q_{\true}(p,q): [0,1]^{2}\to [0,1]$ can be computed as $e^{-\alpha},e^{-\beta}$, where $\alpha,\beta\in[0,\infty]$ are the \emph{minimum} values of the tropical polynomial
-$\mathsf t Q_{\true}(\alpha,\beta) = \min \{ 2\alpha, 2\alpha+\beta, 3\beta\}$.
+\begin{example}
+For our running example $M$, we have $\trop Q_{\true}(x,y)=\min\set{2x,y+2x,3y}$.
+For simplicity, let us suppose that we observed the event ``$M\twoheadrightarrow\true$'', and thus our probabilities are conditioned under this observation.
+It is not difficult to see that $\trop Q_{\true}(x,y)=3y$ iff $y\leq \frac{2}{3}x$, and $\trop Q_{\true}(x,y)=2x$ otherwise.
+Remembering that $3y=\trop P_{rrr}(x,y)$, we can now solve problem 2) for $\omega_0=rrr$:
+via the substitution $x:=-\log p$, $y:=-\log (1-p)$, the condition $y\leq \frac{2}{3}x$ is equivalent to $-\log (1-p)\leq -\frac{2}{3}\log p$, i.e.\ $1-p\geq p^{\frac{2}{3}}$.
+This means that, if $p\in[0,1]$ satisfies $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), then $P_{rrr}(p,1-p) = \max_{\omega=ll, rll, rrr} P_{\omega}(p,1-p)$, i.e.: knowing that $M$ sampled $\true$ in its normal form, the most likely sampled occurrence of $\true$ is the one at the address $rrr$ (the rightmost one) iff $1-p\geq p^{\frac{2}{3}}$.
+Said differently, the maximum likelihood estimator of $p$, with prior knowledge  ``$M\twoheadrightarrow \true$'' and seen the event ``$M\twoheadrightarrow_{\omega_0}\true$'', is any $p$ s.t.\ $1-p\geq p^{\frac{2}{3}}$.
+We have thus solved problem 2) above for $\omega_0=rrr$, under the additional hypothesis that ``$M\twoheadrightarrow\true$''.
+Remembering that $2x=\trop P_{ll}(x,y)$, in this simple case we can even say that for the other values of $p$ (for example, $p=\frac{1}{2}$), the most likely $\true$ to be sampled (knowing that a $\true$ has been sampled) is the one at address $ll$, therefore actually solving problem 2), under the same additional hypothesis, for all instances of the reduction path $\omega$ leading to $\true$ (the path $rll$ being the most likely one for no $p$ at all).
+\end{example}
+%the maximum values of $Q_{\true}(p,q): [0,1]^{2}\to [0,1]$ can be computed as $e^{-\alpha},e^{-\beta}$, where $\alpha,\beta\in[0,\infty]$ are the \emph{minimum} values of the tropical polynomial $\mathsf t Q_{\true}(\alpha,\beta) = \min \{ 2\alpha, 2\alpha+\beta, 3\beta\}$.
 
-Similar arguments could be done by replacing $\true$ and $\false$ by, respectively, a converging and diverging term (e.g.~in a $\mathrm{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.
+As we'll see in Section \ref{section3}, this analysis extends to PCF-style programs.
+For example, the program $M=\mathbf Y(\lambda x.\true \oplus_{p} x)$ yields the power series $Q_{\true}(p,q)=\sum_{n=0}^{\infty}pq^{n}=\frac{p}{1-q}$ that sums all \emph{infinitely many} ways in which $M$ may reduce to $\true$.
+Notice that the tropicalised series $\mathsf tQ_{\true}(-\log p,-\log q)=\inf_{n\in \mathbb N}\{-\log p -n\log q\}=-\log p$ collapses onto a single monomial describing the \emph{unique} most likely reduction path of $M$, namely the one that passes through a coin toss only once.
+%
+%Similar arguments could be done by replacing $\true$ and $\false$ by, respectively, a converging and diverging term (e.g.~in a $\mathrm{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.
 

### 2023-11-30 CSL24/section3fancy.tex
Commit notturno

@@
 $\mathsf{ev}^{!}: \Lawv^{!X\times Y}\times \Lawv^{X}\to \Lawv^{Y}$ given by 
-$b\in Y \mapsto \mathsf{ev}^{!}(F,x)_{b}= \inf_{\mu,b}\{ F_{\mu,b}+ \mu x\}$.
+$b\in Y \mapsto \mathsf{ev}^{!}(F,x)_{b}= \inf_{\mu}\{ F_{\mu,b}+ \mu x\}$.
 So, for instance, supposing the ground type $o$ of $\STLC$ is interpreted as the singleton set $\{*\}$, and recalling the identification $!\{*\}\simeq \mathbb N$, the interpretation of the term $x:o, z:(o\to o\to o) \vdash zxx:o$, involving two consecutive evaluations, yields the tps $\varphi:\Lawv\times \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}\to \Lawv$ given by $\varphi(x,z)=\inf_{n,n'}\{z_{[(n,n')]}+(n+n')x\}$.

### 2023-11-30 CSL24/section4.tex
Commit notturno

@@
 \end{theorem}
-\begin{proof}
-By observing that $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$, we apply the result that for all $f:V\subseteq (\BB R^X,d) \to (\BB R,\absv \cdot)$ concave and locally bounded, with $V$ open and convex and $d$ any metric, $f$ is locally Lipschitz, with the stated Lipschitz constant on $\overline{B_{\delta}(x)}$.
+%\begin{proof}
+%By observing that $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$, we apply the result that for all $f:V\subseteq (\BB R^X,d) \to (\BB R,\absv \cdot)$ concave and locally bounded, with $V$ open and convex and $d$ any metric, $f$ is locally Lipschitz, with the stated Lipschitz constant on $\overline{B_{\delta}(x)}$.
 %
 % Moreover, the Lipschitz constant of $f$ on $\overline{B_{\delta}(x)}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} \absv f$.
+%\end{proof}
+This crucial ingredient of the proof of this result is the following \autoref{th:locLip}, from which \autoref{thmTLSlocLip} immediately follows, since $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$ and all tps are non-negative.
+
+\begin{wrapfigure}{r}{0.5\textwidth}%\begin{figure}
+\begin{tikzpicture}[thick,scale=0.9]
+    \coordinate (u) at (-2,0);
+    \coordinate (v) at (2,0);
+    \coordinate (y) at (-1.2,0);
+    \coordinate (z) at (0,0);
+
+    \draw[name path={u--v}] (u) -- (v);
+    \node [draw,name path=z] at (z) [circle through={(u)}] {};
+
+    \path[name intersections={of=u--v and z,by={int_u,int_v}}]
+      foreach \X in {u,v}{(int_\X) node[below left]{$\X$}};
+
+    \fill   (int_u) circle[radius=2pt]  node [below left] {$u$};
+    \fill   (int_v) circle[radius=2pt]  node [below left] {$v$};
+
+    \fill   (z) circle[radius=0pt]  node [xshift=-1cm, yshift=2.2cm] {$\overline{B_{2\delta}(z)}$};
+
+    \fill   (z) circle[radius=2pt]  node [above] {$z$};
+    \fill   (y) circle[radius=2pt]  node [above] {$y$};
+
+    \coordinate (x) at (-0.7,0.4);
+
+    \draw   (x) circle[radius=1cm] node [above] {$x$};
+    \fill   (x) circle[radius=2pt]  node [xshift=10mm, yshift=10mm] {$\overline{B_{\delta}(x)}$};
+
+    \draw   (x) circle[radius=3cm] node {};
+    \fill   (x) circle[radius=0pt]  node [xshift=-1.7cm, yshift=3cm] {$\overline{B_{3\delta}(x)}$};
+\end{tikzpicture}
+\caption{Drawing of the proof of \autoref{th:locLip}.}
+\label{fig:proof_loc_lip}
+\end{wrapfigure} %%
+
+\begin{theorem}\label{th:locLip}
+Let $f:V\subseteq (\BB R^X,\norm\cdot) \to (\BB R,\absv \cdot)$, with $V$ open and convex and $\norm\cdot$ any norm.
+If $f$ is concave and locally bounded, then $f$ is locally Lipschitz.
+Moreover, the Lipschitz constant of $f$ on $\overline{B_{\delta}(x)}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} \absv f$.
+\end{theorem}
+\begin{proof}
+ Call $\overline{B_{\delta}(x)}:=B_1$, $\overline{B_{3\delta}(x)}:=B_3$.
+ It suffices to show that for all $x\in V$, there is $\delta>0$ s.t.\ $B_3\subseteq \mathrm{interior}(V)$, $K:=\max_{B_3}  \absv f$ exists and $f$ is $\frac{K}{\delta}$-Lipschitz on $B_1$.
+A $\delta$ satisfying the first two conditions exists since $V$ is open and bcause $f$ is locally bounded and $B_3$ is compact.
+We show that the third condition holds for all such $\delta$.
+ For that, fix $y,z\in B_1$ and call $r:=\frac{d(y,z)}{2\delta}\in[0,1]$.
+ We want to show that $\absv{f(y)-f(z)}\leq \frac{K}{\delta}d(y,z)=2Kr$.
+ Wlog $y\neq z$, otherwise there is nothing to prove.
+
+ So $r\neq 0$ and we can consider $u:=\frac{1+r}{r}z-\frac{1}{r}y$, $v:=\frac{1}{r}y-\frac{r-1}{r}z$.
+ We have $u,v\in \overline{B_{2\delta}(z)}=:B_2$.
+ Indeed, $d(u,z)=\norm{u-z}=\norm{\frac{z}{r}+z-\frac{y}{r}-z}=\frac{\norm{z-y}}{r}=2\delta$ and similarly $d(v,z)=2\delta$.
+ Geometrically, those are actually the intersections between $B_2$ and the line generated by $y$ and $z$, see \autoref{fig:proof_loc_lip}.
+ Now we have the convex combinations $z=\frac{1}{1+r}y+\frac{r}{1+r}u$ and $y=(1-r)z+rv$, so the concavity of $f$ entails on one hand:
+ $f(z)\geq \frac{1}{1+r}f(y)+\frac{r}{1+r}f(u)\geq \frac{f(y)}{1+r} - \frac{rK}{1+r}$, i.e.\ $f(y)-f(z)\leq r(K+f(z))\leq 2rK$, and on the other hand:
+ $f(y)\geq (1-r)f(z)+rf(v)\geq f(z)-r(f(z)+K)$, i.e.\ $f(z)-f(y)\leq r(f(z)+K)\leq 2rK$.
+ In the previous inequalities we have used that $f(u),f(v)\geq -K$.
+ This follows from the fact that $u,v\in B_2\subseteq B_3$, as it can be immediately checked, thus $\absv{f(u)},\absv{f(v)}\leq K$.
+ Putting the final inequalities together, we have $\absv{f(y)-f(z)}\leq 2rK$, i.e.\ the thesis.
 \end{proof}

### 2023-11-30 CSL24/section5bis.tex
Commit notturno

@@
 k\cdot x: \BB R\to \BB R$ are all Lipschitz-continuous, with Lipschitz constant $k$, the function $M=\lambda x.x^{2}$ obtained by ``duplicating'' $x$ is not Lipschitz anymore: $M$ is, so to say, \emph{too} sensitive to errors.
-More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of $\STLC$ (yet, several cartesian closed \emph{sub-}categories of $\Met$ do exist, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
+More abstractly, it is well-known that the category $\Met$ of pseudo-metric spaces and non-expansive maps, is \emph{not} cartesian closed, so it is not a model of $\STLC$ (yet, several cartesian closed \emph{sub-}categories of $\Met$ do exist, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
 Still, one might observe that the program $M$ above is actually Lipschitz-continuous, if not globally, at least \emph{locally} (i.e.~over any compact set).

### 2023-11-30 CSL24/section1.tex
commit

@@
 
-On the one hand, one one could mention the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}.
+On the one hand, there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}.
 This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in, e.g., differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.

### 2023-11-30 CSL24/section2bis.tex
Commit notturno
commit

@@
 Let us recall that a quantale is a complete lattice endowed with a continuous monoid action.
-In the case of $\Lawv$ the lattice is defined by the reverse order $\geq$ on $\BB R$, and the monoid action is provided by addition.
+In the case of $\Lawv$, the lattice is defined by the reverse order $\geq$ on $\BB R$, and the monoid action is provided by addition.
 Notice that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$.
@@
 %(tps) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
-in $k$-variables is a function $f:\Lawv^k\to\Lawv$ of shape $f(x)=\inf_{i\in I}\{ix + \matr f(i) \}$, where 
+in $k$ variables is a function $f:\Lawv^k\to\Lawv$ of shape $f(x)=\inf_{i\in I}\{ix + \matr f(i) \}$, where 
 $I\subseteq\mathbb N^{k}$, $i x$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients.
@@
 Therefore, for instance, the tropicalisation (w.r.t.\ any valuation) sends the infinite power series $\sum_{n\geq 2} x^n$ to the tps $\varphi(x)=\inf_n (n+2)x$.
-But the latter always coincides with the tropical polynomial $\varphi(x)=2x$, since $nx\geq 0$ for all $n\in\N$, which is in turn the tropicalisation of the polynomial $x^2$.
+Yet the latter always coincides with the tropical polynomial $\varphi(x)=2x$, since $nx\geq 0$ for all $n\in\N$, which is in turn the tropicalisation of the polynomial $x^2$.
 This shows that tropicalisation is \emph{not} in general an injective operation and in fact, as we show in \autoref{theorem:fepsilon} below, tps (in finitely many variables) have a tendency to collapse, if not globally at least locally, onto tropical polynomials.
@@
 
-\begin{remark}\label{rmk:AC}
+\begin{remark}\label{rmk:AC2}
 If $U\subseteq \N^{K}$ is infinite, then $U$ contains an infinite ascending chain $ m  _{0}\prec  m  _{1} \prec  m  _{2} \prec \dots$.
@@
 In other words, $ n  \in\mathcal F_\epsilon$ iff $\hat f( n  )<+\infty$ and for all $ m  \prec  n  $, one has $\hat f( m  )>\hat f( n  )+\epsilon$.
-
-Suppose that $\mathcal F_\epsilon$ is infinite; then, using Remark~\ref{rmk:AC}, it contains an infinite ascending chain $\set{ m  _0\prec  m  _1\prec\cdots}$.  
+Suppose that $\mathcal F_\epsilon$ is infinite; then, using Remark~\ref{rmk:AC2}, it contains an infinite ascending chain $\set{ m  _0\prec  m  _1\prec\cdots}$.  
 By definition of $\mathcal F_\epsilon$ we have then 
@@
 so that $+\infty>\hat f( m  _0)>\hat f( m  _{i})+i\epsilon\geq i\epsilon$ for all $i\in\N$.
-This contradicts the Archimedean property of $\R$.
+This contradicts the Archimedean property of $\R$. Hence $\mathcal F_{\epsilon}$ is finite.
 
-1).
+\begin{enumerate}
+\item
 We show that if $\mathcal F_\epsilon=\emptyset$, then $\hat f( n  )=+\infty$ for all $ n  \in\N^{K}$.
@@
 We go by induction on the well-founded order $\prec$ over $ n  \in\N^{K}$:
+\begin{itemize}
+\item if $ n  =0^{K}\notin\mathcal F_\epsilon$, then $\hat f( n  )=+\infty$, because there is no $ m  \prec n  $.
 
-- if $ n  =0^{K}\notin\mathcal F_\epsilon$, then $\hat f( n  )=+\infty$, because there is no $ m  \prec n  $.
-
-- if $ n  \notin\mathcal F_\epsilon$, with $ n  \neq 0^{K}$ then either $\hat f( n  )=+\infty$ and we are done, or there is $ m  \prec  n  $ s.t.\ $\hat f( m  )\leq \hat f( n  )+\epsilon$.
+\item if $ n  \notin\mathcal F_\epsilon$, with $ n  \neq 0^{K}$ then either $\hat f( n  )=+\infty$ and we are done, or there is $ m  \prec  n  $ s.t.\ $\hat f( m  )\leq \hat f( n  )+\epsilon$.
 By induction $\hat f( m  )=+\infty$ and, since $\epsilon<+\infty$, this entails $\hat f( n  )=+\infty$.
-
-2).
-If $f( x _0)=+\infty$ with $ x _0\in [\epsilon,\infty)^{K}$, then necessarily $\hat f( n  )=+\infty$ for all $ n  \in\N^{K}$.
+\end{itemize}
+\item If $f( x _0)=+\infty$ with $ x _0\in [\epsilon,\infty)^{K}$, then necessarily $\hat f( n  )=+\infty$ for all $ n  \in\N^{K}$.
 Therefore, no $ n  \in\N^{K}$ belongs to $\mathcal F_\epsilon$.
 
-3).
-We have to show that $f( x )=P_\epsilon( x )$ for all $ x \in [\epsilon,+\infty]^{K}$.
+\item We have to show that $f( x )=P_\epsilon( x )$ for all $ x \in [\epsilon,+\infty]^{K}$.
 By 1), it suffices to show that we can compute $f( x )$ by taking the $\inf$, that is therefore a $\min$, only in $\mathcal F_\epsilon$ (instead of all $\N^{K}$).
@@
 We do it again by induction on $\prec_{1}$:
-
-- if $ n  =0^{K}$, then from $\mathbf  n\notin \mathcal F_{\epsilon}$, by definition of $\mathcal F_\epsilon$, we have $\hat f( n  )=+\infty$ (because there is no $ n  '\prec n  $).
+\begin{itemize}
+\item if $ n  =0^{K}$, then from $\mathbf  n\notin \mathcal F_{\epsilon}$, by definition of $\mathcal F_\epsilon$, we have $\hat f( n  )=+\infty$ (because there is no $ n  '\prec n  $).
 So any element of $\mathcal F_\epsilon\neq\emptyset$ works.
 
-- if $ n  \neq 0^{K}$, then we have two cases:
+\item if $ n  \neq 0^{K}$, then we have two cases:
 either $\hat f( n  )=+\infty$, in which case we are done as before by taking any element of $\mathcal F_\epsilon\neq\emptyset$.
@@
 Therefore this $ m  $ works.
+\end{itemize}
+\end{enumerate}
 \end{proof}

### 2023-11-30 CSL24/section2bisbis.tex
commit

@@
 This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}.
-Typical questions in this case would be, for a fixed $\omega_0$:
+A natural question in this case would be, for a fixed $\omega_0$:
 %
 %The tropical point of view allows now to express two natural questions about this situation:
-\begin{enumerate}
- \item What is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0} \false$''  (similarly for $\true$)?
- I.e., which is the choice of $p,q$ that maximizes the probability $P_{\omega_0}$?
- \item 
-Knowing that $M$ reached a normal form, %produced $\false$ (similarly for $\true$),
- which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0}\false$''? %, knowing that ``$M\twoheadrightarrow \false$''?
-I.e., which is the choice of $p,q$ that makes $\omega_0$ the most likely path %among those leading to $\false$ 
-(i.e.\ that maximizes the conditional probability $\BB P(``M\twoheadrightarrow_{\omega_0} \false'')$?% \mid ``M\twoheadrightarrow \false'')$)?
-\end{enumerate}
+%\begin{enumerate}
+% \item What is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0} \false$''  (similarly for $\true$)?
+% I.e., which is the choice of $p,q$ that maximizes the probability $P_{\omega_0}$?
+% \item 
+knowing that $M$ reached a normal form, say $\true$, %produced $\false$ (similarly for $\true$),
+ what is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0}\true$''? %, knowing that ``$M\twoheadrightarrow \false$''?
+In other words, what is the choice of $p,q$ that maximizes the conditional probability $\BB P(M\twoheadrightarrow_{\omega_0} \true\mid M\twoheadrightarrow \true)$, i.e.~that 
+ makes $\omega_0$ the most likely path among those leading to $\true$?
+% \mid ``M\twoheadrightarrow \false'')$)?
+%\end{enumerate}
+%
+%Answering 1) amounts at solving a constrained maximization problem:
+%\[(p,q)\in\arg\max_{y=1-x} P_{\omega_{0}}(x,y)\] or, equivalently: $p\in\arg\max_{x\in[0,1]} P_{\omega_0}(x,1-x)$,
+%related to the monomial $P_{\omega_{0}}(p,q)$ and for which a tropical transformation does not, in general, help factoring the contraint $q=1-p$ out of the maximization.
+%We will, instead, mainly concentrate on 2).
 
-Answering 1) amounts at solving a contrained maximization problem:
-\[(p,q)\in\arg\max_{y=1-x} P_{\omega_{0}}(x,y)\] or, equivalently: $p\in\arg\max_{x\in[0,1]} P_{\omega_0}(x,1-x)$,
-related to the monomial $P_{\omega_{0}}(p,q)$ and for which a tropical transformation does not, in general, help factoring the contraint $q=1-p$ out of the maximization.
-We will, instead, mainly concentrate on 2).
-This asks for finding a solution to the following contrained maximization problem in the unknown $p\in[0,1]$:
+Answering this question amounts at finding a solution to the following constrained maximization problem in the unknown $p\in[0,1]$:
+%\[
+%P_{\omega_0}(p,q) = \max_\omega P_{\omega}(p,q)
+%
+%
+%\]
+%
+%\[
+%(p,q) \in \arg\max_{y=1-x} P_{\omega_{0}}(x,y)\] or, equivalently: $p\in\arg\max_{x\in[0,1]} P_{\omega_0}(x,1-x)$,
+%
+%\]
+% in the unknown $p\in[0,1]$:
 \[
-P_{\omega_0}(p,1-p) = \max_w P_{\omega}(p,1-p)
+P_{\omega_0}(p,1-p) = \max_\omega P_{\omega}(p,1-p)
 \]
 which is related to the polynomial, say, $Q_{\true}(p,q)$.
-Now, the difficulty raised by the constraint between the two variables of the monomials might be overcome by passing to the associated \emph{tropical} polynomial.
-In fact, since $-\log P_\omega(p,q)=(\trop P)(-\log p,-\log(1-p))$, this is equivalent to the following minimization problem in the unknown $p\in[0,1]$:
-\[
-(\trop P_{\omega_0})(-\log p,-\log(1-p)) = (\trop Q)(-\log p,-\log(1-p))
-\]
-and with a change of variable we can now move the difficult constraint outside the minimization problem, as this is equivalent to finding a solution of the following system of equations in the unknowns $(p,x,y)\in[0,1]\times[0,\infty]\times[0,\infty]$:
-\[
-\trop P_{\omega_0}(x,y) = \trop Q (x,y), \qquad x= -\log p,\qquad y=-\log(1-p).
-\]
-Since studying a tropical polynomial could be a feasable task, the first equation might be easily trasformed into an explicit relation between $x$ and $y$, and this might make it possible to solve the whole system, finally finding our $p$, as the next example shows.
+% the difficulty raised by the constraint between the two variables of the monomials might be overcome by passing to the associated \emph{tropical} polynomial.
+%In fact, 
+Since $-\log P_\omega(p,q)=(\trop P_{\omega})(-\log p,-\log q)$, this is equivalent
+% to the following \emph{minimization} problem in the unknown $p\in[0,1]$:
+%\[
+%(\trop P_{\omega_0})(-\log p,-\log(1-p)) = (\trop Q_{\true})(-\log p,-\log(1-p))
+%\]
+%%With a change of variable we can now move the  constraint outside the minimization problem, as
+%which is in turn equivalent
+ to finding a solution of the following constrained \emph{minimization} problem in the unknowns $(p,x,y)\in[0,1]\times[0,\infty]\times[0,\infty]$:
+\begin{equation}\label{eq:troproot}
+\trop P_{\omega_{0}} (x,y)= \trop Q_{\true}(x,y) , \qquad x= -\log p,\qquad y=-\log(1-p).
+\end{equation}
+Since the first equation can be solved easily (i.e.~in linear time) by computing the tropical roots of $\trop Q_{\true}$, we can obtain an explicit relation between $x$ and $y$ that can be used to solve the whole system, finally finding our $p$, as the next example shows.
 %Problem 1) the maximum values $x,y$ of $P_{rll}(p,q)$ can be computed by finding the \emph{minimum} values of $\mathsf tP_{rll}(-\log p, -\log q)= -2\log p- \log q$ under the contraint $q=1-p$.
@@
 \begin{example}
-For our running example $M$, we have $\trop Q_{\true}(x,y)=\min\set{2x,y+2x,3y}$.
-For simplicity, let us suppose that we observed the event ``$M\twoheadrightarrow\true$'', and thus our probabilities are conditioned under this observation.
-It is not difficult to see that $\trop Q_{\true}(x,y)=3y$ iff $y\leq \frac{2}{3}x$, and $\trop Q_{\true}(x,y)=2x$ otherwise.
-Remembering that $3y=\trop P_{rrr}(x,y)$, we can now solve problem 2) for $\omega_0=rrr$:
+For our running example $M$, 
+let us suppose that we observed the event ``$M\twoheadrightarrow\true$'', and thus our probabilities are conditioned under this observation. We have
+\[\trop Q_{\true}(x,y)=\min\set{\trop P_{ll}(x,y),\trop  P_{rll}(x,y),\trop P_{rrr}(x,y)}=\min
+\set{2x,y+2x,3y}.\]
+The tropical roots of $\trop Q_{\true}(x,y)$ are $(0,0)$ and all points of the form $(x,\frac{2}{3}x)$.
+Recall that these are the points where \autoref{eq:troproot} is satisfied for \emph{at least two} distinct values of $\omega_{0}$.
+From this it follows that $\trop Q_{\true} (x,y)=\trop P_{rrr}(x,y)=3y$ holds iff $y\leq \frac{2}{3}x$, and  $\trop Q_{\true}(x,y)=2x=\trop P_{ll}(x,y)$ otherwise.
+In this way can find the maximum likelihood estimator for $\omega_0=rrr$:
 via the substitution $x:=-\log p$, $y:=-\log (1-p)$, the condition $y\leq \frac{2}{3}x$ is equivalent to $-\log (1-p)\leq -\frac{2}{3}\log p$, i.e.\ $1-p\geq p^{\frac{2}{3}}$.
-This means that, if $p\in[0,1]$ satisfies $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), then $P_{rrr}(p,1-p) = \max_{\omega=ll, rll, rrr} P_{\omega}(p,1-p)$, i.e.: knowing that $M$ sampled $\true$ in its normal form, the most likely sampled occurrence of $\true$ is the one at the address $rrr$ (the rightmost one) iff $1-p\geq p^{\frac{2}{3}}$.
-Said differently, the maximum likelihood estimator of $p$, with prior knowledge  ``$M\twoheadrightarrow \true$'' and seen the event ``$M\twoheadrightarrow_{\omega_0}\true$'', is any $p$ s.t.\ $1-p\geq p^{\frac{2}{3}}$.
-We have thus solved problem 2) above for $\omega_0=rrr$, under the additional hypothesis that ``$M\twoheadrightarrow\true$''.
-Remembering that $2x=\trop P_{ll}(x,y)$, in this simple case we can even say that for the other values of $p$ (for example, $p=\frac{1}{2}$), the most likely $\true$ to be sampled (knowing that a $\true$ has been sampled) is the one at address $ll$, therefore actually solving problem 2), under the same additional hypothesis, for all instances of the reduction path $\omega$ leading to $\true$ (the path $rll$ being the most likely one for no $p$ at all).
+This means that, if $p\in[0,1]$ satisfies $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), then $P_{rrr}(p,1-p) = \max_{\omega=ll, rll, rrr} P_{\omega}(p,1-p)$, i.e.: knowing that $M$ sampled $\true$ in its normal form, the most likely sampled occurrence of $\true$ is the one at the address $rrr$ iff $1-p\geq p^{\frac{2}{3}}$.
+%Said differently, the maximum likelihood estimator of $p$, with prior knowledge  ``$M\twoheadrightarrow \true$'' and seen the event ``$M\twoheadrightarrow_{\omega_0}\true$'', is any $p$ s.t.\ $1-p\geq p^{\frac{2}{3}}$.
+%Observe that the tropical roots of $\trop Q_{\true}(x,y)$ provide a complete description of the relationship between events and their maximum likelihood estimators. 
+%We have thus solved problem 2) above for $\omega_0=rrr$, under the additional hypothesis that ``$M\twoheadrightarrow\true$''.
+%Remembering that $2x=\trop P_{ll}(x,y)$, in this simple case we can even say that for the other values of $p$ (for example, $p=\frac{1}{2}$), the most likely $\true$ to be sampled (knowing that a $\true$ has been sampled) is the one at address $ll$, therefore actually solving problem 2), under the same additional hypothesis, for all instances of the reduction path $\omega$ leading to $\true$ (the path $rll$ being the most likely one for no $p$ at all).
 \end{example}
@@
 For example, the program $M=\mathbf Y(\lambda x.\true \oplus_{p} x)$ yields the power series $Q_{\true}(p,q)=\sum_{n=0}^{\infty}pq^{n}=\frac{p}{1-q}$ that sums all \emph{infinitely many} ways in which $M$ may reduce to $\true$.
-Notice that the tropicalised series $\mathsf tQ_{\true}(-\log p,-\log q)=\inf_{n\in \mathbb N}\{-\log p -n\log q\}=-\log p$ collapses onto a single monomial describing the \emph{unique} most likely reduction path of $M$, namely the one that passes through a coin toss only once.
+Notice that the tropicalised series $\mathsf tQ_{\true}(-\log p,-\log q)=\inf_{n\in \mathbb N}\{-\log p -n\log q\}=-\log p$ collapses onto a single monomial describing the \emph{unique} most likely reduction path of $M$ leading to $\true$, namely the one that passes through a coin toss only once.
 %
@@
 This leads to the tps 
-$\varphi_{M\twoheadrightarrow \true}(\alpha)=\inf_{n\in \mathbb N}\big\{2\alpha+(n+1)\beta, 3\alpha, 3\alpha+(n+1)\beta\big \}= \min\{ 2\alpha+\beta, 3\alpha\}$, which describes all possible behaviors of $M$.
+$\varphi_{M\twoheadrightarrow \true}(\alpha)=\inf_{n\in \mathbb N}\big\{(n+2)\alpha+(n+1)\beta, 3\alpha, (n+3)\alpha+(n+1)\beta\big \}= \min\{ 2\alpha+\beta, 3\alpha\}$, which describes all possible behaviors of $M$.
 Notice that, since $\alpha$ and $\beta$ are always positive, the power series $\varphi_{M\twoheadrightarrow\true}(\alpha)$ is indeed equivalent to the tropical polynomial $\min\{ 2\alpha+\beta, 3\alpha\}$.

### 2023-11-30 CSL24/section3fancy.tex
commit

@@
 
-Now, one can see that the comonad $!$ of $\LREL$ can be ``decomposed'' into a family of ``graded exponentials functors'' $!_n:\LREL\to\LREL$ ($n\in\BB N$), where $!_{n}X$ is the set of multisets on $X$ of cardinality \emph{at most} $n$.
+On the side of $\LREL$, one can check that the comonad $!$ of $\LREL$ can be ``decomposed'' into a family of ``graded exponentials functors'' $!_n:\LREL\to\LREL$ ($n\in\BB N$), where $!_{n}X$ is the set of multisets on $X$ of cardinality \emph{at most} $n$.
 %  lift to functors
@@
 %satisfying the adjunction: $\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
-As such, $(\LREL,(!_n)_{n\in\N})$ yields then a model of $\BSTLC$.
-Remark that arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$.
-Notice that, whenever $\model *$ is finite, the set $\model{A}$ is finite for any type $A$ of $\BSTLC$.
+As such, $(\LREL,(!_n)_{n\in\N})$ yields then a model of $\BSTLC$. More details can be found in the extended version of this article.
+
+Notice that arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$ and that, whenever $\model *$ is finite, the set $\model{A}$ is finite for any type $A$ of $\BSTLC$.
 
@@
 %\begin{proposition}[{\color{red}LEMAY??}]\label{thm:LREL!CCDC}
- $\LREL_!$ is a CC$\partial\lambda$C (see Section \ref{section5bis}) when equipped with $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
+ $\LREL_!$ is a CC$\partial\lambda$C when equipped with $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the isomorphism $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
  The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tps $f:\Lawv^{X}\to \Lawv^{Y}$ into a tps $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by
-$D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a}+x_{a}+\mu y\right\}$.
+$D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a,b}+x_{a}+\mu y\right\}$. %, where $\hat f \in \LREL^{!X\times Y}$ is the matrix associated with $f$.
 One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).

### 2023-11-30 CSL24/section4.tex
commit

@@
 Moreover, tps in finitely many variables are locally equivalent to tropical polynomials (except at some singular points), and are thus also concave and locally Lipschitz-continuous.
-In this section we show that much of these properties extend also to tps with infinitely many variables, as those arising from the tropical relational model, %establish some topological and metric properties of tps, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
-and by a tropical polynomial or power series we mean one with possibly infintely many variables.
+In this section we show that much of these properties extend also to tps with infinitely many variables, as those arising from the tropical relational model.
+%establish some topological and metric properties of tps, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
+%and by a tropical polynomial or power series we mean one with possibly infintely many variables.
 The literature on tropical power series is often recent (e.g.~\cite{Porzio2021}), and several results we prove in this section are, to our knowledge, new.
@@
 
-Looking at Fig~\ref{fig:plot1}, we see that $\varphi$, just like the polynomials $\varphi_{n}$, is non-decreasing and concave.
+Looking at Fig.~\ref{fig:plot1}, we see that $\varphi$, just like the polynomials $\varphi_{n}$, is non-decreasing and concave.
 This is indeed always the case:
@@
 \begin{enumerate}
-\item If a tps $f: \overline{\BB R}_{\geq 0}^{X}\to \overline{\BB R}_{\geq 0}^{Y}$ arises from a matrix $\matr f:X\times Y\to \overline{\BB R}_{\geq 0}$ (i.e.~it is tropical linear), then $f$ is non-expansive (i.e.\ $1$-Lipschitz).
+\item If a tps $f: \overline{\BB R}_{\geq 0}^{X}\to \overline{\BB R}_{\geq 0}^{Y}$ arises from a matrix $\matr f:X\times Y\to \overline{\BB R}_{\geq 0}$ (i.e.~it is linear), then $f$ is non-expansive (i.e.\ $1$-Lipschitz).
 \item If  $f: \overline{\BB R}_{\geq 0}^{X}\to \overline{\BB R}_{\geq 0}^{Y}$ arises from a matrix $\matr f: \ !_{n}X\times Y\to \overline{\BB R}_{\geq 0}$, then $f$ is $n$-Lipschitz-continuous.
@@
 
-For any tropical polynomial $\varphi:\overline{\BB R}_{\geq 0}^{X}\to \overline{\BB R}_{\geq 0}^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \overline{\BB R}_{\geq 0}$ (as a monomial $\mu_ix+c_{i}$ yields a matrix entry on $!_{\#\mu_i}X\times Y$).
+When a tps is expressed as an $\inf$ of finitely many monomials we simply call it a tropical polynomial. 
+For any such tropical polynomial $\varphi:\overline{\BB R}_{\geq 0}^{X}\to \overline{\BB R}_{\geq 0}^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \overline{\BB R}_{\geq 0}$ (as a monomial $\mu_ix+c_{i}$ yields a matrix entry on $!_{\#\mu_i}X\times Y$).
 Hence, using Proposition \ref{prop:troplinear} 2., we have:
@@
 %\end{proof}
-This crucial ingredient of the proof of this result is the following \autoref{th:locLip}, from which \autoref{thmTLSlocLip} immediately follows, since $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$ and all tps are non-negative.
+The fundamental ingredient of the proof of this result is the following \autoref{th:locLip}, from which \autoref{thmTLSlocLip} immediately follows, since $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$ and all tps are non-negative.
 
-\begin{wrapfigure}{r}{0.5\textwidth}%\begin{figure}
-\begin{tikzpicture}[thick,scale=0.9]
+ %%
+
+% \begin{wrapfigure}{r}{0.5\textwidth}%\begin{figure}
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[thick,scale=0.7]
     \coordinate (u) at (-2,0);
@@
 
-    \fill   (z) circle[radius=0pt]  node [xshift=-1cm, yshift=2.2cm] {$\overline{B_{2\delta}(z)}$};
+    \fill   (z) circle[radius=0pt]  node [xshift=-1cm, yshift=1.6cm] {$\overline{B_{2\delta}(z)}$};
 
@@
     \draw   (x) circle[radius=1cm] node [above] {$x$};
-    \fill   (x) circle[radius=2pt]  node [xshift=10mm, yshift=10mm] {$\overline{B_{\delta}(x)}$};
+    \fill   (x) circle[radius=2pt]  node [xshift=10mm, yshift=7mm] {$\overline{B_{\delta}(x)}$};
 
     \draw   (x) circle[radius=3cm] node {};
-    \fill   (x) circle[radius=0pt]  node [xshift=-1.7cm, yshift=3cm] {$\overline{B_{3\delta}(x)}$};
+    \fill   (x) circle[radius=0pt]  node [xshift=-1.7cm, yshift=2cm] {$\overline{B_{3\delta}(x)}$};
 \end{tikzpicture}
+\end{center}
 \caption{Drawing of the proof of \autoref{th:locLip}.}
 \label{fig:proof_loc_lip}
-\end{wrapfigure} %%
+\end{figure}
+%\end{wrapfigure}
 
@@
  It suffices to show that for all $x\in V$, there is $\delta>0$ s.t.\ $B_3\subseteq \mathrm{interior}(V)$, $K:=\max_{B_3}  \absv f$ exists and $f$ is $\frac{K}{\delta}$-Lipschitz on $B_1$.
-A $\delta$ satisfying the first two conditions exists since $V$ is open and bcause $f$ is locally bounded and $B_3$ is compact.
+A $\delta$ satisfying the first two conditions exists since $V$ is open and because $f$ is locally bounded and $B_3$ is compact.
 We show that the third condition holds for all such $\delta$.

### 2023-11-30 CSL24/section5bis.tex
commit

--- 2023-08-02 CSL24/section5bis.tex
@@
 A fundamental insight coming from this line of work is that, if one can somehow   \emph{bound} the number of times that a program may duplicate its input, the resulting program will be \emph{Lipschitz-continuous}:
-if $M$ may duplicate at most $L$ times, then an error $\epsilon$ between two inputs will result in an error less or equal to $L\cdot \epsilon$ in the corresponding outputs \cite{Reed2010, Gaboardi2017}.
+if $M$ may duplicate at most $L$ times, then an error $\epsilon$ between two inputs will result in an error less or equal to $L\cdot \epsilon$ in the corresponding outputs \cite{Reed2010, Gaboardi2017} (yet, this property may fail in a concurrent setting, see e.g.~\cite{Gebler2018}).
 For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the functional input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.
@@
 k\cdot x: \BB R\to \BB R$ are all Lipschitz-continuous, with Lipschitz constant $k$, the function $M=\lambda x.x^{2}$ obtained by ``duplicating'' $x$ is not Lipschitz anymore: $M$ is, so to say, \emph{too} sensitive to errors.
-More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of $\STLC$ (yet, several cartesian closed \emph{sub-}categories of $\Met$ do exist, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
+More abstractly, it is well-known that the category $\Met$ of metric spaces and non-expansive maps, is \emph{not} cartesian closed, so it is not a model of $\STLC$ (yet, several cartesian closed \emph{sub-}categories of $\Met$ do exist, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
 Still, one might observe that the program $M$ above is actually Lipschitz-continuous, if not globally, at least \emph{locally} (i.e.~over any compact set).

### 2023-11-30 CSL24/sectionTaylip.tex
commit

@@
 Suppose $x: A \vdash_{\STLC}M:B$ and $\vdash_{\STLC} N:A$. 
-Then for all $t\in \Te{M}$ such that $\model{t}^{!}(\model{N})\neq \infty$, and $\delta>0$, the tps $\model{ x:A\vdash_{\STLC}M:B}^{!}$ is $\frac{\model{t}^{!}(\model{N}+2\delta)}{\delta}$-Lipschitz over the open ball
+Then for all $t\in \Te{M}$ such that $\model{t}^{!}(\model{N})\neq \infty$, and $\delta>0$, the tps $\model{ x:A\vdash_{\STLC}M:B}^{!}$ is $\frac{\model{t}^{!}(\model{N}+3\delta)}{\delta}$-Lipschitz over the open ball
  $B_{\delta}(\model{N})$.

### 2023-11-30 CSL24/section2bis.tex
Commit

@@
 
-\begin{wrapfigure}{r}{0.5\textwidth}%\begin{figure}
+%\begin{wrapfigure}{r}{0.5\textwidth}%
+\begin{figure}
+\begin{center}
 \begin{tikzpicture}[scale=0.9]
@@
 \end{tikzpicture}
+\end{center}
 \caption{\small Tropical polynomials $\varphi_0,\dots,\varphi_4$ (top to bottom), and their limit tps $\varphi$ (in violet).
 The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x=2^{-(i+1)}$, satisfying $ix+2^{-i}=(i+1)x+2^{-(i+1)}$.}
-\label{fig:plot1}%\end{figure}%
-\end{wrapfigure} %%
+\label{fig:plot1}
+\end{figure}%\end{wrapfigure} %%
 
@@
  There is a well-known relation between tropical polynomial/power series and usual polynomials/power series. 
-Given $Q,L$ semirings with units and zeros, $L$ commutative idempotent, if $L$ is totally ordered by the partial order: $\alpha \preceq \beta$ iff $\alpha +  \beta = \beta$, then one defines a \emph{valuation} \cite{Izhakian2015} to be a map $\mathrm{val}:Q\to L$ s.t.\ $\mathrm{val}(0)=0$, $\mathrm{val}(1)=1$, $\mathrm{val}(ab)=\mathrm{val}(a)\mathrm{val}(b)$, $\mathrm{val}(a+b)\preceq \max\set{\mathrm{val}(a),\mathrm{val}(b)}$.
-For example, the valuation which gives $1\in L$ on all the invertibles of $Q$ and $0$ otherwise, is called the \emph{trivial valuation} $\mathrm{val}_1$.
+%Given $Q,L$ semirings with units and zeros, $L$ commutative idempotent, if $L$ is totally ordered by the partial order: $\alpha \preceq \beta$ iff $\alpha +  \beta = \beta$, then one defines a \emph{valuation} \cite{Izhakian2015} to be a map $\mathrm{val}:Q\to L$ s.t.\ $\mathrm{val}(0)=0$, $\mathrm{val}(1)=1$, $\mathrm{val}(ab)=\mathrm{val}(a)\mathrm{val}(b)$, $\mathrm{val}(a+b)\preceq \max\set{\mathrm{val}(a),\mathrm{val}(b)}$.
+%For example, the valuation which gives $1\in L$ on all the invertibles of $Q$ and $0$ otherwise, is called the \emph{trivial valuation} $\mathrm{val}_1$.
 %Notice that if $\mathrm{val}:Q\to ([0,+\infty],+,\cdot,0,1)$ is a valuation, then $\mathrm{logval}(x):Q\to ([-\infty,+\infty],\min,+,+\infty,0)$, $\mathrm{logval}(x):=-\log \mathrm{val}(x)$ is a valuation.
-
-Now, for a power series $f(x)=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) with coefficients $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$, one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
+For a power series $f(x)=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) defined on $[0,1]$ and coefficients $a_i$ in $[0,1]$, %and for a fixed valuation $\mathrm{val}:Q\to \Lawv$, 
+one defines the \emph{tropicalisation} $\trop f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop f(\alpha):=\inf_{i\in I} \set{-\log(a_i)+i\alpha}$.
 %Remark that tropicalising w.r.t.\ the trivial valuation gives $\trop^{\mathrm{val}_1} \sum_{i\in I} a_i x^i = \inf_{i\in I} \set{i\alpha} = \trop^{\mathrm{val}} \sum_{i\in I\subseteq\N} x^i$ for all valuation $\mathrm{val}$, because in $\Lawv$ the unit is $0$.
@@
 So the trivial valuation sends both the infinite series $\sum_{n}(x^{2}y^{n+1}+x^{3})$ and the polynomial $x^{2}y+x^{3}$ onto the same map.
-Notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$ and the zero is $\infty$.
-Therefore, for instance, the tropicalisation (w.r.t.\ any valuation) sends the infinite power series $\sum_{n\geq 2} x^n$ to the tps $\varphi(x)=\inf_n (n+2)x$.
-Yet the latter always coincides with the tropical polynomial $\varphi(x)=2x$, since $nx\geq 0$ for all $n\in\N$, which is in turn the tropicalisation of the polynomial $x^2$.
+
+%Notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$ and the zero is $\infty$.
+%Therefore,
+For instance, the tropicalisation sends the infinite power series $\sum_{n\geq 2} x^n$ to the tps $\varphi(x)=\inf_n (n+2)x$.
+Yet the latter always coincides with the tropical monomial $\varphi(x)=2x$, since $nx\geq 0$ for all $n\in\N$, which is in turn the tropicalisation of the polynomial $x^2$.
 This shows that tropicalisation is \emph{not} in general an injective operation and in fact, as we show in \autoref{theorem:fepsilon} below, tps (in finitely many variables) have a tendency to collapse, if not globally at least locally, onto tropical polynomials.

### 2023-11-30 CSL24/main.tex
commit

@@
 
-\author{Davide {Barbarossa}}{Dipartimento di Informatica - Scienza e Ingegneria, Universit\`a di Bologna, %[optional: Address], 
+\author{Davide {Barbarossa}}{Universit\`a di Bologna, %[optional: Address], 
 Italy% \and My second affiliation, Country 
@@
 
-\author{Paolo {Pistone}}{Dipartimento di Informatica - Scienza e Ingegneria, Universit\`a di Bologna, %[optional: Address], 
+\author{Paolo {Pistone}}{Universit\`a di Bologna, %[optional: Address], 
 Italy% \and LIP, Universit\'e de Lyon, France 
@@
 
-\relatedversion{\url{https://arxiv.org/abs/2311.15704}} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
+\relatedversion{An extended version is available at \url{https://arxiv.org/abs/2311.15704}.} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
 %\relatedversiondetails[linktext={opt.
@@
 %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
+\EventEditors{Aniello Murano and Alexandra Silva} 
 \EventNoEds{2} 
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
+\EventLongTitle{32nd EACSL Annual Conference on Computer Science Logic (CSL 
+2024)} 
+\EventShortTitle{CSL 2024} 
+\EventAcronym{CSL} 
+\EventYear{2024} 
+\EventDate{February 19--23, 2024} 
+\EventLocation{Naples, Italy} 
 \EventLogo{} 
-\SeriesVolume{42}
-\ArticleNo{23}
+\SeriesVolume{288} 
+\ArticleNo{35}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\appendix
-
-\section{Proofs from \autoref{section2}: Theorem \ref{theorem:fepsilon}}
-\input{appSection3}
-
-\section{Proofs from \autoref{section3}: $(\LREL, !_{n})$ is a model of $\BSTLC$}
-\input{appSection5}
-
-\section{Proofs from \autoref{sec:tls}: the Analysis of Tropical Power Series}\label{appsec:tls}
-\input{app_section4}
-
-\section{Proofs from \autoref{sec:TayLip}: Lipschitz meets Taylor}
-\input{appSection7}
-
-\section{Proofs of \autoref{sec:GMS}: $\Lawv$-modules and Generalized Metric Spaces}
-\input{appSection8}
+%\appendix
+%
+%\section{Proofs from \autoref{section2}: Theorem \ref{theorem:fepsilon}}
+%\input{appSection3}
+%
+%\section{Proofs from \autoref{section3}: $(\LREL, !_{n})$ is a model of $\BSTLC$}
+%\input{appSection5}
+%
+%
+%\section{Proofs from \autoref{sec:tls}: the Analysis of Tropical Power Series}\label{appsec:tls}
+%\input{app_section4}
+%
+%\section{Proofs from \autoref{sec:TayLip}: Lipschitz meets Taylor}
+%\input{appSection7}
+%
+%
+%\section{Proofs of \autoref{sec:GMS}: $\Lawv$-modules and Generalized Metric Spaces}
+%\input{appSection8}
 

### 2023-11-30 CSL24/section2bis.tex
commit

@@
 \begin{center}
-\begin{tikzpicture}[scale=0.9]
+\begin{tikzpicture}[scale=0.7]
 \begin{axis}[samples=250]
@@
 For a power series $f(x)=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) defined on $[0,1]$ and coefficients $a_i$ in $[0,1]$, %and for a fixed valuation $\mathrm{val}:Q\to \Lawv$, 
-one defines the \emph{tropicalisation} $\trop f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop f(\alpha):=\inf_{i\in I} \set{-\log(a_i)+i\alpha}$.
+one can define the \emph{tropicalisation} $\trop f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop f(\alpha):=\inf_{i\in I} \set{-\log(a_i)+i\alpha}$.
 %Remark that tropicalising w.r.t.\ the trivial valuation gives $\trop^{\mathrm{val}_1} \sum_{i\in I} a_i x^i = \inf_{i\in I} \set{i\alpha} = \trop^{\mathrm{val}} \sum_{i\in I\subseteq\N} x^i$ for all valuation $\mathrm{val}$, because in $\Lawv$ the unit is $0$.
@@
 
-For instance, by looking at Fig~\ref{fig:plot1} it appears that, \emph{far from $0$}, $\varphi$ behaves like some of the polynomials $\varphi_{n}$.
+For instance, by looking at Fig.~\ref{fig:plot1} it appears that, \emph{far from $0$}, $\varphi$ behaves like some of the polynomials $\varphi_{n}$.
 In particular, %for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$. More, precisely, 

### 2023-11-30 CSL24/section2bisbis.tex
commit

@@
 This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}.
-A natural question in this case would be, for a fixed $\omega_0$:
+Then, a natural question in this case, for a fixed $\omega_0$, is the following:
 %
@@
 \trop P_{\omega_{0}} (x,y)= \trop Q_{\true}(x,y) , \qquad x= -\log p,\qquad y=-\log(1-p).
+\tag{$\star$}
 \end{equation}
@@
 For our running example $M$, 
-let us suppose that we observed the event ``$M\twoheadrightarrow\true$'', and thus our probabilities are conditioned under this observation. We have
+let us suppose that we observed the event ``$M\twoheadrightarrow\true$'', so that our probabilities are conditioned under this observation. We have
 \[\trop Q_{\true}(x,y)=\min\set{\trop P_{ll}(x,y),\trop  P_{rll}(x,y),\trop P_{rrr}(x,y)}=\min
 \set{2x,y+2x,3y}.\]
-The tropical roots of $\trop Q_{\true}(x,y)$ are $(0,0)$ and all points of the form $(x,\frac{2}{3}x)$.
-Recall that these are the points where \autoref{eq:troproot} is satisfied for \emph{at least two} distinct values of $\omega_{0}$.
+The tropical roots of $\trop Q_{\true}(x,y)$ are all the points of the form $(x,\frac{2}{3}x)$.
+Recall that these are the points where \eqref{eq:troproot} is satisfied for \emph{at least two} distinct values of $\omega_{0}$ (indeed for $\omega_{0}\in \{ll, rrr\}$).
 From this it follows that $\trop Q_{\true} (x,y)=\trop P_{rrr}(x,y)=3y$ holds iff $y\leq \frac{2}{3}x$, and  $\trop Q_{\true}(x,y)=2x=\trop P_{ll}(x,y)$ otherwise.
@@
 via the substitution $x:=-\log p$, $y:=-\log (1-p)$, the condition $y\leq \frac{2}{3}x$ is equivalent to $-\log (1-p)\leq -\frac{2}{3}\log p$, i.e.\ $1-p\geq p^{\frac{2}{3}}$.
-This means that, if $p\in[0,1]$ satisfies $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), then $P_{rrr}(p,1-p) = \max_{\omega=ll, rll, rrr} P_{\omega}(p,1-p)$, i.e.: knowing that $M$ sampled $\true$ in its normal form, the most likely sampled occurrence of $\true$ is the one at the address $rrr$ iff $1-p\geq p^{\frac{2}{3}}$.
+This means that, if $p\in[0,1]$ satisfies $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), then $P_{rrr}(p,1-p) = \max_{\omega=ll, rll, rrr} P_{\omega}(p,1-p)$.
+In other words, knowing that $M$ sampled $\true$ in its normal form, the most likely sampled occurrence of $\true$ is the one at the address $rrr$ iff $1-p\geq p^{\frac{2}{3}}$.
 %Said differently, the maximum likelihood estimator of $p$, with prior knowledge  ``$M\twoheadrightarrow \true$'' and seen the event ``$M\twoheadrightarrow_{\omega_0}\true$'', is any $p$ s.t.\ $1-p\geq p^{\frac{2}{3}}$.

### 2023-11-30 CSL24/sectionLaw.tex
commit

@@
 
-\subsection{$\Lawv$-Modules and Complete $\Lawv$-Categories}
+\subsection{$\Lawv$-Modules and Cocomplete $\Lawv$-Categories}
 
@@
 Moreover, a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is nothing but a non-expansive map $f:X\to Y$, since functoriality reads as $Y(f(x),f(y))\leq X(x,y)$.
-Functors of shape $\Phi: X\times Y^{\op}\to \Lawv$ are called \emph{distributors} and usually noted $\Phi: Y \pfun X$.
-Notice that distributors $\Phi: Y\pfun X$ and $\Psi: Z\pfun Y$ can be composed just like
+Functors of shape $\Phi:  Y^{\op}\times X\to \Lawv$ are called \emph{distributors} and noted $\Phi: X \pfun Y$. Two distributors $\Phi: Y\pfun X$ and $\Psi: Z\pfun Y$ can be composed just like 
 ordinary matrices in $\LREL$: $\Psi\diamond \Phi : Z\pfun X$ is given by
-$(\Psi\diamond \Phi)_{z,x}=\inf_{y\in Y}\Psi(z,y)+\Phi(y,x)$.
+$(\Psi\diamond \Phi)_{x,z}=\inf_{y\in Y}\{\Psi(y,z)+\Phi(x,y)\}$.
 
@@
 A functor $g:Z\to X$ is the \emph{$\Phi$-weighted colimit of $f$ over $X$}, noted $\colim(\Phi,f)$, if for all $z\in Z$ and $x\in X$
-\begin{align}
+\begin{align*}
 X(g(z),x)= \sup_{y\in Y}\left\{X(f(y),x)\menus \Phi(y,z)\right\}
-\end{align}
+\end{align*}
 
@@
 
-Observe that the usual Cauchy completeness for a $\Lawv$-category $X$ follows from its cocompleteness.
+Observe that cocompleteness for a symmetric $\Lawv$-category $X$ implies the usual Cauchy completeness.
 Indeed, a Cauchy sequence $(x_{n})_{n\in \mathbb N}$ in $X$ yields two adjoint distributors $\phi^{*}:1\pfun X$ and $\phi_{*}:X\pfun 1$, where
-$\phi^{*}(x')=\lim_{n\to \infty}X(x_{n},x')$ and
-$\phi_{*}(x')=\lim_{n\to \infty}X(x',x_{n})$. Hence,  
-$\colim(\phi^{*},1_{X}):1\to X$ must be a point $x$ satisfying $0=X(x,x)=\sup_{y\in X}\lim_{n\to \infty}(X(y,x)\menus X(x_{n},y))$, which implies $\lim_{n\to \infty}X(x_{n},x)=0$.
+$\phi^{*}(x')=\lim_{n\to \infty}X(x',x_{n},)$ and
+$\phi_{*}(x')=\lim_{n\to \infty}X(x_{n},x')$. Hence,  
+$\colim(\phi^{*},1_{X}):1\to X$ must be a point $x$ satisfying $0=X(x,x)=\sup_{y\in X}\lim_{n\to \infty}(X(y,x)\menus X(y,x_{n}))$, which implies $\lim_{n\to \infty}X(x_{n},x)=0$.
 
@@
 Conversely, in cocomplete $\Lawv$-categories it is possible to introduce a continuous $\Lawv$-action
-via suitable weighted colimits called \emph{tensors} (cf.~\cite{Stubbe2014}):
-
-\begin{definition}[tensors]
+via suitable weighted colimits called \emph{tensors}:
+\begin{definition}[tensors, cf.~\cite{Stubbe2014}]
 Let $X$ be a $\Lawv$-category, $x\in X$ and $\epsilon \in \Lawv$. The \emph{tensor of $x$ and $\epsilon$}, if it exists, is the colimit $\epsilon \otimes x:= \colim( [\epsilon],\Delta x)$, where
-$[\epsilon]: \{\star\}\pfun \{\star\}$ is the constantly $\epsilon$ distributor
-and $\Delta x:\{\star\}\to X$ is the constant functor. 
+$[\epsilon]:1\pfun1$ is the constantly $\epsilon$ distributor
+and $\Delta x:1\to X$ is the constant functor. 
 \end{definition}
